### 2023-02-11 dev/proposal.qmd
proposal

---
title: Conformal Counterfactual Explanations
subtitle: Research Proposal
abstract: |
    We propose Conformal Counterfactual Explanations: an effortless and rigorous way to produce realistic and faithful Counterfactual Explanations using Conformal Prediction. To address the need for realistic counterfactuals, existing work has primarily relied on separate generative models to learn the data generating process. While this an effective way to produce plausible and model-agnostic counterfactual explanations, it not only introduces an significant engineering overhead, but also reallocates the task of creating realistic model explanations from the model itsel to the generative model. Recent work has shown that there is no need for any of this when working with probabilistic models that explicitly quantify their own uncertainty. Unfortunately, most models used in practice still do not fulfil that basic requirement, in which case we would like to have a way to quantify predictive uncertainty in a post-hoc fashion.
---

## Motivation

Counterfactual Explanations are a powerful, flexible and intuitive way to not only explain black-box models, but also enable affected individuals to challenge them though the means of Algorithmic Recourse.

### From Adversarial Examples to Counterfactual Explanations

Most state-of-the-art approaches to generating Counterfactual Explanations (CE) rely on gradient descent in the feature space. The key idea is to perturb inputs $x\in\mathcal{X}$ into a black-box model $f: \mathcal{X} \mapsto \mathcal{Y}$ in order to change the model output $f(x)$ to some pre-specified target value $t\in\mathcal{Y}$. Formally, this boils down to defining some loss function $\ell(f(x),t)$ and taking gradient steps in the minimizing direction. The so generated counterfactuals are considered valid as soon as the predicted label matches the target label. A stripped down counterfactual is therefore little different from Adversarial Examples [@goodfellow2014explaining].

> You may not like it, but this is what counterfactuals look like

- Show DiCE for weak MLP
- Show Latent for same weak MLP
- Latent can be manipulated: 
    - train biased model
    - train VAE with biased variable removed/attacked (use Boston housing dataset)
    - hypothesis: will generate bias-free explanations

## Introduction to Conformal Prediction

- distribution-free, model-agnostic and scalable approach to predictive uncertainty quantification

### Post-hoc

- Take any fitted model and turn it into a conformal model using calibration data.

### Intrinsic --- Conformal Training [MAYBE]

- Model explicitly trained for conformal prediction.

## Conformal Counterfactuals

- Realistic counterfactuals by minimizing predictive uncertainty [@schut2021generating].
- Problem: restricted to Bayesian models.
- Solution: post-hoc predictive uncertainty quantification. 
- Conformal prediction is instance-based. So is CE. 
- Does the coverage guarantee carry over to counterfactuals?

### Research Questions

- Is CP alone enough to ensure realistic counterfactuals?
- Do counterfactuals improve further as the models get better?
- Do counterfactuals get more realistic as coverage
- What happens as we vary coverage and setsize?
- What happens as we improve the model robustness?
- What happens as we improve the model's ability to incorporate predictive uncertainty (deep ensemble, laplace)?

## Experiments

- Maybe: conformalised Laplace
- Benchmarking:
    - add PROBE into the mix
    - compare travel costs to domain shits.

## References

### 2023-02-12 dev/proposal.qmd
proposal

@@
 
+```{julia}
+using CounterfactualExplanations
+using CounterfactualExplanations.Data: load_mnist
+using CounterfactualExplanations.Models: load_mnist_mlp
+using Images
+using MLDatasets
+using MLDatasets: convert2image
+using Plots
+```
+
 ## Motivation
@@
 The so generated counterfactuals are considered valid as soon as the predicted label matches the target label.
-A stripped down counterfactual is therefore little different from Adversarial Examples [@goodfellow2014explaining].
+A stripped down counterfactual explanation is therefore little different from an adversarial example.
 
 > You may not like it, but this is what counterfactuals look like
+
+```{julia}
+# Data:
+counterfactual_data = load_mnist()
+X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
+input_dim, n_obs = size(counterfactual_data.X)
+M = load_mnist_mlp()
+
+# Target:
+factual_label = 8
+x = reshape(X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
+target = 3
+factual = predict_label(M, counterfactual_data, x)[1]
+
+# Search:
+n_ce = 3
+generator = GenericGenerator()
+ces = generate_counterfactual(x, target, counterfactual_data, M, generator; num_counterfactuals=n_ce)
+```
+
+```{julia}
+image_size = 300
+p1 = plot(
+    convert2image(MNIST, reshape(x,28,28)),
+    axis=nothing, 
+    size=(image_size, image_size),
+    title="Factual"
+)
+plts = [p1]
+
+for i in eachindex(ces)
+    ce = ces[i]
+    plt = plot(
+        convert2image(MNIST, reshape(CounterfactualExplanations.counterfactual(ce),28,28)),
+        axis=nothing, 
+        size=(image_size, image_size),
+        title="Counterfactual $i"
+    )
+    plts = [plts..., plt]
+end
+
+plt = plot(plts...; size=(image_size * (n_ce + 1),image_size), layout=(1,(n_ce + 1)))
+```
+
+The crucial difference between adversarial examples and counterfactuals is one of intent.
+While adversarial examples are typically intened to go unnoticed, counterfactuals in the context of Explainable AI are generally sought to be "plausible" or "realistic".
+To fulfill this latter goal, researchers have come up with a myriad of ways.
+@joshi2019realistic were among the first to suggest that instead of searching counterfactuals in the feature space, we can instead traverse a latent embedding learned by a surrogate generative model.
+This ensures that the generated counterfactuals comply with the (learned) data-generating process (DGB).
+Similarly, @poyiadzi2020face use density ...
 
@@
     - hypothesis: will generate bias-free explanations
+
+::: {#prp-surrogate}
+
+## Avoid Surrogates
+
+Since we are in the business of explaining a black-box model, the task of learning realistic representations of the data should not be reallocated from the model itself to some surrogate model.
+
+:::
 

### 2023-02-13 dev/proposal.qmd
proposal

--- 2023-02-11 dev/proposal.qmd
@@
 
-### From Adversarial Examples to Counterfactual Explanations
+### Counterfactual Explanations or Adversarial Examples?
 
@@
 The so generated counterfactuals are considered valid as soon as the predicted label matches the target label.
-A stripped down counterfactual is therefore little different from Adversarial Examples [@goodfellow2014explaining].
+A stripped down counterfactual explanation is therefore little different from an adversarial example.
 
-> You may not like it, but this is what counterfactuals look like
+> You may not like it, but this is what counterfactuals look like ...
+
+The crucial difference between adversarial examples and counterfactuals is one of intent.
+While adversarial examples are typically intened to go unnoticed, counterfactuals in the context of Explainable AI are generally sought to be "plausible", "realistic" or "feasible".
+To fulfill this latter goal, researchers have come up with a myriad of ways.
+@joshi2019realistic were among the first to suggest that instead of searching counterfactuals in the feature space, we can instead traverse a latent embedding learned by a surrogate generative model.
+Similarly, @poyiadzi2020face use density ...
+Finally, @karimi2021algorithmic argue that counterfactuals should comply with the causal model that generates them [CHECK IF WE CAN PHASE THIS LIKE THIS].
+Other related approaches include ...
+All of these different approaches have a common goal: they aim to ensure that the generated counterfactuals comply with the (learned) data-generating process (DGB).
+
+::: {#def-plausible}
+
+## Plausible Counterfactuals
+
+Since we are in the business of explaining a black-box model, the task of learning realistic representations of the data should not be reallocated from the model itself to some surrogate model.
+
+:::
+
+Formally, if $x \sim \mathcal{X}$ and for the corresponding counterfactual we have $x^{\prime}\sim\mathcal{X}x^{\prime}$, then
+
+![](www/mnist_9to4_latent.png)
 

### 2023-02-13 dev/proposal.qmd
work on proposal and extended abstract

@@
 ---
-title: Conformal Counterfactual Explanations
+title: High-Fidelity Counterfactual Explanations through Conformal Prediction
 subtitle: Research Proposal
@@
     We propose Conformal Counterfactual Explanations: an effortless and rigorous way to produce realistic and faithful Counterfactual Explanations using Conformal Prediction.
-To address the need for realistic counterfactuals, existing work has primarily relied on separate generative models to learn the data generating process.
-While this an effective way to produce plausible and model-agnostic counterfactual explanations, it not only introduces an significant engineering overhead, but also reallocates the task of creating realistic model explanations from the model itsel to the generative model.
+To address the need for realistic counterfactuals, existing work has primarily relied on separate generative models to learn the data-generating process.
+While this is an effective way to produce plausible and model-agnostic counterfactual explanations, it not only introduces a significant engineering overhead but also reallocates the task of creating realistic model explanations from the model itself to the generative model.
 Recent work has shown that there is no need for any of this when working with probabilistic models that explicitly quantify their own uncertainty.
@@
 
+```{julia}
+using CounterfactualExplanations
+using CounterfactualExplanations.Data: load_mnist
+using CounterfactualExplanations.Models: load_mnist_mlp
+using Images
+using MLDatasets
+using MLDatasets: convert2image
+using Plots
+www_path = "dev/www"
+```
+
 ## Motivation
 
-Counterfactual Explanations are a powerful, flexible and intuitive way to not only explain black-box models, but also enable affected individuals to challenge them though the means of Algorithmic Recourse.
+Counterfactual Explanations are a powerful, flexible and intuitive way to not only explain black-box models but also enable affected individuals to challenge them through the means of Algorithmic Recourse.
 
@@
 Formally, this boils down to defining some loss function $\ell(f(x),t)$ and taking gradient steps in the minimizing direction.
-The so generated counterfactuals are considered valid as soon as the predicted label matches the target label.
-A stripped down counterfactual explanation is therefore little different from an adversarial example.
+The so-generated counterfactuals are considered valid as soon as the predicted label matches the target label.
+A stripped-down counterfactual explanation is therefore little different from an adversarial example.
+In @fig-adv, for example, generic counterfactual search as in @wachter2017counterfactual has been applied to MNIST data.
 
-> You may not like it, but this is what counterfactuals look like ...
+```{julia}
+# Data:
+counterfactual_data = load_mnist()
+X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
+input_dim, n_obs = size(counterfactual_data.X)
+M = load_mnist_mlp()
+# Target:
+factual_label = 8
+x = reshape(X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
+target = 3
+factual = predict_label(M, counterfactual_data, x)[1]
+# Search:
+n_ce = 3
+generator = GenericGenerator()
+ces = generate_counterfactual(x, target, counterfactual_data, M, generator; num_counterfactuals=n_ce)
+```
+
+```{julia}
+image_size = 200
+p1 = plot(
+    convert2image(MNIST, reshape(x,28,28)),
+    axis=nothing, 
+    size=(image_size, image_size),
+    title="Factual"
+)
+plts = [p1]
+
+counterfactuals = CounterfactualExplanations.counterfactual(ces)
+phat = target_probs(ces)
+for x in zip(eachslice(counterfactuals; dims=3), eachslice(phat; dims=3))
+    ce, _phat = (x[1],x[2])
+    _title = "p(y=$(target)|x′)=$(round(_phat[1]; digits=3))"
+    plt = plot(
+        convert2image(MNIST, reshape(ce,28,28)),
+        axis=nothing, 
+        size=(image_size, image_size),
+        title=_title
+    )
+    plts = [plts..., plt]
+end
+plt = plot(plts...; size=(image_size * (n_ce + 1),image_size), layout=(1,(n_ce + 1)))
+savefig(plt, joinpath(www_path, "you_may_not_like_it.png"))
+```
+
+![You may not like it, but this is what stripped-down counterfactuals look like.
+Here we have used @wachter2017counterfactual to generate multiple counterfactuals for turning an 8 (eight) into a 3 (three).](www/you_may_not_like_it.png){#fig-adv}
 
 The crucial difference between adversarial examples and counterfactuals is one of intent.
-While adversarial examples are typically intened to go unnoticed, counterfactuals in the context of Explainable AI are generally sought to be "plausible", "realistic" or "feasible".
-To fulfill this latter goal, researchers have come up with a myriad of ways.
+While adversarial examples are typically intended to go unnoticed, counterfactuals in the context of Explainable AI are generally sought to be "plausible", "realistic" or "feasible".
+To fulfil this latter goal, researchers have come up with a myriad of ways.
 @joshi2019realistic were among the first to suggest that instead of searching counterfactuals in the feature space, we can instead traverse a latent embedding learned by a surrogate generative model.
 Similarly, @poyiadzi2020face use density ...
-Finally, @karimi2021algorithmic argue that counterfactuals should comply with the causal model that generates them [CHECK IF WE CAN PHASE THIS LIKE THIS].
+Finally, @karimi2021algorithmic argues that counterfactuals should comply with the causal model that generates them [CHECK IF WE CAN PHASE THIS LIKE THIS].
 Other related approaches include ...
@@
 
-Since we are in the business of explaining a black-box model, the task of learning realistic representations of the data should not be reallocated from the model itself to some surrogate model.
+Formally, if $x \sim \mathcal{X}$ and for the corresponding counterfactual we have $x^{\prime}\sim\mathcal{X}^{\prime}$, then for $x^{\prime}$ to be considered a plausible counterfactual, we need: $\mathcal{X} \approxeq \mathcal{X}^{\prime}$.
 
@@
 
-Formally, if $x \sim \mathcal{X}$ and for the corresponding counterfactual we have $x^{\prime}\sim\mathcal{X}x^{\prime}$, then
+In the context of Algorithmic Recourse, it makes sense to strive for plausible counterfactuals, since anything else would essentially require individuals to move to out-of-distribution states.
+But it is worth noting that our ambition to meet this goal, may have implications on our ability to faithfully explain the behaviour of the underlying black-box model (arguably our principal goal).
+By essentially decoupling the task of learning plausible representations of the data from the model itself, we open ourselves up to vulnerabilities.
+Using a separate generative model to learn $\mathcal{X}$, for example, has very serious implications for the generated counterfactuals.
+@fig-latent compares the results of applying REVISE [@joshi2019realistic] to MNIST data using two different Variational Auto-Encoders: while the counterfactual generated using an expressive (strong) VAE is compelling, the result relying on a less expressive (weak) VAE is not even valid.
+In this latter case, the decoder step of the VAE fails to yield values in $\mathcal{X}$ and hence the counterfactual search in the learned latent space is doomed.
 
-![](www/mnist_9to4_latent.png)
+![Counterfactual explanations for MNIST using a Latent Space generator: turning a nine (9) into a four (4).](www/mnist_9to4_latent.png){#fig-latent}
 
-- Show DiCE for weak MLP
-- Show Latent for same weak MLP
+> Here it would be nice to have another example where we poison the data going into the generative model to hide biases present in the data (e.g. Boston housing).
+
 - Latent can be manipulated: 
@@
     - hypothesis: will generate bias-free explanations
+
+### From Plausible to High-Fidelity Counterfactuals {#sec-fidelity}
+
+In light of the findings, we propose to generally avoid using surrogate models to learn $\mathcal{X}$ in the context of Counterfactual Explanations.
 
@@
 
-## Introduction to Conformal Prediction
+In cases where the use of surrogate models cannot be avoided, we propose to weigh the plausibility of counterfactuals against their fidelity to the black-box model.
+In the context of Explainable AI, fidelity is defined as describing how an explanation approximates the prediction of the black-box model [@molnar2020interpretable].
+Fidelity has become the default metric for evaluating Local Model-Agnostic Models, since they often involve local surrogate models whose predictions need not always match those of the black-box model.
 
-- distribution-free, model-agnostic and scalable approach to predictive uncertainty quantification
+In the case of Counterfactual Explanations, the concept of fidelity has so far been ignored.
+This is not altogether surprising, since by construction and design, Counterfactual Explanations work with the predictions of the black-box model directly: as stated above, a counterfactual $x^{\prime}$ is considered valid if and only if $f(x^{\prime})=t$, where $t$ denote some target outcome.
 
-### Post-hoc
+Does fidelity even make sense in the context of CE, and if so, how can we define it?
+In light of the examples in the previous section, we think it is urgent to introduce a notion of fidelity in this context, that relates to the distributional properties of the generated counterfactuals.
+In particular, we propose that a high-fidelity counterfactual $x^{\prime}$ complies with the class-conditional distribution $\mathcal{X}_{\theta} = p_{\theta}(X|y)$ where $\theta$ denote the black-box model parameters.
 
+::: {#def-fidele}
+
+## High-Fidelity Counterfactuals
+
+Let $\mathcal{X}_{\theta}|y = p_{\theta}(X|y)$ denote the class-conditional distribution of $X$ defined by $\theta$.
+Then for $x^{\prime}$ to be considered a high-fidelity counterfactual, we need: $\mathcal{X}_{\theta}|t \approxeq \mathcal{X}^{\prime}$ where $t$ denotes the target outcome.
+
+:::
+
+In order to assess the fidelity of counterfactuals, we propose the following two-step procedure:
+
+1) Generate samples $X_{\theta}|y$ and $X^{\prime}$ from $\mathcal{X}_{\theta}|t$ and $\mathcal{X}^{\prime}$, respectively.
+2) Compute the Maximum Mean Discrepancy (MMD) between $X_{\theta}|y$ and $X^{\prime}$.
+
+If the computed value is different from zero, we can reject the null-hypothesis of fidelity.
+
+> Two challenges here: 1) implementing the sampling procedure in @grathwohl2020your; 2) it is unclear if MMD is really the right way to measure this.
+
+## Conformal Counterfactual Explanations
+
+In @sec-fidelity, we have advocated for avoiding surrogate models in the context of Counterfactual Explanations.
+In this section, we introduce an alternative way to generate high-fidelity Counterfactual Explanations.
+In particular, we propose Conformal Counterfactual Explanations (CCE), that is Counterfactual Explanations that minimize the predictive uncertainty of conformal models.
+
+### Minimizing Predictive Uncertainty
+
+@schut2021generating demonstrated that the goal of generating realistic (plausible) counterfactuals can also be achieved by seeking counterfactuals that minimize the predictive uncertainty of the underlying black-box model.
+Similarly, @antoran2020getting ...
+
+- Problem: restricted to Bayesian models.
+- Solution: post-hoc predictive uncertainty quantification. In particular, Conformal Prediction.
+
+### Background on Conformal Prediction
+
+- Distribution-free, model-agnostic and scalable approach to predictive uncertainty quantification.
+- Conformal prediction is instance-based. So is CE. 
 - Take any fitted model and turn it into a conformal model using calibration data.
+- Our approach, therefore, relaxes the restriction on the family of black-box models, at the cost of relying on a subset of the data.
+Arguably, data is often abundant and in most applications practitioners tend to hold out a test data set anyway.
 
-### Intrinsic --- Conformal Training [MAYBE]
+> Does the coverage guarantee carry over to counterfactuals?
 
-- Model explicitly trained for conformal prediction.
+### Generating Conformal Counterfactuals
 
-## Conformal Counterfactuals
+While Conformal Prediction has recently grown in popularity, it does introduce a challenge in the context of classification: the predictions of Conformal Classifiers are set-valued and therefore difficult to work with, since they are, for example, non-differentiable.
+Fortunately, @stutz2022learning introduced carefully designed differentiable loss functions that make it possible to evaluate the performance of conformal predictions in training.
+We can leverage these recent advances in the context of gradient-based counterfactual search ...
 
-- Realistic counterfactuals by minimizing predictive uncertainty [@schut2021generating].
-- Problem: restricted to Bayesian models.
-- Solution: post-hoc predictive uncertainty quantification. 
-- Conformal prediction is instance-based. So is CE. 
-- Does the coverage guarantee carry over to counterfactuals?
+> Challenge: still need to implement these loss functions.
+
+## Experiments
 
@@
 - What happens as we improve the model's ability to incorporate predictive uncertainty (deep ensemble, laplace)?
+- What happens if we combine with DiCE, ClaPROAR, Gravitational?
+- What about CE robustness to endogenous shifts [@altmeyer2023endogenous]?
 
-## Experiments
-
-- Maybe: conformalised Laplace
 - Benchmarking:
-    - add PROBE into the mix
+    - add PROBE [@pawelczyk2022probabilistically] into the mix.
     - compare travel costs to domain shits.
 
+> Nice to have: What about using Laplace Approximation, then Conformal Prediction? What about using Conformalised Laplace?
+
 ## References

### 2023-02-14 notebooks/conformal.qmd
uh

```{julia}
using CCE
using ConformalPrediction
using CounterfactualExplanations
using CounterfactualExplanations.Data
using MLJFlux
```

```{julia}
counterfactual_data = load_linearly_separable()
```

```{julia}
clf = NeuralNetworkClassifier()
```

```{julia}
X = table(permutedims(counterfactual_data.X))
y =  counterfactual_data.output_encoder.labels
conf_model = conformal_model(clf; method=:simple_inductive)
mach = machine(conf_model, X, y)
fit!(mach)
```

### 2023-02-15 src/model.jl
penalties

@@ -1,6 +1,7 @@
 using ConformalPrediction
 using CounterfactualExplanations.Models
 using Flux
+using MLJBase
 using MLUtils
 using SliceMap
 using Statistics
@@ -12,10 +13,11 @@ Constructor for models trained in `Flux.jl`.
 """
 struct ConformalModel <: Models.AbstractDifferentiableJuliaModel
     model::ConformalPrediction.ConformalProbabilisticSet
+    fitresult::Any
     likelihood::Symbol
-    function ConformalModel(model, likelihood)
+    function ConformalModel(model, fitresult, likelihood)
         if likelihood ∈ [:classification_binary, :classification_multi]
-            new(model, likelihood)
+            new(model, fitresult, likelihood)
         else
             throw(
                 ArgumentError(
@@ -27,20 +29,57 @@ struct ConformalModel <: Models.AbstractDifferentiableJuliaModel
 end
 
 # Outer constructor method:
-function ConformalModel(model; likelihood::Symbol=:classification_binary)
-    ConformalModel(model, likelihood)
+function ConformalModel(model, fitresult; likelihood::Symbol=:classification_binary)
+    ConformalModel(model, fitresult, likelihood)
 end
 
 # Methods
-function logits(M::ConformalModel, X::AbstractArray)
-    return SliceMap.slicemap(x -> M.model(x), X, dims=(1, 2))
+@doc raw"""
+    Models.logits(M::ConformalModel, X::AbstractArray)
+
+To keep things consistent with the architecture of `CounterfactualExplanations.jl`, this method computes logits $\beta_i x_i$ (i.e. the linear predictions) for a Conformal Classifier. By default, `MLJ.jl` and `ConformalPrediction.jl` return probabilistic predictions. To get the underlying logits, we invert the softmax function. 
+
+Let $\hat{p}_i$ denote the estimated softmax output for feature $i$. Then in the multi-class case the following formula can be applied:
+
+```math
+\beta_i x_i = \log (\hat{p}_i) + \log (\sum_i \exp(\hat{p}_i))
+```
+
+For a short derivation, see here: https://math.stackexchange.com/questions/2786600/invert-the-softmax-function. 
+
+In the binary case logits are fed through the sigmoid function instead of softmax, so we need to further adjust as follows,
+
+```math
+\beta x = \beta_1 x_1 - \beta_0 x_0
+```    
+
[…]

### 2023-02-15 notebooks/conformal.qmd
penalties

@@
 using CounterfactualExplanations.Data
+using Flux
+using MLJBase
 using MLJFlux
+using Plots
 ```
@@
 ```{julia}
-clf = NeuralNetworkClassifier()
+builder = MLJFlux.@builder Chain(
+    Dense(n_in, 32, relu),
+    Dense(32, n_out)
+)
+clf = NeuralNetworkClassifier(builder=builder, epochs=100)
 ```
@@
 ```
+
+```{julia}
+M = CCE.ConformalModel(conf_model, mach.fitresult)
+```

### 2023-02-15 src/ConformalGenerator.jl

@@ -48,6 +48,31 @@ function ConformalGenerator(;
     ConformalGenerator(loss, complexity, λ, decision_threshold, params.opt, params.τ)
 end
 
+# Loss:
+# """
+#     ℓ(generator::ConformalGenerator, counterfactual_explanation::AbstractCounterfactualExplanation)
+
+# The default method to apply the generator loss function to the current counterfactual state for any generator.
+# """
+# function ℓ(
+#     generator::ConformalGenerator,
+#     counterfactual_explanation::AbstractCounterfactualExplanation,
+# )
+
+#     loss_fun =
+#         !isnothing(generator.loss) ? getfield(Losses, generator.loss) :
+#         CounterfactualExplanations.guess_loss(counterfactual_explanation)
+#     @assert !isnothing(loss_fun) "No loss function provided and loss function could not be guessed based on model."
+#     loss = loss_fun(
+#         getfield(Models, :logits)(
+#             counterfactual_explanation.M,
+#             CounterfactualExplanations.decode_state(counterfactual_explanation),
+#         ),
+#         counterfactual_explanation.target_encoded,
+#     )
+#     return loss
+# end
+
 """
     set_size_penalty(
         generator::ConformalGenerator,
@@ -61,21 +86,7 @@ function set_size_penalty(
     counterfactual_explanation::AbstractCounterfactualExplanation,
 )
 
-    x_ = CounterfactualExplanations.decode_state(counterfactual_explanation)
-    M = counterfactual_explanation.M
-    model = isa(M.model, Vector) ? M.model : [M.model]
-    y_ = counterfactual_explanation.target_encoded
 
-    if M.likelihood == :classification_binary
-        loss_type = :logitbinarycrossentropy
-    else
-        loss_type = :logitcrossentropy
-    end
-
-    loss(x, y) =
-        sum([getfield(Flux.Losses, loss_type)(nn(x), y) for nn in model]) / length(model)
-
-    return loss(x_, y_)
 end
 
 # Complexity:

### 2023-02-15 notebooks/conformal.qmd
slowly slowly

@@
 ```{julia}
+contourf(mach.model, mach.fitresult, X, y; plot_set_size=true)
+```
+
+## Counterfactual Explanation
+
+```{julia}
 M = CCE.ConformalModel(conf_model, mach.fitresult)
+generator = CCE.ConformalGenerator()
 ```
+
+```{julia}
+x = select_factual(counterfactual_data,rand(1:size(counterfactual_data.X,2)))
+y = predict_label(M, counterfactual_data, x)[1]
+target = counterfactual_data.y_levels[counterfactual_data.y_levels .!= y][1]
+```

### 2023-02-15 src/ConformalGenerator.jl
slowly slowly

@@ -3,6 +3,7 @@ using CounterfactualExplanations.Generators
 using Flux
 using LinearAlgebra
 using Parameters
+using SliceMap
 using Statistics
 
 mutable struct ConformalGenerator <: AbstractGradientBasedGenerator
@@ -19,7 +20,7 @@ end
     opt::Flux.Optimise.AbstractOptimiser = Descent()
     τ::AbstractFloat = 1e-3
     κ::Real = 1.0
-    Τ::Real = 0.5
+    temp::Real = 0.5
 end
 
 """
@@ -50,31 +51,6 @@ function ConformalGenerator(;
     ConformalGenerator(loss, complexity, λ, decision_threshold, params.opt, params.τ)
 end
 
-# Loss:
-# """
-#     ℓ(generator::ConformalGenerator, counterfactual_explanation::AbstractCounterfactualExplanation)
-
-# The default method to apply the generator loss function to the current counterfactual state for any generator.
-# """
-# function ℓ(
-#     generator::ConformalGenerator,
-#     counterfactual_explanation::AbstractCounterfactualExplanation,
-# )
-
-#     loss_fun =
-#         !isnothing(generator.loss) ? getfield(Losses, generator.loss) :
-#         CounterfactualExplanations.guess_loss(counterfactual_explanation)
-#     @assert !isnothing(loss_fun) "No loss function provided and loss function could not be guessed based on model."
-#     loss = loss_fun(
-#         getfield(Models, :logits)(
-#             counterfactual_explanation.M,
-#             CounterfactualExplanations.decode_state(counterfactual_explanation),
-#         ),
-#         counterfactual_explanation.target_encoded,
-#     )
-#     return loss
-# end
-
 """
     set_size_penalty(
         generator::ConformalGenerator,
@@ -88,6 +64,19 @@ function set_size_penalty(
     counterfactual_explanation::AbstractCounterfactualExplanation,
 )
 
+    conf_model = counterfactual_explanation.M.model
+    fitresult = counterfactual_explanation.M.fitresult
+    X = CounterfactualExplanations.decode_state(counterfactual_explanation)
+    loss = SliceMap.slicemap(X, dims=(1,2)) do x
+        ConformalPrediction.smooth_size_loss(
+            conf_model, fitresult, x;
+            κ = generator.κ,
+            temp = generator.temp
+        )
+    end
+    loss = mean(loss)
+
+    return loss
 
 end
 
[…]

### 2023-02-16 notebooks/conformal.qmd
ufff
implemented loss functions definitely seem to make sense

--- 2023-02-15 notebooks/conformal.qmd
@@
 
+## Counterfactual Explanation
+
 ```{julia}
 M = CCE.ConformalModel(conf_model, mach.fitresult)
+generator = CCE.ConformalGenerator()
 ```
+
+```{julia}
+x = select_factual(counterfactual_data,rand(1:size(counterfactual_data.X,2)))
+y_factual = predict_label(M, counterfactual_data, x)[1]
+target = counterfactual_data.y_levels[counterfactual_data.y_levels .!= y_factual][1]
+```
+
+```{julia}
+p1 = contourf(mach.model, mach.fitresult, X, y; plot_classification_loss=true, target=target, zoom=0)
+p2 = contourf(mach.model, mach.fitresult, X, y; plot_set_loss=true, zoom=0)
+plot(p1, p2, size=(800,320))
+```
+
+```{julia}
+ce = generate_counterfactual(x, target, counterfactual_data, M, generator)
+```

### 2023-02-16 notebooks/conformal.qmd
basic set size loss penalty now for working counterfactual search

@@
 M = CCE.ConformalModel(conf_model, mach.fitresult)
-generator = CCE.ConformalGenerator()
 ```
@@
 ```{julia}
-ce = generate_counterfactual(x, target, counterfactual_data, M, generator)
+#| output: true
+#| echo: false
+# Generators:
+generators = Dict(
+    "Generic (γ=0.5)" => GenericGenerator(opt = opt, decision_threshold=0.5),
+    "Generic (γ=0.9)" => GenericGenerator(opt = opt, decision_threshold=0.9),
+    "Conformal (λ₂=1)" => CCE.ConformalGenerator(opt=opt, λ=[0.1,1]),
+    "Conformal (λ₂=10)" => CCE.ConformalGenerator(opt=opt, λ=[0.1,10]),
+)
+
+counterfactuals = Dict([name => generate_counterfactual(x, target, counterfactual_data, M, gen;) for (name, gen) in generators])
+# Plots:
+plts = []
+for (name,ce) ∈ counterfactuals
+    plt = plot(ce; title=name, colorbar=false, ticks = false, legend=false, zoom=0)
+    plts = vcat(plts..., plt)
+end
+_n = length(generators)
+plot(plts..., size=(_n * 185,200), layout=(1,_n))
 ```

### 2023-02-16 notebooks/conformal.qmd
woohoo

--- 2023-02-16 notebooks/conformal.qmd
@@
 ```{julia}
-ce = generate_counterfactual(x, target, counterfactual_data, M, generator)
+#| output: true
+#| echo: false
+
+ordered_names = [
+    "Generic (γ=0.5)",
+    "Generic (γ=0.9)",
+    "Conformal (λ₂=1)",
+    "Conformal (λ₂=10)"
+]
+
+# Generators:
+generators = Dict(
+    ordered_names[1] => GenericGenerator(opt = opt, decision_threshold=0.5),
+    ordered_names[2] => GenericGenerator(opt = opt, decision_threshold=0.9),
+    ordered_names[3] => CCE.ConformalGenerator(opt=opt, λ=[0.1,1]),
+    ordered_names[4] => CCE.ConformalGenerator(opt=opt, λ=[0.1,10]),
+)
+
+counterfactuals = Dict([name => generate_counterfactual(x, target, counterfactual_data, M, gen; initialization=:identity) for (name, gen) in generators])
+
+# Plots:
+plts = []
+for name ∈ ordered_names
+    ce = counterfactuals[name]
+    plt = plot(ce; title=name, colorbar=false, ticks = false, legend=false, zoom=0)
+    plts = vcat(plts..., plt)
+end
+_n = length(generators)
+img_size = 300
+plot(plts..., size=(_n * img_size,1.05*img_size), layout=(1,_n))
 ```

### 2023-02-16 src/ConformalGenerator.jl
uh

@@ -43,11 +43,11 @@ generator = ConformalGenerator()
 ```
 """
 function ConformalGenerator(;
-    loss::Union{Nothing,Symbol} = nothing,
-    complexity::Function = norm,
-    λ::Union{AbstractFloat,AbstractVector} = [0.1, 1.0],
-    decision_threshold = nothing,
-    kwargs...,
+    loss::Union{Nothing,Symbol}=nothing,
+    complexity::Function=norm,
+    λ::Union{AbstractFloat,AbstractVector}=[0.1, 1.0],
+    decision_threshold=nothing,
+    kwargs...
 )
     params = ConformalGeneratorParams(; kwargs...)
     ConformalGenerator(loss, complexity, λ, decision_threshold, params.opt, params.τ, params.κ, params.temp)
@@ -69,12 +69,12 @@ function set_size_penalty(
     conf_model = counterfactual_explanation.M.model
     fitresult = counterfactual_explanation.M.fitresult
     X = CounterfactualExplanations.decode_state(counterfactual_explanation)
-    loss = SliceMap.slicemap(X, dims=(1,2)) do x
+    loss = SliceMap.slicemap(X, dims=(1, 2)) do x
         x = Matrix(x)
         ConformalPrediction.smooth_size_loss(
             conf_model, fitresult, x;
-            κ = generator.κ,
-            temp = generator.temp
+            κ=generator.κ,
+            temp=generator.temp
         )
     end
     loss = mean(loss)
@@ -107,7 +107,7 @@ function Generators.h(
     else
         Ω = 0
     end
-    
+
     if length(generator.λ) == 1
         penalty = generator.λ * (dist_ .+ Ω)
     else

### 2023-02-16 notebooks/conformal.qmd
uh

@@
 
+# `ConformalGenerator`
+
+In this section, we will look at a simple example involving synthetic data, a black-box model and a generic Conformal Counterfactual Generator.
+
+## Black-box Model
+
+We consider a simple binary classification problem.
+Let $(X_i, Y_i), \ i=1,...,n$ denote our feature-label pairs and let $\mu: \mathcal{X} \mapsto \mathcal{Y}$ denote the mapping from features to labels.
+For illustration purposes, we will use linearly separable data.
+
 ```{julia}
@@
 ```
+
+While we could use a linear classifier in this case, let's pretend we need a black-box model for this task and rely on a small Multi-Layer Perceptron (MLP):
 
@@
 
+We can fit this model to data to produce plug-in predictions.
+
+### Split Conformal Classification
+
+Here we will instead use a specific case of CP called *split conformal prediction* which can then be summarized as follows:^[In other places split conformal prediction is sometimes referred to as *inductive* conformal prediction.]
+
+1. Partition the training into a proper training set and a separate calibration set: $\mathcal{D}_n=\mathcal{D}^{\text{train}} \cup \mathcal{D}^{\text{cali}}$.
+2. Train the machine learning model on the proper training set: $\hat\mu_{i \in \mathcal{D}^{\text{train}}}(X_i,Y_i)$.
+
+The model $\hat\mu_{i \in \mathcal{D}^{\text{train}}}$ can now produce plug-in predictions.
+
+::: callout-note
+
+## Starting Point
+
+Note that this represents the starting point in applications of Algorithmic Recourse: we have some pre-trained classifier $M$ for which we would like to generate plausible Counterfactual Explanations.
+Next, we turn to the calibration step.
+:::
+
+3. Compute nonconformity scores, $\mathcal{S}$, using the calibration data $\mathcal{D}^{\text{cali}}$ and the fitted model $\hat\mu_{i \in \mathcal{D}^{\text{train}}}$. 
+4. For a user-specified desired coverage ratio $(1-\alpha)$ compute the corresponding quantile, $\hat{q}$, of the empirical distribution of nonconformity scores, $\mathcal{S}$.
+5. For the given quantile and test sample $X_{\text{test}}$, form the corresponding conformal prediction set:
+
+$$
+C(X_{\text{test}})=\{y:s(X_{\text{test}},y) \le \hat{q}\}
+$$ {#eq-set}
+
+This is the default procedure used for classification and regression in [`ConformalPrediction.jl`](https://github.com/pat-alt/ConformalPrediction.jl).
+
+Using the package, we can apply Split Conformal Prediction as follows:
+
 ```{julia}
@@
 
+To be clear, all of the calibration steps (3 to 5) are post hoc, and yet none of them involved any changes to the model parameters.
+These are two important characteristics of Split Conformal Prediction (SCP) that make it particularly useful in the context of Algorithmic Recourse.
+Firstly, the fact that SCP involves posthoc calibration steps that happen after training, ensures that we need not place any restrictions on the black-box model itself.
+This stands in contrast to the approach proposed by @schut2021generating in which they essentially restrict the class of models to Bayesian models.
+Secondly, the fact that the model itself is kept entirely intact ensures that the generated counterfactuals maintain fidelity to the model.
+Finally, note that we also have not resorted to a surrogate model to learn more about $X \sim \mathcal{X}$.
+Instead, we have used the fitted model itself and a calibration data set to learn about the model's predictive uncertainty.
+
 ## Counterfactual Explanation
+
+We first wrap our model in a container that makes it compatible with `CounterfactualExplanations.jl`.
+Then we draw a random sample, determine its predicted label $\hat{y}$ and choose the opposite label as our target.
 
@@
 M = CCE.ConformalModel(conf_model, mach.fitresult)
-```
-
-```{julia}
 x = select_factual(counterfactual_data,rand(1:size(counterfactual_data.X,2)))
@@
 
+The generic Conformal Counterfactual Generator works with the following counterfactual search objective,
+
+$$
+x^\prime = \arg \min_{x^\prime}  \ell(M(x^\prime),t) + \lambda \mathbb{I}_{y^\prime = t} \Omega(C_{\theta}(x;\tau)) 
+$$ {#eq-solution}
+
+where
+
+$$
+\Omega(C_{\theta}(x;\tau)) = = \max (0, \sum_k C_{\theta,k}(x;\tau) - \kappa)
+$$ {#eq-size-loss}
+
+is a smooth size penalty for conformal classifiers introduced by @stutz2022learning.
+Here, $C_{\theta,k}(x;\tau)$ is loosely defined as the probability that class $k$ is assigned to the conformal prediction set $C$.
+In the context of Conformal Training, this penalty reduces the **inefficiency** of the conformal predictor.
+In our context, we are not interested in improving the model itself, but rather in producing **plausible** counterfactuals.
+Provided that our counterfactual $x^\prime$ is already inside the target domain ($\mathbb{I}_{y^\prime = t}=1$), penalizing $\Omega(C_{\theta}(x;\tau))$ corresponds to guiding counterfactuals into regions of the target domain that are characterized by low ambiguity: for $\kappa=1$ the conformal prediction set includes only the target label $t$ as $\Omega(C_{\theta}(x;\tau))$.
+Arguably, less ambiguous counterfactuals are more **plausible**.
+Since the search is guided purely by properties of the model itself and (exchangeable) calibration data, counterfactuals also maintain **high fidelity**.
+
 ```{julia}
-p1 = contourf(mach.model, mach.fitresult, X, y; plot_classification_loss=true, target=target, zoom=0)
-p2 = contourf(mach.model, mach.fitresult, X, y; plot_set_loss=true, zoom=0)
+#| output: true
+#| echo: false
+#| label: fig-losses
+#| fig-cap: "Illustration of the smooth size loss and the configurable classification loss."
+
+p1 = contourf(mach.model, mach.fitresult, X, y; plot_set_loss=true, zoom=0)
+p2 = contourf(mach.model, mach.fitresult, X, y; plot_classification_loss=true, target=target, zoom=0)
 plot(p1, p2, size=(800,320))

### 2023-02-28 notebooks/conformal.qmd
uh
some work on configurable loss function

@@
 using CounterfactualExplanations.Data
+using CounterfactualExplanations.Objectives
 using Flux
@@
 
+### Penalizing Size
+
 ```{julia}
@@
 #| echo: false
+#| label: fig-ce
+#| fig-cap: "Comparison of counterfactuals produced using different generators."
 
+opt = Descent(0.01)
 ordered_names = [
@@
 ]
+loss_fun = Objectives.logitbinarycrossentropy
+
+# Generators:
+generators = Dict(
+    ordered_names[1] => GenericGenerator(opt = opt, decision_threshold=0.5),
+    ordered_names[2] => GenericGenerator(opt = opt, decision_threshold=0.9),
+    ordered_names[3] => CCE.ConformalGenerator(loss=loss_fun, opt=opt, λ=[0.1,1]),
+    ordered_names[4] => CCE.ConformalGenerator(loss=loss_fun, opt=opt, λ=[0.1,10]),
+)
+
+counterfactuals = Dict([name => generate_counterfactual(x, target, counterfactual_data, M, gen; initialization=:identity) for (name, gen) in generators])
+
+# Plots:
+plts = []
+for name ∈ ordered_names
+    ce = counterfactuals[name]
+    plt = plot(ce; title=name, colorbar=false, ticks = false, legend=false, zoom=0)
+    plts = vcat(plts..., plt)
+end
+_n = length(generators)
+img_size = 300
+plot(plts..., size=(_n * img_size,1.05*img_size), layout=(1,_n))
+```
+
+### Configurable Classification Loss
+
+```{julia}
+#| output: true
+#| echo: false
+#| label: fig-ce
+#| fig-cap: "Comparison of counterfactuals produced using different generators."
+
+opt = Descent(0.01)
+ordered_names = [
+    "Generic (γ=0.5)",
+    "Generic (γ=0.9)",
+    "Conformal (λ₂=1)",
+    "Conformal (λ₂=10)"
+]
+loss_fun = Objectives.logitbinarycrossentropy
 

### 2023-02-28 src/ConformalGenerator.jl
some work on configurable loss function

@@ -1,5 +1,8 @@
+using CategoricalArrays
 using CounterfactualExplanations
 using CounterfactualExplanations.Generators
+using CounterfactualExplanations.Models: binary_to_onehot
+using CounterfactualExplanations.Objectives
 using Flux
 using LinearAlgebra
 using Parameters
@@ -7,7 +10,7 @@ using SliceMap
 using Statistics
 
 mutable struct ConformalGenerator <: AbstractGradientBasedGenerator
-    loss::Union{Nothing,Symbol} # loss function
+    loss::Union{Nothing,Function} # loss function
     complexity::Function # complexity function
     λ::Union{AbstractFloat,AbstractVector} # strength of penalty
     decision_threshold::Union{Nothing,AbstractFloat}
@@ -26,13 +29,12 @@ end
 end
 
 """
-    ConformalGenerator(
-        ;
-        loss::Symbol=:logitbinarycrossentropy,
+    ConformalGenerator(;
+        loss::Union{Nothing,Function}=conformal_training_loss,
         complexity::Function=norm,
-        λ::AbstractFloat=0.1,
-        opt::Flux.Optimise.AbstractOptimiser=Flux.Optimise.Descent(),
-        τ::AbstractFloat=1e-5
+        λ::Union{AbstractFloat,AbstractVector}=[0.1, 1.0],
+        decision_threshold=nothing,
+        kwargs...
     )
 
 An outer constructor method that instantiates a generic generator.
@@ -43,7 +45,7 @@ generator = ConformalGenerator()
 ```
 """
 function ConformalGenerator(;
-    loss::Union{Nothing,Symbol}=nothing,
+    loss::Union{Nothing,Function}=conformal_training_loss,
     complexity::Function=norm,
     λ::Union{AbstractFloat,AbstractVector}=[0.1, 1.0],
     decision_threshold=nothing,
@@ -51,6 +53,32 @@ function ConformalGenerator(;
 )
     params = ConformalGeneratorParams(; kwargs...)
     ConformalGenerator(loss, complexity, λ, decision_threshold, params.opt, params.τ, params.κ, params.temp)
+end
+
+@doc raw"""
+    conformal_training_loss(counterfactual_explanation::AbstractCounterfactualExplanation; kwargs...)
+
+A configurable classification loss function for Conformal Predictors.
+"""
+function conformal_training_loss(counterfactual_explanation::AbstractCounterfactualExplanation; kwargs...)
+    conf_model = counterfactual_explanation.M.model
[…]

### 2023-02-28 notebooks/conformal.qmd
confused about domain of classification loss

@@
 using Flux
+using LinearAlgebra
 using MLJBase
@@
 #| fig-cap: "Illustration of the smooth size loss and the configurable classification loss."
-
-p1 = contourf(mach.model, mach.fitresult, X, y; plot_set_loss=true, zoom=0)
-p2 = contourf(mach.model, mach.fitresult, X, y; plot_classification_loss=true, target=target, zoom=0)
+temp = 0.5
+p1 = contourf(mach.model, mach.fitresult, X, y; plot_set_loss=true, zoom=0, temp=temp)
+p2 = contourf(mach.model, mach.fitresult, X, y; plot_classification_loss=true, target=target, zoom=0, temp=temp, clim=nothing, loss_matrix=ones(2,2))
 plot(p1, p2, size=(800,320))
@@
 opt = Descent(0.01)
+temp = 0.75
 ordered_names = [
@@
     ordered_names[2] => GenericGenerator(opt = opt, decision_threshold=0.9),
-    ordered_names[3] => CCE.ConformalGenerator(opt=opt, λ=[0.1,1]),
-    ordered_names[4] => CCE.ConformalGenerator(opt=opt, λ=[0.1,10]),
+    ordered_names[3] => CCE.ConformalGenerator(opt=opt, λ=[0.1,1], temp=temp),
+    ordered_names[4] => CCE.ConformalGenerator(opt=opt, λ=[0.1,10], temp=temp),
 )

### 2023-03-17 notebooks/fidelity.qmd
more on sampling

```{julia}
using CCE
using ConformalPrediction
using CounterfactualExplanations
using CounterfactualExplanations.Data
using CounterfactualExplanations.Objectives
using Distributions
using Flux
using JointEnergyModels
using LinearAlgebra
using MLJBase
using MLJFlux
using Plots
```

# Fidelity Measures

```{julia}
# Setup
counterfactual_data = load_multi_class()
M = fit_model(counterfactual_data, :MLP)
target = 4
factual = 1
chosen = rand(findall(predict_label(M, counterfactual_data) .== factual))
x = select_factual(counterfactual_data, chosen)
```

```{julia}
niter = 10
nsamples = 100
plts = []
for target in ce.data.y_levels
    # Search:
    generator = GenericGenerator()
    ce = generate_counterfactual(x, target, counterfactual_data, M, generator)
    sampler = CCE.EnergySampler(ce;niter=niter, nsamples=100)
    Xgen = rand(sampler, nsamples)
    plt = plot(M, counterfactual_data, target=ce.target, xlims=(-5,5),ylims=(-5,5),cbar=false)
    scatter!(Xgen[1,:],Xgen[2,:],alpha=0.5,color=target,shape=:star,label="X|y=$target")
    push!(plts, plt)
end
plot(plts..., layout=(1,length(ce.data.y_levels)), size=(length(ce.data.y_levels)*300,300))
```

### 2023-03-17 notebooks/conformal.qmd
more on sampling

@@
     "Generic (γ=0.5)",
-    "Generic (γ=0.9)",
     "Conformal (λ₂=1)",
@@
 loss_fun = Objectives.logitbinarycrossentropy
+generator = GenericGenerator(opt = opt)
 
@@
 generators = Dict(
-    ordered_names[1] => GenericGenerator(opt = opt, decision_threshold=0.5),
-    ordered_names[2] => GenericGenerator(opt = opt, decision_threshold=0.9),
-    ordered_names[3] => CCE.ConformalGenerator(loss=loss_fun, opt=opt, λ=[0.1,1]),
-    ordered_names[4] => CCE.ConformalGenerator(loss=loss_fun, opt=opt, λ=[0.1,10]),
+    ordered_names[1] => generator,
+    ordered_names[2] => deepcopy(generator) |> gen -> @objective(gen, _ + 0.1distance_l2 + 1.0set_size_penalty),
+    ordered_names[3] => deepcopy(generator) |> gen -> @objective(gen, _ + 0.1distance_l2 + 10.0set_size_penalty),
 )
 
-counterfactuals = Dict([name => generate_counterfactual(x, target, counterfactual_data, M, gen; initialization=:identity) for (name, gen) in generators])
+counterfactuals = Dict([name => generate_counterfactual(x, target, counterfactual_data, M, gen; initialization=:identity, converge_when=:generator_conditions, gradient_tol=1e-3) for (name, gen) in generators])
 

### 2023-03-17 src/penalties.jl

@@ -1,0 +1,43 @@
+using Statistics: mean
+
+"""
+    set_size_penalty(counterfactual_explanation::AbstractCounterfactualExplanation)
+
+Penalty for smooth conformal set size.
+"""
+function set_size_penalty(
+    counterfactual_explanation::AbstractCounterfactualExplanation; 
+    κ::Real=1.0, temp::Real=0.5, agg=mean
+)
+
+    conf_model = counterfactual_explanation.M.model
+    fitresult = counterfactual_explanation.M.fitresult
+    X = CounterfactualExplanations.decode_state(counterfactual_explanation)
+    loss = map(eachslice(X, dims=3)) do x
+        x = Matrix(x)
+        if target_probs(counterfactual_explanation, x)[1] >= 0.5
+            l = ConformalPrediction.smooth_size_loss(
+                conf_model, fitresult, x;
+                κ=κ,
+                temp=temp
+            )[1]
+        else 
+            l = 0.0
+        end
+        return l
+    end
+    loss = agg(loss)
+
+    return loss
+
+end
+
+function distance_from_energy(
+    counterfactual_explanation::AbstractCounterfactualExplanation; 
+    n::Int=100, retrain=false, kwargs...
+)
+    sampler = get!(counterfactual_explanation.params, :energy_sampler) do 
+        CCE.EnergySampler(counterfactual_explanation; kwargs...)
+    end
+    conditional_samples = 
+end

### 2023-03-17 src/sampling.jl

@@ -1,0 +1,43 @@
+using CounterfactualExplanations
+using Distributions
+using JointEnergyModels
+
+(model::AbstractFittedModel)(x) = logits(model, x)
+
+mutable struct EnergySampler
+    ce::CounterfactualExplanation
+    sampler::JointEnergyModels.ConditionalSampler
+    opt::JointEnergyModels.AbstractSamplingRule
+    buffer::AbstractArray
+end
+
+function EnergySampler(
+    ce::CounterfactualExplanation;
+    opt::JointEnergyModels.AbstractSamplingRule=ImproperSGLD(),
+    niter::Int=100,
+    nsamples::Int=1000
+)
+
+    # Setup:
+    model = ce.M
+    data = ce.data
+    K = length(data.y_levels)
+    𝒟x = Normal()
+    𝒟y = Categorical(ones(K) ./ K)
+    sampler = ConditionalSampler(𝒟x, 𝒟y)
+
+    # Fit:
+    i = get_target_index(data.y_levels, ce.target)
+    buffer = sampler(model.model, opt, (size(data.X, 1), nsamples); niter=niter, y=i)
+
+    return EnergySampler(ce, sampler, opt, buffer)
+end
+
+function Base.rand(sampler::EnergySampler, n::Int=100; retrain=false)
+    ntotal = size(sampler.buffer,2)
+    idx = rand(1:ntotal, n)
+    if !retrain
+        X = sampler.buffer[:,idx]
+    end
+    return X
+end

### 2023-03-17 notebooks/fidelity.qmd
bloody hell

@@
 ```{julia}
-niter = 10
+niter = 10000
 nsamples = 100

### 2023-03-17 src/sampling.jl
bloody hell

@@ -13,7 +13,7 @@ end
 
 function EnergySampler(
     ce::CounterfactualExplanation;
-    opt::JointEnergyModels.AbstractSamplingRule=ImproperSGLD(),
+    opt::JointEnergyModels.AbstractSamplingRule=SGLD(),
     niter::Int=100,
     nsamples::Int=1000
 )

### 2023-03-17 notebooks/fidelity.qmd
penalty working

--- 2023-03-17 notebooks/fidelity.qmd
@@
 target = 4
-factual = 1
+factual = 2
 chosen = rand(findall(predict_label(M, counterfactual_data) .== factual))
 x = select_factual(counterfactual_data, chosen)
+
+# Search:
+generator = GenericGenerator(opt=Descent(0.01))
+ce = generate_counterfactual(x, target, counterfactual_data, M, generator)
 ```
@@
 ```{julia}
-niter = 10
+niter = 100
 nsamples = 100
-plts = []
-for target in ce.data.y_levels
-    # Search:
-    generator = GenericGenerator()
-    ce = generate_counterfactual(x, target, counterfactual_data, M, generator)
+
 sampler = CCE.EnergySampler(ce;niter=niter, nsamples=100)
@@
 scatter!(Xgen[1,:],Xgen[2,:],alpha=0.5,color=target,shape=:star,label="X|y=$target")
-    push!(plts, plt)
-end
-plot(plts..., layout=(1,length(ce.data.y_levels)), size=(length(ce.data.y_levels)*300,300))
 ```
+
+```{julia}
+p1 = plot(ce)
+```
+
+```{julia}
+@objective(generator, _ + 0.1distance_l2 + 100.0distance_from_energy)
+ce = generate_counterfactual(x, target, counterfactual_data, M, generator)
+p2 = plot(ce)
+```

### 2023-03-17 src/sampling.jl

@@ -13,7 +13,7 @@ end
 
 function EnergySampler(
     ce::CounterfactualExplanation;
-    opt::JointEnergyModels.AbstractSamplingRule=SGLD(),
+    opt::JointEnergyModels.AbstractSamplingRule=ImproperSGLD(),
     niter::Int=100,
     nsamples::Int=1000
 )

### 2023-03-17 src/penalties.jl
penalty working

@@ -1,3 +1,5 @@
+using ChainRules: ignore_derivatives
+using LinearAlgebra: norm
 using Statistics: mean
 
 """
@@ -33,11 +35,28 @@ function set_size_penalty(
 end
 
 function distance_from_energy(
-    counterfactual_explanation::AbstractCounterfactualExplanation; 
-    n::Int=100, retrain=false, kwargs...
+    counterfactual_explanation::AbstractCounterfactualExplanation;
+    n::Int=100, retrain=false, agg=mean, kwargs...
 )
-    sampler = get!(counterfactual_explanation.params, :energy_sampler) do 
-        CCE.EnergySampler(counterfactual_explanation; kwargs...)
+    conditional_samples = []
+    ignore_derivatives() do
+        _dict = counterfactual_explanation.params
+        if !(:energy_sampler ∈ collect(keys(_dict)))
+            _dict[:energy_sampler] = CCE.EnergySampler(counterfactual_explanation; kwargs...)
+        end
+        sampler = _dict[:energy_sampler]
+        push!(conditional_samples, rand(sampler, n; retrain=retrain))
     end
-    conditional_samples = 
+    x′ = CounterfactualExplanations.counterfactual(counterfactual_explanation)
+    loss = map(eachslice(x′, dims=3)) do x
+        x = Matrix(x)
+        Δ = map(eachcol(conditional_samples[1])) do xsample
+            norm(x - xsample)
+        end
+        return mean(Δ)
+    end
+    loss = agg(loss)
+
+    return loss
+
 end

### 2023-03-18 notebooks/fidelity.qmd
differentiable penalties for fidelity and plausibility

@@
 
+## Binary
+
+```{julia}
+# Setup
+counterfactual_data = load_linearly_separable()
+M = fit_model(counterfactual_data, :DeepEnsemble)
+target = 2
+factual = 1
+chosen = rand(findall(predict_label(M, counterfactual_data) .== factual))
+x = select_factual(counterfactual_data, chosen)
+
+# Search:
+generator = GenericGenerator(opt=Descent(0.01))
+ce = generate_counterfactual(x, target, counterfactual_data, M, generator)
+```
+
+```{julia}
+niter = 100
+nsamples = 100
+
+sampler = CCE.EnergySampler(ce;niter=niter, nsamples=100)
+Xgen = rand(sampler, nsamples)
+plt = plot(M, counterfactual_data, target=ce.target, xlims=(-5,5),ylims=(-5,5),cbar=false)
+scatter!(Xgen[1,:],Xgen[2,:],alpha=0.5,color=target,shape=:star,label="X|y=$target")
+```
+
+## Multi-Class
+
 ```{julia}
@@
 counterfactual_data = load_multi_class()
-M = fit_model(counterfactual_data, :MLP)
+M = fit_model(counterfactual_data, :DeepEnsemble)
 target = 4
@@
 ```{julia}
-@objective(generator, _ + 0.1distance_l2 + 100.0distance_from_energy)
+using CCE: distance_from_energy
+@objective(generator, _ + 0.1distance_l2 + 10.0distance_from_energy)
 ce = generate_counterfactual(x, target, counterfactual_data, M, generator)
@@
 ```
+
+```{julia}
+using CCE: distance_from_targets
+@objective(
+    generator, 
+    _ + 0.1distance_l2 + 1.0distance_from_energy + 10.0distance_from_targets
+)
+ce = generate_counterfactual(x, target, counterfactual_data, M, generator)
+p3 = plot(ce)
+```

### 2023-03-18 src/penalties.jl

@@ -60,3 +60,24 @@ function distance_from_energy(
     return loss
 
 end
+
+function distance_from_targets(
+    counterfactual_explanation::AbstractCounterfactualExplanation;
+    n::Int=100, agg=mean
+)
+    target_samples = counterfactual_explanation.data.X |>
+        X -> X[:,rand(1:end,n)]
+    x′ = CounterfactualExplanations.counterfactual(counterfactual_explanation)
+    loss = map(eachslice(x′, dims=3)) do x
+        x = Matrix(x)
+        Δ = map(eachcol(target_samples)) do xsample
+            norm(x - xsample)
+        end
+        return mean(Δ)
+    end
+    loss = agg(loss)
+
+    return loss
+
+end
+

### 2023-03-18 src/sampling.jl
differentiable penalties for fidelity and plausibility

@@ -1,8 +1,9 @@
 using CounterfactualExplanations
 using Distributions
+using Flux
 using JointEnergyModels
 
-(model::AbstractFittedModel)(x) = logits(model, x)
+(model::AbstractFittedModel)(x) = log.(CounterfactualExplanations.predict_proba(model, nothing, x))
 
 mutable struct EnergySampler
     ce::CounterfactualExplanation
@@ -28,7 +29,7 @@ function EnergySampler(
 
     # Fit:
     i = get_target_index(data.y_levels, ce.target)
-    buffer = sampler(model.model, opt, (size(data.X, 1), nsamples); niter=niter, y=i)
+    buffer = sampler(model, opt, (size(data.X, 1), nsamples); niter=niter, y=i)
 
     return EnergySampler(ce, sampler, opt, buffer)
 end

### 2023-03-18 src/sampling.jl
slowly slowly

@@ -3,42 +3,113 @@ using Distributions
 using Flux
 using JointEnergyModels
 
+"""
+    (model::AbstractFittedModel)(x)
+
+When called on data `x`, softmax logits are returned. In the binary case, outputs are one-hot encoded.
+"""
 (model::AbstractFittedModel)(x) = log.(CounterfactualExplanations.predict_proba(model, nothing, x))
 
+"Base type that stores information relevant to energy-based posterior sampling from `AbstractFittedModel`."
 mutable struct EnergySampler
-    ce::CounterfactualExplanation
+    model::AbstractFittedModel
+    data::CounterfactualData
     sampler::JointEnergyModels.ConditionalSampler
     opt::JointEnergyModels.AbstractSamplingRule
-    buffer::AbstractArray
+    buffer::Union{Nothing,AbstractArray}
+    yidx::Union{Nothing,Any}
 end
 
+"""
+    EnergySampler(
+        model::AbstractFittedModel,
+        data::CounterfactualData,
+        y::Any;
+        opt::JointEnergyModels.AbstractSamplingRule=ImproperSGLD(),
+        niter::Int=100,
+        nsamples::Int=1000
+    )
+
+Constructor for `EnergySampler` that takes a `model`, `data` and conditioning value `y` as inputs.
+"""
 function EnergySampler(
-    ce::CounterfactualExplanation;
+    model::AbstractFittedModel,
+    data::CounterfactualData,
+    y::Any;
     opt::JointEnergyModels.AbstractSamplingRule=ImproperSGLD(),
     niter::Int=100,
     nsamples::Int=1000
 )
 
-    # Setup:
-    model = ce.M
-    data = ce.data
+    @assert y ∈ data.y_levels || y ∈ 1:length(data.y_levels)
     K = length(data.y_levels)
     𝒟x = Normal()
     𝒟y = Categorical(ones(K) ./ K)
     sampler = ConditionalSampler(𝒟x, 𝒟y)
+    yidx = get_target_index(data.y_levels, y)
 
-    # Fit:
-    i = get_target_index(data.y_levels, ce.target)
-    buffer = sampler(model, opt, (size(data.X, 1), nsamples); niter=niter, y=i)
+    # Initiate:
+    energy_sampler = EnergySampler(model, data, sampler, opt, nothing, nothing)
 
-    return EnergySampler(ce, sampler, opt, buffer)
+    # Generate samples:
[…]

### 2023-03-18 notebooks/fidelity.qmd
slowly slowly

@@
 using CCE
+using CCE: set_size_penalty, distance_from_energy, distance_from_targets
 using ConformalPrediction
@@
 using Plots
+img_height = 300
 ```
 
-# Fidelity Measures
+# `ConformalGenerator`
 
-## Binary
+In this section, we will look at a simple example involving synthetic data, a black-box model and a generic Conformal Counterfactual Generator.
+
+## Black-box Model
+
+We consider a simple binary classification problem.
+Let $(X_i, Y_i), \ i=1,...,n$ denote our feature-label pairs and let $\mu: \mathcal{X} \mapsto \mathcal{Y}$ denote the mapping from features to labels.
+For illustration purposes, we will use linearly separable data.
 
 ```{julia}
-# Setup
 counterfactual_data = load_linearly_separable()
-M = fit_model(counterfactual_data, :DeepEnsemble)
-target = 2
-factual = 1
-chosen = rand(findall(predict_label(M, counterfactual_data) .== factual))
-x = select_factual(counterfactual_data, chosen)
-
-# Search:
-generator = GenericGenerator(opt=Descent(0.01))
-ce = generate_counterfactual(x, target, counterfactual_data, M, generator)
 ```
 
+While we could use a linear classifier in this case, let's pretend we need a black-box model for this task and rely on a small Multi-Layer Perceptron (MLP):
+
 ```{julia}
+builder = MLJFlux.@builder Chain(
+    Dense(n_in, 32, relu),
+    Dense(32, n_out)
+)
+clf = NeuralNetworkClassifier(builder=builder, epochs=100)
+```
+
+We can fit this model to data to produce plug-in predictions.
+
+## Conformal Prediction
+
+Here we will instead use a specific case of CP called *split conformal prediction* which can then be summarized as follows:^[In other places split conformal prediction is sometimes referred to as *inductive* conformal prediction.]
+
+1. Partition the training into a proper training set and a separate calibration set: $\mathcal{D}_n=\mathcal{D}^{\text{train}} \cup \mathcal{D}^{\text{cali}}$.
+2. Train the machine learning model on the proper training set: $\hat\mu_{i \in \mathcal{D}^{\text{train}}}(X_i,Y_i)$.
+
+The model $\hat\mu_{i \in \mathcal{D}^{\text{train}}}$ can now produce plug-in predictions.
+
+::: callout-note
+
+## Starting Point
+
+Note that this represents the starting point in applications of Algorithmic Recourse: we have some pre-trained classifier $M$ for which we would like to generate plausible Counterfactual Explanations.
+Next, we turn to the calibration step.
+:::
+
+3. Compute nonconformity scores, $\mathcal{S}$, using the calibration data $\mathcal{D}^{\text{cali}}$ and the fitted model $\hat\mu_{i \in \mathcal{D}^{\text{train}}}$. 
+4. For a user-specified desired coverage ratio $(1-\alpha)$ compute the corresponding quantile, $\hat{q}$, of the empirical distribution of nonconformity scores, $\mathcal{S}$.
+5. For the given quantile and test sample $X_{\text{test}}$, form the corresponding conformal prediction set:
+
+$$
+C(X_{\text{test}})=\{y:s(X_{\text{test}},y) \le \hat{q}\}
+$$ {#eq-set}
+
+This is the default procedure used for classification and regression in [`ConformalPrediction.jl`](https://github.com/pat-alt/ConformalPrediction.jl).
+
+Using the package, we can apply Split Conformal Prediction as follows:
+
+```{julia}
+X = table(permutedims(counterfactual_data.X))
+y =  counterfactual_data.output_encoder.labels
+conf_model = conformal_model(clf; method=:simple_inductive)
+mach = machine(conf_model, X, y)
+fit!(mach)
+```
+
+To be clear, all of the calibration steps (3 to 5) are post hoc, and yet none of them involved any changes to the model parameters.
+These are two important characteristics of Split Conformal Prediction (SCP) that make it particularly useful in the context of Algorithmic Recourse.
+Firstly, the fact that SCP involves posthoc calibration steps that happen after training, ensures that we need not place any restrictions on the black-box model itself.
+This stands in contrast to the approach proposed by @schut2021generating in which they essentially restrict the class of models to Bayesian models.
+Secondly, the fact that the model itself is kept entirely intact ensures that the generated counterfactuals maintain fidelity to the model.
+Finally, note that we also have not resorted to a surrogate model to learn more about $X \sim \mathcal{X}$.
+Instead, we have used the fitted model itself and a calibration data set to learn about the model's predictive uncertainty.
+
+## Differentiable CP
+
+In order to use CP in the context of gradient-based counterfactual search, we need it to be differentiable.
+@stutz2022learning introduce a framework for training differentiable conformal predictors.
+They introduce a configurable loss function as well as smooth set size penalty.
+
+### Smooth Set Size Penalty
+
+Starting with the former, @stutz2022learning propose the following:
+
+$$
+\Omega(C_{\theta}(x;\tau)) = = \max (0, \sum_k C_{\theta,k}(x;\tau) - \kappa)
+$$ {#eq-size-loss}
+
+Here, $C_{\theta,k}(x;\tau)$ is loosely defined as the probability that class $k$ is assigned to the conformal prediction set $C$.
+In the context of Conformal Training, this penalty reduces the **inefficiency** of the conformal predictor.
+
+In our context, we are not interested in improving the model itself, but rather in producing **plausible** counterfactuals.
+Provided that our counterfactual $x^\prime$ is already inside the target domain ($\mathbb{I}_{y^\prime = t}=1$), penalizing $\Omega(C_{\theta}(x;\tau))$ corresponds to guiding counterfactuals into regions of the target domain that are characterized by low ambiguity: for $\kappa=1$ the conformal prediction set includes only the target label $t$ as $\Omega(C_{\theta}(x;\tau))$.
+Arguably, less ambiguous counterfactuals are more **plausible**.
+Since the search is guided purely by properties of the model itself and (exchangeable) calibration data, counterfactuals also maintain **high fidelity**.
+
+The left panel of @fig-losses shows the smooth size penalty in the two-dimensional feature space of our synthetic data.
+
+### Configurable Classification Loss
+
+The right panel of @fig-losses shows the configurable classification loss in the two-dimensional feature space of our synthetic data.
+
+```{julia}
+#| output: true
+#| echo: false
+#| label: fig-losses
+#| fig-cap: "Illustration of the smooth size loss and the configurable classification loss."
+
+temp = 0.5
+p1 = contourf(mach.model, mach.fitresult, X, y; plot_set_loss=true, zoom=0, temp=temp)
+p2 = contourf(mach.model, mach.fitresult, X, y; plot_classification_loss=true, target=target, zoom=0, temp=temp, clim=nothing, loss_matrix=ones(2,2))
+plot(p1, p2, size=(800,320))
+```
+
+## Fidelity and Plausibility
+
+The main evaluation criteria we are interested in are *fidelity* and *plausibility*. Interestingly, we could also consider using these measures as penalties in the counterfactual search.
+
+### Fidelity
+
+We propose to define fidelity as follows:
+
+::: {#def-fidelity}
+
+## High-Fidelity Counterfactuals
+
+Let $\mathcal{X}_{\theta}|y = p_{\theta}(X|y)$ denote the class-conditional distribution of $X$ defined by $\theta$.
+Then for $x^{\prime}$ to be considered a high-fidelity counterfactual, we need: $\mathcal{X}_{\theta}|t \approxeq \mathcal{X}^{\prime}$ where $t$ denotes the target outcome.
+
+:::
+
+We can generate samples from $p_{\theta}(X|y)$ following @grathwohl2020your. In @fig-energy, I have applied the methodology to our synthetic data.
+
+```{julia}
+#| fig-cap: "Energy-based conditional samples."
+#| label: fig-energy
+
 niter = 100
@@
 
-sampler = CCE.EnergySampler(ce;niter=niter, nsamples=100)
+plts = []
+for target ∈ counterfactual_data.y_levels
+    sampler = CCE.EnergySampler(M, counterfactual_data, target; niter=niter, nsamples=100)
     Xgen = rand(sampler, nsamples)
-plt = plot(M, counterfactual_data, target=ce.target, xlims=(-5,5),ylims=(-5,5),cbar=false)
+    plt = plot(M, counterfactual_data; target=target, zoom=-3,cbar=false)
     scatter!(Xgen[1,:],Xgen[2,:],alpha=0.5,color=target,shape=:star,label="X|y=$target")
+    push!(plts, plt)
+end
+plot(plts..., layout=(1,length(plts)), size=(img_height*length(plts),img_height))
+```
+
+As an evaluation metric and penalty, we could use the average distance of the counterfactual $x^{\prime}$ from these generated samples, for example.
+
+### Plausibility
+
+We propose to define plausibility as follows:
+
+::: {#def-plausible}
+
+## Plausible Counterfactuals
+
+Formally, let $\mathcal{X}|t$ denote the conditional distribution of samples in the target class.
+As before, we have $x^{\prime}\sim\mathcal{X}^{\prime}$, then for $x^{\prime}$ to be considered a plausible counterfactual, we need: $\mathcal{X}|t \approxeq \mathcal{X}^{\prime}$.
+
+:::
+
+As an evaluation metric and penalty, we could use the average distance of the counterfactual $x^{\prime}$ from (potentially bootstrapped) training samples in the target class, for example.
+
+## Counterfactual Explanations
+
+Next, let's generate counterfactual explanations for our synthetic data.
+We first wrap our model in a container that makes it compatible with `CounterfactualExplanations.jl`.
+Then we draw a random sample, determine its predicted label $\hat{y}$ and choose the opposite label as our target.
+
+```{julia}
+M = CCE.ConformalModel(conf_model, mach.fitresult)
+x = select_factual(counterfactual_data,rand(1:size(counterfactual_data.X,2)))
+y_factual = predict_label(M, counterfactual_data, x)[1]
+target = counterfactual_data.y_levels[counterfactual_data.y_levels .!= y_factual][1]
+```
+
+The generic Conformal Counterfactual Generator penalises the only the set size only:
+
+$$
+x^\prime = \arg \min_{x^\prime}  \ell(M(x^\prime),t) + \lambda \mathbb{I}_{y^\prime = t} \Omega(C_{\theta}(x;\tau)) 
+$$ {#eq-solution}
+
+```{julia}
+#| output: true
+#| echo: false
+#| label: fig-ce
+#| fig-cap: "Comparison of counterfactuals produced using different generators."
+
+opt = Descent(0.01)
+ordered_names = [
+    "Generic (γ=0.5)",
+    "Conformal (λ₂=1)",
+    "Conformal (λ₂=10)"
+]
+loss_fun = Objectives.logitbinarycrossentropy
+generator = GenericGenerator(opt = opt)
+
+# Generators:
+generators = Dict(
+    ordered_names[1] => generator,
+    ordered_names[2] => deepcopy(generator) |> gen -> @objective(gen, _ + 0.1distance_l2 + 1.0set_size_penalty),
+    ordered_names[3] => deepcopy(generator) |> gen -> @objective(gen, _ + 0.1distance_l2 + 10.0set_size_penalty),
+)
+
+counterfactuals = Dict([name => generate_counterfactual(x, target, counterfactual_data, M, gen; initialization=:identity, converge_when=:generator_conditions, gradient_tol=1e-3) for (name, gen) in generators])
+
+# Plots:
+plts = []
+for name ∈ ordered_names
+    ce = counterfactuals[name]
+    plt = plot(ce; title=name, colorbar=false, ticks = false, legend=false, zoom=0)
+    plts = vcat(plts..., plt)
+end
+_n = length(generators)
+img_size = 300
+plot(plts..., size=(_n * img_size,1.05*img_size), layout=(1,_n))
 ```

### 2023-03-18 src/model.jl

@@ -56,16 +56,16 @@ In the binary case logits are fed through the sigmoid function instead of softma
 which follows from the derivation here: https://stats.stackexchange.com/questions/233658/softmax-vs-sigmoid-function-in-logistic-classifier
 """
 function Models.logits(M::ConformalModel, X::AbstractArray)
-    yhat = SliceMap.slicemap(X, dims=(1, 2)) do x
-        conf_model = M.model
-        fitresult = M.fitresult
-        # x = MLJBase.table(permutedims(x))
-        # p̂ = MMI.predict(conf_model.model, fitresult, MMI.reformat(conf_model.model, x)...)
-        # p̂ = map(p̂) do pp
-        #     L = p̂.decoder.classes
-        #     probas = pdf.(pp, L)
-        #     return probas
-        # end
+    conf_model = M.model
+    fitresult = M.fitresult
+    # x = MLJBase.table(permutedims(x))
+    # p̂ = MMI.predict(conf_model.model, fitresult, MMI.reformat(conf_model.model, x)...)
+    # p̂ = map(p̂) do pp
+    #     L = p̂.decoder.classes
+    #     probas = pdf.(pp, L)
+    #     return probas
+    # end
+    function predict_logits(fitresult, x)
         p̂ = fitresult[1](x)
         if ndims(p̂) == 2
             p̂ = [p̂]
@@ -77,6 +77,13 @@ function Models.logits(M::ConformalModel, X::AbstractArray)
         end
         ŷ = ndims(ŷ) > 1 ? ŷ : permutedims([ŷ])
         return ŷ
+    end
+    if ndims(X) > 2
+        yhat = map(eachslice(X, dims=ndims(X))) do x
+            predict_logits(fitresult, x)
+        end
+    else
+        yhat = predict_logits(fitresult, X)
     end
     return yhat
 end

### 2023-03-18 notebooks/fidelity.qmd
uh

@@
 p1 = contourf(mach.model, mach.fitresult, X, y; plot_set_loss=true, zoom=0, temp=temp)
-p2 = contourf(mach.model, mach.fitresult, X, y; plot_classification_loss=true, target=target, zoom=0, temp=temp, clim=nothing, loss_matrix=ones(2,2))
+p2 = contourf(mach.model, mach.fitresult, X, y; plot_classification_loss=true, target=1, zoom=0, temp=temp, clim=nothing, loss_matrix=ones(2,2))
 plot(p1, p2, size=(800,320))
@@
 
+λ = 10.0
+
 opt = Descent(0.01)
 ordered_names = [
-    "Generic (γ=0.5)",
-    "Conformal (λ₂=1)",
-    "Conformal (λ₂=10)"
+    "Distance x",
+    "Set Size",
+    "Distance X_θ|t",
+    "Distance X|t"
 ]
@@
     ordered_names[1] => generator,
-    ordered_names[2] => deepcopy(generator) |> gen -> @objective(gen, _ + 0.1distance_l2 + 1.0set_size_penalty),
-    ordered_names[3] => deepcopy(generator) |> gen -> @objective(gen, _ + 0.1distance_l2 + 10.0set_size_penalty),
+    ordered_names[2] => deepcopy(generator) |> gen -> @objective(gen, _ + 0.1distance + 10.0set_size_penalty),
+    ordered_names[3] => deepcopy(generator) |> gen -> @objective(gen, _ + 0.1distance + 10.0distance_from_energy),
+    ordered_names[4] => deepcopy(generator) |> gen -> @objective(gen, _ + 0.1distance + 10.0distance_from_targets),
 )
@@
 _n = length(generators)
-img_size = 300
-plot(plts..., size=(_n * img_size,1.05*img_size), layout=(1,_n))
+plot(plts..., size=(_n * img_height,1.05*img_height), layout=(1,_n))
 ```

### 2023-03-18 notebooks/fidelity.qmd
more work on notebook

@@
     ce = counterfactuals[name]
-    plt = plot(ce; title=name, colorbar=false, ticks = false, legend=false, zoom=0)
+    plt = plot(ce; title=name, colorbar=false, ticks = false, legend=false, zoom=-0.5)
+    if :energy_sampler ∈ collect(keys(ce.params))
+        Xgen = ce.params[:energy_sampler].buffer
+        scatter!(Xgen[1,:],Xgen[2,:],alpha=0.5,color=target,shape=:star,label="X|y=$target")
+    end
     plts = vcat(plts..., plt)

### 2023-03-18 src/sampling.jl
more work on notebook

@@ -38,7 +38,7 @@ function EnergySampler(
     y::Any;
     opt::JointEnergyModels.AbstractSamplingRule=ImproperSGLD(),
     niter::Int=100,
-    nsamples::Int=1000
+    nsamples::Int=100
 )
 
     @assert y ∈ data.y_levels || y ∈ 1:length(data.y_levels)
@@ -64,6 +64,7 @@ Generates `n` samples from `EnergySampler` for conditioning value `y`.
 """
 function generate_samples(e::EnergySampler, n::Int, y::Int; niter::Int=100)
     X = e.sampler(e.model, e.opt, (size(e.data.X, 1), n); niter=niter, y=y)
+    X = X[:,map(x -> !any(isnan.(x)), eachcol(X))]
     return X
 end
 

### 2023-03-20 dev/experimentation.qmd
benchmark set up

--- 2023-03-18 notebooks/fidelity.qmd
@@
 
+### Benchmark
+
+```{julia}
+#| output: false
+
+# Data:
+datasets = Dict(
+    :linearly_separable => load_linearly_separable(),
+    :overlapping => load_overlapping(),
+    :moons => load_moons(),
+    :circles => load_circles(),
+)
+
+# Models:
+models = Dict(
+    :MLP => M,
+)
+```
+
+Then we can simply loop over the datasets and eventually concatenate the results like so:
+
+```{julia}
+#| output: false
+
+measure = 
+using CounterfactualExplanations.Evaluation: benchmark
+bmks = []
+for (dataname, dataset) in datasets
+    bmk = benchmark(dataset; models=models, generators=generators, suppress_training=true, data_name=dataname)
+    push!(bmks, bmk)
+end
+bmk = reduce(vcat, bmks)
+```
+
 ## Multi-Class
@@
 ```{julia}
-# Setup
 counterfactual_data = load_multi_class()
-M = fit_model(counterfactual_data, :DeepEnsemble)
-target = 4
-factual = 2
-chosen = rand(findall(predict_label(M, counterfactual_data) .== factual))
-x = select_factual(counterfactual_data, chosen)
-
-# Search:
-generator = GenericGenerator(opt=Descent(0.01))
-ce = generate_counterfactual(x, target, counterfactual_data, M, generator)
 ```
@@
 ```{julia}
+X = table(permutedims(counterfactual_data.X))
+y =  counterfactual_data.output_encoder.labels
+conf_model = conformal_model(clf; method=:simple_inductive)
+mach = machine(conf_model, X, y)
+fit!(mach)
+```
+
+```{julia}
+#| output: true
+#| echo: false
+#| label: fig-losses-multi
+#| fig-cap: "Illustration of the smooth size loss and the configurable classification loss."
+
+temp = 0.5
+p1 = contourf(mach.model, mach.fitresult, X, y; plot_set_loss=true, zoom=0, temp=temp)
+p2 = contourf(mach.model, mach.fitresult, X, y; plot_classification_loss=true, target=1, zoom=0, temp=temp, clim=nothing, loss_matrix=ones(4,4))
+plot(p1, p2, size=(800,320))
+```
+
+```{julia}
+#| fig-cap: "Energy-based conditional samples."
+#| label: fig-energy-multi
+
 niter = 100
@@
 
-sampler = CCE.EnergySampler(ce;niter=niter, nsamples=100)
+plts = []
+for target ∈ counterfactual_data.y_levels
+    sampler = CCE.EnergySampler(M, counterfactual_data, target; niter=niter, nsamples=100)
     Xgen = rand(sampler, nsamples)
-plt = plot(M, counterfactual_data, target=ce.target, xlims=(-5,5),ylims=(-5,5),cbar=false)
+    plt = plot(M, counterfactual_data; target=target, zoom=-3,cbar=false)
     scatter!(Xgen[1,:],Xgen[2,:],alpha=0.5,color=target,shape=:star,label="X|y=$target")
+    push!(plts, plt)
+end
+plot(plts..., layout=(1,length(plts)), size=(img_height*length(plts),img_height))
 ```
@@
 ```{julia}
-p1 = plot(ce)
+M = CCE.ConformalModel(conf_model, mach.fitresult; likelihood=:classification_multi)
+x = select_factual(counterfactual_data,rand(1:size(counterfactual_data.X,2)))
+y_factual = predict_label(M, counterfactual_data, x)[1]
+target = counterfactual_data.y_levels[counterfactual_data.y_levels .!= y_factual][1]
 ```
@@
 ```{julia}
-using CCE: distance_from_energy
-@objective(generator, _ + 0.1distance_l2 + 10.0distance_from_energy)
-ce = generate_counterfactual(x, target, counterfactual_data, M, generator)
-p2 = plot(ce)
+counterfactuals = Dict([name => generate_counterfactual(x, target, counterfactual_data, M, gen; initialization=:identity, converge_when=:generator_conditions, gradient_tol=1e-3) for (name, gen) in generators])
+
+# Plots:
+plts = []
+for name ∈ ordered_names
+    ce = counterfactuals[name]
+    plt = plot(ce; title=name, colorbar=false, ticks = false, legend=false, zoom=-0.5)
+    if :energy_sampler ∈ collect(keys(ce.params))
+        Xgen = ce.params[:energy_sampler].buffer
+        scatter!(Xgen[1,:],Xgen[2,:],alpha=0.5,color=target,shape=:star,label="X|y=$target")
+    end
+    plts = vcat(plts..., plt)
+end
+_n = length(generators)
+plot(plts..., size=(_n * img_height,1.05*img_height), layout=(1,_n))
 ```
-
-```{julia}
-using CCE: distance_from_targets
-@objective(
-    generator, 
-    _ + 0.1distance_l2 + 1.0distance_from_energy + 10.0distance_from_targets
-)
-ce = generate_counterfactual(x, target, counterfactual_data, M, generator)
-p3 = plot(ce)
-```

### 2023-03-20 notebooks/intro.qmd
setting up notebooks

--- 2023-03-20 dev/experimentation.qmd
@@
 ```{julia}
-using CCE
-using CCE: set_size_penalty, distance_from_energy, distance_from_targets
-using ConformalPrediction
-using CounterfactualExplanations
-using CounterfactualExplanations.Data
-using CounterfactualExplanations.Objectives
-using Distributions
-using Flux
-using JointEnergyModels
-using LinearAlgebra
-using MLJBase
-using MLJFlux
-using Plots
-img_height = 300
+include("notebooks/setup.jl")
+eval(setup_notebooks)
 ```
@@
 
+M = CCE.ConformalModel(conf_model, mach.fitresult)
+
 niter = 100
@@
 ```{julia}
-M = CCE.ConformalModel(conf_model, mach.fitresult)
 x = select_factual(counterfactual_data,rand(1:size(counterfactual_data.X,2)))
@@
 nsamples = 100
+M = CCE.ConformalModel(conf_model, mach.fitresult; likelihood=:classification_multi)
 
@@
     Xgen = rand(sampler, nsamples)
-    plt = plot(M, counterfactual_data; target=target, zoom=-3,cbar=false)
+    plt = plot(M, counterfactual_data; target=target, zoom=-0.5,cbar=false)
     scatter!(Xgen[1,:],Xgen[2,:],alpha=0.5,color=target,shape=:star,label="X|y=$target")
@@
 ```{julia}
-M = CCE.ConformalModel(conf_model, mach.fitresult; likelihood=:classification_multi)
+
 x = select_factual(counterfactual_data,rand(1:size(counterfactual_data.X,2)))

### 2023-03-20 notebooks/proposal.qmd
setting up notebooks

--- 2023-02-13 dev/proposal.qmd
@@
 ```{julia}
-using CounterfactualExplanations
-using CounterfactualExplanations.Data: load_mnist
-using CounterfactualExplanations.Models: load_mnist_mlp
-using Images
-using MLDatasets
-using MLDatasets: convert2image
-using Plots
-www_path = "dev/www"
+include("notebooks/setup.jl")
+eval(setup_notebooks)
 ```

### 2023-03-21 paper/paper.tex
uh

\documentclass{article}

% if you need to pass options to natbib, use, e.g.:
%     \PassOptionsToPackage{numbers, compress}{natbib}
% before loading neurips_2022

% ready for submission
\usepackage{neurips_2022}

% to compile a preprint version, e.g., for submission to arXiv, add add the
% [preprint] option:
%     \usepackage[preprint]{neurips_2022}

% to compile a camera-ready version, add the [final] option, e.g.:
%     \usepackage[final]{neurips_2022}

% to avoid loading the natbib package, add option nonatbib:
%    \usepackage[nonatbib]{neurips_2022}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{xcolor}         % colors

\title{High-Fidelity Counterfactual Explanations through Conformal Prediction}

% The \author macro works with any number of authors. There are two commands
% used to separate the names and addresses of multiple authors: \And and \AND.
%
% Using \And between authors leaves it to LaTeX to determine where to break the
% lines. Using \AND forces a line break at that point. So, if LaTeX puts 3 of 4
% authors names on the first line, and the last on the second line, try using
% \AND instead of \And before the third author name.

\author{%
  David S.~Hippocampus\thanks{Use footnote for providing further information
    about author (webpage, alternative address)---\emph{not} for acknowledging
    funding agencies.} \\
  Department of Computer Science\\
  Cranberry-Lemon University\\
  Pittsburgh, PA 15213 \\
  \texttt{hippo@cs.cranberry-lemon.edu} \\
  % examples of more authors
  % \And
  % Coauthor \\
  % Affiliation \\
  % Address \\
  % \texttt{email} \\
  % \AND
  % Coauthor \\
  % Affiliation \\
  % Address \\
  % \texttt{email} \\
  % \And
  % Coauthor \\
  % Affiliation \\
  % Address \\
  % \texttt{email} \\
  % \And
  % Coauthor \\
  % Affiliation \\
  % Address \\
  % \texttt{email} \\
}

\begin{document}

\maketitle

\begin{abstract}
  The abstract paragraph should be indented \nicefrac{1}{2}~inch (3~picas) on
  both the left- and right-hand margins. Use 10~point type, with a vertical
  spacing (leading) of 11~points.  The word \textbf{Abstract} must be centered,
  bold, and in point size 12. Two line spaces precede the abstract. The abstract
  must be limited to one paragraph.
\end{abstract}

\section{Submission of papers to NeurIPS 2022}

Please read the instructions below carefully and follow them faithfully.

\subsection{Style}

Papers to be submitted to NeurIPS 2022 must be prepared according to the
instructions presented here. Papers may only be up to {\bf nine} pages long,
including figures. Additional pages \emph{containing only acknowledgments and
references} are allowed. Papers that exceed the page limit will not be
reviewed, or in any other way considered for presentation at the conference.

The margins in 2022 are the same as those in 2007, which allow for $\sim$$15\%$
more words in the paper compared to earlier years.

Authors are required to use the NeurIPS \LaTeX{} style files obtainable at the
NeurIPS website as indicated below. Please make sure you use the current files
and not previous versions. Tweaking the style files may be grounds for
rejection.

\subsection{Retrieval of style files}

The style files for NeurIPS and other conference information are available on
the World Wide Web at
\begin{center}
  \url{http://www.neurips.cc/}
\end{center}
The file \verb+neurips_2022.pdf+ contains these instructions and illustrates the
various formatting requirements your NeurIPS paper must satisfy.

The only supported style file for NeurIPS 2022 is \verb+neurips_2022.sty+,
rewritten for \LaTeXe{}.  \textbf{Previous style files for \LaTeX{} 2.09,
  Microsoft Word, and RTF are no longer supported!}

The \LaTeX{} style file contains three optional arguments: \verb+final+, which
creates a camera-ready copy, \verb+preprint+, which creates a preprint for
submission to, e.g., arXiv, and \verb+nonatbib+, which will not load the
\verb+natbib+ package for you in case of package clash.

\paragraph{Preprint option}
If you wish to post a preprint of your work online, e.g., on arXiv, using the
NeurIPS style, please use the \verb+preprint+ option. This will create a
nonanonymized version of your work with the text ``Preprint. Work in progress.''
in the footer. This version may be distributed as you see fit. Please \textbf{do
  not} use the \verb+final+ option, which should \textbf{only} be used for
papers accepted to NeurIPS.

At submission time, please omit the \verb+final+ and \verb+preprint+
options. This will anonymize your submission and add line numbers to aid
review. Please do \emph{not} refer to these line numbers in your paper as they
will be removed during generation of camera-ready copies.

The file \verb+neurips_2022.tex+ may be used as a ``shell'' for writing your
paper. All you have to do is replace the author, title, abstract, and text of
the paper with your own.

The formatting instructions contained in these style files are summarized in
Sections \ref{gen_inst}, \ref{headings}, and \ref{others} below.

\section{General formatting instructions}
\label{gen_inst}

The text must be confined within a rectangle 5.5~inches (33~picas) wide and
9~inches (54~picas) long. The left margin is 1.5~inch (9~picas).  Use 10~point
type with a vertical spacing (leading) of 11~points.  Times New Roman is the
preferred typeface throughout, and will be selected for you by default.
Paragraphs are separated by \nicefrac{1}{2}~line space (5.5 points), with no
indentation.

The paper title should be 17~point, initial caps/lower case, bold, centered
between two horizontal rules. The top rule should be 4~points thick and the
bottom rule should be 1~point thick. Allow \nicefrac{1}{4}~inch space above and
below the title to rules. All pages should start at 1~inch (6~picas) from the
top of the page.

For the final version, authors' names are set in boldface, and each name is
centered above the corresponding address. The lead author's name is to be listed
first (left-most), and the co-authors' names (if different address) are set to
follow. If there is only one co-author, list both author and co-author side by
side.

Please pay special attention to the instructions in Section \ref{others}
regarding figures, tables, acknowledgments, and references.

\section{Headings: first level}
\label{headings}

All headings should be lower case (except for first word and proper nouns),
flush left, and bold.

First-level headings should be in 12-point type.

\subsection{Headings: second level}

Second-level headings should be in 10-point type.

\subsubsection{Headings: third level}

Third-level headings should be in 10-point type.

\paragraph{Paragraphs}

There is also a \verb+\paragraph+ command available, which sets the heading in
bold, flush left, and inline with the text, with the heading followed by 1\,em
of space.

\section{Citations, figures, tables, references}
\label{others}

These instructions apply to everyone.

\subsection{Citations within the text}

The \verb+natbib+ package will be loaded for you by default.  Citations may be
author/year or numeric, as long as you maintain internal consistency.  As to the
format of the references themselves, any style is acceptable as long as it is
used consistently.

The documentation for \verb+natbib+ may be found at
\begin{center}
  \url{http://mirrors.ctan.org/macros/latex/contrib/natbib/natnotes.pdf}
\end{center}
Of note is the command \verb+\citet+, which produces citations appropriate for
use in inline text.  For example,
\begin{verbatim}
   \citet{hasselmo} investigated\dots
\end{verbatim}
produces
\begin{quote}
  Hasselmo, et al.\ (1995) investigated\dots
\end{quote}

If you wish to load the \verb+natbib+ package with options, you may add the
following before loading the \verb+neurips_2022+ package:
\begin{verbatim}
   \PassOptionsToPackage{options}{natbib}
\end{verbatim}

If \verb+natbib+ clashes with another package you load, you can add the optional
argument \verb+nonatbib+ when loading the style file:
\begin{verbatim}
   \usepackage[nonatbib]{neurips_2022}
\end{verbatim}

As submission is double blind, refer to your own published work in the third
person. That is, use ``In the previous work of Jones et al.\ [4],'' not ``In our
previous work [4].'' If you cite your other papers that are not widely available
(e.g., a journal paper under review), use anonymous author names in the
citation, e.g., an author of the form ``A.\ Anonymous.''

\subsection{Footnotes}

Footnotes should be used sparingly.  If you do require a footnote, indicate
footnotes with a number\footnote{Sample of the first footnote.} in the
text. Place the footnotes at the bottom of the page on which they appear.
Precede the footnote with a horizontal rule of 2~inches (12~picas).

Note that footnotes are properly typeset \emph{after} punctuation
marks.\footnote{As in this example.}

\subsection{Figures}

\begin{figure}
  \centering
  \fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
  \caption{Sample figure caption.}
\end{figure}

All artwork must be neat, clean, and legible. Lines should be dark enough for
purposes of reproduction. The figure number and caption always appear after the
figure. Place one line space before the figure caption and one line space after
the figure. The figure caption should be lower case (except for first word and
proper nouns); figures are numbered consecutively.

You may use color figures.  However, it is best for the figure captions and the
paper body to be legible if the paper is printed in either black/white or in
color.

\subsection{Tables}

All tables must be centered, neat, clean and legible.  The table number and
title always appear before the table.  See Table~\ref{sample-table}.

Place one line space before the table title, one line space after the
table title, and one line space after the table. The table title must
be lower case (except for first word and proper nouns); tables are
numbered consecutively.

Note that publication-quality tables \emph{do not contain vertical rules.} We
strongly suggest the use of the \verb+booktabs+ package, which allows for
typesetting high-quality, professional tables:
\begin{center}
  \url{https://www.ctan.org/pkg/booktabs}
\end{center}
This package was used to typeset Table~\ref{sample-table}.

\begin{table}
  \caption{Sample table title}
  \label{sample-table}
  \centering
  \begin{tabular}{lll}
    \toprule
    \multicolumn{2}{c}{Part}                   \\
    \cmidrule(r){1-2}
    Name     & Description     & Size ($\mu$m) \\
    \midrule
    Dendrite & Input terminal  & $\sim$100     \\
    Axon     & Output terminal & $\sim$10      \\
    Soma     & Cell body       & up to $10^6$  \\
    \bottomrule
  \end{tabular}
\end{table}

\section{Final instructions}

Do not change any aspects of the formatting parameters in the style files.  In
particular, do not modify the width or length of the rectangle the text should
fit into, and do not change font sizes (except perhaps in the
\textbf{References} section; see below). Please note that pages should be
numbered.

\section{Preparing PDF files}

Please prepare submission files with paper size ``US Letter,'' and not, for
example, ``A4.''

Fonts were the main cause of problems in the past years. Your PDF file must only
contain Type 1 or Embedded TrueType fonts. Here are a few instructions to
achieve this.

\begin{itemize}

\item You should directly generate PDF files using \verb+pdflatex+.

\item You can check which fonts a PDF files uses.  In Acrobat Reader, select the
  menu Files$>$Document Properties$>$Fonts and select Show All Fonts. You can
  also use the program \verb+pdffonts+ which comes with \verb+xpdf+ and is
  available out-of-the-box on most Linux machines.

\item The IEEE has recommendations for generating PDF files whose fonts are also
  acceptable for NeurIPS. Please see
  \url{http://www.emfield.org/icuwb2010/downloads/IEEE-PDF-SpecV32.pdf}

\item \verb+xfig+ "patterned" shapes are implemented with bitmap fonts.  Use
  "solid" shapes instead.

\item The \verb+\bbold+ package almost always uses bitmap fonts.  You should use
  the equivalent AMS Fonts:
\begin{verbatim}
   \usepackage{amsfonts}
\end{verbatim}
followed by, e.g., \verb+\mathbb{R}+, \verb+\mathbb{N}+, or \verb+\mathbb{C}+
for $\mathbb{R}$, $\mathbb{N}$ or $\mathbb{C}$.  You can also use the following
workaround for reals, natural and complex:
\begin{verbatim}
   \newcommand{\RR}{I\!\!R} %real numbers
   \newcommand{\Nat}{I\!\!N} %natural numbers
   \newcommand{\CC}{I\!\!\!\!C} %complex numbers
\end{verbatim}
Note that \verb+amsfonts+ is automatically loaded by the \verb+amssymb+ package.

\end{itemize}

If your file contains type 3 fonts or non embedded TrueType fonts, we will ask
you to fix it.

\subsection{Margins in \LaTeX{}}

Most of the margin problems come from figures positioned by hand using
\verb+\special+ or other commands. We suggest using the command
\verb+\includegraphics+ from the \verb+graphicx+ package. Always specify the
figure width as a multiple of the line width as in the example below:
\begin{verbatim}
   \usepackage[pdftex]{graphicx} ...
   \includegraphics[width=0.8\linewidth]{myfile.pdf}
\end{verbatim}
See Section 4.4 in the graphics bundle documentation
(\url{http://mirrors.ctan.org/macros/latex/required/graphics/grfguide.pdf})

A number of width problems arise when \LaTeX{} cannot properly hyphenate a
line. Please give LaTeX hyphenation hints using the \verb+\-+ command when
necessary.

\begin{ack}
Use unnumbered first level headings for the acknowledgments. All acknowledgments
go at the end of the paper before the list of references. Moreover, you are required to declare
funding (financial activities supporting the submitted work) and competing interests (related financial activities outside the submitted work).
More information about this disclosure can be found at: \url{https://neurips.cc/Conferences/2022/PaperInformation/FundingDisclosure}.

Do {\bf not} include this section in the anonymized submission, only in the final paper. You can use the \texttt{ack} environment provided in the style file to autmoatically hide this section in the anonymized submission.
\end{ack}

\section*{References}

References follow the acknowledgments. Use unnumbered first-level heading for
the references. Any choice of citation style is acceptable as long as you are
consistent. It is permissible to reduce the font size to \verb+small+ (9 point)
when listing the references.
Note that the Reference section does not count towards the page limit.
\medskip

{
\small

[1] Alexander, J.A.\ \& Mozer, M.C.\ (1995) Template-based algorithms for
connectionist rule extraction. In G.\ Tesauro, D.S.\ Touretzky and T.K.\ Leen
(eds.), {\it Advances in Neural Information Processing Systems 7},
pp.\ 609--616. Cambridge, MA: MIT Press.

[2] Bower, J.M.\ \& Beeman, D.\ (1995) {\it The Book of GENESIS: Exploring
  Realistic Neural Models with the GEneral NEural SImulation System.}  New York:
TELOS/Springer--Verlag.

[3] Hasselmo, M.E., Schnell, E.\ \& Barkai, E.\ (1995) Dynamics of learning and
recall at excitatory recurrent synapses and cholinergic modulation in rat
hippocampal region CA3. {\it Journal of Neuroscience} {\bf 15}(7):5249-5262.
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Checklist}

%%% BEGIN INSTRUCTIONS %%%
The checklist follows the references.  Please
read the checklist guidelines carefully for information on how to answer these
questions.  For each question, change the default \answerTODO{} to \answerYes{},
\answerNo{}, or \answerNA{}.  You are strongly encouraged to include a {\bf
justification to your answer}, either by referencing the appropriate section of
your paper or providing a brief inline description.  For example:
\begin{itemize}
  \item Did you include the license to the code and datasets? \answerYes{See Section~\ref{gen_inst}.}
  \item Did you include the license to the code and datasets? \answerNo{The code and the data are proprietary.}
  \item Did you include the license to the code and datasets? \answerNA{}
\end{itemize}
Please do not modify the questions and only use the provided macros for your
answers.  Note that the Checklist section does not count towards the page
limit.  In your paper, please delete this instructions block and only keep the
Checklist section heading above along with the questions/answers below.
%%% END INSTRUCTIONS %%%

\begin{enumerate}

\item For all authors...
\begin{enumerate}
  \item Do the main claims made in the abstract and introduction accurately reflect the paper's contributions and scope?
    \answerTODO{}
  \item Did you describe the limitations of your work?
    \answerTODO{}
  \item Did you discuss any potential negative societal impacts of your work?
    \answerTODO{}
  \item Have you read the ethics review guidelines and ensured that your paper conforms to them?
    \answerTODO{}
\end{enumerate}

\item If you are including theoretical results...
\begin{enumerate}
  \item Did you state the full set of assumptions of all theoretical results?
    \answerTODO{}
        \item Did you include complete proofs of all theoretical results?
    \answerTODO{}
\end{enumerate}

\item If you ran experiments...
\begin{enumerate}
  \item Did you include the code, data, and instructions needed to reproduce the main experimental results (either in the supplemental material or as a URL)?
    \answerTODO{}
  \item Did you specify all the training details (e.g., data splits, hyperparameters, how they were chosen)?
    \answerTODO{}
        \item Did you report error bars (e.g., with respect to the random seed after running experiments multiple times)?
    \answerTODO{}
        \item Did you include the total amount of compute and the type of resources used (e.g., type of GPUs, internal cluster, or cloud provider)?
    \answerTODO{}
\end{enumerate}

\item If you are using existing assets (e.g., code, data, models) or curating/releasing new assets...
\begin{enumerate}
  \item If your work uses existing assets, did you cite the creators?
    \answerTODO{}
  \item Did you mention the license of the assets?
    \answerTODO{}
  \item Did you include any new assets either in the supplemental material or as a URL?
    \answerTODO{}
  \item Did you discuss whether and how consent was obtained from people whose data you're using/curating?
    \answerTODO{}
  \item Did you discuss whether the data you are using/curating contains personally identifiable information or offensive content?
    \answerTODO{}
\end{enumerate}

\item If you used crowdsourcing or conducted research with human subjects...
\begin{enumerate}
  \item Did you include the full text of instructions given to participants and screenshots, if applicable?
    \answerTODO{}
  \item Did you describe any potential participant risks, with links to Institutional Review Board (IRB) approvals, if applicable?
    \answerTODO{}
  \item Did you include the estimated hourly wage paid to participants and the total amount spent on participant compensation?
    \answerTODO{}
\end{enumerate}

\end{enumerate}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\appendix

\section{Appendix}

Optionally include extra information (complete proofs, additional experiments and plots) in the appendix.
This section will often be part of the supplemental material.

\end{document}

### 2023-03-21 paper/paper.tex
set up on overleaf

@@
 more words in the paper compared to earlier years.
+
+\cite{abadie2002instrumental}
 
@@
 Note that the Reference section does not count towards the page limit.
+
 \medskip
 
-{
-\small
-
-[1] Alexander, J.A.\ \& Mozer, M.C.\ (1995) Template-based algorithms for
-connectionist rule extraction. In G.\ Tesauro, D.S.\ Touretzky and T.K.\ Leen
-(eds.), {\it Advances in Neural Information Processing Systems 7},
-pp.\ 609--616. Cambridge, MA: MIT Press.
-
-[2] Bower, J.M.\ \& Beeman, D.\ (1995) {\it The Book of GENESIS: Exploring
-  Realistic Neural Models with the GEneral NEural SImulation System.}  New York:
-TELOS/Springer--Verlag.
-
-[3] Hasselmo, M.E., Schnell, E.\ \& Barkai, E.\ (1995) Dynamics of learning and
-recall at excitatory recurrent synapses and cholinergic modulation in rat
-hippocampal region CA3. {\it Journal of Neuroscience} {\bf 15}(7):5249-5262.
-}
+\bibliography{paper/references}
 

### 2023-03-21 paper/paper.tex
fair to say I hate latex
trying to sort out reference
references sorted. Now starting writing

@@
 \author{%
-  David S.~Hippocampus\thanks{Use footnote for providing further information
+  Patrick Altmeyer\thanks{Use footnote for providing further information
     about author (webpage, alternative address)---\emph{not} for acknowledging
@@
 \begin{abstract}
-  The abstract paragraph should be indented \nicefrac{1}{2}~inch (3~picas) on
-  both the left- and right-hand margins. Use 10~point type, with a vertical
-  spacing (leading) of 11~points.  The word \textbf{Abstract} must be centered,
-  bold, and in point size 12. Two line spaces precede the abstract. The abstract
-  must be limited to one paragraph.
+  We propose Conformal Counterfactual Explanations: an effortless and rigorous way to produce realistic and faithful Counterfactual Explanations for Black-Box models using Conformal Prediction.
+To address the need for realistic counterfactuals, existing work has primarily relied on surrogate models to learn the data-generating process.
+While this is an effective way to produce plausible counterfactual explanations, it effectively reallocates the task of learning realistic representations of the data from the model itself to the surrogate.
+Consequently, the generated explanations may look compelling to human inspectors but not necessarily accurately reflect the behaviour of the Black-Box model.
+Recent work has shown that instead of relying on surrogate models to generate plausible explanations, we can leverage predictive uncertainty for this task.
+Unfortunately, this approach works under the assumption that models can produce predictive uncertainty estimates.
+By leveraging a novel approach to differentiable Conformal Prediction we show that it is possible to relax that assumption.
 \end{abstract}
@@
 more words in the paper compared to earlier years.
-
-\cite{abadie2002instrumental}
 
@@
 
-\bibliography{paper/references}
+\bibliographystyle{plain}
+\bibliography{../bib}
 

### 2023-03-21 notebooks/intro.qmd
references sorted. Now starting writing

@@
 for (dataname, dataset) in datasets
-    bmk = benchmark(dataset; models=models, generators=generators, suppress_training=true, data_name=dataname)
+    bmk = benchmark(dataset; models=models, generators=generators, suppress_training=true, dataname=dataname)
     push!(bmks, bmk)

### 2023-03-21 paper/paper.tex
small improvement of benchmark function
quarto striking, but intro in tex file done

--- 2023-03-21 paper/paper.tex
@@
 % ready for submission
-\usepackage{neurips_2022}
+% \usepackage{neurips_2022}
 
@@
 % [preprint] option:
-%     \usepackage[preprint]{neurips_2022}
+\usepackage[preprint]{neurips_2022}
 
@@
 \begin{abstract}
-  The abstract paragraph should be indented \nicefrac{1}{2}~inch (3~picas) on
-  both the left- and right-hand margins. Use 10~point type, with a vertical
-  spacing (leading) of 11~points.  The word \textbf{Abstract} must be centered,
-  bold, and in point size 12. Two line spaces precede the abstract. The abstract
-  must be limited to one paragraph.
+  We propose Conformal Counterfactual Explanations: an effortless and rigorous way to produce plausible and conformal Counterfactual Explanations for Black Box Models using Conformal Prediction.
+To address the need for plausible explanations, existing work has primarily relied on surrogate models to learn the data-generating process.
+This effectively reallocates the task of learning realistic representations of the data from the model itself to the surrogate.
+Consequently, the generated explanations may look plausible to humans but not necessarily conform with the behaviour of the Black Box Model.
+We formalise this notion through the introduction of new evaluation measures.
+In order to still address the need for plausibility, we build on a recent approach that works by minimizing predictive model uncertainty.
+Using differentiable Conformal Prediction, we relax the previous assumption that the Black Box Model can produce predictive uncertainty estimates.
 \end{abstract}
+
+\section{Introduction}\label{intro}
+
+Counterfactual Explanations are a powerful, flexible and intuitive way to not only explain Black Box Models but also enable affected individuals to challenge them through the means of Algorithmic Recourse.
+Instead of opening the black box, Counterfactual Explanations work under the premise of strategically perturbing model inputs to understand model behaviour \cite{wachter2017counterfactual}.
+Intuitively speaking, we generate explanations in this context by asking simple what-if questions of the following nature: `Our credit risk model currently predicts that this individual's credit profile is too risky to offer them a loan.
+What if they reduced their monthly expenditures by 10\%?
+Will our model then predict that the individual is credit-worthy'?
+
+This is typically implemented by defining a target outcome $t \in \mathcal{Y}$ for some individual $x \in \mathcal{X}$, for which the model $f: \mathcal{X} \mapsto \mathcal{Y}$ initially predicts a different outcome: $f(x)\ne t$.
+Counterfactuals are then searched by minimizing a loss function that compares the predicted model output to the target outcome: $\ell(f(x),t)$.
+Since Counterfactual Explanations (CE) work directly with the Black Box Model, they always have full local fidelity by construction.
+Fidelity is defined as the degree to which explanations approximate the predictions of the Black Box Model.
+This arguably one of the most important evaluation metrics for model explanations, since any explanation that explains a prediction not actually made by the model is useless \cite{molnar2020interpretable}.
+
+In situations where full fidelity is a requirement, CE therefore offers a more appropriate solution to Explainable Artificial Intelligence (XAI) than other popular approaches like LIME \cite{ribeiro2016why} and SHAP \cite{lundberg2017unified}, which involve local surrogate models.
+But even full fidelity is not a sufficient condition for ensuring that an explanation adequately describes the behaviour of a model.
+That is because two very distinct explanations can both lead to the same model prediction, especially when dealing with heavily parameterized models:
+
+\begin{quotation}
+  […] deep neural networks are typically very underspecified by the available data, and […] parameters [therefore] correspond to a diverse variety of compelling explanations for the data.
+  --- \cite[Wilson 2020]{wilson2020case}
+\end{quotation}
+
+When people talk about Black Box Models, this is usually the type of model they have in mind.
+
+In the context of CE, the idea that no two explanations are the same arises almost naturally.
+Even the baseline approach proposed by \cite[Wachter et al.]{wachter2017counterfactual} can yield a diverse set of explanations if counterfactuals are intialised randomly.
+This multiplicity of explanations has not only been acknowledged in the literature but positively embraced: since individuals seeking Algorithmic Recourse (AR) have unique preferences,~\cite[Mothilal et al.]{mothilal2020explaining}, for example, have prescribed \textit{diversity} as an explicit goal for counterfactuals.
+More generally, the literature on CE and AR has brought forward a myriad of desiderata for explanations, which we will discuss in more detail in the following section.
+
+\section{Adversarial Example or Plausible Explanation?}\label{background}
 

### 2023-03-21 paper/paper.tex
not sure why year isn't showing up in citations

@@
 \usepackage{xcolor}         % colors
+\usepackage{amsmath}
 
-\title{High-Fidelity Counterfactual Explanations through Conformal Prediction}
+\title{Conformal Counterfactual Explanations}
 
@@
     funding agencies.} \\
-  Department of Computer Science\\
-  Cranberry-Lemon University\\
-  Pittsburgh, PA 15213 \\
-  \texttt{hippo@cs.cranberry-lemon.edu} \\
+  Faculty of Electrical Engineering, Mathematics and Computer Science\\
+  Delft University of Technology\\
+  2628 XE Delft, The Netherlands \\
+  \texttt{p.altmeyer@tudelft.nl} \\
   % examples of more authors
@@
 Counterfactual Explanations are a powerful, flexible and intuitive way to not only explain Black Box Models but also enable affected individuals to challenge them through the means of Algorithmic Recourse.
-Instead of opening the black box, Counterfactual Explanations work under the premise of strategically perturbing model inputs to understand model behaviour \cite{wachter2017counterfactual}.
+Instead of opening the black box, Counterfactual Explanations work under the premise of strategically perturbing model inputs to understand model behaviour \citep{wachter2017counterfactual}.
 Intuitively speaking, we generate explanations in this context by asking simple what-if questions of the following nature: `Our credit risk model currently predicts that this individual's credit profile is too risky to offer them a loan.
@@
 
-This is typically implemented by defining a target outcome $t \in \mathcal{Y}$ for some individual $x \in \mathcal{X}$, for which the model $f: \mathcal{X} \mapsto \mathcal{Y}$ initially predicts a different outcome: $f(x)\ne t$.
-Counterfactuals are then searched by minimizing a loss function that compares the predicted model output to the target outcome: $\ell(f(x),t)$.
+This is typically implemented by defining a target outcome $t \in \mathcal{Y}$ for some individual $x \in \mathcal{X}$, for which the model $M:\mathcal{X}\mapsto\mathcal{Y}$ initially predicts a different outcome: $M(x)\ne t$.
+Counterfactuals are then searched by minimizing a loss function that compares the predicted model output to the target outcome: $\text{yloss}(M(x),t)$.
 Since Counterfactual Explanations (CE) work directly with the Black Box Model, they always have full local fidelity by construction.
 Fidelity is defined as the degree to which explanations approximate the predictions of the Black Box Model.
-This arguably one of the most important evaluation metrics for model explanations, since any explanation that explains a prediction not actually made by the model is useless \cite{molnar2020interpretable}.
+This arguably one of the most important evaluation metrics for model explanations, since any explanation that explains a prediction not actually made by the model is useless \citep{molnar2020interpretable}.
 
-In situations where full fidelity is a requirement, CE therefore offers a more appropriate solution to Explainable Artificial Intelligence (XAI) than other popular approaches like LIME \cite{ribeiro2016why} and SHAP \cite{lundberg2017unified}, which involve local surrogate models.
+In situations where full fidelity is a requirement, CE therefore offers a more appropriate solution to Explainable Artificial Intelligence (XAI) than other popular approaches like LIME \citep{ribeiro2016why} and SHAP \citep{lundberg2017unified}, which involve local surrogate models.
 But even full fidelity is not a sufficient condition for ensuring that an explanation adequately describes the behaviour of a model.
@@
   […] deep neural networks are typically very underspecified by the available data, and […] parameters [therefore] correspond to a diverse variety of compelling explanations for the data.
-  --- \cite[Wilson 2020]{wilson2020case}
+  --- \citet{wilson2020case}
 \end{quotation}
@@
 In the context of CE, the idea that no two explanations are the same arises almost naturally.
-Even the baseline approach proposed by \cite[Wachter et al.]{wachter2017counterfactual} can yield a diverse set of explanations if counterfactuals are intialised randomly.
-This multiplicity of explanations has not only been acknowledged in the literature but positively embraced: since individuals seeking Algorithmic Recourse (AR) have unique preferences,~\cite[Mothilal et al.]{mothilal2020explaining}, for example, have prescribed \textit{diversity} as an explicit goal for counterfactuals.
+Even the baseline approach proposed by \citet{wachter2017counterfactual} can yield a diverse set of explanations if counterfactuals are initialised randomly.
+This multiplicity of explanations has not only been acknowledged in the literature but positively embraced: since individuals seeking Algorithmic Recourse (AR) have unique preferences,~\citet{mothilal2020explaining}, for example, have prescribed \textit{diversity} as an explicit goal for counterfactuals.
 More generally, the literature on CE and AR has brought forward a myriad of desiderata for explanations, which we will discuss in more detail in the following section.
 
-\section{Adversarial Example or Plausible Explanation?}\label{background}
+\section{Adversarial Example or Valid Explanation?}\label{background}
+
+Most state-of-the-art approaches to generating Counterfactual Explanations (CE) rely on gradient descent to optimize different flavours of the same counterfactual search objective,
+
+\begin{equation} \label{eq:general}
+\begin{aligned}
+\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}} \left\{  {\text{yloss}(M(f(\mathbf{s}^\prime)),y^*)}+ \lambda {\text{cost}(f(\mathbf{s}^\prime)) }  \right\} 
+\end{aligned} 
+\end{equation}
+
+where $\text{yloss}$ denotes the primary loss function already introduced above and $\text{cost}$ is either a single penalty or a collection of penalties that are used to impose constraints through regularization.
+Following the convention in \citet{altmeyer2023endogenous} we use $\mathbf{s}^\prime=\{ s_k\}_K$ to denote the vector $K$-dimensional array of counterfactual states.
+This is to explicitly account for the fact that we can generate multiple counterfactuals, as with DiCE \citep{mothilal2020explaining}, and may choose to traverse a latent representation $\mathcal{Z}$ of the feature space $\mathcal{X}$, as we will discuss further below.
+
+Solutions to Equation~\ref{eq:general} are considered valid as soon as the predicted label matches the target label.
+A stripped-down counterfactual explanation is therefore little different from an adversarial example.
+In, for example, generic counterfactual search as in \citet{wachter2017counterfactual} has been applied to MNIST data.
+
+\begin{figure}
+  \centering
+  \fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
+  \caption{Sample figure caption.}
+\end{figure}
+
+To properly serve both AI practitioners and individuals affected by AI decision-making systems, Counterfactual Explanations should have certain desirable properties, sometimes referred to as \textit{desiderata}.
+Besides diversity, which we already introduced above, some of the most prominent desiderata include sparsity, proximity \citep{wachter2017counterfactual}, actionability~\citep{ustun2019actionable}, plausibility \citep{joshi2019realistic,poyiadzi2020face,schut2021generating}, robustness \citep{upadhyay2021robust,pawelczyk2022probabilistically,altmeyer2023endogenous} and causality~\citep{karimi2021algorithmic}.
+Researchers have come up with various ways to meet these desiderata, which have been surveyed in~\citep{verma2020counterfactual} and~\citep{karimi2020survey}.
 
@@
 
-\bibliographystyle{plain}
+\bibliographystyle{plainnat}
 \bibliography{../bib}

### 2023-03-22 paper/paper.tex
sorted out citations. Had to convert bibliograhy to bibtex
progress on paper

@@
 \usepackage{amsmath}
+\usepackage{graphicx}
+
+% Bibliography
+\bibliographystyle{plainnat}
+\setcitestyle{numbers,square,comma}
 
@@
 A stripped-down counterfactual explanation is therefore little different from an adversarial example.
-In, for example, generic counterfactual search as in \citet{wachter2017counterfactual} has been applied to MNIST data.
-
-\begin{figure}
-  \centering
-  \fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
-  \caption{Sample figure caption.}
-\end{figure}
+In Figure~\ref{fig:adv}, for example, generic counterfactual search as in \citet{wachter2017counterfactual} has been applied to MNIST data.
 
@@
 Researchers have come up with various ways to meet these desiderata, which have been surveyed in~\citep{verma2020counterfactual} and~\citep{karimi2020survey}.
+
+\begin{figure}
+  \centering
+  \begin{minipage}[t]{0.45\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{www/you_may_not_like_it.png}
+    \caption{You may not like it, but this is what stripped-down counterfactuals look like.
+Counterfactuals for turning an 8 (eight) into a 3 (three) using different methodologies.
+Left: Using the generic approach proposed in \citet{wachter2017counterfactual}.
+Center: Using REVISE \citep{joshi2019realistic} with a well-specified surrogate.
+Right: Using REVISE \citep{joshi2019realistic} with a poorly specified surrogate.}\label{fig:adv}
+  \end{minipage}\hfill
+  \begin{minipage}[t]{0.45\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{www/mnist_9to4_latent.png}
+    \caption{Using surrogates introduces a dependency.
+Counterfactuals for turning a 9 (nine) into a 4 (four) using REVISE: original image (left); counterfactual using a well-specified surrogate (center); and counterfactual using a poorly specified surrogate (right).}\label{fig:vae}
+  \end{minipage}
+\end{figure}
+
+\medskip
+
+\bibliography{../bib}
 
@@
 
-\medskip
-
-\bibliographystyle{plainnat}
-\bibliography{../bib}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2023-03-22 notebooks/proposal.qmd
some finetuning of intro examples

@@
 M = load_mnist_mlp()
+
 # Target:
@@
 factual = predict_label(M, counterfactual_data, x)[1]
-# Search:
-n_ce = 3
-generator = GenericGenerator()
-ces = generate_counterfactual(x, target, counterfactual_data, M, generator; num_counterfactuals=n_ce)
+γ = 0.9
+T = 50
 ```
@@
 ```{julia}
-image_size = 200
+# Search:
+generator = GenericGenerator()
+ce_wachter = generate_counterfactual(
+    x, target, counterfactual_data, M, generator; 
+    decision_threshold=γ, max_iter=T,
+    initialization=:identity,
+)
+generator = GreedyGenerator(η=5.0)
+ce_jsma = generate_counterfactual(
+    x, target, counterfactual_data, M, generator; 
+    decision_threshold=γ, max_iter=T,
+    initialization=:identity,
+)
+```
+
+```{julia}
 p1 = plot(
@@
     axis=nothing, 
-    size=(image_size, image_size),
+    size=(img_height, img_height),
     title="Factual"
@@
 
-counterfactuals = CounterfactualExplanations.counterfactual(ces)
-phat = target_probs(ces)
+ces = zip([ce_wachter,ce_jsma])
+counterfactuals = reduce((x,y)->cat(x,y,dims=3),map(ce -> CounterfactualExplanations.counterfactual(ce[1]), ces))
+phat = reduce((x,y) -> cat(x,y,dims=3), map(ce -> target_probs(ce[1]), ces))
 for x in zip(eachslice(counterfactuals; dims=3), eachslice(phat; dims=3))
@@
         axis=nothing, 
-        size=(image_size, image_size),
+        size=(img_height, img_height),
         title=_title
@@
 end
-plt = plot(plts...; size=(image_size * (n_ce + 1),image_size), layout=(1,(n_ce + 1)))
+plt = plot(plts...; size=(img_height*length(plts),img_height), layout=(1,length(plts)))
 savefig(plt, joinpath(www_path, "you_may_not_like_it.png"))
@@
 In this latter case, the decoder step of the VAE fails to yield values in $\mathcal{X}$ and hence the counterfactual search in the learned latent space is doomed.
+
+```{julia}
+using CounterfactualExplanations.Models: load_mnist_vae
+vae = load_mnist_vae()
+vae_weak = load_mnist_vae(;strong=false)
+Serialization.serialize(joinpath(output_path,"mnist_classifier.jls"), M)
+Serialization.serialize(joinpath(output_path,"mnist_vae.jls"), vae)
+Serialization.serialize(joinpath(output_path,"mnist_vae_weak.jls"), vae_weak)
+```
+
+```{julia}
+# Define generator:
+generator = REVISEGenerator(
+  opt = Descent(0.1),
+  λ=0.01
+)
+# Generate recourse:
+counterfactual_data.generative_model = vae # assign generative model
+ce = generate_counterfactual(
+    x, target, counterfactual_data, M, generator; 
+    decision_threshold=γ, max_iter=T,
+    initialization=:identity,
+)
+counterfactual_data = deepcopy(counterfactual_data)
+counterfactual_data.generative_model = vae_weak
+ce_weak = generate_counterfactual(
+    x, target, counterfactual_data, M, generator;
+    decision_threshold=γ, max_iter=T,
+    initialization=:identity,
+)
+```
+
+```{julia}
+ces = zip([ce,ce_weak])
+counterfactuals = reduce((x,y)->cat(x,y,dims=3),map(ce -> CounterfactualExplanations.counterfactual(ce[1]), ces))
+phat = reduce((x,y) -> cat(x,y,dims=3), map(ce -> target_probs(ce[1]), ces))
+plts = [p1]
+for x in zip(eachslice(counterfactuals; dims=3), eachslice(phat; dims=3))
+    ce, _phat = (x[1],x[2])
+    _title = "p(y=$(target)|x′)=$(round(_phat[1]; digits=3))"
+    plt = plot(
+        convert2image(MNIST, reshape(ce,28,28)),
+        axis=nothing, 
+        size=(img_height, img_height),
+        title=_title
+    )
+    plts = [plts..., plt]
+end
+plt = plot(plts...; size=(img_height*length(plts),img_height), layout=(1,length(plts)))
+savefig(plt, joinpath(www_path, "surrogate_gone_wrong.png"))
+```
 

### 2023-03-22 paper/paper.tex
some finetuning of intro examples

--- 2023-03-21 paper/paper.tex
@@
 Researchers have come up with various ways to meet these desiderata, which have been surveyed in~\citep{verma2020counterfactual} and~\citep{karimi2020survey}.
+
+\begin{figure}
+  \centering
+  \begin{minipage}[t]{0.45\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{../www/you_may_not_like_it.png}
+    \caption{You may not like it, but this is what stripped-down counterfactuals look like.
+Counterfactuals for turning an 8 (eight) into a 3 (three): original image (left); counterfactual produced using \citet{wachter2017counterfactual} (center); and a counterfactual produced using JSMA-based approach introduced by \citep{schut2021generating}.}\label{fig:adv}
+  \end{minipage}\hfill
+  \begin{minipage}[t]{0.45\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{../www/surrogate_gone_wrong.png}
+    \caption{Using surrogates can improve plausibility, but also increases vulnerability.
+Counterfactuals for turning an 8 (eight) into a 3 (three): original image (left); counterfactual produced using REVISE \citep{joshi2019realistic} with a well-specified surrogate (center); and a counterfactual produced using REVISE \citep{joshi2019realistic} with a poorly specified surrogate (right).}\label{fig:vae}
+  \end{minipage}
+\end{figure}
+
+\medskip
+
+\bibliography{../bib}
 

### 2023-03-22 paper/paper.tex
Updates from Overleaf
more work on paper

--- 2023-03-21 paper/paper.tex
@@
 
-\section{Adversarial Example or Valid Explanation?}\label{background}
+\section{From Adversarial Examples to Plausible Explanations}\label{background}
 
@@
 A stripped-down counterfactual explanation is therefore little different from an adversarial example.
-In, for example, generic counterfactual search as in \citet{wachter2017counterfactual} has been applied to MNIST data.
+In Figure~\ref{fig:adv}, for example, we have the baseline approach proposed in \citet{wachter2017counterfactual} to MNIST data (centre panel).
+This approach solves Equation~\ref{eq:general} through gradient-descent in the feature space with a penalty for the distance between the factual $x$ and the counterfactual $x^{\prime}$.
+The underlying classifier $M$ is a simple Multi-Layer Perceptron (MLP) with good test accuracy.
+For the generated counterfactual $x^{\prime}$ the model predicts the target label with high confidence (centre panel in Figure~\ref{fig:adv}).
+The explanation is valid by definition, even though it looks a lot like an Adversarial Example \citep{goodfellow2014explaining}.
+\citet{schut2021generating} make the connection between Adversarial Examples and Counterfactual Explanations explicit and propose using a Jacobian-Based Saliency Map Attack to solve Equation~\ref{eq:general}.
+They demonstrate that this approach yields realistic and sparse counterfactuals for Bayesian, adversarially robust classifiers.
+Applying their approach to our simple MNIST classifier does not yield a realistic counterfactual but this one, too, is valid (right panel in Figure~\ref{fig:adv}).
+
+The crucial difference between Adversarial Examples (AE) and Counterfactual Explanations is one of intent.
+While an AE is intended to go unnoticed, a CE should have certain desirable properties.
+The literature has made this explicit by introducing various so-called \textit{desiderata}.
+To properly serve both AI practitioners and individuals affected by AI decision-making systems, counterfactuals should be sparse, proximate~\citep{wachter2017counterfactual}, actionable~\citep{ustun2019actionable}, diverse~\citep{mothilal2020explaining}, plausible~\citep{joshi2019realistic,poyiadzi2020face,schut2021generating}, robust~\citep{upadhyay2021robust,pawelczyk2022probabilistically,altmeyer2023endogenous} and causal~\citep{karimi2021algorithmic} among other things.
+Researchers have come up with various ways to meet these desiderata, which have been surveyed in~\citep{verma2020counterfactual} and~\citep{karimi2020survey}.
+
+To fulfil this latter goal, researchers have come up with a myriad of ways.
+@joshi2019realistic were among the first to suggest that instead of searching counterfactuals in the feature space, we can instead traverse a latent embedding learned by a surrogate generative model.
+Similarly, @poyiadzi2020face use density ...
+Finally, @karimi2021algorithmic argues that counterfactuals should comply with the causal model that generates them [CHECK IF WE CAN PHASE THIS LIKE THIS].
+Other related approaches include ...
+All of these different approaches have a common goal: they aim to ensure that the generated counterfactuals comply with the (learned) data-generating process (DGB).
 
@@
   \centering
-  \fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
-  \caption{Sample figure caption.}
+  \begin{minipage}[t]{0.45\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{../www/you_may_not_like_it.png}
+    \caption{You may not like it, but this is what stripped-down counterfactuals look like.
+Counterfactuals for turning an 8 (eight) into a 3 (three): original image (left); counterfactual produced using \citet{wachter2017counterfactual} (centre); and a counterfactual produced using JSMA-based approach introduced by \citep{schut2021generating}.}\label{fig:adv}
+  \end{minipage}\hfill
+  \begin{minipage}[t]{0.45\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{../www/surrogate_gone_wrong.png}
+    \caption{Using surrogates can improve plausibility, but also increases vulnerability.
+Counterfactuals for turning an 8 (eight) into a 3 (three): original image (left); counterfactual produced using REVISE \citep{joshi2019realistic} with a well-specified surrogate (centre); and a counterfactual produced using REVISE \citep{joshi2019realistic} with a poorly specified surrogate (right).}\label{fig:vae}
+  \end{minipage}
 \end{figure}
 
-To properly serve both AI practitioners and individuals affected by AI decision-making systems, Counterfactual Explanations should have certain desirable properties, sometimes referred to as \textit{desiderata}.
-Besides diversity, which we already introduced above, some of the most prominent desiderata include sparsity, proximity \citep{wachter2017counterfactual}, actionability~\citep{ustun2019actionable}, plausibility \citep{joshi2019realistic,poyiadzi2020face,schut2021generating}, robustness \citep{upadhyay2021robust,pawelczyk2022probabilistically,altmeyer2023endogenous} and causality~\citep{karimi2021algorithmic}.
-Researchers have come up with various ways to meet these desiderata, which have been surveyed in~\citep{verma2020counterfactual} and~\citep{karimi2020survey}.
+\medskip
+
+\bibliography{bib}
 

### 2023-03-22 paper/paper.tex
progress

@@
 \usepackage{amsmath}
+\usepackage{amsthm}
 \usepackage{graphicx}
@@
 \setcitestyle{numbers,square,comma}
+
+% Numbered Environments:
+\newtheorem{definition}{Definition}[section]
 
@@
 
-To fulfil this latter goal, researchers have come up with a myriad of ways.
-@joshi2019realistic were among the first to suggest that instead of searching counterfactuals in the feature space, we can instead traverse a latent embedding learned by a surrogate generative model.
-Similarly, @poyiadzi2020face use density ...
-Finally, @karimi2021algorithmic argues that counterfactuals should comply with the causal model that generates them [CHECK IF WE CAN PHASE THIS LIKE THIS].
-Other related approaches include ...
-All of these different approaches have a common goal: they aim to ensure that the generated counterfactuals comply with the (learned) data-generating process (DGB).
+Finding ways to generate \textit{plausible} counterfactuals has been one of the primary concerns.
+To this end, \citet{joshi2019realistic} were among the first to suggest that instead of searching counterfactuals in the feature space $\mathcal{X}$, we can instead traverse a latent embedding $\mathcal{Z}$ that implicitly codifies the data generating process (DGP) of $x\sim\mathcal{X}$.
+To learn the latent embedding, they introduce a surrogate model.
+In particular, they propose to use the latent embedding of a Variational Autoencoder (VAE) trained to generate samples $x^*\sim \mathcal{G}(z)$ where $\mathcal{G}$ denotes the decoder part of the VAE.
+Provided the surrogate model is well-trained, their proposed approach can yield compelling counterfactual explanations like the one in the centre panel of Figure~ref{fig:vae}.
+
+Others have proposed similar approaches.
+\citet{dombrowski2021diffeomorphic} traverse the base space of a normalizing flow to solve Equation~\ref{eq:general}, essentially relying on a different surrogate model for the generative task.
+\citet{poyiadzi2020face} use density estimators ($\hat{p}: \mathcal{X} \mapsto [0,1]$) to constrain the counterfactual paths.
+\citet{karimi2021algorithmic} argue that counterfactuals should comply with the causal model that generates the data.
+All of these different approaches share a common goal: ensuring that the generated counterfactuals comply with the true and unobserved DGP.
+To summarize this broad objective, we propose the following definition:
+
+\begin{definition}[Plausible Counterfactuals]
+  \label{def:plausible}
+  Let $\mathcal{X}|t$ denote the true conditional distribution of samples in the target class.
+Then for $x^{\prime}$ to be considered a plausible counterfactual, we need: $x^{\prime} \sim \mathcal{X}|t$.
+\end{definition}
+
+Note that Definition~\ref{def:plausible} subsumes the notion of plausible counterfactual paths, since we can simply apply it to each counterfactual state along the path.
 

### 2023-03-22 paper/paper.tex
intro and background done

@@
 To learn the latent embedding, they introduce a surrogate model.
-In particular, they propose to use the latent embedding of a Variational Autoencoder (VAE) trained to generate samples $x^*\sim \mathcal{G}(z)$ where $\mathcal{G}$ denotes the decoder part of the VAE.
-Provided the surrogate model is well-trained, their proposed approach can yield compelling counterfactual explanations like the one in the centre panel of Figure~ref{fig:vae}.
+In particular, they propose to use the latent embedding of a Variational Autoencoder (VAE) trained to generate samples $x^* \leftarrow \mathcal{G}(z)$ where $\mathcal{G}$ denotes the decoder part of the VAE.
+Provided the surrogate model is well-trained, their proposed approach ---REVISE--- can yield compelling counterfactual explanations like the one in the centre panel of Figure~\ref{fig:vae}.
 
@@
 
-Note that Definition~\ref{def:plausible} subsumes the notion of plausible counterfactual paths, since we can simply apply it to each counterfactual state along the path.
+Note that Definition~\ref{def:plausible} is consistent with the notion of plausible counterfactual paths, since we can simply apply it to each counterfactual state along the path.
+
+Surrogate models offer an obvious solution to achieve this objective.
+Unfortunately, surrogates also introduce a dependency: the generated explanations no longer depend exclusively on the Black Box Model itself, but also on the surrogate model.
+This is not necessarily problematic if the primary objective is not to explain the behaviour of the model but to offer recourse to individuals affected by it.
+It may become problematic even in this context if the dependency turns into a vulnerability.
+To illustrate this point, we have used REVISE \citep{joshi2019realistic} with an underfitted VAE to generate the counterfactual in the right panel of Figure~\ref{fig:vae}: in this case, the decoder step of the VAE fails to yield plausible values ($\{x^{\prime} \leftarrow \mathcal{G}(z)\} \not\sim \mathcal{X}|t$) and hence the counterfactual search in the learned latent space is doomed.
 
@@
 \end{figure}
+
+\section{A Framework for Conformal Counterfactual Explanations}\label{cce}
 

### 2023-03-23 paper/paper.tex
uh

@@
 
-Most state-of-the-art approaches to generating Counterfactual Explanations (CE) rely on gradient descent to optimize different flavours of the same counterfactual search objective,
+Most state-of-the-art approaches to generating Counterfactual Explanations rely on gradient descent to optimize different flavours of the same counterfactual search objective,
 
@@
   \label{def:plausible}
-  Let $\mathcal{X}|t$ denote the true conditional distribution of samples in the target class.
-Then for $x^{\prime}$ to be considered a plausible counterfactual, we need: $x^{\prime} \sim \mathcal{X}|t$.
+  Let $\mathcal{X}|y=t$ denote the true conditional distribution of samples in the target class $t$.
+Then for $x^{\prime}$ to be considered a plausible counterfactual, we need: $x^{\prime} \sim \mathcal{X}|y=t$.
 \end{definition}
@@
 It may become problematic even in this context if the dependency turns into a vulnerability.
-To illustrate this point, we have used REVISE \citep{joshi2019realistic} with an underfitted VAE to generate the counterfactual in the right panel of Figure~\ref{fig:vae}: in this case, the decoder step of the VAE fails to yield plausible values ($\{x^{\prime} \leftarrow \mathcal{G}(z)\} \not\sim \mathcal{X}|t$) and hence the counterfactual search in the learned latent space is doomed.
+To illustrate this point, we have used REVISE \citep{joshi2019realistic} with an underfitted VAE to generate the counterfactual in the right panel of Figure~\ref{fig:vae}: in this case, the decoder step of the VAE fails to yield plausible values ($\{x^{\prime} \leftarrow \mathcal{G}(z)\} \not\sim \mathcal{X}|y=t$) and hence the counterfactual search in the learned latent space is doomed.
 
@@
 \section{A Framework for Conformal Counterfactual Explanations}\label{cce}
+
+In Section~\ref{background} we explained that Counterfactual Explanations work directly with Black Box Model, so fidelity is not a concern.
+This may explain why research has primarily focused on other desiderata, most notably plausibility (Definition~\ref{def:plausible}).
+Enquiring about the plausibility of a counterfactual essentially boils down to the following question: `Is this counterfactual consistent with the underlying data'?
+To introduce this section, we posit a related, slightly more nuanced question: `Is this counterfactual consistent with what the model has learned about the underlying data'?
+We will argue that fidelity is not a sufficient evaluation measure to answer this question and propose a novel way to assess if explanations conform with model behaviour.
+Finally, we will introduce a framework for Conformal Counterfactual Explanations, that reconciles the notions of plausibility and model conformity.
+
+\subsection{From Fidelity to Model Conformity}
+
+The word \textit{fidelity} stems from the Latin word `fidelis', which means `faithful, loyal, trustworthy' \citep{mw2023fidelity}.
+As we explained in Section~\ref{background}, model explanations are considered faithful if their corresponding predictions coincide with the predictions made by the model itself.
+Since this definition of faithfulness is not useful in the context of Counterfactual Explanations, we propose an adapted version:
+
+\begin{definition}[Conformal Counterfactuals]
+  \label{def:conformal}
+  Let $\mathcal{X}_{\theta}|y=t = p_{\theta}(x|y=t)$ denote the conditional distribution of $x$ in the target class $t$, where $theta$ denotes the parameters of model $M$.
+Then for $x^{\prime}$ to be considered a conformal counterfactual, we need: $x^{\prime} \sim \mathcal{X}_{\theta}|y=t$.
+\end{definition}
+
+In words, conformal counterfactuals conform with what the predictive model has learned about the input data $x$.
+Since this definition works with distributional properties, it explicitly accounts for the multiplicity of explanations we discussed earlier.
+Except for the posterior conditional distribution $p_{\theta}(x|y=t)$, we already have access to all the ingredients in Definition~\ref{def:conformal}.
+
+How can we quantify $p_{\theta}(x|y=t)$?
+After all, the predictive model $M$ was trained to discriminate outputs conditional on inputs, which is a different conditional distribution: $p_{\theta}(y|x)$.
+Learning the distribution over inputs $p_{\theta}(x|y=t)$ is a generative task that $M$ was not explicitly trained for.
+In the context of Counterfactual Explanations, it is the task that existing approaches have reallocated from the model itself to a surrogate.
+
+Fortunately, recent work by \citet{grathwohl2020your} on Energy Based Models (EBM) has pointed out that there is a generative model hidden within every discriminative model.
+\citet{schut2021generating} were the first to notice and leverage this in the context of CE.
 

### 2023-03-23 paper/paper.tex
progress

@@
   \label{def:conformal}
-  Let $\mathcal{X}_{\theta}|y=t = p_{\theta}(x|y=t)$ denote the conditional distribution of $x$ in the target class $t$, where $theta$ denotes the parameters of model $M$.
-Then for $x^{\prime}$ to be considered a conformal counterfactual, we need: $x^{\prime} \sim \mathcal{X}_{\theta}|y=t$.
+  Let $\mathcal{X}_{\theta}|t = p_{\theta}(x|y=t)$ denote the conditional distribution of $x$ in the target class $t$, where $theta$ denotes the parameters of model $M$.
+Then for $x^{\prime}$ to be considered a conformal counterfactual, we need: $x^{\prime} \sim \mathcal{X}_{\theta}|t$.
 \end{definition}
@@
 \citet{schut2021generating} were the first to notice and leverage this in the context of CE.
+
+\subsection{Conformal Training meets Counterfactual Explanations}
+
+\section{Experiments}
 

### 2023-03-23 paper/paper.tex
progress

@@
 
-This is typically implemented by defining a target outcome $t \in \mathcal{Y}$ for some individual $x \in \mathcal{X}$, for which the model $M:\mathcal{X}\mapsto\mathcal{Y}$ initially predicts a different outcome: $M(x)\ne t$.
-Counterfactuals are then searched by minimizing a loss function that compares the predicted model output to the target outcome: $\text{yloss}(M(x),t)$.
+This is typically implemented by defining a target outcome $t \in \mathcal{Y}$ for some individual $x \in \mathcal{X}$, for which the model $M_{\theta}:\mathcal{X}\mapsto\mathcal{Y}$ initially predicts a different outcome: $M_{\theta}(x)\ne t$.
+Counterfactuals are then searched by minimizing a loss function that compares the predicted model output to the target outcome: $\text{yloss}(M_{\theta}(x),t)$.
 Since Counterfactual Explanations (CE) work directly with the Black Box Model, they always have full local fidelity by construction.
@@
 \begin{aligned}
-\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}} \left\{  {\text{yloss}(M(f(\mathbf{s}^\prime)),y^*)}+ \lambda {\text{cost}(f(\mathbf{s}^\prime)) }  \right\} 
+\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}} \left\{  {\text{yloss}(M_{\theta}(f(\mathbf{s}^\prime)),y^*)}+ \lambda {\text{cost}(f(\mathbf{s}^\prime)) }  \right\} 
 \end{aligned} 
@@
 This approach solves Equation~\ref{eq:general} through gradient-descent in the feature space with a penalty for the distance between the factual $x$ and the counterfactual $x^{\prime}$.
-The underlying classifier $M$ is a simple Multi-Layer Perceptron (MLP) with good test accuracy.
+The underlying classifier $M_{\theta}$ is a simple Multi-Layer Perceptron (MLP) with good test accuracy.
 For the generated counterfactual $x^{\prime}$ the model predicts the target label with high confidence (centre panel in Figure~\ref{fig:adv}).
@@
   \label{def:conformal}
-  Let $\mathcal{X}_{\theta}|t = p_{\theta}(x|y=t)$ denote the conditional distribution of $x$ in the target class $t$, where $theta$ denotes the parameters of model $M$.
+  Let $\mathcal{X}_{\theta}|t = p_{\theta}(x|y=t)$ denote the conditional distribution of $x$ in the target class $t$, where $theta$ denotes the parameters of model $M_{\theta}$.
 Then for $x^{\prime}$ to be considered a conformal counterfactual, we need: $x^{\prime} \sim \mathcal{X}_{\theta}|t$.
@@
 
-How can we quantify $p_{\theta}(x|y=t)$?
-After all, the predictive model $M$ was trained to discriminate outputs conditional on inputs, which is a different conditional distribution: $p_{\theta}(y|x)$.
-Learning the distribution over inputs $p_{\theta}(x|y=t)$ is a generative task that $M$ was not explicitly trained for.
+How can we quantify $p_{\theta}(\mathbf{x}|y=t)$?
+After all, the predictive model $M_{\theta}$ was trained to discriminate outputs conditional on inputs, which is a different conditional distribution: $p_{\theta}(y|x)$.
+Learning the distribution over inputs $p_{\theta}(\mathbf{x}|y=t)$ is a generative task that $M_{\theta}$ was not explicitly trained for.
 In the context of Counterfactual Explanations, it is the task that existing approaches have reallocated from the model itself to a surrogate.
 
-Fortunately, recent work by \citet{grathwohl2020your} on Energy Based Models (EBM) has pointed out that there is a generative model hidden within every discriminative model.
-\citet{schut2021generating} were the first to notice and leverage this in the context of CE.
+Fortunately, recent work by \citet{grathwohl2020your} on Energy Based Models (EBM) has pointed out that there is a `generative model hidden within every standard discriminative model'.
+The authors show that we can draw samples from the posterior conditional distribution $p_{\theta}(\mathbf{x}|y)$ using Stochastic Gradient Langevin Dynamics (SGLD).
+In doing so, it is possible to train classifiers jointly for the discriminative task using standard cross-entropy and the generative task using SGLD.
+They demonstrate empirically that among other things this improves predictive uncertainty quantification for discriminative models.
+
+To see how their proposed conditional sampling strategy can be applied in our context, note that if we fix $y$ to our target value $t$, we can sample from $p_{\theta}(\mathbf{x}|y=t)$ using SGLD as follows,
+
+\begin{equation}\label{eq:sgld}
+  \begin{aligned}
+    \mathbf{x}_{j+1} &\leftarrow \mathbf{x}_j - \frac{\epsilon^2}{2} \mathcal{E}(\mathbf{x}_j|y=t) + \epsilon \mathbf{r}_j, && j=1,...,J
+  \end{aligned}
+\end{equation}
+
+where $\mathbf{r}_j \sim \mathcal{N}(\mathbf{0},\mathbf{I})$ is the stochastic term and the step-size $\epsilon$ is typically polynomially decayed.
+The term $\mathcal{E}(\mathbf{x}_j|y=t)$ denotes the energy function.
+Following \citet{grathwohl2020your} we use $\mathcal{E}(\mathbf{x}_j|y=t)=-M_{\theta}(x)[t]$, that is the negative logit corresponding to the target class label $t$.
+
+While $\mathbf{x}_K$ is only guaranteed to distribute as $p_{\theta}(\mathbf{x}|y=t)$ if $\epsilon \rightarrow 0$ and $J \rightarrow \infty$, the bias introduced for a small finite $\epsilon$ is negligible in practice \citep{murphy2023probabilistic,grathwohl2020your}.
+While \citet{grathwohl2020your} use Equation~\ref{eq:sgld} during training, we are interested in applying the conditional sampling procedure in a post hoc fashion to any standard discriminative model.
+Generating multiple samples in this manner yields an empirical distribution $\hat{\mathcal{X}}_{\theta}|t$, which we can use to assess if a given counterfactual $x^{\prime}$ conforms with the model $M_{\theta}$ (Definition~\ref{def:conformal}).
+
+\textbf{TBD}
+
+\begin{itemize}
+  \item What exact sampler do we use? ImproperSGLD as in \citet{grathwohl2020your} seems to work best.
+  \item How exactly do we plan to quantify plausibility and conformity? Elaborate on measures.
+\end{itemize}
 
 \subsection{Conformal Training meets Counterfactual Explanations}
+
+Now that we have a way of evaluating Counterfactual Explanations in terms of their plausibility and conformity, we are interested in finding a way to generate counterfactuals that are as plausible and conformal as possible.
+We hypothesize that a narrow focus on plausibility may come at the cost of reduced conformity.
+Using a surrogate model for the generative task, for example, may improve plausibility but inadvertently yield counterfactuals that are more consistent with the surrogate than the Black Box Model itself.
+
+One way to ensure model conformity is to rely strictly on the model itself.~\citet{schut2021generating} demonstrate that this restriction need not impede plausibility, since we can rely on predictive uncertainty estimates to guide our counterfactual search.
+By avoiding counterfactual paths that are associated with high predictive uncertainty, we end up generating counterfactuals for which the model $M_{\theta}$ predicts the target label $t$ with high confidence.
+Provided the model is well-calibrated, these counterfactuals are plausible.
+
+Interestingly, \citet{schut2021generating} point to this connection between the generative task and predictive uncertainty quantification
 

### 2023-03-24 paper/paper.tex
uh

@@
 To properly serve both AI practitioners and individuals affected by AI decision-making systems, counterfactuals should be sparse, proximate~\citep{wachter2017counterfactual}, actionable~\citep{ustun2019actionable}, diverse~\citep{mothilal2020explaining}, plausible~\citep{joshi2019realistic,poyiadzi2020face,schut2021generating}, robust~\citep{upadhyay2021robust,pawelczyk2022probabilistically,altmeyer2023endogenous} and causal~\citep{karimi2021algorithmic} among other things.
-Researchers have come up with various ways to meet these desiderata, which have been surveyed in~\citep{verma2020counterfactual} and~\citep{karimi2020survey}.
 
-Finding ways to generate \textit{plausible} counterfactuals has been one of the primary concerns.
-To this end, \citet{joshi2019realistic} were among the first to suggest that instead of searching counterfactuals in the feature space $\mathcal{X}$, we can instead traverse a latent embedding $\mathcal{Z}$ that implicitly codifies the data generating process (DGP) of $x\sim\mathcal{X}$.
+Researchers have come up with various ways to meet these desiderata, which have been extensively surveyed and evaluated in various studies~\citep{verma2020counterfactual,karimi2020survey,pawelczyk2021carla,artelt2021evaluating,guidotti2022counterfactual}.
+Perhaps unsurprisingly, the different desiderata are often positively correlated.
+For example, \citet{artelt2021evaluating} find that plausibility typically also leads to improved robustness.
+Similarly, plausibility has also been connected to causality in the sense that plausible counterfactuals respect causal relationships \citep{mahajan2020preserving}.
+
+Arguably, the plausibility of counterfactuals has been among the primary concerns and some have focused explicitly on this goal.
+\citet{joshi2019realistic}, for example, were among the first to suggest that instead of searching counterfactuals in the feature space $\mathcal{X}$, we can instead traverse a latent embedding $\mathcal{Z}$ that implicitly codifies the data generating process (DGP) of $x\sim\mathcal{X}$.
 To learn the latent embedding, they introduce a surrogate model.
@@
 
+\subsection{Evaluation Measures}\label{evaluation}
+
+Above we have defined plausibility (\ref{def:plausible}) and conformity (\ref{def:conformal}) for Counterfactual Explanations.
+In this subsection, we introduce evaluation measures that facilitate a quantitative evaluation of counterfactuals for these objectives.
+
+Firstly, in order to assess the plausibility of counterfactuals we adapt the implausibility metric proposed in \citet{guidotti2022counterfactual}.
+The authors propose to evaluate plausibility in terms of the distance of the counterfactual $x^{\prime}$ from its nearest neighbour in the target class $t$: the smaller this distance, the more plausible the counterfactual.
+Instead of focusing only on the nearest neighbour of $x^{\prime}$, we suggest computing the average over distances from multiple (possibly all) observed instances in the target class.
+Formally, for a single counterfactual, we have:
+
+\begin{equation}\label{eq:impl}
+  \begin{aligned}
+    \text{impl} = \frac{1}{|x \in \mathcal{X}|t|} \sum_{x \in \mathcal{X}|t} \text{dist}(x^{\prime},x)
+  \end{aligned}
+\end{equation}
+
+This measure is straightforward to compute and should be less sensitive to outliers in the target class than the one based on the nearest neighbour.
+It also gives rise to a very similar evaluation measure for conformity.
+We merely swap out the subsample of individuals in the target class for the empirical distribution of generated conditional samples:
+
+\begin{equation}\label{eq:conf}
+  \begin{aligned}
+    \text{conf} = \frac{1}{|x \in \mathcal{X}_{\theta}|t|} \sum_{x \in \mathcal{X}_{\theta}|t} \text{dist}(x^{\prime},x)
+  \end{aligned}
+\end{equation}
+
+As noted by \citet{guidotti2022counterfactual}, these distance-based measures are simplistic and more complex alternative measures may ultimately be more appropriate for the task.
+For example, we considered using statistical divergence measures instead.
+This would involve generating not one but many counterfactuals and comparing the generated empirical distribution to the target distributions in Definitions~\ref{def:plausible} and~\ref{def:conformal}.
+While this approach is potentially more rigorous, generating enough counterfactuals is not always practical.
+
 \subsection{Conformal Training meets Counterfactual Explanations}
@@
 Provided the model is well-calibrated, these counterfactuals are plausible.
+The authors demonstrate this empirically by evaluating
 

### 2023-03-24 paper/paper.tex
uh

@@
 
-Now that we have a way of evaluating Counterfactual Explanations in terms of their plausibility and conformity, we are interested in finding a way to generate counterfactuals that are as plausible and conformal as possible.
+Now that we have a framework for evaluating Counterfactual Explanations in terms of their plausibility and conformity, we are interested in finding a way to generate counterfactuals that are as plausible and conformal as possible.
 We hypothesize that a narrow focus on plausibility may come at the cost of reduced conformity.
@@
 
-One way to ensure model conformity is to rely strictly on the model itself.~\citet{schut2021generating} demonstrate that this restriction need not impede plausibility, since we can rely on predictive uncertainty estimates to guide our counterfactual search.
+While we do not want seek to discourage the use of surrogate models, we suggest that one way to ensure model conformity is to rely strictly on the model itself.~\citet{schut2021generating} demonstrate that this restriction need not impede plausibility, since we can rely on predictive uncertainty estimates to guide our counterfactual search.
 By avoiding counterfactual paths that are associated with high predictive uncertainty, we end up generating counterfactuals for which the model $M_{\theta}$ predicts the target label $t$ with high confidence.
-Provided the model is well-calibrated, these counterfactuals are plausible.
-The authors demonstrate this empirically by evaluating
+Provided the model is well-calibrated, these counterfactuals are plausible which the authors demonstrate empirically through benchmarks.
 
-Interestingly, \citet{schut2021generating} point to this connection between the generative task and predictive uncertainty quantification
+This idea hinges on ...
 
 \section{Experiments}
+
+\section{Discussion}
+
+Consistent with the findings in \citet{schut2021generating}, we have demonstrated that predictive uncertainty estimates can be leveraged to generate plausible counterfactuals.
+Interestingly, \citet{schut2021generating} point out that this finding --- as intuitive as it is --- may be linked to a positive connection between the generative task and predictive uncertainty quantification.
+In particular, \citet{grathwohl2020your} demonstrate that their proposed method for integrating the generative objective in training yields models that have improved predictive uncertainty quantification.
+Since neither \citet{schut2021generating} nor we have employed any surrogate generative models, our findings seem to indicate that the positive connection found in \citet{grathwohl2020your} is bidirectional.
 

### 2023-03-25 paper/paper.tex
uh

@@
 Counterfactuals are then searched by minimizing a loss function that compares the predicted model output to the target outcome: $\text{yloss}(M_{\theta}(x),t)$.
-Since Counterfactual Explanations (CE) work directly with the Black Box Model, they always have full local fidelity by construction.
+Since Counterfactual Explanations (CE) work directly with the Black Box Model, valid counterfactuals always have full local fidelity by construction \citep{mothilal2020explaining}.
 Fidelity is defined as the degree to which explanations approximate the predictions of the Black Box Model.
-This arguably one of the most important evaluation metrics for model explanations, since any explanation that explains a prediction not actually made by the model is useless \citep{molnar2020interpretable}.
+This is arguably one of the most important evaluation metrics for model explanations, since any explanation that explains a prediction not actually made by the model is useless \citep{molnar2020interpretable}.
 
@@
 
-\section{A Framework for Conformal Counterfactual Explanations}\label{cce}
+\section{Evaluating the Faithfulness of Counterfactuals}\label{conformity}
 
@@
   \item What exact sampler do we use? ImproperSGLD as in \citet{grathwohl2020your} seems to work best.
-  \item How exactly do we plan to quantify plausibility and conformity? Elaborate on measures.
 \end{itemize}
@@
 
-\subsection{Conformal Training meets Counterfactual Explanations}
+\section{A Framework for Conformal Counterfactual Explanations}\label{cce}
 
@@
 Using a surrogate model for the generative task, for example, may improve plausibility but inadvertently yield counterfactuals that are more consistent with the surrogate than the Black Box Model itself.
+We suggest that one way to ensure model conformity is to rely strictly on the model itself.
+In this section, we introduce a novel framework that meets this requirement, works under minimal assumptions and does not impede the plausibility objective: Conformal Counterfactual Explanations.
 
-While we do not want seek to discourage the use of surrogate models, we suggest that one way to ensure model conformity is to rely strictly on the model itself.~\citet{schut2021generating} demonstrate that this restriction need not impede plausibility, since we can rely on predictive uncertainty estimates to guide our counterfactual search.
+\subsection{Predictive Uncertainty Quantification for Counterfactual Explanations}
+
+Our proposed methodolgy is built off of the findings presented in ~\citet{schut2021generating}.
+The authors demonstrate that it is not only possible but remarkably easy to generate plausible counterfactuals for Black Box Models that provide predictive uncertainty estimates.
 By avoiding counterfactual paths that are associated with high predictive uncertainty, we end up generating counterfactuals for which the model $M_{\theta}$ predicts the target label $t$ with high confidence.
@@
 
-This idea hinges on ...
+\textbf{TBD}: maximizing predicted probability is equivalent to minimizing uncertainty ...
+
+The approach proposed by \citet{schut2021generating} hinges on the crucial assumption that the Black Box Model provides predictive uncertainty estimates.
+The authors argue that in light of rapid advances in Bayesian Deep Learning (DL), this assumption is overall less costly than the engineering overhead induced by using surrogate models.
+This is even more true today, as recent work has put Laplace Approximation back on the map for truly effortless Bayesian DL \citep{immer2020improving,daxberger2021laplace,antoran2023sampling}.
+Nonetheless, the need for Bayesian methods may be too restrictive in some cases.
+
+Fortunately, there is a promising alternative approach to predictive uncertainty quantification (UQ) that we will turn to next: Conformal Prediction.
+
+\subsection{Conformal Prediction}
+
+Conformal Prediction (CP): a scalable and statistically rigorous  Not only is CP statistically rigorous and works under minimal distributional assumptions, it is also model-agnostic and can be applied at test time.
+That last part is critical since it allows us to relax the assumption that the Black Box Model needs to learn to generate predictive uncertainty estimates during training.
+In other words, CP promises to provide a way to generate plausible counterfactuals for any model without the need for surrogate models.
+
+\subsection{Conformal Counterfactual Explanations}
 

### 2023-03-27 paper/paper.tex
more progress

@@
 The explanation is valid by definition, even though it looks a lot like an Adversarial Example \citep{goodfellow2014explaining}.
-\citet{schut2021generating} make the connection between Adversarial Examples and Counterfactual Explanations explicit and propose using a Jacobian-Based Saliency Map Attack to solve Equation~\ref{eq:general}.
+\citet{schut2021generating} make the connection between Adversarial Examples and Counterfactual Explanations explicit and propose using a Jacobian-Based Saliency Map Attack (JSMA) to solve Equation~\ref{eq:general}.
 They demonstrate that this approach yields realistic and sparse counterfactuals for Bayesian, adversarially robust classifiers.
@@
 
-\subsection{Predictive Uncertainty Quantification for Counterfactual Explanations}
+\subsection{Plausible Counterfactuals through Minimal Uncertainty}
 
-Our proposed methodolgy is built off of the findings presented in ~\citet{schut2021generating}.
+Our proposed methodology is built on the findings presented in~\citet{schut2021generating}.
 The authors demonstrate that it is not only possible but remarkably easy to generate plausible counterfactuals for Black Box Models that provide predictive uncertainty estimates.
-By avoiding counterfactual paths that are associated with high predictive uncertainty, we end up generating counterfactuals for which the model $M_{\theta}$ predicts the target label $t$ with high confidence.
-Provided the model is well-calibrated, these counterfactuals are plausible which the authors demonstrate empirically through benchmarks.
+Their proposed algorithm solves Equation~\ref{eq:general} by greedily applying JSMA in the feature space with standard cross-entropy loss and no penalty at all.
+They show that this is equivalent to minimizing predictive uncertainty and hence yields counterfactuals for which the model $M_{\theta}$ predicts the target label $t$ with high confidence.
+Provided the model is well-calibrated, these counterfactuals are plausible which the authors demonstrate empirically through benchmarks \citep{schut2021generating}.
 
-\textbf{TBD}: maximizing predicted probability is equivalent to minimizing uncertainty ...
-
-The approach proposed by \citet{schut2021generating} hinges on the crucial assumption that the Black Box Model provides predictive uncertainty estimates.
+Unfortunately, this idea hinges on the crucial assumption that the Black Box Model provides predictive uncertainty estimates.
 The authors argue that in light of rapid advances in Bayesian Deep Learning (DL), this assumption is overall less costly than the engineering overhead induced by using surrogate models.
@@
 
-Fortunately, there is a promising alternative approach to predictive uncertainty quantification (UQ) that we will turn to next: Conformal Prediction.
+In looking for ways to lift that restriction, we found a promising alternative candidate for predictive uncertainty quantification (UQ) that we will briefly introduce next: Conformal Prediction.
 
@@
 
-Conformal Prediction (CP): a scalable and statistically rigorous  Not only is CP statistically rigorous and works under minimal distributional assumptions, it is also model-agnostic and can be applied at test time.
-That last part is critical since it allows us to relax the assumption that the Black Box Model needs to learn to generate predictive uncertainty estimates during training.
-In other words, CP promises to provide a way to generate plausible counterfactuals for any model without the need for surrogate models.
+Conformal Prediction (CP) is a scalable and statistically rigorous approach to predictive UQ that works under minimal distributional assumptions \citep{angelopoulos2021gentle}.
+It has recently gained popularity in the Machine Learning community \citep{angelopoulos2021gentle,manokhin2022awesome}.
+Crucially for our intended application, CP is model-agnostic and can be applied at test time.
+This allows us to relax the assumption that the Black Box Model needs to learn to generate predictive uncertainty estimates during training.
+In other words, CP promises to provide a way to generate plausible counterfactuals for any standard discriminative model without the need for surrogate models.
+
+...
 
 \subsection{Conformal Counterfactual Explanations}
+
+The fact that conformal classifiers produce set-valued predictions introduces a challenge: it is not immediately obvious how to use such classifiers in the context of gradient-based counterfactual search.
+Put differently, it is not clear how to use set-valued outputs $M_{\theta}(x)$ in Equation~\ref{eq:general}.
+Fortunately, \citet{stutz2022learning} have recently proposed a framework for Conformal Training that also hinges on differentiability.
+To evaluate the performance of conformal classifiers during training, they introduce a custom loss function as well as a smooth set size penalty.
+
+...
 

### 2023-03-27 paper/paper.tex
progress

@@
 
-...
+Intuitively, CP works under the premise of turning heuristic notions of uncertainty into rigorous uncertainty estimates by repeatedly sifting through the data.
+It can be used to generate prediction intervals for regression models and prediction sets for classification models \citep{altmeyer2022conformal}.
+Since the literature on CE and AR is typically concerned with classification problems, we focus on the latter.
+A particular variant of CP called Split Conformal Prediction (SCP) is well-suited for our purposes because it imposes only minimal restrictions on model training.
+
+Specifically, SCP involves splitting the data $\mathcal{D}_n=\{(X_i,Y_i)\}_{i=1,...,n}$ into a proper training set $\mathcal{D}_{\text{train}}$ and a calibration set $\mathcal{D}_{\text{cal}}$.
+The former is used to train the classifier in any conventional fashion: $\widehat{M}_{\theta}(X_i,Y_i)$, $i\in\mathcal{D}_{\text{train}}$.
+The latter is then used to compute so-called nonconformity scores: $\mathcal{S}=\{s(X_i,Y_i)\}_{i \in \mathcal{D}_{\text{cal}}}$ where $s: (\mathcal{X},\mathcal{Y}) \mapsto \mathbb{R}$ is referred to as \textit{score function}.
+In the context of classification, a common choice for the score function is just $s_i=1-\widehat{M}_{\theta}(X_i)[Y_i]$, that is one minus the logit corresponding to the observed label $Y_i$ \citep{angelopoulos2021gentle}.
+
+Finally, classification sets are formed as follows,
+
+\begin{equation}\label{eq:scp}
+  \begin{aligned}
+    C_{\theta}(X_i;\alpha)=\{y: s(X_i,y) \le \hat{q}\}
+  \end{aligned}
+\end{equation}
+
+where $\hat{q}$ denotes the $(1-\alpha)$-quantile of $\mathcal{S}$ and $\alpha$ is a predetermined error rate.
+As the size of the calibration set increases, the probability that the classification set $C(X_{\text{test}})$ for a newly arrived sample $X_i$ does not cover the true test label $Y_i$ approaches $\alpha$ \citep{angelopoulos2021gentle}.
+
+Observe from Equation~\ref{eq:scp} that Conformal Prediction works on an instance-level basis, much like Counterfactual Explanations are local.
+The prediction set for an individual instance $X_i$ depends only on the characteristics of that sample and the specified error rate.
+Intuitively, the set is more likely to include multiple labels for samples that are difficult to classify, so the set size is indicative of predictive uncertainty.
+To see why this effect is exacerbated by small choices for $\alpha$ consider the case of $\alpha=0$, which requires that the true label is covered by the prediction set with probability equal to one.
 
@@
 The fact that conformal classifiers produce set-valued predictions introduces a challenge: it is not immediately obvious how to use such classifiers in the context of gradient-based counterfactual search.
-Put differently, it is not clear how to use set-valued outputs $M_{\theta}(x)$ in Equation~\ref{eq:general}.
+Put differently, it is not clear how to use prediction sets in Equation~\ref{eq:general}.
 Fortunately, \citet{stutz2022learning} have recently proposed a framework for Conformal Training that also hinges on differentiability.
 To evaluate the performance of conformal classifiers during training, they introduce a custom loss function as well as a smooth set size penalty.
-
-...
+Their key idea lies in forming soft assignment scores for each class to be included in the prediction set: $C_{\theta,y}(X_i;\alpha):=\sigma\left((s(X_i,y)-\alpha) T^{-1}\right)$ for $y\in\{1,...,K\}$ where $\sigma$ is the sigmoid function and $T$ is a temperature hyper-parameter.
 

### 2023-03-27 paper/paper.tex
slowly slowly

@@
 Fortunately, \citet{stutz2022learning} have recently proposed a framework for Conformal Training that also hinges on differentiability.
-To evaluate the performance of conformal classifiers during training, they introduce a custom loss function as well as a smooth set size penalty.
-Their key idea lies in forming soft assignment scores for each class to be included in the prediction set: $C_{\theta,y}(X_i;\alpha):=\sigma\left((s(X_i,y)-\alpha) T^{-1}\right)$ for $y\in\{1,...,K\}$ where $\sigma$ is the sigmoid function and $T$ is a temperature hyper-parameter.
+Specifically, they show how Stochastic Gradient Descent can be used to train classifiers not only for the discriminative task but also for additional objectives related to Conformal Prediction.
+One such objective is \textit{efficiency}: for a given target error rate $alpha$, the efficiency of a conformal classifier improves as its average prediction set size decreases.
+To this end, the authors introduce a smooth set size penalty,
+
+\begin{equation}\label{eq:setsize}
+  \begin{aligned}
+    \Omega(C_{\theta}(x;\alpha))&=\max \left(0, \sum_{y\in\mathcal{Y}}C_{\theta,y}(X_i;\alpha) - \kappa \right)
+  \end{aligned}
+\end{equation}
+
+where $\kappa \in \{0,1\}$ is a hyper-parameter and $C_{\theta,y}(X_i;\alpha)$ can be interpreted as the probability of label $y$ being included in the prediction set.
+Formally, it is defined as $C_{\theta,y}(X_i;\alpha):=\sigma\left((s(X_i,y)-\alpha) T^{-1}\right)$ for $y\in\{1,...,K\}$ where $\sigma$ is the sigmoid function and $T$ is a hyper-parameter used for temperature scaling \citep{stutz2022learning}.
+
+We propose to include this penalty in the counterfactual search objective (Equation~\ref{eq:general})
 

### 2023-03-27 paper/paper.tex
first draft for everything excluding experiments and conclusion is done

@@
 
-We propose to include this penalty in the counterfactual search objective (Equation~\ref{eq:general})
+Penalizing the set size in this way is in principal enough to train efficient conformal classifiers \citep{stutz2022learning}.
+As we explained above, the set size is also closely linked to predictive uncertainty at the local level.
+This makes the smooth penalty defined in Equation~\ref{eq:setsize} useful in the context of meeting our objective of generating plausible counterfactuals.
+In particular, we adapt Equation~\ref{eq:general} to define the baseline objective for Conformal Counterfactual Explanations (CCE):
+
+\begin{equation}\label{eq:cce}
+  \begin{aligned}
+    \mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}} \left\{  {\text{yloss}(M_{\theta}(f(\mathbf{s}^\prime)),y^*)}+ \lambda \Omega(C_{\theta}(f(\mathbf{s}^\prime);\alpha)) \right\} 
+  \end{aligned}
+\end{equation}
+
+Since we can still retrieve unperturbed logits from our conformal classifier $M_{\theta}$, we are still free work with any loss function of our choice.
+For example, we could use standard cross-entropy for $\text{yloss}$.
+
+In order to generate prediction sets $C_{\theta}(f(\mathbf{s}^\prime);\alpha)$ for any Black Box Model we merely need to perform a single calibration pass through a holdout set $\mathcal{D}_{\text{cal}}$.
+Arguably, data is typically abundant and in most applications practitioners tend to hold out a test data set anyway.
+Our proposed approach for CCE therefore removes the restriction on the family of predictive models, at the small cost of reserving a subset of the available data for calibration.
 

### 2023-03-29 paper/paper.tex
first draft for everything excluding experiments and conclusion is done

@@
 
-This is typically implemented by defining a target outcome $t \in \mathcal{Y}$ for some individual $x \in \mathcal{X}$, for which the model $M_{\theta}:\mathcal{X}\mapsto\mathcal{Y}$ initially predicts a different outcome: $M_{\theta}(x)\ne t$.
-Counterfactuals are then searched by minimizing a loss function that compares the predicted model output to the target outcome: $\text{yloss}(M_{\theta}(x),t)$.
+This is typically implemented by defining a target outcome $\mathbf{y}^* \in \mathcal{Y}$ for some individual $\mathbf{x} \in \mathcal{X}=\mathbb{R}^D$ described by $D$ attributes, for which the model $M_{\theta}:\mathcal{X}\mapsto\mathcal{Y}$ initially predicts a different outcome: $M_{\theta}(\mathbf{x})\ne \mathbf{y}^*$.
+Counterfactuals are then searched by minimizing a loss function that compares the predicted model output to the target outcome: $\text{yloss}(M_{\theta}(\mathbf{x}),\mathbf{y}^*)$.
 Since Counterfactual Explanations (CE) work directly with the Black Box Model, valid counterfactuals always have full local fidelity by construction \citep{mothilal2020explaining}.
@@
 
-Most state-of-the-art approaches to generating Counterfactual Explanations rely on gradient descent to optimize different flavours of the same counterfactual search objective,
+In this section, we provide some background on Counterfactual Explanations and our motivation for this work.
+To start off, we briefly introduce the methodology uncerlying most state-of-the-art (SOTA) counterfactual generators.
+
+While Counterfactual Explanations can be generated for arbitrary regression models \citep{spooner2021counterfactual}, existing work has primarily focused on classification problems.
+Let $\mathcal{Y}=(0,1)^K$ denote the one-hot-encoded output domain with $K$ classes.
+Then most SOTA counterfactual generators rely on gradient descent to optimize different flavours of the following counterfactual search objective:
 
@@
 \begin{aligned}
-\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}} \left\{  {\text{yloss}(M_{\theta}(f(\mathbf{s}^\prime)),y^*)}+ \lambda {\text{cost}(f(\mathbf{s}^\prime)) }  \right\} 
+\mathbf{Z}^\prime &= \arg \min_{\mathbf{Z}^\prime \in \mathcal{Z}^M} \left\{  {\text{yloss}(M_{\theta}(f(\mathbf{Z}^\prime)),\mathbf{y}^*)}+ \lambda {\text{cost}(f(\mathbf{Z}^\prime)) }  \right\} 
 \end{aligned} 
@@
 
-where $\text{yloss}$ denotes the primary loss function already introduced above and $\text{cost}$ is either a single penalty or a collection of penalties that are used to impose constraints through regularization.
-Following the convention in \citet{altmeyer2023endogenous} we use $\mathbf{s}^\prime=\{ s_k\}_K$ to denote the vector $K$-dimensional array of counterfactual states.
-This is to explicitly account for the fact that we can generate multiple counterfactuals, as with DiCE \citep{mothilal2020explaining}, and may choose to traverse a latent representation $\mathcal{Z}$ of the feature space $\mathcal{X}$, as we will discuss further below.
+Here $\text{yloss}$ denotes the primary loss function already introduced above and $\text{cost}$ is either a single penalty or a collection of penalties that are used to impose constraints through regularization.
+Following the convention in \citet{altmeyer2023endogenous} we use $\mathbf{Z}^\prime=\{ \mathbf{z}_m\}_M$ to denote the vector $M$-dimensional array of counterfactual states.
+This is to explicitly account for the fact that we can generate multiple counterfactuals $M$, as with DiCE \citep{mothilal2020explaining}, and may choose to traverse a latent representation $\mathcal{Z}$ of the feature space $\mathcal{X}$, as we will discuss further below.
 
@@
 In Figure~\ref{fig:adv}, for example, we have the baseline approach proposed in \citet{wachter2017counterfactual} to MNIST data (centre panel).
-This approach solves Equation~\ref{eq:general} through gradient-descent in the feature space with a penalty for the distance between the factual $x$ and the counterfactual $x^{\prime}$.
+This approach solves Equation~\ref{eq:general} through gradient-descent in the feature space with a penalty for the distance between the factual $\mathbf{x}$ and the counterfactual $\mathbf{x}^{\prime}$.
 The underlying classifier $M_{\theta}$ is a simple Multi-Layer Perceptron (MLP) with good test accuracy.
-For the generated counterfactual $x^{\prime}$ the model predicts the target label with high confidence (centre panel in Figure~\ref{fig:adv}).
+For the generated counterfactual $\mathbf{x}^{\prime}$ the model predicts the target label with high confidence (centre panel in Figure~\ref{fig:adv}).
 The explanation is valid by definition, even though it looks a lot like an Adversarial Example \citep{goodfellow2014explaining}.
@@
 Arguably, the plausibility of counterfactuals has been among the primary concerns and some have focused explicitly on this goal.
-\citet{joshi2019realistic}, for example, were among the first to suggest that instead of searching counterfactuals in the feature space $\mathcal{X}$, we can instead traverse a latent embedding $\mathcal{Z}$ that implicitly codifies the data generating process (DGP) of $x\sim\mathcal{X}$.
+\citet{joshi2019realistic}, for example, were among the first to suggest that instead of searching counterfactuals in the feature space $\mathcal{X}$, we can instead traverse a latent embedding $\mathcal{Z}$ that implicitly codifies the data generating process (DGP) of $\mathbf{x}\sim\mathcal{X}$.
 To learn the latent embedding, they introduce a surrogate model.
-In particular, they propose to use the latent embedding of a Variational Autoencoder (VAE) trained to generate samples $x^* \leftarrow \mathcal{G}(z)$ where $\mathcal{G}$ denotes the decoder part of the VAE.
+In particular, they propose to use the latent embedding of a Variational Autoencoder (VAE) trained to generate samples $\mathbf{x}^* \leftarrow \mathcal{G}(\mathbf{z})$ where $\mathcal{G}$ denotes the decoder part of the VAE.
 Provided the surrogate model is well-trained, their proposed approach ---REVISE--- can yield compelling counterfactual explanations like the one in the centre panel of Figure~\ref{fig:vae}.
@@
   \label{def:plausible}
-  Let $\mathcal{X}|y=t$ denote the true conditional distribution of samples in the target class $t$.
-Then for $x^{\prime}$ to be considered a plausible counterfactual, we need: $x^{\prime} \sim \mathcal{X}|y=t$.
+  Let $\mathcal{X}|\mathbf{y}^*$ denote the true conditional distribution of samples in the target class $\mathbf{y}^*$.
+Then for $\mathbf{x}^{\prime}$ to be considered a plausible counterfactual, we need: $\mathbf{x}^{\prime} \sim \mathcal{X}|\mathbf{y}^*$.
 \end{definition}
@@
 It may become problematic even in this context if the dependency turns into a vulnerability.
-To illustrate this point, we have used REVISE \citep{joshi2019realistic} with an underfitted VAE to generate the counterfactual in the right panel of Figure~\ref{fig:vae}: in this case, the decoder step of the VAE fails to yield plausible values ($\{x^{\prime} \leftarrow \mathcal{G}(z)\} \not\sim \mathcal{X}|y=t$) and hence the counterfactual search in the learned latent space is doomed.
+To illustrate this point, we have used REVISE \citep{joshi2019realistic} with an underfitted VAE to generate the counterfactual in the right panel of Figure~\ref{fig:vae}: in this case, the decoder step of the VAE fails to yield plausible values ($\{\mathbf{x}^{\prime} \leftarrow \mathcal{G}(\mathbf{z})\} \not\sim \mathcal{X}|\mathbf{y}^*$) and hence the counterfactual search in the learned latent space is doomed.
 
@@
   \label{def:conformal}
-  Let $\mathcal{X}_{\theta}|t = p_{\theta}(x|y=t)$ denote the conditional distribution of $x$ in the target class $t$, where $theta$ denotes the parameters of model $M_{\theta}$.
-Then for $x^{\prime}$ to be considered a conformal counterfactual, we need: $x^{\prime} \sim \mathcal{X}_{\theta}|t$.
+  Let $\mathcal{X}_{\theta}|\mathbf{y}^* = p_{\theta}(x|\mathbf{y}^*)$ denote the conditional distribution of $\mathbf{x}$ in the target class $\mathbf{y}^*$, where $\theta$ denotes the parameters of model $M_{\theta}$.
+Then for $\mathbf{x}^{\prime}$ to be considered a conformal counterfactual, we need: $\mathbf{x}^{\prime} \sim \mathcal{X}_{\theta}|\mathbf{y}^*$.
 \end{definition}
 
-In words, conformal counterfactuals conform with what the predictive model has learned about the input data $x$.
+In words, conformal counterfactuals conform with what the predictive model has learned about the input data $\mathbf{x}$.
 Since this definition works with distributional properties, it explicitly accounts for the multiplicity of explanations we discussed earlier.
-Except for the posterior conditional distribution $p_{\theta}(x|y=t)$, we already have access to all the ingredients in Definition~\ref{def:conformal}.
+Except for the posterior conditional distribution $p_{\theta}(\mathbf{x}|\mathbf{y}^*)$, we already have access to all the ingredients in Definition~\ref{def:conformal}.
 
-How can we quantify $p_{\theta}(\mathbf{x}|y=t)$?
-After all, the predictive model $M_{\theta}$ was trained to discriminate outputs conditional on inputs, which is a different conditional distribution: $p_{\theta}(y|x)$.
-Learning the distribution over inputs $p_{\theta}(\mathbf{x}|y=t)$ is a generative task that $M_{\theta}$ was not explicitly trained for.
+How can we quantify $p_{\theta}(\mathbf{x}|\mathbf{y}^*)$?
+After all, the predictive model $M_{\theta}$ was trained to discriminate outputs conditional on inputs, which is a different conditional distribution: $p_{\theta}(\mathbf{y}|\mathbf{x})$.
+Learning the distribution over inputs $p_{\theta}(\mathbf{x}|\mathbf{y}^*)$ is a generative task that $M_{\theta}$ was not explicitly trained for.
 In the context of Counterfactual Explanations, it is the task that existing approaches have reallocated from the model itself to a surrogate.
@@
 Fortunately, recent work by \citet{grathwohl2020your} on Energy Based Models (EBM) has pointed out that there is a `generative model hidden within every standard discriminative model'.
-The authors show that we can draw samples from the posterior conditional distribution $p_{\theta}(\mathbf{x}|y)$ using Stochastic Gradient Langevin Dynamics (SGLD).
+The authors show that we can draw samples from the posterior conditional distribution $p_{\theta}(\mathbf{x}|\mathbf{y})$ using Stochastic Gradient Langevin Dynamics (SGLD).
 In doing so, it is possible to train classifiers jointly for the discriminative task using standard cross-entropy and the generative task using SGLD.
@@
 
-To see how their proposed conditional sampling strategy can be applied in our context, note that if we fix $y$ to our target value $t$, we can sample from $p_{\theta}(\mathbf{x}|y=t)$ using SGLD as follows,
+To see how their proposed conditional sampling strategy can be applied in our context, note that if we fix $\mathbf{y}$ to our target value $\mathbf{y}^*$, we can sample from $p_{\theta}(\mathbf{x}|\mathbf{y}^*)$ using SGLD as follows,
 
@@
   \begin{aligned}
-    \mathbf{x}_{j+1} &\leftarrow \mathbf{x}_j - \frac{\epsilon^2}{2} \mathcal{E}(\mathbf{x}_j|y=t) + \epsilon \mathbf{r}_j, && j=1,...,J
+    \mathbf{x}_{j+1} &\leftarrow \mathbf{x}_j - \frac{\epsilon^2}{2} \mathcal{E}(\mathbf{x}_j|\mathbf{y}^*) + \epsilon \mathbf{r}_j, && j=1,...,J
   \end{aligned}
@@
 where $\mathbf{r}_j \sim \mathcal{N}(\mathbf{0},\mathbf{I})$ is the stochastic term and the step-size $\epsilon$ is typically polynomially decayed.
-The term $\mathcal{E}(\mathbf{x}_j|y=t)$ denotes the energy function.
-Following \citet{grathwohl2020your} we use $\mathcal{E}(\mathbf{x}_j|y=t)=-M_{\theta}(x)[t]$, that is the negative logit corresponding to the target class label $t$.
+The term $\mathcal{E}(\mathbf{x}_j|\mathbf{y}^*)$ denotes the energy function where as in \citep{grathwohl2020your} we use $\mathcal{E}(\mathbf{x}_j|\mathbf{y}^*)=-M_{\theta}(\mathbf{x}_j)[\mathbf{y}^*]$, that is the negative logit corresponding to the target class label $\mathbf{y}^*$.
 
-While $\mathbf{x}_K$ is only guaranteed to distribute as $p_{\theta}(\mathbf{x}|y=t)$ if $\epsilon \rightarrow 0$ and $J \rightarrow \infty$, the bias introduced for a small finite $\epsilon$ is negligible in practice \citep{murphy2023probabilistic,grathwohl2020your}.
+While $\mathbf{x}_J$ is only guaranteed to distribute as $p_{\theta}(\mathbf{x}|\mathbf{y}^*)$ if $\epsilon \rightarrow 0$ and $J \rightarrow \infty$, the bias introduced for a small finite $\epsilon$ is negligible in practice \citep{murphy2023probabilistic,grathwohl2020your}.
 While \citet{grathwohl2020your} use Equation~\ref{eq:sgld} during training, we are interested in applying the conditional sampling procedure in a post hoc fashion to any standard discriminative model.
-Generating multiple samples in this manner yields an empirical distribution $\hat{\mathcal{X}}_{\theta}|t$, which we can use to assess if a given counterfactual $x^{\prime}$ conforms with the model $M_{\theta}$ (Definition~\ref{def:conformal}).
+Generating multiple samples in this manner yields an empirical distribution $\hat{\mathcal{X}}_{\theta}|\mathbf{y}^*$, which we can use to assess if a given counterfactual $\mathbf{x}^{\prime}$ conforms with the model $M_{\theta}$ (Definition~\ref{def:conformal}).
 
@@
 Firstly, in order to assess the plausibility of counterfactuals we adapt the implausibility metric proposed in \citet{guidotti2022counterfactual}.
-The authors propose to evaluate plausibility in terms of the distance of the counterfactual $x^{\prime}$ from its nearest neighbour in the target class $t$: the smaller this distance, the more plausible the counterfactual.
-Instead of focusing only on the nearest neighbour of $x^{\prime}$, we suggest computing the average over distances from multiple (possibly all) observed instances in the target class.
+The authors propose to evaluate plausibility in terms of the distance of the counterfactual $\mathbf{x}^{\prime}$ from its nearest neighbour in the target class $\mathbf{y}^*$: the smaller this distance, the more plausible the counterfactual.
+Instead of focusing only on the nearest neighbour of $\mathbf{x}^{\prime}$, we suggest computing the average over distances from multiple (possibly all) observed instances in the target class.
 Formally, for a single counterfactual, we have:
@@
   \begin{aligned}
-    \text{impl} = \frac{1}{|x \in \mathcal{X}|t|} \sum_{x \in \mathcal{X}|t} \text{dist}(x^{\prime},x)
+    \text{impl} = \frac{1}{\lvert\mathbf{x} \in \mathcal{X}|\mathbf{y}^*\rvert} \sum_{\mathbf{x} \in \mathcal{X}|\mathbf{y}^*} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
   \end{aligned}
@@
   \begin{aligned}
-    \text{conf} = \frac{1}{|x \in \mathcal{X}_{\theta}|t|} \sum_{x \in \mathcal{X}_{\theta}|t} \text{dist}(x^{\prime},x)
+    \text{conf} = \frac{1}{\lvert\mathbf{x} \in \mathcal{X}_{\theta}|\mathbf{y}^*\rvert} \sum_{\mathbf{x} \in \mathcal{X}_{\theta}|\mathbf{y}^*} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
   \end{aligned}
@@
 Their proposed algorithm solves Equation~\ref{eq:general} by greedily applying JSMA in the feature space with standard cross-entropy loss and no penalty at all.
-They show that this is equivalent to minimizing predictive uncertainty and hence yields counterfactuals for which the model $M_{\theta}$ predicts the target label $t$ with high confidence.
+They show that this is equivalent to minimizing predictive uncertainty and hence yields counterfactuals for which the model $M_{\theta}$ predicts the target label $\mathbf{y}^*$ with high confidence.
 Provided the model is well-calibrated, these counterfactuals are plausible which the authors demonstrate empirically through benchmarks \citep{schut2021generating}.
@@
 
-Specifically, SCP involves splitting the data $\mathcal{D}_n=\{(X_i,Y_i)\}_{i=1,...,n}$ into a proper training set $\mathcal{D}_{\text{train}}$ and a calibration set $\mathcal{D}_{\text{cal}}$.
-The former is used to train the classifier in any conventional fashion: $\widehat{M}_{\theta}(X_i,Y_i)$, $i\in\mathcal{D}_{\text{train}}$.
-The latter is then used to compute so-called nonconformity scores: $\mathcal{S}=\{s(X_i,Y_i)\}_{i \in \mathcal{D}_{\text{cal}}}$ where $s: (\mathcal{X},\mathcal{Y}) \mapsto \mathbb{R}$ is referred to as \textit{score function}.
-In the context of classification, a common choice for the score function is just $s_i=1-\widehat{M}_{\theta}(X_i)[Y_i]$, that is one minus the logit corresponding to the observed label $Y_i$ \citep{angelopoulos2021gentle}.
+Specifically, SCP involves splitting the data $\mathcal{D}_n=\{(\mathbf{x}_i,\mathbf{y}_i)\}_{i=1,...,n}$ into a proper training set $\mathcal{D}_{\text{train}}$ and a calibration set $\mathcal{D}_{\text{cal}}$.
+The former is used to train the classifier in any conventional fashion.
+The latter is then used to compute so-called nonconformity scores: $\mathcal{S}=\{s(\mathbf{x}_i,\mathbf{y}_i)\}_{i \in \mathcal{D}_{\text{cal}}}$ where $s: (\mathcal{X},\mathcal{Y}) \mapsto \mathbb{R}$ is referred to as \textit{score function}.
+In the context of classification, a common choice for the score function is just $s_i=1-M_{\theta}(\mathbf{x}_i)[\mathbf{y}_i]$, that is one minus the softmax output corresponding to the observed label $\mathbf{y}_i$ \citep{angelopoulos2021gentle}.
 
@@
   \begin{aligned}
-    C_{\theta}(X_i;\alpha)=\{y: s(X_i,y) \le \hat{q}\}
+    C_{\theta}(\mathbf{x}_i;\alpha)=\{\mathbf{y}: s(\mathbf{x}_i,\mathbf{y}) \le \hat{q}\}
   \end{aligned}
@@
 where $\hat{q}$ denotes the $(1-\alpha)$-quantile of $\mathcal{S}$ and $\alpha$ is a predetermined error rate.
-As the size of the calibration set increases, the probability that the classification set $C(X_{\text{test}})$ for a newly arrived sample $X_i$ does not cover the true test label $Y_i$ approaches $\alpha$ \citep{angelopoulos2021gentle}.
+As the size of the calibration set increases, the probability that the classification set $C(\mathbf{x}_{\text{test}})$ for a newly arrived sample $\mathbf{x}_{\text{test}}$ does not cover the true test label $\mathbf{y}_{\text{test}}$ approaches $\alpha$ \citep{angelopoulos2021gentle}.
 
 Observe from Equation~\ref{eq:scp} that Conformal Prediction works on an instance-level basis, much like Counterfactual Explanations are local.
-The prediction set for an individual instance $X_i$ depends only on the characteristics of that sample and the specified error rate.
+The prediction set for an individual instance $\mathbf{x}_i$ depends only on the characteristics of that sample and the specified error rate.
 Intuitively, the set is more likely to include multiple labels for samples that are difficult to classify, so the set size is indicative of predictive uncertainty.
@@
   \begin{aligned}
-    \Omega(C_{\theta}(x;\alpha))&=\max \left(0, \sum_{y\in\mathcal{Y}}C_{\theta,y}(X_i;\alpha) - \kappa \right)
+    \Omega(C_{\theta}(\mathbf{x};\alpha))&=\max \left(0, \sum_{\mathbf{y}\in\mathcal{Y}}C_{\theta,\mathbf{y}}(\mathbf{x}_i;\alpha) - \kappa \right)
   \end{aligned}
@@
 
-where $\kappa \in \{0,1\}$ is a hyper-parameter and $C_{\theta,y}(X_i;\alpha)$ can be interpreted as the probability of label $y$ being included in the prediction set.
-Formally, it is defined as $C_{\theta,y}(X_i;\alpha):=\sigma\left((s(X_i,y)-\alpha) T^{-1}\right)$ for $y\in\{1,...,K\}$ where $\sigma$ is the sigmoid function and $T$ is a hyper-parameter used for temperature scaling \citep{stutz2022learning}.
+where $\kappa \in \{0,1\}$ is a hyper-parameter and $C_{\theta,\mathbf{y}}(\mathbf{x}_i;\alpha)$ can be interpreted as the probability of label $\mathbf{y}$ being included in the prediction set.
+Formally, it is defined as $C_{\theta,\mathbf{y}}(\mathbf{x}_i;\alpha):=\sigma\left((s(\mathbf{x}_i,\mathbf{y})-\alpha) T^{-1}\right)$ for $\mathbf{y}\in\mathcal{Y}$ where $\sigma$ is the sigmoid function and $T$ is a hyper-parameter used for temperature scaling \citep{stutz2022learning}.
 
@@
   \begin{aligned}
-    \mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}} \left\{  {\text{yloss}(M_{\theta}(f(\mathbf{s}^\prime)),y^*)}+ \lambda \Omega(C_{\theta}(f(\mathbf{s}^\prime);\alpha)) \right\} 
+    \mathbf{Z}^\prime &= \arg \min_{\mathbf{Z}^\prime \in \mathcal{Z}^M} \left\{  {\text{yloss}(M_{\theta}(f(\mathbf{Z}^\prime)),\mathbf{y}^*)}+ \lambda \Omega(C_{\theta}(f(\mathbf{Z}^\prime);\alpha)) \right\}
   \end{aligned}
@@
 
-Since we can still retrieve unperturbed logits from our conformal classifier $M_{\theta}$, we are still free work with any loss function of our choice.
+Since we can still retrieve unperturbed softmax outputs from our conformal classifier $M_{\theta}$, we are free to work with any loss function of our choice.
 For example, we could use standard cross-entropy for $\text{yloss}$.
 
-In order to generate prediction sets $C_{\theta}(f(\mathbf{s}^\prime);\alpha)$ for any Black Box Model we merely need to perform a single calibration pass through a holdout set $\mathcal{D}_{\text{cal}}$.
+In order to generate prediction sets $C_{\theta}(f(\mathbf{Z}^\prime);\alpha)$ for any Black Box Model we merely need to perform a single calibration pass through a holdout set $\mathcal{D}_{\text{cal}}$.
 Arguably, data is typically abundant and in most applications practitioners tend to hold out a test data set anyway.

### 2023-03-29 paper/paper.tex
moved to 2023 template

@@
 %     \PassOptionsToPackage{numbers, compress}{natbib}
-% before loading neurips_2022
+% before loading neurips_2023
 
 % ready for submission
-% \usepackage{neurips_2022}
+% \usepackage{neurips_2023}
 
@@
 % [preprint] option:
-\usepackage[preprint]{neurips_2022}
+\usepackage[preprint]{neurips_2023}
 
 % to compile a camera-ready version, add the [final] option, e.g.:
-%     \usepackage[final]{neurips_2022}
+%     \usepackage[final]{neurips_2023}
 
 % to avoid loading the natbib package, add option nonatbib:
-%    \usepackage[nonatbib]{neurips_2022}
+%    \usepackage[nonatbib]{neurips_2023}
 
@@
 \usepackage{xcolor}         % colors
+
 \usepackage{amsmath}
@@
 
-\section{Submission of papers to NeurIPS 2022}
+\section{Submission of papers to NeurIPS 2023}
 
@@
 
-Papers to be submitted to NeurIPS 2022 must be prepared according to the
+Papers to be submitted to NeurIPS 2023 must be prepared according to the
 instructions presented here. Papers may only be up to {\bf nine} pages long,
@@
 
-The margins in 2022 are the same as those in 2007, which allow for $\sim$$15\%$
-more words in the paper compared to earlier years.
+The margins in 2023 are the same as those in previous years.
 
@@
 The style files for NeurIPS and other conference information are available on
-the World Wide Web at
+the website at
 \begin{center}
@@
 \end{center}
-The file \verb+neurips_2022.pdf+ contains these instructions and illustrates the
+The file \verb+neurips_2023.pdf+ contains these instructions and illustrates the
 various formatting requirements your NeurIPS paper must satisfy.
 
-The only supported style file for NeurIPS 2022 is \verb+neurips_2022.sty+,
+The only supported style file for NeurIPS 2023 is \verb+neurips_2023.sty+,
 rewritten for \LaTeXe{}.  \textbf{Previous style files for \LaTeX{} 2.09,
@@
 nonanonymized version of your work with the text ``Preprint. Work in progress.''
-in the footer. This version may be distributed as you see fit. Please \textbf{do
+in the footer. This version may be distributed as you see fit, as long as you do not say which conference it was submitted to. Please \textbf{do
   not} use the \verb+final+ option, which should \textbf{only} be used for
@@
 
-The file \verb+neurips_2022.tex+ may be used as a ``shell'' for writing your
+The file \verb+neurips_2023.tex+ may be used as a ``shell'' for writing your
 paper. All you have to do is replace the author, title, abstract, and text of
@@
 If you wish to load the \verb+natbib+ package with options, you may add the
-following before loading the \verb+neurips_2022+ package:
+following before loading the \verb+neurips_2023+ package:
 \begin{verbatim}
@@
 \begin{verbatim}
-   \usepackage[nonatbib]{neurips_2022}
+   \usepackage[nonatbib]{neurips_2023}
 \end{verbatim}
@@
 (e.g., a journal paper under review), use anonymous author names in the
-citation, e.g., an author of the form ``A.\ Anonymous.''
+citation, e.g., an author of the form ``A.\ Anonymous'' and include a copy of the anonymized paper in the supplementary material.
 
@@
 
-\section{Final instructions}
+\subsection{Math}
+Note that display math in bare TeX commands will not create correct line numbers for submission.
+Please use LaTeX (or AMSTeX) commands for unnumbered display math.
+(You really shouldn't be using \$\$ anyway; see \url{https://tex.stackexchange.com/questions/503/why-is-preferable-to} and \url{https://tex.stackexchange.com/questions/40492/what-are-the-differences-between-align-equation-and-displaymath} for more information.)
+
+\subsection{Final instructions}
 
@@
   available out-of-the-box on most Linux machines.
-
-\item The IEEE has recommendations for generating PDF files whose fonts are also
-  acceptable for NeurIPS. Please see
-  \url{http://www.emfield.org/icuwb2010/downloads/IEEE-PDF-SpecV32.pdf}
 
@@
 funding (financial activities supporting the submitted work) and competing interests (related financial activities outside the submitted work).
-More information about this disclosure can be found at: \url{https://neurips.cc/Conferences/2022/PaperInformation/FundingDisclosure}.
+More information about this disclosure can be found at: \url{https://neurips.cc/Conferences/2023/PaperInformation/FundingDisclosure}.
 
@@
 
+\section{Supplementary Material}
+
+Authors may wish to optionally include extra information (complete proofs, additional experiments and plots) in the appendix.
+All such materials should be part of the supplemental material (submitted separately) and should NOT be included in the main submission.
+
 \section*{References}
 
-References follow the acknowledgments. Use unnumbered first-level heading for
+References follow the acknowledgments in the camera-ready paper. Use unnumbered first-level heading for
 the references. Any choice of citation style is acceptable as long as you are
@@
 Note that the Reference section does not count towards the page limit.
+\medskip
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section*{Checklist}
+{
+\small
 
-%%% BEGIN INSTRUCTIONS %%%
-The checklist follows the references.  Please
-read the checklist guidelines carefully for information on how to answer these
-questions.  For each question, change the default \answerTODO{} to \answerYes{},
-\answerNo{}, or \answerNA{}.  You are strongly encouraged to include a {\bf
-justification to your answer}, either by referencing the appropriate section of
-your paper or providing a brief inline description.  For example:
-\begin{itemize}
-  \item Did you include the license to the code and datasets? \answerYes{See Section~\ref{gen_inst}.}
-  \item Did you include the license to the code and datasets? \answerNo{The code and the data are proprietary.}
-  \item Did you include the license to the code and datasets? \answerNA{}
-\end{itemize}
-Please do not modify the questions and only use the provided macros for your
-answers.  Note that the Checklist section does not count towards the page
-limit.  In your paper, please delete this instructions block and only keep the
-Checklist section heading above along with the questions/answers below.
-%%% END INSTRUCTIONS %%%
+[1] Alexander, J.A.\ \& Mozer, M.C.\ (1995) Template-based algorithms for
+connectionist rule extraction. In G.\ Tesauro, D.S.\ Touretzky and T.K.\ Leen
+(eds.), {\it Advances in Neural Information Processing Systems 7},
+pp.\ 609--616. Cambridge, MA: MIT Press.
 
-\begin{enumerate}
+[2] Bower, J.M.\ \& Beeman, D.\ (1995) {\it The Book of GENESIS: Exploring
+  Realistic Neural Models with the GEneral NEural SImulation System.}  New York:
+TELOS/Springer--Verlag.
 
-\item For all authors...
-\begin{enumerate}
-  \item Do the main claims made in the abstract and introduction accurately reflect the paper's contributions and scope?
-    \answerTODO{}
-  \item Did you describe the limitations of your work?
-    \answerTODO{}
-  \item Did you discuss any potential negative societal impacts of your work?
-    \answerTODO{}
-  \item Have you read the ethics review guidelines and ensured that your paper conforms to them?
-    \answerTODO{}
-\end{enumerate}
-
-\item If you are including theoretical results...
-\begin{enumerate}
-  \item Did you state the full set of assumptions of all theoretical results?
-    \answerTODO{}
-        \item Did you include complete proofs of all theoretical results?
-    \answerTODO{}
-\end{enumerate}
-
-\item If you ran experiments...
-\begin{enumerate}
-  \item Did you include the code, data, and instructions needed to reproduce the main experimental results (either in the supplemental material or as a URL)?
-    \answerTODO{}
-  \item Did you specify all the training details (e.g., data splits, hyperparameters, how they were chosen)?
-    \answerTODO{}
-        \item Did you report error bars (e.g., with respect to the random seed after running experiments multiple times)?
-    \answerTODO{}
-        \item Did you include the total amount of compute and the type of resources used (e.g., type of GPUs, internal cluster, or cloud provider)?
-    \answerTODO{}
-\end{enumerate}
-
-\item If you are using existing assets (e.g., code, data, models) or curating/releasing new assets...
-\begin{enumerate}
-  \item If your work uses existing assets, did you cite the creators?
-    \answerTODO{}
-  \item Did you mention the license of the assets?
-    \answerTODO{}
-  \item Did you include any new assets either in the supplemental material or as a URL?
-    \answerTODO{}
-  \item Did you discuss whether and how consent was obtained from people whose data you're using/curating?
-    \answerTODO{}
-  \item Did you discuss whether the data you are using/curating contains personally identifiable information or offensive content?
-    \answerTODO{}
-\end{enumerate}
-
-\item If you used crowdsourcing or conducted research with human subjects...
-\begin{enumerate}
-  \item Did you include the full text of instructions given to participants and screenshots, if applicable?
-    \answerTODO{}
-  \item Did you describe any potential participant risks, with links to Institutional Review Board (IRB) approvals, if applicable?
-    \answerTODO{}
-  \item Did you include the estimated hourly wage paid to participants and the total amount spent on participant compensation?
-    \answerTODO{}
-\end{enumerate}
-
-\end{enumerate}
+[3] Hasselmo, M.E., Schnell, E.\ \& Barkai, E.\ (1995) Dynamics of learning and
+recall at excitatory recurrent synapses and cholinergic modulation in rat
+hippocampal region CA3. {\it Journal of Neuroscience} {\bf 15}(7):5249-5262.
+}
 
@@
 
-\appendix
-
-\section{Appendix}
-
-Optionally include extra information (complete proofs, additional experiments and plots) in the appendix.
-This section will often be part of the supplemental material.
-
 \end{document}

### 2023-03-30 notebooks/intro.qmd
progress

@@
 #| output: false
-
-measure = 
 using CounterfactualExplanations.Evaluation: benchmark
@@
 y =  counterfactual_data.output_encoder.labels
-conf_model = conformal_model(clf; method=:simple_inductive)
+conf_model = conformal_model(clf; method=:simple_inductive, coverage=0.99)
 mach = machine(conf_model, X, y)
@@
 
-temp = 0.5
-p1 = contourf(mach.model, mach.fitresult, X, y; plot_set_loss=true, zoom=0, temp=temp)
-p2 = contourf(mach.model, mach.fitresult, X, y; plot_classification_loss=true, target=1, zoom=0, temp=temp, clim=nothing, loss_matrix=ones(4,4))
+temp = 0.05
+p1 = contourf(mach.model, mach.fitresult, X, y; plot_set_loss=true, zoom=-1, temp=temp)
+p2 = contourf(mach.model, mach.fitresult, X, y; plot_classification_loss=true, zoom=-1, temp=temp, clim=nothing, loss_matrix=ones(4,4))
 plot(p1, p2, size=(800,320))
@@
 ```{julia}
+#| output: true
+#| echo: false
 #| fig-cap: "Energy-based conditional samples."
@@
 ```{julia}
+#| output: true
+#| echo: false
+#| fig-cap: "Comparison of counterfactuals produced using different generators."
+#| label: fig-ce-multi
+
 counterfactuals = Dict([name => generate_counterfactual(x, target, counterfactual_data, M, gen; initialization=:identity, converge_when=:generator_conditions, gradient_tol=1e-3) for (name, gen) in generators])

### 2023-03-31 notebooks/intro.qmd
ufff

@@
 y =  counterfactual_data.output_encoder.labels
-conf_model = conformal_model(clf; method=:simple_inductive)
+conf_model = conformal_model(clf; method=:trainable_simple_inductive)
 mach = machine(conf_model, X, y)
@@
 p1 = contourf(mach.model, mach.fitresult, X, y; plot_set_loss=true, zoom=0, temp=temp)
-p2 = contourf(mach.model, mach.fitresult, X, y; plot_classification_loss=true, target=1, zoom=0, temp=temp, clim=nothing, loss_matrix=ones(2,2))
+p2 = contourf(mach.model, mach.fitresult, X, y; plot_classification_loss=true, zoom=0, temp=temp, clim=nothing, loss_matrix=ones(2,2))
 plot(p1, p2, size=(800,320))
@@
 plts = []
-for target ∈ counterfactual_data.y_levels
+for (i,target) ∈ enumerate(counterfactual_data.y_levels)
     sampler = CCE.EnergySampler(M, counterfactual_data, target; niter=niter, nsamples=100)
@@
     plt = plot(M, counterfactual_data; target=target, zoom=-3,cbar=false)
-    scatter!(Xgen[1,:],Xgen[2,:],alpha=0.5,color=target,shape=:star,label="X|y=$target")
+    scatter!(Xgen[1,:],Xgen[2,:],alpha=0.5,color=i,shape=:star,label="X|y=$target")
     push!(plts, plt)
@@
 
-temp = 0.05
-p1 = contourf(mach.model, mach.fitresult, X, y; plot_set_loss=true, zoom=-1, temp=temp)
+temp = 0.5
+p1 = contourf(mach.model, mach.fitresult, X, y; plot_set_loss=true, zoom=-1, temp=temp, clim=nothing)
 p2 = contourf(mach.model, mach.fitresult, X, y; plot_classification_loss=true, zoom=-1, temp=temp, clim=nothing, loss_matrix=ones(4,4))

### 2023-03-31 src/CCE.jl

@@ -10,4 +10,7 @@ include("generator.jl")
 include("sampling.jl")
 # include("ConformalGenerator.jl")
 
+using MLJFlux
+MLJFlux.reformat(X, ::Type{<:AbstractMatrix}) = permutedims(X)
+
 end

### 2023-03-31 notebooks/intro.qmd
no clue why quarto is bugging

@@
 y =  counterfactual_data.output_encoder.labels
-conf_model = conformal_model(clf; method=:trainable_simple_inductive)
+conf_model = conformal_model(clf; method=:simple_inductive)
 mach = machine(conf_model, X, y)

### 2023-04-01 notebooks/intro.qmd
work on benchmarks

@@
 
-### Benchmark
-
-```{julia}
-#| output: false
-
-# Data:
-datasets = Dict(
-    :linearly_separable => load_linearly_separable(),
-    :overlapping => load_overlapping(),
-    :moons => load_moons(),
-    :circles => load_circles(),
-)
-
-# Models:
-models = Dict(
-    :MLP => M,
-)
-```
-
-Then we can simply loop over the datasets and eventually concatenate the results like so:
-
-```{julia}
-#| output: false
-using CounterfactualExplanations.Evaluation: benchmark
-bmks = []
-for (dataname, dataset) in datasets
-    bmk = benchmark(dataset; models=models, generators=generators, suppress_training=true, dataname=dataname)
-    push!(bmks, bmk)
-end
-bmk = reduce(vcat, bmks)
-```
-
 ## Multi-Class
@@
 ```
+
+## Benchmarks
+
+```{julia}
+#| output: false
+
+# Data:
+datasets = Dict(
+    :linearly_separable => load_linearly_separable(),
+    :overlapping => load_overlapping(),
+    :moons => load_moons(),
+    :circles => load_circles(),
+)
+
+# Models:
+models = Dict(
+    :MLP => M,
+)
+```
+
+Then we can simply loop over the datasets and eventually concatenate the results like so:
+
+```{julia}
+#| output: false
+using CounterfactualExplanations.Evaluation: benchmark
+bmks = []
+measures = [
+    CounterfactualExplanations.distance,
+    CCE.distance_from_energy,
+    CCE.distance_from_targets
+]
+for (dataname, dataset) in datasets
+    bmk = benchmark(
+        dataset; 
+        models=models, generators=generators, 
+        measure=measures,
+        suppress_training=true, dataname=dataname,
+        n_individuals=10
+    )
+    push!(bmks, bmk)
+end
+bmk = reduce(vcat, bmks)
+```
+
+```{julia}
+#| output: true
+#| echo: false
+#| label: fig-benchmark
+#| fig-cap: "Benchmark results for the different generators."
+
+df = bmk()
+plt = data(df) * visual(BoxPlot) * 
+    mapping(:generator, :value, row=:variable, col=:dataname, color=:generator)
+draw(plt, axis=(xlabel="", xticksvisible=false, xticklabelsvisible=false, width=200, height=180))
+```

### 2023-04-03 notebooks/intro.qmd
uh

--- 2023-03-31 notebooks/intro.qmd
@@
 temp = 0.5
-p1 = contourf(mach.model, mach.fitresult, X, y; plot_set_loss=true, zoom=0, temp=temp)
-p2 = contourf(mach.model, mach.fitresult, X, y; plot_classification_loss=true, zoom=0, temp=temp, clim=nothing, loss_matrix=ones(2,2))
-plot(p1, p2, size=(800,320))
+p1 = Plots.contourf(mach.model, mach.fitresult, X, y; plot_set_loss=true, zoom=0, temp=temp)
+p2 = Plots.contourf(mach.model, mach.fitresult, X, y; plot_classification_loss=true, zoom=0, temp=temp, clim=nothing, loss_matrix=ones(2,2))
+Plots.plot(p1, p2, size=(800,320))
 ```
@@
     Xgen = rand(sampler, nsamples)
-    plt = plot(M, counterfactual_data; target=target, zoom=-3,cbar=false)
-    scatter!(Xgen[1,:],Xgen[2,:],alpha=0.5,color=i,shape=:star,label="X|y=$target")
+    plt = Plots.plot(M, counterfactual_data; target=target, zoom=-3,cbar=false)
+    Plots.scatter!(Xgen[1,:],Xgen[2,:],alpha=0.5,color=i,shape=:star,label="X|y=$target")
     push!(plts, plt)
 end
-plot(plts..., layout=(1,length(plts)), size=(img_height*length(plts),img_height))
+Plots.plot(plts..., layout=(1,length(plts)), size=(img_height*length(plts),img_height))
 ```
@@
 generators = Dict(
-    ordered_names[1] => generator,
+    ordered_names[1] => generator |> gen -> @objective(gen, _ + 1.0distance),
     ordered_names[2] => deepcopy(generator) |> gen -> @objective(gen, _ + 0.1distance + 10.0set_size_penalty),
@@
     ce = counterfactuals[name]
-    plt = plot(ce; title=name, colorbar=false, ticks = false, legend=false, zoom=-0.5)
+    plt = Plots.plot(ce; title=name, colorbar=false, ticks = false, legend=false, zoom=-0.5)
     if :energy_sampler ∈ collect(keys(ce.params))
         Xgen = ce.params[:energy_sampler].buffer
-        scatter!(Xgen[1,:],Xgen[2,:],alpha=0.5,color=target,shape=:star,label="X|y=$target")
+        Plots.scatter!(Xgen[1,:],Xgen[2,:],alpha=0.5,color=target,shape=:star,label="X|y=$target")
     end
@@
 _n = length(generators)
-plot(plts..., size=(_n * img_height,1.05*img_height), layout=(1,_n))
-```
-
-### Benchmark
-
-```{julia}
-#| output: false
-
-# Data:
-datasets = Dict(
-    :linearly_separable => load_linearly_separable(),
-    :overlapping => load_overlapping(),
-    :moons => load_moons(),
-    :circles => load_circles(),
-)
-
-# Models:
-models = Dict(
-    :MLP => M,
-)
-```
-
-Then we can simply loop over the datasets and eventually concatenate the results like so:
-
-```{julia}
-#| output: false
-using CounterfactualExplanations.Evaluation: benchmark
-bmks = []
-for (dataname, dataset) in datasets
-    bmk = benchmark(dataset; models=models, generators=generators, suppress_training=true, dataname=dataname)
-    push!(bmks, bmk)
-end
-bmk = reduce(vcat, bmks)
+Plots.plot(plts..., size=(_n * img_height,1.05*img_height), layout=(1,_n))
 ```
@@
 temp = 0.5
-p1 = contourf(mach.model, mach.fitresult, X, y; plot_set_loss=true, zoom=-1, temp=temp, clim=nothing)
-p2 = contourf(mach.model, mach.fitresult, X, y; plot_classification_loss=true, zoom=-1, temp=temp, clim=nothing, loss_matrix=ones(4,4))
-plot(p1, p2, size=(800,320))
+p1 = Plots.contourf(mach.model, mach.fitresult, X, y; plot_set_loss=true, zoom=-1, temp=temp, clim=nothing)
+p2 = Plots.contourf(mach.model, mach.fitresult, X, y; plot_classification_loss=true, zoom=-1, temp=temp, clim=nothing, loss_matrix=ones(4,4))
+Plots.plot(p1, p2, size=(800,320))
 ```
@@
     Xgen = rand(sampler, nsamples)
-    plt = plot(M, counterfactual_data; target=target, zoom=-0.5,cbar=false)
-    scatter!(Xgen[1,:],Xgen[2,:],alpha=0.5,color=target,shape=:star,label="X|y=$target")
+    plt = Plots.plot(M, counterfactual_data; target=target, zoom=-0.5,cbar=false)
+    Plots.scatter!(Xgen[1,:],Xgen[2,:],alpha=0.5,color=target,shape=:star,label="X|y=$target")
     push!(plts, plt)
 end
-plot(plts..., layout=(1,length(plts)), size=(img_height*length(plts),img_height))
+Plots.plot(plts..., layout=(1,length(plts)), size=(img_height*length(plts),img_height))
 ```
@@
 _n = length(generators)
-plot(plts..., size=(_n * img_height,1.05*img_height), layout=(1,_n))
+Plots.plot(plts..., size=(_n * img_height,1.05*img_height), layout=(1,_n))
 ```
+
+## Benchmarks
+
+```{julia}
+#| output: false
+
+# Data:
+datasets = Dict(
+    :linearly_separable => load_linearly_separable(),
+    :overlapping => load_overlapping(),
+    :moons => load_moons(),
+    :circles => load_circles(),
+    :multi_class => load_multi_class(),
+)
+
+# Untrained Models:
+models = Dict(
+    :cov75 => CCE.ConformalModel(conformal_model(clf; method=:simple_inductive, coverage=0.75)),
+    :cov80 => CCE.ConformalModel(conformal_model(clf; method=:simple_inductive, coverage=0.80)),
+    :cov90 => CCE.ConformalModel(conformal_model(clf; method=:simple_inductive, coverage=0.90)),
+    :cov99 => CCE.ConformalModel(conformal_model(clf; method=:simple_inductive, coverage=0.99)),
+)
+```
+
+Then we can simply loop over the datasets and eventually concatenate the results like so:
+
+```{julia}
+#| output: false
+using CounterfactualExplanations.Evaluation: benchmark
+bmks = []
+measures = [
+    CounterfactualExplanations.distance,
+    CCE.distance_from_energy,
+    CCE.distance_from_targets
+]
+for (dataname, dataset) in datasets
+    bmk = benchmark(
+        dataset; 
+        models=deepcopy(models), 
+        generators=generators, 
+        measure=measures,
+        suppress_training=false, dataname=dataname,
+        n_individuals=10
+    )
+    push!(bmks, bmk)
+end
+bmk = reduce(vcat, bmks)
+```
+
+```{julia}
+@chain bmk() begin
+    @group_by(model, dataname)
+    @select(model, dataname, ce)
+    @mutate(performace = CounterfactualExplanations.model_evaluation(ce.M, ce.data))
+    @summarize(model=unique(model), dataname=unique(dataname), performace=unique(performance))
+    @ungroup
+end
+```
+
+```{julia}
+#| output: true
+#| echo: false
+#| label: fig-benchmark
+#| fig-cap: "Benchmark results for the different generators."
+#| fig-subcap: 
+#|  - "Circles."
+#|  - "Linearly Separable."
+#|  - "Moons."
+#|  - "Multi-class."
+#|  - "Overlapping."
+
+df = bmk()
+for dataname ∈ sort(unique(df.dataname))
+    df_ = df[df.dataname .== dataname, :]
+    plt = data(df_) * visual(BoxPlot) * 
+        mapping(:generator, :value, row=:variable, col=:model, color=:generator)
+    plt = draw(
+        plt, axis=(xlabel="", xticksvisible=false, xticklabelsvisible=false, width=200, height=180), 
+        # facet=(; linkyaxes=:none)
+    )   
+    display(plt)
+end
+```

### 2023-04-03 notebooks/setup.jl

@@ -8,6 +8,7 @@ setup_notebooks = quote
     using CairoMakie
     using CCE
     using CCE: set_size_penalty, distance_from_energy, distance_from_targets
+    using Chain: @chain
     using ConformalPrediction
     using CounterfactualExplanations
     using CounterfactualExplanations.Data
@@ -26,9 +27,10 @@ setup_notebooks = quote
     using Plots
     using Random
     using Serialization
+    using Tidier
 
     # Setup:
-    theme(:wong)
+    Plots.theme(:wong)
     Random.seed!(2023)
     www_path = "www"
     output_path = "artifacts"

### 2023-04-04 notebooks/intro.qmd
pff

@@
 ```{julia}
-builder = MLJFlux.@builder Chain(
+builder = MLJFlux.@builder Flux.Chain(
     Dense(n_in, 32, relu),
@@
 #| label: fig-energy
+#| output: true
 
@@
     ordered_names[1] => generator |> gen -> @objective(gen, _ + 1.0distance),
-    ordered_names[2] => deepcopy(generator) |> gen -> @objective(gen, _ + 0.1distance + 10.0set_size_penalty),
+    ordered_names[2] => deepcopy(generator) |> gen -> @objective(gen, _ + 0.1distance + 1.0set_size_penalty),
     ordered_names[3] => deepcopy(generator) |> gen -> @objective(gen, _ + 0.1distance + 10.0distance_from_energy),
@@
 y =  counterfactual_data.output_encoder.labels
-conf_model = conformal_model(clf; method=:simple_inductive, coverage=0.99)
+```
+
+```{julia}
+#| output: true
+#| echo: false
+#| label: fig-pen-multi
+#| fig-cap: "Illustration of the smooth size loss."
+
+Plots.gr_cbar_width[] = 0.01
+cvgs = [0.5, 0.75, 0.90, 0.95, 0.99]
+temps = [0.05, 0.1, 0.5, 1.0, 2.0]
+plts = []
+
+for cov in cvgs
+    # Train:
+    conf_model = conformal_model(clf; method=:simple_inductive, coverage=cov)
     mach = machine(conf_model, X, y)
     fit!(mach)
+
+    for temp in temps
+        # Plot:
+        plt = Plots.contourf(
+            mach.model, mach.fitresult, X, y; 
+            plot_set_loss=true, zoom=-1, temp=temp, clim=nothing, size=(2img_height,img_height),
+            title="cov=$cov, temp=$temp", cbar=true, ticks=false, legend=false,
+            κ = 0.0,
+        )
+        push!(plts, plt)
+    end
+end
+
+Plots.plot(plts..., layout=(length(cvgs),length(cvgs)), size=(2img_height*length(cvgs), img_height*length(cvgs)))
 ```
@@
 #| label: fig-losses-multi
-#| fig-cap: "Illustration of the smooth size loss and the configurable classification loss."
+#| fig-cap: "Illustration of the configurable classification loss."
 
-temp = 0.5
-p1 = Plots.contourf(mach.model, mach.fitresult, X, y; plot_set_loss=true, zoom=-1, temp=temp, clim=nothing)
-p2 = Plots.contourf(mach.model, mach.fitresult, X, y; plot_classification_loss=true, zoom=-1, temp=temp, clim=nothing, loss_matrix=ones(4,4))
-Plots.plot(p1, p2, size=(800,320))
+Plots.gr_cbar_width[] = 0.01
+cvgs = [0.5, 0.75, 0.90, 0.95, 0.99]
+temps = [0.05, 0.1, 0.5, 1.0, 2.0]
+plts = []
+
+for cov in cvgs
+    # Train:
+    conf_model = conformal_model(clf; method=:simple_inductive, coverage=cov)
+    mach = machine(conf_model, X, y)
+    fit!(mach)
+
+    for temp in temps
+        # Plot:
+        plt = Plots.contourf(
+            mach.model, mach.fitresult, X, y; 
+            plot_classification_loss=true, zoom=0, temp=temp, clim=nothing, loss_matrix=ones(4,4)
+        )
+        push!(plts, plt)
+    end
+end
+
+Plots.plot(plts..., layout=(length(cvgs),length(cvgs)), size=(2img_height*length(cvgs), img_height*length(cvgs)))
 ```
@@
 ```{julia}
+f(ce) = CounterfactualExplanations.model_evaluation(ce.M, ce.data)
 @chain bmk() begin
-    @group_by(model, dataname)
-    @select(model, dataname, ce)
-    @mutate(performace = CounterfactualExplanations.model_evaluation(ce.M, ce.data))
-    @summarize(model=unique(model), dataname=unique(dataname), performace=unique(performance))
+    @group_by(model, generator, dataname, variable)
+    @select(model, generator, dataname, ce, value)
+    @mutate(performance = f(ce))
+    @summarize(model=unique(model), generator=unique(generator), dataname=unique(dataname), performace=unique(performance), value=mean(value))
     @ungroup
+    @filter(dataname == :multi_class)
+    @filter(model == :cov99)
+    @filter(variable == "distance")
 end

### 2023-04-04 notebooks/mnist.qmd
pff

--- 2023-03-22 notebooks/proposal.qmd
@@
----
-title: High-Fidelity Counterfactual Explanations through Conformal Prediction
-subtitle: Research Proposal
-abstract: |
-    We propose Conformal Counterfactual Explanations: an effortless and rigorous way to produce realistic and faithful Counterfactual Explanations using Conformal Prediction.
-To address the need for realistic counterfactuals, existing work has primarily relied on separate generative models to learn the data-generating process.
-While this is an effective way to produce plausible and model-agnostic counterfactual explanations, it not only introduces a significant engineering overhead but also reallocates the task of creating realistic model explanations from the model itself to the generative model.
-Recent work has shown that there is no need for any of this when working with probabilistic models that explicitly quantify their own uncertainty.
-Unfortunately, most models used in practice still do not fulfil that basic requirement, in which case we would like to have a way to quantify predictive uncertainty in a post-hoc fashion.
----
-
 ```{julia}
@@
 
-## Motivation
-
-Counterfactual Explanations are a powerful, flexible and intuitive way to not only explain black-box models but also enable affected individuals to challenge them through the means of Algorithmic Recourse.
-
-### Counterfactual Explanations or Adversarial Examples?
-
-Most state-of-the-art approaches to generating Counterfactual Explanations (CE) rely on gradient descent in the feature space.
-The key idea is to perturb inputs $x\in\mathcal{X}$ into a black-box model $f: \mathcal{X} \mapsto \mathcal{Y}$ in order to change the model output $f(x)$ to some pre-specified target value $t\in\mathcal{Y}$.
-Formally, this boils down to defining some loss function $\ell(f(x),t)$ and taking gradient steps in the minimizing direction.
-The so-generated counterfactuals are considered valid as soon as the predicted label matches the target label.
-A stripped-down counterfactual explanation is therefore little different from an adversarial example.
-In @fig-adv, for example, generic counterfactual search as in @wachter2017counterfactual has been applied to MNIST data.
-
 ```{julia}
@@
 X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
+labels = counterfactual_data.output_encoder.labels
 input_dim, n_obs = size(counterfactual_data.X)
@@
 ```{julia}
+builder = MLJFlux.@builder M.model
+clf = NeuralNetworkClassifier(builder=builder, epochs=100)
+conf_model = conformal_model(clf; method=:simple_inductive)
+mach = machine(conf_model, X, labels)
+fit!(mach)
+```
+
+```{julia}
 # Search:
@@
 )
-generator = GreedyGenerator(η=5.0)
-ce_jsma = generate_counterfactual(
+generator = deepcopy(generator) |> gen -> @objective(gen, _ + 0.001distance + 1.0set_size_penalty)
+ce_conformal = generate_counterfactual(
     x, target, counterfactual_data, M, generator; 
@@
 ```
-
-![You may not like it, but this is what stripped-down counterfactuals look like.
-Here we have used @wachter2017counterfactual to generate multiple counterfactuals for turning an 8 (eight) into a 3 (three).](www/you_may_not_like_it.png){#fig-adv}
-
-The crucial difference between adversarial examples and counterfactuals is one of intent.
-While adversarial examples are typically intended to go unnoticed, counterfactuals in the context of Explainable AI are generally sought to be "plausible", "realistic" or "feasible".
-To fulfil this latter goal, researchers have come up with a myriad of ways.
-@joshi2019realistic were among the first to suggest that instead of searching counterfactuals in the feature space, we can instead traverse a latent embedding learned by a surrogate generative model.
-Similarly, @poyiadzi2020face use density ...
-Finally, @karimi2021algorithmic argues that counterfactuals should comply with the causal model that generates them [CHECK IF WE CAN PHASE THIS LIKE THIS].
-Other related approaches include ...
-All of these different approaches have a common goal: they aim to ensure that the generated counterfactuals comply with the (learned) data-generating process (DGB).
-
-::: {#def-plausible}
-
-## Plausible Counterfactuals
-
-Formally, if $x \sim \mathcal{X}$ and for the corresponding counterfactual we have $x^{\prime}\sim\mathcal{X}^{\prime}$, then for $x^{\prime}$ to be considered a plausible counterfactual, we need: $\mathcal{X} \approxeq \mathcal{X}^{\prime}$.
-
-:::
-
-In the context of Algorithmic Recourse, it makes sense to strive for plausible counterfactuals, since anything else would essentially require individuals to move to out-of-distribution states.
-But it is worth noting that our ambition to meet this goal, may have implications on our ability to faithfully explain the behaviour of the underlying black-box model (arguably our principal goal).
-By essentially decoupling the task of learning plausible representations of the data from the model itself, we open ourselves up to vulnerabilities.
-Using a separate generative model to learn $\mathcal{X}$, for example, has very serious implications for the generated counterfactuals.
-@fig-latent compares the results of applying REVISE [@joshi2019realistic] to MNIST data using two different Variational Auto-Encoders: while the counterfactual generated using an expressive (strong) VAE is compelling, the result relying on a less expressive (weak) VAE is not even valid.
-In this latter case, the decoder step of the VAE fails to yield values in $\mathcal{X}$ and hence the counterfactual search in the learned latent space is doomed.
-
-```{julia}
-using CounterfactualExplanations.Models: load_mnist_vae
-vae = load_mnist_vae()
-vae_weak = load_mnist_vae(;strong=false)
-Serialization.serialize(joinpath(output_path,"mnist_classifier.jls"), M)
-Serialization.serialize(joinpath(output_path,"mnist_vae.jls"), vae)
-Serialization.serialize(joinpath(output_path,"mnist_vae_weak.jls"), vae_weak)
-```
-
-```{julia}
-# Define generator:
-generator = REVISEGenerator(
-  opt = Descent(0.1),
-  λ=0.01
-)
-# Generate recourse:
-counterfactual_data.generative_model = vae # assign generative model
-ce = generate_counterfactual(
-    x, target, counterfactual_data, M, generator; 
-    decision_threshold=γ, max_iter=T,
-    initialization=:identity,
-)
-counterfactual_data = deepcopy(counterfactual_data)
-counterfactual_data.generative_model = vae_weak
-ce_weak = generate_counterfactual(
-    x, target, counterfactual_data, M, generator;
-    decision_threshold=γ, max_iter=T,
-    initialization=:identity,
-)
-```
-
-```{julia}
-ces = zip([ce,ce_weak])
-counterfactuals = reduce((x,y)->cat(x,y,dims=3),map(ce -> CounterfactualExplanations.counterfactual(ce[1]), ces))
-phat = reduce((x,y) -> cat(x,y,dims=3), map(ce -> target_probs(ce[1]), ces))
-plts = [p1]
-for x in zip(eachslice(counterfactuals; dims=3), eachslice(phat; dims=3))
-    ce, _phat = (x[1],x[2])
-    _title = "p(y=$(target)|x′)=$(round(_phat[1]; digits=3))"
-    plt = plot(
-        convert2image(MNIST, reshape(ce,28,28)),
-        axis=nothing, 
-        size=(img_height, img_height),
-        title=_title
-    )
-    plts = [plts..., plt]
-end
-plt = plot(plts...; size=(img_height*length(plts),img_height), layout=(1,length(plts)))
-savefig(plt, joinpath(www_path, "surrogate_gone_wrong.png"))
-```
-
-![Counterfactual explanations for MNIST using a Latent Space generator: turning a nine (9) into a four (4).](www/mnist_9to4_latent.png){#fig-latent}
-
-> Here it would be nice to have another example where we poison the data going into the generative model to hide biases present in the data (e.g. Boston housing).
-
-- Latent can be manipulated: 
-    - train biased model
-    - train VAE with biased variable removed/attacked (use Boston housing dataset)
-    - hypothesis: will generate bias-free explanations
-
-### From Plausible to High-Fidelity Counterfactuals {#sec-fidelity}
-
-In light of the findings, we propose to generally avoid using surrogate models to learn $\mathcal{X}$ in the context of Counterfactual Explanations.
-
-::: {#prp-surrogate}
-
-## Avoid Surrogates
-
-Since we are in the business of explaining a black-box model, the task of learning realistic representations of the data should not be reallocated from the model itself to some surrogate model.
-
-:::
-
-In cases where the use of surrogate models cannot be avoided, we propose to weigh the plausibility of counterfactuals against their fidelity to the black-box model.
-In the context of Explainable AI, fidelity is defined as describing how an explanation approximates the prediction of the black-box model [@molnar2020interpretable].
-Fidelity has become the default metric for evaluating Local Model-Agnostic Models, since they often involve local surrogate models whose predictions need not always match those of the black-box model.
-
-In the case of Counterfactual Explanations, the concept of fidelity has so far been ignored.
-This is not altogether surprising, since by construction and design, Counterfactual Explanations work with the predictions of the black-box model directly: as stated above, a counterfactual $x^{\prime}$ is considered valid if and only if $f(x^{\prime})=t$, where $t$ denote some target outcome.
-
-Does fidelity even make sense in the context of CE, and if so, how can we define it?
-In light of the examples in the previous section, we think it is urgent to introduce a notion of fidelity in this context, that relates to the distributional properties of the generated counterfactuals.
-In particular, we propose that a high-fidelity counterfactual $x^{\prime}$ complies with the class-conditional distribution $\mathcal{X}_{\theta} = p_{\theta}(X|y)$ where $\theta$ denote the black-box model parameters.
-
-::: {#def-fidele}
-
-## High-Fidelity Counterfactuals
-
-Let $\mathcal{X}_{\theta}|y = p_{\theta}(X|y)$ denote the class-conditional distribution of $X$ defined by $\theta$.
-Then for $x^{\prime}$ to be considered a high-fidelity counterfactual, we need: $\mathcal{X}_{\theta}|t \approxeq \mathcal{X}^{\prime}$ where $t$ denotes the target outcome.
-
-:::
-
-In order to assess the fidelity of counterfactuals, we propose the following two-step procedure:
-
-1) Generate samples $X_{\theta}|y$ and $X^{\prime}$ from $\mathcal{X}_{\theta}|t$ and $\mathcal{X}^{\prime}$, respectively.
-2) Compute the Maximum Mean Discrepancy (MMD) between $X_{\theta}|y$ and $X^{\prime}$.
-
-If the computed value is different from zero, we can reject the null-hypothesis of fidelity.
-
-> Two challenges here: 1) implementing the sampling procedure in @grathwohl2020your; 2) it is unclear if MMD is really the right way to measure this.
-
-## Conformal Counterfactual Explanations
-
-In @sec-fidelity, we have advocated for avoiding surrogate models in the context of Counterfactual Explanations.
-In this section, we introduce an alternative way to generate high-fidelity Counterfactual Explanations.
-In particular, we propose Conformal Counterfactual Explanations (CCE), that is Counterfactual Explanations that minimize the predictive uncertainty of conformal models.
-
-### Minimizing Predictive Uncertainty
-
-@schut2021generating demonstrated that the goal of generating realistic (plausible) counterfactuals can also be achieved by seeking counterfactuals that minimize the predictive uncertainty of the underlying black-box model.
-Similarly, @antoran2020getting ...
-
-- Problem: restricted to Bayesian models.
-- Solution: post-hoc predictive uncertainty quantification. In particular, Conformal Prediction.
-
-### Background on Conformal Prediction
-
-- Distribution-free, model-agnostic and scalable approach to predictive uncertainty quantification.
-- Conformal prediction is instance-based. So is CE. 
-- Take any fitted model and turn it into a conformal model using calibration data.
-- Our approach, therefore, relaxes the restriction on the family of black-box models, at the cost of relying on a subset of the data.
-Arguably, data is often abundant and in most applications practitioners tend to hold out a test data set anyway.
-
-> Does the coverage guarantee carry over to counterfactuals?
-
-### Generating Conformal Counterfactuals
-
-While Conformal Prediction has recently grown in popularity, it does introduce a challenge in the context of classification: the predictions of Conformal Classifiers are set-valued and therefore difficult to work with, since they are, for example, non-differentiable.
-Fortunately, @stutz2022learning introduced carefully designed differentiable loss functions that make it possible to evaluate the performance of conformal predictions in training.
-We can leverage these recent advances in the context of gradient-based counterfactual search ...
-
-> Challenge: still need to implement these loss functions.
-
-## Experiments
-
-### Research Questions
-
-- Is CP alone enough to ensure realistic counterfactuals?
-- Do counterfactuals improve further as the models get better?
-- Do counterfactuals get more realistic as coverage
-- What happens as we vary coverage and setsize?
-- What happens as we improve the model robustness?
-- What happens as we improve the model's ability to incorporate predictive uncertainty (deep ensemble, laplace)?
-- What happens if we combine with DiCE, ClaPROAR, Gravitational?
-- What about CE robustness to endogenous shifts [@altmeyer2023endogenous]?
-
-- Benchmarking:
-    - add PROBE [@pawelczyk2022probabilistically] into the mix.
-    - compare travel costs to domain shits.
-
-> Nice to have: What about using Laplace Approximation, then Conformal Prediction? What about using Conformalised Laplace?
-
-## References

### 2023-04-04 notebooks/setup.jl
pff

@@ -34,6 +34,6 @@ setup_notebooks = quote
     Random.seed!(2023)
     www_path = "www"
     output_path = "artifacts"
-    img_height = 300
+    img_height = 300;
 
 end;

### 2023-04-04 src/penalties.jl

@@ -9,7 +9,7 @@ Penalty for smooth conformal set size.
 """
 function set_size_penalty(
     counterfactual_explanation::AbstractCounterfactualExplanation; 
-    κ::Real=1.0, temp::Real=0.5, agg=mean
+    κ::Real=0.0, temp::Real=0.05, agg=mean
 )
 
     conf_model = counterfactual_explanation.M.model

### 2023-04-04 src/ConformalGenerator.jl
work on granular results

@@ -1,149 +1,0 @@
-using CategoricalArrays
-using CounterfactualExplanations
-using CounterfactualExplanations.Generators
-using CounterfactualExplanations.Models: binary_to_onehot
-using CounterfactualExplanations.Objectives
-using Flux
-using LinearAlgebra
-using Parameters
-using SliceMap
-using Statistics
-
-mutable struct ConformalGenerator <: AbstractGradientBasedGenerator
-    loss::Union{Nothing,Function} # loss function
-    complexity::Function # complexity function
-    λ::Union{AbstractFloat,AbstractVector} # strength of penalty
-    decision_threshold::Union{Nothing,AbstractFloat}
-    opt::Flux.Optimise.AbstractOptimiser # optimizer
-    τ::AbstractFloat # tolerance for convergence
-    κ::Real
-    temp::Real
-end
-
-# API streamlining:
-@with_kw struct ConformalGeneratorParams
-    opt::Flux.Optimise.AbstractOptimiser = Descent()
-    τ::AbstractFloat = 1e-3
-    κ::Real = 1.0
-    temp::Real = 0.5
-end
-
-"""
-    ConformalGenerator(;
-        loss::Union{Nothing,Function}=conformal_training_loss,
-        complexity::Function=norm,
-        λ::Union{AbstractFloat,AbstractVector}=[0.1, 1.0],
-        decision_threshold=nothing,
-        kwargs...
-    )
-
-An outer constructor method that instantiates a generic generator.
-
-# Examples
-```julia-repl
-generator = ConformalGenerator()
-```
-"""
-function ConformalGenerator(;
-    loss::Union{Nothing,Function}=conformal_training_loss,
-    complexity::Function=norm,
-    λ::Union{AbstractFloat,AbstractVector}=[0.1, 1.0],
-    decision_threshold=nothing,
-    kwargs...
-)
-    params = ConformalGeneratorParams(; kwargs...)
-    ConformalGenerator(loss, complexity, λ, decision_threshold, params.opt, params.τ, params.κ, params.temp)
-end
-
-@doc raw"""
-    conformal_training_loss(counterfactual_explanation::AbstractCounterfactualExplanation; kwargs...)
-
-A configurable classification loss function for Conformal Predictors.
-"""
-function conformal_training_loss(counterfactual_explanation::AbstractCounterfactualExplanation; kwargs...)
-    conf_model = counterfactual_explanation.M.model
-    fitresult = counterfactual_explanation.M.fitresult
[…]

### 2023-04-04 src/generator.jl

@@ -1,0 +1,11 @@
+using CounterfactualExplanations.Objectives
+
+"Constructor for `CCEGenerator`."
+function CCEGenerator(; λ::Union{AbstractFloat,Vector{<:AbstractFloat}}=[0.1, 1.0], κ::Real=0.0, temp::Real=0.05, kwargs...)
+    function _set_size_penalty(ce::AbstractCounterfactualExplanation)
+        return CCE.set_size_penalty(ce; κ=κ, temp=temp)
+    end
+    _penalties = [Objectives.distance_l2, _set_size_penalty]
+    λ = λ isa AbstractFloat ? [0.0, λ] : λ
+    return Generator(; penalty=_penalties, λ=λ, kwargs...)
+end

### 2023-04-04 notebooks/synthetic.qmd
work on granular results

```{julia}
include("notebooks/setup.jl")
eval(setup_notebooks);
```

# Synthetic data

```{julia}
#| output: false

# Data:
datasets = Dict(
    :linearly_separable => load_linearly_separable(),
    :overlapping => load_overlapping(),
    :moons => load_moons(),
    :circles => load_circles(),
    :multi_class => load_multi_class(),
)

# Hyperparameters:
cvgs = [0.5, 0.75, 0.90, 0.95, 0.99]
temps = [0.05, 0.1, 0.5, 1.0, 2.0]
Λ = [0.1, 1.0, 5.0]

# Classifiers:
epochs = 100
link_fun = sigmoid
logreg = NeuralNetworkClassifier(builder=MLJFlux.Linear(σ=link_fun), epochs=epochs)
mlp = NeuralNetworkClassifier(builder=MLJFlux.MLP(hidden=(32,), σ=link_fun), epochs=epochs)
ensmbl = EnsembleModel(model=mlp, n=5)
classifiers = Dict(
    :logreg => logreg,
    :mlp => mlp,
    # :ensmbl => ensmbl,
)

# Search parameters:
target = 2
factual = 1
```

```{julia}
results = DataFrame()
for (dataname, data) in datasets

    # Data:
    X = table(permutedims(data.X))
    y = data.output_encoder.labels

    for (clf_name, clf) in classifiers, cov in cvgs

        # Classifier and coverage:
        conf_model = conformal_model(clf; method=:simple_inductive, coverage=cov)
        mach = machine(conf_model, X, y)
        fit!(mach)
        M = CCE.ConformalModel(mach.model, mach.fitresult)

        # Set up CCE:
        yhat = predict_label(M, data)
        factual_label = data.y_levels[factual]
        target_label = data.y_levels[target]
        x = select_factual(data,rand(findall(yhat .== factual_label)))

        for λ in Λ, temp in temps

            # CCE for given classifier, coverage, temperature and λ:
            generator = CCEGenerator(temp=temp, λ=λ)
            @assert predict_label(M, data, x) != target_label
            ce = try
                generate_counterfactual(
                    x, target_label, data, M, generator;
                    initialization=:identity,
                )
            catch
                missing
            end

            _results = DataFrame(
                dataset = dataname,
                classifier = clf_name,
                coverage = cov,
                temperature = temp,
                λ = λ,
                counterfactual = ce,
                factual = factual_label,
                target = target_label,
            )
            append!(results, _results)

        end

    end

end
```

```{julia}

```

### 2023-04-04 src/CCE.jl

@@ -8,9 +8,10 @@ include("penalties.jl")
 include("losses.jl")
 include("generator.jl")
 include("sampling.jl")
-# include("ConformalGenerator.jl")
 
 using MLJFlux
-MLJFlux.reformat(X, ::Type{<:AbstractMatrix}) = permutedims(X)
+MLJFlux.reformat(X, ::Type{<:AbstractMatrix}) = X'
+
+export CCEGenerator, EnergySampler, set_size_penalty, distance_from_energy
 
 end

### 2023-04-04 notebooks/synthetic.qmd
uh

@@
 classifiers = Dict(
-    :logreg => logreg,
+    # :logreg => logreg,
     :mlp => mlp,
@@
                     initialization=:identity,
+                    converge_when=:generator_conditions,
                 )

### 2023-04-04 notebooks/synthetic.qmd
progres on granular examination

--- 2023-04-04 notebooks/synthetic.qmd
@@
 # Hyperparameters:
-cvgs = [0.5, 0.75, 0.90, 0.95, 0.99]
-temps = [0.05, 0.1, 0.5, 1.0, 2.0]
-Λ = [0.1, 1.0, 5.0]
+cvgs = [0.5, 0.75, 0.95]
+temps = [0.01, 0.1, 1.0]
+Λ = [0.0, 0.1, 1.0]
+l2_λ = 0.1
 
 # Classifiers:
-epochs = 100
-link_fun = sigmoid
+epochs = 250
+link_fun = relu
 logreg = NeuralNetworkClassifier(builder=MLJFlux.Linear(σ=link_fun), epochs=epochs)
@@
 factual = 1
+max_iter = 50
+gradient_tol = 1e-2
+opt = Descent(0.01)
 ```
@@
     y = data.output_encoder.labels
+    x = select_factual(data,rand(1:size(data.X,2)))
 
@@
         # Set up CCE:
-        yhat = predict_label(M, data)
-        factual_label = data.y_levels[factual]
-        target_label = data.y_levels[target]
-        x = select_factual(data,rand(findall(yhat .== factual_label)))
+        factual_label = predict_label(M, data, x)[1]
+        target_label = data.y_levels[data.y_levels .!= factual_label][1]
 
@@
             # CCE for given classifier, coverage, temperature and λ:
-            generator = CCEGenerator(temp=temp, λ=λ)
+            generator = CCEGenerator(temp=temp, λ=[l2_λ,λ], opt=opt)
             @assert predict_label(M, data, x) != target_label
@@
                     initialization=:identity,
+                    converge_when=:generator_conditions,
+                    gradient_tol=gradient_tol,
+                    max_iter=max_iter,
                 )
@@
                 λ = λ,
-                counterfactual = ce,
+                ce = ce,
                 factual = factual_label,
@@
 ```{julia}
+function plot_ce(results; dataset=:multi_class, classifier=:mlp, λ=0.1, img_height=300, zoom=-0.2)
+    df_plot = results[results.dataset .== dataset,:] |>
+        res -> res[res.classifier .== classifier,:] |>
+        res -> res[res.λ .== λ,:]
+    plts = map(eachrow(df_plot)) do row
+        Plots.plot(
+            row.ce,
+            title="cov: $(row.coverage), temp: $(row.temperature)",
+            cbar=false,
+            zoom=zoom,
+            legend=false,
+        )
+    end
+    nrow = length(cvgs)
+    ncol = length(temps)
+    _layout = (nrow, ncol)
+    Plots.plot(
+        plts..., 
+        size=img_height.*reverse(_layout), layout=_layout,
+        plot_title="λ: $λ, dataset: $dataset, classifier: $classifier",
+    )
+end
+```
 
+```{julia}
+plot_ce(results; λ=Λ[1], dataset=:moons)
 ```

### 2023-04-04 src/generator.jl
progres on granular examination

@@ -1,7 +1,7 @@
 using CounterfactualExplanations.Objectives
 
 "Constructor for `CCEGenerator`."
-function CCEGenerator(; λ::Union{AbstractFloat,Vector{<:AbstractFloat}}=[0.1, 1.0], κ::Real=0.0, temp::Real=0.05, kwargs...)
+function CCEGenerator(; λ::Union{AbstractFloat,Vector{<:AbstractFloat}}=[0.1, 1.0], κ::Real=1.0, temp::Real=0.05, kwargs...)
     function _set_size_penalty(ce::AbstractCounterfactualExplanation)
         return CCE.set_size_penalty(ce; κ=κ, temp=temp)
     end

### 2023-04-04 notebooks/synthetic.qmd
more work on synthetic

@@
 ```{julia}
+#| echo: false
+
 results = DataFrame()
@@
 ```{julia}
+#| echo: false
+
 function plot_ce(results; dataset=:multi_class, classifier=:mlp, λ=0.1, img_height=300, zoom=-0.2)
@@
 ```{julia}
-plot_ce(results; λ=Λ[1], dataset=:moons)
+#| output: true
+#| echo: false
+
+for dataset in keys(datasets)
+    Markdown.parse("""### $dataset""")
+    for classifier in keys(classifiers)
+        Markdown.parse("""#### $classifier""")
+        Markdown.parse("""::: {.panel-tabset}""")
+        for λ in Λ
+            Markdown.parse("""##### λ: $λ""")
+            display(plot_ce(results; dataset=dataset, classifier=classifier, λ=λ))
+        end
+        Markdown.parse(""":::""")
+    end
+end
 ```
+
+## Benchmark
+
+```{julia}
+generators = Dict(
+    :wachter => GenericGenerator(opt=opt, λ=l2_λ),
+    :revise => REVISEGenerator(opt=opt, λ=l2_λ),
+    :greedy => GreedyGenerator(opt=opt),
+)
+
+# Untrained Models:
+models = Dict("cov$(cov)" => CCE.ConformalModel(conformal_model(mlp; method=:simple_inductive, coverage=cov)) for cov in cvgs)
+```
+
+```{julia}
+bmks = []
+measures = [
+    CounterfactualExplanations.distance,
+    CCE.distance_from_energy,
+    CCE.distance_from_targets,
+    CounterfactualExplanations.validity,
+]
+for (dataname, dataset) in datasets
+    for λ in Λ, temp in temps
+        _generators = deepcopy(generators)
+        _generators[:cce] = CCEGenerator(temp=temp, λ=[l2_λ,λ], opt=opt)
+        _generators[:energy] = CCE.EnergyDrivenGenerator(λ=[l2_λ,λ], opt=opt)
+        _generators[:target] = CCE.TargetDrivenGenerator(λ=[l2_λ,λ], opt=opt)
+        bmk = benchmark(
+            dataset; 
+            models=deepcopy(models), 
+            generators=_generators, 
+            measure=measures,
+            suppress_training=false, dataname=dataname,
+            n_individuals=5,
+            initialization=:identity,
+        )
+        bmk.evaluation.λ .= λ
+        bmk.evaluation.temperature .= temp
+        push!(bmks, bmk)
+    end
+end
+bmk = reduce(vcat, bmks)
+```
+
+```{julia}
+CSV.write(joinpath(output_path, "synthetic_benchmark.csv"), bmk())
+```
+
+```{julia}
+#| output: true
+#| echo: false
+
+df = bmk()
+for dataname ∈ sort(unique(df.dataname))
+    Markdown.parse("""### $dataset""")
+    df_ = df[df.dataname .== dataname, :]
+    for λ in Λ, temp in temps
+        Markdown.parse("""#### λ: $λ""")
+        df_plot = df_[df_.λ .== λ, :]
+        df_plot = df_plot[df_plot.temperature .== temp, :]
+        plt = AlgebraOfGraphics.data(df_plot) * visual(BoxPlot) * 
+            mapping(:generator, :value, row=:variable, col=:model, color=:generator)
+        plt = draw(
+            plt, axis=(xlabel="", xticksvisible=false, xticklabelsvisible=false, width=200, height=180), 
+            facet=(; linkyaxes=:minimal)
+        )   
+        display(plt)
+    end
+end
+```

### 2023-04-04 src/generator.jl

@@ -9,3 +9,17 @@ function CCEGenerator(; λ::Union{AbstractFloat,Vector{<:AbstractFloat}}=[0.1, 1
     λ = λ isa AbstractFloat ? [0.0, λ] : λ
     return Generator(; penalty=_penalties, λ=λ, kwargs...)
 end
+
+"Constructor for `EnergyDrivenGenerator`."
+function EnergyDrivenGenerator(; λ::Union{AbstractFloat,Vector{<:AbstractFloat}}=[0.1, 1.0], kwargs...)
+    _penalties = [Objectives.distance_l2, CCE.distance_from_energy]
+    λ = λ isa AbstractFloat ? [0.0, λ] : λ
+    return Generator(; penalty=_penalties, λ=λ, kwargs...)
+end
+
+"Constructor for `TargetDrivenGenerator`."
+function TargetDrivenGenerator(; λ::Union{AbstractFloat,Vector{<:AbstractFloat}}=[0.1, 1.0], kwargs...)
+    _penalties = [Objectives.distance_l2, CCE.distance_from_targets]
+    λ = λ isa AbstractFloat ? [0.0, λ] : λ
+    return Generator(; penalty=_penalties, λ=λ, kwargs...)
+end

### 2023-04-04 notebooks/setup.jl
more work on synthetic

@@ -12,13 +12,16 @@ setup_notebooks = quote
     using ConformalPrediction
     using CounterfactualExplanations
     using CounterfactualExplanations.Data
+    using CounterfactualExplanations.Evaluation: benchmark
     using CounterfactualExplanations.Models: load_mnist_mlp
     using CounterfactualExplanations.Objectives
+    using CSV
     using Distributions
     using Flux
     using Images
     using JointEnergyModels
     using LinearAlgebra
+    using Markdown
     using MLDatasets
     using MLDatasets: convert2image
     using MLJBase
@@ -34,7 +37,7 @@ setup_notebooks = quote
     Plots.theme(:wong)
     Random.seed!(2023)
     www_path = "www"
-    output_path = "artifacts"
+    output_path = "artifacts/results"
     img_height = 300
 
 end;

### 2023-04-05 notebooks/synthetic.qmd
more work on granular stuff

@@
 temps = [0.01, 0.1, 1.0]
-Λ = [0.0, 0.1, 1.0]
+Λ = [0.0, 0.1, 1.0, 10.0]
 l2_λ = 0.1
@@
 ```{julia}
+# Benchmark generators:
 generators = Dict(
@@
     :revise => REVISEGenerator(opt=opt, λ=l2_λ),
-    :greedy => GreedyGenerator(opt=opt),
+    :greedy => GreedyGenerator(),
 )
@@
 # Untrained Models:
-models = Dict("cov$(cov)" => CCE.ConformalModel(conformal_model(mlp; method=:simple_inductive, coverage=cov)) for cov in cvgs)
-```
+models = Dict(Symbol("cov$(Int(100*cov))") => CCE.ConformalModel(conformal_model(mlp; method=:simple_inductive, coverage=cov)) for cov in cvgs)
 
-```{julia}
-bmks = []
+# Measures:
 measures = [
@@
 ]
+```
+
+### Single CE
+
+```{julia}
+#| echo: false
+
+_temp = 0.01
+results = DataFrame()
+for (dataname, data) in datasets
+
+    # Data:
+    X = table(permutedims(data.X))
+    y = data.output_encoder.labels
+    x = select_factual(data,rand(1:size(data.X,2)))
+
+    for (modelname, M) in deepcopy(models)
+
+        # Model training:
+        M = train(M, data)
+        # Set up CCE:
+        factual_label = predict_label(M, data, x)[1]
+        target_label = data.y_levels[data.y_levels .!= factual_label][1]
+
+        for λ in Λ
+
+            # Generators:
+            _generators = deepcopy(generators)
+            _generators[:cce] = CCEGenerator(temp=_temp, λ=[l2_λ,λ], opt=opt)
+            _generators[:energy] = CCE.EnergyDrivenGenerator(λ=[l2_λ,λ], opt=opt)
+            _generators[:target] = CCE.TargetDrivenGenerator(λ=[l2_λ,λ], opt=opt)
+
+            for (gen_name, gen) in _generators
+
+                # CCE for given models, λ and generator:
+                @assert predict_label(M, data, x) != target_label
+                ce = try
+                    generate_counterfactual(
+                        x, target_label, data, M, gen;
+                        initialization=:identity,
+                        converge_when=:generator_conditions,
+                        gradient_tol=gradient_tol,
+                        max_iter=max_iter,
+                    )
+                catch
+                    missing
+                end
+
+                if !ismissing(ce)
+                    eval = DataFrame(evaluate(ce, measure=measures, output_format=:Dict))
+                else
+                    eval = DataFrame(Dict(Symbol(fun) => missing for fun in measures))
+                end
+
+                _results = DataFrame(
+                    dataset = dataname,
+                    model = modelname,
+                    λ = λ,
+                    generator = gen_name,
+                    ce = ce,
+                    factual = factual_label,
+                    target = target_label,
+                )
+
+                _results = crossjoin(_results, eval; makeunique=true)
+                append!(results, _results)
+
+            end
+        end
+    end
+end
+```
+
+```{julia}
+#| echo: false
+
+function plot_benchmark(results; dataset=:multi_class, modelname=:cov95, img_height=300, zoom=-0.2)
+    df_plot = results[results.dataset .== dataset,:] |>
+        res -> res[res.model .== modelname,:]
+    plts = map(eachrow(df_plot)) do row
+        Plots.plot(
+            row.ce,
+            title="λ: $(row.λ), gen: $(row.generator)",
+            cbar=false,
+            zoom=zoom,
+            legend=false,
+        )
+    end
+    ncol = length(unique(df_plot.generator))
+    nrow = length(unique(df_plot.λ))
+    _layout = (nrow, ncol)
+    Plots.plot(
+        plts..., 
+        size=img_height.*reverse(_layout), layout=_layout,
+        plot_title="dataset: $dataset, model: $modelname",
+    )
+end
+```
+
+```{julia}
+#| output: true
+#| echo: false
+
+df = @pivot_longer(results, distance:distance_from_targets)
+for dataname ∈ sort(unique(df.dataset))
+    Markdown.parse("""### $dataname""")
+    df_ = df[df.dataset .== dataname, :]
+    for model in unique(df_.model)
+        Markdown.parse("""#### model: $model""")
+        df_plot = df_[df_.model .== model, :]
+        df_plot = @mutate(df_plot, lambda = string("λ: ", round(λ, digits=2)))
+        plt = AlgebraOfGraphics.data(df_plot) * visual(BarPlot) * 
+            mapping(:generator, :value, row=:variable, col=:lambda, color=:generator)
+        plt = draw(
+            plt, axis=(xlabel="", xticksvisible=false, xticklabelsvisible=false, width=200, height=180), 
+            facet=(; linkyaxes=:minimal)
+        )   
+        # plt.figure[0, :] = Label(
+        #     plt.figure, "data: $dataname, model: $model", 
+        #     fontsize=20, tellwidth=false
+        # )
+        display(plt)
+    end
+end
+```
+
+### Full Benchmark
+
+```{julia}
+bmks = []
 for (dataname, dataset) in datasets
@@
 for dataname ∈ sort(unique(df.dataname))
-    Markdown.parse("""### $dataset""")
+    Markdown.parse("""### $dataname""")
     df_ = df[df.dataname .== dataname, :]

### 2023-04-05 src/penalties.jl

@@ -36,7 +36,7 @@ end
 
 function distance_from_energy(
     counterfactual_explanation::AbstractCounterfactualExplanation;
-    n::Int=100, from_buffer=true, agg=mean, kwargs...
+    n::Int=10000, from_buffer=true, agg=mean, kwargs...
 )
     conditional_samples = []
     ignore_derivatives() do
@@ -63,9 +63,10 @@ end
 
 function distance_from_targets(
     counterfactual_explanation::AbstractCounterfactualExplanation;
-    n::Int=100, agg=mean
+    n::Int=10000, agg=mean
 )
-    target_samples = counterfactual_explanation.data.X |>
+    target_idx = counterfactual_explanation.data.output_encoder.labels .== counterfactual_explanation.target
+    target_samples = counterfactual_explanation.data.X[:,target_idx] |>
         X -> X[:,rand(1:end,n)]
     x′ = CounterfactualExplanations.counterfactual(counterfactual_explanation)
     loss = map(eachslice(x′, dims=3)) do x

### 2023-04-05 notebooks/mnist.qmd

@@
 
+# MNIST
+
 ```{julia}
 # Data:
-counterfactual_data = load_mnist()
+counterfactual_data = load_mnist(1000)
 X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
+X = table(permutedims(X))
 labels = counterfactual_data.output_encoder.labels
 input_dim, n_obs = size(counterfactual_data.X)
-M = load_mnist_mlp()
-
-# Target:
-factual_label = 8
-x = reshape(X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
-target = 3
-factual = predict_label(M, counterfactual_data, x)[1]
-γ = 0.9
-T = 50
 ```
@@
 ```{julia}
-builder = MLJFlux.@builder M.model
-clf = NeuralNetworkClassifier(builder=builder, epochs=100)
+epochs = 100
+clf = NeuralNetworkClassifier(builder=MLJFlux.MLP(hidden=(32,), σ=relu), epochs=epochs)
 conf_model = conformal_model(clf; method=:simple_inductive)
@@
 ```{julia}
-# Search:
+M = CCE.ConformalModel(mach.model, mach.fitresult)
+```
+
+```{julia}
+test_data = load_mnist_test()
+f1 = CounterfactualExplanations.Models.model_evaluation(M, test_data)
+println("F1 score (test): $(round(f1,digits=3))")
+```
+
+```{julia}
+# Set up search:
+factual_label = 8
+x = reshape(counterfactual_data.X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
+target = 3
+factual = predict_label(M, counterfactual_data, x)[1]
+γ = 0.9
+T = 100
+
+# Generate counterfactual using generic generator:
 generator = GenericGenerator()
@@
 )
-generator = deepcopy(generator) |> gen -> @objective(gen, _ + 0.001distance + 1.0set_size_penalty)
+
+# Generate counterfactual using CCE generator:
+generator = CCEGenerator(λ=[0.0,10.0], temp=0.01, opt=CounterfactualExplanations.Generators.JSMADescent(η=5.0))
 ce_conformal = generate_counterfactual(
@@
     initialization=:identity,
+    converge_when=:generator_conditions,
 )
@@
 ```{julia}
-p1 = plot(
+p1 = Plots.plot(
     convert2image(MNIST, reshape(x,28,28)),
@@
 
-ces = zip([ce_wachter,ce_jsma])
+ces = zip([ce_wachter,ce_conformal])
 counterfactuals = reduce((x,y)->cat(x,y,dims=3),map(ce -> CounterfactualExplanations.counterfactual(ce[1]), ces))
@@
     _title = "p(y=$(target)|x′)=$(round(_phat[1]; digits=3))"
-    plt = plot(
+    plt = Plots.plot(
         convert2image(MNIST, reshape(ce,28,28)),
@@
 end
-plt = plot(plts...; size=(img_height*length(plts),img_height), layout=(1,length(plts)))
-savefig(plt, joinpath(www_path, "you_may_not_like_it.png"))
+plt = Plots.plot(plts...; size=(img_height*length(plts),img_height), layout=(1,length(plts)))
+display(plt)
+savefig(plt, joinpath(www_path, "cce_mnist.png"))
 ```

### 2023-04-05 notebooks/mnist.qmd
ufff

@@
 # Data:
-counterfactual_data = load_mnist(1000)
+counterfactual_data = load_mnist()
 X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
@@
 epochs = 100
-clf = NeuralNetworkClassifier(builder=MLJFlux.MLP(hidden=(32,), σ=relu), epochs=epochs)
+clf = NeuralNetworkClassifier(
+    builder=MLJFlux.MLP(hidden=(32,), σ=relu), 
+    epochs=epochs,
+    batch_size=Int(round(n_obs/10))
+)
 conf_model = conformal_model(clf; method=:simple_inductive)
@@
 ```{julia}
+using CounterfactualExplanations.DataPreprocessing: undersample
+dt_reduced = undersample(counterfactual_data, 100)
+# dt_reduced = counterfactual_data
+```
+
+```{julia}
 # Set up search:
-factual_label = 8
+factual_label = 9
 x = reshape(counterfactual_data.X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
-target = 3
+target = 4
 factual = predict_label(M, counterfactual_data, x)[1]
@@
 ce_wachter = generate_counterfactual(
-    x, target, counterfactual_data, M, generator; 
+    x, target, dt_reduced, M, generator; 
     decision_threshold=γ, max_iter=T,
@@
 # Generate counterfactual using CCE generator:
-generator = CCEGenerator(λ=[0.0,10.0], temp=0.01, opt=CounterfactualExplanations.Generators.JSMADescent(η=5.0))
+generator = CCEGenerator(
+    λ=[0.0,10.0], 
+    temp=0.01, 
+    # opt=CounterfactualExplanations.Generators.JSMADescent(η=5.0),
+)
 ce_conformal = generate_counterfactual(
-    x, target, counterfactual_data, M, generator; 
+    x, target, dt_reduced, M, generator; 
     decision_threshold=γ, max_iter=T,

### 2023-04-05 notebooks/mnist.qmd
pfff

@@
 )
-conf_model = conformal_model(clf; method=:simple_inductive)
+conf_model = conformal_model(clf; method=:simple_inductive, coverage=.99)
 mach = machine(conf_model, X, labels)
@@
 dt_reduced = undersample(counterfactual_data, 100)
-# dt_reduced = counterfactual_data
+dt_reduced = counterfactual_data
 ```

### 2023-04-06 notebooks/mnist.qmd
bloody jupyter seriously

@@
 )
-conf_model = conformal_model(clf; method=:simple_inductive, coverage=.99)
+conf_model = conformal_model(clf; method=:adaptive_inductive, coverage=.99)
 mach = machine(conf_model, X, labels)
@@
 # Set up search:
-factual_label = 9
+factual_label = 8
 x = reshape(counterfactual_data.X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
-target = 4
+target = 3
 factual = predict_label(M, counterfactual_data, x)[1]
@@
 generator = CCEGenerator(
-    λ=[0.0,10.0], 
+    λ=[0.0,100.0], 
     temp=0.01, 
@@
 )
-```
 
-```{julia}
+# Plot:
 p1 = Plots.plot(

### 2023-04-06 notebooks/mnist.qmd
resnet

--- 2023-04-05 notebooks/mnist.qmd
@@
 
+First, let's create a couple of image classifier architectures:
+
 ```{julia}
 epochs = 100
-clf = NeuralNetworkClassifier(
+batchsize = Int(round(n_obs/10))
+
+# Simple MLP:
+mlp = NeuralNetworkClassifier(
     builder=MLJFlux.MLP(hidden=(32,), σ=relu), 
     epochs=epochs,
-    batch_size=Int(round(n_obs/10))
+    batch_size=batchsize,
 )
-conf_model = conformal_model(clf; method=:simple_inductive)
-mach = machine(conf_model, X, labels)
-fit!(mach)
+
+# Deep Ensemble:
+mlp_ens = EnsembleModel(model=mlp, n=5)
+
+# ResNet:
+using MLJFlux, Metalhead
+builder = MLJFlux.image_builder(ResNet, 18, pretrain=true)
+resnet = ImageClassifier(
+    builder=builder,
+    epochs=epochs,
+    batch_size=batchsize,
+)
+
+# ResNet Ensemble:
+resnet_ens = EnsembleModel(model=resnet, n=5)
 ```
@@
 ```{julia}
+
+conf_model = conformal_model(mlp; method=:adaptive_inductive, coverage=.99)
+mach = machine(conf_model, X, labels)
+fit!(mach)
 M = CCE.ConformalModel(mach.model, mach.fitresult)
@@
 println("F1 score (test): $(round(f1,digits=3))")
-```
-
-```{julia}
-using CounterfactualExplanations.DataPreprocessing: undersample
-dt_reduced = undersample(counterfactual_data, 100)
-# dt_reduced = counterfactual_data
 ```
@@
     temp=0.01, 
-    # opt=CounterfactualExplanations.Generators.JSMADescent(η=5.0),
+    # opt=CounterfactualExplanations.Generators.JSMADescent(η=0.5),
 )
@@
 ```
+
+## Benchmark
+
+```{julia}
+# Benchmark generators:
+generators = Dict(
+    :wachter => GenericGenerator(opt=opt, λ=l2_λ),
+    :revise => REVISEGenerator(opt=opt, λ=l2_λ),
+    :greedy => GreedyGenerator(),
+)
+
+# Conformal Models:
+
+# Measures:
+measures = [
+    CounterfactualExplanations.distance,
+    CCE.distance_from_energy,
+    CCE.distance_from_targets,
+    CounterfactualExplanations.validity,
+]
+```

### 2023-04-06 src/model.jl

@@ -2,9 +2,13 @@ using ConformalPrediction
 using CounterfactualExplanations.Models
 using Flux
 using MLJBase
+using MLJEnsembles
+using MLJFlux
 using MLUtils
 using SliceMap
 using Statistics
+
+const CompatibleAtomicModel = Union{<:MLJFlux.MLJFluxProbabilistic,MLJEnsembles.ProbabilisticEnsembleModel{<:MLJFlux.MLJFluxProbabilistic}}
 
 """
     ConformalModel <: Models.AbstractDifferentiableJuliaModel

### 2023-04-18 notebooks/mnist.qmd
almost there

@@
 # Data:
-counterfactual_data = load_mnist()
+n_obs = 1000
+counterfactual_data = load_mnist(n_obs)
 X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
@@
 input_dim, n_obs = size(counterfactual_data.X)
+n_digits = Int(sqrt(input_dim))
+output_dim = length(unique(labels))
 ```
@@
 ```{julia}
+# Model parameters:
 epochs = 100
-batchsize = Int(round(n_obs/10))
+batch_size = Int(round(n_obs/10))
+n_hidden = 32
+activation = Flux.swish
+builder = MLJFlux.MLP(hidden=(n_hidden,), σ=activation)
+α = [0.33,1.0,1e-1]
 
@@
 mlp = NeuralNetworkClassifier(
-    builder=MLJFlux.MLP(hidden=(32,), σ=relu), 
+    builder=builder, 
     epochs=epochs,
-    batch_size=batchsize,
+    batch_size=batch_size,
+)
+
+# Joint Energy Model:
+𝒟x = Uniform(-1,1)
+𝒟y = Categorical(ones(output_dim) ./ output_dim)
+sampler = ConditionalSampler(𝒟x, 𝒟y, input_size=(input_dim,), batch_size=batch_size)
+jem = JointEnergyClassifier(
+    sampler;
+    builder=builder,
+    batch_size=batch_size,
+    finaliser=Flux.softmax,
+    loss=Flux.Losses.crossentropy,
+    jem_training_params=(α=α,verbosity=10,),
+    sampling_steps=20,
 )
@@
 mlp_ens = EnsembleModel(model=mlp, n=5)
-
-# ResNet:
-using MLJFlux, Metalhead
-builder = MLJFlux.image_builder(ResNet, 18, pretrain=true)
-resnet = ImageClassifier(
-    builder=builder,
-    epochs=epochs,
-    batch_size=batchsize,
-)
-
-# ResNet Ensemble:
-resnet_ens = EnsembleModel(model=resnet, n=5)
 ```
@@
 ```{julia}
-
-conf_model = conformal_model(mlp; method=:adaptive_inductive, coverage=.99)
+cov = .9
+conf_model = conformal_model(jem; method=:adaptive_inductive, coverage=cov)
 mach = machine(conf_model, X, labels)
@@
 M = CCE.ConformalModel(mach.model, mach.fitresult)
+```
+
+```{julia}
+jem = mach.model.model.jem
+n_iter = 5000
+_w = 1500
+plts = []
+neach = 10
+for i in 1:10
+    x = jem.sampler(jem.chain, jem.sampling_rule; niter=n_iter, n_samples=neach, y=i)
+    plts_i = []
+    for j in 1:size(x, 2)
+        xj = x[:,j]
+        xj = reshape(xj, (n_digits, n_digits))
+        plts_i = [plts_i..., Plots.heatmap(rotl90(xj), axis=nothing, cb=false)]
+    end
+    plt = Plots.plot(plts_i..., size=(_w,0.10*_w), layout=(1,10))
+    plts = [plts..., plt]
+end
+Plots.plot(plts..., size=(_w,_w), layout=(10,1))
 ```
@@
 ce_wachter = generate_counterfactual(
-    x, target, dt_reduced, M, generator; 
+    x, target, counterfactual_data, M, generator; 
     decision_threshold=γ, max_iter=T,
@@
 ce_conformal = generate_counterfactual(
-    x, target, dt_reduced, M, generator; 
+    x, target, counterfactual_data, M, generator; 
     decision_threshold=γ, max_iter=T,

### 2023-04-18 notebooks/mnist.qmd
shoot me

@@
 activation = Flux.swish
-builder = MLJFlux.MLP(hidden=(n_hidden,), σ=activation)
-α = [0.33,1.0,1e-1]
+builder = MLJFlux.@builder Flux.Chain(
+    Dense(n_in, n_hidden),
+    BatchNorm(n_hidden, activation),
+    Dense(n_hidden, n_out),
+    BatchNorm(n_out)
+)
+# builder = MLJFlux.MLP(
+#     hidden=(
+#         n_hidden,
+#         n_hidden,
+#         n_hidden,
+#     ), 
+#     σ=activation
+# )
+α = [1.0,1.0,1e-1]
 
@@
 # Joint Energy Model:
-𝒟x = Uniform(-1,1)
+𝒟x = Uniform(0,1)
 𝒟y = Categorical(ones(output_dim) ./ output_dim)
@@
     batch_size=batch_size,
-    finaliser=Flux.softmax,
-    loss=Flux.Losses.crossentropy,
+    finaliser=x -> x,
+    loss=Flux.Losses.logitcrossentropy,
     jem_training_params=(α=α,verbosity=10,),
     sampling_steps=20,
+    epochs=epochs,
 )
@@
 ```{julia}
-cov = .9
-conf_model = conformal_model(jem; method=:adaptive_inductive, coverage=cov)
+cov = .90
+conf_model = conformal_model(jem; method=:simple_inductive, coverage=cov)
 mach = machine(conf_model, X, labels)
@@
 jem = mach.model.model.jem
-n_iter = 5000
+n_iter = 100
 _w = 1500
@@
 end
-Plots.plot(plts..., size=(_w,_w), layout=(10,1))
+plt = Plots.plot(plts..., size=(_w,_w), layout=(10,1))
+display(plt)
 ```
@@
 ```{julia}
+Random.seed!(1234)
+
 # Set up search:
-factual_label = 9
+factual_label = 2
 x = reshape(counterfactual_data.X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
-target = 4
+target = 0
 factual = predict_label(M, counterfactual_data, x)[1]
-γ = 0.9
+γ = 0.5
 T = 100
@@
 generator = CCEGenerator(
-    λ=[0.0,10.0], 
-    temp=0.01, 
-    # opt=CounterfactualExplanations.Generators.JSMADescent(η=0.5),
+    λ=[0.0,1.0], 
+    temp=0.5, 
+    # opt=CounterfactualExplanations.Generators.JSMADescent(η=1.0),
 )

### 2023-04-18 src/model.jl

@@ -86,7 +86,11 @@ function Models.logits(M::ConformalModel, X::AbstractArray)
             p̂ = [p̂]
         end
         p̂ = reduce(hcat, p̂)
-        ŷ = reduce(hcat, (map(p -> log.(p) .+ log(sum(exp.(p))), eachcol(p̂))))
+        if all(0.0 .<= vec(p̂) .<= 1.0)
+            ŷ = reduce(hcat, (map(p -> log.(p) .+ log(sum(exp.(p))), eachcol(p̂))))
+        else
+            ŷ = p̂
+        end
         if M.likelihood == :classification_binary
             ŷ = reduce(hcat, (map(y -> y[2] - y[1], eachcol(ŷ))))
         end

### 2023-04-19 notebooks/mnist.qmd
:cry:

@@
 # Data:
-n_obs = 1000
+n_obs = 10000
 counterfactual_data = load_mnist(n_obs)
@@
 epochs = 100
-batch_size = Int(round(n_obs/10))
+batch_size = minimum([Int(round(n_obs/10)), 100])
 n_hidden = 32
-activation = Flux.swish
+activation = Flux.relu
 builder = MLJFlux.@builder Flux.Chain(
@@
 )
+# builder = MLJFlux.Short(n_hidden=n_hidden, dropout=0.2, σ=activation)
 # builder = MLJFlux.MLP(
@@
 # )
-α = [1.0,1.0,1e-1]
+α = [1.0,1.0,5e-1]
 
@@
     loss=Flux.Losses.logitcrossentropy,
-    jem_training_params=(α=α,verbosity=10,),
+    jem_training_params=(
+        α=α,verbosity=10,
+        use_gen_loss=false,
+        use_reg_loss=false,
+    ),
     sampling_steps=20,
@@
 # Set up search:
-factual_label = 2
+factual_label = 9
 x = reshape(counterfactual_data.X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
-target = 0
+target = 4
 factual = predict_label(M, counterfactual_data, x)[1]
 γ = 0.5
-T = 100
+T = 1
 

### 2023-04-19 notebooks/mnist.qmd
:cry:

@@
     Dense(n_hidden, n_out),
-    BatchNorm(n_out)
 )
@@
 # )
-α = [1.0,1.0,5e-1]
+α = [1.0,1.0,1e-2]
 
@@
         α=α,verbosity=10,
-        use_gen_loss=false,
-        use_reg_loss=false,
+        # use_gen_loss=false,
+        # use_reg_loss=false,
     ),

### 2023-04-19 notebooks/mnist.qmd
batchnorm sorted

@@
 epochs = 100
-batch_size = minimum([Int(round(n_obs/10)), 100])
-n_hidden = 32
+batch_size = minimum([Int(round(n_obs/10)), 128])
+n_hidden = 200
 activation = Flux.relu
@@
     Dense(n_in, n_hidden),
+    BatchNorm(n_hidden, activation),
+    Dense(n_hidden, n_hidden),
     BatchNorm(n_hidden, activation),
@@
 𝒟y = Categorical(ones(output_dim) ./ output_dim)
-sampler = ConditionalSampler(𝒟x, 𝒟y, input_size=(input_dim,), batch_size=batch_size)
+sampler = ConditionalSampler(
+    𝒟x, 𝒟y, 
+    input_size=(input_dim,), 
+    batch_size=10
+)
 jem = JointEnergyClassifier(
@@
 cov = .90
-conf_model = conformal_model(jem; method=:simple_inductive, coverage=cov)
+conf_model = conformal_model(mlp; method=:adaptive_inductive, coverage=cov)
 mach = machine(conf_model, X, labels)
@@
 ```{julia}
+if mach.model.model isa JointEnergyModels.JointEnergyClassifier
     jem = mach.model.model.jem
@@
     display(plt)
+end
 ```
@@
 # Set up search:
-factual_label = 9
+factual_label = 4
 x = reshape(counterfactual_data.X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
-target = 4
+target = 9
 factual = predict_label(M, counterfactual_data, x)[1]
 γ = 0.5
-T = 1
+T = 100
 
@@
 
+# CCE:
+λ=[0.0,1.0]
+temp=0.5
+
 # Generate counterfactual using CCE generator:
 generator = CCEGenerator(
-    λ=[0.0,1.0], 
-    temp=0.5, 
-    # opt=CounterfactualExplanations.Generators.JSMADescent(η=1.0),
+    λ=λ, 
+    temp=temp, 
 )
 ce_conformal = generate_counterfactual(
+    x, target, counterfactual_data, M, generator; 
+    decision_threshold=γ, max_iter=T,
+    initialization=:identity,
+    converge_when=:generator_conditions,
+)
+
+# Generate counterfactual using CCE generator:
+generator = CCEGenerator(
+    λ=λ, 
+    temp=temp, 
+    opt=CounterfactualExplanations.Generators.JSMADescent(η=1.0),
+)
+ce_conformal_jsma = generate_counterfactual(
     x, target, counterfactual_data, M, generator; 
@@
 
-ces = zip([ce_wachter,ce_conformal])
+ces = zip([ce_wachter,ce_conformal,ce_conformal_jsma])
+_names = ["Wachter", "CCE", "CCE-JSMA"]
 counterfactuals = reduce((x,y)->cat(x,y,dims=3),map(ce -> CounterfactualExplanations.counterfactual(ce[1]), ces))
 phat = reduce((x,y) -> cat(x,y,dims=3), map(ce -> target_probs(ce[1]), ces))
-for x in zip(eachslice(counterfactuals; dims=3), eachslice(phat; dims=3))
-    ce, _phat = (x[1],x[2])
-    _title = "p(y=$(target)|x′)=$(round(_phat[1]; digits=3))"
+for x in zip(eachslice(counterfactuals; dims=3), _names, eachslice(phat; dims=3))
+    ce, _name, _phat = (x[1],x[2],x[3])
+    _title = "$_name (p̂=$(round(_phat[1]; digits=3)))"
     plt = Plots.plot(

### 2023-04-19 src/model.jl
batchnorm sorted

@@ -52,6 +52,7 @@ function ConformalModel(model, fitresult=nothing; likelihood::Union{Nothing,Symb
     end
 
     # Construct model:
+    testmode!(fitresult[1])
     M = ConformalModel(model, fitresult, likelihood)
     return M
 end

### 2023-04-19 notebooks/mnist.qmd
:sob

@@
 cov = .90
-conf_model = conformal_model(mlp; method=:adaptive_inductive, coverage=cov)
+conf_model = conformal_model(jem; method=:adaptive_inductive, coverage=cov)
 mach = machine(conf_model, X, labels)
@@
 ```{julia}
-Random.seed!(1234)
+# Random.seed!(1234)
 
 # Set up search:
-factual_label = 4
+factual_label = 8
 x = reshape(counterfactual_data.X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
-target = 9
+target = 3
 factual = predict_label(M, counterfactual_data, x)[1]
@@
 # Generate counterfactual using generic generator:
-generator = GenericGenerator()
+generator = GenericGenerator(opt=Flux.Optimise.Adam(),)
 ce_wachter = generate_counterfactual(
+    x, target, counterfactual_data, M, generator; 
+    decision_threshold=γ, max_iter=T,
+    initialization=:identity,
+)
+
+generator = GreedyGenerator(η=1.0)
+ce_jsma = generate_counterfactual(
     x, target, counterfactual_data, M, generator; 
@@
 λ=[0.0,1.0]
-temp=0.5
+temp=0.01
 
@@
     temp=temp, 
+    opt=Flux.Optimise.Adam(),
 )
@@
 
-ces = zip([ce_wachter,ce_conformal,ce_conformal_jsma])
-_names = ["Wachter", "CCE", "CCE-JSMA"]
-counterfactuals = reduce((x,y)->cat(x,y,dims=3),map(ce -> CounterfactualExplanations.counterfactual(ce[1]), ces))
-phat = reduce((x,y) -> cat(x,y,dims=3), map(ce -> target_probs(ce[1]), ces))
-for x in zip(eachslice(counterfactuals; dims=3), _names, eachslice(phat; dims=3))
-    ce, _name, _phat = (x[1],x[2],x[3])
+ces = [ce_wachter, ce_conformal, ce_jsma, ce_conformal_jsma]
+_names = ["Wachter", "CCE", "JSMA", "CCE-JSMA"]
+for x in zip(ces, _names)
+    ce, _name = (x[1],x[2])
+    x = CounterfactualExplanations.counterfactual(ce)
+    _phat = target_probs(ce)
     _title = "$_name (p̂=$(round(_phat[1]; digits=3)))"
     plt = Plots.plot(
-        convert2image(MNIST, reshape(ce,28,28)),
+        convert2image(MNIST, reshape(x,28,28)),
         axis=nothing, 

### 2023-04-20 notebooks/proposal.qmd
uh

@@
 In this section, we introduce an alternative way to generate high-fidelity Counterfactual Explanations.
-In particular, we propose Conformal Counterfactual Explanations (CCE), that is Counterfactual Explanations that minimize the predictive uncertainty of conformal models.
+In particular, we propose Conformal Counterfactual Explanations (ECCCE), that is Counterfactual Explanations that minimize the predictive uncertainty of conformal models.
 

### 2023-04-20 notebooks/synthetic.qmd
uh

@@
         fit!(mach)
-        M = CCE.ConformalModel(mach.model, mach.fitresult)
+        M = ECCCE.ConformalModel(mach.model, mach.fitresult)
 
-        # Set up CCE:
+        # Set up ECCCE:
         factual_label = predict_label(M, data, x)[1]
@@
 
-            # CCE for given classifier, coverage, temperature and λ:
-            generator = CCEGenerator(temp=temp, λ=[l2_λ,λ], opt=opt)
+            # ECCCE for given classifier, coverage, temperature and λ:
+            generator = ECCCEGenerator(temp=temp, λ=[l2_λ,λ], opt=opt)
             @assert predict_label(M, data, x) != target_label
@@
 # Untrained Models:
-models = Dict(Symbol("cov$(Int(100*cov))") => CCE.ConformalModel(conformal_model(mlp; method=:simple_inductive, coverage=cov)) for cov in cvgs)
+models = Dict(Symbol("cov$(Int(100*cov))") => ECCCE.ConformalModel(conformal_model(mlp; method=:simple_inductive, coverage=cov)) for cov in cvgs)
 
@@
     CounterfactualExplanations.distance,
-    CCE.distance_from_energy,
-    CCE.distance_from_targets,
+    ECCCE.distance_from_energy,
+    ECCCE.distance_from_targets,
     CounterfactualExplanations.validity,
@@
         M = train(M, data)
-        # Set up CCE:
+        # Set up ECCCE:
         factual_label = predict_label(M, data, x)[1]
@@
             _generators = deepcopy(generators)
-            _generators[:cce] = CCEGenerator(temp=_temp, λ=[l2_λ,λ], opt=opt)
-            _generators[:energy] = CCE.EnergyDrivenGenerator(λ=[l2_λ,λ], opt=opt)
-            _generators[:target] = CCE.TargetDrivenGenerator(λ=[l2_λ,λ], opt=opt)
+            _generators[:cce] = ECCCEGenerator(temp=_temp, λ=[l2_λ,λ], opt=opt)
+            _generators[:energy] = ECCCE.EnergyDrivenGenerator(λ=[l2_λ,λ], opt=opt)
+            _generators[:target] = ECCCE.TargetDrivenGenerator(λ=[l2_λ,λ], opt=opt)
 
@@
 
-                # CCE for given models, λ and generator:
+                # ECCCE for given models, λ and generator:
                 @assert predict_label(M, data, x) != target_label
@@
         _generators = deepcopy(generators)
-        _generators[:cce] = CCEGenerator(temp=temp, λ=[l2_λ,λ], opt=opt)
-        _generators[:energy] = CCE.EnergyDrivenGenerator(λ=[l2_λ,λ], opt=opt)
-        _generators[:target] = CCE.TargetDrivenGenerator(λ=[l2_λ,λ], opt=opt)
+        _generators[:cce] = ECCCEGenerator(temp=temp, λ=[l2_λ,λ], opt=opt)
+        _generators[:energy] = ECCCE.EnergyDrivenGenerator(λ=[l2_λ,λ], opt=opt)
+        _generators[:target] = ECCCE.TargetDrivenGenerator(λ=[l2_λ,λ], opt=opt)
         bmk = benchmark(

### 2023-04-20 src/ECCCE.jl

@@ -1,4 +1,4 @@
-module CCE
+module ECCCE
 
 using CounterfactualExplanations
 import MLJModelInterface as MMI
@@ -9,6 +9,6 @@ include("losses.jl")
 include("generator.jl")
 include("sampling.jl")
 
-export CCEGenerator, EnergySampler, set_size_penalty, distance_from_energy
+export ECCCEGenerator, EnergySampler, set_size_penalty, distance_from_energy
 
 end

### 2023-04-20 src/generator.jl

@@ -1,25 +1,43 @@
 using CounterfactualExplanations.Objectives
 
-"Constructor for `CCEGenerator`."
-function CCEGenerator(; λ::Union{AbstractFloat,Vector{<:AbstractFloat}}=[0.1, 1.0], κ::Real=1.0, temp::Real=0.05, kwargs...)
+"Constructor for `ECCCEGenerator`."
+function ECCCEGenerator(; λ::Union{AbstractFloat,Vector{<:AbstractFloat}}=[0.1, 1.0], κ::Real=1.0, temp::Real=0.05, kwargs...)
     function _set_size_penalty(ce::AbstractCounterfactualExplanation)
-        return CCE.set_size_penalty(ce; κ=κ, temp=temp)
+        return ECCCE.set_size_penalty(ce; κ=κ, temp=temp)
     end
     _penalties = [Objectives.distance_l2, _set_size_penalty]
     λ = λ isa AbstractFloat ? [0.0, λ] : λ
     return Generator(; penalty=_penalties, λ=λ, kwargs...)
 end
 
+"Constructor for `ECECCCEGenerator`: Energy Constrained Conformal Counterfactual Explanation Generator."
+function ECECCCEGenerator(; 
+    λ::Union{AbstractFloat,Vector{<:AbstractFloat}}=[0.1, 1.0, 1.0], 
+    κ::Real=1.0, 
+    temp::Real=0.5, 
+    η::Union{Nothing,Real}=nothing,
+    n::Union{Nothing,Int}=nothing,
+    opt::Flux.Optimise.AbstractOptimiser=CounterfactualExplanations.Generators.JSMADescent(η=η,n=n),
+    kwargs...
+)
+    function _set_size_penalty(ce::AbstractCounterfactualExplanation)
+        return ECCCE.set_size_penalty(ce; κ=κ, temp=temp)
+    end
+    _penalties = [Objectives.distance_l2, _set_size_penalty, ECCCE.distance_from_energy]
+    λ = λ isa AbstractFloat ? [0.0, λ, λ] : λ
+    return Generator(; penalty=_penalties, λ=λ, opt=opt, kwargs...)
+end
+
 "Constructor for `EnergyDrivenGenerator`."
 function EnergyDrivenGenerator(; λ::Union{AbstractFloat,Vector{<:AbstractFloat}}=[0.1, 1.0], kwargs...)
-    _penalties = [Objectives.distance_l2, CCE.distance_from_energy]
[…]

### 2023-04-20 notebooks/intro.qmd
uh

@@
 
-M = CCE.ConformalModel(conf_model, mach.fitresult)
+M = ECCCE.ConformalModel(conf_model, mach.fitresult)
 
@@
 for (i,target) ∈ enumerate(counterfactual_data.y_levels)
-    sampler = CCE.EnergySampler(M, counterfactual_data, target; niter=niter, nsamples=100)
+    sampler = ECCCE.EnergySampler(M, counterfactual_data, target; niter=niter, nsamples=100)
     Xgen = rand(sampler, nsamples)
@@
 nsamples = 100
-M = CCE.ConformalModel(conf_model, mach.fitresult; likelihood=:classification_multi)
+M = ECCCE.ConformalModel(conf_model, mach.fitresult; likelihood=:classification_multi)
 
@@
 for target ∈ counterfactual_data.y_levels
-    sampler = CCE.EnergySampler(M, counterfactual_data, target; niter=niter, nsamples=100)
+    sampler = ECCCE.EnergySampler(M, counterfactual_data, target; niter=niter, nsamples=100)
     Xgen = rand(sampler, nsamples)
@@
 models = Dict(
-    :cov75 => CCE.ConformalModel(conformal_model(clf; method=:simple_inductive, coverage=0.75)),
-    :cov80 => CCE.ConformalModel(conformal_model(clf; method=:simple_inductive, coverage=0.80)),
-    :cov90 => CCE.ConformalModel(conformal_model(clf; method=:simple_inductive, coverage=0.90)),
-    :cov99 => CCE.ConformalModel(conformal_model(clf; method=:simple_inductive, coverage=0.99)),
+    :cov75 => ECCCE.ConformalModel(conformal_model(clf; method=:simple_inductive, coverage=0.75)),
+    :cov80 => ECCCE.ConformalModel(conformal_model(clf; method=:simple_inductive, coverage=0.80)),
+    :cov90 => ECCCE.ConformalModel(conformal_model(clf; method=:simple_inductive, coverage=0.90)),
+    :cov99 => ECCCE.ConformalModel(conformal_model(clf; method=:simple_inductive, coverage=0.99)),
 )
@@
     CounterfactualExplanations.distance,
-    CCE.distance_from_energy,
-    CCE.distance_from_targets
+    ECCCE.distance_from_energy,
+    ECCCE.distance_from_targets
 ]

### 2023-04-20 notebooks/mnist.qmd
uh

@@
 ```{julia}
+function pre_process(x; noise::Float32=0.03f0)
+    ϵ = Float32.(randn(size(x)) * noise)
+    x = @.(2 * x - 1) .+ ϵ
+    return x
+end
+```
+
+```{julia}
 # Data:
@@
 X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
+X = pre_process.(X)
 X = table(permutedims(X))
@@
 batch_size = minimum([Int(round(n_obs/10)), 128])
-n_hidden = 200
+n_hidden = 32
 activation = Flux.relu
-builder = MLJFlux.@builder Flux.Chain(
-    Dense(n_in, n_hidden),
-    BatchNorm(n_hidden, activation),
-    Dense(n_hidden, n_hidden),
-    BatchNorm(n_hidden, activation),
-    Dense(n_hidden, n_out),
-)
-# builder = MLJFlux.Short(n_hidden=n_hidden, dropout=0.2, σ=activation)
+# builder = MLJFlux.@builder Flux.Chain(
+#     Dense(n_in, n_hidden, activation),
+#     Dense(n_hidden, n_hidden, activation),
+#     Dense(n_hidden, n_hidden, activation),
+#     # BatchNorm(n_hidden, activation),
+#     # Dense(n_hidden, n_hidden),
+#     # BatchNorm(n_hidden, activation),
+#     Dense(n_hidden, n_out),
+# )
+builder = MLJFlux.Short(n_hidden=n_hidden, dropout=0.2, σ=activation)
 # builder = MLJFlux.MLP(
@@
 # Joint Energy Model:
-𝒟x = Uniform(0,1)
+𝒟x = Uniform(-1,1)
 𝒟y = Categorical(ones(output_dim) ./ output_dim)
@@
 ```{julia}
-cov = .90
+cov = .95
 conf_model = conformal_model(jem; method=:adaptive_inductive, coverage=cov)
@@
 fit!(mach)
-M = CCE.ConformalModel(mach.model, mach.fitresult)
+M = ECCCE.ConformalModel(mach.model, mach.fitresult)
 ```
@@
 
-# CCE:
+# ECCCE:
 λ=[0.0,1.0]
@@
 
-# Generate counterfactual using CCE generator:
-generator = CCEGenerator(
+# Generate counterfactual using ECCCE generator:
+generator = ECCCEGenerator(
     λ=λ, 
@@
 ces = [ce_wachter, ce_conformal, ce_jsma, ce_conformal_jsma]
-_names = ["Wachter", "CCE", "JSMA", "CCE-JSMA"]
+_names = ["Wachter", "ECCCE", "JSMA", "ECCCE-JSMA"]
 for x in zip(ces, _names)

### 2023-04-20 paper/paper.tex
title
uh

@@
 
-\title{Conformal Counterfactual Explanations}
+\title{Plausibility isn't all you need: Conformal Counterfactual Explanations}
 
@@
 This makes the smooth penalty defined in Equation~\ref{eq:setsize} useful in the context of meeting our objective of generating plausible counterfactuals.
-In particular, we adapt Equation~\ref{eq:general} to define the baseline objective for Conformal Counterfactual Explanations (CCE):
+In particular, we adapt Equation~\ref{eq:general} to define the baseline objective for Conformal Counterfactual Explanations (ECCCE):
 
@@
 Arguably, data is typically abundant and in most applications practitioners tend to hold out a test data set anyway.
-Our proposed approach for CCE therefore removes the restriction on the family of predictive models, at the small cost of reserving a subset of the available data for calibration.
+Our proposed approach for ECCCE therefore removes the restriction on the family of predictive models, at the small cost of reserving a subset of the available data for calibration.
 

### 2023-04-20 src/penalties.jl

@@ -42,7 +42,7 @@ function distance_from_energy(
     ignore_derivatives() do
         _dict = ce.params
         if !(:energy_sampler ∈ collect(keys(_dict)))
-            _dict[:energy_sampler] = CCE.EnergySampler(ce; kwargs...)
+            _dict[:energy_sampler] = ECCCE.EnergySampler(ce; kwargs...)
         end
         sampler = _dict[:energy_sampler]
         push!(conditional_samples, rand(sampler, n; from_buffer=from_buffer))

### 2023-04-21 notebooks/synthetic.qmd
ECCCo

--- 2023-04-05 notebooks/synthetic.qmd
@@
         fit!(mach)
-        M = CCE.ConformalModel(mach.model, mach.fitresult)
+        M = ECCCo.ConformalModel(mach.model, mach.fitresult)
 
-        # Set up CCE:
+        # Set up ECCCo:
         factual_label = predict_label(M, data, x)[1]
@@
 
-            # CCE for given classifier, coverage, temperature and λ:
-            generator = CCEGenerator(temp=temp, λ=[l2_λ,λ], opt=opt)
+            # ECCCo for given classifier, coverage, temperature and λ:
+            generator = ECCCoGenerator(temp=temp, λ=[l2_λ,λ], opt=opt)
             @assert predict_label(M, data, x) != target_label
@@
 # Untrained Models:
-models = Dict(Symbol("cov$(Int(100*cov))") => CCE.ConformalModel(conformal_model(mlp; method=:simple_inductive, coverage=cov)) for cov in cvgs)
+models = Dict(Symbol("cov$(Int(100*cov))") => ECCCo.ConformalModel(conformal_model(mlp; method=:simple_inductive, coverage=cov)) for cov in cvgs)
 
@@
     CounterfactualExplanations.distance,
-    CCE.distance_from_energy,
-    CCE.distance_from_targets,
+    ECCCo.distance_from_energy,
+    ECCCo.distance_from_targets,
     CounterfactualExplanations.validity,
@@
         M = train(M, data)
-        # Set up CCE:
+        # Set up ECCCo:
         factual_label = predict_label(M, data, x)[1]
@@
             _generators = deepcopy(generators)
-            _generators[:cce] = CCEGenerator(temp=_temp, λ=[l2_λ,λ], opt=opt)
-            _generators[:energy] = CCE.EnergyDrivenGenerator(λ=[l2_λ,λ], opt=opt)
-            _generators[:target] = CCE.TargetDrivenGenerator(λ=[l2_λ,λ], opt=opt)
+            _generators[:cce] = ECCCoGenerator(temp=_temp, λ=[l2_λ,λ], opt=opt)
+            _generators[:energy] = ECCCo.EnergyDrivenGenerator(λ=[l2_λ,λ], opt=opt)
+            _generators[:target] = ECCCo.TargetDrivenGenerator(λ=[l2_λ,λ], opt=opt)
 
@@
 
-                # CCE for given models, λ and generator:
+                # ECCCo for given models, λ and generator:
                 @assert predict_label(M, data, x) != target_label
@@
         _generators = deepcopy(generators)
-        _generators[:cce] = CCEGenerator(temp=temp, λ=[l2_λ,λ], opt=opt)
-        _generators[:energy] = CCE.EnergyDrivenGenerator(λ=[l2_λ,λ], opt=opt)
-        _generators[:target] = CCE.TargetDrivenGenerator(λ=[l2_λ,λ], opt=opt)
+        _generators[:cce] = ECCCoGenerator(temp=temp, λ=[l2_λ,λ], opt=opt)
+        _generators[:energy] = ECCCo.EnergyDrivenGenerator(λ=[l2_λ,λ], opt=opt)
+        _generators[:target] = ECCCo.TargetDrivenGenerator(λ=[l2_λ,λ], opt=opt)
         bmk = benchmark(

### 2023-04-21 src/penalties.jl

@@ -42,7 +42,7 @@ function distance_from_energy(
     ignore_derivatives() do
         _dict = ce.params
         if !(:energy_sampler ∈ collect(keys(_dict)))
-            _dict[:energy_sampler] = ECCCE.EnergySampler(ce; kwargs...)
+            _dict[:energy_sampler] = ECCCo.EnergySampler(ce; kwargs...)
         end
         sampler = _dict[:energy_sampler]
         push!(conditional_samples, rand(sampler, n; from_buffer=from_buffer))

### 2023-04-21 test/runtests.jl

@@ -1,6 +1,6 @@
-using ECCCE
+using ECCCo
 using Test
 
-@testset "ECCCE.jl" begin
+@testset "ECCCo.jl" begin
     # Write your tests here.
 end

### 2023-04-21 notebooks/intro.qmd
ECCCo

--- 2023-04-04 notebooks/intro.qmd
@@
 
-M = CCE.ConformalModel(conf_model, mach.fitresult)
+M = ECCCo.ConformalModel(conf_model, mach.fitresult)
 
@@
 for (i,target) ∈ enumerate(counterfactual_data.y_levels)
-    sampler = CCE.EnergySampler(M, counterfactual_data, target; niter=niter, nsamples=100)
+    sampler = ECCCo.EnergySampler(M, counterfactual_data, target; niter=niter, nsamples=100)
     Xgen = rand(sampler, nsamples)
@@
 nsamples = 100
-M = CCE.ConformalModel(conf_model, mach.fitresult; likelihood=:classification_multi)
+M = ECCCo.ConformalModel(conf_model, mach.fitresult; likelihood=:classification_multi)
 
@@
 for target ∈ counterfactual_data.y_levels
-    sampler = CCE.EnergySampler(M, counterfactual_data, target; niter=niter, nsamples=100)
+    sampler = ECCCo.EnergySampler(M, counterfactual_data, target; niter=niter, nsamples=100)
     Xgen = rand(sampler, nsamples)
@@
 models = Dict(
-    :cov75 => CCE.ConformalModel(conformal_model(clf; method=:simple_inductive, coverage=0.75)),
-    :cov80 => CCE.ConformalModel(conformal_model(clf; method=:simple_inductive, coverage=0.80)),
-    :cov90 => CCE.ConformalModel(conformal_model(clf; method=:simple_inductive, coverage=0.90)),
-    :cov99 => CCE.ConformalModel(conformal_model(clf; method=:simple_inductive, coverage=0.99)),
+    :cov75 => ECCCo.ConformalModel(conformal_model(clf; method=:simple_inductive, coverage=0.75)),
+    :cov80 => ECCCo.ConformalModel(conformal_model(clf; method=:simple_inductive, coverage=0.80)),
+    :cov90 => ECCCo.ConformalModel(conformal_model(clf; method=:simple_inductive, coverage=0.90)),
+    :cov99 => ECCCo.ConformalModel(conformal_model(clf; method=:simple_inductive, coverage=0.99)),
 )
@@
     CounterfactualExplanations.distance,
-    CCE.distance_from_energy,
-    CCE.distance_from_targets
+    ECCCo.distance_from_energy,
+    ECCCo.distance_from_targets
 ]

### 2023-04-21 notebooks/mnist.qmd
ECCCo

@@
 n_hidden = 32
-activation = Flux.relu
+activation = Flux.swish
 # builder = MLJFlux.@builder Flux.Chain(
@@
 # )
-builder = MLJFlux.Short(n_hidden=n_hidden, dropout=0.2, σ=activation)
+builder = MLJFlux.Short(n_hidden=n_hidden, dropout=0.1, σ=activation)
 # builder = MLJFlux.MLP(
@@
 # )
-α = [1.0,1.0,1e-2]
+α = [1.0,1.0,1e-1]
 
@@
 fit!(mach)
-M = ECCCE.ConformalModel(mach.model, mach.fitresult)
+M = ECCCo.ConformalModel(mach.model, mach.fitresult)
 ```
@@
     jem = mach.model.model.jem
-    n_iter = 100
+    n_iter = 500
     _w = 1500
@@
 
-# ECCCE:
+# ECCCo:
 λ=[0.0,1.0]
@@
 
-# Generate counterfactual using ECCCE generator:
-generator = ECCCEGenerator(
+# Generate counterfactual using ECCCo generator:
+generator = CCEGenerator(
     λ=λ, 
@@
 ces = [ce_wachter, ce_conformal, ce_jsma, ce_conformal_jsma]
-_names = ["Wachter", "ECCCE", "JSMA", "ECCCE-JSMA"]
+_names = ["Wachter", "ECCCo", "JSMA", "ECCCo-JSMA"]
+for x in zip(ces, _names)
+    ce, _name = (x[1],x[2])
+    x = CounterfactualExplanations.counterfactual(ce)
+    _phat = target_probs(ce)
+    _title = "$_name (p̂=$(round(_phat[1]; digits=3)))"
+    plt = Plots.plot(
+        convert2image(MNIST, reshape(x,28,28)),
+        axis=nothing, 
+        size=(img_height, img_height),
+        title=_title
+    )
+    plts = [plts..., plt]
+end
+plt = Plots.plot(plts...; size=(img_height*length(plts),img_height), layout=(1,length(plts)))
+display(plt)
+savefig(plt, joinpath(www_path, "cce_mnist.png"))
+```
+
+```{julia}
+# Random.seed!(1234)
+
+# Set up search:
+factual_label = 8
+x = reshape(counterfactual_data.X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
+target = 3
+factual = predict_label(M, counterfactual_data, x)[1]
+γ = 0.5
+T = 100
+
+# Generate counterfactual using generic generator:
+generator = GenericGenerator(opt=Flux.Optimise.Adam(),)
+ce_wachter = generate_counterfactual(
+    x, target, counterfactual_data, M, generator; 
+    decision_threshold=γ, max_iter=T,
+    initialization=:identity,
+)
+
+generator = GreedyGenerator(η=1.0)
+ce_jsma = generate_counterfactual(
+    x, target, counterfactual_data, M, generator; 
+    decision_threshold=γ, max_iter=T,
+    initialization=:identity,
+)
+
+# ECCCo:
+λ=[0.0,1.0,1.0]
+temp=0.01
+
+# Generate counterfactual using ECCCo generator:
+generator = ECCCoGenerator(
+    λ=λ, 
+    temp=temp, 
+    opt=Flux.Optimise.Adam(),
+)
+ce_conformal = generate_counterfactual(
+    x, target, counterfactual_data, M, generator; 
+    decision_threshold=γ, max_iter=T,
+    initialization=:identity,
+    converge_when=:generator_conditions,
+)
+
+# Generate counterfactual using ECCCo generator:
+generator = ECCCoGenerator(
+    λ=λ, 
+    temp=temp, 
+    opt=CounterfactualExplanations.Generators.JSMADescent(η=1.0),
+)
+ce_conformal_jsma = generate_counterfactual(
+    x, target, counterfactual_data, M, generator; 
+    decision_threshold=γ, max_iter=T,
+    initialization=:identity,
+    converge_when=:generator_conditions,
+)
+
+# Plot:
+p1 = Plots.plot(
+    convert2image(MNIST, reshape(x,28,28)),
+    axis=nothing, 
+    size=(img_height, img_height),
+    title="Factual"
+)
+plts = [p1]
+
+ces = [ce_wachter, ce_conformal, ce_jsma, ce_conformal_jsma]
+_names = ["Wachter", "ECCCo", "JSMA", "ECCCo-JSMA"]
 for x in zip(ces, _names)
@@
     CounterfactualExplanations.distance,
-    ECCCE.distance_from_energy,
-    ECCCE.distance_from_targets,
+    ECCCo.distance_from_energy,
+    ECCCo.distance_from_targets,
     CounterfactualExplanations.validity,

### 2023-04-21 notebooks/proposal.qmd
ECCCo

--- 2023-03-22 notebooks/proposal.qmd
@@
 In this section, we introduce an alternative way to generate high-fidelity Counterfactual Explanations.
-In particular, we propose Conformal Counterfactual Explanations (CCE), that is Counterfactual Explanations that minimize the predictive uncertainty of conformal models.
+In particular, we propose Conformal Counterfactual Explanations (ECCCo), that is Counterfactual Explanations that minimize the predictive uncertainty of conformal models.
 

### 2023-04-21 paper/paper.tex
ECCCo

--- 2023-03-29 paper/paper.tex
@@
 
-\title{Conformal Counterfactual Explanations}
+\title{ECCCos from the Black Box: Letting Models speak for Themselves}
 
@@
 This makes the smooth penalty defined in Equation~\ref{eq:setsize} useful in the context of meeting our objective of generating plausible counterfactuals.
-In particular, we adapt Equation~\ref{eq:general} to define the baseline objective for Conformal Counterfactual Explanations (CCE):
+In particular, we adapt Equation~\ref{eq:general} to define the baseline objective for Conformal Counterfactual Explanations (ECCCo):
 
@@
 Arguably, data is typically abundant and in most applications practitioners tend to hold out a test data set anyway.
-Our proposed approach for CCE therefore removes the restriction on the family of predictive models, at the small cost of reserving a subset of the available data for calibration.
+Our proposed approach for ECCCo therefore removes the restriction on the family of predictive models, at the small cost of reserving a subset of the available data for calibration.
 

### 2023-04-21 src/generator.jl

@@ -1,17 +1,17 @@
 using CounterfactualExplanations.Objectives
 
-"Constructor for `ECCCEGenerator`."
-function ECCCEGenerator(; λ::Union{AbstractFloat,Vector{<:AbstractFloat}}=[0.1, 1.0], κ::Real=1.0, temp::Real=0.05, kwargs...)
+"Constructor for `ECCCoGenerator`."
+function ECCCoGenerator(; λ::Union{AbstractFloat,Vector{<:AbstractFloat}}=[0.1, 1.0], κ::Real=1.0, temp::Real=0.05, kwargs...)
     function _set_size_penalty(ce::AbstractCounterfactualExplanation)
-        return ECCCE.set_size_penalty(ce; κ=κ, temp=temp)
+        return ECCCo.set_size_penalty(ce; κ=κ, temp=temp)
     end
     _penalties = [Objectives.distance_l2, _set_size_penalty]
     λ = λ isa AbstractFloat ? [0.0, λ] : λ
     return Generator(; penalty=_penalties, λ=λ, kwargs...)
 end
 
-"Constructor for `ECECCCEGenerator`: Energy Constrained Conformal Counterfactual Explanation Generator."
-function ECECCCEGenerator(; 
+"Constructor for `ECECCCoGenerator`: Energy Constrained Conformal Counterfactual Explanation Generator."
+function ECECCCoGenerator(; 
     λ::Union{AbstractFloat,Vector{<:AbstractFloat}}=[0.1, 1.0, 1.0], 
     κ::Real=1.0, 
     temp::Real=0.5, 
@@ -21,23 +21,23 @@ function ECECCCEGenerator(;
     kwargs...
 )
     function _set_size_penalty(ce::AbstractCounterfactualExplanation)
-        return ECCCE.set_size_penalty(ce; κ=κ, temp=temp)
+        return ECCCo.set_size_penalty(ce; κ=κ, temp=temp)
     end
-    _penalties = [Objectives.distance_l2, _set_size_penalty, ECCCE.distance_from_energy]
+    _penalties = [Objectives.distance_l2, _set_size_penalty, ECCCo.distance_from_energy]
     λ = λ isa AbstractFloat ? [0.0, λ, λ] : λ
     return Generator(; penalty=_penalties, λ=λ, opt=opt, kwargs...)
 end
 
 "Constructor for `EnergyDrivenGenerator`."
[…]

### 2023-04-21 src/penalties.jl
trying to sort out the sampling :sob:

@@ -36,20 +36,19 @@ end
 
 function distance_from_energy(
     ce::AbstractCounterfactualExplanation;
-    n::Int=10000, from_buffer=true, agg=mean, kwargs...
+    n::Int=1, niter=200, from_buffer=true, agg=mean, kwargs...
 )
     conditional_samples = []
     ignore_derivatives() do
         _dict = ce.params
         if !(:energy_sampler ∈ collect(keys(_dict)))
-            _dict[:energy_sampler] = ECCCo.EnergySampler(ce; kwargs...)
+            _dict[:energy_sampler] = ECCCo.EnergySampler(ce; niter=niter, nsamples=1000, kwargs...)
         end
         sampler = _dict[:energy_sampler]
         push!(conditional_samples, rand(sampler, n; from_buffer=from_buffer))
     end
     x′ = CounterfactualExplanations.counterfactual(ce)
-    loss = map(eachslice(x′, dims=3)) do x
-        x = Matrix(x)
+    loss = map(eachslice(x′, dims=ndims(x′))) do x
         Δ = map(eachcol(conditional_samples[1])) do xsample
             norm(x - xsample)
         end
@@ -63,7 +62,7 @@ end
 
 function distance_from_targets(
     ce::AbstractCounterfactualExplanation;
-    n::Int=10000, agg=mean
+    n::Int=1000, agg=mean
 )
     target_idx = ce.data.output_encoder.labels .== ce.target
     target_samples = ce.data.X[:,target_idx] |>

### 2023-04-21 notebooks/mnist.qmd
trying to sort out the sampling :sob:

--- 2023-04-19 notebooks/mnist.qmd
@@
 ```{julia}
+function pre_process(x; noise::Float32=0.03f0)
+    ϵ = Float32.(randn(size(x)) * noise)
+    x = @.(2 * x - 1) .+ ϵ
+    return x
+end
+```
+
+```{julia}
 # Data:
-n_obs = 10000
+n_obs = 1000
 counterfactual_data = load_mnist(n_obs)
+counterfactual_data.X = pre_process.(counterfactual_data.X)
 X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
@@
 batch_size = minimum([Int(round(n_obs/10)), 128])
-n_hidden = 200
-activation = Flux.relu
+n_hidden = 50
+activation = Flux.swish
 builder = MLJFlux.@builder Flux.Chain(
-    Dense(n_in, n_hidden),
-    BatchNorm(n_hidden, activation),
-    Dense(n_hidden, n_hidden),
-    BatchNorm(n_hidden, activation),
+
+    Dense(n_in, n_hidden, activation),
+    # Dense(n_hidden, n_hidden, activation),
+    # Dense(n_hidden, n_hidden, activation),
+
+    # Dense(n_in, n_hidden),
+    # BatchNorm(n_hidden, activation),
+    # Dense(n_hidden, n_hidden),
+    # BatchNorm(n_hidden, activation),
+
     Dense(n_hidden, n_out),
 )
-# builder = MLJFlux.Short(n_hidden=n_hidden, dropout=0.2, σ=activation)
+# builder = MLJFlux.Short(n_hidden=n_hidden, dropout=0.1, σ=activation)
 # builder = MLJFlux.MLP(
@@
 # Deep Ensemble:
-mlp_ens = EnsembleModel(model=mlp, n=5)
+mlp_ens = EnsembleModel(model=mlp, n=50)
 ```
@@
 fit!(mach)
-M = CCE.ConformalModel(mach.model, mach.fitresult)
+M = ECCCo.ConformalModel(mach.model, mach.fitresult)
 ```
@@
     jem = mach.model.model.jem
-    n_iter = 100
+    n_iter = 200
     _w = 1500
@@
 test_data = load_mnist_test()
+test_data.X = pre_process.(test_data.X)
 f1 = CounterfactualExplanations.Models.model_evaluation(M, test_data)
@@
 
-# CCE:
+# ECCCo:
 λ=[0.0,1.0]
-temp=0.01
+temp=0.5
 
@@
 
+```{julia}
+# Random.seed!(1234)
+
+# Set up search:
+factual_label = 8
+x = reshape(counterfactual_data.X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
+target = 3
+factual = predict_label(M, counterfactual_data, x)[1]
+γ = 0.5
+T = 100
+
+η=0.1
+
+# Generate counterfactual using generic generator:
+generator = GenericGenerator(opt=Flux.Optimise.Adam(),)
+ce_wachter = generate_counterfactual(
+    x, target, counterfactual_data, M, generator; 
+    decision_threshold=γ, max_iter=T,
+    initialization=:identity,
+)
+
+generator = GreedyGenerator(η=η)
+ce_jsma = generate_counterfactual(
+    x, target, counterfactual_data, M, generator; 
+    decision_threshold=γ, max_iter=T,
+    initialization=:identity,
+)
+
+# ECCCo:
+λ=[0.0,0.0,10.0]
+temp=0.5
+
+# Generate counterfactual using ECCCo generator:
+generator = ECCCoGenerator(
+    λ=λ, 
+    temp=temp, 
+    opt=Flux.Optimise.Adam(),
+)
+ce_conformal = generate_counterfactual(
+    x, target, counterfactual_data, M, generator; 
+    decision_threshold=γ, max_iter=T,
+    initialization=:identity,
+    converge_when=:generator_conditions,
+)
+
+# Generate counterfactual using ECCCo generator:
+generator = ECCCoGenerator(
+    λ=λ, 
+    temp=temp, 
+    opt=CounterfactualExplanations.Generators.JSMADescent(η=η),
+)
+ce_conformal_jsma = generate_counterfactual(
+    x, target, counterfactual_data, M, generator; 
+    decision_threshold=γ, max_iter=T,
+    initialization=:identity,
+    converge_when=:generator_conditions,
+)
+
+# Plot:
+p1 = Plots.plot(
+    convert2image(MNIST, reshape(x,28,28)),
+    axis=nothing, 
+    size=(img_height, img_height),
+    title="Factual"
+)
+plts = [p1]
+
+ces = [ce_wachter, ce_conformal, ce_jsma, ce_conformal_jsma]
+_names = ["Wachter", "ECCCo", "JSMA", "ECCCo-JSMA"]
+for x in zip(ces, _names)
+    ce, _name = (x[1],x[2])
+    x = CounterfactualExplanations.counterfactual(ce)
+    _phat = target_probs(ce)
+    _title = "$_name (p̂=$(round(_phat[1]; digits=3)))"
+    plt = Plots.plot(
+        convert2image(MNIST, reshape(x,28,28)),
+        axis=nothing, 
+        size=(img_height, img_height),
+        title=_title
+    )
+    plts = [plts..., plt]
+end
+plt = Plots.plot(plts...; size=(img_height*length(plts),img_height), layout=(1,length(plts)))
+display(plt)
+savefig(plt, joinpath(www_path, "eccco_mnist.png"))
+```
+
+```{julia}
+if M.model.model isa JointEnergyModels.JointEnergyClassifier
+    jem = M.model.model.jem
+    n_iter = 200
+    _w = 1500
+    plts = []
+    neach = 10
+    for i in 1:10
+        x = jem.sampler(jem.chain, jem.sampling_rule; niter=n_iter, n_samples=neach, y=i)
+        plts_i = []
+        for j in 1:size(x, 2)
+            xj = x[:,j]
+            xj = reshape(xj, (n_digits, n_digits))
+            plts_i = [plts_i..., Plots.heatmap(rotl90(xj), axis=nothing, cb=false)]
+        end
+        plt = Plots.plot(plts_i..., size=(_w,0.10*_w), layout=(1,10))
+        plts = [plts..., plt]
+    end
+    plt = Plots.plot(plts..., size=(_w,_w), layout=(10,1))
+    display(plt)
+end
+```
+
 ## Benchmark

### 2023-04-21 src/ECCCo.jl

@@ -9,6 +9,7 @@ include("losses.jl")
 include("generator.jl")
 include("sampling.jl")
 
-export ECCCoGenerator, EnergySampler, set_size_penalty, distance_from_energy
+export CCEGenerator, ECCCoGenerator, EnergySampler
+export set_size_penalty, distance_from_energy
 
 end

### 2023-04-22 notebooks/mnist.qmd
seem to finally be getting somewhere

@@
 cov = .95
-conf_model = conformal_model(jem; method=:adaptive_inductive, coverage=cov)
+conf_model = conformal_model(jem; method=:simple_inductive, coverage=cov)
 mach = machine(conf_model, X, labels)
@@
 println("F1 score (test): $(round(f1,digits=3))")
+```
+
+```{julia}
+𝒟x = Uniform(-1,1)
+𝒟y = Categorical(ones(K) ./ K)
+sampler = UnconditionalSampler(𝒟x; input_size=(D,))
+conditional_sampler = ConditionalSampler(𝒟x, 𝒟y; input_size=(D,))
+opt = ImproperSGLD()
+n_iter = 100
+```
+
+```{julia}
+# Random.seed!(1234)
+
+# Set up search:
+factual_label = 8
+x = reshape(counterfactual_data.X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
+target = 3
+factual = predict_label(M, counterfactual_data, x)[1]
+γ = 0.5
+T = 100
+
+η=1.0
+
+# Generate counterfactual using generic generator:
+generator = GenericGenerator()
+ce_wachter = generate_counterfactual(
+    x, target, counterfactual_data, M, generator; 
+    decision_threshold=γ, max_iter=T,
+    initialization=:identity,
+)
+
+generator = GreedyGenerator(η=η)
+ce_jsma = generate_counterfactual(
+    x, target, counterfactual_data, M, generator; 
+    decision_threshold=γ, max_iter=T,
+    initialization=:identity,
+)
+
+# ECCCo:
+λ=[0.0,1.0,10.0]
+temp=0.01
+
+# Generate counterfactual using ECCCo generator:
+generator = ECCCoGenerator(
+    λ=λ, 
+    temp=temp, 
+)
+ce_conformal = generate_counterfactual(
+    x, target, counterfactual_data, M, generator; 
+    decision_threshold=γ, max_iter=T,
+    initialization=:identity,
+    converge_when=:generator_conditions,
+)
+
+# Generate counterfactual using ECCCo generator:
+generator = ECCCoGenerator(
+    λ=λ, 
+    temp=temp, 
+    opt=CounterfactualExplanations.Generators.JSMADescent(η=η),
+)
+ce_conformal_jsma = generate_counterfactual(
+    x, target, counterfactual_data, M, generator; 
+    decision_threshold=γ, max_iter=T,
+    initialization=:identity,
+    converge_when=:generator_conditions,
+)
+
+# Plot:
+p1 = Plots.plot(
+    convert2image(MNIST, reshape(x,28,28)),
+    axis=nothing, 
+    size=(img_height, img_height),
+    title="Factual"
+)
+plts = [p1]
+
+ces = [ce_wachter, ce_conformal, ce_jsma, ce_conformal_jsma]
+_names = ["Wachter", "ECCCo", "JSMA", "ECCCo-JSMA"]
+for x in zip(ces, _names)
+    ce, _name = (x[1],x[2])
+    x = CounterfactualExplanations.counterfactual(ce)
+    _phat = target_probs(ce)
+    _title = "$_name (p̂=$(round(_phat[1]; digits=3)))"
+    plt = Plots.plot(
+        convert2image(MNIST, reshape(x,28,28)),
+        axis=nothing, 
+        size=(img_height, img_height),
+        title=_title
+    )
+    plts = [plts..., plt]
+end
+plt = Plots.plot(plts...; size=(img_height*length(plts),img_height), layout=(1,length(plts)))
+display(plt)
+savefig(plt, joinpath(www_path, "eccco_mnist.png"))
 ```
@@
 ```{julia}
-# Random.seed!(1234)
-
-# Set up search:
-factual_label = 8
-x = reshape(counterfactual_data.X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
-target = 3
-factual = predict_label(M, counterfactual_data, x)[1]
-γ = 0.5
-T = 100
-
-η=0.1
-
-# Generate counterfactual using generic generator:
-generator = GenericGenerator(opt=Flux.Optimise.Adam(),)
-ce_wachter = generate_counterfactual(
-    x, target, counterfactual_data, M, generator; 
-    decision_threshold=γ, max_iter=T,
-    initialization=:identity,
-)
-
-generator = GreedyGenerator(η=η)
-ce_jsma = generate_counterfactual(
-    x, target, counterfactual_data, M, generator; 
-    decision_threshold=γ, max_iter=T,
-    initialization=:identity,
-)
-
-# ECCCo:
-λ=[0.0,0.0,10.0]
-temp=0.5
-
-# Generate counterfactual using ECCCo generator:
-generator = ECCCoGenerator(
-    λ=λ, 
-    temp=temp, 
-    opt=Flux.Optimise.Adam(),
-)
-ce_conformal = generate_counterfactual(
-    x, target, counterfactual_data, M, generator; 
-    decision_threshold=γ, max_iter=T,
-    initialization=:identity,
-    converge_when=:generator_conditions,
-)
-
-# Generate counterfactual using ECCCo generator:
-generator = ECCCoGenerator(
-    λ=λ, 
-    temp=temp, 
-    opt=CounterfactualExplanations.Generators.JSMADescent(η=η),
-)
-ce_conformal_jsma = generate_counterfactual(
-    x, target, counterfactual_data, M, generator; 
-    decision_threshold=γ, max_iter=T,
-    initialization=:identity,
-    converge_when=:generator_conditions,
-)
-
-# Plot:
-p1 = Plots.plot(
-    convert2image(MNIST, reshape(x,28,28)),
-    axis=nothing, 
-    size=(img_height, img_height),
-    title="Factual"
-)
-plts = [p1]
-
-ces = [ce_wachter, ce_conformal, ce_jsma, ce_conformal_jsma]
-_names = ["Wachter", "ECCCo", "JSMA", "ECCCo-JSMA"]
-for x in zip(ces, _names)
-    ce, _name = (x[1],x[2])
-    x = CounterfactualExplanations.counterfactual(ce)
-    _phat = target_probs(ce)
-    _title = "$_name (p̂=$(round(_phat[1]; digits=3)))"
-    plt = Plots.plot(
-        convert2image(MNIST, reshape(x,28,28)),
-        axis=nothing, 
-        size=(img_height, img_height),
-        title=_title
-    )
-    plts = [plts..., plt]
-end
-plt = Plots.plot(plts...; size=(img_height*length(plts),img_height), layout=(1,length(plts)))
-display(plt)
-savefig(plt, joinpath(www_path, "eccco_mnist.png"))
-```
-
-```{julia}
 if M.model.model isa JointEnergyModels.JointEnergyClassifier

### 2023-04-22 src/sampling.jl
seem to finally be getting somewhere

@@ -57,10 +57,8 @@ function EnergySampler(
     # Initiate:
     energy_sampler = EnergySampler(model, data, sampler, opt, nothing, nothing)
 
-    # Generate samples:
-    chain = model.model.model.jem.chain
-    rule = model.model.model.jem.sampling_rule
-    energy_sampler.sampler(chain, rule; niter=niter, n_samples=nsamples, y=yidx)
+    # Generate conditional samples:
+    generate_samples!(energy_sampler, nsamples, yidx; niter=niter)
 
     return energy_sampler
 end
@@ -87,18 +85,42 @@ function EnergySampler(
 end
 
 """
+    generate_samples(e::EnergySampler, n::Int, y::Int; niter::Int=100)
+
+Generates `n` samples from `EnergySampler` for conditioning value `y`.
+"""
+function generate_samples(e::EnergySampler, n::Int, y::Int; niter::Int=100)
+
+    # Generate samples:
+    chain = e.model.fitresult[1]
+    rule = e.opt
+    xsamples = e.sampler(chain, rule; niter=niter, n_samples=n, y=y)
+
+    return xsamples
+end
+
+"""
+    generate_samples!(e::EnergySampler, n::Int, y::Int; niter::Int=100)
+
+Generates `n` samples from `EnergySampler` for conditioning value `y`. Assigns samples and conditioning value to `EnergySampler`.
+"""
+function generate_samples!(e::EnergySampler, n::Int, y::Int; niter::Int=100)
+    e.buffer = generate_samples(e, n, y; niter=niter)
+    e.yidx = y
+end
+
+"""
     Base.rand(sampler::EnergySampler, n::Int=100; retrain=false)
 
 Overloads the `rand` method to randomly draw `n` samples from `EnergySampler`.
 """
 function Base.rand(sampler::EnergySampler, n::Int=100; from_buffer=true, niter::Int=100)
-    ntotal = size(sampler.sampler.buffer)[end]
+    ntotal = size(sampler.buffer, 2)
     idx = rand(1:ntotal, n)
     if from_buffer
-        X = sampler.sampler.buffer[:, idx]
+        X = sampler.buffer[:, idx]
     else
-        chain = sampler.model.fitresult[1]
-        X = sampler.sampler(chain, sampler.opt; niter=niter, n_samples=n, y=sampler.yidx)
+        X = generate_samples(sampler, n, sampler.yidx; niter=niter)
     end
     return X
 end

### 2023-04-22 notebooks/mnist.qmd
getting somewhere

@@
 batch_size = minimum([Int(round(n_obs/10)), 128])
-n_hidden = 50
+n_hidden = 128
 activation = Flux.swish
@@
     input_size=(input_dim,), 
-    batch_size=10
+    batch_size=1
 )
@@
 ```{julia}
-𝒟x = Uniform(-1,1)
-𝒟y = Categorical(ones(K) ./ K)
-sampler = UnconditionalSampler(𝒟x; input_size=(D,))
-conditional_sampler = ConditionalSampler(𝒟x, 𝒟y; input_size=(D,))
-opt = ImproperSGLD()
-n_iter = 100
-```
-
-```{julia}
-# Random.seed!(1234)
+Random.seed!(1234)
 
 # Set up search:
-factual_label = 8
+factual_label = 9
 x = reshape(counterfactual_data.X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
-target = 3
+target = 4
 factual = predict_label(M, counterfactual_data, x)[1]
 γ = 0.5
-T = 100
+T = 250
 
@@
 # ECCCo:
-λ=[0.0,1.0,10.0]
-temp=0.01
+λ=[0.1,0.1,0.1]
+temp=0.1
 

### 2023-04-23 notebooks/mnist.qmd
working my boi :sunglasses:

@@
 n_hidden = 128
-activation = Flux.swish
+activation = Flux.relu
 builder = MLJFlux.@builder Flux.Chain(
@@
 # )
-α = [1.0,1.0,1e-2]
+α = [1.0,1.0,5e-3]
 
@@
     batch_size=batch_size,
+    finaliser=x -> x,
+    loss=Flux.Losses.logitcrossentropy,
 )
+
+# Deep Ensemble:
+mlp_ens = EnsembleModel(model=mlp, n=5)
 
@@
 
-# Deep Ensemble:
-mlp_ens = EnsembleModel(model=mlp, n=50)
+# Deep Ensemble of Joint Energy Models:
+jem_ens = EnsembleModel(model=jem, n=5)
 ```
@@
 cov = .95
-conf_model = conformal_model(jem; method=:simple_inductive, coverage=cov)
+conf_model = conformal_model(jem_ens; method=:adaptive_inductive, coverage=cov)
 mach = machine(conf_model, X, labels)
@@
 ```{julia}
-if mach.model.model isa JointEnergyModels.JointEnergyClassifier
-    jem = mach.model.model.jem
+if mach.model.model isa JointEnergyClassifier
+    sampler = mach.model.model.jem.sampler
+else
+    K = length(counterfactual_data.y_levels)
+    input_size = size(selectdim(counterfactual_data.X, ndims(counterfactual_data.X), 1))
+    𝒟x = Uniform(extrema(counterfactual_data.X)...)
+    𝒟y = Categorical(ones(K) ./ K)
+    sampler = ConditionalSampler(𝒟x, 𝒟y; input_size=input_size)
+end
+opt = ImproperSGLD()
+f(x) = logits(M, x)
+
 n_iter = 200
@@
 for i in 1:10
-        x = jem.sampler(jem.chain, jem.sampling_rule; niter=n_iter, n_samples=neach, y=i)
+    x = sampler(f, opt; niter=n_iter, n_samples=neach, y=i)
     plts_i = []
@@
 display(plt)
-end
+
 ```
@@
 # Set up search:
-factual_label = 9
+factual_label = 4
 x = reshape(counterfactual_data.X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
-target = 4
+target = 5
 factual = predict_label(M, counterfactual_data, x)[1]
-γ = 0.5
-T = 250
+γ = 0.9
+T = 100
 
@@
     temp=temp, 
+    opt=Flux.Optimise.Adam(0.01),
 )

### 2023-04-23 paper/paper.tex
trying to sort out the sampling :sob:
working my boi :sunglasses:

@@
 
-\title{ECCCos from the Black Box: Letting Models speak for Themselves}
+\title{ECCoEs from the Black Box: Letting Models speak for Themselves}
 
@@
 
-\subsection{Conformal Counterfactual Explanations}
+\subsection{ECCoE: Energy-Constrained Conformal Counterfactual Explanation}
 
@@
 \section{Experiments}
+
+\begin{itemize}
+  \item BatchNorm does not seem compatible with JEM
+  \item Coverage and temperature impacts CCE in somewhat unpredictable ways
+  \item It seems that models that are not explicitly trained for generative task, still learn it implictly
+  \item Batch size seems to impact quality of generated samples
+\end{itemize}
 

### 2023-04-23 src/model.jl

@@ -32,14 +32,46 @@ struct ConformalModel <: Models.AbstractDifferentiableJuliaModel
     end
 end
 
-# Outer constructor method:
+"""
+    _get_chains(fitresult)
+
+Private function that extracts the chains from a fitted model.
+"""
+function _get_chains(fitresult)
+    if fitresult isa MLJEnsembles.WrappedEnsemble
+        chains = map(res -> res[1], fitresult.ensemble)
+    else
+        chains = [fitresult[1]]
+    end
+    return chains
+end
+
+"""
+    _outdim(fitresult)
+
+Private function that extracts the output dimension from a fitted model.
+"""
+function _outdim(fitresult)
+    if fitresult isa MLJEnsembles.WrappedEnsemble
+        outdim = length(fitresult.ensemble[1][2])
+    else
+        outdim = length(fitresult[2])
+    end
+    return outdim
+end
+
+"""
+    ConformalModel(model, fitresult=nothing; likelihood::Union{Nothing,Symbol}=nothing)
+
+Outer constructor for `ConformalModel`. If `fitresult` is not specified, the model is not fitted and `likelihood` is inferred from the model. If `fitresult` is specified, `likelihood` is inferred from the output dimension of the model. If `likelihood` is not specified, it defaults to `:classification_binary`.
+"""
 function ConformalModel(model, fitresult=nothing; likelihood::Union{Nothing,Symbol}=nothing)
 
     # Check if model is fitted and infer likelihood:
     if isnothing(fitresult)
         @info "Conformal Model is not fitted."
     else
-        outdim = length(fitresult[2])
+        outdim = _outdim(fitresult)
         _likelihood = outdim == 2 ? :classification_binary : :classification_multi
         @assert likelihood == _likelihood || isnothing(likelihood) "Specification of `likelihood` does not match the output dimension of the model."
         likelihood = _likelihood
@@ -52,7 +84,7 @@ function ConformalModel(model, fitresult=nothing; likelihood::Union{Nothing,Symb
     end
 
     # Construct model:
-    testmode!(fitresult[1])
+    testmode!.(_get_chains(fitresult))
     M = ConformalModel(model, fitresult, likelihood)
     return M
 end
@@ -82,7 +114,9 @@ which follows from the derivation here: https://stats.stackexchange.com/question
 function Models.logits(M::ConformalModel, X::AbstractArray)
[…]

### 2023-04-23 src/sampling.jl

@@ -92,9 +92,10 @@ Generates `n` samples from `EnergySampler` for conditioning value `y`.
 function generate_samples(e::EnergySampler, n::Int, y::Int; niter::Int=100)
 
     # Generate samples:
-    chain = e.model.fitresult[1]
+    # chain = e.model.fitresult[1]
+    f(x) = logits(e.model, x)
     rule = e.opt
-    xsamples = e.sampler(chain, rule; niter=niter, n_samples=n, y=y)
+    xsamples = e.sampler(f, rule; niter=niter, n_samples=n, y=y)
 
     return xsamples
 end

### 2023-04-24 notebooks/dev_mnist.qmd
MNIST complete run (examples)

--- 2023-04-23 notebooks/mnist.qmd
@@
 # Set up search:
-factual_label = 4
+factual_label = 9
 x = reshape(counterfactual_data.X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
-target = 5
+target = 7
 factual = predict_label(M, counterfactual_data, x)[1]
@@
 # Generate counterfactual using generic generator:
-generator = GenericGenerator()
+generator = GenericGenerator(opt=Flux.Optimise.Adam(0.01),)
 ce_wachter = generate_counterfactual(

### 2023-04-24 paper/paper.tex
MNIST complete run (examples)

@@
 \begin{abstract}
-  We propose Conformal Counterfactual Explanations: an effortless and rigorous way to produce plausible and conformal Counterfactual Explanations for Black Box Models using Conformal Prediction.
+  We propose Eccoe: an effortless and rigorous way to produce plausible and conformal Counterfactual Explanations for Black Box Models using Conformal Prediction.
 To address the need for plausible explanations, existing work has primarily relied on surrogate models to learn the data-generating process.
@@
 
-Penalizing the set size in this way is in principal enough to train efficient conformal classifiers \citep{stutz2022learning}.
+Penalizing the set size in this way is in principle enough to train efficient conformal classifiers \citep{stutz2022learning}.
 As we explained above, the set size is also closely linked to predictive uncertainty at the local level.
 This makes the smooth penalty defined in Equation~\ref{eq:setsize} useful in the context of meeting our objective of generating plausible counterfactuals.
-In particular, we adapt Equation~\ref{eq:general} to define the baseline objective for Conformal Counterfactual Explanations (ECCCo):
+In particular, we adapt Equation~\ref{eq:general} to define the objective for our proposed Energy-Constrained Conformal Counterfactual Explanations (Eccoe):
 
-\begin{equation}\label{eq:cce}
+\begin{equation}\label{eq:eccoe}
   \begin{aligned}

### 2023-04-24 notebooks/mnist.qmd
paper updates

@@
 # MNIST
+
+## Anecdotal Evidence
+
+### Examples in Introduction
+
+#### Wachter and JSMA
+
+```{julia}
+# Data:
+counterfactual_data = load_mnist()
+X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
+input_dim, n_obs = size(counterfactual_data.X)
+M = load_mnist_mlp()
+
+# Target:
+factual_label = 8
+x = reshape(X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
+target = 3
+factual = predict_label(M, counterfactual_data, x)[1]
+γ = 0.9
+T = 50
+```
+
+```{julia}
+# Search:
+opt = Flux.Optimise.Adam(0.01)
+generator = GenericGenerator(opt=opt)
+ce_wachter = generate_counterfactual(
+    x, target, counterfactual_data, M, generator; 
+    decision_threshold=γ, max_iter=T,
+    initialization=:identity,
+)
+generator = GreedyGenerator(η=1.0)
+ce_jsma = generate_counterfactual(
+    x, target, counterfactual_data, M, generator; 
+    decision_threshold=γ, max_iter=T,
+    initialization=:identity,
+)
+```
+
+```{julia}
+p1 = Plots.plot(
+    convert2image(MNIST, reshape(x,28,28)),
+    axis=nothing, 
+    size=(img_height, img_height),
+    title="Factual"
+)
+plts = [p1]
+
+ces = zip([ce_wachter,ce_jsma])
+counterfactuals = reduce((x,y)->cat(x,y,dims=3),map(ce -> CounterfactualExplanations.counterfactual(ce[1]), ces))
+phat = reduce((x,y) -> cat(x,y,dims=3), map(ce -> target_probs(ce[1]), ces))
+for x in zip(eachslice(counterfactuals; dims=3), eachslice(phat; dims=3), ["Wachter","JSMA"])
+    ce, _phat, _name = (x[1],x[2],x[3])
+    _title = "$(_name) (p=$(round(_phat[1]; digits=2)))"
+    plt = Plots.plot(
+        convert2image(MNIST, reshape(ce,28,28)),
+        axis=nothing, 
+        size=(img_height, img_height),
+        title=_title
+    )
+    plts = [plts..., plt]
+end
+plt = Plots.plot(plts...; size=(img_height*length(plts),img_height), layout=(1,length(plts)))
+display(plt)
+savefig(plt, joinpath(www_path, "you_may_not_like_it.png"))
+```
+
+#### REVISE
+
+```{julia}
+using CounterfactualExplanations.Models: load_mnist_vae
+vae = load_mnist_vae()
+vae_weak = load_mnist_vae(;strong=false)
+Serialization.serialize(joinpath(output_path,"mnist_classifier.jls"), M)
+Serialization.serialize(joinpath(output_path,"mnist_vae.jls"), vae)
+Serialization.serialize(joinpath(output_path,"mnist_vae_weak.jls"), vae_weak)
+```
+
+```{julia}
+# Define generator:
+generator = REVISEGenerator(
+  opt = opt,
+  λ=0.01
+)
+# Generate recourse:
+counterfactual_data.generative_model = vae # assign generative model
+ce_strong = generate_counterfactual(
+    x, target, counterfactual_data, M, generator; 
+    decision_threshold=γ, max_iter=T,
+    initialization=:identity,
+)
+counterfactual_data = deepcopy(counterfactual_data)
+counterfactual_data.generative_model = vae_weak
+ce_weak = generate_counterfactual(
+    x, target, counterfactual_data, M, generator;
+    decision_threshold=γ, max_iter=T,
+    initialization=:identity,
+)
+```
+
+```{julia}
+ces = zip([ce_strong,ce_weak])
+counterfactuals = reduce((x,y)->cat(x,y,dims=3),map(ce -> CounterfactualExplanations.counterfactual(ce[1]), ces))
+phat = reduce((x,y) -> cat(x,y,dims=3), map(ce -> target_probs(ce[1]), ces))
+plts = [p1]
+for x in zip(eachslice(counterfactuals; dims=3), eachslice(phat; dims=3), ["Strong VAE","Weak VAE"])
+    ce, _phat, _name = (x[1],x[2],x[3])
+    _title = "$(_name) (p=$(round(_phat[1]; digits=2)))"
+    plt = Plots.plot(
+        convert2image(MNIST, reshape(ce,28,28)),
+        axis=nothing, 
+        size=(img_height, img_height),
+        title=_title
+    )
+    plts = [plts..., plt]
+end
+plt = Plots.plot(plts...; size=(img_height*length(plts),img_height), layout=(1,length(plts)))
+display(plt)
+savefig(plt, joinpath(www_path, "surrogate_gone_wrong.png"))
+```
+
+### ECCo
 
@@
 # Data:
-n_obs = 1000
-counterfactual_data = load_mnist(n_obs)
 counterfactual_data.X = pre_process.(counterfactual_data.X)
@@
 builder = MLJFlux.@builder Flux.Chain(
-
     Dense(n_in, n_hidden, activation),
-    # Dense(n_hidden, n_hidden, activation),
-    # Dense(n_hidden, n_hidden, activation),
-
-    # Dense(n_in, n_hidden),
-    # BatchNorm(n_hidden, activation),
-    # Dense(n_hidden, n_hidden),
-    # BatchNorm(n_hidden, activation),
-
     Dense(n_hidden, n_out),
 )
-# builder = MLJFlux.Short(n_hidden=n_hidden, dropout=0.1, σ=activation)
-# builder = MLJFlux.MLP(
-#     hidden=(
-#         n_hidden,
-#         n_hidden,
-#         n_hidden,
-#     ), 
-#     σ=activation
-# )
-α = [1.0,1.0,5e-3]
+n_ens = 5                                   # number of models in ensemble
+_loss = Flux.Losses.logitcrossentropy       # loss function
+_finaliser = x -> x                         # finaliser function
+```
 
+```{julia}
+# JEM parameters:
+𝒟x = Uniform(-1,1)
+𝒟y = Categorical(ones(output_dim) ./ output_dim)
+sampler = ConditionalSampler(
+    𝒟x, 𝒟y, 
+    input_size=(input_dim,), 
+    batch_size=1,
+)
+α = [1.0,1.0,1e-2]      # penalty strengths
+```
+
+```{julia}
 # Simple MLP:
@@
     batch_size=batch_size,
-    finaliser=x -> x,
-    loss=Flux.Losses.logitcrossentropy,
+    finaliser=_finaliser,
+    loss=_loss,
 )
@@
 # Deep Ensemble:
-mlp_ens = EnsembleModel(model=mlp, n=5)
+mlp_ens = EnsembleModel(model=mlp, n=n_ens)
 
 # Joint Energy Model:
-𝒟x = Uniform(-1,1)
-𝒟y = Categorical(ones(output_dim) ./ output_dim)
-sampler = ConditionalSampler(
-    𝒟x, 𝒟y, 
-    input_size=(input_dim,), 
-    batch_size=1
-)
 jem = JointEnergyClassifier(
@@
     builder=builder,
+    epochs=epochs,
     batch_size=batch_size,
-    finaliser=x -> x,
-    loss=Flux.Losses.logitcrossentropy,
+    finaliser=_finaliser,
+    loss=_loss,
     jem_training_params=(
         α=α,verbosity=10,
-        # use_gen_loss=false,
-        # use_reg_loss=false,
     ),
     sampling_steps=20,
-    epochs=epochs,
 )
 
+# JEM with adversarial training:
+jem_adv = deepcopy(jem)
+# jem_adv.adv_training = true
+
 # Deep Ensemble of Joint Energy Models:
-jem_ens = EnsembleModel(model=jem, n=5)
+jem_ens = EnsembleModel(model=jem, n=n_ens)
+
+# Deep Ensemble of Joint Energy Models with adversarial training:
+# jem_ens_plus = EnsembleModel(model=jem_adv, n=n_ens)
+
+# Dictionary of models:
+models = Dict(
+    "MLP" => mlp,
+    "MLP Ensemble" => mlp_ens,
+    "JEM" => jem,
+    "JEM Ensemble" => jem_ens,
+    # "JEM Ensemble+" => jem_ens_plus,
+)
 ```
@@
 ```{julia}
-cov = .95
-conf_model = conformal_model(jem_ens; method=:adaptive_inductive, coverage=cov)
-mach = machine(conf_model, X, labels)
+# Train models:
+function _train(model, X=X, y=labels; cov=.95, method=:simple_inductive)
+    conf_model = conformal_model(jem_ens; method=method, coverage=cov)
+    mach = machine(conf_model, X, y)
     fit!(mach)
     M = ECCCo.ConformalModel(mach.model, mach.fitresult)
+    return M
+end
+model_dict = Dict(mod_name => _train(mod) for (mod_name, mod) in models)
 ```

### 2023-04-24 paper/paper.tex
paper updates

@@
 
-\title{ECCoEs from the Black Box: Letting Models speak for Themselves}
+\title{ECCos from the Black Box: Letting Models speak for Themselves}
 
@@
 \begin{abstract}
-  We propose Eccoe: an effortless and rigorous way to produce plausible and conformal Counterfactual Explanations for Black Box Models using Conformal Prediction.
-To address the need for plausible explanations, existing work has primarily relied on surrogate models to learn the data-generating process.
+  Counterfactual Explanations offer an intuitive and straightforward way to explain Black Box Models but they are not unique.
+To address the need for plausible explanations, existing work has primarily relied on surrogate models to learn how the input data is distributed.
 This effectively reallocates the task of learning realistic representations of the data from the model itself to the surrogate.
 Consequently, the generated explanations may look plausible to humans but not necessarily conform with the behaviour of the Black Box Model.
-We formalise this notion through the introduction of new evaluation measures.
-In order to still address the need for plausibility, we build on a recent approach that works by minimizing predictive model uncertainty.
-Using differentiable Conformal Prediction, we relax the previous assumption that the Black Box Model can produce predictive uncertainty estimates.
+We formalise this notion of model conformity through the introduction of tailored evaluation measures and propose a novel algorithmic framework for generating \textbf{E}nergy-Constrained \textbf{C}onformal \textbf{Co}unterfactuals that are only as plausible as the model permits.
+To do so, \textbf{ECCo} leverages recent advances in energy-based modelling and predictive uncertainty quantification through conformal inference.
+Through illustrative examples and extensive empirical studies, we demonstrate that ECCos reconcile the need for plausibility and model conformity.
 \end{abstract}
@@
 
-Counterfactual Explanations are a powerful, flexible and intuitive way to not only explain Black Box Models but also enable affected individuals to challenge them through the means of Algorithmic Recourse.
+Counterfactual Explanations provide a powerful, flexible and intuitive way to not only explain Black Box Models but also enable affected individuals to challenge them through the means of Algorithmic Recourse.
 Instead of opening the black box, Counterfactual Explanations work under the premise of strategically perturbing model inputs to understand model behaviour \citep{wachter2017counterfactual}.
@@
 
-\section{From Adversarial Examples to Plausible Explanations}\label{background}
+\section{Background and Related Work}\label{background}
 
@@
 To start off, we briefly introduce the methodology uncerlying most state-of-the-art (SOTA) counterfactual generators.
+
+\subsection{Gradient-Based Counterfactual Search}\label{gradient}
 
@@
 
+\subsection{From Adversial Examples to Plausible Explanations}
+
 The crucial difference between Adversarial Examples (AE) and Counterfactual Explanations is one of intent.
@@
 Similarly, plausibility has also been connected to causality in the sense that plausible counterfactuals respect causal relationships \citep{mahajan2020preserving}.
+
+\subsubsection{Plausibility through Surrogates}
 
@@
 
-\section{Evaluating the Faithfulness of Counterfactuals}\label{conformity}
+\subsubsection{Plausibility through Minimal Predictive Uncertainty}
 
-In Section~\ref{background} we explained that Counterfactual Explanations work directly with Black Box Model, so fidelity is not a concern.
-This may explain why research has primarily focused on other desiderata, most notably plausibility (Definition~\ref{def:plausible}).
-Enquiring about the plausibility of a counterfactual essentially boils down to the following question: `Is this counterfactual consistent with the underlying data'?
-To introduce this section, we posit a related, slightly more nuanced question: `Is this counterfactual consistent with what the model has learned about the underlying data'?
-We will argue that fidelity is not a sufficient evaluation measure to answer this question and propose a novel way to assess if explanations conform with model behaviour.
-Finally, we will introduce a framework for Conformal Counterfactual Explanations, that reconciles the notions of plausibility and model conformity.
+\citet{schut2021generating} show that to meet the plausibility objective we need not explicitly model the input distribution.
+Pointing to the undesirable engineering overhead induced by surrogate models, they propose that we rely on the implicit minimisation of predictive uncertainty instead.
+Their proposed methodology solves Equation~\ref{eq:general} by greedily applying JSMA in the feature space with standard cross-entropy loss and no penalty at all.
+They demonstrate theoretically and empirically that their approach yields counterfactuals for which the model $M_{\theta}$ predicts the target label $\mathbf{y}^*$ with high confidence.
+Provided the model is well-specified, these counterfactuals are plausible.
 
-\subsection{From Fidelity to Model Conformity}
+Unfortunately, this idea hinges on the assumption that the Black Box Model provides well-calibrated predictive uncertainty estimates.
+The authors argue that in light of rapid advances in Bayesian Deep Learning (DL), this assumption is overall less costly than the engineering overhead induced by using surrogate models.
+This is even more true today, as recent work has put Laplace Approximation back on the map for truly effortless Bayesian DL \citep{immer2020improving,daxberger2021laplace,antoran2023sampling}.
+Nonetheless, the need for Bayesian methods may be too restrictive in some cases.
+Our proposed methodology, which we will turn to next, relaxes this restriction.
 
-The word \textit{fidelity} stems from the Latin word `fidelis', which means `faithful, loyal, trustworthy' \citep{mw2023fidelity}.
-As we explained in Section~\ref{background}, model explanations are considered faithful if their corresponding predictions coincide with the predictions made by the model itself.
-Since this definition of faithfulness is not useful in the context of Counterfactual Explanations, we propose an adapted version:
+\section{Energy-Constrained Conformal Counterfactuals}\label{ecco}
 
-\begin{definition}[Conformal Counterfactuals]
-  \label{def:conformal}
-  Let $\mathcal{X}_{\theta}|\mathbf{y}^* = p_{\theta}(x|\mathbf{y}^*)$ denote the conditional distribution of $\mathbf{x}$ in the target class $\mathbf{y}^*$, where $\theta$ denotes the parameters of model $M_{\theta}$.
-Then for $\mathbf{x}^{\prime}$ to be considered a conformal counterfactual, we need: $\mathbf{x}^{\prime} \sim \mathcal{X}_{\theta}|\mathbf{y}^*$.
-\end{definition}
+The primary objective of this work has been to develop a methodology for generating maximally plausible counterfactuals under minimal intervention.
+Our proposed framework is based on the premise that explanations should be plausible but not plausible at all costs.
+Energy-Constrained Conformal Counterfactuals (ECCo) achieve this goal in two ways: firstly, they rely on the Black Box itself for the generative task; and, secondly, they involve an approach to predictive uncertainty quantification that is model-agnostic.
 
-In words, conformal counterfactuals conform with what the predictive model has learned about the input data $\mathbf{x}$.
-Since this definition works with distributional properties, it explicitly accounts for the multiplicity of explanations we discussed earlier.
-Except for the posterior conditional distribution $p_{\theta}(\mathbf{x}|\mathbf{y}^*)$, we already have access to all the ingredients in Definition~\ref{def:conformal}.
+\subsection{Quantifying the Model's Generative Property}
 
@@
 
-Fortunately, recent work by \citet{grathwohl2020your} on Energy Based Models (EBM) has pointed out that there is a `generative model hidden within every standard discriminative model'.
+Recent work by \citet{grathwohl2020your} on Energy Based Models (EBM) has pointed out that there is a `generative model hidden within every standard discriminative model'.
 The authors show that we can draw samples from the posterior conditional distribution $p_{\theta}(\mathbf{x}|\mathbf{y})$ using Stochastic Gradient Langevin Dynamics (SGLD).
@@
 
-\textbf{TBD}
-
-\begin{itemize}
-  \item What exact sampler do we use? ImproperSGLD as in \citet{grathwohl2020your} seems to work best.
-\end{itemize}
-
-\subsection{Evaluation Measures}\label{evaluation}
-
-Above we have defined plausibility (\ref{def:plausible}) and conformity (\ref{def:conformal}) for Counterfactual Explanations.
-In this subsection, we introduce evaluation measures that facilitate a quantitative evaluation of counterfactuals for these objectives.
-
-Firstly, in order to assess the plausibility of counterfactuals we adapt the implausibility metric proposed in \citet{guidotti2022counterfactual}.
-The authors propose to evaluate plausibility in terms of the distance of the counterfactual $\mathbf{x}^{\prime}$ from its nearest neighbour in the target class $\mathbf{y}^*$: the smaller this distance, the more plausible the counterfactual.
-Instead of focusing only on the nearest neighbour of $\mathbf{x}^{\prime}$, we suggest computing the average over distances from multiple (possibly all) observed instances in the target class.
-Formally, for a single counterfactual, we have:
-
-\begin{equation}\label{eq:impl}
-  \begin{aligned}
-    \text{impl} = \frac{1}{\lvert\mathbf{x} \in \mathcal{X}|\mathbf{y}^*\rvert} \sum_{\mathbf{x} \in \mathcal{X}|\mathbf{y}^*} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
-  \end{aligned}
-\end{equation}
-
-This measure is straightforward to compute and should be less sensitive to outliers in the target class than the one based on the nearest neighbour.
-It also gives rise to a very similar evaluation measure for conformity.
-We merely swap out the subsample of individuals in the target class for the empirical distribution of generated conditional samples:
-
-\begin{equation}\label{eq:conf}
-  \begin{aligned}
-    \text{conf} = \frac{1}{\lvert\mathbf{x} \in \mathcal{X}_{\theta}|\mathbf{y}^*\rvert} \sum_{\mathbf{x} \in \mathcal{X}_{\theta}|\mathbf{y}^*} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
-  \end{aligned}
-\end{equation}
-
-As noted by \citet{guidotti2022counterfactual}, these distance-based measures are simplistic and more complex alternative measures may ultimately be more appropriate for the task.
-For example, we considered using statistical divergence measures instead.
-This would involve generating not one but many counterfactuals and comparing the generated empirical distribution to the target distributions in Definitions~\ref{def:plausible} and~\ref{def:conformal}.
-While this approach is potentially more rigorous, generating enough counterfactuals is not always practical.
-
-\section{A Framework for Conformal Counterfactual Explanations}\label{cce}
-
-Now that we have a framework for evaluating Counterfactual Explanations in terms of their plausibility and conformity, we are interested in finding a way to generate counterfactuals that are as plausible and conformal as possible.
-We hypothesize that a narrow focus on plausibility may come at the cost of reduced conformity.
-Using a surrogate model for the generative task, for example, may improve plausibility but inadvertently yield counterfactuals that are more consistent with the surrogate than the Black Box Model itself.
-We suggest that one way to ensure model conformity is to rely strictly on the model itself.
-In this section, we introduce a novel framework that meets this requirement, works under minimal assumptions and does not impede the plausibility objective: Conformal Counterfactual Explanations.
-
-\subsection{Plausible Counterfactuals through Minimal Uncertainty}
-
-Our proposed methodology is built on the findings presented in~\citet{schut2021generating}.
-The authors demonstrate that it is not only possible but remarkably easy to generate plausible counterfactuals for Black Box Models that provide predictive uncertainty estimates.
-Their proposed algorithm solves Equation~\ref{eq:general} by greedily applying JSMA in the feature space with standard cross-entropy loss and no penalty at all.
-They show that this is equivalent to minimizing predictive uncertainty and hence yields counterfactuals for which the model $M_{\theta}$ predicts the target label $\mathbf{y}^*$ with high confidence.
-Provided the model is well-calibrated, these counterfactuals are plausible which the authors demonstrate empirically through benchmarks \citep{schut2021generating}.
-
-Unfortunately, this idea hinges on the crucial assumption that the Black Box Model provides predictive uncertainty estimates.
-The authors argue that in light of rapid advances in Bayesian Deep Learning (DL), this assumption is overall less costly than the engineering overhead induced by using surrogate models.
-This is even more true today, as recent work has put Laplace Approximation back on the map for truly effortless Bayesian DL \citep{immer2020improving,daxberger2021laplace,antoran2023sampling}.
-Nonetheless, the need for Bayesian methods may be too restrictive in some cases.
-
-In looking for ways to lift that restriction, we found a promising alternative candidate for predictive uncertainty quantification (UQ) that we will briefly introduce next: Conformal Prediction.
-
-\subsection{Conformal Prediction}
+\subsection{Quantifying the Model's Predictive Uncertainty}
 
@@
 Our proposed approach for ECCCo therefore removes the restriction on the family of predictive models, at the small cost of reserving a subset of the available data for calibration.
+
+\section{Evaluation Framework}\label{conformity}
+
+In Section~\ref{background} we explained that Counterfactual Explanations work directly with Black Box Model, so fidelity is not a concern.
+This may explain why research has primarily focused on other desiderata, most notably plausibility (Definition~\ref{def:plausible}).
+Enquiring about the plausibility of a counterfactual essentially boils down to the following question: `Is this counterfactual consistent with the underlying data'?
+To introduce this section, we posit a related, slightly more nuanced question: `Is this counterfactual consistent with what the model has learned about the underlying data'?
+We will argue that fidelity is not a sufficient evaluation measure to answer this question and propose a novel way to assess if explanations conform with model behaviour.
+Finally, we will introduce a framework for Conformal Counterfactual Explanations, that reconciles the notions of plausibility and model conformity.
+
+\subsection{From Fidelity to Model Conformity}
+
+The word \textit{fidelity} stems from the Latin word `fidelis', which means `faithful, loyal, trustworthy' \citep{mw2023fidelity}.
+As we explained in Section~\ref{background}, model explanations are considered faithful if their corresponding predictions coincide with the predictions made by the model itself.
+Since this definition of faithfulness is not useful in the context of Counterfactual Explanations, we propose an adapted version:
+
+\begin{definition}[Conformal Counterfactuals]
+  \label{def:conformal}
+  Let $\mathcal{X}_{\theta}|\mathbf{y}^* = p_{\theta}(x|\mathbf{y}^*)$ denote the conditional distribution of $\mathbf{x}$ in the target class $\mathbf{y}^*$, where $\theta$ denotes the parameters of model $M_{\theta}$.
+Then for $\mathbf{x}^{\prime}$ to be considered a conformal counterfactual, we need: $\mathbf{x}^{\prime} \sim \mathcal{X}_{\theta}|\mathbf{y}^*$.
+\end{definition}
+
+In words, conformal counterfactuals conform with what the predictive model has learned about the input data $\mathbf{x}$.
+Since this definition works with distributional properties, it explicitly accounts for the multiplicity of explanations we discussed earlier.
+Except for the posterior conditional distribution $p_{\theta}(\mathbf{x}|\mathbf{y}^*)$, we already have access to all the ingredients in Definition~\ref{def:conformal}.
+
+\textbf{TBD}
+
+\begin{itemize}
+  \item What exact sampler do we use? ImproperSGLD as in \citet{grathwohl2020your} seems to work best.
+\end{itemize}
+
+\subsection{Evaluation Measures}\label{evaluation}
+
+Above we have defined plausibility (\ref{def:plausible}) and conformity (\ref{def:conformal}) for Counterfactual Explanations.
+In this subsection, we introduce evaluation measures that facilitate a quantitative evaluation of counterfactuals for these objectives.
+
+Firstly, in order to assess the plausibility of counterfactuals we adapt the implausibility metric proposed in \citet{guidotti2022counterfactual}.
+The authors propose to evaluate plausibility in terms of the distance of the counterfactual $\mathbf{x}^{\prime}$ from its nearest neighbour in the target class $\mathbf{y}^*$: the smaller this distance, the more plausible the counterfactual.
+Instead of focusing only on the nearest neighbour of $\mathbf{x}^{\prime}$, we suggest computing the average over distances from multiple (possibly all) observed instances in the target class.
+Formally, for a single counterfactual, we have:
+
+\begin{equation}\label{eq:impl}
+  \begin{aligned}
+    \text{impl} = \frac{1}{\lvert\mathbf{x} \in \mathcal{X}|\mathbf{y}^*\rvert} \sum_{\mathbf{x} \in \mathcal{X}|\mathbf{y}^*} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
+  \end{aligned}
+\end{equation}
+
+This measure is straightforward to compute and should be less sensitive to outliers in the target class than the one based on the nearest neighbour.
+It also gives rise to a very similar evaluation measure for conformity.
+We merely swap out the subsample of individuals in the target class for the empirical distribution of generated conditional samples:
+
+\begin{equation}\label{eq:conf}
+  \begin{aligned}
+    \text{conf} = \frac{1}{\lvert\mathbf{x} \in \mathcal{X}_{\theta}|\mathbf{y}^*\rvert} \sum_{\mathbf{x} \in \mathcal{X}_{\theta}|\mathbf{y}^*} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
+  \end{aligned}
+\end{equation}
+
+As noted by \citet{guidotti2022counterfactual}, these distance-based measures are simplistic and more complex alternative measures may ultimately be more appropriate for the task.
+For example, we considered using statistical divergence measures instead.
+This would involve generating not one but many counterfactuals and comparing the generated empirical distribution to the target distributions in Definitions~\ref{def:plausible} and~\ref{def:conformal}.
+While this approach is potentially more rigorous, generating enough counterfactuals is not always practical.
 

### 2023-04-24 notebooks/mnist.qmd
more work on paper

@@
 
-### ECCo
+### ECCCo
 
@@
 # Data:
+n_obs = 1000
+counterfactual_data = load_mnist(n_obs)
 counterfactual_data.X = pre_process.(counterfactual_data.X)

### 2023-04-24 paper/paper.tex
more work on paper

@@
 
-\title{ECCos from the Black Box: Letting Models speak for Themselves}
+\title{ECCCos from the Black Box: Letting Models speak for Themselves}
 
@@
 Consequently, the generated explanations may look plausible to humans but not necessarily conform with the behaviour of the Black Box Model.
-We formalise this notion of model conformity through the introduction of tailored evaluation measures and propose a novel algorithmic framework for generating \textbf{E}nergy-Constrained \textbf{C}onformal \textbf{Co}unterfactuals that are only as plausible as the model permits.
-To do so, \textbf{ECCo} leverages recent advances in energy-based modelling and predictive uncertainty quantification through conformal inference.
+We formalise this notion of model conformity through the introduction of tailored evaluation measures and propose a novel algorithmic framework for generating \textbf{E}nergy-\textbf{C}onstrained \textbf{C}onformal \textbf{Co}unterfactuals that are only as plausible as the model permits.
+To do so, \textbf{ECCCo} leverages recent advances in energy-based modelling and predictive uncertainty quantification through conformal inference.
 Through illustrative examples and extensive empirical studies, we demonstrate that ECCos reconcile the need for plausibility and model conformity.
@@
 Provided the model is well-specified, these counterfactuals are plausible.
-
 Unfortunately, this idea hinges on the assumption that the Black Box Model provides well-calibrated predictive uncertainty estimates.
-The authors argue that in light of rapid advances in Bayesian Deep Learning (DL), this assumption is overall less costly than the engineering overhead induced by using surrogate models.
-This is even more true today, as recent work has put Laplace Approximation back on the map for truly effortless Bayesian DL \citep{immer2020improving,daxberger2021laplace,antoran2023sampling}.
-Nonetheless, the need for Bayesian methods may be too restrictive in some cases.
 Our proposed methodology, which we will turn to next, relaxes this restriction.
 
-\section{Energy-Constrained Conformal Counterfactuals}\label{ecco}
+\section{Methodological Framework}\label{meth}
 
@@
 Our proposed framework is based on the premise that explanations should be plausible but not plausible at all costs.
-Energy-Constrained Conformal Counterfactuals (ECCo) achieve this goal in two ways: firstly, they rely on the Black Box itself for the generative task; and, secondly, they involve an approach to predictive uncertainty quantification that is model-agnostic.
+Energy-Constrained Conformal Counterfactuals (ECCCo) achieve this goal in two ways: firstly, they rely on the Black Box itself for the generative task; and, secondly, they involve an approach to predictive uncertainty quantification that is model-agnostic.
 
@@
 
-How can we quantify $p_{\theta}(\mathbf{x}|\mathbf{y}^*)$?
-After all, the predictive model $M_{\theta}$ was trained to discriminate outputs conditional on inputs, which is a different conditional distribution: $p_{\theta}(\mathbf{y}|\mathbf{x})$.
-Learning the distribution over inputs $p_{\theta}(\mathbf{x}|\mathbf{y}^*)$ is a generative task that $M_{\theta}$ was not explicitly trained for.
-In the context of Counterfactual Explanations, it is the task that existing approaches have reallocated from the model itself to a surrogate.
-
 Recent work by \citet{grathwohl2020your} on Energy Based Models (EBM) has pointed out that there is a `generative model hidden within every standard discriminative model'.
 The authors show that we can draw samples from the posterior conditional distribution $p_{\theta}(\mathbf{x}|\mathbf{y})$ using Stochastic Gradient Langevin Dynamics (SGLD).
-In doing so, it is possible to train classifiers jointly for the discriminative task using standard cross-entropy and the generative task using SGLD.
+The authors use this insight to train classifiers jointly for the discriminative task using standard cross-entropy and the generative task using SGLD.
 They demonstrate empirically that among other things this improves predictive uncertainty quantification for discriminative models.
+Our findings in this work suggest that Joint Energy Models (JEM) also tend to yield more plausible Counterfactual Explanations.
+Based on the definition of plausible counterfactuals (Definition~\ref{def:plausible}) this is not surprising.
 
-To see how their proposed conditional sampling strategy can be applied in our context, note that if we fix $\mathbf{y}$ to our target value $\mathbf{y}^*$, we can sample from $p_{\theta}(\mathbf{x}|\mathbf{y}^*)$ using SGLD as follows,
+Crucially for our purpose, one can apply their proposed sampling strategy during inference to essentially any standard discriminative model.
+Even models that are not explicitly trained for the joint objective learn about the distribution of inputs $X$ by learning to make conditional predictions about the output $y$.
+We can leverage this observation to quantify the generative property of the Black Box model itself.
+In particular, note that if we fix $\mathbf{y}$ to our target value $\mathbf{y}^*$, we can sample from $p_{\theta}(\mathbf{x}|\mathbf{y}^*)$ using SGLD as follows,
 
@@
 where $\mathbf{r}_j \sim \mathcal{N}(\mathbf{0},\mathbf{I})$ is the stochastic term and the step-size $\epsilon$ is typically polynomially decayed.
-The term $\mathcal{E}(\mathbf{x}_j|\mathbf{y}^*)$ denotes the energy function where as in \citep{grathwohl2020your} we use $\mathcal{E}(\mathbf{x}_j|\mathbf{y}^*)=-M_{\theta}(\mathbf{x}_j)[\mathbf{y}^*]$, that is the negative logit corresponding to the target class label $\mathbf{y}^*$.
-
-While $\mathbf{x}_J$ is only guaranteed to distribute as $p_{\theta}(\mathbf{x}|\mathbf{y}^*)$ if $\epsilon \rightarrow 0$ and $J \rightarrow \infty$, the bias introduced for a small finite $\epsilon$ is negligible in practice \citep{murphy2023probabilistic,grathwohl2020your}.
-While \citet{grathwohl2020your} use Equation~\ref{eq:sgld} during training, we are interested in applying the conditional sampling procedure in a post hoc fashion to any standard discriminative model.
-Generating multiple samples in this manner yields an empirical distribution $\hat{\mathcal{X}}_{\theta}|\mathbf{y}^*$, which we can use to assess if a given counterfactual $\mathbf{x}^{\prime}$ conforms with the model $M_{\theta}$ (Definition~\ref{def:conformal}).
+The term $\mathcal{E}(\mathbf{x}_j|\mathbf{y}^*)$ denotes the energy function where we use $\mathcal{E}(\mathbf{x}_j|\mathbf{y}^*)=-M_{\theta}(\mathbf{x}_j)[\mathbf{y}^*]$, that is the negative logit corresponding to the target class label $\mathbf{y}^*$.
+Generating multiple samples in this manner yields an empirical distribution $\hat{\mathcal{X}}_{\theta}|\mathbf{y}^*$ that we use in our search for plausible counterfactuals, as discussed in more detail below.
+Appendix~\ref{app-jem} provides additional implementation details for any tasks related to energy-based modelling.
 
@@
 
-Conformal Prediction (CP) is a scalable and statistically rigorous approach to predictive UQ that works under minimal distributional assumptions \citep{angelopoulos2021gentle}.
-It has recently gained popularity in the Machine Learning community \citep{angelopoulos2021gentle,manokhin2022awesome}.
-Crucially for our intended application, CP is model-agnostic and can be applied at test time.
-This allows us to relax the assumption that the Black Box Model needs to learn to generate predictive uncertainty estimates during training.
-In other words, CP promises to provide a way to generate plausible counterfactuals for any standard discriminative model without the need for surrogate models.
+To quantify the model's predictive uncertainty we use Conformal Prediction (CP), an approach that has recently gained popularity in the Machine Learning community \citep{angelopoulos2021gentle,manokhin2022awesome}.
+Crucially for our intended application, CP is model-agnostic and can be applied during inference without placing any restrictions on model training.
+Intuitively, CP works under the premise of turning heuristic notions of uncertainty into rigorous uncertainty estimates by repeatedly sifting through the training data or a dedicated calibration dataset.
+Conformal classifiers produce prediction sets for individual inputs that include all output labels that can be reasonably attributed to the input.
+These sets tend to be larger for inputs that do not conform with the training data and are therefore characterized by high predictive uncertainty.
 
-Intuitively, CP works under the premise of turning heuristic notions of uncertainty into rigorous uncertainty estimates by repeatedly sifting through the data.
-It can be used to generate prediction intervals for regression models and prediction sets for classification models \citep{altmeyer2022conformal}.
-Since the literature on CE and AR is typically concerned with classification problems, we focus on the latter.
-A particular variant of CP called Split Conformal Prediction (SCP) is well-suited for our purposes because it imposes only minimal restrictions on model training.
-
-Specifically, SCP involves splitting the data $\mathcal{D}_n=\{(\mathbf{x}_i,\mathbf{y}_i)\}_{i=1,...,n}$ into a proper training set $\mathcal{D}_{\text{train}}$ and a calibration set $\mathcal{D}_{\text{cal}}$.
-The former is used to train the classifier in any conventional fashion.
-The latter is then used to compute so-called nonconformity scores: $\mathcal{S}=\{s(\mathbf{x}_i,\mathbf{y}_i)\}_{i \in \mathcal{D}_{\text{cal}}}$ where $s: (\mathcal{X},\mathcal{Y}) \mapsto \mathbb{R}$ is referred to as \textit{score function}.
-In the context of classification, a common choice for the score function is just $s_i=1-M_{\theta}(\mathbf{x}_i)[\mathbf{y}_i]$, that is one minus the softmax output corresponding to the observed label $\mathbf{y}_i$ \citep{angelopoulos2021gentle}.
-
-Finally, classification sets are formed as follows,
-
-\begin{equation}\label{eq:scp}
-  \begin{aligned}
-    C_{\theta}(\mathbf{x}_i;\alpha)=\{\mathbf{y}: s(\mathbf{x}_i,\mathbf{y}) \le \hat{q}\}
-  \end{aligned}
-\end{equation}
-
-where $\hat{q}$ denotes the $(1-\alpha)$-quantile of $\mathcal{S}$ and $\alpha$ is a predetermined error rate.
-As the size of the calibration set increases, the probability that the classification set $C(\mathbf{x}_{\text{test}})$ for a newly arrived sample $\mathbf{x}_{\text{test}}$ does not cover the true test label $\mathbf{y}_{\text{test}}$ approaches $\alpha$ \citep{angelopoulos2021gentle}.
-
-Observe from Equation~\ref{eq:scp} that Conformal Prediction works on an instance-level basis, much like Counterfactual Explanations are local.
-The prediction set for an individual instance $\mathbf{x}_i$ depends only on the characteristics of that sample and the specified error rate.
-Intuitively, the set is more likely to include multiple labels for samples that are difficult to classify, so the set size is indicative of predictive uncertainty.
-To see why this effect is exacerbated by small choices for $\alpha$ consider the case of $\alpha=0$, which requires that the true label is covered by the prediction set with probability equal to one.
-
-\subsection{ECCoE: Energy-Constrained Conformal Counterfactual Explanation}
-
-The fact that conformal classifiers produce set-valued predictions introduces a challenge: it is not immediately obvious how to use such classifiers in the context of gradient-based counterfactual search.
-Put differently, it is not clear how to use prediction sets in Equation~\ref{eq:general}.
-Fortunately, \citet{stutz2022learning} have recently proposed a framework for Conformal Training that also hinges on differentiability.
-Specifically, they show how Stochastic Gradient Descent can be used to train classifiers not only for the discriminative task but also for additional objectives related to Conformal Prediction.
-One such objective is \textit{efficiency}: for a given target error rate $alpha$, the efficiency of a conformal classifier improves as its average prediction set size decreases.
-To this end, the authors introduce a smooth set size penalty,
+In order to generate counterfactuals that are associated with low predictive uncertainty, we use a smooth set size penalty introduced by \citet{stutz2022learning} in the context of conformal training:
 
@@
 
-where $\kappa \in \{0,1\}$ is a hyper-parameter and $C_{\theta,\mathbf{y}}(\mathbf{x}_i;\alpha)$ can be interpreted as the probability of label $\mathbf{y}$ being included in the prediction set.
-Formally, it is defined as $C_{\theta,\mathbf{y}}(\mathbf{x}_i;\alpha):=\sigma\left((s(\mathbf{x}_i,\mathbf{y})-\alpha) T^{-1}\right)$ for $\mathbf{y}\in\mathcal{Y}$ where $\sigma$ is the sigmoid function and $T$ is a hyper-parameter used for temperature scaling \citep{stutz2022learning}.
+Here, $\kappa \in \{0,1\}$ is a hyper-parameter and $C_{\theta,\mathbf{y}}(\mathbf{x}_i;\alpha)$ can be interpreted as the probability of label $\mathbf{y}$ being included in the prediction set.
 
-Penalizing the set size in this way is in principle enough to train efficient conformal classifiers \citep{stutz2022learning}.
-As we explained above, the set size is also closely linked to predictive uncertainty at the local level.
-This makes the smooth penalty defined in Equation~\ref{eq:setsize} useful in the context of meeting our objective of generating plausible counterfactuals.
-In particular, we adapt Equation~\ref{eq:general} to define the objective for our proposed Energy-Constrained Conformal Counterfactual Explanations (Eccoe):
+In order to compute this penalty for any Black Box Model we merely need to perform a single calibration pass through a holdout set $\mathcal{D}_{\text{cal}}$.
+Arguably, data is typically abundant and in most applications, practitioners tend to hold out a test data set anyway.
+Consequently, CP removes the restriction on the family of predictive models, at the small cost of reserving a subset of the available data for calibration.
+Details concerning our implementation of Conformal Prediction can be found in Appendix~\ref{app-cp}.
 
-\begin{equation}\label{eq:eccoe}
+\subsection{Energy-Constrained Conformal Counterfactuals (ECCCo)}
+
+Our framework for generating ECCCos combines the ideas introduced in the previous two subsections. Formally, we extend Equation~\ref{eq:general} as follows,
+
+\begin{equation} \label{eq:eccco}
   \begin{aligned}
-    \mathbf{Z}^\prime &= \arg \min_{\mathbf{Z}^\prime \in \mathcal{Z}^M} \left\{  {\text{yloss}(M_{\theta}(f(\mathbf{Z}^\prime)),\mathbf{y}^*)}+ \lambda \Omega(C_{\theta}(f(\mathbf{Z}^\prime);\alpha)) \right\}
+  \mathbf{Z}^\prime &= \arg \min_{\mathbf{Z}^\prime \in \mathcal{Z}^M} \left\{  {\text{yloss}(M_{\theta}(f(\mathbf{Z}^\prime)),\mathbf{y}^*)}+ \lambda {\text{cost}(f(\mathbf{Z}^\prime)) }  \right\} 
   \end{aligned} 
 \end{equation}
-
-Since we can still retrieve unperturbed softmax outputs from our conformal classifier $M_{\theta}$, we are free to work with any loss function of our choice.
-For example, we could use standard cross-entropy for $\text{yloss}$.
-
-In order to generate prediction sets $C_{\theta}(f(\mathbf{Z}^\prime);\alpha)$ for any Black Box Model we merely need to perform a single calibration pass through a holdout set $\mathcal{D}_{\text{cal}}$.
-Arguably, data is typically abundant and in most applications practitioners tend to hold out a test data set anyway.
-Our proposed approach for ECCCo therefore removes the restriction on the family of predictive models, at the small cost of reserving a subset of the available data for calibration.
 
@@
 \bibliography{bib}
+
+\appendix
+\section*{Appendices}
+\renewcommand{\thesubsection}{\Alph{subsection}}
+
+\subsection{JEM}\label{app-jem}
+
+While $\mathbf{x}_J$ is only guaranteed to distribute as $p_{\theta}(\mathbf{x}|\mathbf{y}^*)$ if $\epsilon \rightarrow 0$ and $J \rightarrow \infty$, the bias introduced for a small finite $\epsilon$ is negligible in practice \citep{murphy2023probabilistic,grathwohl2020your}.
+While \citet{grathwohl2020your} use Equation~\ref{eq:sgld} during training, we are interested in applying the conditional sampling procedure in a post hoc fashion to any standard discriminative model.
+
+\subsection{Conformal Prediction}\label{app-cp}
+
+The fact that conformal classifiers produce set-valued predictions introduces a challenge: it is not immediately obvious how to use such classifiers in the context of gradient-based counterfactual search.
+Put differently, it is not clear how to use prediction sets in Equation~\ref{eq:general}.
+Fortunately, \citet{stutz2022learning} have recently proposed a framework for Conformal Training that also hinges on differentiability.
+Specifically, they show how Stochastic Gradient Descent can be used to train classifiers not only for the discriminative task but also for additional objectives related to Conformal Prediction.
+One such objective is \textit{efficiency}: for a given target error rate $alpha$, the efficiency of a conformal classifier improves as its average prediction set size decreases.
+To this end, the authors introduce a smooth set size penalty defined in Equation~\ref{eq:setsize}
+
+Formally, it is defined as $C_{\theta,\mathbf{y}}(\mathbf{x}_i;\alpha):=\sigma\left((s(\mathbf{x}_i,\mathbf{y})-\alpha) T^{-1}\right)$ for $\mathbf{y}\in\mathcal{Y}$ where $\sigma$ is the sigmoid function and $T$ is a hyper-parameter used for temperature scaling \citep{stutz2022learning}.
+
+Intuitively, CP works under the premise of turning heuristic notions of uncertainty into rigorous uncertainty estimates by repeatedly sifting through the data.
+It can be used to generate prediction intervals for regression models and prediction sets for classification models \citep{altmeyer2022conformal}.
+Since the literature on CE and AR is typically concerned with classification problems, we focus on the latter.
+A particular variant of CP called Split Conformal Prediction (SCP) is well-suited for our purposes because it imposes only minimal restrictions on model training.
+
+Specifically, SCP involves splitting the data $\mathcal{D}_n=\{(\mathbf{x}_i,\mathbf{y}_i)\}_{i=1,...,n}$ into a proper training set $\mathcal{D}_{\text{train}}$ and a calibration set $\mathcal{D}_{\text{cal}}$.
+The former is used to train the classifier in any conventional fashion.
+The latter is then used to compute so-called nonconformity scores: $\mathcal{S}=\{s(\mathbf{x}_i,\mathbf{y}_i)\}_{i \in \mathcal{D}_{\text{cal}}}$ where $s: (\mathcal{X},\mathcal{Y}) \mapsto \mathbb{R}$ is referred to as \textit{score function}.
+In the context of classification, a common choice for the score function is just $s_i=1-M_{\theta}(\mathbf{x}_i)[\mathbf{y}_i]$, that is one minus the softmax output corresponding to the observed label $\mathbf{y}_i$ \citep{angelopoulos2021gentle}.
+
+Finally, classification sets are formed as follows,
+
+\begin{equation}\label{eq:scp}
+  \begin{aligned}
+    C_{\theta}(\mathbf{x}_i;\alpha)=\{\mathbf{y}: s(\mathbf{x}_i,\mathbf{y}) \le \hat{q}\}
+  \end{aligned}
+\end{equation}
+
+where $\hat{q}$ denotes the $(1-\alpha)$-quantile of $\mathcal{S}$ and $\alpha$ is a predetermined error rate.
+As the size of the calibration set increases, the probability that the classification set $C(\mathbf{x}_{\text{test}})$ for a newly arrived sample $\mathbf{x}_{\text{test}}$ does not cover the true test label $\mathbf{y}_{\text{test}}$ approaches $\alpha$ \citep{angelopoulos2021gentle}.
+
+Observe from Equation~\ref{eq:scp} that Conformal Prediction works on an instance-level basis, much like Counterfactual Explanations are local.
+The prediction set for an individual instance $\mathbf{x}_i$ depends only on the characteristics of that sample and the specified error rate.
+Intuitively, the set is more likely to include multiple labels for samples that are difficult to classify, so the set size is indicative of predictive uncertainty.
+To see why this effect is exacerbated by small choices for $\alpha$ consider the case of $\alpha=0$, which requires that the true label is covered by the prediction set with probability equal to one.
 

### 2023-04-24 notebooks/mnist.qmd
mnist

@@
 # Train models:
-function _train(model, X=X, y=labels; cov=.95, method=:simple_inductive)
-    conf_model = conformal_model(jem_ens; method=method, coverage=cov)
+function _train(model, X=X, y=labels; cov=.95, method=:simple_inductive, mod_name="model")
+    conf_model = conformal_model(model; method=method, coverage=cov)
     mach = machine(conf_model, X, y)
+    @info "Begin training $mod_name."
     fit!(mach)
+    @info "Finished training $mod_name."
     M = ECCCo.ConformalModel(mach.model, mach.fitresult)
@@
 end
-model_dict = Dict(mod_name => _train(mod) for (mod_name, mod) in models)
+model_dict = Dict(mod_name => _train(mod; mod_name=mod_name) for (mod_name, mod) in models)
+Serialization.serialize(joinpath(output_path,"mnist_models.jls"), model_dict)
 ```
@@
 ```{julia}
-if mach.model.model isa JointEnergyClassifier
-    sampler = mach.model.model.jem.sampler
+for (mod_name, mod) in model_dict
+
+    # Plot:
+    if mod.model.model isa JointEnergyClassifier
+        sampler = mod.model.model.jem.sampler
+    elseif mod.model.model.model isa JointEnergyClassifier
+        sampler = mod.model.model.model.jem.sampler
     else
@@
     opt = ImproperSGLD()
-f(x) = logits(M, x)
+    f(x) = logits(mod, x)
 
@@
     end
-plt = Plots.plot(plts..., size=(_w,_w), layout=(10,1))
+    plt = Plots.plot(plts..., size=(_w,_w), layout=(10,1), plot_title=mod_name)
     display(plt)
 
-```
-
-```{julia}
+    # Test performance:
     test_data = load_mnist_test()
@@
     println("F1 score (test): $(round(f1,digits=3))")
+end
 ```

### 2023-04-25 src/sampling.jl
MNIST

@@ -43,8 +43,8 @@ function EnergySampler(
 
     @assert y ∈ data.y_levels || y ∈ 1:length(data.y_levels)
 
-    if model.model.model isa JointEnergyClassifier
-        sampler = model.model.model.jem.sampler
+    if ECCCo._has_sampler(model)
+        sampler = ECCCo._get_sampler(model)
     else
         K = length(data.y_levels)
         input_size = size(selectdim(data.X, ndims(data.X), 1))
@@ -57,8 +57,10 @@ function EnergySampler(
     # Initiate:
     energy_sampler = EnergySampler(model, data, sampler, opt, nothing, nothing)
 
-    # Generate conditional samples:
-    generate_samples!(energy_sampler, nsamples, yidx; niter=niter)
+    # Generate conditional samples (one at a time):
+    for i in 1:nsamples
+        generate_samples!(energy_sampler, 1, yidx; niter=niter)
+    end
 
     return energy_sampler
 end
@@ -106,7 +108,11 @@ end
 Generates `n` samples from `EnergySampler` for conditioning value `y`. Assigns samples and conditioning value to `EnergySampler`.
 """
 function generate_samples!(e::EnergySampler, n::Int, y::Int; niter::Int=100)
-    e.buffer = generate_samples(e, n, y; niter=niter)
+    if isnothing(e.buffer)
+        e.buffer = generate_samples(e, n, y; niter=niter)
+    else
+        e.buffer = cat(e.buffer, generate_samples(e, n, y; niter=niter), dims=ndims(e.buffer))
+    end
     e.yidx = y
 end
 

### 2023-04-25 notebooks/mnist.qmd
MNIST

@@
 ```{julia}
+# Plot generated samples:
+
 for (mod_name, mod) in model_dict
-
-    # Plot:
-    if mod.model.model isa JointEnergyClassifier
-        sampler = mod.model.model.jem.sampler
-    elseif mod.model.model.model isa JointEnergyClassifier
-        sampler = mod.model.model.model.jem.sampler
+    if ECCCo._has_sampler(mod)
+        sampler = ECCCo._get_sampler(mod)
     else
@@
     plt = Plots.plot(plts..., size=(_w,_w), layout=(10,1), plot_title=mod_name)
+    savefig(plt, joinpath(output_images_path, "mnist_generated_$(mod_name).png"))
     display(plt)
+end
+```
 
+```{julia}
+# Evaluate models:
+
+measure = Dict(
+    :f1score => multiclass_f1score, 
+    :acc => accuracy, 
+    :precision => multiclass_precision
+)
+model_performance = DataFrame()
+for (mod_name, mod) in model_dict
     # Test performance:
@@
     test_data.X = pre_process.(test_data.X)
-    f1 = CounterfactualExplanations.Models.model_evaluation(M, test_data)
-    println("F1 score (test): $(round(f1,digits=3))")
+    _perf = CounterfactualExplanations.Models.model_evaluation(mod, test_data, measure=collect(values(measure)))
+    _perf = DataFrame([[p] for p in _perf], collect(keys(measure)))
+    _perf.mod_name .= mod_name
+    model_performance = vcat(model_performance, _perf)
 end
+Serialization.serialize(joinpath(output_path,"mnist_model_performance.jls"), model_performance)
+CSV.write(joinpath(output_path, "mnist_model_performance.csv"), model_performance)
+model_performance
+```
+
+```{julia}
+Random.seed!(123)
+
+# Set up search:
+factual = 8
+x = reshape(counterfactual_data.X[:,rand(findall(labels.==factual))],input_dim,1)
+target = 3
+γ = 0.9
+T = 100
+
+# ECCCo:
+λ=[0.5,0.1,0.5]
+temp=0.1
+η=0.01
+
+# Generate counterfactuals using ECCCo generator:
+generator = ECCCoGenerator(
+    λ=λ, 
+    temp=temp, 
+    opt=Flux.Optimise.Adam(η),
+)
+
+ces = Dict()
+for (mod_name, mod) in model_dict
+    ce = generate_counterfactual(
+        x, target, counterfactual_data, mod, generator; 
+        decision_threshold=γ, max_iter=T,
+        initialization=:identity,
+        converge_when=:generator_conditions,
+    )
+    ces[mod_name] = ce
+end
+
+# Plot:
+p1 = Plots.plot(
+    convert2image(MNIST, reshape(x,28,28)),
+    axis=nothing, 
+    size=(img_height, img_height),
+    title="Factual"
+)
+plts = [p1]
+
+for (_name,ce) in ces
+    x = CounterfactualExplanations.counterfactual(ce)
+    _phat = target_probs(ce)
+    _title = "$_name (p̂=$(round(_phat[1]; digits=3)))"
+    plt = Plots.plot(
+        convert2image(MNIST, reshape(x,28,28)),
+        axis=nothing, 
+        size=(img_height, img_height),
+        title=_title
+    )
+    plts = [plts..., plt]
+end
+plt = Plots.plot(plts...; size=(img_height*length(plts),img_height), layout=(1,length(plts)))
+display(plt)
 ```

### 2023-04-25 notebooks/setup.jl

@@ -39,6 +39,7 @@ setup_notebooks = quote
     Random.seed!(2023)
     www_path = "www"
     output_path = "artifacts/results"
+    output_images_path = "artifacts/results/images"
     img_height = 300
 
 end;

### 2023-04-25 notebooks/mnist.qmd
work on mnist example

--- 2023-04-24 notebooks/mnist.qmd
@@
 ```{julia}
+Random.seed!(1234)
+
 # Data:
@@
 factual_label = 8
-x = reshape(X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
+x_factual = reshape(X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
 target = 3
-factual = predict_label(M, counterfactual_data, x)[1]
+factual = predict_label(M, counterfactual_data, x_factual)[1]
 γ = 0.9
-T = 50
+
+# Training params:
+T = 100
+opt = Flux.Optimise.Adam(0.01)
 ```
@@
 # Search:
-opt = Flux.Optimise.Adam(0.01)
 generator = GenericGenerator(opt=opt)
 ce_wachter = generate_counterfactual(
-    x, target, counterfactual_data, M, generator; 
+    x_factual, target, counterfactual_data, M, generator; 
     decision_threshold=γ, max_iter=T,
@@
 ce_jsma = generate_counterfactual(
-    x, target, counterfactual_data, M, generator; 
+    x_factual, target, counterfactual_data, M, generator; 
     decision_threshold=γ, max_iter=T,
@@
 p1 = Plots.plot(
-    convert2image(MNIST, reshape(x,28,28)),
+    convert2image(MNIST, reshape(x_factual,28,28)),
     axis=nothing, 
@@
 display(plt)
-savefig(plt, joinpath(www_path, "you_may_not_like_it.png"))
+savefig(plt, joinpath(output_images_path, "you_may_not_like_it.png"))
 ```
@@
   opt = opt,
-  λ=0.01
+  λ=0.1
 )
@@
 ce_strong = generate_counterfactual(
-    x, target, counterfactual_data, M, generator; 
+    x_factual, target, counterfactual_data, M, generator; 
     decision_threshold=γ, max_iter=T,
     initialization=:identity,
+    converge_when=:generator_conditions,
 )
@@
 ce_weak = generate_counterfactual(
-    x, target, counterfactual_data, M, generator;
+    x_factual, target, counterfactual_data, M, generator;
     decision_threshold=γ, max_iter=T,
     initialization=:identity,
+    converge_when=:generator_conditions,
 )
@@
 display(plt)
-savefig(plt, joinpath(www_path, "surrogate_gone_wrong.png"))
+savefig(plt, joinpath(output_images_path, "surrogate_gone_wrong.png"))
 ```
@@
 ```{julia}
+# Hyper:
+_retrain = false
+_regen = false
+
 # Data:
-n_obs = 1000
+n_obs = 10000
 counterfactual_data = load_mnist(n_obs)
@@
 X = table(permutedims(X))
+x_factual = reshape(pre_process(x_factual, noise=0.0f0), input_dim, 1)
 labels = counterfactual_data.output_encoder.labels
@@
 end
-model_dict = Dict(mod_name => _train(mod) for (mod_name, mod) in models)
+if _retrain
+    model_dict = Dict(mod_name => _train(mod; mod_name=mod_name) for (mod_name, mod) in models)
+    Serialization.serialize(joinpath(output_path,"mnist_models.jls"), model_dict)
+else
+    model_dict = Serialization.deserialize(joinpath(output_path,"mnist_models.jls"))
+end
 ```
@@
 ```{julia}
-if mach.model.model isa JointEnergyClassifier
-    sampler = mach.model.model.jem.sampler
+# Plot generated samples:
+if _regen 
+    for (mod_name, mod) in model_dict
+        if ECCCo._has_sampler(mod)
+            sampler = ECCCo._get_sampler(mod)
         else
@@
         opt = ImproperSGLD()
-f(x) = logits(M, x)
+        f(x) = logits(mod, x)
 
@@
         end
-plt = Plots.plot(plts..., size=(_w,_w), layout=(10,1))
+        plt = Plots.plot(plts..., size=(_w,_w), layout=(10,1), plot_title=mod_name)
+        savefig(plt, joinpath(output_images_path, "mnist_generated_$(mod_name).png"))
         display(plt)
-
+    end
+end
 ```
@@
 ```{julia}
+# Evaluate models:
+
+measure = Dict(
+    :f1score => multiclass_f1score, 
+    :acc => accuracy, 
+    :precision => multiclass_precision
+)
+model_performance = DataFrame()
+for (mod_name, mod) in model_dict
+    # Test performance:
     test_data = load_mnist_test()
-test_data.X = pre_process.(test_data.X)
-f1 = CounterfactualExplanations.Models.model_evaluation(M, test_data)
-println("F1 score (test): $(round(f1,digits=3))")
+    test_data.X = pre_process.(test_data.X, noise=0.0f0)
+    _perf = CounterfactualExplanations.Models.model_evaluation(mod, test_data, measure=collect(values(measure)))
+    _perf = DataFrame([[p] for p in _perf], collect(keys(measure)))
+    _perf.mod_name .= mod_name
+    model_performance = vcat(model_performance, _perf)
+end
+Serialization.serialize(joinpath(output_path,"mnist_model_performance.jls"), model_performance)
+CSV.write(joinpath(output_path, "mnist_model_performance.csv"), model_performance)
+model_performance
+```
+
+```{julia}
+# ECCCo:
+λ=[0.5,0.1,0.5]
+temp=0.1
+η=0.01
+
+# Generate counterfactuals using ECCCo generator:
+generator = ECCCoGenerator(
+    λ=λ, 
+    temp=temp, 
+    opt=Flux.Optimise.Adam(η),
+)
+
+ces = Dict()
+for (mod_name, mod) in model_dict
+    ce = generate_counterfactual(
+        x_factual, target, counterfactual_data, mod, generator; 
+        decision_threshold=γ, max_iter=T,
+        initialization=:identity,
+        converge_when=:generator_conditions,
+    )
+    ces[mod_name] = ce
+end
+plt_order = sortperm(collect(keys(ces)))
+
+# Plot:
+p1 = Plots.plot(
+    convert2image(MNIST, reshape(x_factual,28,28)),
+    axis=nothing, 
+    size=(img_height, img_height),
+    title="Factual"
+)
+
+plts = []
+for (_name,ce) in ces
+    _x = CounterfactualExplanations.counterfactual(ce)
+    _phat = target_probs(ce)
+    _title = "$_name (p̂=$(round(_phat[1]; digits=3)))"
+    plt = Plots.plot(
+        convert2image(MNIST, reshape(_x,28,28)),
+        axis=nothing, 
+        size=(img_height, img_height),
+        title=_title
+    )
+    plts = [plts..., plt]
+end
+plts = plts[plt_order]
+plts = [p1, plts...]
+plt = Plots.plot(plts...; size=(img_height*length(plts),img_height), layout=(1,length(plts)))
+display(plt)
+savefig(plt, joinpath(output_images_path, "mnist_eccco.png"))
 ```
@@
 factual_label = 9
-x = reshape(counterfactual_data.X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
+x_factual = reshape(counterfactual_data.X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
 target = 7
-factual = predict_label(M, counterfactual_data, x)[1]
+factual = predict_label(M, counterfactual_data, x_factual)[1]
 γ = 0.9
@@
 ce_wachter = generate_counterfactual(
-    x, target, counterfactual_data, M, generator; 
+    x_factual, target, counterfactual_data, M, generator; 
     decision_threshold=γ, max_iter=T,
@@
 ce_conformal = generate_counterfactual(
-    x, target, counterfactual_data, M, generator; 
+    x_factual, target, counterfactual_data, M, generator; 
     decision_threshold=γ, max_iter=T,
@@
 ce_conformal_jsma = generate_counterfactual(
-    x, target, counterfactual_data, M, generator; 
+    x_factual, target, counterfactual_data, M, generator; 
     decision_threshold=γ, max_iter=T,
@@
     ce, _name = (x[1],x[2])
-    x = CounterfactualExplanations.counterfactual(ce)
+    x_factual = CounterfactualExplanations.counterfactual(ce)
     _phat = target_probs(ce)
@@
     plt = Plots.plot(
-        convert2image(MNIST, reshape(x,28,28)),
+        convert2image(MNIST, reshape(x_factual,28,28)),
         axis=nothing, 
@@
 factual_label = 8
-x = reshape(counterfactual_data.X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
+x_factual = reshape(counterfactual_data.X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
 target = 3
-factual = predict_label(M, counterfactual_data, x)[1]
+factual = predict_label(M, counterfactual_data, x_factual)[1]
 γ = 0.5

### 2023-04-25 paper/paper.tex
MNIST
work on mnist example

@@
     \centering
-    \includegraphics[width=\textwidth]{../www/you_may_not_like_it.png}
+    \includegraphics[width=\textwidth]{../artifacts/results/images/you_may_not_like_it.png}
     \caption{You may not like it, but this is what stripped-down counterfactuals look like.
@@
     \centering
-    \includegraphics[width=\textwidth]{../www/surrogate_gone_wrong.png}
+    \includegraphics[width=\textwidth]{../artifacts/results/images/surrogate_gone_wrong.png}
     \caption{Using surrogates can improve plausibility, but also increases vulnerability.
@@
 
+\begin{figure}
+  \includegraphics[width=\textwidth]{../artifacts/results/images/mnist_eccco.png}
+  \caption{ECCCos from Black Boxes. Counterfactuals for turning an 8 (eight) into a 3 (three): original image (left); }\label{fig:eccco}
+\end{figure}
+
 \section{Evaluation Framework}\label{conformity}
@@
   \item It seems that models that are not explicitly trained for generative task, still learn it implictly
-  \item Batch size seems to impact quality of generated samples
+  \item Batch size seems to impact quality of generated samples (at inference, but not so much during JEM training)
+  \item ECCCo is sensitive to optimizer (Adam works well), learning rate and distance metric (l1 works well)
+  \item SGLD takes time 
+  \item REVISE has benefit of lower dimensional space
 \end{itemize}

### 2023-04-25 src/generator.jl
more work on mnist

@@ -28,7 +28,7 @@ function ECCCoGenerator(;
     function _set_size_penalty(ce::AbstractCounterfactualExplanation)
         return ECCCo.set_size_penalty(ce; κ=κ, temp=temp)
     end
-    _penalties = [Objectives.distance_l1, _set_size_penalty, ECCCo.distance_from_energy]
+    _penalties = [Objectives.distance_l2, _set_size_penalty, ECCCo.distance_from_energy]
     λ = λ isa AbstractFloat ? [0.0, λ, λ] : λ
     return Generator(; penalty=_penalties, λ=λ, opt=opt, kwargs...)
 end

### 2023-04-25 notebooks/eccco.qmd
work on paper

--- 2023-04-22 notebooks/mnist.qmd
@@
 ```
-
-# MNIST
-
-```{julia}
-function pre_process(x; noise::Float32=0.03f0)
-    ϵ = Float32.(randn(size(x)) * noise)
-    x = @.(2 * x - 1) .+ ϵ
-    return x
-end
-```
-
-```{julia}
-# Data:
-n_obs = 1000
-counterfactual_data = load_mnist(n_obs)
-counterfactual_data.X = pre_process.(counterfactual_data.X)
-X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
-X = table(permutedims(X))
-labels = counterfactual_data.output_encoder.labels
-input_dim, n_obs = size(counterfactual_data.X)
-n_digits = Int(sqrt(input_dim))
-output_dim = length(unique(labels))
-```
-
-First, let's create a couple of image classifier architectures:
-
-```{julia}
-# Model parameters:
-epochs = 100
-batch_size = minimum([Int(round(n_obs/10)), 128])
-n_hidden = 128
-activation = Flux.swish
-builder = MLJFlux.@builder Flux.Chain(
-
-    Dense(n_in, n_hidden, activation),
-    # Dense(n_hidden, n_hidden, activation),
-    # Dense(n_hidden, n_hidden, activation),
-
-    # Dense(n_in, n_hidden),
-    # BatchNorm(n_hidden, activation),
-    # Dense(n_hidden, n_hidden),
-    # BatchNorm(n_hidden, activation),
-
-    Dense(n_hidden, n_out),
-)
-# builder = MLJFlux.Short(n_hidden=n_hidden, dropout=0.1, σ=activation)
-# builder = MLJFlux.MLP(
-#     hidden=(
-#         n_hidden,
-#         n_hidden,
-#         n_hidden,
-#     ), 
-#     σ=activation
-# )
-α = [1.0,1.0,1e-2]
-
-# Simple MLP:
-mlp = NeuralNetworkClassifier(
-    builder=builder, 
-    epochs=epochs,
-    batch_size=batch_size,
-)
-
-# Joint Energy Model:
-𝒟x = Uniform(-1,1)
-𝒟y = Categorical(ones(output_dim) ./ output_dim)
-sampler = ConditionalSampler(
-    𝒟x, 𝒟y, 
-    input_size=(input_dim,), 
-    batch_size=1
-)
-jem = JointEnergyClassifier(
-    sampler;
-    builder=builder,
-    batch_size=batch_size,
-    finaliser=x -> x,
-    loss=Flux.Losses.logitcrossentropy,
-    jem_training_params=(
-        α=α,verbosity=10,
-        # use_gen_loss=false,
-        # use_reg_loss=false,
-    ),
-    sampling_steps=20,
-    epochs=epochs,
-)
-
-# Deep Ensemble:
-mlp_ens = EnsembleModel(model=mlp, n=50)
-```
-
-```{julia}
-cov = .95
-conf_model = conformal_model(jem; method=:simple_inductive, coverage=cov)
-mach = machine(conf_model, X, labels)
-fit!(mach)
-M = ECCCo.ConformalModel(mach.model, mach.fitresult)
-```
-
-```{julia}
-if mach.model.model isa JointEnergyModels.JointEnergyClassifier
-    jem = mach.model.model.jem
-    n_iter = 200
-    _w = 1500
-    plts = []
-    neach = 10
-    for i in 1:10
-        x = jem.sampler(jem.chain, jem.sampling_rule; niter=n_iter, n_samples=neach, y=i)
-        plts_i = []
-        for j in 1:size(x, 2)
-            xj = x[:,j]
-            xj = reshape(xj, (n_digits, n_digits))
-            plts_i = [plts_i..., Plots.heatmap(rotl90(xj), axis=nothing, cb=false)]
-        end
-        plt = Plots.plot(plts_i..., size=(_w,0.10*_w), layout=(1,10))
-        plts = [plts..., plt]
-    end
-    plt = Plots.plot(plts..., size=(_w,_w), layout=(10,1))
-    display(plt)
-end
-```
-
-```{julia}
-test_data = load_mnist_test()
-test_data.X = pre_process.(test_data.X)
-f1 = CounterfactualExplanations.Models.model_evaluation(M, test_data)
-println("F1 score (test): $(round(f1,digits=3))")
-```
-
-```{julia}
-Random.seed!(1234)
-
-# Set up search:
-factual_label = 9
-x = reshape(counterfactual_data.X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
-target = 4
-factual = predict_label(M, counterfactual_data, x)[1]
-γ = 0.5
-T = 250
-
-η=1.0
-
-# Generate counterfactual using generic generator:
-generator = GenericGenerator()
-ce_wachter = generate_counterfactual(
-    x, target, counterfactual_data, M, generator; 
-    decision_threshold=γ, max_iter=T,
-    initialization=:identity,
-)
-
-generator = GreedyGenerator(η=η)
-ce_jsma = generate_counterfactual(
-    x, target, counterfactual_data, M, generator; 
-    decision_threshold=γ, max_iter=T,
-    initialization=:identity,
-)
-
-# ECCCo:
-λ=[0.1,0.1,0.1]
-temp=0.1
-
-# Generate counterfactual using ECCCo generator:
-generator = ECCCoGenerator(
-    λ=λ, 
-    temp=temp, 
-)
-ce_conformal = generate_counterfactual(
-    x, target, counterfactual_data, M, generator; 
-    decision_threshold=γ, max_iter=T,
-    initialization=:identity,
-    converge_when=:generator_conditions,
-)
-
-# Generate counterfactual using ECCCo generator:
-generator = ECCCoGenerator(
-    λ=λ, 
-    temp=temp, 
-    opt=CounterfactualExplanations.Generators.JSMADescent(η=η),
-)
-ce_conformal_jsma = generate_counterfactual(
-    x, target, counterfactual_data, M, generator; 
-    decision_threshold=γ, max_iter=T,
-    initialization=:identity,
-    converge_when=:generator_conditions,
-)
-
-# Plot:
-p1 = Plots.plot(
-    convert2image(MNIST, reshape(x,28,28)),
-    axis=nothing, 
-    size=(img_height, img_height),
-    title="Factual"
-)
-plts = [p1]
-
-ces = [ce_wachter, ce_conformal, ce_jsma, ce_conformal_jsma]
-_names = ["Wachter", "ECCCo", "JSMA", "ECCCo-JSMA"]
-for x in zip(ces, _names)
-    ce, _name = (x[1],x[2])
-    x = CounterfactualExplanations.counterfactual(ce)
-    _phat = target_probs(ce)
-    _title = "$_name (p̂=$(round(_phat[1]; digits=3)))"
-    plt = Plots.plot(
-        convert2image(MNIST, reshape(x,28,28)),
-        axis=nothing, 
-        size=(img_height, img_height),
-        title=_title
-    )
-    plts = [plts..., plt]
-end
-plt = Plots.plot(plts...; size=(img_height*length(plts),img_height), layout=(1,length(plts)))
-display(plt)
-savefig(plt, joinpath(www_path, "eccco_mnist.png"))
-```
-
-```{julia}
-# Random.seed!(1234)
-
-# Set up search:
-factual_label = 8
-x = reshape(counterfactual_data.X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
-target = 3
-factual = predict_label(M, counterfactual_data, x)[1]
-γ = 0.5
-T = 100
-
-# Generate counterfactual using generic generator:
-generator = GenericGenerator(opt=Flux.Optimise.Adam(),)
-ce_wachter = generate_counterfactual(
-    x, target, counterfactual_data, M, generator; 
-    decision_threshold=γ, max_iter=T,
-    initialization=:identity,
-)
-
-generator = GreedyGenerator(η=1.0)
-ce_jsma = generate_counterfactual(
-    x, target, counterfactual_data, M, generator; 
-    decision_threshold=γ, max_iter=T,
-    initialization=:identity,
-)
-
-# ECCCo:
-λ=[0.0,1.0]
-temp=0.5
-
-# Generate counterfactual using CCE generator:
-generator = CCEGenerator(
-    λ=λ, 
-    temp=temp, 
-    opt=Flux.Optimise.Adam(),
-)
-ce_conformal = generate_counterfactual(
-    x, target, counterfactual_data, M, generator; 
-    decision_threshold=γ, max_iter=T,
-    initialization=:identity,
-    converge_when=:generator_conditions,
-)
-
-# Generate counterfactual using CCE generator:
-generator = CCEGenerator(
-    λ=λ, 
-    temp=temp, 
-    opt=CounterfactualExplanations.Generators.JSMADescent(η=1.0),
-)
-ce_conformal_jsma = generate_counterfactual(
-    x, target, counterfactual_data, M, generator; 
-    decision_threshold=γ, max_iter=T,
-    initialization=:identity,
-    converge_when=:generator_conditions,
-)
-
-# Plot:
-p1 = Plots.plot(
-    convert2image(MNIST, reshape(x,28,28)),
-    axis=nothing, 
-    size=(img_height, img_height),
-    title="Factual"
-)
-plts = [p1]
-
-ces = [ce_wachter, ce_conformal, ce_jsma, ce_conformal_jsma]
-_names = ["Wachter", "CCE", "JSMA", "CCE-JSMA"]
-for x in zip(ces, _names)
-    ce, _name = (x[1],x[2])
-    x = CounterfactualExplanations.counterfactual(ce)
-    _phat = target_probs(ce)
-    _title = "$_name (p̂=$(round(_phat[1]; digits=3)))"
-    plt = Plots.plot(
-        convert2image(MNIST, reshape(x,28,28)),
-        axis=nothing, 
-        size=(img_height, img_height),
-        title=_title
-    )
-    plts = [plts..., plt]
-end
-plt = Plots.plot(plts...; size=(img_height*length(plts),img_height), layout=(1,length(plts)))
-display(plt)
-savefig(plt, joinpath(www_path, "cce_mnist.png"))
-```
-
-```{julia}
-if M.model.model isa JointEnergyModels.JointEnergyClassifier
-    jem = M.model.model.jem
-    n_iter = 200
-    _w = 1500
-    plts = []
-    neach = 10
-    for i in 1:10
-        x = jem.sampler(jem.chain, jem.sampling_rule; niter=n_iter, n_samples=neach, y=i)
-        plts_i = []
-        for j in 1:size(x, 2)
-            xj = x[:,j]
-            xj = reshape(xj, (n_digits, n_digits))
-            plts_i = [plts_i..., Plots.heatmap(rotl90(xj), axis=nothing, cb=false)]
-        end
-        plt = Plots.plot(plts_i..., size=(_w,0.10*_w), layout=(1,10))
-        plts = [plts..., plt]
-    end
-    plt = Plots.plot(plts..., size=(_w,_w), layout=(10,1))
-    display(plt)
-end
-```
-
-## Benchmark
-
-```{julia}
-# Benchmark generators:
-generators = Dict(
-    :wachter => GenericGenerator(opt=opt, λ=l2_λ),
-    :revise => REVISEGenerator(opt=opt, λ=l2_λ),
-    :greedy => GreedyGenerator(),
-)
-
-# Conformal Models:
-
-# Measures:
-measures = [
-    CounterfactualExplanations.distance,
-    ECCCo.distance_from_energy,
-    ECCCo.distance_from_targets,
-    CounterfactualExplanations.validity,
-]
-```

### 2023-04-25 paper/paper.tex
more work on mnist
work on paper

@@
 % ready for submission
-% \usepackage{neurips_2023}
+\usepackage{neurips_2023}
 
@@
 % [preprint] option:
-\usepackage[preprint]{neurips_2023}
+% \usepackage[preprint]{neurips_2023}
 
@@
 \usepackage{graphicx}
+\usepackage{algorithm}
+\usepackage{algpseudocode}
 
@@
 Unfortunately, this idea hinges on the assumption that the Black Box Model provides well-calibrated predictive uncertainty estimates.
-Our proposed methodology, which we will turn to next, relaxes this restriction.
+
+\subsection{From Fidelity to Model Conformity}
+
+Above we explained that since Counterfactual Explanations work directly with the Black Box model, the fidelity of explanations as we defined it earlier is not a concern.
+This may explain why research has primarily focused on other desiderata, most notably plausibility (Definition~\ref{def:plausible}).
+Enquiring about the plausibility of a counterfactual essentially boils down to the following question: `Is this counterfactual consistent with the underlying data'?
+To introduce this section, we posit a related, slightly more nuanced question: `Is this counterfactual consistent with what the model has learned about the underlying data'?
+We will argue that fidelity is not a sufficient evaluation measure to answer this question and propose a novel way to assess if Counterfactual Explanations conform with model behaviour.
+
+The word \textit{fidelity} stems from the Latin word `fidelis', which means `faithful, loyal, trustworthy' \citep{mw2023fidelity}.
+As we explained in Section~\ref{background}, model explanations are generally considered faithful if their corresponding predictions coincide with the predictions made by the model itself.
+Since this definition of faithfulness is not useful in the context of Counterfactual Explanations, we propose an adapted version:
+
+\begin{definition}[Conformal Counterfactuals]
+  \label{def:conformal}
+  Let $\mathcal{X}_{\theta}|\mathbf{y}^* = p_{\theta}(x|\mathbf{y}^*)$ denote the conditional distribution of $\mathbf{x}$ in the target class $\mathbf{y}^*$, where $\theta$ denotes the parameters of model $M_{\theta}$.
+Then for $\mathbf{x}^{\prime}$ to be considered a conformal counterfactual, we need: $\mathbf{x}^{\prime} \sim \mathcal{X}_{\theta}|\mathbf{y}^*$.
+\end{definition}
+
+In words, conformal counterfactuals conform with what the predictive model has learned about the input data $\mathbf{x}$.
+Since this definition works with distributional properties, it explicitly accounts for the multiplicity of explanations we discussed earlier.
+To assess counterfactuals with respect to Definition~\ref{def:conformal}, we need to be able to quantify the posterior conditional distribution $p_{\theta}(\mathbf{x}|\mathbf{y}^*)$.
+This is very much at the core of our proposed methodological framework, which reconciles the notions of plausibility and model conformity and which we will introduce next.
 
@@
   \begin{aligned}
-  \mathbf{Z}^\prime &= \arg \min_{\mathbf{Z}^\prime \in \mathcal{Z}^M} \left\{  {\text{yloss}(M_{\theta}(f(\mathbf{Z}^\prime)),\mathbf{y}^*)}+ \lambda {\text{cost}(f(\mathbf{Z}^\prime)) }  \right\} 
+  \mathbf{Z}^\prime= \arg \min_{\mathbf{Z}^\prime \in \mathcal{Z}^M}  &\{  {\text{yloss}(M_{\theta}(f(\mathbf{Z}^\prime)),\mathbf{y}^*)}+ \lambda_{1} {\text{dist}(f(\mathbf{Z}^\prime),\mathbf{x}) } \\
+  &+ \lambda_2 \text{dist}(f(\mathbf{Z}^\prime),\hat{\mathbf{x}}_{\theta}) + \lambda_3 \Omega(C_{\theta}(f(\mathbf{Z}^\prime);\alpha)) \} 
   \end{aligned} 
@@
 
-\begin{figure}
-  \includegraphics[width=\textwidth]{../artifacts/results/images/mnist_eccco.png}
-  \caption{ECCCos from Black Boxes. Counterfactuals for turning an 8 (eight) into a 3 (three): original image (left); }\label{fig:eccco}
-\end{figure}
+where $\hat{\mathbf{x}}_{\theta}$ denotes samples generated using SGLD (Equation~\ref{eq:sgld}) and $\text{dist}(\cdot)$ is a generic term for a distance metric.
+Our default choice for $\text{dist}(\cdot)$ is the Manhatten Distance since it enforces sparsity.
+
+The first two terms in Equation~\ref{eq:eccco} correspond to the counterfactual search objective defined in~\citet{wachter2017counterfactual} which merely penalises the distance of counterfactuals from their factual values.
+The additional two penalties in ECCCo ensure that counterfactuals conform with the model's generative property and lead to minimally uncertain predictions, respectively.
+The hyperparameters $\lambda_1, ..., \lambda_3$ can be used to balance the different objectives: for example, we may choose to incur larger deviations from the factual in favour of conformity with the model's generative property by choosing lower values of $\lambda_1$ and relatively higher values of $\lambda_2$.
+
+\begin{algorithm}
+  \caption{An algorithm with caption}\label{alg:cap}
+  \begin{minipage}[t]{0.45\textwidth}
+  \begin{algorithmic}
+    \Require $n \geq 0$
+    \Ensure $y = x^n$
+    \State $y \gets 1$
+    \State $X \gets x$
+    \State $N \gets n$
+    \While{$N \neq 0$}
+    \If{$N$ is even}
+        \State $X \gets X \times X$
+        \State $N \gets \frac{N}{2}$  \Comment{This is a comment}
+    \ElsIf{$N$ is odd}
+        \State $y \gets y \times X$
+        \State $N \gets N - 1$
+    \EndIf
+    \EndWhile
+  \end{algorithmic}
+  \end{minipage}
+\end{algorithm}
 
 \section{Evaluation Framework}\label{conformity}
-
-In Section~\ref{background} we explained that Counterfactual Explanations work directly with Black Box Model, so fidelity is not a concern.
-This may explain why research has primarily focused on other desiderata, most notably plausibility (Definition~\ref{def:plausible}).
-Enquiring about the plausibility of a counterfactual essentially boils down to the following question: `Is this counterfactual consistent with the underlying data'?
-To introduce this section, we posit a related, slightly more nuanced question: `Is this counterfactual consistent with what the model has learned about the underlying data'?
-We will argue that fidelity is not a sufficient evaluation measure to answer this question and propose a novel way to assess if explanations conform with model behaviour.
-Finally, we will introduce a framework for Conformal Counterfactual Explanations, that reconciles the notions of plausibility and model conformity.
-
-\subsection{From Fidelity to Model Conformity}
-
-The word \textit{fidelity} stems from the Latin word `fidelis', which means `faithful, loyal, trustworthy' \citep{mw2023fidelity}.
-As we explained in Section~\ref{background}, model explanations are considered faithful if their corresponding predictions coincide with the predictions made by the model itself.
-Since this definition of faithfulness is not useful in the context of Counterfactual Explanations, we propose an adapted version:
-
-\begin{definition}[Conformal Counterfactuals]
-  \label{def:conformal}
-  Let $\mathcal{X}_{\theta}|\mathbf{y}^* = p_{\theta}(x|\mathbf{y}^*)$ denote the conditional distribution of $\mathbf{x}$ in the target class $\mathbf{y}^*$, where $\theta$ denotes the parameters of model $M_{\theta}$.
-Then for $\mathbf{x}^{\prime}$ to be considered a conformal counterfactual, we need: $\mathbf{x}^{\prime} \sim \mathcal{X}_{\theta}|\mathbf{y}^*$.
-\end{definition}
-
-In words, conformal counterfactuals conform with what the predictive model has learned about the input data $\mathbf{x}$.
-Since this definition works with distributional properties, it explicitly accounts for the multiplicity of explanations we discussed earlier.
-Except for the posterior conditional distribution $p_{\theta}(\mathbf{x}|\mathbf{y}^*)$, we already have access to all the ingredients in Definition~\ref{def:conformal}.
-
-\textbf{TBD}
-
-\begin{itemize}
-  \item What exact sampler do we use? ImproperSGLD as in \citet{grathwohl2020your} seems to work best.
-\end{itemize}
 
@@
 
+\begin{figure}
+  \includegraphics[width=\textwidth]{../artifacts/results/images/mnist_eccco.png}
+  \caption{ECCCos from Black Boxes. Counterfactuals for turning an 8 (eight) into a 3 (three): original image (left); }\label{fig:eccco}
+\end{figure}
+
 \begin{itemize}
@@
   \item Batch size seems to impact quality of generated samples (at inference, but not so much during JEM training)
-  \item ECCCo is sensitive to optimizer (Adam works well), learning rate and distance metric (l1 works well)
+  \item ECCCo is sensitive to optimizer (Adam works well), learning rate and distance metric (l1 currently only one that works)
   \item SGLD takes time 

### 2023-04-26 paper/paper.tex
added pseudo code

@@
 \usepackage{amsthm}
+\usepackage{caption}
 \usepackage{graphicx}
@@
 
-\begin{algorithm}
-  \caption{An algorithm with caption}\label{alg:cap}
-  \begin{minipage}[t]{0.45\textwidth}
-  \begin{algorithmic}
-    \Require $n \geq 0$
-    \Ensure $y = x^n$
-    \State $y \gets 1$
-    \State $X \gets x$
-    \State $N \gets n$
-    \While{$N \neq 0$}
-    \If{$N$ is even}
-        \State $X \gets X \times X$
-        \State $N \gets \frac{N}{2}$  \Comment{This is a comment}
-    \ElsIf{$N$ is odd}
-        \State $y \gets y \times X$
-        \State $N \gets N - 1$
-    \EndIf
+\medskip
+
+\renewcommand{\algorithmicrequire}{\textbf{Input:}}
+\renewcommand{\algorithmicensure}{\textbf{Output:}}
+
+\begin{minipage}[c]{0.45\textwidth}
+  \centering
+  \includegraphics[width=\textwidth]{../artifacts/results/images/surrogate_gone_wrong.png}
+  \captionof{figure}{Using surrogates can improve plausibility, but also increases vulnerability.
+Counterfactuals for turning an 8 (eight) into a 3 (three): original image (left); counterfactual produced using REVISE \citep{joshi2019realistic} with a well-specified surrogate (centre); and a counterfactual produced using REVISE \citep{joshi2019realistic} with a poorly specified surrogate (right).} \label{fig:vae}
+\end{minipage}  
+\hfill
+\begin{minipage}[c]{0.45\textwidth}
+\captionof{algorithm}{Generating ECCCos (For more details, see Appendix~\ref{app:eccco})}\label{alg:eccco}
+\begin{algorithmic}[1]
+  \Require $\mathbf{x}, \mathbf{y}^*, M_{\theta}, f, \Lambda, \alpha, \mathcal{D}, T, \eta, m, M$ \linebreak where $M_{\theta}(\mathbf{x})\neq\mathbf{y}^*$
+  \Ensure $\mathbf{x}^\prime$
+  \State Initialize $\mathbf{z}^\prime \gets f^{-1}(\mathbf{x})$
+  \State Run \textit{SCP} for $M_{\theta}$ using $\mathcal{D}$
+  \State Generate buffer $\mathcal{B}$ of $M$ conditional samples $\hat{\mathbf{x}}_{\theta}|\mathbf{y}^*$ using SGLD (Equation~\ref{eq:sgld})
+  \State Initialize $t \gets 0$
+  \While{\textit{not converged} or $t < T$}
+  \State $\hat{\mathbf{x}}_{\theta, t} \gets \text{rand}(\mathcal{B},m)$
+  \State $\mathbf{z}^\prime \gets \mathbf{z}^\prime - \eta \nabla_{\mathbf{z}^\prime} \mathcal{L}(\mathbf{z}^\prime,\mathbf{y}^*,\hat{\mathbf{x}}_{\theta, t})$
+  \State $t \gets t+1$
   \EndWhile
+  \State $\mathbf{x}^\prime \gets f(\mathbf{z}^\prime)$
 \end{algorithmic}
 \end{minipage}
-\end{algorithm}
+
+\medskip
 
@@
 To see why this effect is exacerbated by small choices for $\alpha$ consider the case of $\alpha=0$, which requires that the true label is covered by the prediction set with probability equal to one.
+
+\subsection{Conformal Prediction}\label{app:eccco}
 

### 2023-04-26 notebooks/eccco.qmd
some work on illustrative example

@@
 ```
+
+```{julia}
+# Counteractual data:
+counterfactual_data = load_circles(2000)
+X = counterfactual_data.X
+y = counterfactual_data.y
+labels = counterfactual_data.output_encoder.labels
+input_dim, nobs = size(X)
+batch_size = Int(round(nobs/10))
+```
+
+```{julia}
+𝒟x = Normal()
+𝒟y = Categorical(ones(2) ./ 2)
+sampler = ConditionalSampler(𝒟x, 𝒟y, input_size=size(X)[1:end-1], batch_size=batch_size)
+clf = JointEnergyClassifier(
+    sampler;
+    builder=MLJFlux.MLP(hidden=(32, 32, 32,), σ=Flux.relu),
+    batch_size=batch_size,
+    finaliser=x -> x,
+    loss=Flux.Losses.logitcrossentropy,
+    jem_training_params=(
+        α=[1.0,1.0,1e-1],
+        verbosity=10,
+    ),
+)
+```
+
+```{julia}
+method = :simple_inductive
+cov = .95
+conf_model = conformal_model(clf; method=method, coverage=cov)
+mach = machine(conf_model, table(permutedims(X)), labels)
+fit!(mach)
+```
+
+```{julia}
+#| echo: false
+
+jem = mach.model.model.jem
+batch_size = mach.model.model.batch_size
+X = Float32.(matrix(X))
+if typeof(jem.sampler) <: ConditionalSampler
+
+    plts = []
+    for target in 1:2
+        X̂ = generate_conditional_samples(jem, batch_size, target; niter=200) 
+        ex = extrema(hcat(X,X̂), dims=2)
+        xlims = ex[1]
+        ylims = ex[2]
+        x1 = range(1.0f0.*xlims...,length=100)
+        x2 = range(1.0f0.*ylims...,length=100)
+        plt = Plots.contour(
+            x1, x2, (x, y) -> softmax(jem([x, y]))[target], 
+            fill=true, alpha=0.5, title="Target: $target", cbar=true,
+            xlims=xlims,
+            ylims=ylims,
+        )
+        Plots.scatter!(X[1,:], X[2,:], color=Int.(labels.refs), group=Int.(labels.refs), alpha=0.5)
+        Plots.scatter!(
+            X̂[1,:], X̂[2,:], 
+            color=repeat([target], size(X̂,2)), 
+            group=repeat([target], size(X̂,2)), 
+            shape=:star5, ms=10
+        )
+        push!(plts, plt)
+    end
+    plt = Plots.plot(plts..., layout=(1, 2), size=(2*500, 400))
+    display(plt)
+end
+```
+
+```{julia}
+M = ECCCo.ConformalModel(mach.model, mach.fitresult)
+factual_label = 0
+x_factual = reshape(X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
+target = 1
+factual = predict_label(M, counterfactual_data, x_factual)[1]
+
+# ECCCo:
+λ=[0.5,0.1,0.5]
+temp=0.5
+η=0.01
+
+# Generate counterfactuals using ECCCo generator:
+generator = ECCCoGenerator(
+    λ=λ, 
+    temp=temp, 
+    opt=Flux.Optimise.Adam(η),
+)
+
+ce = CounterfactualExplanation(x_factual, target, counterfactual_data, M, generator)
+```

### 2023-04-26 notebooks/mnist.qmd
more work on mnist
reran model with batchsize 10

@@
 # Search:
-generator = GenericGenerator(opt=opt)
+generic_generator = GenericGenerator(opt=opt)
 ce_wachter = generate_counterfactual(
-    x_factual, target, counterfactual_data, M, generator; 
+    x_factual, target, counterfactual_data, M, generic_generator; 
     decision_threshold=γ, max_iter=T,
@@
 )
-generator = GreedyGenerator(η=1.0)
+greedy_generator = GreedyGenerator(η=1.0)
 ce_jsma = generate_counterfactual(
-    x_factual, target, counterfactual_data, M, generator; 
+    x_factual, target, counterfactual_data, M, greedy_generator; 
     decision_threshold=γ, max_iter=T,
@@
 # Define generator:
-generator = REVISEGenerator(
+revise_generator = REVISEGenerator(
   opt = opt,
@@
 ce_strong = generate_counterfactual(
-    x_factual, target, counterfactual_data, M, generator; 
+    x_factual, target, counterfactual_data, M, revise_generator; 
     decision_threshold=γ, max_iter=T,
@@
 )
-counterfactual_data = deepcopy(counterfactual_data)
-counterfactual_data.generative_model = vae_weak
+counterfactual_data_weak = deepcopy(counterfactual_data)
+counterfactual_data_weak.generative_model = vae_weak
 ce_weak = generate_counterfactual(
-    x_factual, target, counterfactual_data, M, generator;
+    x_factual, target, counterfactual_data_weak, M, revise_generator;
     decision_threshold=γ, max_iter=T,
@@
     input_size=(input_dim,), 
-    batch_size=1,
+    batch_size=10,
 )
@@
 
+### Different Models
+
 ```{julia}
@@
 λ=[0.5,0.1,0.5]
-temp=0.1
+temp=0.5
 η=0.01
@@
 # Generate counterfactuals using ECCCo generator:
-generator = ECCCoGenerator(
+eccco_generator = ECCCoGenerator(
     λ=λ, 
@@
     ce = generate_counterfactual(
-        x_factual, target, counterfactual_data, mod, generator; 
+        x_factual, target, counterfactual_data, mod, eccco_generator; 
         decision_threshold=γ, max_iter=T,
@@
 
+### Different Generators
+
 ```{julia}
-Random.seed!(1234)
+# Setup:
+model = model_dict["JEM Ensemble"]
 
-# Set up search:
-factual_label = 9
-x_factual = reshape(counterfactual_data.X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
-target = 7
-factual = predict_label(M, counterfactual_data, x_factual)[1]
-γ = 0.9
-T = 100
+# Benchmark generators:
+generator_dict = Dict(
+    :wachter => generic_generator,
+    :revise => revise_generator,
+    :greedy => greedy_generator,
+    :eccco => eccco_generator,
+)
 
-η=1.0
-
-# Generate counterfactual using generic generator:
-generator = GenericGenerator(opt=Flux.Optimise.Adam(0.01),)
-ce_wachter = generate_counterfactual(
-    x_factual, target, counterfactual_data, M, generator; 
+ces = Dict()
+for (gen_name, gen) in generator_dict
+    ce = generate_counterfactual(
+        x_factual, target, counterfactual_data, model, gen; 
         decision_threshold=γ, max_iter=T,
@@
     )
-
-generator = GreedyGenerator(η=η)
-ce_jsma = generate_counterfactual(
-    x_factual, target, counterfactual_data, M, generator; 
-    decision_threshold=γ, max_iter=T,
-    initialization=:identity,
-    converge_when=:generator_conditions,
-)
-
-# ECCCo:
-λ=[0.1,0.1,0.1]
-temp=0.1
-
-# Generate counterfactual using ECCCo generator:
-generator = ECCCoGenerator(
-    λ=λ, 
-    temp=temp, 
-    opt=Flux.Optimise.Adam(0.01),
-)
-ce_conformal = generate_counterfactual(
-    x_factual, target, counterfactual_data, M, generator; 
-    decision_threshold=γ, max_iter=T,
-    initialization=:identity,
-    converge_when=:generator_conditions,
-)
-
-# Generate counterfactual using ECCCo generator:
-generator = ECCCoGenerator(
-    λ=λ, 
-    temp=temp, 
-    opt=CounterfactualExplanations.Generators.JSMADescent(η=η),
-)
-ce_conformal_jsma = generate_counterfactual(
-    x_factual, target, counterfactual_data, M, generator; 
-    decision_threshold=γ, max_iter=T,
-    initialization=:identity,
-    converge_when=:generator_conditions,
-)
+    ces[gen_name] = ce
+end
+plt_order = sortperm(collect(keys(ces)))
 
@@
 )
-plts = [p1]
 
-ces = [ce_wachter, ce_conformal, ce_jsma, ce_conformal_jsma]
-_names = ["Wachter", "ECCCo", "JSMA", "ECCCo-JSMA"]
-for x in zip(ces, _names)
-    ce, _name = (x[1],x[2])
-    x_factual = CounterfactualExplanations.counterfactual(ce)
+plts = []
+for (_name,ce) in ces
+    _x = CounterfactualExplanations.counterfactual(ce)
     _phat = target_probs(ce)
@@
     plt = Plots.plot(
-        convert2image(MNIST, reshape(x_factual,28,28)),
+        convert2image(MNIST, reshape(_x,28,28)),
         axis=nothing, 
@@
 end
+plts = plts[plt_order]
+plts = [p1, plts...]
 plt = Plots.plot(plts...; size=(img_height*length(plts),img_height), layout=(1,length(plts)))
 display(plt)
-savefig(plt, joinpath(www_path, "eccco_mnist.png"))
-```
-
-```{julia}
-# Random.seed!(1234)
-
-# Set up search:
-factual_label = 8
-x_factual = reshape(counterfactual_data.X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
-target = 3
-factual = predict_label(M, counterfactual_data, x_factual)[1]
-γ = 0.5
-T = 100
-
-# Generate counterfactual using generic generator:
-generator = GenericGenerator(opt=Flux.Optimise.Adam(),)
-ce_wachter = generate_counterfactual(
-    x_factual, target, counterfactual_data, M, generator; 
-    decision_threshold=γ, max_iter=T,
-    initialization=:identity,
-)
-
-generator = GreedyGenerator(η=1.0)
-ce_jsma = generate_counterfactual(
-    x_factual, target, counterfactual_data, M, generator; 
-    decision_threshold=γ, max_iter=T,
-    initialization=:identity,
-)
-
-# ECCCo:
-λ=[0.0,1.0]
-temp=0.5
-
-# Generate counterfactual using CCE generator:
-generator = CCEGenerator(
-    λ=λ, 
-    temp=temp, 
-    opt=Flux.Optimise.Adam(),
-)
-ce_conformal = generate_counterfactual(
-    x_factual, target, counterfactual_data, M, generator; 
-    decision_threshold=γ, max_iter=T,
-    initialization=:identity,
-    converge_when=:generator_conditions,
-)
-
-# Generate counterfactual using CCE generator:
-generator = CCEGenerator(
-    λ=λ, 
-    temp=temp, 
-    opt=CounterfactualExplanations.Generators.JSMADescent(η=1.0),
-)
-ce_conformal_jsma = generate_counterfactual(
-    x_factual, target, counterfactual_data, M, generator; 
-    decision_threshold=γ, max_iter=T,
-    initialization=:identity,
-    converge_when=:generator_conditions,
-)
-
-# Plot:
-p1 = Plots.plot(
-    convert2image(MNIST, reshape(x_factual,28,28)),
-    axis=nothing, 
-    size=(img_height, img_height),
-    title="Factual"
-)
-plts = [p1]
-
-ces = [ce_wachter, ce_conformal, ce_jsma, ce_conformal_jsma]
-_names = ["Wachter", "CCE", "JSMA", "CCE-JSMA"]
-for x in zip(ces, _names)
-    ce, _name = (x[1],x[2])
-    x = CounterfactualExplanations.counterfactual(ce)
-    _phat = target_probs(ce)
-    _title = "$_name (p̂=$(round(_phat[1]; digits=3)))"
-    plt = Plots.plot(
-        convert2image(MNIST, reshape(x,28,28)),
-        axis=nothing, 
-        size=(img_height, img_height),
-        title=_title
-    )
-    plts = [plts..., plt]
-end
-plt = Plots.plot(plts...; size=(img_height*length(plts),img_height), layout=(1,length(plts)))
-display(plt)
-savefig(plt, joinpath(www_path, "cce_mnist.png"))
-```
-
-```{julia}
-if M.model.model isa JointEnergyModels.JointEnergyClassifier
-    jem = M.model.model.jem
-    n_iter = 200
-    _w = 1500
-    plts = []
-    neach = 10
-    for i in 1:10
-        x = jem.sampler(jem.chain, jem.sampling_rule; niter=n_iter, n_samples=neach, y=i)
-        plts_i = []
-        for j in 1:size(x, 2)
-            xj = x[:,j]
-            xj = reshape(xj, (n_digits, n_digits))
-            plts_i = [plts_i..., Plots.heatmap(rotl90(xj), axis=nothing, cb=false)]
-        end
-        plt = Plots.plot(plts_i..., size=(_w,0.10*_w), layout=(1,10))
-        plts = [plts..., plt]
-    end
-    plt = Plots.plot(plts..., size=(_w,_w), layout=(10,1))
-    display(plt)
-end
+savefig(plt, joinpath(output_images_path, "mnist_eccco_benchmark.png"))
 ```

### 2023-04-26 paper/paper.tex
reran model with batchsize 10

@@
 Here $\text{yloss}$ denotes the primary loss function already introduced above and $\text{cost}$ is either a single penalty or a collection of penalties that are used to impose constraints through regularization.
-Following the convention in \citet{altmeyer2023endogenous} we use $\mathbf{Z}^\prime=\{ \mathbf{z}_m\}_M$ to denote the vector $M$-dimensional array of counterfactual states.
-This is to explicitly account for the fact that we can generate multiple counterfactuals $M$, as with DiCE \citep{mothilal2020explaining}, and may choose to traverse a latent representation $\mathcal{Z}$ of the feature space $\mathcal{X}$, as we will discuss further below.
+Following the convention in \citet{altmeyer2023endogenous} we use $\mathbf{Z}^\prime=\{ \mathbf{z}_m\}_M$ to denote the $M$-dimensional array of counterfactual states.
+This is to explicitly account for the fact that we can generate multiple counterfactuals $M$, as with DiCE \citep{mothilal2020explaining}, and may choose to traverse a latent encoding $\mathcal{Z}$ of the feature space $\mathcal{X}$ where we denote $f^{-1}: \mathcal{X} \mapsto \mathcal{Z}$.
+Encodings may involve simple feature transformations or more advanced techniques involving generative models, as we will discuss further below.
 
@@
 Consequently, CP removes the restriction on the family of predictive models, at the small cost of reserving a subset of the available data for calibration.
+This particular case of conformal prediction is referred to as Split Conformal Prediction (SCP) as it involves splitting the training data into a proper training dataset and a calibration dataset.
 Details concerning our implementation of Conformal Prediction can be found in Appendix~\ref{app-cp}.
@@
 The hyperparameters $\lambda_1, ..., \lambda_3$ can be used to balance the different objectives: for example, we may choose to incur larger deviations from the factual in favour of conformity with the model's generative property by choosing lower values of $\lambda_1$ and relatively higher values of $\lambda_2$.
+Figure~\ref{fig:eccco} illustrates this balancing act for an example involving synthetic data: vector fields indicate the direction of gradients with respect to the different components our proposed objective function (Equation~\ref{eq:eccco}).
+
+The entire procedure for Generating ECCCos is described in Algorithm~\ref{alg:eccco}.
+For the sake of simplicity and without loss of generality, we limit our attention to generating a single counterfactual $\mathbf{x}^\prime=f(\mathbf{z}^\prime)$ where in contrast to Equation~\ref{eq:eccco} $\mathbf{z}^\prime$ denotes a $1$-dimensional array containing a single counterfactual state.
+That state is initialized by passing the factual $\mathbf{x}$ through the encoder $f^{-1}$ which in our case corresponds to a simple feature transformer, rather than the encoder part of VAE as in REVISE~\citep{joshi2019realistic}.
 
@@
 
-\begin{minipage}[c]{0.45\textwidth}
+\begin{minipage}[c]{0.40\textwidth}
   \centering
-  \includegraphics[width=\textwidth]{../artifacts/results/images/surrogate_gone_wrong.png}
-  \captionof{figure}{Using surrogates can improve plausibility, but also increases vulnerability.
-Counterfactuals for turning an 8 (eight) into a 3 (three): original image (left); counterfactual produced using REVISE \citep{joshi2019realistic} with a well-specified surrogate (centre); and a counterfactual produced using REVISE \citep{joshi2019realistic} with a poorly specified surrogate (right).} \label{fig:vae}
+  \includegraphics[width=\textwidth]{../artifacts/results/images/eccco_illustration.png}
+  \captionof{figure}{Vector fields indicating the direction of gradients with respect to the different components of the ECCCo objective (Equation~\ref{eq:eccco}).} \label{fig:eccco}
 \end{minipage}  
 \hfill
-\begin{minipage}[c]{0.45\textwidth}
+\begin{minipage}[c]{0.50\textwidth}
 \captionof{algorithm}{Generating ECCCos (For more details, see Appendix~\ref{app:eccco})}\label{alg:eccco}
@@
   \State Initialize $\mathbf{z}^\prime \gets f^{-1}(\mathbf{x})$
+  \State Generate buffer $\mathcal{B}$ of $M$ conditional samples $\hat{\mathbf{x}}_{\theta}|\mathbf{y}^*$ using SGLD (Equation~\ref{eq:sgld})
   \State Run \textit{SCP} for $M_{\theta}$ using $\mathcal{D}$
-  \State Generate buffer $\mathcal{B}$ of $M$ conditional samples $\hat{\mathbf{x}}_{\theta}|\mathbf{y}^*$ using SGLD (Equation~\ref{eq:sgld})
   \State Initialize $t \gets 0$

### 2023-04-26 notebooks/mnist.qmd
attempting a run without normalising digits to between -1 and 1

@@
     ϵ = Float32.(randn(size(x)) * noise)
-    x = @.(2 * x - 1) .+ ϵ
+    # x = @.(2 * x - 1) 
+    x += ϵ
     return x
@@
 # Hyper:
-_retrain = false
-_regen = false
+_retrain = true
+_regen = true
 
@@
 # JEM parameters:
-𝒟x = Uniform(-1,1)
+𝒟x = Uniform(0,1)
 𝒟y = Categorical(ones(output_dim) ./ output_dim)
@@
     test_data = load_mnist_test()
-    test_data.X = pre_process.(test_data.X, noise=0.0f0)
+    # test_data.X = pre_process.(test_data.X, noise=0.0f0)
     _perf = CounterfactualExplanations.Models.model_evaluation(mod, test_data, measure=collect(values(measure)))

### 2023-04-27 paper/paper.tex
attempting a run without normalising digits to between -1 and 1

@@
 
-The entire procedure for Generating ECCCos is described in Algorithm~\ref{alg:eccco}.
-For the sake of simplicity and without loss of generality, we limit our attention to generating a single counterfactual $\mathbf{x}^\prime=f(\mathbf{z}^\prime)$ where in contrast to Equation~\ref{eq:eccco} $\mathbf{z}^\prime$ denotes a $1$-dimensional array containing a single counterfactual state.
-That state is initialized by passing the factual $\mathbf{x}$ through the encoder $f^{-1}$ which in our case corresponds to a simple feature transformer, rather than the encoder part of VAE as in REVISE~\citep{joshi2019realistic}.
-
 \medskip
@@
 \begin{algorithmic}[1]
-  \Require $\mathbf{x}, \mathbf{y}^*, M_{\theta}, f, \Lambda, \alpha, \mathcal{D}, T, \eta, m, M$ \linebreak where $M_{\theta}(\mathbf{x})\neq\mathbf{y}^*$
+  \Require $\mathbf{x}, \mathbf{y}^*, M_{\theta}, f, \Lambda, \alpha, \mathcal{D}, T, \eta, n_{\mathcal{B}}, N_{\mathcal{B}}$ \linebreak where $M_{\theta}(\mathbf{x})\neq\mathbf{y}^*$
   \Ensure $\mathbf{x}^\prime$
   \State Initialize $\mathbf{z}^\prime \gets f^{-1}(\mathbf{x})$
-  \State Generate buffer $\mathcal{B}$ of $M$ conditional samples $\hat{\mathbf{x}}_{\theta}|\mathbf{y}^*$ using SGLD (Equation~\ref{eq:sgld})
+  \State Generate buffer $\mathcal{B}$ of $N_{\mathcal{B}}$ conditional samples $\hat{\mathbf{x}}_{\theta}|\mathbf{y}^*$ using SGLD (Equation~\ref{eq:sgld})
   \State Run \textit{SCP} for $M_{\theta}$ using $\mathcal{D}$
@@
   \While{\textit{not converged} or $t < T$}
-  \State $\hat{\mathbf{x}}_{\theta, t} \gets \text{rand}(\mathcal{B},m)$
+  \State $\hat{\mathbf{x}}_{\theta, t} \gets \text{rand}(\mathcal{B},n_{\mathcal{B}})$
   \State $\mathbf{z}^\prime \gets \mathbf{z}^\prime - \eta \nabla_{\mathbf{z}^\prime} \mathcal{L}(\mathbf{z}^\prime,\mathbf{y}^*,\hat{\mathbf{x}}_{\theta, t})$
@@
 
-\section{Evaluation Framework}\label{conformity}
+\begin{minipage}[c]{\textwidth}
+  \includegraphics[width=\textwidth]{../artifacts/results/images/mnist_eccco.png}
+  \captionof{figure}{Original image (left) and ECCCos for turning an 8 (eight) into a 3 (three) for different Black Boxes from left to right: Multi-Layer Perceptron (MLP), Ensemble of MLPs, Joint Energy Model (JEM), Ensemble of JEMs.}\label{fig:eccco-mnist}
+\end{minipage}
+
+\medskip
+
+The entire procedure for Generating ECCCos is described in Algorithm~\ref{alg:eccco}.
+For the sake of simplicity and without loss of generality, we limit our attention to generating a single counterfactual $\mathbf{x}^\prime=f(\mathbf{z}^\prime)$ where in contrast to Equation~\ref{eq:eccco} $\mathbf{z}^\prime$ denotes a $1$-dimensional array containing a single counterfactual state.
+That state is initialized by passing the factual $\mathbf{x}$ through the encoder $f^{-1}$ which in our case corresponds to a simple feature transformer, rather than the encoder part of VAE as in REVISE~\citep{joshi2019realistic}.
+Next, we generate a buffer of $N_{\mathcal{B}}$ conditional samples $\hat{\mathbf{x}}_{\theta}|\mathbf{y}^*$ using SGLD (Equation~\ref{eq:sgld}) and conformalise the model $M_{\theta}$ through Split Conformal Prediction on training data $\mathcal{D}$.
+
+Finally, we search counterfactuals through gradient descent.
+Let $\mathcal{L}(\mathbf{z}^\prime,\mathbf{y}^*,\hat{\mathbf{x}}_{\theta, t})$ denote our loss function defined in Equation~\ref{eq:eccco}.
+Then in each iteration, we first randomly draw $n_{\mathcal{B}}$ samples from the buffer $\mathcal{B}$ before updating the counterfactual state $\mathbf{z}^\prime$ by moving in the negative direction of that loss function.
+The search terminates once the convergence criterium is met or the maximum number of iterations $T$ has been exhausted.
+Note that the choice of convergence criterium has important implications on the final counterfactual.
+For more detail on this see Appendix~\ref{app:eccco}).
+
+Figure~\ref{fig:eccco-mnist} presents ECCCos for the MNIST example from Section~\ref{background} for various Black Box models of increasing complexity from left to right: a simple Multi-Layer Perceptron (MLP); an Ensemble of MLPs, each of the same architecture as the single MLP; a Joint Energy Model (JEM) based on the same MLP architecture; and finally, an Ensemble of these JEMs.
+Since Deep Ensembles have an improved capacity for predictive uncertainty quantification and JEMs are explicitly trained to learn plausible representations of the input data, it is intuitive to see that the plausibility of counterfactuals visibly improves from left to right.
+
+\section{Experiments}\label{conformity}
 
@@
 
-\begin{figure}
-  \includegraphics[width=\textwidth]{../artifacts/results/images/mnist_eccco.png}
-  \caption{ECCCos from Black Boxes. Counterfactuals for turning an 8 (eight) into a 3 (three): original image (left); }\label{fig:eccco}
-\end{figure}
-
 \begin{itemize}
@@
   \item REVISE has benefit of lower dimensional space
+  \item For MNIST it seems that ECCCo is better at reducing pixel values than increasing them (better at erasing than writing)
 \end{itemize}

### 2023-04-27 notebooks/mnist.qmd

--- 2023-04-26 notebooks/mnist.qmd
@@
     ϵ = Float32.(randn(size(x)) * noise)
-    x = @.(2 * x - 1) .+ ϵ
+    x = @.(2 * x - 1) 
+    x += ϵ
     return x
@@
 ```{julia}
+plt_order = ["MLP", "MLP Ensemble", "JEM", "JEM Ensemble"]
+
 # ECCCo:
@@
 end
-plt_order = sortperm(collect(keys(ces)))
+_plt_order = map(x -> findall(collect(keys(model_dict)) .== x)[1], plt_order)
 
@@
 end
-plts = plts[plt_order]
+plts = plts[_plt_order]
 plts = [p1, plts...]

### 2023-04-27 notebooks/mnist.qmd
for 0 to 1 counterfactuals are looking a bit to homogenous

--- 2023-04-26 notebooks/mnist.qmd
@@
     ϵ = Float32.(randn(size(x)) * noise)
-    x = @.(2 * x - 1) .+ ϵ
+    x += ϵ
     return x
@@
 # Data:
-n_obs = 10000
+n_obs = 1000
 counterfactual_data = load_mnist(n_obs)
@@
 )
-α = [1.0,1.0,1e-2]      # penalty strengths
+α = [1.0,1.0,1e-1]      # penalty strengths
 ```
@@
     test_data = load_mnist_test()
-    test_data.X = pre_process.(test_data.X, noise=0.0f0)
     _perf = CounterfactualExplanations.Models.model_evaluation(mod, test_data, measure=collect(values(measure)))
@@
 ```{julia}
-# ECCCo:
-λ=[0.5,0.1,0.5]
-temp=0.5
-η=0.01
+function _plot_eccco_mnist(
+    x::Union{AbstractArray, Int}=x_factual, target::Int=target;
+    λ=[0.1,0.1,0.1],
+    temp=0.1,η=0.01,
+    plt_order = ["MLP", "MLP Ensemble", "JEM", "JEM Ensemble"],
+    opt = Flux.Optimise.Adam(η),
+    rng::Union{Int,AbstractRNG}=Random.GLOBAL_RNG,
+)
+
+    # Setup:
+    Random.seed!(rng)
+    if x isa Int
+        x = reshape(counterfactual_data.X[:,rand(findall(labels.==x))],input_dim,1)
+    end
 
@@
         temp=temp, 
-    opt=Flux.Optimise.Adam(η),
+        opt=opt,
     )
@@
         ce = generate_counterfactual(
-        x_factual, target, counterfactual_data, mod, eccco_generator; 
+            x, target, counterfactual_data, mod, eccco_generator; 
             decision_threshold=γ, max_iter=T,
@@
     end
-plt_order = sortperm(collect(keys(ces)))
+    _plt_order = map(x -> findall(collect(keys(model_dict)) .== x)[1], plt_order)
 
@@
     p1 = Plots.plot(
-    convert2image(MNIST, reshape(x_factual,28,28)),
+        convert2image(MNIST, reshape(x,28,28)),
         axis=nothing, 
@@
     end
-plts = plts[plt_order]
+    plts = plts[_plt_order]
     plts = [p1, plts...]
     plt = Plots.plot(plts...; size=(img_height*length(plts),img_height), layout=(1,length(plts)))
+
+    return plt, eccco_generator
+end
+```
+
+```{julia}
+plt, eccco_generator = _plot_eccco_mnist()
 display(plt)
 savefig(plt, joinpath(output_images_path, "mnist_eccco.png"))
+```
+
+### All digits
+
+```{julia}
+function plot_mnist(
+    factual::Int,target::Int;
+    generator::AbstractGenerator,
+    model::AbstractFittedModel=model_dict["JEM Ensemble"],
+    data::CounterfactualData=counterfactual_data,
+    rng::Union{Int,AbstractRNG}=Random.GLOBAL_RNG,
+    _plot_title::Bool=true,
+    kwargs...,
+)
+    decision_threshold = !isdefined(kwargs, :decision_threshold) ? 0.9 : decision_threshold
+    max_iter = !isdefined(kwargs, :max_iter) ? 100 : max_iter
+    initialization = !isdefined(kwargs, :initialization) ? :identity : initialization
+    converge_when = !isdefined(kwargs, :converge_when) ? :generator_conditions : converge_when
+
+    x = reshape(data.X[:,rand(findall(predict_label(model, data).==factual))],input_dim,1)
+    ce = generate_counterfactual(
+        x, target, data, model, generator; 
+        decision_threshold=decision_threshold, max_iter=max_iter,
+        initialization=initialization,
+        converge_when=converge_when,
+        kwargs...
+    )
+
+    _title = _plot_title ? "$(factual) -> $(target)" : ""
+
+    _x = CounterfactualExplanations.counterfactual(ce)
+    plt = Plots.plot(
+        convert2image(MNIST, reshape(_x,28,28)),
+        axis=nothing, 
+        size=(img_height, img_height),
+        title=_title
+    )
+    return plt
+end
+```
+
+```{julia}
+if _regen
+    function plot_all_digits(rng=1;verbose=true,kwargs...)
+        plts = []
+        for i in 0:9
+            for j in 0:9
+                @info "Generating counterfactual for $(i) -> $(j)"
+                plt = plot_mnist(i,j;kwargs...,rng=rng)
+                !verbose || display(plt)
+                plts = [plts..., plt]
+            end
+        end
+        plt = Plots.plot(plts...; size=(img_height*10,img_height*10), layout=(10,10))
+        return plt
+    end
+    plt = plot_all_digits(generator=eccco_generator)
+    savefig(plt, joinpath(output_images_path, "mnist_eccco_all_digits.png"))
+end
 ```
@@
 ```{julia}
-# Benchmark generators:
-generators = Dict(
-    :wachter => GenericGenerator(opt=opt, λ=l2_λ),
-    :revise => REVISEGenerator(opt=opt, λ=l2_λ),
-    :greedy => GreedyGenerator(),
-)
-
-# Conformal Models:
-
 # Measures:
@@
     CounterfactualExplanations.validity,
+    CounterfactualExplanations.redudancy,
 ]

### 2023-04-27 src/generator.jl

@@ -28,7 +28,7 @@ function ECCCoGenerator(;
     function _set_size_penalty(ce::AbstractCounterfactualExplanation)
         return ECCCo.set_size_penalty(ce; κ=κ, temp=temp)
     end
-    _penalties = [Objectives.distance_l2, _set_size_penalty, ECCCo.distance_from_energy]
+    _penalties = [Objectives.distance_l1, _set_size_penalty, ECCCo.distance_from_energy]
     λ = λ isa AbstractFloat ? [0.0, λ, λ] : λ
     return Generator(; penalty=_penalties, λ=λ, opt=opt, kwargs...)
 end

### 2023-04-27 notebooks/mnist.qmd
mnist benchmark set up

@@
 # Data:
-n_obs = 1000
+n_obs = 10000
 counterfactual_data = load_mnist(n_obs)
 counterfactual_data.X = pre_process.(counterfactual_data.X)
+counterfactual_data.generative_model = vae
 X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
@@
 )
-α = [1.0,1.0,1e-1]      # penalty strengths
+α = [1.0,1.0,1e-2]      # penalty strengths
 ```
@@
 # Plot generated samples:
+n_regen = 150
 if _regen 
@@
 
-        n_iter = 200
         _w = 1500
@@
         for i in 1:10
-            x = sampler(f, opt; niter=n_iter, n_samples=neach, y=i)
+            x = sampler(f, opt; niter=n_regen, n_samples=neach, y=i)
             plts_i = []
@@
     x::Union{AbstractArray, Int}=x_factual, target::Int=target;
-    λ=[0.1,0.1,0.1],
+    λ=[0.5,0.1,0.5],
     temp=0.1,η=0.01,
@@
     ECCCo.distance_from_targets,
-    CounterfactualExplanations.validity,
-    CounterfactualExplanations.redudancy,
+    CounterfactualExplanations.Evaluation.validity,
+    CounterfactualExplanations.Evaluation.redundancy,
 ]
+
+bmk = benchmark(
+    counterfactual_data; 
+    models=model_dict, 
+    generators=generator_dict, 
+    measure=measures,
+    suppress_training=true, dataname="MNIST",
+    n_individuals=100,
+    initialization=:identity,
+)
+
+Serialization.serialize(joinpath(output_path, "mnist_benchmark.jls"), bmk)
 ```
+
+```{julia}
+@chain bmk() begin
+    @group_by(dataname, generator, model, variable)
+    @summarize(mean=mean(value),sd=std(value))
+    @ungroup
+    @filter(variable == "distance_from_targets")
+end
+```

### 2023-04-27 src/penalties.jl

@@ -38,7 +38,7 @@ end
 
 function distance_from_energy(
     ce::AbstractCounterfactualExplanation;
-    n::Int=10, niter=100, from_buffer=true, agg=mean, kwargs...
+    n::Int=10, niter=60, from_buffer=true, agg=mean, kwargs...
 )
     conditional_samples = []
     ignore_derivatives() do
@@ -47,7 +47,7 @@ function distance_from_energy(
             _dict[:energy_sampler] = ECCCo.EnergySampler(ce; niter=niter, nsamples=n, kwargs...)
         end
         sampler = _dict[:energy_sampler]
-        push!(conditional_samples, rand(sampler, 100; from_buffer=from_buffer))
+        push!(conditional_samples, rand(sampler, n; from_buffer=from_buffer))
     end
     x′ = CounterfactualExplanations.counterfactual(ce)
     loss = map(eachslice(x′, dims=ndims(x′))) do x
@@ -64,16 +64,15 @@ end
 
 function distance_from_targets(
     ce::AbstractCounterfactualExplanation;
-    n::Int=1000, agg=mean
+    n::Int=100, agg=mean
 )
     target_idx = ce.data.output_encoder.labels .== ce.target
     target_samples = ce.data.X[:,target_idx] |>
         X -> X[:,rand(1:end,n)]
     x′ = CounterfactualExplanations.counterfactual(ce)
-    loss = map(eachslice(x′, dims=3)) do x
-        x = Matrix(x)
+    loss = map(eachslice(x′, dims=ndims(x′))) do x
         Δ = map(eachcol(target_samples)) do xsample
-            norm(x - xsample)
+            norm(x - xsample, 1)
         end
         return mean(Δ)
     end

### 2023-04-28 paper/paper.tex
paper

@@
 To do so, \textbf{ECCCo} leverages recent advances in energy-based modelling and predictive uncertainty quantification through conformal inference.
-Through illustrative examples and extensive empirical studies, we demonstrate that ECCos reconcile the need for plausibility and model conformity.
+Through illustrative examples and extensive empirical studies, we demonstrate that ECCCos reconcile the need for plausibility and model conformity.
 \end{abstract}
@@
 Since Deep Ensembles have an improved capacity for predictive uncertainty quantification and JEMs are explicitly trained to learn plausible representations of the input data, it is intuitive to see that the plausibility of counterfactuals visibly improves from left to right.
+This provides some first anecdotal evidence that ECCCos achieve plausibility while maintaining faithfulness to the Black Box.
 
 \section{Experiments}\label{conformity}
+
+In this section, we bolster our anecdotal findings from the previous section through rigorous empirical analysis.
+We first briefly describe our evaluation framework and data, before presenting and discussing our results.
 

### 2023-04-28 notebooks/mnist.qmd
mnist benchmark

@@
     suppress_training=true, dataname="MNIST",
-    n_individuals=100,
+    n_individuals=5,
+    factual=0, target=1,
     initialization=:identity,
@@
     @ungroup
-    @filter(variable == "distance_from_targets")
+    @filter(variable == "distance_from_energy")
 end

### 2023-04-28 notebooks/mnist.qmd

@@
 display(plt)
-savefig(plt, joinpath(output_images_path, "mnist_eccco_benchmark.png"))
+savefig(plt, joinpath(output_images_path, "mnist_all_generators.png"))
 ```
@@
 ```
+
+```{julia}
+df = @chain bmk() begin
+    @filter(variable in [
+        "distance_from_energy",
+        "distance_from_targets",
+        "distance",])
+    @mutate(variable = ifelse.(variable .== "distance_from_energy", "Non-Conformity", variable))
+    @mutate(variable = ifelse.(variable .== "distance_from_targets", "Implausibility", variable))
+    @mutate(variable = ifelse.(variable .== "distance", "Cost", variable))
+end
+plt = AlgebraOfGraphics.data(df) * visual(BoxPlot) * 
+    mapping(:generator, :value, row=:variable, col=:model, color=:generator)
+plt = draw(
+    plt, axis=(xlabel="", xticksvisible=false, xticklabelsvisible=false, width=150, height=120), 
+    facet=(; linkyaxes=:minimal)
+)   
+display(plt)
+save(joinpath(output_images_path, "mnist_benchmark.png"), plt, px_per_unit=5)
+```

### 2023-04-28 paper/paper.tex

@@
   \includegraphics[width=\textwidth]{../artifacts/results/images/eccco_illustration.png}
-  \captionof{figure}{Vector fields indicating the direction of gradients with respect to the different components of the ECCCo objective (Equation~\ref{eq:eccco}).} \label{fig:eccco}
+  \captionof{figure}{[PLACEHOLDER] Vector fields indicating the direction of gradients with respect to the different components of the ECCCo objective (Equation~\ref{eq:eccco}).} \label{fig:eccco}
 \end{minipage}  
@@
   \includegraphics[width=\textwidth]{../artifacts/results/images/mnist_eccco.png}
-  \captionof{figure}{Original image (left) and ECCCos for turning an 8 (eight) into a 3 (three) for different Black Boxes from left to right: Multi-Layer Perceptron (MLP), Ensemble of MLPs, Joint Energy Model (JEM), Ensemble of JEMs.}\label{fig:eccco-mnist}
+  \captionof{figure}{[SUBJECTO TO CHANGE] Original image (left) and ECCCos for turning an 8 (eight) into a 3 (three) for different Black Boxes from left to right: Multi-Layer Perceptron (MLP), Ensemble of MLPs, Joint Energy Model (JEM), Ensemble of JEMs.}\label{fig:eccco-mnist}
 \end{minipage}
@@
 
-The entire procedure for Generating ECCCos is described in Algorithm~\ref{alg:eccco}.
+The entire procedure for generating ECCCos is described in Algorithm~\ref{alg:eccco}.
 For the sake of simplicity and without loss of generality, we limit our attention to generating a single counterfactual $\mathbf{x}^\prime=f(\mathbf{z}^\prime)$ where in contrast to Equation~\ref{eq:eccco} $\mathbf{z}^\prime$ denotes a $1$-dimensional array containing a single counterfactual state.
@@
 
-\section{Experiments}\label{conformity}
+\section{Empirical Analysis}\label{emp}
 
@@
 
-\section{Experiments}
+\subsection{Data}
+
+\subsection{Results}
+
+\begin{figure}
+  \includegraphics[width=\textwidth]{../artifacts/results/images/mnist_benchmark.png}
+  \caption{[SUBJECTO TO CHANGE] Original image (left) and ECCCos for turning an 8 (eight) into a 3 (three) for different Black Boxes from left to right: Multi-Layer Perceptron (MLP), Ensemble of MLPs, Joint Energy Model (JEM), Ensemble of JEMs.}\label{fig:mnist-benchmark}
+\end{figure}
+
+\section{Discussion}
+
+\subsection{Key Insights}
+
+Consistent with the findings in \citet{schut2021generating}, we have demonstrated that predictive uncertainty estimates can be leveraged to generate plausible counterfactuals.
+Interestingly, \citet{schut2021generating} point out that this finding --- as intuitive as it is --- may be linked to a positive connection between the generative task and predictive uncertainty quantification.
+In particular, \citet{grathwohl2020your} demonstrate that their proposed method for integrating the generative objective in training yields models that have improved predictive uncertainty quantification.
+Since neither \citet{schut2021generating} nor we have employed any surrogate generative models, our findings seem to indicate that the positive connection found in \citet{grathwohl2020your} is bidirectional.
+
+\subsection{Limitations}
 
@@
 
-\section{Discussion}
-
-Consistent with the findings in \citet{schut2021generating}, we have demonstrated that predictive uncertainty estimates can be leveraged to generate plausible counterfactuals.
-Interestingly, \citet{schut2021generating} point out that this finding --- as intuitive as it is --- may be linked to a positive connection between the generative task and predictive uncertainty quantification.
-In particular, \citet{grathwohl2020your} demonstrate that their proposed method for integrating the generative objective in training yields models that have improved predictive uncertainty quantification.
-Since neither \citet{schut2021generating} nor we have employed any surrogate generative models, our findings seem to indicate that the positive connection found in \citet{grathwohl2020your} is bidirectional.
+\section{Conclusion}
 

### 2023-04-30 paper/paper.tex
paper

--- 2023-04-28 paper/paper.tex
@@
 Counterfactual Explanations provide a powerful, flexible and intuitive way to not only explain Black Box Models but also enable affected individuals to challenge them through the means of Algorithmic Recourse.
-Instead of opening the black box, Counterfactual Explanations work under the premise of strategically perturbing model inputs to understand model behaviour \citep{wachter2017counterfactual}.
+Instead of opening the black box, Counterfactual Explanations work under the premise of strategically perturbing model inputs to understand model behaviour~\citep{wachter2017counterfactual}.
 Intuitively speaking, we generate explanations in this context by asking simple what-if questions of the following nature: `Our credit risk model currently predicts that this individual's credit profile is too risky to offer them a loan.
@@
 Counterfactuals are then searched by minimizing a loss function that compares the predicted model output to the target outcome: $\text{yloss}(M_{\theta}(\mathbf{x}),\mathbf{y}^*)$.
-Since Counterfactual Explanations (CE) work directly with the Black Box Model, valid counterfactuals always have full local fidelity by construction \citep{mothilal2020explaining}.
+Since Counterfactual Explanations (CE) work directly with the Black Box Model, valid counterfactuals always have full local fidelity by construction~\citep{mothilal2020explaining}.
 Fidelity is defined as the degree to which explanations approximate the predictions of the Black Box Model.
-This is arguably one of the most important evaluation metrics for model explanations, since any explanation that explains a prediction not actually made by the model is useless \citep{molnar2020interpretable}.
+This is arguably one of the most important evaluation metrics for model explanations, since any explanation that explains a prediction not actually made by the model is useless~\citep{molnar2020interpretable}.
 
-In situations where full fidelity is a requirement, CE therefore offers a more appropriate solution to Explainable Artificial Intelligence (XAI) than other popular approaches like LIME \citep{ribeiro2016why} and SHAP \citep{lundberg2017unified}, which involve local surrogate models.
+In situations where full fidelity is a requirement, CE therefore offers a more appropriate solution to Explainable Artificial Intelligence (XAI) than other popular approaches like LIME~\citep{ribeiro2016why} and SHAP~\citep{lundberg2017unified}, which involve local surrogate models.
 But even full fidelity is not a sufficient condition for ensuring that an explanation adequately describes the behaviour of a model.

### 2023-04-30 paper/paper.tex
addressed cynthia's comments

--- 2023-04-28 paper/paper.tex
@@
 \begin{abstract}
-  Counterfactual Explanations offer an intuitive and straightforward way to explain Black Box Models but they are not unique.
+  Counterfactual Explanations offer an intuitive and straightforward way to explain black-box models but they are not unique.
 To address the need for plausible explanations, existing work has primarily relied on surrogate models to learn how the input data is distributed.
 This effectively reallocates the task of learning realistic representations of the data from the model itself to the surrogate.
-Consequently, the generated explanations may look plausible to humans but not necessarily conform with the behaviour of the Black Box Model.
+Consequently, the generated explanations may look plausible to humans but not necessarily conform with the behaviour of the black-box model.
 We formalise this notion of model conformity through the introduction of tailored evaluation measures and propose a novel algorithmic framework for generating \textbf{E}nergy-\textbf{C}onstrained \textbf{C}onformal \textbf{Co}unterfactuals that are only as plausible as the model permits.
@@
 
-Counterfactual Explanations provide a powerful, flexible and intuitive way to not only explain Black Box Models but also enable affected individuals to challenge them through the means of Algorithmic Recourse.
-Instead of opening the black box, Counterfactual Explanations work under the premise of strategically perturbing model inputs to understand model behaviour \citep{wachter2017counterfactual}.
+Counterfactual Explanations provide a powerful, flexible and intuitive way to not only explain black-box models but also enable affected individuals to challenge them through the means of Algorithmic Recourse.
+Instead of opening the black box, Counterfactual Explanations work under the premise of strategically perturbing model inputs to understand model behaviour~\citep{wachter2017counterfactual}.
 Intuitively speaking, we generate explanations in this context by asking simple what-if questions of the following nature: `Our credit risk model currently predicts that this individual's credit profile is too risky to offer them a loan.
@@
 Counterfactuals are then searched by minimizing a loss function that compares the predicted model output to the target outcome: $\text{yloss}(M_{\theta}(\mathbf{x}),\mathbf{y}^*)$.
-Since Counterfactual Explanations (CE) work directly with the Black Box Model, valid counterfactuals always have full local fidelity by construction \citep{mothilal2020explaining}.
-Fidelity is defined as the degree to which explanations approximate the predictions of the Black Box Model.
-This is arguably one of the most important evaluation metrics for model explanations, since any explanation that explains a prediction not actually made by the model is useless \citep{molnar2020interpretable}.
+Since Counterfactual Explanations (CE) work directly with the black-box model, valid counterfactuals always have full local fidelity by construction~\citep{mothilal2020explaining}.
+Fidelity is defined as the degree to which explanations approximate the predictions of the black-box model.
+This is arguably one of the most important evaluation metrics for model explanations, since any explanation that explains a prediction not actually made by the model is useless~\citep{molnar2020interpretable}.
 
-In situations where full fidelity is a requirement, CE therefore offers a more appropriate solution to Explainable Artificial Intelligence (XAI) than other popular approaches like LIME \citep{ribeiro2016why} and SHAP \citep{lundberg2017unified}, which involve local surrogate models.
-But even full fidelity is not a sufficient condition for ensuring that an explanation adequately describes the behaviour of a model.
-That is because two very distinct explanations can both lead to the same model prediction, especially when dealing with heavily parameterized models:
-
-\begin{quotation}
-  […] deep neural networks are typically very underspecified by the available data, and […] parameters [therefore] correspond to a diverse variety of compelling explanations for the data.
-  --- \citet{wilson2020case}
-\end{quotation}
-
-When people talk about Black Box Models, this is usually the type of model they have in mind.
+In situations where full fidelity is a requirement, CE therefore offers a more appropriate solution to Explainable Artificial Intelligence (XAI) than other popular approaches like LIME~\citep{ribeiro2016why} and SHAP~\citep{lundberg2017unified}, which involve local surrogate models.
+But even full fidelity is not a sufficient condition for ensuring that an explanation faithfully describes the behaviour of a model.
+That is because multiple very distinct explanations can all lead to the same model prediction, especially when dealing with heavily parameterized models like deep neural networks which are typically underspecified by the available data~\citep{wilson2020case}.
 
 In the context of CE, the idea that no two explanations are the same arises almost naturally.
-Even the baseline approach proposed by \citet{wachter2017counterfactual} can yield a diverse set of explanations if counterfactuals are initialised randomly.
-This multiplicity of explanations has not only been acknowledged in the literature but positively embraced: since individuals seeking Algorithmic Recourse (AR) have unique preferences,~\citet{mothilal2020explaining}, for example, have prescribed \textit{diversity} as an explicit goal for counterfactuals.
-More generally, the literature on CE and AR has brought forward a myriad of desiderata for explanations, which we will discuss in more detail in the following section.
+A key focus in the literature has therefore been to identify those explanations and algorithmic recourses that are deemed most appropriate based on a myriad of desiderata such as sparsity, actionability and plausibility.
+In this work, we draw closer attention to the insufficiency of model fidelity as an evaluation metric for the faithfulness of counterfactual explanations.
+Our key contributions are as follows: firstly, we introduce a new notion of faithfulness that is suitable for counterfactuals and propose a novel evaluation measure that draws inspiration from recent advances in Energy-Based Modelling (EBM); secondly, we a novel algorithmic approach for generating Energy-Constrained Conformal Counterfactuals (ECCCo) that explicitly address the need for faithfulness; finally, we provide illustrative examples and extensive empirical evidence demonstrating that ECCCos faithfully explain model behaviour without sacrificing existing desidarata like plausibility and sparsity.
 
@@
 In this section, we provide some background on Counterfactual Explanations and our motivation for this work.
-To start off, we briefly introduce the methodology uncerlying most state-of-the-art (SOTA) counterfactual generators.
+To start, we briefly introduce the methodology underlying most state-of-the-art (SOTA) counterfactual generators.
 
@@
 
-While Counterfactual Explanations can be generated for arbitrary regression models \citep{spooner2021counterfactual}, existing work has primarily focused on classification problems.
+While Counterfactual Explanations can be generated for arbitrary regression models~\citep{spooner2021counterfactual}, existing work has primarily focused on classification problems.
 Let $\mathcal{Y}=(0,1)^K$ denote the one-hot-encoded output domain with $K$ classes.
@@
 \begin{aligned}
-\mathbf{Z}^\prime &= \arg \min_{\mathbf{Z}^\prime \in \mathcal{Z}^M} \left\{  {\text{yloss}(M_{\theta}(f(\mathbf{Z}^\prime)),\mathbf{y}^*)}+ \lambda {\text{cost}(f(\mathbf{Z}^\prime)) }  \right\} 
+\mathbf{Z}^\prime &= \arg \min_{\mathbf{Z}^\prime \in \mathcal{Z}^L} \left\{  {\text{yloss}(M_{\theta}(f(\mathbf{Z}^\prime)),\mathbf{y}^*)}+ \lambda {\text{cost}(f(\mathbf{Z}^\prime)) }  \right\} 
 \end{aligned} 
@@
 Here $\text{yloss}$ denotes the primary loss function already introduced above and $\text{cost}$ is either a single penalty or a collection of penalties that are used to impose constraints through regularization.
-Following the convention in \citet{altmeyer2023endogenous} we use $\mathbf{Z}^\prime=\{ \mathbf{z}_m\}_M$ to denote the $M$-dimensional array of counterfactual states.
-This is to explicitly account for the fact that we can generate multiple counterfactuals $M$, as with DiCE \citep{mothilal2020explaining}, and may choose to traverse a latent encoding $\mathcal{Z}$ of the feature space $\mathcal{X}$ where we denote $f^{-1}: \mathcal{X} \mapsto \mathcal{Z}$.
+Equation~\ref{eq:general} restates the baseline approach to gradient-based counterfactual search proposed by~\citet{wachter2017counterfactual} in general form where $\mathbf{Z}^\prime=\{ \mathbf{z}_l\}_L$ denotes an $L$-dimensional array of counterfactual states~\citep{altmeyer2023endogenous}.
+This is to explicitly account for the multiplicity of explanations and the fact that we may choose to generate multiple counterfactuals and traverse a latent encoding $\mathcal{Z}$ of the feature space $\mathcal{X}$ where we denote $f^{-1}: \mathcal{X} \mapsto \mathcal{Z}$.
 Encodings may involve simple feature transformations or more advanced techniques involving generative models, as we will discuss further below.
+The baseline approach, which we will simply refer to as \textbf{Wachter}~\citep{wachter2017counterfactual}, searches a single counterfactual directly in the feature space and penalises its distance between the original factual.
 
@@
 A stripped-down counterfactual explanation is therefore little different from an adversarial example.
-In Figure~\ref{fig:adv}, for example, we have the baseline approach proposed in \citet{wachter2017counterfactual} to MNIST data (centre panel).
-This approach solves Equation~\ref{eq:general} through gradient-descent in the feature space with a penalty for the distance between the factual $\mathbf{x}$ and the counterfactual $\mathbf{x}^{\prime}$.
-The underlying classifier $M_{\theta}$ is a simple Multi-Layer Perceptron (MLP) with good test accuracy.
+In Figure~\ref{fig:adv}, for example, we have applied Wachter to MNIST data (centre panel) where the underlying classifier $M_{\theta}$ is a simple Multi-Layer Perceptron (MLP) with above 90 percent test accuracy.
 For the generated counterfactual $\mathbf{x}^{\prime}$ the model predicts the target label with high confidence (centre panel in Figure~\ref{fig:adv}).
-The explanation is valid by definition, even though it looks a lot like an Adversarial Example \citep{goodfellow2014explaining}.
+The explanation is valid by definition, even though it looks a lot like an Adversarial Example~\citep{goodfellow2014explaining}.
 \citet{schut2021generating} make the connection between Adversarial Examples and Counterfactual Explanations explicit and propose using a Jacobian-Based Saliency Map Attack (JSMA) to solve Equation~\ref{eq:general}.
@@
 While an AE is intended to go unnoticed, a CE should have certain desirable properties.
-The literature has made this explicit by introducing various so-called \textit{desiderata}.
-To properly serve both AI practitioners and individuals affected by AI decision-making systems, counterfactuals should be sparse, proximate~\citep{wachter2017counterfactual}, actionable~\citep{ustun2019actionable}, diverse~\citep{mothilal2020explaining}, plausible~\citep{joshi2019realistic,poyiadzi2020face,schut2021generating}, robust~\citep{upadhyay2021robust,pawelczyk2022probabilistically,altmeyer2023endogenous} and causal~\citep{karimi2021algorithmic} among other things.
+The literature has made this explicit by introducing various so-called \textit{desiderata} that counterfactuals should meet in order to properly serve both AI practitioners and individuals affected by AI decision-making systems.
+The list of desiderate includes but is not limited to the following: sparsity, proximity~\citep{wachter2017counterfactual}, actionability~\citep{ustun2019actionable}, diversity~\citep{mothilal2020explaining}, plausibility~\citep{joshi2019realistic,poyiadzi2020face,schut2021generating}, robustness~\citep{upadhyay2021robust,pawelczyk2022probabilistically,altmeyer2023endogenous} and causality~\citep{karimi2021algorithmic}.
 
@@
 For example, \citet{artelt2021evaluating} find that plausibility typically also leads to improved robustness.
-Similarly, plausibility has also been connected to causality in the sense that plausible counterfactuals respect causal relationships \citep{mahajan2020preserving}.
+Similarly, plausibility has also been connected to causality in the sense that plausible counterfactuals respect causal relationships~\citep{mahajan2020preserving}.
 
@@
 Arguably, the plausibility of counterfactuals has been among the primary concerns and some have focused explicitly on this goal.
-\citet{joshi2019realistic}, for example, were among the first to suggest that instead of searching counterfactuals in the feature space $\mathcal{X}$, we can instead traverse a latent embedding $\mathcal{Z}$ that implicitly codifies the data generating process (DGP) of $\mathbf{x}\sim\mathcal{X}$.
+\citet{joshi2019realistic}, for example, were among the first to suggest that instead of searching counterfactuals in the feature space $\mathcal{X}$, we can instead traverse a latent embedding $\mathcal{Z}$ (Equation~\ref{eq:general}) that implicitly codifies the data generating process (DGP) of $\mathbf{x}\sim\mathcal{X}$.
 To learn the latent embedding, they introduce a surrogate model.
@@
 \citet{dombrowski2021diffeomorphic} traverse the base space of a normalizing flow to solve Equation~\ref{eq:general}, essentially relying on a different surrogate model for the generative task.
-\citet{poyiadzi2020face} use density estimators ($\hat{p}: \mathcal{X} \mapsto [0,1]$) to constrain the counterfactual paths.
+\citet{poyiadzi2020face} use density estimators ($\hat{p}: \mathcal{X} \mapsto [0,1]$) to constrain the counterfactuals to dense regions in the feature space.
 \citet{karimi2021algorithmic} argue that counterfactuals should comply with the causal model that generates the data.
@@
 
-Note that Definition~\ref{def:plausible} is consistent with the notion of plausible counterfactual paths, since we can simply apply it to each counterfactual state along the path.
-
 Surrogate models offer an obvious solution to achieve this objective.
-Unfortunately, surrogates also introduce a dependency: the generated explanations no longer depend exclusively on the Black Box Model itself, but also on the surrogate model.
+Unfortunately, surrogates also introduce a dependency: the generated explanations no longer depend exclusively on the black-box model itself, but also on the surrogate model.
 This is not necessarily problematic if the primary objective is not to explain the behaviour of the model but to offer recourse to individuals affected by it.
 It may become problematic even in this context if the dependency turns into a vulnerability.
-To illustrate this point, we have used REVISE \citep{joshi2019realistic} with an underfitted VAE to generate the counterfactual in the right panel of Figure~\ref{fig:vae}: in this case, the decoder step of the VAE fails to yield plausible values ($\{\mathbf{x}^{\prime} \leftarrow \mathcal{G}(\mathbf{z})\} \not\sim \mathcal{X}|\mathbf{y}^*$) and hence the counterfactual search in the learned latent space is doomed.
+To illustrate this point, we have used REVISE~\citep{joshi2019realistic} with an underfitted VAE to generate the counterfactual in the right panel of Figure~\ref{fig:vae}: in this case, the decoder step of the VAE fails to yield plausible values ($\{\mathbf{x}^{\prime} \leftarrow \mathcal{G}(\mathbf{z})\} \not\sim \mathcal{X}|\mathbf{y}^*$) and hence the counterfactual search in the learned latent space is doomed.
 
@@
     \includegraphics[width=\textwidth]{../artifacts/results/images/you_may_not_like_it.png}
-    \caption{You may not like it, but this is what stripped-down counterfactuals look like.
-Counterfactuals for turning an 8 (eight) into a 3 (three): original image (left); counterfactual produced using \citet{wachter2017counterfactual} (centre); and a counterfactual produced using JSMA-based approach introduced by \citep{schut2021generating}.}\label{fig:adv}
+    \caption{Explanations or Adversarial Examples?
+Counterfactuals for turning an 8 (eight) into a 3 (three): original image (left); counterfactual produced using~\citet{wachter2017counterfactual} (centre); and a counterfactual produced using the approach introduced by~\citep{schut2021generating} that uses Jacobian-Based Saliency Map Attacks to solve Equation~\ref{eq:general}.}\label{fig:adv}
   \end{minipage}\hfill
@@
     \caption{Using surrogates can improve plausibility, but also increases vulnerability.
-Counterfactuals for turning an 8 (eight) into a 3 (three): original image (left); counterfactual produced using REVISE \citep{joshi2019realistic} with a well-specified surrogate (centre); and a counterfactual produced using REVISE \citep{joshi2019realistic} with a poorly specified surrogate (right).}\label{fig:vae}
+Counterfactuals for turning an 8 (eight) into a 3 (three): original image (left); counterfactual produced using REVISE~\citep{joshi2019realistic} with a well-specified surrogate (centre); and a counterfactual produced using REVISE~\citep{joshi2019realistic} with a poorly specified surrogate (right).}\label{fig:vae}
   \end{minipage}
@@
 Provided the model is well-specified, these counterfactuals are plausible.
-Unfortunately, this idea hinges on the assumption that the Black Box Model provides well-calibrated predictive uncertainty estimates.
+Unfortunately, this idea hinges on the assumption that the black-box model provides well-calibrated predictive uncertainty estimates.
 
@@
 Enquiring about the plausibility of a counterfactual essentially boils down to the following question: `Is this counterfactual consistent with the underlying data'?
-To introduce this section, we posit a related, slightly more nuanced question: `Is this counterfactual consistent with what the model has learned about the underlying data'?
+We posit a related, slightly more nuanced question: `Is this counterfactual consistent with what the model has learned about the underlying data'?
 We will argue that fidelity is not a sufficient evaluation measure to answer this question and propose a novel way to assess if Counterfactual Explanations conform with model behaviour.
 
-The word \textit{fidelity} stems from the Latin word `fidelis', which means `faithful, loyal, trustworthy' \citep{mw2023fidelity}.
+The word \textit{fidelity} stems from the Latin word `fidelis', which means `faithful, loyal, trustworthy'~\citep{mw2023fidelity}.
 As we explained in Section~\ref{background}, model explanations are generally considered faithful if their corresponding predictions coincide with the predictions made by the model itself.
@@
 
-Recent work by \citet{grathwohl2020your} on Energy Based Models (EBM) has pointed out that there is a `generative model hidden within every standard discriminative model'.
+Recent work by~\citet{grathwohl2020your} on Energy Based Models (EBM) has pointed out that there is a `generative model hidden within every standard discriminative model'.
 The authors show that we can draw samples from the posterior conditional distribution $p_{\theta}(\mathbf{x}|\mathbf{y})$ using Stochastic Gradient Langevin Dynamics (SGLD).
@@
 
-To quantify the model's predictive uncertainty we use Conformal Prediction (CP), an approach that has recently gained popularity in the Machine Learning community \citep{angelopoulos2021gentle,manokhin2022awesome}.
+To quantify the model's predictive uncertainty we use Conformal Prediction (CP), an approach that has recently gained popularity in the Machine Learning community~\citep{angelopoulos2021gentle,manokhin2022awesome}.
 Crucially for our intended application, CP is model-agnostic and can be applied during inference without placing any restrictions on model training.
@@
 
-In order to generate counterfactuals that are associated with low predictive uncertainty, we use a smooth set size penalty introduced by \citet{stutz2022learning} in the context of conformal training:
+In order to generate counterfactuals that are associated with low predictive uncertainty, we use a smooth set size penalty introduced by~\citet{stutz2022learning} in the context of conformal training:
 
@@
 
-In order to compute this penalty for any Black Box Model we merely need to perform a single calibration pass through a holdout set $\mathcal{D}_{\text{cal}}$.
+In order to compute this penalty for any black-box model we merely need to perform a single calibration pass through a holdout set $\mathcal{D}_{\text{cal}}$.
 Arguably, data is typically abundant and in most applications, practitioners tend to hold out a test data set anyway.
@@
 where $\hat{\mathbf{x}}_{\theta}$ denotes samples generated using SGLD (Equation~\ref{eq:sgld}) and $\text{dist}(\cdot)$ is a generic term for a distance metric.
-Our default choice for $\text{dist}(\cdot)$ is the Manhatten Distance since it enforces sparsity.
+Our default choice for $\text{dist}(\cdot)$ is the L1 Norm, or Manhattan distance, since it induces sparsity.
 
@@
   \State $\hat{\mathbf{x}}_{\theta, t} \gets \text{rand}(\mathcal{B},n_{\mathcal{B}})$
-  \State $\mathbf{z}^\prime \gets \mathbf{z}^\prime - \eta \nabla_{\mathbf{z}^\prime} \mathcal{L}(\mathbf{z}^\prime,\mathbf{y}^*,\hat{\mathbf{x}}_{\theta, t})$
+  \State $\mathbf{z}^\prime \gets \mathbf{z}^\prime - \eta \nabla_{\mathbf{z}^\prime} \mathcal{L}(\mathbf{z}^\prime,\mathbf{y}^*,\hat{\mathbf{x}}_{\theta, t}; \Lambda, \alpha)$
   \State $t \gets t+1$
@@
 The search terminates once the convergence criterium is met or the maximum number of iterations $T$ has been exhausted.
-Note that the choice of convergence criterium has important implications on the final counterfactual.
-For more detail on this see Appendix~\ref{app:eccco}).
+Note that the choice of convergence criterium has important implications on the final counterfactual (for more detail on this see Appendix~\ref{app:eccco}).
 
-Figure~\ref{fig:eccco-mnist} presents ECCCos for the MNIST example from Section~\ref{background} for various Black Box models of increasing complexity from left to right: a simple Multi-Layer Perceptron (MLP); an Ensemble of MLPs, each of the same architecture as the single MLP; a Joint Energy Model (JEM) based on the same MLP architecture; and finally, an Ensemble of these JEMs.
+Figure~\ref{fig:eccco-mnist} presents ECCCos for the MNIST example from Section~\ref{background} for various black-box models of increasing complexity from left to right: a simple Multi-Layer Perceptron (MLP); an Ensemble of MLPs, each of the same architecture as the single MLP; a Joint Energy Model (JEM) based on the same MLP architecture; and finally, an Ensemble of these JEMs.
 Since Deep Ensembles have an improved capacity for predictive uncertainty quantification and JEMs are explicitly trained to learn plausible representations of the input data, it is intuitive to see that the plausibility of counterfactuals visibly improves from left to right.
@@
 
-Above we have defined plausibility (\ref{def:plausible}) and conformity (\ref{def:conformal}) for Counterfactual Explanations.
+Above we have defined plausibility (Definition~\ref{def:plausible}) and conformity (Definition~\ref{def:conformal}) for Counterfactual Explanations.
 In this subsection, we introduce evaluation measures that facilitate a quantitative evaluation of counterfactuals for these objectives.
 
-Firstly, in order to assess the plausibility of counterfactuals we adapt the implausibility metric proposed in \citet{guidotti2022counterfactual}.
+Firstly, in order to assess the plausibility of counterfactuals we adapt the implausibility metric proposed in~\citet{guidotti2022counterfactual}.
 The authors propose to evaluate plausibility in terms of the distance of the counterfactual $\mathbf{x}^{\prime}$ from its nearest neighbour in the target class $\mathbf{y}^*$: the smaller this distance, the more plausible the counterfactual.
@@
 
-As noted by \citet{guidotti2022counterfactual}, these distance-based measures are simplistic and more complex alternative measures may ultimately be more appropriate for the task.
+As noted by~\citet{guidotti2022counterfactual}, these distance-based measures are simplistic and more complex alternative measures may ultimately be more appropriate for the task.
 For example, we considered using statistical divergence measures instead.
@@
 
-Consistent with the findings in \citet{schut2021generating}, we have demonstrated that predictive uncertainty estimates can be leveraged to generate plausible counterfactuals.
+Consistent with the findings in~\citet{schut2021generating}, we have demonstrated that predictive uncertainty estimates can be leveraged to generate plausible counterfactuals.
 Interestingly, \citet{schut2021generating} point out that this finding --- as intuitive as it is --- may be linked to a positive connection between the generative task and predictive uncertainty quantification.
 In particular, \citet{grathwohl2020your} demonstrate that their proposed method for integrating the generative objective in training yields models that have improved predictive uncertainty quantification.
-Since neither \citet{schut2021generating} nor we have employed any surrogate generative models, our findings seem to indicate that the positive connection found in \citet{grathwohl2020your} is bidirectional.
+Since neither~\citet{schut2021generating} nor we have employed any surrogate generative models, our findings seem to indicate that the positive connection found in~\citet{grathwohl2020your} is bidirectional.
 
@@
 
-While $\mathbf{x}_J$ is only guaranteed to distribute as $p_{\theta}(\mathbf{x}|\mathbf{y}^*)$ if $\epsilon \rightarrow 0$ and $J \rightarrow \infty$, the bias introduced for a small finite $\epsilon$ is negligible in practice \citep{murphy2023probabilistic,grathwohl2020your}.
-While \citet{grathwohl2020your} use Equation~\ref{eq:sgld} during training, we are interested in applying the conditional sampling procedure in a post hoc fashion to any standard discriminative model.
+While $\mathbf{x}_J$ is only guaranteed to distribute as $p_{\theta}(\mathbf{x}|\mathbf{y}^*)$ if $\epsilon \rightarrow 0$ and $J \rightarrow \infty$, the bias introduced for a small finite $\epsilon$ is negligible in practice~\citep{murphy2023probabilistic,grathwohl2020your}.
+While~\citet{grathwohl2020your} use Equation~\ref{eq:sgld} during training, we are interested in applying the conditional sampling procedure in a post-hoc fashion to any standard discriminative model.
 
@@
 Specifically, they show how Stochastic Gradient Descent can be used to train classifiers not only for the discriminative task but also for additional objectives related to Conformal Prediction.
-One such objective is \textit{efficiency}: for a given target error rate $alpha$, the efficiency of a conformal classifier improves as its average prediction set size decreases.
-To this end, the authors introduce a smooth set size penalty defined in Equation~\ref{eq:setsize}
+One such objective is \textit{efficiency}: for a given target error rate $\alpha$, the efficiency of a conformal classifier improves as its average prediction set size decreases.
+To this end, the authors introduce a smooth set size penalty defined in Equation~\ref{eq:setsize} in the body of this paper
 
-Formally, it is defined as $C_{\theta,\mathbf{y}}(\mathbf{x}_i;\alpha):=\sigma\left((s(\mathbf{x}_i,\mathbf{y})-\alpha) T^{-1}\right)$ for $\mathbf{y}\in\mathcal{Y}$ where $\sigma$ is the sigmoid function and $T$ is a hyper-parameter used for temperature scaling \citep{stutz2022learning}.
+Formally, it is defined as $C_{\theta,\mathbf{y}}(\mathbf{x}_i;\alpha):=\sigma\left((s(\mathbf{x}_i,\mathbf{y})-\alpha) T^{-1}\right)$ for $\mathbf{y}\in\mathcal{Y}$, where $\sigma$ is the sigmoid function and $T$ is a hyper-parameter used for temperature scaling~\citep{stutz2022learning}.
 
 Intuitively, CP works under the premise of turning heuristic notions of uncertainty into rigorous uncertainty estimates by repeatedly sifting through the data.
-It can be used to generate prediction intervals for regression models and prediction sets for classification models \citep{altmeyer2022conformal}.
+It can be used to generate prediction intervals for regression models and prediction sets for classification models~\citep{altmeyer2022conformal}.
 Since the literature on CE and AR is typically concerned with classification problems, we focus on the latter.
-A particular variant of CP called Split Conformal Prediction (SCP) is well-suited for our purposes because it imposes only minimal restrictions on model training.
+A particular variant of CP called Split Conformal Prediction (SCP) is well-suited for our purposes, because it imposes only minimal restrictions on model training.
 
@@
 The latter is then used to compute so-called nonconformity scores: $\mathcal{S}=\{s(\mathbf{x}_i,\mathbf{y}_i)\}_{i \in \mathcal{D}_{\text{cal}}}$ where $s: (\mathcal{X},\mathcal{Y}) \mapsto \mathbb{R}$ is referred to as \textit{score function}.
-In the context of classification, a common choice for the score function is just $s_i=1-M_{\theta}(\mathbf{x}_i)[\mathbf{y}_i]$, that is one minus the softmax output corresponding to the observed label $\mathbf{y}_i$ \citep{angelopoulos2021gentle}.
+In the context of classification, a common choice for the score function is just $s_i=1-M_{\theta}(\mathbf{x}_i)[\mathbf{y}_i]$, that is one minus the softmax output corresponding to the observed label $\mathbf{y}_i$~\citep{angelopoulos2021gentle}.
 
@@
 where $\hat{q}$ denotes the $(1-\alpha)$-quantile of $\mathcal{S}$ and $\alpha$ is a predetermined error rate.
-As the size of the calibration set increases, the probability that the classification set $C(\mathbf{x}_{\text{test}})$ for a newly arrived sample $\mathbf{x}_{\text{test}}$ does not cover the true test label $\mathbf{y}_{\text{test}}$ approaches $\alpha$ \citep{angelopoulos2021gentle}.
+As the size of the calibration set increases, the probability that the classification set $C(\mathbf{x}_{\text{test}})$ for a newly arrived sample $\mathbf{x}_{\text{test}}$ does not cover the true test label $\mathbf{y}_{\text{test}}$ approaches $\alpha$~\citep{angelopoulos2021gentle}.
 
@@
 Intuitively, the set is more likely to include multiple labels for samples that are difficult to classify, so the set size is indicative of predictive uncertainty.
-To see why this effect is exacerbated by small choices for $\alpha$ consider the case of $\alpha=0$, which requires that the true label is covered by the prediction set with probability equal to one.
+To see why this effect is exacerbated by small choices for $\alpha$ consider the case of $\alpha=0$, which requires that the true label is covered by the prediction set with probability equal to 1.
 

### 2023-04-30 notebooks/mnist.qmd

@@
 )
-
-Serialization.serialize(joinpath(output_path, "mnist_benchmark.jls"), bmk)
+CSV.write(joinpath(output_path, "mnist_benchmark.csv"), bmk())
 ```

### 2023-05-01 notebooks/mnist.qmd
bleh

@@
   opt = opt,
-  λ=0.1
+  λ=0.0
 )

### 2023-05-01 paper/paper.tex

--- 2023-04-30 paper/paper.tex
@@
 
-The entire procedure for generating ECCCos is described in Algorithm~\ref{alg:eccco}.
+The entire procedure for Generating ECCCos is described in Algorithm~\ref{alg:eccco}.
 For the sake of simplicity and without loss of generality, we limit our attention to generating a single counterfactual $\mathbf{x}^\prime=f(\mathbf{z}^\prime)$ where in contrast to Equation~\ref{eq:eccco} $\mathbf{z}^\prime$ denotes a $1$-dimensional array containing a single counterfactual state.
@@
 Finally, we search counterfactuals through gradient descent.
-Let $\mathcal{L}(\mathbf{z}^\prime,\mathbf{y}^*,\hat{\mathbf{x}}_{\theta, t})$ denote our loss function defined in Equation~\ref{eq:eccco}.
+Let $\mathcal{L}(\mathbf{z}^\prime,\mathbf{y}^*,\hat{\mathbf{x}}_{\theta, t}; \Lambda, \alpha)$ denote our loss function defined in Equation~\ref{eq:eccco}.
 Then in each iteration, we first randomly draw $n_{\mathcal{B}}$ samples from the buffer $\mathcal{B}$ before updating the counterfactual state $\mathbf{z}^\prime$ by moving in the negative direction of that loss function.

### 2023-05-02 notebooks/eccco.qmd
setting up experiments for real-world data

@@
 # Counteractual data:
-counterfactual_data = load_circles(2000)
+counterfactual_data = load_circles(1000)
 X = counterfactual_data.X
@@
     sampler;
-    builder=MLJFlux.MLP(hidden=(32, 32, 32,), σ=Flux.relu),
+    builder=MLJFlux.MLP(hidden=(32, 32, 32,), σ=Flux.swish),
     batch_size=batch_size,
@@
     jem_training_params=(
-        α=[1.0,1.0,1e-1],
+        α=[1.0,1.0,1e-2],
         verbosity=10,
@@
 M = ECCCo.ConformalModel(mach.model, mach.fitresult)
-factual_label = 0
+factual_label =  levels(labels)[1]
 x_factual = reshape(X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
-target = 1
+target =  levels(labels)[2]
 factual = predict_label(M, counterfactual_data, x_factual)[1]

### 2023-05-02 notebooks/mnist.qmd
setting up experiments for real-world data

@@
 ```{julia}
-function pre_process(x; noise::Float32=0.03f0)
+function pre_process(x; noise::Float32=1/256f0)
     ϵ = Float32.(randn(size(x)) * noise)
@@
 # Hyper:
-_retrain = false
-_regen = false
+_retrain = true
+_regen = true
 
 # Data:
-n_obs = 10000
+n_obs = 1000
 counterfactual_data = load_mnist(n_obs)
@@
 n_hidden = 128
-activation = Flux.relu
+activation = Flux.swish
 builder = MLJFlux.@builder Flux.Chain(
@@
 )
-α = [1.0,1.0,1e-2]      # penalty strengths
+α = [1.0,1.0,5e-2]      # penalty strengths
 ```
@@
     ),
-    sampling_steps=20,
+    sampling_steps=50,
 )
@@
 )
+
+Serialization.serialize(joinpath(output_path,"mnist_architectures.jls"), models)
 ```
@@
 # Plot generated samples:
-n_regen = 150
+n_regen = 120
 if _regen 

### 2023-05-02 notebooks/real_world.qmd
setting up experiments for real-world data

--- 2023-05-01 notebooks/mnist.qmd
@@
 
-# MNIST
-
-## Anecdotal Evidence
-
-### Examples in Introduction
-
-#### Wachter and JSMA
-
-```{julia}
-Random.seed!(1234)
-
-# Data:
-counterfactual_data = load_mnist()
-X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
-input_dim, n_obs = size(counterfactual_data.X)
-M = load_mnist_mlp()
-
-# Target:
-factual_label = 8
-x_factual = reshape(X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
-target = 3
-factual = predict_label(M, counterfactual_data, x_factual)[1]
-γ = 0.9
-
-# Training params:
-T = 100
-opt = Flux.Optimise.Adam(0.01)
-```
-
-```{julia}
-# Search:
-generic_generator = GenericGenerator(opt=opt)
-ce_wachter = generate_counterfactual(
-    x_factual, target, counterfactual_data, M, generic_generator; 
-    decision_threshold=γ, max_iter=T,
-    initialization=:identity,
-)
-greedy_generator = GreedyGenerator(η=1.0)
-ce_jsma = generate_counterfactual(
-    x_factual, target, counterfactual_data, M, greedy_generator; 
-    decision_threshold=γ, max_iter=T,
-    initialization=:identity,
-)
-```
-
-```{julia}
-p1 = Plots.plot(
-    convert2image(MNIST, reshape(x_factual,28,28)),
-    axis=nothing, 
-    size=(img_height, img_height),
-    title="Factual"
-)
-plts = [p1]
-
-ces = zip([ce_wachter,ce_jsma])
-counterfactuals = reduce((x,y)->cat(x,y,dims=3),map(ce -> CounterfactualExplanations.counterfactual(ce[1]), ces))
-phat = reduce((x,y) -> cat(x,y,dims=3), map(ce -> target_probs(ce[1]), ces))
-for x in zip(eachslice(counterfactuals; dims=3), eachslice(phat; dims=3), ["Wachter","JSMA"])
-    ce, _phat, _name = (x[1],x[2],x[3])
-    _title = "$(_name) (p=$(round(_phat[1]; digits=2)))"
-    plt = Plots.plot(
-        convert2image(MNIST, reshape(ce,28,28)),
-        axis=nothing, 
-        size=(img_height, img_height),
-        title=_title
-    )
-    plts = [plts..., plt]
-end
-plt = Plots.plot(plts...; size=(img_height*length(plts),img_height), layout=(1,length(plts)))
-display(plt)
-savefig(plt, joinpath(output_images_path, "you_may_not_like_it.png"))
-```
-
-#### REVISE
-
-```{julia}
-using CounterfactualExplanations.Models: load_mnist_vae
-vae = load_mnist_vae()
-vae_weak = load_mnist_vae(;strong=false)
-Serialization.serialize(joinpath(output_path,"mnist_classifier.jls"), M)
-Serialization.serialize(joinpath(output_path,"mnist_vae.jls"), vae)
-Serialization.serialize(joinpath(output_path,"mnist_vae_weak.jls"), vae_weak)
-```
-
-```{julia}
-# Define generator:
-revise_generator = REVISEGenerator(
-  opt = opt,
-  λ=0.0
-)
-# Generate recourse:
-counterfactual_data.generative_model = vae # assign generative model
-ce_strong = generate_counterfactual(
-    x_factual, target, counterfactual_data, M, revise_generator; 
-    decision_threshold=γ, max_iter=T,
-    initialization=:identity,
-    converge_when=:generator_conditions,
-)
-counterfactual_data_weak = deepcopy(counterfactual_data)
-counterfactual_data_weak.generative_model = vae_weak
-ce_weak = generate_counterfactual(
-    x_factual, target, counterfactual_data_weak, M, revise_generator;
-    decision_threshold=γ, max_iter=T,
-    initialization=:identity,
-    converge_when=:generator_conditions,
-)
-```
-
-```{julia}
-ces = zip([ce_strong,ce_weak])
-counterfactuals = reduce((x,y)->cat(x,y,dims=3),map(ce -> CounterfactualExplanations.counterfactual(ce[1]), ces))
-phat = reduce((x,y) -> cat(x,y,dims=3), map(ce -> target_probs(ce[1]), ces))
-plts = [p1]
-for x in zip(eachslice(counterfactuals; dims=3), eachslice(phat; dims=3), ["Strong VAE","Weak VAE"])
-    ce, _phat, _name = (x[1],x[2],x[3])
-    _title = "$(_name) (p=$(round(_phat[1]; digits=2)))"
-    plt = Plots.plot(
-        convert2image(MNIST, reshape(ce,28,28)),
-        axis=nothing, 
-        size=(img_height, img_height),
-        title=_title
-    )
-    plts = [plts..., plt]
-end
-plt = Plots.plot(plts...; size=(img_height*length(plts),img_height), layout=(1,length(plts)))
-display(plt)
-savefig(plt, joinpath(output_images_path, "surrogate_gone_wrong.png"))
-```
-
-### ECCCo
-
-```{julia}
-function pre_process(x; noise::Float32=0.03f0)
-    ϵ = Float32.(randn(size(x)) * noise)
-    x += ϵ
-    return x
-end
-```
+# Real-World Data
 
@@
 n_obs = 10000
-counterfactual_data = load_mnist(n_obs)
-counterfactual_data.X = pre_process.(counterfactual_data.X)
-counterfactual_data.generative_model = vae
+datasets = load_tabular_data(n_obs; drop=:credit_default)
 X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
@@
 batch_size = minimum([Int(round(n_obs/10)), 128])
-n_hidden = 128
+n_hidden = 64
 activation = Flux.relu

### 2023-05-02 notebooks/cal_housing.qmd
added code for real-world data experiments

```{julia}
include("notebooks/setup.jl")
eval(setup_notebooks)
```

# Real-World Data

```{julia}
# Hyper:
_retrain = true

# Data:
n_obs = 10000
counterfactual_data = load_california_housing(n_obs)
X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
X = table(permutedims(X))
labels = counterfactual_data.output_encoder.labels
input_dim, n_obs = size(counterfactual_data.X)
output_dim = length(unique(labels))
```

First, let's create a couple of image classifier architectures:

```{julia}
# Model parameters:
epochs = 100
batch_size = minimum([Int(round(n_obs/10)), 128])
n_hidden = 64
activation = Flux.relu
builder = MLJFlux.@builder Flux.Chain(
    Dense(n_in, n_hidden, activation),
    Dense(n_hidden, n_out),
)
n_ens = 5                                   # number of models in ensemble
_loss = Flux.Losses.logitcrossentropy       # loss function
_finaliser = x -> x                         # finaliser function
```

```{julia}
# JEM parameters:
𝒟x = Normal()
𝒟y = Categorical(ones(output_dim) ./ output_dim)
sampler = ConditionalSampler(
    𝒟x, 𝒟y, 
    input_size=(input_dim,), 
    batch_size=10,
)
α = [1.0,1.0,1e-2]      # penalty strengths
```

```{julia}
# Simple MLP:
mlp = NeuralNetworkClassifier(
    builder=builder, 
    epochs=epochs,
    batch_size=batch_size,
    finaliser=_finaliser,
    loss=_loss,
)

# Deep Ensemble:
mlp_ens = EnsembleModel(model=mlp, n=n_ens)

# Joint Energy Model:
jem = JointEnergyClassifier(
    sampler;
    builder=builder,
    epochs=epochs,
    batch_size=batch_size,
    finaliser=_finaliser,
    loss=_loss,
    jem_training_params=(
        α=α,verbosity=10,
    ),
    sampling_steps=20,
)

# JEM with adversarial training:
jem_adv = deepcopy(jem)
# jem_adv.adv_training = true

# Deep Ensemble of Joint Energy Models:
jem_ens = EnsembleModel(model=jem, n=n_ens)

# Deep Ensemble of Joint Energy Models with adversarial training:
# jem_ens_plus = EnsembleModel(model=jem_adv, n=n_ens)

# Dictionary of models:
models = Dict(
    "MLP" => mlp,
    "MLP Ensemble" => mlp_ens,
    "JEM" => jem,
    "JEM Ensemble" => jem_ens,
    # "JEM Ensemble+" => jem_ens_plus,
)
```

```{julia}
# Train models:
function _train(model, X=X, y=labels; cov=.95, method=:simple_inductive, mod_name="model")
    conf_model = conformal_model(model; method=method, coverage=cov)
    mach = machine(conf_model, X, y)
    @info "Begin training $mod_name."
    fit!(mach)
    @info "Finished training $mod_name."
    M = ECCCo.ConformalModel(mach.model, mach.fitresult)
    return M
end
if _retrain
    model_dict = Dict(mod_name => _train(mod; mod_name=mod_name) for (mod_name, mod) in models)
    Serialization.serialize(joinpath(output_path,"cal_housing_models.jls"), model_dict)
else
    model_dict = Serialization.deserialize(joinpath(output_path,"cal_housing_models.jls"))
end
```

```{julia}
# # Evaluate models:

# measure = Dict(
#     :f1score => multiclass_f1score, 
#     :acc => accuracy, 
#     :precision => multiclass_precision
# )
# model_performance = DataFrame()
# for (mod_name, mod) in model_dict
#     # Test performance:
#     test_data = load_mnist_test()
#     _perf = CounterfactualExplanations.Models.model_evaluation(mod, test_data, measure=collect(values(measure)))
#     _perf = DataFrame([[p] for p in _perf], collect(keys(measure)))
#     _perf.mod_name .= mod_name
#     model_performance = vcat(model_performance, _perf)
# end
# Serialization.serialize(joinpath(output_path,"cal_housing_model_performance.jls"), model_performance)
# CSV.write(joinpath(output_path, "cal_housing_model_performance.csv"), model_performance)
# model_performance
```

## Benchmark

```{julia}
# Benchmark generators:
generator_dict = Dict(
    :wachter => WachterGenerator(),
    :revise => REVISEGenerator(),
    :greedy => GreedyGenerator(),
    :eccco => ECCCoGenerator(),
)

# Measures:
measures = [
    CounterfactualExplanations.distance,
    ECCCo.distance_from_energy,
    ECCCo.distance_from_targets,
    CounterfactualExplanations.Evaluation.validity,
    CounterfactualExplanations.Evaluation.redundancy,
]

bmk = benchmark(
    counterfactual_data; 
    models=model_dict, 
    generators=generator_dict, 
    measure=measures,
    suppress_training=true, dataname="California Housing",
    n_individuals=5,
    factual=0, target=1,
    initialization=:identity,
)
CSV.write(joinpath(output_path, "cal_housing_benchmark.csv"), bmk())
```

```{julia}
@chain bmk() begin
    @group_by(dataname, generator, model, variable)
    @summarize(mean=mean(value),sd=std(value))
    @ungroup
    @filter(variable == "distance_from_energy")
end
```

```{julia}
df = @chain bmk() begin
    @filter(variable in [
        "distance_from_energy",
        "distance_from_targets",
        "distance",])
    @mutate(variable = ifelse.(variable .== "distance_from_energy", "Non-Conformity", variable))
    @mutate(variable = ifelse.(variable .== "distance_from_targets", "Implausibility", variable))
    @mutate(variable = ifelse.(variable .== "distance", "Cost", variable))
end
plt = AlgebraOfGraphics.data(df) * visual(BoxPlot) * 
    mapping(:generator, :value, row=:variable, col=:model, color=:generator)
plt = draw(
    plt, axis=(xlabel="", xticksvisible=false, xticklabelsvisible=false, width=150, height=120), 
    facet=(; linkyaxes=:minimal)
)   
display(plt)
save(joinpath(output_images_path, "cal_housing_benchmark.png"), plt, px_per_unit=5)
```

### 2023-05-02 notebooks/gmsc.qmd
added code for real-world data experiments

--- 2023-05-02 notebooks/cal_housing.qmd
@@
     model_dict = Dict(mod_name => _train(mod; mod_name=mod_name) for (mod_name, mod) in models)
-    Serialization.serialize(joinpath(output_path,"cal_housing_models.jls"), model_dict)
+    Serialization.serialize(joinpath(output_path,"gmsc_models.jls"), model_dict)
 else
-    model_dict = Serialization.deserialize(joinpath(output_path,"cal_housing_models.jls"))
+    model_dict = Serialization.deserialize(joinpath(output_path,"gmsc_models.jls"))
 end
@@
 # end
-# Serialization.serialize(joinpath(output_path,"cal_housing_model_performance.jls"), model_performance)
-# CSV.write(joinpath(output_path, "cal_housing_model_performance.csv"), model_performance)
+# Serialization.serialize(joinpath(output_path,"gmsc_model_performance.jls"), model_performance)
+# CSV.write(joinpath(output_path, "gmsc_model_performance.csv"), model_performance)
 # model_performance
@@
     measure=measures,
-    suppress_training=true, dataname="California Housing",
+    suppress_training=true, dataname="Californian Housing",
     n_individuals=5,
@@
 )
-CSV.write(joinpath(output_path, "cal_housing_benchmark.csv"), bmk())
+CSV.write(joinpath(output_path, "gmsc_benchmark.csv"), bmk())
 ```
@@
 display(plt)
-save(joinpath(output_images_path, "cal_housing_benchmark.png"), plt, px_per_unit=5)
+save(joinpath(output_images_path, "gmsc_benchmark.png"), plt, px_per_unit=5)
 ```

### 2023-05-02 notebooks/gmsc.qmd
experiments

@@
 )
-CSV.write(joinpath(output_path, "gmsc_benchmark.csv"), bmk())
+CSV.write(joinpath(output_path, "gmsc_benchmark.csv"), bmk.evaluation)
 ```
@@
 ```{julia}
-@chain bmk() begin
+@chain bmk.evaluation begin
     @group_by(dataname, generator, model, variable)
@@
 ```{julia}
-df = @chain bmk() begin
+df = @chain bmk.evaluation begin
     @filter(variable in [
@@
         "distance_from_targets",
-        "distance",])
+        # "distance",
+    ])
     @mutate(variable = ifelse.(variable .== "distance_from_energy", "Non-Conformity", variable))
     @mutate(variable = ifelse.(variable .== "distance_from_targets", "Implausibility", variable))
-    @mutate(variable = ifelse.(variable .== "distance", "Cost", variable))
+    # @mutate(variable = ifelse.(variable .== "distance", "Cost", variable))
 end

### 2023-05-03 notebooks/cal_housing.qmd
:cry

@@
 # Data:
-n_obs = 10000
-counterfactual_data = load_california_housing(n_obs)
+test_size = 0.2
+n_obs = Int(1000 / (1.0 - test_size))
+counterfactual_data, test_data = train_test_split(load_california_housing(n_obs); test_size=test_size)
 X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
@@
 batch_size = minimum([Int(round(n_obs/10)), 128])
-n_hidden = 64
+n_hidden = 128
 activation = Flux.relu
@@
     Dense(n_in, n_hidden, activation),
+    Dense(n_hidden, n_hidden, activation),
     Dense(n_hidden, n_out),
@@
 ```{julia}
-# # Evaluate models:
+# Evaluate models:
 
-# measure = Dict(
-#     :f1score => multiclass_f1score, 
-#     :acc => accuracy, 
-#     :precision => multiclass_precision
-# )
-# model_performance = DataFrame()
-# for (mod_name, mod) in model_dict
-#     # Test performance:
-#     test_data = load_mnist_test()
-#     _perf = CounterfactualExplanations.Models.model_evaluation(mod, test_data, measure=collect(values(measure)))
-#     _perf = DataFrame([[p] for p in _perf], collect(keys(measure)))
-#     _perf.mod_name .= mod_name
-#     model_performance = vcat(model_performance, _perf)
-# end
-# Serialization.serialize(joinpath(output_path,"cal_housing_model_performance.jls"), model_performance)
-# CSV.write(joinpath(output_path, "cal_housing_model_performance.csv"), model_performance)
-# model_performance
+measure = Dict(
+    :f1score => multiclass_f1score, 
+    :acc => accuracy, 
+    :precision => multiclass_precision
+)
+model_performance = DataFrame()
+for (mod_name, mod) in model_dict
+    # Test performance:
+    _perf = CounterfactualExplanations.Models.model_evaluation(mod, test_data, measure=collect(values(measure)))
+    _perf = DataFrame([[p] for p in _perf], collect(keys(measure)))
+    _perf.mod_name .= mod_name
+    model_performance = vcat(model_performance, _perf)
+end
+Serialization.serialize(joinpath(output_path,"cal_housing_model_performance.jls"), model_performance)
+CSV.write(joinpath(output_path, "cal_housing_model_performance.csv"), model_performance)
+model_performance
 ```
@@
 df = @chain bmk() begin
-    @filter(variable in [
-        "distance_from_energy",
-        "distance_from_targets",
-        "distance",])
     @mutate(variable = ifelse.(variable .== "distance_from_energy", "Non-Conformity", variable))
@@
     @mutate(variable = ifelse.(variable .== "distance", "Cost", variable))
+    @mutate(variable = ifelse.(variable .== "redundancy", "Redundancy", variable))
+    @mutate(variable = ifelse.(variable .== "Validity", "Validity", variable))
 end
@@
     plt, axis=(xlabel="", xticksvisible=false, xticklabelsvisible=false, width=150, height=120), 
-    facet=(; linkyaxes=:minimal)
+    facet=(; linkyaxes=:none)
 )   

### 2023-05-03 src/generator.jl

@@ -17,7 +17,7 @@ end
 
 "Constructor for `ECECCCoGenerator`: Energy Constrained Conformal Counterfactual Explanation Generator."
 function ECCCoGenerator(; 
-    λ::Union{AbstractFloat,Vector{<:AbstractFloat}}=[0.1, 1.0, 1.0], 
+    λ::Union{AbstractFloat,Vector{<:AbstractFloat}}=[0.1, 0.1, 0.1], 
     κ::Real=1.0, 
     temp::Real=0.5, 
     η::Union{Nothing,Real}=nothing,
@@ -28,7 +28,7 @@ function ECCCoGenerator(;
     function _set_size_penalty(ce::AbstractCounterfactualExplanation)
         return ECCCo.set_size_penalty(ce; κ=κ, temp=temp)
     end
-    _penalties = [Objectives.distance_l1, _set_size_penalty, ECCCo.distance_from_energy]
+    _penalties = [Objectives.distance_mad, _set_size_penalty, ECCCo.distance_from_energy]
     λ = λ isa AbstractFloat ? [0.0, λ, λ] : λ
     return Generator(; penalty=_penalties, λ=λ, opt=opt, kwargs...)
 end

### 2023-05-04 notebooks/circles.qmd
work on tuning JEM

```{julia}
include("notebooks/setup.jl")
eval(setup_notebooks)
```

# Circles Data

```{julia}
# Hyper:
_retrain = true

# Data:
test_size = 0.2
n_obs = Int(1000 / (1.0 - test_size))
counterfactual_data, test_data = train_test_split(load_circles(n_obs); test_size=test_size)
X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
X = table(permutedims(X))
labels = counterfactual_data.output_encoder.labels
input_dim, n_obs = size(counterfactual_data.X)
output_dim = length(unique(labels))
```

First, let's create a couple of image classifier architectures:

```{julia}
# Model parameters:
epochs = 100
batch_size = minimum([Int(round(n_obs/10)), 128])
n_hidden = 32
activation = Flux.swish
builder = MLJFlux.MLP(
    hidden=(n_hidden, n_hidden, n_hidden,), 
    σ=activation
)
n_ens = 5                                   # number of models in ensemble
_loss = Flux.Losses.logitcrossentropy       # loss function
_finaliser = x -> x                         # finaliser function
```

```{julia}
# JEM parameters:
𝒟x = Normal()
𝒟y = Categorical(ones(output_dim) ./ output_dim)
sampler = ConditionalSampler(
    𝒟x, 𝒟y, 
    input_size=(input_dim,), 
    batch_size=batch_size,
)
α = [1.0,1.0,1e-2]      # penalty strengths
```

```{julia}
# Simple MLP:
mlp = NeuralNetworkClassifier(
    builder=builder, 
    epochs=epochs,
    batch_size=batch_size,
    finaliser=_finaliser,
    loss=_loss,
)

# Deep Ensemble:
mlp_ens = EnsembleModel(model=mlp, n=n_ens)

# Joint Energy Model:
jem = JointEnergyClassifier(
    sampler;
    builder=builder,
    epochs=epochs,
    batch_size=batch_size,
    finaliser=_finaliser,
    loss=_loss,
    jem_training_params=(
        α=α,verbosity=10,
    ),
    sampling_steps=20,
)

# JEM with adversarial training:
jem_adv = deepcopy(jem)
# jem_adv.adv_training = true

# Deep Ensemble of Joint Energy Models:
jem_ens = EnsembleModel(model=jem, n=n_ens)

# Deep Ensemble of Joint Energy Models with adversarial training:
# jem_ens_plus = EnsembleModel(model=jem_adv, n=n_ens)

# Dictionary of models:
models = Dict(
    "MLP" => mlp,
    "MLP Ensemble" => mlp_ens,
    "JEM" => jem,
    "JEM Ensemble" => jem_ens,
    # "JEM Ensemble+" => jem_ens_plus,
)
```

```{julia}
# Train models:
function _train(model, X=X, y=labels; cov=.95, method=:simple_inductive, mod_name="model")
    conf_model = conformal_model(model; method=method, coverage=cov)
    mach = machine(conf_model, X, y)
    @info "Begin training $mod_name."
    fit!(mach)
    @info "Finished training $mod_name."
    M = ECCCo.ConformalModel(mach.model, mach.fitresult)
    return M
end
if _retrain
    model_dict = Dict(mod_name => _train(mod; mod_name=mod_name) for (mod_name, mod) in models)
    Serialization.serialize(joinpath(output_path,"moons_models.jls"), model_dict)
else
    model_dict = Serialization.deserialize(joinpath(output_path,"moons_models.jls"))
end
```

```{julia}
# Evaluate models:

measure = Dict(
    :f1score => multiclass_f1score, 
    :acc => accuracy, 
    :precision => multiclass_precision
)
model_performance = DataFrame()
for (mod_name, mod) in model_dict
    # Test performance:
    _perf = CounterfactualExplanations.Models.model_evaluation(mod, test_data, measure=collect(values(measure)))
    _perf = DataFrame([[p] for p in _perf], collect(keys(measure)))
    _perf.mod_name .= mod_name
    model_performance = vcat(model_performance, _perf)
end
Serialization.serialize(joinpath(output_path,"moons_model_performance.jls"), model_performance)
CSV.write(joinpath(output_path, "moons_model_performance.csv"), model_performance)
model_performance
```

```{julia}
n_regen = 256
n_each = batch_size
for (mod_name, mod) in model_dict
    K = length(counterfactual_data.y_levels)
    input_size = size(selectdim(counterfactual_data.X, ndims(counterfactual_data.X), 1))
    𝒟x = Uniform(extrema(counterfactual_data.X)...)
    𝒟y = Categorical(ones(K) ./ K)
    sampler = ConditionalSampler(𝒟x, 𝒟y; input_size=input_size, prob_buffer=0.0)
    opt = ImproperSGLD()
    plts = []
    for target in levels(labels)
        X̂ = sampler(f, opt; niter=n_regen, n_samples=n_each, y=target)
        ex = extrema(hcat(MLJFlux.reformat(X),X̂), dims=2)
        xlims = ex[1]
        ylims = ex[2]
        x1 = range(1.0f0.*xlims...,length=100)
        x2 = range(1.0f0.*ylims...,length=100)
        target_idx = findall(levels(labels) .== target)[1]
        f(x) = target_idx == 2 ? probs(mod, x) : 1 .- probs(mod, x)
        plt = Plots.contour(
            x1, x2, (x, y) -> f([x, y][:,:])[1], 
            fill=true, alpha=0.5, title="Target: $target", cbar=true,
            xlims=xlims,
            ylims=ylims,
        )
        Plots.scatter!(
            MLJFlux.reformat(X)[1,:], MLJFlux.reformat(X)[2,:], 
            color=Int.(labels.refs).-1, group=Int.(labels.refs).-1, alpha=0.5
        )
        Plots.scatter!(
            X̂[1,:], X̂[2,:], 
            color=repeat([target], size(X̂,2)), 
            group=repeat([target], size(X̂,2)), 
            shape=:star5, ms=10
        )
        savefig(plt, joinpath(output_images_path, "circles_generated_$(mod_name).png"))
        push!(plts, plt)
    end
    plt = Plots.plot(plts..., layout=(1, 2), size=(2*500, 400), plot_title=mod_name)
    display(plt)
end
```

## Benchmark

```{julia}
# Benchmark generators:
generator_dict = Dict(
    :wachter => WachterGenerator(),
    :revise => REVISEGenerator(),
    :greedy => GreedyGenerator(),
    :eccco => ECCCoGenerator(),
)

# Measures:
measures = [
    CounterfactualExplanations.distance,
    ECCCo.distance_from_energy,
    ECCCo.distance_from_targets,
    CounterfactualExplanations.Evaluation.validity,
    CounterfactualExplanations.Evaluation.redundancy,
]

bmk = benchmark(
    counterfactual_data; 
    models=model_dict, 
    generators=generator_dict, 
    measure=measures,
    suppress_training=true, dataname="California Housing",
    n_individuals=5,
    factual=0, target=1,
    initialization=:identity,
)
CSV.write(joinpath(output_path, "moons_benchmark.csv"), bmk())
```

```{julia}
@chain bmk() begin
    @group_by(dataname, generator, model, variable)
    @summarize(mean=mean(value),sd=std(value))
    @ungroup
    @filter(variable == "distance_from_energy")
end
```

```{julia}
df = @chain bmk() begin
    @mutate(variable = ifelse.(variable .== "distance_from_energy", "Non-Conformity", variable))
    @mutate(variable = ifelse.(variable .== "distance_from_targets", "Implausibility", variable))
    @mutate(variable = ifelse.(variable .== "distance", "Cost", variable))
    @mutate(variable = ifelse.(variable .== "redundancy", "Redundancy", variable))
    @mutate(variable = ifelse.(variable .== "Validity", "Validity", variable))
end
plt = AlgebraOfGraphics.data(df) * visual(BoxPlot) * 
    mapping(:generator, :value, row=:variable, col=:model, color=:generator)
plt = draw(
    plt, axis=(xlabel="", xticksvisible=false, xticklabelsvisible=false, width=150, height=120), 
    facet=(; linkyaxes=:none)
)   
display(plt)
save(joinpath(output_images_path, "moons_benchmark.png"), plt, px_per_unit=5)
```

### 2023-05-04 notebooks/gmsc.qmd
work on tuning JEM

--- 2023-05-02 notebooks/gmsc.qmd
@@
 
-# Real-World Data
+# GMSC
 
@@
 # Data:
-n_obs = 10000
-counterfactual_data = load_california_housing(n_obs)
+test_size = 0.2
+n_obs = Int(1000 / (1.0 - test_size))
+counterfactual_data, test_data = train_test_split(load_gmsc(n_obs); test_size=test_size)
 X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
@@
 batch_size = minimum([Int(round(n_obs/10)), 128])
-n_hidden = 64
+n_hidden = 256
 activation = Flux.relu
@@
     Dense(n_in, n_hidden, activation),
+    Dense(n_hidden, n_hidden, activation),
+    Dense(n_hidden, n_hidden, activation),
+    Dense(n_hidden, n_hidden, activation),
     Dense(n_hidden, n_out),
@@
     ),
-    sampling_steps=20,
+    sampling_steps=60,
 )
@@
 ```{julia}
-# # Evaluate models:
+# Evaluate models:
 
-# measure = Dict(
-#     :f1score => multiclass_f1score, 
-#     :acc => accuracy, 
-#     :precision => multiclass_precision
-# )
-# model_performance = DataFrame()
-# for (mod_name, mod) in model_dict
-#     # Test performance:
-#     test_data = load_mnist_test()
-#     _perf = CounterfactualExplanations.Models.model_evaluation(mod, test_data, measure=collect(values(measure)))
-#     _perf = DataFrame([[p] for p in _perf], collect(keys(measure)))
-#     _perf.mod_name .= mod_name
-#     model_performance = vcat(model_performance, _perf)
-# end
-# Serialization.serialize(joinpath(output_path,"gmsc_model_performance.jls"), model_performance)
-# CSV.write(joinpath(output_path, "gmsc_model_performance.csv"), model_performance)
-# model_performance
+measure = Dict(
+    :f1score => multiclass_f1score, 
+    :acc => accuracy, 
+    :precision => multiclass_precision
+)
+model_performance = DataFrame()
+for (mod_name, mod) in model_dict
+    # Test performance:
+    _perf = CounterfactualExplanations.Models.model_evaluation(mod, test_data, measure=collect(values(measure)))
+    _perf = DataFrame([[p] for p in _perf], collect(keys(measure)))
+    _perf.mod_name .= mod_name
+    model_performance = vcat(model_performance, _perf)
+end
+Serialization.serialize(joinpath(output_path,"gmsc_model_performance.jls"), model_performance)
+CSV.write(joinpath(output_path, "gmsc_model_performance.csv"), model_performance)
+model_performance
 ```
@@
     measure=measures,
-    suppress_training=true, dataname="Californian Housing",
+    suppress_training=true, dataname="California Housing",
     n_individuals=5,

### 2023-05-04 notebooks/mnist.qmd
work on tuning JEM

@@
 ```{julia}
-function pre_process(x; noise::Float32=1/256f0)
+function pre_process(x; noise::Float32=0.03f0)
     ϵ = Float32.(randn(size(x)) * noise)
@@
 )
-α = [1.0,1.0,5e-2]      # penalty strengths
+α = [1.0,1.0,1e-1]      # penalty strengths
 ```
@@
     ),
-    sampling_steps=50,
+    sampling_steps=20,
 )
@@
 # Plot generated samples:
-n_regen = 120
+n_regen = 200
 if _regen 
     for (mod_name, mod) in model_dict
-        if ECCCo._has_sampler(mod)
-            sampler = ECCCo._get_sampler(mod)
-        else
         K = length(counterfactual_data.y_levels)
@@
         𝒟y = Categorical(ones(K) ./ K)
-            sampler = ConditionalSampler(𝒟x, 𝒟y; input_size=input_size)
-        end
+        sampler = ConditionalSampler(𝒟x, 𝒟y; input_size=input_size, prob_buffer=0.0)
         opt = ImproperSGLD()
@@
     x::Union{AbstractArray, Int}=x_factual, target::Int=target;
-    λ=[0.5,0.1,0.5],
+    λ=[0.1,0.1,0.1],
     temp=0.1,η=0.01,
@@
     rng::Union{Int,AbstractRNG}=Random.GLOBAL_RNG,
+    T::Int = 100,
 )
@@
 ```{julia}
-plt, eccco_generator = _plot_eccco_mnist()
+plt, eccco_generator = _plot_eccco_mnist(opt=JSMADescent(η=1.0))
 display(plt)

### 2023-05-04 notebooks/moons.qmd
work on tuning JEM

--- 2023-05-04 notebooks/circles.qmd
@@
 
-# Circles Data
+# Moons Data
 
@@
 n_obs = Int(1000 / (1.0 - test_size))
-counterfactual_data, test_data = train_test_split(load_circles(n_obs); test_size=test_size)
+counterfactual_data, test_data = train_test_split(load_moons(n_obs); test_size=test_size)
 X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
@@
 batch_size = minimum([Int(round(n_obs/10)), 128])
-n_hidden = 32
-activation = Flux.swish
-builder = MLJFlux.MLP(
-    hidden=(n_hidden, n_hidden, n_hidden,), 
-    σ=activation
+n_hidden = 128
+activation = Flux.relu
+builder = MLJFlux.@builder Flux.Chain(
+    Dense(n_in, n_hidden, activation),
+    Dense(n_hidden, n_hidden, activation),
+    Dense(n_hidden, n_out),
 )
@@
     input_size=(input_dim,), 
-    batch_size=batch_size,
+    batch_size=10,
 )
-α = [1.0,1.0,1e-2]      # penalty strengths
+α = [1.0,1.0,1e-1]      # penalty strengths
 ```
@@
 
-```{julia}
-n_regen = 256
-n_each = batch_size
-for (mod_name, mod) in model_dict
-    K = length(counterfactual_data.y_levels)
-    input_size = size(selectdim(counterfactual_data.X, ndims(counterfactual_data.X), 1))
-    𝒟x = Uniform(extrema(counterfactual_data.X)...)
-    𝒟y = Categorical(ones(K) ./ K)
-    sampler = ConditionalSampler(𝒟x, 𝒟y; input_size=input_size, prob_buffer=0.0)
-    opt = ImproperSGLD()
-    plts = []
-    for target in levels(labels)
-        X̂ = sampler(f, opt; niter=n_regen, n_samples=n_each, y=target)
-        ex = extrema(hcat(MLJFlux.reformat(X),X̂), dims=2)
-        xlims = ex[1]
-        ylims = ex[2]
-        x1 = range(1.0f0.*xlims...,length=100)
-        x2 = range(1.0f0.*ylims...,length=100)
-        target_idx = findall(levels(labels) .== target)[1]
-        f(x) = target_idx == 2 ? probs(mod, x) : 1 .- probs(mod, x)
-        plt = Plots.contour(
-            x1, x2, (x, y) -> f([x, y][:,:])[1], 
-            fill=true, alpha=0.5, title="Target: $target", cbar=true,
-            xlims=xlims,
-            ylims=ylims,
-        )
-        Plots.scatter!(
-            MLJFlux.reformat(X)[1,:], MLJFlux.reformat(X)[2,:], 
-            color=Int.(labels.refs).-1, group=Int.(labels.refs).-1, alpha=0.5
-        )
-        Plots.scatter!(
-            X̂[1,:], X̂[2,:], 
-            color=repeat([target], size(X̂,2)), 
-            group=repeat([target], size(X̂,2)), 
-            shape=:star5, ms=10
-        )
-        savefig(plt, joinpath(output_images_path, "circles_generated_$(mod_name).png"))
-        push!(plts, plt)
-    end
-    plt = Plots.plot(plts..., layout=(1, 2), size=(2*500, 400), plot_title=mod_name)
-    display(plt)
-end
-```
-
 ## Benchmark

### 2023-05-04 notebooks/setup.jl

@@ -14,7 +14,8 @@ setup_notebooks = quote
     using CounterfactualExplanations.Data
     using CounterfactualExplanations.DataPreprocessing: train_test_split
     using CounterfactualExplanations.Evaluation: benchmark, evaluate
-    using CounterfactualExplanations.Models: load_mnist_mlp, train
+    using CounterfactualExplanations.Generators: JSMADescent
+    using CounterfactualExplanations.Models: load_mnist_mlp, train, probs
     using CounterfactualExplanations.Objectives
     using CSV
     using DataFrames

### 2023-05-04 src/generator.jl

@@ -28,7 +28,7 @@ function ECCCoGenerator(;
     function _set_size_penalty(ce::AbstractCounterfactualExplanation)
         return ECCCo.set_size_penalty(ce; κ=κ, temp=temp)
     end
-    _penalties = [Objectives.distance_mad, _set_size_penalty, ECCCo.distance_from_energy]
+    _penalties = [Objectives.distance_l1, _set_size_penalty, ECCCo.distance_from_energy]
     λ = λ isa AbstractFloat ? [0.0, λ, λ] : λ
     return Generator(; penalty=_penalties, λ=λ, opt=opt, kwargs...)
 end

### 2023-05-04 src/model.jl

@@ -61,17 +61,21 @@ function _outdim(fitresult)
 end
 
 """
-    _get_sampler(model::ConformalModel)
+    _get_sampler(model::AbstractFittedModel)
 
 Private helper function that extracts the sampler from a fitted model.
 """
-function _get_sampler(model::ConformalModel)
+function _get_sampler(model::AbstractFittedModel)
     _mod = model.model
-    if _mod.model isa MLJEnsembles.EitherEnsembleModel
-        _mod = _mod.model
-    end
-    if _mod.model isa JointEnergyClassifier
-        sampler = _mod.model.sampler
+    if hasfield(typeof(_mod), :model)
+        if _mod.model isa MLJEnsembles.EitherEnsembleModel
+            _mod = _mod.model
+        end
+        if _mod.model isa JointEnergyClassifier
+            sampler = _mod.model.sampler
+        else
+            sampler = false
+        end
     else
         sampler = false
     end
@@ -79,11 +83,11 @@ function _get_sampler(model::ConformalModel)
 end
 
 """
-    _has_sampler(model::ConformalModel)
+    _has_sampler(model::AbstractFittedModel)
 
 Private helper function that checks if a fitted model has a sampler.
 """
-function _has_sampler(model::ConformalModel)
+function _has_sampler(model::AbstractFittedModel)
     return !(_get_sampler(model) isa Bool)
 end
 

### 2023-05-04 notebooks/circles.qmd
finding that sweet sweet architecture

@@
 n_obs = Int(1000 / (1.0 - test_size))
-counterfactual_data, test_data = train_test_split(load_circles(n_obs); test_size=test_size)
+counterfactual_data, test_data = train_test_split(load_circles(n_obs; noise=0.05, factor=0.5); test_size=test_size)
 X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
@@
 if _retrain
-    model_dict = Dict(mod_name => _train(mod; mod_name=mod_name) for (mod_name, mod) in models)
+    model_dict = Dict(mod_name => _train(model; mod_name=mod_name) for (mod_name, model) in models)
     Serialization.serialize(joinpath(output_path,"moons_models.jls"), model_dict)
@@
 model_performance = DataFrame()
-for (mod_name, mod) in model_dict
+for (mod_name, model) in model_dict
     # Test performance:
-    _perf = CounterfactualExplanations.Models.model_evaluation(mod, test_data, measure=collect(values(measure)))
+    _perf = CounterfactualExplanations.Models.model_evaluation(model, test_data, measure=collect(values(measure)))
     _perf = DataFrame([[p] for p in _perf], collect(keys(measure)))
@@
 ```{julia}
-n_regen = 256
+n_regen = 200
 n_each = batch_size
-for (mod_name, mod) in model_dict
+for (mod_name, model) in model_dict
     K = length(counterfactual_data.y_levels)
@@
     𝒟y = Categorical(ones(K) ./ K)
-    sampler = ConditionalSampler(𝒟x, 𝒟y; input_size=input_size, prob_buffer=0.0)
+    sampler = ConditionalSampler(𝒟x, 𝒟y; input_size=input_size)
     opt = ImproperSGLD()
@@
     for target in levels(labels)
-        X̂ = sampler(f, opt; niter=n_regen, n_samples=n_each, y=target)
+        target_idx = findall(levels(labels) .== target)[1]
+        f(x) = logits(model, x)
+        X̂ = sampler(f, opt; niter=n_regen, n_samples=n_each, y=target_idx)
         ex = extrema(hcat(MLJFlux.reformat(X),X̂), dims=2)
@@
         x2 = range(1.0f0.*ylims...,length=100)
-        target_idx = findall(levels(labels) .== target)[1]
-        f(x) = target_idx == 2 ? probs(mod, x) : 1 .- probs(mod, x)
+        p(x) = probs(model, x)
         plt = Plots.contour(
-            x1, x2, (x, y) -> f([x, y][:,:])[1], 
+            x1, x2, (x, y) -> p([x, y][:,:])[target_idx], 
             fill=true, alpha=0.5, title="Target: $target", cbar=true,
@@
     measure=measures,
-    suppress_training=true, dataname="California Housing",
+    suppress_training=true, dataname="Moons",
     n_individuals=5,
-    factual=0, target=1,
+    target=0, factual=1,
     initialization=:identity,

### 2023-05-04 notebooks/moons.qmd
finding that sweet sweet architecture

--- 2023-05-04 notebooks/circles.qmd
@@
 n_obs = Int(1000 / (1.0 - test_size))
-counterfactual_data, test_data = train_test_split(load_circles(n_obs); test_size=test_size)
+counterfactual_data, test_data = train_test_split(load_moons(n_obs; noise=0.05, factor=0.5); test_size=test_size)
 X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
@@
         )
-        savefig(plt, joinpath(output_images_path, "circles_generated_$(mod_name).png"))
+        savefig(plt, joinpath(output_images_path, "moons_generated_$(mod_name).png"))
         push!(plts, plt)

### 2023-05-04 notebooks/poc.qmd
finding that sweet sweet architecture

```{julia}
include("notebooks/setup.jl")
eval(setup_notebooks)
```

```{julia}
# Counteractual data:
counterfactual_data = load_moons(2500; noise=0.1)
X = counterfactual_data.X
y = counterfactual_data.y
labels = counterfactual_data.output_encoder.labels
input_dim, nobs = size(X)
batch_size = Int(round(nobs/10))
epochs = 500
```

```{julia}
Plots.plot()
display(Plots.scatter!(counterfactual_data))
```

```{julia}
𝒟x = Normal()
𝒟y = Categorical(ones(2) ./ 2)
sampler = ConditionalSampler(𝒟x, 𝒟y, input_size=size(X)[1:end-1], batch_size=25)
n_hidden = 32
clf = JointEnergyClassifier(
    sampler;
    builder=MLJFlux.MLP(
        hidden=(n_hidden, n_hidden, n_hidden), 
        σ=Flux.relu
    ),
    batch_size=batch_size,
    finaliser=x -> x,
    loss=Flux.Losses.logitcrossentropy,
    jem_training_params=(
        α=[1.0,1.0,1e-1],
        verbosity=50,
    ),
    epochs=epochs,
    sampling_steps=30,
)
```

```{julia}
method = :simple_inductive
cov = .95
conf_model = conformal_model(clf; method=method, coverage=cov)
mach = machine(conf_model, table(permutedims(X)), labels)
fit!(mach)
```

```{julia}
#| echo: false
niter = 1000
jem = mach.model.model.jem
batch_size = mach.model.model.batch_size
X = Float32.(matrix(X))
if typeof(jem.sampler) <: ConditionalSampler

    plts = []
    for target in 1:2
        X̂ = generate_conditional_samples(jem, batch_size, target; niter=niter) 
        ex = extrema(hcat(X,X̂), dims=2)
        xlims = ex[1]
        ylims = ex[2]
        x1 = range(1.0f0.*xlims...,length=100)
        x2 = range(1.0f0.*ylims...,length=100)
        plt = Plots.contour(
            x1, x2, (x, y) -> softmax(jem([x, y]))[target], 
            fill=true, alpha=0.5, title="Target: $target", cbar=true,
            xlims=xlims,
            ylims=ylims,
        )
        Plots.scatter!(X[1,:], X[2,:], color=Int.(labels.refs), group=Int.(labels.refs), alpha=0.5)
        Plots.scatter!(
            X̂[1,:], X̂[2,:], 
            color=repeat([target], size(X̂,2)), 
            group=repeat([target], size(X̂,2)), 
            shape=:star5, ms=10
        )
        push!(plts, plt)
    end
    plt = Plots.plot(plts..., layout=(1, 2), size=(2*500, 400))
    display(plt)
end
```

```{julia}
M = ECCCo.ConformalModel(mach.model, mach.fitresult)
factual_label =  levels(labels)[2]
x_factual = reshape(X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
target =  levels(labels)[1]
factual = predict_label(M, counterfactual_data, x_factual)[1]

generator_dict = Dict(
    :wachter => WachterGenerator(),
    :revise => REVISEGenerator(),
    :greedy => GreedyGenerator(),
    :eccco => ECCCoGenerator(),
)

ces = []
plts = []
for (name, generator) in generator_dict
    ce = generate_counterfactual(
        x_factual, target, counterfactual_data, M, generator;
        initialization=:identity, 
        converge_when=:generator_conditions,
    )
    plt = Plots.plot(ce, title=name)
    push!(plts, plt)
    push!(ces, ce)
end
display(Plots.plot(plts...))
```

### 2023-05-04 src/penalties.jl

@@ -50,11 +50,11 @@ function distance_from_energy(
         push!(conditional_samples, rand(sampler, n; from_buffer=from_buffer))
     end
 
-    loss = map(eachcol(conditional_samples[1])) do xsample
-        distance_l1(ce; from=xsample)
+    _loss = map(eachcol(conditional_samples[1])) do xsample
+        distance_l1(ce; from=xsample, agg=agg)
     end
-    loss = agg(loss)
-    return loss
+    _loss = reduce((x,y) -> x + y, _loss) / n       # aggregate over samples
+    return _loss
 
 end
 

### 2023-05-05 notebooks/poc.qmd
good specs for moons :moon:

@@
 𝒟y = Categorical(ones(2) ./ 2)
-sampler = ConditionalSampler(𝒟x, 𝒟y, input_size=size(X)[1:end-1], batch_size=25)
+sampler = ConditionalSampler(𝒟x, 𝒟y, input_size=size(X)[1:end-1], batch_size=10)
 n_hidden = 32
@@
     epochs=epochs,
-    sampling_steps=30,
+    sampling_steps=50,
 )
@@
 ```{julia}
-M = ECCCo.ConformalModel(mach.model, mach.fitresult)
-factual_label =  levels(labels)[2]
-x_factual = reshape(X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
-target =  levels(labels)[1]
-factual = predict_label(M, counterfactual_data, x_factual)[1]
+# M = ECCCo.ConformalModel(mach.model, mach.fitresult)
+# factual_label =  levels(labels)[2]
+# x_factual = reshape(X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
+# target =  levels(labels)[1]
+# factual = predict_label(M, counterfactual_data, x_factual)[1]
 
-generator_dict = Dict(
-    :wachter => WachterGenerator(),
-    :revise => REVISEGenerator(),
-    :greedy => GreedyGenerator(),
-    :eccco => ECCCoGenerator(),
-)
+# generator_dict = Dict(
+#     :wachter => WachterGenerator(),
+#     :revise => REVISEGenerator(),
+#     :greedy => GreedyGenerator(),
+#     :eccco => ECCCoGenerator(),
+# )
 
-ces = []
-plts = []
-for (name, generator) in generator_dict
-    ce = generate_counterfactual(
-        x_factual, target, counterfactual_data, M, generator;
-        initialization=:identity, 
-        converge_when=:generator_conditions,
-    )
-    plt = Plots.plot(ce, title=name)
-    push!(plts, plt)
-    push!(ces, ce)
-end
-display(Plots.plot(plts...))
+# ces = []
+# plts = []
+# for (name, generator) in generator_dict
+#     ce = generate_counterfactual(
+#         x_factual, target, counterfactual_data, M, generator;
+#         initialization=:identity, 
+#         converge_when=:generator_conditions,
+#     )
+#     plt = Plots.plot(ce, title=name)
+#     push!(plts, plt)
+#     push!(ces, ce)
+# end
+# display(Plots.plot(plts...))
 ```

### 2023-05-05 notebooks/poc.qmd
more work on moons for poc

@@
 ```{julia}
-# M = ECCCo.ConformalModel(mach.model, mach.fitresult)
-# factual_label =  levels(labels)[2]
-# x_factual = reshape(X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
-# target =  levels(labels)[1]
-# factual = predict_label(M, counterfactual_data, x_factual)[1]
+Random.seed!(1234)
 
-# generator_dict = Dict(
-#     :wachter => WachterGenerator(),
-#     :revise => REVISEGenerator(),
-#     :greedy => GreedyGenerator(),
-#     :eccco => ECCCoGenerator(),
-# )
+λ₁ = 0.1
+λ₂ = 0.5
+λ₃ = 0.5
+Λ = [λ₁, λ₂, λ₃]
 
-# ces = []
-# plts = []
-# for (name, generator) in generator_dict
-#     ce = generate_counterfactual(
-#         x_factual, target, counterfactual_data, M, generator;
-#         initialization=:identity, 
-#         converge_when=:generator_conditions,
-#     )
-#     plt = Plots.plot(ce, title=name)
-#     push!(plts, plt)
-#     push!(ces, ce)
-# end
-# display(Plots.plot(plts...))
+M = ECCCo.ConformalModel(mach.model, mach.fitresult)
+factual_label =  levels(labels)[1]
+x_factual = reshape(X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
+target =  levels(labels)[2]
+factual = predict_label(M, counterfactual_data, x_factual)[1]
+γ = 0.5
+
+generator_dict = OrderedDict(
+    "Wachter (γ=$γ)" => WachterGenerator(λ = λ₁),
+    "Schut" => GreedyGenerator(λ = λ₁),
+    "REVISE" => REVISEGenerator(λ = λ₁),
+    "ECCCo" => ECCCoGenerator(λ = Λ),
+)
+
+ces = Dict{Any,Any}()
+plts = []
+for (name, generator) in generator_dict
+    conv_when = name == "Wachter (γ=$γ)" ? :decision_threshold : :generator_conditions
+    ce = generate_counterfactual(
+        x_factual, target, counterfactual_data, M, generator;
+        initialization=:identity, 
+        converge_when=conv_when,
+        decision_threshold=γ,
+    )
+    plt = Plots.plot(ce, title=name, alpha=0.2, cbar=false, axis=nothing)
+    if name == "ECCCo"
+        _X = distance_from_energy(ce, return_conditionals=true)
+        Plots.scatter!(
+            _X[1,:],_X[2,:], color=:purple, shape=:star5, 
+            ms=10, label="x̂|$target", alpha=0.5
+        )
+    end
+    push!(plts, plt)
+    ces[name] = ce
+end
+plt = Plots.plot(plts..., size=(500,520))
+display(plt)
+savefig(plt, joinpath(output_images_path, "poc.png"))
 ```
+
+```{julia}
+ce = ces["ECCCo"]
+```

### 2023-05-05 src/sampling.jl

@@ -55,12 +55,10 @@ function EnergySampler(
     yidx = get_target_index(data.y_levels, y)
 
     # Initiate:
-    energy_sampler = EnergySampler(model, data, sampler, opt, nothing, nothing)
+    energy_sampler = EnergySampler(model, data, sampler, opt, nothing, yidx)
 
-    # Generate conditional samples (one at a time):
-    for i in 1:nsamples
-        generate_samples!(energy_sampler, 1, yidx; niter=niter)
-    end
+    # Generate conditional:
+    generate_samples!(energy_sampler, nsamples, yidx; niter=niter)
 
     return energy_sampler
 end
@@ -130,4 +128,17 @@ function Base.rand(sampler::EnergySampler, n::Int=100; from_buffer=true, niter::
         X = generate_samples(sampler, n, sampler.yidx; niter=niter)
     end
     return X
+end
+
+"""
+    get_lowest_energy_sample(sampler::EnergySampler; n::Int=5)
+
+Chooses the samples with the lowest energy (i.e. highest probability) from `EnergySampler`.
+"""
+function get_lowest_energy_sample(sampler::EnergySampler; n::Int=5)
+    X = sampler.buffer
+    model = sampler.model
+    y = sampler.yidx
+    x = selectdim(X, ndims(X), energy(sampler.sampler, model, X, y; agg=x -> partialsortperm(x, 1:n)))
+    return x
 end

### 2023-05-05 notebooks/circles.qmd
results checking out now for synthetic data

@@
     "MLP" => mlp,
-    "MLP Ensemble" => mlp_ens,
+    # "MLP Ensemble" => mlp_ens,
     "JEM" => jem,
-    "JEM Ensemble" => jem_ens,
+    # "JEM Ensemble" => jem_ens,
     # "JEM Ensemble+" => jem_ens_plus,
@@
     model_dict = Dict(mod_name => _train(model; mod_name=mod_name) for (mod_name, model) in models)
-    Serialization.serialize(joinpath(output_path,"moons_models.jls"), model_dict)
+    Serialization.serialize(joinpath(output_path,"circles_models.jls"), model_dict)
 else
-    model_dict = Serialization.deserialize(joinpath(output_path,"moons_models.jls"))
+    model_dict = Serialization.deserialize(joinpath(output_path,"circles_models.jls"))
 end
@@
 end
-Serialization.serialize(joinpath(output_path,"moons_model_performance.jls"), model_performance)
-CSV.write(joinpath(output_path, "moons_model_performance.csv"), model_performance)
+Serialization.serialize(joinpath(output_path,"circles_model_performance.jls"), model_performance)
+CSV.write(joinpath(output_path, "circles_model_performance.csv"), model_performance)
 model_performance
@@
 ```{julia}
+λ₁ = 0.25
+λ₂ = 0.75
+λ₃ = 0.75
+Λ = [λ₁, λ₂, λ₃]
+
 # Benchmark generators:
 generator_dict = Dict(
-    :wachter => WachterGenerator(),
-    :revise => REVISEGenerator(),
-    :greedy => GreedyGenerator(),
-    :eccco => ECCCoGenerator(),
+    "Wachter" => WachterGenerator(λ=λ₁),
+    "REVISE" => REVISEGenerator(λ=λ₁),
+    "Schut" => GreedyGenerator(),
+    "ECCCo" => ECCCoGenerator(λ=Λ),
 )
+```
 
+### POC
+
+```{julia}
+Random.seed!(2023)
+
+M = model_dict["JEM"]
+X = X isa Matrix ? X : Float32.(permutedims(matrix(X)))
+factual_label =  levels(labels)[1]
+x_factual = reshape(X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
+target =  levels(labels)[2]
+factual = predict_label(M, counterfactual_data, x_factual)[1]
+
+ces = Dict{Any,Any}()
+plts = []
+for (name, generator) in generator_dict
+    ce = generate_counterfactual(
+        x_factual, target, counterfactual_data, M, generator;
+        initialization=:identity, 
+        converge_when=:generator_conditions,
+    )
+    plt = Plots.plot(ce, title=name, alpha=0.2, cbar=false, axis=nothing)
+    if name == "ECCCo"
+        _X = distance_from_energy(ce, return_conditionals=true)
+        Plots.scatter!(
+            _X[1,:],_X[2,:], color=:purple, shape=:star5, 
+            ms=10, label="x̂|$target", alpha=0.5
+        )
+    end
+    push!(plts, plt)
+    ces[name] = ce
+end
+plt = Plots.plot(plts..., size=(500,520))
+display(plt)
+savefig(plt, joinpath(output_images_path, "circles_poc.png"))
+```
+
+### Complete Benchmark
+
+```{julia}
 # Measures:
@@
     CounterfactualExplanations.Evaluation.redundancy,
+    ECCCo.set_size_penalty
 ]
@@
     initialization=:identity,
+    converge_when=:generator_conditions,
 )
-CSV.write(joinpath(output_path, "moons_benchmark.csv"), bmk())
+CSV.write(joinpath(output_path, "circles_benchmark.csv"), bmk())
 ```
@@
 display(plt)
-save(joinpath(output_images_path, "moons_benchmark.png"), plt, px_per_unit=5)
+save(joinpath(output_images_path, "circles_benchmark.png"), plt, px_per_unit=5)
 ```

### 2023-05-05 notebooks/moons.qmd
results checking out now for synthetic data

@@
 test_size = 0.2
-n_obs = Int(1000 / (1.0 - test_size))
-counterfactual_data, test_data = train_test_split(load_moons(n_obs; noise=0.05, factor=0.5); test_size=test_size)
+n_obs = Int(2500 / (1.0 - test_size))
+counterfactual_data, test_data = train_test_split(load_moons(n_obs); test_size=test_size)
 X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
@@
 # Model parameters:
-epochs = 100
+epochs = 500
 batch_size = minimum([Int(round(n_obs/10)), 128])
 n_hidden = 32
-activation = Flux.swish
+activation = Flux.relu
 builder = MLJFlux.MLP(
@@
     ),
-    sampling_steps=20,
+    sampling_steps=30,
 )
@@
 ```{julia}
-n_regen = 200
+n_regen = 500
 n_each = batch_size
@@
 ```{julia}
+λ₁ = 0.25
+λ₂ = 0.75
+λ₃ = 0.75
+Λ = [λ₁, λ₂, λ₃]
+
 # Benchmark generators:
 generator_dict = Dict(
-    :wachter => WachterGenerator(),
-    :revise => REVISEGenerator(),
-    :greedy => GreedyGenerator(),
-    :eccco => ECCCoGenerator(),
+    "Wachter" => WachterGenerator(λ=λ₁),
+    "REVISE" => REVISEGenerator(λ=λ₁),
+    "Schut" => GreedyGenerator(),
+    "ECCCo" => ECCCoGenerator(λ=Λ),
 )
+```
 
+### POC
+
+```{julia}
+Random.seed!(2023)
+
+M = model_dict["JEM"]
+X = X isa Matrix ? X : Float32.(permutedims(matrix(X)))
+factual_label =  levels(labels)[1]
+x_factual = reshape(X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
+target =  levels(labels)[2]
+factual = predict_label(M, counterfactual_data, x_factual)[1]
+
+ces = Dict{Any,Any}()
+plts = []
+for (name, generator) in generator_dict
+    ce = generate_counterfactual(
+        x_factual, target, counterfactual_data, M, generator;
+        initialization=:identity, 
+        converge_when=:generator_conditions,
+    )
+    plt = Plots.plot(ce, title=name, alpha=0.2, cbar=false, axis=nothing)
+    if name == "ECCCo"
+        _X = distance_from_energy(ce, return_conditionals=true)
+        Plots.scatter!(
+            _X[1,:],_X[2,:], color=:purple, shape=:star5, 
+            ms=10, label="x̂|$target", alpha=0.5
+        )
+    end
+    push!(plts, plt)
+    ces[name] = ce
+end
+plt = Plots.plot(plts..., size=(500,520))
+display(plt)
+savefig(plt, joinpath(output_images_path, "moons_poc.png"))
+```
+
+```{julia}
+evaluate(collect(values(ces)))
+```
+
+```{julia}
 # Measures:
@@
     n_individuals=5,
-    target=0, factual=1,
+    target=1, factual=0,
     initialization=:identity,
+    converge_when=:generator_conditions,
 )

### 2023-05-05 notebooks/poc.qmd
results checking out now for synthetic data

@@
 fit!(mach)
+Serialization.serialize(joinpath(output_path,"poc_model.jls"), mach)
 ```

### 2023-05-05 src/penalties.jl

@@ -1,7 +1,6 @@
 using ChainRules: ignore_derivatives
 using Distances
 using Flux
-using JointEnergyModels.Samplers: get_lowest_energy_sample
 using LinearAlgebra: norm
 using Statistics: mean
 
@@ -39,10 +38,10 @@ end
 
 function distance_from_energy(
     ce::AbstractCounterfactualExplanation;
-    n::Int=50, niter=500, from_buffer=true, agg=mean, 
+    n::Int=100, niter=500, from_buffer=true, agg=mean, 
     choose_lowest_energy=true,
     choose_random=false,
-    nmin::Int=25,
+    nmin::Int=50,
     return_conditionals=false,
     kwargs...
 )

### 2023-05-05 notebooks/cal_housing.qmd
:star-struck: :star-struck: :star-struck:

@@
 test_size = 0.2
-n_obs = Int(1000 / (1.0 - test_size))
+n_obs = Int(10000 / (1.0 - test_size))
 counterfactual_data, test_data = train_test_split(load_california_housing(n_obs); test_size=test_size)
@@
     ),
-    sampling_steps=50,
+    sampling_steps=30,
 )
@@
 ```{julia}
+λ₁ = 0.25
+λ₂ = 0.75
+λ₃ = 0.75
+Λ = [λ₁, λ₂, λ₃]
+
 # Benchmark generators:
 generator_dict = Dict(
-    :wachter => WachterGenerator(),
-    :revise => REVISEGenerator(),
-    :greedy => GreedyGenerator(),
-    :eccco => ECCCoGenerator(),
+    "Wachter" => WachterGenerator(λ=λ₁),
+    "REVISE" => REVISEGenerator(λ=λ₁),
+    "Schut" => GreedyGenerator(),
+    "ECCCo" => ECCCoGenerator(λ=Λ),
 )
+```
 
+```{julia}
 # Measures:

### 2023-05-05 notebooks/gmsc.qmd
:star-struck: :star-struck: :star-struck:

@@
 batch_size = minimum([Int(round(n_obs/10)), 128])
-n_hidden = 256
+n_hidden = 128
 activation = Flux.relu
@@
     Dense(n_in, n_hidden, activation),
-    Dense(n_hidden, n_hidden, activation),
-    Dense(n_hidden, n_hidden, activation),
     Dense(n_hidden, n_hidden, activation),
@@
     ),
-    sampling_steps=60,
+    sampling_steps=30,
 )
@@
     measure=measures,
-    suppress_training=true, dataname="California Housing",
+    suppress_training=true, dataname="GMSC",
     n_individuals=5,

### 2023-05-05 notebooks/poc.qmd
:star-struck: :star-struck: :star-struck:

@@
 # Counteractual data:
-counterfactual_data = load_moons(2500; noise=0.1)
+counterfactual_data = load_linearly_separable(1000)
 X = counterfactual_data.X
@@
 batch_size = Int(round(nobs/10))
-epochs = 500
+epochs = 100
 ```
@@
 𝒟y = Categorical(ones(2) ./ 2)
-sampler = ConditionalSampler(𝒟x, 𝒟y, input_size=size(X)[1:end-1], batch_size=10)
-n_hidden = 32
+sampler = ConditionalSampler(𝒟x, 𝒟y, input_size=size(X)[1:end-1], batch_size=50)
+n_hidden = 16
 clf = JointEnergyClassifier(
@@
         hidden=(n_hidden, n_hidden, n_hidden), 
-        σ=Flux.relu
+        σ=Flux.swish
     ),
@@
         α=[1.0,1.0,1e-1],
-        verbosity=50,
+        verbosity=10,
     ),
     epochs=epochs,
-    sampling_steps=50,
+    sampling_steps=30,
 )
@@
 M = ECCCo.ConformalModel(mach.model, mach.fitresult)
-factual_label =  levels(labels)[1]
+factual_label =  levels(labels)[2]
 x_factual = reshape(X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
-target =  levels(labels)[2]
+target =  levels(labels)[1]
 factual = predict_label(M, counterfactual_data, x_factual)[1]
-γ = 0.5
 
@@
 for (name, generator) in generator_dict
-    conv_when = name == "Wachter (γ=$γ)" ? :decision_threshold : :generator_conditions
     ce = generate_counterfactual(
@@
         initialization=:identity, 
-        converge_when=conv_when,
+        converge_when=:generator_conditions,
         decision_threshold=γ,
     )
-    plt = Plots.plot(ce, title=name, alpha=0.2, cbar=false, axis=nothing)
+    plt = Plots.plot(ce, title=name, alpha=0.2, cbar=false, axis=nothing, length_out=10)
     if name == "ECCCo"

### 2023-05-05 paper/paper.tex
more work on moons for poc
results checking out now for synthetic data
:star-struck: :star-struck: :star-struck:

@@
 
-\title{ECCCos from the Black Box: Letting Models speak for Themselves}
+\title{ECCCos from the Black Box: Faithful Counterfactual Explanations through Energy Based Modeling and Conformal Prediction}
 
@@
 This effectively reallocates the task of learning realistic representations of the data from the model itself to the surrogate.
-Consequently, the generated explanations may look plausible to humans but not necessarily conform with the behaviour of the black-box model.
-We formalise this notion of model conformity through the introduction of tailored evaluation measures and propose a novel algorithmic framework for generating \textbf{E}nergy-\textbf{C}onstrained \textbf{C}onformal \textbf{Co}unterfactuals that are only as plausible as the model permits.
-To do so, \textbf{ECCCo} leverages recent advances in energy-based modelling and predictive uncertainty quantification through conformal inference.
-Through illustrative examples and extensive empirical studies, we demonstrate that ECCCos reconcile the need for plausibility and model conformity.
+Consequently, the generated explanations may look plausible to humans but not necessarily faithfully describe the behaviour of the black-box model.
+We formalise this notion of faithfulness through the introduction of a tailored evaluation metric and propose a novel algorithmic framework for generating \textbf{E}nergy-\textbf{C}onstrained \textbf{C}onformal \textbf{Co}unterfactuals that are only as plausible as the model permits.
+To do so, \textbf{ECCCo} leverages recent advances with respect to energy-based models and conformal inference.
+Through extensive empirical studies, we demonstrate that ECCCos reconcile the need for plausibility and faithfulness.
+Our findings also highlight that black-box models that are trained jointly to discriminate outputs and generate inputs yield more plausible explanations than pure discriminative models.
 \end{abstract}
@@
   \centering
-  \includegraphics[width=\textwidth]{../artifacts/results/images/eccco_illustration.png}
+  \includegraphics[width=\textwidth]{../artifacts/results/images/poc.png}
   \captionof{figure}{[PLACEHOLDER] Vector fields indicating the direction of gradients with respect to the different components of the ECCCo objective (Equation~\ref{eq:eccco}).} \label{fig:eccco}
@@
   \item For MNIST it seems that ECCCo is better at reducing pixel values than increasing them (better at erasing than writing)
+  \item JEMs are more difficult to train
 \end{itemize}

### 2023-05-05 notebooks/poc.qmd
some intial work on illustration

@@
             _X[1,:],_X[2,:], color=:purple, shape=:star5, 
-            ms=10, label="x̂|$target", alpha=0.5
+            ms=10, label="x̂|$target", alpha=0.1
         )
@@
 ce = ces["ECCCo"]
+using CounterfactualExplanations.Generators: ℓ
+
+# loss
+function f1(x)
+    _ce = deepcopy(ce)
+    _ce.s′ = x
+    return - ℓ(_ce.generator, _ce)
+end
+
+# cost
+function f2(x)
+    _ce = deepcopy(ce)
+    _ce.s′ = x
+    λ = _ce.generator.λ[1]
+    _loss = _ce.generator.penalty[1]
+    return - λ * _loss(_ce)
+end
+
+# set size
+function f3(x)
+    _ce = deepcopy(ce)
+    _ce.s′ = x
+    λ = _ce.generator.λ[2]
+    _loss = _ce.generator.penalty[2]
+    return - λ * _loss(_ce)
+end
+
+# distance from energy
+function f4(x)
+    _ce = deepcopy(ce)
+    _ce.s′ = x
+    λ = _ce.generator.λ[3]
+    _loss = _ce.generator.penalty[3]
+    return - λ * _loss(_ce)
+end
+
+# Helper function:
+meshgrid(x, y) = (repeat(x, outer=length(y)), repeat(y, inner=length(x)))
+
+xlims, ylims = extrema(X, dims=2)
+xrange = range(xlims..., length=10)
+yrange = range(ylims..., length=10)
+
+x1, x2 = meshgrid(xrange, yrange)
+inputs = zip(x, y)
+
+u = []
+v = []
+
+scale = 0.1
+for (x, y) in inputs
+    push!(u, scale * gradient(f2, [x, y][:,:])[1][1])
+    push!(v, scale * gradient(f2, [x, y][:,:])[1][2])
+end
+
+Plots.plot(xlims=xlims, ylims=ylims)
+Plots.scatter!(counterfactual_data)
+Plots.quiver!(x1, x2, quiver=(u, v))
 ```

### 2023-05-05 src/penalties.jl

@@ -14,10 +14,12 @@ function set_size_penalty(
     κ::Real=0.0, temp::Real=0.05, agg=mean
 )
 
+    _loss = 0.0
+
     conf_model = ce.M.model
     fitresult = ce.M.fitresult
     X = CounterfactualExplanations.decode_state(ce)
-    loss = map(eachslice(X, dims=ndims(X))) do x
+    _loss = map(eachslice(X, dims=ndims(X))) do x
         x = ndims(x) == 1 ? x[:,:] : x
         if target_probs(ce, x)[1] >= 0.5
             l = ConformalPrediction.smooth_size_loss(
@@ -30,9 +32,9 @@ function set_size_penalty(
         end
         return l
     end
-    loss = agg(loss)
+    _loss = agg(_loss)
 
-    return loss
+    return _loss
 
 end
 
@@ -45,6 +47,8 @@ function distance_from_energy(
     return_conditionals=false,
     kwargs...
 )
+
+    _loss = 0.0
 
     @assert choose_lowest_energy ⊻ choose_random || !choose_lowest_energy && !choose_random "Must choose either lowest energy or random samples or neither."
 

### 2023-05-07 notebooks/results.qmd
work on table

```{julia}
include("notebooks/setup.jl")
eval(setup_notebooks)
```

# Results

```{julia}
df = DataFrame()
for _file in filter(endswith("_benchmark.csv"), readdir(output_path)) 
    x = joinpath(output_path, _file)
    _df = CSV.read(x, DataFrame)
    df = vcat(df, _df, cols=:intersect)
end
synth = ("Moons", "Circles")
df.source .= ifelse.(df.dataname .∈ [synth], :synthetic, :real) 
```

```{julia}
tab = @chain df begin
    @group_by(dataname, generator, model, variable, source)
    @summarize(mean=mean(value),sd=std(value))
    @ungroup
    @filter(variable ∈ ["distance_from_energy", "distance_from_targets"])
    @mutate(variable = ifelse.(variable .== "distance_from_energy", "Non-Conformity", variable))
    @mutate(variable = ifelse.(variable .== "distance_from_targets", "Implausibility", variable))
end
```

### 2023-05-07 notebooks/tables.Rmd
work on table

```{r}
library(data.table)
library(kableExtra)
```

```{r}
files <- list.files("artifacts/results/")
dt <- lapply(files[grepl("_benchmark.csv", files)], function(x) {
    fread(file.path("artifacts/results/", x))
})
dt <- Reduce(function(x,y) {rbind(x,y, fill=TRUE)}, dt)
dt[,ce:=NULL]
synth <- c("Moons", "Circles")
dt[,source:=ifelse(dataname %in% synth, "synthetic", "real")]
```

```{r}
measures <- c("distance_from_energy", "distance_from_targets")
tab <- dt[
    variable %in% measures,
    .(value=sprintf("%1.2f\n(%1.2f)", mean(value), sd(value))),
    .(dataname, generator, model, variable, source)
]
tab[generator == "eccco", generator:="ECCCo"]
tab[generator == "revise", generator:="REVISE"]
tab[generator == "generic", generator:="Wachter"]
tab[generator == "wachter", generator:="Wachter"]
tab[generator == "greedy", generator:="Schut"]
tab <- dcast(tab, model + variable + generator ~ dataname)
kbl(tab, align = "c") %>%
  kable_paper(full_width = F) %>%
  column_spec(1, bold = T) %>%
  collapse_rows(columns = 1:3, valign = "top")
```

### 2023-05-08 notebooks/circles.qmd
code for tables mostly sorted

@@
     measure=measures,
-    suppress_training=true, dataname="Moons",
+    suppress_training=true, dataname="Circles",
     n_individuals=5,

### 2023-05-08 notebooks/mnist.qmd
code for tables mostly sorted

@@
 ```{julia}
-Random.seed!(1234)
+Random.seed!(2023)
 
@@
 T = 100
-opt = Flux.Optimise.Adam(0.01)
+opt = Flux.Optimise.Adam(0.1)
 ```
@@
 # Search:
-generic_generator = GenericGenerator(opt=opt)
+generic_generator = GenericGenerator(opt=Flux.Optimise.Descent(1.0))
 ce_wachter = generate_counterfactual(
@@
 )
-greedy_generator = GreedyGenerator(η=1.0)
+greedy_generator = GreedyGenerator(η=2.0)
 ce_jsma = generate_counterfactual(
@@
 # JEM parameters:
-𝒟x = Uniform(0,1)
+𝒟x = Uniform(-1.0,1.0)
 𝒟y = Categorical(ones(output_dim) ./ output_dim)
@@
 # Plot generated samples:
-n_regen = 200
+n_regen = 500
 if _regen 

### 2023-05-08 notebooks/tables.Rmd
code for tables mostly sorted

@@
 tab <- dt[
-    variable %in% measures,
-    .(value=sprintf("%1.2f\n(%1.2f)", mean(value), sd(value))),
-    .(dataname, generator, model, variable, source)
+    variable %in% measures & source=="real",
+    .(
+      value=sprintf("%1.2f (%1.2f)", mean(value), sd(value)),
+      val = mean(value)
+    ),
+    .(dataname, generator, model, variable)
 ]
+tab[,min_val:=val==min(val),.(model, dataname, variable)]
+tab$value <- cell_spec(tab$value, "latex", bold=tab$min_val)
+tab[,val:=NULL]
+tab[,min_val:=NULL]
 tab[generator == "eccco", generator:="ECCCo"]
@@
 tab[generator == "greedy", generator:="Schut"]
-tab <- dcast(tab, model + variable + generator ~ dataname)
-kbl(tab, align = "c") %>%
+n_data <- length(unique(tab$dataname))
+tab <- dcast(tab, model + generator ~ dataname + variable)
+col_names <- c(
+  "Model",
+  "Generator",
+  rep(c("Non-conformity ↓","Implausibility ↓"),n_data)
+)
+kbl(
+  tab, caption = "Results for real-world datasets. Standard deviation in parentheses. \\label{tab:results-real}", 
+  align = "c", col.names=col_names, booktabs = T, escape=F, 
+  format="latex", linesep = c("", "", "", "\\addlinespace") 
+) %>%
+  kable_styling(latex_options = c("scale_down")) %>%
+  add_header_above(c(" " = 2, "California Housing" = 2, "GMSC" = 2, "MNIST" = 2)) %>%
+  column_spec(1, bold = T) %>%
+  collapse_rows(columns = 1:2, latex_hline = "major", valign = "middle") %>%
+  save_kable("paper/contents/table_real.tex")
+```
+
+```{r}
+measures <- c("distance_from_energy", "distance_from_targets")
+tab <- dt[
+    variable %in% measures & source=="synthetic",
+    .(
+      value=sprintf("%1.2f (%1.2f)", mean(value), sd(value)),
+      val = mean(value)
+    ),
+    .(dataname, generator, model, variable)
+]
+tab[,min_val:=val==min(val),.(model, dataname, variable)]
+tab$value <- cell_spec(tab$value, "latex", bold=tab$min_val)
+tab[,val:=NULL]
+tab[,min_val:=NULL]
+tab[generator == "eccco", generator:="ECCCo"]
+tab[generator == "revise", generator:="REVISE"]
+tab[generator == "generic", generator:="Wachter"]
+tab[generator == "wachter", generator:="Wachter"]
+tab[generator == "greedy", generator:="Schut"]
+n_data <- length(unique(tab$dataname))
+tab <- dcast(tab, model + generator ~ dataname + variable)
+col_names <- c(
+  "Model",
+  "Generator",
+  rep(c("Non-conformity ↓","Implausibility ↓"),n_data)
+)
+kbl(
+  tab, caption = "Results for synthetic datasets. Standard deviation in parentheses. \\label{tab:results-synth}", 
+  align = "c", col.names=col_names, booktabs = T, escape=F, 
+  format="latex", linesep = c("", "", "", "\\addlinespace") 
+) %>%
+  # kable_styling(latex_options = c("scale_down")) %>%
   kable_paper(full_width = F) %>%
+  add_header_above(c(" " = 2, "Moons" = 2)) %>%
   column_spec(1, bold = T) %>%
-  collapse_rows(columns = 1:3, valign = "top")
+  collapse_rows(columns = 1:2, latex_hline = "major", valign = "middle") %>%
+  save_kable("paper/contents/table_synth.tex")
 ```

### 2023-05-08 paper/paper.tex
work on table
code for tables mostly sorted

@@
 \usepackage{algpseudocode}
+\usepackage{import}
+\usepackage{booktabs}
+\usepackage{longtable}
+\usepackage{array}
+\usepackage{multirow}
 
@@
 
-\title{ECCCos from the Black Box: Faithful Counterfactual Explanations through Energy Based Modeling and Conformal Prediction}
+\title{ECCCos from the Black Box:\\
+Faithful Explanations through\\
+Energy-Constrained Conformal Counterfactuals}
 
@@
   Counterfactual Explanations offer an intuitive and straightforward way to explain black-box models but they are not unique.
-To address the need for plausible explanations, existing work has primarily relied on surrogate models to learn how the input data is distributed.
+To identify the most plausible explanations, existing work has primarily relied on surrogate models to learn how the input data is distributed.
 This effectively reallocates the task of learning realistic representations of the data from the model itself to the surrogate.
@@
 We formalise this notion of faithfulness through the introduction of a tailored evaluation metric and propose a novel algorithmic framework for generating \textbf{E}nergy-\textbf{C}onstrained \textbf{C}onformal \textbf{Co}unterfactuals that are only as plausible as the model permits.
-To do so, \textbf{ECCCo} leverages recent advances with respect to energy-based models and conformal inference.
-Through extensive empirical studies, we demonstrate that ECCCos reconcile the need for plausibility and faithfulness.
-Our findings also highlight that black-box models that are trained jointly to discriminate outputs and generate inputs yield more plausible explanations than pure discriminative models.
+Through extensive empirical studies involving multiple synthetic and real-world datasets, we demonstrate that ECCCos reconcile the need for plausibility and faithfulness.
+In particular, we show that it is possible to achieve state-of-the-art plausibility for any black-box model with gradient access without the need for surrogate models.
+To do so, \textbf{ECCCo} relies solely on properties defining the black-box model itself by leveraging recent advances in energy-based modelling and conformal inference.
+The empirical findings also highlight that black-box models that are trained jointly to discriminate outputs and generate inputs tend to yield more plausible explanations than pure discriminative models.
+Our framework is intuitive, flexible and open-sourced.
+By highlighting the need for faithfulness in the context of Counterfactual Explanations, we believe that in the short term, our work will enable researchers and practitioners to better distinguish trustworthy from unreliable models.
+We anticipate that ECCCo can serve as a baseline for future research directed at providing plausible but faithful Counterfactual Explanations.
 \end{abstract}
@@
 
-\begin{figure}
-  \includegraphics[width=\textwidth]{../artifacts/results/images/mnist_benchmark.png}
-  \caption{[SUBJECTO TO CHANGE] Original image (left) and ECCCos for turning an 8 (eight) into a 3 (three) for different Black Boxes from left to right: Multi-Layer Perceptron (MLP), Ensemble of MLPs, Joint Energy Model (JEM), Ensemble of JEMs.}\label{fig:mnist-benchmark}
-\end{figure}
+See Table~\ref{tab:results-real}
+
+\import{contents/}{table_synth.tex}
+
+\import{contents/}{table_real.tex}
 

### 2023-05-08 notebooks/mnist.qmd
code for tables mostly sorted

@@
 ```{julia}
-Random.seed!(2023)
+Random.seed!(1234)
 
@@
 # Search:
-generic_generator = GenericGenerator(opt=Flux.Optimise.Descent(1.0))
+generic_generator = GenericGenerator(opt=Flux.Optimise.Adam(1.0))
 ce_wachter = generate_counterfactual(

### 2023-05-08 notebooks/circles.qmd
added lin sep

--- 2023-05-05 notebooks/circles.qmd
@@
     "ECCCo" => ECCCoGenerator(λ=Λ),
+    "ECCCo (no CP)" => ECCCoGenerator(λ=[λ₁, 0.0, λ₃]),
+    "ECCCo (no EBM)" => ECCCoGenerator(λ=[λ₁, λ₂, 0.0]),
 )
@@
 ```{julia}
-Random.seed!(2023)
-
 M = model_dict["JEM"]
@@
 
+bmks = []
+for target in sort(unique(labels))
+    for factual in sort(unique(labels))
+        if factual == target
+            continue
+        end
         bmk = benchmark(
@@
             measure=measures,
-    suppress_training=true, dataname="Moons",
+            suppress_training=true, dataname="Circles",
             n_individuals=5,
-    target=0, factual=1,
+            target=target, factual=factual,
             initialization=:identity,
@@
         )
+        push!(bmks, bmk)
+    end
+end
+bmk = reduce(vcat, bmks)
 CSV.write(joinpath(output_path, "circles_benchmark.csv"), bmk())
-```
-
-```{julia}
-@chain bmk() begin
-    @group_by(dataname, generator, model, variable)
-    @summarize(mean=mean(value),sd=std(value))
-    @ungroup
-    @filter(variable == "distance_from_energy")
-end
 ```

### 2023-05-08 notebooks/linearly_separable.qmd
added lin sep

--- 2023-05-08 notebooks/circles.qmd
@@
 n_obs = Int(1000 / (1.0 - test_size))
-counterfactual_data, test_data = train_test_split(load_circles(n_obs; noise=0.05, factor=0.5); test_size=test_size)
+counterfactual_data, test_data = train_test_split(load_linearly_separable(n_obs); test_size=test_size)
 X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
@@
 batch_size = minimum([Int(round(n_obs/10)), 128])
-n_hidden = 32
+n_hidden = 16
 activation = Flux.swish

### 2023-05-08 paper/paper.tex
minor updates to abstract

@@
 \begin{abstract}
-  Counterfactual Explanations offer an intuitive and straightforward way to explain black-box models but they are not unique.
-To identify the most plausible explanations, existing work has primarily relied on surrogate models to learn how the input data is distributed.
+  Counterfactual Explanations offer an intuitive and straightforward way to explain black-box models and offer Algorithmic Recourse to individuals.
+To address the need for plausible explanations, existing work has primarily relied on surrogate models to learn how the input data is distributed.
 This effectively reallocates the task of learning realistic representations of the data from the model itself to the surrogate.
-Consequently, the generated explanations may look plausible to humans but not necessarily faithfully describe the behaviour of the black-box model.
+Consequently, the generated explanations may seem plausible to humans but need not necessarily faithfully describe the behaviour of the black-box model.
 We formalise this notion of faithfulness through the introduction of a tailored evaluation metric and propose a novel algorithmic framework for generating \textbf{E}nergy-\textbf{C}onstrained \textbf{C}onformal \textbf{Co}unterfactuals that are only as plausible as the model permits.
-Through extensive empirical studies involving multiple synthetic and real-world datasets, we demonstrate that ECCCos reconcile the need for plausibility and faithfulness.
-In particular, we show that it is possible to achieve state-of-the-art plausibility for any black-box model with gradient access without the need for surrogate models.
-To do so, \textbf{ECCCo} relies solely on properties defining the black-box model itself by leveraging recent advances in energy-based modelling and conformal inference.
-The empirical findings also highlight that black-box models that are trained jointly to discriminate outputs and generate inputs tend to yield more plausible explanations than pure discriminative models.
-Our framework is intuitive, flexible and open-sourced.
+Through extensive empirical studies involving multiple synthetic and real-world datasets, we demonstrate that \textbf{ECCCo} reconciles the need for plausibility and faithfulness.
+In particular, we show that it is possible to achieve state-of-the-art plausibility for models with gradient access without the need for surrogate models.
+To do so, ECCCo relies solely on properties defining the black-box model itself by leveraging recent advances in energy-based modelling and conformal inference.
+Through this work, we also shine new light on the explanatory properties of Joint Energy Models.
+Our framework is intuitive, flexible and fully open-sourced.
 By highlighting the need for faithfulness in the context of Counterfactual Explanations, we believe that in the short term, our work will enable researchers and practitioners to better distinguish trustworthy from unreliable models.
-We anticipate that ECCCo can serve as a baseline for future research directed at providing plausible but faithful Counterfactual Explanations.
+We further anticipate that ECCCo can serve as a baseline for future research directed at providing plausible but faithful Counterfactual Explanations.
 \end{abstract}

### 2023-05-09 notebooks/cal_housing.qmd
added lin sep
more results

@@
 ```{julia}
-include("notebooks/setup.jl")
+include("$(pwd())/notebooks/setup.jl")
 eval(setup_notebooks)
@@
 generator_dict = Dict(
-    "Wachter" => WachterGenerator(λ=λ₁),
-    "REVISE" => REVISEGenerator(λ=λ₁),
+    "Wachter" => WachterGenerator(),
+    "REVISE" => REVISEGenerator(),
     "Schut" => GreedyGenerator(),
@@
             suppress_training=true, dataname="California Housing",
-    n_individuals=5,
-    factual=0, target=1,
+            n_individuals=10,
+            target=target, factual=factual,
             initialization=:identity,
@@
         )
+        push!(bmks, bmk)
+    end
+end
+bmk = reduce(vcat, bmks)
 CSV.write(joinpath(output_path, "cal_housing_benchmark.csv"), bmk())
-```
-
-```{julia}
-@chain bmk() begin
-    @group_by(dataname, generator, model, variable)
-    @summarize(mean=mean(value),sd=std(value))
-    @ungroup
-    @filter(variable == "distance_from_energy")
-end
 ```

### 2023-05-09 notebooks/circles.qmd
more results

@@
 ```{julia}
-λ₁ = 0.25
-λ₂ = 0.75
-λ₃ = 0.75
+λ₁ = 0.1
+λ₂ = 0.25
+λ₃ = 0.25
 Λ = [λ₁, λ₂, λ₃]
@@
 ```{julia}
+Random.seed!(2023)
+
 M = model_dict["JEM"]
 X = X isa Matrix ? X : Float32.(permutedims(matrix(X)))
-factual_label =  levels(labels)[1]
+factual_label =  levels(labels)[2]
 x_factual = reshape(X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
-target =  levels(labels)[2]
+target =  levels(labels)[1]
 factual = predict_label(M, counterfactual_data, x_factual)[1]
@@
     plt = Plots.plot(ce, title=name, alpha=0.2, cbar=false, axis=nothing)
-    if name == "ECCCo"
+    if contains(name, "ECCCo")
         _X = distance_from_energy(ce, return_conditionals=true)
@@
 end
-plt = Plots.plot(plts..., size=(500,520))
+plt = Plots.plot(plts..., size=(650,500))
 display(plt)
@@
             suppress_training=true, dataname="Circles",
-            n_individuals=5,
+            n_individuals=25,
             target=target, factual=factual,

### 2023-05-09 notebooks/fashion_mnist.qmd
more results

```{julia}
include("$(pwd())/notebooks/setup.jl")
eval(setup_notebooks)
```

# FashionMNIST

## Anecdotal Evidence

### Examples in Introduction

#### Wachter and JSMA

```{julia}
Random.seed!(2023)

# Data:
counterfactual_data = load_fashion_mnist()
X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
input_dim, n_obs = size(counterfactual_data.X)
M = load_fashion_mnist_mlp()

# Target:
factual_label = 9
x_factual = reshape(X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
target = 7
factual = predict_label(M, counterfactual_data, x_factual)[1]
γ = 0.9

# Training params:
T = 100
```

```{julia}
# Search:
generic_generator = WachterGenerator()
ce_wachter = generate_counterfactual(
    x_factual, target, counterfactual_data, M, generic_generator; 
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
)
greedy_generator = GreedyGenerator(η=2.0)
ce_jsma = generate_counterfactual(
    x_factual, target, counterfactual_data, M, greedy_generator; 
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
)
```

```{julia}
p1 = Plots.plot(
    convert2image(FashionMNIST, reshape(x_factual,28,28)),
    axis=nothing, 
    size=(img_height, img_height),
    title="Factual"
)
plts = [p1]

ces = zip([ce_wachter,ce_jsma])
counterfactuals = reduce((x,y)->cat(x,y,dims=3),map(ce -> CounterfactualExplanations.counterfactual(ce[1]), ces))
phat = reduce((x,y) -> cat(x,y,dims=3), map(ce -> target_probs(ce[1]), ces))
for x in zip(eachslice(counterfactuals; dims=3), eachslice(phat; dims=3), ["Wachter","JSMA"])
    ce, _phat, _name = (x[1],x[2],x[3])
    _title = "$(_name) (p=$(round(_phat[1]; digits=2)))"
    plt = Plots.plot(
        convert2image(FashionMNIST, reshape(ce,28,28)),
        axis=nothing, 
        size=(img_height, img_height),
        title=_title
    )
    plts = [plts..., plt]
end
plt = Plots.plot(plts...; size=(img_height*length(plts),img_height), layout=(1,length(plts)))
display(plt)
savefig(plt, joinpath(output_images_path, "fashion_you_may_not_like_it.png"))
```

#### REVISE

```{julia}
using CounterfactualExplanations.Models: load_fashion_mnist_vae
vae = load_fashion_mnist_vae()
vae_weak = load_fashion_mnist_vae(;strong=false)
Serialization.serialize(joinpath(output_path,"fashion_mnist_classifier.jls"), M)
Serialization.serialize(joinpath(output_path,"fashion_mnist_vae.jls"), vae)
Serialization.serialize(joinpath(output_path,"fashion_mnist_vae_weak.jls"), vae_weak)
```

```{julia}
# Define generator:
revise_generator = REVISEGenerator(
    opt = Flux.Optimise.Descent(0.25),
    λ=0.0,
)
# Generate recourse:
counterfactual_data.generative_model = vae # assign generative model
ce_strong = generate_counterfactual(
    x_factual, target, counterfactual_data, M, revise_generator; 
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
    converge_when=:generator_conditions,
)
counterfactual_data_weak = deepcopy(counterfactual_data)
counterfactual_data_weak.generative_model = vae_weak
ce_weak = generate_counterfactual(
    x_factual, target, counterfactual_data_weak, M, revise_generator;
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
    converge_when=:generator_conditions,
)
```

```{julia}
ces = zip([ce_strong,ce_weak])
counterfactuals = reduce((x,y)->cat(x,y,dims=3),map(ce -> CounterfactualExplanations.counterfactual(ce[1]), ces))
phat = reduce((x,y) -> cat(x,y,dims=3), map(ce -> target_probs(ce[1]), ces))
plts = [p1]
for x in zip(eachslice(counterfactuals; dims=3), eachslice(phat; dims=3), ["Strong VAE","Weak VAE"])
    ce, _phat, _name = (x[1],x[2],x[3])
    _title = "$(_name) (p=$(round(_phat[1]; digits=2)))"
    plt = Plots.plot(
        convert2image(FashionMNIST, reshape(ce,28,28)),
        axis=nothing, 
        size=(img_height, img_height),
        title=_title
    )
    plts = [plts..., plt]
end
plt = Plots.plot(plts...; size=(img_height*length(plts),img_height), layout=(1,length(plts)))
display(plt)
savefig(plt, joinpath(output_images_path, "fashion_surrogate_gone_wrong.png"))
```

### ECCCo

```{julia}
function pre_process(x; noise::Float32=0.03f0)
    ϵ = Float32.(randn(size(x)) * noise)
    x += ϵ
    return x
end
```

```{julia}
# Hyper:
_retrain = true
_regen = true

# Data:
n_obs = 10000
counterfactual_data = load_fashion_mnist(n_obs)
counterfactual_data.X = pre_process.(counterfactual_data.X)
counterfactual_data.generative_model = vae
X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
X = table(permutedims(X))
x_factual = reshape(pre_process(x_factual, noise=0.0f0), input_dim, 1)
labels = counterfactual_data.output_encoder.labels
input_dim, n_obs = size(counterfactual_data.X)
n_digits = Int(sqrt(input_dim))
output_dim = length(unique(labels))
```

First, let's create a couple of image classifier architectures:

```{julia}
# Model parameters:
epochs = 100
batch_size = minimum([Int(round(n_obs/10)), 128])
n_hidden = 128
activation = Flux.swish
builder = MLJFlux.@builder Flux.Chain(
    Dense(n_in, n_hidden, activation),
    Dense(n_hidden, n_out),
)
n_ens = 5                                   # number of models in ensemble
_loss = Flux.Losses.logitcrossentropy       # loss function
_finaliser = x -> x                         # finaliser function
```

```{julia}
# JEM parameters:
𝒟x = Uniform(-1.0,1.0)
𝒟y = Categorical(ones(output_dim) ./ output_dim)
sampler = ConditionalSampler(
    𝒟x, 𝒟y, 
    input_size=(input_dim,), 
    batch_size=10,
)
α = [1.0,1.0,2e-2]      # penalty strengths
```

```{julia}
# Simple MLP:
mlp = NeuralNetworkClassifier(
    builder=builder, 
    epochs=epochs,
    batch_size=batch_size,
    finaliser=_finaliser,
    loss=_loss,
)

# Deep Ensemble:
mlp_ens = EnsembleModel(model=mlp, n=n_ens)

# Joint Energy Model:
jem = JointEnergyClassifier(
    sampler;
    builder=builder,
    epochs=epochs,
    batch_size=batch_size,
    finaliser=_finaliser,
    loss=_loss,
    jem_training_params=(
        α=α,verbosity=10,
    ),
    sampling_steps=25,
)

# JEM with adversarial training:
jem_adv = deepcopy(jem)
# jem_adv.adv_training = true

# Deep Ensemble of Joint Energy Models:
jem_ens = EnsembleModel(model=jem, n=n_ens)

# Deep Ensemble of Joint Energy Models with adversarial training:
# jem_ens_plus = EnsembleModel(model=jem_adv, n=n_ens)

# Dictionary of models:
models = Dict(
    "MLP" => mlp,
    "MLP Ensemble" => mlp_ens,
    "JEM" => jem,
    "JEM Ensemble" => jem_ens,
    # "JEM Ensemble+" => jem_ens_plus,
)

Serialization.serialize(joinpath(output_path,"fashion_mnist_architectures.jls"), models)
```

```{julia}
# Train models:
function _train(model, X=X, y=labels; cov=.95, method=:simple_inductive, mod_name="model")
    conf_model = conformal_model(model; method=method, coverage=cov)
    mach = machine(conf_model, X, y)
    @info "Begin training $mod_name."
    fit!(mach)
    @info "Finished training $mod_name."
    M = ECCCo.ConformalModel(mach.model, mach.fitresult)
    return M
end
if _retrain
    model_dict = Dict(mod_name => _train(mod; mod_name=mod_name) for (mod_name, mod) in models)
    Serialization.serialize(joinpath(output_path,"fashion_mnist_models.jls"), model_dict)
else
    model_dict = Serialization.deserialize(joinpath(output_path,"fashion_mnist_models.jls"))
end
```

```{julia}
# Plot generated samples:
n_regen = 500
if _regen 
    for (mod_name, mod) in model_dict
        K = length(counterfactual_data.y_levels)
        input_size = size(selectdim(counterfactual_data.X, ndims(counterfactual_data.X), 1))
        𝒟x = Uniform(extrema(counterfactual_data.X)...)
        𝒟y = Categorical(ones(K) ./ K)
        sampler = ConditionalSampler(𝒟x, 𝒟y; input_size=input_size, prob_buffer=0.0)
        opt = ImproperSGLD()
        f(x) = logits(mod, x)

        _w = 1500
        plts = []
        neach = 10
        for i in 1:10
            x = sampler(f, opt; niter=n_regen, n_samples=neach, y=i)
            plts_i = []
            for j in 1:size(x, 2)
                xj = x[:,j]
                xj = reshape(xj, (n_digits, n_digits))
                plts_i = [plts_i..., Plots.heatmap(rotl90(xj), axis=nothing, cb=false)]
            end
            plt = Plots.plot(plts_i..., size=(_w,0.10*_w), layout=(1,10))
            plts = [plts..., plt]
        end
        plt = Plots.plot(plts..., size=(_w,_w), layout=(10,1), plot_title=mod_name)
        savefig(plt, joinpath(output_images_path, "fashion_mnist_generated_$(mod_name).png"))
        display(plt)
    end
end
```

```{julia}
# Evaluate models:

measure = Dict(
    :f1score => multiclass_f1score, 
    :acc => accuracy, 
    :precision => multiclass_precision
)
model_performance = DataFrame()
for (mod_name, mod) in model_dict
    # Test performance:
    test_data = load_fashion_mnist_test()
    _perf = CounterfactualExplanations.Models.model_evaluation(mod, test_data, measure=collect(values(measure)))
    _perf = DataFrame([[p] for p in _perf], collect(keys(measure)))
    _perf.mod_name .= mod_name
    model_performance = vcat(model_performance, _perf)
end
Serialization.serialize(joinpath(output_path,"fashion_mnist_model_performance.jls"), model_performance)
CSV.write(joinpath(output_path, "fashion_mnist_model_performance.csv"), model_performance)
model_performance
```

### Different Models

```{julia}
function _plot_eccco_fashion_mnist(
    x::Union{AbstractArray, Int}=x_factual, target::Int=target;
    λ=[0.1,0.25,0.25],
    temp=0.1,
    plt_order = ["MLP", "MLP Ensemble", "JEM", "JEM Ensemble"],
    opt = nothing,
    rng::Union{Int,AbstractRNG}=1234,
    T::Int = 100,
)

    # Setup:
    Random.seed!(rng)
    if x isa Int
        x = reshape(counterfactual_data.X[:,rand(findall(labels.==x))],input_dim,1)
    end

    # Generate counterfactuals using ECCCo generator:
    eccco_generator = ECCCoGenerator(
        λ=λ, 
        temp=temp, 
        opt=opt,
    )

    ces = Dict()
    for (mod_name, mod) in model_dict
        ce = generate_counterfactual(
            x, target, counterfactual_data, mod, eccco_generator; 
            decision_threshold=γ, max_iter=T,
            initialization=:identity,
            converge_when=:generator_conditions,
        )
        ces[mod_name] = ce
    end
    _plt_order = map(x -> findall(collect(keys(model_dict)) .== x)[1], plt_order)

    # Plot:
    p1 = Plots.plot(
        convert2image(FashionMNIST, reshape(x,28,28)),
        axis=nothing, 
        size=(img_height, img_height),
        title="Factual"
    )

    plts = []
    for (_name,ce) in ces
        _x = CounterfactualExplanations.counterfactual(ce)
        _phat = target_probs(ce)
        _title = "$_name (p̂=$(round(_phat[1]; digits=3)))"
        plt = Plots.plot(
            convert2image(FashionMNIST, reshape(_x,28,28)),
            axis=nothing, 
            size=(img_height, img_height),
            title=_title
        )
        plts = [plts..., plt]
    end
    plts = plts[_plt_order]
    plts = [p1, plts...]
    plt = Plots.plot(plts...; size=(img_height*length(plts),img_height), layout=(1,length(plts)))

    return plt, eccco_generator
end
```

```{julia}
plt, eccco_generator = _plot_eccco_fashion_mnist()
display(plt)
savefig(plt, joinpath(output_images_path, "fashion_mnist_eccco.png"))
```

### All digits

```{julia}
function plot_fashion_mnist(
    factual::Int,target::Int;
    generator::AbstractGenerator,
    model::AbstractFittedModel=model_dict["JEM Ensemble"],
    data::CounterfactualData=counterfactual_data,
    rng::Union{Int,AbstractRNG}=Random.GLOBAL_RNG,
    _plot_title::Bool=true,
    show_factual::Bool=false,
    kwargs...,
)
    Random.seed!(rng)

    decision_threshold = !isdefined(kwargs, :decision_threshold) ? 0.9 : decision_threshold
    max_iter = !isdefined(kwargs, :max_iter) ? 100 : max_iter
    initialization = !isdefined(kwargs, :initialization) ? :identity : initialization
    converge_when = !isdefined(kwargs, :converge_when) ? :generator_conditions : converge_when

    x = reshape(data.X[:,rand(findall(predict_label(model, data).==factual))],input_dim,1)
    ce = generate_counterfactual(
        x, target, data, model, generator; 
        decision_threshold=decision_threshold, max_iter=max_iter,
        initialization=initialization,
        converge_when=converge_when,
        kwargs...
    )

    _title = _plot_title ? "$(factual) -> $(target)" : ""

    _x = CounterfactualExplanations.counterfactual(ce)
    plt = Plots.plot(
        convert2image(FashionMNIST, reshape(_x,28,28)),
        axis=nothing, 
        size=(img_height, img_height),
        title=_title
    )
    if show_factual
        plt_factual = Plots.plot(
            convert2image(FashionMNIST, reshape(x,28,28)),
            axis=nothing, 
            size=(img_height, img_height),
            title="Factual"
        )
        plt = Plots.plot(plt_factual, plt; size=(img_height*2,img_height), layout=(1,2))
    end

    return plt
end
```

```{julia}
if _regen
    function plot_all_digits(rng=123;verbose=true,kwargs...)
        plts = []
        for i in 0:9
            for j in 0:9
                @info "Generating counterfactual for $(i) -> $(j)"
                plt = plot_fashion_mnist(i,j;kwargs...,rng=rng)
                !verbose || display(plt)
                plts = [plts..., plt]
            end
        end
        plt = Plots.plot(plts...; size=(img_height*10,img_height*10), layout=(10,10))
        return plt
    end
    plt = plot_all_digits(generator=eccco_generator)
    savefig(plt, joinpath(output_images_path, "fashion_mnist_eccco_all_digits.png"))
end
```

## Benchmark

```{julia}
λ₁ = 0.1
λ₂ = 0.25
λ₃ = 0.25
Λ = [λ₁, λ₂, λ₃]

# Benchmark generators:
generator_dict = Dict(
    "Wachter" => WachterGenerator(λ=λ₁),
    "REVISE" => REVISEGenerator(λ=λ₁),
    "Schut" => GreedyGenerator(),
    "ECCCo" => ECCCoGenerator(λ=Λ),
)
```

```{julia}
# Measures:
measures = [
    CounterfactualExplanations.distance,
    ECCCo.distance_from_energy,
    ECCCo.distance_from_targets,
    CounterfactualExplanations.Evaluation.validity,
    CounterfactualExplanations.Evaluation.redundancy,
]

bmks = []
for target in sort(unique(labels))
    for factual in sort(unique(labels))
        if factual == target
            continue
        end
        bmk = benchmark(
            counterfactual_data; 
            models=model_dict, 
            generators=generator_dict, 
            measure=measures,
            suppress_training=true, dataname="FashionMNIST",
            n_individuals=5,
            target=target, factual=factual,
            initialization=:identity,
            converge_when=:generator_conditions,
        )
        push!(bmks, bmk)
    end
end
bmk = reduce(vcat, bmks)

CSV.write(joinpath(output_path, "fashion_mnist_benchmark.csv"), bmk())
```

```{julia}
df = @chain bmk() begin
    @mutate(variable = ifelse.(variable .== "distance_from_energy", "Non-Conformity", variable))
    @mutate(variable = ifelse.(variable .== "distance_from_targets", "Implausibility", variable))
    @mutate(variable = ifelse.(variable .== "distance", "Cost", variable))
    @mutate(variable = ifelse.(variable .== "redundancy", "Redundancy", variable))
    @mutate(variable = ifelse.(variable .== "Validity", "Validity", variable))
end
plt = AlgebraOfGraphics.data(df) * visual(BoxPlot) * 
    mapping(:generator, :value, row=:variable, col=:model, color=:generator)
plt = draw(
    plt, axis=(xlabel="", xticksvisible=false, xticklabelsvisible=false, width=150, height=120), 
    facet=(; linkyaxes=:minimal)
)   
display(plt)
save(joinpath(output_images_path, "fashion_mnist_benchmark.png"), plt, px_per_unit=5)
```

### 2023-05-09 notebooks/gmsc.qmd
more results

@@
     input_size=(input_dim,), 
-    batch_size=10,
+    batch_size=50,
 )
-α = [1.0,1.0,5e-2]      # penalty strengths
+α = [1.0,1.0,1e-2]      # penalty strengths
 ```
@@
     CounterfactualExplanations.Evaluation.redundancy,
+    ECCCo.set_size_penalty
 ]
 
+bmks = []
+for target in sort(unique(labels))
+    for factual in sort(unique(labels))
+        if factual == target
+            continue
+        end
         bmk = benchmark(
@@
             suppress_training=true, dataname="GMSC",
-    n_individuals=5,
-    factual=0, target=1,
+            n_individuals=10,
+            target=target, factual=factual,
             initialization=:identity,
+            converge_when=:generator_conditions,
         )
+        push!(bmks, bmk)
+    end
+end
+bmk = reduce(vcat, bmks)
 CSV.write(joinpath(output_path, "gmsc_benchmark.csv"), bmk())
-```
-
-```{julia}
-@chain bmk() begin
-    @group_by(dataname, generator, model, variable)
-    @summarize(mean=mean(value),sd=std(value))
-    @ungroup
-    @filter(variable == "distance_from_energy")
-end
 ```

### 2023-05-09 notebooks/linearly_separable.qmd
more results

@@
 
-# Circles Data
+# Linearly Separable Data
 
@@
 n_obs = Int(1000 / (1.0 - test_size))
-counterfactual_data, test_data = train_test_split(load_linearly_separable(n_obs); test_size=test_size)
+counterfactual_data, test_data = train_test_split(
+    load_blobs(n_obs; cluster_std=0.1, center_box=(-1. => 1.)); 
+    test_size=test_size
+)
 X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
@@
     model_dict = Dict(mod_name => _train(model; mod_name=mod_name) for (mod_name, model) in models)
-    Serialization.serialize(joinpath(output_path,"circles_models.jls"), model_dict)
+    Serialization.serialize(joinpath(output_path,"linearly_separable_models.jls"), model_dict)
 else
-    model_dict = Serialization.deserialize(joinpath(output_path,"circles_models.jls"))
+    model_dict = Serialization.deserialize(joinpath(output_path,"linearly_separable_models.jls"))
 end
@@
 end
-Serialization.serialize(joinpath(output_path,"circles_model_performance.jls"), model_performance)
-CSV.write(joinpath(output_path, "circles_model_performance.csv"), model_performance)
+Serialization.serialize(joinpath(output_path,"linearly_separable_model_performance.jls"), model_performance)
+CSV.write(joinpath(output_path, "linearly_separable_model_performance.csv"), model_performance)
 model_performance
@@
         ylims = ex[2]
-        x1 = range(1.0f0.*xlims...,length=100)
-        x2 = range(1.0f0.*ylims...,length=100)
+        x1 = range(1.0f0.*xlims...,length=10)
+        x2 = range(1.0f0.*ylims...,length=10)
         p(x) = probs(model, x)
@@
             MLJFlux.reformat(X)[1,:], MLJFlux.reformat(X)[2,:], 
-            color=Int.(labels.refs).-1, group=Int.(labels.refs).-1, alpha=0.5
+            color=Int.(labels.refs), group=Int.(labels.refs), alpha=0.5
         )
@@
             X̂[1,:], X̂[2,:], 
-            color=repeat([target], size(X̂,2)), 
-            group=repeat([target], size(X̂,2)), 
+            color=repeat([target_idx], size(X̂,2)), 
+            group=repeat([target_idx], size(X̂,2)), 
             shape=:star5, ms=10
         )
-        savefig(plt, joinpath(output_images_path, "circles_generated_$(mod_name).png"))
+        savefig(plt, joinpath(output_images_path, "linearly_separable_generated_$(mod_name).png"))
         push!(plts, plt)
@@
 ```{julia}
+Random.seed!(2023)
+
 M = model_dict["JEM"]
@@
 end
-plt = Plots.plot(plts..., size=(500,520))
+plt = Plots.plot(plts..., size=(650,500))
 display(plt)
-savefig(plt, joinpath(output_images_path, "circles_poc.png"))
+savefig(plt, joinpath(output_images_path, "linearly_separable_poc.png"))
 ```
@@
             measure=measures,
-            suppress_training=true, dataname="Circles",
-            n_individuals=5,
+            suppress_training=true, dataname="Linearly Separable",
+            n_individuals=25,
             target=target, factual=factual,
@@
 bmk = reduce(vcat, bmks)
-CSV.write(joinpath(output_path, "circles_benchmark.csv"), bmk())
+CSV.write(joinpath(output_path, "linearly_separable_benchmark.csv"), bmk())
 ```
@@
 display(plt)
-save(joinpath(output_images_path, "circles_benchmark.png"), plt, px_per_unit=5)
+save(joinpath(output_images_path, "linearly_separable_benchmark.png"), plt, px_per_unit=5)
 ```

### 2023-05-09 notebooks/mnist.qmd
more results

--- 2023-05-08 notebooks/mnist.qmd
@@
 # Target:
-factual_label = 8
+factual_label = 9
 x_factual = reshape(X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
-target = 3
+target = 7
 factual = predict_label(M, counterfactual_data, x_factual)[1]
@@
 T = 100
-opt = Flux.Optimise.Adam(0.1)
 ```
@@
 # Search:
-generic_generator = GenericGenerator(opt=Flux.Optimise.Descent(1.0))
+generic_generator = WachterGenerator()
 ce_wachter = generate_counterfactual(
@@
 revise_generator = REVISEGenerator(
-  opt = opt,
-  λ=0.0
+    opt = Flux.Optimise.Descent(0.25),
+    λ=0.0,
 )
@@
 )
-α = [1.0,1.0,1e-1]      # penalty strengths
+α = [1.0,1.0,2e-2]      # penalty strengths
 ```
@@
     ),
-    sampling_steps=20,
+    sampling_steps=25,
 )
@@
     x::Union{AbstractArray, Int}=x_factual, target::Int=target;
-    λ=[0.1,0.1,0.1],
-    temp=0.1,η=0.01,
+    λ=[0.1,0.25,0.25],
+    temp=0.1,
     plt_order = ["MLP", "MLP Ensemble", "JEM", "JEM Ensemble"],
-    opt = Flux.Optimise.Adam(η),
-    rng::Union{Int,AbstractRNG}=Random.GLOBAL_RNG,
+    opt = nothing,
+    rng::Union{Int,AbstractRNG}=1234,
     T::Int = 100,
@@
 ```{julia}
-plt, eccco_generator = _plot_eccco_mnist(opt=JSMADescent(η=1.0))
+plt, eccco_generator = _plot_eccco_mnist()
 display(plt)
@@
     _plot_title::Bool=true,
+    show_factual::Bool=false,
     kwargs...,
 )
+    Random.seed!(rng)
+
     decision_threshold = !isdefined(kwargs, :decision_threshold) ? 0.9 : decision_threshold
@@
     )
+    if show_factual
+        plt_factual = Plots.plot(
+            convert2image(MNIST, reshape(x,28,28)),
+            axis=nothing, 
+            size=(img_height, img_height),
+            title="Factual"
+        )
+        plt = Plots.plot(plt_factual, plt; size=(img_height*2,img_height), layout=(1,2))
+    end
+
     return plt
@@
 if _regen
-    function plot_all_digits(rng=1;verbose=true,kwargs...)
+    function plot_all_digits(rng=123;verbose=true,kwargs...)
         plts = []
@@
 
-### Different Generators
+## Benchmark
 
 ```{julia}
-# Setup:
-model = model_dict["JEM Ensemble"]
+λ₁ = 0.1
+λ₂ = 0.25
+λ₃ = 0.25
+Λ = [λ₁, λ₂, λ₃]
 
@@
 generator_dict = Dict(
-    :wachter => generic_generator,
-    :revise => revise_generator,
-    :greedy => greedy_generator,
-    :eccco => eccco_generator,
+    "Wachter" => WachterGenerator(λ=λ₁),
+    "REVISE" => REVISEGenerator(λ=λ₁),
+    "Schut" => GreedyGenerator(),
+    "ECCCo" => ECCCoGenerator(λ=Λ),
 )
-
-ces = Dict()
-for (gen_name, gen) in generator_dict
-    ce = generate_counterfactual(
-        x_factual, target, counterfactual_data, model, gen; 
-        decision_threshold=γ, max_iter=T,
-        initialization=:identity,
-        converge_when=:generator_conditions,
-    )
-    ces[gen_name] = ce
-end
-plt_order = sortperm(collect(keys(ces)))
-
-# Plot:
-p1 = Plots.plot(
-    convert2image(MNIST, reshape(x_factual,28,28)),
-    axis=nothing, 
-    size=(img_height, img_height),
-    title="Factual"
-)
-
-plts = []
-for (_name,ce) in ces
-    _x = CounterfactualExplanations.counterfactual(ce)
-    _phat = target_probs(ce)
-    _title = "$_name (p̂=$(round(_phat[1]; digits=3)))"
-    plt = Plots.plot(
-        convert2image(MNIST, reshape(_x,28,28)),
-        axis=nothing, 
-        size=(img_height, img_height),
-        title=_title
-    )
-    plts = [plts..., plt]
-end
-plts = plts[plt_order]
-plts = [p1, plts...]
-plt = Plots.plot(plts...; size=(img_height*length(plts),img_height), layout=(1,length(plts)))
-display(plt)
-savefig(plt, joinpath(output_images_path, "mnist_all_generators.png"))
 ```
-
-## Benchmark
 
@@
             n_individuals=5,
-    factual=0, target=1,
+            target=target, factual=factual,
             initialization=:identity,
+            converge_when=:generator_conditions,
         )
+        push!(bmks, bmk)
+    end
+end
+bmk = reduce(vcat, bmks)
+
 CSV.write(joinpath(output_path, "mnist_benchmark.csv"), bmk())
-```
-
-```{julia}
-@chain bmk() begin
-    @group_by(dataname, generator, model, variable)
-    @summarize(mean=mean(value),sd=std(value))
-    @ungroup
-    @filter(variable == "distance_from_energy")
-end
 ```

### 2023-05-09 notebooks/moons.qmd
more results

@@
 end
-plt = Plots.plot(plts..., size=(500,520))
+plt = Plots.plot(plts..., size=(650,500))
 display(plt)
 savefig(plt, joinpath(output_images_path, "moons_poc.png"))
-```
-
-```{julia}
-evaluate(collect(values(ces)))
 ```
@@
             suppress_training=true, dataname="Moons",
-    n_individuals=5,
-    target=1, factual=0,
+            n_individuals=25,
+            target=target, factual=factual,
             initialization=:identity,
@@
         )
+        push!(bmks, bmk)
+    end
+end
+bmk = reduce(vcat, bmks)
 CSV.write(joinpath(output_path, "moons_benchmark.csv"), bmk())
-```
-
-```{julia}
-@chain bmk() begin
-    @group_by(dataname, generator, model, variable)
-    @summarize(mean=mean(value),sd=std(value))
-    @ungroup
-    @filter(variable == "distance_from_energy")
-end
 ```

### 2023-05-09 notebooks/poc.qmd
work on table
more results

@@
 # Counteractual data:
-counterfactual_data = load_linearly_separable(1000)
+n_obs = 1000
+counterfactual_data = load_blobs(n_obs; cluster_std=0.1, center_box=(-1. => 1.))
 X = counterfactual_data.X
@@
 λ₁ = 0.1
-λ₂ = 0.5
-λ₃ = 0.5
+λ₂ = 0.4
+λ₃ = 0.4
 Λ = [λ₁, λ₂, λ₃]
@@
 
+opt = Flux.Optimise.Descent(0.01)
+
 generator_dict = OrderedDict(
-    "Wachter (γ=$γ)" => WachterGenerator(λ = λ₁),
+    "Wachter" => WachterGenerator(λ = λ₁, opt=opt),
     "Schut" => GreedyGenerator(λ = λ₁),
-    "REVISE" => REVISEGenerator(λ = λ₁),
-    "ECCCo" => ECCCoGenerator(λ = Λ),
+    "REVISE" => REVISEGenerator(λ = λ₁, opt=opt),
+    "ECCCo" => ECCCoGenerator(λ = Λ, opt=opt),
 )
@@
         converge_when=:generator_conditions,
-        decision_threshold=γ,
     )
-    plt = Plots.plot(ce, title=name, alpha=0.2, cbar=false, axis=nothing, length_out=10)
+    plt = Plots.plot(
+        ce, title=name, alpha=0.2, cbar=false, 
+        axis=nothing, length_out=10, contour_alpha=1.0,
+    )
     if name == "ECCCo"
@@
 for (x, y) in inputs
-    push!(u, scale * gradient(f2, [x, y][:,:])[1][1])
-    push!(v, scale * gradient(f2, [x, y][:,:])[1][2])
+    push!(u, scale * gradient(f1, [x, y][:,:])[1][1])
+    push!(v, scale * gradient(f1, [x, y][:,:])[1][2])
 end

### 2023-05-09 notebooks/tables.Rmd
more results

@@
 dt[,ce:=NULL]
-synth <- c("Moons", "Circles")
+synth <- c("Moons", "Circles", "Linearly Separable")
 dt[,source:=ifelse(dataname %in% synth, "synthetic", "real")]
@@
       value=sprintf("%1.2f (%1.2f)", mean(value), sd(value)),
-      val = mean(value)
+      val = mean(value),
+      std = sd(value)
     ),
@@
 tab[,min_val:=val==min(val),.(model, dataname, variable)]
+tab[,two_std_wachter:=val+2*std<val[generator=="Wachter"],.(model, dataname, variable)]
+tab[,one_std_wachter:=val+1*std<val[generator=="Wachter"],.(model, dataname, variable)]
+tab[one_std_wachter==T,value:=paste0(value,"*")]
+tab[two_std_wachter==T,value:=paste0(value,"*")]
 tab$value <- cell_spec(tab$value, "latex", bold=tab$min_val)
 tab[,val:=NULL]
+tab[,std:=NULL]
 tab[,min_val:=NULL]
-tab[generator == "eccco", generator:="ECCCo"]
-tab[generator == "revise", generator:="REVISE"]
-tab[generator == "generic", generator:="Wachter"]
-tab[generator == "wachter", generator:="Wachter"]
-tab[generator == "greedy", generator:="Schut"]
+tab[,two_std_wachter:=NULL]
+tab[,one_std_wachter:=NULL]
 n_data <- length(unique(tab$dataname))
@@
 kbl(
-  tab, caption = "Results for synthetic datasets. Standard deviation in parentheses. \\label{tab:results-synth}", 
+  tab, caption = "Results for synthetic datasets.
+Standard deviations across samples are shown in parentheses.
+Best outcomes are highlighted in bold.
+Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (Wachter).
+\\label{tab:results-synth} \\newline",
   align = "c", col.names=col_names, booktabs = T, escape=F, 
@@
 ) %>%
-  # kable_styling(latex_options = c("scale_down")) %>%
+  kable_styling(latex_options = c("scale_down")) %>%
   kable_paper(full_width = F) %>%
-  add_header_above(c(" " = 2, "Moons" = 2)) %>%
+  add_header_above(c(" " = 2, "Circles" = 2, "Linearly Separable" = 2, "Moons" = 2)) %>%
   column_spec(1, bold = T) %>%

### 2023-05-09 paper/paper.tex
more results

@@
   \item Batch size seems to impact quality of generated samples (at inference, but not so much during JEM training)
-  \item ECCCo is sensitive to optimizer (Adam works well), learning rate and distance metric (l1 currently only one that works)
   \item SGLD takes time 
@@
   \item JEMs are more difficult to train
+  \item There is a tradeoff: higher cost vs. higher conformity/plausibility
+  \item Results are sensitive to choices of penalty strength and step size
+  \item Counterfactuals may end up looking fairly homogenous
+  \item For MNIST data we found CP to have little effect
+  \item JEMs themselves are sensitive to scale
+  \item ECCCo can backfire, in case generative property of model is poor
 \end{itemize}

### 2023-05-09 src/generator.jl

@@ -4,7 +4,7 @@ using CounterfactualExplanations.Objectives
 function CCEGenerator(; 
     λ::Union{AbstractFloat,Vector{<:AbstractFloat}}=[0.1, 1.0], 
     κ::Real=1.0, 
-    temp::Real=0.5, 
+    temp::Real=0.1, 
     kwargs...
 )
     function _set_size_penalty(ce::AbstractCounterfactualExplanation)
@@ -17,14 +17,15 @@ end
 
 "Constructor for `ECECCCoGenerator`: Energy Constrained Conformal Counterfactual Explanation Generator."
 function ECCCoGenerator(; 
-    λ::Union{AbstractFloat,Vector{<:AbstractFloat}}=[0.1, 0.1, 0.1], 
+    λ::Union{AbstractFloat,Vector{<:AbstractFloat}}=[0.2,0.4,0.4], 
     κ::Real=1.0, 
-    temp::Real=0.5, 
-    η::Union{Nothing,Real}=nothing,
-    n::Union{Nothing,Int}=nothing,
-    opt::Flux.Optimise.AbstractOptimiser=CounterfactualExplanations.Generators.JSMADescent(η=η,n=n),
+    temp::Real=0.1, 
+    opt::Union{Nothing,Flux.Optimise.AbstractOptimiser}=nothing,
     kwargs...
 )
+    if isnothing(opt)
+        opt = CounterfactualExplanations.Generators.Descent(0.1)
+    end
     function _set_size_penalty(ce::AbstractCounterfactualExplanation)
         return ECCCo.set_size_penalty(ce; κ=κ, temp=temp)
     end

### 2023-05-09 src/sampling.jl

@@ -43,15 +43,11 @@ function EnergySampler(
 
     @assert y ∈ data.y_levels || y ∈ 1:length(data.y_levels)
 
-    if ECCCo._has_sampler(model)
-        sampler = ECCCo._get_sampler(model)
-    else
-        K = length(data.y_levels)
-        input_size = size(selectdim(data.X, ndims(data.X), 1))
-        𝒟x = Uniform(extrema(data.X)...)
-        𝒟y = Categorical(ones(K) ./ K)
-        sampler = ConditionalSampler(𝒟x, 𝒟y; input_size=input_size)
-    end
+    K = length(data.y_levels)
+    input_size = size(selectdim(data.X, ndims(data.X), 1))
+    𝒟x = Uniform(extrema(data.X)...)
+    𝒟y = Categorical(ones(K) ./ K)
+    sampler = ConditionalSampler(𝒟x, 𝒟y; input_size=input_size)
     yidx = get_target_index(data.y_levels, y)
 
     # Initiate:

### 2023-05-10 notebooks/mnist.qmd
paper

@@
     CounterfactualExplanations.Evaluation.redundancy,
+    ECCCo.set_size_penalty,
 ]

### 2023-05-10 notebooks/tables.Rmd
paper

@@
 
+## Main tables
+
 ```{r}
@@
 kbl(
-  tab, caption = "Results for real-world datasets. Standard deviation in parentheses. \\label{tab:results-real}", 
+  tab, caption = "Results for real-world datasets.
+Standard deviations across samples are shown in parentheses.
+Best outcomes are highlighted in bold.
+Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (Wachter).
+\\label{tab:results-real} \\newline",
   align = "c", col.names=col_names, booktabs = T, escape=F, 
@@
   kable_styling(latex_options = c("scale_down")) %>%
+  kable_paper(full_width = F) %>%
   add_header_above(c(" " = 2, "California Housing" = 2, "GMSC" = 2, "MNIST" = 2)) %>%

### 2023-05-10 paper/paper.tex
paper

@@
 \newtheorem{definition}{Definition}[section]
+\newtheorem{question}{Research Question}[section]
 
@@
 
-\subsection{From Fidelity to Model Conformity}
+\subsection{From Fidelity to Model Faithfulness}
 
@@
 
-\begin{definition}[Conformal Counterfactuals]
-  \label{def:conformal}
-  Let $\mathcal{X}_{\theta}|\mathbf{y}^* = p_{\theta}(x|\mathbf{y}^*)$ denote the conditional distribution of $\mathbf{x}$ in the target class $\mathbf{y}^*$, where $\theta$ denotes the parameters of model $M_{\theta}$.
+\begin{definition}[Faithful Counterfactuals]
+  \label{def:faithful}
+  Let $\mathcal{X}_{\theta}|\mathbf{y}^* = p_{\theta}(\mathbf{X}_{\mathbf{y}^*})$ denote the conditional distribution of $\mathbf{x}$ in the target class $\mathbf{y}^*$, where $\theta$ denotes the parameters of model $M_{\theta}$.
 Then for $\mathbf{x}^{\prime}$ to be considered a conformal counterfactual, we need: $\mathbf{x}^{\prime} \sim \mathcal{X}_{\theta}|\mathbf{y}^*$.
@@
 Since this definition works with distributional properties, it explicitly accounts for the multiplicity of explanations we discussed earlier.
-To assess counterfactuals with respect to Definition~\ref{def:conformal}, we need to be able to quantify the posterior conditional distribution $p_{\theta}(\mathbf{x}|\mathbf{y}^*)$.
-This is very much at the core of our proposed methodological framework, which reconciles the notions of plausibility and model conformity and which we will introduce next.
+To assess counterfactuals with respect to Definition~\ref{def:faithful}, we need to be able to quantify the posterior conditional distribution $p_{\theta}(\mathbf{x}|\mathbf{y}^*)$.
+This is very much at the core of our proposed methodological framework, which reconciles the notions of plausibility and model faithfulness and which we will introduce next.
 
@@
 The term $\mathcal{E}(\mathbf{x}_j|\mathbf{y}^*)$ denotes the energy function where we use $\mathcal{E}(\mathbf{x}_j|\mathbf{y}^*)=-M_{\theta}(\mathbf{x}_j)[\mathbf{y}^*]$, that is the negative logit corresponding to the target class label $\mathbf{y}^*$.
-Generating multiple samples in this manner yields an empirical distribution $\hat{\mathcal{X}}_{\theta}|\mathbf{y}^*$ that we use in our search for plausible counterfactuals, as discussed in more detail below.
+Generating multiple samples in this manner yields an empirical distribution $\hat{\mathbf{X}}_{\theta,\mathbf{y}^*}$ that we use in our search for plausible counterfactuals, as discussed in more detail below.
 Appendix~\ref{app-jem} provides additional implementation details for any tasks related to energy-based modelling.
@@
 The additional two penalties in ECCCo ensure that counterfactuals conform with the model's generative property and lead to minimally uncertain predictions, respectively.
-The hyperparameters $\lambda_1, ..., \lambda_3$ can be used to balance the different objectives: for example, we may choose to incur larger deviations from the factual in favour of conformity with the model's generative property by choosing lower values of $\lambda_1$ and relatively higher values of $\lambda_2$.
+The hyperparameters $\lambda_1, ..., \lambda_3$ can be used to balance the different objectives: for example, we may choose to incur larger deviations from the factual in favour of faithfulness with the model's generative property by choosing lower values of $\lambda_1$ and relatively higher values of $\lambda_2$.
 Figure~\ref{fig:eccco} illustrates this balancing act for an example involving synthetic data: vector fields indicate the direction of gradients with respect to the different components our proposed objective function (Equation~\ref{eq:eccco}).
@@
 
-In this section, we bolster our anecdotal findings from the previous section through rigorous empirical analysis.
-We first briefly describe our evaluation framework and data, before presenting and discussing our results.
+In this section, we present our empirical analysis and findings. Our goal is to shed line on the following questions:
 
-\subsection{Evaluation Measures}\label{evaluation}
+\begin{question}[Feasibility]\label{rq:feasibility}
+  Is it feasible to generate plausible Counterfactual Explanations through ECCCo without relying on surrogate models?
+\end{question}
 
-Above we have defined plausibility (Definition~\ref{def:plausible}) and conformity (Definition~\ref{def:conformal}) for Counterfactual Explanations.
-In this subsection, we introduce evaluation measures that facilitate a quantitative evaluation of counterfactuals for these objectives.
+\begin{question}[Drivers]\label{rq:drivers}
+  Subject to feasibility, what drives the performance of ECCCo?
+Is it sufficient to rely on energy-based modelling to quantify the model's generative property?
+Is it sufficient to rely on conformal prediction to quantify the model's uncertainty?
+\end{question}
 
-Firstly, in order to assess the plausibility of counterfactuals we adapt the implausibility metric proposed in~\citet{guidotti2022counterfactual}.
-The authors propose to evaluate plausibility in terms of the distance of the counterfactual $\mathbf{x}^{\prime}$ from its nearest neighbour in the target class $\mathbf{y}^*$: the smaller this distance, the more plausible the counterfactual.
-Instead of focusing only on the nearest neighbour of $\mathbf{x}^{\prime}$, we suggest computing the average over distances from multiple (possibly all) observed instances in the target class.
-Formally, for a single counterfactual, we have:
+In the following, we first briefly describe our evaluation framework and data, before presenting and discussing our results.
+
+\subsection{Key Evaluation Measures}\label{evaluation}
+
+Above we have defined plausibility (Definition~\ref{def:plausible}) and faithfulness (Definition~\ref{def:faithful}) for Counterfactual Explanations.
+These are the main criteria we use to evaluate counterfactuals in this study.
+In order to quantify the plausibility of counterfactuals we use a slightly adapted version of the implausibility metric proposed in~\citet{guidotti2022counterfactual}.
+Formally, for a single counterfactual, we define implausibility as follows,
 
@@
   \begin{aligned}
-    \text{impl} = \frac{1}{\lvert\mathbf{x} \in \mathcal{X}|\mathbf{y}^*\rvert} \sum_{\mathbf{x} \in \mathcal{X}|\mathbf{y}^*} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
+    \text{impl} = \frac{1}{\lvert\mathbf{x} \in \mathbf{X}_{\mathbf{y}^*}\rvert} \sum_{\mathbf{x} \in \mathbf{X}_{\mathbf{y}^*}} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
   \end{aligned}
@@
 
-This measure is straightforward to compute and should be less sensitive to outliers in the target class than the one based on the nearest neighbour.
-It also gives rise to a very similar evaluation measure for conformity.
-We merely swap out the subsample of individuals in the target class for the empirical distribution of generated conditional samples:
+where $\mathbf{X}_{\mathbf{y}^*}$ is a subsample of the training data in the target class $\mathbf{y}^*$.
+This gives rise to a very similar evaluation measure for unfaithfulness.
+We merely swap out the subsample of individuals in the target class for a subset $\hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^*}$ of the generated conditional samples:
 
-\begin{equation}\label{eq:conf}
+\begin{equation}\label{eq:faith}
   \begin{aligned}
-    \text{conf} = \frac{1}{\lvert\mathbf{x} \in \mathcal{X}_{\theta}|\mathbf{y}^*\rvert} \sum_{\mathbf{x} \in \mathcal{X}_{\theta}|\mathbf{y}^*} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
+    \text{unfaith} = \frac{1}{\lvert\mathbf{x} \in \hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^*}\rvert} \sum_{\mathbf{x} \in \hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^*}} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
   \end{aligned}
@@
 
+Specifically, we form this subset based on the $n_E$ generated samples associated with the lowest energy.
+
+While we focus on these key evaluation metrics in the body of this paper, we also sporadically discuss outcomes with respect to other common measures used to evaluate the validity, proximity and sparsity of counterfactuals.
+Details can be found in Appendix~\ref{app:results}.
+
 As noted by~\citet{guidotti2022counterfactual}, these distance-based measures are simplistic and more complex alternative measures may ultimately be more appropriate for the task.
 For example, we considered using statistical divergence measures instead.
-This would involve generating not one but many counterfactuals and comparing the generated empirical distribution to the target distributions in Definitions~\ref{def:plausible} and~\ref{def:conformal}.
+This would involve generating not one but many counterfactuals and comparing the generated empirical distribution to the target distributions in Definitions~\ref{def:plausible} and~\ref{def:faithful}.
 While this approach is potentially more rigorous, generating enough counterfactuals is not always practical.
@@
   \item JEMs are more difficult to train
-  \item There is a tradeoff: higher cost vs. higher conformity/plausibility
+  \item There is a tradeoff: higher cost vs. higher faithfulness/plausibility
   \item Results are sensitive to choices of penalty strength and step size
@@
 
-\section{Submission of papers to NeurIPS 2023}
-
-Please read the instructions below carefully and follow them faithfully.
-
-\subsection{Style}
-
-Papers to be submitted to NeurIPS 2023 must be prepared according to the
-instructions presented here. Papers may only be up to {\bf nine} pages long,
-including figures. Additional pages \emph{containing only acknowledgments and
-references} are allowed. Papers that exceed the page limit will not be
-reviewed, or in any other way considered for presentation at the conference.
-
-The margins in 2023 are the same as those in previous years.
-
-Authors are required to use the NeurIPS \LaTeX{} style files obtainable at the
-NeurIPS website as indicated below. Please make sure you use the current files
-and not previous versions. Tweaking the style files may be grounds for
-rejection.
-
-\subsection{Retrieval of style files}
-
-The style files for NeurIPS and other conference information are available on
-the website at
-\begin{center}
-  \url{http://www.neurips.cc/}
-\end{center}
-The file \verb+neurips_2023.pdf+ contains these instructions and illustrates the
-various formatting requirements your NeurIPS paper must satisfy.
-
-The only supported style file for NeurIPS 2023 is \verb+neurips_2023.sty+,
-rewritten for \LaTeXe{}.  \textbf{Previous style files for \LaTeX{} 2.09,
-  Microsoft Word, and RTF are no longer supported!}
-
-The \LaTeX{} style file contains three optional arguments: \verb+final+, which
-creates a camera-ready copy, \verb+preprint+, which creates a preprint for
-submission to, e.g., arXiv, and \verb+nonatbib+, which will not load the
-\verb+natbib+ package for you in case of package clash.
-
-\paragraph{Preprint option}
-If you wish to post a preprint of your work online, e.g., on arXiv, using the
-NeurIPS style, please use the \verb+preprint+ option. This will create a
-nonanonymized version of your work with the text ``Preprint. Work in progress.''
-in the footer. This version may be distributed as you see fit, as long as you do not say which conference it was submitted to. Please \textbf{do
-  not} use the \verb+final+ option, which should \textbf{only} be used for
-papers accepted to NeurIPS.
-
-At submission time, please omit the \verb+final+ and \verb+preprint+
-options. This will anonymize your submission and add line numbers to aid
-review. Please do \emph{not} refer to these line numbers in your paper as they
-will be removed during generation of camera-ready copies.
-
-The file \verb+neurips_2023.tex+ may be used as a ``shell'' for writing your
-paper. All you have to do is replace the author, title, abstract, and text of
-the paper with your own.
-
-The formatting instructions contained in these style files are summarized in
-Sections \ref{gen_inst}, \ref{headings}, and \ref{others} below.
-
-\section{General formatting instructions}
-\label{gen_inst}
-
-The text must be confined within a rectangle 5.5~inches (33~picas) wide and
-9~inches (54~picas) long. The left margin is 1.5~inch (9~picas).  Use 10~point
-type with a vertical spacing (leading) of 11~points.  Times New Roman is the
-preferred typeface throughout, and will be selected for you by default.
-Paragraphs are separated by \nicefrac{1}{2}~line space (5.5 points), with no
-indentation.
-
-The paper title should be 17~point, initial caps/lower case, bold, centered
-between two horizontal rules. The top rule should be 4~points thick and the
-bottom rule should be 1~point thick. Allow \nicefrac{1}{4}~inch space above and
-below the title to rules. All pages should start at 1~inch (6~picas) from the
-top of the page.
-
-For the final version, authors' names are set in boldface, and each name is
-centered above the corresponding address. The lead author's name is to be listed
-first (left-most), and the co-authors' names (if different address) are set to
-follow. If there is only one co-author, list both author and co-author side by
-side.
-
-Please pay special attention to the instructions in Section \ref{others}
-regarding figures, tables, acknowledgments, and references.
-
-\section{Headings: first level}
-\label{headings}
-
-All headings should be lower case (except for first word and proper nouns),
-flush left, and bold.
-
-First-level headings should be in 12-point type.
-
-\subsection{Headings: second level}
-
-Second-level headings should be in 10-point type.
-
-\subsubsection{Headings: third level}
-
-Third-level headings should be in 10-point type.
-
-\paragraph{Paragraphs}
-
-There is also a \verb+\paragraph+ command available, which sets the heading in
-bold, flush left, and inline with the text, with the heading followed by 1\,em
-of space.
-
-\section{Citations, figures, tables, references}
-\label{others}
-
-These instructions apply to everyone.
-
-\subsection{Citations within the text}
-
-The \verb+natbib+ package will be loaded for you by default.  Citations may be
-author/year or numeric, as long as you maintain internal consistency.  As to the
-format of the references themselves, any style is acceptable as long as it is
-used consistently.
-
-The documentation for \verb+natbib+ may be found at
-\begin{center}
-  \url{http://mirrors.ctan.org/macros/latex/contrib/natbib/natnotes.pdf}
-\end{center}
-Of note is the command \verb+\citet+, which produces citations appropriate for
-use in inline text.  For example,
-\begin{verbatim}
-   \citet{hasselmo} investigated\dots
-\end{verbatim}
-produces
-\begin{quote}
-  Hasselmo, et al.\ (1995) investigated\dots
-\end{quote}
-
-If you wish to load the \verb+natbib+ package with options, you may add the
-following before loading the \verb+neurips_2023+ package:
-\begin{verbatim}
-   \PassOptionsToPackage{options}{natbib}
-\end{verbatim}
-
-If \verb+natbib+ clashes with another package you load, you can add the optional
-argument \verb+nonatbib+ when loading the style file:
-\begin{verbatim}
-   \usepackage[nonatbib]{neurips_2023}
-\end{verbatim}
-
-As submission is double blind, refer to your own published work in the third
-person. That is, use ``In the previous work of Jones et al.\ [4],'' not ``In our
-previous work [4].'' If you cite your other papers that are not widely available
-(e.g., a journal paper under review), use anonymous author names in the
-citation, e.g., an author of the form ``A.\ Anonymous'' and include a copy of the anonymized paper in the supplementary material.
-
-\subsection{Footnotes}
-
-Footnotes should be used sparingly.  If you do require a footnote, indicate
-footnotes with a number\footnote{Sample of the first footnote.} in the
-text. Place the footnotes at the bottom of the page on which they appear.
-Precede the footnote with a horizontal rule of 2~inches (12~picas).
-
-Note that footnotes are properly typeset \emph{after} punctuation
-marks.\footnote{As in this example.}
-
-\subsection{Figures}
-
-\begin{figure}
-  \centering
-  \fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
-  \caption{Sample figure caption.}
-\end{figure}
-
-All artwork must be neat, clean, and legible. Lines should be dark enough for
-purposes of reproduction. The figure number and caption always appear after the
-figure. Place one line space before the figure caption and one line space after
-the figure. The figure caption should be lower case (except for first word and
-proper nouns); figures are numbered consecutively.
-
-You may use color figures.  However, it is best for the figure captions and the
-paper body to be legible if the paper is printed in either black/white or in
-color.
-
-\subsection{Tables}
-
-All tables must be centered, neat, clean and legible.  The table number and
-title always appear before the table.  See Table~\ref{sample-table}.
-
-Place one line space before the table title, one line space after the
-table title, and one line space after the table. The table title must
-be lower case (except for first word and proper nouns); tables are
-numbered consecutively.
-
-Note that publication-quality tables \emph{do not contain vertical rules.} We
-strongly suggest the use of the \verb+booktabs+ package, which allows for
-typesetting high-quality, professional tables:
-\begin{center}
-  \url{https://www.ctan.org/pkg/booktabs}
-\end{center}
-This package was used to typeset Table~\ref{sample-table}.
-
-\begin{table}
-  \caption{Sample table title}
-  \label{sample-table}
-  \centering
-  \begin{tabular}{lll}
-    \toprule
-    \multicolumn{2}{c}{Part}                   \\
-    \cmidrule(r){1-2}
-    Name     & Description     & Size ($\mu$m) \\
-    \midrule
-    Dendrite & Input terminal  & $\sim$100     \\
-    Axon     & Output terminal & $\sim$10      \\
-    Soma     & Cell body       & up to $10^6$  \\
-    \bottomrule
-  \end{tabular}
-\end{table}
-
-\subsection{Math}
-Note that display math in bare TeX commands will not create correct line numbers for submission.
-Please use LaTeX (or AMSTeX) commands for unnumbered display math.
-(You really shouldn't be using \$\$ anyway; see \url{https://tex.stackexchange.com/questions/503/why-is-preferable-to} and \url{https://tex.stackexchange.com/questions/40492/what-are-the-differences-between-align-equation-and-displaymath} for more information.)
-
-\subsection{Final instructions}
-
-Do not change any aspects of the formatting parameters in the style files.  In
-particular, do not modify the width or length of the rectangle the text should
-fit into, and do not change font sizes (except perhaps in the
-\textbf{References} section; see below). Please note that pages should be
-numbered.
-
-\section{Preparing PDF files}
-
-Please prepare submission files with paper size ``US Letter,'' and not, for
-example, ``A4.''
-
-Fonts were the main cause of problems in the past years. Your PDF file must only
-contain Type 1 or Embedded TrueType fonts. Here are a few instructions to
-achieve this.
-
-\begin{itemize}
-
-\item You should directly generate PDF files using \verb+pdflatex+.
-
-\item You can check which fonts a PDF files uses.  In Acrobat Reader, select the
-  menu Files$>$Document Properties$>$Fonts and select Show All Fonts. You can
-  also use the program \verb+pdffonts+ which comes with \verb+xpdf+ and is
-  available out-of-the-box on most Linux machines.
-
-\item \verb+xfig+ "patterned" shapes are implemented with bitmap fonts.  Use
-  "solid" shapes instead.
-
-\item The \verb+\bbold+ package almost always uses bitmap fonts.  You should use
-  the equivalent AMS Fonts:
-\begin{verbatim}
-   \usepackage{amsfonts}
-\end{verbatim}
-followed by, e.g., \verb+\mathbb{R}+, \verb+\mathbb{N}+, or \verb+\mathbb{C}+
-for $\mathbb{R}$, $\mathbb{N}$ or $\mathbb{C}$.  You can also use the following
-workaround for reals, natural and complex:
-\begin{verbatim}
-   \newcommand{\RR}{I\!\!R} %real numbers
-   \newcommand{\Nat}{I\!\!N} %natural numbers
-   \newcommand{\CC}{I\!\!\!\!C} %complex numbers
-\end{verbatim}
-Note that \verb+amsfonts+ is automatically loaded by the \verb+amssymb+ package.
-
-\end{itemize}
-
-If your file contains type 3 fonts or non embedded TrueType fonts, we will ask
-you to fix it.
-
-\subsection{Margins in \LaTeX{}}
-
-Most of the margin problems come from figures positioned by hand using
-\verb+\special+ or other commands. We suggest using the command
-\verb+\includegraphics+ from the \verb+graphicx+ package. Always specify the
-figure width as a multiple of the line width as in the example below:
-\begin{verbatim}
-   \usepackage[pdftex]{graphicx} ...
-   \includegraphics[width=0.8\linewidth]{myfile.pdf}
-\end{verbatim}
-See Section 4.4 in the graphics bundle documentation
-(\url{http://mirrors.ctan.org/macros/latex/required/graphics/grfguide.pdf})
-
-A number of width problems arise when \LaTeX{} cannot properly hyphenate a
-line. Please give LaTeX hyphenation hints using the \verb+\-+ command when
-necessary.
+\subsection{Results}\label{app:results}
 
@@
 
-\section{Supplementary Material}
-
-Authors may wish to optionally include extra information (complete proofs, additional experiments and plots) in the appendix.
-All such materials should be part of the supplemental material (submitted separately) and should NOT be included in the main submission.
-
-\section*{References}
-
-References follow the acknowledgments in the camera-ready paper. Use unnumbered first-level heading for
-the references. Any choice of citation style is acceptable as long as you are
-consistent. It is permissible to reduce the font size to \verb+small+ (9 point)
-when listing the references.
-Note that the Reference section does not count towards the page limit.
-\medskip
-
-{
-\small
-
-[1] Alexander, J.A.\ \& Mozer, M.C.\ (1995) Template-based algorithms for
-connectionist rule extraction. In G.\ Tesauro, D.S.\ Touretzky and T.K.\ Leen
-(eds.), {\it Advances in Neural Information Processing Systems 7},
-pp.\ 609--616. Cambridge, MA: MIT Press.
-
-[2] Bower, J.M.\ \& Beeman, D.\ (1995) {\it The Book of GENESIS: Exploring
-  Realistic Neural Models with the GEneral NEural SImulation System.}  New York:
-TELOS/Springer--Verlag.
-
-[3] Hasselmo, M.E., Schnell, E.\ \& Barkai, E.\ (1995) Dynamics of learning and
-recall at excitatory recurrent synapses and cholinergic modulation in rat
-hippocampal region CA3. {\it Journal of Neuroscience} {\bf 15}(7):5249-5262.
-}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \end{document}

### 2023-05-10 notebooks/tables.Rmd
just results, limitations and conclusion left

@@
 dt[,source:=ifelse(dataname %in% synth, "synthetic", "real")]
+```
+
+```{r}
+tab <- dt[
+    ,
+    .(
+      value=sprintf("%1.2f (%1.2f)", mean(value), sd(value)),
+      val = mean(value),
+      std = sd(value)
+    ),
+    .(dataname, generator, model, variable, source)
+]
+
+tab$top_val = F
+tab$one_std_wachter = F
+tab$two_std_wachter = F
+
+# Measures to be minimized:
+min_measures <- c(
+  "distance",
+  "distance_from_energy", 
+  "distance_from_targets",
+  "set_size_penalty"
+)
+tab[variable %in% min_measures,top_val:=val==min(val),.(model, dataname, variable)]
+tab[variable %in% min_measures,top_val:=ifelse(rep(all(top_val),length(top_val)),F,top_val),.(model, dataname, variable)]
+tab[variable %in% min_measures,two_std_wachter:=val+2*std<val[generator=="Wachter"],.(model, dataname, variable)]
+tab[variable %in% min_measures,one_std_wachter:=val+1*std<val[generator=="Wachter"],.(model, dataname, variable)]
+
+# Measures to be maximized:
+max_measures <- c(
+  "validity",
+  "redundancy"
+)
+tab[variable %in% max_measures,top_val:=val==max(val),.(model, dataname, variable)]
+tab[variable %in% max_measures,top_val:=ifelse(rep(all(top_val),length(top_val)),F,top_val),.(model, dataname, variable)]
+tab[variable %in% max_measures,two_std_wachter:=val-2*std>val[generator=="Wachter"],.(model, dataname, variable)]
+tab[variable %in% max_measures,one_std_wachter:=val-1*std>val[generator=="Wachter"],.(model, dataname, variable)]
+
+# Add conditional formatting:
+tab[one_std_wachter==T,value:=paste0(value,"*")]
+tab[two_std_wachter==T,value:=paste0(value,"*")]
+tab$value <- cell_spec(tab$value, "latex", bold=tab$top_val)
+
+# Remove redundant columns:
+tab[,val:=NULL]
+tab[,std:=NULL]
+tab[,top_val:=NULL]
+tab[,two_std_wachter:=NULL]
+tab[,one_std_wachter:=NULL]
+```
+
+## Full table
+
+```{r}
+tab_full <- dcast(tab, dataname + model + generator ~ variable)
+col_names <- c(
+  "Model",
+  "Data",
+  "Generator",
+  "Cost ↓", 
+  "Unfaithfulness ↓", 
+  "Implausibility ↓", 
+  "Redundancy ↑",
+  "Uncertainty ↓",
+  "Validity ↑"
+)
+kbl(
+  tab_full, caption = "All results for all datasets.
+Standard deviations across samples are shown in parentheses.
+Best outcomes are highlighted in bold.
+Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (Wachter).
+\\label{tab:results-full} \\newline",
+  align = "c", col.names=col_names, booktabs = F, escape=F, 
+  format="latex"
+) %>%
+  kable_styling(latex_options = c("scale_down")) %>%
+  kable_paper(full_width = F) %>%
+  column_spec(1, bold = T) %>%
+  collapse_rows(columns = 1:3, latex_hline = "custom", valign = "middle", custom_latex_hline = 1:2) %>%
+  save_kable("paper/contents/table_all.tex")
 ```
@@
   "Generator",
-  rep(c("Non-conformity ↓","Implausibility ↓"),n_data)
+  rep(c("Unfaithfulness ↓","Implausibility ↓"),n_data)
 )

### 2023-05-10 paper/paper.tex
just results, limitations and conclusion left

@@
 
-In the following, we first briefly describe our evaluation framework and data, before presenting and discussing our results.
+In the following, we first briefly describe our evaluation framework and experimental setup, before presenting and discussing our results.
 
@@
 
-As noted by~\citet{guidotti2022counterfactual}, these distance-based measures are simplistic and more complex alternative measures may ultimately be more appropriate for the task.
-For example, we considered using statistical divergence measures instead.
-This would involve generating not one but many counterfactuals and comparing the generated empirical distribution to the target distributions in Definitions~\ref{def:plausible} and~\ref{def:faithful}.
-While this approach is potentially more rigorous, generating enough counterfactuals is not always practical.
+\subsection{Experimental Setup}
 
-\subsection{Data}
+To assess and benchmark the performance of ECCCo against the state of the art, we generate multiple counterfactuals for different black-box models and datasets.
+In particular, we compare ECCCo to the following counterfactual generators that were introduced above: firstly; \textbf{Schut}~\citep{schut2021generating}, which minimizes predictive uncertainty; secondly, \textbf{REVISE}~\citep{joshi2019realistic}, which uses a VAE as its surrogate model; and, finally, \textbf{Wachter}~\citep{wachter2017counterfactual}, which serves as our baseline.
+
+We use both synthetic and real-world datasets from different domains, all of which are publically available.
+The synthetic datasets include: a dataset containing two \textbf{Linearly Separable} Gaussian clusters ($n=1000$), as well as the common \textbf{Circles} ($n=1000$) and \textbf{Moons} ($n=2500$) data.
+As for real-world data, we use two well-known datasets from the vision domain: firstly, following~\citet{schut2021generating} we use \textbf{MNIST}~\citep{lecun1998mnist} and, secondly, we add \textbf{Fashion-MNIST}~\citep{xiao2017fashion}.
+Both datasets involve 60000 images of 28x28 pixels each in 10 different classes.
+Finally, we also study the Give Me Some Credit (\textbf{GMSC}) dataset, a tabular dataset from the social sciences domain that has been used extensively in the literature on Algorithmic Recourse~\citep{pawelczyk2021carla}.
+It consists of 11 numeric features that can be used to predict the binary outcome variable indicating whether or not retail borrowers experience financial distress~\citep{kaggle2011give}.
+
+As for the example in Section~\ref{meth}, we use simple neural networks (\textbf{MLP}), ensembles of neural networks (\textbf{MLP Ensemble}), Joint Energy Models (\textbf{JEM}) and ensembles of JEMs (\textbf{JEM Ensemble}) to model our real-world datasets.
+For the synthetic datasets, we found that the use of ensembles was not necessary.
+Full details concerning our hyperparameter choices, training procedures and model performance can be found in Appendix~\ref{app:setup}.
 
@@
 
+\subsection{Experimental Setup}\label{app:setup}
 \subsection{Results}\label{app:results}
+
+\import{contents/}{table_all.tex}
 

### 2023-05-10 notebooks/tables.Rmd
results shaping up

@@
 synth <- c("Moons", "Circles", "Linearly Separable")
-dt[,source:=ifelse(dataname %in% synth, "synthetic", "real")]
+dt[,source:=ifelse(dataname %in% synth, "synthetic", "real-world")]
 ```
@@
 ```{r}
+# Choices:
 measures <- c("distance_from_energy", "distance_from_targets")
-tab <- dt[
-    variable %in% measures & source=="real",
-    .(
-      value=sprintf("%1.2f (%1.2f)", mean(value), sd(value)),
-      val = mean(value),
-      std = sd(value)
-    ),
-    .(dataname, generator, model, variable)
-]
-tab[,min_val:=val==min(val),.(model, dataname, variable)]
-tab[,two_std_wachter:=val+2*std<val[generator=="Wachter"],.(model, dataname, variable)]
-tab[,one_std_wachter:=val+1*std<val[generator=="Wachter"],.(model, dataname, variable)]
-tab[one_std_wachter==T,value:=paste0(value,"*")]
-tab[two_std_wachter==T,value:=paste0(value,"*")]
-tab$value <- cell_spec(tab$value, "latex", bold=tab$min_val)
-tab[,val:=NULL]
-tab[,std:=NULL]
-tab[,min_val:=NULL]
-tab[,two_std_wachter:=NULL]
-tab[,one_std_wachter:=NULL]
-n_data <- length(unique(tab$dataname))
-tab <- dcast(tab, model + generator ~ dataname + variable)
+measure_names <- c("Unfaithfulness ↓","Implausibility ↓")
+chosen_source <- "real-world"
+chosen_data <- c(
+  "MNIST",
+  "GMSC"
+)
+tab_i <- tab
+
+# Logic:
+tab_i <- tab_i[variable %in% measures]
+tab_i <- tab_i[source == chosen_source]
+tab_i <- tab_i[dataname %in% chosen_data]
+tab_i[,dataname:=factor(dataname,levels=chosen_data)]
+tab_i <- dcast(tab_i, model + generator ~ dataname + variable)
 col_names <- c(
@@
   "Generator",
-  rep(c("Unfaithfulness ↓","Implausibility ↓"),n_data)
+  rep(measure_names,length(chosen_data))
 )
-kbl(
-  tab, caption = "Results for real-world datasets.
+caption <- sprintf(
+  "Results for %s datasets.
 Standard deviations across samples are shown in parentheses.
@@
 Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (Wachter).
-\\label{tab:results-real} \\newline",
+\\label{tab:results-%s} \\newline",
+  chosen_source,
+  chosen_source
+)
+file_name <- sprintf(
+  "paper/contents/table-%s.tex",
+  chosen_source
+)
+sub_header <- rep(length(measures), length(chosen_data))
+names(sub_header) <- chosen_data
+header <- c(
+  " " = 2, sub_header
+)
+line_sep <- c(rep("",length(measures)-1),"\\addlinespace")
+kbl(
+  tab_i, caption = caption, 
   align = "c", col.names=col_names, booktabs = T, escape=F, 
-  format="latex", linesep = c("", "", "", "\\addlinespace") 
+  format="latex", linesep = line_sep 
 ) %>%
@@
   kable_paper(full_width = F) %>%
-  add_header_above(c(" " = 2, "California Housing" = 2, "GMSC" = 2, "MNIST" = 2)) %>%
+  add_header_above(header) %>%
   column_spec(1, bold = T) %>%
   collapse_rows(columns = 1:2, latex_hline = "major", valign = "middle") %>%
-  save_kable("paper/contents/table_real.tex")
+  save_kable(file_name)
 ```
@@
 ```{r}
+# Choices:
 measures <- c("distance_from_energy", "distance_from_targets")
-tab <- dt[
-    variable %in% measures & source=="synthetic",
-    .(
-      value=sprintf("%1.2f (%1.2f)", mean(value), sd(value)),
-      val = mean(value),
-      std = sd(value)
-    ),
-    .(dataname, generator, model, variable)
-]
-tab[,min_val:=val==min(val),.(model, dataname, variable)]
-tab[,two_std_wachter:=val+2*std<val[generator=="Wachter"],.(model, dataname, variable)]
-tab[,one_std_wachter:=val+1*std<val[generator=="Wachter"],.(model, dataname, variable)]
-tab[one_std_wachter==T,value:=paste0(value,"*")]
-tab[two_std_wachter==T,value:=paste0(value,"*")]
-tab$value <- cell_spec(tab$value, "latex", bold=tab$min_val)
-tab[,val:=NULL]
-tab[,std:=NULL]
-tab[,min_val:=NULL]
-tab[,two_std_wachter:=NULL]
-tab[,one_std_wachter:=NULL]
-n_data <- length(unique(tab$dataname))
-tab <- dcast(tab, model + generator ~ dataname + variable)
+measure_names <- c("Unfaithfulness ↓","Implausibility ↓")
+chosen_source <- "synthetic"
+chosen_data <- c(
+  "Linearly Separable",
+  "Moons",
+  "Circles"
+)
+tab_i <- tab
+
+# Logic:
+tab_i <- tab_i[variable %in% measures]
+tab_i <- tab_i[source == chosen_source]
+tab_i <- tab_i[dataname %in% chosen_data]
+tab_i[,dataname:=factor(dataname,levels=chosen_data)]
+tab_i <- dcast(tab_i, model + generator ~ dataname + variable)
 col_names <- c(
@@
   "Generator",
-  rep(c("Unfaithfulness ↓","Implausibility ↓"),n_data)
+  rep(measure_names,length(chosen_data))
 )
-kbl(
-  tab, caption = "Results for synthetic datasets.
+caption <- sprintf(
+  "Results for %s datasets.
 Standard deviations across samples are shown in parentheses.
@@
 Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (Wachter).
-\\label{tab:results-synth} \\newline",
+\\label{tab:results-%s} \\newline",
+  chosen_source,
+  chosen_source
+)
+file_name <- sprintf(
+  "paper/contents/table-%s.tex",
+  chosen_source
+)
+sub_header <- rep(length(measures), length(chosen_data))
+names(sub_header) <- chosen_data
+header <- c(
+  " " = 2, sub_header
+)
+line_sep <- c(rep("",length(measures)-1),"\\addlinespace")
+kbl(
+  tab_i, caption = caption, 
   align = "c", col.names=col_names, booktabs = T, escape=F, 
-  format="latex", linesep = c("", "", "", "\\addlinespace") 
+  format="latex", linesep = line_sep 
 ) %>%
@@
   kable_paper(full_width = F) %>%
-  add_header_above(c(" " = 2, "Circles" = 2, "Linearly Separable" = 2, "Moons" = 2)) %>%
+  add_header_above(header) %>%
   column_spec(1, bold = T) %>%
   collapse_rows(columns = 1:2, latex_hline = "major", valign = "middle") %>%
-  save_kable("paper/contents/table_synth.tex")
+  save_kable(file_name)
 ```

### 2023-05-10 paper/paper.tex
results shaping up

@@
 
-This is typically implemented by defining a target outcome $\mathbf{y}^* \in \mathcal{Y}$ for some individual $\mathbf{x} \in \mathcal{X}=\mathbb{R}^D$ described by $D$ attributes, for which the model $M_{\theta}:\mathcal{X}\mapsto\mathcal{Y}$ initially predicts a different outcome: $M_{\theta}(\mathbf{x})\ne \mathbf{y}^*$.
-Counterfactuals are then searched by minimizing a loss function that compares the predicted model output to the target outcome: $\text{yloss}(M_{\theta}(\mathbf{x}),\mathbf{y}^*)$.
+This is typically implemented by defining a target outcome $\mathbf{y}^+ \in \mathcal{Y}$ for some individual $\mathbf{x} \in \mathcal{X}=\mathbb{R}^D$ described by $D$ attributes, for which the model $M_{\theta}:\mathcal{X}\mapsto\mathcal{Y}$ initially predicts a different outcome: $M_{\theta}(\mathbf{x})\ne \mathbf{y}^+$.
+Counterfactuals are then searched by minimizing a loss function that compares the predicted model output to the target outcome: $\text{yloss}(M_{\theta}(\mathbf{x}),\mathbf{y}^+)$.
 Since Counterfactual Explanations (CE) work directly with the black-box model, valid counterfactuals always have full local fidelity by construction~\citep{mothilal2020explaining}.
@@
 In this work, we draw closer attention to the insufficiency of model fidelity as an evaluation metric for the faithfulness of counterfactual explanations.
-Our key contributions are as follows: firstly, we introduce a new notion of faithfulness that is suitable for counterfactuals and propose a novel evaluation measure that draws inspiration from recent advances in Energy-Based Modelling (EBM); secondly, we a novel algorithmic approach for generating Energy-Constrained Conformal Counterfactuals (ECCCo) that explicitly address the need for faithfulness; finally, we provide illustrative examples and extensive empirical evidence demonstrating that ECCCos faithfully explain model behaviour without sacrificing existing desidarata like plausibility and sparsity.
+Our key contributions are as follows: firstly, we introduce a new notion of faithfulness that is suitable for counterfactuals and propose a novel evaluation metric that draws inspiration from recent advances in Energy-Based Modelling (EBM); secondly, we a novel algorithmic approach for generating Energy-Constrained Conformal Counterfactuals (ECCCo) that explicitly address the need for faithfulness; finally, we provide illustrative examples and extensive empirical evidence demonstrating that ECCCos faithfully explain model behaviour without sacrificing existing desidarata like plausibility and sparsity.
 
@@
 \begin{aligned}
-\mathbf{Z}^\prime &= \arg \min_{\mathbf{Z}^\prime \in \mathcal{Z}^L} \left\{  {\text{yloss}(M_{\theta}(f(\mathbf{Z}^\prime)),\mathbf{y}^*)}+ \lambda {\text{cost}(f(\mathbf{Z}^\prime)) }  \right\} 
+\mathbf{Z}^\prime &= \arg \min_{\mathbf{Z}^\prime \in \mathcal{Z}^L} \left\{  {\text{yloss}(M_{\theta}(f(\mathbf{Z}^\prime)),\mathbf{y}^+)}+ \lambda {\text{cost}(f(\mathbf{Z}^\prime)) }  \right\} 
 \end{aligned} 
@@
   \label{def:plausible}
-  Let $\mathcal{X}|\mathbf{y}^*$ denote the true conditional distribution of samples in the target class $\mathbf{y}^*$.
-Then for $\mathbf{x}^{\prime}$ to be considered a plausible counterfactual, we need: $\mathbf{x}^{\prime} \sim \mathcal{X}|\mathbf{y}^*$.
+  Let $\mathcal{X}|\mathbf{y}^+$ denote the true conditional distribution of samples in the target class $\mathbf{y}^+$.
+Then for $\mathbf{x}^{\prime}$ to be considered a plausible counterfactual, we need: $\mathbf{x}^{\prime} \sim \mathcal{X}|\mathbf{y}^+$.
 \end{definition}
@@
 It may become problematic even in this context if the dependency turns into a vulnerability.
-To illustrate this point, we have used REVISE~\citep{joshi2019realistic} with an underfitted VAE to generate the counterfactual in the right panel of Figure~\ref{fig:vae}: in this case, the decoder step of the VAE fails to yield plausible values ($\{\mathbf{x}^{\prime} \leftarrow \mathcal{G}(\mathbf{z})\} \not\sim \mathcal{X}|\mathbf{y}^*$) and hence the counterfactual search in the learned latent space is doomed.
+To illustrate this point, we have used REVISE~\citep{joshi2019realistic} with an underfitted VAE to generate the counterfactual in the right panel of Figure~\ref{fig:vae}: in this case, the decoder step of the VAE fails to yield plausible values ($\{\mathbf{x}^{\prime} \leftarrow \mathcal{G}(\mathbf{z})\} \not\sim \mathcal{X}|\mathbf{y}^+$) and hence the counterfactual search in the learned latent space is doomed.
 
@@
 Their proposed methodology solves Equation~\ref{eq:general} by greedily applying JSMA in the feature space with standard cross-entropy loss and no penalty at all.
-They demonstrate theoretically and empirically that their approach yields counterfactuals for which the model $M_{\theta}$ predicts the target label $\mathbf{y}^*$ with high confidence.
+They demonstrate theoretically and empirically that their approach yields counterfactuals for which the model $M_{\theta}$ predicts the target label $\mathbf{y}^+$ with high confidence.
 Provided the model is well-specified, these counterfactuals are plausible.
@@
 We posit a related, slightly more nuanced question: `Is this counterfactual consistent with what the model has learned about the underlying data'?
-We will argue that fidelity is not a sufficient evaluation measure to answer this question and propose a novel way to assess if Counterfactual Explanations conform with model behaviour.
+We will argue that fidelity is not a sufficient evaluation metric to answer this question and propose a novel way to assess if Counterfactual Explanations conform with model behaviour.
 
@@
   \label{def:faithful}
-  Let $\mathcal{X}_{\theta}|\mathbf{y}^* = p_{\theta}(\mathbf{X}_{\mathbf{y}^*})$ denote the conditional distribution of $\mathbf{x}$ in the target class $\mathbf{y}^*$, where $\theta$ denotes the parameters of model $M_{\theta}$.
-Then for $\mathbf{x}^{\prime}$ to be considered a conformal counterfactual, we need: $\mathbf{x}^{\prime} \sim \mathcal{X}_{\theta}|\mathbf{y}^*$.
+  Let $\mathcal{X}_{\theta}|\mathbf{y}^+ = p_{\theta}(\mathbf{X}_{\mathbf{y}^+})$ denote the conditional distribution of $\mathbf{x}$ in the target class $\mathbf{y}^+$, where $\theta$ denotes the parameters of model $M_{\theta}$.
+Then for $\mathbf{x}^{\prime}$ to be considered a conformal counterfactual, we need: $\mathbf{x}^{\prime} \sim \mathcal{X}_{\theta}|\mathbf{y}^+$.
 \end{definition}
@@
 Since this definition works with distributional properties, it explicitly accounts for the multiplicity of explanations we discussed earlier.
-To assess counterfactuals with respect to Definition~\ref{def:faithful}, we need to be able to quantify the posterior conditional distribution $p_{\theta}(\mathbf{x}|\mathbf{y}^*)$.
+To assess counterfactuals with respect to Definition~\ref{def:faithful}, we need to be able to quantify the posterior conditional distribution $p_{\theta}(\mathbf{x}|\mathbf{y}^+)$.
 This is very much at the core of our proposed methodological framework, which reconciles the notions of plausibility and model faithfulness and which we will introduce next.
@@
 We can leverage this observation to quantify the generative property of the Black Box model itself.
-In particular, note that if we fix $\mathbf{y}$ to our target value $\mathbf{y}^*$, we can sample from $p_{\theta}(\mathbf{x}|\mathbf{y}^*)$ using SGLD as follows,
+In particular, note that if we fix $\mathbf{y}$ to our target value $\mathbf{y}^+$, we can sample from $p_{\theta}(\mathbf{x}|\mathbf{y}^+)$ using SGLD as follows,
 
@@
   \begin{aligned}
-    \mathbf{x}_{j+1} &\leftarrow \mathbf{x}_j - \frac{\epsilon^2}{2} \mathcal{E}(\mathbf{x}_j|\mathbf{y}^*) + \epsilon \mathbf{r}_j, && j=1,...,J
+    \mathbf{x}_{j+1} &\leftarrow \mathbf{x}_j - \frac{\epsilon^2}{2} \mathcal{E}(\mathbf{x}_j|\mathbf{y}^+) + \epsilon \mathbf{r}_j, && j=1,...,J
   \end{aligned}
@@
 where $\mathbf{r}_j \sim \mathcal{N}(\mathbf{0},\mathbf{I})$ is the stochastic term and the step-size $\epsilon$ is typically polynomially decayed.
-The term $\mathcal{E}(\mathbf{x}_j|\mathbf{y}^*)$ denotes the energy function where we use $\mathcal{E}(\mathbf{x}_j|\mathbf{y}^*)=-M_{\theta}(\mathbf{x}_j)[\mathbf{y}^*]$, that is the negative logit corresponding to the target class label $\mathbf{y}^*$.
-Generating multiple samples in this manner yields an empirical distribution $\hat{\mathbf{X}}_{\theta,\mathbf{y}^*}$ that we use in our search for plausible counterfactuals, as discussed in more detail below.
+The term $\mathcal{E}(\mathbf{x}_j|\mathbf{y}^+)$ denotes the energy function where we use $\mathcal{E}(\mathbf{x}_j|\mathbf{y}^+)=-M_{\theta}(\mathbf{x}_j)[\mathbf{y}^+]$, that is the negative logit corresponding to the target class label $\mathbf{y}^+$.
+Generating multiple samples in this manner yields an empirical distribution $\hat{\mathbf{X}}_{\theta,\mathbf{y}^+}$ that we use in our search for plausible counterfactuals, as discussed in more detail below.
 Appendix~\ref{app-jem} provides additional implementation details for any tasks related to energy-based modelling.
@@
   \begin{aligned}
-  \mathbf{Z}^\prime= \arg \min_{\mathbf{Z}^\prime \in \mathcal{Z}^M}  &\{  {\text{yloss}(M_{\theta}(f(\mathbf{Z}^\prime)),\mathbf{y}^*)}+ \lambda_{1} {\text{dist}(f(\mathbf{Z}^\prime),\mathbf{x}) } \\
+  \mathbf{Z}^\prime= \arg \min_{\mathbf{Z}^\prime \in \mathcal{Z}^M}  &\{  {\text{yloss}(M_{\theta}(f(\mathbf{Z}^\prime)),\mathbf{y}^+)}+ \lambda_{1} {\text{dist}(f(\mathbf{Z}^\prime),\mathbf{x}) } \\
   &+ \lambda_2 \text{dist}(f(\mathbf{Z}^\prime),\hat{\mathbf{x}}_{\theta}) + \lambda_3 \Omega(C_{\theta}(f(\mathbf{Z}^\prime);\alpha)) \} 
@@
 \begin{algorithmic}[1]
-  \Require $\mathbf{x}, \mathbf{y}^*, M_{\theta}, f, \Lambda, \alpha, \mathcal{D}, T, \eta, n_{\mathcal{B}}, N_{\mathcal{B}}$ \linebreak where $M_{\theta}(\mathbf{x})\neq\mathbf{y}^*$
+  \Require $\mathbf{x}, \mathbf{y}^+, M_{\theta}, f, \Lambda, \alpha, \mathcal{D}, T, \eta, n_{\mathcal{B}}, N_{\mathcal{B}}$ \linebreak where $M_{\theta}(\mathbf{x})\neq\mathbf{y}^+$
   \Ensure $\mathbf{x}^\prime$
   \State Initialize $\mathbf{z}^\prime \gets f^{-1}(\mathbf{x})$
-  \State Generate buffer $\mathcal{B}$ of $N_{\mathcal{B}}$ conditional samples $\hat{\mathbf{x}}_{\theta}|\mathbf{y}^*$ using SGLD (Equation~\ref{eq:sgld})
+  \State Generate buffer $\mathcal{B}$ of $N_{\mathcal{B}}$ conditional samples $\hat{\mathbf{x}}_{\theta}|\mathbf{y}^+$ using SGLD (Equation~\ref{eq:sgld})
   \State Run \textit{SCP} for $M_{\theta}$ using $\mathcal{D}$
@@
   \State $\hat{\mathbf{x}}_{\theta, t} \gets \text{rand}(\mathcal{B},n_{\mathcal{B}})$
-  \State $\mathbf{z}^\prime \gets \mathbf{z}^\prime - \eta \nabla_{\mathbf{z}^\prime} \mathcal{L}(\mathbf{z}^\prime,\mathbf{y}^*,\hat{\mathbf{x}}_{\theta, t}; \Lambda, \alpha)$
+  \State $\mathbf{z}^\prime \gets \mathbf{z}^\prime - \eta \nabla_{\mathbf{z}^\prime} \mathcal{L}(\mathbf{z}^\prime,\mathbf{y}^+,\hat{\mathbf{x}}_{\theta, t}; \Lambda, \alpha)$
   \State $t \gets t+1$
@@
 That state is initialized by passing the factual $\mathbf{x}$ through the encoder $f^{-1}$ which in our case corresponds to a simple feature transformer, rather than the encoder part of VAE as in REVISE~\citep{joshi2019realistic}.
-Next, we generate a buffer of $N_{\mathcal{B}}$ conditional samples $\hat{\mathbf{x}}_{\theta}|\mathbf{y}^*$ using SGLD (Equation~\ref{eq:sgld}) and conformalise the model $M_{\theta}$ through Split Conformal Prediction on training data $\mathcal{D}$.
+Next, we generate a buffer of $N_{\mathcal{B}}$ conditional samples $\hat{\mathbf{x}}_{\theta}|\mathbf{y}^+$ using SGLD (Equation~\ref{eq:sgld}) and conformalise the model $M_{\theta}$ through Split Conformal Prediction on training data $\mathcal{D}$.
 
 Finally, we search counterfactuals through gradient descent.
-Let $\mathcal{L}(\mathbf{z}^\prime,\mathbf{y}^*,\hat{\mathbf{x}}_{\theta, t}; \Lambda, \alpha)$ denote our loss function defined in Equation~\ref{eq:eccco}.
+Let $\mathcal{L}(\mathbf{z}^\prime,\mathbf{y}^+,\hat{\mathbf{x}}_{\theta, t}; \Lambda, \alpha)$ denote our loss function defined in Equation~\ref{eq:eccco}.
 Then in each iteration, we first randomly draw $n_{\mathcal{B}}$ samples from the buffer $\mathcal{B}$ before updating the counterfactual state $\mathbf{z}^\prime$ by moving in the negative direction of that loss function.
@@
 
-\subsection{Key Evaluation Measures}\label{evaluation}
+\subsection{Key Evaluation Metrics}\label{evaluation}
 
@@
   \begin{aligned}
-    \text{impl} = \frac{1}{\lvert\mathbf{x} \in \mathbf{X}_{\mathbf{y}^*}\rvert} \sum_{\mathbf{x} \in \mathbf{X}_{\mathbf{y}^*}} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
+    \text{impl} = \frac{1}{\lvert\mathbf{x} \in \mathbf{X}_{\mathbf{y}^+}\rvert} \sum_{\mathbf{x} \in \mathbf{X}_{\mathbf{y}^+}} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
   \end{aligned}
@@
 
-where $\mathbf{X}_{\mathbf{y}^*}$ is a subsample of the training data in the target class $\mathbf{y}^*$.
-This gives rise to a very similar evaluation measure for unfaithfulness.
-We merely swap out the subsample of individuals in the target class for a subset $\hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^*}$ of the generated conditional samples:
+where $\mathbf{X}_{\mathbf{y}^+}$ is a subsample of the training data in the target class $\mathbf{y}^+$.
+This gives rise to a very similar evaluation metric for unfaithfulness.
+We merely swap out the subsample of individuals in the target class for a subset $\hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^+}$ of the generated conditional samples:
 
@@
   \begin{aligned}
-    \text{unfaith} = \frac{1}{\lvert\mathbf{x} \in \hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^*}\rvert} \sum_{\mathbf{x} \in \hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^*}} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
+    \text{unfaith} = \frac{1}{\lvert\mathbf{x} \in \hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^+}\rvert} \sum_{\mathbf{x} \in \hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^+}} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
   \end{aligned}
@@
 
-We use both synthetic and real-world datasets from different domains, all of which are publically available.
-The synthetic datasets include: a dataset containing two \textbf{Linearly Separable} Gaussian clusters ($n=1000$), as well as the common \textbf{Circles} ($n=1000$) and \textbf{Moons} ($n=2500$) data.
-As for real-world data, we use two well-known datasets from the vision domain: firstly, following~\citet{schut2021generating} we use \textbf{MNIST}~\citep{lecun1998mnist} and, secondly, we add \textbf{Fashion-MNIST}~\citep{xiao2017fashion}.
-Both datasets involve 60000 images of 28x28 pixels each in 10 different classes.
-Finally, we also study the Give Me Some Credit (\textbf{GMSC}) dataset, a tabular dataset from the social sciences domain that has been used extensively in the literature on Algorithmic Recourse~\citep{pawelczyk2021carla}.
-It consists of 11 numeric features that can be used to predict the binary outcome variable indicating whether or not retail borrowers experience financial distress~\citep{kaggle2011give}.
+We use both synthetic and real-world datasets from different domains, all of which are publically available and commonly used to train and benchmark classification algorithms.
+The synthetic datasets include: a dataset containing two \textbf{Linearly Separable} Gaussian clusters ($n=1000$), as well as the well-known \textbf{Circles} ($n=1000$) and \textbf{Moons} ($n=2500$) data.
+As for real-world data, we follow~\citet{schut2021generating} and use the \textbf{MNIST}~\citep{lecun1998mnist} image dataset.
+It is composed of 60000 images of 28x28 pixels each showing handwritten digits from 0 to 9 such as the examples shown above.
+From the social sciences domain, we include Give Me Some Credit (\textbf{GMSC})~\citep{kaggle2011give}: a tabular dataset that has been studied extensively in the literature on Algorithmic Recourse~\citep{pawelczyk2021carla}.
+It consists of 11 numeric features that can be used to predict the binary outcome variable indicating whether or not retail borrowers experience financial distress.
 
-As for the example in Section~\ref{meth}, we use simple neural networks (\textbf{MLP}), ensembles of neural networks (\textbf{MLP Ensemble}), Joint Energy Models (\textbf{JEM}) and ensembles of JEMs (\textbf{JEM Ensemble}) to model our real-world datasets.
+As with the example in Section~\ref{meth}, we use simple neural networks (\textbf{MLP}), ensembles of neural networks (\textbf{MLP Ensemble}), Joint Energy Models (\textbf{JEM}) and ensembles of JEMs (\textbf{JEM Ensemble}) to model our real-world datasets.
 For the synthetic datasets, we found that the use of ensembles was not necessary.
-Full details concerning our hyperparameter choices, training procedures and model performance can be found in Appendix~\ref{app:setup}.
+
+To account for stochasticity, we generate multiple counterfactuals for each possible target class, generator, model and dataset.
+Specifically, we randomly sample $n^{-}$ times from the subset of individuals for which the given model predicts the non-target class $\mathbf{y}^{-}$ given the current target.
+We set $n^{-}=25$ for all of our synthetic datasets, $n^{-}=10$ for GMSC and $n^{-}=5$ for MNIST.
+Note that in the latter case, we still end up generating disproportionately more counterfactuals because all 10 digits appear both as factuals and targets for each other.
+Full details concerning our parameter choices, training procedures and model performance can be found in Appendix~\ref{app:setup}.
 
@@
 
-See Table~\ref{tab:results-real}
+Table~\ref{tab:results-synthetic} shows the key results for the synthetic datasets separated by model (first columns) and generator (second column).
+The numerical columns show the average values of our key evaluation metrics computed across all counterfactuals.
+Standard deviations are shown in parentheses.
+In bold we have highlighted the best outcome for each model and metric.
+To provide some sense of the statistical significance of our findings, we have added asterisks to indicate that a given value is at least one ($*$) or two ($**$) standard deviations lower than the baseline (Wachter).
 
-\import{contents/}{table_synth.tex}
+\import{contents/}{table-synthetic.tex}
 
-\import{contents/}{table_real.tex}
+Starting with the high-level results for our Linearly Separable data, we find that ECCCo produces the most faithful counterfactuals for both black-box models.
+It also produces the most plausible counterfactuals for the Joint Energy Model, which was explicitly trained to learn plausible representations of the input data.
+This high-level pattern is broadly consistent across all datasets and supportive of our narrative, so it is worth highlighting here, before we discuss it below: ECCCos consistently achieve high faithfulness, which---subject to the quality of the black-box model itself---coincides with high plausibility.
 
-\section{Discussion}
+Zooming in on the granular details for the Linearly Separable data, note that the list of generators in Table~\ref{tab:results-synthetic} includes `ECCCo (no CP)' and `ECCCo (no EBM)' in addition to `ECCCo' and our benchmark generators.
+These have been added to gain some sense of the degree to which the two components underlying ECCCo---namely energy-based modelling (EBM) and conformal prediction (CP)---drive the results.
+Specifically, `ECCCo (no CP)' involves no set size penalty ($\lambda_3=0$ in Equation~\ref{eq:eccco}), while `ECCCo (no EBM)' does not penalise the distance to samples generated through SGLD ($\lambda_2=0$ in Equation~\ref{eq:eccco}).
+The corresponding results indicate that the positive results are dominated by the effect of quantifying and leveraging the model's generative property (EBM) in our search for counterfactuals.
+Conformal Prediction alone only leads to marginally improved faithfulness and plausibility relative to the benchmark generators for our JEM.
 
-\subsection{Key Insights}
+As a final key observation for the Linearly Separable data we note that for the MLP, increased faithfulness comes at the cost of reduced plausibility.
+Specifically, this means that counterfactuals generated through ECCCo end up further away from individuals in the target class than those produced by our benchmark generators.
+
+The findings for the Moons dataset are broadly in line with the findings so far: for the JEM, ECCCo yields significantly more faithful and plausible counterfactuals than all other generators.
+For the MLP, faithfulness is maintained but not plausibility.
+For comparison, REVISE yields fairly plausible counterfactuals in both cases, but it does so at the cost of faithfulness.
+For the Circles data, ...
+
+Moving on to our real-world datasets, the results are shown in Table~\ref{tab:results-real-world}.
+Once again the findings indicate that the plausibility of ECCCos is positively correlated with the capacity of the black-box model to distinguish plausible from implausible inputs.
+The case is very clear for MNIST: ECCCos are consistently more faithful than the corresponding counterfactuals produced by any of the benchmark generators and their plausibility gradually improves through ensembling and Joint Energy modelling.
+For the JEM Ensemble, ECCCo is essentially on par with REVISE and does significantly better than the baseline generator.
+
+The results for GMSC ...
+
+\import{contents/}{table-real-world.tex}
+
+\subsection{Discussion of Results}\label{disc}
 
@@
 
-While $\mathbf{x}_J$ is only guaranteed to distribute as $p_{\theta}(\mathbf{x}|\mathbf{y}^*)$ if $\epsilon \rightarrow 0$ and $J \rightarrow \infty$, the bias introduced for a small finite $\epsilon$ is negligible in practice~\citep{murphy2023probabilistic,grathwohl2020your}.
+While $\mathbf{x}_J$ is only guaranteed to distribute as $p_{\theta}(\mathbf{x}|\mathbf{y}^+)$ if $\epsilon \rightarrow 0$ and $J \rightarrow \infty$, the bias introduced for a small finite $\epsilon$ is negligible in practice~\citep{murphy2023probabilistic,grathwohl2020your}.
 While~\citet{grathwohl2020your} use Equation~\ref{eq:sgld} during training, we are interested in applying the conditional sampling procedure in a post-hoc fashion to any standard discriminative model.

### 2023-05-10 paper/paper.tex
first draft doen

@@
 This effectively reallocates the task of learning realistic representations of the data from the model itself to the surrogate.
-Consequently, the generated explanations may seem plausible to humans but need not necessarily faithfully describe the behaviour of the black-box model.
+Consequently, the generated explanations may seem plausible to humans but need not necessarily describe the behaviour of the black-box model faithfully.
 We formalise this notion of faithfulness through the introduction of a tailored evaluation metric and propose a novel algorithmic framework for generating \textbf{E}nergy-\textbf{C}onstrained \textbf{C}onformal \textbf{Co}unterfactuals that are only as plausible as the model permits.
-Through extensive empirical studies involving multiple synthetic and real-world datasets, we demonstrate that \textbf{ECCCo} reconciles the need for plausibility and faithfulness.
-In particular, we show that it is possible to achieve state-of-the-art plausibility for models with gradient access without the need for surrogate models.
-To do so, ECCCo relies solely on properties defining the black-box model itself by leveraging recent advances in energy-based modelling and conformal inference.
-Through this work, we also shine new light on the explanatory properties of Joint Energy Models.
-Our framework is intuitive, flexible and fully open-sourced.
+Through extensive empirical studies involving multiple synthetic and real-world datasets, we demonstrate that \textbf{ECCCos} reconcile the need for plausibility and faithfulness.
+In particular, we show that it is possible to achieve state-of-the-art plausibility without the need for surrogate models.
+To do so, our framework relies solely on properties defining the black-box model itself by leveraging recent advances in energy-based modelling and conformal inference.
+While this work is limited in scope and our proposed methodology is only readily applicable to models with gradient access, we anticipate that ECCCo can serve as a baseline for future research directed at providing plausible but faithful Counterfactual Explanations.
 By highlighting the need for faithfulness in the context of Counterfactual Explanations, we believe that in the short term, our work will enable researchers and practitioners to better distinguish trustworthy from unreliable models.
-We further anticipate that ECCCo can serve as a baseline for future research directed at providing plausible but faithful Counterfactual Explanations.
 \end{abstract}
@@
     \caption{Explanations or Adversarial Examples?
-Counterfactuals for turning an 8 (eight) into a 3 (three): original image (left); counterfactual produced using~\citet{wachter2017counterfactual} (centre); and a counterfactual produced using the approach introduced by~\citep{schut2021generating} that uses Jacobian-Based Saliency Map Attacks to solve Equation~\ref{eq:general}.}\label{fig:adv}
+Counterfactuals for turning a 9 (nine) into a 7 (seven): original image (left); counterfactual produced using~\citet{wachter2017counterfactual} (centre); and a counterfactual produced using the approach introduced by~\citep{schut2021generating} that uses Jacobian-Based Saliency Map Attacks to solve Equation~\ref{eq:general}.}\label{fig:adv}
   \end{minipage}\hfill
@@
     \caption{Using surrogates can improve plausibility, but also increases vulnerability.
-Counterfactuals for turning an 8 (eight) into a 3 (three): original image (left); counterfactual produced using REVISE~\citep{joshi2019realistic} with a well-specified surrogate (centre); and a counterfactual produced using REVISE~\citep{joshi2019realistic} with a poorly specified surrogate (right).}\label{fig:vae}
+Counterfactuals for turning an 9 (nine) into a 7 (seven): original image (left); counterfactual produced using REVISE~\citep{joshi2019realistic} with a well-specified surrogate (centre); and a counterfactual produced using REVISE~\citep{joshi2019realistic} with a poorly specified surrogate (right).}\label{fig:vae}
   \end{minipage}
@@
   \includegraphics[width=\textwidth]{../artifacts/results/images/poc.png}
-  \captionof{figure}{[PLACEHOLDER] Vector fields indicating the direction of gradients with respect to the different components of the ECCCo objective (Equation~\ref{eq:eccco}).} \label{fig:eccco}
+  \captionof{figure}{[PLACEHOLDER; may swap for:] Vector fields indicating the direction of gradients with respect to the different components of the ECCCo objective (Equation~\ref{eq:eccco}).} \label{fig:eccco}
 \end{minipage}  
@@
   \includegraphics[width=\textwidth]{../artifacts/results/images/mnist_eccco.png}
-  \captionof{figure}{[SUBJECTO TO CHANGE] Original image (left) and ECCCos for turning an 8 (eight) into a 3 (three) for different Black Boxes from left to right: Multi-Layer Perceptron (MLP), Ensemble of MLPs, Joint Energy Model (JEM), Ensemble of JEMs.}\label{fig:eccco-mnist}
+  \captionof{figure}{Original image (left) and ECCCos for turning a 9 (nine) into a 7 (seven) for different black-box models from left to right: Multi-Layer Perceptron (MLP), Ensemble of MLPs, Joint Energy Model (JEM), Ensemble of JEMs.}\label{fig:eccco-mnist}
 \end{minipage}
@@
 Starting with the high-level results for our Linearly Separable data, we find that ECCCo produces the most faithful counterfactuals for both black-box models.
-It also produces the most plausible counterfactuals for the Joint Energy Model, which was explicitly trained to learn plausible representations of the input data.
-This high-level pattern is broadly consistent across all datasets and supportive of our narrative, so it is worth highlighting here, before we discuss it below: ECCCos consistently achieve high faithfulness, which---subject to the quality of the black-box model itself---coincides with high plausibility.
+This is not surprising, since ECCCo directly enforces faithfulness through regularization.
+Crucially though, ECCCo also produces the most plausible counterfactuals for the Joint Energy Model, which was explicitly trained to learn plausible representations of the input data.
+This high-level pattern is broadly consistent across all datasets and supportive of our narrative, so it is worth highlighting it here: ECCCos consistently achieve high faithfulness, which---subject to the quality of the black-box model itself---coincides with high plausibility.
 
@@
 
-\subsection{Discussion of Results}\label{disc}
+To conclude this section, we summarize our findings with reference to the opening questions.
+Concerning the feasibility of our proposed methodology (Research Question~\ref{rq:feasibility}), our findings clearly demonstrate that it is indeed possible to generate plausible counterfactuals without the need for surrogate models.
+A related important finding is that ECCCo never sacrifices faithfulness for plausibility: any plausible ECCCo also faithfully describes model behaviour.
+This mitigates the risk of generating plausible explanations for models that are, in fact, highly susceptible to implausible counterfactuals as well.
+Our findings here indicate that ECCCo achieves this result primarily by leveraging the model's generative property.
+We think that further work is needed, however, to definitively answer Research Question~\ref{rq:drivers}, on which we elaborate in the following section.
 
-Consistent with the findings in~\citet{schut2021generating}, we have demonstrated that predictive uncertainty estimates can be leveraged to generate plausible counterfactuals.
-Interestingly, \citet{schut2021generating} point out that this finding --- as intuitive as it is --- may be linked to a positive connection between the generative task and predictive uncertainty quantification.
-In particular, \citet{grathwohl2020your} demonstrate that their proposed method for integrating the generative objective in training yields models that have improved predictive uncertainty quantification.
-Since neither~\citet{schut2021generating} nor we have employed any surrogate generative models, our findings seem to indicate that the positive connection found in~\citet{grathwohl2020your} is bidirectional.
+\section{Limitations}
 
-\subsection{Limitations}
+Even though we have taken considerable measures to study our proposed methodology carefully, this work is limited in scope, which caveats our findings.
+In particular, we have found that the performance of ECCCo is sensitive to hyperparameter choices.
+In order to achieve faithfulness, we generally had to penalise the distance from generated samples slightly more than the distance from factual values.
+This choice is associated with relatively higher costs to individuals since the proposed recourses typically involve more substantial feature changes than for our benchmark generators.
 
-\begin{itemize}
-  \item BatchNorm does not seem compatible with JEM
-  \item Coverage and temperature impacts CCE in somewhat unpredictable ways
-  \item It seems that models that are not explicitly trained for generative task, still learn it implictly
-  \item Batch size seems to impact quality of generated samples (at inference, but not so much during JEM training)
-  \item SGLD takes time 
-  \item REVISE has benefit of lower dimensional space
-  \item For MNIST it seems that ECCCo is better at reducing pixel values than increasing them (better at erasing than writing)
-  \item JEMs are more difficult to train
-  \item There is a tradeoff: higher cost vs. higher faithfulness/plausibility
-  \item Results are sensitive to choices of penalty strength and step size
-  \item Counterfactuals may end up looking fairly homogenous
-  \item For MNIST data we found CP to have little effect
-  \item JEMs themselves are sensitive to scale
-  \item ECCCo can backfire, in case generative property of model is poor
-\end{itemize}
+Conversely, we have not found that penalising prediction set sizes disproportionately strongly had any discernable effect on our results.
+As discussed above, we also struggled to achieve good results by relying on conformal prediction alone.
+We want to caveat this finding by acknowledging that the role of CP in this context needs to be investigated more thoroughly through future work.
+Our suggested approach involving a smooth set size penalty may be insufficient in this context.
+
+The fact that our findings are primarily driven by applying ideas from energy-based modelling presents a challenge in itself: while our approach is readily applicable to models with gradient access like deep neural networks, more work is needed to generalise our methodology to other popular machine learning models such as gradient-boosted trees.
+Relatedly, we have encountered common challenges associated with energy-based modelling during our experiments including sensitivity to scale, training instabilities and sensitivity to hyperparameters.
+We have also struggled to apply our proposed approach to low-dimensional tabular data.
 
 \section{Conclusion}
+
+This work leverages recent advances in energy-based modelling and conformal prediction in the context of Explainable Artificial Intelligence.
+We have proposed a new way to generate Counterfactual Explanations that are maximally faithful to the black-model they aim to explain.
+Our proposed counterfactual generator, ECCCo, produces plausible counterfactual if and only if the black-model itself has learned realistic representations of the data.
+This should enable researchers and practitioners to use counterfactuals in order to discern trustworthy models from unreliable ones.
+While the scope of this work limits its generalizability, we believe that ECCCo offers a solid baseline for future work on faithful Counterfactual Explanations.
+
+\begin{ack}
+
+Some of the members of TU Delft were partially funded by ICAI AI for Fintech Research, an ING — TU Delft
+collaboration.
+
+\end{ack}
 
@@
 
-\begin{ack}
-Use unnumbered first level headings for the acknowledgments. All acknowledgments
-go at the end of the paper before the list of references. Moreover, you are required to declare
-funding (financial activities supporting the submitted work) and competing interests (related financial activities outside the submitted work).
-More information about this disclosure can be found at: \url{https://neurips.cc/Conferences/2023/PaperInformation/FundingDisclosure}.
-
-Do {\bf not} include this section in the anonymized submission, only in the final paper.
-You can use the \texttt{ack} environment provided in the style file to autmoatically hide this section in the anonymized submission.
-\end{ack}
-
 \end{document}

### 2023-05-11 notebooks/fashion_mnist.qmd
fashion mnist good generative, poor predictive, n1000

@@
 # Data:
-n_obs = 10000
+n_obs = 1000
 counterfactual_data = load_fashion_mnist(n_obs)
@@
 batch_size = minimum([Int(round(n_obs/10)), 128])
-n_hidden = 128
+n_hidden = 512
 activation = Flux.swish
@@
     Dense(n_in, n_hidden, activation),
+    Dense(n_hidden, n_hidden, activation),
+    Dense(n_hidden, n_hidden, activation),
     Dense(n_hidden, n_out),
@@
 )
-α = [1.0,1.0,2e-2]      # penalty strengths
+α = [1.0,1.0,1e-2]      # penalty strengths
 ```
@@
     ),
-    sampling_steps=25,
+    sampling_steps=30,
 )
@@
     x::Union{AbstractArray, Int}=x_factual, target::Int=target;
-    λ=[0.1,0.25,0.25],
+    λ=[0.1,0.3,0.3],
     temp=0.1,
@@
 λ₁ = 0.1
-λ₂ = 0.25
-λ₃ = 0.25
+λ₂ = 0.3
+λ₃ = 0.3
 Λ = [λ₁, λ₂, λ₃]

### 2023-05-11 paper/paper.tex
fashion mnist good generative, poor predictive, n1000

@@
 To address the need for plausible explanations, existing work has primarily relied on surrogate models to learn how the input data is distributed.
-This effectively reallocates the task of learning realistic representations of the data from the model itself to the surrogate.
+This effectively reallocates the task of learning realistic explanations for the data from the model itself to the surrogate.
 Consequently, the generated explanations may seem plausible to humans but need not necessarily describe the behaviour of the black-box model faithfully.
-We formalise this notion of faithfulness through the introduction of a tailored evaluation metric and propose a novel algorithmic framework for generating \textbf{E}nergy-\textbf{C}onstrained \textbf{C}onformal \textbf{Co}unterfactuals that are only as plausible as the model permits.
-Through extensive empirical studies involving multiple synthetic and real-world datasets, we demonstrate that \textbf{ECCCos} reconcile the need for plausibility and faithfulness.
-In particular, we show that it is possible to achieve state-of-the-art plausibility without the need for surrogate models.
-To do so, our framework relies solely on properties defining the black-box model itself by leveraging recent advances in energy-based modelling and conformal inference.
-While this work is limited in scope and our proposed methodology is only readily applicable to models with gradient access, we anticipate that ECCCo can serve as a baseline for future research directed at providing plausible but faithful Counterfactual Explanations.
-By highlighting the need for faithfulness in the context of Counterfactual Explanations, we believe that in the short term, our work will enable researchers and practitioners to better distinguish trustworthy from unreliable models.
+We formalise this notion of faithfulness through the introduction of a tailored evaluation metric and propose a novel algorithmic framework for generating \textbf{E}nergy-\textbf{C}onstrained \textbf{C}onformal \textbf{Co}unterfactuals (ECCCos) that are only as plausible as the model permits.
+Through extensive empirical studies involving multiple synthetic and real-world datasets, we demonstrate that ECCCos reconcile the need for plausibility and faithfulness.
+In particular, we show that it is possible to achieve state-of-the-art plausibility for models with gradient access without the need for surrogate models.
+To do so, our framework relies solely on properties defining the black-box model itself by leveraging recent advances in energy-based modelling and conformal prediction.
+To our knowledge, this is the first venture in this direction for generating faithful Counterfactual Explanations.
+Thus, we anticipate that ECCCos can serve as a baseline for future research.
+We believe that our work opens avenues for researchers and practitioners seeking tools to better distinguish trustworthy from unreliable models.
 \end{abstract}

### 2023-05-12 notebooks/fashion_mnist.qmd
fmnist

@@
 # Model parameters:
-epochs = 100
+epochs = 150
 batch_size = minimum([Int(round(n_obs/10)), 128])
@@
     ),
-    sampling_steps=30,
+    sampling_steps=35,
 )

### 2023-05-12 paper/paper.tex
circles and gmsc

@@
 The findings for the Moons dataset are broadly in line with the findings so far: for the JEM, ECCCo yields significantly more faithful and plausible counterfactuals than all other generators.
-For the MLP, faithfulness is maintained but not plausibility.
-For comparison, REVISE yields fairly plausible counterfactuals in both cases, but it does so at the cost of faithfulness.
-For the Circles data, ...
+For the MLP, faithfulness is maintained but counterfactuals are not plausible.
+By comparison, REVISE yields fairly plausible counterfactuals in both cases, but it does so at the cost of faithfulness.
+We also observe that the best results for ECCCo are achieved when using both penalties.
+Once again though, the generative component (EBM) has a stronger impact on the positive results for the JEM.
+a
+
+For the Circles data, the most faithful counterfactuals are generated by ECCCo.
+While it appears that REVISE generates the most plausible counterfactuals in this case, we note that they are valid only half of the time (see Appendix~\ref{app:results} for a complete overview of all evaluation metrics).
+It turns out that in this case, the underlying VAE with default parameters has not adequately learned the data-generating process.
+Of course, it is possible to achieve better generative performance through hyperparameter tuning.
+But this example serves to illustrate that REVISE depends strictly on the quality of the surrogate model.
+Independent of the outcome for REVISE, however, the results do not seem to indicate that ECCCo significantly improves our plausibility metric for the Circles data.
+
+\textit{NOTE: visual inspection of the results shows that ECCCo still does exactly what it's supposed to do.
+I think the outcome here demonstrates the limitations of the implausibility metric: computing average distances to randomly drawn samples from the target class does not really make perfect sense when samples are distributed as a circle.
+But this story may be too convoluted to add here?}
 
@@
 Once again the findings indicate that the plausibility of ECCCos is positively correlated with the capacity of the black-box model to distinguish plausible from implausible inputs.
-The case is very clear for MNIST: ECCCos are consistently more faithful than the corresponding counterfactuals produced by any of the benchmark generators and their plausibility gradually improves through ensembling and Joint Energy modelling.
+The case is very clear for MNIST: ECCCos are consistently more faithful than the corresponding counterfactuals produced by any of the benchmark generators and their plausibility gradually improves through ensembling and joint-energy modelling.
 For the JEM Ensemble, ECCCo is essentially on par with REVISE and does significantly better than the baseline generator.
+We also note that ECCCo is the only generator that consistently achieves full validity for all models (Appendix~\ref{app:results}).
+Interestingly, ECCCo also yields lower-cost outcomes than the baseline generator for the JEMs.
 
-The results for GMSC ...
+For the tabular credit dataset (GMSC) we have struggled to get good generative and discriminative performance for our JEMs.
+Consequently, it is not surprising to find that ECCCo never achieves state-of-the-art plausibility, although it does improve outcomes compared to the baseline (Wachter).
+With respect to faithfulness, ECCCo once again consistently outperforms all other counterfactual generators.
 

### 2023-05-14 notebooks/linearly_separable.qmd
some working on cutting down pages

@@
 n_ens = 5                                   # number of models in ensemble
-_loss = Flux.Losses.logitcrossentropy       # loss function
-_finaliser = x -> x                         # finaliser function
+_loss = Flux.Losses.crossentropy       # loss function
+_finaliser = Flux.softmax                   # finaliser function
 ```
@@
     input_size=(input_dim,), 
-    batch_size=batch_size,
+    batch_size=50,
 )
@@
 ```{julia}
-n_regen = 200
+n_regen = 1000
 n_each = batch_size
@@
 
+```{julia}
+#| output: true
+#| echo: false
+#| label: fig-losses
+#| fig-cap: "Illustration of the smooth size loss and the configurable classification loss."
+
+X_plot = matrix(X)
+temp = 0.1
+
+for (mod_name, model) in model_dict
+    p1 = Plots.contourf(model.model, model.fitresult, X_plot, labels; plot_set_loss=true, zoom=0, temp=temp)
+    p2 = Plots.contourf(model.model, model.fitresult, X_plot, labels; plot_classification_loss=true, zoom=0, temp=temp, clim=nothing, loss_matrix=ones(2,2))
+    display(Plots.plot(p1, p2, size=(800,320), plot_title=mod_name))
+end
+```
+
 ## Benchmark
@@
 
+opt = Flux.Optimise.Descent(0.05)
+
 # Benchmark generators:
 generator_dict = Dict(
-    "Wachter" => WachterGenerator(λ=λ₁),
-    "REVISE" => REVISEGenerator(λ=λ₁),
+    "Wachter" => WachterGenerator(λ=λ₁, opt=opt),
+    "REVISE" => REVISEGenerator(λ=λ₁, opt=opt),
     "Schut" => GreedyGenerator(),
-    "ECCCo" => ECCCoGenerator(λ=Λ),
-    "ECCCo (no CP)" => ECCCoGenerator(λ=[λ₁, 0.0, λ₃]),
-    "ECCCo (no EBM)" => ECCCoGenerator(λ=[λ₁, λ₂, 0.0]),
+    "ECCCo" => ECCCoGenerator(λ=Λ, opt=opt),
+    "ECCCo (no CP)" => ECCCoGenerator(λ=[λ₁, 0.0, λ₃], opt=opt),
+    "ECCCo (no EBM)" => ECCCoGenerator(λ=[λ₁, λ₂, 0.0], opt=opt),
 )
@@
 
-M = model_dict["JEM"]
+M = model_dict["MLP"]
 X = X isa Matrix ? X : Float32.(permutedims(matrix(X)))
-factual_label =  levels(labels)[1]
+factual_label =  levels(labels)[2]
 x_factual = reshape(X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
-target =  levels(labels)[2]
+target =  levels(labels)[1]
 factual = predict_label(M, counterfactual_data, x_factual)[1]

### 2023-05-14 notebooks/mnist.qmd
some working on cutting down pages

@@
 n_ens = 5                                   # number of models in ensemble
-_loss = Flux.Losses.logitcrossentropy       # loss function
-_finaliser = x -> x                         # finaliser function
+_loss = Flux.Losses.crossentropy       # loss function
+_finaliser = Flux.softmax                        # finaliser function
 ```

### 2023-05-14 notebooks/poc.qmd
some working on cutting down pages

@@
 end
+```
+
+```{julia}
+#| output: true
+#| echo: false
+#| label: fig-losses
+#| fig-cap: "Illustration of the smooth size loss and the configurable classification loss."
+
+temp = 0.25
+p1 = Plots.contourf(mach.model, mach.fitresult, permutedims(X), labels; plot_set_loss=true, zoom=0, temp=temp)
+p2 = Plots.contourf(mach.model, mach.fitresult, permutedims(X), labels; plot_classification_loss=true, zoom=0, temp=temp, clim=nothing, loss_matrix=ones(2,2))
+display(Plots.plot(p1, p2, size=(800,320)))
 ```

### 2023-05-14 paper/paper.tex
some working on cutting down pages

@@
 
-Counterfactual Explanations provide a powerful, flexible and intuitive way to not only explain black-box models but also enable affected individuals to challenge them through the means of Algorithmic Recourse.
-Instead of opening the black box, Counterfactual Explanations work under the premise of strategically perturbing model inputs to understand model behaviour~\citep{wachter2017counterfactual}.
-Intuitively speaking, we generate explanations in this context by asking simple what-if questions of the following nature: `Our credit risk model currently predicts that this individual's credit profile is too risky to offer them a loan.
-What if they reduced their monthly expenditures by 10\%?
-Will our model then predict that the individual is credit-worthy'?
+Counterfactual Explanations (CE) provide a powerful, flexible and intuitive way to not only explain black-box models but also help affected individuals through the means of Algorithmic Recourse.
+Instead of opening the Black Box, CE works under the premise of strategically perturbing model inputs to understand model behaviour~\citep{wachter2017counterfactual}.
+Intuitively speaking, we generate explanations in this context by asking what-if questions of the following nature: `Our credit risk model currently predicts that this individual is not credit-worthy.
+What if they reduced their monthly expenditures by 10\%?'
 
@@
 Counterfactuals are then searched by minimizing a loss function that compares the predicted model output to the target outcome: $\text{yloss}(M_{\theta}(\mathbf{x}),\mathbf{y}^+)$.
-Since Counterfactual Explanations (CE) work directly with the black-box model, valid counterfactuals always have full local fidelity by construction~\citep{mothilal2020explaining}.
-Fidelity is defined as the degree to which explanations approximate the predictions of the black-box model.
-This is arguably one of the most important evaluation metrics for model explanations, since any explanation that explains a prediction not actually made by the model is useless~\citep{molnar2020interpretable}.
+Since Counterfactual Explanations work directly with the black-box model, valid counterfactuals always have full local fidelity by construction where fidelity is defined as the degree to which explanations approximate the predictions of a black-box model~\citep{mothilal2020explaining,molnar2020interpretable}.
 
-In situations where full fidelity is a requirement, CE therefore offers a more appropriate solution to Explainable Artificial Intelligence (XAI) than other popular approaches like LIME~\citep{ribeiro2016why} and SHAP~\citep{lundberg2017unified}, which involve local surrogate models.
+In situations where full fidelity is a requirement, CE offers a more appropriate solution to Explainable Artificial Intelligence (XAI) than other popular approaches like LIME~\citep{ribeiro2016why} and SHAP~\citep{lundberg2017unified}, which involve local surrogate models.
 But even full fidelity is not a sufficient condition for ensuring that an explanation faithfully describes the behaviour of a model.
-That is because multiple very distinct explanations can all lead to the same model prediction, especially when dealing with heavily parameterized models like deep neural networks which are typically underspecified by the available data~\citep{wilson2020case}.
+That is because multiple very distinct explanations can all lead to the same model prediction, especially when dealing with heavily parameterized models like deep neural networks which are typically underspecified by the data~\citep{wilson2020case}.
 
 In the context of CE, the idea that no two explanations are the same arises almost naturally.
-A key focus in the literature has therefore been to identify those explanations and algorithmic recourses that are deemed most appropriate based on a myriad of desiderata such as sparsity, actionability and plausibility.
-In this work, we draw closer attention to the insufficiency of model fidelity as an evaluation metric for the faithfulness of counterfactual explanations.
-Our key contributions are as follows: firstly, we introduce a new notion of faithfulness that is suitable for counterfactuals and propose a novel evaluation metric that draws inspiration from recent advances in Energy-Based Modelling (EBM); secondly, we a novel algorithmic approach for generating Energy-Constrained Conformal Counterfactuals (ECCCo) that explicitly address the need for faithfulness; finally, we provide illustrative examples and extensive empirical evidence demonstrating that ECCCos faithfully explain model behaviour without sacrificing existing desidarata like plausibility and sparsity.
+A key focus in the literature has therefore been to identify those explanations and algorithmic recourses that are most appropriate based on a myriad of desiderata such as sparsity, actionability and plausibility.
+In this work, we draw closer attention to model faithfulness as a desideratum for counterfactuals.
+Our key contributions are as follows: firstly, we propose a definition of faithfulness that is suitable for counterfactuals; secondly, we introduce a novel algorithmic approach for generating Energy-Constrained Conformal Counterfactuals (ECCCos) that explicitly addresses the need for faithfulness; finally, we provide extensive empirical evidence demonstrating that ECCCos faithfully explain model behaviour without sacrificing plausibility.
 
@@
 
-In this section, we provide some background on Counterfactual Explanations and our motivation for this work.
-To start, we briefly introduce the methodology underlying most state-of-the-art (SOTA) counterfactual generators.
+In this section, we provide some background on CE and our motivation for this work. To start, we briefly introduce the methodology underlying most state-of-the-art (SOTA) counterfactual generators.
 
@@
 Let $\mathcal{Y}=(0,1)^K$ denote the one-hot-encoded output domain with $K$ classes.
-Then most SOTA counterfactual generators rely on gradient descent to optimize different flavours of the following counterfactual search objective:
+Then most counterfactual generators rely on gradient descent to optimize different flavours of the following counterfactual search objective:
 
@@
 Encodings may involve simple feature transformations or more advanced techniques involving generative models, as we will discuss further below.
-The baseline approach, which we will simply refer to as \textbf{Wachter}~\citep{wachter2017counterfactual}, searches a single counterfactual directly in the feature space and penalises its distance between the original factual.
+The baseline approach, which we will simply refer to as \textbf{Wachter}~\citep{wachter2017counterfactual}, searches a single counterfactual directly in the feature space and penalises its distance to the original factual.
 
@@
 In particular, they propose to use the latent embedding of a Variational Autoencoder (VAE) trained to generate samples $\mathbf{x}^* \leftarrow \mathcal{G}(\mathbf{z})$ where $\mathcal{G}$ denotes the decoder part of the VAE.
-Provided the surrogate model is well-trained, their proposed approach ---REVISE--- can yield compelling counterfactual explanations like the one in the centre panel of Figure~\ref{fig:vae}.
+Provided the surrogate model is well-trained, their proposed approach ---\textbf{REVISE}--- can yield plausible explanations like the one in the centre panel of Figure~\ref{fig:vae}.
 
@@
 Provided the model is well-specified, these counterfactuals are plausible.
-Unfortunately, this idea hinges on the assumption that the black-box model provides well-calibrated predictive uncertainty estimates.
+This idea hinges on the assumption that the black-box model provides well-calibrated predictive uncertainty estimates.
 
@@
 This may explain why research has primarily focused on other desiderata, most notably plausibility (Definition~\ref{def:plausible}).
-Enquiring about the plausibility of a counterfactual essentially boils down to the following question: `Is this counterfactual consistent with the underlying data'?
-We posit a related, slightly more nuanced question: `Is this counterfactual consistent with what the model has learned about the underlying data'?
-We will argue that fidelity is not a sufficient evaluation metric to answer this question and propose a novel way to assess if Counterfactual Explanations conform with model behaviour.
+Enquiring about the plausibility of a counterfactual essentially boils down to the following question: `Is this counterfactual consistent with the data'?
+We ask a related question: `Is this counterfactual consistent with what the model has learned about the data'?
+To answer this question and propose a novel way to assess if explanations conform with model behaviour.
 
@@
 As we explained in Section~\ref{background}, model explanations are generally considered faithful if their corresponding predictions coincide with the predictions made by the model itself.
-Since this definition of faithfulness is not useful in the context of Counterfactual Explanations, we propose an adapted version:
+Since this definition of faithfulness is not useful in the context of CE, we propose the following definition:
 
@@
 
-In words, conformal counterfactuals conform with what the predictive model has learned about the input data $\mathbf{x}$.
-Since this definition works with distributional properties, it explicitly accounts for the multiplicity of explanations we discussed earlier.
 To assess counterfactuals with respect to Definition~\ref{def:faithful}, we need to be able to quantify the posterior conditional distribution $p_{\theta}(\mathbf{x}|\mathbf{y}^+)$.
-This is very much at the core of our proposed methodological framework, which reconciles the notions of plausibility and model faithfulness and which we will introduce next.
+This is very much at the core of our proposed methodological framework, which we will introduce next.
 
@@
 
-Recent work by~\citet{grathwohl2020your} on Energy Based Models (EBM) has pointed out that there is a `generative model hidden within every standard discriminative model'.
-The authors show that we can draw samples from the posterior conditional distribution $p_{\theta}(\mathbf{x}|\mathbf{y})$ using Stochastic Gradient Langevin Dynamics (SGLD).
-The authors use this insight to train classifiers jointly for the discriminative task using standard cross-entropy and the generative task using SGLD.
-They demonstrate empirically that among other things this improves predictive uncertainty quantification for discriminative models.
-Our findings in this work suggest that Joint Energy Models (JEM) also tend to yield more plausible Counterfactual Explanations.
-Based on the definition of plausible counterfactuals (Definition~\ref{def:plausible}) this is not surprising.
+Recent work by~\citet{grathwohl2020your} on Energy-Based Models (EBM) has shown that there is a `generative model hidden within every standard discriminative model'.
+They use Stochastic Gradient Langevin Dynamics (SGLD) to estimate the posterior conditional distribution $p_{\theta}(\mathbf{x}|\mathbf{y})$ using and leverage this to train classifiers jointly for the discriminative and generative task.
 
-Crucially for our purpose, one can apply their proposed sampling strategy during inference to essentially any standard discriminative model.
+Crucially for our purpose, SGLD can be applied during inference to essentially any standard discriminative model.
 Even models that are not explicitly trained for the joint objective learn about the distribution of inputs $X$ by learning to make conditional predictions about the output $y$.
-We can leverage this observation to quantify the generative property of the Black Box model itself.
+We can leverage this observation to quantify the generative property of models.
 In particular, note that if we fix $\mathbf{y}$ to our target value $\mathbf{y}^+$, we can sample from $p_{\theta}(\mathbf{x}|\mathbf{y}^+)$ using SGLD as follows,
@@
 where $\mathbf{r}_j \sim \mathcal{N}(\mathbf{0},\mathbf{I})$ is the stochastic term and the step-size $\epsilon$ is typically polynomially decayed.
-The term $\mathcal{E}(\mathbf{x}_j|\mathbf{y}^+)$ denotes the energy function where we use $\mathcal{E}(\mathbf{x}_j|\mathbf{y}^+)=-M_{\theta}(\mathbf{x}_j)[\mathbf{y}^+]$, that is the negative logit corresponding to the target class label $\mathbf{y}^+$.
+The term $\mathcal{E}(\mathbf{x}_j|\mathbf{y}^+)$ denotes the model energy conditioned on the target class label $\mathbf{y}^+$.
 Generating multiple samples in this manner yields an empirical distribution $\hat{\mathbf{X}}_{\theta,\mathbf{y}^+}$ that we use in our search for plausible counterfactuals, as discussed in more detail below.
@@
 where $\hat{\mathbf{x}}_{\theta}$ denotes samples generated using SGLD (Equation~\ref{eq:sgld}) and $\text{dist}(\cdot)$ is a generic term for a distance metric.
-Our default choice for $\text{dist}(\cdot)$ is the L1 Norm, or Manhattan distance, since it induces sparsity.
+Our default choice for $\text{dist}(\cdot)$ is the L1 Norm since it induces sparsity.
 
@@
 
-\begin{minipage}[c]{0.40\textwidth}
+\begin{minipage}[c]{0.45\textwidth}
   \centering
-  \includegraphics[width=\textwidth]{../artifacts/results/images/poc.png}
+  \includegraphics[width=0.8\textwidth]{../artifacts/results/images/poc.png}
   \captionof{figure}{[PLACEHOLDER; may swap for:] Vector fields indicating the direction of gradients with respect to the different components of the ECCCo objective (Equation~\ref{eq:eccco}).} \label{fig:eccco}
@@
 \hfill
-\begin{minipage}[c]{0.50\textwidth}
+\begin{minipage}[c]{0.45\textwidth}
 \captionof{algorithm}{Generating ECCCos (For more details, see Appendix~\ref{app:eccco})}\label{alg:eccco}
@@
 \begin{minipage}[c]{\textwidth}
-  \includegraphics[width=\textwidth]{../artifacts/results/images/mnist_eccco.png}
+  \centering
+  \includegraphics[width=0.8\textwidth]{../artifacts/results/images/mnist_eccco.png}
   \captionof{figure}{Original image (left) and ECCCos for turning a 9 (nine) into a 7 (seven) for different black-box models from left to right: Multi-Layer Perceptron (MLP), Ensemble of MLPs, Joint Energy Model (JEM), Ensemble of JEMs.}\label{fig:eccco-mnist}
@@
 
-In this section, we present our empirical analysis and findings. Our goal is to shed line on the following questions:
+Our goal in this section is to shed light on the following research questions:
 
@@
 
-In the following, we first briefly describe our evaluation framework and experimental setup, before presenting and discussing our results.
+We first briefly describe our experimental setup, before presenting our main results.
 
@@
 In order to quantify the plausibility of counterfactuals we use a slightly adapted version of the implausibility metric proposed in~\citet{guidotti2022counterfactual}.
-Formally, for a single counterfactual, we define implausibility as follows,
+Formally, we define implausibility as follows,
 
@@
 
-Specifically, we form this subset based on the $n_E$ generated samples associated with the lowest energy.
+Specifically, we form this subset based on the $n_E$ generated samples with the lowest energy.
 
@@
 The synthetic datasets include: a dataset containing two \textbf{Linearly Separable} Gaussian clusters ($n=1000$), as well as the well-known \textbf{Circles} ($n=1000$) and \textbf{Moons} ($n=2500$) data.
-As for real-world data, we follow~\citet{schut2021generating} and use the \textbf{MNIST}~\citep{lecun1998mnist} image dataset.
-It is composed of 60000 images of 28x28 pixels each showing handwritten digits from 0 to 9 such as the examples shown above.
+As for real-world data, we follow~\citet{schut2021generating} and use the \textbf{MNIST}~\citep{lecun1998mnist} dataset containing images of handwritten digits such as the examples shown above.
 From the social sciences domain, we include Give Me Some Credit (\textbf{GMSC})~\citep{kaggle2011give}: a tabular dataset that has been studied extensively in the literature on Algorithmic Recourse~\citep{pawelczyk2021carla}.
@@
 
-As with the example in Section~\ref{meth}, we use simple neural networks (\textbf{MLP}), ensembles of neural networks (\textbf{MLP Ensemble}), Joint Energy Models (\textbf{JEM}) and ensembles of JEMs (\textbf{JEM Ensemble}) to model our real-world datasets.
-For the synthetic datasets, we found that the use of ensembles was not necessary.
-
+As with the example in Section~\ref{meth}, we use simple neural networks (\textbf{MLP}) and Joint Energy Models (\textbf{JEM}).
+For the more complex real-world datasets we also use ensembling in each case.
 To account for stochasticity, we generate multiple counterfactuals for each possible target class, generator, model and dataset.
@@
 We set $n^{-}=25$ for all of our synthetic datasets, $n^{-}=10$ for GMSC and $n^{-}=5$ for MNIST.
-Note that in the latter case, we still end up generating disproportionately more counterfactuals because all 10 digits appear both as factuals and targets for each other.
 Full details concerning our parameter choices, training procedures and model performance can be found in Appendix~\ref{app:setup}.
@@
 
-\import{contents/}{table-synthetic.tex}
-
 Starting with the high-level results for our Linearly Separable data, we find that ECCCo produces the most faithful counterfactuals for both black-box models.
@@
 Crucially though, ECCCo also produces the most plausible counterfactuals for the Joint Energy Model, which was explicitly trained to learn plausible representations of the input data.
-This high-level pattern is broadly consistent across all datasets and supportive of our narrative, so it is worth highlighting it here: ECCCos consistently achieve high faithfulness, which---subject to the quality of the black-box model itself---coincides with high plausibility.
+This high-level pattern is broadly consistent across all datasets and supportive of our narrative, so it is worth highlighting: ECCCos consistently achieve high faithfulness, which---subject to the quality of the model itself---coincides with high plausibility.
 
@@
 Conformal Prediction alone only leads to marginally improved faithfulness and plausibility relative to the benchmark generators for our JEM.
-
-As a final key observation for the Linearly Separable data we note that for the MLP, increased faithfulness comes at the cost of reduced plausibility.
+As a final observation for the Linearly Separable data we note that for the MLP, increased faithfulness comes at the cost of reduced plausibility.
 Specifically, this means that counterfactuals generated through ECCCo end up further away from individuals in the target class than those produced by our benchmark generators.
@@
 Once again though, the generative component (EBM) has a stronger impact on the positive results for the JEM.
-a
+
+\import{contents/}{table-synthetic.tex}
 
@@
 Independent of the outcome for REVISE, however, the results do not seem to indicate that ECCCo significantly improves our plausibility metric for the Circles data.
-
-\textit{NOTE: visual inspection of the results shows that ECCCo still does exactly what it's supposed to do.
-I think the outcome here demonstrates the limitations of the implausibility metric: computing average distances to randomly drawn samples from the target class does not really make perfect sense when samples are distributed as a circle.
-But this story may be too convoluted to add here?}
 
@@
 Consequently, it is not surprising to find that ECCCo never achieves state-of-the-art plausibility, although it does improve outcomes compared to the baseline (Wachter).
-With respect to faithfulness, ECCCo once again consistently outperforms all other counterfactual generators.
-
-\import{contents/}{table-real-world.tex}
+Concerning faithfulness, ECCCo once again consistently outperforms all other generators.
 
 To conclude this section, we summarize our findings with reference to the opening questions.
-Concerning the feasibility of our proposed methodology (Research Question~\ref{rq:feasibility}), our findings clearly demonstrate that it is indeed possible to generate plausible counterfactuals without the need for surrogate models.
+Concerning the feasibility of our proposed methodology (Research Question~\ref{rq:feasibility}), our findings demonstrate that it is indeed possible to generate plausible counterfactuals without the need for surrogate models.
 A related important finding is that ECCCo never sacrifices faithfulness for plausibility: any plausible ECCCo also faithfully describes model behaviour.
@@
 We think that further work is needed, however, to definitively answer Research Question~\ref{rq:drivers}, on which we elaborate in the following section.
+
+\import{contents/}{table-real-world.tex}
 

### 2023-05-14 notebooks/linearly_separable.qmd
CP loss suddenly sorted out looool

@@
 builder = MLJFlux.MLP(
-    hidden=(n_hidden, n_hidden, n_hidden,), 
-    σ=activation
+    hidden=(n_hidden, n_hidden, n_hidden), 
+    σ=Flux.swish
 )
@@
 # Train models:
-function _train(model, X=X, y=labels; cov=.95, method=:simple_inductive, mod_name="model")
+function _train(model, X=X, y=labels; cov=0.95, method=:simple_inductive, mod_name="model")
     conf_model = conformal_model(model; method=method, coverage=cov)
@@
 for (mod_name, model) in model_dict
+    p0 = Plots.contourf(model.model, model.fitresult, X_plot, labels; plot_set_size=true, zoom=0, temp=temp)
     p1 = Plots.contourf(model.model, model.fitresult, X_plot, labels; plot_set_loss=true, zoom=0, temp=temp)
     p2 = Plots.contourf(model.model, model.fitresult, X_plot, labels; plot_classification_loss=true, zoom=0, temp=temp, clim=nothing, loss_matrix=ones(2,2))
-    display(Plots.plot(p1, p2, size=(800,320), plot_title=mod_name))
+    display(Plots.plot(p0, p1, p2, size=(1400,320), plot_title=mod_name, layout=(1,3)))
 end
@@
 λ₁ = 0.1
-λ₂ = 0.25
-λ₃ = 0.25
+λ₂ = 0.4
+λ₃ = 1.0
 Λ = [λ₁, λ₂, λ₃]
 
-opt = Flux.Optimise.Descent(0.05)
+opt = Flux.Optimise.Descent(0.01)
 
@@
 
-M = model_dict["MLP"]
+M = model_dict["JEM"]
 X = X isa Matrix ? X : Float32.(permutedims(matrix(X)))
@@
     )
-    plt = Plots.plot(ce, title=name, alpha=0.2, cbar=false, axis=nothing)
+    plt = Plots.plot(
+        ce, title=name, alpha=0.2, 
+        cbar=false, 
+        # axis=nothing,
+    )
     if contains(name, "ECCCo")

### 2023-05-14 notebooks/mnist.qmd
CP loss suddenly sorted out looool

@@
 # Hyper:
-_retrain = true
-_regen = true
+_retrain = false
+_regen = false
 
@@
     x::Union{AbstractArray, Int}=x_factual, target::Int=target;
-    λ=[0.1,0.25,0.25],
+    λ=[0.1,0.4,0.4],
     temp=0.1,
@@
     T::Int = 100,
+    use_class_loss::Bool = true,
 )
@@
         opt=opt,
+        use_class_loss=use_class_loss,
     )
@@
     plts = []
-    for (_name,ce) in ces
+    letters = collect('a':'z')[1:length(ces)]
+    _count = 1
+
+    for (_name,ce) in collect(ces)[_plt_order]
         _x = CounterfactualExplanations.counterfactual(ce)
         _phat = target_probs(ce)
-        _title = "$_name (p̂=$(round(_phat[1]; digits=3)))"
+        _title = "($(letters[_count]))"
         plt = Plots.plot(
             convert2image(MNIST, reshape(_x,28,28)),
-            axis=nothing, 
+            axis=([], false), 
             size=(img_height, img_height),
@@
         plts = [plts..., plt]
+        _count += 1
     end
-    plts = plts[_plt_order]
-    plts = [p1, plts...]
-    plt = Plots.plot(plts...; size=(img_height*length(plts),img_height), layout=(1,length(plts)))
+    # plts = plts[_plt_order]
+    # plts = [p1, plts...]
+    plt = Plots.plot(plts...; size=(img_height,img_height))
 
-    return plt, eccco_generator
+    return plt, eccco_generator, ces
 end
@@
 ```{julia}
-plt, eccco_generator = _plot_eccco_mnist()
+plt, eccco_generator, ces = _plot_eccco_mnist()
 display(plt)

### 2023-05-14 notebooks/moons.qmd
CP loss suddenly sorted out looool

@@
 n_ens = 5                                   # number of models in ensemble
-_loss = Flux.Losses.logitcrossentropy       # loss function
-_finaliser = x -> x                         # finaliser function
+_loss = Flux.Losses.crossentropy       # loss function
+_finaliser = Flux.softmax                      # finaliser function
 ```
@@
 
+```{julia}
+#| output: true
+#| echo: false
+#| label: fig-losses
+#| fig-cap: "Illustration of the smooth size loss and the configurable classification loss."
+
+X_plot = matrix(X)
+temp = 0.1
+
+for (mod_name, model) in model_dict
+    p0 = Plots.contourf(model.model, model.fitresult, X_plot, labels; plot_set_size=true, zoom=0, temp=temp)
+    p1 = Plots.contourf(model.model, model.fitresult, X_plot, labels; plot_set_loss=true, zoom=0, temp=temp)
+    p2 = Plots.contourf(model.model, model.fitresult, X_plot, labels; plot_classification_loss=true, zoom=0, temp=temp, clim=nothing, loss_matrix=ones(2,2))
+    display(Plots.plot(p0, p1, p2, size=(1400,320), plot_title=mod_name, layout=(1,3)))
+end
+```
+
 ## Benchmark
@@
 λ₁ = 0.1
-λ₂ = 0.25
-λ₃ = 0.25
+λ₂ = 0.5
+λ₃ = 0.5
 Λ = [λ₁, λ₂, λ₃]
+
+opt = Flux.Optimise.Descent(0.01)
+use_class_loss = true
 
@@
 generator_dict = Dict(
-    "Wachter" => WachterGenerator(λ=λ₁),
-    "REVISE" => REVISEGenerator(λ=λ₁),
+    "Wachter" => WachterGenerator(λ=λ₁, opt=opt),
+    "REVISE" => REVISEGenerator(λ=λ₁, opt=opt),
     "Schut" => GreedyGenerator(),
-    "ECCCo" => ECCCoGenerator(λ=Λ),
-    "ECCCo (no CP)" => ECCCoGenerator(λ=[λ₁, 0.0, λ₃]),
-    "ECCCo (no EBM)" => ECCCoGenerator(λ=[λ₁, λ₂, 0.0]),
+    "ECCCo" => ECCCoGenerator(λ=Λ, opt=opt, use_class_loss=use_class_loss),
+    "ECCCo (no CP)" => ECCCoGenerator(λ=[λ₁, 0.0, λ₃], opt=opt, use_class_loss=use_class_loss),
+    "ECCCo (no EBM)" => ECCCoGenerator(λ=[λ₁, λ₂, 0.0], opt=opt, use_class_loss=use_class_loss),
 )

### 2023-05-14 notebooks/poc.qmd
CP loss suddenly sorted out looool

@@
     batch_size=batch_size,
-    finaliser=x -> x,
-    loss=Flux.Losses.logitcrossentropy,
+    finaliser=Flux.softmax,
+    loss=Flux.Losses.crossentropy,
     jem_training_params=(
@@
 x1, x2 = meshgrid(xrange, yrange)
-inputs = zip(x, y)
+inputs = zip(x1, x2)
 

### 2023-05-14 paper/paper.tex
CP loss suddenly sorted out looool

@@
 
+\section{Methodological Framework}\label{meth}
+
+The primary objective of this work has been to develop a methodology for generating maximally plausible counterfactuals under minimal intervention.
+Our proposed framework is based on the premise that explanations should be plausible but not plausible at all costs.
+Energy-Constrained Conformal Counterfactuals (ECCCo) achieve this goal in two ways: firstly, they rely on the Black Box itself for the generative task; and, secondly, they involve an approach to predictive uncertainty quantification that is model-agnostic.
+In this section, we first formalise our notion of faithfulness in the context of CE.
+We then explain how exactly ECCCos are generated.
+
 \subsection{From Fidelity to Model Faithfulness}
 
-Above we explained that since Counterfactual Explanations work directly with the Black Box model, the fidelity of explanations as we defined it earlier is not a concern.
+Above we explained that since Counterfactual Explanations work directly with the black-box model, the fidelity of explanations as we defined it earlier is not a concern.
 This may explain why research has primarily focused on other desiderata, most notably plausibility (Definition~\ref{def:plausible}).
@@
 This is very much at the core of our proposed methodological framework, which we will introduce next.
-
-\section{Methodological Framework}\label{meth}
-
-The primary objective of this work has been to develop a methodology for generating maximally plausible counterfactuals under minimal intervention.
-Our proposed framework is based on the premise that explanations should be plausible but not plausible at all costs.
-Energy-Constrained Conformal Counterfactuals (ECCCo) achieve this goal in two ways: firstly, they rely on the Black Box itself for the generative task; and, secondly, they involve an approach to predictive uncertainty quantification that is model-agnostic.
 
@@
 The hyperparameters $\lambda_1, ..., \lambda_3$ can be used to balance the different objectives: for example, we may choose to incur larger deviations from the factual in favour of faithfulness with the model's generative property by choosing lower values of $\lambda_1$ and relatively higher values of $\lambda_2$.
-Figure~\ref{fig:eccco} illustrates this balancing act for an example involving synthetic data: vector fields indicate the direction of gradients with respect to the different components our proposed objective function (Equation~\ref{eq:eccco}).
 
@@
   \centering
-  \includegraphics[width=0.8\textwidth]{../artifacts/results/images/poc.png}
-  \captionof{figure}{[PLACEHOLDER; may swap for:] Vector fields indicating the direction of gradients with respect to the different components of the ECCCo objective (Equation~\ref{eq:eccco}).} \label{fig:eccco}
+  \includegraphics[width=0.8\textwidth]{../artifacts/results/images/mnist_eccco.png}
+  \captionof{figure}{ECCCos for turning a 9 (nine) into a 7 (seven) for different black-box models from left to right: Multi-Layer Perceptron (a), Ensemble of MLPs (b), Joint Energy Model (c), Ensemble of JEMs (d).}\label{fig:eccco-mnist}
 \end{minipage}  
@@
 
-\begin{minipage}[c]{\textwidth}
-  \centering
-  \includegraphics[width=0.8\textwidth]{../artifacts/results/images/mnist_eccco.png}
-  \captionof{figure}{Original image (left) and ECCCos for turning a 9 (nine) into a 7 (seven) for different black-box models from left to right: Multi-Layer Perceptron (MLP), Ensemble of MLPs, Joint Energy Model (JEM), Ensemble of JEMs.}\label{fig:eccco-mnist}
-\end{minipage}
+% \begin{minipage}[c]{\textwidth}
+%   \centering
+%   \includegraphics[width=0.8\textwidth]{../artifacts/results/images/mnist_eccco.png}
+%   \captionof{figure}{Original image (left) and ECCCos for turning a 9 (nine) into a 7 (seven) for different black-box models from left to right: Multi-Layer Perceptron (MLP), Ensemble of MLPs, Joint Energy Model (JEM), Ensemble of JEMs.}\label{fig:eccco-mnist}
+% \end{minipage}
 

### 2023-05-14 src/losses.jl

@@ -1,0 +1,30 @@
+using CounterfactualExplanations
+using Statistics: mean
+
+@doc raw"""
+    conformal_training_loss(ce::AbstractCounterfactualExplanation; temp::Real=0.1, agg=mean, kwargs...)
+
+A configurable classification loss function for Conformal Predictors.
+"""
+function conformal_training_loss(ce::AbstractCounterfactualExplanation; temp::Real=0.1, agg=mean, kwargs...)
+    conf_model = ce.M.model
+    fitresult = ce.M.fitresult
+    X = CounterfactualExplanations.decode_state(ce)
+    y = ce.target_encoded[:, :, 1]
+    if ce.M.likelihood == :classification_binary
+        y = binary_to_onehot(y)
+    end
+    y = permutedims(y)
+
+    n_classes = length(ce.data.y_levels)
+    loss_mat = ones(n_classes, n_classes)
+    loss = map(eachslice(X, dims=ndims(X))) do x
+        x = ndims(x) == 1 ? x[:,:]' : x
+        ConformalPrediction.classification_loss(
+            conf_model, fitresult, x, y;
+            temp=temp, loss_matrix = loss_mat,
+        )
+    end
+    loss = agg(loss)[1]
+    return loss
+end

### 2023-05-14 src/model.jl

@@ -101,11 +101,6 @@ function ConformalModel(model, fitresult=nothing; likelihood::Union{Nothing,Symb
     # Check if model is fitted and infer likelihood:
     if isnothing(fitresult)
         @info "Conformal Model is not fitted."
-    # else
-    #     outdim = _outdim(fitresult)
-    #     _likelihood = outdim == 2 ? :classification_binary : :classification_multi
-    #     @assert likelihood == _likelihood || isnothing(likelihood) "Specification of `likelihood` does not match the output dimension of the model."
-    #     likelihood = _likelihood
     end
 
     # Default to binary classification, if not specified or inferred:
@@ -119,6 +114,13 @@ function ConformalModel(model, fitresult=nothing; likelihood::Union{Nothing,Symb
     M = ConformalModel(model, fitresult, likelihood)
     return M
 end
+
+"""
+    get_logits(f::Flux.Chain, x)
+
+Helper function to return logits in case final layer is an activation function.
+"""
+get_logits(f::Flux.Chain, x) = f[end] isa Function ? f[1:end-1](x) : f(x)
 
 # Methods
 @doc raw"""
@@ -145,18 +147,13 @@ which follows from the derivation here: https://stats.stackexchange.com/question
 function Models.logits(M::ConformalModel, X::AbstractArray)
     fitresult = M.fitresult
     function predict_logits(fitresult, x)
-        p̂ = MLUtils.stack(map(chain -> chain(x),_get_chains(fitresult))) |> 
-            p -> mean(p, dims=ndims(p)) |>
-            p -> MLUtils.unstack(p, dims=ndims(p))[1]
-        if ndims(p̂) == 2
-            p̂ = [p̂]
+        ŷ = MLUtils.stack(map(chain -> get_logits(chain,x),_get_chains(fitresult))) |> 
+            y -> mean(y, dims=ndims(y)) |>
+            y -> MLUtils.unstack(y, dims=ndims(y))[1]
+        if ndims(ŷ) == 2
+            ŷ = [ŷ]
         end
-        p̂ = reduce(hcat, p̂)
-        if all(0.0 .<= vec(p̂) .<= 1.0)
-            ŷ = reduce(hcat, (map(p -> log.(p) .+ log(sum(exp.(p))), eachcol(p̂))))
-        else
-            ŷ = p̂
-        end
+        ŷ = reduce(hcat, ŷ)
         if M.likelihood == :classification_binary
[…]

### 2023-05-14 src/penalties.jl

@@ -11,7 +11,7 @@ Penalty for smooth conformal set size.
 """
 function set_size_penalty(
     ce::AbstractCounterfactualExplanation; 
-    κ::Real=0.0, temp::Real=0.05, agg=mean
+    κ::Real=1.0, temp::Real=0.1, agg=mean
 )
 
     _loss = 0.0

### 2023-05-14 src/sampling.jl

@@ -88,7 +88,7 @@ Generates `n` samples from `EnergySampler` for conditioning value `y`.
 function generate_samples(e::EnergySampler, n::Int, y::Int; niter::Int=100)
 
     # Generate samples:
-    # chain = e.model.fitresult[1]
+    # f = e.model.fitresult[1]
     f(x) = logits(e.model, x)
     rule = e.opt
     xsamples = e.sampler(f, rule; niter=niter, n_samples=n, y=y)

### 2023-05-14 notebooks/circles.qmd
rerun

@@
 ```{julia}
-λ₁ = 0.1
-λ₂ = 0.25
-λ₃ = 0.25
+λ₁ = 0.25
+λ₂ = 0.75
+λ₃ = 0.75
 Λ = [λ₁, λ₂, λ₃]
+
+opt = Flux.Optimise.Descent(0.01)
+use_class_loss = false
 

### 2023-05-14 notebooks/gmsc.qmd
rerun

@@
 test_size = 0.2
-n_obs = Int(1000 / (1.0 - test_size))
+n_obs = Int(10000 / (1.0 - test_size))
 counterfactual_data, test_data = train_test_split(load_gmsc(n_obs); test_size=test_size)
@@
 n_ens = 5                                   # number of models in ensemble
-_loss = Flux.Losses.logitcrossentropy       # loss function
-_finaliser = x -> x                         # finaliser function
+_loss = Flux.Losses.crossentropy       # loss function
+_finaliser =  Flux.softmax                       # finaliser function
 ```
@@
 ```{julia}
-λ₁ = 0.25
-λ₂ = 0.75
-λ₃ = 0.75
+λ₁ = 0.1
+λ₂ = 0.5
+λ₃ = 0.5
 Λ = [λ₁, λ₂, λ₃]
+
+opt = Flux.Optimise.Descent(0.01)
+use_class_loss = true
 
@@
 generator_dict = Dict(
-    "Wachter" => WachterGenerator(),
-    "REVISE" => REVISEGenerator(),
+    "Wachter" => WachterGenerator(λ=λ₁, opt=opt),
+    "REVISE" => REVISEGenerator(λ=λ₁, opt=opt),
     "Schut" => GreedyGenerator(),
-    "ECCCo" => ECCCoGenerator(λ=Λ),
+    "ECCCo" => ECCCoGenerator(λ=Λ, opt=opt, use_class_loss=use_class_loss),
 )

### 2023-05-14 notebooks/linearly_separable.qmd
rerun

@@
 ```{julia}
-λ₁ = 0.1
-λ₂ = 0.4
-λ₃ = 1.0
+λ₁ = 0.25
+λ₂ = 0.75
+λ₃ = 0.75
 Λ = [λ₁, λ₂, λ₃]
@@
 opt = Flux.Optimise.Descent(0.01)
+use_class_loss = false
 
@@
     "Schut" => GreedyGenerator(),
-    "ECCCo" => ECCCoGenerator(λ=Λ, opt=opt),
-    "ECCCo (no CP)" => ECCCoGenerator(λ=[λ₁, 0.0, λ₃], opt=opt),
-    "ECCCo (no EBM)" => ECCCoGenerator(λ=[λ₁, λ₂, 0.0], opt=opt),
+    "ECCCo" => ECCCoGenerator(λ=Λ, opt=opt, use_class_loss=use_class_loss),
+    "ECCCo (no CP)" => ECCCoGenerator(λ=[λ₁, 0.0, λ₃], opt=opt, use_class_loss=use_class_loss),
+    "ECCCo (no EBM)" => ECCCoGenerator(λ=[λ₁, λ₂, 0.0], opt=opt, use_class_loss=use_class_loss),
 )
@@
 X = X isa Matrix ? X : Float32.(permutedims(matrix(X)))
-factual_label =  levels(labels)[2]
+factual_label =  levels(labels)[1]
 x_factual = reshape(X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
-target =  levels(labels)[1]
+target =  levels(labels)[2]
 factual = predict_label(M, counterfactual_data, x_factual)[1]

### 2023-05-14 notebooks/mnist.qmd
rerun

@@
 # Hyper:
-_retrain = false
-_regen = false
+_retrain = true
+_regen = true
 
 # Data:
-n_obs = 1000
+n_obs = 10000
 counterfactual_data = load_mnist(n_obs)

### 2023-05-15 notebooks/gmsc.qmd
uh
rerun

@@
 test_size = 0.2
-n_obs = Int(10000 / (1.0 - test_size))
+n_obs = Int(50000 / (1.0 - test_size))
 counterfactual_data, test_data = train_test_split(load_gmsc(n_obs); test_size=test_size)
@@
 epochs = 100
-batch_size = minimum([Int(round(n_obs/10)), 128])
+batch_size = minimum([Int(round(n_obs/10)), 250])
 n_hidden = 128
-activation = Flux.relu
+activation = Flux.swish
 builder = MLJFlux.@builder Flux.Chain(
@@
     input_size=(input_dim,), 
-    batch_size=50,
+    batch_size=10,
 )
@@
 λ₁ = 0.1
-λ₂ = 0.5
-λ₃ = 0.5
+λ₂ = 0.1
+λ₃ = 1.0
 Λ = [λ₁, λ₂, λ₃]
 
-opt = Flux.Optimise.Descent(0.01)
-use_class_loss = true
+opt = Flux.Optimise.Descent(0.05)
+use_class_loss = false
 

### 2023-05-15 notebooks/mnist.qmd
rerun

@@
     x::Union{AbstractArray, Int}=x_factual, target::Int=target;
-    λ=[0.1,0.4,0.4],
+    λ=[0.25,0.75,0.75],
     temp=0.1,
@@
 ```{julia}
-λ₁ = 0.1
-λ₂ = 0.25
-λ₃ = 0.25
-Λ = [λ₁, λ₂, λ₃]
+Λ = eccco_generator.λ
 
@@
 generator_dict = Dict(
-    "Wachter" => WachterGenerator(λ=λ₁),
-    "REVISE" => REVISEGenerator(λ=λ₁),
+    "Wachter" => WachterGenerator(λ=Λ[1]),
+    "REVISE" => REVISEGenerator(λ=Λ[1]),
     "Schut" => GreedyGenerator(),
-    "ECCCo" => ECCCoGenerator(λ=Λ),
+    "ECCCo" => eccco_generator,
 )

### 2023-05-15 paper/paper.tex
incorporating comments

@@
 But even full fidelity is not a sufficient condition for ensuring that an explanation faithfully describes the behaviour of a model.
-That is because multiple very distinct explanations can all lead to the same model prediction, especially when dealing with heavily parameterized models like deep neural networks which are typically underspecified by the data~\citep{wilson2020case}.
+That is because multiple very distinct explanations can all lead to the same model prediction, especially when dealing with heavily parameterized models like deep neural networks, which are typically underspecified by the data~\citep{wilson2020case}.
 
@@
 A key focus in the literature has therefore been to identify those explanations and algorithmic recourses that are most appropriate based on a myriad of desiderata such as sparsity, actionability and plausibility.
-In this work, we draw closer attention to model faithfulness as a desideratum for counterfactuals.
-Our key contributions are as follows: firstly, we propose a definition of faithfulness that is suitable for counterfactuals; secondly, we introduce a novel algorithmic approach for generating Energy-Constrained Conformal Counterfactuals (ECCCos) that explicitly addresses the need for faithfulness; finally, we provide extensive empirical evidence demonstrating that ECCCos faithfully explain model behaviour without sacrificing plausibility.
+In this work, we draw closer attention to model faithfulness rather than fidelity as a desideratum for counterfactuals.
+Our key contributions are as follows:
 
-\section{Background and Related Work}\label{background}
+\begin{itemize}
+  \item We demonstrate that a narrow focus on plausibility can yield counterfactuals that are unfaithful to model behaviour despite having full fidelity (Section~\ref{fidelity}).
+  \item We propose a definition of faithfulness that gives rise to a more suitable evaluation metric for counterfactuals than fidelity (Section~\ref{faithfulness}).
+  \item We introduce a novel algorithmic approach for generating Energy-Constrained Conformal Counterfactuals (ECCCos) in Section~\ref{meth}.
+  \item We provide extensive empirical evidence demonstrating that ECCCos faithfully explain model behaviour without sacrificing plausibility (Section~\ref{emp}).
+\end{itemize}
 
-In this section, we provide some background on CE and our motivation for this work. To start, we briefly introduce the methodology underlying most state-of-the-art (SOTA) counterfactual generators.
+Thus, we believe that our work opens avenues for researchers and practitioners seeking tools to better distinguish trustworthy from unreliable models.
 
-\subsection{Gradient-Based Counterfactual Search}\label{gradient}
+\section{Background}\label{background}
 
@@
 The baseline approach, which we will simply refer to as \textbf{Wachter}~\citep{wachter2017counterfactual}, searches a single counterfactual directly in the feature space and penalises its distance to the original factual.
+\section{Why Fidelity is not Enough}\label{fidelity}
 
+Counterfactual Explanations have full fidelity by construction as long as they are valid.
 Solutions to Equation~\ref{eq:general} are considered valid as soon as the predicted label matches the target label.
@@
 For the generated counterfactual $\mathbf{x}^{\prime}$ the model predicts the target label with high confidence (centre panel in Figure~\ref{fig:adv}).
-The explanation is valid by definition, even though it looks a lot like an Adversarial Example~\citep{goodfellow2014explaining}.
+The explanation is valid by definition and therefore has full fidelity, even though it looks a lot like an Adversarial Example~\citep{goodfellow2014explaining}.
 \citet{schut2021generating} make the connection between Adversarial Examples and Counterfactual Explanations explicit and propose using a Jacobian-Based Saliency Map Attack (JSMA) to solve Equation~\ref{eq:general}.
-They demonstrate that this approach yields realistic and sparse counterfactuals for Bayesian, adversarially robust classifiers.
+They demonstrate that this approach yields plausible counterfactuals for Bayesian, adversarially robust classifiers.
 Applying their approach to our simple MNIST classifier does not yield a realistic counterfactual but this one, too, is valid (right panel in Figure~\ref{fig:adv}).
-
-\subsection{From Adversial Examples to Plausible Explanations}
-
-The crucial difference between Adversarial Examples (AE) and Counterfactual Explanations is one of intent.
-While an AE is intended to go unnoticed, a CE should have certain desirable properties.
-The literature has made this explicit by introducing various so-called \textit{desiderata} that counterfactuals should meet in order to properly serve both AI practitioners and individuals affected by AI decision-making systems.
-The list of desiderate includes but is not limited to the following: sparsity, proximity~\citep{wachter2017counterfactual}, actionability~\citep{ustun2019actionable}, diversity~\citep{mothilal2020explaining}, plausibility~\citep{joshi2019realistic,poyiadzi2020face,schut2021generating}, robustness~\citep{upadhyay2021robust,pawelczyk2022probabilistically,altmeyer2023endogenous} and causality~\citep{karimi2021algorithmic}.
-
-Researchers have come up with various ways to meet these desiderata, which have been extensively surveyed and evaluated in various studies~\citep{verma2020counterfactual,karimi2020survey,pawelczyk2021carla,artelt2021evaluating,guidotti2022counterfactual}.
-Perhaps unsurprisingly, the different desiderata are often positively correlated.
-For example, \citet{artelt2021evaluating} find that plausibility typically also leads to improved robustness.
-Similarly, plausibility has also been connected to causality in the sense that plausible counterfactuals respect causal relationships~\citep{mahajan2020preserving}.
-
-\subsubsection{Plausibility through Surrogates}
-
-Arguably, the plausibility of counterfactuals has been among the primary concerns and some have focused explicitly on this goal.
-\citet{joshi2019realistic}, for example, were among the first to suggest that instead of searching counterfactuals in the feature space $\mathcal{X}$, we can instead traverse a latent embedding $\mathcal{Z}$ (Equation~\ref{eq:general}) that implicitly codifies the data generating process (DGP) of $\mathbf{x}\sim\mathcal{X}$.
-To learn the latent embedding, they introduce a surrogate model.
-In particular, they propose to use the latent embedding of a Variational Autoencoder (VAE) trained to generate samples $\mathbf{x}^* \leftarrow \mathcal{G}(\mathbf{z})$ where $\mathcal{G}$ denotes the decoder part of the VAE.
-Provided the surrogate model is well-trained, their proposed approach ---\textbf{REVISE}--- can yield plausible explanations like the one in the centre panel of Figure~\ref{fig:vae}.
-
-Others have proposed similar approaches.
-\citet{dombrowski2021diffeomorphic} traverse the base space of a normalizing flow to solve Equation~\ref{eq:general}, essentially relying on a different surrogate model for the generative task.
-\citet{poyiadzi2020face} use density estimators ($\hat{p}: \mathcal{X} \mapsto [0,1]$) to constrain the counterfactuals to dense regions in the feature space.
-\citet{karimi2021algorithmic} argue that counterfactuals should comply with the causal model that generates the data.
-All of these different approaches share a common goal: ensuring that the generated counterfactuals comply with the true and unobserved DGP.
-To summarize this broad objective, we propose the following definition:
-
-\begin{definition}[Plausible Counterfactuals]
-  \label{def:plausible}
-  Let $\mathcal{X}|\mathbf{y}^+$ denote the true conditional distribution of samples in the target class $\mathbf{y}^+$.
-Then for $\mathbf{x}^{\prime}$ to be considered a plausible counterfactual, we need: $\mathbf{x}^{\prime} \sim \mathcal{X}|\mathbf{y}^+$.
-\end{definition}
-
-Surrogate models offer an obvious solution to achieve this objective.
-Unfortunately, surrogates also introduce a dependency: the generated explanations no longer depend exclusively on the black-box model itself, but also on the surrogate model.
-This is not necessarily problematic if the primary objective is not to explain the behaviour of the model but to offer recourse to individuals affected by it.
-It may become problematic even in this context if the dependency turns into a vulnerability.
-To illustrate this point, we have used REVISE~\citep{joshi2019realistic} with an underfitted VAE to generate the counterfactual in the right panel of Figure~\ref{fig:vae}: in this case, the decoder step of the VAE fails to yield plausible values ($\{\mathbf{x}^{\prime} \leftarrow \mathcal{G}(\mathbf{z})\} \not\sim \mathcal{X}|\mathbf{y}^+$) and hence the counterfactual search in the learned latent space is doomed.
 
@@
 
-\subsubsection{Plausibility through Minimal Predictive Uncertainty}
+The crucial difference between Adversarial Examples (AE) and Counterfactual Explanations is one of intent.
+While an AE is intended to go unnoticed, a CE should have certain desirable properties.
+The literature has made this explicit by introducing various so-called \textit{desiderata} that counterfactuals should meet in order to properly serve both AI practitioners and individuals affected by AI decision-making systems.
+The list of desiderate includes but is not limited to the following: sparsity, proximity~\citep{wachter2017counterfactual}, actionability~\citep{ustun2019actionable}, diversity~\citep{mothilal2020explaining}, plausibility~\citep{joshi2019realistic,poyiadzi2020face,schut2021generating}, robustness~\citep{upadhyay2021robust,pawelczyk2022probabilistically,altmeyer2023endogenous} and causality~\citep{karimi2021algorithmic}.
 
-\citet{schut2021generating} show that to meet the plausibility objective we need not explicitly model the input distribution.
-Pointing to the undesirable engineering overhead induced by surrogate models, they propose that we rely on the implicit minimisation of predictive uncertainty instead.
-Their proposed methodology solves Equation~\ref{eq:general} by greedily applying JSMA in the feature space with standard cross-entropy loss and no penalty at all.
-They demonstrate theoretically and empirically that their approach yields counterfactuals for which the model $M_{\theta}$ predicts the target label $\mathbf{y}^+$ with high confidence.
-Provided the model is well-specified, these counterfactuals are plausible.
-This idea hinges on the assumption that the black-box model provides well-calibrated predictive uncertainty estimates.
+Researchers have come up with various ways to meet these desiderata, which have been extensively surveyed and evaluated in various studies~\citep{verma2020counterfactual,karimi2020survey,pawelczyk2021carla,artelt2021evaluating,guidotti2022counterfactual}.
+Perhaps unsurprisingly, the different desiderata are often positively correlated.
+For example, \citet{artelt2021evaluating} find that plausibility typically also leads to improved robustness.
+Similarly, plausibility has also been connected to causality in the sense that plausible counterfactuals respect causal relationships~\citep{mahajan2020preserving}.
+Consequently, the plausibility of counterfactuals has been among the primary concerns for researchers.
+
+Achieving plausibility is equivalent to ensuring that the generated counterfactuals comply with the true and unobserved data-generating process (DGP). Formally, this can be defined as follows:
+
+\begin{definition}[Plausible Counterfactuals]
+  \label{def:plausible}
+  Let $\mathcal{X}|\mathbf{y}^+$ denote the true conditional distribution of samples in the target class $\mathbf{y}^+$.
+Then for $\mathbf{x}^{\prime}$ to be considered a plausible counterfactual, we need: $\mathbf{x}^{\prime} \sim \mathcal{X}|\mathbf{y}^+$.
+\end{definition}
+
+To generate plausible counterfactuals, we need to be able to quantify the DGP: $\mathcal{X}|\mathbf{y}^+$.
+One straightforward way to do this is to use surrogate models for the task.
+\citet{joshi2019realistic}, for example, suggest that instead of searching counterfactuals in the feature space $\mathcal{X}$, we can instead traverse a latent embedding $\mathcal{Z}$ (Equation~\ref{eq:general}) that implicitly codifies the DGP.
+To learn the latent embedding, they propose using a generative model such as a Variational Autoencoder (VAE).
+Provided the surrogate model is well-trained, their proposed approach ---\textbf{REVISE}--- can yield plausible explanations like the one in the centre panel of Figure~\ref{fig:vae}.
+Others have proposed similar approaches: \citet{dombrowski2021diffeomorphic} traverse the base space of a normalizing flow to solve Equation~\ref{eq:general}; \citet{poyiadzi2020face} use density estimators ($\hat{p}: \mathcal{X} \mapsto [0,1]$) to constrain the counterfactuals to dense regions in the feature space; and, finally, \citet{karimi2021algorithmic} assume knowledge about the structural causal model that generates the data.
+
+While surrogates offer an easy way to address the need for plausibility, they also introduce friction: the generated explanations no longer depend exclusively on the black-box model itself.
+The counterfactual in the centre panel of Figure~\ref{fig:vae}, for example, looks more plausible than the corresponding counterfactuals generated by \textbf{Wachter} and \textbf{Schut}.
+But is it also more faithful to the behaviour of our MNIST classifier?
+Fidelity cannot help us to make that judgement, because all of these counterfactuals have full fidelity.
+To bridge this gap, we introduce a new notion of faithfulness in the following section.
+
+\section{A new Notion of Faithfulness}\label{faithfulness}
+
+Analogous to Definition~\ref{def:plausible}, we propose to define faithfulness in the context of Counterfactual Explanations as follows:
+
+\begin{definition}[Faithful Counterfactuals]
+  \label{def:faithful}
+  Let $\mathcal{X}_{\theta}|\mathbf{y}^+ = p_{\theta}(\mathbf{X}_{\mathbf{y}^+})$ denote the conditional distribution of $\mathbf{x}$ in the target class $\mathbf{y}^+$, where $\theta$ denotes the parameters of model $M_{\theta}$.
+Then for $\mathbf{x}^{\prime}$ to be considered a conformal counterfactual, we need: $\mathbf{x}^{\prime} \sim \mathcal{X}_{\theta}|\mathbf{y}^+$.
+\end{definition}
+
+In doing this, we merge in and nuance the concept of plausibility (Definition~\ref{def:plausible}) where the notion of `consistent with the data' becomes `consistent with what the model has learned about the data'.
+
+\subsection{Quantifying the Model's Generative Property}
+
+To assess counterfactuals with respect to Definition~\ref{def:faithful}, we need a way to quantify the posterior conditional distribution $p_{\theta}(\mathbf{x}|\mathbf{y}^+)$.
+To this end, we draw on recent advances in Energy-Based Modelling (EBM), a subdomain of machine learning that is concerned with generative or hybrid modelling~\citep{grathwohl2020your,}.
+In particular, note that if we fix $\mathbf{y}$ to our target value $\mathbf{y}^+$, we can conditionally draw from $p_{\theta}(\mathbf{x}|\mathbf{y}^+)$ using Stochastic Gradient Langevin Dynamics (SGLD) as follows,
+
+\begin{equation}\label{eq:sgld}
+  \begin{aligned}
+    \mathbf{x}_{j+1} &\leftarrow \mathbf{x}_j - \frac{\epsilon^2}{2} \mathcal{E}(\mathbf{x}_j|\mathbf{y}^+) + \epsilon \mathbf{r}_j, && j=1,...,J
+  \end{aligned}
+\end{equation}
+
+where $\mathbf{r}_j \sim \mathcal{N}(\mathbf{0},\mathbf{I})$ is the stochastic term and the step-size $\epsilon$ is typically polynomially decayed~\citep{welling2011bayesian}.
+The term $\mathcal{E}(\mathbf{x}_j|\mathbf{y}^+)$ denotes the model energy conditioned on the target class label $\mathbf{y}^+$.
+To allow for faster sampling, we follow the common practice of choosing the step-size $\epsilon$ and the standard deviation of $\mathbf{r}_j$ separately.
+While $\mathbf{x}_J$ is only guaranteed to distribute as $p_{\theta}(\mathbf{x}|\mathbf{y}^*)$ if $\epsilon \rightarrow 0$ and $J \rightarrow \infty$, the bias introduced for a small finite $\epsilon$ is negligible in practice \citep{murphy2023probabilistic,grathwohl2020your}.
+Appendix~\ref{app-jem} provides additional implementation details for any tasks related to energy-based modelling.
+
+Generating multiple samples using SGLD thus yields an empirical distribution $\hat{\mathbf{X}}_{\theta,\mathbf{y}^+}$ that approximates what the model has learned about the input data.
+While in the context of Energy-Based Modelling, this is usually done during training, we propose to use SGLD during inference in order to evaluate and generate faithful model explanations.
+
+\subsection{Evaluating Plausibility and Faithfulness}
+
+The parallels between our definitions of plausibility and faithfulness imply that we can also use similar evaluation metrics in both cases.
+Since existing work has focused heavily on plausibility, it offers a useful starting point.
+In particular,~\citet{guidotti2022counterfactual} have proposed an implausibility metric that measures the distance of the counterfactual from its nearest neighbour in the target class.
+As this distance is reduced, counterfactuals get more plausible under the assumption that the nearest neighbour itself is plausible in the sense of Definition~\ref{def:plausible}.
+In this work, we use the following adapted implausibility metric that relaxes this assumption,
+
+\begin{equation}\label{eq:impl}
+  \begin{aligned}
+    \text{impl} = \frac{1}{\lvert\mathbf{x} \in \mathbf{X}_{\mathbf{y}^+}\rvert} \sum_{\mathbf{x} \in \mathbf{X}_{\mathbf{y}^+}} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
+  \end{aligned}
+\end{equation}
+
+where $\mathbf{X}_{\mathbf{y}^+}$ is a subsample of the training data in the target class $\mathbf{y}^+$.
+
+This gives rise to a very similar evaluation metric for unfaithfulness.
+We merely swap out the subsample of individuals in the target class for a subset $\hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^+}$ of the generated conditional samples:
+
+\begin{equation}\label{eq:faith}
+  \begin{aligned}
+    \text{unfaith} = \frac{1}{\lvert\mathbf{x} \in \hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^+}\rvert} \sum_{\mathbf{x} \in \hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^+}} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
+  \end{aligned}
+\end{equation}
+
+Specifically, we form this subset based on the $n_E$ generated samples with the lowest energy.
 
@@
 
-\subsection{From Fidelity to Model Faithfulness}
+\subsection{Quantifying the Model's Predictive Uncertainty}
 
-Above we explained that since Counterfactual Explanations work directly with the black-box model, the fidelity of explanations as we defined it earlier is not a concern.
-This may explain why research has primarily focused on other desiderata, most notably plausibility (Definition~\ref{def:plausible}).
-Enquiring about the plausibility of a counterfactual essentially boils down to the following question: `Is this counterfactual consistent with the data'?
-We ask a related question: `Is this counterfactual consistent with what the model has learned about the data'?
-To answer this question and propose a novel way to assess if explanations conform with model behaviour.
-
-The word \textit{fidelity} stems from the Latin word `fidelis', which means `faithful, loyal, trustworthy'~\citep{mw2023fidelity}.
-As we explained in Section~\ref{background}, model explanations are generally considered faithful if their corresponding predictions coincide with the predictions made by the model itself.
-Since this definition of faithfulness is not useful in the context of CE, we propose the following definition:
-
-\begin{definition}[Faithful Counterfactuals]
-  \label{def:faithful}
-  Let $\mathcal{X}_{\theta}|\mathbf{y}^+ = p_{\theta}(\mathbf{X}_{\mathbf{y}^+})$ denote the conditional distribution of $\mathbf{x}$ in the target class $\mathbf{y}^+$, where $\theta$ denotes the parameters of model $M_{\theta}$.
-Then for $\mathbf{x}^{\prime}$ to be considered a conformal counterfactual, we need: $\mathbf{x}^{\prime} \sim \mathcal{X}_{\theta}|\mathbf{y}^+$.
-\end{definition}
-
-To assess counterfactuals with respect to Definition~\ref{def:faithful}, we need to be able to quantify the posterior conditional distribution $p_{\theta}(\mathbf{x}|\mathbf{y}^+)$.
-This is very much at the core of our proposed methodological framework, which we will introduce next.
-
-\subsection{Quantifying the Model's Generative Property}
-
-Recent work by~\citet{grathwohl2020your} on Energy-Based Models (EBM) has shown that there is a `generative model hidden within every standard discriminative model'.
-They use Stochastic Gradient Langevin Dynamics (SGLD) to estimate the posterior conditional distribution $p_{\theta}(\mathbf{x}|\mathbf{y})$ using and leverage this to train classifiers jointly for the discriminative and generative task.
-
-Crucially for our purpose, SGLD can be applied during inference to essentially any standard discriminative model.
-Even models that are not explicitly trained for the joint objective learn about the distribution of inputs $X$ by learning to make conditional predictions about the output $y$.
-We can leverage this observation to quantify the generative property of models.
-In particular, note that if we fix $\mathbf{y}$ to our target value $\mathbf{y}^+$, we can sample from $p_{\theta}(\mathbf{x}|\mathbf{y}^+)$ using SGLD as follows,
-
-\begin{equation}\label{eq:sgld}
-  \begin{aligned}
-    \mathbf{x}_{j+1} &\leftarrow \mathbf{x}_j - \frac{\epsilon^2}{2} \mathcal{E}(\mathbf{x}_j|\mathbf{y}^+) + \epsilon \mathbf{r}_j, && j=1,...,J
-  \end{aligned}
-\end{equation}
-
-where $\mathbf{r}_j \sim \mathcal{N}(\mathbf{0},\mathbf{I})$ is the stochastic term and the step-size $\epsilon$ is typically polynomially decayed.
-The term $\mathcal{E}(\mathbf{x}_j|\mathbf{y}^+)$ denotes the model energy conditioned on the target class label $\mathbf{y}^+$.
-Generating multiple samples in this manner yields an empirical distribution $\hat{\mathbf{X}}_{\theta,\mathbf{y}^+}$ that we use in our search for plausible counterfactuals, as discussed in more detail below.
-Appendix~\ref{app-jem} provides additional implementation details for any tasks related to energy-based modelling.
-
-\subsection{Quantifying the Model's Predictive Uncertainty}
+\citet{schut2021generating} show that to meet the plausibility objective we need not explicitly model the input distribution.
+Pointing to the undesirable engineering overhead induced by surrogate models, they propose that we rely on the implicit minimisation of predictive uncertainty instead.
+Their proposed methodology solves Equation~\ref{eq:general} by greedily applying JSMA in the feature space with standard cross-entropy loss and no penalty at all.
+They demonstrate theoretically and empirically that their approach yields counterfactuals for which the model $M_{\theta}$ predicts the target label $\mathbf{y}^+$ with high confidence.
+Provided the model is well-specified, these counterfactuals are plausible.
+This idea hinges on the assumption that the black-box model provides well-calibrated predictive uncertainty estimates.
 
@@
 \subsection{Key Evaluation Metrics}\label{evaluation}
-
-Above we have defined plausibility (Definition~\ref{def:plausible}) and faithfulness (Definition~\ref{def:faithful}) for Counterfactual Explanations.
-These are the main criteria we use to evaluate counterfactuals in this study.
-In order to quantify the plausibility of counterfactuals we use a slightly adapted version of the implausibility metric proposed in~\citet{guidotti2022counterfactual}.
-Formally, we define implausibility as follows,
-
-\begin{equation}\label{eq:impl}
-  \begin{aligned}
-    \text{impl} = \frac{1}{\lvert\mathbf{x} \in \mathbf{X}_{\mathbf{y}^+}\rvert} \sum_{\mathbf{x} \in \mathbf{X}_{\mathbf{y}^+}} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
-  \end{aligned}
-\end{equation}
-
-where $\mathbf{X}_{\mathbf{y}^+}$ is a subsample of the training data in the target class $\mathbf{y}^+$.
-This gives rise to a very similar evaluation metric for unfaithfulness.
-We merely swap out the subsample of individuals in the target class for a subset $\hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^+}$ of the generated conditional samples:
-
-\begin{equation}\label{eq:faith}
-  \begin{aligned}
-    \text{unfaith} = \frac{1}{\lvert\mathbf{x} \in \hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^+}\rvert} \sum_{\mathbf{x} \in \hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^+}} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
-  \end{aligned}
-\end{equation}
-
-Specifically, we form this subset based on the $n_E$ generated samples with the lowest energy.
 

### 2023-05-15 notebooks/mnist.qmd
:yawning_face:

@@
 savefig(plt, joinpath(output_images_path, "surrogate_gone_wrong.png"))
+```
+
+```{julia}
+ces = zip([ce_wachter, ce_jsma, ce_strong])
+counterfactuals = reduce((x,y)->cat(x,y,dims=3),map(ce -> CounterfactualExplanations.counterfactual(ce[1]), ces))
+phat = reduce((x,y) -> cat(x,y,dims=3), map(ce -> target_probs(ce[1]), ces))
+plts = [p1]
+for x in zip(eachslice(counterfactuals; dims=3), eachslice(phat; dims=3), ["Wachter","Schut","REVISE"])
+    ce, _phat, _name = (x[1],x[2],x[3])
+    _title = "$(_name) (p=$(round(_phat[1]; digits=2)))"
+    plt = Plots.plot(
+        convert2image(MNIST, reshape(ce,28,28)),
+        axis=nothing, 
+        size=(img_height, img_height),
+        title=_title
+    )
+    plts = [plts..., plt]
+end
+plt = Plots.plot(plts...; size=(img_height*length(plts),img_height), layout=(1,length(plts)))
+display(plt)
+savefig(plt, joinpath(output_images_path, "mnist_motivation.png"))
 ```

### 2023-05-15 paper/paper.tex
:yawning_face:

@@
 \begin{itemize}
-  \item We demonstrate that a narrow focus on plausibility can yield counterfactuals that are unfaithful to model behaviour despite having full fidelity (Section~\ref{fidelity}).
-  \item We propose a definition of faithfulness that gives rise to a more suitable evaluation metric for counterfactuals than fidelity (Section~\ref{faithfulness}).
+  \item We show that fidelity is an insufficient evaluation metric for counterfactuals (Section~\ref{fidelity}) and propose a definition of faithfulness that gives rise to more suitable metrics (Section~\ref{faithfulness}).
   \item We introduce a novel algorithmic approach for generating Energy-Constrained Conformal Counterfactuals (ECCCos) in Section~\ref{meth}.
@@
 
-Here $\text{yloss}$ denotes the primary loss function already introduced above and $\text{cost}$ is either a single penalty or a collection of penalties that are used to impose constraints through regularization.
-Equation~\ref{eq:general} restates the baseline approach to gradient-based counterfactual search proposed by~\citet{wachter2017counterfactual} in general form where $\mathbf{Z}^\prime=\{ \mathbf{z}_l\}_L$ denotes an $L$-dimensional array of counterfactual states~\citep{altmeyer2023endogenous}.
-This is to explicitly account for the multiplicity of explanations and the fact that we may choose to generate multiple counterfactuals and traverse a latent encoding $\mathcal{Z}$ of the feature space $\mathcal{X}$ where we denote $f^{-1}: \mathcal{X} \mapsto \mathcal{Z}$.
-Encodings may involve simple feature transformations or more advanced techniques involving generative models, as we will discuss further below.
+Here $\text{yloss}$ denotes the primary loss function, $f(\cdot)$ is a function that maps from the counterfactual state space to the feature space and $\text{cost}$ is either a single penalty or a collection of penalties that are used to impose constraints through regularization.
+Equation~\ref{eq:general} restates the baseline approach to gradient-based counterfactual search proposed by~\citet{wachter2017counterfactual} in general form as introduced by~\citet{altmeyer2023endogenous}.
+To explicitly account for the multiplicity of explanations $\mathbf{Z}^\prime=\{ \mathbf{z}_l\}_L$ denotes an $L$-dimensional array of counterfactual states.
+
 The baseline approach, which we will simply refer to as \textbf{Wachter}~\citep{wachter2017counterfactual}, searches a single counterfactual directly in the feature space and penalises its distance to the original factual.
-\section{Why Fidelity is not Enough}\label{fidelity}
+In this case, $f(\cdot)$ is simply the identity function and $\mathcal{Z}$ corresponds to the feature space itself.
+Many derivative works of~\citet{wachter2017counterfactual} have proposed new flavours of Equation~\ref{eq:general}, each of them designed to address specific \textit{desiderata} that counterfactuals ought to meet in order to properly serve both AI practitioners and individuals affected by algorithmic decision-making systems.
+The list of desiderata includes but is not limited to the following: sparsity, proximity~\citep{wachter2017counterfactual}, actionability~\citep{ustun2019actionable}, diversity~\citep{mothilal2020explaining}, plausibility~\citep{joshi2019realistic,poyiadzi2020face,schut2021generating}, robustness~\citep{upadhyay2021robust,pawelczyk2022probabilistically,altmeyer2023endogenous} and causality~\citep{karimi2021algorithmic}.
+Different counterfactual generators addressing these needs have been extensively surveyed and evaluated in various studies~\citep{verma2020counterfactual,karimi2020survey,pawelczyk2021carla,artelt2021evaluating,guidotti2022counterfactual}.
 
-Counterfactual Explanations have full fidelity by construction as long as they are valid.
-Solutions to Equation~\ref{eq:general} are considered valid as soon as the predicted label matches the target label.
-A stripped-down counterfactual explanation is therefore little different from an adversarial example.
-In Figure~\ref{fig:adv}, for example, we have applied Wachter to MNIST data (centre panel) where the underlying classifier $M_{\theta}$ is a simple Multi-Layer Perceptron (MLP) with above 90 percent test accuracy.
-For the generated counterfactual $\mathbf{x}^{\prime}$ the model predicts the target label with high confidence (centre panel in Figure~\ref{fig:adv}).
-The explanation is valid by definition and therefore has full fidelity, even though it looks a lot like an Adversarial Example~\citep{goodfellow2014explaining}.
-\citet{schut2021generating} make the connection between Adversarial Examples and Counterfactual Explanations explicit and propose using a Jacobian-Based Saliency Map Attack (JSMA) to solve Equation~\ref{eq:general}.
-They demonstrate that this approach yields plausible counterfactuals for Bayesian, adversarially robust classifiers.
-Applying their approach to our simple MNIST classifier does not yield a realistic counterfactual but this one, too, is valid (right panel in Figure~\ref{fig:adv}).
-
-\begin{figure}
-  \centering
-  \begin{minipage}[t]{0.45\textwidth}
-    \centering
-    \includegraphics[width=\textwidth]{../artifacts/results/images/you_may_not_like_it.png}
-    \caption{Explanations or Adversarial Examples?
-Counterfactuals for turning a 9 (nine) into a 7 (seven): original image (left); counterfactual produced using~\citet{wachter2017counterfactual} (centre); and a counterfactual produced using the approach introduced by~\citep{schut2021generating} that uses Jacobian-Based Saliency Map Attacks to solve Equation~\ref{eq:general}.}\label{fig:adv}
-  \end{minipage}\hfill
-  \begin{minipage}[t]{0.45\textwidth}
-    \centering
-    \includegraphics[width=\textwidth]{../artifacts/results/images/surrogate_gone_wrong.png}
-    \caption{Using surrogates can improve plausibility, but also increases vulnerability.
-Counterfactuals for turning an 9 (nine) into a 7 (seven): original image (left); counterfactual produced using REVISE~\citep{joshi2019realistic} with a well-specified surrogate (centre); and a counterfactual produced using REVISE~\citep{joshi2019realistic} with a poorly specified surrogate (right).}\label{fig:vae}
-  \end{minipage}
-\end{figure}
-
-The crucial difference between Adversarial Examples (AE) and Counterfactual Explanations is one of intent.
-While an AE is intended to go unnoticed, a CE should have certain desirable properties.
-The literature has made this explicit by introducing various so-called \textit{desiderata} that counterfactuals should meet in order to properly serve both AI practitioners and individuals affected by AI decision-making systems.
-The list of desiderate includes but is not limited to the following: sparsity, proximity~\citep{wachter2017counterfactual}, actionability~\citep{ustun2019actionable}, diversity~\citep{mothilal2020explaining}, plausibility~\citep{joshi2019realistic,poyiadzi2020face,schut2021generating}, robustness~\citep{upadhyay2021robust,pawelczyk2022probabilistically,altmeyer2023endogenous} and causality~\citep{karimi2021algorithmic}.
-
-Researchers have come up with various ways to meet these desiderata, which have been extensively surveyed and evaluated in various studies~\citep{verma2020counterfactual,karimi2020survey,pawelczyk2021carla,artelt2021evaluating,guidotti2022counterfactual}.
 Perhaps unsurprisingly, the different desiderata are often positively correlated.
@@
 Consequently, the plausibility of counterfactuals has been among the primary concerns for researchers.
-
-Achieving plausibility is equivalent to ensuring that the generated counterfactuals comply with the true and unobserved data-generating process (DGP). Formally, this can be defined as follows:
+Achieving plausibility is equivalent to ensuring that the generated counterfactuals comply with the true and unobserved data-generating process (DGP).
+We define plausibility formally in this work as follows:
 
@@
 To learn the latent embedding, they propose using a generative model such as a Variational Autoencoder (VAE).
-Provided the surrogate model is well-trained, their proposed approach ---\textbf{REVISE}--- can yield plausible explanations like the one in the centre panel of Figure~\ref{fig:vae}.
+Provided the surrogate model is well-trained, their proposed approach called \textbf{REVISE} can yield plausible explanations.
 Others have proposed similar approaches: \citet{dombrowski2021diffeomorphic} traverse the base space of a normalizing flow to solve Equation~\ref{eq:general}; \citet{poyiadzi2020face} use density estimators ($\hat{p}: \mathcal{X} \mapsto [0,1]$) to constrain the counterfactuals to dense regions in the feature space; and, finally, \citet{karimi2021algorithmic} assume knowledge about the structural causal model that generates the data.
 
-While surrogates offer an easy way to address the need for plausibility, they also introduce friction: the generated explanations no longer depend exclusively on the black-box model itself.
-The counterfactual in the centre panel of Figure~\ref{fig:vae}, for example, looks more plausible than the corresponding counterfactuals generated by \textbf{Wachter} and \textbf{Schut}.
+A competing approach towards plausibility that is also closely related to this work instead relies on the black-box model itself.
+\citet{schut2021generating} show that to meet the plausibility objective we need not explicitly model the input distribution.
+Pointing to the undesirable engineering overhead induced by surrogate models, they propose that we rely on the implicit minimisation of predictive uncertainty instead.
+Their proposed methodology, which we will refer to as \textbf{Schut}, solves Equation~\ref{eq:general} by greedily applying JSMA in the feature space with standard cross-entropy loss and no penalty at all.
+The authors demonstrate theoretically and empirically that their approach yields counterfactuals for which the model $M_{\theta}$ predicts the target label $\mathbf{y}^+$ with high confidence.
+Provided the model is well-specified, these counterfactuals are plausible.
+This idea hinges on the assumption that the black-box model provides well-calibrated predictive uncertainty estimates.
+
+\section{Why Fidelity is not Enough}\label{fidelity}
+
+As discussed in the introduction, any valid Counterfactual Explanation also has full fidelity by construction: solutions to Equation~\ref{eq:general} are considered valid as soon as the label predicted by the model matches the target class.
+So while fidelity always applies, counterfactuals that address the various desiderata introduced above can look vastly different from each other.
+The following motivating example illustrates this point and demonstrates why fidelity is an insufficient evaluation metric to assess the faithfulness of Counterfactual Explanations.
+
+We have trained a simple image classifier $M_{\theta}$ on the well-known MNIST dataset~\citep{lecun1998mnist}: a Multi-Layer Perceptron (MLP) with above 90 percent test accuracy.
+No measures have been taken to improve the model's adversarial robustness or its capacity for predictive uncertainty quantification.
+The far left panel of Figure ~\ref{fig:motiv} shows a random sample drawn from the dataset.
+The underlying classifier correctly predicts the label `nine' for this image.
+For the given factual image and model, we have used Wachter, Schut and REVISE to generate one counterfactual each in the target class `seven'.
+The perturbed images are shown next to the factual image from left to right in Figure ~\ref{fig:motiv}.
+Captions on top of the individual images indicate the generator along with the predicted probability that the image belongs to the target class.
+In all three cases that probability is above 90 percent and yet the counterfactuals look very different from each other.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.75\textwidth]{../artifacts/results/images/mnist_motivation.png}
+  \caption{Counterfactuals for turning a 9 (nine) into a 7 (seven): original image (left); then from left to right the counterfactuals generated using Wachter, Schut and REVISE}\label{fig:motiv}
+\end{figure}
+
+Since Wachter is only concerned with proximity, the generated counterfactual is almost indistinguishable from the factual.
+The approach by Schut expects a well-calibrated model that can generate predictive uncertainty estimates.
+Since this is not the case, the generated counterfactual looks like an adversarial example.
+Finally, the counterfactual generated by REVISE looks much more plausible than the other two.
 But is it also more faithful to the behaviour of our MNIST classifier?
+That is much less clear because the surrogate used by REVISE introduces friction: the generated explanations no longer depend exclusively on the black-box model itself.
+
+So which of the counterfactuals most faithfully explains the behaviour of our image classifier?
 Fidelity cannot help us to make that judgement, because all of these counterfactuals have full fidelity.
@@
 where $\mathbf{r}_j \sim \mathcal{N}(\mathbf{0},\mathbf{I})$ is the stochastic term and the step-size $\epsilon$ is typically polynomially decayed~\citep{welling2011bayesian}.
-The term $\mathcal{E}(\mathbf{x}_j|\mathbf{y}^+)$ denotes the model energy conditioned on the target class label $\mathbf{y}^+$.
+The term $\mathcal{E}(\mathbf{x}_j|\mathbf{y}^+)$ denotes the model energy conditioned on the target class label $\mathbf{y}^+$ which we specify as the negative logit corresponding to the target class label $\mathbf{y}^*$.
 To allow for faster sampling, we follow the common practice of choosing the step-size $\epsilon$ and the standard deviation of $\mathbf{r}_j$ separately.
@@
 Generating multiple samples using SGLD thus yields an empirical distribution $\hat{\mathbf{X}}_{\theta,\mathbf{y}^+}$ that approximates what the model has learned about the input data.
-While in the context of Energy-Based Modelling, this is usually done during training, we propose to use SGLD during inference in order to evaluate and generate faithful model explanations.
+While in the context of Energy-Based Modelling, this is usually done during training, we propose to repurpose this approach during inference in order to evaluate and generate faithful model explanations.
 
@@
 
-\section{Methodological Framework}\label{meth}
+\section{Energy-Constrained Conformal Counterfactuals (ECCCo)}\label{meth}
 
-The primary objective of this work has been to develop a methodology for generating maximally plausible counterfactuals under minimal intervention.
-Our proposed framework is based on the premise that explanations should be plausible but not plausible at all costs.
-Energy-Constrained Conformal Counterfactuals (ECCCo) achieve this goal in two ways: firstly, they rely on the Black Box itself for the generative task; and, secondly, they involve an approach to predictive uncertainty quantification that is model-agnostic.
-In this section, we first formalise our notion of faithfulness in the context of CE.
-We then explain how exactly ECCCos are generated.
+In this section, we describe our proposed framework for generating Energy-Constrained Conformal Counterfactuals (ECCCos).
+It is based on the premise that counterfactuals should be faithful, first and foremost.
+Plausibility, as a secondary concern, is then still attainable but only to the degree that the black-box model itself has learned plausible explanations for the underlying data.
 
-\subsection{Quantifying the Model's Predictive Uncertainty}
+We begin by stating our proposed objective function, which involves tailored loss and penalty functions that we will explain in the following.
+In particular, we extend Equation~\ref{eq:general} as follows:
 
-\citet{schut2021generating} show that to meet the plausibility objective we need not explicitly model the input distribution.
-Pointing to the undesirable engineering overhead induced by surrogate models, they propose that we rely on the implicit minimisation of predictive uncertainty instead.
-Their proposed methodology solves Equation~\ref{eq:general} by greedily applying JSMA in the feature space with standard cross-entropy loss and no penalty at all.
-They demonstrate theoretically and empirically that their approach yields counterfactuals for which the model $M_{\theta}$ predicts the target label $\mathbf{y}^+$ with high confidence.
-Provided the model is well-specified, these counterfactuals are plausible.
-This idea hinges on the assumption that the black-box model provides well-calibrated predictive uncertainty estimates.
+\begin{equation} \label{eq:eccco}
+  \begin{aligned}
+  \mathbf{Z}^\prime= \arg \min_{\mathbf{Z}^\prime \in \mathcal{Z}^M}  &\{  {\text{yloss}(M_{\theta}(f(\mathbf{Z}^\prime)),\mathbf{y}^+)}+ \lambda_{1} {\text{dist}(f(\mathbf{Z}^\prime),\mathbf{x}) } \\
+  &+ \lambda_2 \text{dist}(f(\mathbf{Z}^\prime),\hat{\mathbf{x}}_{\theta}) + \lambda_3 \Omega(C_{\theta}(f(\mathbf{Z}^\prime);\alpha)) \} 
+  \end{aligned} 
+\end{equation}
 
-To quantify the model's predictive uncertainty we use Conformal Prediction (CP), an approach that has recently gained popularity in the Machine Learning community~\citep{angelopoulos2021gentle,manokhin2022awesome}.
+The first penalty term involving $\lambda_1$ induces proximity like in~\citet{wachter2017counterfactual}.
+Our default choice for $\text{dist}(\cdot)$ is the L1 Norm due to its sparsity-inducing properties.
+The second penalty term involving $\lambda_2$ constrains the energy of the generated counterfactual by penalising its distance from the lowest-energy conditional samples as defined in Equation~\ref{eq:faith}.
+Intuitively, this component induces faithfulness which coincides with plausibility to the extent that the model $M_{\theta}$ has learned the true posterior conditional distribution of inputs: $p_{\theta}(\mathbf{X}_{\mathbf{y}^+}) \rightarrow p(\mathbf{X}_{\mathbf{y}^+})$.
+
+The third and final penalty term involving $\lambda_3$ introduces a new but familiar concept: it ensures that the generated counterfactual is associated with low predictive uncertainty.
+As mentioned above,~\citet{schut2021generating} have shown that plausible counterfactuals can be generated implicitly through predictive uncertainty minimization.
+Unfortunately, this relies on the assumption that the model itself can provide predictive uncertainty estimates, which may be too restrictive in practice.
+
+To relax this assumption, we leverage recent advances in Conformal Prediction (CP), an approach to predictive uncertainty quantification that has recently gained popularity~\citep{angelopoulos2021gentle,manokhin2022awesome}.
 Crucially for our intended application, CP is model-agnostic and can be applied during inference without placing any restrictions on model training.
@@
 This particular case of conformal prediction is referred to as Split Conformal Prediction (SCP) as it involves splitting the training data into a proper training dataset and a calibration dataset.
-Details concerning our implementation of Conformal Prediction can be found in Appendix~\ref{app-cp}.
 
-\subsection{Energy-Constrained Conformal Counterfactuals (ECCCo)}
-
-Our framework for generating ECCCos combines the ideas introduced in the previous two subsections. Formally, we extend Equation~\ref{eq:general} as follows,
-
-\begin{equation} \label{eq:eccco}
-  \begin{aligned}
-  \mathbf{Z}^\prime= \arg \min_{\mathbf{Z}^\prime \in \mathcal{Z}^M}  &\{  {\text{yloss}(M_{\theta}(f(\mathbf{Z}^\prime)),\mathbf{y}^+)}+ \lambda_{1} {\text{dist}(f(\mathbf{Z}^\prime),\mathbf{x}) } \\
-  &+ \lambda_2 \text{dist}(f(\mathbf{Z}^\prime),\hat{\mathbf{x}}_{\theta}) + \lambda_3 \Omega(C_{\theta}(f(\mathbf{Z}^\prime);\alpha)) \} 
-  \end{aligned} 
-\end{equation}
-
-where $\hat{\mathbf{x}}_{\theta}$ denotes samples generated using SGLD (Equation~\ref{eq:sgld}) and $\text{dist}(\cdot)$ is a generic term for a distance metric.
-Our default choice for $\text{dist}(\cdot)$ is the L1 Norm since it induces sparsity.
-
-The first two terms in Equation~\ref{eq:eccco} correspond to the counterfactual search objective defined in~\citet{wachter2017counterfactual} which merely penalises the distance of counterfactuals from their factual values.
-The additional two penalties in ECCCo ensure that counterfactuals conform with the model's generative property and lead to minimally uncertain predictions, respectively.
-The hyperparameters $\lambda_1, ..., \lambda_3$ can be used to balance the different objectives: for example, we may choose to incur larger deviations from the factual in favour of faithfulness with the model's generative property by choosing lower values of $\lambda_1$ and relatively higher values of $\lambda_2$.
+In addition to the smooth set size penalty, we have also experimented with the use of a tailored function for $\text{yloss}(\cdot)$ that enforces that only the target label $\mathbf{y}^+$ is included in the prediction set ~\citet{stutz2022learning}.
+Further details are described Appendix~\ref{app-cp}.
 
@@
 
-\begin{minipage}[c]{0.45\textwidth}
+\begin{minipage}[c]{0.50\textwidth}
   \centering
-  \includegraphics[width=0.8\textwidth]{../artifacts/results/images/mnist_eccco.png}
-  \captionof{figure}{ECCCos for turning a 9 (nine) into a 7 (seven) for different black-box models from left to right: Multi-Layer Perceptron (a), Ensemble of MLPs (b), Joint Energy Model (c), Ensemble of JEMs (d).}\label{fig:eccco-mnist}
+  \includegraphics[width=0.8\textwidth]{../artifacts/results/images/poc.png}
+  \captionof{figure}{An example involving linearly separable synthetic data and illustrating how ECCCo compares to Wachter, Schut and REVISE.
+The factual class is 2 and the target class is 1.
+Contours indicate the predicted probability given by a Joint Energy Model that the counterfactual belongs to the target class.}\label{fig:poc}
 \end{minipage}  
 \hfill
-\begin{minipage}[c]{0.45\textwidth}
+\begin{minipage}[c]{0.40\textwidth}
 \captionof{algorithm}{Generating ECCCos (For more details, see Appendix~\ref{app:eccco})}\label{alg:eccco}
@@
 
-Figure~\ref{fig:eccco-mnist} presents ECCCos for the MNIST example from Section~\ref{background} for various black-box models of increasing complexity from left to right: a simple Multi-Layer Perceptron (MLP); an Ensemble of MLPs, each of the same architecture as the single MLP; a Joint Energy Model (JEM) based on the same MLP architecture; and finally, an Ensemble of these JEMs.
-Since Deep Ensembles have an improved capacity for predictive uncertainty quantification and JEMs are explicitly trained to learn plausible representations of the input data, it is intuitive to see that the plausibility of counterfactuals visibly improves from left to right.
-This provides some first anecdotal evidence that ECCCos achieve plausibility while maintaining faithfulness to the Black Box.
+Figure~\ref{fig:poc} illustrates how ECCCos compare to counterfactuals generated using Wachter, Schut and REVISE.
+The example involves synthetically generated linearly separable data that belong to one of two classes.
+Contours indicate the predicted probabilities of a Joint Energy Model that has been jointly trained to predict the output class and generate inputs~\citet{grathwohl2020your}.
+We have drawn a random sample from the factual class 1 and used each generator to produce a counterfactual in the target class 2.
+Both Wachter and Schut yield valid counterfactuals but fail to achieve plausibility in the sense that the generated counterfactuals are far away from the densely populated region in the target class.
+Conversely, ECCCo yields a faithful and plausible counterfactual in the neighbourhood of the generated conditional samples.
+REVISE fails to yield a valid counterfactual because the underlying surrogate has failed to learn the DGP.
 
@@
 
-\begin{question}[Feasibility]\label{rq:feasibility}
-  Is it feasible to generate plausible Counterfactual Explanations through ECCCo without relying on surrogate models?
+\begin{question}[Faithfulness]\label{rq:feasibility}
+  Are ECCCos more faithful to the black-box model than counterfactuals produced by benchmark generators?
 \end{question}
 
-\begin{question}[Drivers]\label{rq:drivers}
-  Subject to feasibility, what drives the performance of ECCCo?
-Is it sufficient to rely on energy-based modelling to quantify the model's generative property?
-Is it sufficient to rely on conformal prediction to quantify the model's uncertainty?
+\begin{question}[Plausibility]\label{rq:plausibility}
+  How do ECCCos compare to state-of-the-art generators with respect to plausibility?
 \end{question}
@@
 To assess and benchmark the performance of ECCCo against the state of the art, we generate multiple counterfactuals for different black-box models and datasets.
-In particular, we compare ECCCo to the following counterfactual generators that were introduced above: firstly; \textbf{Schut}~\citep{schut2021generating}, which minimizes predictive uncertainty; secondly, \textbf{REVISE}~\citep{joshi2019realistic}, which uses a VAE as its surrogate model; and, finally, \textbf{Wachter}~\citep{wachter2017counterfactual}, which serves as our baseline.
+In particular, we compare ECCCo to the following counterfactual generators that were introduced above: firstly; Schut~\citep{schut2021generating}, which works under the premise of minimizing predictive uncertainty; secondly, REVISE~\citep{joshi2019realistic}, which uses a VAE as its surrogate model; and, finally, Wachter~\citep{wachter2017counterfactual}, which serves as our baseline.
 

### 2023-05-15 paper/paper.tex
f me

@@
 
-\begin{question}[Faithfulness]\label{rq:feasibility}
-  Are ECCCos more faithful to the black-box model than counterfactuals produced by benchmark generators?
+\begin{question}[Faithfulness]\label{rq:faithfulness}
+  Are ECCCos more faithful than counterfactuals produced by our benchmark generators?
 \end{question}
@@
 
-\subsection{Key Evaluation Metrics}\label{evaluation}
-
-While we focus on these key evaluation metrics in the body of this paper, we also sporadically discuss outcomes with respect to other common measures used to evaluate the validity, proximity and sparsity of counterfactuals.
-Details can be found in Appendix~\ref{app:results}.
-
 \subsection{Experimental Setup}
@@
 To assess and benchmark the performance of ECCCo against the state of the art, we generate multiple counterfactuals for different black-box models and datasets.
-In particular, we compare ECCCo to the following counterfactual generators that were introduced above: firstly; Schut~\citep{schut2021generating}, which works under the premise of minimizing predictive uncertainty; secondly, REVISE~\citep{joshi2019realistic}, which uses a VAE as its surrogate model; and, finally, Wachter~\citep{wachter2017counterfactual}, which serves as our baseline.
+In particular, we compare ECCCo to the following counterfactual generators that were introduced above: firstly; Schut~\citep{schut2021generating}, which works under the premise of minimizing predictive uncertainty; secondly, REVISE~\citep{joshi2019realistic}, which is state-of-the-art with respect to plausibility; and, finally, Wachter~\citep{wachter2017counterfactual}, which serves as our baseline.
+We also consider two variations of ECCCo: `ECCCo (no CP)' involves no set size penalty ($\lambda_3=0$ in Equation~\ref{eq:eccco}), while `ECCCo (no EBM)' does not penalise the distance to samples generated through SGLD ($\lambda_2=0$ in Equation~\ref{eq:eccco}).
+These have been added to gain some sense of the degree to which the two components underlying ECCCo---namely energy-based modelling (EBM) and conformal prediction (CP)---drive the results.
 
 We use both synthetic and real-world datasets from different domains, all of which are publically available and commonly used to train and benchmark classification algorithms.
-The synthetic datasets include: a dataset containing two \textbf{Linearly Separable} Gaussian clusters ($n=1000$), as well as the well-known \textbf{Circles} ($n=1000$) and \textbf{Moons} ($n=2500$) data.
+We synthetically generate a dataset containing two \textbf{Linearly Separable} Gaussian clusters ($n=1000$), as well as the well-known \textbf{Circles} ($n=1000$) and \textbf{Moons} ($n=2500$) data.
+Since these data are generated by distributions of varying degrees of complexity, they allow us to gauge how complexity affects the different generators.
+
 As for real-world data, we follow~\citet{schut2021generating} and use the \textbf{MNIST}~\citep{lecun1998mnist} dataset containing images of handwritten digits such as the examples shown above.
@@
 
-As with the example in Section~\ref{meth}, we use simple neural networks (\textbf{MLP}) and Joint Energy Models (\textbf{JEM}).
+For the predictive modelling tasks, we use simple neural networks (\textbf{MLP}) and Joint Energy Models (\textbf{JEM}).
 For the more complex real-world datasets we also use ensembling in each case.
+Both joint-energy modelling and ensembling are associated with generative properties and adversarial robustness, so we expect this to be positively correlated with the plausibility of ECCCos.
 To account for stochasticity, we generate multiple counterfactuals for each possible target class, generator, model and dataset.
@@
 
-\subsection{Results}
+\subsection{Results for Synthetic Data}
 
@@
 In bold we have highlighted the best outcome for each model and metric.
-To provide some sense of the statistical significance of our findings, we have added asterisks to indicate that a given value is at least one ($*$) or two ($**$) standard deviations lower than the baseline (Wachter).
+To provide some sense of effect sizes, we have added asterisks to indicate that a given value is at least one ($*$) or two ($**$) standard deviations lower than the baseline (Wachter).
 
 Starting with the high-level results for our Linearly Separable data, we find that ECCCo produces the most faithful counterfactuals for both black-box models.
-This is not surprising, since ECCCo directly enforces faithfulness through regularization.
-Crucially though, ECCCo also produces the most plausible counterfactuals for the Joint Energy Model, which was explicitly trained to learn plausible representations of the input data.
-This high-level pattern is broadly consistent across all datasets and supportive of our narrative, so it is worth highlighting: ECCCos consistently achieve high faithfulness, which---subject to the quality of the model itself---coincides with high plausibility.
-
-Zooming in on the granular details for the Linearly Separable data, note that the list of generators in Table~\ref{tab:results-synthetic} includes `ECCCo (no CP)' and `ECCCo (no EBM)' in addition to `ECCCo' and our benchmark generators.
-These have been added to gain some sense of the degree to which the two components underlying ECCCo---namely energy-based modelling (EBM) and conformal prediction (CP)---drive the results.
-Specifically, `ECCCo (no CP)' involves no set size penalty ($\lambda_3=0$ in Equation~\ref{eq:eccco}), while `ECCCo (no EBM)' does not penalise the distance to samples generated through SGLD ($\lambda_2=0$ in Equation~\ref{eq:eccco}).
-The corresponding results indicate that the positive results are dominated by the effect of quantifying and leveraging the model's generative property (EBM) in our search for counterfactuals.
-Conformal Prediction alone only leads to marginally improved faithfulness and plausibility relative to the benchmark generators for our JEM.
-As a final observation for the Linearly Separable data we note that for the MLP, increased faithfulness comes at the cost of reduced plausibility.
-Specifically, this means that counterfactuals generated through ECCCo end up further away from individuals in the target class than those produced by our benchmark generators.
+This is consistent with our design since ECCCo directly enforces faithfulness through regularization.
+Crucially though, ECCCo also produces the most plausible counterfactuals for both models.
+This dataset is so simple that even the MLP has learned plausible explanations of the input data.
+Zooming in on the granular details for the Linearly Separable data, the results for `ECCCo (no CP)' and `ECCCo (no EBM)' indicate that the positive results are dominated by the effect of quantifying and leveraging the model's generative property (EBM).
+Conformal Prediction alone only leads to marginally improved faithfulness and plausibility relative to the benchmark generators.
 
@@
 For the MLP, faithfulness is maintained but counterfactuals are not plausible.
-By comparison, REVISE yields fairly plausible counterfactuals in both cases, but it does so at the cost of faithfulness.
+This high-level pattern is broadly consistent other more complex datasets and supportive of our narrative, so it is worth highlighting: ECCCos consistently achieve high faithfulness, which---subject to the quality of the model itself---coincides with high plausibility.
+By comparison, REVISE yields the most plausible counterfactuals for the MLP, but it does so at the cost of faithfulness.
 We also observe that the best results for ECCCo are achieved when using both penalties.
@@
 
+For the Circles data, it appears that REVISE performs well but we note that it generates valid counterfactuals only half of the time (see Appendix~\ref{app:results} for a complete overview of all evaluation metrics).
+It turns out that in this case, the underlying VAE with default parameters has not adequately learned the data-generating process.
+Of course, it is possible to achieve better generative performance through hyperparameter tuning but this example serves to illustrate that REVISE depends strongly on the quality of the surrogate model.
+Independent of the outcome for REVISE, however, the results do not seem to indicate that ECCCo significantly improves faithfulness and plausibility for the Circles data.
+We think this points to a limitation of our evaluation metrics rather than ECCCo itself: computing average distances fails to account for the `wraparound' effect associated with circular data~\citep{gill2010circular}.
+
 \import{contents/}{table-synthetic.tex}
 
-For the Circles data, the most faithful counterfactuals are generated by ECCCo.
-While it appears that REVISE generates the most plausible counterfactuals in this case, we note that they are valid only half of the time (see Appendix~\ref{app:results} for a complete overview of all evaluation metrics).
-It turns out that in this case, the underlying VAE with default parameters has not adequately learned the data-generating process.
-Of course, it is possible to achieve better generative performance through hyperparameter tuning.
-But this example serves to illustrate that REVISE depends strictly on the quality of the surrogate model.
-Independent of the outcome for REVISE, however, the results do not seem to indicate that ECCCo significantly improves our plausibility metric for the Circles data.
+\subsection{Results for Real-World Data}
 
-Moving on to our real-world datasets, the results are shown in Table~\ref{tab:results-real-world}.
+The results for our real-world datasets are shown in Table~\ref{tab:results-real-world}.
 Once again the findings indicate that the plausibility of ECCCos is positively correlated with the capacity of the black-box model to distinguish plausible from implausible inputs.
@@
 
-For the tabular credit dataset (GMSC) we have struggled to get good generative and discriminative performance for our JEMs.
-Consequently, it is not surprising to find that ECCCo never achieves state-of-the-art plausibility, although it does improve outcomes compared to the baseline (Wachter).
-Concerning faithfulness, ECCCo once again consistently outperforms all other generators.
+For the tabular credit dataset (GMSC) it is inherently challenging to use deep neural networks in order to achieve good discriminative performance~\citep{borisov2021deep,grinsztajn2022why} and discriminative performance~\citep{liu2023goggle}, respectively.
+In order to achieve high plausibility, ECCCo effectively requires classifiers to achieve good performance for both tasks.
+Since this is a challenging task even for Joint Energy Models, it is not surprising to find that even though ECCCo once again achieves state-of-the-art faithfulness, it is outperformed by REVISE and Schut with respect to plausibility.
+
+\subsection{Key Takeways}
 
 To conclude this section, we summarize our findings with reference to the opening questions.
-Concerning the feasibility of our proposed methodology (Research Question~\ref{rq:feasibility}), our findings demonstrate that it is indeed possible to generate plausible counterfactuals without the need for surrogate models.
-A related important finding is that ECCCo never sacrifices faithfulness for plausibility: any plausible ECCCo also faithfully describes model behaviour.
-This mitigates the risk of generating plausible explanations for models that are, in fact, highly susceptible to implausible counterfactuals as well.
-Our findings here indicate that ECCCo achieves this result primarily by leveraging the model's generative property.
-We think that further work is needed, however, to definitively answer Research Question~\ref{rq:drivers}, on which we elaborate in the following section.
+The results have clearly demonstrated that ECCCo consistently achieves state-of-the-art faithfulness, as it was designed to do (Research Question~\ref{rq:faithfulness}).
+A related important finding is that ECCCo yields highly plausible explanations provided that they faithfully describe model behaviour (Research Question~\ref{rq:plausibility}).
+Our findings here also indicate that ECCCo achieves this result primarily by leveraging the model's generative property.
 
@@
 Conversely, we have not found that penalising prediction set sizes disproportionately strongly had any discernable effect on our results.
-As discussed above, we also struggled to achieve good results by relying on conformal prediction alone.
-We want to caveat this finding by acknowledging that the role of CP in this context needs to be investigated more thoroughly through future work.
-Our suggested approach involving a smooth set size penalty may be insufficient in this context.
+Our results indicate that Conformal Prediction alone is often not sufficient to achieve faithfulness and plausibility, although we acknowlege that this needs to be investigated more thoroughly through future work.
 
-The fact that our findings are primarily driven by applying ideas from energy-based modelling presents a challenge in itself: while our approach is readily applicable to models with gradient access like deep neural networks, more work is needed to generalise our methodology to other popular machine learning models such as gradient-boosted trees.
-Relatedly, we have encountered common challenges associated with energy-based modelling during our experiments including sensitivity to scale, training instabilities and sensitivity to hyperparameters.
-We have also struggled to apply our proposed approach to low-dimensional tabular data.
+Furthermore, while our approach is readily applicable to models with gradient access like deep neural networks, more work is needed to generalise our methodology to other popular machine learning models such as gradient-boosted trees.
+Relatedly, common challenges associated with energy-based modelling during our experiments including sensitivity to scale, training instabilities and sensitivity to hyperparameters also apply to ECCCo.
 
@@
 We have proposed a new way to generate Counterfactual Explanations that are maximally faithful to the black-model they aim to explain.
-Our proposed counterfactual generator, ECCCo, produces plausible counterfactual if and only if the black-model itself has learned realistic representations of the data.
+Our proposed counterfactual generator, ECCCo, produces plausible counterfactual if and only if the black-model itself has learned realistic representations of the data, which we demonstrate through rigorous empirical analysis.
 This should enable researchers and practitioners to use counterfactuals in order to discern trustworthy models from unreliable ones.

### 2023-05-16 notebooks/poc.qmd
new illustrative chart

@@
 ```{julia}
+using Colors
+col_pal = palette(:seaborn_colorblind)
+Random.seed!(1234)
+using CounterfactualExplanations.Generators: ∇
+
+λ₁ = 0.1
+λ₂ = 1.0
+λ₃ = 0.5
+Λ = [λ₁, λ₂, λ₃]
+η = 0.01
+
+M = ECCCo.ConformalModel(mach.model, mach.fitresult)
+factual_label =  levels(labels)[2]
+x_factual = reshape(X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
+target =  levels(labels)[1]
+factual = predict_label(M, counterfactual_data, x_factual)[1]
+
+opt = Flux.Optimise.Descent(η)
+
+generator_dict = OrderedDict(
+    "Wachter" => WachterGenerator(λ = 0.3, opt=opt),
+    "ECCCo (no EBM)" => ECCCoGenerator(λ = [λ₁,λ₂,0.0], opt=opt),
+    "ECCCo (no CP)" => ECCCoGenerator(λ = [λ₁,0.0,λ₃], opt=opt),
+    "ECCCo" => ECCCoGenerator(λ = Λ, opt=opt),
+)
+
+# Gradient field:
+function loss_grads(generator, model, ce, x)
+    x = Float32.(x)
+    _ce = deepcopy(ce)
+    _ce.s′ = x
+    return ∇(generator,M,_ce)
+end
+
+meshgrid(x, y) = (repeat(x, outer=length(y)), repeat(y, inner=length(x)))
+
+xlims, ylims = extrema(X, dims=2)
+xrange = range(xlims..., length=10)
+yrange = range(ylims..., length=10)
+x1, x2 = meshgrid(xrange, yrange)
+inputs = zip(x1, x2)
+
+# Plot:
+ces = Dict{Any,Any}()
+plts = []
+for (name, generator) in generator_dict
+
+    # CE:
+    ce = generate_counterfactual(
+        x_factual, target, counterfactual_data, M, generator;
+        initialization=:identity, 
+        converge_when=:generator_conditions,
+    )
+
+    # Main plot (path):
+    plt = Plots.plot(
+        ce, title=name, alpha=0.1, cbar=false, 
+        axis=nothing, length_out=10, contour_alpha=0.0,
+        legend = false,
+        palette = col_pal,
+    )
+
+    # Generated samples:
+    if name ∈ ["ECCCo","ECCCo (no CP)"]
+        _X = distance_from_energy(ce, return_conditionals=true)
+        Plots.scatter!(
+            _X[1,:],_X[2,:], color=col_pal[end-1], shape=:star5, 
+            ms=10, label="x̂|$target", alpha=0.1
+        )
+    end
+
+    # Gradient field:
+    u = []
+    v = []
+    for (x, y) in inputs
+        g = -loss_grads(generator, M, ce, [x, y][:,:])
+        push!(u, η * g[1])
+        push!(v, η * g[2])
+    end
+    Plots.quiver!(x1, x2, quiver=(u, v), color=col_pal[5])
+
+    push!(plts, plt)
+    ces[name] = ce
+end
+plt = Plots.plot(plts..., size=(500,520))
+display(plt)
+savefig(plt, joinpath(output_images_path, "poc_gradient_fields.png"))
+```
+
+```{julia}
 ce = ces["ECCCo"]
@@
 
-scale = 0.1
 for (x, y) in inputs

### 2023-05-16 paper/paper.tex
new illustrative chart

@@
 We synthetically generate a dataset containing two \textbf{Linearly Separable} Gaussian clusters ($n=1000$), as well as the well-known \textbf{Circles} ($n=1000$) and \textbf{Moons} ($n=2500$) data.
-Since these data are generated by distributions of varying degrees of complexity, they allow us to gauge how complexity affects the different generators.
+Since these data are generated by distributions of varying degrees of complexity, they allow us to assess how the generators and our proposed evaluation metrics handle this.
 

### 2023-05-16 notebooks/gmsc.qmd
rerun

@@
 # Hyper:
-_retrain = true
+_retrain = false
 
@@
 λ₁ = 0.1
-λ₂ = 0.1
-λ₃ = 1.0
+λ₂ = 0.5
+λ₃ = 0.5
 Λ = [λ₁, λ₂, λ₃]
@@
     "Schut" => GreedyGenerator(),
-    "ECCCo" => ECCCoGenerator(λ=Λ, opt=opt, use_class_loss=use_class_loss),
+    "ECCCo" => ECCCoGenerator(
+        λ=Λ, opt=opt, use_class_loss=use_class_loss,
+        nsamples=10, nmin=10,
+    ),
 )

### 2023-05-16 notebooks/mnist.qmd
rerun

--- 2023-05-15 notebooks/mnist.qmd
@@
     x::Union{AbstractArray, Int}=x_factual, target::Int=target;
-    λ=[0.25,0.75,0.75],
+    λ=[0.1,0.25,0.25],
     temp=0.1,
@@
         use_class_loss=use_class_loss,
+        nsamples=10,
+        nmin=10,
     )
@@
 ```{julia}
-if _regen
+_regen_all_digits = false
+if _regen_all_digits
     function plot_all_digits(rng=123;verbose=true,kwargs...)

### 2023-05-16 src/generator.jl

@@ -22,6 +22,8 @@ function ECCCoGenerator(;
     temp::Real=0.1, 
     opt::Union{Nothing,Flux.Optimise.AbstractOptimiser}=nothing,
     use_class_loss::Bool=false,
+    nsamples::Int=50,
+    nmin::Int=25,
     kwargs...
 )
 
@@ -41,7 +43,13 @@ function ECCCoGenerator(;
     function _set_size_penalty(ce::AbstractCounterfactualExplanation)
         return ECCCo.set_size_penalty(ce; κ=κ, temp=temp)
     end
-    _penalties = [Objectives.distance_l1, _set_size_penalty, ECCCo.distance_from_energy]
+
+    # Energy penalty
+    function _energy_penalty(ce::AbstractCounterfactualExplanation)
+        return ECCCo.distance_from_energy(ce; n=nsamples, nmin=nmin, kwargs...)
+    end
+
+    _penalties = [Objectives.distance_l1, _set_size_penalty, _energy_penalty]
     λ = λ isa AbstractFloat ? [0.0, λ, λ] : λ
 
     # Generator

### 2023-05-16 paper/paper.tex

@@
 
-\begin{minipage}[c]{0.50\textwidth}
+\begin{minipage}[c]{0.45\textwidth}
   \centering
-  \includegraphics[width=0.8\textwidth]{../artifacts/results/images/poc.png}
-  \captionof{figure}{An example involving linearly separable synthetic data and illustrating how ECCCo compares to Wachter, Schut and REVISE.
-The factual class is 2 and the target class is 1.
-Contours indicate the predicted probability given by a Joint Energy Model that the counterfactual belongs to the target class.}\label{fig:poc}
+  \includegraphics[width=1.0\textwidth]{../artifacts/results/images/poc_gradient_fields.png}
+  \captionof{figure}{Gradient fields and counterfactual paths for different generators.
+The objective is to generate a counterfual in the `blue' class for a sample from the `orange' class.
+Bright yellow stars indicate conditional samples generated through SGLD.
+The underlying classifier is a Joint Energy Model.}\label{fig:poc}
 \end{minipage}  
 \hfill
-\begin{minipage}[c]{0.40\textwidth}
+\begin{minipage}[c]{0.45\textwidth}
 \captionof{algorithm}{Generating ECCCos (For more details, see Appendix~\ref{app:eccco})}\label{alg:eccco}
@@
 \end{minipage}
-
-\medskip
-
-% \begin{minipage}[c]{\textwidth}
-%   \centering
-%   \includegraphics[width=0.8\textwidth]{../artifacts/results/images/mnist_eccco.png}
-%   \captionof{figure}{Original image (left) and ECCCos for turning a 9 (nine) into a 7 (seven) for different black-box models from left to right: Multi-Layer Perceptron (MLP), Ensemble of MLPs, Joint Energy Model (JEM), Ensemble of JEMs.}\label{fig:eccco-mnist}
-% \end{minipage}
 

### 2023-05-16 notebooks/mnist.qmd
improved tables

@@
     "REVISE" => REVISEGenerator(λ=Λ[1]),
-    "Schut" => GreedyGenerator(),
+    "Schut" => greedy_generator,
     "ECCCo" => eccco_generator,

### 2023-05-16 notebooks/tables.Rmd
improved tables

@@
 dt[,source:=ifelse(dataname %in% synth, "synthetic", "real-world")]
+reported_data <- c(
+  "MNIST",
+  "GMSC",
+  "Linearly Separable",
+  "Moons",
+  "Circles"
+)
+dt <- dt[dataname %in% reported_data]
+dt[,non_valid:=variable=="validity" & value==0.0,.(sample,dataname,generator,model,target,factual,source)]
+dt[,non_valid:=any(non_valid==TRUE),.(sample,dataname,generator,model,target,factual,source)]
+dt_valid <- dt[non_valid==FALSE]
 ```
@@
     .(
-      value=sprintf("%1.2f (%1.2f)", mean(value), sd(value)),
+      value=sprintf("%1.2f ± %1.2f", mean(value), sd(value)),
       val = mean(value),
@@
 # Add conditional formatting:
+tab$value <- cell_spec(tab$value, "latex", bold=tab$top_val)
 tab[one_std_wachter==T,value:=paste0(value,"*")]
+tab[one_std_wachter==F,value:=paste0(value,"\\hphantom{*}")]
 tab[two_std_wachter==T,value:=paste0(value,"*")]
-tab$value <- cell_spec(tab$value, "latex", bold=tab$top_val)
+tab[two_std_wachter==F,value:=paste0(value,"\\hphantom{*}")]
 
@@
 tab[,one_std_wachter:=NULL]
+```
+
+```{r}
+tab_valid <- dt_valid[
+    ,
+    .(
+      value=sprintf("%1.2f ± %1.2f", mean(value), sd(value)),
+      val = mean(value),
+      std = sd(value)
+    ),
+    .(dataname, generator, model, variable, source)
+]
+
+tab_valid$top_val = F
+tab_valid$one_std_wachter = F
+tab_valid$two_std_wachter = F
+
+# Measures to be minimized:
+min_measures <- c(
+  "distance",
+  "distance_from_energy", 
+  "distance_from_targets",
+  "set_size_penalty"
+)
+tab_valid[variable %in% min_measures,top_val:=val==min(val),.(model, dataname, variable)]
+tab_valid[variable %in% min_measures,top_val:=ifelse(rep(all(top_val),length(top_val)),F,top_val),.(model, dataname, variable)]
+tab_valid[variable %in% min_measures,two_std_wachter:=val+2*std<val[generator=="Wachter"],.(model, dataname, variable)]
+tab_valid[variable %in% min_measures,one_std_wachter:=val+1*std<val[generator=="Wachter"],.(model, dataname, variable)]
+
+# Measures to be maximized:
+max_measures <- c(
+  "validity",
+  "redundancy"
+)
+tab_valid[variable %in% max_measures,top_val:=val==max(val),.(model, dataname, variable)]
+tab_valid[variable %in% max_measures,top_val:=ifelse(rep(all(top_val),length(top_val)),F,top_val),.(model, dataname, variable)]
+tab_valid[variable %in% max_measures,two_std_wachter:=val-2*std>val[generator=="Wachter"],.(model, dataname, variable)]
+tab_valid[variable %in% max_measures,one_std_wachter:=val-1*std>val[generator=="Wachter"],.(model, dataname, variable)]
+
+# Add conditional formatting:
+tab_valid$value <- cell_spec(tab_valid$value, "latex", bold=tab_valid$top_val)
+tab_valid[one_std_wachter==T,value:=paste0(value,"*")]
+tab_valid[one_std_wachter==F,value:=paste0(value,"\\hphantom{*}")]
+tab_valid[two_std_wachter==T,value:=paste0(value,"*")]
+tab_valid[two_std_wachter==F,value:=paste0(value,"\\hphantom{*}")]
+
+# Remove redundant columns:
+tab_valid[,val:=NULL]
+tab_valid[,std:=NULL]
+tab_valid[,top_val:=NULL]
+tab_valid[,two_std_wachter:=NULL]
+tab_valid[,one_std_wachter:=NULL]
+```
+
+## Main tables
+
+```{r}
+# Choices:
+measures <- c("distance_from_energy", "distance_from_targets")
+measure_names <- c("Unfaithfulness ↓","Implausibility ↓")
+chosen_source <- "real-world"
+chosen_data <- c(
+  "MNIST",
+  "GMSC"
+)
+tab_i <- tab
+
+# Logic:
+tab_i <- tab_i[variable %in% measures]
+tab_i <- tab_i[source == chosen_source]
+tab_i <- tab_i[dataname %in% chosen_data]
+tab_i[,dataname:=factor(dataname,levels=chosen_data)]
+tab_i <- dcast(tab_i, model + generator ~ dataname + variable)
+col_names <- c(
+  "Model",
+  "Generator",
+  rep(measure_names,length(chosen_data))
+)
+caption <- sprintf(
+  "Results for %s datasets: sample averages plus/minus one standard deviation.
+Best outcomes are highlighted in bold.
+Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (Wachter).
+\\label{tab:results-%s} \\newline",
+  chosen_source,
+  chosen_source
+)
+file_name <- sprintf(
+  "paper/contents/table-%s.tex",
+  chosen_source
+)
+sub_header <- rep(length(measures), length(chosen_data))
+names(sub_header) <- chosen_data
+header <- c(
+  " " = 2, sub_header
+)
+line_sep <- c(rep("",length(measures)-1),"\\addlinespace")
+algin_cols <- c(rep('l',2),rep('c',ncol(tab_i)-2))
+kbl(
+  tab_i, caption = caption, 
+  align = algin_cols, col.names=col_names, booktabs = T, escape=F, 
+  format="latex", linesep = line_sep 
+) %>%
+  kable_styling(latex_options = c("scale_down")) %>%
+  kable_paper(full_width = F) %>%
+  add_header_above(header) %>%
+  collapse_rows(columns = 1:2, latex_hline = "major", valign = "middle") %>%
+  save_kable(file_name)
+```
+
+```{r}
+# Choices:
+measures <- c("distance_from_energy", "distance_from_targets")
+measure_names <- c("Unfaithfulness ↓","Implausibility ↓")
+chosen_source <- "synthetic"
+chosen_data <- c(
+  "Linearly Separable",
+  "Moons",
+  "Circles"
+)
+tab_i <- tab
+
+# Logic:
+tab_i <- tab_i[variable %in% measures]
+tab_i <- tab_i[source == chosen_source]
+tab_i <- tab_i[dataname %in% chosen_data]
+tab_i[,dataname:=factor(dataname,levels=chosen_data)]
+tab_i <- dcast(tab_i, model + generator ~ dataname + variable)
+col_names <- c(
+  "Model",
+  "Generator",
+  rep(measure_names,length(chosen_data))
+)
+caption <- sprintf(
+  "Results for %s datasets: sample averages plus/minus one standard deviation.
+Best outcomes are highlighted in bold.
+Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (Wachter).
+\\label{tab:results-%s} \\newline",
+  chosen_source,
+  chosen_source
+)
+file_name <- sprintf(
+  "paper/contents/table-%s.tex",
+  chosen_source
+)
+sub_header <- rep(length(measures), length(chosen_data))
+names(sub_header) <- chosen_data
+header <- c(
+  " " = 2, sub_header
+)
+line_sep <- c(rep("",length(measures)-1),"\\addlinespace")
+algin_cols <- c(rep('l',2),rep('c',ncol(tab_i)-2))
+kbl(
+  tab_i, caption = caption, 
+  align = algin_cols, col.names=col_names, booktabs = T, escape=F, 
+  format="latex", linesep = line_sep 
+) %>%
+  kable_styling(latex_options = c("scale_down")) %>%
+  kable_paper(full_width = F) %>%
+  add_header_above(header) %>%
+  collapse_rows(columns = 1:2, latex_hline = "major", valign = "middle") %>%
+  save_kable(file_name)
 ```
@@
 
-## Main tables
+## Full table (valid only)
 
 ```{r}
-# Choices:
-measures <- c("distance_from_energy", "distance_from_targets")
-measure_names <- c("Unfaithfulness ↓","Implausibility ↓")
-chosen_source <- "real-world"
-chosen_data <- c(
-  "MNIST",
-  "GMSC"
-)
-tab_i <- tab
-
-# Logic:
-tab_i <- tab_i[variable %in% measures]
-tab_i <- tab_i[source == chosen_source]
-tab_i <- tab_i[dataname %in% chosen_data]
-tab_i[,dataname:=factor(dataname,levels=chosen_data)]
-tab_i <- dcast(tab_i, model + generator ~ dataname + variable)
+tab_full <- dcast(tab_valid, dataname + model + generator ~ variable)
 col_names <- c(
   "Model",
+  "Data",
   "Generator",
-  rep(measure_names,length(chosen_data))
+  "Cost ↓", 
+  "Unfaithfulness ↓", 
+  "Implausibility ↓", 
+  "Redundancy ↑",
+  "Uncertainty ↓",
+  "Validity ↑"
 )
-caption <- sprintf(
-  "Results for %s datasets.
+kbl(
+  tab_full, caption = "All results for all datasets including only valid counterfactuals.
 Standard deviations across samples are shown in parentheses.
@@
 Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (Wachter).
-\\label{tab:results-%s} \\newline",
-  chosen_source,
-  chosen_source
-)
-file_name <- sprintf(
-  "paper/contents/table-%s.tex",
-  chosen_source
-)
-sub_header <- rep(length(measures), length(chosen_data))
-names(sub_header) <- chosen_data
-header <- c(
-  " " = 2, sub_header
-)
-line_sep <- c(rep("",length(measures)-1),"\\addlinespace")
-kbl(
-  tab_i, caption = caption, 
-  align = "c", col.names=col_names, booktabs = T, escape=F, 
-  format="latex", linesep = line_sep 
+\\label{tab:results-full} \\newline",
+  align = "c", col.names=col_names, booktabs = F, escape=F, 
+  format="latex"
 ) %>%
@@
   kable_paper(full_width = F) %>%
-  add_header_above(header) %>%
   column_spec(1, bold = T) %>%
-  collapse_rows(columns = 1:2, latex_hline = "major", valign = "middle") %>%
-  save_kable(file_name)
+  collapse_rows(columns = 1:3, latex_hline = "custom", valign = "middle", custom_latex_hline = 1:2) %>%
+  save_kable("paper/contents/table_all_valid.tex")
 ```
-
-```{r}
-# Choices:
-measures <- c("distance_from_energy", "distance_from_targets")
-measure_names <- c("Unfaithfulness ↓","Implausibility ↓")
-chosen_source <- "synthetic"
-chosen_data <- c(
-  "Linearly Separable",
-  "Moons",
-  "Circles"
-)
-tab_i <- tab
-
-# Logic:
-tab_i <- tab_i[variable %in% measures]
-tab_i <- tab_i[source == chosen_source]
-tab_i <- tab_i[dataname %in% chosen_data]
-tab_i[,dataname:=factor(dataname,levels=chosen_data)]
-tab_i <- dcast(tab_i, model + generator ~ dataname + variable)
-col_names <- c(
-  "Model",
-  "Generator",
-  rep(measure_names,length(chosen_data))
-)
-caption <- sprintf(
-  "Results for %s datasets.
-Standard deviations across samples are shown in parentheses.
-Best outcomes are highlighted in bold.
-Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (Wachter).
-\\label{tab:results-%s} \\newline",
-  chosen_source,
-  chosen_source
-)
-file_name <- sprintf(
-  "paper/contents/table-%s.tex",
-  chosen_source
-)
-sub_header <- rep(length(measures), length(chosen_data))
-names(sub_header) <- chosen_data
-header <- c(
-  " " = 2, sub_header
-)
-line_sep <- c(rep("",length(measures)-1),"\\addlinespace")
-kbl(
-  tab_i, caption = caption, 
-  align = "c", col.names=col_names, booktabs = T, escape=F, 
-  format="latex", linesep = line_sep 
-) %>%
-  kable_styling(latex_options = c("scale_down")) %>%
-  kable_paper(full_width = F) %>%
-  add_header_above(header) %>%
-  column_spec(1, bold = T) %>%
-  collapse_rows(columns = 1:2, latex_hline = "major", valign = "middle") %>%
-  save_kable(file_name)
-```

### 2023-05-16 notebooks/mnist.qmd
Some polishing and addressed Cynthia's comments

--- 2023-05-16 notebooks/mnist.qmd
@@
     convert2image(MNIST, reshape(x_factual,28,28)),
-    axis=nothing, 
+    axis=([], false), 
     size=(img_height, img_height),
@@
         convert2image(MNIST, reshape(ce,28,28)),
-        axis=nothing, 
+        axis=([], false), 
         size=(img_height, img_height),
@@
 savefig(plt, joinpath(output_images_path, "surrogate_gone_wrong.png"))
+```
+
+```{julia}
+ces = zip([ce_wachter, ce_jsma, ce_strong])
+counterfactuals = reduce((x,y)->cat(x,y,dims=3),map(ce -> CounterfactualExplanations.counterfactual(ce[1]), ces))
+phat = reduce((x,y) -> cat(x,y,dims=3), map(ce -> target_probs(ce[1]), ces))
+plts = [p1]
+for x in zip(eachslice(counterfactuals; dims=3), eachslice(phat; dims=3), ["Wachter","Schut","REVISE"])
+    ce, _phat, _name = (x[1],x[2],x[3])
+    _title = "$(_name) (p=$(round(_phat[1]; digits=2)))"
+    plt = Plots.plot(
+        convert2image(MNIST, reshape(ce,28,28)),
+        axis=([], false), 
+        size=(img_height, img_height),
+        title=_title
+    )
+    plts = [plts..., plt]
+end
+plt = Plots.plot(plts...; size=(0.8*panel_height*length(plts),0.8*panel_height), layout=(1,length(plts)), dpi=400)
+display(plt)
+savefig(plt, joinpath(output_images_path, "mnist_motivation.png"))
 ```

### 2023-05-16 notebooks/poc.qmd
Some polishing and addressed Cynthia's comments

@@
 λ₁ = 0.1
-λ₂ = 1.0
+λ₂ = 0.5
 λ₃ = 0.5
@@
 
+function arrow0!(x, y, u, v; as=0.2, lw=1, lc=:black, la=1)
+    nuv = sqrt(u^2 + v^2)
+    v1, v2 = [u;v] / nuv,  [-v;u] / nuv
+    v4 = (3*v1 + v2)/3.1623  # sqrt(10) to get unit vector
+    v5 = v4 - 2*(v4'*v2)*v2
+    v4, v5 = as*nuv*v4, as*nuv*v5
+    println(v4)
+    println(v5)
+    Plots.plot!([x,x+u], [y,y+v], lw=lw, lc=lc, la=la)
+    Plots.plot!([x+u,x+u-v5[1]], [y+v,y+v-v5[2]], lw=lw, lc=lc, la=la)
+    Plots.plot!([x+u,x+u-v4[1]], [y+v,y+v-v4[2]], lw=lw, lc=lc, la=la)
+end
+
+GR.setarrowsize(0.5)
 # Plot:
@@
     Plots.quiver!(x1, x2, quiver=(u, v), color=col_pal[5])
-
+    # arrow0!.(x1, x2, u, v; as=0.2, lw=1.0,lc=col_pal[5], la=1)
     push!(plts, plt)
@@
 end
-plt = Plots.plot(plts..., size=(500,520))
+plt = Plots.plot(plts...; size=(panel_height*length(plts),panel_height), layout=(1,length(plts)), dpi=300)
+# plt = Plots.plot(plts..., size=(1000,250), layout=(1,4), dpi=300)
 display(plt)

### 2023-05-16 paper/paper.tex
improved tables
sorted citation order
Some polishing and addressed Cynthia's comments

@@
 % Bibliography
-\bibliographystyle{plainnat}
+\bibliographystyle{unsrtnat}
 \setcitestyle{numbers,square,comma}
@@
 Counterfactuals are then searched by minimizing a loss function that compares the predicted model output to the target outcome: $\text{yloss}(M_{\theta}(\mathbf{x}),\mathbf{y}^+)$.
-Since Counterfactual Explanations work directly with the black-box model, valid counterfactuals always have full local fidelity by construction where fidelity is defined as the degree to which explanations approximate the predictions of a black-box model~\citep{mothilal2020explaining,molnar2020interpretable}.
+Since CE work directly with the black-box model, valid counterfactuals always have full local fidelity by construction where fidelity is defined as the degree to which explanations approximate the predictions of a black-box model~\citep{mothilal2020explaining,molnar2020interpretable}.
 
-In situations where full fidelity is a requirement, CE offers a more appropriate solution to Explainable Artificial Intelligence (XAI) than other popular approaches like LIME~\citep{ribeiro2016why} and SHAP~\citep{lundberg2017unified}, which involve local surrogate models.
+In situations where full fidelity is a requirement, CE offer a more appropriate solution to Explainable Artificial Intelligence (XAI) than other popular approaches like LIME~\citep{ribeiro2016why} and SHAP~\citep{lundberg2017unified}, which involve local surrogate models.
 But even full fidelity is not a sufficient condition for ensuring that an explanation faithfully describes the behaviour of a model.
@@
 
-While Counterfactual Explanations can be generated for arbitrary regression models~\citep{spooner2021counterfactual}, existing work has primarily focused on classification problems.
+While CE can also be generated for arbitrary regression models~\citep{spooner2021counterfactual}, existing work has primarily focused on classification problems.
 Let $\mathcal{Y}=(0,1)^K$ denote the one-hot-encoded output domain with $K$ classes.
@@
 
-The baseline approach, which we will simply refer to as \textbf{Wachter}~\citep{wachter2017counterfactual}, searches a single counterfactual directly in the feature space and penalises its distance to the original factual.
+The baseline approach, which we will simply refer to as \textbf{Wachter}, searches a single counterfactual directly in the feature space and penalises its distance to the original factual.
 In this case, $f(\cdot)$ is simply the identity function and $\mathcal{Z}$ corresponds to the feature space itself.
@@
   \label{def:plausible}
-  Let $\mathcal{X}|\mathbf{y}^+$ denote the true conditional distribution of samples in the target class $\mathbf{y}^+$.
+  Let $\mathcal{X}|\mathbf{y}^+= p(\mathbf{x}|\mathbf{y}^+)$ denote the true conditional distribution of samples in the target class $\mathbf{y}^+$.
 Then for $\mathbf{x}^{\prime}$ to be considered a plausible counterfactual, we need: $\mathbf{x}^{\prime} \sim \mathcal{X}|\mathbf{y}^+$.
@@
 
-As discussed in the introduction, any valid Counterfactual Explanation also has full fidelity by construction: solutions to Equation~\ref{eq:general} are considered valid as soon as the label predicted by the model matches the target class.
+As discussed in the introduction, any valid counterfactual also has full fidelity by construction: solutions to Equation~\ref{eq:general} are considered valid as soon as the label predicted by the model matches the target class.
 So while fidelity always applies, counterfactuals that address the various desiderata introduced above can look vastly different from each other.
-The following motivating example illustrates this point and demonstrates why fidelity is an insufficient evaluation metric to assess the faithfulness of Counterfactual Explanations.
 
-We have trained a simple image classifier $M_{\theta}$ on the well-known MNIST dataset~\citep{lecun1998mnist}: a Multi-Layer Perceptron (MLP) with above 90 percent test accuracy.
+To demonstrate this with an example, we have trained a simple image classifier $M_{\theta}$ on the well-known MNIST dataset~\citep{lecun1998mnist}: a Multi-Layer Perceptron (MLP) with above 90 percent test accuracy.
 No measures have been taken to improve the model's adversarial robustness or its capacity for predictive uncertainty quantification.
@@
 The underlying classifier correctly predicts the label `nine' for this image.
-For the given factual image and model, we have used Wachter, Schut and REVISE to generate one counterfactual each in the target class `seven'.
+For the given factual image and model, we have used \textbf{Wachter}, \textbf{Schut} and \textbf{REVISE} to generate one counterfactual each in the target class `seven'.
 The perturbed images are shown next to the factual image from left to right in Figure ~\ref{fig:motiv}.
@@
   \centering
-  \includegraphics[width=0.75\textwidth]{../artifacts/results/images/mnist_motivation.png}
-  \caption{Counterfactuals for turning a 9 (nine) into a 7 (seven): original image (left); then from left to right the counterfactuals generated using Wachter, Schut and REVISE}\label{fig:motiv}
+  \includegraphics[width=0.8\textwidth]{../artifacts/results/images/mnist_motivation.png}
+  \caption{Counterfactuals for turning a 9 (nine) into a 7 (seven): original image (left); then from left to right the counterfactuals generated using \textbf{Wachter}, \textbf{Schut} and \textbf{REVISE}.}\label{fig:motiv}
 \end{figure}
 
-Since Wachter is only concerned with proximity, the generated counterfactual is almost indistinguishable from the factual.
-The approach by Schut expects a well-calibrated model that can generate predictive uncertainty estimates.
+Since \textbf{Wachter} is only concerned with proximity, the generated counterfactual is almost indistinguishable from the factual.
+The approach by~\citet{schut2021generating} expects a well-calibrated model that can generate predictive uncertainty estimates.
 Since this is not the case, the generated counterfactual looks like an adversarial example.
-Finally, the counterfactual generated by REVISE looks much more plausible than the other two.
+Finally, the counterfactual generated by \textbf{REVISE} looks much more plausible than the other two.
 But is it also more faithful to the behaviour of our MNIST classifier?
-That is much less clear because the surrogate used by REVISE introduces friction: the generated explanations no longer depend exclusively on the black-box model itself.
+That is much less clear because the surrogate used by \textbf{REVISE} introduces friction: the generated explanations no longer depend exclusively on the black-box model itself.
 
@@
 Fidelity cannot help us to make that judgement, because all of these counterfactuals have full fidelity.
-To bridge this gap, we introduce a new notion of faithfulness in the following section.
+Thus, fidelity is an insufficient evaluation metric to assess the faithfulness of CE.
 
-\section{A new Notion of Faithfulness}\label{faithfulness}
+\section{A New Notion of Faithfulness}\label{faithfulness}
 
-Analogous to Definition~\ref{def:plausible}, we propose to define faithfulness in the context of Counterfactual Explanations as follows:
+Considering the limitations of fidelity as demonstrated in the previous section, analogous to Definition~\ref{def:plausible}, we introduce a new notion of faithfulness in the context of CE:
 
@@
   \label{def:faithful}
-  Let $\mathcal{X}_{\theta}|\mathbf{y}^+ = p_{\theta}(\mathbf{X}_{\mathbf{y}^+})$ denote the conditional distribution of $\mathbf{x}$ in the target class $\mathbf{y}^+$, where $\theta$ denotes the parameters of model $M_{\theta}$.
-Then for $\mathbf{x}^{\prime}$ to be considered a conformal counterfactual, we need: $\mathbf{x}^{\prime} \sim \mathcal{X}_{\theta}|\mathbf{y}^+$.
+  Let $\mathcal{X}_{\theta}|\mathbf{y}^+ = p_{\theta}(\mathbf{x}|\mathbf{y}^+)$ denote the conditional distribution of $\mathbf{x}$ in the target class $\mathbf{y}^+$, where $\theta$ denotes the parameters of model $M_{\theta}$.
+Then for $\mathbf{x}^{\prime}$ to be considered a faithful counterfactual, we need: $\mathbf{x}^{\prime} \sim \mathcal{X}_{\theta}|\mathbf{y}^+$.
 \end{definition}
@@
 To assess counterfactuals with respect to Definition~\ref{def:faithful}, we need a way to quantify the posterior conditional distribution $p_{\theta}(\mathbf{x}|\mathbf{y}^+)$.
-To this end, we draw on recent advances in Energy-Based Modelling (EBM), a subdomain of machine learning that is concerned with generative or hybrid modelling~\citep{grathwohl2020your,}.
+To this end, we draw on recent advances in Energy-Based Modelling (EBM), a subdomain of machine learning that is concerned with generative or hybrid modelling~\citep{grathwohl2020your,du2020implicit}.
 In particular, note that if we fix $\mathbf{y}$ to our target value $\mathbf{y}^+$, we can conditionally draw from $p_{\theta}(\mathbf{x}|\mathbf{y}^+)$ using Stochastic Gradient Langevin Dynamics (SGLD) as follows,
@@
   \begin{aligned}
-    \text{impl} = \frac{1}{\lvert\mathbf{x} \in \mathbf{X}_{\mathbf{y}^+}\rvert} \sum_{\mathbf{x} \in \mathbf{X}_{\mathbf{y}^+}} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
+    \text{impl} = \frac{1}{\lvert\mathbf{X}_{\mathbf{y}^+}\rvert} \sum_{\mathbf{x} \in \mathbf{X}_{\mathbf{y}^+}} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
   \end{aligned}
@@
   \begin{aligned}
-    \text{unfaith} = \frac{1}{\lvert\mathbf{x} \in \hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^+}\rvert} \sum_{\mathbf{x} \in \hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^+}} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
+    \text{unfaith} = \frac{1}{\lvert \hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^+}\rvert} \sum_{\mathbf{x} \in \hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^+}} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
   \end{aligned}
@@
 
-In this section, we describe our proposed framework for generating Energy-Constrained Conformal Counterfactuals (ECCCos).
-It is based on the premise that counterfactuals should be faithful, first and foremost.
-Plausibility, as a secondary concern, is then still attainable but only to the degree that the black-box model itself has learned plausible explanations for the underlying data.
+In this section, we describe \textbf{ECCCo}, our proposed framework for generating Energy-Constrained Conformal Counterfactuals (ECCCos).
+It is based on the premise that counterfactuals should first and foremost be faithful.
+Plausibility, as a secondary concern, is then still attainable, but only to the degree that the black-box model itself has learned plausible explanations for the underlying data.
 
@@
 The second penalty term involving $\lambda_2$ constrains the energy of the generated counterfactual by penalising its distance from the lowest-energy conditional samples as defined in Equation~\ref{eq:faith}.
-Intuitively, this component induces faithfulness which coincides with plausibility to the extent that the model $M_{\theta}$ has learned the true posterior conditional distribution of inputs: $p_{\theta}(\mathbf{X}_{\mathbf{y}^+}) \rightarrow p(\mathbf{X}_{\mathbf{y}^+})$.
+Intuitively, this component induces faithfulness which coincides with plausibility to the extent that the model $M_{\theta}$ has learned the true posterior conditional distribution of inputs: $p_{\theta}(\mathbf{x}_{\mathbf{y}^+}) \rightarrow p(\mathbf{x}_{\mathbf{y}^+})$.
 
@@
 In addition to the smooth set size penalty, we have also experimented with the use of a tailored function for $\text{yloss}(\cdot)$ that enforces that only the target label $\mathbf{y}^+$ is included in the prediction set ~\citet{stutz2022learning}.
-Further details are described Appendix~\ref{app-cp}.
+Further details are provided in Appendix~\ref{app-cp}.
 
-\medskip
+\begin{figure}
+  \centering
+  \includegraphics[width=1.0\textwidth]{../artifacts/results/images/poc_gradient_fields.png}
+  \caption{Gradient fields and counterfactual paths for different generators.
+The objective is to generate a counterfual in the `blue' class for a sample from the `orange' class.
+Bright yellow stars indicate conditional samples generated through SGLD.
+The underlying classifier is a Joint Energy Model.}\label{fig:poc}
+\end{figure}
 
@@
 
-\begin{minipage}[c]{0.45\textwidth}
-  \centering
-  \includegraphics[width=1.0\textwidth]{../artifacts/results/images/poc_gradient_fields.png}
-  \captionof{figure}{Gradient fields and counterfactual paths for different generators.
-The objective is to generate a counterfual in the `blue' class for a sample from the `orange' class.
-Bright yellow stars indicate conditional samples generated through SGLD.
-The underlying classifier is a Joint Energy Model.}\label{fig:poc}
-\end{minipage}  
-\hfill
-\begin{minipage}[c]{0.45\textwidth}
-\captionof{algorithm}{Generating ECCCos (For more details, see Appendix~\ref{app:eccco})}\label{alg:eccco}
+\begin{algorithm}
+  \caption{The \textbf{ECCCo} generator}\label{alg:eccco}
   \begin{algorithmic}[1]
-  \Require $\mathbf{x}, \mathbf{y}^+, M_{\theta}, f, \Lambda, \alpha, \mathcal{D}, T, \eta, n_{\mathcal{B}}, N_{\mathcal{B}}$ \linebreak where $M_{\theta}(\mathbf{x})\neq\mathbf{y}^+$
+    \Require $\mathbf{x}, \mathbf{y}^+, M_{\theta}, f, \Lambda, \alpha, \mathcal{D}, T, \eta, n_{\mathcal{B}}$ where $M_{\theta}(\mathbf{x})\neq\mathbf{y}^+$
     \Ensure $\mathbf{x}^\prime$
-  \State Initialize $\mathbf{z}^\prime \gets f^{-1}(\mathbf{x})$
-  \State Generate buffer $\mathcal{B}$ of $N_{\mathcal{B}}$ conditional samples $\hat{\mathbf{x}}_{\theta}|\mathbf{y}^+$ using SGLD (Equation~\ref{eq:sgld})
-  \State Run \textit{SCP} for $M_{\theta}$ using $\mathcal{D}$
+    \State Initialize $\mathbf{z}^\prime \gets f^{-1}(\mathbf{x})$ \Comment{Map to counterfactual state space.}
+    \State Generate $\left\{\hat{\mathbf{x}}_{\theta,\mathbf{y}^+}\right\}_{n_{\mathcal{B}}} \gets p_{\theta}(\mathbf{x}_{\mathbf{y}^+})$ \Comment{Generate $n_{\mathcal{B}}$ samples using SGLD (Equation~\ref{eq:sgld}).}
+    \State Run \textit{SCP} for $M_{\theta}$ using $\mathcal{D}$ \Comment{Calibrate model through Split Conformal Prediction.}
     \State Initialize $t \gets 0$
-  \While{\textit{not converged} or $t < T$}
-  \State $\hat{\mathbf{x}}_{\theta, t} \gets \text{rand}(\mathcal{B},n_{\mathcal{B}})$
-  \State $\mathbf{z}^\prime \gets \mathbf{z}^\prime - \eta \nabla_{\mathbf{z}^\prime} \mathcal{L}(\mathbf{z}^\prime,\mathbf{y}^+,\hat{\mathbf{x}}_{\theta, t}; \Lambda, \alpha)$
+    \While{\textit{not converged} or $t < T$} \Comment{For convergence conditions see Appendix~\ref{app:eccco}.}
+    \State $\mathbf{z}^\prime \gets \mathbf{z}^\prime - \eta \nabla_{\mathbf{z}^\prime} \mathcal{L}(\mathbf{z}^\prime,\mathbf{y}^+,\hat{\mathbf{x}}_{\theta, t}; \Lambda, \alpha)$ \Comment{Take gradient step of size $\eta$.}
     \State $t \gets t+1$
     \EndWhile
-  \State $\mathbf{x}^\prime \gets f(\mathbf{z}^\prime)$
+    \State $\mathbf{x}^\prime \gets f(\mathbf{z}^\prime)$ \Comment{Map back to feature space.}
   \end{algorithmic}
-\end{minipage}
+\end{algorithm}
 
-\medskip
-
-The entire procedure for generating ECCCos is described in Algorithm~\ref{alg:eccco}.
+Algorithm~\ref{alg:eccco} describes how exactly \textbf{ECCCo} works.
 For the sake of simplicity and without loss of generality, we limit our attention to generating a single counterfactual $\mathbf{x}^\prime=f(\mathbf{z}^\prime)$ where in contrast to Equation~\ref{eq:eccco} $\mathbf{z}^\prime$ denotes a $1$-dimensional array containing a single counterfactual state.
-That state is initialized by passing the factual $\mathbf{x}$ through the encoder $f^{-1}$ which in our case corresponds to a simple feature transformer, rather than the encoder part of VAE as in REVISE~\citep{joshi2019realistic}.
+That state is initialized by passing the factual $\mathbf{x}$ through the encoder $f^{-1}$ which in our case corresponds to a simple feature transformer, rather than the encoder part of VAE as in \textbf{REVISE}.
 Next, we generate a buffer of $N_{\mathcal{B}}$ conditional samples $\hat{\mathbf{x}}_{\theta}|\mathbf{y}^+$ using SGLD (Equation~\ref{eq:sgld}) and conformalise the model $M_{\theta}$ through Split Conformal Prediction on training data $\mathcal{D}$.
@@
 The search terminates once the convergence criterium is met or the maximum number of iterations $T$ has been exhausted.
-Note that the choice of convergence criterium has important implications on the final counterfactual (for more detail on this see Appendix~\ref{app:eccco}).
+Note that the choice of convergence criterium has important implications on the final counterfactual.
+Additional details concerning Algorithm~\ref{alg:eccco} can be found in Appendix~\ref{app:eccco}.
 
-Figure~\ref{fig:poc} illustrates how ECCCos compare to counterfactuals generated using Wachter, Schut and REVISE.
+Figure~\ref{fig:poc} illustrates how ECCCos compare to counterfactuals generated using \textbf{Wachter}, \textbf{Schut} and \textbf{REVISE}.
 The example involves synthetically generated linearly separable data that belong to one of two classes.
@@
 We have drawn a random sample from the factual class 1 and used each generator to produce a counterfactual in the target class 2.
-Both Wachter and Schut yield valid counterfactuals but fail to achieve plausibility in the sense that the generated counterfactuals are far away from the densely populated region in the target class.
+Both \textbf{Wachter} and \textbf{Schut} yield valid counterfactuals but fail to achieve plausibility in the sense that the generated counterfactuals are far away from the densely populated region in the target class.
 Conversely, ECCCo yields a faithful and plausible counterfactual in the neighbourhood of the generated conditional samples.
-REVISE fails to yield a valid counterfactual because the underlying surrogate has failed to learn the DGP.
+\textbf{REVISE} fails to yield a valid counterfactual because the underlying surrogate has failed to learn the DGP.
 
@@
 To assess and benchmark the performance of ECCCo against the state of the art, we generate multiple counterfactuals for different black-box models and datasets.
-In particular, we compare ECCCo to the following counterfactual generators that were introduced above: firstly; Schut~\citep{schut2021generating}, which works under the premise of minimizing predictive uncertainty; secondly, REVISE~\citep{joshi2019realistic}, which is state-of-the-art with respect to plausibility; and, finally, Wachter~\citep{wachter2017counterfactual}, which serves as our baseline.
-We also consider two variations of ECCCo: `ECCCo (no CP)' involves no set size penalty ($\lambda_3=0$ in Equation~\ref{eq:eccco}), while `ECCCo (no EBM)' does not penalise the distance to samples generated through SGLD ($\lambda_2=0$ in Equation~\ref{eq:eccco}).
-These have been added to gain some sense of the degree to which the two components underlying ECCCo---namely energy-based modelling (EBM) and conformal prediction (CP)---drive the results.
+In particular, we compare ECCCo to the following counterfactual generators that were introduced above: firstly; \textbf{Schut}, which works under the premise of minimizing predictive uncertainty; secondly, \textbf{REVISE}, which is state-of-the-art with respect to plausibility; and, finally, \textbf{Wachter}, which serves as our baseline.
+We also consider two variations of our propsed generator: \textbf{ECCCo (no CP)} involves no set size penalty ($\lambda_3=0$ in Equation~\ref{eq:eccco}), while \textbf{ECCCo (no EBM)} does not penalise the distance to samples generated through SGLD ($\lambda_2=0$ in Equation~\ref{eq:eccco}).
+These have been added to gain some sense of the degree to which the two components underlying ECCCo---namely, energy-based modelling (EBM) and conformal prediction (CP)---drive the results.
 
-We use both synthetic and real-world datasets from different domains, all of which are publically available and commonly used to train and benchmark classification algorithms.
+We use both synthetic and real-world datasets from different domains, all of which are publicly available and commonly used to train and benchmark classification algorithms.
 We synthetically generate a dataset containing two \textbf{Linearly Separable} Gaussian clusters ($n=1000$), as well as the well-known \textbf{Circles} ($n=1000$) and \textbf{Moons} ($n=2500$) data.
@@
 
-As for real-world data, we follow~\citet{schut2021generating} and use the \textbf{MNIST}~\citep{lecun1998mnist} dataset containing images of handwritten digits such as the examples shown above.
+As for real-world data, we follow~\citet{schut2021generating} and use the \textbf{MNIST}~\citep{lecun1998mnist} dataset containing images of handwritten digits such as the example shown above in Figure~\ref{fig:motiv}.
 From the social sciences domain, we include Give Me Some Credit (\textbf{GMSC})~\citep{kaggle2011give}: a tabular dataset that has been studied extensively in the literature on Algorithmic Recourse~\citep{pawelczyk2021carla}.
@@
 
-Table~\ref{tab:results-synthetic} shows the key results for the synthetic datasets separated by model (first columns) and generator (second column).
+Table~\ref{tab:results-synthetic} shows the key results for the synthetic datasets separated by model (first column) and generator (second column).
 The numerical columns show the average values of our key evaluation metrics computed across all counterfactuals.
 Standard deviations are shown in parentheses.
-In bold we have highlighted the best outcome for each model and metric.
-To provide some sense of effect sizes, we have added asterisks to indicate that a given value is at least one ($*$) or two ($**$) standard deviations lower than the baseline (Wachter).
+We have highlighted the best outcome for each model and metric in bold.
+To provide some sense of effect sizes, we have added asterisks to indicate that a given value is at least one ($*$) or two ($**$) standard deviations lower than the baseline (\textbf{Wachter}).
 
@@
 This high-level pattern is broadly consistent other more complex datasets and supportive of our narrative, so it is worth highlighting: ECCCos consistently achieve high faithfulness, which---subject to the quality of the model itself---coincides with high plausibility.
-By comparison, REVISE yields the most plausible counterfactuals for the MLP, but it does so at the cost of faithfulness.
+By comparison, \textbf{REVISE} yields the most plausible counterfactuals for the MLP, but it does so at the cost of faithfulness.
 We also observe that the best results for ECCCo are achieved when using both penalties.
@@
 
-For the Circles data, it appears that REVISE performs well but we note that it generates valid counterfactuals only half of the time (see Appendix~\ref{app:results} for a complete overview of all evaluation metrics).
+For the Circles data, it appears that \textbf{REVISE} performs well, but we note that it generates valid counterfactuals only half of the time (see Appendix~\ref{app:results} for a complete overview of all evaluation metrics).
 It turns out that in this case, the underlying VAE with default parameters has not adequately learned the data-generating process.
-Of course, it is possible to achieve better generative performance through hyperparameter tuning but this example serves to illustrate that REVISE depends strongly on the quality of the surrogate model.
-Independent of the outcome for REVISE, however, the results do not seem to indicate that ECCCo significantly improves faithfulness and plausibility for the Circles data.
+Of course, it is possible to achieve better generative performance through hyperparameter tuning but this example serves to illustrate that \textbf{REVISE} depends strongly on the quality of the surrogate model.
+Independent of the outcome for \textbf{REVISE}, however, the results do not seem to indicate that ECCCo significantly improves faithfulness and plausibility for the Circles data.
 We think this points to a limitation of our evaluation metrics rather than ECCCo itself: computing average distances fails to account for the `wraparound' effect associated with circular data~\citep{gill2010circular}.
@@
 The case is very clear for MNIST: ECCCos are consistently more faithful than the corresponding counterfactuals produced by any of the benchmark generators and their plausibility gradually improves through ensembling and joint-energy modelling.
-For the JEM Ensemble, ECCCo is essentially on par with REVISE and does significantly better than the baseline generator.
+For the JEM Ensemble, ECCCo is essentially on par with \textbf{REVISE} and does significantly better than the baseline generator.
 We also note that ECCCo is the only generator that consistently achieves full validity for all models (Appendix~\ref{app:results}).
@@
 In order to achieve high plausibility, ECCCo effectively requires classifiers to achieve good performance for both tasks.
-Since this is a challenging task even for Joint Energy Models, it is not surprising to find that even though ECCCo once again achieves state-of-the-art faithfulness, it is outperformed by REVISE and Schut with respect to plausibility.
+Since this is a challenging task even for Joint Energy Models, it is not surprising to find that even though ECCCo once again achieves state-of-the-art faithfulness, it is outperformed by \textbf{REVISE} and \textbf{Schut} with respect to plausibility.
 
@@
 
-Even though we have taken considerable measures to study our proposed methodology carefully, this work is limited in scope, which caveats our findings.
+Even though we have taken considerable measures to study our proposed methodology carefully, limitations can still be identified.
 In particular, we have found that the performance of ECCCo is sensitive to hyperparameter choices.
 In order to achieve faithfulness, we generally had to penalise the distance from generated samples slightly more than the distance from factual values.
-This choice is associated with relatively higher costs to individuals since the proposed recourses typically involve more substantial feature changes than for our benchmark generators.
+This choice is associated with relatively higher costs to individuals since the proposed recourse typically involves more substantial feature changes than for our benchmark generators.
 
-Conversely, we have not found that penalising prediction set sizes disproportionately strongly had any discernable effect on our results.
-Our results indicate that Conformal Prediction alone is often not sufficient to achieve faithfulness and plausibility, although we acknowlege that this needs to be investigated more thoroughly through future work.
+Conversely, we have not found that disproportionately strongly penalising prediction set sizes had any discernable effect.
+Our results indicate that Conformal Prediction alone is often not sufficient to achieve faithfulness and plausibility, although we acknowledge that this needs to be investigated more thoroughly through future work.
 
@@
 This work leverages recent advances in energy-based modelling and conformal prediction in the context of Explainable Artificial Intelligence.
-We have proposed a new way to generate Counterfactual Explanations that are maximally faithful to the black-model they aim to explain.
+We have proposed a new way to generate counterfactuals that are maximally faithful to the black-model they aim to explain.
 Our proposed counterfactual generator, ECCCo, produces plausible counterfactual if and only if the black-model itself has learned realistic representations of the data, which we demonstrate through rigorous empirical analysis.
 This should enable researchers and practitioners to use counterfactuals in order to discern trustworthy models from unreliable ones.
-While the scope of this work limits its generalizability, we believe that ECCCo offers a solid baseline for future work on faithful Counterfactual Explanations.
+While the scope of this work limits its generalizability, we believe that ECCCo offers a solid baseline for future work on faithful CE.
 
@@
 \end{ack}
-
-\medskip
 
@@
 
-Observe from Equation~\ref{eq:scp} that Conformal Prediction works on an instance-level basis, much like Counterfactual Explanations are local.
+Observe from Equation~\ref{eq:scp} that Conformal Prediction works on an instance-level basis, much like CE are local.
 The prediction set for an individual instance $\mathbf{x}_i$ depends only on the characteristics of that sample and the specified error rate.
@@
 
+\import{contents/}{table_all_valid.tex}
+
 \end{document}

### 2023-05-16 paper/paper.tex
all comments addressed

@@
 
-The baseline approach, which we will simply refer to as \textbf{Wachter}, searches a single counterfactual directly in the feature space and penalises its distance to the original factual.
+The baseline approach, which we will simply refer to as \textit{Wachter}, searches a single counterfactual directly in the feature space and penalises its distance to the original factual.
 In this case, $f(\cdot)$ is simply the identity function and $\mathcal{Z}$ corresponds to the feature space itself.
@@
 To learn the latent embedding, they propose using a generative model such as a Variational Autoencoder (VAE).
-Provided the surrogate model is well-trained, their proposed approach called \textbf{REVISE} can yield plausible explanations.
+Provided the surrogate model is well-trained, their proposed approach called \textit{REVISE} can yield plausible explanations.
 Others have proposed similar approaches: \citet{dombrowski2021diffeomorphic} traverse the base space of a normalizing flow to solve Equation~\ref{eq:general}; \citet{poyiadzi2020face} use density estimators ($\hat{p}: \mathcal{X} \mapsto [0,1]$) to constrain the counterfactuals to dense regions in the feature space; and, finally, \citet{karimi2021algorithmic} assume knowledge about the structural causal model that generates the data.
@@
 Pointing to the undesirable engineering overhead induced by surrogate models, they propose that we rely on the implicit minimisation of predictive uncertainty instead.
-Their proposed methodology, which we will refer to as \textbf{Schut}, solves Equation~\ref{eq:general} by greedily applying JSMA in the feature space with standard cross-entropy loss and no penalty at all.
+Their proposed methodology, which we will refer to as \textit{Schut}, solves Equation~\ref{eq:general} by greedily applying JSMA in the feature space with standard cross-entropy loss and no penalty at all.
 The authors demonstrate theoretically and empirically that their approach yields counterfactuals for which the model $M_{\theta}$ predicts the target label $\mathbf{y}^+$ with high confidence.
@@
 
-To demonstrate this with an example, we have trained a simple image classifier $M_{\theta}$ on the well-known MNIST dataset~\citep{lecun1998mnist}: a Multi-Layer Perceptron (MLP) with above 90 percent test accuracy.
+To demonstrate this with an example, we have trained a simple image classifier $M_{\theta}$ on the well-known \textit{MNIST} dataset~\citep{lecun1998mnist}: a Multi-Layer Perceptron (\textit{MLP}) with above 90 percent test accuracy.
 No measures have been taken to improve the model's adversarial robustness or its capacity for predictive uncertainty quantification.
@@
 The underlying classifier correctly predicts the label `nine' for this image.
-For the given factual image and model, we have used \textbf{Wachter}, \textbf{Schut} and \textbf{REVISE} to generate one counterfactual each in the target class `seven'.
+For the given factual image and model, we have used \textit{Wachter}, \textit{Schut} and \textit{REVISE} to generate one counterfactual each in the target class `seven'.
 The perturbed images are shown next to the factual image from left to right in Figure ~\ref{fig:motiv}.
@@
   \includegraphics[width=0.8\textwidth]{../artifacts/results/images/mnist_motivation.png}
-  \caption{Counterfactuals for turning a 9 (nine) into a 7 (seven): original image (left); then from left to right the counterfactuals generated using \textbf{Wachter}, \textbf{Schut} and \textbf{REVISE}.}\label{fig:motiv}
+  \caption{Counterfactuals for turning a 9 (nine) into a 7 (seven): original image (left); then from left to right the counterfactuals generated using \textit{Wachter}, \textit{Schut} and \textit{REVISE}.}\label{fig:motiv}
 \end{figure}
 
-Since \textbf{Wachter} is only concerned with proximity, the generated counterfactual is almost indistinguishable from the factual.
+Since \textit{Wachter} is only concerned with proximity, the generated counterfactual is almost indistinguishable from the factual.
 The approach by~\citet{schut2021generating} expects a well-calibrated model that can generate predictive uncertainty estimates.
 Since this is not the case, the generated counterfactual looks like an adversarial example.
-Finally, the counterfactual generated by \textbf{REVISE} looks much more plausible than the other two.
-But is it also more faithful to the behaviour of our MNIST classifier?
-That is much less clear because the surrogate used by \textbf{REVISE} introduces friction: the generated explanations no longer depend exclusively on the black-box model itself.
+Finally, the counterfactual generated by \textit{REVISE} looks much more plausible than the other two.
+But is it also more faithful to the behaviour of our \textit{MNIST} classifier?
+That is much less clear because the surrogate used by \textit{REVISE} introduces friction: the generated explanations no longer depend exclusively on the black-box model itself.
 
@@
 As this distance is reduced, counterfactuals get more plausible under the assumption that the nearest neighbour itself is plausible in the sense of Definition~\ref{def:plausible}.
-In this work, we use the following adapted implausibility metric that relaxes this assumption,
+In this work, we use the following adapted implausibility metric,
 
@@
   \begin{aligned}
-    \text{impl} = \frac{1}{\lvert\mathbf{X}_{\mathbf{y}^+}\rvert} \sum_{\mathbf{x} \in \mathbf{X}_{\mathbf{y}^+}} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
+    \text{impl}(\mathbf{x}^{\prime},\mathbf{X}_{\mathbf{y}^+}) = \frac{1}{\lvert\mathbf{X}_{\mathbf{y}^+}\rvert} \sum_{\mathbf{x} \in \mathbf{X}_{\mathbf{y}^+}} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
   \end{aligned}
@@
 
-where $\mathbf{X}_{\mathbf{y}^+}$ is a subsample of the training data in the target class $\mathbf{y}^+$.
+where $\mathbf{x}^{\prime}$ denotes the counterfactual and $\mathbf{X}_{\mathbf{y}^+}$ is a subsample of the training data in the target class $\mathbf{y}^+$.
+By averaging over multiple samples in this manner, we avoid the need to make any assumption about the nearest neighbour of $\mathbf{x}^{\prime}$.
 
-This gives rise to a very similar evaluation metric for unfaithfulness.
+Equation~\ref{eq:impl} gives rise to a similar evaluation metric for unfaithfulness.
 We merely swap out the subsample of individuals in the target class for a subset $\hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^+}$ of the generated conditional samples:
@@
   \begin{aligned}
-    \text{unfaith} = \frac{1}{\lvert \hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^+}\rvert} \sum_{\mathbf{x} \in \hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^+}} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
+    \text{unfaith}(\mathbf{x}^{\prime},\hat{\mathbf{X}}^{n_E}) = \frac{1}{\lvert \hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^+}\rvert} \sum_{\mathbf{x} \in \hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^+}} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
   \end{aligned}
@@
 
-\section{Energy-Constrained Conformal Counterfactuals (ECCCo)}\label{meth}
+\section{Energy-Constrained Conformal Counterfactuals}\label{meth}
 
-In this section, we describe \textbf{ECCCo}, our proposed framework for generating Energy-Constrained Conformal Counterfactuals (ECCCos).
+In this section, we describe \textit{ECCCo}, our proposed framework for generating Energy-Constrained Conformal Counterfactuals (ECCCos).
 It is based on the premise that counterfactuals should first and foremost be faithful.
@@
   \mathbf{Z}^\prime= \arg \min_{\mathbf{Z}^\prime \in \mathcal{Z}^M}  &\{  {\text{yloss}(M_{\theta}(f(\mathbf{Z}^\prime)),\mathbf{y}^+)}+ \lambda_{1} {\text{dist}(f(\mathbf{Z}^\prime),\mathbf{x}) } \\
-  &+ \lambda_2 \text{dist}(f(\mathbf{Z}^\prime),\hat{\mathbf{x}}_{\theta}) + \lambda_3 \Omega(C_{\theta}(f(\mathbf{Z}^\prime);\alpha)) \} 
+  &+ \lambda_2 \text{unfaith}(f(\mathbf{Z}^\prime),\hat{\mathbf{X}}^{n_E}) + \lambda_3 \Omega(C_{\theta}(f(\mathbf{Z}^\prime);\alpha)) \} 
   \end{aligned} 
@@
 Conformal classifiers produce prediction sets for individual inputs that include all output labels that can be reasonably attributed to the input.
-These sets tend to be larger for inputs that do not conform with the training data and are therefore characterized by high predictive uncertainty.
+These sets tend to be larger for inputs that do not conform with the training data and are characterized by high predictive uncertainty.
 
@@
 Here, $\kappa \in \{0,1\}$ is a hyper-parameter and $C_{\theta,\mathbf{y}}(\mathbf{x}_i;\alpha)$ can be interpreted as the probability of label $\mathbf{y}$ being included in the prediction set.
-
 In order to compute this penalty for any black-box model we merely need to perform a single calibration pass through a holdout set $\mathcal{D}_{\text{cal}}$.
@@
 This particular case of conformal prediction is referred to as Split Conformal Prediction (SCP) as it involves splitting the training data into a proper training dataset and a calibration dataset.
-
 In addition to the smooth set size penalty, we have also experimented with the use of a tailored function for $\text{yloss}(\cdot)$ that enforces that only the target label $\mathbf{y}^+$ is included in the prediction set ~\citet{stutz2022learning}.
@@
 \begin{algorithm}
-  \caption{The \textbf{ECCCo} generator}\label{alg:eccco}
+  \caption{The \textit{ECCCo} generator}\label{alg:eccco}
   \begin{algorithmic}[1]
-    \Require $\mathbf{x}, \mathbf{y}^+, M_{\theta}, f, \Lambda, \alpha, \mathcal{D}, T, \eta, n_{\mathcal{B}}$ where $M_{\theta}(\mathbf{x})\neq\mathbf{y}^+$
+    \Require $\mathbf{x}, \mathbf{y}^+, M_{\theta}, f, \Lambda=[\lambda_1,\lambda_2,\lambda_3], \alpha, \mathcal{D}, T, \eta, n_{\mathcal{B}}, n_E$ where $M_{\theta}(\mathbf{x})\neq\mathbf{y}^+$
     \Ensure $\mathbf{x}^\prime$
@@
     \State Generate $\left\{\hat{\mathbf{x}}_{\theta,\mathbf{y}^+}\right\}_{n_{\mathcal{B}}} \gets p_{\theta}(\mathbf{x}_{\mathbf{y}^+})$ \Comment{Generate $n_{\mathcal{B}}$ samples using SGLD (Equation~\ref{eq:sgld}).}
+    \State Store $\hat{\mathbf{X}}^{n_E} \gets \left\{\hat{\mathbf{x}}_{\theta,\mathbf{y}^+}\right\}_{n_{\mathcal{B}}}$ \Comment{Choose $n_E$ lowest-energy samples.}
     \State Run \textit{SCP} for $M_{\theta}$ using $\mathcal{D}$ \Comment{Calibrate model through Split Conformal Prediction.}
@@
     \While{\textit{not converged} or $t < T$} \Comment{For convergence conditions see Appendix~\ref{app:eccco}.}
-    \State $\mathbf{z}^\prime \gets \mathbf{z}^\prime - \eta \nabla_{\mathbf{z}^\prime} \mathcal{L}(\mathbf{z}^\prime,\mathbf{y}^+,\hat{\mathbf{x}}_{\theta, t}; \Lambda, \alpha)$ \Comment{Take gradient step of size $\eta$.}
+    \State $\mathbf{z}^\prime \gets \mathbf{z}^\prime - \eta \nabla_{\mathbf{z}^\prime} \mathcal{L}(\mathbf{z}^\prime,\mathbf{y}^+,\hat{\mathbf{X}}^{n_E}; \Lambda, \alpha)$ \Comment{Take gradient step of size $\eta$.}
     \State $t \gets t+1$
@@
 
-Algorithm~\ref{alg:eccco} describes how exactly \textbf{ECCCo} works.
-For the sake of simplicity and without loss of generality, we limit our attention to generating a single counterfactual $\mathbf{x}^\prime=f(\mathbf{z}^\prime)$ where in contrast to Equation~\ref{eq:eccco} $\mathbf{z}^\prime$ denotes a $1$-dimensional array containing a single counterfactual state.
-That state is initialized by passing the factual $\mathbf{x}$ through the encoder $f^{-1}$ which in our case corresponds to a simple feature transformer, rather than the encoder part of VAE as in \textbf{REVISE}.
-Next, we generate a buffer of $N_{\mathcal{B}}$ conditional samples $\hat{\mathbf{x}}_{\theta}|\mathbf{y}^+$ using SGLD (Equation~\ref{eq:sgld}) and conformalise the model $M_{\theta}$ through Split Conformal Prediction on training data $\mathcal{D}$.
+To provide some further intuition about our objective defined in Equation~\ref{eq:eccco}, Figure~\ref{fig:poc} illustrates how the different components affect the counterfactual search for a synthetic dataset.
+The underlying classifier is a Joint Energy Model (\textit{JEM}) that was trained to predict the output class (`blue' or `orange') and generate class-conditional samples~\citep{grathwohl2020your}.
+We have used four different generator flavours to produce a counterfactual in the `blue' class for a sample from the `orange' class: \textit{Wachter}, which only uses the first penalty ($\lambda_2=\lambda_3=0$); \textit{ECCCo (no CP)}, which involves no set size penalty ($\lambda_3=0$); \textit{ECCCo (no EBM)}, which does not constrain energy ($\lambda_2=0$); and, finally, \textit{ECCCo}, which involves all penalties defined in Equation~\ref{eq:eccco}.
+Arrows indicate (negative) gradients with respect to the objective function at different points in the feature space.
 
-Finally, we search counterfactuals through gradient descent.
-Let $\mathcal{L}(\mathbf{z}^\prime,\mathbf{y}^+,\hat{\mathbf{x}}_{\theta, t}; \Lambda, \alpha)$ denote our loss function defined in Equation~\ref{eq:eccco}.
-Then in each iteration, we first randomly draw $n_{\mathcal{B}}$ samples from the buffer $\mathcal{B}$ before updating the counterfactual state $\mathbf{z}^\prime$ by moving in the negative direction of that loss function.
+While \textit{Wachter} generates a valid counterfactual, it ends up close to the original starting point consistent with its objective.
+\textit{ECCCo (no EBM)} pushes the counterfactual further into the target domain to minimize predictive uncertainty, but the outcome is still not plausible.
+The counterfactual produced by \textit{ECCCo (no CP)} is attracted by the generated samples shown in bright yellow.
+Since the \textit{JEM} has learned the conditional input distribution reasonably well in this case, the counterfactuals are both faithful and plausible.
+Finally, the outcome for \textit{ECCCo} looks similar, but the additional smooth set size penalty leads to somewhat faster convergence.
+
+Algorithm~\ref{alg:eccco} describes how exactly \textit{ECCCo} works.
+For the sake of simplicity and without loss of generality, we limit our attention to generating a single counterfactual $\mathbf{x}^\prime=f(\mathbf{z}^\prime)$.
+That counterfactual state $\mathbf{z}^\prime$ is initialized by passing the factual $\mathbf{x}$ through a simple feature transformer $f^{-1}$.
+Next, we generate $n_{\mathcal{B}}$ conditional samples $\hat{\mathbf{x}}_{\theta,\mathbf{y}^+}$ using SGLD (Equation~\ref{eq:sgld}) and store the $n_E$ instances with the lowest energy.
+We then calibrate the model $M_{\theta}$ through Split Conformal Prediction.
+Finally, we search counterfactuals through gradient descent where $\mathcal{L}(\mathbf{z}^\prime,\mathbf{y}^+,\hat{\mathbf{X}}^{n_E}; \Lambda, \alpha)$ denotes our loss function defined in Equation~\ref{eq:eccco}.
 The search terminates once the convergence criterium is met or the maximum number of iterations $T$ has been exhausted.
-Note that the choice of convergence criterium has important implications on the final counterfactual.
-Additional details concerning Algorithm~\ref{alg:eccco} can be found in Appendix~\ref{app:eccco}.
-
-Figure~\ref{fig:poc} illustrates how ECCCos compare to counterfactuals generated using \textbf{Wachter}, \textbf{Schut} and \textbf{REVISE}.
-The example involves synthetically generated linearly separable data that belong to one of two classes.
-Contours indicate the predicted probabilities of a Joint Energy Model that has been jointly trained to predict the output class and generate inputs~\citet{grathwohl2020your}.
-We have drawn a random sample from the factual class 1 and used each generator to produce a counterfactual in the target class 2.
-Both \textbf{Wachter} and \textbf{Schut} yield valid counterfactuals but fail to achieve plausibility in the sense that the generated counterfactuals are far away from the densely populated region in the target class.
-Conversely, ECCCo yields a faithful and plausible counterfactual in the neighbourhood of the generated conditional samples.
-\textbf{REVISE} fails to yield a valid counterfactual because the underlying surrogate has failed to learn the DGP.
+Note that the choice of convergence criterium has important implications on the final counterfactual which we explain in Appendix~\ref{app:eccco}.
 
@@
 
-\begin{question}[Plausibility]\label{rq:plausibility}
-  How do ECCCos compare to state-of-the-art generators with respect to plausibility?
+\begin{question}[Joint Objective]\label{rq:plausibility}
+  Compared to our benchmark generators, how do ECCCos balance the two key objectives of faithfulness and plausibility?
 \end{question}
 
-We first briefly describe our experimental setup, before presenting our main results.
+The second question is motivated by the intuition that faithfulness and plausibility should coincide for models that have learned plausible explanations of the data.
+Next, we first briefly describe our experimental setup before presenting our main results.
 
@@
 
-To assess and benchmark the performance of ECCCo against the state of the art, we generate multiple counterfactuals for different black-box models and datasets.
-In particular, we compare ECCCo to the following counterfactual generators that were introduced above: firstly; \textbf{Schut}, which works under the premise of minimizing predictive uncertainty; secondly, \textbf{REVISE}, which is state-of-the-art with respect to plausibility; and, finally, \textbf{Wachter}, which serves as our baseline.
-We also consider two variations of our propsed generator: \textbf{ECCCo (no CP)} involves no set size penalty ($\lambda_3=0$ in Equation~\ref{eq:eccco}), while \textbf{ECCCo (no EBM)} does not penalise the distance to samples generated through SGLD ($\lambda_2=0$ in Equation~\ref{eq:eccco}).
-These have been added to gain some sense of the degree to which the two components underlying ECCCo---namely, energy-based modelling (EBM) and conformal prediction (CP)---drive the results.
+To assess and benchmark the performance of our proposed generator against the state of the art, we generate multiple counterfactuals for different models and datasets.
+In particular, we compare \textit{ECCCo} and its variants to the following counterfactual generators that were introduced above: firstly; \textit{Schut}, which works under the premise of minimizing predictive uncertainty; secondly, \textit{REVISE}, which is state-of-the-art with respect to plausibility; and, finally, \textit{Wachter}, which serves as our baseline.
 
 We use both synthetic and real-world datasets from different domains, all of which are publicly available and commonly used to train and benchmark classification algorithms.
-We synthetically generate a dataset containing two \textbf{Linearly Separable} Gaussian clusters ($n=1000$), as well as the well-known \textbf{Circles} ($n=1000$) and \textbf{Moons} ($n=2500$) data.
+We synthetically generate a dataset containing two \textit{Linearly Separable} Gaussian clusters ($n=1000$), as well as the well-known \textit{Circles} ($n=1000$) and \textit{Moons} ($n=2500$) data.
 Since these data are generated by distributions of varying degrees of complexity, they allow us to assess how the generators and our proposed evaluation metrics handle this.
 
-As for real-world data, we follow~\citet{schut2021generating} and use the \textbf{MNIST}~\citep{lecun1998mnist} dataset containing images of handwritten digits such as the example shown above in Figure~\ref{fig:motiv}.
-From the social sciences domain, we include Give Me Some Credit (\textbf{GMSC})~\citep{kaggle2011give}: a tabular dataset that has been studied extensively in the literature on Algorithmic Recourse~\citep{pawelczyk2021carla}.
-It consists of 11 numeric features that can be used to predict the binary outcome variable indicating whether or not retail borrowers experience financial distress.
+As for real-world data, we follow~\citet{schut2021generating} and use the \textit{MNIST}~\citep{lecun1998mnist} dataset containing images of handwritten digits such as the example shown above in Figure~\ref{fig:motiv}.
+From the social sciences domain, we include Give Me Some Credit (\textit{GMSC})~\citep{kaggle2011give}: a tabular dataset that has been studied extensively in the literature on Algorithmic Recourse~\citep{pawelczyk2021carla}.
+It consists of 11 numeric features that can be used to predict the binary outcome variable indicating whether retail borrowers experience financial distress.
 
-For the predictive modelling tasks, we use simple neural networks (\textbf{MLP}) and Joint Energy Models (\textbf{JEM}).
+For the predictive modelling tasks, we use simple neural networks (\textit{MLP}) and Joint Energy Models (\textit{JEM}).
 For the more complex real-world datasets we also use ensembling in each case.
@@
 Specifically, we randomly sample $n^{-}$ times from the subset of individuals for which the given model predicts the non-target class $\mathbf{y}^{-}$ given the current target.
-We set $n^{-}=25$ for all of our synthetic datasets, $n^{-}=10$ for GMSC and $n^{-}=5$ for MNIST.
+We set $n^{-}=25$ for all of our synthetic datasets, $n^{-}=10$ for \textit{GMSC} and $n^{-}=5$ for \textit{MNIST}.
 Full details concerning our parameter choices, training procedures and model performance can be found in Appendix~\ref{app:setup}.
@@
 We have highlighted the best outcome for each model and metric in bold.
-To provide some sense of effect sizes, we have added asterisks to indicate that a given value is at least one ($*$) or two ($**$) standard deviations lower than the baseline (\textbf{Wachter}).
+To provide some sense of effect sizes, we have added asterisks to indicate that a given value is at least one ($*$) or two ($**$) standard deviations lower than the baseline (\textit{Wachter}).
 
-Starting with the high-level results for our Linearly Separable data, we find that ECCCo produces the most faithful counterfactuals for both black-box models.
-This is consistent with our design since ECCCo directly enforces faithfulness through regularization.
-Crucially though, ECCCo also produces the most plausible counterfactuals for both models.
-This dataset is so simple that even the MLP has learned plausible explanations of the input data.
-Zooming in on the granular details for the Linearly Separable data, the results for `ECCCo (no CP)' and `ECCCo (no EBM)' indicate that the positive results are dominated by the effect of quantifying and leveraging the model's generative property (EBM).
+Starting with the high-level results for our \textit{Linearly Separable} data, we find that \textit{ECCCo} produces the most faithful counterfactuals for both black-box models.
+This is consistent with our design since \textit{ECCCo} directly enforces faithfulness through regularization.
+Crucially though, \textit{ECCCo} also produces the most plausible counterfactuals for both models.
+This dataset is so simple that even the \textit{MLP} has learned plausible explanations of the input data.
+Zooming in on the granular details for the \textit{Linearly Separable} data, the results for \textit{ECCCo (no CP)} and \textit{ECCCo (no EBM)} indicate that the positive results are dominated by the effect of quantifying and leveraging the model's generative property (EBM).
 Conformal Prediction alone only leads to marginally improved faithfulness and plausibility relative to the benchmark generators.
 
-The findings for the Moons dataset are broadly in line with the findings so far: for the JEM, ECCCo yields significantly more faithful and plausible counterfactuals than all other generators.
-For the MLP, faithfulness is maintained but counterfactuals are not plausible.
+The findings for the \textit{Moons} dataset are broadly in line with the findings so far: for the \textit{JEM}, \textit{ECCCo} yields significantly more faithful and plausible counterfactuals than all other generators.
+For the \textit{MLP}, faithfulness is maintained but counterfactuals are not plausible.
 This high-level pattern is broadly consistent other more complex datasets and supportive of our narrative, so it is worth highlighting: ECCCos consistently achieve high faithfulness, which---subject to the quality of the model itself---coincides with high plausibility.
-By comparison, \textbf{REVISE} yields the most plausible counterfactuals for the MLP, but it does so at the cost of faithfulness.
-We also observe that the best results for ECCCo are achieved when using both penalties.
-Once again though, the generative component (EBM) has a stronger impact on the positive results for the JEM.
+By comparison, \textit{REVISE} yields the most plausible counterfactuals for the \textit{MLP}, but it does so at the cost of faithfulness.
+We also observe that the best results for \textit{ECCCo} are achieved when using both penalties.
+Once again though, the generative component (EBM) has a stronger impact on the positive results for the \textit{JEM}.
 
-For the Circles data, it appears that \textbf{REVISE} performs well, but we note that it generates valid counterfactuals only half of the time (see Appendix~\ref{app:results} for a complete overview of all evaluation metrics).
-It turns out that in this case, the underlying VAE with default parameters has not adequately learned the data-generating process.
-Of course, it is possible to achieve better generative performance through hyperparameter tuning but this example serves to illustrate that \textbf{REVISE} depends strongly on the quality of the surrogate model.
-Independent of the outcome for \textbf{REVISE}, however, the results do not seem to indicate that ECCCo significantly improves faithfulness and plausibility for the Circles data.
-We think this points to a limitation of our evaluation metrics rather than ECCCo itself: computing average distances fails to account for the `wraparound' effect associated with circular data~\citep{gill2010circular}.
+For the \textit{Circles} data, it appears that \textit{REVISE} performs well, but we note that it generates valid counterfactuals only half of the time (see Appendix~\ref{app:results} for a complete overview of all evaluation metrics).
+It turns out that the underlying VAE with default parameters has not adequately learned the data-generating process.
+Of course, it is possible to achieve better generative performance through hyperparameter tuning but this example serves to illustrate that \textit{REVISE} depends strongly on the quality of the surrogate model.
+Independent of the outcome for \textit{REVISE}, however, the results do not seem to indicate that \textit{ECCCo} significantly improves faithfulness and plausibility for the \textit{Circles} data.
+We think this points to a limitation of our evaluation metrics rather than \textit{ECCCo} itself: computing average distances fails to account for the `wraparound' effect associated with circular data~\citep{gill2010circular}.
 
@@
 Once again the findings indicate that the plausibility of ECCCos is positively correlated with the capacity of the black-box model to distinguish plausible from implausible inputs.
-The case is very clear for MNIST: ECCCos are consistently more faithful than the corresponding counterfactuals produced by any of the benchmark generators and their plausibility gradually improves through ensembling and joint-energy modelling.
-For the JEM Ensemble, ECCCo is essentially on par with \textbf{REVISE} and does significantly better than the baseline generator.
-We also note that ECCCo is the only generator that consistently achieves full validity for all models (Appendix~\ref{app:results}).
-Interestingly, ECCCo also yields lower-cost outcomes than the baseline generator for the JEMs.
+The case is very clear for \textit{MNIST}: ECCCos are consistently more faithful than the corresponding counterfactuals produced by any of the benchmark generators and their plausibility gradually improves through ensembling and joint-energy modelling.
+For the \textit{JEM Ensemble}, \textit{ECCCo} is essentially on par with \textit{REVISE} and does significantly better than the baseline generator.
+We also note that \textit{ECCCo} is the only generator that consistently achieves full validity for all models (Appendix~\ref{app:results}).
 
-For the tabular credit dataset (GMSC) it is inherently challenging to use deep neural networks in order to achieve good discriminative performance~\citep{borisov2021deep,grinsztajn2022why} and discriminative performance~\citep{liu2023goggle}, respectively.
-In order to achieve high plausibility, ECCCo effectively requires classifiers to achieve good performance for both tasks.
-Since this is a challenging task even for Joint Energy Models, it is not surprising to find that even though ECCCo once again achieves state-of-the-art faithfulness, it is outperformed by \textbf{REVISE} and \textbf{Schut} with respect to plausibility.
+For the tabular credit dataset (\textit{GMSC}) it is inherently challenging to use deep neural networks in order to achieve good discriminative performance~\citep{borisov2021deep,grinsztajn2022why} and discriminative performance~\citep{liu2023goggle}, respectively.
+In order to achieve high plausibility, \textit{ECCCo} effectively requires classifiers to achieve good performance for both tasks.
+Since this is a challenging task even for Joint Energy Models, it is not surprising to find that even though \textit{ECCCo} once again achieves state-of-the-art faithfulness, it is outperformed by \textit{REVISE} and \textit{Schut} with respect to plausibility.
 
@@
 To conclude this section, we summarize our findings with reference to the opening questions.
-The results have clearly demonstrated that ECCCo consistently achieves state-of-the-art faithfulness, as it was designed to do (Research Question~\ref{rq:faithfulness}).
-A related important finding is that ECCCo yields highly plausible explanations provided that they faithfully describe model behaviour (Research Question~\ref{rq:plausibility}).
-Our findings here also indicate that ECCCo achieves this result primarily by leveraging the model's generative property.
+The results have clearly demonstrated that \textit{ECCCo} consistently achieves state-of-the-art faithfulness, as it was designed to do (Research Question~\ref{rq:faithfulness}).
+A related important finding is that \textit{ECCCo} yields highly plausible explanations provided that they faithfully describe model behaviour (Research Question~\ref{rq:plausibility}).
+Our findings here also indicate that \textit{ECCCo} achieves this result primarily by leveraging the model's generative property.
 
@@
 Even though we have taken considerable measures to study our proposed methodology carefully, limitations can still be identified.
-In particular, we have found that the performance of ECCCo is sensitive to hyperparameter choices.
+In particular, we have found that the performance of \textit{ECCCo} is sensitive to hyperparameter choices.
 In order to achieve faithfulness, we generally had to penalise the distance from generated samples slightly more than the distance from factual values.
@@
 Furthermore, while our approach is readily applicable to models with gradient access like deep neural networks, more work is needed to generalise our methodology to other popular machine learning models such as gradient-boosted trees.
-Relatedly, common challenges associated with energy-based modelling during our experiments including sensitivity to scale, training instabilities and sensitivity to hyperparameters also apply to ECCCo.
+Relatedly, common challenges associated with energy-based modelling during our experiments including sensitivity to scale, training instabilities and sensitivity to hyperparameters also apply to \textit{ECCCo}.
 
@@
 This work leverages recent advances in energy-based modelling and conformal prediction in the context of Explainable Artificial Intelligence.
-We have proposed a new way to generate counterfactuals that are maximally faithful to the black-model they aim to explain.
-Our proposed counterfactual generator, ECCCo, produces plausible counterfactual if and only if the black-model itself has learned realistic representations of the data, which we demonstrate through rigorous empirical analysis.
+We have proposed a new way to generate counterfactuals that are maximally faithful to the black-box model they aim to explain.
+Our proposed generator, \textit{ECCCo}, produces plausible counterfactuals if and only if the black-box model itself has learned realistic representations of the data, which we demonstrate through rigorous empirical analysis.
 This should enable researchers and practitioners to use counterfactuals in order to discern trustworthy models from unreliable ones.
-While the scope of this work limits its generalizability, we believe that ECCCo offers a solid baseline for future work on faithful CE.
+While the scope of this work limits its generalizability, we believe that \textit{ECCCo} offers a solid baseline for future work on faithful CE.
 

### 2023-05-16 paper/paper.tex
nearly there

@@
 For the \textit{MLP}, faithfulness is maintained but counterfactuals are not plausible.
-This high-level pattern is broadly consistent other more complex datasets and supportive of our narrative, so it is worth highlighting: ECCCos consistently achieve high faithfulness, which---subject to the quality of the model itself---coincides with high plausibility.
+This high-level pattern is broadly consistent with other more complex datasets and supportive of our narrative, so it is worth highlighting: ECCCos consistently achieve high faithfulness, which---subject to the quality of the model itself---coincides with high plausibility.
 By comparison, \textit{REVISE} yields the most plausible counterfactuals for the \textit{MLP}, but it does so at the cost of faithfulness.
@@
 To conclude this section, we summarize our findings with reference to the opening questions.
-The results have clearly demonstrated that \textit{ECCCo} consistently achieves state-of-the-art faithfulness, as it was designed to do (Research Question~\ref{rq:faithfulness}).
+The results clearly demonstrate that \textit{ECCCo} consistently achieves state-of-the-art faithfulness, as it was designed to do (Research Question~\ref{rq:faithfulness}).
 A related important finding is that \textit{ECCCo} yields highly plausible explanations provided that they faithfully describe model behaviour (Research Question~\ref{rq:plausibility}).

### 2023-05-16 paper/paper.tex
only proof-reading now

@@
 
-Thus, we believe that our work opens avenues for researchers and practitioners seeking tools to better distinguish trustworthy from unreliable models.
+To our knowledge, this is the first venture in this direction for generating faithful counterfactuals.
+Thus, we anticipate that ECCCos can serve as a baseline for future research.
+We believe that our work opens avenues for researchers and practitioners seeking tools to better distinguish trustworthy from unreliable models.
 
@@
 Our default choice for $\text{dist}(\cdot)$ is the L1 Norm due to its sparsity-inducing properties.
-The second penalty term involving $\lambda_2$ constrains the energy of the generated counterfactual by penalising its distance from the lowest-energy conditional samples as defined in Equation~\ref{eq:faith}.
-Intuitively, this component induces faithfulness which coincides with plausibility to the extent that the model $M_{\theta}$ has learned the true posterior conditional distribution of inputs: $p_{\theta}(\mathbf{x}_{\mathbf{y}^+}) \rightarrow p(\mathbf{x}_{\mathbf{y}^+})$.
-
-The third and final penalty term involving $\lambda_3$ introduces a new but familiar concept: it ensures that the generated counterfactual is associated with low predictive uncertainty.
+The second penalty term involving $\lambda_2$ induces faithfulness by constraining the energy of the generated counterfactual where $\text{unfaith}(\cdot)$ corresponds to the metric defined in Equation~\ref{eq:faith}.
+The third and final penalty term involving $\lambda_3$ introduces a new concept: it ensures that the generated counterfactual is associated with low predictive uncertainty.
 As mentioned above,~\citet{schut2021generating} have shown that plausible counterfactuals can be generated implicitly through predictive uncertainty minimization.
@@
 Zooming in on the granular details for the \textit{Linearly Separable} data, the results for \textit{ECCCo (no CP)} and \textit{ECCCo (no EBM)} indicate that the positive results are dominated by the effect of quantifying and leveraging the model's generative property (EBM).
-Conformal Prediction alone only leads to marginally improved faithfulness and plausibility relative to the benchmark generators.
+Conformal Prediction alone only leads to marginally improved faithfulness and plausibility.
 
-The findings for the \textit{Moons} dataset are broadly in line with the findings so far: for the \textit{JEM}, \textit{ECCCo} yields significantly more faithful and plausible counterfactuals than all other generators.
+The findings for the \textit{Moons} dataset are broadly in line with the findings so far: for the \textit{JEM}, \textit{ECCCo} yields substantially more faithful and plausible counterfactuals than all other generators.
 For the \textit{MLP}, faithfulness is maintained but counterfactuals are not plausible.
@@
 
-For the \textit{Circles} data, it appears that \textit{REVISE} performs well, but we note that it generates valid counterfactuals only half of the time (see Appendix~\ref{app:results} for a complete overview of all evaluation metrics).
-It turns out that the underlying VAE with default parameters has not adequately learned the data-generating process.
-Of course, it is possible to achieve better generative performance through hyperparameter tuning but this example serves to illustrate that \textit{REVISE} depends strongly on the quality of the surrogate model.
-Independent of the outcome for \textit{REVISE}, however, the results do not seem to indicate that \textit{ECCCo} significantly improves faithfulness and plausibility for the \textit{Circles} data.
+For the \textit{Circles} data, it appears that \textit{REVISE} performs well, but we note that it generates valid counterfactuals only half of the time (see Appendix~\ref{app:results} for a complete overview including additional common evaluation metrics).
+The underlying VAE with default parameters has not adequately learned the data-generating process.
+Of course, it is possible to improve generative performance through hyperparameter tuning but this example serves to illustrate that \textit{REVISE} depends on the quality of its surrogate.
+Independent of the outcome for \textit{REVISE}, however, the results do not seem to indicate that \textit{ECCCo} substantially improves faithfulness and plausibility for the \textit{Circles} data.
 We think this points to a limitation of our evaluation metrics rather than \textit{ECCCo} itself: computing average distances fails to account for the `wraparound' effect associated with circular data~\citep{gill2010circular}.
@@
 Once again the findings indicate that the plausibility of ECCCos is positively correlated with the capacity of the black-box model to distinguish plausible from implausible inputs.
-The case is very clear for \textit{MNIST}: ECCCos are consistently more faithful than the corresponding counterfactuals produced by any of the benchmark generators and their plausibility gradually improves through ensembling and joint-energy modelling.
-For the \textit{JEM Ensemble}, \textit{ECCCo} is essentially on par with \textit{REVISE} and does significantly better than the baseline generator.
-We also note that \textit{ECCCo} is the only generator that consistently achieves full validity for all models (Appendix~\ref{app:results}).
+The case is very clear for \textit{MNIST}: ECCCos are consistently more faithful than the counterfactuals produced by our benchmark generators and their plausibility gradually improves through ensembling and joint-energy modelling.
+Interestingly, faithfulness also gradually improves for \textit{REVISE}.
+This indicates that as our models improve, their generative capacity approaches that of the surrogate VAE used by \textit{REVISE}.
+The VAE still outperforms our classifiers in this regard, as evident from the fact that \textit{ECCCo} never quite reaches the same level of plausibility as \textit{REVISE}.
+With reference to Appendix~\ref{app:results} we note that the results for \textit{Schut} need to be discounted as it rarely produces valid counterfactuals for \textit{MNIST}.
+Relatedly, we find that \textit{ECCCo} is the only generator that consistently achieves full validity.
+Finally, it is worth noting that \textit{ECCCo} produces counterfactual images with the lowest average predictive uncertainty for all models.
 
@@
 A related important finding is that \textit{ECCCo} yields highly plausible explanations provided that they faithfully describe model behaviour (Research Question~\ref{rq:plausibility}).
-Our findings here also indicate that \textit{ECCCo} achieves this result primarily by leveraging the model's generative property.
+\textit{ECCCo} achieves this result primarily by leveraging the model's generative property.
 
@@
 In order to achieve faithfulness, we generally had to penalise the distance from generated samples slightly more than the distance from factual values.
-This choice is associated with relatively higher costs to individuals since the proposed recourse typically involves more substantial feature changes than for our benchmark generators.
 
-Conversely, we have not found that disproportionately strongly penalising prediction set sizes had any discernable effect.
-Our results indicate that Conformal Prediction alone is often not sufficient to achieve faithfulness and plausibility, although we acknowledge that this needs to be investigated more thoroughly through future work.
+Conversely, we have not found that strongly penalising prediction set sizes had any discernable effect.
+Our results indicate that CP alone is often not sufficient to achieve faithfulness and plausibility, although we acknowledge that this needs to be investigated more thoroughly through future work.
 
-Furthermore, while our approach is readily applicable to models with gradient access like deep neural networks, more work is needed to generalise our methodology to other popular machine learning models such as gradient-boosted trees.
-Relatedly, common challenges associated with energy-based modelling during our experiments including sensitivity to scale, training instabilities and sensitivity to hyperparameters also apply to \textit{ECCCo}.
+While our approach is readily applicable to models with gradient access like deep neural networks, more work is needed to generalise it to other machine learning models such as decision trees.
+Relatedly, common challenges associated with energy-based modelling including sensitivity to scale, training instabilities and sensitivity to hyperparameters also apply to \textit{ECCCo}.
 

### 2023-05-16 paper/paper.tex
minor things

@@
 \author{%
-  Patrick Altmeyer\thanks{Use footnote for providing further information
-    about author (webpage, alternative address)---\emph{not} for acknowledging
-    funding agencies.} \\
+  Patrick Altmeyer\thanks{See also: https://www.paltmeyer.com/} \\
   Faculty of Electrical Engineering, Mathematics and Computer Science\\
@@
   \texttt{p.altmeyer@tudelft.nl} \\
-  % examples of more authors
-  % \And
-  % Coauthor \\
-  % Affiliation \\
-  % Address \\
-  % \texttt{email} \\
-  % \AND
-  % Coauthor \\
-  % Affiliation \\
-  % Address \\
-  % \texttt{email} \\
-  % \And
-  % Coauthor \\
-  % Affiliation \\
-  % Address \\
-  % \texttt{email} \\
-  % \And
-  % Coauthor \\
-  % Affiliation \\
-  % Address \\
-  % \texttt{email} \\
+  \And
+  Mojtaba Farmanbar \\
+  ING Netherlands \\
+  1102 CT Amsterdam, The Netherlands \\
+  \texttt{mojtaba.farmanbar@ing.com} \\
+  \AND
+  Arie van Deursen \\
+  Delft University of Technology\\
+  2628 XE Delft, The Netherlands \\
+  \texttt{arie.vandeursen@tudelft.nl} \\
+  \And
+  Cynthia C. S. Liem \\
+  Delft University of Technology\\
+  2628 XE Delft, The Netherlands \\
+  \texttt{c.c.s.liem@tudelft.nl} \\
 }
@@
 We formalise this notion of faithfulness through the introduction of a tailored evaluation metric and propose a novel algorithmic framework for generating \textbf{E}nergy-\textbf{C}onstrained \textbf{C}onformal \textbf{Co}unterfactuals (ECCCos) that are only as plausible as the model permits.
-Through extensive empirical studies involving multiple synthetic and real-world datasets, we demonstrate that ECCCos reconcile the need for plausibility and faithfulness.
-In particular, we show that it is possible to achieve state-of-the-art plausibility for models with gradient access without the need for surrogate models.
-To do so, our framework relies solely on properties defining the black-box model itself by leveraging recent advances in energy-based modelling and conformal prediction.
+Through extensive empirical studies, we demonstrate that ECCCos reconcile the need for faithfulness and plausibility.
+In particular, we show that for models with gradient access, it is possible to achieve state-of-the-art performance without the need for surrogate models.
+To do so, our framework relies solely on properties defining the black-box model itself by leveraging recent advances in Energy-Based Modelling and Conformal Prediction.
 To our knowledge, this is the first venture in this direction for generating faithful Counterfactual Explanations.
@@
 While $\mathbf{x}_J$ is only guaranteed to distribute as $p_{\theta}(\mathbf{x}|\mathbf{y}^*)$ if $\epsilon \rightarrow 0$ and $J \rightarrow \infty$, the bias introduced for a small finite $\epsilon$ is negligible in practice \citep{murphy2023probabilistic,grathwohl2020your}.
-Appendix~\ref{app-jem} provides additional implementation details for any tasks related to energy-based modelling.
+Appendix~\ref{app-jem} provides additional implementation details for any tasks related to Energy-Based Modelling.
 
@@
 Consequently, CP removes the restriction on the family of predictive models, at the small cost of reserving a subset of the available data for calibration.
-This particular case of conformal prediction is referred to as Split Conformal Prediction (SCP) as it involves splitting the training data into a proper training dataset and a calibration dataset.
+This particular case of Conformal Prediction is referred to as Split Conformal Prediction (SCP) as it involves splitting the training data into a proper training dataset and a calibration dataset.
 In addition to the smooth set size penalty, we have also experimented with the use of a tailored function for $\text{yloss}(\cdot)$ that enforces that only the target label $\mathbf{y}^+$ is included in the prediction set ~\citet{stutz2022learning}.
@@
 While our approach is readily applicable to models with gradient access like deep neural networks, more work is needed to generalise it to other machine learning models such as decision trees.
-Relatedly, common challenges associated with energy-based modelling including sensitivity to scale, training instabilities and sensitivity to hyperparameters also apply to \textit{ECCCo}.
+Relatedly, common challenges associated with Energy-Based Modelling including sensitivity to scale, training instabilities and sensitivity to hyperparameters also apply to \textit{ECCCo}.
 
@@
 
-This work leverages recent advances in energy-based modelling and conformal prediction in the context of Explainable Artificial Intelligence.
+This work leverages recent advances in Energy-Based Modelling and Conformal Prediction in the context of Explainable Artificial Intelligence.
 We have proposed a new way to generate counterfactuals that are maximally faithful to the black-box model they aim to explain.

### 2023-05-16 notebooks/poc.qmd
done

@@
             _X[1,:],_X[2,:], color=col_pal[end-1], shape=:star5, 
-            ms=10, label="x̂|$target", alpha=0.1
+            ms=10, label="x̂|$target", alpha=0.5
         )

### 2023-05-16 paper/paper.tex
done

@@
   \item We introduce a novel algorithmic approach for generating Energy-Constrained Conformal Counterfactuals (ECCCos) in Section~\ref{meth}.
-  \item We provide extensive empirical evidence demonstrating that ECCCos faithfully explain model behaviour without sacrificing plausibility (Section~\ref{emp}).
+  \item We provide extensive empirical evidence demonstrating that ECCCos faithfully explain model behaviour and attain plausibility only when appropriate (Section~\ref{emp}).
 \end{itemize}
@@
 
-Here $\text{yloss}$ denotes the primary loss function, $f(\cdot)$ is a function that maps from the counterfactual state space to the feature space and $\text{cost}$ is either a single penalty or a collection of penalties that are used to impose constraints through regularization.
+Here $\text{yloss}(\cdot)$ denotes the primary loss function, $f(\cdot)$ is a function that maps from the counterfactual state space to the feature space and $\text{cost}(\cdot)$ is either a single penalty or a collection of penalties that are used to impose constraints through regularization.
 Equation~\ref{eq:general} restates the baseline approach to gradient-based counterfactual search proposed by~\citet{wachter2017counterfactual} in general form as introduced by~\citet{altmeyer2023endogenous}.
-To explicitly account for the multiplicity of explanations $\mathbf{Z}^\prime=\{ \mathbf{z}_l\}_L$ denotes an $L$-dimensional array of counterfactual states.
+To explicitly account for the multiplicity of explanations, $\mathbf{Z}^\prime=\{ \mathbf{z}_l\}_L$ denotes an $L$-dimensional array of counterfactual states.
 
@@
 To learn the latent embedding, they propose using a generative model such as a Variational Autoencoder (VAE).
-Provided the surrogate model is well-trained, their proposed approach called \textit{REVISE} can yield plausible explanations.
+Provided the surrogate model is well-specified, their proposed approach called \textit{REVISE} can yield plausible explanations.
 Others have proposed similar approaches: \citet{dombrowski2021diffeomorphic} traverse the base space of a normalizing flow to solve Equation~\ref{eq:general}; \citet{poyiadzi2020face} use density estimators ($\hat{p}: \mathcal{X} \mapsto [0,1]$) to constrain the counterfactuals to dense regions in the feature space; and, finally, \citet{karimi2021algorithmic} assume knowledge about the structural causal model that generates the data.
@@
 Pointing to the undesirable engineering overhead induced by surrogate models, they propose that we rely on the implicit minimisation of predictive uncertainty instead.
-Their proposed methodology, which we will refer to as \textit{Schut}, solves Equation~\ref{eq:general} by greedily applying JSMA in the feature space with standard cross-entropy loss and no penalty at all.
+Their proposed methodology, which we will refer to as \textit{Schut}, solves Equation~\ref{eq:general} by greedily applying Jacobian-Based Saliency Map Attacks (JSMA) in the feature space with cross-entropy loss and no penalty at all.
 The authors demonstrate theoretically and empirically that their approach yields counterfactuals for which the model $M_{\theta}$ predicts the target label $\mathbf{y}^+$ with high confidence.
@@
 where $\mathbf{r}_j \sim \mathcal{N}(\mathbf{0},\mathbf{I})$ is the stochastic term and the step-size $\epsilon$ is typically polynomially decayed~\citep{welling2011bayesian}.
-The term $\mathcal{E}(\mathbf{x}_j|\mathbf{y}^+)$ denotes the model energy conditioned on the target class label $\mathbf{y}^+$ which we specify as the negative logit corresponding to the target class label $\mathbf{y}^*$.
+The term $\mathcal{E}(\mathbf{x}_j|\mathbf{y}^+)$ denotes the model energy conditioned on the target class label $\mathbf{y}^+$ which we specify as the negative logit corresponding to that label~\citep{grathwohl2020your}.
 To allow for faster sampling, we follow the common practice of choosing the step-size $\epsilon$ and the standard deviation of $\mathbf{r}_j$ separately.
-While $\mathbf{x}_J$ is only guaranteed to distribute as $p_{\theta}(\mathbf{x}|\mathbf{y}^*)$ if $\epsilon \rightarrow 0$ and $J \rightarrow \infty$, the bias introduced for a small finite $\epsilon$ is negligible in practice \citep{murphy2023probabilistic,grathwohl2020your}.
+While $\mathbf{x}_J$ is only guaranteed to distribute as $p_{\theta}(\mathbf{x}|\mathbf{y}^+)$ if $\epsilon \rightarrow 0$ and $J \rightarrow \infty$, the bias introduced for a small finite $\epsilon$ is negligible in practice \citep{murphy2023probabilistic,grathwohl2020your}.
 Appendix~\ref{app-jem} provides additional implementation details for any tasks related to Energy-Based Modelling.
@@
 Generating multiple samples using SGLD thus yields an empirical distribution $\hat{\mathbf{X}}_{\theta,\mathbf{y}^+}$ that approximates what the model has learned about the input data.
-While in the context of Energy-Based Modelling, this is usually done during training, we propose to repurpose this approach during inference in order to evaluate and generate faithful model explanations.
+While in the context of EBM, this is usually done during training, we propose to repurpose this approach during inference in order to evaluate and generate faithful model explanations.
 
@@
 where $\mathbf{x}^{\prime}$ denotes the counterfactual and $\mathbf{X}_{\mathbf{y}^+}$ is a subsample of the training data in the target class $\mathbf{y}^+$.
-By averaging over multiple samples in this manner, we avoid the need to make any assumption about the nearest neighbour of $\mathbf{x}^{\prime}$.
+By averaging over multiple samples in this manner, we avoid the risk that the nearest neighbour of $\mathbf{x}^{\prime}$ itself is not plausible according to Definition~\ref{def:plausible} (e.g an outlier).
 
@@
   \begin{aligned}
-    \text{unfaith}(\mathbf{x}^{\prime},\hat{\mathbf{X}}^{n_E}) = \frac{1}{\lvert \hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^+}\rvert} \sum_{\mathbf{x} \in \hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^+}} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
+    \text{unfaith}(\mathbf{x}^{\prime},\hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^+}) = \frac{1}{\lvert \hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^+}\rvert} \sum_{\mathbf{x} \in \hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^+}} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
   \end{aligned}
@@
   \mathbf{Z}^\prime= \arg \min_{\mathbf{Z}^\prime \in \mathcal{Z}^M}  &\{  {\text{yloss}(M_{\theta}(f(\mathbf{Z}^\prime)),\mathbf{y}^+)}+ \lambda_{1} {\text{dist}(f(\mathbf{Z}^\prime),\mathbf{x}) } \\
-  &+ \lambda_2 \text{unfaith}(f(\mathbf{Z}^\prime),\hat{\mathbf{X}}^{n_E}) + \lambda_3 \Omega(C_{\theta}(f(\mathbf{Z}^\prime);\alpha)) \} 
+  &+ \lambda_2 \text{unfaith}(f(\mathbf{Z}^\prime),\hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^+}) + \lambda_3 \Omega(C_{\theta}(f(\mathbf{Z}^\prime);\alpha)) \} 
   \end{aligned} 
@@
 This particular case of Conformal Prediction is referred to as Split Conformal Prediction (SCP) as it involves splitting the training data into a proper training dataset and a calibration dataset.
-In addition to the smooth set size penalty, we have also experimented with the use of a tailored function for $\text{yloss}(\cdot)$ that enforces that only the target label $\mathbf{y}^+$ is included in the prediction set ~\citet{stutz2022learning}.
+In addition to the smooth set size penalty, we have also experimented with the use of a tailored function for $\text{yloss}(\cdot)$ that enforces that only the target label $\mathbf{y}^+$ is included in the prediction set~\citep{stutz2022learning}.
 Further details are provided in Appendix~\ref{app-cp}.
@@
     \State Generate $\left\{\hat{\mathbf{x}}_{\theta,\mathbf{y}^+}\right\}_{n_{\mathcal{B}}} \gets p_{\theta}(\mathbf{x}_{\mathbf{y}^+})$ \Comment{Generate $n_{\mathcal{B}}$ samples using SGLD (Equation~\ref{eq:sgld}).}
-    \State Store $\hat{\mathbf{X}}^{n_E} \gets \left\{\hat{\mathbf{x}}_{\theta,\mathbf{y}^+}\right\}_{n_{\mathcal{B}}}$ \Comment{Choose $n_E$ lowest-energy samples.}
+    \State Store $\hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^+} \gets \left\{\hat{\mathbf{x}}_{\theta,\mathbf{y}^+}\right\}_{n_{\mathcal{B}}}$ \Comment{Choose $n_E$ lowest-energy samples.}
     \State Run \textit{SCP} for $M_{\theta}$ using $\mathcal{D}$ \Comment{Calibrate model through Split Conformal Prediction.}
@@
     \While{\textit{not converged} or $t < T$} \Comment{For convergence conditions see Appendix~\ref{app:eccco}.}
-    \State $\mathbf{z}^\prime \gets \mathbf{z}^\prime - \eta \nabla_{\mathbf{z}^\prime} \mathcal{L}(\mathbf{z}^\prime,\mathbf{y}^+,\hat{\mathbf{X}}^{n_E}; \Lambda, \alpha)$ \Comment{Take gradient step of size $\eta$.}
+    \State $\mathbf{z}^\prime \gets \mathbf{z}^\prime - \eta \nabla_{\mathbf{z}^\prime} \mathcal{L}(\mathbf{z}^\prime,\mathbf{y}^+,\hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^+}; \Lambda, \alpha)$ \Comment{Take gradient step of size $\eta$.}
     \State $t \gets t+1$
@@
 The underlying classifier is a Joint Energy Model (\textit{JEM}) that was trained to predict the output class (`blue' or `orange') and generate class-conditional samples~\citep{grathwohl2020your}.
-We have used four different generator flavours to produce a counterfactual in the `blue' class for a sample from the `orange' class: \textit{Wachter}, which only uses the first penalty ($\lambda_2=\lambda_3=0$); \textit{ECCCo (no CP)}, which involves no set size penalty ($\lambda_3=0$); \textit{ECCCo (no EBM)}, which does not constrain energy ($\lambda_2=0$); and, finally, \textit{ECCCo}, which involves all penalties defined in Equation~\ref{eq:eccco}.
+We have used four different generator flavours to produce a counterfactual in the `blue' class for a sample from the `orange' class: \textit{Wachter}, which only uses the first penalty ($\lambda_2=\lambda_3=0$); \textit{ECCCo (no EBM)}, which does not constrain energy ($\lambda_2=0$); \textit{ECCCo (no CP)}, which involves no set size penalty ($\lambda_3=0$); and, finally, \textit{ECCCo}, which involves all penalties defined in Equation~\ref{eq:eccco}.
 Arrows indicate (negative) gradients with respect to the objective function at different points in the feature space.
@@
 For the sake of simplicity and without loss of generality, we limit our attention to generating a single counterfactual $\mathbf{x}^\prime=f(\mathbf{z}^\prime)$.
-That counterfactual state $\mathbf{z}^\prime$ is initialized by passing the factual $\mathbf{x}$ through a simple feature transformer $f^{-1}$.
+The counterfactual state $\mathbf{z}^\prime$ is initialized by passing the factual $\mathbf{x}$ through a simple feature transformer $f^{-1}$.
 Next, we generate $n_{\mathcal{B}}$ conditional samples $\hat{\mathbf{x}}_{\theta,\mathbf{y}^+}$ using SGLD (Equation~\ref{eq:sgld}) and store the $n_E$ instances with the lowest energy.
 We then calibrate the model $M_{\theta}$ through Split Conformal Prediction.
-Finally, we search counterfactuals through gradient descent where $\mathcal{L}(\mathbf{z}^\prime,\mathbf{y}^+,\hat{\mathbf{X}}^{n_E}; \Lambda, \alpha)$ denotes our loss function defined in Equation~\ref{eq:eccco}.
+Finally, we search counterfactuals through gradient descent where $\mathcal{L}(\mathbf{z}^\prime,\mathbf{y}^+,\hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^+}; \Lambda, \alpha)$ denotes our loss function defined in Equation~\ref{eq:eccco}.
 The search terminates once the convergence criterium is met or the maximum number of iterations $T$ has been exhausted.
@@
 
-\begin{question}[Joint Objective]\label{rq:plausibility}
+\begin{question}[Balancing Objectives]\label{rq:plausibility}
   Compared to our benchmark generators, how do ECCCos balance the two key objectives of faithfulness and plausibility?
@@
 For the more complex real-world datasets we also use ensembling in each case.
-Both joint-energy modelling and ensembling are associated with generative properties and adversarial robustness, so we expect this to be positively correlated with the plausibility of ECCCos.
-To account for stochasticity, we generate multiple counterfactuals for each possible target class, generator, model and dataset.
+Both joint-energy modelling and ensembling have been associated with improved generative properties and adversarial robustness~\citep{grathwohl2020your,lakshminarayanan2016simple}, so we expect this to be positively correlated with the plausibility of ECCCos.
+To account for stochasticity, we generate multiple counterfactuals for each target class, generator, model and dataset.
 Specifically, we randomly sample $n^{-}$ times from the subset of individuals for which the given model predicts the non-target class $\mathbf{y}^{-}$ given the current target.
@@
 Table~\ref{tab:results-synthetic} shows the key results for the synthetic datasets separated by model (first column) and generator (second column).
-The numerical columns show the average values of our key evaluation metrics computed across all counterfactuals.
-Standard deviations are shown in parentheses.
+The numerical columns show sample averages and standard deviations of our key evaluation metrics computed across all counterfactuals.
 We have highlighted the best outcome for each model and metric in bold.
@@
 
-For the tabular credit dataset (\textit{GMSC}) it is inherently challenging to use deep neural networks in order to achieve good discriminative performance~\citep{borisov2021deep,grinsztajn2022why} and discriminative performance~\citep{liu2023goggle}, respectively.
+For the tabular credit dataset (\textit{GMSC}) it is inherently challenging to use deep neural networks in order to achieve good discriminative performance~\citep{borisov2021deep,grinsztajn2022why} and generative performance~\citep{liu2023goggle}, respectively.
 In order to achieve high plausibility, \textit{ECCCo} effectively requires classifiers to achieve good performance for both tasks.
@@
 We have proposed a new way to generate counterfactuals that are maximally faithful to the black-box model they aim to explain.
-Our proposed generator, \textit{ECCCo}, produces plausible counterfactuals if and only if the black-box model itself has learned realistic representations of the data, which we demonstrate through rigorous empirical analysis.
+Our proposed generator, \textit{ECCCo}, produces plausible counterfactuals if and only if the black-box model itself has learned realistic explanations for the data, which we have demonstrated through rigorous empirical analysis.
 This should enable researchers and practitioners to use counterfactuals in order to discern trustworthy models from unreliable ones.
-While the scope of this work limits its generalizability, we believe that \textit{ECCCo} offers a solid baseline for future work on faithful CE.
+While the scope of this work limits its generalizability, we believe that \textit{ECCCo} offers a solid baseline for future work on faithful Counterfactual Explanations.
 
@@
 
-\subsection{Conformal Prediction}\label{app:eccco}
+\subsection{ECCCo}\label{app:eccco}
 

### 2023-05-17 paper/paper.tex
work on appendix

--- 2023-05-16 paper/paper.tex
@@
 
-To our knowledge, this is the first venture in this direction for generating faithful counterfactuals.
-Thus, we anticipate that ECCCos can serve as a baseline for future research.
-We believe that our work opens avenues for researchers and practitioners seeking tools to better distinguish trustworthy from unreliable models.
+Thus, we believe that our work opens avenues for researchers and practitioners seeking tools to better distinguish trustworthy from unreliable models.
 
@@
 While $\mathbf{x}_J$ is only guaranteed to distribute as $p_{\theta}(\mathbf{x}|\mathbf{y}^*)$ if $\epsilon \rightarrow 0$ and $J \rightarrow \infty$, the bias introduced for a small finite $\epsilon$ is negligible in practice \citep{murphy2023probabilistic,grathwohl2020your}.
-Appendix~\ref{app-jem} provides additional implementation details for any tasks related to energy-based modelling.
+Appendix~\ref{app:jem} provides additional implementation details for any tasks related to energy-based modelling.
 
@@
 Our default choice for $\text{dist}(\cdot)$ is the L1 Norm due to its sparsity-inducing properties.
-The second penalty term involving $\lambda_2$ induces faithfulness by constraining the energy of the generated counterfactual where $\text{unfaith}(\cdot)$ corresponds to the metric defined in Equation~\ref{eq:faith}.
-The third and final penalty term involving $\lambda_3$ introduces a new concept: it ensures that the generated counterfactual is associated with low predictive uncertainty.
+The second penalty term involving $\lambda_2$ constrains the energy of the generated counterfactual by penalising its distance from the lowest-energy conditional samples as defined in Equation~\ref{eq:faith}.
+Intuitively, this component induces faithfulness which coincides with plausibility to the extent that the model $M_{\theta}$ has learned the true posterior conditional distribution of inputs: $p_{\theta}(\mathbf{x}_{\mathbf{y}^+}) \rightarrow p(\mathbf{x}_{\mathbf{y}^+})$.
+
+The third and final penalty term involving $\lambda_3$ introduces a new but familiar concept: it ensures that the generated counterfactual is associated with low predictive uncertainty.
 As mentioned above,~\citet{schut2021generating} have shown that plausible counterfactuals can be generated implicitly through predictive uncertainty minimization.
@@
 In addition to the smooth set size penalty, we have also experimented with the use of a tailored function for $\text{yloss}(\cdot)$ that enforces that only the target label $\mathbf{y}^+$ is included in the prediction set ~\citet{stutz2022learning}.
-Further details are provided in Appendix~\ref{app-cp}.
+Further details are provided in Appendix~\ref{app:cp}.
 
@@
 Zooming in on the granular details for the \textit{Linearly Separable} data, the results for \textit{ECCCo (no CP)} and \textit{ECCCo (no EBM)} indicate that the positive results are dominated by the effect of quantifying and leveraging the model's generative property (EBM).
-Conformal Prediction alone only leads to marginally improved faithfulness and plausibility.
+Conformal Prediction alone only leads to marginally improved faithfulness and plausibility relative to the benchmark generators.
 
-The findings for the \textit{Moons} dataset are broadly in line with the findings so far: for the \textit{JEM}, \textit{ECCCo} yields substantially more faithful and plausible counterfactuals than all other generators.
+The findings for the \textit{Moons} dataset are broadly in line with the findings so far: for the \textit{JEM}, \textit{ECCCo} yields significantly more faithful and plausible counterfactuals than all other generators.
 For the \textit{MLP}, faithfulness is maintained but counterfactuals are not plausible.
@@
 
-For the \textit{Circles} data, it appears that \textit{REVISE} performs well, but we note that it generates valid counterfactuals only half of the time (see Appendix~\ref{app:results} for a complete overview including additional common evaluation metrics).
-The underlying VAE with default parameters has not adequately learned the data-generating process.
-Of course, it is possible to improve generative performance through hyperparameter tuning but this example serves to illustrate that \textit{REVISE} depends on the quality of its surrogate.
-Independent of the outcome for \textit{REVISE}, however, the results do not seem to indicate that \textit{ECCCo} substantially improves faithfulness and plausibility for the \textit{Circles} data.
+For the \textit{Circles} data, it appears that \textit{REVISE} performs well, but we note that it generates valid counterfactuals only half of the time (see Appendix~\ref{app:results} for a complete overview of all evaluation metrics).
+It turns out that the underlying VAE with default parameters has not adequately learned the data-generating process.
+Of course, it is possible to achieve better generative performance through hyperparameter tuning but this example serves to illustrate that \textit{REVISE} depends strongly on the quality of the surrogate model.
+Independent of the outcome for \textit{REVISE}, however, the results do not seem to indicate that \textit{ECCCo} significantly improves faithfulness and plausibility for the \textit{Circles} data.
 We think this points to a limitation of our evaluation metrics rather than \textit{ECCCo} itself: computing average distances fails to account for the `wraparound' effect associated with circular data~\citep{gill2010circular}.
@@
 Once again the findings indicate that the plausibility of ECCCos is positively correlated with the capacity of the black-box model to distinguish plausible from implausible inputs.
-The case is very clear for \textit{MNIST}: ECCCos are consistently more faithful than the counterfactuals produced by our benchmark generators and their plausibility gradually improves through ensembling and joint-energy modelling.
-Interestingly, faithfulness also gradually improves for \textit{REVISE}.
-This indicates that as our models improve, their generative capacity approaches that of the surrogate VAE used by \textit{REVISE}.
-The VAE still outperforms our classifiers in this regard, as evident from the fact that \textit{ECCCo} never quite reaches the same level of plausibility as \textit{REVISE}.
-With reference to Appendix~\ref{app:results} we note that the results for \textit{Schut} need to be discounted as it rarely produces valid counterfactuals for \textit{MNIST}.
-Relatedly, we find that \textit{ECCCo} is the only generator that consistently achieves full validity.
-Finally, it is worth noting that \textit{ECCCo} produces counterfactual images with the lowest average predictive uncertainty for all models.
+The case is very clear for \textit{MNIST}: ECCCos are consistently more faithful than the corresponding counterfactuals produced by any of the benchmark generators and their plausibility gradually improves through ensembling and joint-energy modelling.
+For the \textit{JEM Ensemble}, \textit{ECCCo} is essentially on par with \textit{REVISE} and does significantly better than the baseline generator.
+We also note that \textit{ECCCo} is the only generator that consistently achieves full validity for all models (Appendix~\ref{app:results}).
 
@@
 A related important finding is that \textit{ECCCo} yields highly plausible explanations provided that they faithfully describe model behaviour (Research Question~\ref{rq:plausibility}).
-\textit{ECCCo} achieves this result primarily by leveraging the model's generative property.
+Our findings here also indicate that \textit{ECCCo} achieves this result primarily by leveraging the model's generative property.
 
@@
 In order to achieve faithfulness, we generally had to penalise the distance from generated samples slightly more than the distance from factual values.
+This choice is associated with relatively higher costs to individuals since the proposed recourse typically involves more substantial feature changes than for our benchmark generators.
 
-Conversely, we have not found that strongly penalising prediction set sizes had any discernable effect.
-Our results indicate that CP alone is often not sufficient to achieve faithfulness and plausibility, although we acknowledge that this needs to be investigated more thoroughly through future work.
+Conversely, we have not found that disproportionately strongly penalising prediction set sizes had any discernable effect.
+Our results indicate that Conformal Prediction alone is often not sufficient to achieve faithfulness and plausibility, although we acknowledge that this needs to be investigated more thoroughly through future work.
 
-While our approach is readily applicable to models with gradient access like deep neural networks, more work is needed to generalise it to other machine learning models such as decision trees.
-Relatedly, common challenges associated with energy-based modelling including sensitivity to scale, training instabilities and sensitivity to hyperparameters also apply to \textit{ECCCo}.
+Furthermore, while our approach is readily applicable to models with gradient access like deep neural networks, more work is needed to generalise our methodology to other popular machine learning models such as gradient-boosted trees.
+Relatedly, common challenges associated with energy-based modelling during our experiments including sensitivity to scale, training instabilities and sensitivity to hyperparameters also apply to \textit{ECCCo}.
 
@@
 
-\subsection{JEM}\label{app-jem}
+The following appendices provide additional details that are relevant to the paper.
+Appendices~\ref{app:jem} and~\ref{app:cp} explain any tasks related to Energy-Based Modelling and Predictive Uncertainty Quantification through Conformal Prediction, respectively.
+Appendix~\ref{app:eccco} provides additional technical and implementation details about our proposed generator, \textit{ECCCo}, including references to our open-sourced code base.
+A complete overview of our experimental setup detailing our parameter choices, training procedures and initial black-box model performance can be found in Appendix~\ref{app:setup}.
+Finally, Appendix~\ref{app:results} reports all of our experimental results in more detail.
 
-While $\mathbf{x}_J$ is only guaranteed to distribute as $p_{\theta}(\mathbf{x}|\mathbf{y}^+)$ if $\epsilon \rightarrow 0$ and $J \rightarrow \infty$, the bias introduced for a small finite $\epsilon$ is negligible in practice~\citep{murphy2023probabilistic,grathwohl2020your}.
-While~\citet{grathwohl2020your} use Equation~\ref{eq:sgld} during training, we are interested in applying the conditional sampling procedure in a post-hoc fashion to any standard discriminative model.
+\subsection{Energy-Based Modelling}\label{app:jem}
 
-\subsection{Conformal Prediction}\label{app-cp}
+Since we were not able to identify any existing open-source software for Energy-Based Modelling that would be flexible enough to cater to our needs, we have developed a Julia package from scratch.
+In our development we have heavily drawn on the existing literature:~\citet{du2020implicit} describe best practices for using EBM for generative modelling;~\citet{grathwohl2020your} explain how EBM can be used to train classifiers jointly for the discriminative and generative tasks.
+We have used the same package for training and inference, but there are some important differences between the two cases that are worth highlighting here.
+
+\subsubsection{Training: Joint Energy Models}
+
+To train our Joint Energy Models we broadly follow the approach outlined in~\citet{grathwohl2020your}.
+These models are trained to optimize a hybrid objective that involves a standard classification loss component $L_{\text{clf}}(\theta)=-\log p_{\theta}(\mathbf{y}|\mathbf{x})$ (e.g.
+crossentropy loss) as well as a generative loss component $L_{\text{gen}}(\theta)=-\log p_{\theta}(\mathbf{x})$.
+
+\begin{equation}\label{eq:jem-loss}
+  \begin{aligned}
+    L(\theta) &= L_{\text{clf}}(\theta) + L_{\text{gen}}(\theta) + \lambda L_{\text{reg}}(\theta) 
+  \end{aligned}
+\end{equation}
+
+To draw samples from $p_{\theta}(\mathbf{x})$, we rely exclusively on the conditional sampling approach described in~\citet{grathwohl2020your} for both training and inference: we first sample $\mathbf{y}\sim p(\mathbf{y})=$ where  While our package also supports unconditional sampling, we found co since in the context of CE we are interested in conditioning on the target class anyway.
+During training As mentioned in the body of the paper, we rely on a biased sampler involving separately specified values for the step size $\epsilon$ and the standard deviation $\sigma$ of the stochastic term involving $\mathbf{r}$.
+Formally, our biased sampler performs updates as follows:
+
+\begin{equation}\label{eq:biased-sgld}
+  \begin{aligned}
+    \mathbf{x}_{j+1} &\leftarrow \mathbf{x}_j - \frac{\epsilon}{2} \mathcal{E}(\mathbf{x}_j|\mathbf{y}^+) + \sigma \mathbf{r}_j, && j=1,...,J
+  \end{aligned}
+\end{equation}
+
+Consistent with ~\citet{grathwohl2020your}, we have specified $\epsilon=2$ and $\sigma=0.01$ as the default values for all of our experiments.
+
+\subsubsection{Inference: Quantifying Models' Generative Property}
+
+At inference time, we assume no prior knowledge about the model's generative property.
+
+\subsection{Conformal Prediction}\label{app:cp}
 

### 2023-05-17 notebooks/tables.Rmd

@@
 )
-tab_i <- tab
+tab_i <- tab_valid
 
@@
 caption <- sprintf(
-  "Results for %s datasets: sample averages plus/minus one standard deviation.
+  "Results for %s datasets: sample averages +/- one standard deviation over all valid counterfactuals.
 Best outcomes are highlighted in bold.
@@
 kbl(
-  tab_full, caption = "All results for all datasets.
-Standard deviations across samples are shown in parentheses.
+  tab_full, caption = "All results for all datasets: sample averages +/- one standard deviation over all counterfactuals.
 Best outcomes are highlighted in bold.
@@
 kbl(
-  tab_full, caption = "All results for all datasets including only valid counterfactuals.
-Standard deviations across samples are shown in parentheses.
+  tab_full, caption = "All results for all datasets: sample averages +/- one standard deviation over all valid counterfactuals.
 Best outcomes are highlighted in bold.

### 2023-05-17 paper/paper.tex

--- 2023-05-16 paper/paper.tex
@@
 To this end, we draw on recent advances in Energy-Based Modelling (EBM), a subdomain of machine learning that is concerned with generative or hybrid modelling~\citep{grathwohl2020your,du2020implicit}.
-In particular, note that if we fix $\mathbf{y}$ to our target value $\mathbf{y}^+$, we can conditionally draw from $p_{\theta}(\mathbf{x}|\mathbf{y}^+)$ using Stochastic Gradient Langevin Dynamics (SGLD) as follows,
+In particular, note that if we fix $\mathbf{y}$ to our target value $\mathbf{y}^+$, we can conditionally draw from $p_{\theta}(\mathbf{x}|\mathbf{y}^+)$ by randomly initializing $\mathbf{x}_0$ and then using Stochastic Gradient Langevin Dynamics (SGLD) as follows,
 
@@
 where $\mathbf{r}_j \sim \mathcal{N}(\mathbf{0},\mathbf{I})$ is the stochastic term and the step-size $\epsilon$ is typically polynomially decayed~\citep{welling2011bayesian}.
-The term $\mathcal{E}(\mathbf{x}_j|\mathbf{y}^+)$ denotes the model energy conditioned on the target class label $\mathbf{y}^+$ which we specify as the negative logit corresponding to that label~\citep{grathwohl2020your}.
+The term $\mathcal{E}(\mathbf{x}_j|\mathbf{y}^+)$ denotes the model energy conditioned on the target class label $\mathbf{y}^+$ which we specify as the negative logit corresponding to the target class label $\mathbf{y}^*$.
 To allow for faster sampling, we follow the common practice of choosing the step-size $\epsilon$ and the standard deviation of $\mathbf{r}_j$ separately.
-While $\mathbf{x}_J$ is only guaranteed to distribute as $p_{\theta}(\mathbf{x}|\mathbf{y}^+)$ if $\epsilon \rightarrow 0$ and $J \rightarrow \infty$, the bias introduced for a small finite $\epsilon$ is negligible in practice \citep{murphy2023probabilistic,grathwohl2020your}.
-Appendix~\ref{app-jem} provides additional implementation details for any tasks related to Energy-Based Modelling.
+While $\mathbf{x}_J$ is only guaranteed to distribute as $p_{\theta}(\mathbf{x}|\mathbf{y}^*)$ if $\epsilon \rightarrow 0$ and $J \rightarrow \infty$, the bias introduced for a small finite $\epsilon$ is negligible in practice \citep{murphy2023probabilistic,grathwohl2020your}.
+Appendix~\ref{app:jem} provides additional implementation details for any tasks related to energy-based modelling.
 
@@
 Consequently, CP removes the restriction on the family of predictive models, at the small cost of reserving a subset of the available data for calibration.
-This particular case of Conformal Prediction is referred to as Split Conformal Prediction (SCP) as it involves splitting the training data into a proper training dataset and a calibration dataset.
-In addition to the smooth set size penalty, we have also experimented with the use of a tailored function for $\text{yloss}(\cdot)$ that enforces that only the target label $\mathbf{y}^+$ is included in the prediction set~\citep{stutz2022learning}.
-Further details are provided in Appendix~\ref{app-cp}.
+This particular case of conformal prediction is referred to as Split Conformal Prediction (SCP) as it involves splitting the training data into a proper training dataset and a calibration dataset.
+In addition to the smooth set size penalty, we have also experimented with the use of a tailored function for $\text{yloss}(\cdot)$ that enforces that only the target label $\mathbf{y}^+$ is included in the prediction set ~\citet{stutz2022learning}.
+Further details are provided in Appendix~\ref{app:cp}.
 
@@
 
-\subsection{JEM}\label{app-jem}
+The following appendices provide additional details that are relevant to the paper.
+Appendices~\ref{app:jem} and~\ref{app:cp} explain any tasks related to Energy-Based Modelling and Predictive Uncertainty Quantification through Conformal Prediction, respectively.
+Appendix~\ref{app:eccco} provides additional technical and implementation details about our proposed generator, \textit{ECCCo}, including references to our open-sourced code base.
+A complete overview of our experimental setup detailing our parameter choices, training procedures and initial black-box model performance can be found in Appendix~\ref{app:setup}.
+Finally, Appendix~\ref{app:results} reports all of our experimental results in more detail.
 
-While $\mathbf{x}_J$ is only guaranteed to distribute as $p_{\theta}(\mathbf{x}|\mathbf{y}^+)$ if $\epsilon \rightarrow 0$ and $J \rightarrow \infty$, the bias introduced for a small finite $\epsilon$ is negligible in practice~\citep{murphy2023probabilistic,grathwohl2020your}.
-While~\citet{grathwohl2020your} use Equation~\ref{eq:sgld} during training, we are interested in applying the conditional sampling procedure in a post-hoc fashion to any standard discriminative model.
+\subsection{Energy-Based Modelling}\label{app:jem}
 
-\subsection{Conformal Prediction}\label{app-cp}
+Since we were not able to identify any existing open-source software for Energy-Based Modelling that would be flexible enough to cater to our needs, we have developed a Julia package from scratch.
+In our development we have heavily drawn on the existing literature:~\citet{du2020implicit} describe best practices for using EBM for generative modelling;~\citet{grathwohl2020your} explain how EBM can be used to train classifiers jointly for the discriminative and generative tasks.
+We have used the same package for training and inference, but there are some important differences between the two cases that are worth highlighting here.
 
-The fact that conformal classifiers produce set-valued predictions introduces a challenge: it is not immediately obvious how to use such classifiers in the context of gradient-based counterfactual search.
-Put differently, it is not clear how to use prediction sets in Equation~\ref{eq:general}.
-Fortunately, \citet{stutz2022learning} have recently proposed a framework for Conformal Training that also hinges on differentiability.
-Specifically, they show how Stochastic Gradient Descent can be used to train classifiers not only for the discriminative task but also for additional objectives related to Conformal Prediction.
-One such objective is \textit{efficiency}: for a given target error rate $\alpha$, the efficiency of a conformal classifier improves as its average prediction set size decreases.
-To this end, the authors introduce a smooth set size penalty defined in Equation~\ref{eq:setsize} in the body of this paper
+\subsubsection{Training: Joint Energy Models}
 
-Formally, it is defined as $C_{\theta,\mathbf{y}}(\mathbf{x}_i;\alpha):=\sigma\left((s(\mathbf{x}_i,\mathbf{y})-\alpha) T^{-1}\right)$ for $\mathbf{y}\in\mathcal{Y}$, where $\sigma$ is the sigmoid function and $T$ is a hyper-parameter used for temperature scaling~\citep{stutz2022learning}.
+To train our Joint Energy Models we broadly follow the approach outlined in~\citet{grathwohl2020your}.
+These models are trained to optimize a hybrid objective that involves a standard classification loss component $L_{\text{clf}}(\theta)=-\log p_{\theta}(\mathbf{y}|\mathbf{x})$ (e.g.
+crossentropy loss) as well as a generative loss component $L_{\text{gen}}(\theta)=-\log p_{\theta}(\mathbf{x})$.
+
+To draw samples from $p_{\theta}(\mathbf{x})$, we rely exclusively on the conditional sampling approach described in~\citet{grathwohl2020your} for both training and inference: we first draw $\mathbf{y}\sim p(\mathbf{y})$ and then sample $\mathbf{x} \sim p_{\theta}(\mathbf{x}|\mathbf{y})$~\citep{grathwohl2020your} via Equation~\ref{eq:sgld} with energy $\mathcal{E}(\mathbf{x}|\mathbf{y})=\mu_{\theta}(\mathbf{x})[\mathbf{y}]$ where $\mu_{\theta}: \mathcal{X} \mapsto \mathbb{R}^K$ returns the linear predictions (logits) of our classifier $M_{\theta}$.
+While our package also supports unconditional sampling, we found conditional sampling to work well.
+It is also well aligned with CE, since in this context we are interested in conditioning on the target class.
+
+As mentioned in the body of the paper, we rely on a biased sampler involving separately specified values for the step size $\epsilon$ and the standard deviation $\sigma$ of the stochastic term involving $\mathbf{r}$.
+Formally, our biased sampler performs updates as follows:
+
+\begin{equation}\label{eq:biased-sgld}
+  \begin{aligned}
+    \hat{\mathbf{x}}_{j+1} &\leftarrow \hat{\mathbf{x}}_j - \frac{\epsilon}{2} \mathcal{E}(\hat{\mathbf{x}}_j|\mathbf{y}^+) + \sigma \mathbf{r}_j, && j=1,...,J
+  \end{aligned}
+\end{equation}
+
+Consistent with~\citet{grathwohl2020your}, we have specified $\epsilon=2$ and $\sigma=0.01$ as the default values for all of our experiments.
+The number of total SGLD steps $J$ varies by dataset.
+Following best practices, we initialize $\mathbf{x}_0$ randomly in 5\% of all cases and sample from a buffer in all other cases.
+The buffer itself is randomly initialised and gradually grows to a maximum of 10,000 samples during training as $\hat{\mathbf{x}}_{J}$ is stored in each epoch~\citep{du2020implicit,grathwohl2020your}.
+
+It is important to realise that sampling is done during each training epoch, which makes training Joint Energy Models significantly harder than conventional neural classifiers.
+In each epoch the generated (batch of) sample(s) $\hat{\mathbf{x}}_{J}$ is used as part of the generative loss component, which compares its energy to that of observed samples $\mathbf{x}$: $L_{\text{gen}}(\theta)=\mu_{\theta}(\mathbf{x})[\mathbf{y}]-\mu_{\theta}(\hat{\mathbf{x}}_{J})[\mathbf{y}]$.
+Our full training objective can be summarized as follows,
+
+\begin{equation}\label{eq:jem-loss}
+  \begin{aligned}
+    L(\theta) &= L_{\text{clf}}(\theta) + L_{\text{gen}}(\theta) + \lambda L_{\text{reg}}(\theta) 
+  \end{aligned}
+\end{equation}
+
+where $L_{\text{reg}}(\theta)$ is a Ridge penalty (L2 norm) that regularises energy magnitudes for both observed and generated samples~\citep{du2020implicit}.
+We have used varying degrees of regularization depending on the dataset.
+
+Contrary to existing work, we have not typically used the entire minibatch of training data for the generative loss component but found that using a subset of the minibatch was often sufficient in attaining decent generative performance.
+This has helped to reduce the computational burden for our models, which should make it easier for others to reproduce our findings.
+
+\subsubsection{Inference: Quantifying Models' Generative Property}
+
+At inference time, we assume no prior knowledge about the model's generative property.
+This means that we do not tab into the existing buffer of generated samples for our Joint Energy Models, but instead generate conditional samples from scratch.
+While have relied on the default values $\epsilon=2$ and $\sigma=0.01$ also during inference, the number of total SGLD steps was set to $J=500$ in all cases, so significantly higher than during training.
+For all of our synthetic datasets and models, we generated 50 conditional samples and then formed subsets containing the $n_{E}=25$ lowest-energy samples.
+While in practice it would be sufficient to do this once for each model and dataset, we have chosen to perform sampling separately for each individual counterfactual in our experiment to account for stochasticity.
+To help reduce the computational burden for our real-world datasets we have generated only 10 conditional samples each time and used all of them in our counterfactual search.
+Using more samples, as we originally did, had no substantial impact on our results.
+
+\subsection{Conformal Prediction}\label{app:cp}
+
+In this Appendix we provide some more background on CP and explain in some more detail how we have used recent advances in Conformal Training for our purposes.
+
+\subsubsection{Background on CP}
 
@@
 
+\subsubsection{Differentiability}
+
+The fact that conformal classifiers produce set-valued predictions introduces a challenge: it is not immediately obvious how to use such classifiers in the context of gradient-based counterfactual search.
+Put differently, it is not clear how to use prediction sets in Equation~\ref{eq:general}.
+Fortunately, \citet{stutz2022learning} have recently proposed a framework for Conformal Training that also hinges on differentiability.
+Specifically, they show how Stochastic Gradient Descent can be used to train classifiers not only for the discriminative task but also for additional objectives related to Conformal Prediction.
+One such objective is \textit{efficiency}: for a given target error rate $\alpha$, the efficiency of a conformal classifier improves as its average prediction set size decreases.
+To this end, the authors introduce a smooth set size penalty defined in Equation~\ref{eq:setsize} in the body of this paper.
+Formally, it is defined as $C_{\theta,\mathbf{y}}(\mathbf{x}_i;\alpha):=\sigma\left((s(\mathbf{x}_i,\mathbf{y})-\alpha) T^{-1}\right)$ for $\mathbf{y}\in\mathcal{Y}$, where $\sigma$ is the sigmoid function and $T$ is a hyper-parameter used for temperature scaling~\citep{stutz2022learning}.
+
+In addition to the smooth set size penalty,~\citet{stutz2022learning} also propose a configurable classification loss function, that can be used to enforce coverage.
+For \textit{MNIST} data, we found that using this function generally improved the visual quality of the generated counterfactuals, so we used it in our experiments involving real-world data.
+For the synthetic dataset, visual inspection of the counterfactuals showed that using the configurable loss function sometimes led to overshooting: counterfactuals would end up deep inside the target domain but far away from the observed samples.
+For this reason we instead relied on standard crossentropy loss for our synthetic datasets.
+As we have noted in the body of the paper, more experimental work is certainly needed in this context.
+
 \subsection{ECCCo}\label{app:eccco}
+
+In this section, we briefly discuss convergence conditions for CE and provide details concerning the actual implementation of our framework in Julia.
+
+\subsubsection{A Note on Convergence}
+
+Convergence is not typically discussed much in the context of CE, even though it has important implications on outcomes.
+One intuitive way to specify convergence is in terms of threshold probabilities: once the predicted probability $p(\mathbf{y}^+|\mathbf{x}^{\prime})$ exceeds some user-defined threshold $\gamma$ such that the counterfactual is valid, we could consider the search to have converged.
+In the binary case, for example, convergence could be defined as $p(\mathbf{y}^+|\mathbf{x}^{\prime})>0.5$ in this sense.
+Note, however, how this can be expected to yield counterfactuals in the proximity of the decision boundary, a region characterized by high aleatoric uncertainty.
+In other words, counterfactuals generated in this way would generally not be plausible.
+To avoid this from happening, we specify convergence in terms of gradients approaching zero for all our experiments and all of our generators.
+This is allows us to get a cleaner read on how the different counterfactual search objectives affect counterfactual outcomes.
+
+\subsubsection{\texttt{ECCCo.jl}}
+
+Our code base is integrated into a larger ecosystem of \texttt{Julia} packages that we are actively developing.
 

### 2023-05-17 paper/paper.tex
just missing more tables

@@
 
-Since we were not able to identify any existing open-source software for Energy-Based Modelling that would be flexible enough to cater to our needs, we have developed a Julia package from scratch.
+Since we were not able to identify any existing open-source software for Energy-Based Modelling that would be flexible enough to cater to our needs, we have developed a \texttt{Julia} package from scratch.
+The package has been open-sourced, but to avoid compromising the double-blind review process, we refrain from providing more information at this stage.
 In our development we have heavily drawn on the existing literature:~\citet{du2020implicit} describe best practices for using EBM for generative modelling;~\citet{grathwohl2020your} explain how EBM can be used to train classifiers jointly for the discriminative and generative tasks.
@@
 In this section, we briefly discuss convergence conditions for CE and provide details concerning the actual implementation of our framework in Julia.  
-
 \subsubsection{A Note on Convergence}
@@
 
-Our code base is integrated into a larger ecosystem of \texttt{Julia} packages that we are actively developing.
+The core part of our code base is integrated into a larger ecosystem of \texttt{Julia} packages that we are actively developing and maintaining.
+To avoid compromising the double-blind review process, we only provide a link to an anonymized repository at this stage: \url{https://anonymous.4open.science/r/ECCCo-1252/README.md}.
 
 \subsection{Experimental Setup}\label{app:setup}
+
 \subsection{Results}\label{app:results}

### 2023-05-17 notebooks/mnist.qmd
uh

@@
 # Hyper:
-_retrain = false
-_regen = false
+_retrain = true
+_regen = true
 
 # Data:
-n_obs = 10000
+n_obs = nothing
 counterfactual_data = load_mnist(n_obs)

### 2023-05-17 notebooks/circles.qmd
appendix

@@
 )
+```
+
+```{julia}
+params = DataFrame(
+    Dict(
+        :n_obs => Int.(round(n_obs/10)*10),
+        :epochs => epochs,
+        :batch_size => batch_size,
+        :n_hidden => n_hidden,
+        :n_layers => length(builder.hidden),
+        :activation => string(activation),
+        :n_ens => n_ens,
+        :lambda => string(α[3]),
+        :jem_sampling_steps => jem.sampling_steps,
+        :sgld_batch_size => sampler.batch_size,
+        :dataname => "Circles",
+    )
+)
+CSV.write(joinpath(params_path, "circles.csv"), params)
 ```

### 2023-05-17 notebooks/linearly_separable.qmd
appendix

@@
 )
+```
+
+```{julia}
+params = DataFrame(
+    Dict(
+        :n_obs => Int.(round(n_obs/10)*10),
+        :epochs => epochs,
+        :batch_size => batch_size,
+        :n_hidden => n_hidden,
+        :n_layers => length(builder.hidden),
+        :activation => string(activation),
+        :n_ens => n_ens,
+        :lambda => string(α[3]),
+        :jem_sampling_steps => jem.sampling_steps,
+        :sgld_batch_size => sampler.batch_size,
+        :dataname => "Linearly Separable",
+    )
+)
+CSV.write(joinpath(params_path, "linearly_separable.csv"), params)
 ```

### 2023-05-17 notebooks/mnist.qmd
appendix

--- 2023-05-16 notebooks/mnist.qmd
@@
 ```{julia}
+params = DataFrame(
+    Dict(
+        :n_obs => Int.(round(n_obs/10)*10),
+        :epochs => epochs,
+        :batch_size => batch_size,
+        :n_hidden => n_hidden,
+        :n_layers => length(model_dict["MLP"].fitresult[1][1])-1,
+        :activation => string(activation),
+        :n_ens => n_ens,
+        :lambda => string(α[3]),
+        :jem_sampling_steps => jem.sampling_steps,
+        :sgld_batch_size => sampler.batch_size,
+        :dataname => "MNIST",
+    )
+)
+CSV.write(joinpath(params_path, "mnist.csv"), params)
+```
+
+```{julia}
 # Plot generated samples:

### 2023-05-17 notebooks/moons.qmd
rerun
appendix

@@
 )
+```
+
+```{julia}
+params = DataFrame(
+    Dict(
+        :n_obs => Int.(round(n_obs/10)*10),
+        :epochs => epochs,
+        :batch_size => batch_size,
+        :n_hidden => n_hidden,
+        :n_layers => length(builder.hidden),
+        :activation => string(activation),
+        :n_ens => n_ens,
+        :lambda => string(α[3]),
+        :jem_sampling_steps => jem.sampling_steps,
+        :sgld_batch_size => sampler.batch_size,
+        :dataname => "Moons",
+    )
+)
+CSV.write(joinpath(params_path, "moons.csv"), params)
 ```
@@
 ```{julia}
-λ₁ = 0.1
-λ₂ = 0.5
-λ₃ = 0.5
+λ₁ = 0.25
+λ₂ = 0.75
+λ₃ = 0.75
 Λ = [λ₁, λ₂, λ₃]
 
-opt = Flux.Optimise.Descent(0.01)
-use_class_loss = true
+opt = Flux.Optimise.Descent(0.05)
+use_class_loss = false
 

### 2023-05-17 notebooks/poc.qmd
appendix

@@
 
-temp = 0.25
+temp = 0.3
+p0 = Plots.contourf(mach.model, mach.fitresult, permutedims(X), labels; plot_set_size=true, zoom=0, temp=temp)
 p1 = Plots.contourf(mach.model, mach.fitresult, permutedims(X), labels; plot_set_loss=true, zoom=0, temp=temp)
 p2 = Plots.contourf(mach.model, mach.fitresult, permutedims(X), labels; plot_classification_loss=true, zoom=0, temp=temp, clim=nothing, loss_matrix=ones(2,2))
-display(Plots.plot(p1, p2, size=(800,320)))
+plt = display(Plots.plot(p0, p1, p2, size=(1400,320), layout=(1,3)))
+savefig(joinpath(output_images_path, "poc_set_size.png"))
 ```

### 2023-05-17 notebooks/tables.Rmd
appendix

@@
 )
+algin_cols <- c(rep('l',3),rep('c',ncol(tab_full)-3))
 kbl(
@@
 \\label{tab:results-full} \\newline",
-  align = "c", col.names=col_names, booktabs = F, escape=F, 
+  align = "c", col.names=col_names, booktabs = T, escape=F, 
   format="latex"
@@
   kable_paper(full_width = F) %>%
-  column_spec(1, bold = T) %>%
-  collapse_rows(columns = 1:3, latex_hline = "custom", valign = "middle", custom_latex_hline = 1:2) %>%
+  collapse_rows(columns = 1:3, latex_hline = "custom", valign = "top", custom_latex_hline = 1:2) %>%
   save_kable("paper/contents/table_all.tex")
@@
   kable_paper(full_width = F) %>%
-  column_spec(1, bold = T) %>%
-  collapse_rows(columns = 1:3, latex_hline = "custom", valign = "middle", custom_latex_hline = 1:2) %>%
+  collapse_rows(columns = 1:3, latex_hline = "custom", valign = "top", custom_latex_hline = 1:2) %>%
   save_kable("paper/contents/table_all_valid.tex")

### 2023-05-17 paper/paper.tex
appendix

@@
 These models are trained to optimize a hybrid objective that involves a standard classification loss component $L_{\text{clf}}(\theta)=-\log p_{\theta}(\mathbf{y}|\mathbf{x})$ (e.g.
-crossentropy loss) as well as a generative loss component $L_{\text{gen}}(\theta)=-\log p_{\theta}(\mathbf{x})$.
+cross-entropy loss) as well as a generative loss component $L_{\text{gen}}(\theta)=-\log p_{\theta}(\mathbf{x})$.
 
@@
 This has helped to reduce the computational burden for our models, which should make it easier for others to reproduce our findings.
+Figures~\ref{fig:mnist-gen} and~\ref{fig:moons-gen} show generated samples for our \textit{MNIST} and \textit{Moons} data, to provide a sense of their generative property.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.75\textwidth]{../artifacts/results/images/mnist_generated_JEM Ensemble.png}
+  \caption{Conditionally generated \textit{MNIST} images for our JEM Ensemble.}\label{fig:mnist-gen}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.5\textwidth]{../artifacts/results/images/moons_generated_JEM.png}
+  \caption{Conditionally generated samples (stars) for our \textit{Moons} data using a JEM.}\label{fig:moons-gen}
+\end{figure}
 
@@
 For the synthetic dataset, visual inspection of the counterfactuals showed that using the configurable loss function sometimes led to overshooting: counterfactuals would end up deep inside the target domain but far away from the observed samples.
-For this reason we instead relied on standard crossentropy loss for our synthetic datasets.
+For this reason, we instead relied on standard cross-entropy loss for our synthetic datasets.
 As we have noted in the body of the paper, more experimental work is certainly needed in this context.
+Figure~\ref{fig:cp-diff} shows the prediction set size (left), smooth set size loss (centre) and configurable classification loss (right) for a JEM trained on our \textit{Linearly Separable} data.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=1.0\textwidth]{../artifacts/results/images/poc_set_size.png}
+  \caption{Prediction set size (left), smooth set size loss (centre) and configurable classification loss (right) for a JEM trained on our \textit{Linearly Separable} data.}\label{fig:cp-diff}
+\end{figure}
 

### 2023-05-17 notebooks/setup.jl

@@ -41,6 +41,7 @@ setup_notebooks = quote
     Plots.theme(:wong)
     Random.seed!(2023)
     www_path = "www"
+    params_path = "artifacts/params"
     output_path = "artifacts/results"
     output_images_path = "artifacts/results/images"
     img_height = 300

### 2023-05-17 notebooks/gmsc.qmd
appendix
ufff

@@
 # Hyper:
-_retrain = false
+_retrain = true
 
@@
 test_size = 0.2
-n_obs = Int(50000 / (1.0 - test_size))
-counterfactual_data, test_data = train_test_split(load_gmsc(n_obs); test_size=test_size)
+counterfactual_data, test_data = train_test_split(load_gmsc(); test_size=test_size)
 X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
@@
 ```{julia}
+generator_params = DataFrame(
+    Dict(
+        :λ1 => λ₁,
+        :λ2 => λ₂,
+        :λ3 => λ₃,
+        :opt => string(typeof(opt)),
+        :eta => opt.eta,
+        :dataname => "GMSC",
+    )
+)
+CSV.write(joinpath(params_path, "generator/gmsc.csv"), generator_params)
+```
+
+```{julia}
 # Measures:

### 2023-05-17 notebooks/mnist.qmd
ufff

@@
     show_factual::Bool=false,
+    img_height::Int=180,
     kwargs...,
@@
         convert2image(MNIST, reshape(_x,28,28)),
-        axis=nothing, 
+        axis=([], false),
         size=(img_height, img_height),
@@
             convert2image(MNIST, reshape(x,28,28)),
-            axis=nothing, 
+            axis=([], false),
             size=(img_height, img_height),
@@
 ```{julia}
-_regen_all_digits = false
+_regen_all_digits = true
 if _regen_all_digits
-    function plot_all_digits(rng=123;verbose=true,kwargs...)
+    function plot_all_digits(rng=123;verbose=true,img_height=180,kwargs...)
         plts = []
@@
                 @info "Generating counterfactual for $(i) -> $(j)"
-                plt = plot_mnist(i,j;kwargs...,rng=rng)
+                plt = plot_mnist(i,j;kwargs...,rng=rng, img_height=img_height)
                 !verbose || display(plt)
@@
         end
-        plt = Plots.plot(plts...; size=(img_height*10,img_height*10), layout=(10,10))
+        plt = Plots.plot(plts...; size=(img_height*10,img_height*10), layout=(10,10), dpi=300)
         return plt
@@
 generator_dict = Dict(
-    "Wachter" => WachterGenerator(λ=Λ[1]),
-    "REVISE" => REVISEGenerator(λ=Λ[1]),
+    "Wachter" => WachterGenerator(λ=Λ[1], opt=eccco_generator.opt),
+    "REVISE" => REVISEGenerator(λ=Λ[1], opt=eccco_generator.opt),
     "Schut" => greedy_generator,
@@
 )
+```
+
+```{julia}
+generator_params = DataFrame(
+    Dict(
+        :λ1 => Λ[1],
+        :λ2 => Λ[2],
+        :λ3 => Λ[3],
+        :opt => string(typeof(eccco_generator.opt)),
+        :eta => eccco_generator.opt.eta,
+        :dataname => "MNIST",
+    )
+)
+CSV.write(joinpath(params_path, "generator/mnist.csv"), generator_params)
 ```

### 2023-05-17 notebooks/tables.Rmd
ufff

@@
 Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (Wachter).
-\\label{tab:results-full} \\newline",
+\\label{tab:results-full-valid} \\newline",
   align = "c", col.names=col_names, booktabs = T, escape=F, 
@@
 ```
+
+## EBM
+
+```{r}
+files <- list.files("artifacts/params/")
+dt <- lapply(files, function(x) {
+    fread(file.path("artifacts/params/", x))
+})
+dt <- Reduce(function(x,y) {rbind(x,y, fill=TRUE)}, dt)
+setcolorder(
+  dt, 
+  c(
+    "dataname", "n_obs",
+    "n_hidden", "n_layers", "activation", "n_ens",
+    "epochs", "batch_size",
+    "jem_sampling_steps", "sgld_batch_size", "lambda"
+  )
+)
+dt_ebm <- dt[,.(dataname, jem_sampling_steps, sgld_batch_size, lambda)]
+col_names <- c(
+  "Dataset",
+  "SGLD Steps", "Batch Size", "$\\lambda$"
+)
+kbl(
+  dt_ebm, caption = "EBM hyperparemeter choices for our experiments. \\label{tab:ebmparams} \\newline", 
+  align = "r", col.names=col_names, booktabs = T, escape=F, 
+  format="latex"
+) %>%
+  kable_styling(font_size = 8) %>%
+  kable_paper(full_width = F) %>%
+  save_kable("paper/contents/table_ebm_params.tex")
+```
+
+## Experimental setup
+
+```{r}
+dt_exp <- dt[,.(dataname, n_obs, n_hidden, n_layers, activation, n_ens, epochs, batch_size)]
+col_names <- c(
+  "Dataset", "Sample Size",
+  "Hidden Units", "Hidden Layers", "Activation", "Ensemble Size",
+  "Epochs", "Batch Size"
+)
+header <- c(" " = 2, "Network Architecture" = 4, "Training" = 2)
+kbl(
+  dt_exp, caption = "Paremeter choices for our experiments. \\label{tab:params} \\newline", 
+  align = "r", col.names=col_names, booktabs = T, escape=F, 
+  format="latex"
+) %>%
+  kable_styling(latex_options = c("scale_down")) %>%
+  kable_paper(full_width = F) %>%
+  add_header_above(header) %>%
+  save_kable("paper/contents/table_params.tex")
+```

### 2023-05-17 paper/paper.tex
ufff

@@
 Consistent with~\citet{grathwohl2020your}, we have specified $\epsilon=2$ and $\sigma=0.01$ as the default values for all of our experiments.
-The number of total SGLD steps $J$ varies by dataset.
+The number of total SGLD steps $J$ varies by dataset (Table~\ref{tab:ebmparams}).
 Following best practices, we initialize $\mathbf{x}_0$ randomly in 5\% of all cases and sample from a buffer in all other cases.
@@
 where $L_{\text{reg}}(\theta)$ is a Ridge penalty (L2 norm) that regularises energy magnitudes for both observed and generated samples~\citep{du2020implicit}.
-We have used varying degrees of regularization depending on the dataset.
+We have used varying degrees of regularization depending on the dataset ($\lambda$ in Table~\ref{tab:ebmparams}).
 
-Contrary to existing work, we have not typically used the entire minibatch of training data for the generative loss component but found that using a subset of the minibatch was often sufficient in attaining decent generative performance.
+Contrary to existing work, we have not typically used the entire minibatch of training data for the generative loss component but found that using a subset of the minibatch was often sufficient in attaining decent generative performance (Table~\ref{tab:ebmparams}).
 This has helped to reduce the computational burden for our models, which should make it easier for others to reproduce our findings.
 Figures~\ref{fig:mnist-gen} and~\ref{fig:moons-gen} show generated samples for our \textit{MNIST} and \textit{Moons} data, to provide a sense of their generative property.
+
+\import{contents/}{table_ebm_params.tex}
 
@@
 \end{figure}
-
 \subsubsection{Inference: Quantifying Models' Generative Property}
@@
 
+Table~\ref{tab:params} provides an overview of all parameters related to our experiments.
+The \textit{GMSC} data were randomly undersampled for balancing purposes and all features were standardized.
+\textit{MNIST} data was also randomly undersampled for reasons outlined below.
+Pixel values were preprocessed to fall in the range of $[-1,1]$ and a small Gaussian noise component ($\sigma=0.03$) was added to training samples following common practices in the EBM literature.
+
+\subsubsection{Compute}
+
+To enable others to easily replicate our experiments, we have chosen to work with small neural network architectures and randomly undersampled large datasets where appropriate.
+All of our experiments could then be run locally on a personal machine.
+The longest runtimes we experienced for model training and counterfactual benchmarking were on the order of 8-12 hours (\textit{MNIST} data).
+For the synthetic data, all experiments could be completed in less than an hour.
+
+We have summarised the system information below:
+
+\textbf{Software}:
+
+\begin{itemize}
+  \item System Version: macOS 13.3.1
+  \item Kernel Version: Darwin 22.4.0
+  \item Boot Volume: Macintosh HD
+  \item Boot Mode: Normal
+\end{itemize}
+
+\textbf{Hardware}:
+
+\begin{itemize}
+  \item Model Name: MacBook Pro
+  \item Model Identifier: MacBookPro16,1
+  \item Processor Name: 8-Core Intel Core i9
+  \item Processor Speed: 2.3 GHz
+  \item Number of Processors: 1
+  \item Total Number of Cores: 8
+  \item L2 Cache (per Core): 256 KB
+  \item L3 Cache: 16 MB
+  \item Hyper-Threading Technology: Enabled
+  \item Memory: 32 GB
+\end{itemize}
+
+\import{contents/}{table_params.tex}
 \subsection{Results}\label{app:results}
+
+Figure~\ref{fig:mnist-eccco} shows examples of counterfactuals for \textit{MNIST} data where the underlying model is our \textit{JEM Ensemble}.
+Original images are shown on the diagonal and the corresponding counterfactuals are plotted across rows.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.75\textwidth]{../artifacts/results/images/mnist_eccco_all_digits.png}
+  \caption{Counterfactuals for \textit{MNIST} data and our \textit{JEM Ensemble}.
+Original images are shown on the diagonal with the corresponding counterfactuals plotted across rows.}\label{fig:mnist-eccco}
+\end{figure}
+
+Table~\ref{tab:results-full} reports all of the evaluation metrics we have computed.
+Table~\ref{tab:results-full} reports the same metric for the subset of valid counterfactuals.
+The `Unfaithfulness' and `Implausibility' metrics have been discussed extensively in the body of the paper.
+The `Cost' metric relates to the distance between the factual and the counterfactual.
+The `Redundancy' metric measures sparsity as the percentage of features that remain unperturbed (higher is better).
+The `Uncertainty' metric is just the average value of the smooth set size penalty (Equation~\ref{eq:setsize}).
+Finally, `Validity' is the percentage of valid counterfactuals.
 

### 2023-05-17 notebooks/circles.qmd
ufff
uh

@@
     _perf.mod_name .= mod_name
+    _perf.dataname .= "Circles"
     model_performance = vcat(model_performance, _perf)
@@
 )
+```
+
+```{julia}
+generator_params = DataFrame(
+    Dict(
+        :λ1 => λ₁,
+        :λ2 => λ₂,
+        :λ3 => λ₃,
+        :opt => string(typeof(opt)),
+        :eta => opt.eta,
+        :dataname => "Circles",
+    )
+)
+CSV.write(joinpath(params_path, "generator/circles.csv"), generator_params)
 ```

### 2023-05-17 notebooks/gmsc.qmd
uh

@@
 test_size = 0.2
-counterfactual_data, test_data = train_test_split(load_gmsc(); test_size=test_size)
+counterfactual_data, test_data = train_test_split(load_gmsc(nothing); test_size=test_size)
 X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
@@
 ```{julia}
+params = DataFrame(
+    Dict(
+        :n_obs => Int.(round(n_obs/10)*10),
+        :epochs => epochs,
+        :batch_size => batch_size,
+        :n_hidden => n_hidden,
+        :n_layers => length(model_dict["MLP"].fitresult[1][1])-1,
+        :activation => string(activation),
+        :n_ens => n_ens,
+        :lambda => string(α[3]),
+        :jem_sampling_steps => jem.sampling_steps,
+        :sgld_batch_size => sampler.batch_size,
+        :dataname => "GMSC",
+    )
+)
+CSV.write(joinpath(params_path, "gmsc.csv"), params)
+```
+
+```{julia}
 # Evaluate models:
@@
     _perf.mod_name .= mod_name
+    _perf.dataname .= "GMSC"
     model_performance = vcat(model_performance, _perf)

### 2023-05-17 notebooks/linearly_separable.qmd
ufff
uh

@@
     _perf.mod_name .= mod_name
+    _perf.dataname .= "Linearly Separable"
     model_performance = vcat(model_performance, _perf)
@@
 )
+```
+
+```{julia}
+generator_params = DataFrame(
+    Dict(
+        :λ1 => λ₁,
+        :λ2 => λ₂,
+        :λ3 => λ₃,
+        :opt => string(typeof(opt)),
+        :eta => opt.eta,
+        :dataname => "Linearly Separable",
+    )
+)
+CSV.write(joinpath(params_path, "generator/linearly_separable.csv"), generator_params)
 ```

### 2023-05-17 notebooks/mnist.qmd
uh

@@
     _perf.mod_name .= mod_name
+    _perf.dataname .= "MNIST"
     model_performance = vcat(model_performance, _perf)

### 2023-05-17 notebooks/moons.qmd
ufff
uh

@@
     _perf.mod_name .= mod_name
+    _perf.dataname .= "Moons"
     model_performance = vcat(model_performance, _perf)
@@
 
+```{julia}
+generator_params = DataFrame(
+    Dict(
+        :λ1 => λ₁,
+        :λ2 => λ₂,
+        :λ3 => λ₃,
+        :opt => string(typeof(opt)),
+        :eta => opt.eta,
+        :dataname => "Moons",
+    )
+)
+CSV.write(joinpath(params_path, "generator/moons.csv"), generator_params)
+```
+
 ### POC

### 2023-05-17 notebooks/tables.Rmd
uh

@@
 files <- list.files("artifacts/params/")
-dt <- lapply(files, function(x) {
+dt <- lapply(files[grepl(".csv", files)], function(x) {
     fread(file.path("artifacts/params/", x))
@@
 )
+dt[,dataname:=factor(dataname, levels=c("Linearly Separable", "Moons", "Circles", "MNIST", "GMSC"))]
+dt <- dt[order(dataname)]
 dt_ebm <- dt[,.(dataname, jem_sampling_steps, sgld_batch_size, lambda)]
@@
 ```
+
+```{r}
+files <- list.files("artifacts/params/generator")
+dt <- lapply(files, function(x) {
+    fread(file.path("artifacts/params/generator", x))
+})
+dt <- Reduce(function(x,y) {rbind(x,y, fill=TRUE)}, dt)
+dt <- dt[,.(dataname,eta,λ1,λ3,λ3)]
+dt[,dataname:=factor(dataname, levels=c("Linearly Separable", "Moons", "Circles", "MNIST", "GMSC"))]
+dt <- dt[order(dataname)]
+col_names <- c(
+  "Dataset",
+  "$\\eta$", "$\\lambda_1$", "$\\lambda_2$", "$\\lambda_3$"
+)
+kbl(
+  dt, caption = "Generator hyperparameters. \\label{tab:genparams} \\newline", 
+  align = "r", col.names=col_names, booktabs = T, escape=F, 
+  format="latex"
+) %>%
+  kable_styling(font_size = 8) %>%
+  kable_paper(full_width = F) %>%
+  save_kable("paper/contents/table_gen_params.tex")
+```
+
+```{r}
+files <- list.files("artifacts/results/")
+dt <- lapply(files[grepl("_model_performance.csv", files)], function(x) {
+    fread(file.path("artifacts/results/", x))
+})
+dt <- Reduce(function(x,y) {rbind(x,y, fill=TRUE)}, dt)
+dt[,dataname:=factor(dataname, levels=c("Linearly Separable", "Moons", "Circles", "MNIST", "GMSC"))]
+dt <- dt[order(dataname,mod_name)]
+setcolorder(
+  dt, 
+  c(
+    "dataname", "mod_name",
+    "acc", "precision", "f1score"
+  )
+)
+col_names <- c("Dataset", "Model", "Accuracy", "Precision", "F1-Score")
+kbl(
+  dt, caption = "Various standard performance metrics for our different models grouped by dataset. \\label{tab:perf} \\newline", 
+  align = "r", col.names=col_names, booktabs = T, escape=F, 
+  format="latex", digits=2
+) %>%
+  kable_styling(font_size = 8) %>%
+  kable_paper(full_width = F) %>%
+  add_header_above(c(" "=2, "Performance Metrics" = 3)) %>%
+  collapse_rows(columns = 1, latex_hline = "custom", valign = "top", custom_latex_hline = 1) %>%
+  save_kable("paper/contents/table_perf.tex")
+```

### 2023-05-17 paper/paper.tex
uh

@@
 \textit{MNIST} data was also randomly undersampled for reasons outlined below.
-Pixel values were preprocessed to fall in the range of $[-1,1]$ and a small Gaussian noise component ($\sigma=0.03$) was added to training samples following common practices in the EBM literature.
+Pixel values were preprocessed to fall in the range of $[-1,1]$ and a small Gaussian noise component ($\sigma=0.03$) was added to training samples following common practice in the EBM literature.
+Table~\ref{tab:perf} shows standard evaluation metrics measuring the predictive performance of our different models grouped by dataset.
+These measures were computed over test datasets.
+
+Table~\ref{tab:genparams} summarises our hyperparameter choices for the counterfactual generators where $\eta$ denotes the learning rate used for Stochastic Gradient Descent (SGD) and $\lambda_1$, $\lambda_2$, $\lambda_3$ represent the chosen penalty strengths (Equations~\ref{eq:general} and~\ref{eq:eccco}).
+Here $\lambda_1$ also refers to the chosen penalty for the distance from factual values that applies to both \textit{Wachter} and \textit{REVISE}, but not \textit{Schut} which is penalty-free.
+\textit{Schut} is also the only generator that uses JSMA instead of SGD for optimization.
+
+\import{contents/}{table_params.tex}
+
+\import{contents/}{table_perf.tex}
+
+\import{contents/}{table_gen_params.tex}
 
@@
 
-\import{contents/}{table_params.tex}
 \subsection{Results}\label{app:results}
@@
   \centering
-  \includegraphics[width=0.75\textwidth]{../artifacts/results/images/mnist_eccco_all_digits.png}
+  \includegraphics[width=0.9\textwidth]{../artifacts/results/images/mnist_eccco_all_digits.png}
   \caption{Counterfactuals for \textit{MNIST} data and our \textit{JEM Ensemble}.

### 2023-05-17 paper/paper.tex
appendix also done

@@
 
+\pagebreak
+
 \appendix
@@
 This means that we do not tab into the existing buffer of generated samples for our Joint Energy Models, but instead generate conditional samples from scratch.
-While have relied on the default values $\epsilon=2$ and $\sigma=0.01$ also during inference, the number of total SGLD steps was set to $J=500$ in all cases, so significantly higher than during training.
+While we have relied on the default values $\epsilon=2$ and $\sigma=0.01$ also during inference, the number of total SGLD steps was set to $J=500$ in all cases, so significantly higher than during training.
 For all of our synthetic datasets and models, we generated 50 conditional samples and then formed subsets containing the $n_{E}=25$ lowest-energy samples.
-While in practice it would be sufficient to do this once for each model and dataset, we have chosen to perform sampling separately for each individual counterfactual in our experiment to account for stochasticity.
+While in practice it would be sufficient to do this once for each model and dataset, we have chosen to perform sampling separately for each individual counterfactual in our experiments to account for stochasticity.
 To help reduce the computational burden for our real-world datasets we have generated only 10 conditional samples each time and used all of them in our counterfactual search.
@@
 
-In this Appendix we provide some more background on CP and explain in some more detail how we have used recent advances in Conformal Training for our purposes.
+In this Appendix~\ref{app:cp} we provide some more background on CP and explain in some more detail how we have used recent advances in Conformal Training for our purposes.
 
@@
 Intuitively, CP works under the premise of turning heuristic notions of uncertainty into rigorous uncertainty estimates by repeatedly sifting through the data.
-It can be used to generate prediction intervals for regression models and prediction sets for classification models~\citep{altmeyer2022conformal}.
+It can be used to generate prediction intervals for regression models and prediction sets for classification models.
 Since the literature on CE and AR is typically concerned with classification problems, we focus on the latter.
@@
 As we have noted in the body of the paper, more experimental work is certainly needed in this context.
-Figure~\ref{fig:cp-diff} shows the prediction set size (left), smooth set size loss (centre) and configurable classification loss (right) for a JEM trained on our \textit{Linearly Separable} data.
+Figure~\ref{fig:cp-diff} shows the prediction set size (left), smooth set size loss (centre) and configurable classification loss (right) for a \textit{JEM} trained on our \textit{Linearly Separable} data.
 
@@
 
-In this section, we briefly discuss convergence conditions for CE and provide details concerning the actual implementation of our framework in Julia.  
+In this section, we briefly discuss convergence conditions for CE and provide details concerning the actual implementation of our framework in \texttt{Julia}.  
 \subsubsection{A Note on Convergence}
@@
 Pixel values were preprocessed to fall in the range of $[-1,1]$ and a small Gaussian noise component ($\sigma=0.03$) was added to training samples following common practice in the EBM literature.
+All of our models were trained through mini-batch training using the Adam optimiser (\citet{kingma2017adam}).
 Table~\ref{tab:perf} shows standard evaluation metrics measuring the predictive performance of our different models grouped by dataset.
-These measures were computed over test datasets.
+These measures were computed on test data.
 
@@
 
-To enable others to easily replicate our experiments, we have chosen to work with small neural network architectures and randomly undersampled large datasets where appropriate.
+To enable others to easily replicate our experiments, we have chosen to work with small neural network architectures and randomly undersampled the \textit{MNIST} dataset (maintaining class balance).
 All of our experiments could then be run locally on a personal machine.
@@
 
-We have summarised the system information below:
+We have summarised our system information below:
 
@@
   \item Kernel Version: Darwin 22.4.0
-  \item Boot Volume: Macintosh HD
-  \item Boot Mode: Normal
 \end{itemize}
@@
 Table~\ref{tab:results-full} reports all of the evaluation metrics we have computed.
-Table~\ref{tab:results-full} reports the same metric for the subset of valid counterfactuals.
+Table~\ref{tab:results-full-valid} reports the same metrics for the subset of valid counterfactuals.
 The `Unfaithfulness' and `Implausibility' metrics have been discussed extensively in the body of the paper.
 The `Cost' metric relates to the distance between the factual and the counterfactual.
-The `Redundancy' metric measures sparsity as the percentage of features that remain unperturbed (higher is better).
+The `Redundancy' metric measures sparsity in is defined as the percentage of features that remain unperturbed (higher is better).
 The `Uncertainty' metric is just the average value of the smooth set size penalty (Equation~\ref{eq:setsize}).

### 2023-05-17 notebooks/tables.Rmd
uh

@@
 )
-tab_i <- tab_valid
+tab_i <- tab
 
@@
 caption <- sprintf(
-  "Results for %s datasets: sample averages +/- one standard deviation over all valid counterfactuals.
+  "Results for %s datasets: sample averages +/- one standard deviation across counterfactuals.
 Best outcomes are highlighted in bold.

### 2023-05-19 notebooks/mnist.qmd
adam also working nicely

@@
 epochs = 100
-batch_size = minimum([Int(round(n_obs/10)), 128])
+batch_size = minimum([Int(round(n_obs/10)), 100])
 n_hidden = 128
@@
 ```{julia}
-plt, eccco_generator, ces = _plot_eccco_mnist()
+opt = Flux.Optimise.Adam(0.01)
+Λ = [0.1,0.25,0.25]
+use_class_loss = true
+
+plt, eccco_generator, ces = _plot_eccco_mnist(opt=opt, λ=Λ, use_class_loss=use_class_loss)
 display(plt)

### 2023-05-19 notebooks/mnist.qmd
polishing repo

--- 2023-05-17 notebooks/mnist.qmd
@@
     end
-    # plts = plts[_plt_order]
-    # plts = [p1, plts...]
     plt = Plots.plot(plts...; size=(img_height,img_height))
@@
 ```{julia}
-_regen_all_digits = true
+_regen_all_digits = false
 if _regen_all_digits

### 2023-05-19 README.qmd
README

---
format: 
  commonmark:
    variant: -raw_html+tex_math_dollars
    wrap: none
    mermaid-format: png
crossref:
  fig-prefix: Figure
  tbl-prefix: Table
bibliography: https://raw.githubusercontent.com/pat-alt/bib/main/bib.bib
output: asis
jupyter: julia-1.8
execute: 
  freeze: auto
  eval: true
  echo: true
  output: false
---

# ECCCo

*Counterfactual Explanations and Algorithmic Recourse in Julia.*

This work is currently undergoing peer review. This README is therefore only meant to provide reviewers access to the code base. The code base will be made public after the review process.

## Inspecting the Package Code

This code base is structured as a Julia package. The package code is located in the `src` folder.

## Inspecting the Code for Experiments

We used [Quarto] notebooks for prototyping and running experiments. The notebooks are located in the `notebooks` folder, separated by dataset:

- [Linearly Separable](notebooks/linearly_separable.qmd)
- [Moons](notebooks/moons.qmd)
- [Circles](notebooks/circles.qmd)
- [MNIST](notebooks/mnist.qmd)
- [GMSC](notebooks/gmsc.qmd)

Instead of looking at the notebooks directly, you may choose to browse the HTML book contained inside the `docs` folder. The book is automatically generated from the notebooks and includes all code chunks and their outputs. It is currently not possible to view the book online, but you can download the `docs` folder and open the `index.html` file in your browser.

## Reproducing the Results

To reproduce the results, you need to install the package, which will automatically install all dependencies. Since the package is not publicly registered and you are looking at an anonymous repository that [cannot be cloned](https://anonymous.4open.science/faq#download), unfortunately, it is not possible to easily install the package and reproduce the results at this stage of the review process.

However, provided that the package is indeed installed, you can reproduce the results by running the experiments in the `experiments` folder. The experiments are structured as Julia scripts. You can run the experiments by executing the following command:

### 2023-05-19 experiments/run_experiments.jl

@@ -1,0 +1,7 @@
+
+include("linearly_separable.jl")
+include("moons.jl")
+include("circles.jl")
+include("mnist.jl")
+include("gmsc.jl")
+

### 2023-06-01 paper/paper.tex

@@
 \author{%
-  Patrick Altmeyer\thanks{See also: https://www.paltmeyer.com/} \\
-  Faculty of Electrical Engineering, Mathematics and Computer Science\\
-  Delft University of Technology\\
-  2628 XE Delft, The Netherlands \\
-  \texttt{p.altmeyer@tudelft.nl} \\
+  Anonymous Author\thanks{See also: } \\
+  Faculty \\
+  University \\
+  Address \\
+  \texttt{email} \\
   \And
-  Mojtaba Farmanbar \\
-  ING Netherlands \\
-  1102 CT Amsterdam, The Netherlands \\
-  \texttt{mojtaba.farmanbar@ing.com} \\
-  \AND
-  Arie van Deursen \\
-  Delft University of Technology\\
-  2628 XE Delft, The Netherlands \\
-  \texttt{arie.vandeursen@tudelft.nl} \\
-  \And
-  Cynthia C. S. Liem \\
-  Delft University of Technology\\
-  2628 XE Delft, The Netherlands \\
-  \texttt{c.c.s.liem@tudelft.nl} \\
+  Anonymous Author\thanks{See also: } \\
+  Faculty \\
+  University \\
+  Address \\
+  \texttt{email} \\
 }

### 2023-08-04 notebooks/mnist.qmd
additional models

@@
 
+#### Additional Models (not in paper)
+
+```{julia}
+add_retrain = true
+
+# Deep Ensemble:
+mlp_large_ens = EnsembleModel(model=mlp, n=50)
+
+add_models = Dict(
+    "Large MLP Ensemble" => mlp_large_ens,
+)
+
+if add_retrain
+    add_model_dict = Dict(mod_name => _train(mod; mod_name=mod_name) for (mod_name, mod) in add_models)
+    large_model_dict = merge(model_dict, add_model_dict)
+    Serialization.serialize(joinpath(output_path,"mnist_models_large.jls"), large_model_dict)
+else
+    large_model_dict = Serialization.deserialize(joinpath(output_path,"mnist_models_large.jls"))
+end
+```
+
+```{julia}
+plt_additional_models, _, _ces_ = _plot_eccco_mnist(
+    plt_order = ["MLP", "MLP Ensemble", "Large Ensemble" "JEM", "JEM Ensemble"]
+)
+display(plt_additional_models)
+savefig(plt, joinpath(output_images_path, "mnist_eccco_additional.png"))
+```
+
 ### All digits

### 2023-08-05 notebooks/gmsc.qmd
tried large ensemble (good), large regularized ensemble (poor). First steps for CNN

@@
 
+## Additional feature preprocessing
+
+```{julia}
+pre_proc = false
+if pre_proc
+    Xmat = matrix(X)
+
+    function threshold(x)
+        σ = std(x)
+        return -2σ .< x .< 2σ
+    end
+    mask = reduce(hcat, map(x -> threshold(x), eachcol(Xmat)))
+    keep_rows = map(x -> all(x), eachrow(mask))
+    Xmat = Xmat[keep_rows,:]
+    labels = labels[keep_rows]
+
+    function boxcox(x)
+        transf = MLJ.UnivariateBoxCoxTransformer()
+        x = exp.(x)
+        mach = machine(transf, x)
+        fit!(mach)
+        z = MLJ.transform(mach, x)
+        return z
+    end
+    Xmat = reduce(hcat, map(x -> boxcox(x), eachcol(Xmat)))
+    X = table(Float32.(Xmat))
+end
+```
+
 First, let's create a couple of image classifier architectures:
@@
 batch_size = minimum([Int(round(n_obs/10)), 250])
-n_hidden = 128
-activation = Flux.swish
+n_hidden = 32
+activation = Flux.relu
 builder = MLJFlux.@builder Flux.Chain(

### 2023-08-05 notebooks/mnist.qmd
tried large ensemble (good), large regularized ensemble (poor). First steps for CNN

@@
 # Model parameters:
-epochs = 100
+epochs = 10
 batch_size = minimum([Int(round(n_obs/10)), 128])
@@
     use_class_loss::Bool = true,
+    model_dict=model_dict,
+    wide::Bool = false,
 )
@@
     end
+    if wide
+        plt = Plots.plot(plts...; size=(img_height*length(plts),img_height), layout=(1,length(plts)))
+    else
         plt = Plots.plot(plts...; size=(img_height,img_height))
+    end
 
@@
 
+An MLP with regularization:
+
+```{julia}
+mlp_reg = NeuralNetworkClassifier(
+    builder=builder, 
+    epochs=epochs,
+    batch_size=batch_size,
+    finaliser=_finaliser,
+    loss=_loss,
+    lambda=1.0,
+    alpha=0.5,
+)
+```
+
+A CNN:
+
+```{julia}
+mutable struct MyConvBuilder
+	filter_size::Int
+	channels1::Int
+	channels2::Int
+	channels3::Int
+end
+
+function MLJFlux.build(b::MyConvBuilder, rng, n_in, n_out, n_channels)
+
+	k, c1, c2, c3 = b.filter_size, b.channels1, b.channels2, b.channels3
+
+	mod(k, 2) == 1 || error("`filter_size` must be odd. ")
+
+	# padding to preserve image size on convolution:
+	p = div(k - 1, 2)
+
+	front = Chain(
+			   Conv((k, k), n_channels => c1, pad=(p, p), relu),
+			   MaxPool((2, 2)),
+			   Conv((k, k), c1 => c2, pad=(p, p), relu),
+			   MaxPool((2, 2)),
+			   Conv((k, k), c2 => c3, pad=(p, p), relu),
+			   MaxPool((2 ,2)),
+			   flatten)
+	d = Flux.outputsize(front, (n_in..., n_channels, 1)) |> first
+	return Chain(front, Dense(d, n_out))
+end
+
+clf = ImageClassifier(builder=MyConvBuilder(3, 16, 32, 32),
+					  epochs=10,
+					  loss=Flux.crossentropy)
+function toimg(X)
+    X = MLUtils.stack(map(x -> reshape(x, 28,28), eachrow(matrix(X))))
+    X = coerce(X, GrayImage)
+    return X
+end
+clf = (X -> toimg(X)) |> clf
+```
+
 ```{julia}
@@
 mlp_large_ens = EnsembleModel(model=mlp, n=50)
+mlp_drop_large_ens = EnsembleModel(model=mlp_reg, n=50)
 
 add_models = Dict(
-    "Large MLP Ensemble" => mlp_large_ens,
+    "Large Ensemble" => mlp_large_ens,
+    "Large Ensemble Reg" => mlp_drop_large_ens,
 )
@@
 plt_additional_models, _, _ces_ = _plot_eccco_mnist(
-    plt_order = ["MLP", "MLP Ensemble", "Large Ensemble" "JEM", "JEM Ensemble"]
+    plt_order = ["MLP", "MLP Ensemble", "Large Ensemble", "Large Ensemble Reg", "JEM", "JEM Ensemble"],
+    model_dict=large_model_dict,
+    wide = true,
 )

### 2023-08-05 .CondaPkg/env/lib/tclConfig.sh

@@ -1,0 +1,169 @@
+# tclConfig.sh --
+#
+# This shell script (for sh) is generated automatically by Tcl's
+# configure script.  It will create shell variables for most of
+# the configuration options discovered by the configure script.
+# This script is intended to be included by the configure scripts
+# for Tcl extensions so that they don't have to figure this all
+# out for themselves.
+#
+# The information in this file is specific to a single platform.
+
+# Tcl's version number.
+TCL_VERSION='8.6'
+TCL_MAJOR_VERSION='8'
+TCL_MINOR_VERSION='6'
+TCL_PATCH_LEVEL='.12'
+
+# C compiler to use for compilation.
+TCL_CC='x86_64-apple-darwin13.4.0-clang'
+
+# -D flags for use with the C compiler.
+TCL_DEFS='-DPACKAGE_NAME=\"tcl\" -DPACKAGE_TARNAME=\"tcl\" -DPACKAGE_VERSION=\"8.6\" -DPACKAGE_STRING=\"tcl\ 8.6\" -DPACKAGE_BUGREPORT=\"\" -DSTDC_HEADERS=1 -DHAVE_SYS_TYPES_H=1 -DHAVE_SYS_STAT_H=1 -DHAVE_STDLIB_H=1 -DHAVE_STRING_H=1 -DHAVE_MEMORY_H=1 -DHAVE_STRINGS_H=1 -DHAVE_INTTYPES_H=1 -DHAVE_STDINT_H=1 -DHAVE_UNISTD_H=1 -DNO_VALUES_H=1 -DHAVE_SYS_PARAM_H=1 -DUSE_THREAD_ALLOC=1 -D_REENTRANT=1 -D_THREAD_SAFE=1 -DHAVE_PTHREAD_ATTR_SETSTACKSIZE=1 -DHAVE_PTHREAD_ATFORK=1 -DTCL_THREADS=1 -DTCL_CFGVAL_ENCODING=\"iso8859-1\" -DHAVE_ZLIB=1 -DMODULE_SCOPE=extern\ __attribute__\(\(__visibility__\(\"hidden\"\)\)\) -DHAVE_HIDDEN=1 -DMAC_OSX_TCL=1 -DHAVE_COREFOUNDATION=1 -DTCL_CFG_DO64BIT=1 -DHAVE_CAST_TO_
[…]

### 2023-08-05 .CondaPkg/env/lib/tkConfig.sh

@@ -1,0 +1,97 @@
+# tkConfig.sh --
+#
+# This shell script (for sh) is generated automatically by Tk's
+# configure script.  It will create shell variables for most of
+# the configuration options discovered by the configure script.
+# This script is intended to be included by the configure scripts
+# for Tk extensions so that they don't have to figure this all
+# out for themselves.  This file does not duplicate information
+# already provided by tclConfig.sh, so you may need to use that
+# file in addition to this one.
+#
+# The information in this file is specific to a single platform.
+
+# Tk's version number.
+TK_VERSION='8.6'
+TK_MAJOR_VERSION='8'
+TK_MINOR_VERSION='6'
+TK_PATCH_LEVEL='.12'
+
+# -D flags for use with the C compiler.
[…]

### 2023-08-05 notebooks/.CondaPkg/env/include/openssl/configuration.h

@@ -1,0 +1,140 @@
+/*
+ * WARNING: do not edit!
+ * Generated by configdata.pm from Configurations/common0.tmpl, Configurations/unix-Makefile.tmpl
+ * via Makefile.in
+ *
+ * Copyright 2016-2021 The OpenSSL Project Authors. All Rights Reserved.
+ *
+ * Licensed under the Apache License 2.0 (the "License").  You may not use
+ * this file except in compliance with the License.  You can obtain a copy
+ * in the file LICENSE in the source distribution or at
+ * https://www.openssl.org/source/license.html
+ */
+
+#ifndef OPENSSL_CONFIGURATION_H
+# define OPENSSL_CONFIGURATION_H
+# pragma once
+
+# ifdef  __cplusplus
+extern "C" {
+# endif
+
+# ifdef OPENSSL_ALGORITHM_DEFINES
+#  error OPENSSL_ALGORITHM_DEFINES no longer supported
+# endif
+
+/*
+ * OpenSSL was configured with the following options:
+ */
+
+# ifndef OPENSSL_SYS_MACOSX
+#  define OPENSSL_SYS_MACOSX 1
+# endif
+# define OPENSSL_CONFIGURED_API 30100
+# ifndef OPENSSL_RAND_SEED_OS
+#  define OPENSSL_RAND_SEED_OS
+# endif
+# ifndef OPENSSL_THREADS
+#  define OPENSSL_THREADS
+# endif
+# ifndef OPENSSL_NO_ACVP_TESTS
+#  define OPENSSL_NO_ACVP_TESTS
+# endif
+# ifndef OPENSSL_NO_AFALGENG
+#  define OPENSSL_NO_AFALGENG
+# endif
+# ifndef OPENSSL_NO_ASAN
+#  define OPENSSL_NO_ASAN
+# endif
+# ifndef OPENSSL_NO_CRYPTO_MDEBUG
+#  define OPENSSL_NO_CRYPTO_MDEBUG
+# endif
+# ifndef OPENSSL_NO_CRYPTO_MDEBUG_BACKTRACE
+#  define OPENSSL_NO_CRYPTO_MDEBUG_BACKTRACE
+# endif
+# ifndef OPENSSL_NO_DEVCRYPTOENG
[…]

### 2023-08-05 src/model.jl

@@ -11,11 +11,11 @@ using Statistics
 const CompatibleAtomicModel = Union{<:MLJFlux.MLJFluxProbabilistic,MLJEnsembles.ProbabilisticEnsembleModel{<:MLJFlux.MLJFluxProbabilistic}}
 
 """
-    ConformalModel <: Models.AbstractDifferentiableJuliaModel
+    ConformalModel <: Models.AbstractDifferentiableModel
 
 Constructor for models trained in `Flux.jl`. 
 """
-struct ConformalModel <: Models.AbstractDifferentiableJuliaModel
+struct ConformalModel <: Models.AbstractDifferentiableModel
     model::ConformalPrediction.ConformalProbabilisticSet
     fitresult::Any
     likelihood::Union{Nothing,Symbol}

### 2023-08-05 src/penalties.jl

@@ -22,7 +22,7 @@ function set_size_penalty(
     _loss = map(eachslice(X, dims=ndims(X))) do x
         x = ndims(x) == 1 ? x[:,:] : x
         if target_probs(ce, x)[1] >= 0.5
-            l = ConformalPrediction.smooth_size_loss(
+            l = ConformalPrediction.ConformalTraining.smooth_size_loss(
                 conf_model, fitresult, x';
                 κ=κ,
                 temp=temp

### 2023-08-08 notebooks/mnist.qmd
added lenet CNN

--- 2023-08-04 notebooks/mnist.qmd
@@
 _retrain = false
-_regen = false
+_regen = true
 
@@
 
+LeNet-5:
+
+```{julia}
+mutable struct LeNetBuilder
+	filter_size::Int
+	channels1::Int
+	channels2::Int
+end
+
+function MLJFlux.build(b::LeNetBuilder, rng, n_in, n_out, n_channels)
+
+	k, c1, c2 = b.filter_size, b.channels1, b.channels2
+
+	mod(k, 2) == 1 || error("`filter_size` must be odd. ")
+
+	# padding to preserve image size on convolution:
+	p = div(k - 1, 2)
+
+	front = Flux.Chain(
+			   Conv((k, k), n_channels => c1, pad=(p, p), relu),
+			   MaxPool((2, 2)),
+			   Conv((k, k), c1 => c2, pad=(p, p), relu),
+			   MaxPool((2, 2)),
+			   Flux.flatten)
+	d = Flux.outputsize(front, (n_in..., n_channels, 1)) |> first
+    back = Flux.Chain(
+        Dense(256, 120, relu),
+        Dense(120, 84, relu),
+        Dense(84, n_out),
+    )
+	return Flux.Chain(front, back)
+end
+
+# Final model:
+lenet = ImageClassifier(
+    builder=LeNetBuilder(5, 6, 16),
+    epochs=epochs,
+    batch_size=batch_size,
+    finaliser=_finaliser,
+    loss=_loss,
+)
+
+# Convert to image:
+function toimg(X)
+    X = matrix(X) |> x -> reshape(x, 28, 28, :)
+    X = coerce(X, GrayImage)
+    return X
+end
+
+lenet = (X -> toimg(X)) |> lenet
+```
+
 ```{julia}
@@
 add_models = Dict(
-    "Large MLP Ensemble" => mlp_large_ens,
+    "Large Ensemble" => mlp_large_ens,
+    "LeNet-5" => lenet,
 )
@@
 ```{julia}
+_plt_order = [
+    "MLP", 
+    "MLP Ensemble", 
+    "Large Ensemble", 
+    "JEM", 
+    "JEM Ensemble",
+]
 plt_additional_models, _, _ces_ = _plot_eccco_mnist(
-    plt_order = ["MLP", "MLP Ensemble", "Large Ensemble" "JEM", "JEM Ensemble"]
+    plt_order = _plt_order,
+    model_dict=large_model_dict,
+    wide = true,
 )
 display(plt_additional_models)
-savefig(plt, joinpath(output_images_path, "mnist_eccco_additional.png"))
+savefig(plt_additional_models, joinpath(output_images_path, "mnist_eccco_additional.png"))
 ```

### 2023-08-08 notebooks/setup.jl

@@ -24,6 +24,7 @@ setup_notebooks = quote
     using Flux
     using Images
     using JointEnergyModels
+    using LaplaceRedux: LaplaceApproximation
     using LinearAlgebra
     using Markdown
     using MLDatasets

### 2023-08-08 notebooks/mnist.qmd
issues with formatting for MLJFlux

@@
     for (mod_name, mod) in model_dict
+        if mod.model.model isa ProbabilisticPipeline
+            x = mod.model.model.f(x)
+            _data = counterfactual_data
+            _data.X = mod.model.model.f(_data.X)
+        end
         ce = generate_counterfactual(
@@
         MaxPool((2, 2)),
-			   Flux.flatten)
+        Flux.flatten
+    )
 	d = Flux.outputsize(front, (n_in..., n_channels, 1)) |> first
     back = Flux.Chain(
-        Dense(256, 120, relu),
+        Dense(d, 120, relu),
         Dense(120, 84, relu),
@@
     )
-	return Flux.Chain(front, back)
+
+    chain = Flux.Chain(front, back)
+
+	return chain
 end
@@
 add_models = Dict(
-    "Large Ensemble" => mlp_large_ens,
     "LeNet-5" => lenet,
+    "Large Ensemble (n=50)" => mlp_large_ens,
 )
@@
     "MLP Ensemble", 
-    "Large Ensemble", 
+    "Large Ensemble (n=50)", 
+    "LeNet-5",
     "JEM", 

### 2023-08-08 src/model.jl

@@ -40,6 +40,8 @@ Private function that extracts the chains from a fitted model.
 function _get_chains(fitresult)
     if fitresult isa MLJEnsembles.WrappedEnsemble
         chains = map(res -> res[1], fitresult.ensemble)
+    elseif fitresult isa MLJBase.Signature
+        chains = [fitted_params(fitresult)[:image_classifier][1]]      # for piped image classifiers
     else
         chains = [fitresult[1]]
     end

### 2023-08-08 notebooks/mnist.qmd
LeNet implemented

--- 2023-08-08 notebooks/mnist.qmd
@@
 
-function MLJFlux.build(b::LeNetBuilder, rng, n_in, n_out, n_channels)
+preproc(X) = reshape(X, (28, 28, 1, :))
+
+function MLJFlux.build(b::LeNetBuilder, rng, n_in, n_out)
+
+    _n_in = Int(sqrt(n_in))
 
@@
 
+    preproc(x) = reshape(x, (_n_in, _n_in, 1, :))
+
 	front = Flux.Chain(
-			   Conv((k, k), n_channels => c1, pad=(p, p), relu),
+        Conv((k, k), 1 => c1, pad=(p, p), relu),
         MaxPool((2, 2)),
@@
         MaxPool((2, 2)),
-			   Flux.flatten)
-	d = Flux.outputsize(front, (n_in..., n_channels, 1)) |> first
+        Flux.flatten
+    )
+	d = Flux.outputsize(front, (_n_in, _n_in, 1, 1)) |> first
     back = Flux.Chain(
-        Dense(256, 120, relu),
+        Dense(d, 120, relu),
         Dense(120, 84, relu),
@@
     )
-	return Flux.Chain(front, back)
+
+    chain = Flux.Chain(preproc, front, back)
+
+	return chain
 end
@@
 # Final model:
-lenet = ImageClassifier(
+lenet = NeuralNetworkClassifier(
     builder=LeNetBuilder(5, 6, 16),
@@
 )
-
-# Convert to image:
-function toimg(X)
-    X = matrix(X) |> x -> reshape(x, 28, 28, :)
-    X = coerce(X, GrayImage)
-    return X
-end
-
-lenet = (X -> toimg(X)) |> lenet
 ```
@@
 
+# LeNet-5 Ensemble:
+lenet_ens = EnsembleModel(model=lenet, n=5)
+
 add_models = Dict(
-    "Large Ensemble" => mlp_large_ens,
     "LeNet-5" => lenet,
+    "LeNet-5 Ensemble" => lenet_ens,
+    "Large Ensemble (n=50)" => mlp_large_ens,
 )
@@
     "MLP Ensemble", 
-    "Large Ensemble", 
+    "Large Ensemble (n=50)", 
+    "LeNet-5",
+    "LeNet-5 Ensemble",
     "JEM", 

### 2023-08-08 notebooks/mnist.qmd
only robust NN left

@@
     wide::Bool = false,
+    img_height::Int = img_height,
 )
@@
 
-# LeNet-5 Ensemble:
-lenet_ens = EnsembleModel(model=lenet, n=5)
-
 add_models = Dict(
     "LeNet-5" => lenet,
-    "LeNet-5 Ensemble" => lenet_ens,
     "Large Ensemble (n=50)" => mlp_large_ens,
@@
     "LeNet-5",
-    "LeNet-5 Ensemble",
     "JEM", 
@@
 savefig(plt_additional_models, joinpath(output_images_path, "mnist_eccco_additional.png"))
+```
+
+```{julia}
+
 ```

### 2023-08-08 notebooks/mnist.qmd
robust net also added

@@
 
+Robust Neural network:
+
+```{julia}
+mutable struct RobustNetBuilder
+	n_hidden::Int
+	lipschitz_bound::Float32
+end
+
+function MLJFlux.build(b::RobustNetBuilder, rng, n_in, n_out)
+	n_hidden, γ = b.n_hidden, b.lipschitz_bound
+    _n_hidden = fill(n_hidden,1)
+    model_ps = DenseLBDNParams{Float32}(n_in, _n_hidden, n_out, γ; rng)
+    chain = Flux.Chain(DiffLBDN(model_ps))
+	return chain
+end
+
+# Final model:
+rob_net = NeuralNetworkClassifier(
+    builder=RobustNetBuilder(128, 5.0),
+    epochs=epochs,
+    batch_size=batch_size,
+    finaliser=_finaliser,
+    loss=_loss,
+)
+```
+
+Training all of them:
+
 ```{julia}
@@
     "LeNet-5" => lenet,
+    "RobustNet" => rob_net,
     "Large Ensemble (n=50)" => mlp_large_ens,
@@
 ```{julia}
+# Evaluate models:
+
+measure = Dict(
+    :f1score => multiclass_f1score, 
+    :acc => accuracy, 
+    :precision => multiclass_precision
+)
+model_performance = DataFrame()
+for (mod_name, mod) in large_model_dict
+    # Test performance:
+    test_data = load_mnist_test()
+    _perf = CounterfactualExplanations.Models.model_evaluation(mod, test_data, measure=collect(values(measure)))
+    _perf = DataFrame([[p] for p in _perf], collect(keys(measure)))
+    _perf.mod_name .= mod_name
+    _perf.dataname .= "MNIST"
+    model_performance = vcat(model_performance, _perf)
+end
+Serialization.serialize(joinpath(output_path,"mnist_model_performance_additional.jls"), model_performance)
+CSV.write(joinpath(output_path, "mnist_model_performance_additional.csv"), model_performance)
+model_performance
+```
+
+```{julia}
 _plt_order = [
@@
     "LeNet-5",
+    "RobustNet",
     "JEM", 
@@
 savefig(plt_additional_models, joinpath(output_images_path, "mnist_eccco_additional.png"))
-```
-
-```{julia}
-
 ```

### 2023-08-08 .CondaPkg/env/include/openssl/configuration.h

@@ -1,140 +1,0 @@
-/*
- * WARNING: do not edit!
- * Generated by configdata.pm from Configurations/common0.tmpl, Configurations/unix-Makefile.tmpl
- * via Makefile.in
- *
- * Copyright 2016-2021 The OpenSSL Project Authors. All Rights Reserved.
- *
- * Licensed under the Apache License 2.0 (the "License").  You may not use
- * this file except in compliance with the License.  You can obtain a copy
- * in the file LICENSE in the source distribution or at
- * https://www.openssl.org/source/license.html
- */
-
-#ifndef OPENSSL_CONFIGURATION_H
-# define OPENSSL_CONFIGURATION_H
-# pragma once
-
-# ifdef  __cplusplus
-extern "C" {
-# endif
-
-# ifdef OPENSSL_ALGORITHM_DEFINES
-#  error OPENSSL_ALGORITHM_DEFINES no longer supported
-# endif
-
-/*
- * OpenSSL was configured with the following options:
- */
-
-# ifndef OPENSSL_SYS_MACOSX
-#  define OPENSSL_SYS_MACOSX 1
-# endif
-# define OPENSSL_CONFIGURED_API 30100
-# ifndef OPENSSL_RAND_SEED_OS
-#  define OPENSSL_RAND_SEED_OS
-# endif
-# ifndef OPENSSL_THREADS
-#  define OPENSSL_THREADS
-# endif
-# ifndef OPENSSL_NO_ACVP_TESTS
-#  define OPENSSL_NO_ACVP_TESTS
-# endif
-# ifndef OPENSSL_NO_AFALGENG
-#  define OPENSSL_NO_AFALGENG
-# endif
-# ifndef OPENSSL_NO_ASAN
-#  define OPENSSL_NO_ASAN
-# endif
-# ifndef OPENSSL_NO_CRYPTO_MDEBUG
-#  define OPENSSL_NO_CRYPTO_MDEBUG
-# endif
-# ifndef OPENSSL_NO_CRYPTO_MDEBUG_BACKTRACE
-#  define OPENSSL_NO_CRYPTO_MDEBUG_BACKTRACE
-# endif
-# ifndef OPENSSL_NO_DEVCRYPTOENG
[…]

### 2023-08-08 .CondaPkg/env/lib/tclConfig.sh

@@ -1,169 +1,0 @@
-# tclConfig.sh --
-#
-# This shell script (for sh) is generated automatically by Tcl's
-# configure script.  It will create shell variables for most of
-# the configuration options discovered by the configure script.
-# This script is intended to be included by the configure scripts
-# for Tcl extensions so that they don't have to figure this all
-# out for themselves.
-#
-# The information in this file is specific to a single platform.
-
-# Tcl's version number.
-TCL_VERSION='8.6'
-TCL_MAJOR_VERSION='8'
-TCL_MINOR_VERSION='6'
-TCL_PATCH_LEVEL='.12'
-
-# C compiler to use for compilation.
-TCL_CC='x86_64-apple-darwin13.4.0-clang'
-
-# -D flags for use with the C compiler.
-TCL_DEFS='-DPACKAGE_NAME=\"tcl\" -DPACKAGE_TARNAME=\"tcl\" -DPACKAGE_VERSION=\"8.6\" -DPACKAGE_STRING=\"tcl\ 8.6\" -DPACKAGE_BUGREPORT=\"\" -DSTDC_HEADERS=1 -DHAVE_SYS_TYPES_H=1 -DHAVE_SYS_STAT_H=1 -DHAVE_STDLIB_H=1 -DHAVE_STRING_H=1 -DHAVE_MEMORY_H=1 -DHAVE_STRINGS_H=1 -DHAVE_INTTYPES_H=1 -DHAVE_STDINT_H=1 -DHAVE_UNISTD_H=1 -DNO_VALUES_H=1 -DHAVE_SYS_PARAM_H=1 -DUSE_THREAD_ALLOC=1 -D_REENTRANT=1 -D_THREAD_SAFE=1 -DHAVE_PTHREAD_ATTR_SETSTACKSIZE=1 -DHAVE_PTHREAD_ATFORK=1 -DTCL_THREADS=1 -DTCL_CFGVAL_ENCODING=\"iso8859-1\" -DHAVE_ZLIB=1 -DMODULE_SCOPE=extern\ __attribute__\(\(__visibility__\(\"hidden\"\)\)\) -DHAVE_HIDDEN=1 -DMAC_OSX_TCL=1 -DHAVE_COREFOUNDATION=1 -DTCL_CFG_DO64BIT=1 -DHAVE_CAST_TO_
[…]

### 2023-08-08 .CondaPkg/env/lib/tkConfig.sh

@@ -1,97 +1,0 @@
-# tkConfig.sh --
-#
-# This shell script (for sh) is generated automatically by Tk's
-# configure script.  It will create shell variables for most of
-# the configuration options discovered by the configure script.
-# This script is intended to be included by the configure scripts
-# for Tk extensions so that they don't have to figure this all
-# out for themselves.  This file does not duplicate information
-# already provided by tclConfig.sh, so you may need to use that
-# file in addition to this one.
-#
-# The information in this file is specific to a single platform.
-
-# Tk's version number.
-TK_VERSION='8.6'
-TK_MAJOR_VERSION='8'
-TK_MINOR_VERSION='6'
-TK_PATCH_LEVEL='.12'
-
-# -D flags for use with the C compiler.
[…]

### 2023-08-08 notebooks/setup.jl

@@ -35,6 +35,7 @@ setup_notebooks = quote
     using MLUtils
     using Plots
     using Random
+    using RobustNeuralNetworks
     using Serialization
     using StatsPlots
     using Tidier

### 2023-08-08 notebooks/mnist.qmd
all coding stuff done

@@
 _retrain = false
-_regen = true
+_regen = false
 
@@
     img_height::Int = img_height,
+    plot_factual::Bool = false,
+    kwrgs...,
 )
@@
     if x isa Int
-        x = reshape(counterfactual_data.X[:,rand(findall(labels.==x))],input_dim,1)
+        x_fact = counterfactual_data.X[:,rand(findall(labels.==x))][:,:]
+    else
+        x_fact = x
     end
@@
         ce = generate_counterfactual(
-            x, target, counterfactual_data, mod, eccco_generator; 
+            x_fact, target, counterfactual_data, mod, eccco_generator; 
             decision_threshold=γ, max_iter=T,
@@
     p1 = Plots.plot(
-        convert2image(MNIST, reshape(x,28,28)),
-        axis=nothing, 
+        convert2image(MNIST, reshape(x_fact,28,28)),
+        axis=([], false),
         size=(img_height, img_height),
@@
 
+    if plot_factual
+        plts = [p1]
+    else
         plts = []
+    end
     letters = collect('a':'z')[1:length(ces)]
@@
     if wide
-        plt = Plots.plot(plts...; size=(img_height*length(plts),img_height), layout=(1,length(plts)))
+        plt = Plots.plot(plts...; size=(img_height*length(plts),img_height), layout=(1,length(plts)), kwrgs...)
     else
-        plt = Plots.plot(plts...; size=(img_height,img_height))
+        plt = Plots.plot(plts...; size=(img_height,img_height), kwrgs...)
     end
@@
 	n_hidden, γ = b.n_hidden, b.lipschitz_bound
-    _n_hidden = fill(n_hidden,1)
+    _n_hidden = fill(n_hidden,2)
     model_ps = DenseLBDNParams{Float32}(n_in, _n_hidden, n_out, γ; rng)
@@
 rob_net = NeuralNetworkClassifier(
-    builder=RobustNetBuilder(128, 5.0),
-    epochs=epochs,
+    builder=RobustNetBuilder(60, 5.0),
+    epochs=600,
     batch_size=batch_size,
@@
     wide = true,
+    plot_factual = true,
+    img_height = 150,
 )
@@
 savefig(plt_additional_models, joinpath(output_images_path, "mnist_eccco_additional.png"))
+```
+
+```{julia}
+combos = [
+    (7,2),
+    (0,8),
+    (8,3),
+    (9,4),
+]
+
+plts = [plt_additional_models]
+
+for (factual, target) in combos
+    plt, _, _ = _plot_eccco_mnist(
+        plt_order = _plt_order,
+        model_dict = large_model_dict,
+        x = factual,
+        target = target,
+        wide = true,
+        plot_factual = true,
+        img_height = 150
+    )
+    display(plt)
+    push!(plts, plt)
+end
 ```

### 2023-08-08 notebooks/mnist.qmd
robustnn not as good as expected unfortunately

@@
 	front = Flux.Chain(
-        Conv((k, k), 1 => c1, pad=(p, p), relu),
+        Conv((k, k), 1 => c1, pad=(p, p), sigmoid),
         MaxPool((2, 2)),
-        Conv((k, k), c1 => c2, pad=(p, p), relu),
+        Conv((k, k), c1 => c2, pad=(p, p), sigmoid),
         MaxPool((2, 2)),
@@
     back = Flux.Chain(
-        Dense(d, 120, relu),
-        Dense(120, 84, relu),
+        Dense(d, 120, sigmoid),
+        Dense(120, 84, sigmoid),
         Dense(84, n_out),
@@
 
+# CNN Ensemble:
+lenet_ens = EnsembleModel(model=lenet, n=5)
+
 add_models = Dict(
     "LeNet-5" => lenet,
-    "RobustNet" => rob_net,
+    "LeNet-5 Ensemble" => lenet_ens,
+    # "RobustNet" => rob_net,
     "Large Ensemble (n=50)" => mlp_large_ens,
@@
     "LeNet-5",
-    "RobustNet",
+    "LeNet-5 Ensemble",
+    # "RobustNet",
     "JEM", 
@@
     plt, _, _ = _plot_eccco_mnist(
+        factual, target; 
         plt_order = _plt_order,
         model_dict = large_model_dict,
-        x = factual,
-        target = target,
         wide = true,

### 2023-08-08 notebooks/mnist.qmd
all run

--- 2023-08-08 notebooks/mnist.qmd
@@
 
+# CNN Ensemble:
+lenet_ens = EnsembleModel(model=lenet, n=5)
+
 add_models = Dict(
     "LeNet-5" => lenet,
-    "RobustNet" => rob_net,
+    # "LeNet-5 Ensemble" => lenet_ens,
+    # "RobustNet" => rob_net,
     "Large Ensemble (n=50)" => mlp_large_ens,
@@
     "LeNet-5",
-    "RobustNet",
+    # "LeNet-5 Ensemble",
+    # "RobustNet",
     "JEM", 

### 2023-08-08 notebooks/mnist.qmd
all run
60000

@@
 # Model parameters:
-epochs = 10
+epochs = 25
 batch_size = minimum([Int(round(n_obs/10)), 128])
@@
 )
-α = [1.0,1.0,1e-2]      # penalty strengths
+α = [1.0,1.0,25e-3]      # penalty strengths
 ```
@@
     builder=LeNetBuilder(5, 6, 16),
-    epochs=epochs,
+    epochs=50,
     batch_size=batch_size,
@@
     (8,3),
-    (9,4),
+    (9,7),
+    (4,9),
+    (2,1),
+    (3,5),
+    (5,8),
+    (6,0),
+    (1,7),
 ]
 
-plts = [plt_additional_models]
+plts = []
 

### 2023-08-09 notebooks/mnist.qmd
revision of rebuttals

@@
     plot_factual::Bool = false,
+    generator::Union{Nothing,CounterfactualExplanations.AbstractGenerator}=nothing,
+    test_data::Bool = false,
     kwrgs...,
@@
 
+    if isnothing(generator)
         # Generate counterfactuals using ECCCo generator:
-    eccco_generator = ECCCoGenerator(
+        generator = ECCCoGenerator(
             λ=λ, 
@@
         )
+    end
+
+    if test_data
+        data = load_mnist_test()
+    else
+        data = counterfactual_data
+    end
 
@@
         ce = generate_counterfactual(
-            x_fact, target, counterfactual_data, mod, eccco_generator; 
+            x_fact, target, data, mod, generator; 
             decision_threshold=γ, max_iter=T,
@@
 
-    return plt, eccco_generator, ces
+    return plt, generator, ces
 end
@@
 plt_additional_models, _, _ces_ = _plot_eccco_mnist(
+    λ = [0.125,0.25,0.25],
     plt_order = _plt_order,
@@
 display(plt_additional_models)
-savefig(plt_additional_models, joinpath(output_images_path, "mnist_eccco_additional.png"))
+# savefig(plt_additional_models, joinpath(output_images_path, "mnist_eccco_additional.png"))
 ```
@@
     push!(plts, plt)
+end
+```
+
+```{julia}
+λ = [0.1,0.25,0.25]
+
+wachter = WachterGenerator(
+    λ=λ[1], 
+    opt=eccco_generator.opt
+)
+
+combos = [
+    (9,7),
+    # (9,7),
+    # (9,7),
+    (6,0),
+    # (6,0),
+    # (6,0),
+    (4,9),
+    # (4,9),
+    # (4,9),
+    (8,3),
+    # (8,3),
+    # (8,3),
+    (5,6),
+    # (5,6),
+    # (5,6),
+]
+
+plts_eccco = []
+plts_wachter = []
+ces_eccco = []
+ces_wachter = []
+rngs = []
+
+for (factual, target) in combos
+
+    rng = rand(1:10000)
+
+    # ECCCo:
+    plt, _, ces = _plot_eccco_mnist(
+        factual, target; 
+        λ = λ,
+        plt_order = _plt_order,
+        model_dict = large_model_dict,
+        wide = true,
+        plot_factual = true,
+        rng = rng,
+        img_height = 150
+    )
+    display(plt)
+    push!(plts_eccco, plt)
+    push!(ces_eccco, reduce(hcat, target_probs.(values(ces))))
+
+    # Wachter:
+    plt, _, ces = _plot_eccco_mnist(
+        factual, target; 
+        plt_order = _plt_order,
+        model_dict = large_model_dict,
+        wide = true,
+        plot_factual = true,
+        rng = rng,
+        img_height = 150,
+        generator = wachter,
+    )
+    display(plt)
+    push!(plts_wachter, plt)
+    push!(ces_wachter, reduce(hcat, target_probs.(values(ces))))
+
+    push!(rngs, rng)
+
 end

### 2023-08-09 notebooks/fmnist.qmd
now some progress on fmnist as well

--- 2023-05-10 notebooks/fashion_mnist.qmd
@@
 
-```{julia}
-# Search:
-generic_generator = WachterGenerator()
-ce_wachter = generate_counterfactual(
-    x_factual, target, counterfactual_data, M, generic_generator; 
-    decision_threshold=γ, max_iter=T,
-    initialization=:identity,
-)
-greedy_generator = GreedyGenerator(η=2.0)
-ce_jsma = generate_counterfactual(
-    x_factual, target, counterfactual_data, M, greedy_generator; 
-    decision_threshold=γ, max_iter=T,
-    initialization=:identity,
-)
-```
-
-```{julia}
-p1 = Plots.plot(
-    convert2image(FashionMNIST, reshape(x_factual,28,28)),
-    axis=nothing, 
-    size=(img_height, img_height),
-    title="Factual"
-)
-plts = [p1]
-
-ces = zip([ce_wachter,ce_jsma])
-counterfactuals = reduce((x,y)->cat(x,y,dims=3),map(ce -> CounterfactualExplanations.counterfactual(ce[1]), ces))
-phat = reduce((x,y) -> cat(x,y,dims=3), map(ce -> target_probs(ce[1]), ces))
-for x in zip(eachslice(counterfactuals; dims=3), eachslice(phat; dims=3), ["Wachter","JSMA"])
-    ce, _phat, _name = (x[1],x[2],x[3])
-    _title = "$(_name) (p=$(round(_phat[1]; digits=2)))"
-    plt = Plots.plot(
-        convert2image(FashionMNIST, reshape(ce,28,28)),
-        axis=nothing, 
-        size=(img_height, img_height),
-        title=_title
-    )
-    plts = [plts..., plt]
-end
-plt = Plots.plot(plts...; size=(img_height*length(plts),img_height), layout=(1,length(plts)))
-display(plt)
-savefig(plt, joinpath(output_images_path, "fashion_you_may_not_like_it.png"))
-```
-
-#### REVISE
-
-```{julia}
-using CounterfactualExplanations.Models: load_fashion_mnist_vae
-vae = load_fashion_mnist_vae()
-vae_weak = load_fashion_mnist_vae(;strong=false)
-Serialization.serialize(joinpath(output_path,"fashion_mnist_classifier.jls"), M)
-Serialization.serialize(joinpath(output_path,"fashion_mnist_vae.jls"), vae)
-Serialization.serialize(joinpath(output_path,"fashion_mnist_vae_weak.jls"), vae_weak)
-```
-
-```{julia}
-# Define generator:
-revise_generator = REVISEGenerator(
-    opt = Flux.Optimise.Descent(0.25),
-    λ=0.0,
-)
-# Generate recourse:
-counterfactual_data.generative_model = vae # assign generative model
-ce_strong = generate_counterfactual(
-    x_factual, target, counterfactual_data, M, revise_generator; 
-    decision_threshold=γ, max_iter=T,
-    initialization=:identity,
-    converge_when=:generator_conditions,
-)
-counterfactual_data_weak = deepcopy(counterfactual_data)
-counterfactual_data_weak.generative_model = vae_weak
-ce_weak = generate_counterfactual(
-    x_factual, target, counterfactual_data_weak, M, revise_generator;
-    decision_threshold=γ, max_iter=T,
-    initialization=:identity,
-    converge_when=:generator_conditions,
-)
-```
-
-```{julia}
-ces = zip([ce_strong,ce_weak])
-counterfactuals = reduce((x,y)->cat(x,y,dims=3),map(ce -> CounterfactualExplanations.counterfactual(ce[1]), ces))
-phat = reduce((x,y) -> cat(x,y,dims=3), map(ce -> target_probs(ce[1]), ces))
-plts = [p1]
-for x in zip(eachslice(counterfactuals; dims=3), eachslice(phat; dims=3), ["Strong VAE","Weak VAE"])
-    ce, _phat, _name = (x[1],x[2],x[3])
-    _title = "$(_name) (p=$(round(_phat[1]; digits=2)))"
-    plt = Plots.plot(
-        convert2image(FashionMNIST, reshape(ce,28,28)),
-        axis=nothing, 
-        size=(img_height, img_height),
-        title=_title
-    )
-    plts = [plts..., plt]
-end
-plt = Plots.plot(plts...; size=(img_height*length(plts),img_height), layout=(1,length(plts)))
-display(plt)
-savefig(plt, joinpath(output_images_path, "fashion_surrogate_gone_wrong.png"))
-```
-
 ### ECCCo
@@
 counterfactual_data.X = pre_process.(counterfactual_data.X)
-counterfactual_data.generative_model = vae
 X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
@@
 # Model parameters:
-epochs = 100
+epochs = 10
 batch_size = minimum([Int(round(n_obs/10)), 128])
-n_hidden = 128
-activation = Flux.swish
+n_hidden = 512
+activation = Flux.relu
 builder = MLJFlux.@builder Flux.Chain(
     Dense(n_in, n_hidden, activation),
+    # Dense(n_hidden, n_hidden, activation),
+    # Dense(n_hidden, n_hidden, activation),
     Dense(n_hidden, n_out),
@@
 _finaliser = x -> x                         # finaliser function
-```
-
-```{julia}
-# JEM parameters:
-𝒟x = Uniform(-1.0,1.0)
-𝒟y = Categorical(ones(output_dim) ./ output_dim)
-sampler = ConditionalSampler(
-    𝒟x, 𝒟y, 
-    input_size=(input_dim,), 
-    batch_size=10,
-)
-α = [1.0,1.0,2e-2]      # penalty strengths
 ```
@@
 
-# Joint Energy Model:
-jem = JointEnergyClassifier(
-    sampler;
-    builder=builder,
-    epochs=epochs,
-    batch_size=batch_size,
-    finaliser=_finaliser,
-    loss=_loss,
-    jem_training_params=(
-        α=α,verbosity=10,
-    ),
-    sampling_steps=25,
-)
-
-# JEM with adversarial training:
-jem_adv = deepcopy(jem)
-# jem_adv.adv_training = true
-
-# Deep Ensemble of Joint Energy Models:
-jem_ens = EnsembleModel(model=jem, n=n_ens)
-
-# Deep Ensemble of Joint Energy Models with adversarial training:
-# jem_ens_plus = EnsembleModel(model=jem_adv, n=n_ens)
-
 # Dictionary of models:
 models = Dict(
-    "MLP" => mlp,
-    "MLP Ensemble" => mlp_ens,
-    "JEM" => jem,
-    "JEM Ensemble" => jem_ens,
-    # "JEM Ensemble+" => jem_ens_plus,
+    # "MLP" => mlp,
+    "Ensemble" => mlp_ens,
 )
@@
 
-        _w = 1500
+        _w = 1000
         plts = []
-        neach = 10
+        neach = 1
         for i in 1:10
@@
         end
-        plt = Plots.plot(plts..., size=(_w,_w), layout=(10,1), plot_title=mod_name)
+        plt = Plots.plot(plts..., size=(_w,_w/10), layout=(1,10), plot_title=mod_name)
         savefig(plt, joinpath(output_images_path, "fashion_mnist_generated_$(mod_name).png"))
@@
 
-### Different Models
-
 ```{julia}
-function _plot_eccco_fashion_mnist(
+function plot_fmnist(
     x::Union{AbstractArray, Int}=x_factual, target::Int=target;
-    λ=[0.1,0.25,0.25],
-    temp=0.1,
-    plt_order = ["MLP", "MLP Ensemble", "JEM", "JEM Ensemble"],
-    opt = nothing,
-    rng::Union{Int,AbstractRNG}=1234,
+    generator = ECCCoGenerator(),
+    rng::Union{Int,AbstractRNG}=Random.GLOBAL_RNG,
     T::Int = 100,
+    use_class_loss::Bool = true,
+    model=model_dict["Ensemble"],
+    img_height::Int = img_height,
+    test_data::Bool = false,
+    kwrgs...,
 )
@@
     if x isa Int
-        x = reshape(counterfactual_data.X[:,rand(findall(labels.==x))],input_dim,1)
+        x_fact = counterfactual_data.X[:,rand(findall(labels.==x))][:,:]
+    else
+        x_fact = x
     end
 
-    # Generate counterfactuals using ECCCo generator:
-    eccco_generator = ECCCoGenerator(
-        λ=λ, 
-        temp=temp, 
-        opt=opt,
-    )
+    if test_data
+        data = load_mnist_test()
+    else
+        data = counterfactual_data
+    end
 
-    ces = Dict()
-    for (mod_name, mod) in model_dict
     ce = generate_counterfactual(
-            x, target, counterfactual_data, mod, eccco_generator; 
+        x_fact, target, data, model, generator; 
         decision_threshold=γ, max_iter=T,
@@
     )
-        ces[mod_name] = ce
-    end
-    _plt_order = map(x -> findall(collect(keys(model_dict)) .== x)[1], plt_order)
 
@@
     p1 = Plots.plot(
-        convert2image(FashionMNIST, reshape(x,28,28)),
-        axis=nothing, 
+        convert2image(MNIST, reshape(x_fact,28,28)),
+        axis=([], false),
         size=(img_height, img_height),
@@
 
-    plts = []
-    for (_name,ce) in ces
     _x = CounterfactualExplanations.counterfactual(ce)
-        _phat = target_probs(ce)
-        _title = "$_name (p̂=$(round(_phat[1]; digits=3)))"
-        plt = Plots.plot(
-            convert2image(FashionMNIST, reshape(_x,28,28)),
-            axis=nothing, 
+    p2 = Plots.plot(
+        convert2image(MNIST, reshape(_x,28,28)),
+        axis=([], false), 
         size=(img_height, img_height),
-            title=_title
+        title="Counterfactual"
     )
-        plts = [plts..., plt]
+
+    plt = Plots.plot(p1, p2; size=(img_height*2,img_height), layout=(1,2), kwrgs...)
+
+    return plt
 end
-    plts = plts[_plt_order]
-    plts = [p1, plts...]
-    plt = Plots.plot(plts...; size=(img_height*length(plts),img_height), layout=(1,length(plts)))
 
-    return plt, eccco_generator
-end
 ```
@@
 ```{julia}
-plt, eccco_generator = _plot_eccco_fashion_mnist()
-display(plt)
-savefig(plt, joinpath(output_images_path, "fashion_mnist_eccco.png"))
+rng = rand(1:10000)
+factual = 3
+plt = plot_fmnist(3, 1, rng=rng)
 ```
-
-### All digits
-
-```{julia}
-function plot_fashion_mnist(
-    factual::Int,target::Int;
-    generator::AbstractGenerator,
-    model::AbstractFittedModel=model_dict["JEM Ensemble"],
-    data::CounterfactualData=counterfactual_data,
-    rng::Union{Int,AbstractRNG}=Random.GLOBAL_RNG,
-    _plot_title::Bool=true,
-    show_factual::Bool=false,
-    kwargs...,
-)
-    Random.seed!(rng)
-
-    decision_threshold = !isdefined(kwargs, :decision_threshold) ? 0.9 : decision_threshold
-    max_iter = !isdefined(kwargs, :max_iter) ? 100 : max_iter
-    initialization = !isdefined(kwargs, :initialization) ? :identity : initialization
-    converge_when = !isdefined(kwargs, :converge_when) ? :generator_conditions : converge_when
-
-    x = reshape(data.X[:,rand(findall(predict_label(model, data).==factual))],input_dim,1)
-    ce = generate_counterfactual(
-        x, target, data, model, generator; 
-        decision_threshold=decision_threshold, max_iter=max_iter,
-        initialization=initialization,
-        converge_when=converge_when,
-        kwargs...
-    )
-
-    _title = _plot_title ? "$(factual) -> $(target)" : ""
-
-    _x = CounterfactualExplanations.counterfactual(ce)
-    plt = Plots.plot(
-        convert2image(FashionMNIST, reshape(_x,28,28)),
-        axis=nothing, 
-        size=(img_height, img_height),
-        title=_title
-    )
-    if show_factual
-        plt_factual = Plots.plot(
-            convert2image(FashionMNIST, reshape(x,28,28)),
-            axis=nothing, 
-            size=(img_height, img_height),
-            title="Factual"
-        )
-        plt = Plots.plot(plt_factual, plt; size=(img_height*2,img_height), layout=(1,2))
-    end
-
-    return plt
-end
-```
-
-```{julia}
-if _regen
-    function plot_all_digits(rng=123;verbose=true,kwargs...)
-        plts = []
-        for i in 0:9
-            for j in 0:9
-                @info "Generating counterfactual for $(i) -> $(j)"
-                plt = plot_fashion_mnist(i,j;kwargs...,rng=rng)
-                !verbose || display(plt)
-                plts = [plts..., plt]
-            end
-        end
-        plt = Plots.plot(plts...; size=(img_height*10,img_height*10), layout=(10,10))
-        return plt
-    end
-    plt = plot_all_digits(generator=eccco_generator)
-    savefig(plt, joinpath(output_images_path, "fashion_mnist_eccco_all_digits.png"))
-end
-```
-
-## Benchmark
-
-```{julia}
-λ₁ = 0.1
-λ₂ = 0.25
-λ₃ = 0.25
-Λ = [λ₁, λ₂, λ₃]
-
-# Benchmark generators:
-generator_dict = Dict(
-    "Wachter" => WachterGenerator(λ=λ₁),
-    "REVISE" => REVISEGenerator(λ=λ₁),
-    "Schut" => GreedyGenerator(),
-    "ECCCo" => ECCCoGenerator(λ=Λ),
-)
-```
-
-```{julia}
-# Measures:
-measures = [
-    CounterfactualExplanations.distance,
-    ECCCo.distance_from_energy,
-    ECCCo.distance_from_targets,
-    CounterfactualExplanations.Evaluation.validity,
-    CounterfactualExplanations.Evaluation.redundancy,
-    ECCCo.set_size_penalty
-]
-
-bmks = []
-for target in sort(unique(labels))
-    for factual in sort(unique(labels))
-        if factual == target
-            continue
-        end
-        bmk = benchmark(
-            counterfactual_data; 
-            models=model_dict, 
-            generators=generator_dict, 
-            measure=measures,
-            suppress_training=true, dataname="FashionMNIST",
-            n_individuals=5,
-            target=target, factual=factual,
-            initialization=:identity,
-            converge_when=:generator_conditions,
-        )
-        push!(bmks, bmk)
-    end
-end
-bmk = reduce(vcat, bmks)
-
-CSV.write(joinpath(output_path, "fashion_mnist_benchmark.csv"), bmk())
-```
-
-```{julia}
-df = @chain bmk() begin
-    @mutate(variable = ifelse.(variable .== "distance_from_energy", "Non-Conformity", variable))
-    @mutate(variable = ifelse.(variable .== "distance_from_targets", "Implausibility", variable))
-    @mutate(variable = ifelse.(variable .== "distance", "Cost", variable))
-    @mutate(variable = ifelse.(variable .== "redundancy", "Redundancy", variable))
-    @mutate(variable = ifelse.(variable .== "Validity", "Validity", variable))
-end
-plt = AlgebraOfGraphics.data(df) * visual(BoxPlot) * 
-    mapping(:generator, :value, row=:variable, col=:model, color=:generator)
-plt = draw(
-    plt, axis=(xlabel="", xticksvisible=false, xticklabelsvisible=false, width=150, height=120), 
-    facet=(; linkyaxes=:minimal)
-)   
-display(plt)
-save(joinpath(output_images_path, "fashion_mnist_benchmark.png"), plt, px_per_unit=5)
-```

### 2023-08-09 notebooks/mnist.qmd
rebuttals nearly revised
now some progress on fmnist as well

@@
 plt_additional_models, _, _ces_ = _plot_eccco_mnist(
-    λ = [0.125,0.25,0.25],
+    λ = [0.1,0.21,0.21],
     plt_order = _plt_order,
@@
 ```{julia}
-combos = [
-    (7,2),
-    (0,8),
-    (8,3),
-    (9,7),
-    (4,9),
-    (2,1),
-    (3,5),
-    (5,8),
-    (6,0),
-    (1,7),
-]
+Random.seed!(2023)
 
-plts = []
-
-for (factual, target) in combos
-    plt, _, _ = _plot_eccco_mnist(
-        factual, target; 
-        plt_order = _plt_order,
-        model_dict = large_model_dict,
-        wide = true,
-        plot_factual = true,
-        img_height = 150
-    )
-    display(plt)
-    push!(plts, plt)
-end
-```
-
-```{julia}
-λ = [0.1,0.25,0.25]
+λ = [0.1,0.215,0.215]
 
@@
     (9,7),
-    # (9,7),
-    # (9,7),
     (6,0),
-    # (6,0),
-    # (6,0),
-    (4,9),
-    # (4,9),
-    # (4,9),
-    (8,3),
-    # (8,3),
-    # (8,3),
-    (5,6),
-    # (5,6),
-    # (5,6),
+    (4,1),
+    (2,3),
+    (5,8),
 ]
+
+n_each = 2
+combos = reduce(vcat, [fill(c, n_each) for c in combos])
 
@@
 
+end
+```
+
+```{julia}
+final_plts = []
+for (i, (factual, target)) in enumerate(combos)
+    p1 = plts_eccco[i]
+    p2 = plts_wachter[i]
+    plt = Plots.plot(
+        p1, p2, layout=(2,1), plot_title="Factual: $factual, Target: $target",
+        size = (1200, 400),
+    )
+    display(plt)
+    savefig(plt, joinpath("dev/rebuttal/www", "mnist_$(factual)to$(target)_$(i).png"))
+    push!(final_plts, plt)
 end

### 2023-08-09 dev/rebuttal/support.qmd
nearly there

---
format: 
    pdf:
        fontsize: 10pt
---

![Turning a 6 into a 0. Top row: ECCCo. Bottom row: Wachter. The different underlying models across columns: (a) MLP, (b) Small Ensemble $n=5$, (c) Large Ensemble $n=50$, (d) LeNet-5, (e) JEM, (f) JEM Ensemble.](www/mnist_6to0_7.png){#fig-mnist-1 width="80%"}

![Turning a 9 into a 7. Rows and columns as in @fig-mnist-1.](www/mnist_9to7_1.png){#fig-mnist-2 width="80%"}

<!-- ![Turning a 4 into a 1. Top row: ECCCo. Bottom row: Wachter. The different underlying models across columns: (a) MLP, (b) Small Ensemble $n=5$, (c) Large Ensemble $n=50$, (d) LeNet-5, (e) JEM, (f) JEM Ensemble.](www/mnist_4to1_11.png){#fig-mnist-3 width="75%"} -->

### 2023-08-09 notebooks/mnist.qmd
done

@@
 ```{julia}
-Random.seed!(2023)
+Random.seed!(123)
 
-λ = [0.1,0.215,0.215]
+λ = [0.1,0.22,0.22]
 
@@
     (9,7),
+    (7,2),
     (6,0),
@@
     (5,8),
+    (6,5),
+    (1,7),
+    (1,4),
+    (0,3)
 ]
 
-n_each = 2
+n_each = 3
 combos = reduce(vcat, [fill(c, n_each) for c in combos])
@@
     plt = Plots.plot(
-        p1, p2, layout=(2,1), plot_title="Factual: $factual, Target: $target",
+        p1, p2, layout=(2,1),
         size = (1200, 400),

### 2023-08-10 dev/rebuttal/support.qmd

@@
         fontsize: 10pt
+        geometry: 
+            - top=10mm
+            - bottom=10mm
+            - left=10mm
+            - right=10mm
 ---
 
-![Turning a 6 into a 0.
-Top row: ECCCo.
+\pagenumbering{gobble}
+
+::: {#fig-fmnist layout="[[-1,10,-1], [-1,10,-1], [-1,10,-1], [20,-1,20,-1,20]]"}
+
+![Turning a 9 into a 7.](www/mnist_9to7_1.png){#fig-mnist-1 width="50%"}
+
+![Turning a 2 into a 3. ](www/mnist_2to3_14.png){#fig-mnist-2 width="50%"}
+
+![Turning a 2 into a 3.](www/mnist_4to1_12.png){#fig-mnist-3 width="50%"}
+
+![Turning a dress (left) into a trouser (right): a gap between the legs appears as expected.](www/fmnist_dress.png){#fig-fmnist-dress width="5cm"}
+
+![Turning a pullover (left) into a coat (right): it looks like a V-neck has formed.](www/fmnist_pullover.png){#fig-fmnist-pullover width="5cm"}
+
+![Turning a boot (left) into a sandal (right): pixels are darkened in the right places.](www/fmnist_boot.png){#fig-fmnist-boot width="5cm"}
+
+Qualitative examples for MNIST and Fashion-MNIST.
+**MNIST** (@fig-mnist-1 to @fig-mnist-3): Top row: ECCCo.
 Bottom row: Wachter.
-The different underlying models across columns: (a) MLP, (b) Small Ensemble $n=5$, (c) Large Ensemble $n=50$, (d) LeNet-5, (e) JEM, (f) JEM Ensemble.](www/mnist_6to0_7.png){#fig-mnist-1 width="80%"}
-
-![Turning a 9 into a 7. Rows and columns as in @fig-mnist-1.](www/mnist_9to7_1.png){#fig-mnist-2 width="80%"}
-
-<!-- ![Turning a 4 into a 1.
-Top row: ECCCo.
-Bottom row: Wachter.
-The different underlying models across columns: (a) MLP, (b) Small Ensemble $n=5$, (c) Large Ensemble $n=50$, (d) LeNet-5, (e) JEM, (f) JEM Ensemble.](www/mnist_4to1_11.png){#fig-mnist-3 width="75%"} -->
+The different underlying models across columns: (a) MLP, (b) Small Ensemble $n=5$, (c) Large Ensemble $n=50$, (d) LeNet-5, (e) JEM, (f) JEM Ensemble.
+**Fashion-MNIST** (@fig-fmnist-dress to @fig-fmnist-boot): The underlying classifier is a small ensemble of 5 MLPs with one hidden layer.
+Counterfactuals are generated by ECCCo.
+:::

### 2023-08-10 notebooks/fmnist.qmd

@@
 ```{julia}
+label_names = [
+    "T-shirt/top",
+    "Trouser",
+    "Pullover",
+    "Dress",
+    "Coat",
+    "Sandal",
+    "Shirt",
+    "Sneaker",
+    "Bag",
+    "Boot",
+]
+
 function plot_fmnist(
-    x::Union{AbstractArray, Int}=x_factual, target::Int=target;
+    x::Int=3, targets::Vector{Int}=[1];
     generator = ECCCoGenerator(),
@@
     test_data::Bool = false,
+    init::Symbol = :identity,
+    noise::Float32 = 0.0f0,
     kwrgs...,
@@
     Random.seed!(rng)
-    if x isa Int
     x_fact = counterfactual_data.X[:,rand(findall(labels.==x))][:,:]
-    else
-        x_fact = x
-    end
+    x_fact = pre_process(x_fact, noise=noise)
 
@@
 
-    ce = generate_counterfactual(
-        x_fact, target, data, model, generator; 
-        decision_threshold=γ, max_iter=T,
-        initialization=:identity,
-        converge_when=:generator_conditions,
-    )
-
     # Plot:
@@
         size=(img_height, img_height),
-        title="Factual"
+        title="$(label_names[x+1]) (factual)",
+        dpi=300
     )
 
+    plts = [p1]
+
+    for t in targets
+        ce = generate_counterfactual(
+            x_fact, t, data, model, generator; 
+            decision_threshold=γ, max_iter=T,
+            initialization=init,
+            converge_when=:generator_conditions,
+        )
         _x = CounterfactualExplanations.counterfactual(ce)
-    p2 = Plots.plot(
+        plt = Plots.plot(
             convert2image(MNIST, reshape(_x,28,28)),
@@
             size=(img_height, img_height),
-        title="Counterfactual"
+            title="$(label_names[t+1])", 
+            dpi=300
         )
+        plts = [plts..., plt]
+    end
 
-    plt = Plots.plot(p1, p2; size=(img_height*2,img_height), layout=(1,2), kwrgs...)
+    plt = Plots.plot(plts..., size=(img_height*length(plts),img_height), layout=(1,length(plts)), kwrgs...)
 
@@
 ```{julia}
-rng = rand(1:10000)
-factual = 3
-plt = plot_fmnist(3, 1, rng=rng)
+rng = 42
+gen = ECCCoGenerator(
+    λ=[0.1,0.30,0.30], 
+    nsamples=25,
+    nmin=10,
+)
+plts = []
+
+plt = plot_fmnist(3, [1]; rng=rng, generator=gen, img_height=200)
+display(plt)
+push!(plts, plt)
+savefig(plt, "dev/rebuttal/www/fmnist_dress.png")
+
+plt = plot_fmnist(9, [5]; rng=rng, generator=gen, img_height=200)
+display(plt)
+push!(plts, plt)
+savefig(plt, "dev/rebuttal/www/fmnist_boot.png")
+
+plt = plot_fmnist(2, [4]; rng=rng, generator=gen, img_height=200)
+display(plt)
+push!(plts, plt)
+savefig(plt, "dev/rebuttal/www/fmnist_pullover.png")
 ```

### 2023-08-10 notebooks/prototyping.qmd
:fire:

--- 2023-05-14 notebooks/linearly_separable.qmd
@@
 ```{julia}
-# Simple MLP:
-mlp = NeuralNetworkClassifier(
-    builder=builder, 
-    epochs=epochs,
-    batch_size=batch_size,
-    finaliser=_finaliser,
-    loss=_loss,
-)
-
-# Deep Ensemble:
-mlp_ens = EnsembleModel(model=mlp, n=n_ens)
-
 # Joint Energy Model:
-jem = JointEnergyClassifier(
+model = JointEnergyClassifier(
     sampler;
@@
 )
-
-# JEM with adversarial training:
-jem_adv = deepcopy(jem)
-# jem_adv.adv_training = true
-
-# Deep Ensemble of Joint Energy Models:
-jem_ens = EnsembleModel(model=jem, n=n_ens)
-
-# Deep Ensemble of Joint Energy Models with adversarial training:
-# jem_ens_plus = EnsembleModel(model=jem_adv, n=n_ens)
-
-# Dictionary of models:
-models = Dict(
-    "MLP" => mlp,
-    # "MLP Ensemble" => mlp_ens,
-    "JEM" => jem,
-    # "JEM Ensemble" => jem_ens,
-    # "JEM Ensemble+" => jem_ens_plus,
-)
 ```
@@
 ```{julia}
-# Train models:
-function _train(model, X=X, y=labels; cov=0.95, method=:simple_inductive, mod_name="model")
-    conf_model = conformal_model(model; method=method, coverage=cov)
-    mach = machine(conf_model, X, y)
-    @info "Begin training $mod_name."
+conf_model = conformal_model(model; method=:simple_inductive, coverage=0.95)
+mach = machine(conf_model, X, labels)
+@info "Begin training model."
 fit!(mach)
-    @info "Finished training $mod_name."
+@info "Finished training model."
 M = ECCCo.ConformalModel(mach.model, mach.fitresult)
-    return M
-end
-if _retrain
-    model_dict = Dict(mod_name => _train(model; mod_name=mod_name) for (mod_name, model) in models)
-    Serialization.serialize(joinpath(output_path,"linearly_separable_models.jls"), model_dict)
-else
-    model_dict = Serialization.deserialize(joinpath(output_path,"linearly_separable_models.jls"))
-end
 ```
-
-```{julia}
-# Evaluate models:
-
-measure = Dict(
-    :f1score => multiclass_f1score, 
-    :acc => accuracy, 
-    :precision => multiclass_precision
-)
-model_performance = DataFrame()
-for (mod_name, model) in model_dict
-    # Test performance:
-    _perf = CounterfactualExplanations.Models.model_evaluation(model, test_data, measure=collect(values(measure)))
-    _perf = DataFrame([[p] for p in _perf], collect(keys(measure)))
-    _perf.mod_name .= mod_name
-    model_performance = vcat(model_performance, _perf)
-end
-Serialization.serialize(joinpath(output_path,"linearly_separable_model_performance.jls"), model_performance)
-CSV.write(joinpath(output_path, "linearly_separable_model_performance.csv"), model_performance)
-model_performance
-```
-
-```{julia}
-n_regen = 1000
-n_each = batch_size
-for (mod_name, model) in model_dict
-    K = length(counterfactual_data.y_levels)
-    input_size = size(selectdim(counterfactual_data.X, ndims(counterfactual_data.X), 1))
-    𝒟x = Uniform(extrema(counterfactual_data.X)...)
-    𝒟y = Categorical(ones(K) ./ K)
-    sampler = ConditionalSampler(𝒟x, 𝒟y; input_size=input_size)
-    opt = ImproperSGLD()
-    plts = []
-    for target in levels(labels)
-        target_idx = findall(levels(labels) .== target)[1]
-        f(x) = logits(model, x)
-        X̂ = sampler(f, opt; niter=n_regen, n_samples=n_each, y=target_idx)
-        ex = extrema(hcat(MLJFlux.reformat(X),X̂), dims=2)
-        xlims = ex[1]
-        ylims = ex[2]
-        x1 = range(1.0f0.*xlims...,length=10)
-        x2 = range(1.0f0.*ylims...,length=10)
-        p(x) = probs(model, x)
-        plt = Plots.contour(
-            x1, x2, (x, y) -> p([x, y][:,:])[target_idx], 
-            fill=true, alpha=0.5, title="Target: $target", cbar=true,
-            xlims=xlims,
-            ylims=ylims,
-        )
-        Plots.scatter!(
-            MLJFlux.reformat(X)[1,:], MLJFlux.reformat(X)[2,:], 
-            color=Int.(labels.refs), group=Int.(labels.refs), alpha=0.5
-        )
-        Plots.scatter!(
-            X̂[1,:], X̂[2,:], 
-            color=repeat([target_idx], size(X̂,2)), 
-            group=repeat([target_idx], size(X̂,2)), 
-            shape=:star5, ms=10
-        )
-        savefig(plt, joinpath(output_images_path, "linearly_separable_generated_$(mod_name).png"))
-        push!(plts, plt)
-    end
-    plt = Plots.plot(plts..., layout=(1, 2), size=(2*500, 400), plot_title=mod_name)
-    display(plt)
-end
-```
-
-```{julia}
-#| output: true
-#| echo: false
-#| label: fig-losses
-#| fig-cap: "Illustration of the smooth size loss and the configurable classification loss."
-
-X_plot = matrix(X)
-temp = 0.1
-
-for (mod_name, model) in model_dict
-    p0 = Plots.contourf(model.model, model.fitresult, X_plot, labels; plot_set_size=true, zoom=0, temp=temp)
-    p1 = Plots.contourf(model.model, model.fitresult, X_plot, labels; plot_set_loss=true, zoom=0, temp=temp)
-    p2 = Plots.contourf(model.model, model.fitresult, X_plot, labels; plot_classification_loss=true, zoom=0, temp=temp, clim=nothing, loss_matrix=ones(2,2))
-    display(Plots.plot(p0, p1, p2, size=(1400,320), plot_title=mod_name, layout=(1,3)))
-end
-```
-
-## Benchmark
 
@@
 generator_dict = Dict(
-    "Wachter" => WachterGenerator(λ=λ₁, opt=opt),
-    "REVISE" => REVISEGenerator(λ=λ₁, opt=opt),
-    "Schut" => GreedyGenerator(),
     "ECCCo" => ECCCoGenerator(λ=Λ, opt=opt, use_class_loss=use_class_loss),
-    "ECCCo (no CP)" => ECCCoGenerator(λ=[λ₁, 0.0, λ₃], opt=opt, use_class_loss=use_class_loss),
-    "ECCCo (no EBM)" => ECCCoGenerator(λ=[λ₁, λ₂, 0.0], opt=opt, use_class_loss=use_class_loss),
+    "ECCCo (energy delta)" => ECCCoGenerator(λ=Λ, opt=opt, use_class_loss=use_class_loss, use_energy_delta=true),
 )
 ```
-
-### POC
 
@@
 
-M = model_dict["JEM"]
 X = X isa Matrix ? X : Float32.(permutedims(matrix(X)))
@@
         cbar=false, 
-        # axis=nothing,
     )
@@
 end
-plt = Plots.plot(plts..., size=(650,500))
+plt = Plots.plot(plts..., size=(800,350))
 display(plt)
-savefig(plt, joinpath(output_images_path, "linearly_separable_poc.png"))
 ```
-
-### Complete Benchmark
-
-```{julia}
-# Measures:
-measures = [
-    CounterfactualExplanations.distance,
-    ECCCo.distance_from_energy,
-    ECCCo.distance_from_targets,
-    CounterfactualExplanations.Evaluation.validity,
-    CounterfactualExplanations.Evaluation.redundancy,
-    ECCCo.set_size_penalty
-]
-
-bmks = []
-for target in sort(unique(labels))
-    for factual in sort(unique(labels))
-        if factual == target
-            continue
-        end
-        bmk = benchmark(
-            counterfactual_data; 
-            models=model_dict, 
-            generators=generator_dict, 
-            measure=measures,
-            suppress_training=true, dataname="Linearly Separable",
-            n_individuals=25,
-            target=target, factual=factual,
-            initialization=:identity,
-            converge_when=:generator_conditions,
-        )
-        push!(bmks, bmk)
-    end
-end
-bmk = reduce(vcat, bmks)
-CSV.write(joinpath(output_path, "linearly_separable_benchmark.csv"), bmk())
-```
-
-```{julia}
-df = @chain bmk() begin
-    @mutate(variable = ifelse.(variable .== "distance_from_energy", "Non-Conformity", variable))
-    @mutate(variable = ifelse.(variable .== "distance_from_targets", "Implausibility", variable))
-    @mutate(variable = ifelse.(variable .== "distance", "Cost", variable))
-    @mutate(variable = ifelse.(variable .== "redundancy", "Redundancy", variable))
-    @mutate(variable = ifelse.(variable .== "Validity", "Validity", variable))
-end
-plt = AlgebraOfGraphics.data(df) * visual(BoxPlot) * 
-    mapping(:generator, :value, row=:variable, col=:model, color=:generator)
-plt = draw(
-    plt, axis=(xlabel="", xticksvisible=false, xticklabelsvisible=false, width=150, height=120), 
-    facet=(; linkyaxes=:none)
-)   
-display(plt)
-save(joinpath(output_images_path, "linearly_separable_benchmark.png"), plt, px_per_unit=5)
-```

### 2023-08-10 notebooks/mnist.qmd
why did my brain decide to come up with this last night and not 3 months ago :fire: :cry:

@@
 ```{julia}
+function plot_mnist(ce; size=(img_height, img_height), kwrgs...)
+    x = CounterfactualExplanations.counterfactual(ce)
+    phat = target_probs(ce)
+    plt = Plots.plot(
+        convert2image(MNIST, reshape(x,28,28));
+        axis=([], false), 
+        size=size,
+        kwrgs...,
+    )
+end
+
 function _plot_eccco_mnist(
@@
     test_data::Bool = false,
+    use_energy_delta::Bool = false,
     kwrgs...,
@@
             nmin=10,
+            use_energy_delta=use_energy_delta,
         )
@@
 savefig(plt, joinpath(output_images_path, "mnist_eccco.png"))
+```
+
+#### Energy Delta (not in paper)
+
+```{julia}
+λ_delta = [0.1,0.1,2.5]
+λ = [0.1,0.25,0.25]
+plts = []
+for i in 0:9
+    plt, _, _ = _plot_eccco_mnist(x_factual, i; λ = λ, plot_title="Distance")
+    plt_delta, _, _ = _plot_eccco_mnist(x_factual, i; λ = λ_delta, use_energy_delta=true, plot_title="Energy Delta")
+    plt = Plots.plot(plt, plt_delta; size=(img_height*2,img_height), layout=(1,2))
+    display(plt)
+    push!(plts, plt)
+end
 ```

### 2023-08-10 notebooks/mnist.qmd
uh

@@
 ```{julia}
+plt, gen_delta, ces = _plot_eccco_mnist(λ = [0.1,0.1,3.0], use_energy_delta=true)
+display(plt)
+savefig(plt, joinpath(output_images_path, "mnist_eccco_energy_delta.png"))
+```
+
+```{julia}
 λ_delta = [0.1,0.1,2.5]
@@
 
+#### Energy Delta (not in paper)
+
+```{julia}
+_regen_all_digits = true
+if _regen_all_digits
+    function plot_all_digits(rng=123;verbose=true,img_height=180,kwargs...)
+        plts = []
+        for i in 0:9
+            for j in 0:9
+                @info "Generating counterfactual for $(i) -> $(j)"
+                plt = plot_mnist(i,j;kwargs...,rng=rng, img_height=img_height)
+                !verbose || display(plt)
+                plts = [plts..., plt]
+            end
+        end
+        plt = Plots.plot(plts...; size=(img_height*10,img_height*10), layout=(10,10), dpi=300)
+        return plt
+    end
+    plt = plot_all_digits(generator=gen_delta)
+    savefig(plt, joinpath(output_images_path, "mnist_eccco_all_digits-delta.png"))
+end
+```
+
 ## Benchmark

### 2023-08-23 notebooks/prototyping.qmd
that's done for now

@@
 epochs = 100
-batch_size = minimum([Int(round(n_obs/10)), 128])
+bs = minimum([Int(round(n_obs/10)), 128])
 n_hidden = 16
@@
     epochs=epochs,
-    batch_size=batch_size,
+    batch_size=bs,
     finaliser=_finaliser,
@@
 X = X isa Matrix ? X : Float32.(permutedims(matrix(X)))
-factual_label =  levels(labels)[1]
+factual_label =  levels(labels)[2]
 x_factual = reshape(X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
-target =  levels(labels)[2]
+target =  levels(labels)[1]
 factual = predict_label(M, counterfactual_data, x_factual)[1]

### 2023-08-23 notebooks/mnist.qmd

@@
 
-Robust Neural network:
-
-```{julia}
-mutable struct RobustNetBuilder
-	n_hidden::Int
-	lipschitz_bound::Float32
-end
-
-function MLJFlux.build(b::RobustNetBuilder, rng, n_in, n_out)
-	n_hidden, γ = b.n_hidden, b.lipschitz_bound
-    _n_hidden = fill(n_hidden,2)
-    model_ps = DenseLBDNParams{Float32}(n_in, _n_hidden, n_out, γ; rng)
-    chain = Flux.Chain(DiffLBDN(model_ps))
-	return chain
-end
-
-# Final model:
-rob_net = NeuralNetworkClassifier(
-    builder=RobustNetBuilder(60, 5.0),
-    epochs=600,
-    batch_size=batch_size,
-    finaliser=_finaliser,
-    loss=_loss,
-)
-```
-
 Training all of them:

### 2023-08-23 notebooks/prototyping.qmd
need to regularize energy delta

@@
 ```{julia}
-include("$(pwd())/notebooks/setup.jl")
-eval(setup_notebooks)
+include("$(pwd())/experiments/setup_env.jl")
 ```
@@
 ```{julia}
-# Hyper:
-_retrain = false
+dataname = "linearly_separable"
+outcome = Serialization.deserialize(joinpath(DEFAULT_OUTPUT_PATH, "$(dataname)_outcome.jls"))
 
-# Data:
-test_size = 0.2
-n_obs = Int(1000 / (1.0 - test_size))
-counterfactual_data, test_data = train_test_split(
-    load_blobs(n_obs; cluster_std=0.1, center_box=(-1. => 1.)); 
-    test_size=test_size
-)
-X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
-X = table(permutedims(X))
-labels = counterfactual_data.output_encoder.labels
-input_dim, n_obs = size(counterfactual_data.X)
-output_dim = length(unique(labels))
-```
-
-First, let's create a couple of image classifier architectures:
-
-```{julia}
-# Model parameters:
-epochs = 100
-bs = minimum([Int(round(n_obs/10)), 128])
-n_hidden = 16
-activation = Flux.swish
-builder = MLJFlux.MLP(
-    hidden=(n_hidden, n_hidden, n_hidden), 
-    σ=Flux.swish
-)
-n_ens = 5                                   # number of models in ensemble
-_loss = Flux.Losses.crossentropy       # loss function
-_finaliser = Flux.softmax                   # finaliser function
-```
-
-```{julia}
-# JEM parameters:
-𝒟x = Normal()
-𝒟y = Categorical(ones(output_dim) ./ output_dim)
-sampler = ConditionalSampler(
-    𝒟x, 𝒟y, 
-    input_size=(input_dim,), 
-    batch_size=50,
-)
-α = [1.0,1.0,1e-1]      # penalty strengths
-```
-
-```{julia}
-# Joint Energy Model:
-model = JointEnergyClassifier(
-    sampler;
-    builder=builder,
-    epochs=epochs,
-    batch_size=bs,
-    finaliser=_finaliser,
-    loss=_loss,
-    jem_training_params=(
-        α=α,verbosity=10,
-    ),
-    sampling_steps=30,
-)
-```
-
-```{julia}
-conf_model = conformal_model(model; method=:simple_inductive, coverage=0.95)
-mach = machine(conf_model, X, labels)
-@info "Begin training model."
-fit!(mach)
-@info "Finished training model."
-M = ECCCo.ConformalModel(mach.model, mach.fitresult)
-```
-
-```{julia}
-λ₁ = 0.25
-λ₂ = 0.75
-λ₃ = 0.75
-Λ = [λ₁, λ₂, λ₃]
-
-opt = Flux.Optimise.Descent(0.01)
-use_class_loss = false
-
-# Benchmark generators:
-generator_dict = Dict(
-    "ECCCo" => ECCCoGenerator(λ=Λ, opt=opt, use_class_loss=use_class_loss),
-    "ECCCo (energy delta)" => ECCCoGenerator(λ=Λ, opt=opt, use_class_loss=use_class_loss, use_energy_delta=true),
-)
+# Unpack
+exp = outcome.exp
+model_dict = outcome.model_dict
+generator_dict = outcome.generator_dict
+bmk = outcome.bmk
 ```
@@
 
+# Unpack
+counterfactual_data = exp.counterfactual_data
+X, labels = counterfactual_data.X, counterfactual_data.output_encoder.labels
+M = model_dict["MLP"]
+gen = filter(((k,v),) -> k in ["ECCCo", "ECCCo-Δ"], generator_dict)
+
+# Prepare search:
 X = X isa Matrix ? X : Float32.(permutedims(matrix(X)))
 factual_label =  levels(labels)[2]
-x_factual = reshape(X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
+x_factual = X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))] |>
+    x -> x[:,:]
 target =  levels(labels)[1]
@@
 plts = []
-for (name, generator) in generator_dict
+for (name, generator) in gen
     ce = generate_counterfactual(

### 2023-08-23 experiments/models/default_models.jl

@@ -31,7 +31,7 @@ Builds a dictionary of default models for training.
 function default_models(;
     sampler::AbstractSampler,
     builder::MLJFlux.Builder=default_builder(),
-    epochs::Int=25,
+    epochs::Int=100,
     batch_size::Int=128,
     finaliser::Function=Flux.softmax,
     loss::Function=Flux.Losses.crossentropy,

### 2023-08-23 src/penalties.jl

@@ -76,14 +76,19 @@ function energy_delta(
     xproposed = CounterfactualExplanations.decode_state(ce)     # current state
     t = get_target_index(ce.data.y_levels, ce.target)
     E(x) = -logits(ce.M, x)[t,:]                                # negative logits for target class
-    _loss = E(xproposed) .- E(xgenerated)
 
-    _loss = reduce((x, y) -> x + y, _loss) / n                  # aggregate over samples
+    # Generative loss:
+    gen_loss = E(xproposed) .- E(xgenerated)
+    gen_loss = reduce((x, y) -> x + y, gen_loss) / n                  # aggregate over samples
+
+    # Regularization loss:
+    reg_loss = E(xgenerated).^2 .+ E(xproposed).^2
+    reg_loss = reduce((x, y) -> x + y, reg_loss) / n                  # aggregate over samples
 
     if return_conditionals
         return conditional_samples[1]
     end
-    return _loss
+    return gen_loss + 0.1reg_loss
 
 end
 

### 2023-09-05 notebooks/circles.qmd
multi-threading for model training

@@
 _loss = Flux.Losses.crossentropy       # loss function
-_finaliser = Flux.softmax                      # finaliser function
+_finaliser = Flux.softmax                      # finalizer function
 ```
@@
     batch_size=batch_size,
-    finaliser=_finaliser,
+    finalizer=_finaliser,
     loss=_loss,

### 2023-09-05 notebooks/gmsc.qmd
multi-threading for model training

@@
 _loss = Flux.Losses.crossentropy            # loss function
-_finaliser =  Flux.softmax                  # finaliser function
+_finaliser =  Flux.softmax                  # finalizer function
 ```

### 2023-09-05 notebooks/linearly_separable.qmd
multi-threading for model training

@@
 _loss = Flux.Losses.crossentropy       # loss function
-_finaliser = Flux.softmax                   # finaliser function
+_finaliser = Flux.softmax                   # finalizer function
 ```

### 2023-09-05 notebooks/mnist.qmd
multi-threading for model training

@@
 _loss = Flux.Losses.crossentropy       # loss function
-_finaliser = Flux.softmax                        # finaliser function
+_finaliser = Flux.softmax                        # finalizer function
 ```

### 2023-09-05 experiments/benchmarking/benchmarking.jl

@@ -4,6 +4,8 @@ function default_generators(;
     use_variants::Bool=true,
     use_class_loss::Bool=false,
     opt=Flux.Optimise.Descent(0.01),
+    nsamples::Union{Nothing,Int}=nothing,
+    nmin::Union{Nothing,Int}=nothing,
 )
 
     @info "Begin benchmarking counterfactual explanations."
@@ -15,20 +17,20 @@ function default_generators(;
             "Wachter" => WachterGenerator(λ=λ₁, opt=opt),
             "REVISE" => REVISEGenerator(λ=λ₁, opt=opt),
             "Schut" => GreedyGenerator(),
-            "ECCCo" => ECCCoGenerator(λ=Λ, opt=opt, use_class_loss=use_class_loss),
-            "ECCCo (no CP)" => ECCCoGenerator(λ=[λ₁, 0.0, λ₃], opt=opt, use_class_loss=use_class_loss),
-            "ECCCo (no EBM)" => ECCCoGenerator(λ=[λ₁, λ₂, 0.0], opt=opt, use_class_loss=use_class_loss),
-            "ECCCo-Δ" => ECCCoGenerator(λ=Λ_Δ, opt=opt, use_class_loss=use_class_loss, use_energy_delta=true),
-            "ECCCo-Δ (no CP)" => ECCCoGenerator(λ=[λ₁_Δ, 0.0, λ₃_Δ], opt=opt, use_class_loss=use_class_loss, use_energy_delta=true),
-            "ECCCo-Δ (no EBM)" => ECCCoGenerator(λ=[λ₁_Δ, λ₂_Δ, 0.0], opt=opt, use_class_loss=use_class_loss, use_energy_delta=true),
+            "ECCCo" => ECCCoGenerator(λ=Λ, opt=opt, use_class_loss=use_class_loss, nsamples=nsamples, nmin=nmin),
+            "ECCCo (no CP)" => ECCCoGenerator(λ=[λ₁, 0.0, λ₃], opt=opt, use_class_loss=use_class_loss, nsamples=nsamples, nmin=nmin),
+            "ECCCo (no EBM)" => ECCCoGenerator(λ=[λ₁, λ₂, 0.0], opt=opt, use_class_loss=use_class_loss, nsamples=nsamples, nmin=nmin),
+            "ECCCo-Δ" => ECCCoGenerator(λ=Λ_Δ, opt=opt, use_class_loss=use_class_loss, use_energy_delta=true, nsamples=nsamples, nmin=nmin),
[…]

### 2023-09-05 experiments/circles.jl

@@ -5,7 +5,6 @@ run_experiment(
     dataname="Circles",
     n_hidden=32,
     α=[1.0, 1.0, 1e-2],
-    sampling_batch_size=nothing,
     sampling_steps=20,
     Λ=[0.25, 0.75, 0.75],
     opt=Flux.Optimise.Descent(0.01),

### 2023-09-05 experiments/models/models.jl

@@ -12,6 +12,7 @@ function prepare_models(exp::Experiment)
         if isnothing(exp.builder)
             builder = default_builder()
         end
+        # Default models:
         if isnothing(exp.models)
             @info "Using default models."
             models = default_models(;
@@ -19,7 +20,26 @@ function prepare_models(exp::Experiment)
                 builder=builder,
                 batch_size=batch_size(exp),
                 sampling_steps=exp.sampling_steps,
+                α=exp.α,
+                n_ens=exp.n_ens,
+                use_ensembling=exp.use_ensembling,
+                finaliser=exp.finaliser,
+                loss=exp.loss,
             )
+        end
+        # Additional models:
+        if !isnothing(exp.additional_models)
+            @info "Using additional models."
+            add_models = Dict{Any,Any}()
+            for (k, mod) in exp.additional_models
+                add_models[k] = mod(;
+                    batch_size=batch_size(exp),
+                    finaliser=exp.finaliser,
+                    loss=exp.loss,
+                )
+            end
+            add_models = Dict(k => mod(;batch_size=batch_size(exp), ) for (k, mod) in exp.additional_models)
+            models = merge!(models, exp.additional_models)
         end
         @info "Training models."
         model_dict = train_models(models, X, labels; cov=exp.coverage)

### 2023-09-05 experiments/post_processing.jl

@@ -36,7 +36,7 @@ function meta_model(outcome::ExperimentOutcome; save_output::Bool=false)
             :n_layers => length(model_dict["MLP"].fitresult[1][1]) - 1,
             :activation => string(exp.activation),
             :n_ens => exp.n_ens,
-            :lambda => exp.string(exp.α[3]),
+            :lambda => string(exp.α[3]),
             :jem_sampling_steps => exp.sampling_steps,
         )
     )
@@ -65,7 +65,7 @@ function meta_generators(outcome::ExperimentOutcome; save_output::Bool=false)
         Dict(
             :opt => string(typeof(opt)),
             :eta => opt.eta,
-            :dataname => dataname,
+            :dataname => exp.dataname,
             :lambda_1 => string(Λ[1]),
             :lambda_2 => string(Λ[2]),
             :lambda_3 => string(Λ[3]),

### 2023-09-05 experiments/run_experiments.jl

@@ -6,6 +6,7 @@ if "run-all" in ARGS
     datanames = ["linearly_separable", "moons", "circles", "mnist", "gmsc"]
 else
     datanames = [ARGS[findall(contains.(ARGS, "data="))][1] |> x -> replace(x, "data=" => "")]
+    datanames = replace.(split(datanames[1], ","), " " => "")
 end
 
 # Linearly Separable

### 2023-09-05 experiments/setup_env.jl

@@ -21,8 +21,11 @@ using Logging
 using MLJBase: multiclass_f1score, accuracy, multiclass_precision, table, machine, fit!
 using MLJEnsembles
 using MLJFlux
+using Random
 using Serialization
 using TidierData
+
+Random.seed!(2023)
 
 # Parallelization:
 plz = nothing

### 2023-09-11 notebooks/prototyping.qmd
slowly slowly

@@
 # Unpack
-exp = outcome.exp
+exper = outcome.exper
 model_dict = outcome.model_dict
@@
 # Unpack
-counterfactual_data = exp.counterfactual_data
+counterfactual_data = exper.counterfactual_data
 X, labels = counterfactual_data.X, counterfactual_data.output_encoder.labels

### 2023-09-11 experiments/benchmarking/benchmarking.jl

@@ -44,25 +44,25 @@ end
 
 Run the benchmarking procedure.
 """
-function run_benchmark(exp::Experiment, model_dict::Dict)
+function run_benchmark(exper::Experiment, model_dict::Dict)
 
-    n_individuals = exp.n_individuals
-    dataname = exp.dataname
-    counterfactual_data = exp.counterfactual_data
-    generator_dict = exp.generators
-    measures = exp.ce_measures
-    parallelizer = exp.parallelizer
+    n_individuals = exper.n_individuals
+    dataname = exper.dataname
+    counterfactual_data = exper.counterfactual_data
+    generator_dict = exper.generators
+    measures = exper.ce_measures
+    parallelizer = exper.parallelizer
 
     # Benchmark generators:
     if isnothing(generator_dict)
         generator_dict = default_generators(;
-            Λ=exp.Λ,
-            Λ_Δ=exp.Λ_Δ,
-            use_variants=exp.use_variants,
-            use_class_loss=exp.use_class_loss,
-            opt=exp.opt,
-            nsamples=exp.nsamples,
-            nmin=exp.nmin,
+            Λ=exper.Λ,
+            Λ_Δ=exper.Λ_Δ,
+            use_variants=exper.use_variants,
+            use_class_loss=exper.use_class_loss,
+            opt=exper.opt,
+            nsamples=exper.nsamples,
+            nmin=exper.nmin,
         )
     end
 

### 2023-09-11 experiments/experiment.jl

@@ -35,65 +35,68 @@ Base.@kwdef struct Experiment
     nmin::Union{Nothing,Int} = nothing
     finaliser::Function = Flux.softmax
     loss::Function = Flux.Losses.crossentropy
+    train_parallel::Bool = false
 end
 
 "A container to hold the results of an experiment."
 mutable struct ExperimentOutcome
-    exp::Experiment
+    exper::Experiment
     model_dict::Union{Nothing, Dict}
     generator_dict::Union{Nothing, Dict}
     bmk::Union{Nothing, Benchmark}
 end
 
 """
-    train_models!(outcome::ExperimentOutcome, exp::Experiment)
+    train_models!(outcome::ExperimentOutcome, exper::Experiment)
 
-Train the models specified by `exp` and store them in `outcome`.
+Train the models specified by `exper` and store them in `outcome`.
 """
-function train_models!(outcome::ExperimentOutcome, exp::Experiment)
-    model_dict = prepare_models(exp)
+function train_models!(outcome::ExperimentOutcome, exper::Experiment)
+    model_dict = prepare_models(exper)
     outcome.model_dict = model_dict
-    meta_model_performance(outcome)
+    if !(is_multi_processed(exper) && MPI.Comm_rank(exper.parallelizer.comm) != 0)
+        meta_model_performance(outcome)
+    end
 end
 
 """
-    benchmark!(outcome::ExperimentOutcome, exp::Experiment)
+    benchmark!(outcome::ExperimentOutcome, exper::Experiment)
 
-Benchmark the models specified by `exp` and store the results in `outcome`.
+Benchmark the models specified by `exper` and store the results in `outcome`.
 """
-function benchmark!(outcome::ExperimentOutcome, exp::Experiment)
-    bmk, generator_dict = run_benchmark(exp, outcome.model_dict)
+function benchmark!(outcome::ExperimentOutcome, exper::Experiment)
+    bmk, generator_dict = run_benchmark(exper, outcome.model_dict)
     outcome.bmk = bmk
     outcome.generator_dict = generator_dict
 end
 
 """
-    run_experiment(exp::Experiment)
+    run_experiment(exper::Experiment)
 
-Run the experiment specified by `exp`.
+Run the experiment specified by `exper`.
 """
-function run_experiment(exp::Experiment; save_output::Bool=true, only_models::Bool=ONLY_MODELS)
+function run_experiment(exper::Experiment; save_output::Bool=true, only_models::Bool=ONLY_MODELS)
     
     # Setup
[…]

### 2023-09-11 src/model.jl

@@ -138,7 +138,7 @@ To keep things consistent with the architecture of `CounterfactualExplanations.j
 Let $\hat{p}_i$ denote the estimated softmax output for feature $i$. Then in the multi-class case the following formula can be applied:
 
 ```math
-\beta_i x_i = \log (\hat{p}_i) + \log (\sum_i \exp(\hat{p}_i))
+\beta_i x_i = \log (\hat{p}_i) + \log (\sum_i \exper(\hat{p}_i))
 ```
 
 For a short derivation, see here: https://math.stackexchange.com/questions/2786600/invert-the-softmax-function. 

### 2023-09-12 AAAI/anonymous-submission-latex-2024.tex
fuck latex seriously

--- 2023-06-01 paper/paper.tex
@@
-\documentclass{article}
+%File: anonymous-submission-latex-2024.tex
+\documentclass[letterpaper]{article} % DO NOT CHANGE THIS
+\usepackage[submission]{aaai24}  % DO NOT CHANGE THIS
+\usepackage{times}  % DO NOT CHANGE THIS
+\usepackage{helvet}  % DO NOT CHANGE THIS
+\usepackage{courier}  % DO NOT CHANGE THIS
+\usepackage[hyphens]{url}  % DO NOT CHANGE THIS
+\usepackage{graphicx} % DO NOT CHANGE THIS
+\urlstyle{rm} % DO NOT CHANGE THIS
+\def\UrlFont{\rm}  % DO NOT CHANGE THIS
+\usepackage{natbib}  % DO NOT CHANGE THIS AND DO NOT ADD ANY OPTIONS TO IT
+\usepackage{caption} % DO NOT CHANGE THIS AND DO NOT ADD ANY OPTIONS TO IT
+\frenchspacing  % DO NOT CHANGE THIS
+\setlength{\pdfpagewidth}{8.5in} % DO NOT CHANGE THIS
+\setlength{\pdfpageheight}{11in} % DO NOT CHANGE THIS
+%
+% These are recommended to typeset algorithms but not required. See the subsubsection on algorithms. Remove them if you don't have algorithms in your paper.
+\usepackage{algorithm}
+\usepackage{algorithmic}
 
-% if you need to pass options to natbib, use, e.g.:
-%     \PassOptionsToPackage{numbers, compress}{natbib}
-% before loading neurips_2023
-
-% ready for submission
-\usepackage{neurips_2023}
-
-% to compile a preprint version, e.g., for submission to arXiv, add add the
-% [preprint] option:
-% \usepackage[preprint]{neurips_2023}
-
-% to compile a camera-ready version, add the [final] option, e.g.:
-%     \usepackage[final]{neurips_2023}
-
-% to avoid loading the natbib package, add option nonatbib:
-%    \usepackage[nonatbib]{neurips_2023}
+%
+% These are are recommended to typeset listings but not required. See the subsubsection on listing. Remove this block if you don't have listings in your paper.
+\usepackage{newfloat}
+\usepackage{listings}
+\DeclareCaptionStyle{ruled}{labelfont=normalfont,labelsep=colon,strut=off} % DO NOT CHANGE THIS
+\lstset{%
+	basicstyle={\footnotesize\ttfamily},% footnotesize acceptable for monospace
+	numbers=left,numberstyle=\footnotesize,xleftmargin=2em,% show line numbers, remove this entire line if you don't want the numbers.
+	aboveskip=0pt,belowskip=0pt,%
+	showstringspaces=false,tabsize=2,breaklines=true}
+\floatstyle{ruled}
+\newfloat{listing}{tb}{lst}{}
+\floatname{listing}{Listing}
+%
+% Keep the \pdfinfo as shown here. There's no need
+% for you to add the /Title and /Author tags.
+\pdfinfo{
+/TemplateVersion (2024.1)
+}
 
@@
 \usepackage{graphicx}
-\usepackage{algorithm}
 \usepackage{algpseudocode}
@@
 
-% Bibliography
-\bibliographystyle{unsrtnat}
-\setcitestyle{numbers,square,comma}
-
 % Numbered Environments:
@@
 
-\title{ECCCos from the Black Box:\\
-Faithful Explanations through\\
-Energy-Constrained Conformal Counterfactuals}
+% DISALLOWED PACKAGES
+% \usepackage{authblk} -- This package is specifically forbidden
+% \usepackage{balance} -- This package is specifically forbidden
+% \usepackage{color (if used in text)
+% \usepackage{CJK} -- This package is specifically forbidden
+% \usepackage{float} -- This package is specifically forbidden
+% \usepackage{flushend} -- This package is specifically forbidden
+% \usepackage{fontenc} -- This package is specifically forbidden
+% \usepackage{fullpage} -- This package is specifically forbidden
+% \usepackage{geometry} -- This package is specifically forbidden
+% \usepackage{grffile} -- This package is specifically forbidden
+% \usepackage{hyperref} -- This package is specifically forbidden
+% \usepackage{navigator} -- This package is specifically forbidden
+% (or any other package that embeds links such as navigator or hyperref)
+% \indentfirst} -- This package is specifically forbidden
+% \layout} -- This package is specifically forbidden
+% \multicol} -- This package is specifically forbidden
+% \nameref} -- This package is specifically forbidden
+% \usepackage{savetrees} -- This package is specifically forbidden
+% \usepackage{setspace} -- This package is specifically forbidden
+% \usepackage{stfloats} -- This package is specifically forbidden
+% \usepackage{tabu} -- This package is specifically forbidden
+% \usepackage{titlesec} -- This package is specifically forbidden
+% \usepackage{tocbibind} -- This package is specifically forbidden
+% \usepackage{ulem} -- This package is specifically forbidden
+% \usepackage{wrapfig} -- This package is specifically forbidden
+% DISALLOWED COMMANDS
+% \nocopyright -- Your paper will not be published if you use this command
+% \addtolength -- This command may not be used
+% \balance -- This command may not be used
+% \baselinestretch -- Your paper will not be published if you use this command
+% \clearpage -- No page breaks of any kind may be used for the final version of your paper
+% \columnsep -- This command may not be used
+% \newpage -- No page breaks of any kind may be used for the final version of your paper
+% \pagebreak -- No page breaks of any kind may be used for the final version of your paperr
+% \pagestyle -- This command may not be used
+% \tiny -- This is not an acceptable font size.
+% \vspace{- -- No negative value may be used in proximity of a caption, figure, table, section, subsection, subsubsection, or reference
+% \vskip{- -- No negative value may be used to alter spacing above or below a caption, figure, table, section, subsection, subsubsection, or reference
 
-% The \author macro works with any number of authors. There are two commands
-% used to separate the names and addresses of multiple authors: \And and \AND.
+\setcounter{secnumdepth}{0} %May be changed to 1 or 2 if section numbers are desired.
+
+% The file aaai24.sty is the style file for AAAI Press
+% proceedings, working notes, and technical reports.
 %
-% Using \And between authors leaves it to LaTeX to determine where to break the
-% lines. Using \AND forces a line break at that point. So, if LaTeX puts 3 of 4
-% authors names on the first line, and the last on the second line, try using
-% \AND instead of \And before the third author name.
 
-\author{%
-  Anonymous Author\thanks{See also: } \\
-  Faculty \\
-  University \\
-  Address \\
-  \texttt{email} \\
-  \And
-  Anonymous Author\thanks{See also: } \\
-  Faculty \\
-  University \\
-  Address \\
-  \texttt{email} \\
+% Title
+
+% Your title must be in mixed case, not sentence case.
+% That means all verbs (including short verbs like be, is, using,and go),
+% nouns, adverbs, adjectives should be capitalized, including both words in hyphenated terms, while
+% articles, conjunctions, and prepositions are lower case unless they
+% directly follow a colon or long dash
+\title{AAAI Press Anonymous Submission\\Instructions for Authors Using \LaTeX{}}
+\author{
+    %Authors
+    % All authors must be in the same font size and format.
+    Written by AAAI Press Staff\textsuperscript{\rm 1}\thanks{With help from the AAAI Publications Committee.}\\
+    AAAI Style Contributions by Pater Patel Schneider,
+    Sunil Issar,\\
+    J. Scott Penberthy,
+    George Ferguson,
+    Hans Guesgen,
+    Francisco Cruz\equalcontrib,
+    Marc Pujol-Gonzalez\equalcontrib
 }
+\affiliations{
+    %Afiliations
+    \textsuperscript{\rm 1}Association for the Advancement of Artificial Intelligence\\
+    % If you have multiple authors and multiple affiliations
+    % use superscripts in text and roman font to identify them.
+    % For example,
+
+    % Sunil Issar\textsuperscript{\rm 2},
+    % J. Scott Penberthy\textsuperscript{\rm 3},
+    % George Ferguson\textsuperscript{\rm 4},
+    % Hans Guesgen\textsuperscript{\rm 5}
+    % Note that the comma should be placed after the superscript
+
+    1900 Embarcadero Road, Suite 101\\
+    Palo Alto, California 94303-3310 USA\\
+    % email address must be in roman text type, not monospace or sans serif
+    proceedings-questions@aaai.org
+%
+% See more examples next
+}
+
+%Example, Single Author, ->> remove \iffalse,\fi and place them surrounding AAAI title to use it
+\iffalse
+\title{My Publication Title --- Single Author}
+\author {
+    Author Name
+}
+\affiliations{
+    Affiliation\\
+    Affiliation Line 2\\
+    name@example.com
+}
+\fi
+
+\iffalse
+%Example, Multiple Authors, ->> remove \iffalse,\fi and place them surrounding AAAI title to use it
+\title{My Publication Title --- Multiple Authors}
+\author {
+    % Authors
+    First Author Name\textsuperscript{\rm 1},
+    Second Author Name\textsuperscript{\rm 2},
+    Third Author Name\textsuperscript{\rm 1}
+}
+\affiliations {
+    % Affiliations
+    \textsuperscript{\rm 1}Affiliation 1\\
+    \textsuperscript{\rm 2}Affiliation 2\\
+    firstAuthor@affiliation1.com, secondAuthor@affilation2.com, thirdAuthor@affiliation1.com
+}
+\fi
+
+% REMOVE THIS: bibentry
+% This is only needed to show inline citations in the guidelines document. You should not need it and can safely delete it.
+\usepackage{bibentry}
+% END REMOVE bibentry
 
@@
 
-\begin{ack}
+  \section{Acknowledgments}
 
@@
 
-\end{ack}
-
   \bibliography{bib}
 
-\pagebreak
-
-\appendix
-\section*{Appendices}
-\renewcommand{\thesubsection}{\Alph{subsection}}
-
-The following appendices provide additional details that are relevant to the paper.
-Appendices~\ref{app:jem} and~\ref{app:cp} explain any tasks related to Energy-Based Modelling and Predictive Uncertainty Quantification through Conformal Prediction, respectively.
-Appendix~\ref{app:eccco} provides additional technical and implementation details about our proposed generator, \textit{ECCCo}, including references to our open-sourced code base.
-A complete overview of our experimental setup detailing our parameter choices, training procedures and initial black-box model performance can be found in Appendix~\ref{app:setup}.
-Finally, Appendix~\ref{app:results} reports all of our experimental results in more detail.
-
-\subsection{Energy-Based Modelling}\label{app:jem}
-
-Since we were not able to identify any existing open-source software for Energy-Based Modelling that would be flexible enough to cater to our needs, we have developed a \texttt{Julia} package from scratch.
-The package has been open-sourced, but to avoid compromising the double-blind review process, we refrain from providing more information at this stage.
-In our development we have heavily drawn on the existing literature:~\citet{du2020implicit} describe best practices for using EBM for generative modelling;~\citet{grathwohl2020your} explain how EBM can be used to train classifiers jointly for the discriminative and generative tasks.
-We have used the same package for training and inference, but there are some important differences between the two cases that are worth highlighting here.
-
-\subsubsection{Training: Joint Energy Models}
-
-To train our Joint Energy Models we broadly follow the approach outlined in~\citet{grathwohl2020your}.
-These models are trained to optimize a hybrid objective that involves a standard classification loss component $L_{\text{clf}}(\theta)=-\log p_{\theta}(\mathbf{y}|\mathbf{x})$ (e.g.
-cross-entropy loss) as well as a generative loss component $L_{\text{gen}}(\theta)=-\log p_{\theta}(\mathbf{x})$.
-
-To draw samples from $p_{\theta}(\mathbf{x})$, we rely exclusively on the conditional sampling approach described in~\citet{grathwohl2020your} for both training and inference: we first draw $\mathbf{y}\sim p(\mathbf{y})$ and then sample $\mathbf{x} \sim p_{\theta}(\mathbf{x}|\mathbf{y})$~\citep{grathwohl2020your} via Equation~\ref{eq:sgld} with energy $\mathcal{E}(\mathbf{x}|\mathbf{y})=\mu_{\theta}(\mathbf{x})[\mathbf{y}]$ where $\mu_{\theta}: \mathcal{X} \mapsto \mathbb{R}^K$ returns the linear predictions (logits) of our classifier $M_{\theta}$.
-While our package also supports unconditional sampling, we found conditional sampling to work well.
-It is also well aligned with CE, since in this context we are interested in conditioning on the target class.
-
-As mentioned in the body of the paper, we rely on a biased sampler involving separately specified values for the step size $\epsilon$ and the standard deviation $\sigma$ of the stochastic term involving $\mathbf{r}$.
-Formally, our biased sampler performs updates as follows:
-
-\begin{equation}\label{eq:biased-sgld}
-  \begin{aligned}
-    \hat{\mathbf{x}}_{j+1} &\leftarrow \hat{\mathbf{x}}_j - \frac{\epsilon}{2} \mathcal{E}(\hat{\mathbf{x}}_j|\mathbf{y}^+) + \sigma \mathbf{r}_j, && j=1,...,J
-  \end{aligned}
-\end{equation}
-
-Consistent with~\citet{grathwohl2020your}, we have specified $\epsilon=2$ and $\sigma=0.01$ as the default values for all of our experiments.
-The number of total SGLD steps $J$ varies by dataset (Table~\ref{tab:ebmparams}).
-Following best practices, we initialize $\mathbf{x}_0$ randomly in 5\% of all cases and sample from a buffer in all other cases.
-The buffer itself is randomly initialised and gradually grows to a maximum of 10,000 samples during training as $\hat{\mathbf{x}}_{J}$ is stored in each epoch~\citep{du2020implicit,grathwohl2020your}.
-
-It is important to realise that sampling is done during each training epoch, which makes training Joint Energy Models significantly harder than conventional neural classifiers.
-In each epoch the generated (batch of) sample(s) $\hat{\mathbf{x}}_{J}$ is used as part of the generative loss component, which compares its energy to that of observed samples $\mathbf{x}$: $L_{\text{gen}}(\theta)=\mu_{\theta}(\mathbf{x})[\mathbf{y}]-\mu_{\theta}(\hat{\mathbf{x}}_{J})[\mathbf{y}]$.
-Our full training objective can be summarized as follows,
-
-\begin{equation}\label{eq:jem-loss}
-  \begin{aligned}
-    L(\theta) &= L_{\text{clf}}(\theta) + L_{\text{gen}}(\theta) + \lambda L_{\text{reg}}(\theta) 
-  \end{aligned}
-\end{equation}
-
-where $L_{\text{reg}}(\theta)$ is a Ridge penalty (L2 norm) that regularises energy magnitudes for both observed and generated samples~\citep{du2020implicit}.
-We have used varying degrees of regularization depending on the dataset ($\lambda$ in Table~\ref{tab:ebmparams}).
-
-Contrary to existing work, we have not typically used the entire minibatch of training data for the generative loss component but found that using a subset of the minibatch was often sufficient in attaining decent generative performance (Table~\ref{tab:ebmparams}).
-This has helped to reduce the computational burden for our models, which should make it easier for others to reproduce our findings.
-Figures~\ref{fig:mnist-gen} and~\ref{fig:moons-gen} show generated samples for our \textit{MNIST} and \textit{Moons} data, to provide a sense of their generative property.
-
-\import{contents/}{table_ebm_params.tex}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=0.75\textwidth]{../artifacts/results/images/mnist_generated_JEM Ensemble.png}
-  \caption{Conditionally generated \textit{MNIST} images for our JEM Ensemble.}\label{fig:mnist-gen}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=0.5\textwidth]{../artifacts/results/images/moons_generated_JEM.png}
-  \caption{Conditionally generated samples (stars) for our \textit{Moons} data using a JEM.}\label{fig:moons-gen}
-\end{figure}
-\subsubsection{Inference: Quantifying Models' Generative Property}
-
-At inference time, we assume no prior knowledge about the model's generative property.
-This means that we do not tab into the existing buffer of generated samples for our Joint Energy Models, but instead generate conditional samples from scratch.
-While we have relied on the default values $\epsilon=2$ and $\sigma=0.01$ also during inference, the number of total SGLD steps was set to $J=500$ in all cases, so significantly higher than during training.
-For all of our synthetic datasets and models, we generated 50 conditional samples and then formed subsets containing the $n_{E}=25$ lowest-energy samples.
-While in practice it would be sufficient to do this once for each model and dataset, we have chosen to perform sampling separately for each individual counterfactual in our experiments to account for stochasticity.
-To help reduce the computational burden for our real-world datasets we have generated only 10 conditional samples each time and used all of them in our counterfactual search.
-Using more samples, as we originally did, had no substantial impact on our results.
-
-\subsection{Conformal Prediction}\label{app:cp}
-
-In this Appendix~\ref{app:cp} we provide some more background on CP and explain in some more detail how we have used recent advances in Conformal Training for our purposes.
-
-\subsubsection{Background on CP}
-
-Intuitively, CP works under the premise of turning heuristic notions of uncertainty into rigorous uncertainty estimates by repeatedly sifting through the data.
-It can be used to generate prediction intervals for regression models and prediction sets for classification models.
-Since the literature on CE and AR is typically concerned with classification problems, we focus on the latter.
-A particular variant of CP called Split Conformal Prediction (SCP) is well-suited for our purposes, because it imposes only minimal restrictions on model training.
-
-Specifically, SCP involves splitting the data $\mathcal{D}_n=\{(\mathbf{x}_i,\mathbf{y}_i)\}_{i=1,...,n}$ into a proper training set $\mathcal{D}_{\text{train}}$ and a calibration set $\mathcal{D}_{\text{cal}}$.
-The former is used to train the classifier in any conventional fashion.
-The latter is then used to compute so-called nonconformity scores: $\mathcal{S}=\{s(\mathbf{x}_i,\mathbf{y}_i)\}_{i \in \mathcal{D}_{\text{cal}}}$ where $s: (\mathcal{X},\mathcal{Y}) \mapsto \mathbb{R}$ is referred to as \textit{score function}.
-In the context of classification, a common choice for the score function is just $s_i=1-M_{\theta}(\mathbf{x}_i)[\mathbf{y}_i]$, that is one minus the softmax output corresponding to the observed label $\mathbf{y}_i$~\citep{angelopoulos2021gentle}.
-
-Finally, classification sets are formed as follows,
-
-\begin{equation}\label{eq:scp}
-  \begin{aligned}
-    C_{\theta}(\mathbf{x}_i;\alpha)=\{\mathbf{y}: s(\mathbf{x}_i,\mathbf{y}) \le \hat{q}\}
-  \end{aligned}
-\end{equation}
-
-where $\hat{q}$ denotes the $(1-\alpha)$-quantile of $\mathcal{S}$ and $\alpha$ is a predetermined error rate.
-As the size of the calibration set increases, the probability that the classification set $C(\mathbf{x}_{\text{test}})$ for a newly arrived sample $\mathbf{x}_{\text{test}}$ does not cover the true test label $\mathbf{y}_{\text{test}}$ approaches $\alpha$~\citep{angelopoulos2021gentle}.
-
-Observe from Equation~\ref{eq:scp} that Conformal Prediction works on an instance-level basis, much like CE are local.
-The prediction set for an individual instance $\mathbf{x}_i$ depends only on the characteristics of that sample and the specified error rate.
-Intuitively, the set is more likely to include multiple labels for samples that are difficult to classify, so the set size is indicative of predictive uncertainty.
-To see why this effect is exacerbated by small choices for $\alpha$ consider the case of $\alpha=0$, which requires that the true label is covered by the prediction set with probability equal to 1.
-
-\subsubsection{Differentiability}
-
-The fact that conformal classifiers produce set-valued predictions introduces a challenge: it is not immediately obvious how to use such classifiers in the context of gradient-based counterfactual search.
-Put differently, it is not clear how to use prediction sets in Equation~\ref{eq:general}.
-Fortunately, \citet{stutz2022learning} have recently proposed a framework for Conformal Training that also hinges on differentiability.
-Specifically, they show how Stochastic Gradient Descent can be used to train classifiers not only for the discriminative task but also for additional objectives related to Conformal Prediction.
-One such objective is \textit{efficiency}: for a given target error rate $\alpha$, the efficiency of a conformal classifier improves as its average prediction set size decreases.
-To this end, the authors introduce a smooth set size penalty defined in Equation~\ref{eq:setsize} in the body of this paper.
-Formally, it is defined as $C_{\theta,\mathbf{y}}(\mathbf{x}_i;\alpha):=\sigma\left((s(\mathbf{x}_i,\mathbf{y})-\alpha) T^{-1}\right)$ for $\mathbf{y}\in\mathcal{Y}$, where $\sigma$ is the sigmoid function and $T$ is a hyper-parameter used for temperature scaling~\citep{stutz2022learning}.
-
-In addition to the smooth set size penalty,~\citet{stutz2022learning} also propose a configurable classification loss function, that can be used to enforce coverage.
-For \textit{MNIST} data, we found that using this function generally improved the visual quality of the generated counterfactuals, so we used it in our experiments involving real-world data.
-For the synthetic dataset, visual inspection of the counterfactuals showed that using the configurable loss function sometimes led to overshooting: counterfactuals would end up deep inside the target domain but far away from the observed samples.
-For this reason, we instead relied on standard cross-entropy loss for our synthetic datasets.
-As we have noted in the body of the paper, more experimental work is certainly needed in this context.
-Figure~\ref{fig:cp-diff} shows the prediction set size (left), smooth set size loss (centre) and configurable classification loss (right) for a \textit{JEM} trained on our \textit{Linearly Separable} data.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=1.0\textwidth]{../artifacts/results/images/poc_set_size.png}
-  \caption{Prediction set size (left), smooth set size loss (centre) and configurable classification loss (right) for a JEM trained on our \textit{Linearly Separable} data.}\label{fig:cp-diff}
-\end{figure}
-
-\subsection{ECCCo}\label{app:eccco}
-
-In this section, we briefly discuss convergence conditions for CE and provide details concerning the actual implementation of our framework in \texttt{Julia}.  
-\subsubsection{A Note on Convergence}
-
-Convergence is not typically discussed much in the context of CE, even though it has important implications on outcomes.
-One intuitive way to specify convergence is in terms of threshold probabilities: once the predicted probability $p(\mathbf{y}^+|\mathbf{x}^{\prime})$ exceeds some user-defined threshold $\gamma$ such that the counterfactual is valid, we could consider the search to have converged.
-In the binary case, for example, convergence could be defined as $p(\mathbf{y}^+|\mathbf{x}^{\prime})>0.5$ in this sense.
-Note, however, how this can be expected to yield counterfactuals in the proximity of the decision boundary, a region characterized by high aleatoric uncertainty.
-In other words, counterfactuals generated in this way would generally not be plausible.
-To avoid this from happening, we specify convergence in terms of gradients approaching zero for all our experiments and all of our generators.
-This is allows us to get a cleaner read on how the different counterfactual search objectives affect counterfactual outcomes.
-
-\subsubsection{\texttt{ECCCo.jl}}
-
-The core part of our code base is integrated into a larger ecosystem of \texttt{Julia} packages that we are actively developing and maintaining.
-To avoid compromising the double-blind review process, we only provide a link to an anonymized repository at this stage: \url{https://anonymous.4open.science/r/ECCCo-1252/README.md}.
-
-\subsection{Experimental Setup}\label{app:setup}
-
-Table~\ref{tab:params} provides an overview of all parameters related to our experiments.
-The \textit{GMSC} data were randomly undersampled for balancing purposes and all features were standardized.
-\textit{MNIST} data was also randomly undersampled for reasons outlined below.
-Pixel values were preprocessed to fall in the range of $[-1,1]$ and a small Gaussian noise component ($\sigma=0.03$) was added to training samples following common practice in the EBM literature.
-All of our models were trained through mini-batch training using the Adam optimiser (\citet{kingma2017adam}).
-Table~\ref{tab:perf} shows standard evaluation metrics measuring the predictive performance of our different models grouped by dataset.
-These measures were computed on test data.
-
-Table~\ref{tab:genparams} summarises our hyperparameter choices for the counterfactual generators where $\eta$ denotes the learning rate used for Stochastic Gradient Descent (SGD) and $\lambda_1$, $\lambda_2$, $\lambda_3$ represent the chosen penalty strengths (Equations~\ref{eq:general} and~\ref{eq:eccco}).
-Here $\lambda_1$ also refers to the chosen penalty for the distance from factual values that applies to both \textit{Wachter} and \textit{REVISE}, but not \textit{Schut} which is penalty-free.
-\textit{Schut} is also the only generator that uses JSMA instead of SGD for optimization.
-
-\import{contents/}{table_params.tex}
-
-\import{contents/}{table_perf.tex}
-
-\import{contents/}{table_gen_params.tex}
-
-\subsubsection{Compute}
-
-To enable others to easily replicate our experiments, we have chosen to work with small neural network architectures and randomly undersampled the \textit{MNIST} dataset (maintaining class balance).
-All of our experiments could then be run locally on a personal machine.
-The longest runtimes we experienced for model training and counterfactual benchmarking were on the order of 8-12 hours (\textit{MNIST} data).
-For the synthetic data, all experiments could be completed in less than an hour.
-
-We have summarised our system information below:
-
-\textbf{Software}:
-
-\begin{itemize}
-  \item System Version: macOS 13.3.1
-  \item Kernel Version: Darwin 22.4.0
-\end{itemize}
-
-\textbf{Hardware}:
-
-\begin{itemize}
-  \item Model Name: MacBook Pro
-  \item Model Identifier: MacBookPro16,1
-  \item Processor Name: 8-Core Intel Core i9
-  \item Processor Speed: 2.3 GHz
-  \item Number of Processors: 1
-  \item Total Number of Cores: 8
-  \item L2 Cache (per Core): 256 KB
-  \item L3 Cache: 16 MB
-  \item Hyper-Threading Technology: Enabled
-  \item Memory: 32 GB
-\end{itemize}
-
-\subsection{Results}\label{app:results}
-
-Figure~\ref{fig:mnist-eccco} shows examples of counterfactuals for \textit{MNIST} data where the underlying model is our \textit{JEM Ensemble}.
-Original images are shown on the diagonal and the corresponding counterfactuals are plotted across rows.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=0.9\textwidth]{../artifacts/results/images/mnist_eccco_all_digits.png}
-  \caption{Counterfactuals for \textit{MNIST} data and our \textit{JEM Ensemble}.
-Original images are shown on the diagonal with the corresponding counterfactuals plotted across rows.}\label{fig:mnist-eccco}
-\end{figure}
-
-Table~\ref{tab:results-full} reports all of the evaluation metrics we have computed.
-Table~\ref{tab:results-full-valid} reports the same metrics for the subset of valid counterfactuals.
-The `Unfaithfulness' and `Implausibility' metrics have been discussed extensively in the body of the paper.
-The `Cost' metric relates to the distance between the factual and the counterfactual.
-The `Redundancy' metric measures sparsity in is defined as the percentage of features that remain unperturbed (higher is better).
-The `Uncertainty' metric is just the average value of the smooth set size penalty (Equation~\ref{eq:setsize}).
-Finally, `Validity' is the percentage of valid counterfactuals.
-
-\import{contents/}{table_all.tex}
-
-\import{contents/}{table_all_valid.tex}
-
 \end{document}

### 2023-09-12 AAAI/paper.tex
latex template set up

--- 2023-09-12 AAAI/anonymous-submission-latex-2024.tex
@@
 \usepackage{algorithm}
-\usepackage{algorithmic}
+% \usepackage{algorithmic}
 
@@
 % These are are recommended to typeset listings but not required. See the subsubsection on listing. Remove this block if you don't have listings in your paper.
-\usepackage{newfloat}
-\usepackage{listings}
-\DeclareCaptionStyle{ruled}{labelfont=normalfont,labelsep=colon,strut=off} % DO NOT CHANGE THIS
-\lstset{%
-	basicstyle={\footnotesize\ttfamily},% footnotesize acceptable for monospace
-	numbers=left,numberstyle=\footnotesize,xleftmargin=2em,% show line numbers, remove this entire line if you don't want the numbers.
-	aboveskip=0pt,belowskip=0pt,%
-	showstringspaces=false,tabsize=2,breaklines=true}
-\floatstyle{ruled}
-\newfloat{listing}{tb}{lst}{}
-\floatname{listing}{Listing}
+% \usepackage{newfloat}
+% \usepackage{listings}
+% \DeclareCaptionStyle{ruled}{labelfont=normalfont,labelsep=colon,strut=off} % DO NOT CHANGE THIS
+% \lstset{%
+% 	basicstyle={\footnotesize\ttfamily},% footnotesize acceptable for monospace
+% 	numbers=left,numberstyle=\footnotesize,xleftmargin=2em,% show line numbers, remove this entire line if you don't want the numbers.
+% 	aboveskip=0pt,belowskip=0pt,%
+% 	showstringspaces=false,tabsize=2,breaklines=true}
+% \floatstyle{ruled}
+% \newfloat{listing}{tb}{lst}{}
+% \floatname{listing}{Listing}
 %
@@
 
-\usepackage[utf8]{inputenc} % allow utf-8 input
-\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
-\usepackage{hyperref}       % hyperlinks
-\usepackage{url}            % simple URL typesetting
-\usepackage{booktabs}       % professional-quality tables
 \usepackage{amsfonts}       % blackboard math symbols
-\usepackage{nicefrac}       % compact symbols for 1/2, etc.
-\usepackage{microtype}      % microtypography
-\usepackage{xcolor}         % colors
-
 \usepackage{amsmath}
@@
 \newtheorem{question}{Research Question}[section]
+
+% Bibliography
+% \bibliographystyle{unsrtnat}
+% \setcitestyle{numbers,square,comma}
+
+% Algorithm
+\renewcommand{\algorithmicrequire}{\textbf{Input:}}
+\renewcommand{\algorithmicensure}{\textbf{Output:}}
 
@@
     \centering
-    \includegraphics[width=0.8\textwidth]{../artifacts/results/images/mnist_motivation.png}
+    \includegraphics[width=0.4\textwidth]{../artifacts/results/images/mnist_motivation.png}
     \caption{Counterfactuals for turning a 9 (nine) into a 7 (seven): original image (left); then from left to right the counterfactuals generated using \textit{Wachter}, \textit{Schut} and \textit{REVISE}.}\label{fig:motiv}
@@
     \centering
-    \includegraphics[width=1.0\textwidth]{../artifacts/results/images/poc_gradient_fields.png}
+    \includegraphics[width=0.5\textwidth]{../artifacts/results/images/poc_gradient_fields.png}
     \caption{Gradient fields and counterfactual paths for different generators.

### 2023-09-12 paper/appendix.tex
latex cleaned up

--- 2023-06-01 paper/paper.tex
@@
-\documentclass{article}
-
-% if you need to pass options to natbib, use, e.g.:
-%     \PassOptionsToPackage{numbers, compress}{natbib}
-% before loading neurips_2023
-
-% ready for submission
-\usepackage{neurips_2023}
-
-% to compile a preprint version, e.g., for submission to arXiv, add add the
-% [preprint] option:
-% \usepackage[preprint]{neurips_2023}
-
-% to compile a camera-ready version, add the [final] option, e.g.:
-%     \usepackage[final]{neurips_2023}
-
-% to avoid loading the natbib package, add option nonatbib:
-%    \usepackage[nonatbib]{neurips_2023}
-
-\usepackage[utf8]{inputenc} % allow utf-8 input
-\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
-\usepackage{hyperref}       % hyperlinks
-\usepackage{url}            % simple URL typesetting
-\usepackage{booktabs}       % professional-quality tables
-\usepackage{amsfonts}       % blackboard math symbols
-\usepackage{nicefrac}       % compact symbols for 1/2, etc.
-\usepackage{microtype}      % microtypography
-\usepackage{xcolor}         % colors
-
-\usepackage{amsmath}
-\usepackage{amsthm}
-\usepackage{caption}
-\usepackage{graphicx}
-\usepackage{algorithm}
-\usepackage{algpseudocode}
-\usepackage{import}
-\usepackage{booktabs}
-\usepackage{longtable}
-\usepackage{array}
-\usepackage{multirow}
-
-% Bibliography
-\bibliographystyle{unsrtnat}
-\setcitestyle{numbers,square,comma}
-
-% Numbered Environments:
-\newtheorem{definition}{Definition}[section]
-\newtheorem{question}{Research Question}[section]
-
-\title{ECCCos from the Black Box:\\
-Faithful Explanations through\\
-Energy-Constrained Conformal Counterfactuals}
-
-% The \author macro works with any number of authors. There are two commands
-% used to separate the names and addresses of multiple authors: \And and \AND.
-%
-% Using \And between authors leaves it to LaTeX to determine where to break the
-% lines. Using \AND forces a line break at that point. So, if LaTeX puts 3 of 4
-% authors names on the first line, and the last on the second line, try using
-% \AND instead of \And before the third author name.
-
-\author{%
-  Anonymous Author\thanks{See also: } \\
-  Faculty \\
-  University \\
-  Address \\
-  \texttt{email} \\
-  \And
-  Anonymous Author\thanks{See also: } \\
-  Faculty \\
-  University \\
-  Address \\
-  \texttt{email} \\
-}
-
-\begin{document}
-
-\maketitle
-
-\begin{abstract}
-  Counterfactual Explanations offer an intuitive and straightforward way to explain black-box models and offer Algorithmic Recourse to individuals.
-To address the need for plausible explanations, existing work has primarily relied on surrogate models to learn how the input data is distributed.
-This effectively reallocates the task of learning realistic explanations for the data from the model itself to the surrogate.
-Consequently, the generated explanations may seem plausible to humans but need not necessarily describe the behaviour of the black-box model faithfully.
-We formalise this notion of faithfulness through the introduction of a tailored evaluation metric and propose a novel algorithmic framework for generating \textbf{E}nergy-\textbf{C}onstrained \textbf{C}onformal \textbf{Co}unterfactuals (ECCCos) that are only as plausible as the model permits.
-Through extensive empirical studies, we demonstrate that ECCCos reconcile the need for faithfulness and plausibility.
-In particular, we show that for models with gradient access, it is possible to achieve state-of-the-art performance without the need for surrogate models.
-To do so, our framework relies solely on properties defining the black-box model itself by leveraging recent advances in Energy-Based Modelling and Conformal Prediction.
-To our knowledge, this is the first venture in this direction for generating faithful Counterfactual Explanations.
-Thus, we anticipate that ECCCos can serve as a baseline for future research.
-We believe that our work opens avenues for researchers and practitioners seeking tools to better distinguish trustworthy from unreliable models.
-\end{abstract}
-
-\section{Introduction}\label{intro}
-
-Counterfactual Explanations (CE) provide a powerful, flexible and intuitive way to not only explain black-box models but also help affected individuals through the means of Algorithmic Recourse.
-Instead of opening the Black Box, CE works under the premise of strategically perturbing model inputs to understand model behaviour~\citep{wachter2017counterfactual}.
-Intuitively speaking, we generate explanations in this context by asking what-if questions of the following nature: `Our credit risk model currently predicts that this individual is not credit-worthy.
-What if they reduced their monthly expenditures by 10\%?'
-
-This is typically implemented by defining a target outcome $\mathbf{y}^+ \in \mathcal{Y}$ for some individual $\mathbf{x} \in \mathcal{X}=\mathbb{R}^D$ described by $D$ attributes, for which the model $M_{\theta}:\mathcal{X}\mapsto\mathcal{Y}$ initially predicts a different outcome: $M_{\theta}(\mathbf{x})\ne \mathbf{y}^+$.
-Counterfactuals are then searched by minimizing a loss function that compares the predicted model output to the target outcome: $\text{yloss}(M_{\theta}(\mathbf{x}),\mathbf{y}^+)$.
-Since CE work directly with the black-box model, valid counterfactuals always have full local fidelity by construction where fidelity is defined as the degree to which explanations approximate the predictions of a black-box model~\citep{mothilal2020explaining,molnar2020interpretable}.
-
-In situations where full fidelity is a requirement, CE offer a more appropriate solution to Explainable Artificial Intelligence (XAI) than other popular approaches like LIME~\citep{ribeiro2016why} and SHAP~\citep{lundberg2017unified}, which involve local surrogate models.
-But even full fidelity is not a sufficient condition for ensuring that an explanation faithfully describes the behaviour of a model.
-That is because multiple very distinct explanations can all lead to the same model prediction, especially when dealing with heavily parameterized models like deep neural networks, which are typically underspecified by the data~\citep{wilson2020case}.
-
-In the context of CE, the idea that no two explanations are the same arises almost naturally.
-A key focus in the literature has therefore been to identify those explanations and algorithmic recourses that are most appropriate based on a myriad of desiderata such as sparsity, actionability and plausibility.
-In this work, we draw closer attention to model faithfulness rather than fidelity as a desideratum for counterfactuals.
-Our key contributions are as follows:
-
-\begin{itemize}
-  \item We show that fidelity is an insufficient evaluation metric for counterfactuals (Section~\ref{fidelity}) and propose a definition of faithfulness that gives rise to more suitable metrics (Section~\ref{faithfulness}).
-  \item We introduce a novel algorithmic approach for generating Energy-Constrained Conformal Counterfactuals (ECCCos) in Section~\ref{meth}.
-  \item We provide extensive empirical evidence demonstrating that ECCCos faithfully explain model behaviour and attain plausibility only when appropriate (Section~\ref{emp}).
-\end{itemize}
-
-To our knowledge, this is the first venture in this direction for generating faithful counterfactuals.
-Thus, we anticipate that ECCCos can serve as a baseline for future research.
-We believe that our work opens avenues for researchers and practitioners seeking tools to better distinguish trustworthy from unreliable models.
-
-\section{Background}\label{background}
-
-While CE can also be generated for arbitrary regression models~\citep{spooner2021counterfactual}, existing work has primarily focused on classification problems.
-Let $\mathcal{Y}=(0,1)^K$ denote the one-hot-encoded output domain with $K$ classes.
-Then most counterfactual generators rely on gradient descent to optimize different flavours of the following counterfactual search objective:
-
-\begin{equation} \label{eq:general}
-\begin{aligned}
-\mathbf{Z}^\prime &= \arg \min_{\mathbf{Z}^\prime \in \mathcal{Z}^L} \left\{  {\text{yloss}(M_{\theta}(f(\mathbf{Z}^\prime)),\mathbf{y}^+)}+ \lambda {\text{cost}(f(\mathbf{Z}^\prime)) }  \right\} 
-\end{aligned} 
-\end{equation}
-
-Here $\text{yloss}(\cdot)$ denotes the primary loss function, $f(\cdot)$ is a function that maps from the counterfactual state space to the feature space and $\text{cost}(\cdot)$ is either a single penalty or a collection of penalties that are used to impose constraints through regularization.
-Equation~\ref{eq:general} restates the baseline approach to gradient-based counterfactual search proposed by~\citet{wachter2017counterfactual} in general form as introduced by~\citet{altmeyer2023endogenous}.
-To explicitly account for the multiplicity of explanations, $\mathbf{Z}^\prime=\{ \mathbf{z}_l\}_L$ denotes an $L$-dimensional array of counterfactual states.
-
-The baseline approach, which we will simply refer to as \textit{Wachter}, searches a single counterfactual directly in the feature space and penalises its distance to the original factual.
-In this case, $f(\cdot)$ is simply the identity function and $\mathcal{Z}$ corresponds to the feature space itself.
-Many derivative works of~\citet{wachter2017counterfactual} have proposed new flavours of Equation~\ref{eq:general}, each of them designed to address specific \textit{desiderata} that counterfactuals ought to meet in order to properly serve both AI practitioners and individuals affected by algorithmic decision-making systems.
-The list of desiderata includes but is not limited to the following: sparsity, proximity~\citep{wachter2017counterfactual}, actionability~\citep{ustun2019actionable}, diversity~\citep{mothilal2020explaining}, plausibility~\citep{joshi2019realistic,poyiadzi2020face,schut2021generating}, robustness~\citep{upadhyay2021robust,pawelczyk2022probabilistically,altmeyer2023endogenous} and causality~\citep{karimi2021algorithmic}.
-Different counterfactual generators addressing these needs have been extensively surveyed and evaluated in various studies~\citep{verma2020counterfactual,karimi2020survey,pawelczyk2021carla,artelt2021evaluating,guidotti2022counterfactual}.
-
-Perhaps unsurprisingly, the different desiderata are often positively correlated.
-For example, \citet{artelt2021evaluating} find that plausibility typically also leads to improved robustness.
-Similarly, plausibility has also been connected to causality in the sense that plausible counterfactuals respect causal relationships~\citep{mahajan2020preserving}.
-Consequently, the plausibility of counterfactuals has been among the primary concerns for researchers.
-Achieving plausibility is equivalent to ensuring that the generated counterfactuals comply with the true and unobserved data-generating process (DGP).
-We define plausibility formally in this work as follows:
-
-\begin{definition}[Plausible Counterfactuals]
-  \label{def:plausible}
-  Let $\mathcal{X}|\mathbf{y}^+= p(\mathbf{x}|\mathbf{y}^+)$ denote the true conditional distribution of samples in the target class $\mathbf{y}^+$.
-Then for $\mathbf{x}^{\prime}$ to be considered a plausible counterfactual, we need: $\mathbf{x}^{\prime} \sim \mathcal{X}|\mathbf{y}^+$.
-\end{definition}
-
-To generate plausible counterfactuals, we need to be able to quantify the DGP: $\mathcal{X}|\mathbf{y}^+$.
-One straightforward way to do this is to use surrogate models for the task.
-\citet{joshi2019realistic}, for example, suggest that instead of searching counterfactuals in the feature space $\mathcal{X}$, we can instead traverse a latent embedding $\mathcal{Z}$ (Equation~\ref{eq:general}) that implicitly codifies the DGP.
-To learn the latent embedding, they propose using a generative model such as a Variational Autoencoder (VAE).
-Provided the surrogate model is well-specified, their proposed approach called \textit{REVISE} can yield plausible explanations.
-Others have proposed similar approaches: \citet{dombrowski2021diffeomorphic} traverse the base space of a normalizing flow to solve Equation~\ref{eq:general}; \citet{poyiadzi2020face} use density estimators ($\hat{p}: \mathcal{X} \mapsto [0,1]$) to constrain the counterfactuals to dense regions in the feature space; and, finally, \citet{karimi2021algorithmic} assume knowledge about the structural causal model that generates the data.
-
-A competing approach towards plausibility that is also closely related to this work instead relies on the black-box model itself.
-\citet{schut2021generating} show that to meet the plausibility objective we need not explicitly model the input distribution.
-Pointing to the undesirable engineering overhead induced by surrogate models, they propose that we rely on the implicit minimisation of predictive uncertainty instead.
-Their proposed methodology, which we will refer to as \textit{Schut}, solves Equation~\ref{eq:general} by greedily applying Jacobian-Based Saliency Map Attacks (JSMA) in the feature space with cross-entropy loss and no penalty at all.
-The authors demonstrate theoretically and empirically that their approach yields counterfactuals for which the model $M_{\theta}$ predicts the target label $\mathbf{y}^+$ with high confidence.
-Provided the model is well-specified, these counterfactuals are plausible.
-This idea hinges on the assumption that the black-box model provides well-calibrated predictive uncertainty estimates.
-
-\section{Why Fidelity is not Enough}\label{fidelity}
-
-As discussed in the introduction, any valid counterfactual also has full fidelity by construction: solutions to Equation~\ref{eq:general} are considered valid as soon as the label predicted by the model matches the target class.
-So while fidelity always applies, counterfactuals that address the various desiderata introduced above can look vastly different from each other.
-
-To demonstrate this with an example, we have trained a simple image classifier $M_{\theta}$ on the well-known \textit{MNIST} dataset~\citep{lecun1998mnist}: a Multi-Layer Perceptron (\textit{MLP}) with above 90 percent test accuracy.
-No measures have been taken to improve the model's adversarial robustness or its capacity for predictive uncertainty quantification.
-The far left panel of Figure ~\ref{fig:motiv} shows a random sample drawn from the dataset.
-The underlying classifier correctly predicts the label `nine' for this image.
-For the given factual image and model, we have used \textit{Wachter}, \textit{Schut} and \textit{REVISE} to generate one counterfactual each in the target class `seven'.
-The perturbed images are shown next to the factual image from left to right in Figure ~\ref{fig:motiv}.
-Captions on top of the individual images indicate the generator along with the predicted probability that the image belongs to the target class.
-In all three cases that probability is above 90 percent and yet the counterfactuals look very different from each other.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=0.8\textwidth]{../artifacts/results/images/mnist_motivation.png}
-  \caption{Counterfactuals for turning a 9 (nine) into a 7 (seven): original image (left); then from left to right the counterfactuals generated using \textit{Wachter}, \textit{Schut} and \textit{REVISE}.}\label{fig:motiv}
-\end{figure}
-
-Since \textit{Wachter} is only concerned with proximity, the generated counterfactual is almost indistinguishable from the factual.
-The approach by~\citet{schut2021generating} expects a well-calibrated model that can generate predictive uncertainty estimates.
-Since this is not the case, the generated counterfactual looks like an adversarial example.
-Finally, the counterfactual generated by \textit{REVISE} looks much more plausible than the other two.
-But is it also more faithful to the behaviour of our \textit{MNIST} classifier?
-That is much less clear because the surrogate used by \textit{REVISE} introduces friction: the generated explanations no longer depend exclusively on the black-box model itself.
-
-So which of the counterfactuals most faithfully explains the behaviour of our image classifier?
-Fidelity cannot help us to make that judgement, because all of these counterfactuals have full fidelity.
-Thus, fidelity is an insufficient evaluation metric to assess the faithfulness of CE.
-
-\section{A New Notion of Faithfulness}\label{faithfulness}
-
-Considering the limitations of fidelity as demonstrated in the previous section, analogous to Definition~\ref{def:plausible}, we introduce a new notion of faithfulness in the context of CE:
-
-\begin{definition}[Faithful Counterfactuals]
-  \label{def:faithful}
-  Let $\mathcal{X}_{\theta}|\mathbf{y}^+ = p_{\theta}(\mathbf{x}|\mathbf{y}^+)$ denote the conditional distribution of $\mathbf{x}$ in the target class $\mathbf{y}^+$, where $\theta$ denotes the parameters of model $M_{\theta}$.
-Then for $\mathbf{x}^{\prime}$ to be considered a faithful counterfactual, we need: $\mathbf{x}^{\prime} \sim \mathcal{X}_{\theta}|\mathbf{y}^+$.
-\end{definition}
-
-In doing this, we merge in and nuance the concept of plausibility (Definition~\ref{def:plausible}) where the notion of `consistent with the data' becomes `consistent with what the model has learned about the data'.
-
-\subsection{Quantifying the Model's Generative Property}
-
-To assess counterfactuals with respect to Definition~\ref{def:faithful}, we need a way to quantify the posterior conditional distribution $p_{\theta}(\mathbf{x}|\mathbf{y}^+)$.
-To this end, we draw on recent advances in Energy-Based Modelling (EBM), a subdomain of machine learning that is concerned with generative or hybrid modelling~\citep{grathwohl2020your,du2020implicit}.
-In particular, note that if we fix $\mathbf{y}$ to our target value $\mathbf{y}^+$, we can conditionally draw from $p_{\theta}(\mathbf{x}|\mathbf{y}^+)$ by randomly initializing $\mathbf{x}_0$ and then using Stochastic Gradient Langevin Dynamics (SGLD) as follows,
-
-\begin{equation}\label{eq:sgld}
-  \begin{aligned}
-    \mathbf{x}_{j+1} &\leftarrow \mathbf{x}_j - \frac{\epsilon^2}{2} \mathcal{E}(\mathbf{x}_j|\mathbf{y}^+) + \epsilon \mathbf{r}_j, && j=1,...,J
-  \end{aligned}
-\end{equation}
-
-where $\mathbf{r}_j \sim \mathcal{N}(\mathbf{0},\mathbf{I})$ is the stochastic term and the step-size $\epsilon$ is typically polynomially decayed~\citep{welling2011bayesian}.
-The term $\mathcal{E}(\mathbf{x}_j|\mathbf{y}^+)$ denotes the model energy conditioned on the target class label $\mathbf{y}^+$ which we specify as the negative logit corresponding to the target class label $\mathbf{y}^*$.
-To allow for faster sampling, we follow the common practice of choosing the step-size $\epsilon$ and the standard deviation of $\mathbf{r}_j$ separately.
-While $\mathbf{x}_J$ is only guaranteed to distribute as $p_{\theta}(\mathbf{x}|\mathbf{y}^*)$ if $\epsilon \rightarrow 0$ and $J \rightarrow \infty$, the bias introduced for a small finite $\epsilon$ is negligible in practice \citep{murphy2023probabilistic,grathwohl2020your}.
-Appendix~\ref{app:jem} provides additional implementation details for any tasks related to energy-based modelling.
-
-Generating multiple samples using SGLD thus yields an empirical distribution $\hat{\mathbf{X}}_{\theta,\mathbf{y}^+}$ that approximates what the model has learned about the input data.
-While in the context of EBM, this is usually done during training, we propose to repurpose this approach during inference in order to evaluate and generate faithful model explanations.
-
-\subsection{Evaluating Plausibility and Faithfulness}
-
-The parallels between our definitions of plausibility and faithfulness imply that we can also use similar evaluation metrics in both cases.
-Since existing work has focused heavily on plausibility, it offers a useful starting point.
-In particular,~\citet{guidotti2022counterfactual} have proposed an implausibility metric that measures the distance of the counterfactual from its nearest neighbour in the target class.
-As this distance is reduced, counterfactuals get more plausible under the assumption that the nearest neighbour itself is plausible in the sense of Definition~\ref{def:plausible}.
-In this work, we use the following adapted implausibility metric,
-
-\begin{equation}\label{eq:impl}
-  \begin{aligned}
-    \text{impl}(\mathbf{x}^{\prime},\mathbf{X}_{\mathbf{y}^+}) = \frac{1}{\lvert\mathbf{X}_{\mathbf{y}^+}\rvert} \sum_{\mathbf{x} \in \mathbf{X}_{\mathbf{y}^+}} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
-  \end{aligned}
-\end{equation}
-
-where $\mathbf{x}^{\prime}$ denotes the counterfactual and $\mathbf{X}_{\mathbf{y}^+}$ is a subsample of the training data in the target class $\mathbf{y}^+$.
-By averaging over multiple samples in this manner, we avoid the risk that the nearest neighbour of $\mathbf{x}^{\prime}$ itself is not plausible according to Definition~\ref{def:plausible} (e.g an outlier).
-
-Equation~\ref{eq:impl} gives rise to a similar evaluation metric for unfaithfulness.
-We merely swap out the subsample of individuals in the target class for a subset $\hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^+}$ of the generated conditional samples:
-
-\begin{equation}\label{eq:faith}
-  \begin{aligned}
-    \text{unfaith}(\mathbf{x}^{\prime},\hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^+}) = \frac{1}{\lvert \hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^+}\rvert} \sum_{\mathbf{x} \in \hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^+}} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
-  \end{aligned}
-\end{equation}
-
-Specifically, we form this subset based on the $n_E$ generated samples with the lowest energy.
-
-\section{Energy-Constrained Conformal Counterfactuals}\label{meth}
-
-In this section, we describe \textit{ECCCo}, our proposed framework for generating Energy-Constrained Conformal Counterfactuals (ECCCos).
-It is based on the premise that counterfactuals should first and foremost be faithful.
-Plausibility, as a secondary concern, is then still attainable, but only to the degree that the black-box model itself has learned plausible explanations for the underlying data.
-
-We begin by stating our proposed objective function, which involves tailored loss and penalty functions that we will explain in the following.
-In particular, we extend Equation~\ref{eq:general} as follows:
-
-\begin{equation} \label{eq:eccco}
-  \begin{aligned}
-  \mathbf{Z}^\prime= \arg \min_{\mathbf{Z}^\prime \in \mathcal{Z}^M}  &\{  {\text{yloss}(M_{\theta}(f(\mathbf{Z}^\prime)),\mathbf{y}^+)}+ \lambda_{1} {\text{dist}(f(\mathbf{Z}^\prime),\mathbf{x}) } \\
-  &+ \lambda_2 \text{unfaith}(f(\mathbf{Z}^\prime),\hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^+}) + \lambda_3 \Omega(C_{\theta}(f(\mathbf{Z}^\prime);\alpha)) \} 
-  \end{aligned} 
-\end{equation}
-
-The first penalty term involving $\lambda_1$ induces proximity like in~\citet{wachter2017counterfactual}.
-Our default choice for $\text{dist}(\cdot)$ is the L1 Norm due to its sparsity-inducing properties.
-The second penalty term involving $\lambda_2$ induces faithfulness by constraining the energy of the generated counterfactual where $\text{unfaith}(\cdot)$ corresponds to the metric defined in Equation~\ref{eq:faith}.
-The third and final penalty term involving $\lambda_3$ introduces a new concept: it ensures that the generated counterfactual is associated with low predictive uncertainty.
-As mentioned above,~\citet{schut2021generating} have shown that plausible counterfactuals can be generated implicitly through predictive uncertainty minimization.
-Unfortunately, this relies on the assumption that the model itself can provide predictive uncertainty estimates, which may be too restrictive in practice.
-
-To relax this assumption, we leverage recent advances in Conformal Prediction (CP), an approach to predictive uncertainty quantification that has recently gained popularity~\citep{angelopoulos2021gentle,manokhin2022awesome}.
-Crucially for our intended application, CP is model-agnostic and can be applied during inference without placing any restrictions on model training.
-Intuitively, CP works under the premise of turning heuristic notions of uncertainty into rigorous uncertainty estimates by repeatedly sifting through the training data or a dedicated calibration dataset.
-Conformal classifiers produce prediction sets for individual inputs that include all output labels that can be reasonably attributed to the input.
-These sets tend to be larger for inputs that do not conform with the training data and are characterized by high predictive uncertainty.
-
-In order to generate counterfactuals that are associated with low predictive uncertainty, we use a smooth set size penalty introduced by~\citet{stutz2022learning} in the context of conformal training:
-
-\begin{equation}\label{eq:setsize}
-  \begin{aligned}
-    \Omega(C_{\theta}(\mathbf{x};\alpha))&=\max \left(0, \sum_{\mathbf{y}\in\mathcal{Y}}C_{\theta,\mathbf{y}}(\mathbf{x}_i;\alpha) - \kappa \right)
-  \end{aligned}
-\end{equation}
-
-Here, $\kappa \in \{0,1\}$ is a hyper-parameter and $C_{\theta,\mathbf{y}}(\mathbf{x}_i;\alpha)$ can be interpreted as the probability of label $\mathbf{y}$ being included in the prediction set.
-In order to compute this penalty for any black-box model we merely need to perform a single calibration pass through a holdout set $\mathcal{D}_{\text{cal}}$.
-Arguably, data is typically abundant and in most applications, practitioners tend to hold out a test data set anyway.
-Consequently, CP removes the restriction on the family of predictive models, at the small cost of reserving a subset of the available data for calibration.
-This particular case of conformal prediction is referred to as Split Conformal Prediction (SCP) as it involves splitting the training data into a proper training dataset and a calibration dataset.
-In addition to the smooth set size penalty, we have also experimented with the use of a tailored function for $\text{yloss}(\cdot)$ that enforces that only the target label $\mathbf{y}^+$ is included in the prediction set ~\citet{stutz2022learning}.
-Further details are provided in Appendix~\ref{app:cp}.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=1.0\textwidth]{../artifacts/results/images/poc_gradient_fields.png}
-  \caption{Gradient fields and counterfactual paths for different generators.
-The objective is to generate a counterfual in the `blue' class for a sample from the `orange' class.
-Bright yellow stars indicate conditional samples generated through SGLD.
-The underlying classifier is a Joint Energy Model.}\label{fig:poc}
-\end{figure}
-
-\renewcommand{\algorithmicrequire}{\textbf{Input:}}
-\renewcommand{\algorithmicensure}{\textbf{Output:}}
-
-\begin{algorithm}
-  \caption{The \textit{ECCCo} generator}\label{alg:eccco}
-  \begin{algorithmic}[1]
-    \Require $\mathbf{x}, \mathbf{y}^+, M_{\theta}, f, \Lambda=[\lambda_1,\lambda_2,\lambda_3], \alpha, \mathcal{D}, T, \eta, n_{\mathcal{B}}, n_E$ where $M_{\theta}(\mathbf{x})\neq\mathbf{y}^+$
-    \Ensure $\mathbf{x}^\prime$
-    \State Initialize $\mathbf{z}^\prime \gets f^{-1}(\mathbf{x})$ \Comment{Map to counterfactual state space.}
-    \State Generate $\left\{\hat{\mathbf{x}}_{\theta,\mathbf{y}^+}\right\}_{n_{\mathcal{B}}} \gets p_{\theta}(\mathbf{x}_{\mathbf{y}^+})$ \Comment{Generate $n_{\mathcal{B}}$ samples using SGLD (Equation~\ref{eq:sgld}).}
-    \State Store $\hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^+} \gets \left\{\hat{\mathbf{x}}_{\theta,\mathbf{y}^+}\right\}_{n_{\mathcal{B}}}$ \Comment{Choose $n_E$ lowest-energy samples.}
-    \State Run \textit{SCP} for $M_{\theta}$ using $\mathcal{D}$ \Comment{Calibrate model through Split Conformal Prediction.}
-    \State Initialize $t \gets 0$
-    \While{\textit{not converged} or $t < T$} \Comment{For convergence conditions see Appendix~\ref{app:eccco}.}
-    \State $\mathbf{z}^\prime \gets \mathbf{z}^\prime - \eta \nabla_{\mathbf{z}^\prime} \mathcal{L}(\mathbf{z}^\prime,\mathbf{y}^+,\hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^+}; \Lambda, \alpha)$ \Comment{Take gradient step of size $\eta$.}
-    \State $t \gets t+1$
-    \EndWhile
-    \State $\mathbf{x}^\prime \gets f(\mathbf{z}^\prime)$ \Comment{Map back to feature space.}
-  \end{algorithmic}
-\end{algorithm}
-
-To provide some further intuition about our objective defined in Equation~\ref{eq:eccco}, Figure~\ref{fig:poc} illustrates how the different components affect the counterfactual search for a synthetic dataset.
-The underlying classifier is a Joint Energy Model (\textit{JEM}) that was trained to predict the output class (`blue' or `orange') and generate class-conditional samples~\citep{grathwohl2020your}.
-We have used four different generator flavours to produce a counterfactual in the `blue' class for a sample from the `orange' class: \textit{Wachter}, which only uses the first penalty ($\lambda_2=\lambda_3=0$); \textit{ECCCo (no EBM)}, which does not constrain energy ($\lambda_2=0$); \textit{ECCCo (no CP)}, which involves no set size penalty ($\lambda_3=0$); and, finally, \textit{ECCCo}, which involves all penalties defined in Equation~\ref{eq:eccco}.
-Arrows indicate (negative) gradients with respect to the objective function at different points in the feature space.
-
-While \textit{Wachter} generates a valid counterfactual, it ends up close to the original starting point consistent with its objective.
-\textit{ECCCo (no EBM)} pushes the counterfactual further into the target domain to minimize predictive uncertainty, but the outcome is still not plausible.
-The counterfactual produced by \textit{ECCCo (no CP)} is attracted by the generated samples shown in bright yellow.
-Since the \textit{JEM} has learned the conditional input distribution reasonably well in this case, the counterfactuals are both faithful and plausible.
-Finally, the outcome for \textit{ECCCo} looks similar, but the additional smooth set size penalty leads to somewhat faster convergence.
-
-Algorithm~\ref{alg:eccco} describes how exactly \textit{ECCCo} works.
-For the sake of simplicity and without loss of generality, we limit our attention to generating a single counterfactual $\mathbf{x}^\prime=f(\mathbf{z}^\prime)$.
-The counterfactual state $\mathbf{z}^\prime$ is initialized by passing the factual $\mathbf{x}$ through a simple feature transformer $f^{-1}$.
-Next, we generate $n_{\mathcal{B}}$ conditional samples $\hat{\mathbf{x}}_{\theta,\mathbf{y}^+}$ using SGLD (Equation~\ref{eq:sgld}) and store the $n_E$ instances with the lowest energy.
-We then calibrate the model $M_{\theta}$ through Split Conformal Prediction.
-Finally, we search counterfactuals through gradient descent where $\mathcal{L}(\mathbf{z}^\prime,\mathbf{y}^+,\hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^+}; \Lambda, \alpha)$ denotes our loss function defined in Equation~\ref{eq:eccco}.
-The search terminates once the convergence criterium is met or the maximum number of iterations $T$ has been exhausted.
-Note that the choice of convergence criterium has important implications on the final counterfactual which we explain in Appendix~\ref{app:eccco}.
-
-\section{Empirical Analysis}\label{emp}
-
-Our goal in this section is to shed light on the following research questions:
-
-\begin{question}[Faithfulness]\label{rq:faithfulness}
-  Are ECCCos more faithful than counterfactuals produced by our benchmark generators?
-\end{question}
-
-\begin{question}[Balancing Objectives]\label{rq:plausibility}
-  Compared to our benchmark generators, how do ECCCos balance the two key objectives of faithfulness and plausibility?
-\end{question}
-
-The second question is motivated by the intuition that faithfulness and plausibility should coincide for models that have learned plausible explanations of the data.
-Next, we first briefly describe our experimental setup before presenting our main results.
-
-\subsection{Experimental Setup}
-
-To assess and benchmark the performance of our proposed generator against the state of the art, we generate multiple counterfactuals for different models and datasets.
-In particular, we compare \textit{ECCCo} and its variants to the following counterfactual generators that were introduced above: firstly; \textit{Schut}, which works under the premise of minimizing predictive uncertainty; secondly, \textit{REVISE}, which is state-of-the-art with respect to plausibility; and, finally, \textit{Wachter}, which serves as our baseline.
-
-We use both synthetic and real-world datasets from different domains, all of which are publicly available and commonly used to train and benchmark classification algorithms.
-We synthetically generate a dataset containing two \textit{Linearly Separable} Gaussian clusters ($n=1000$), as well as the well-known \textit{Circles} ($n=1000$) and \textit{Moons} ($n=2500$) data.
-Since these data are generated by distributions of varying degrees of complexity, they allow us to assess how the generators and our proposed evaluation metrics handle this.
-
-As for real-world data, we follow~\citet{schut2021generating} and use the \textit{MNIST}~\citep{lecun1998mnist} dataset containing images of handwritten digits such as the example shown above in Figure~\ref{fig:motiv}.
-From the social sciences domain, we include Give Me Some Credit (\textit{GMSC})~\citep{kaggle2011give}: a tabular dataset that has been studied extensively in the literature on Algorithmic Recourse~\citep{pawelczyk2021carla}.
-It consists of 11 numeric features that can be used to predict the binary outcome variable indicating whether retail borrowers experience financial distress.
-
-For the predictive modelling tasks, we use simple neural networks (\textit{MLP}) and Joint Energy Models (\textit{JEM}).
-For the more complex real-world datasets we also use ensembling in each case.
-Both joint-energy modelling and ensembling have been associated with improved generative properties and adversarial robustness~\citep{grathwohl2020your,lakshminarayanan2016simple}, so we expect this to be positively correlated with the plausibility of ECCCos.
-To account for stochasticity, we generate multiple counterfactuals for each target class, generator, model and dataset.
-Specifically, we randomly sample $n^{-}$ times from the subset of individuals for which the given model predicts the non-target class $\mathbf{y}^{-}$ given the current target.
-We set $n^{-}=25$ for all of our synthetic datasets, $n^{-}=10$ for \textit{GMSC} and $n^{-}=5$ for \textit{MNIST}.
-Full details concerning our parameter choices, training procedures and model performance can be found in Appendix~\ref{app:setup}.
-
-\subsection{Results for Synthetic Data}
-
-Table~\ref{tab:results-synthetic} shows the key results for the synthetic datasets separated by model (first column) and generator (second column).
-The numerical columns show sample averages and standard deviations of our key evaluation metrics computed across all counterfactuals.
-We have highlighted the best outcome for each model and metric in bold.
-To provide some sense of effect sizes, we have added asterisks to indicate that a given value is at least one ($*$) or two ($**$) standard deviations lower than the baseline (\textit{Wachter}).
-
-Starting with the high-level results for our \textit{Linearly Separable} data, we find that \textit{ECCCo} produces the most faithful counterfactuals for both black-box models.
-This is consistent with our design since \textit{ECCCo} directly enforces faithfulness through regularization.
-Crucially though, \textit{ECCCo} also produces the most plausible counterfactuals for both models.
-This dataset is so simple that even the \textit{MLP} has learned plausible explanations of the input data.
-Zooming in on the granular details for the \textit{Linearly Separable} data, the results for \textit{ECCCo (no CP)} and \textit{ECCCo (no EBM)} indicate that the positive results are dominated by the effect of quantifying and leveraging the model's generative property (EBM).
-Conformal Prediction alone only leads to marginally improved faithfulness and plausibility.
-
-The findings for the \textit{Moons} dataset are broadly in line with the findings so far: for the \textit{JEM}, \textit{ECCCo} yields substantially more faithful and plausible counterfactuals than all other generators.
-For the \textit{MLP}, faithfulness is maintained but counterfactuals are not plausible.
-This high-level pattern is broadly consistent with other more complex datasets and supportive of our narrative, so it is worth highlighting: ECCCos consistently achieve high faithfulness, which---subject to the quality of the model itself---coincides with high plausibility.
-By comparison, \textit{REVISE} yields the most plausible counterfactuals for the \textit{MLP}, but it does so at the cost of faithfulness.
-We also observe that the best results for \textit{ECCCo} are achieved when using both penalties.
-Once again though, the generative component (EBM) has a stronger impact on the positive results for the \textit{JEM}.
-
-For the \textit{Circles} data, it appears that \textit{REVISE} performs well, but we note that it generates valid counterfactuals only half of the time (see Appendix~\ref{app:results} for a complete overview including additional common evaluation metrics).
-The underlying VAE with default parameters has not adequately learned the data-generating process.
-Of course, it is possible to improve generative performance through hyperparameter tuning but this example serves to illustrate that \textit{REVISE} depends on the quality of its surrogate.
-Independent of the outcome for \textit{REVISE}, however, the results do not seem to indicate that \textit{ECCCo} substantially improves faithfulness and plausibility for the \textit{Circles} data.
-We think this points to a limitation of our evaluation metrics rather than \textit{ECCCo} itself: computing average distances fails to account for the `wraparound' effect associated with circular data~\citep{gill2010circular}.
-
-\import{contents/}{table-synthetic.tex}
-
-\subsection{Results for Real-World Data}
-
-The results for our real-world datasets are shown in Table~\ref{tab:results-real-world}.
-Once again the findings indicate that the plausibility of ECCCos is positively correlated with the capacity of the black-box model to distinguish plausible from implausible inputs.
-The case is very clear for \textit{MNIST}: ECCCos are consistently more faithful than the counterfactuals produced by our benchmark generators and their plausibility gradually improves through ensembling and joint-energy modelling.
-Interestingly, faithfulness also gradually improves for \textit{REVISE}.
-This indicates that as our models improve, their generative capacity approaches that of the surrogate VAE used by \textit{REVISE}.
-The VAE still outperforms our classifiers in this regard, as evident from the fact that \textit{ECCCo} never quite reaches the same level of plausibility as \textit{REVISE}.
-With reference to Appendix~\ref{app:results} we note that the results for \textit{Schut} need to be discounted as it rarely produces valid counterfactuals for \textit{MNIST}.
-Relatedly, we find that \textit{ECCCo} is the only generator that consistently achieves full validity.
-Finally, it is worth noting that \textit{ECCCo} produces counterfactual images with the lowest average predictive uncertainty for all models.
-
-For the tabular credit dataset (\textit{GMSC}) it is inherently challenging to use deep neural networks in order to achieve good discriminative performance~\citep{borisov2021deep,grinsztajn2022why} and generative performance~\citep{liu2023goggle}, respectively.
-In order to achieve high plausibility, \textit{ECCCo} effectively requires classifiers to achieve good performance for both tasks.
-Since this is a challenging task even for Joint Energy Models, it is not surprising to find that even though \textit{ECCCo} once again achieves state-of-the-art faithfulness, it is outperformed by \textit{REVISE} and \textit{Schut} with respect to plausibility.
-
-\subsection{Key Takeways}
-
-To conclude this section, we summarize our findings with reference to the opening questions.
-The results clearly demonstrate that \textit{ECCCo} consistently achieves state-of-the-art faithfulness, as it was designed to do (Research Question~\ref{rq:faithfulness}).
-A related important finding is that \textit{ECCCo} yields highly plausible explanations provided that they faithfully describe model behaviour (Research Question~\ref{rq:plausibility}).
-\textit{ECCCo} achieves this result primarily by leveraging the model's generative property.
-
-\import{contents/}{table-real-world.tex}
-
-\section{Limitations}
-
-Even though we have taken considerable measures to study our proposed methodology carefully, limitations can still be identified.
-In particular, we have found that the performance of \textit{ECCCo} is sensitive to hyperparameter choices.
-In order to achieve faithfulness, we generally had to penalise the distance from generated samples slightly more than the distance from factual values.
-
-Conversely, we have not found that strongly penalising prediction set sizes had any discernable effect.
-Our results indicate that CP alone is often not sufficient to achieve faithfulness and plausibility, although we acknowledge that this needs to be investigated more thoroughly through future work.
-
-While our approach is readily applicable to models with gradient access like deep neural networks, more work is needed to generalise it to other machine learning models such as decision trees.
-Relatedly, common challenges associated with Energy-Based Modelling including sensitivity to scale, training instabilities and sensitivity to hyperparameters also apply to \textit{ECCCo}.
-
-\section{Conclusion}
-
-This work leverages recent advances in Energy-Based Modelling and Conformal Prediction in the context of Explainable Artificial Intelligence.
-We have proposed a new way to generate counterfactuals that are maximally faithful to the black-box model they aim to explain.
-Our proposed generator, \textit{ECCCo}, produces plausible counterfactuals if and only if the black-box model itself has learned realistic explanations for the data, which we have demonstrated through rigorous empirical analysis.
-This should enable researchers and practitioners to use counterfactuals in order to discern trustworthy models from unreliable ones.
-While the scope of this work limits its generalizability, we believe that \textit{ECCCo} offers a solid baseline for future work on faithful Counterfactual Explanations.
-
-\begin{ack}
-
-Some of the members of TU Delft were partially funded by ICAI AI for Fintech Research, an ING — TU Delft
-collaboration.
-
-\end{ack}
-
-\bibliography{bib}
-
-\pagebreak
-
 \appendix
@@
   \centering
-  \includegraphics[width=0.75\textwidth]{../artifacts/results/images/mnist_generated_JEM Ensemble.png}
+  \includegraphics[width=0.75\linewidth]{../artifacts/results/images/mnist_generated_JEM Ensemble.png}
   \caption{Conditionally generated \textit{MNIST} images for our JEM Ensemble.}\label{fig:mnist-gen}
@@
   \centering
-  \includegraphics[width=0.5\textwidth]{../artifacts/results/images/moons_generated_JEM.png}
+  \includegraphics[width=0.5\linewidth]{../artifacts/results/images/moons_generated_JEM.png}
   \caption{Conditionally generated samples (stars) for our \textit{Moons} data using a JEM.}\label{fig:moons-gen}
@@
   \centering
-  \includegraphics[width=1.0\textwidth]{../artifacts/results/images/poc_set_size.png}
+  \includegraphics[width=1.0\linewidth]{../artifacts/results/images/poc_set_size.png}
   \caption{Prediction set size (left), smooth set size loss (centre) and configurable classification loss (right) for a JEM trained on our \textit{Linearly Separable} data.}\label{fig:cp-diff}
@@
   \centering
-  \includegraphics[width=0.9\textwidth]{../artifacts/results/images/mnist_eccco_all_digits.png}
+  \includegraphics[width=0.9\linewidth]{../artifacts/results/images/mnist_eccco_all_digits.png}
   \caption{Counterfactuals for \textit{MNIST} data and our \textit{JEM Ensemble}.
@@
 \import{contents/}{table_all_valid.tex}
-
-\end{document}

### 2023-09-12 paper/body.tex
latex cleaned up

--- 2023-09-12 AAAI/paper.tex
@@
-%File: anonymous-submission-latex-2024.tex
-\documentclass[letterpaper]{article} % DO NOT CHANGE THIS
-\usepackage[submission]{aaai24}  % DO NOT CHANGE THIS
-\usepackage{times}  % DO NOT CHANGE THIS
-\usepackage{helvet}  % DO NOT CHANGE THIS
-\usepackage{courier}  % DO NOT CHANGE THIS
-\usepackage[hyphens]{url}  % DO NOT CHANGE THIS
-\usepackage{graphicx} % DO NOT CHANGE THIS
-\urlstyle{rm} % DO NOT CHANGE THIS
-\def\UrlFont{\rm}  % DO NOT CHANGE THIS
-\usepackage{natbib}  % DO NOT CHANGE THIS AND DO NOT ADD ANY OPTIONS TO IT
-\usepackage{caption} % DO NOT CHANGE THIS AND DO NOT ADD ANY OPTIONS TO IT
-\frenchspacing  % DO NOT CHANGE THIS
-\setlength{\pdfpagewidth}{8.5in} % DO NOT CHANGE THIS
-\setlength{\pdfpageheight}{11in} % DO NOT CHANGE THIS
-%
-% These are recommended to typeset algorithms but not required. See the subsubsection on algorithms. Remove them if you don't have algorithms in your paper.
-\usepackage{algorithm}
-% \usepackage{algorithmic}
-
-%
-% These are are recommended to typeset listings but not required. See the subsubsection on listing. Remove this block if you don't have listings in your paper.
-% \usepackage{newfloat}
-% \usepackage{listings}
-% \DeclareCaptionStyle{ruled}{labelfont=normalfont,labelsep=colon,strut=off} % DO NOT CHANGE THIS
-% \lstset{%
-% 	basicstyle={\footnotesize\ttfamily},% footnotesize acceptable for monospace
-% 	numbers=left,numberstyle=\footnotesize,xleftmargin=2em,% show line numbers, remove this entire line if you don't want the numbers.
-% 	aboveskip=0pt,belowskip=0pt,%
-% 	showstringspaces=false,tabsize=2,breaklines=true}
-% \floatstyle{ruled}
-% \newfloat{listing}{tb}{lst}{}
-% \floatname{listing}{Listing}
-%
-% Keep the \pdfinfo as shown here. There's no need
-% for you to add the /Title and /Author tags.
-\pdfinfo{
-/TemplateVersion (2024.1)
-}
-
-\usepackage{amsfonts}       % blackboard math symbols
-\usepackage{amsmath}
-\usepackage{amsthm}
-\usepackage{caption}
-\usepackage{graphicx}
-\usepackage{algpseudocode}
-\usepackage{import}
-\usepackage{booktabs}
-\usepackage{longtable}
-\usepackage{array}
-\usepackage{multirow}
-
-% Numbered Environments:
-\newtheorem{definition}{Definition}[section]
-\newtheorem{question}{Research Question}[section]
-
-% Bibliography
-% \bibliographystyle{unsrtnat}
-% \setcitestyle{numbers,square,comma}
-
-% Algorithm
-\renewcommand{\algorithmicrequire}{\textbf{Input:}}
-\renewcommand{\algorithmicensure}{\textbf{Output:}}
-
-% DISALLOWED PACKAGES
-% \usepackage{authblk} -- This package is specifically forbidden
-% \usepackage{balance} -- This package is specifically forbidden
-% \usepackage{color (if used in text)
-% \usepackage{CJK} -- This package is specifically forbidden
-% \usepackage{float} -- This package is specifically forbidden
-% \usepackage{flushend} -- This package is specifically forbidden
-% \usepackage{fontenc} -- This package is specifically forbidden
-% \usepackage{fullpage} -- This package is specifically forbidden
-% \usepackage{geometry} -- This package is specifically forbidden
-% \usepackage{grffile} -- This package is specifically forbidden
-% \usepackage{hyperref} -- This package is specifically forbidden
-% \usepackage{navigator} -- This package is specifically forbidden
-% (or any other package that embeds links such as navigator or hyperref)
-% \indentfirst} -- This package is specifically forbidden
-% \layout} -- This package is specifically forbidden
-% \multicol} -- This package is specifically forbidden
-% \nameref} -- This package is specifically forbidden
-% \usepackage{savetrees} -- This package is specifically forbidden
-% \usepackage{setspace} -- This package is specifically forbidden
-% \usepackage{stfloats} -- This package is specifically forbidden
-% \usepackage{tabu} -- This package is specifically forbidden
-% \usepackage{titlesec} -- This package is specifically forbidden
-% \usepackage{tocbibind} -- This package is specifically forbidden
-% \usepackage{ulem} -- This package is specifically forbidden
-% \usepackage{wrapfig} -- This package is specifically forbidden
-% DISALLOWED COMMANDS
-% \nocopyright -- Your paper will not be published if you use this command
-% \addtolength -- This command may not be used
-% \balance -- This command may not be used
-% \baselinestretch -- Your paper will not be published if you use this command
-% \clearpage -- No page breaks of any kind may be used for the final version of your paper
-% \columnsep -- This command may not be used
-% \newpage -- No page breaks of any kind may be used for the final version of your paper
-% \pagebreak -- No page breaks of any kind may be used for the final version of your paperr
-% \pagestyle -- This command may not be used
-% \tiny -- This is not an acceptable font size.
-% \vspace{- -- No negative value may be used in proximity of a caption, figure, table, section, subsection, subsubsection, or reference
-% \vskip{- -- No negative value may be used to alter spacing above or below a caption, figure, table, section, subsection, subsubsection, or reference
-
-\setcounter{secnumdepth}{0} %May be changed to 1 or 2 if section numbers are desired.
-
-% The file aaai24.sty is the style file for AAAI Press
-% proceedings, working notes, and technical reports.
-%
-
-% Title
-
-% Your title must be in mixed case, not sentence case.
-% That means all verbs (including short verbs like be, is, using,and go),
-% nouns, adverbs, adjectives should be capitalized, including both words in hyphenated terms, while
-% articles, conjunctions, and prepositions are lower case unless they
-% directly follow a colon or long dash
-\title{AAAI Press Anonymous Submission\\Instructions for Authors Using \LaTeX{}}
-\author{
-    %Authors
-    % All authors must be in the same font size and format.
-    Written by AAAI Press Staff\textsuperscript{\rm 1}\thanks{With help from the AAAI Publications Committee.}\\
-    AAAI Style Contributions by Pater Patel Schneider,
-    Sunil Issar,\\
-    J. Scott Penberthy,
-    George Ferguson,
-    Hans Guesgen,
-    Francisco Cruz\equalcontrib,
-    Marc Pujol-Gonzalez\equalcontrib
-}
-\affiliations{
-    %Afiliations
-    \textsuperscript{\rm 1}Association for the Advancement of Artificial Intelligence\\
-    % If you have multiple authors and multiple affiliations
-    % use superscripts in text and roman font to identify them.
-    % For example,
-
-    % Sunil Issar\textsuperscript{\rm 2},
-    % J. Scott Penberthy\textsuperscript{\rm 3},
-    % George Ferguson\textsuperscript{\rm 4},
-    % Hans Guesgen\textsuperscript{\rm 5}
-    % Note that the comma should be placed after the superscript
-
-    1900 Embarcadero Road, Suite 101\\
-    Palo Alto, California 94303-3310 USA\\
-    % email address must be in roman text type, not monospace or sans serif
-    proceedings-questions@aaai.org
-%
-% See more examples next
-}
-
-%Example, Single Author, ->> remove \iffalse,\fi and place them surrounding AAAI title to use it
-\iffalse
-\title{My Publication Title --- Single Author}
-\author {
-    Author Name
-}
-\affiliations{
-    Affiliation\\
-    Affiliation Line 2\\
-    name@example.com
-}
-\fi
-
-\iffalse
-%Example, Multiple Authors, ->> remove \iffalse,\fi and place them surrounding AAAI title to use it
-\title{My Publication Title --- Multiple Authors}
-\author {
-    % Authors
-    First Author Name\textsuperscript{\rm 1},
-    Second Author Name\textsuperscript{\rm 2},
-    Third Author Name\textsuperscript{\rm 1}
-}
-\affiliations {
-    % Affiliations
-    \textsuperscript{\rm 1}Affiliation 1\\
-    \textsuperscript{\rm 2}Affiliation 2\\
-    firstAuthor@affiliation1.com, secondAuthor@affilation2.com, thirdAuthor@affiliation1.com
-}
-\fi
-
-% REMOVE THIS: bibentry
-% This is only needed to show inline citations in the guidelines document. You should not need it and can safely delete it.
-\usepackage{bibentry}
-% END REMOVE bibentry
-
-\begin{document}
-
 \maketitle
@@
   \centering
-    \includegraphics[width=0.4\textwidth]{../artifacts/results/images/mnist_motivation.png}
+  \includegraphics[width=0.8\linewidth]{../artifacts/results/images/mnist_motivation.png}
   \caption{Counterfactuals for turning a 9 (nine) into a 7 (seven): original image (left); then from left to right the counterfactuals generated using \textit{Wachter}, \textit{Schut} and \textit{REVISE}.}\label{fig:motiv}
@@
   \centering
-    \includegraphics[width=0.5\textwidth]{../artifacts/results/images/poc_gradient_fields.png}
+  \includegraphics[width=1.0\linewidth]{../artifacts/results/images/poc_gradient_fields.png}
   \caption{Gradient fields and counterfactual paths for different generators.
@@
 \end{figure}
-
-  \renewcommand{\algorithmicrequire}{\textbf{Input:}}
-  \renewcommand{\algorithmicensure}{\textbf{Output:}}
 
@@
 While the scope of this work limits its generalizability, we believe that \textit{ECCCo} offers a solid baseline for future work on faithful Counterfactual Explanations.
-
-  \section{Acknowledgments}
-
-  Some of the members of TU Delft were partially funded by ICAI AI for Fintech Research, an ING — TU Delft
-  collaboration.
-
-  \bibliography{bib}
-
-\end{document}

### 2023-09-13 paper/body.tex
mata me

@@
 Counterfactual Explanations (CE) provide a powerful, flexible and intuitive way to not only explain black-box models but also help affected individuals through the means of Algorithmic Recourse.
-Instead of opening the Black Box, CE works under the premise of strategically perturbing model inputs to understand model behaviour~\citep{wachter2017counterfactual}.
+Instead of opening the black box, Counterfactual Explanations work under the premise of strategically perturbing model inputs to understand model behaviour~\citep{wachter2017counterfactual}.
 Intuitively speaking, we generate explanations in this context by asking what-if questions of the following nature: `Our credit risk model currently predicts that this individual is not credit-worthy.
@@
 Counterfactuals are then searched by minimizing a loss function that compares the predicted model output to the target outcome: $\text{yloss}(M_{\theta}(\mathbf{x}),\mathbf{y}^+)$.
-Since CE work directly with the black-box model, valid counterfactuals always have full local fidelity by construction where fidelity is defined as the degree to which explanations approximate the predictions of a black-box model~\citep{mothilal2020explaining,molnar2020interpretable}.
+Since Counterfactual Explanations work directly with the black-box model, valid counterfactuals always have full local fidelity by construction where fidelity is defined as the degree to which explanations approximate the predictions of a black-box model~\citep{mothilal2020explaining,molnar2020interpretable}.
 

### 2023-09-13 notebooks/tables.Rmd
mostly doen

@@
   kable_styling(latex_options = c("scale_down")) %>%
-  kable_paper(full_width = F) %>%
+  kable_paper(full_width = T) %>%
   add_header_above(header) %>%
@@
   kable_styling(latex_options = c("scale_down")) %>%
-  kable_paper(full_width = F) %>%
+  kable_paper(full_width = T) %>%
   collapse_rows(columns = 1:3, latex_hline = "custom", valign = "top", custom_latex_hline = 1:2) %>%
@@
   kable_styling(font_size = 8) %>%
-  kable_paper(full_width = F) %>%
+  kable_paper(full_width = T) %>%
   save_kable("paper/contents/table_ebm_params.tex")
@@
   kable_styling(font_size = 8) %>%
-  kable_paper(full_width = F) %>%
+  kable_paper(full_width = T) %>%
   save_kable("paper/contents/table_gen_params.tex")
@@
   kable_styling(font_size = 8) %>%
-  kable_paper(full_width = F) %>%
+  kable_paper(full_width = T) %>%
   add_header_above(c(" "=2, "Performance Metrics" = 3)) %>%

### 2023-09-13 paper/body.tex
mostly doen

--- 2023-09-12 paper/body.tex
@@
 \begin{abstract}
-  Counterfactual Explanations offer an intuitive and straightforward way to explain black-box models and offer Algorithmic Recourse to individuals.
+  Counterfactual explanations offer an intuitive and straightforward way to explain black-box models and offer algorithmic recourse to individuals.
 To address the need for plausible explanations, existing work has primarily relied on surrogate models to learn how the input data is distributed.
@@
 In particular, we show that for models with gradient access, it is possible to achieve state-of-the-art performance without the need for surrogate models.
-To do so, our framework relies solely on properties defining the black-box model itself by leveraging recent advances in Energy-Based Modelling and Conformal Prediction.
-To our knowledge, this is the first venture in this direction for generating faithful Counterfactual Explanations.
+To do so, our framework relies solely on properties defining the black-box model itself by leveraging recent advances in energy-based modelling and conformal prediction.
+To our knowledge, this is the first venture in this direction for generating faithful counterfactual explanations.
 Thus, we anticipate that ECCCos can serve as a baseline for future research.
@@
 
-Counterfactual Explanations (CE) provide a powerful, flexible and intuitive way to not only explain black-box models but also help affected individuals through the means of Algorithmic Recourse.
-Instead of opening the Black Box, CE works under the premise of strategically perturbing model inputs to understand model behaviour~\citep{wachter2017counterfactual}.
+Counterfactual explanations provide a powerful, flexible and intuitive way to not only explain black-box models but also help affected individuals through the means of algorithmic recourse.
+Instead of opening the black box, counterfactual explanations work under the premise of strategically perturbing model inputs to understand model behaviour~\citep{wachter2017counterfactual}.
 Intuitively speaking, we generate explanations in this context by asking what-if questions of the following nature: `Our credit risk model currently predicts that this individual is not credit-worthy.
@@
 Counterfactuals are then searched by minimizing a loss function that compares the predicted model output to the target outcome: $\text{yloss}(M_{\theta}(\mathbf{x}),\mathbf{y}^+)$.
-Since CE work directly with the black-box model, valid counterfactuals always have full local fidelity by construction where fidelity is defined as the degree to which explanations approximate the predictions of a black-box model~\citep{mothilal2020explaining,molnar2020interpretable}.
+Since counterfactual explanations work directly with the black-box model, valid counterfactuals always have full local fidelity by construction where fidelity is defined as the degree to which explanations approximate the predictions of a black-box model~\citep{mothilal2020explaining,molnar2020interpretable}.
 
-In situations where full fidelity is a requirement, CE offer a more appropriate solution to Explainable Artificial Intelligence (XAI) than other popular approaches like LIME~\citep{ribeiro2016why} and SHAP~\citep{lundberg2017unified}, which involve local surrogate models.
+In situations where full fidelity is a requirement, counterfactual explanations offer a more appropriate solution to Explainable Artificial Intelligence (XAI) than other popular approaches like LIME~\citep{ribeiro2016why} and SHAP~\citep{lundberg2017unified}, which involve local surrogate models.
 But even full fidelity is not a sufficient condition for ensuring that an explanation faithfully describes the behaviour of a model.
@@
 
-In the context of CE, the idea that no two explanations are the same arises almost naturally.
-A key focus in the literature has therefore been to identify those explanations and algorithmic recourses that are most appropriate based on a myriad of desiderata such as sparsity, actionability and plausibility.
+In the context of counterfactuals, the idea that no two explanations are the same arises almost naturally.
+A key focus in the literature has therefore been to identify those explanations and algorithmic recourses that are most appropriate based on a myriad of desiderata such as closeness~\citep{wachter2017counterfactual}, sparsity~\citep{schut2021generating}, actionability~\citep{ustun2019actionable} and plausibility~\citep{joshi2019realistic}.
+
 In this work, we draw closer attention to model faithfulness rather than fidelity as a desideratum for counterfactuals.
+We define faithfulness as the degree to which counterfactuals are consistent with what the model has learned about the data.
 Our key contributions are as follows:
@@
   \item We show that fidelity is an insufficient evaluation metric for counterfactuals (Section~\ref{fidelity}) and propose a definition of faithfulness that gives rise to more suitable metrics (Section~\ref{faithfulness}).
-  \item We introduce a novel algorithmic approach for generating Energy-Constrained Conformal Counterfactuals (ECCCos) in Section~\ref{meth}.
+  \item We introduce a novel algorithmic approach aimed at generating Energy-Constrained Conformal Counterfactuals (ECCCos) that faithfully explain model behaviour in Section~\ref{meth}.
   \item We provide extensive empirical evidence demonstrating that ECCCos faithfully explain model behaviour and attain plausibility only when appropriate (Section~\ref{emp}).
@@
 
-While CE can also be generated for arbitrary regression models~\citep{spooner2021counterfactual}, existing work has primarily focused on classification problems.
+While counterfactual explanations (CE) can also be generated for arbitrary regression models~\citep{spooner2021counterfactual}, existing work has primarily focused on classification problems.
 Let $\mathcal{Y}=(0,1)^K$ denote the one-hot-encoded output domain with $K$ classes.
@@
 
-Perhaps unsurprisingly, the different desiderata are often positively correlated.
+The notion of plausibility is central to all of the desiderata.
 For example, \citet{artelt2021evaluating} find that plausibility typically also leads to improved robustness.
 Similarly, plausibility has also been connected to causality in the sense that plausible counterfactuals respect causal relationships~\citep{mahajan2020preserving}.
+
 Consequently, the plausibility of counterfactuals has been among the primary concerns for researchers.
@@
 
-To generate plausible counterfactuals, we need to be able to quantify the DGP: $\mathcal{X}|\mathbf{y}^+$.
+To generate plausible counterfactuals, we first need to quantify the conditional distribution of samples in the target class ($\mathcal{X}|\mathbf{y}^+$).
+We can then ensure that we generate counterfactuals that comply with that distribution.
+
 One straightforward way to do this is to use surrogate models for the task.
@@
 
-\section{Why Fidelity is not Enough}\label{fidelity}
+\section{Why Fidelity is not Enough: A Motivational Example}\label{fidelity}
 
@@
 To assess counterfactuals with respect to Definition~\ref{def:faithful}, we need a way to quantify the posterior conditional distribution $p_{\theta}(\mathbf{x}|\mathbf{y}^+)$.
-To this end, we draw on recent advances in Energy-Based Modelling (EBM), a subdomain of machine learning that is concerned with generative or hybrid modelling~\citep{grathwohl2020your,du2020implicit}.
+To this end, we draw on recent advances in energy-based modelling (EBM), a subdomain of machine learning that is concerned with generative or hybrid modelling~\citep{grathwohl2020your,du2020implicit}.
 In particular, note that if we fix $\mathbf{y}$ to our target value $\mathbf{y}^+$, we can conditionally draw from $p_{\theta}(\mathbf{x}|\mathbf{y}^+)$ by randomly initializing $\mathbf{x}_0$ and then using Stochastic Gradient Langevin Dynamics (SGLD) as follows,
@@
   \begin{aligned}
-    \mathbf{x}_{j+1} &\leftarrow \mathbf{x}_j - \frac{\epsilon^2}{2} \mathcal{E}(\mathbf{x}_j|\mathbf{y}^+) + \epsilon \mathbf{r}_j, && j=1,...,J
+    \mathbf{x}_{j+1} &\leftarrow \mathbf{x}_j - \frac{\epsilon_j^2}{2} \mathcal{E}(\mathbf{x}_j|\mathbf{y}^+) + \epsilon_j \mathbf{r}_j, && j=1,...,J
   \end{aligned}
@@
 
-where $\mathbf{r}_j \sim \mathcal{N}(\mathbf{0},\mathbf{I})$ is the stochastic term and the step-size $\epsilon$ is typically polynomially decayed~\citep{welling2011bayesian}.
-The term $\mathcal{E}(\mathbf{x}_j|\mathbf{y}^+)$ denotes the model energy conditioned on the target class label $\mathbf{y}^+$ which we specify as the negative logit corresponding to the target class label $\mathbf{y}^*$.
-To allow for faster sampling, we follow the common practice of choosing the step-size $\epsilon$ and the standard deviation of $\mathbf{r}_j$ separately.
-While $\mathbf{x}_J$ is only guaranteed to distribute as $p_{\theta}(\mathbf{x}|\mathbf{y}^*)$ if $\epsilon \rightarrow 0$ and $J \rightarrow \infty$, the bias introduced for a small finite $\epsilon$ is negligible in practice \citep{murphy2023probabilistic,grathwohl2020your}.
+where $\mathbf{r}_j \sim \mathcal{N}(\mathbf{0},\mathbf{I})$ is the stochastic term and the step-size $\epsilon_j$ is typically polynomially decayed~\citep{welling2011bayesian}.
+The term $\mathcal{E}(\mathbf{x}_j|\mathbf{y}^+)$ denotes the model energy conditioned on the target class label $\mathbf{y}^+$ which we specify as the negative logit corresponding to the target class label $\mathbf{y}^{+}$.
+To allow for faster sampling, we follow the common practice of choosing the step-size $\epsilon_j$ and the standard deviation of $\mathbf{r}_j$ separately.
+While $\mathbf{x}_J$ is only guaranteed to distribute as $p_{\theta}(\mathbf{x}|\mathbf{y}^{+})$ if $\epsilon \rightarrow 0$ and $J \rightarrow \infty$, the bias introduced for a small finite $\epsilon$ is negligible in practice \citep{murphy2023probabilistic,grathwohl2020your}.
 Appendix~\ref{app:jem} provides additional implementation details for any tasks related to energy-based modelling.
 
-Generating multiple samples using SGLD thus yields an empirical distribution $\hat{\mathbf{X}}_{\theta,\mathbf{y}^+}$ that approximates what the model has learned about the input data.
+Generating multiple samples using SGLD thus yields an empirical distribution $\widehat{\mathbf{X}}_{\theta,\mathbf{y}^+}$ that approximates what the model has learned about the input data.
 While in the context of EBM, this is usually done during training, we propose to repurpose this approach during inference in order to evaluate and generate faithful model explanations.
@@
 Equation~\ref{eq:impl} gives rise to a similar evaluation metric for unfaithfulness.
-We merely swap out the subsample of individuals in the target class for a subset $\hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^+}$ of the generated conditional samples:
+We merely swap out the subsample of individuals in the target class for a subset $\widehat{\mathbf{X}}_{\mathbf{y}^+}$ of the generated conditional samples:
 
@@
   \begin{aligned}
-    \text{unfaith}(\mathbf{x}^{\prime},\hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^+}) = \frac{1}{\lvert \hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^+}\rvert} \sum_{\mathbf{x} \in \hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^+}} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
+    \text{unfaith}(\mathbf{x}^{\prime},\widehat{\mathbf{X}}_{\mathbf{y}^+}) = \frac{1}{n_E} \sum_{\mathbf{x} \in \widehat{\mathbf{X}}_{\mathbf{y}^+}} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
   \end{aligned}
@@
 
-In this section, we describe \textit{ECCCo}, our proposed framework for generating Energy-Constrained Conformal Counterfactuals (ECCCos).
+Given our proposed notion of faithfulness, we now describe \textit{ECCCo}, our proposed framework for generating Energy-Constrained Conformal Counterfactuals (ECCCos).
 It is based on the premise that counterfactuals should first and foremost be faithful.
@@
   \begin{aligned}
-  \mathbf{Z}^\prime= \arg \min_{\mathbf{Z}^\prime \in \mathcal{Z}^M}  &\{  {\text{yloss}(M_{\theta}(f(\mathbf{Z}^\prime)),\mathbf{y}^+)}+ \lambda_{1} {\text{dist}(f(\mathbf{Z}^\prime),\mathbf{x}) } \\
-  &+ \lambda_2 \text{unfaith}(f(\mathbf{Z}^\prime),\hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^+}) + \lambda_3 \Omega(C_{\theta}(f(\mathbf{Z}^\prime);\alpha)) \} 
+  \mathbf{Z}^\prime &= \arg \min_{\mathbf{Z}^\prime \in \mathcal{Z}^L} \{  {\text{yloss}(M_{\theta}(f(\mathbf{Z}^\prime)),\mathbf{y}^+)}+ \lambda_{1} {\text{dist}(f(\mathbf{Z}^\prime),\mathbf{x}) } \\
+  &+ \lambda_2 \text{unfaith}(f(\mathbf{Z}^\prime),\widehat{\mathbf{X}}_{\mathbf{y}^+}) + \lambda_3 \Omega(C_{\theta}(f(\mathbf{Z}^\prime);\alpha)) \} 
   \end{aligned} 
@@
 The third and final penalty term involving $\lambda_3$ introduces a new concept: it ensures that the generated counterfactual is associated with low predictive uncertainty.
-As mentioned above,~\citet{schut2021generating} have shown that plausible counterfactuals can be generated implicitly through predictive uncertainty minimization.
+As mentioned in Section~\ref{background},~\citet{schut2021generating} have shown that plausible counterfactuals can be generated implicitly through predictive uncertainty minimization.
 Unfortunately, this relies on the assumption that the model itself can provide predictive uncertainty estimates, which may be too restrictive in practice.
 
-To relax this assumption, we leverage recent advances in Conformal Prediction (CP), an approach to predictive uncertainty quantification that has recently gained popularity~\citep{angelopoulos2021gentle,manokhin2022awesome}.
+To relax this assumption, we leverage recent advances in conformal prediction (CP), an approach to predictive uncertainty quantification that has recently gained popularity~\citep{angelopoulos2021gentle,manokhin2022awesome}.
 Crucially for our intended application, CP is model-agnostic and can be applied during inference without placing any restrictions on model training.
@@
 Consequently, CP removes the restriction on the family of predictive models, at the small cost of reserving a subset of the available data for calibration.
-This particular case of conformal prediction is referred to as Split Conformal Prediction (SCP) as it involves splitting the training data into a proper training dataset and a calibration dataset.
-In addition to the smooth set size penalty, we have also experimented with the use of a tailored function for $\text{yloss}(\cdot)$ that enforces that only the target label $\mathbf{y}^+$ is included in the prediction set ~\citet{stutz2022learning}.
+This particular case of conformal prediction is referred to as *split conformal prediction* (SCP) as it involves splitting the training data into a proper training dataset and a calibration dataset.
 Further details are provided in Appendix~\ref{app:cp}.
@@
 
-\begin{algorithm}
+\begin{algorithm*}
   \caption{The \textit{ECCCo} generator}\label{alg:eccco}
@@
     \State Generate $\left\{\hat{\mathbf{x}}_{\theta,\mathbf{y}^+}\right\}_{n_{\mathcal{B}}} \gets p_{\theta}(\mathbf{x}_{\mathbf{y}^+})$ \Comment{Generate $n_{\mathcal{B}}$ samples using SGLD (Equation~\ref{eq:sgld}).}
-    \State Store $\hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^+} \gets \left\{\hat{\mathbf{x}}_{\theta,\mathbf{y}^+}\right\}_{n_{\mathcal{B}}}$ \Comment{Choose $n_E$ lowest-energy samples.}
-    \State Run \textit{SCP} for $M_{\theta}$ using $\mathcal{D}$ \Comment{Calibrate model through Split Conformal Prediction.}
+    \State Store $\widehat{\mathbf{X}}_{\mathbf{y}^+} \gets \left\{\hat{\mathbf{x}}_{\theta,\mathbf{y}^+}\right\}_{n_{\mathcal{B}}}$ \Comment{Choose $n_E$ lowest-energy samples.}
+    \State Run \textit{SCP} for $M_{\theta}$ using $\mathcal{D}$ \Comment{Calibrate model through split conformal prediction.}
     \State Initialize $t \gets 0$
     \While{\textit{not converged} or $t < T$} \Comment{For convergence conditions see Appendix~\ref{app:eccco}.}
-    \State $\mathbf{z}^\prime \gets \mathbf{z}^\prime - \eta \nabla_{\mathbf{z}^\prime} \mathcal{L}(\mathbf{z}^\prime,\mathbf{y}^+,\hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^+}; \Lambda, \alpha)$ \Comment{Take gradient step of size $\eta$.}
+    \State $\mathbf{z}^\prime \gets \mathbf{z}^\prime - \eta \nabla_{\mathbf{z}^\prime} \mathcal{L}(\mathbf{z}^\prime,\mathbf{y}^+,\widehat{\mathbf{X}}_{\mathbf{y}^+}; \Lambda, \alpha)$ \Comment{Take gradient step of size $\eta$.}
     \State $t \gets t+1$
@@
   \end{algorithmic}
-\end{algorithm}
+\end{algorithm*}
 
-To provide some further intuition about our objective defined in Equation~\ref{eq:eccco}, Figure~\ref{fig:poc} illustrates how the different components affect the counterfactual search for a synthetic dataset.
+Figure~\ref{fig:poc} illustrates how the different components in Equation~\ref{eq:eccco} affect the counterfactual search for a synthetic dataset.
 The underlying classifier is a Joint Energy Model (\textit{JEM}) that was trained to predict the output class (`blue' or `orange') and generate class-conditional samples~\citep{grathwohl2020your}.
@@
 Next, we generate $n_{\mathcal{B}}$ conditional samples $\hat{\mathbf{x}}_{\theta,\mathbf{y}^+}$ using SGLD (Equation~\ref{eq:sgld}) and store the $n_E$ instances with the lowest energy.
-We then calibrate the model $M_{\theta}$ through Split Conformal Prediction.
-Finally, we search counterfactuals through gradient descent where $\mathcal{L}(\mathbf{z}^\prime,\mathbf{y}^+,\hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^+}; \Lambda, \alpha)$ denotes our loss function defined in Equation~\ref{eq:eccco}.
+We then calibrate the model $M_{\theta}$ through split conformal prediction.
+Finally, we search counterfactuals through gradient descent where $\mathcal{L}(\mathbf{z}^\prime,\mathbf{y}^+,\widehat{\mathbf{X}}_{\mathbf{y}^+}; \Lambda, \alpha)$ denotes our loss function defined in Equation~\ref{eq:eccco}.
 The search terminates once the convergence criterium is met or the maximum number of iterations $T$ has been exhausted.
@@
 \begin{question}[Faithfulness]\label{rq:faithfulness}
-  Are ECCCos more faithful than counterfactuals produced by our benchmark generators?
+  To what extent are ECCCos more faithful than counterfactuals produced by state-of-the-art generators?
 \end{question}
@@
 \begin{question}[Balancing Objectives]\label{rq:plausibility}
-  Compared to our benchmark generators, how do ECCCos balance the two key objectives of faithfulness and plausibility?
+  Compared to state-of-the-art generators, how do ECCCos balance the two key objectives of faithfulness and plausibility?
 \end{question}
@@
 The second question is motivated by the intuition that faithfulness and plausibility should coincide for models that have learned plausible explanations of the data.
-Next, we first briefly describe our experimental setup before presenting our main results.
 
@@
 As for real-world data, we follow~\citet{schut2021generating} and use the \textit{MNIST}~\citep{lecun1998mnist} dataset containing images of handwritten digits such as the example shown above in Figure~\ref{fig:motiv}.
-From the social sciences domain, we include Give Me Some Credit (\textit{GMSC})~\citep{kaggle2011give}: a tabular dataset that has been studied extensively in the literature on Algorithmic Recourse~\citep{pawelczyk2021carla}.
+From the social sciences domain, we include Give Me Some Credit (\textit{GMSC})~\citep{kaggle2011give}: a tabular dataset that has been studied extensively in the literature on algorithmic recourse~\citep{pawelczyk2021carla}.
 It consists of 11 numeric features that can be used to predict the binary outcome variable indicating whether retail borrowers experience financial distress.
@@
 Zooming in on the granular details for the \textit{Linearly Separable} data, the results for \textit{ECCCo (no CP)} and \textit{ECCCo (no EBM)} indicate that the positive results are dominated by the effect of quantifying and leveraging the model's generative property (EBM).
-Conformal Prediction alone only leads to marginally improved faithfulness and plausibility.
+Conformal prediction alone only leads to marginally improved faithfulness and plausibility.
 
@@
 While our approach is readily applicable to models with gradient access like deep neural networks, more work is needed to generalise it to other machine learning models such as decision trees.
-Relatedly, common challenges associated with Energy-Based Modelling including sensitivity to scale, training instabilities and sensitivity to hyperparameters also apply to \textit{ECCCo}.
+Relatedly, common challenges associated with energy-based modelling including sensitivity to scale, training instabilities and sensitivity to hyperparameters also apply to \textit{ECCCo}.
 
@@
 
-This work leverages recent advances in Energy-Based Modelling and Conformal Prediction in the context of Explainable Artificial Intelligence.
+This work leverages recent advances in energy-based modelling and conformal prediction in the context of Explainable Artificial Intelligence.
 We have proposed a new way to generate counterfactuals that are maximally faithful to the black-box model they aim to explain.
@@
 This should enable researchers and practitioners to use counterfactuals in order to discern trustworthy models from unreliable ones.
-While the scope of this work limits its generalizability, we believe that \textit{ECCCo} offers a solid baseline for future work on faithful Counterfactual Explanations.
+While the scope of this work limits its generalizability, we believe that \textit{ECCCo} offers a solid baseline for future work on faithful counterfactual explanations.

### 2023-09-13 paper/body.tex
slowly slowly

@@
 Counterfactuals are then searched by minimizing a loss function that compares the predicted model output to the target outcome: $\text{yloss}(M_{\theta}(\mathbf{x}),\mathbf{y}^+)$.
-Since counterfactual explanations work directly with the black-box model, valid counterfactuals always have full local fidelity by construction where fidelity is defined as the degree to which explanations approximate the predictions of a black-box model~\citep{mothilal2020explaining,molnar2020interpretable}.
+Since counterfactual explanations work directly with the black-box model, valid counterfactuals always have full local fidelity by construction where fidelity is defined as the degree to which explanations approximate the predictions of a black-box model~\citep{mothilal2020explaining,molnar2022interpretable}.
 
@@
 For example, \citet{artelt2021evaluating} find that plausibility typically also leads to improved robustness.
-Similarly, plausibility has also been connected to causality in the sense that plausible counterfactuals respect causal relationships~\citep{mahajan2020preserving}.
+Similarly, plausibility has also been connected to causality in the sense that plausible counterfactuals respect causal relationships~\citep{mahajan2019preserving}.
 
@@
 To assess counterfactuals with respect to Definition~\ref{def:faithful}, we need a way to quantify the posterior conditional distribution $p_{\theta}(\mathbf{x}|\mathbf{y}^+)$.
-To this end, we draw on recent advances in energy-based modelling (EBM), a subdomain of machine learning that is concerned with generative or hybrid modelling~\citep{grathwohl2020your,du2020implicit}.
+To this end, we draw on recent advances in energy-based modelling (EBM), a subdomain of machine learning that is concerned with generative or hybrid modelling~\citep{grathwohl2020your,du2019implicit}.
 In particular, note that if we fix $\mathbf{y}$ to our target value $\mathbf{y}^+$, we can conditionally draw from $p_{\theta}(\mathbf{x}|\mathbf{y}^+)$ by randomly initializing $\mathbf{x}_0$ and then using Stochastic Gradient Langevin Dynamics (SGLD) as follows,
@@
 
-For the tabular credit dataset (\textit{GMSC}) it is inherently challenging to use deep neural networks in order to achieve good discriminative performance~\citep{borisov2021deep,grinsztajn2022why} and generative performance~\citep{liu2023goggle}, respectively.
+For the tabular credit dataset (\textit{GMSC}) it is inherently challenging to use deep neural networks in order to achieve good discriminative performance~\citep{borisov2022deep,grinsztajn2022why} and generative performance~\citep{liu2023goggle}, respectively.
 In order to achieve high plausibility, \textit{ECCCo} effectively requires classifiers to achieve good performance for both tasks.

### 2023-09-13 paper/body.tex
just missing section 6

@@
 Consequently, the generated explanations may seem plausible to humans but need not necessarily describe the behaviour of the black-box model faithfully.
-We formalise this notion of faithfulness through the introduction of a tailored evaluation metric and propose a novel algorithmic framework for generating \textbf{E}nergy-\textbf{C}onstrained \textbf{C}onformal \textbf{Co}unterfactuals (ECCCos) that are only as plausible as the model permits.
-Through extensive empirical studies, we demonstrate that ECCCos reconcile the need for faithfulness and plausibility.
+We formalise this notion of faithfulness through the introduction of a tailored evaluation metric and propose a novel algorithmic framework for generating \textbf{E}nergy-\textbf{C}onstrained \textbf{C}onformal \textbf{Co}unterfactuals that are only as plausible as the model permits.
+Through extensive empirical studies, we demonstrate that \textit{ECCCo} reconciles the need for faithfulness and plausibility.
 In particular, we show that for models with gradient access, it is possible to achieve state-of-the-art performance without the need for surrogate models.
@@
 To our knowledge, this is the first venture in this direction for generating faithful counterfactual explanations.
-Thus, we anticipate that ECCCos can serve as a baseline for future research.
+Thus, we anticipate that \textit{ECCCo} can serve as a baseline for future research.
 We believe that our work opens avenues for researchers and practitioners seeking tools to better distinguish trustworthy from unreliable models.
@@
   \item We show that fidelity is an insufficient evaluation metric for counterfactuals (Section~\ref{fidelity}) and propose a definition of faithfulness that gives rise to more suitable metrics (Section~\ref{faithfulness}).
-  \item We introduce a novel algorithmic approach aimed at generating Energy-Constrained Conformal Counterfactuals (ECCCos) that faithfully explain model behaviour in Section~\ref{meth}.
-  \item We provide extensive empirical evidence demonstrating that ECCCos faithfully explain model behaviour and attain plausibility only when appropriate (Section~\ref{emp}).
+  \item We introduce a ECCCo: a novel algorithmic approach aimed at generating Energy-Constrained Conformal Counterfactuals that faithfully explain model behaviour in Section~\ref{meth}.
+  \item We provide extensive empirical evidence demonstrating that ecccos faithfully explain model behaviour and attain plausibility only when appropriate (Section~\ref{emp}).
 \end{itemize}
@@
 To our knowledge, this is the first venture in this direction for generating faithful counterfactuals.
-Thus, we anticipate that ECCCos can serve as a baseline for future research.
+Thus, we anticipate that \textit{ECCCo} can serve as a baseline for future research.
 We believe that our work opens avenues for researchers and practitioners seeking tools to better distinguish trustworthy from unreliable models.
@@
 
-\section{A New Notion of Faithfulness}\label{faithfulness}
+\section{Faithful first, Plausible second}\label{faithfulness}
 
@@
 
+\subsection{Quantifying the Model's Predictive Uncertainty}
+
+Faithful counterfactuals can be expected to also be plausible if the learned conditional distribution $\mathcal{X}_{\theta}|\mathbf{y}^+$ (Defintion~\ref{def:faithful}) is close to the true conditional distribution $\mathcal{X}|\mathbf{y}^+$ (Definition~\ref{def:plausible}).
+We can further improve plausibility of counterfactuals without the need for surrogate models that may interfer with faithfulness by minimizing predictive uncertainty~\citep{schut2021generating}.
+Unfortunately, this approach relies on the assumption that the model itself can provide predictive uncertainty estimates, which may be too restrictive in practice.
+
+To relax this assumption, we use conformal prediction (CP), an approach to predictive uncertainty quantification that has recently gained popularity~\citep{angelopoulos2021gentle,manokhin2022awesome}.
+Crucially for our intended application, CP is model-agnostic and can be applied during inference without placing any restrictions on model training.
+Intuitively, CP works under the premise of turning heuristic notions of uncertainty into rigorous uncertainty estimates by repeatedly sifting through the training data or a dedicated calibration dataset.
+
+Conformal classifiers produce prediction sets for individual inputs that include all output labels that can be reasonably attributed to the input. Finally, classification sets are formed as follows,
+
+\begin{equation}\label{eq:scp}
+  \begin{aligned}
+    C_{\theta}(\mathbf{x}_i;\alpha)=\{\mathbf{y}: s(\mathbf{x}_i,\mathbf{y}) \le \hat{q}\}
+  \end{aligned}
+\end{equation}
+
+where $\hat{q}$ denotes the $(1-\alpha)$-quantile of $\mathcal{S}$ and $\alpha$ is a predetermined error rate.
+
+These sets tend to be larger for inputs that do not conform with the training data and are characterized by high predictive uncertainty.
+To leverage this notion of predictive uncertainty in the context of gradient-based counterfactual search, we use a smooth set size penalty introduced by~\citet{stutz2022learning}:
+
+\begin{equation}\label{eq:setsize}
+  \begin{aligned}
+    \Omega(C_{\theta}(\mathbf{x};\alpha))&=\max \left(0, \sum_{\mathbf{y}\in\mathcal{Y}}C_{\theta,\mathbf{y}}(\mathbf{x}_i;\alpha) - \kappa \right)
+  \end{aligned}
+\end{equation}
+
+Here, $\kappa \in \{0,1\}$ is a hyper-parameter and $C_{\theta,\mathbf{y}}(\mathbf{x}_i;\alpha)$ can be interpreted as the probability of label $\mathbf{y}$ being included in the prediction set (see Appendix~\ref{app:cp-diff} for details).
+In order to compute this penalty for any black-box model we merely need to perform a single calibration pass through a holdout set $\mathcal{D}_{\text{cal}}$.
+Arguably, data is typically abundant and in most applications, practitioners tend to hold out a test data set anyway.
+Consequently, CP removes the restriction on the family of predictive models, at the small cost of reserving a subset of the available data for calibration.
+This particular case of conformal prediction is referred to as *split conformal prediction* (SCP) as it involves splitting the training data into a proper training dataset and a calibration dataset.
+
 \subsection{Evaluating Plausibility and Faithfulness}
@@
 
-Given our proposed notion of faithfulness, we now describe \textit{ECCCo}, our proposed framework for generating Energy-Constrained Conformal Counterfactuals (ECCCos).
+Given our proposed notion of faithfulness, we now describe \textit{ECCCo}, our proposed framework for generating Energy-Constrained Conformal Counterfactuals.
 It is based on the premise that counterfactuals should first and foremost be faithful.
@@
 The second penalty term involving $\lambda_2$ induces faithfulness by constraining the energy of the generated counterfactual where $\text{unfaith}(\cdot)$ corresponds to the metric defined in Equation~\ref{eq:faith}.
-The third and final penalty term involving $\lambda_3$ introduces a new concept: it ensures that the generated counterfactual is associated with low predictive uncertainty.
-As mentioned in Section~\ref{background},~\citet{schut2021generating} have shown that plausible counterfactuals can be generated implicitly through predictive uncertainty minimization.
-Unfortunately, this relies on the assumption that the model itself can provide predictive uncertainty estimates, which may be too restrictive in practice.
-
-To relax this assumption, we leverage recent advances in conformal prediction (CP), an approach to predictive uncertainty quantification that has recently gained popularity~\citep{angelopoulos2021gentle,manokhin2022awesome}.
-Crucially for our intended application, CP is model-agnostic and can be applied during inference without placing any restrictions on model training.
-Intuitively, CP works under the premise of turning heuristic notions of uncertainty into rigorous uncertainty estimates by repeatedly sifting through the training data or a dedicated calibration dataset.
-Conformal classifiers produce prediction sets for individual inputs that include all output labels that can be reasonably attributed to the input.
-These sets tend to be larger for inputs that do not conform with the training data and are characterized by high predictive uncertainty.
-
-In order to generate counterfactuals that are associated with low predictive uncertainty, we use a smooth set size penalty introduced by~\citet{stutz2022learning} in the context of conformal training:
-
-\begin{equation}\label{eq:setsize}
-  \begin{aligned}
-    \Omega(C_{\theta}(\mathbf{x};\alpha))&=\max \left(0, \sum_{\mathbf{y}\in\mathcal{Y}}C_{\theta,\mathbf{y}}(\mathbf{x}_i;\alpha) - \kappa \right)
-  \end{aligned}
-\end{equation}
-
-Here, $\kappa \in \{0,1\}$ is a hyper-parameter and $C_{\theta,\mathbf{y}}(\mathbf{x}_i;\alpha)$ can be interpreted as the probability of label $\mathbf{y}$ being included in the prediction set.
-In order to compute this penalty for any black-box model we merely need to perform a single calibration pass through a holdout set $\mathcal{D}_{\text{cal}}$.
-Arguably, data is typically abundant and in most applications, practitioners tend to hold out a test data set anyway.
-Consequently, CP removes the restriction on the family of predictive models, at the small cost of reserving a subset of the available data for calibration.
-This particular case of conformal prediction is referred to as *split conformal prediction* (SCP) as it involves splitting the training data into a proper training dataset and a calibration dataset.
-Further details are provided in Appendix~\ref{app:cp}.
+The third and final penalty term involving $\lambda_3$ ensures that the generated counterfactual is associated with low predictive uncertainty.
 
@@
 \begin{question}[Faithfulness]\label{rq:faithfulness}
-  To what extent are ECCCos more faithful than counterfactuals produced by state-of-the-art generators?
+  To what extent are counterfactuals generated by \textit{ECCCo} more faithful than those produced by state-of-the-art generators?
 \end{question}
@@
 \begin{question}[Balancing Objectives]\label{rq:plausibility}
-  Compared to state-of-the-art generators, how do ECCCos balance the two key objectives of faithfulness and plausibility?
+  Compared to state-of-the-art generators, how does \textit{ECCCo} balance the two key objectives of faithfulness and plausibility?
 \end{question}
@@
 For the more complex real-world datasets we also use ensembling in each case.
-Both joint-energy modelling and ensembling have been associated with improved generative properties and adversarial robustness~\citep{grathwohl2020your,lakshminarayanan2016simple}, so we expect this to be positively correlated with the plausibility of ECCCos.
+Both joint-energy modelling and ensembling have been associated with improved generative properties and adversarial robustness~\citep{grathwohl2020your,lakshminarayanan2016simple}, so we expect this to be positively correlated with the plausibility of \textit{ECCCo}.
 To account for stochasticity, we generate multiple counterfactuals for each target class, generator, model and dataset.
@@
 For the \textit{MLP}, faithfulness is maintained but counterfactuals are not plausible.
-This high-level pattern is broadly consistent with other more complex datasets and supportive of our narrative, so it is worth highlighting: ECCCos consistently achieve high faithfulness, which---subject to the quality of the model itself---coincides with high plausibility.
+This high-level pattern is broadly consistent with other more complex datasets and supportive of our narrative, so it is worth highlighting: \textit{ECCCo} consistently achieves high faithfulness, which---subject to the quality of the model itself---coincides with high plausibility.
 By comparison, \textit{REVISE} yields the most plausible counterfactuals for the \textit{MLP}, but it does so at the cost of faithfulness.
@@
 The results for our real-world datasets are shown in Table~\ref{tab:results-real-world}.
-Once again the findings indicate that the plausibility of ECCCos is positively correlated with the capacity of the black-box model to distinguish plausible from implausible inputs.
-The case is very clear for \textit{MNIST}: ECCCos are consistently more faithful than the counterfactuals produced by our benchmark generators and their plausibility gradually improves through ensembling and joint-energy modelling.
+Once again the findings indicate that the plausibility attained by \textit{ECCCo} is positively correlated with the capacity of the black-box model to distinguish plausible from implausible inputs.
+The case is very clear for \textit{MNIST}: \textit{ECCCo} consistently generates more faithful counterfactuals than other generators and plausibility gradually improves through ensembling and joint-energy modelling.
 Interestingly, faithfulness also gradually improves for \textit{REVISE}.
@@
 
-For the tabular credit dataset (\textit{GMSC}) it is inherently challenging to use deep neural networks in order to achieve good discriminative performance~\citep{borisov2022deep,grinsztajn2022why} and generative performance~\citep{liu2023goggle}, respectively.
+For the tabular credit dataset (\textit{GMSC}) it is inherently challenging to use deep neural networks in order to achieve good discriminative performance~\citep{borisov2022deep,grinsztajn2022why} and generative performance~\citep{liu2022goggle}, respectively.
 In order to achieve high plausibility, \textit{ECCCo} effectively requires classifiers to achieve good performance for both tasks.
@@
 Even though we have taken considerable measures to study our proposed methodology carefully, limitations can still be identified.
-In particular, we have found that the performance of \textit{ECCCo} is sensitive to hyperparameter choices.
-In order to achieve faithfulness, we generally had to penalise the distance from generated samples slightly more than the distance from factual values.
 
-Conversely, we have not found that strongly penalising prediction set sizes had any discernable effect.
-Our results indicate that CP alone is often not sufficient to achieve faithfulness and plausibility, although we acknowledge that this needs to be investigated more thoroughly through future work.
+\subsection{Evaluation Metrics}
+
+Our proposed distance-based evaluation metrics for plausibility and faithfulness may not be universally applicable to all types of data.
+In any case, they depend on choosing a distance metric on a case-by-base basis for different datasets.
+Arguably, commonly used metrics for measuring other desiderata such as closeness suffer from the same pitfall.
+We therefore think that future work on counterfactual explanations could benefit from defining universal evaluation metrics.
+
+\subsection{Experiments}
+
+While we have employed various datasets in our experiments that are commonly used in the related literature, we acknowledge that additional real-world data and application is needed to test \textit{ECCCo} and improve upon the ideas we have presented in this work.
+One challenge in this context is that counterfactual explanations do not scale very well to high-dimensional input data like images~\citep{samoilescu2021model,chen2021seven}.
+Consequently, we have limited ourselves to studying small image datasets only.
+
+\subsection{Generalizability}
 
@@
 Relatedly, common challenges associated with energy-based modelling including sensitivity to scale, training instabilities and sensitivity to hyperparameters also apply to \textit{ECCCo}.
+
+\subsection{Ablation Studies}
+
+In our experiments we have used ablation to understand the roles of the different components of \textit{ECCCo}.
+Our results here indicate that conformal prediction alone is often not sufficient to achieve faithfulness and plausibility.
+To test this initial finding more throughly, future work could benefit from more extensive abalation studies that thoroughly tune hyperparameters and investigate different approaches to conformal prediction.
 

### 2023-09-19 paper/body.tex
added energy delta

@@
 Equation~\ref{eq:impl} gives rise to a similar evaluation metric for unfaithfulness.
-We merely swap out the subsample of individuals in the target class for a subset $\widehat{\mathbf{X}}_{\mathbf{y}^+}$ of the generated conditional samples:
+We swap out the subsample of observed individuals in the target class for the set of samples generated through SGLD ($\widehat{\mathbf{X}}_{\mathbf{y}^+}$):
 
@@
   \begin{aligned}
-    \text{unfaith}(\mathbf{x}^{\prime},\widehat{\mathbf{X}}_{\mathbf{y}^+}) = \frac{1}{n_E} \sum_{\mathbf{x} \in \widehat{\mathbf{X}}_{\mathbf{y}^+}} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
+    \text{unfaith}(\mathbf{x}^{\prime},\widehat{\mathbf{X}}_{\theta,\mathbf{y}^+}) = \frac{1}{\lvert \widehat{\mathbf{X}}_{\theta,\mathbf{y}^+} \rvert} \sum_{\mathbf{x} \in \widehat{\mathbf{X}}_{\theta,\mathbf{y}^+}} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
   \end{aligned}
@@
 
-Specifically, we form this subset based on the $n_E$ generated samples with the lowest energy.
+Our default choice for the $\text{dist}(\cdot)$ in both cases is the L1 Norm.
 
@@
   \mathbf{Z}^\prime &= \arg \min_{\mathbf{Z}^\prime \in \mathcal{Z}^L} \{  {\text{yloss}(M_{\theta}(f(\mathbf{Z}^\prime)),\mathbf{y}^+)}+ \lambda_{1} {\text{dist}(f(\mathbf{Z}^\prime),\mathbf{x}) } \\
-  &+ \lambda_2 \text{unfaith}(f(\mathbf{Z}^\prime),\widehat{\mathbf{X}}_{\mathbf{y}^+}) + \lambda_3 \Omega(C_{\theta}(f(\mathbf{Z}^\prime);\alpha)) \} 
+  &+ \lambda_2 \Delta\mathcal{E}(\mathbf{Z}^\prime,\widehat{\mathbf{X}}_{\theta,\mathbf{y}^+}) + \lambda_3 \Omega(C_{\theta}(f(\mathbf{Z}^\prime);\alpha)) \} 
   \end{aligned} 
@@
 Our default choice for $\text{dist}(\cdot)$ is the L1 Norm due to its sparsity-inducing properties.
-The second penalty term involving $\lambda_2$ induces faithfulness by constraining the energy of the generated counterfactual where $\text{unfaith}(\cdot)$ corresponds to the metric defined in Equation~\ref{eq:faith}.
+The second penalty term involving $\lambda_2$ induces faithfulness by constraining the energy of the generated counterfactual where we have:
+
+\begin{equation} \label{eq:energy-delta}
+  \begin{aligned}
+    \Delta\mathcal{E}&=\mathcal{E}(f(\mathbf{Z}^\prime)|\mathbf{y}^+)-\mathcal{E}(x|\mathbf{y}^+) &&& x \sim \widehat{\mathbf{X}}_{\theta,\mathbf{y}^+}
+  \end{aligned}
+\end{equation}
+
+In particular, this penalty ensures that the energy of the generated counterfactual is in balance with the energy of the generated conditional samples ($\widehat{\mathbf{X}}_{\theta,\mathbf{y}^+}$).
 The third and final penalty term involving $\lambda_3$ ensures that the generated counterfactual is associated with low predictive uncertainty.
@@
     \State Generate $\left\{\hat{\mathbf{x}}_{\theta,\mathbf{y}^+}\right\}_{n_{\mathcal{B}}} \gets p_{\theta}(\mathbf{x}_{\mathbf{y}^+})$ \Comment{Generate $n_{\mathcal{B}}$ samples using SGLD (Equation~\ref{eq:sgld}).}
-    \State Store $\widehat{\mathbf{X}}_{\mathbf{y}^+} \gets \left\{\hat{\mathbf{x}}_{\theta,\mathbf{y}^+}\right\}_{n_{\mathcal{B}}}$ \Comment{Choose $n_E$ lowest-energy samples.}
+    \State Store $\widehat{\mathbf{X}}_{\theta,\mathbf{y}^+} \gets \left\{\hat{\mathbf{x}}_{\theta,\mathbf{y}^+}\right\}_{n_{\mathcal{B}}}$ \Comment{Choose $n_E$ lowest-energy samples.}
     \State Run \textit{SCP} for $M_{\theta}$ using $\mathcal{D}$ \Comment{Calibrate model through split conformal prediction.}
@@
     \While{\textit{not converged} or $t < T$} \Comment{For convergence conditions see Appendix~\ref{app:eccco}.}
-    \State $\mathbf{z}^\prime \gets \mathbf{z}^\prime - \eta \nabla_{\mathbf{z}^\prime} \mathcal{L}(\mathbf{z}^\prime,\mathbf{y}^+,\widehat{\mathbf{X}}_{\mathbf{y}^+}; \Lambda, \alpha)$ \Comment{Take gradient step of size $\eta$.}
+    \State $\mathbf{z}^\prime \gets \mathbf{z}^\prime - \eta \nabla_{\mathbf{z}^\prime} \mathcal{L}(\mathbf{z}^\prime,\mathbf{y}^+,\widehat{\mathbf{X}}_{\theta,\mathbf{y}^+}; \Lambda, \alpha)$ \Comment{Take gradient step of size $\eta$.}
     \State $t \gets t+1$
@@
 We then calibrate the model $M_{\theta}$ through split conformal prediction.
-Finally, we search counterfactuals through gradient descent where $\mathcal{L}(\mathbf{z}^\prime,\mathbf{y}^+,\widehat{\mathbf{X}}_{\mathbf{y}^+}; \Lambda, \alpha)$ denotes our loss function defined in Equation~\ref{eq:eccco}.
+Finally, we search counterfactuals through gradient descent where $\mathcal{L}(\mathbf{z}^\prime,\mathbf{y}^+,\widehat{\mathbf{X}}_{\theta,\mathbf{y}^+}; \Lambda, \alpha)$ denotes our loss function defined in Equation~\ref{eq:eccco}.
 The search terminates once the convergence criterium is met or the maximum number of iterations $T$ has been exhausted.

### 2023-09-19 paper/body.tex
s

@@
 
-Our default choice for the $\text{dist}(\cdot)$ in both cases is the L1 Norm.
+Our default choice for the $\text{dist}(\cdot)$ in both cases is the L1 Norm, which is consistent with the common choice for evaluating counterfactuals with respect to the closeness desideratum~\citep{pawelczyk2021carla,pawelczyk2022probabilistically}.
 

### 2023-09-20 paper/appendix.tex
slowly slowly
just missing section 6
more shit in the app

@@
 Following best practices, we initialize $\mathbf{x}_0$ randomly in 5\% of all cases and sample from a buffer in all other cases.
-The buffer itself is randomly initialised and gradually grows to a maximum of 10,000 samples during training as $\hat{\mathbf{x}}_{J}$ is stored in each epoch~\citep{du2020implicit,grathwohl2020your}.
+The buffer itself is randomly initialised and gradually grows to a maximum of 10,000 samples during training as $\hat{\mathbf{x}}_{J}$ is stored in each epoch~\citep{du2019implicit,grathwohl2020your}.
 
@@
   \begin{aligned}
-    L(\theta) &= L_{\text{clf}}(\theta) + L_{\text{gen}}(\theta) + \lambda L_{\text{reg}}(\theta) 
+    L_{\text{JEM}}(\theta) &= L_{\text{clf}}(\theta) + L_{\text{gen}}(\theta) + \lambda L_{\text{reg}}(\theta) 
   \end{aligned}
@@
 
-\subsubsection{Differentiability}
+\subsubsection{Differentiability}\label{app:cp-diff}
 
@@
 
-\subsection{ECCCo}\label{app:eccco}
+\subsection{\textit{ECCCo}}\label{app:eccco}
 
-In this section, we briefly discuss convergence conditions for CE and provide details concerning the actual implementation of our framework in \texttt{Julia}.  
+In this section, we explain \textit{ECCCo} in some more detail, briefly discuss convergence conditions for counterfactual explanations and provide details concerning the actual implementation of our framework in \texttt{Julia}.
+
+\subsubsection{More detail on our generator}
+
+The counterfactual search objective for \textit{ECCCo} was introduced in Equation~\ref{eq:eccco} in the body of the paper. We restate this equation here for reference:
+
+\begin{equation} \label{eq:eccco-app}
+  \begin{aligned}
+  \mathbf{Z}^\prime &= \arg \min_{\mathbf{Z}^\prime \in \mathcal{Z}^L} \{  {\text{yloss}(M_{\theta}(f(\mathbf{Z}^\prime)),\mathbf{y}^+)}+ \lambda_{1} {\text{dist}(f(\mathbf{Z}^\prime),\mathbf{x}) } \\
+  &+ \lambda_2 \Delta\mathcal{E}(\mathbf{Z}^\prime,\widehat{\mathbf{X}}_{\theta,\mathbf{y}^+}) + \lambda_3 \Omega(C_{\theta}(f(\mathbf{Z}^\prime);\alpha)) \} 
+  \end{aligned} 
+\end{equation}
+
+We can make the connection to energy-based modeling more explicit by restating this equation in terms $L_{\text{JEM}}(\theta)$, which we defined in Equation~\ref{eq:jem-loss}.
+In particular, note that for $\lambda_2=1$ and $\lambda L_{\text{reg}}(\theta)=0$ we have
+
+\begin{equation} \label{eq:eccco-jem}
+  \begin{aligned}
+  \mathbf{Z}^\prime &= \arg \min_{\mathbf{Z}^\prime \in \mathcal{Z}^L} \{  {L_{\text{JEM}}(\theta;M_{\theta}(f(\mathbf{Z}^\prime)),\mathbf{y}^+)}+ \lambda_{1} {\text{dist}(f(\mathbf{Z}^\prime),\mathbf{x}) }  + \lambda_3 \Omega(C_{\theta}(f(\mathbf{Z}^\prime);\alpha)) \}
+  \end{aligned} 
+\end{equation}
+
+since $\Delta\mathcal{E}(\cdot)$ is equivalent to the generative loss function $L_{\text{gen}}(\cdot)$.
+In fact, this is also true for $\lambda L_{\text{reg}}(\theta)\ne0$ since we use the Ridge penalty $L_{\text{reg}}(\theta)$ in the counterfactual search just like we do in joint-energy training.
+This detail was omitted from the body of the paper for the sake of simplicity.
+
+Aside from the additional penalties in Equation~\ref{eq:eccco-app}, the only key difference between our counterfactual search objective and the joint-energy training objective is the parameter that is being optimized.
+In joint-energy training we optimize the objective with respect to the network weights $\theta$:
+
+\begin{equation}\label{eq:jem-grad}
+  \begin{aligned}
+    \nabla_{\theta} L_{\text{JEM}}(\theta) &=\nabla_{\theta} L_{\text{clf}}(\theta) + \nabla_{\theta}L_{\text{gen}}(\theta) + \lambda \nabla_{\theta} L_{\text{reg}}(\theta) 
+  \end{aligned}
+\end{equation}
+
+we take these parameters as fixed in our counterfactual search.
+
 \subsubsection{A Note on Convergence}
@@
 Pixel values were preprocessed to fall in the range of $[-1,1]$ and a small Gaussian noise component ($\sigma=0.03$) was added to training samples following common practice in the EBM literature.
-All of our models were trained through mini-batch training using the Adam optimiser (\citet{kingma2017adam}).
+All of our models were trained through mini-batch training using the Adam optimiser (\citet{kingma2014adam}).
 Table~\ref{tab:perf} shows standard evaluation metrics measuring the predictive performance of our different models grouped by dataset.

### 2023-09-20 paper/appendix.tex
shoot me
more work on the appendix

@@
 The package has been open-sourced, but to avoid compromising the double-blind review process, we refrain from providing more information at this stage.
-In our development we have heavily drawn on the existing literature:~\citet{du2020implicit} describe best practices for using EBM for generative modelling;~\citet{grathwohl2020your} explain how EBM can be used to train classifiers jointly for the discriminative and generative tasks.
+In our development we have heavily drawn on the existing literature:~\citet{du2019implicit} describe best practices for using EBM for generative modelling;~\citet{grathwohl2020your} explain how EBM can be used to train classifiers jointly for the discriminative and generative tasks.
 We have used the same package for training and inference, but there are some important differences between the two cases that are worth highlighting here.
@@
 
-To train our Joint Energy Models we broadly follow the approach outlined in~\citet{grathwohl2020your}.
-These models are trained to optimize a hybrid objective that involves a standard classification loss component $L_{\text{clf}}(\theta)=-\log p_{\theta}(\mathbf{y}|\mathbf{x})$ (e.g.
+To train our Joint Energy Models we broadly follow the approach outlined in~\citet{grathwohl2020your}. Formally, JEMs are defined by the following joint distribution:
+
+\begin{equation}
+  \begin{aligned}
+    \log p\_{\theta}(\mathbf{x},\mathbf{y}) &= \log p_{\theta}(\mathbf{y}|\mathbf{x}) + \log p_{\theta}(\mathbf{x})
+  \end{aligned}
+\end{equation}
+
+Training therefore involves a standard classification loss component $L_{\text{clf}}(\theta)=-\log p_{\theta}(\mathbf{y}|\mathbf{x})$ (e.g.
 cross-entropy loss) as well as a generative loss component $L_{\text{gen}}(\theta)=-\log p_{\theta}(\mathbf{x})$.
+Analogous to how we defined the conditional distribution over inputs in Definition~\ref{def:faithful}, $p_{\theta}(\mathbf{x})$ denotes the unconditional distribution over inputs.
+The model gradient of this component of the loss function can be expressed as follows:
 
-To draw samples from $p_{\theta}(\mathbf{x})$, we rely exclusively on the conditional sampling approach described in~\citet{grathwohl2020your} for both training and inference: we first draw $\mathbf{y}\sim p(\mathbf{y})$ and then sample $\mathbf{x} \sim p_{\theta}(\mathbf{x}|\mathbf{y})$~\citep{grathwohl2020your} via Equation~\ref{eq:sgld} with energy $\mathcal{E}(\mathbf{x}|\mathbf{y})=\mu_{\theta}(\mathbf{x})[\mathbf{y}]$ where $\mu_{\theta}: \mathcal{X} \mapsto \mathbb{R}^K$ returns the linear predictions (logits) of our classifier $M_{\theta}$.
+\begin{equation}\label{eq:gen-true}
+  \begin{aligned}
+    \nabla_{\theta}L_{\text{gen}}(\theta)&=-\nabla_{\theta}\log p\_{\theta}(\mathbf{x})=-\left(\mathbb{E}_{p(\mathbf{x})} \left\{  \nabla_{\theta} \mathcal{E}_{\theta}(\mathbf{x}) \right\} - \mathbb{E}_{p_{\theta}(\mathbf{x})} \left\{  \nabla_{\theta} \mathcal{E}_{\theta}(\mathbf{x}) \right\} \right)
+  \end{aligned}
+\end{equation}
+
+To draw samples from $p_{\theta}(\mathbf{x})$, we rely exclusively on the conditional sampling approach described in~\citet{grathwohl2020your} for both training and inference: we first draw $\mathbf{y}\sim p(\mathbf{y})$ and then sample $\mathbf{x} \sim p_{\theta}(\mathbf{x}|\mathbf{y})$~\citep{grathwohl2020your} via Equation~\ref{eq:sgld} with energy $\mathcal{E}_{\theta}(\mathbf{x}|\mathbf{y})=\mu_{\theta}(\mathbf{x})[\mathbf{y}]$ where $\mu_{\theta}: \mathcal{X} \mapsto \mathbb{R}^K$ returns the linear predictions (logits) of our classifier $M_{\theta}$.
 While our package also supports unconditional sampling, we found conditional sampling to work well.
@@
   \begin{aligned}
-    \hat{\mathbf{x}}_{j+1} &\leftarrow \hat{\mathbf{x}}_j - \frac{\epsilon}{2} \mathcal{E}(\hat{\mathbf{x}}_j|\mathbf{y}^+) + \sigma \mathbf{r}_j, && j=1,...,J
+    \hat{\mathbf{x}}_{j+1} &\leftarrow \hat{\mathbf{x}}_j - \frac{\epsilon}{2} \mathcal{E}_{\theta}(\hat{\mathbf{x}}_j|\mathbf{y}^+) + \sigma \mathbf{r}_j, && j=1,...,J
   \end{aligned}
@@
 It is important to realise that sampling is done during each training epoch, which makes training Joint Energy Models significantly harder than conventional neural classifiers.
-In each epoch the generated (batch of) sample(s) $\hat{\mathbf{x}}_{J}$ is used as part of the generative loss component, which compares its energy to that of observed samples $\mathbf{x}$: $L_{\text{gen}}(\theta)=\mu_{\theta}(\mathbf{x})[\mathbf{y}]-\mu_{\theta}(\hat{\mathbf{x}}_{J})[\mathbf{y}]$.
+In each epoch the generated (batch of) sample(s) $\hat{\mathbf{x}}_{J}$ is used as part of the generative loss component, which compares its energy to that of observed samples $\mathbf{x}$:
+
+\begin{equation}\label{eq:gen-loss}
+  \begin{aligned}
+    L_{\text{gen}}(\theta)&\approx\mu_{\theta}(\mathbf{x})[\mathbf{y}]-\mu_{\theta}(\hat{\mathbf{x}}_{J})[\mathbf{y}]
+  \end{aligned}
+\end{equation}
+
 Our full training objective can be summarized as follows,
@@
 
-where $L_{\text{reg}}(\theta)$ is a Ridge penalty (L2 norm) that regularises energy magnitudes for both observed and generated samples~\citep{du2020implicit}.
+where $L_{\text{reg}}(\theta)$ is a Ridge penalty (L2 norm) that regularises energy magnitudes for both observed and generated samples~\citep{du2019implicit}.
 We have used varying degrees of regularization depending on the dataset ($\lambda$ in Table~\ref{tab:ebmparams}).
@@
   \mathbf{Z}^\prime &= \arg \min_{\mathbf{Z}^\prime \in \mathcal{Z}^L} \{  {\text{yloss}(M_{\theta}(f(\mathbf{Z}^\prime)),\mathbf{y}^+)}+ \lambda_{1} {\text{dist}(f(\mathbf{Z}^\prime),\mathbf{x}) } \\
-  &+ \lambda_2 \Delta\mathcal{E}(\mathbf{Z}^\prime,\widehat{\mathbf{X}}_{\theta,\mathbf{y}^+}) + \lambda_3 \Omega(C_{\theta}(f(\mathbf{Z}^\prime);\alpha)) \} 
+  &+ \lambda_2 \Delta\mathcal{E}_{\theta}(\mathbf{Z}^\prime,\widehat{\mathbf{X}}_{\theta,\mathbf{y}^+}) + \lambda_3 \Omega(C_{\theta}(f(\mathbf{Z}^\prime);\alpha)) \} 
   \end{aligned} 
@@
 
-since $\Delta\mathcal{E}(\cdot)$ is equivalent to the generative loss function $L_{\text{gen}}(\cdot)$.
+since $\Delta\mathcal{E}_{\theta}(\cdot)$ is equivalent to the generative loss function $L_{\text{gen}}(\cdot)$.
 In fact, this is also true for $\lambda L_{\text{reg}}(\theta)\ne0$ since we use the Ridge penalty $L_{\text{reg}}(\theta)$ in the counterfactual search just like we do in joint-energy training.
@@
 Aside from the additional penalties in Equation~\ref{eq:eccco-app}, the only key difference between our counterfactual search objective and the joint-energy training objective is the parameter that is being optimized.
-In joint-energy training we optimize the objective with respect to the network weights $\theta$:
+In joint-energy training we optimize the objective with respect to the network weights $\theta$.
+Recall that $\mathcal{E}_{\theta}(\mathbf{x}|\mathbf{y})=\mu_{\theta}(\mathbf{x})[\mathbf{y}]$.
+Then the partial gradient with respect to the generative loss component can be expressed as follows:
 
@@
   \begin{aligned}
-    \nabla_{\theta} L_{\text{JEM}}(\theta) &=\nabla_{\theta} L_{\text{clf}}(\theta) + \nabla_{\theta}L_{\text{gen}}(\theta) + \lambda \nabla_{\theta} L_{\text{reg}}(\theta) 
+    \nabla_{\theta}L_{\text{gen}}(\theta) &= \nabla_{\theta}\mu_{\theta}(\mathbf{x})[\mathbf{y}]- \nabla_{\theta}\mu_{\theta}(\hat{\mathbf{x}}_{J})[\mathbf{y}]
   \end{aligned}
@@
 
-we take these parameters as fixed in our counterfactual search.
+During the counterfactual search, we take the network parameters as fixed and instead optimize with respect to the counterfactual itself,
+
+\begin{equation}\label{eq:ce-grad}
+  \begin{aligned}
+    \nabla_{\mathbf{x}}L_{\text{gen}}(\theta) &= \nabla_{\mathbf{x}}\mu_{\theta}(\mathbf{x})[\mathbf{y}]- \nabla_{\mathbf{x}}\mu_{\theta}(\hat{\mathbf{x}}_{J})[\mathbf{y}]
+  \end{aligned}
+\end{equation}
+
+where we omit the notion of a latent search space to make the comparison easier.
+Intuitively, taking iterative gradient steps according to Equation~\ref{eq:ce-grad} has the effect of decreasing the energy of the counterfactual until it is in balance with the energy of conditional samples generated through SGLD.
 

### 2023-09-20 paper/body.tex
shoot me
more work on the appendix

--- 2023-09-19 paper/body.tex
@@
   \begin{aligned}
-    \mathbf{x}_{j+1} &\leftarrow \mathbf{x}_j - \frac{\epsilon_j^2}{2} \mathcal{E}(\mathbf{x}_j|\mathbf{y}^+) + \epsilon_j \mathbf{r}_j, && j=1,...,J
+    \mathbf{x}_{j+1} &\leftarrow \mathbf{x}_j - \frac{\epsilon_j^2}{2} \mathcal{E}_{\theta}(\mathbf{x}_j|\mathbf{y}^+) + \epsilon_j \mathbf{r}_j, && j=1,...,J
   \end{aligned}
@@
 where $\mathbf{r}_j \sim \mathcal{N}(\mathbf{0},\mathbf{I})$ is the stochastic term and the step-size $\epsilon_j$ is typically polynomially decayed~\citep{welling2011bayesian}.
-The term $\mathcal{E}(\mathbf{x}_j|\mathbf{y}^+)$ denotes the model energy conditioned on the target class label $\mathbf{y}^+$ which we specify as the negative logit corresponding to the target class label $\mathbf{y}^{+}$.
+The term $\mathcal{E}_{\theta}(\mathbf{x}_j|\mathbf{y}^+)$ denotes the model energy conditioned on the target class label $\mathbf{y}^+$ which we specify as the negative logit corresponding to the target class label $\mathbf{y}^{+}$.
 To allow for faster sampling, we follow the common practice of choosing the step-size $\epsilon_j$ and the standard deviation of $\mathbf{r}_j$ separately.
@@
 
-Our default choice for the $\text{dist}(\cdot)$ in both cases is the L1 Norm.
+Our default choice for the $\text{dist}(\cdot)$ function in both cases is the Euclidean Norm.
 
@@
   \begin{aligned}
-    \Delta\mathcal{E}&=\mathcal{E}(f(\mathbf{Z}^\prime)|\mathbf{y}^+)-\mathcal{E}(x|\mathbf{y}^+) &&& x \sim \widehat{\mathbf{X}}_{\theta,\mathbf{y}^+}
+    \Delta\mathcal{E}_{\theta}&=\mathcal{E}_{\theta}(f(\mathbf{Z}^\prime)|\mathbf{y}^+)-\mathcal{E}_{\theta}(x|\mathbf{y}^+) &&& x \sim \widehat{\mathbf{X}}_{\theta,\mathbf{y}^+}
   \end{aligned}

### 2023-09-21 paper/appendix.tex
more work on appendix

@@
   \begin{aligned}
-    \log p\_{\theta}(\mathbf{x},\mathbf{y}) &= \log p_{\theta}(\mathbf{y}|\mathbf{x}) + \log p_{\theta}(\mathbf{x})
+    \log p_{\theta}(\mathbf{x},\mathbf{y}) &= \log p_{\theta}(\mathbf{y}|\mathbf{x}) + \log p_{\theta}(\mathbf{x})
   \end{aligned}
@@
   \begin{aligned}
-    \nabla_{\theta}L_{\text{gen}}(\theta)&=-\nabla_{\theta}\log p\_{\theta}(\mathbf{x})=-\left(\mathbb{E}_{p(\mathbf{x})} \left\{  \nabla_{\theta} \mathcal{E}_{\theta}(\mathbf{x}) \right\} - \mathbb{E}_{p_{\theta}(\mathbf{x})} \left\{  \nabla_{\theta} \mathcal{E}_{\theta}(\mathbf{x}) \right\} \right)
+    \nabla_{\theta}L_{\text{gen}}(\theta)&=-\nabla_{\theta}\log p_{\theta}(\mathbf{x})=-\left(\mathbb{E}_{p(\mathbf{x})} \left\{  \nabla_{\theta} \mathcal{E}_{\theta}(\mathbf{x}) \right\} - \mathbb{E}_{p_{\theta}(\mathbf{x})} \left\{  \nabla_{\theta} \mathcal{E}_{\theta}(\mathbf{x}) \right\} \right)
   \end{aligned}
@@
   \mathbf{Z}^\prime &= \arg \min_{\mathbf{Z}^\prime \in \mathcal{Z}^L} \{  {\text{yloss}(M_{\theta}(f(\mathbf{Z}^\prime)),\mathbf{y}^+)}+ \lambda_{1} {\text{dist}(f(\mathbf{Z}^\prime),\mathbf{x}) } \\
-  &+ \lambda_2 \Delta\mathcal{E}_{\theta}(\mathbf{Z}^\prime,\widehat{\mathbf{X}}_{\theta,\mathbf{y}^+}) + \lambda_3 \Omega(C_{\theta}(f(\mathbf{Z}^\prime);\alpha)) \} 
+  &+ \lambda_2 \mathcal{E}_{\theta}(\mathbf{Z}^\prime,\widehat{\mathbf{X}}_{\theta,\mathbf{y}^+}) + \lambda_3 \Omega(C_{\theta}(f(\mathbf{Z}^\prime);\alpha)) \} 
   \end{aligned} 
@@
 
-We can make the connection to energy-based modeling more explicit by restating this equation in terms $L_{\text{JEM}}(\theta)$, which we defined in Equation~\ref{eq:jem-loss}.
-In particular, note that for $\lambda_2=1$ and $\lambda L_{\text{reg}}(\theta)=0$ we have
+We can make the connection to energy-based modeling more explicit by restating the counterfactual search objective in terms $L_{\text{JEM}}(\theta)$, which we defined in Equation~\ref{eq:jem-loss}.
+In particular, consider the following counterfactual search objective,
 
@@
 
-since $\Delta\mathcal{E}_{\theta}(\cdot)$ is equivalent to the generative loss function $L_{\text{gen}}(\cdot)$.
-In fact, this is also true for $\lambda L_{\text{reg}}(\theta)\ne0$ since we use the Ridge penalty $L_{\text{reg}}(\theta)$ in the counterfactual search just like we do in joint-energy training.
-This detail was omitted from the body of the paper for the sake of simplicity.
+where we have simply used the JEM loss function as $\text{yloss}(M_{\theta}(f(\mathbf{Z}^\prime)),\mathbf{y}^+)$.
 
-Aside from the additional penalties in Equation~\ref{eq:eccco-app}, the only key difference between our counterfactual search objective and the joint-energy training objective is the parameter that is being optimized.
+Now note that aside from the additional penalties in Equation~\ref{eq:eccco-app}, the only key difference between our counterfactual search objective and the joint-energy training objective is the parameter that is being optimized.
 In joint-energy training we optimize the objective with respect to the network weights $\theta$.
 Recall that $\mathcal{E}_{\theta}(\mathbf{x}|\mathbf{y})=\mu_{\theta}(\mathbf{x})[\mathbf{y}]$.
-Then the partial gradient with respect to the generative loss component can be expressed as follows:
+Then the partial gradient with respect to the generative loss component of $L_{\text{JEM}}(\theta)$ can be expressed as follows:
 
@@
 
-During the counterfactual search, we take the network parameters as fixed and instead optimize with respect to the counterfactual itself,
+During the counterfactual search, we take the network parameters as fixed and instead optimize with respect to the counterfactual itself\footnote{Here we omit the notion of a latent search space to make the comparison easier.},
 
@@
   \begin{aligned}
-    \nabla_{\mathbf{x}}L_{\text{gen}}(\theta) &= \nabla_{\mathbf{x}}\mu_{\theta}(\mathbf{x})[\mathbf{y}]- \nabla_{\mathbf{x}}\mu_{\theta}(\hat{\mathbf{x}}_{J})[\mathbf{y}]
+    \nabla_{\mathbf{x}}L_{\text{gen}}(\theta) &= \nabla_{\mathbf{x}}\mu_{\theta}(\mathbf{x})[\mathbf{y}^+]- \nabla_{\mathbf{x}}\mu_{\theta}(\hat{\mathbf{x}}_{J})[\mathbf{y}^+]=\nabla_{\mathbf{x}}\mu_{\theta}(\mathbf{x})[\mathbf{y}^+]=\nabla_{\mathbf{x}}\mathcal{E}_{\theta}(\mathbf{x}|\mathbf{y}^+)
   \end{aligned}
@@
 
-where we omit the notion of a latent search space to make the comparison easier.
-Intuitively, taking iterative gradient steps according to Equation~\ref{eq:ce-grad} has the effect of decreasing the energy of the counterfactual until it is in balance with the energy of conditional samples generated through SGLD.
+where the second term is equal to zero because $\mu_{\theta}(\hat{\mathbf{x}}_{J})[\mathbf{y}]$ is invariant with respect to $\mathbf{x}$.
+Since this term has zero gradients, we can remove it from the loss function altogether.
+For the regularization loss component of $L_{\text{JEM}}(\theta)$ we can proceed analogously such that we can rewrite Equation~\ref{eq:eccco-jem} as follows:
+
+\begin{equation} \label{eq:eccco-jem-2}
+  \begin{aligned}
+  \mathbf{Z}^\prime =& \arg \min_{\mathbf{Z}^\prime \in \mathcal{Z}^L} \{  {\text{yloss}(M_{\theta}(f(\mathbf{Z}^\prime)),\mathbf{y}^+) + \mathcal{E}_{\theta}(f(\mathbf{Z}^\prime)|\mathbf{y}^+) + || \mathcal{E}_{\theta}(f(\mathbf{Z}^\prime)|\mathbf{y}^+) ||_2^2} \\ &+ \lambda_{1} {\text{dist}(f(\mathbf{Z}^\prime),\mathbf{x}) }  + \lambda_3 \Omega(C_{\theta}(f(\mathbf{Z}^\prime);\alpha)) \}
+  \end{aligned} 
+\end{equation}
+
+Now we notice that Equation~\ref{eq:eccco-jem-2} is equivalent to Equation~\ref{eq:eccco-app} for $\lambda_2=1$.
+For the sake of simplicity, we omitted the regularization component from Equation~\ref{eq:eccco} in the main text.
+Intuitively, taking iterative gradient steps according to Equation~\ref{eq:ce-grad} has the effect of constraining the energy of the counterfactual until.
+The generative property of the underlying model implicitly enters this equation through $\theta$.
 

### 2023-09-22 paper/body.tex
more on author response

@@
   \begin{aligned}
-  \mathbf{Z}^\prime &= \arg \min_{\mathbf{Z}^\prime \in \mathcal{Z}^L} \{  {\text{yloss}(M_{\theta}(f(\mathbf{Z}^\prime)),\mathbf{y}^+)}+ \lambda_{1} {\text{dist}(f(\mathbf{Z}^\prime),\mathbf{x}) } \\
-  &+ \lambda_2 \Delta\mathcal{E}_{\theta}(\mathbf{Z}^\prime,\widehat{\mathbf{X}}_{\theta,\mathbf{y}^+}) + \lambda_3 \Omega(C_{\theta}(f(\mathbf{Z}^\prime);\alpha)) \} 
+  \mathbf{Z}^\prime =& \arg \min_{\mathbf{Z}^\prime \in \mathcal{Z}^L} \{  {\text{yloss}(M_{\theta}(f(\mathbf{Z}^\prime)),\mathbf{y}^+)}+ \lambda_{1} {\text{dist}(f(\mathbf{Z}^\prime),\mathbf{x}) } \\
+  &+ \lambda_2 \mathcal{E}_{\theta}(f(\mathbf{Z}^\prime)) + \lambda_3 \Omega(C_{\theta}(f(\mathbf{Z}^\prime);\alpha)) \} 
   \end{aligned} 
@@
 Our default choice for $\text{dist}(\cdot)$ is the L1 Norm due to its sparsity-inducing properties.
-The second penalty term involving $\lambda_2$ induces faithfulness by constraining the energy of the generated counterfactual where we have:
-
-\begin{equation} \label{eq:energy-delta}
-  \begin{aligned}
-    \Delta\mathcal{E}_{\theta}&=\mathcal{E}_{\theta}(f(\mathbf{Z}^\prime)|\mathbf{y}^+)-\mathcal{E}_{\theta}(x|\mathbf{y}^+) &&& x \sim \widehat{\mathbf{X}}_{\theta,\mathbf{y}^+}
-  \end{aligned}
-\end{equation}
-
-In particular, this penalty ensures that the energy of the generated counterfactual is in balance with the energy of the generated conditional samples ($\widehat{\mathbf{X}}_{\theta,\mathbf{y}^+}$).
+The second penalty term involving $\lambda_2$ induces faithfulness by constraining the energy of the generated counterfactual.
 The third and final penalty term involving $\lambda_3$ ensures that the generated counterfactual is associated with low predictive uncertainty.

### 2023-09-22 paper/appendix.tex
updates to methodoloy section

@@
 
-\subsubsection{More detail on our generator}
+\subsubsection{Deriving the search objective}
 
@@
 The generative property of the underlying model implicitly enters this equation through $\theta$.
+
+\subsubsection{The \textit{ECCCo} algorithm}
+
+Algorithm~\ref{alg:eccco} describes how exactly \textit{ECCCo} works.
+For the sake of simplicity and without loss of generality, we limit our attention to generating a single counterfactual $\mathbf{x}^\prime=f(\mathbf{z}^\prime)$.
+The counterfactual state $\mathbf{z}^\prime$ is initialized by passing the factual $\mathbf{x}$ through a simple feature transformer $f^{-1}$.
+Next, we generate $n_{\mathcal{B}}$ conditional samples $\hat{\mathbf{x}}_{\theta,\mathbf{y}^+}$ using SGLD (Equation~\ref{eq:sgld}) and store the $n_E$ instances with the lowest energy.
+We then calibrate the model $M_{\theta}$ through split conformal prediction.
+Finally, we search counterfactuals through gradient descent where $\mathcal{L}(\mathbf{z}^\prime,\mathbf{y}^+,\widehat{\mathbf{X}}_{\theta,\mathbf{y}^+}; \Lambda, \alpha)$ denotes our loss function defined in Equation~\ref{eq:eccco}.
+The search terminates once the convergence criterium is met or the maximum number of iterations $T$ has been exhausted.
+Note that the choice of convergence criterium has important implications on the final counterfactual which we explain in Appendix~\ref{app:eccco}.
+
+\begin{algorithm*}
+  \caption{The \textit{ECCCo} generator}\label{alg:eccco}
+  \begin{algorithmic}[1]
+    \Require $\mathbf{x}, \mathbf{y}^+, M_{\theta}, f, \Lambda=[\lambda_1,\lambda_2,\lambda_3], \alpha, \mathcal{D}, T, \eta, n_{\mathcal{B}}, n_E$ where $M_{\theta}(\mathbf{x})\neq\mathbf{y}^+$
+    \Ensure $\mathbf{x}^\prime$
+    \State Initialize $\mathbf{z}^\prime \gets f^{-1}(\mathbf{x})$ \Comment{Map to counterfactual state space.}
+    \State Generate $\left\{\hat{\mathbf{x}}_{\theta,\mathbf{y}^+}\right\}_{n_{\mathcal{B}}} \gets p_{\theta}(\mathbf{x}_{\mathbf{y}^+})$ \Comment{Generate $n_{\mathcal{B}}$ samples using SGLD (Equation~\ref{eq:sgld}).}
+    \State Store $\widehat{\mathbf{X}}_{\theta,\mathbf{y}^+} \gets \left\{\hat{\mathbf{x}}_{\theta,\mathbf{y}^+}\right\}_{n_{\mathcal{B}}}$ \Comment{Choose $n_E$ lowest-energy samples.}
+    \State Run \textit{SCP} for $M_{\theta}$ using $\mathcal{D}$ \Comment{Calibrate model through split conformal prediction.}
+    \State Initialize $t \gets 0$
+    \While{\textit{not converged} or $t < T$} \Comment{For convergence conditions see Appendix~\ref{app:eccco}.}
+    \State $\mathbf{z}^\prime \gets \mathbf{z}^\prime - \eta \nabla_{\mathbf{z}^\prime} \mathcal{L}(\mathbf{z}^\prime,\mathbf{y}^+,\widehat{\mathbf{X}}_{\theta,\mathbf{y}^+}; \Lambda, \alpha)$ \Comment{Take gradient step of size $\eta$.}
+    \State $t \gets t+1$
+    \EndWhile
+    \State $\mathbf{x}^\prime \gets f(\mathbf{z}^\prime)$ \Comment{Map back to feature space.}
+  \end{algorithmic}
+\end{algorithm*}
 

### 2023-09-22 paper/body.tex
updates to methodoloy section

@@
 
+The objective defined in Equation~\ref{eq:eccco} is closely related to the training objective of energy-based models, which involves a generative loss component that measures the difference in energies of observed and generated samples.
+The key observation in our context is that we can rely solely on the energy of the counterfactual itself.
+By simply constraining $\mathcal{E}_{\theta}(f(\mathbf{Z}^\prime))$ we can induce counterfactuals that faithfully represent the generative property of the underlying model, which is implicitly captured by its parameters $\theta$.
+Importantly, this means that we do not need to generate conditional samples through SGLD at all.
+For a detailed discussion of this please see Appendix~\ref{app:eccco}.
+
 \begin{figure}
@@
 \end{figure}
-
-\begin{algorithm*}
-  \caption{The \textit{ECCCo} generator}\label{alg:eccco}
-  \begin{algorithmic}[1]
-    \Require $\mathbf{x}, \mathbf{y}^+, M_{\theta}, f, \Lambda=[\lambda_1,\lambda_2,\lambda_3], \alpha, \mathcal{D}, T, \eta, n_{\mathcal{B}}, n_E$ where $M_{\theta}(\mathbf{x})\neq\mathbf{y}^+$
-    \Ensure $\mathbf{x}^\prime$
-    \State Initialize $\mathbf{z}^\prime \gets f^{-1}(\mathbf{x})$ \Comment{Map to counterfactual state space.}
-    \State Generate $\left\{\hat{\mathbf{x}}_{\theta,\mathbf{y}^+}\right\}_{n_{\mathcal{B}}} \gets p_{\theta}(\mathbf{x}_{\mathbf{y}^+})$ \Comment{Generate $n_{\mathcal{B}}$ samples using SGLD (Equation~\ref{eq:sgld}).}
-    \State Store $\widehat{\mathbf{X}}_{\theta,\mathbf{y}^+} \gets \left\{\hat{\mathbf{x}}_{\theta,\mathbf{y}^+}\right\}_{n_{\mathcal{B}}}$ \Comment{Choose $n_E$ lowest-energy samples.}
-    \State Run \textit{SCP} for $M_{\theta}$ using $\mathcal{D}$ \Comment{Calibrate model through split conformal prediction.}
-    \State Initialize $t \gets 0$
-    \While{\textit{not converged} or $t < T$} \Comment{For convergence conditions see Appendix~\ref{app:eccco}.}
-    \State $\mathbf{z}^\prime \gets \mathbf{z}^\prime - \eta \nabla_{\mathbf{z}^\prime} \mathcal{L}(\mathbf{z}^\prime,\mathbf{y}^+,\widehat{\mathbf{X}}_{\theta,\mathbf{y}^+}; \Lambda, \alpha)$ \Comment{Take gradient step of size $\eta$.}
-    \State $t \gets t+1$
-    \EndWhile
-    \State $\mathbf{x}^\prime \gets f(\mathbf{z}^\prime)$ \Comment{Map back to feature space.}
-  \end{algorithmic}
-\end{algorithm*}
 
@@
 Finally, the outcome for \textit{ECCCo} looks similar, but the additional smooth set size penalty leads to somewhat faster convergence.
-
-Algorithm~\ref{alg:eccco} describes how exactly \textit{ECCCo} works.
-For the sake of simplicity and without loss of generality, we limit our attention to generating a single counterfactual $\mathbf{x}^\prime=f(\mathbf{z}^\prime)$.
-The counterfactual state $\mathbf{z}^\prime$ is initialized by passing the factual $\mathbf{x}$ through a simple feature transformer $f^{-1}$.
-Next, we generate $n_{\mathcal{B}}$ conditional samples $\hat{\mathbf{x}}_{\theta,\mathbf{y}^+}$ using SGLD (Equation~\ref{eq:sgld}) and store the $n_E$ instances with the lowest energy.
-We then calibrate the model $M_{\theta}$ through split conformal prediction.
-Finally, we search counterfactuals through gradient descent where $\mathcal{L}(\mathbf{z}^\prime,\mathbf{y}^+,\widehat{\mathbf{X}}_{\theta,\mathbf{y}^+}; \Lambda, \alpha)$ denotes our loss function defined in Equation~\ref{eq:eccco}.
-The search terminates once the convergence criterium is met or the maximum number of iterations $T$ has been exhausted.
-Note that the choice of convergence criterium has important implications on the final counterfactual which we explain in Appendix~\ref{app:eccco}.
 

### 2023-09-23 paper/body.tex
methodology rewritten

@@
 
-We begin by stating our proposed objective function, which involves tailored loss and penalty functions that we will explain in the following.
-In particular, we extend Equation~\ref{eq:general} as follows:
+We begin by substituting the loss function in Equation~\ref{eq:general},
+
+\begin{equation} \label{eq:eccco-start}
+  \begin{aligned}
+  \mathbf{Z}^\prime =& \arg \min_{\mathbf{Z}^\prime \in \mathcal{Z}^L} \{  {L_{\text{JEM}}(f(\mathbf{Z}^\prime);M_{\theta},\mathbf{y}^+)}+ \lambda {\text{cost}(f(\mathbf{Z}^\prime)) } \} 
+  \end{aligned} 
+\end{equation}
+
+where $L_{\text{JEM}}(f(\mathbf{Z}^\prime);M_{\theta},\mathbf{y}^+)$ is a hybrid loss function used in joint-energy modelling evaluated at a given counterfactual state for a given model and target outcome:
+
+\begin{equation}
+  \begin{aligned}
+    L_{\text{JEM}}(f(\mathbf{Z}^\prime); \cdot) = L_{\text{clf}}(f(\mathbf{Z}^\prime); \cdot) + L_{\text{gen}}(f(\mathbf{Z}^\prime); \cdot)
+  \end{aligned}
+\end{equation}
+
+The first term, $L_{\text{clf}}$, is any standard classification loss function such as cross-entropy loss.
+The second term, $L_{\text{gen}}$, is used to measure loss with respect to the generative task\footnote{In practice, regularization loss is typically also added.
+We follow this convention but have omitted the term here for simplicity.}.
+In the context of joint-energy training, $L_{\text{gen}}$ induces changes in model parameters $\theta$ that decrease the energy of observed samples and increase the energy of samples generated through SGLD~\citep{du2019implicit}.
+
+The key observation in our context is that we can rely solely on decreasing the energy of the counterfactual itself.
+This is sufficient to capture the generative property of the underlying model since it is implicitly captured by its parameters $\theta$.
+Importantly, this means that we do not need to generate conditional samples through SGLD during our counterfactual search at all as we explain in Appendix~\ref{app:eccco}.
+
+This observation leads to the following simple objective function for \textit{ECCCo}:
 
@@
   \begin{aligned}
-  \mathbf{Z}^\prime =& \arg \min_{\mathbf{Z}^\prime \in \mathcal{Z}^L} \{  {\text{yloss}(M_{\theta}(f(\mathbf{Z}^\prime)),\mathbf{y}^+)}+ \lambda_{1} {\text{dist}(f(\mathbf{Z}^\prime),\mathbf{x}) } \\
-  &+ \lambda_2 \mathcal{E}_{\theta}(f(\mathbf{Z}^\prime)) + \lambda_3 \Omega(C_{\theta}(f(\mathbf{Z}^\prime);\alpha)) \} 
+  \mathbf{Z}^\prime =& \arg \min_{\mathbf{Z}^\prime \in \mathcal{Z}^L} \{  {L_{\text{clf}}(f(\mathbf{Z}^\prime);M_{\theta},\mathbf{y}^+)}+ \lambda_1 {\text{cost}(f(\mathbf{Z}^\prime)) } \\
+  &+ \lambda_2 \mathcal{E}_{\theta}(f(\mathbf{Z}^\prime)|\mathbf{y}^+) + \lambda_3 \Omega(C_{\theta}(f(\mathbf{Z}^\prime);\alpha)) \} 
   \end{aligned} 
@@
 
-The first penalty term involving $\lambda_1$ induces proximity like in~\citet{wachter2017counterfactual}.
-Our default choice for $\text{dist}(\cdot)$ is the L1 Norm due to its sparsity-inducing properties.
+The first penalty term involving $\lambda_1$ induces closeness like in~\citet{wachter2017counterfactual}.
 The second penalty term involving $\lambda_2$ induces faithfulness by constraining the energy of the generated counterfactual.
 The third and final penalty term involving $\lambda_3$ ensures that the generated counterfactual is associated with low predictive uncertainty.
-
-The objective defined in Equation~\ref{eq:eccco} is closely related to the training objective of energy-based models, which involves a generative loss component that measures the difference in energies of observed and generated samples.
-The key observation in our context is that we can rely solely on the energy of the counterfactual itself.
-By simply constraining $\mathcal{E}_{\theta}(f(\mathbf{Z}^\prime))$ we can induce counterfactuals that faithfully represent the generative property of the underlying model, which is implicitly captured by its parameters $\theta$.
-Importantly, this means that we do not need to generate conditional samples through SGLD at all.
-For a detailed discussion of this please see Appendix~\ref{app:eccco}.
 

### 2023-09-24 paper/tables.Rmd
saving the best

--- 2023-05-17 notebooks/tables.Rmd
@@
 ```{r}
-files <- list.files("artifacts/results/")
+res_path <- "results/"
+files <- list.files(res_path)
 dt <- lapply(files[grepl("_benchmark.csv", files)], function(x) {
-    fread(file.path("artifacts/results/", x))
+    fread(file.path(res_path, x))
 })

### 2023-09-24 experiments/california_housing.jl

@@ -36,16 +36,26 @@ params = (
 # Best grid search params:
 append_best_params!(params, dataname)
 
-if !GRID_SEARCH
-    run_experiment(
-        counterfactual_data, test_data;
-        dataname=dataname,
-        params...
-    )
-else
+if GRID_SEARCH
     grid_search(
         counterfactual_data, test_data;
         dataname=dataname,
-        tuning_params=tuning_params
+        tuning_params=tuning_params,
+        params...
+    )
+elseif FROM_GRID_SEARCH
+    outcomes_file_path = joinpath(
+        DEFAULT_OUTPUT_PATH,
+        "grid_search",
+        "$(replace(lowercase(dataname), " " => "_")).jls",
+    )
+    save_best(outcomes_file_path)
+    bmk2csv(dataname)
+else
+    run_experiment(
+        counterfactual_data, test_data;
+        dataname=dataname,
+        model_tuning_params=model_tuning_params,
+        params...
     )
 end

### 2023-09-24 experiments/gmsc.jl

@@ -37,16 +37,26 @@ params = (
 # Best grid search params:
 append_best_params!(params, dataname)
 
-if !GRID_SEARCH
-    run_experiment(
-        counterfactual_data, test_data;
-        dataname=dataname,
-        params...
-    )
-else
+if GRID_SEARCH
     grid_search(
         counterfactual_data, test_data;
         dataname=dataname,
-        tuning_params=tuning_params
+        tuning_params=tuning_params,
+        params...
+    )
+elseif FROM_GRID_SEARCH
+    outcomes_file_path = joinpath(
+        DEFAULT_OUTPUT_PATH,
+        "grid_search",
+        "$(replace(lowercase(dataname), " " => "_")).jls",
+    )
+    save_best(outcomes_file_path)
+    bmk2csv(dataname)
+else
+    run_experiment(
+        counterfactual_data, test_data;
+        dataname=dataname,
+        model_tuning_params=model_tuning_params,
+        params...
     )
 end

### 2023-09-24 experiments/linearly_separable.jl

@@ -39,6 +39,7 @@ elseif FROM_GRID_SEARCH
         "$(replace(lowercase(dataname), " " => "_")).jls",
     )
     save_best(outcomes_file_path)
+    bmk2csv(dataname)
 else
     run_experiment(
         counterfactual_data, test_data;

### 2023-09-24 experiments/moons.jl

@@ -24,17 +24,26 @@ params = (
 # Best grid search params:
 append_best_params!(params, dataname)
 
-if !GRID_SEARCH
-    run_experiment(
-        counterfactual_data, test_data;
-        dataname=dataname,
-        params...
-    )
-else
+if GRID_SEARCH
     grid_search(
         counterfactual_data, test_data;
         dataname=dataname,
         tuning_params=tuning_params,
+        params...
+    )
+elseif FROM_GRID_SEARCH
+    outcomes_file_path = joinpath(
+        DEFAULT_OUTPUT_PATH,
+        "grid_search",
+        "$(replace(lowercase(dataname), " " => "_")).jls",
+    )
+    save_best(outcomes_file_path)
+    bmk2csv(dataname)
+else
+    run_experiment(
+        counterfactual_data, test_data;
+        dataname=dataname,
+        model_tuning_params=model_tuning_params,
         params...
     )
 end

### 2023-09-24 experiments/run_experiments.jl

@@ -1,14 +1,15 @@
 include("setup_env.jl");
 
 # User inputs:
+all_data_sets = ["linearly_separable", "moons", "circles", "mnist", "fmnist", "gmsc", "german_credit", "california_housing"]
 if "run-all" in ARGS
-    datanames = ["linearly_separable", "moons", "circles", "mnist", "fmnist", "gmsc"]
+    datanames = all_data_sets
 elseif any(contains.(ARGS, "data="))
     datanames = [ARGS[findall(contains.(ARGS, "data="))][1] |> x -> replace(x, "data=" => "")]
     datanames = replace.(split(datanames[1], ","), " " => "")
 else
     @warn "No dataset specified, defaulting to all."
-    datanames = ["linearly_separable", "moons", "circles", "mnist", "fmnist", "gmsc"]
+    datanames = all_data_sets
 end
 
 # Linearly Separable

### 2023-09-24 experiments/save_best.jl

@@ -1,6 +1,9 @@
 function save_best(outcomes_file_path::String)
     # Just load the best model from the grid search:
-    @assert isfile(outcomes_file_path)
+    if !isfile(outcomes_file_path)
+        @info "No grid search file found at $(outcomes_file_path)."
+        return
+    end
     outcomes = Serialization.deserialize(outcomes_file_path)
     outcome = best_absolute_outcome_eccco_Δ(outcomes).outcome
     exper = outcome.exper
@@ -12,4 +15,17 @@ function save_best(outcomes_file_path::String)
     Serialization.serialize(joinpath(output_path, "$(exper.save_name)_bmk.jls"), outcome.bmk)
     Serialization.serialize(joinpath(output_path, "$(exper.save_name)_models.jls"), outcome.model_dict)
     meta(outcome; save_output=true, params_path=params_path)
+end
+
+function bmk2csv(dataname::String)
+    bmk_path = joinpath(
+        DEFAULT_OUTPUT_PATH,
+        "$(replace(lowercase(dataname), " " => "_"))_bmk.jls",
+    )
+    bmk = Serialization.deserialize(bmk_path)
+    csv_path = joinpath(
+        DEFAULT_OUTPUT_PATH,
+        "$(replace(lowercase(dataname), " " => "_"))_bmk.csv",
+    )
+    CSV.write(csv_path, bmk()[:,Not(:ce)])
 end

### 2023-09-24 experiments/save_best.jl
uh

@@ -27,5 +27,10 @@ function bmk2csv(dataname::String)
         DEFAULT_OUTPUT_PATH,
         "$(replace(lowercase(dataname), " " => "_"))_bmk.csv",
     )
-    CSV.write(csv_path, bmk()[:,Not(:ce)])
+    bmk = bmk()
+    if "ce" ∈ names(bmk)
+        CSV.write(csv_path, bmk[:,Not(:ce)])
+    else
+        CSV.write(csv_path, bmk)
+    end
 end

### 2023-09-24 paper/tables.Rmd

@@
 files <- list.files(res_path)
-dt <- lapply(files[grepl("_benchmark.csv", files)], function(x) {
+dt <- lapply(files[grepl("_bmk.csv", files)], function(x) {
     fread(file.path(res_path, x))
@@
 synth <- c("Moons", "Circles", "Linearly Separable")
-dt[,source:=ifelse(dataname %in% synth, "synthetic", "real-world")]
-reported_data <- c(
-  "MNIST",
-  "GMSC",
-  "Linearly Separable",
-  "Moons",
-  "Circles"
-)
-dt <- dt[dataname %in% reported_data]
+tabular <- c("GMSC", "German Credit", "California Housing")
+dt[,source:=ifelse(dataname %in% synth, "synthetic", ifelse(dataname %in% tabular, "tabular", "vision"))]
 dt[,non_valid:=variable=="validity" & value==0.0,.(sample,dataname,generator,model,target,factual,source)]

### 2023-09-24 notebooks/fmnist.qmd

@@
 _loss = Flux.Losses.logitcrossentropy       # loss function
-_finaliser = x -> x                         # finaliser function
+_finaliser = x -> x                         # finalizer function
 ```

### 2023-09-24 notebooks/gmsc.qmd

@@
         transf = MLJ.UnivariateBoxCoxTransformer()
-        x = exp.(x)
+        x = exper.(x)
         mach = machine(transf, x)

### 2023-09-24 notebooks/poc.qmd

@@
     batch_size=batch_size,
-    finaliser=Flux.softmax,
+    finalizer=Flux.softmax,
     loss=Flux.Losses.crossentropy,

### 2023-09-24 paper/tables.Rmd

@@
 dt[,source:=ifelse(dataname %in% synth, "synthetic", ifelse(dataname %in% tabular, "tabular", "vision"))]
+```
+
+```{r}
+# Generator names
+dt[,generator:=factor(generator)]
+levels(dt$generator)[match("ECCCo",levels(dt$generator))] <- "ECCCo-L1"
+levels(dt$generator)[match("ECCCo (no CP)",levels(dt$generator))] <- "ECCCo-L1 (no CP)"
+levels(dt$generator)[match("ECCCo (no EBM)",levels(dt$generator))] <- "ECCCo-L1 (no EBM)"
+levels(dt$generator)[match("ECCCo-Δ",levels(dt$generator))] <- "ECCCo"
+levels(dt$generator)[match("ECCCo-Δ (latent)",levels(dt$generator))] <- "ECCCo+"
+levels(dt$generator)[match("ECCCo-Δ (no CP)",levels(dt$generator))] <- "ECCCo (no CP)"
+levels(dt$generator)[match("ECCCo-Δ (no EBM)",levels(dt$generator))] <- "ECCCo (no EBM)"
+```
+
+```{r}
+# Adjust measure names
+dt[source=="vision" & variable=="distance_from_targets_ssim", variable:="implausibility"]
+dt[source=="vision" & variable=="distance_from_energy_ssim", variable:="unfaithfulness"]
+dt[source!="vision" & variable=="distance_from_targets_l2", variable:="implausibility"]
+dt[source!="vision" & variable=="distance_from_energy_l2", variable:="unfaithfulness"]
+```
+
+```{r}
 dt[,non_valid:=variable=="validity" & value==0.0,.(sample,dataname,generator,model,target,factual,source)]
@@
   "distance",
+  "implausibility",
+  "unfaithfulness",
   "distance_from_energy", 
+  "distance_from_energy_l2", 
   "distance_from_targets",
+  "distance_from_targets_l2",
   "set_size_penalty"
@@
 # Choices:
-measures <- c("distance_from_energy", "distance_from_targets")
-measure_names <- c("Unfaithfulness ↓","Implausibility ↓")
-chosen_source <- "real-world"
+measures <- c(
+  "unfaithfulness",
+  "implausibility"
+)
+measure_names <- c(
+  "Unfaithfulness ↓",
+  "Implausibility ↓"
+)
+# Order:
 chosen_data <- c(
+  "Linearly Separable",
+  "GMSC",
   "MNIST",
-  "GMSC"
 )
+chosen_model <- c(
+  "MLP",
+  "JEM",
+  "LeNet-5"
+)
+
+tab_i <- tab
+
+# Logic:
+tab_i <- tab_i[variable %in% measures]
+tab_i <- tab_i[dataname %in% chosen_data]
+tab_i <- tab_i[model %in% chosen_model]
+tab_i[,dataname:=factor(dataname,levels=chosen_data)]
+tab_i <- dcast(tab_i, model + generator ~ dataname + variable)
+col_names <- c(
+  "Model",
+  "Generator",
+  rep(measure_names,length(chosen_data))
+)
+caption <-  "Results for datasets from different domains: sample averages +/- one standard deviation across counterfactuals.
+Best outcomes are highlighted in bold.
+Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (Wachter).
+\\label{tab:results-main} \\newline"
+file_name <- "paper/contents/table-main.tex"
+sub_header <- rep(length(measures), length(chosen_data))
+names(sub_header) <- chosen_data
+header <- c(
+  " " = 2, sub_header
+)
+line_sep <- c(rep("",length(measures)-1),"\\addlinespace")
+algin_cols <- c(rep('l',2),rep('c',ncol(tab_i)-2))
+kbl(
+  tab_i, caption = caption, 
+  align = algin_cols, col.names=col_names, booktabs = T, escape=F, 
+  format="latex", linesep = line_sep 
+) %>%
+  kable_styling(latex_options = c("scale_down")) %>%
+  kable_paper(full_width = T) %>%
+  add_header_above(header) %>%
+  collapse_rows(columns = 1:2, latex_hline = "major", valign = "middle") %>%
+  save_kable(file_name)
+```
+
+```{r}
+# Choices:
+measures <- c(
+  "distance_from_energy_l2",
+  "distance_from_targets_l2"
+)
+measure_names <- c(
+  "Unfaithfulness ↓",
+  "Implausibility ↓"
+)
+chosen_source <- "tabular"
+# Order:
+chosen_data <- c(
+  "GMSC",
+  "California Housing",
+  "German Credit"
+)
+chosen_model <- "MLP"
+
 tab_i <- tab
@@
 tab_i <- tab_i[dataname %in% chosen_data]
+tab_i <- tab_i[model == chosen_model]
 tab_i[,dataname:=factor(dataname,levels=chosen_data)]
@@
 # Choices:
-measures <- c("distance_from_energy", "distance_from_targets")
-measure_names <- c("Unfaithfulness ↓","Implausibility ↓")
+# Choices:
+measures <- c(
+  "distance_from_energy_l2",
+  "distance_from_targets_l2"
+)
+measure_names <- c(
+  "Unfaithfulness ↓",
+  "Implausibility ↓"
+)
 chosen_source <- "synthetic"
@@
   kable_styling(latex_options = c("scale_down")) %>%
-  kable_paper(full_width = F) %>%
+  kable_paper(full_width = T) %>%
+  add_header_above(header) %>%
+  collapse_rows(columns = 1:2, latex_hline = "major", valign = "middle") %>%
+  save_kable(file_name)
+```
+
+```{r}
+# Choices:
+measures <- c(
+  "distance_from_energy_l2",
+  "distance_from_targets_l2"
+)
+measure_names <- c(
+  "Unfaithfulness ↓",
+  "Implausibility ↓"
+)
+chosen_source <- "vision"
+# Order:
+chosen_data <- c(
+  "MNIST",
+  "Fashion MNIST",
+)
+chosen_model <- "LeNet-5"
+
+tab_i <- tab
+
+# Logic:
+tab_i <- tab_i[variable %in% measures]
+tab_i <- tab_i[source == chosen_source]
+tab_i <- tab_i[dataname %in% chosen_data]
+tab_i <- tab_i[model == chosen_model]
+tab_i[,dataname:=factor(dataname,levels=chosen_data)]
+tab_i <- dcast(tab_i, model + generator ~ dataname + variable)
+col_names <- c(
+  "Model",
+  "Generator",
+  rep(measure_names,length(chosen_data))
+)
+caption <- sprintf(
+  "Results for %s datasets: sample averages +/- one standard deviation across counterfactuals.
+Best outcomes are highlighted in bold.
+Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (Wachter).
+\\label{tab:results-%s} \\newline",
+  chosen_source,
+  chosen_source
+)
+file_name <- sprintf(
+  "paper/contents/table-%s.tex",
+  chosen_source
+)
+sub_header <- rep(length(measures), length(chosen_data))
+names(sub_header) <- chosen_data
+header <- c(
+  " " = 2, sub_header
+)
+line_sep <- c(rep("",length(measures)-1),"\\addlinespace")
+algin_cols <- c(rep('l',2),rep('c',ncol(tab_i)-2))
+kbl(
+  tab_i, caption = caption, 
+  align = algin_cols, col.names=col_names, booktabs = T, escape=F, 
+  format="latex", linesep = line_sep 
+) %>%
+  kable_styling(latex_options = c("scale_down")) %>%
+  kable_paper(full_width = T) %>%
   add_header_above(header) %>%

### 2023-09-24 paper/tables.Rmd
kill me

@@
 ```{r}
+generators <- unique(dt$generator)[sapply(unique(dt$generator), function(x) {!grepl("L1",x)})]
 tab <- dt[
+    generator %in% generators,
+    .(
+      value=sprintf("%1.2f ± %1.2f", mean(value), sd(value)),
+      val = mean(value),
+      std = sd(value)
+    ),
+    .(dataname, generator, model, variable, source)
+]
+
+tab$top_val = F
+tab$one_std_wachter = F
+tab$two_std_wachter = F
+
+# Measures to be minimized:
+min_measures <- c(
+  "distance",
+  "implausibility",
+  "unfaithfulness",
+  "distance_from_energy", 
+  "distance_from_energy_l2", 
+  "distance_from_targets",
+  "distance_from_targets_l2",
+  "set_size_penalty"
+)
+tab[variable %in% min_measures,top_val:=val==min(val),.(model, dataname, variable)]
+tab[variable %in% min_measures,top_val:=ifelse(rep(all(top_val),length(top_val)),F,top_val),.(model, dataname, variable)]
+tab[variable %in% min_measures,two_std_wachter:=val+2*std<val[generator=="Wachter"],.(model, dataname, variable)]
+tab[variable %in% min_measures,one_std_wachter:=val+1*std<val[generator=="Wachter"],.(model, dataname, variable)]
+
+# Measures to be maximized:
+max_measures <- c(
+  "validity",
+  "redundancy"
+)
+tab[variable %in% max_measures,top_val:=val==max(val),.(model, dataname, variable)]
+tab[variable %in% max_measures,top_val:=ifelse(rep(all(top_val),length(top_val)),F,top_val),.(model, dataname, variable)]
+tab[variable %in% max_measures,two_std_wachter:=val-2*std>val[generator=="Wachter"],.(model, dataname, variable)]
+tab[variable %in% max_measures,one_std_wachter:=val-1*std>val[generator=="Wachter"],.(model, dataname, variable)]
+
+# Add conditional formatting:
+tab$value <- cell_spec(tab$value, "latex", bold=tab$top_val)
+tab[one_std_wachter==T,value:=paste0(value,"*")]
+tab[one_std_wachter==F,value:=paste0(value,"\\hphantom{*}")]
+tab[two_std_wachter==T,value:=paste0(value,"*")]
+tab[two_std_wachter==F,value:=paste0(value,"\\hphantom{*}")]
+
+# Remove redundant columns:
+tab[,val:=NULL]
+tab[,std:=NULL]
+tab[,top_val:=NULL]
+tab[,two_std_wachter:=NULL]
+tab[,one_std_wachter:=NULL]
+```
+
+```{r}
+tab_full <- dt[
     ,
@@
 ```{r}
+generators <- unique(dt$generator)[sapply(unique(dt$generator), function(x) {!grepl("L1",x)})]
 tab_valid <- dt_valid[
-    ,
+    generator %in% generators,
     .(
@@
   "GMSC",
-  "MNIST",
+  "MNIST"
 )
@@
 tab_i <- tab_i[variable %in% measures]
+tab_i[,variable:=factor(variable, levels=measures)]
 tab_i <- tab_i[dataname %in% chosen_data]
 tab_i <- tab_i[model %in% chosen_model]
+tab_i[,model:=factor(model,levels=chosen_model)]
 tab_i[,dataname:=factor(dataname,levels=chosen_data)]
 tab_i <- dcast(tab_i, model + generator ~ dataname + variable)
+tab_i[is.na(tab_i)] <- ""
 col_names <- c(
@@
   kable_styling(latex_options = c("scale_down")) %>%
-  kable_paper(full_width = T) %>%
+  kable_paper(full_width = F) %>%
   add_header_above(header) %>%
@@
 measures <- c(
-  "distance_from_energy_l2",
-  "distance_from_targets_l2"
+  "unfaithfulness",
+  "implausibility",
+  "distance"
 )
@@
   "Unfaithfulness ↓",
-  "Implausibility ↓"
+  "Implausibility ↓",
+  "Cost ↓"
 )
@@
   "GMSC",
-  "California Housing",
-  "German Credit"
+  "California Housing"
 )
-chosen_model <- "MLP"
+chosen_model <- c(
+  "MLP",
+  "JEM"
+)
 
@@
 tab_i <- tab_i[variable %in% measures]
-tab_i <- tab_i[source == chosen_source]
+tab_i[,variable:=factor(variable, levels=measures)]
 tab_i <- tab_i[dataname %in% chosen_data]
-tab_i <- tab_i[model == chosen_model]
+tab_i <- tab_i[model %in% chosen_model]
+tab_i[,model:=factor(model,levels=chosen_model)]
 tab_i[,dataname:=factor(dataname,levels=chosen_data)]
@@
   kable_styling(latex_options = c("scale_down")) %>%
-  kable_paper(full_width = T) %>%
+  kable_paper(full_width = F) %>%
+  add_header_above(header) %>%
+  collapse_rows(columns = 1:2, latex_hline = "major", valign = "middle") %>%
+  save_kable(file_name)
+```
+
+```{r}
+# Choices:
+measures <- c(
+  "unfaithfulness",
+  "implausibility"
+)
+measure_names <- c(
+  "Unfaithfulness ↓",
+  "Implausibility ↓"
+)
+chosen_source <- "vision"
+# Order:
+chosen_data <- c(
+  "MNIST"
+)
+chosen_model <- c(
+  "MLP",
+  "LeNet-5"
+)
+
+tab_i <- tab
+
+# Logic:
+tab_i <- tab_i[variable %in% measures]
+tab_i[,variable:=factor(variable, levels=measures)]
+tab_i <- tab_i[dataname %in% chosen_data]
+tab_i <- tab_i[model %in% chosen_model]
+tab_i[,model:=factor(model,levels=chosen_model)]
+tab_i[,dataname:=factor(dataname,levels=chosen_data)]
+tab_i <- dcast(tab_i, model + generator ~ dataname + variable)
+col_names <- c(
+  "Model",
+  "Generator",
+  rep(measure_names,length(chosen_data))
+)
+caption <- sprintf(
+  "Results for %s dataset: sample averages +/- one standard deviation across counterfactuals.
+Best outcomes are highlighted in bold.
+Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (Wachter).
+\\label{tab:results-%s} \\newline",
+  chosen_source,
+  chosen_source
+)
+file_name <- sprintf(
+  "paper/contents/table-%s.tex",
+  chosen_source
+)
+sub_header <- rep(length(measures), length(chosen_data))
+names(sub_header) <- chosen_data
+header <- c(
+  " " = 2, sub_header
+)
+line_sep <- c(rep("",length(measures)-1),"\\addlinespace")
+algin_cols <- c(rep('l',2),rep('c',ncol(tab_i)-2))
+kbl(
+  tab_i, caption = caption, 
+  align = algin_cols, col.names=col_names, booktabs = T, escape=F, 
+  format="latex", linesep = line_sep 
+) %>%
+  kable_styling(latex_options = c("scale_down")) %>%
+  kable_paper(full_width = F) %>%
   add_header_above(header) %>%

### 2023-09-24 paper/body.tex
kill me
git add .

@@
 
-We use both synthetic and real-world datasets from different domains, all of which are publicly available and commonly used to train and benchmark classification algorithms.
-We synthetically generate a dataset containing two \textit{Linearly Separable} Gaussian clusters ($n=1000$), as well as the well-known \textit{Circles} ($n=1000$) and \textit{Moons} ($n=2500$) data.
-Since these data are generated by distributions of varying degrees of complexity, they allow us to assess how the generators and our proposed evaluation metrics handle this.
-
-As for real-world data, we follow~\citet{schut2021generating} and use the \textit{MNIST}~\citep{lecun1998mnist} dataset containing images of handwritten digits such as the example shown above in Figure~\ref{fig:motiv}.
-From the social sciences domain, we include Give Me Some Credit (\textit{GMSC})~\citep{kaggle2011give}: a tabular dataset that has been studied extensively in the literature on algorithmic recourse~\citep{pawelczyk2021carla}.
-It consists of 11 numeric features that can be used to predict the binary outcome variable indicating whether retail borrowers experience financial distress.
-
-For the predictive modelling tasks, we use simple neural networks (\textit{MLP}) and Joint Energy Models (\textit{JEM}).
-For the more complex real-world datasets we also use ensembling in each case.
+For the predictive modelling tasks, we use multi-layer perceptrons (\textit{MLP}), deep ensembles, joint energy models (\textit{JEM}) and convolutional neural networks.
 Both joint-energy modelling and ensembling have been associated with improved generative properties and adversarial robustness~\citep{grathwohl2020your,lakshminarayanan2016simple}, so we expect this to be positively correlated with the plausibility of \textit{ECCCo}.
 To account for stochasticity, we generate multiple counterfactuals for each target class, generator, model and dataset.
-Specifically, we randomly sample $n^{-}$ times from the subset of individuals for which the given model predicts the non-target class $\mathbf{y}^{-}$ given the current target.
-We set $n^{-}=25$ for all of our synthetic datasets, $n^{-}=10$ for \textit{GMSC} and $n^{-}=5$ for \textit{MNIST}.
 Full details concerning our parameter choices, training procedures and model performance can be found in Appendix~\ref{app:setup}.
 
-\subsection{Results for Synthetic Data}
+We perform benchmarks on eight different datasets from different domains.
+From the credit and finance domain we include three tabular datasets: Give Me Some Credit (\textit{GMSC})~\citep{kaggle2011give}, \textit{German Credit} and \textit{California Housing} (REFRENCES).
+All of these are commonly used in the related literature~\citep{karimi2020survey,altmeyer2023endogenous,pawelczyk2021carla}.
+Following~\citet{schut2021generating} we also include two image datasets: \textit{MNIST}~\citep{lecun1998mnist} and \textit{Fashion MNIST}~\citep{xiao2017fashion}.
+These are small image datasets but they allow us to easily visually gauge how plausible counteractuals are.
+(EXPLAIN WHY NO LARGER?) Finally, we also benchmark our generators on multiple synthetic datasets.
+Detailed descriptions and results for all datasets can be found in the appendix (POINTER).
+In the following we will focus on the most relevant results highlighted in Tables~\ref{tab:results-tabular} and~\ref{tab:results-vision}.
 
-Table~\ref{tab:results-synthetic} shows the key results for the synthetic datasets separated by model (first column) and generator (second column).
-The numerical columns show sample averages and standard deviations of our key evaluation metrics computed across all counterfactuals.
-We have highlighted the best outcome for each model and metric in bold.
-To provide some sense of effect sizes, we have added asterisks to indicate that a given value is at least one ($*$) or two ($**$) standard deviations lower than the baseline (\textit{Wachter}).
+\import{contents/}{table-tabular.tex}
 
-Starting with the high-level results for our \textit{Linearly Separable} data, we find that \textit{ECCCo} produces the most faithful counterfactuals for both black-box models.
-This is consistent with our design since \textit{ECCCo} directly enforces faithfulness through regularization.
-Crucially though, \textit{ECCCo} also produces the most plausible counterfactuals for both models.
-This dataset is so simple that even the \textit{MLP} has learned plausible explanations of the input data.
-Zooming in on the granular details for the \textit{Linearly Separable} data, the results for \textit{ECCCo (no CP)} and \textit{ECCCo (no EBM)} indicate that the positive results are dominated by the effect of quantifying and leveraging the model's generative property (EBM).
-Conformal prediction alone only leads to marginally improved faithfulness and plausibility.
+\import{contents/}{table-vision.tex}
 
-The findings for the \textit{Moons} dataset are broadly in line with the findings so far: for the \textit{JEM}, \textit{ECCCo} yields substantially more faithful and plausible counterfactuals than all other generators.
-For the \textit{MLP}, faithfulness is maintained but counterfactuals are not plausible.
-This high-level pattern is broadly consistent with other more complex datasets and supportive of our narrative, so it is worth highlighting: \textit{ECCCo} consistently achieves high faithfulness, which---subject to the quality of the model itself---coincides with high plausibility.
-By comparison, \textit{REVISE} yields the most plausible counterfactuals for the \textit{MLP}, but it does so at the cost of faithfulness.
-We also observe that the best results for \textit{ECCCo} are achieved when using both penalties.
-Once again though, the generative component (EBM) has a stronger impact on the positive results for the \textit{JEM}.
+\subsection{Faithfulness}
 
-For the \textit{Circles} data, it appears that \textit{REVISE} performs well, but we note that it generates valid counterfactuals only half of the time (see Appendix~\ref{app:results} for a complete overview including additional common evaluation metrics).
-The underlying VAE with default parameters has not adequately learned the data-generating process.
-Of course, it is possible to improve generative performance through hyperparameter tuning but this example serves to illustrate that \textit{REVISE} depends on the quality of its surrogate.
-Independent of the outcome for \textit{REVISE}, however, the results do not seem to indicate that \textit{ECCCo} substantially improves faithfulness and plausibility for the \textit{Circles} data.
-We think this points to a limitation of our evaluation metrics rather than \textit{ECCCo} itself: computing average distances fails to account for the `wraparound' effect associated with circular data~\citep{gill2010circular}.
+Our key finding with respect to Research Question~\ref{rq:faithfulness} is as follows:
 
-\import{contents/}{table-synthetic.tex}
+\begin{quote}
+  We find strong empirical evidence that \textit{ECCCo} consistently achieves state-of-the-art faithfulness, as it was designed to do (Research Question~\ref{rq:faithfulness}).
+\end{quote}
 
-\subsection{Results for Real-World Data}
+\subsection{Balancing Objectives}
 
-The results for our real-world datasets are shown in Table~\ref{tab:results-real-world}.
-Once again the findings indicate that the plausibility attained by \textit{ECCCo} is positively correlated with the capacity of the black-box model to distinguish plausible from implausible inputs.
-The case is very clear for \textit{MNIST}: \textit{ECCCo} consistently generates more faithful counterfactuals than other generators and plausibility gradually improves through ensembling and joint-energy modelling.
-Interestingly, faithfulness also gradually improves for \textit{REVISE}.
-This indicates that as our models improve, their generative capacity approaches that of the surrogate VAE used by \textit{REVISE}.
-The VAE still outperforms our classifiers in this regard, as evident from the fact that \textit{ECCCo} never quite reaches the same level of plausibility as \textit{REVISE}.
-With reference to Appendix~\ref{app:results} we note that the results for \textit{Schut} need to be discounted as it rarely produces valid counterfactuals for \textit{MNIST}.
-Relatedly, we find that \textit{ECCCo} is the only generator that consistently achieves full validity.
-Finally, it is worth noting that \textit{ECCCo} produces counterfactual images with the lowest average predictive uncertainty for all models.
+With respect to balance the objectives of faithfulness and plausibility (Research Question~\ref{rq:plausibility}) we find the following:
 
-For the tabular credit dataset (\textit{GMSC}) it is inherently challenging to use deep neural networks in order to achieve good discriminative performance~\citep{borisov2022deep,grinsztajn2022why} and generative performance~\citep{liu2022goggle}, respectively.
-In order to achieve high plausibility, \textit{ECCCo} effectively requires classifiers to achieve good performance for both tasks.
-Since this is a challenging task even for Joint Energy Models, it is not surprising to find that even though \textit{ECCCo} once again achieves state-of-the-art faithfulness, it is outperformed by \textit{REVISE} and \textit{Schut} with respect to plausibility.
+\begin{quote}
+  The empirical evidence presented here suggests that \textit{ECCCo} consistently achieves near state-of-the-art plausibility without sacrificing faithfulness (Research Question~\ref{rq:plausibility}).
+\end{quote}
 
-\subsection{Key Takeways}
-
-To conclude this section, we summarize our findings with reference to the opening questions.
-The results clearly demonstrate that \textit{ECCCo} consistently achieves state-of-the-art faithfulness, as it was designed to do (Research Question~\ref{rq:faithfulness}).
-A related important finding is that \textit{ECCCo} yields highly plausible explanations provided that they faithfully describe model behaviour (Research Question~\ref{rq:plausibility}).
 \textit{ECCCo} achieves this result primarily by leveraging the model's generative property.
-
-\import{contents/}{table-real-world.tex}
 

### 2023-09-24 paper/tables.Rmd
git add .

@@
   "unfaithfulness",
-  "implausibility",
-  "distance"
+  "implausibility"
 )
@@
   "Unfaithfulness ↓",
-  "Implausibility ↓",
-  "Cost ↓"
+  "Implausibility ↓"
 )
@@
   "GMSC",
-  "California Housing"
+  "California Housing",
+  "German Credit"
 )
@@
   "MLP",
-  "JEM"
+  "JEM",
 )

### 2023-09-24 experiments/setup_env.jl

@@ -171,7 +171,7 @@ DEFAULT_GENERATOR_TUNING_LARGE = (
         [0.1, 0.1, 0.2],
         [0.2, 0.2, 0.2],
     ],
-    reg_strength=[0.0, 0.1, 0.5],
+    reg_strength=[0.0, 0.1, 0.5, 1.0],
     opt=[
         Flux.Optimise.Descent(0.05),
         Flux.Optimise.Descent(0.01),

### 2023-09-24 paper/body.tex
progres

@@
 Our default choice for the $\text{dist}(\cdot)$ function in both cases is the Euclidean Norm.
+Depending on the type of input data other choices may be more adequate, which we discuss further in Section~\ref{emp:setup}.
 
@@
 
-\subsection{Experimental Setup}
+\subsection{Experimental Setup}\label{emp:setup}
 
@@
 Detailed descriptions and results for all datasets can be found in the appendix (POINTER).
+
 In the following we will focus on the most relevant results highlighted in Tables~\ref{tab:results-tabular} and~\ref{tab:results-vision}.
@@
 \import{contents/}{table-vision.tex}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=1.0\linewidth]{../www/mnist_benchmark.png}
+  \caption{Counterfactuals for turning a 3 into a 5: factual (left); then from left to right the counterfactuals generated using \textit{ECCCo}, \textit{ECCCo+}, \textit{REVISE}, \textit{Schut} and \textit{Wachter}.}\label{fig:mnist-bmk}
+\end{figure}
 
@@
 \textit{ECCCo} achieves this result primarily by leveraging the model's generative property.
+
+\subsection{Additional Observations}
 

### 2023-09-25 paper/body.tex
mostly done

@@
 
+Concerning feature autoencoding ($f: \mathcal{Z} \mapsto \mathcal{X}$), \textit{ECCCo} does not rely on latent space search to achieve its primary objective of faithfulness.
+By default, we choose $f(\cdot)$ to be the identity function as in \textit{Wachter}.
+This is generally also enough to achieve plausibility, provided the model has learned plausible explanations for the data.
+In some cases, plausibility can be improved further by mapping counterfactuals to a lower-dimensional latent space.
+In the following, we refer to this approach as \textit{ECCCo+}: that is, \textit{ECCCo} plus dimensionality reduction.
+
 \begin{figure}
@@
 To assess and benchmark the performance of our proposed generator against the state of the art, we generate multiple counterfactuals for different models and datasets.
-In particular, we compare \textit{ECCCo} and its variants to the following counterfactual generators that were introduced above: firstly; \textit{Schut}, which works under the premise of minimizing predictive uncertainty; secondly, \textit{REVISE}, which is state-of-the-art with respect to plausibility; and, finally, \textit{Wachter}, which serves as our baseline.
+In particular, we compare \textit{ECCCo} and its variants to the following counterfactual generators that were introduced above: firstly; \textit{Schut}, which works under the premise of minimizing predictive uncertainty; secondly, \textit{REVISE}, which is state-of-the-art (SOTA) with respect to plausibility; and, finally, \textit{Wachter}, which serves as our baseline.
+In the case of \textit{ECCCo+}, we use Principal Component Analysis for dimensionality reduction: the traversed latent space $\mathcal{Z}$ is spanned by the first $n_z$ principal components where we choose $n_z$ to be equal to the latent dimension of the VAE that is used by \textit{REVISE}.
 
@@
 
-We perform benchmarks on eight different datasets from different domains.
+We perform benchmarks on eight datasets from different domains.
 From the credit and finance domain we include three tabular datasets: Give Me Some Credit (\textit{GMSC})~\citep{kaggle2011give}, \textit{German Credit} and \textit{California Housing} (REFRENCES).
@@
 Following~\citet{schut2021generating} we also include two image datasets: \textit{MNIST}~\citep{lecun1998mnist} and \textit{Fashion MNIST}~\citep{xiao2017fashion}.
-These are small image datasets but they allow us to easily visually gauge how plausible counteractuals are.
+These are small image datasets but they allow us to easily visually gauge how plausible counterfactuals are.
 (EXPLAIN WHY NO LARGER?) Finally, we also benchmark our generators on multiple synthetic datasets.
@@
 
-In the following we will focus on the most relevant results highlighted in Tables~\ref{tab:results-tabular} and~\ref{tab:results-vision}.
+In the following, we will focus on the most relevant results highlighted in Tables~\ref{tab:results-tabular} and~\ref{tab:results-vision}.
+The tables show sample averages along with standard deviations for our key evaluation metrics for the \textit{California Housing} and \textit{GMSC} datasets (Table~\ref{tab:results-tabular}) and the \textit{MNIST} dataset (Table~\ref{tab:results-vision}).
+For each metric, the best outcomes are highlighted in bold.
+Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (Wachter).
+For the tabular datasets, we use the default Euclidian distance to measure unfaithfulness and implausibility as defined in Equations~\ref{eq:faith} and~\ref{eq:impl}, respectively.
+For the vision datasets, we instead rely on measuring the structural dissimilarity between images~\citep{wang2003multiscale}.
+
+\subsection{Faithfulness}
 
@@
 
-\import{contents/}{table-vision.tex}
+Overall, we find strong empirical evidence suggesting that \textit{ECCCo} consistently achieves state-of-the-art faithfulness.
+Across all models and datasets highlighted here, all variations of \textit{ECCCo} consistently outperform all other generators with respect to faithfulness, in many cases substantially.
+This pattern is mostly robust across all other benchmark datasets (Tables~\ref{tab:results-linearly-separable} to~\ref{tab:results-fashion-mnist}).
+
+In particular, we note that the best results are generally obtained when using the full \textit{ECCCo} objective (Equation~\ref{eq:eccco}).
+In other words, constraining both energy and predictive uncertainty typically yields the most faithful counterfactuals.
+We expected the former to play a more significant role in this context and that is typically what we find across all datasets.
+For example, the results for \textit{GMSC} in Table~\ref{tab:results-tabular} indicate that faithfulness can be improved substantially by relying solely on the energy constraint (\textit{ECCCo (no CP)}).
+In some cases though, as for the \textit{California Housing} dataset, \textit{ECCCo (no EBM)} actually outperforms \textit{ECCCo (no CP)}.
+This indicates that predictive uncertainty minimization plays an important role in achieving faithfulness.
+
+We also generally find that the highest degree of faithfulness is obtained when the counterfactual search is performed directly in the feature space $\mathcal{X}$.
+While \textit{ECCCo+} typically attains high levels of faithfulness compared to most other generators, it is consistently outperformed by \textit{ECCCo}.
+The case is even stronger for \textit{REVISE}, which performs worst out of all generators for faithfulness on the \textit{GMSC} dataset and better only than \textit{Wachter} on \textit{California Housing}.
+
+These findings are consistent with the notion that surrogate models may inhibit faithfulness.
+Even though dimensionality reduction through PCA in the case of \textit{ECCCo+} can be considered a relatively mild form of intervention, the first $n_z$ principal components fail to capture some of the variation in the data, that the underlying model itself may be sensitive to.
+This notion is illustrated nicely in Figure~\ref{fig:mnist-bmk}, where the counterfactual produced by \textit{ECCCo} is somewhat noisier and grainier than the one produced by \textit{ECCCo+}.
+
+In conclusion, we recommend in light of the findings here to use the full \textit{ECCCo} search objective whenever model faithfulness is a key priority.
+
+\subsection{Balancing Objectives}
 
@@
 
-\subsection{Faithfulness}
+% Sample UUID of factual: UUID("1ceb06f0-5949-11ee-0e9c-dd49ccfec8c3")
 
-Our key finding with respect to Research Question~\ref{rq:faithfulness} is as follows:
+\import{contents/}{table-vision.tex}
 
-\begin{quote}
-  We find strong empirical evidence that \textit{ECCCo} consistently achieves state-of-the-art faithfulness, as it was designed to do (Research Question~\ref{rq:faithfulness}).
-\end{quote}
+Overall, we find strong empirical evidence suggesting that \textit{ECCCo} can achieve near state-of-the-art plausibility without sacrificing faithfulness.
+Figure~\ref{fig:mnist-bmk} shows one such example taken from the \textit{MNIST} benchmark.
+In this case, the objective was to turn the factual three (far left) into a five.
+The underlying model is a LeNet-5 convolutional neural network.
+The different images show the counterfactuals produced by the generators, of which all but the one produced by \textit{Schut} are valid.
+Both variations of \textit{ECCCo} produce plausible counterfactuals.
 
-\subsection{Balancing Objectives}
+Looking at the benchmark results presented in Tables~\ref{tab:results-tabular} and~\ref{tab:results-vision} we firstly note that although \textit{REVISE} generally performs best, \textit{ECCCo} and in particular \textit{ECCCo+} often approach SOTA performance.
+Zooming in on the details we observe that \textit{ECCCo} and its variations do particularly well, whenever the underlying model has been explicitly trained to learn plausible representations of the data.
+For both tabular datasets in Table~\ref{tab:results-tabular}, \textit{ECCCo} improves plausibility substantially compared to the baseline.
+This broad pattern is mostly consistent for all other datasets, although there are notable exceptions for which \textit{ECCCo} takes the lead on both plausibility and faithfulness (see for example Tables~\ref{tab:results-moons} and~\ref{tab:results-german-credit}).
 
-With respect to balance the objectives of faithfulness and plausibility (Research Question~\ref{rq:plausibility}) we find the following:
+While we maintain that generally speaking plausibility should hinge on the quality of the model, our results also indicate that it is possible to trade off some degree of faithfulness for plausibility if needed: \textit{ECCCo+} generally outperforms other variants of \textit{ECCCo} in this context at the small cost of slightly reduced faithfulness.
+For the vision datasets especially, we find that  \textit{ECCCo+} is consistently second only to \textit{REVISE} for all models and regularly substantially better than the baseline.
+Looking at the \textit{California Housing} data, latent space search markedly improves plausibility without sacrificing faithfulness: for the \textit{JEM} Ensemble, \textit{ECCCo+} performs substantially better than the baseline and only marginally worse than \textit{REVISE}.
+Importantly, \textit{ECCCo+} does not attain plausibility at all costs: for the MLP, plausibility is still very low but this seems to faithfully represent what the model has learned.
 
-\begin{quote}
-  The empirical evidence presented here suggests that \textit{ECCCo} consistently achieves near state-of-the-art plausibility without sacrificing faithfulness (Research Question~\ref{rq:plausibility}).
-\end{quote}
-
-\textit{ECCCo} achieves this result primarily by leveraging the model's generative property.
+We conclude that \textit{ECCCo} offers us a way to balance the objectives of faithfulness and plausibility. \textit{ECCCo+} can be used to tilt the scale in favour of plausibility if needed.
 

### 2023-09-25 paper/tables.Rmd
mostly done

--- 2023-09-24 paper/tables.Rmd
@@
 ```{r}
+generators <- unique(dt$generator)[sapply(unique(dt$generator), function(x) {!grepl("L1",x)})]
 tab <- dt[
-    ,
+    generator %in% generators,
     .(
@@
   "unfaithfulness",
-  "implausibility"
+  "implausibility",
+  "set_size_penalty"
 )
@@
   "Unfaithfulness ↓",
-  "Implausibility ↓"
+  "Implausibility ↓",
+  "Uncertainty ↓"
 )
+chosen_source <- "tabular"
 # Order:
 chosen_data <- c(
-  "Linearly Separable",
-  "GMSC",
-  "MNIST",
+  "California Housing",
+  "GMSC"
 )
@@
   "MLP",
-  "JEM",
-  "LeNet-5"
+  "JEM Ensemble"
 )
@@
 tab_i <- tab_i[variable %in% measures]
+tab_i[,variable:=factor(variable, levels=measures)]
 tab_i <- tab_i[dataname %in% chosen_data]
 tab_i <- tab_i[model %in% chosen_model]
-tab_i[,dataname:=factor(dataname,levels=chosen_data)]
-tab_i <- dcast(tab_i, model + generator ~ dataname + variable)
-col_names <- c(
-  "Model",
-  "Generator",
-  rep(measure_names,length(chosen_data))
-)
-caption <-  "Results for datasets from different domains: sample averages +/- one standard deviation across counterfactuals.
-Best outcomes are highlighted in bold.
-Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (Wachter).
-\\label{tab:results-main} \\newline"
-file_name <- "paper/contents/table-main.tex"
-sub_header <- rep(length(measures), length(chosen_data))
-names(sub_header) <- chosen_data
-header <- c(
-  " " = 2, sub_header
-)
-line_sep <- c(rep("",length(measures)-1),"\\addlinespace")
-algin_cols <- c(rep('l',2),rep('c',ncol(tab_i)-2))
-kbl(
-  tab_i, caption = caption, 
-  align = algin_cols, col.names=col_names, booktabs = T, escape=F, 
-  format="latex", linesep = line_sep 
-) %>%
-  kable_styling(latex_options = c("scale_down")) %>%
-  kable_paper(full_width = T) %>%
-  add_header_above(header) %>%
-  collapse_rows(columns = 1:2, latex_hline = "major", valign = "middle") %>%
-  save_kable(file_name)
-```
-
-```{r}
-# Choices:
-measures <- c(
-  "distance_from_energy_l2",
-  "distance_from_targets_l2"
-)
-measure_names <- c(
-  "Unfaithfulness ↓",
-  "Implausibility ↓"
-)
-chosen_source <- "tabular"
-# Order:
-chosen_data <- c(
-  "GMSC",
-  "California Housing",
-  "German Credit"
-)
-chosen_model <- "MLP"
-
-tab_i <- tab
-
-# Logic:
-tab_i <- tab_i[variable %in% measures]
-tab_i <- tab_i[source == chosen_source]
-tab_i <- tab_i[dataname %in% chosen_data]
-tab_i <- tab_i[model == chosen_model]
+tab_i[,model:=factor(model,levels=chosen_model)]
 tab_i[,dataname:=factor(dataname,levels=chosen_data)]
@@
   align = algin_cols, col.names=col_names, booktabs = T, escape=F, 
-  format="latex", linesep = line_sep 
+  format="latex", linesep = line_sep, table.env="table*"
 ) %>%
-  kable_styling(latex_options = c("scale_down")) %>%
-  kable_paper(full_width = T) %>%
+  kable_styling(
+    latex_options = c("scale_down")
+  ) %>%
+  kable_paper(full_width = F) %>%
   add_header_above(header) %>%
@@
 # Choices:
-# Choices:
 measures <- c(
-  "distance_from_energy_l2",
-  "distance_from_targets_l2"
+  "unfaithfulness",
+  "implausibility"
 )
@@
 )
-chosen_source <- "synthetic"
+chosen_source <- "vision"
+# Order:
 chosen_data <- c(
-  "Linearly Separable",
-  "Moons",
-  "Circles"
+  "MNIST"
 )
+chosen_model <- c(
+  "MLP",
+  "LeNet-5"
+)
+
 tab_i <- tab
@@
 tab_i <- tab_i[variable %in% measures]
-tab_i <- tab_i[source == chosen_source]
+tab_i[,variable:=factor(variable, levels=measures)]
 tab_i <- tab_i[dataname %in% chosen_data]
+tab_i <- tab_i[model %in% chosen_model]
+tab_i[,model:=factor(model,levels=chosen_model)]
 tab_i[,dataname:=factor(dataname,levels=chosen_data)]
@@
 caption <- sprintf(
-  "Results for %s datasets: sample averages +/- one standard deviation across counterfactuals.
-Best outcomes are highlighted in bold.
-Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (Wachter).
-\\label{tab:results-%s} \\newline",
+  "Results for %s dataset. Formatting details are the same as in Table~\\ref{tab:results-tabular}. \\label{tab:results-%s} \\newline",
   chosen_source,
@@
 ) %>%
-  kable_styling(latex_options = c("scale_down")) %>%
-  kable_paper(full_width = T) %>%
+  kable_styling(
+    latex_options = c("scale_down")
+  ) %>%
+  kable_paper(full_width = F) %>%
   add_header_above(header) %>%
@@
 
+## Full tables
+
 ```{r}
-# Choices:
+tab <- dt[
+    ,
+    .(
+      value=sprintf("%1.2f ± %1.2f", mean(value), sd(value)),
+      val = mean(value),
+      std = sd(value)
+    ),
+    .(dataname, generator, model, variable, source)
+]
+
+tab$top_val = F
+tab$one_std_wachter = F
+tab$two_std_wachter = F
+
+# Measures to be minimized:
+min_measures <- c(
+  "distance",
+  "implausibility",
+  "unfaithfulness",
+  "distance_from_energy", 
+  "distance_from_energy_l2", 
+  "distance_from_targets",
+  "distance_from_targets_l2",
+  "set_size_penalty"
+)
+tab[variable %in% min_measures,top_val:=val==min(val),.(model, dataname, variable)]
+tab[variable %in% min_measures,top_val:=ifelse(rep(all(top_val),length(top_val)),F,top_val),.(model, dataname, variable)]
+tab[variable %in% min_measures,two_std_wachter:=val+2*std<val[generator=="Wachter"],.(model, dataname, variable)]
+tab[variable %in% min_measures,one_std_wachter:=val+1*std<val[generator=="Wachter"],.(model, dataname, variable)]
+
+# Measures to be maximized:
+max_measures <- c(
+  "validity",
+  "redundancy"
+)
+tab[variable %in% max_measures,top_val:=val==max(val),.(model, dataname, variable)]
+tab[variable %in% max_measures,top_val:=ifelse(rep(all(top_val),length(top_val)),F,top_val),.(model, dataname, variable)]
+tab[variable %in% max_measures,two_std_wachter:=val-2*std>val[generator=="Wachter"],.(model, dataname, variable)]
+tab[variable %in% max_measures,one_std_wachter:=val-1*std>val[generator=="Wachter"],.(model, dataname, variable)]
+
+# Add conditional formatting:
+tab$value <- cell_spec(tab$value, "latex", bold=tab$top_val)
+tab[one_std_wachter==T,value:=paste0(value,"*")]
+tab[one_std_wachter==F,value:=paste0(value,"\\hphantom{*}")]
+tab[two_std_wachter==T,value:=paste0(value,"*")]
+tab[two_std_wachter==F,value:=paste0(value,"\\hphantom{*}")]
+
+# Remove redundant columns:
+tab[,val:=NULL]
+tab[,std:=NULL]
+tab[,top_val:=NULL]
+tab[,two_std_wachter:=NULL]
+tab[,one_std_wachter:=NULL]
+
+# Measures:
 measures <- c(
-  "distance_from_energy_l2",
-  "distance_from_targets_l2"
+  "unfaithfulness",
+  "implausibility",
+  "set_size_penalty",
+  "distance",
+  "redundancy",
+  "validity"
 )
@@
   "Unfaithfulness ↓",
-  "Implausibility ↓"
+  "Implausibility ↓",
+  "Uncertainty ↓",
+  "Cost ↓",
+  "Redundancy ↑",
+  "Validity ↑"
 )
-chosen_source <- "vision"
-# Order:
-chosen_data <- c(
-  "MNIST",
-  "Fashion MNIST",
+tab <- tab[variable %in% measures]
+tab[,variable:=factor(variable, levels=measures)]
+```
+
+```{r}
+tab_valid <- dt_valid[
+    ,
+    .(
+      value=sprintf("%1.2f ± %1.2f", mean(value), sd(value)),
+      val = mean(value),
+      std = sd(value)
+    ),
+    .(dataname, generator, model, variable, source)
+]
+
+tab_valid$top_val = F
+tab_valid$one_std_wachter = F
+tab_valid$two_std_wachter = F
+
+# Measures to be minimized:
+min_measures <- c(
+  "distance",
+  "implausibility",
+  "unfaithfulness",
+  "distance_from_energy", 
+  "distance_from_energy_l2", 
+  "distance_from_targets",
+  "distance_from_targets_l2",
+  "set_size_penalty"
 )
-chosen_model <- "LeNet-5"
+tab_valid[variable %in% min_measures,top_val:=val==min(val),.(model, dataname, variable)]
+tab_valid[variable %in% min_measures,top_val:=ifelse(rep(all(top_val),length(top_val)),F,top_val),.(model, dataname, variable)]
+tab_valid[variable %in% min_measures,two_std_wachter:=val+2*std<val[generator=="Wachter"],.(model, dataname, variable)]
+tab_valid[variable %in% min_measures,one_std_wachter:=val+1*std<val[generator=="Wachter"],.(model, dataname, variable)]
 
-tab_i <- tab
+# Measures to be maximized:
+max_measures <- c(
+  "validity",
+  "redundancy"
+)
+tab_valid[variable %in% max_measures,top_val:=val==max(val),.(model, dataname, variable)]
+tab_valid[variable %in% max_measures,top_val:=ifelse(rep(all(top_val),length(top_val)),F,top_val),.(model, dataname, variable)]
+tab_valid[variable %in% max_measures,two_std_wachter:=val-2*std>val[generator=="Wachter"],.(model, dataname, variable)]
+tab_valid[variable %in% max_measures,one_std_wachter:=val-1*std>val[generator=="Wachter"],.(model, dataname, variable)]
 
-# Logic:
-tab_i <- tab_i[variable %in% measures]
-tab_i <- tab_i[source == chosen_source]
-tab_i <- tab_i[dataname %in% chosen_data]
-tab_i <- tab_i[model == chosen_model]
-tab_i[,dataname:=factor(dataname,levels=chosen_data)]
-tab_i <- dcast(tab_i, model + generator ~ dataname + variable)
+# Add conditional formatting:
+tab_valid$value <- cell_spec(tab_valid$value, "latex", bold=tab_valid$top_val)
+tab_valid[one_std_wachter==T,value:=paste0(value,"*")]
+tab_valid[one_std_wachter==F,value:=paste0(value,"\\hphantom{*}")]
+tab_valid[two_std_wachter==T,value:=paste0(value,"*")]
+tab_valid[two_std_wachter==F,value:=paste0(value,"\\hphantom{*}")]
+
+# Remove redundant columns:
+tab_valid[,val:=NULL]
+tab_valid[,std:=NULL]
+tab_valid[,top_val:=NULL]
+tab_valid[,two_std_wachter:=NULL]
+tab_valid[,one_std_wachter:=NULL]
+
+# Measures:
+measures <- c(
+  "unfaithfulness",
+  "implausibility",
+  "set_size_penalty",
+  "distance",
+  "redundancy",
+  "validity"
+)
+measure_names <- c(
+  "Unfaithfulness ↓",
+  "Implausibility ↓",
+  "Uncertainty ↓",
+  "Cost ↓",
+  "Redundancy ↑",
+  "Validity ↑"
+)
+tab_valid <- tab_valid[variable %in% measures]
+tab_valid[,variable:=factor(variable, levels=measures)]
+```
+
+### Tables (all)
+
+```{r}
+for (name in unique(tab$dataname)) {
+  data_indicator <- gsub(" ", "-", tolower(unique(name)))
+  # Choices:
+  tab_full <- dcast(tab[dataname==name], model + generator ~ variable)
   col_names <- c(
@@
     "Generator",
-  rep(measure_names,length(chosen_data))
+    measure_names
   )
-caption <- sprintf(
-  "Results for %s datasets: sample averages +/- one standard deviation across counterfactuals.
+  algin_cols <- c(rep('l',3),rep('c',ncol(tab_full)-3))
+  file_name <- sprintf(
+    "paper/contents/table-%s.tex", 
+    data_indicator
+  )
+  cap <- sprintf(
+    "All results for %s dataset: sample averages +/- one standard deviation over all counterfactuals.
 Best outcomes are highlighted in bold.
@@
 \\label{tab:results-%s} \\newline",
-  chosen_source,
-  chosen_source
+    name,
+    data_indicator
   )
-file_name <- sprintf(
-  "paper/contents/table-%s.tex",
-  chosen_source
-)
-sub_header <- rep(length(measures), length(chosen_data))
-names(sub_header) <- chosen_data
-header <- c(
-  " " = 2, sub_header
-)
-line_sep <- c(rep("",length(measures)-1),"\\addlinespace")
-algin_cols <- c(rep('l',2),rep('c',ncol(tab_i)-2))
   kbl(
-  tab_i, caption = caption, 
-  align = algin_cols, col.names=col_names, booktabs = T, escape=F, 
-  format="latex", linesep = line_sep 
-) %>%
-  kable_styling(latex_options = c("scale_down")) %>%
-  kable_paper(full_width = T) %>%
-  add_header_above(header) %>%
-  collapse_rows(columns = 1:2, latex_hline = "major", valign = "middle") %>%
-  save_kable(file_name)
-```
-
-## Full table
-
-```{r}
-tab_full <- dcast(tab, dataname + model + generator ~ variable)
-col_names <- c(
-  "Model",
-  "Data",
-  "Generator",
-  "Cost ↓", 
-  "Unfaithfulness ↓", 
-  "Implausibility ↓", 
-  "Redundancy ↑",
-  "Uncertainty ↓",
-  "Validity ↑"
-)
-algin_cols <- c(rep('l',3),rep('c',ncol(tab_full)-3))
-kbl(
-  tab_full, caption = "All results for all datasets: sample averages +/- one standard deviation over all counterfactuals.
-Best outcomes are highlighted in bold.
-Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (Wachter).
-\\label{tab:results-full} \\newline",
+    tab_full, caption = cap, 
     align = "c", col.names=col_names, booktabs = T, escape=F, 
@@
     kable_paper(full_width = F) %>%
-  collapse_rows(columns = 1:3, latex_hline = "custom", valign = "top", custom_latex_hline = 1:2) %>%
-  save_kable("paper/contents/table_all.tex")
+    collapse_rows(columns = 1:2, latex_hline = "custom", valign = "top") %>%
+    save_kable(file_name)
+}
+
 ```
@@
 ```{r}
-tab_full <- dcast(tab_valid, dataname + model + generator ~ variable)
+for (name in unique(tab_valid$dataname)) {
+  data_indicator <- gsub(" ", "-", tolower(unique(name)))
+  # Choices:
+  tab_full <- dcast(tab_valid[dataname==name], model + generator ~ variable)
   col_names <- c(
     "Model",
-  "Data",
     "Generator",
-  "Cost ↓", 
-  "Unfaithfulness ↓", 
-  "Implausibility ↓", 
-  "Redundancy ↑",
-  "Uncertainty ↓",
-  "Validity ↑"
+    measure_names
   )
   algin_cols <- c(rep('l',3),rep('c',ncol(tab_full)-3))
-kbl(
-  tab_full, caption = "All results for all datasets: sample averages +/- one standard deviation over all valid counterfactuals.
+  file_name <- sprintf(
+    "paper/contents/table-%s-valid.tex", 
+    data_indicator
+  )
+  cap <- sprintf(
+    "All results for %s dataset: sample averages +/- one standard deviation over all valid counterfactuals.
 Best outcomes are highlighted in bold.
 Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (Wachter).
-\\label{tab:results-full-valid} \\newline",
+\\label{tab:results-%s} \\newline",
+    name,
+    data_indicator
+  )
+  kbl(
+    tab_full, caption = cap, 
     align = "c", col.names=col_names, booktabs = T, escape=F, 
@@
     kable_paper(full_width = F) %>%
-  collapse_rows(columns = 1:3, latex_hline = "custom", valign = "top", custom_latex_hline = 1:2) %>%
-  save_kable("paper/contents/table_all_valid.tex")
+    collapse_rows(columns = 1:2, latex_hline = "custom", valign = "top") %>%
+    save_kable(file_name)
+}
 ```

### 2023-09-25 experiments/grid_search.jl

@@ -130,7 +130,7 @@ function best_absolute_outcome(
     for (params, outcome) in outcomes
 
         # Setup
-        evaluation = outcome.bmk.evaluation
+        evaluation = deepcopy(outcome.bmk.evaluation)
         exper = outcome.exper
         generator_dict = outcome.generator_dict
         model_dict = outcome.model_dict

### 2023-09-25 experiments/post_processing/plotting.jl

@@ -1,4 +1,4 @@
-function choose_random_mnist(outcome::ExperimentOutcome; model::String="MLP", img_height=200, seed=966)
+function choose_random_mnist(outcome::ExperimentOutcome; model::String="LeNet-5", img_height=125, seed=966)
 
     # Set seed:
     if !isnothing(seed)
@@ -30,10 +30,12 @@ function choose_random_mnist(outcome::ExperimentOutcome; model::String="MLP", im
         title="Factual",
     )
     plts = [p1]
+    ces = []
 
     # Counterfactuals:
     for (i, generator) in enumerate(generators)
-        img = CounterfactualExplanations.counterfactual(df.ce[i]) |> ECCCo.convert2mnist
+        ce = df.ce[i]
+        img = CounterfactualExplanations.counterfactual(ce) |> ECCCo.convert2mnist
         p = Plots.plot(
             img,
             axis=([], false),
@@ -41,20 +43,22 @@ function choose_random_mnist(outcome::ExperimentOutcome; model::String="MLP", im
             title="$generator",
         )
         push!(plts, p)
+        push!(ces, ce)
     end
 
     plt = Plots.plot(
         plts...,
         layout=(1, n_generators + 1), 
-        size=(img_height * (n_generators + 1), img_height)
+        size=(img_height * (n_generators + 1), img_height),
+        dpi=300
     )
     display(plt)
 
-    return plt, df.target[1], seed
+    return plt, df.target[1], seed, ces, df.sample[1]
 
 end
 
-function plot_random_eccco(outcome::ExperimentOutcome; generator="ECCCo-Δ", img_height=200, seed=966)
+function plot_random_eccco(outcome::ExperimentOutcome; ce=nothing, generator="ECCCo-Δ", img_height=200, seed=966)
     # Set seed:
     if !isnothing(seed)
         Random.seed!(seed)
@@ -62,7 +66,7 @@ function plot_random_eccco(outcome::ExperimentOutcome; generator="ECCCo-Δ", img
 
     # Get output:
     bmk = outcome.bmk()
-    ce = rand(bmk.ce)
+    ce = isnothing(ce) ? rand(bmk.ce) : ce
     gen = outcome.generator_dict[generator]
     models = outcome.model_dict
     x = CounterfactualExplanations.counterfactual(ce)
@@ -95,7 +99,8 @@ function plot_random_eccco(outcome::ExperimentOutcome; generator="ECCCo-Δ", img
     plt = Plots.plot(
[…]

### 2023-09-25 paper/body.tex
Cynthia's comments addressed

@@
 
-Counterfactual explanations provide a powerful, flexible and intuitive way to not only explain black-box models but also help affected individuals through the means of algorithmic recourse.
+Counterfactual explanations provide a powerful, flexible and intuitive way to not only explain black-box models but also offer the possibility of algorithmic recourse to affected individuals.
 Instead of opening the black box, counterfactual explanations work under the premise of strategically perturbing model inputs to understand model behaviour~\citep{wachter2017counterfactual}.
@@
 Counterfactuals are then searched by minimizing a loss function that compares the predicted model output to the target outcome: $\text{yloss}(M_{\theta}(\mathbf{x}),\mathbf{y}^+)$.
-Since counterfactual explanations work directly with the black-box model, valid counterfactuals always have full local fidelity by construction where fidelity is defined as the degree to which explanations approximate the predictions of a black-box model~\citep{mothilal2020explaining,molnar2022interpretable}.
+Since counterfactual explanations work directly with the black-box model, valid counterfactuals always have full local fidelity by construction where fidelity is defined as the degree to which explanations approximate the predictions of a black-box model~\citep{molnar2022interpretable}.
 
 In situations where full fidelity is a requirement, counterfactual explanations offer a more appropriate solution to Explainable Artificial Intelligence (XAI) than other popular approaches like LIME~\citep{ribeiro2016why} and SHAP~\citep{lundberg2017unified}, which involve local surrogate models.
-But even full fidelity is not a sufficient condition for ensuring that an explanation faithfully describes the behaviour of a model.
-That is because multiple very distinct explanations can all lead to the same model prediction, especially when dealing with heavily parameterized models like deep neural networks, which are typically underspecified by the data~\citep{wilson2020case}.
-
+But even full fidelity is not a sufficient condition for ensuring that an explanation \textit{faithfully} describes the behaviour of a model.
+That is because multiple distinct explanations can lead to the same model prediction, especially when dealing with heavily parameterized models like deep neural networks, which are underspecified by the data~\citep{wilson2020case}.
 In the context of counterfactuals, the idea that no two explanations are the same arises almost naturally.
-A key focus in the literature has therefore been to identify those explanations and algorithmic recourses that are most appropriate based on a myriad of desiderata such as closeness~\citep{wachter2017counterfactual}, sparsity~\citep{schut2021generating}, actionability~\citep{ustun2019actionable} and plausibility~\citep{joshi2019realistic}.
+A key focus in the literature has therefore been to identify those explanations that are most appropriate based on a myriad of desiderata such as closeness~\citep{wachter2017counterfactual}, sparsity~\citep{schut2021generating}, actionability~\citep{ustun2019actionable} and plausibility~\citep{joshi2019realistic}.
 
@@
 We define faithfulness as the degree to which counterfactuals are consistent with what the model has learned about the data.
-Our key contributions are as follows:
-
-\begin{itemize}
-  \item We show that fidelity is an insufficient evaluation metric for counterfactuals (Section~\ref{fidelity}) and propose a definition of faithfulness that gives rise to more suitable metrics (Section~\ref{faithfulness}).
-  \item We introduce a ECCCo: a novel algorithmic approach aimed at generating Energy-Constrained Conformal Counterfactuals that faithfully explain model behaviour in Section~\ref{meth}.
-  \item We provide extensive empirical evidence demonstrating that ecccos faithfully explain model behaviour and attain plausibility only when appropriate (Section~\ref{emp}).
-\end{itemize}
+Our key contributions are as follows: first, we show that fidelity is an insufficient evaluation metric for counterfactuals (Section~\ref{fidelity}) and propose a definition of faithfulness that gives rise to more suitable metrics (Section~\ref{faithfulness}).
+Next, we introduce a \textit{ECCCo}: a novel algorithmic approach aimed at generating energy-constrained conformal counterfactuals that faithfully explain model behaviour in Section~\ref{meth}.
+Finally, we provide extensive empirical evidence demonstrating that \textit{ECCCo} faithfully explains model behaviour and attains plausibility only when appropriate (Section~\ref{emp}).
 
@@
 Many derivative works of~\citet{wachter2017counterfactual} have proposed new flavours of Equation~\ref{eq:general}, each of them designed to address specific \textit{desiderata} that counterfactuals ought to meet in order to properly serve both AI practitioners and individuals affected by algorithmic decision-making systems.
-The list of desiderata includes but is not limited to the following: sparsity, proximity~\citep{wachter2017counterfactual}, actionability~\citep{ustun2019actionable}, diversity~\citep{mothilal2020explaining}, plausibility~\citep{joshi2019realistic,poyiadzi2020face,schut2021generating}, robustness~\citep{upadhyay2021robust,pawelczyk2022probabilistically,altmeyer2023endogenous} and causality~\citep{karimi2021algorithmic}.
+The list of desiderata includes but is not limited to the following: sparsity, closeness~\citep{wachter2017counterfactual}, actionability~\citep{ustun2019actionable}, diversity~\citep{mothilal2020explaining}, plausibility~\citep{joshi2019realistic,poyiadzi2020face,schut2021generating}, robustness~\citep{upadhyay2021robust,pawelczyk2022probabilistically,altmeyer2023endogenous} and causality~\citep{karimi2021algorithmic}.
 Different counterfactual generators addressing these needs have been extensively surveyed and evaluated in various studies~\citep{verma2020counterfactual,karimi2020survey,pawelczyk2021carla,artelt2021evaluating,guidotti2022counterfactual}.
@@
 Similarly, plausibility has also been connected to causality in the sense that plausible counterfactuals respect causal relationships~\citep{mahajan2019preserving}.
-
 Consequently, the plausibility of counterfactuals has been among the primary concerns for researchers.
@@
 To learn the latent embedding, they propose using a generative model such as a Variational Autoencoder (VAE).
-Provided the surrogate model is well-specified, their proposed approach called \textit{REVISE} can yield plausible explanations.
+Provided the surrogate model is well-specified, their proposed approach \textit{REVISE} can yield plausible explanations.
 Others have proposed similar approaches: \citet{dombrowski2021diffeomorphic} traverse the base space of a normalizing flow to solve Equation~\ref{eq:general}; \citet{poyiadzi2020face} use density estimators ($\hat{p}: \mathcal{X} \mapsto [0,1]$) to constrain the counterfactuals to dense regions in the feature space; and, finally, \citet{karimi2021algorithmic} assume knowledge about the structural causal model that generates the data.
@@
 
-To demonstrate this with an example, we have trained a simple image classifier $M_{\theta}$ on the well-known \textit{MNIST} dataset~\citep{lecun1998mnist}: a Multi-Layer Perceptron (\textit{MLP}) with above 90 percent test accuracy.
+To demonstrate this with an example, we have trained a simple image classifier $M_{\theta}$ on the well-known \textit{MNIST} dataset~\citep{lecun1998mnist}: a Multi-Layer Perceptron (\textit{MLP}) with test set accuracy $> 0.9$.
 No measures have been taken to improve the model's adversarial robustness or its capacity for predictive uncertainty quantification.
@@
 The perturbed images are shown next to the factual image from left to right in Figure ~\ref{fig:motiv}.
-Captions on top of the individual images indicate the generator along with the predicted probability that the image belongs to the target class.
-In all three cases that probability is above 90 percent and yet the counterfactuals look very different from each other.
+Captions on top of the images indicate the generator along with the predicted probability that the image belongs to the target class.
+In all cases, that probability is very high, while the counterfactuals look very different.
 
@@
   \includegraphics[width=0.8\linewidth]{../artifacts/results/images/mnist_motivation.png}
-  \caption{Counterfactuals for turning a 9 (nine) into a 7 (seven): original image (left); then from left to right the counterfactuals generated using \textit{Wachter}, \textit{Schut} and \textit{REVISE}.}\label{fig:motiv}
+  \caption{Counterfactuals for turning a 9 (nine) into a 7 (seven): original image (left), then the counterfactuals generated using \textit{Wachter}, \textit{Schut} and \textit{REVISE}.}\label{fig:motiv}
 \end{figure}
 
-Since \textit{Wachter} is only concerned with proximity, the generated counterfactual is almost indistinguishable from the factual.
+Since \textit{Wachter} is only concerned with closeness, the generated counterfactual is almost indistinguishable from the factual.
 The approach by~\citet{schut2021generating} expects a well-calibrated model that can generate predictive uncertainty estimates.
@@
 While $\mathbf{x}_J$ is only guaranteed to distribute as $p_{\theta}(\mathbf{x}|\mathbf{y}^{+})$ if $\epsilon \rightarrow 0$ and $J \rightarrow \infty$, the bias introduced for a small finite $\epsilon$ is negligible in practice \citep{murphy2023probabilistic,grathwohl2020your}.
-Appendix~\ref{app:jem} provides additional implementation details for any tasks related to energy-based modelling.
 
 Generating multiple samples using SGLD thus yields an empirical distribution $\widehat{\mathbf{X}}_{\theta,\mathbf{y}^+}$ that approximates what the model has learned about the input data.
-While in the context of EBM, this is usually done during training, we propose to repurpose this approach during inference in order to evaluate and generate faithful model explanations.
+While in the context of EBM, this is usually done during training, we propose to repurpose this approach during inference in order to evaluate the faithfulness of model explanations.
+The technical appendix provides additional implementation details for any tasks related to energy-based modelling.
 
@@
 Faithful counterfactuals can be expected to also be plausible if the learned conditional distribution $\mathcal{X}_{\theta}|\mathbf{y}^+$ (Defintion~\ref{def:faithful}) is close to the true conditional distribution $\mathcal{X}|\mathbf{y}^+$ (Definition~\ref{def:plausible}).
-We can further improve plausibility of counterfactuals without the need for surrogate models that may interfer with faithfulness by minimizing predictive uncertainty~\citep{schut2021generating}.
-Unfortunately, this approach relies on the assumption that the model itself can provide predictive uncertainty estimates, which may be too restrictive in practice.
+We can further improve the plausibility of counterfactuals without the need for surrogate models that may interfere with faithfulness by minimizing predictive uncertainty~\citep{schut2021generating}.
+Unfortunately, this idea relies on the assumption that the model itself provides predictive uncertainty estimates, which may be too restrictive in practice.
 
@@
 Crucially for our intended application, CP is model-agnostic and can be applied during inference without placing any restrictions on model training.
-Intuitively, CP works under the premise of turning heuristic notions of uncertainty into rigorous uncertainty estimates by repeatedly sifting through the training data or a dedicated calibration dataset.
+It works under the premise of turning heuristic notions of uncertainty into rigorous estimates by repeatedly sifting through the training data or a dedicated calibration dataset.
 
@@
 
-Here, $\kappa \in \{0,1\}$ is a hyper-parameter and $C_{\theta,\mathbf{y}}(\mathbf{x}_i;\alpha)$ can be interpreted as the probability of label $\mathbf{y}$ being included in the prediction set (see Appendix~\ref{app:cp-diff} for details).
-In order to compute this penalty for any black-box model we merely need to perform a single calibration pass through a holdout set $\mathcal{D}_{\text{cal}}$.
+Here, $\kappa \in \{0,1\}$ is a hyper-parameter and $C_{\theta,\mathbf{y}}(\mathbf{x}_i;\alpha)$ can be interpreted as the probability of label $\mathbf{y}$ being included in the prediction set (see appendix for details).
+In order to compute this penalty for any black-box model, we merely need to perform a single calibration pass through a holdout set $\mathcal{D}_{\text{cal}}$.
 Arguably, data is typically abundant and in most applications, practitioners tend to hold out a test data set anyway.
 Consequently, CP removes the restriction on the family of predictive models, at the small cost of reserving a subset of the available data for calibration.
-This particular case of conformal prediction is referred to as *split conformal prediction* (SCP) as it involves splitting the training data into a proper training dataset and a calibration dataset.
+This particular case of conformal prediction is referred to as \textit{split conformal prediction} (SCP) as it involves splitting the training data into a proper training dataset and a calibration dataset.
 
@@
 It is based on the premise that counterfactuals should first and foremost be faithful.
-Plausibility, as a secondary concern, is then still attainable, but only to the degree that the black-box model itself has learned plausible explanations for the underlying data.
+Plausibility, as a secondary concern, is then still attainable to the degree that the black-box model itself has learned plausible explanations for the underlying data.
 
@@
 This is sufficient to capture the generative property of the underlying model since it is implicitly captured by its parameters $\theta$.
-Importantly, this means that we do not need to generate conditional samples through SGLD during our counterfactual search at all as we explain in Appendix~\ref{app:eccco}.
+Importantly, this means that we do not need to generate conditional samples through SGLD during our counterfactual search at all as we explain in the technical appendix.
 
@@
 
-\begin{figure}
+\begin{figure*}
   \centering
-  \includegraphics[width=1.0\linewidth]{../artifacts/results/images/poc_gradient_fields.png}
+  \includegraphics[width=0.75\linewidth]{../artifacts/results/images/poc_gradient_fields.png}
   \caption{Gradient fields and counterfactual paths for different generators.
-The objective is to generate a counterfual in the `blue' class for a sample from the `orange' class.
+The objective is to generate a counterfactual in the blue class for a sample from the orange class.
 Bright yellow stars indicate conditional samples generated through SGLD.
 The underlying classifier is a Joint Energy Model.}\label{fig:poc}
-\end{figure}
+\end{figure*}
 
 Figure~\ref{fig:poc} illustrates how the different components in Equation~\ref{eq:eccco} affect the counterfactual search for a synthetic dataset.
-The underlying classifier is a Joint Energy Model (\textit{JEM}) that was trained to predict the output class (`blue' or `orange') and generate class-conditional samples~\citep{grathwohl2020your}.
-We have used four different generator flavours to produce a counterfactual in the `blue' class for a sample from the `orange' class: \textit{Wachter}, which only uses the first penalty ($\lambda_2=\lambda_3=0$); \textit{ECCCo (no EBM)}, which does not constrain energy ($\lambda_2=0$); \textit{ECCCo (no CP)}, which involves no set size penalty ($\lambda_3=0$); and, finally, \textit{ECCCo}, which involves all penalties defined in Equation~\ref{eq:eccco}.
+The underlying classifier is a Joint Energy Model (\textit{JEM}) that was trained to predict the output class (blue or orange) and generate class-conditional samples~\citep{grathwohl2020your}.
+We have used four different generator flavours to produce a counterfactual in the blue class for a sample from the orange class: \textit{Wachter}, which only uses the first penalty ($\lambda_2=\lambda_3=0$); \textit{ECCCo (no EBM)}, which does not constrain energy ($\lambda_2=0$); \textit{ECCCo (no CP)}, which involves no set size penalty ($\lambda_3=0$); and, finally, \textit{ECCCo}, which involves all penalties defined in Equation~\ref{eq:eccco}.
 Arrows indicate (negative) gradients with respect to the objective function at different points in the feature space.
@@
 \textit{ECCCo (no EBM)} pushes the counterfactual further into the target domain to minimize predictive uncertainty, but the outcome is still not plausible.
-The counterfactual produced by \textit{ECCCo (no CP)} is attracted by the generated samples shown in bright yellow.
+The counterfactual produced by \textit{ECCCo (no CP)} is energy-constrained.
 Since the \textit{JEM} has learned the conditional input distribution reasonably well in this case, the counterfactuals are both faithful and plausible.
@@
 To account for stochasticity, we generate multiple counterfactuals for each target class, generator, model and dataset.
-Full details concerning our parameter choices, training procedures and model performance can be found in Appendix~\ref{app:setup}.
+Full details concerning our parameter choices, training procedures and model performance can be found in the appendix.
 
@@
 Across all models and datasets highlighted here, all variations of \textit{ECCCo} consistently outperform all other generators with respect to faithfulness, in many cases substantially.
-This pattern is mostly robust across all other benchmark datasets (Tables~\ref{tab:results-linearly-separable} to~\ref{tab:results-fashion-mnist}).
+This pattern is mostly robust across all other benchmark datasets (Tables~\ref{tab:results-linearly-separable} to~\ref{tab:results-fashion-mnist} in the technical appendix).
 
@@
 Our results here indicate that conformal prediction alone is often not sufficient to achieve faithfulness and plausibility.
-To test this initial finding more throughly, future work could benefit from more extensive abalation studies that thoroughly tune hyperparameters and investigate different approaches to conformal prediction.
+To test this initial finding more thoroughly, future work could benefit from more extensive ablation studies that thoroughly tune hyperparameters and investigate different approaches to conformal prediction.
 
@@
 This should enable researchers and practitioners to use counterfactuals in order to discern trustworthy models from unreliable ones.
-While the scope of this work limits its generalizability, we believe that \textit{ECCCo} offers a solid baseline for future work on faithful counterfactual explanations.
+While the scope of this work limits its generalizability, we believe that \textit{ECCCo} offers a solid base for future work on faithful counterfactual explanations.

### 2023-09-25 paper/body.tex
uh

@@
 To assess counterfactuals with respect to Definition~\ref{def:faithful}, we need a way to quantify the posterior conditional distribution $p_{\theta}(\mathbf{x}|\mathbf{y}^+)$.
-To this end, we draw on recent advances in energy-based modelling (EBM), a subdomain of machine learning that is concerned with generative or hybrid modelling~\citep{grathwohl2020your,du2019implicit}.
+To this end, we draw on ideas from energy-based modelling (EBM), a subdomain of machine learning that is concerned with generative or hybrid modelling~\citep{grathwohl2020your,du2019implicit}.
 In particular, note that if we fix $\mathbf{y}$ to our target value $\mathbf{y}^+$, we can conditionally draw from $p_{\theta}(\mathbf{x}|\mathbf{y}^+)$ by randomly initializing $\mathbf{x}_0$ and then using Stochastic Gradient Langevin Dynamics (SGLD) as follows,
@@
 To allow for faster sampling, we follow the common practice of choosing the step-size $\epsilon_j$ and the standard deviation of $\mathbf{r}_j$ separately.
-While $\mathbf{x}_J$ is only guaranteed to distribute as $p_{\theta}(\mathbf{x}|\mathbf{y}^{+})$ if $\epsilon \rightarrow 0$ and $J \rightarrow \infty$, the bias introduced for a small finite $\epsilon$ is negligible in practice \citep{murphy2023probabilistic,grathwohl2020your}.
+While $\mathbf{x}_J$ is only guaranteed to distribute as $p_{\theta}(\mathbf{x}|\mathbf{y}^{+})$ if $\epsilon \rightarrow 0$ and $J \rightarrow \infty$, the bias introduced for a small finite $\epsilon$ is negligible in practice \citep{murphy2023probabilistic}.
 
@@
 
-Conformal classifiers produce prediction sets for individual inputs that include all output labels that can be reasonably attributed to the input. Finally, classification sets are formed as follows,
+Conformal classifiers produce prediction sets for individual inputs that include all output labels that can be reasonably attributed to the input. These sets are formed as follows,
 
@@
 where $\hat{q}$ denotes the $(1-\alpha)$-quantile of $\mathcal{S}$ and $\alpha$ is a predetermined error rate.
-
 These sets tend to be larger for inputs that do not conform with the training data and are characterized by high predictive uncertainty.
@@
 The third and final penalty term involving $\lambda_3$ ensures that the generated counterfactual is associated with low predictive uncertainty.
+To tune theses hyperparameters we have relied on grid search.
 
@@
 
-\begin{question}[Balancing Objectives]\label{rq:plausibility}
+\begin{question}[Balancing Desiderata]\label{rq:plausibility}
   Compared to state-of-the-art generators, how does \textit{ECCCo} balance the two key objectives of faithfulness and plausibility?
@@
 In particular, we compare \textit{ECCCo} and its variants to the following counterfactual generators that were introduced above: firstly; \textit{Schut}, which works under the premise of minimizing predictive uncertainty; secondly, \textit{REVISE}, which is state-of-the-art (SOTA) with respect to plausibility; and, finally, \textit{Wachter}, which serves as our baseline.
-In the case of \textit{ECCCo+}, we use Principal Component Analysis for dimensionality reduction: the traversed latent space $\mathcal{Z}$ is spanned by the first $n_z$ principal components where we choose $n_z$ to be equal to the latent dimension of the VAE that is used by \textit{REVISE}.
+In the case of \textit{ECCCo+}, we use principal component analysis (PCA) for dimensionality reduction: the latent space $\mathcal{Z}$ is spanned by the first $n_z$ principal components where we choose $n_z$ to be equal to the latent dimension of the VAE used by \textit{REVISE}.
 
-For the predictive modelling tasks, we use multi-layer perceptrons (\textit{MLP}), deep ensembles, joint energy models (\textit{JEM}) and convolutional neural networks.
+For the predictive modelling tasks, we use multi-layer perceptrons (\textit{MLP}), deep ensembles, joint energy models (\textit{JEM}) and convolutional neural networks (\textit{CNN}).
 Both joint-energy modelling and ensembling have been associated with improved generative properties and adversarial robustness~\citep{grathwohl2020your,lakshminarayanan2016simple}, so we expect this to be positively correlated with the plausibility of \textit{ECCCo}.
@@
 We perform benchmarks on eight datasets from different domains.
-From the credit and finance domain we include three tabular datasets: Give Me Some Credit (\textit{GMSC})~\citep{kaggle2011give}, \textit{German Credit} and \textit{California Housing} (REFRENCES).
+From the credit and finance domain we include three tabular datasets: Give Me Some Credit (\textit{GMSC})~\citep{kaggle2011give}, \textit{German Credit}~\citet{hoffman1994german} and \textit{California Housing}~\citet{pace1997sparse}.
 All of these are commonly used in the related literature~\citep{karimi2020survey,altmeyer2023endogenous,pawelczyk2021carla}.
 Following~\citet{schut2021generating} we also include two image datasets: \textit{MNIST}~\citep{lecun1998mnist} and \textit{Fashion MNIST}~\citep{xiao2017fashion}.
-These are small image datasets but they allow us to easily visually gauge how plausible counterfactuals are.
-(EXPLAIN WHY NO LARGER?) Finally, we also benchmark our generators on multiple synthetic datasets.
-Detailed descriptions and results for all datasets can be found in the appendix (POINTER).
+Detailed descriptions and results for all datasets can be found in the appendix.
 
@@
 For the tabular datasets, we use the default Euclidian distance to measure unfaithfulness and implausibility as defined in Equations~\ref{eq:faith} and~\ref{eq:impl}, respectively.
-For the vision datasets, we instead rely on measuring the structural dissimilarity between images~\citep{wang2003multiscale}.
+The third metric presented~\ref{tab:results-tabular} in Table quantifies the predictive uncertainty of the counterfactual as measured by Equation~\ref{eq:setsize}.
+For the vision datasets, we rely on measuring the structural dissimilarity between images for our unfaithfulness and implausibility metrics~\citep{wang2003multiscale}.
 
@@
 
-\subsection{Balancing Objectives}
+\subsection{Balancing Desiderata}
 
@@
 Overall, we find strong empirical evidence suggesting that \textit{ECCCo} can achieve near state-of-the-art plausibility without sacrificing faithfulness.
-Figure~\ref{fig:mnist-bmk} shows one such example taken from the \textit{MNIST} benchmark.
-In this case, the objective was to turn the factual three (far left) into a five.
-The underlying model is a LeNet-5 convolutional neural network.
+Figure~\ref{fig:mnist-bmk} shows one such example taken from the \textit{MNIST} benchmark where the objective is to turn the factual three (far left) into a five.
+The underlying model is a LeNet-5 \textit{CNN}.
 The different images show the counterfactuals produced by the generators, of which all but the one produced by \textit{Schut} are valid.
@@
 For both tabular datasets in Table~\ref{tab:results-tabular}, \textit{ECCCo} improves plausibility substantially compared to the baseline.
-This broad pattern is mostly consistent for all other datasets, although there are notable exceptions for which \textit{ECCCo} takes the lead on both plausibility and faithfulness (see for example Tables~\ref{tab:results-moons} and~\ref{tab:results-german-credit}).
+This broad pattern is mostly consistent for all other datasets, although there are notable exceptions for which \textit{ECCCo} takes the lead on both plausibility and faithfulness (see, for example, Tables~\ref{tab:results-moons} and~\ref{tab:results-german-credit} in the appendix).
 
@@
 
-\subsection{Additional Observations}
+\subsection{Additional Desiderata}
+
+While we have deliberately focused on our key metrics of interest so far, it is worth briefly considering other common desiderata for counterfactuals.
+With reference to the right-most columns for each dataset in Table~\ref{tab:results-tabular}, we firstly note that \textit{ECCCo} typically reduces predictive uncertainty as intended.
+Consistent with its design, \textit{Schut} performs well on this metric even though it does not explicitly address uncertainty as measured by conformal prediction set sizes.
+
+Another commonly discussed desideratum is closeness~\citep{wachter2017counterfactual}: counterfactuals that are closer to their factuals are associated with smaller costs to individuals in the context of algorithmic recourse.
+As evident from the additional tables in the appendix, the closeness desideratum tends to be negatively correlated with plausibility and faithfulness.
+Consequently, both \textit{REVISE} and \textit{ECCCo} generally yield more costly counterfactuals than the baseline.
+Nonetheless, \textit{ECCCo} does not seem to stretch costs unnecessarily: in Figure~\ref{fig:mnist-bmk} useful parts of the factual three are clearly retained.
 
@@
 
-Even though we have taken considerable measures to study our proposed methodology carefully, limitations can still be identified.
+Despite having taken considerable measures to study our methodology carefully, limitations can still be identified.
 
-\subsection{Evaluation Metrics}
-
-Our proposed distance-based evaluation metrics for plausibility and faithfulness may not be universally applicable to all types of data.
-In any case, they depend on choosing a distance metric on a case-by-base basis for different datasets.
+Firstly, we recognise that our proposed distance-based evaluation metrics for plausibility and faithfulness may not be universally applicable to all types of data.
+In any case, they depend on choosing a distance metric on a case-by-case basis, as we have done in this work.
 Arguably, commonly used metrics for measuring other desiderata such as closeness suffer from the same pitfall.
@@
 
-\subsection{Experiments}
+Relatedly, we note that our proposed metric for measuring faithfulness depends on the availability of samples generated through SGLD, which in turn requires gradient access for models.
+This means it cannot be used to evaluate non-differentiable classifiers.
+Consequently, we also have not applied \textit{ECCCo} to some machine learning models commonly used for classification such as decision trees.
+Since \textit{ECCCo} itself does not rely on SGLD, its defining penalty functions are indeed applicable to gradient-free counterfactual generators.
+This is an interesting avenue for future research.
 
-While we have employed various datasets in our experiments that are commonly used in the related literature, we acknowledge that additional real-world data and application is needed to test \textit{ECCCo} and improve upon the ideas we have presented in this work.
-One challenge in this context is that counterfactual explanations do not scale very well to high-dimensional input data like images~\citep{samoilescu2021model,chen2021seven}.
-Consequently, we have limited ourselves to studying small image datasets only.
+Next, common challenges associated with energy-based modelling including sensitivity to scale, training instabilities and sensitivity to hyperparameters also apply to \textit{ECCCo} to some extent.
+In grid searches for optimal hyperparameters, we have noticed that unless properly regularized, \textit{ECCCo} is sometimes prone to overshoot for the energy constraint.
 
-\subsection{Generalizability}
-
-While our approach is readily applicable to models with gradient access like deep neural networks, more work is needed to generalise it to other machine learning models such as decision trees.
-Relatedly, common challenges associated with energy-based modelling including sensitivity to scale, training instabilities and sensitivity to hyperparameters also apply to \textit{ECCCo}.
-
-\subsection{Ablation Studies}
-
-In our experiments we have used ablation to understand the roles of the different components of \textit{ECCCo}.
-Our results here indicate that conformal prediction alone is often not sufficient to achieve faithfulness and plausibility.
-To test this initial finding more thoroughly, future work could benefit from more extensive ablation studies that thoroughly tune hyperparameters and investigate different approaches to conformal prediction.
+Finally, while we have used ablation to understand the roles of the different components of \textit{ECCCo}, the scope of this work has prevented us from investigating the role of conformal prediction in this context more thoroughly.
+We have exclusively relied on split conformal prediction and have used fixed values for the predetermined error rate and other hyperparameters.
+Future work could benefit from more extensive ablation studies that tune hyperparameters and investigate different approaches to conformal prediction.
 

### 2023-09-25 paper/body.tex
tired

@@
   \includegraphics[width=1.0\linewidth]{../www/mnist_benchmark.png}
-  \caption{Counterfactuals for turning a 3 into a 5: factual (left); then from left to right the counterfactuals generated using \textit{ECCCo}, \textit{ECCCo+}, \textit{REVISE}, \textit{Schut} and \textit{Wachter}.}\label{fig:mnist-bmk}
+  \caption{Counterfactuals for turning a 3 into a 5: factual (left), then the counterfactuals generated by \textit{ECCCo}, \textit{ECCCo+}, \textit{REVISE}, \textit{Schut} and \textit{Wachter}.}\label{fig:mnist-bmk}
 \end{figure}

### 2023-09-25 paper/aaai/author_response.qmd
author response

---
title: 'Author Response'
format: commonmark
bibliography: ../bib.bib
---

# Author Response

Following the reviews we received for NeurIPS we have taken substantial measures to address reviewer concerns both during the initial rebuttal period and beyond.

## Additional Datasets

A common concern across reviewers was limited evaluation on real-world datasets. While the scope of our initial experiments was already well in line with the existing related literature, we agreed with the reviewers and have added two additional tabular datasets from the social sciences domain as well as one additional dataset from the vision domain. As mentioned in the paper, we have resorted to datasets commonly used in the literature.

### A note on image datasets

Related work on plausibility of counterfactuals has largely relied on small image datasets like *MNIST* [@dhurandhar2018explanations,@schut2021generating]. This may be due to the fact that generating counterfactuals for high-dimensional input data is computationally very challenging. An exception to this rule is the work on *REVISE* [@joshi2019realistic], which uses a larger image dataset. *REVISE* is suitable for this task, because it maps counterfactuals to a lower-dimensional latent space. Similarly, our proposed *ECCCo+* should also be applicable to high-dimensional input data. In our benchmarks, however, we include other generators that search directly in the input space. Since our benchmarks required us to generate a very large number of counterfactuals, it was not at this time feasible to include larger image datasets. That is despite our best efforts to optimize the code and parallelize the computations through multi-threading and multi-processing on a high-performance computing cluster.

## Constraining Energy Directly

In our initial work we used our unfaithfulness metric directly as a penalty term in *ECCCo*'s counterfactual search objective. This generally achieves the highest levels of faithfulness but it has several disadvantages, some of which were pointed out by the reviewers. Our new approach constrains the energy directly, which is more theoretically grounded and leads to better results across the board. It also addresses the following reviewer concerns:

### Results are biased with respect to unfaithfulness metric

One reviewer raised concern about the fact the using the unfaithfulness metric as a penalty biases the results. This is a valid concern which we have addressed now.

### Counterfactuals looked homogeneous

Another reviewer pointed out that the counterfactuals generated by *ECCCo* looked homogeneous, which is also a valid concern. The observed homogeneity most likely stemmed form the fact that the samples generated through SGLD for the underlying models were fairly homogenous. With our new approach we no longer rely on SGLD samples and the homogeneity issue is no longer present.

### Closeness criterium violated

A related concern was that large perturbations induced by *ECCCo* seemed to violate the closeness criterium. As we discuss in the paper, our findings do not suggest that *ECCCo* yields unnecessarily costly counterfactuals. Indeed, with reference to the vision data, *ECCCo* seems to keep useful parts of the factual largely in tact, which reduces costs. As we already argued during the rebuttal and in the paper, additional costs cannot be avoided entirely when faithfulness and plausibility are prioritized. This applies to *ECCCo* as much as to other generators like *REVISE*.

### Sampling Overhead

1. Applied to additional commonly used tabular real-world datasets
2. Constraining energy directly
   1. Better results across the board, in particular for image data
   2. Derived from JEM loss function -> more theoretically grounded
   3. No sampling overhead.
   4. Energy does not depend on differentiability.
   5. Benchmarks no longer biased with respect to unfaithfulness metric (addressing reviewer concern).
3. Counterfactual explanations do not scale well to high-dimensional input data
   1. We have added native support for multi-processing and multi-threading.
   2. We have run more extensive experiments including fine-tuning hyperparameter choices.
   3. For image data we use PCA to map counterfactuals to a smaller dimenionsional latent space, which not only reduces costs of gradient computations but also leads to higher plausibility.
   4. PCA is much less costly and interventionist than a VAE: pricipal component merely represent variation in the data; nothing else about the data is learned by the surrogate. 
      1. ECCCo-$\Delta$ (latent) remains faithful, although not as faithful as standard ECCCo-$\Delta$.
4. We have revisited the mathematical notation.
5. We have moved the introduction of conformal prediction forward and added more detail in line with reviewer feedback.
6. We have extended the limitations section. 
7. Distance metric
   1. We have revisited the distance metrics and decided to use the L2 Norm for plausibility and faithfulness
   2. Orginially, we used the L1 Norm in line with how the the closeness criterium is commonly evaluated. But in this context the L1 Norm implicitly addresses the desire for sparsity.
   3. In the case of image data, we investigated various additional distance metrics:
      1. Cosine similarity
      2. Euclidean distance
      3. Ultimately we chose to rely on structural dissimilarity.

### 2023-09-25 paper/body.tex
author response

@@
 All of these are commonly used in the related literature~\citep{karimi2020survey,altmeyer2023endogenous,pawelczyk2021carla}.
-Following~\citet{schut2021generating} we also include two image datasets: \textit{MNIST}~\citep{lecun1998mnist} and \textit{Fashion MNIST}~\citep{xiao2017fashion}.
+Following related literature~\citep{schut2021generating,dhurandhar2018explanations} we also include two image datasets: \textit{MNIST}~\citep{lecun1998mnist} and \textit{Fashion MNIST}~\citep{xiao2017fashion}.
 Detailed descriptions and results for all datasets can be found in the appendix.

### 2023-09-25 paper/aaai/author_response.qmd

@@
 title: 'Author Response'
-format: commonmark
+format: 
+   pdf:
+      number-sections: true
+      number-depth: 1
 bibliography: ../bib.bib
 ---
-
-# Author Response
 
@@
 
-Related work on plausibility of counterfactuals has largely relied on small image datasets like *MNIST* [@dhurandhar2018explanations,@schut2021generating].
+Related work on plausibility of counterfactuals has largely relied on small image datasets like *MNIST* [@dhurandhar2018explanations;@schut2021generating;@delaney2023counterfactual].
 This may be due to the fact that generating counterfactuals for high-dimensional input data is computationally very challenging.
@@
 Our new approach constrains the energy directly, which is more theoretically grounded and leads to better results across the board.
-It also addresses the following reviewer concerns:
+Since it does not depend on generating samples through SGLD, our new approach is much more computationally efficient as well.
+Additionally, it also addresses the following reviewer concerns:
 
-### Results are biased with respect to unfaithfulness metric
+### Results were biased with respect to unfaithfulness metric
 
@@
 
-### Closeness criterium violated
+### Closeness criterium was violated
 
@@
 
-### Sampling Overhead
+### Generalizability
 
-1. Applied to additional commonly used tabular real-world datasets
-2. Constraining energy directly
-   1. Better results across the board, in particular for image data
-   2. Derived from JEM loss function -> more theoretically grounded
-   3. No sampling overhead.
-   4. Energy does not depend on differentiability.
-   5. Benchmarks no longer biased with respect to unfaithfulness metric (addressing reviewer concern).
-3. Counterfactual explanations do not scale well to high-dimensional input data
-   1. We have added native support for multi-processing and multi-threading.
-   2. We have run more extensive experiments including fine-tuning hyperparameter choices.
-   3. For image data we use PCA to map counterfactuals to a smaller dimenionsional latent space, which not only reduces costs of gradient computations but also leads to higher plausibility.
-   4. PCA is much less costly and interventionist than a VAE: pricipal component merely represent variation in the data; nothing else about the data is learned by the surrogate. 
-      1. ECCCo-$\Delta$ (latent) remains faithful, although not as faithful as standard ECCCo-$\Delta$.
-4. We have revisited the mathematical notation.
-5. We have moved the introduction of conformal prediction forward and added more detail in line with reviewer feedback.
-6. We have extended the limitations section. 
-7. Distance metric
-   1. We have revisited the distance metrics and decided to use the L2 Norm for plausibility and faithfulness
-   2. Orginially, we used the L1 Norm in line with how the the closeness criterium is commonly evaluated. But in this context the L1 Norm implicitly addresses the desire for sparsity.
-   3. In the case of image data, we investigated various additional distance metrics:
-      1. Cosine similarity
-      2. Euclidean distance
-      3. Ultimately we chose to rely on structural dissimilarity.
+This was not an explicit concern but some reviewers wondered if *ECCCo* could also be applied to non-differentiable models.
+While our initial approach that relied on SGLD samples was not suitable for non-differentiable models, our new approach is.
+This is because none of its penalties rely on differentiability.
+Of course, we still framed *ECCCo* in terms of gradient-based optimization, but the proposed penalties could be applied to other, non-gradient-based counterfactual generators as well such as *FeatureTweak*, for example [@tolomei2017interpretable].
+
+## Mathematical notation and concepts
+
+One reviewer complained about the mathematical notation, a concern that was not shared by any of the other reviewers.
+Nonetheless, we have revisited the notation and hope that it is now more clear.
+That same reviewer also raised concern about our definitions of plausibility and faithfulness that rely on distributional properties.
+We have extensively argued our case during the rebuttal and pointed to a potential reviwer misunderstanding in this context.
+None of the other reviewers found any issue with our definitions and we have made no changes in this regard.
+We did, however, make a minor change with respect to the related evaluation metrics.
+We are now more careful about our choice of the distance function.
+In particular, we investigated various distance metrics for image data and decided to rely on structural dissimilarity.
+For all other data we use the L2 Norm, where we previously used the L1 Norm.
+This has no impact on the results, but there was no obvious reason to use the L1 Norm in the first place other than the fact that it is typically used to assess closeness.
+
+## Conformal prediction was introduced too suddenly
+
+One reviewer pointed out that conformal prediction was introduced too suddenly. We have moved the introduction of conformal prediction forward and added more detail in line with reviewer feedback.
+
+## Limitations section
+
+We have extended the limitations section to address reviewer concerns.
+
+## Other improvements
+
+As discussed above, counterfactual explanations do not scale very well to high-dimensional input data.
+The NeurIPS feedback has motivated us to work on this issue by enabling intuitive support for multi-threading and multi-processing to our code.
+This has not only allowed us to include additional datasets but also to run extensive experiments to fine-tune hyperparameter choices.
+All of our code will be open-sourced as a package and we hope that it will be as useful to the community as was to us during our research.
+
+## References

### 2023-09-26 paper/aaai/author_response.qmd
tables updated

@@
 
+## Beyond JEMs
+
+Another common concern was that *ECCCo* primiarly achieved good results for JEMs.
+
 ## Mathematical notation and concepts
 
-One reviewer complained about the mathematical notation, a concern that was not shared by any of the other reviewers.
+One reviewer took issue with our mathematical notation, a concern that was not shared by any of the other reviewers.
 Nonetheless, we have revisited the notation and hope that it is now more clear.

### 2023-09-26 paper/tables.Rmd
Cynthia's comments addressed
tables updated

@@
 Best outcomes are highlighted in bold.
-Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (Wachter).
+Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (\\textit{Wachter}).
 \\label{tab:results-%s} \\newline",
@@
 ```{r}
-files <- list.files("artifacts/params/")
-dt <- lapply(files[grepl(".csv", files)], function(x) {
-    fread(file.path("artifacts/params/", x))
+files <- list.files(paste0(res_path,"params"))
+dt <- lapply(files[sapply(files, function(x) grepl("model_params.csv",x))], function(x) {
+    fread(file.path(paste0(res_path,"params"), x))
 })
@@
 )
-dt[,dataname:=factor(dataname, levels=c("Linearly Separable", "Moons", "Circles", "MNIST", "GMSC"))]
+dataset_order = c(
+  "Linearly Separable", "Moons", "Circles", 
+  "California Housing", "GMSC", "German Credit",
+  "MNIST", "Fashion MNIST"
+)
+dt[,dataname:=factor(dataname, levels=dataset_order)]
 dt <- dt[order(dataname)]
@@
   align = "r", col.names=col_names, booktabs = T, escape=F, 
-  format="latex"
+  format="latex", linesep = ""
 ) %>%
@@
 ```{r}
-files <- list.files("artifacts/params/generator")
-dt <- lapply(files, function(x) {
-    fread(file.path("artifacts/params/generator", x))
+dt <- lapply(files[sapply(files, function(x) grepl("generator_params.csv",x))], function(x) {
+    fread(file.path(paste0(res_path,"params"), x))
 })
 dt <- Reduce(function(x,y) {rbind(x,y, fill=TRUE)}, dt)
-dt <- dt[,.(dataname,eta,λ1,λ3,λ3)]
-dt[,dataname:=factor(dataname, levels=c("Linearly Separable", "Moons", "Circles", "MNIST", "GMSC"))]
+dt <- dt[,.(dataname,eta,lambda_1,lambda_2,lambda_3)]
+dt[,dataname:=factor(dataname, levels=dataset_order)]
 dt <- dt[order(dataname)]
@@
 ```{r}
-files <- list.files("artifacts/results/")
 dt <- lapply(files[grepl("_model_performance.csv", files)], function(x) {
-    fread(file.path("artifacts/results/", x))
+    fread(file.path(paste0(res_path,"params"), x))
 })
 dt <- Reduce(function(x,y) {rbind(x,y, fill=TRUE)}, dt)
-dt[,dataname:=factor(dataname, levels=c("Linearly Separable", "Moons", "Circles", "MNIST", "GMSC"))]
+dt[,dataname:=factor(dataname, levels=dataset_order)]
 dt <- dt[order(dataname,mod_name)]
@@
   align = "r", col.names=col_names, booktabs = T, escape=F,
-  format="latex", digits=2
+  format="latex", digits=2, linesep = ""
 ) %>%

### 2023-09-26 experiments/post_processing/plotting.jl

@@ -1,3 +1,5 @@
+using Plots
+
 function choose_random_mnist(outcome::ExperimentOutcome; model::String="LeNet-5", img_height=125, seed=966)
 
     # Set seed:
@@ -105,4 +107,57 @@ function plot_random_eccco(outcome::ExperimentOutcome; ce=nothing, generator="EC
     display(plt)
 
     return plt, target, seed
+end
+
+function plot_all_mnist(gen, model, data=load_mnist_test(); img_height=150, seed=123, maxoutdim=64)
+
+    # Set seed:
+    if !isnothing(seed)
+        Random.seed!(seed)
+    end
+
+    # Dimensionality reduction:
+    data.dt = MultivariateStats.fit(MultivariateStats.PCA, data.X; maxoutdim=maxoutdim)
+
+    targets = [0, 1, 2, 3, 4, 5, 6, 7, 8, 9]
+    factuals = targets
+    plts = []
+
+    for factual in factuals
+        chosen = rand(findall(predict_label(model, data) .== factual))
+        x = select_factual(data, chosen)
+        for target in targets
+            if factual != target
+                @info "Generating counterfactual for $(factual) -> $(target)"
+                ce = generate_counterfactual(
+                    x, target, data, model, gen; 
+                    initialization=:identity, converge_when=:generator_conditions
+                )
+                plt = Plots.plot(
+                    CounterfactualExplanations.counterfactual(ce) |> ECCCo.convert2mnist,
+                    axis=([], false),
+                    size=(img_height, img_height),
+                    title="$factual → $target",
+                )
+            else
+                plt = Plots.plot(
+                    x |> ECCCo.convert2mnist,
+                    axis=([], false),
+                    size=(img_height, img_height),
+                    title="Factual",
+                )
+            end
+            push!(plts, plt)
+        end
+    end
+
+    plt = Plots.plot(
+        plts...,
+        layout=(length(factuals), length(targets)),
+        size=(img_height * length(targets), img_height * length(factuals)),
+        dpi=300
+    )
+
+    return plt
+
 end

### 2023-09-26 paper/appendix.tex
mostly done
nearly there now

@@
   \begin{aligned}
-    \hat{\mathbf{x}}_{j+1} &\leftarrow \hat{\mathbf{x}}_j - \frac{\epsilon}{2} \mathcal{E}_{\theta}(\hat{\mathbf{x}}_j|\mathbf{y}^+) + \sigma \mathbf{r}_j, && j=1,...,J
+    \hat{\mathbf{x}}_{j+1} &\leftarrow \hat{\mathbf{x}}_j - \frac{\phi}{2} \mathcal{E}_{\theta}(\hat{\mathbf{x}}_j|\mathbf{y}^+) + \sigma \mathbf{r}_j, && j=1,...,J
   \end{aligned}
@@
 
-Consistent with~\citet{grathwohl2020your}, we have specified $\epsilon=2$ and $\sigma=0.01$ as the default values for all of our experiments.
+Consistent with~\citet{grathwohl2020your}, we have specified $\phi=2$ and $\sigma=0.01$ as the default values for all of our experiments.
+Here we have deliberately departed slightly from the notation in Equation~\ref{eq:sgld} to emphasize that we use fixed values for $\phi$ and $\sigma$, consistent with the related literature.
 The number of total SGLD steps $J$ varies by dataset (Table~\ref{tab:ebmparams}).
@@
 
-\subsubsection{Compute}
+\subsection{Compute}
 
 To enable others to easily replicate our experiments, we have chosen to work with small neural network architectures and randomly undersampled the \textit{MNIST} dataset (maintaining class balance).
-All of our experiments could then be run locally on a personal machine.
-The longest runtimes we experienced for model training and counterfactual benchmarking were on the order of 8-12 hours (\textit{MNIST} data).
-For the synthetic data, all experiments could be completed in less than an hour.
+All of our final benchmarks could then be run locally on a personal machine, but grid searches and the sheer number of datasets required us to move to high-performance computing clusters to conduct our experiments efficiently.
 
+\subsubsection{Local runs}
+
+The longest runtimes we experienced for model training and counterfactual benchmarking for a single benchmark were on the order of 12-24 hours (\textit{MNIST} data).
+For the synthetic data, single benchmarks could be completed in less than an hour.
 We have summarised our system information below:
@@
 
+\subsubsection{HPC}
+
+We used two large clusters, which we do not cite or mention by name at this point to not interfer with the double-blind review process.
+All of our grid searches were multi-processed on 100 CPUs at 4-8GB or memory each.
+
 \subsection{Results}\label{app:results}
@@
 
-Table~\ref{tab:results-full} reports all of the evaluation metrics we have computed.
-Table~\ref{tab:results-full-valid} reports the same metrics for the subset of valid counterfactuals.
+Tables~\ref{tab:results-linearly-separable} to~\ref{tab:results-fashion-mnist} reports all of the evaluation metrics we have computed.
+Tables~\ref{tab:results-linearly-separable-valid} to~\ref{tab:results-fashion-mnist-valid} reports the same metrics for the subset of valid counterfactuals.
 The `Unfaithfulness' and `Implausibility' metrics have been discussed extensively in the body of the paper.
-The `Cost' metric relates to the distance between the factual and the counterfactual.
+The `Cost' metric relates to the distance between the factual and the counterfactual and is measured using the L1 Norm.
 The `Redundancy' metric measures sparsity in is defined as the percentage of features that remain unperturbed (higher is better).
@@
 
-\import{contents/}{table_all.tex}
+\import{contents/}{table-linearly-separable.tex}
 
-\import{contents/}{table_all_valid.tex}
+\import{contents/}{table-circles.tex}
+
+\import{contents/}{table-moons.tex}
+
+\import{contents/}{table-california-housing.tex}
+
+\import{contents/}{table-gmsc.tex}
+
+\import{contents/}{table-german-credit.tex}
+
+\import{contents/}{table-mnist.tex}
+
+\import{contents/}{table-fashion-mnist.tex}
+
+\import{contents/}{table-linearly-separable-valid.tex}
+
+\import{contents/}{table-circles-valid.tex}
+
+\import{contents/}{table-moons-valid.tex}
+
+\import{contents/}{table-california-housing-valid.tex}
+
+\import{contents/}{table-gmsc-valid.tex}
+
+\import{contents/}{table-german-credit-valid.tex}
+
+\import{contents/}{table-mnist-valid.tex}
+
+\import{contents/}{table-fashion-mnist-valid.tex}

### 2023-09-26 src/penalties.jl

@@ -64,9 +64,6 @@ function energy_delta(
     reg_loss = norm(E(xproposed))^2
     reg_loss = reduce((x, y) -> x + y, reg_loss) / length(reg_loss)                  # aggregate over samples
 
-    # Decay:
-    λ = exp(-decay * ce.search[:iteration_count])
-
     return gen_loss + reg_strength * reg_loss
 
 end

### 2023-09-26 paper/appendix.tex
all images

@@
 For the sake of simplicity and without loss of generality, we limit our attention to generating a single counterfactual $\mathbf{x}^\prime=f(\mathbf{z}^\prime)$.
+The counterfactual state $\mathbf{z}^\prime$ is initialized at the factual $\mathbf{x}$.
+Other forms of initialization are also suitable but not considered here.
+For example, one may choose at a small random perturbation to all features~\citep{slack2021counterfactual}.
+Next, we calibrate the model $M_{\theta}$ through split conformal prediction.
+Finally, we search counterfactuals through gradient descent where $\mathcal{L}(\mathbf{z}^\prime,\mathbf{y}^+,\widehat{\mathbf{X}}_{\theta,\mathbf{y}^+}; \Lambda, \alpha)$ denotes our loss function defined in Equation~\ref{eq:eccco}.
+The search terminates once the convergence criterium is met or the maximum number of iterations $T$ has been exhausted.
+Note that the choice of convergence criterium has important implications on the final counterfactual which we explain below.
+
+\begin{algorithm*}[h]
+  \caption{The \textit{ECCCo} generator}\label{alg:eccco}
+  \begin{algorithmic}[1]
+    \Require $\mathbf{x}, \mathbf{y}^+, M_{\theta}, \Lambda=[\lambda_1,\lambda_2,\lambda_3], \alpha, \mathcal{D}, T$ where $M_{\theta}(\mathbf{x})\neq\mathbf{y}^+$
+    \Ensure $\mathbf{x}^\prime$
+    \State Initialize $\mathbf{z}^\prime \gets \mathbf{x}$ 
+    \State Run \textit{SCP} for $M_{\theta}$ using $\mathcal{D}$ \Comment{Calibrate model through split conformal prediction.}
+    \State Initialize $t \gets 0$
+    \While{\textit{not converged} or $t < T$} \Comment{For convergence conditions see below.}
+    \State $\mathbf{z}^\prime \gets \mathbf{z}^\prime - \eta \nabla_{\mathbf{z}^\prime} \mathcal{L}(\mathbf{z}^\prime,\mathbf{y}^+,\widehat{\mathbf{X}}_{\theta,\mathbf{y}^+}; \Lambda, \alpha)$ \Comment{Take gradient step of size $\eta$.}
+    \State $t \gets t+1$
+    \EndWhile
+    \State $\mathbf{x}^\prime \gets \mathbf{z}^\prime$ \Comment{Map back to feature space.}
+  \end{algorithmic}
+\end{algorithm*}
+
+\subsubsection{The \textit{ECCCo+} algorithm}
+
+Algorithm~\ref{alg:eccco-plus} describes how exactly \textit{ECCCo+} works. The only difference to \textit{ECCCo} is that we encode and decode features using PCA.
+
+\begin{algorithm*}[h]
+  \caption{The \textit{ECCCo+} generator}\label{alg:eccco-plus}
+  \begin{algorithmic}[1]
+    \Require $\mathbf{x}, \mathbf{y}^+, M_{\theta}, f, \Lambda=[\lambda_1,\lambda_2,\lambda_3], \alpha, \mathcal{D}, T$ where $M_{\theta}(\mathbf{x})\neq\mathbf{y}^+$
+    \Ensure $\mathbf{x}^\prime$
+    \State Initialize $\mathbf{z}^\prime \gets f^{-1}(\mathbf{x})$ \Comment{Map to counterfactual state space.}
+    \State Run \textit{SCP} for $M_{\theta}$ using $\mathcal{D}$ \Comment{Calibrate model through split conformal prediction.}
+    \State Initialize $t \gets 0$
+    \While{\textit{not converged} or $t < T$} \Comment{For convergence conditions see below.}
+    \State $\mathbf{z}^\prime \gets \mathbf{z}^\prime - \eta \nabla_{\mathbf{z}^\prime} \mathcal{L}(\mathbf{z}^\prime,\mathbf{y}^+,\widehat{\mathbf{X}}_{\theta,\mathbf{y}^+}; \Lambda, \alpha)$ \Comment{Take gradient step of size $\eta$.}
+    \State $t \gets t+1$
+    \EndWhile
+    \State $\mathbf{x}^\prime \gets f(\mathbf{z}^\prime)$ \Comment{Map back to feature space.}
+  \end{algorithmic}
+\end{algorithm*}
+
+\subsubsection{The \textit{ECCCo-L1} algorithm}
+
+Algorithm~\ref{alg:eccco} describes how exactly \textit{ECCCo} works.
+For the sake of simplicity and without loss of generality, we limit our attention to generating a single counterfactual $\mathbf{x}^\prime=f(\mathbf{z}^\prime)$.
 The counterfactual state $\mathbf{z}^\prime$ is initialized by passing the factual $\mathbf{x}$ through a simple feature transformer $f^{-1}$.
@@
 
-\begin{algorithm*}
-  \caption{The \textit{ECCCo} generator}\label{alg:eccco}
+\begin{algorithm*}[h]
+  \caption{The \textit{ECCCo-L1} generator}\label{alg:eccco-l1}
   \begin{algorithmic}[1]
@@
 
-\subsubsection{A Note on Convergence}
+\subsubsection{A Note on Convergence}\label{convergence}
 

### 2023-09-26 paper/appendix.tex
uh

@@
     \While{\textit{not converged} or $t < T$} \Comment{For convergence conditions see below.}
-    \State $\mathbf{z}^\prime \gets \mathbf{z}^\prime - \eta \nabla_{\mathbf{z}^\prime} \mathcal{L}(\mathbf{z}^\prime,\mathbf{y}^+,\widehat{\mathbf{X}}_{\theta,\mathbf{y}^+}; \Lambda, \alpha)$ \Comment{Take gradient step of size $\eta$.}
+    \State $\mathbf{z}^\prime \gets \mathbf{z}^\prime - \eta \nabla_{\mathbf{z}^\prime} \mathcal{L}(\mathbf{z}^\prime,\mathbf{y}^+; \Lambda, \alpha)$ \Comment{Take gradient step of size $\eta$.}
     \State $t \gets t+1$
     \EndWhile
-    \State $\mathbf{x}^\prime \gets \mathbf{z}^\prime$ \Comment{Map back to feature space.}
+    \State $\mathbf{x}^\prime \gets \mathbf{z}^\prime$
   \end{algorithmic}
@@
 
-Algorithm~\ref{alg:eccco-plus} describes how exactly \textit{ECCCo+} works. The only difference to \textit{ECCCo} is that we encode and decode features using PCA.
+Algorithm~\ref{alg:eccco-plus} describes how exactly \textit{ECCCo+} works.
+The only difference to \textit{ECCCo} is that we encode and decode features using PCA.
+In particular, we let $f^{-1}(\mathbf{x})$ denote the projection of $\mathbf{x}$ to its first $n_z$ principal components.
+Conversely, $f(\mathbf{z}^\prime)$ maps back from the projection to the feature space.
 
@@
     \While{\textit{not converged} or $t < T$} \Comment{For convergence conditions see below.}
-    \State $\mathbf{z}^\prime \gets \mathbf{z}^\prime - \eta \nabla_{\mathbf{z}^\prime} \mathcal{L}(\mathbf{z}^\prime,\mathbf{y}^+,\widehat{\mathbf{X}}_{\theta,\mathbf{y}^+}; \Lambda, \alpha)$ \Comment{Take gradient step of size $\eta$.}
+    \State $\mathbf{z}^\prime \gets \mathbf{z}^\prime - \eta \nabla_{\mathbf{z}^\prime} \mathcal{L}(\mathbf{z}^\prime,\mathbf{y}^+; \Lambda, \alpha)$ \Comment{Take gradient step of size $\eta$.}
     \State $t \gets t+1$
@@
 
-Algorithm~\ref{alg:eccco} describes how exactly \textit{ECCCo} works.
-For the sake of simplicity and without loss of generality, we limit our attention to generating a single counterfactual $\mathbf{x}^\prime=f(\mathbf{z}^\prime)$.
-The counterfactual state $\mathbf{z}^\prime$ is initialized by passing the factual $\mathbf{x}$ through a simple feature transformer $f^{-1}$.
+Algorithm~\ref{alg:eccco-l1} describes a variation of \textit{ECCCo} that we initally considered but ultimately discarded.
+For the sake of completeness we have included this approach here in the appendix.
+It generally yields very faithful counterfactuals but it is computationally much more expensive and struggles with plausibility.
+
+Instead of constraining energy directly, this approach works under the premise of penalizing the distance between the counterfactual and samples generated through SGLD.
+The counterfactual state $\mathbf{z}^\prime$ is initialized as in Algorithm~\ref{alg:eccco}.
 Next, we generate $n_{\mathcal{B}}$ conditional samples $\hat{\mathbf{x}}_{\theta,\mathbf{y}^+}$ using SGLD (Equation~\ref{eq:sgld}) and store the $n_E$ instances with the lowest energy.
 We then calibrate the model $M_{\theta}$ through split conformal prediction.
-Finally, we search counterfactuals through gradient descent where $\mathcal{L}(\mathbf{z}^\prime,\mathbf{y}^+,\widehat{\mathbf{X}}_{\theta,\mathbf{y}^+}; \Lambda, \alpha)$ denotes our loss function defined in Equation~\ref{eq:eccco}.
-The search terminates once the convergence criterium is met or the maximum number of iterations $T$ has been exhausted.
-Note that the choice of convergence criterium has important implications on the final counterfactual which we explain in Appendix~\ref{app:eccco}.
+Finally, we search counterfactuals through gradient descent where $\mathcal{L}(\mathbf{z}^\prime,\mathbf{y}^+,\widehat{\mathbf{X}}_{\theta,\mathbf{y}^+}; \Lambda, \alpha)$ denotes our loss function defined in Equation~\ref{eq:eccco}, but instead of constraining energy directly we use Equaqtion~\ref{eq:faith} (unfaithfulness metric) as a penalty.
 
@@
     \Ensure $\mathbf{x}^\prime$
-    \State Initialize $\mathbf{z}^\prime \gets f^{-1}(\mathbf{x})$ \Comment{Map to counterfactual state space.}
+    \State Initialize $\mathbf{z}^\prime \gets \mathbf{x}$ 
     \State Generate $\left\{\hat{\mathbf{x}}_{\theta,\mathbf{y}^+}\right\}_{n_{\mathcal{B}}} \gets p_{\theta}(\mathbf{x}_{\mathbf{y}^+})$ \Comment{Generate $n_{\mathcal{B}}$ samples using SGLD (Equation~\ref{eq:sgld}).}
@@
     \State Initialize $t \gets 0$
-    \While{\textit{not converged} or $t < T$} \Comment{For convergence conditions see Appendix~\ref{app:eccco}.}
+    \While{\textit{not converged} or $t < T$} \Comment{For convergence conditions see below.}
     \State $\mathbf{z}^\prime \gets \mathbf{z}^\prime - \eta \nabla_{\mathbf{z}^\prime} \mathcal{L}(\mathbf{z}^\prime,\mathbf{y}^+,\widehat{\mathbf{X}}_{\theta,\mathbf{y}^+}; \Lambda, \alpha)$ \Comment{Take gradient step of size $\eta$.}
@@
     \EndWhile
-    \State $\mathbf{x}^\prime \gets f(\mathbf{z}^\prime)$ \Comment{Map back to feature space.}
+    \State $\mathbf{x}^\prime \gets \mathbf{z}^\prime$
   \end{algorithmic}
@@
 
-Figure~\ref{fig:mnist-eccco} shows examples of counterfactuals for \textit{MNIST} data where the underlying model is our \textit{JEM Ensemble}.
+Figures~\ref{fig:mnist-eccco-lenet} to~\ref{fig:mnist-eccco-jem-ens} show examples of counterfactuals for \textit{MNIST} generated by \textit{ECCCo+} for our different models.
 Original images are shown on the diagonal and the corresponding counterfactuals are plotted across rows.
+Figures~\ref{fig:mnist-revise-lenet} to~\ref{fig:mnist-revise-jem-ens} show the same examples but for \textit{REVISE}.
+Both counterfactual generators have access to the same optimizer.
+While the results for \textit{REVISE} look fairly poor here, we have observed better results for optizers with higher step sizes.
 
@@
   \centering
-  \includegraphics[width=0.9\linewidth]{../artifacts/results/images/mnist_eccco_all_digits.png}
-  \caption{Counterfactuals for \textit{MNIST} data and our \textit{JEM Ensemble}.
-Original images are shown on the diagonal with the corresponding counterfactuals plotted across rows.}\label{fig:mnist-eccco}
+  \includegraphics[width=1.0\linewidth]{../www/mnist_all_lenet_eccco.png}
+  \caption{Counterfactuals for \textit{MNIST} data generated by \textit{ECCCo+}.
+The underlying model is a LeNet-5 \textit{CNN}.
+Original images are shown on the diagonal with the corresponding counterfactuals plotted across rows.}\label{fig:mnist-eccco-lenet}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=1.0\linewidth]{../www/mnist_all_mlp_eccco.png}
+  \caption{Counterfactuals for \textit{MNIST} data generated by \textit{ECCCo+}.
+The underlying model is an \textit{MLP}.
+Original images are shown on the diagonal with the corresponding counterfactuals plotted across rows.}\label{fig:mnist-eccco-mlp}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=1.0\linewidth]{../www/mnist_all_mlp_ens_eccco.png}
+  \caption{Counterfactuals for \textit{MNIST} data generated by \textit{ECCCo+}.
+The underlying model is an \textit{MLP} ensemble.
+Original images are shown on the diagonal with the corresponding counterfactuals plotted across rows.}\label{fig:mnist-eccco-mlp-ens}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=1.0\linewidth]{../www/mnist_all_jem_eccco.png}
+  \caption{Counterfactuals for \textit{MNIST} data generated by \textit{ECCCo+}.
+The underlying model is a \textit{JEM}.
+Original images are shown on the diagonal with the corresponding counterfactuals plotted across rows.}\label{fig:mnist-eccco-jem}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=1.0\linewidth]{../www/mnist_all_jem_ens_eccco.png}
+  \caption{Counterfactuals for \textit{MNIST} data generated by \textit{ECCCo+}.
+The underlying model is a \textit{JEM} ensemble.
+Original images are shown on the diagonal with the corresponding counterfactuals plotted across rows.}\label{fig:mnist-eccco-jem-ens}
+\end{figure}
+
+% REVISE
+
+\begin{figure}
+  \centering
+  \includegraphics[width=1.0\linewidth]{../www/mnist_all_lenet_revise.png}
+  \caption{Counterfactuals for \textit{MNIST} data generated by \textit{REVISE}.
+The underlying model is a LeNet-5 \textit{CNN}.
+Original images are shown on the diagonal with the corresponding counterfactuals plotted across rows.}\label{fig:mnist-revise-lenet}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=1.0\linewidth]{../www/mnist_all_mlp_revise.png}
+  \caption{Counterfactuals for \textit{MNIST} data generated by \textit{REVISE}.
+The underlying model is an \textit{MLP}.
+Original images are shown on the diagonal with the corresponding counterfactuals plotted across rows.}\label{fig:mnist-revise-mlp}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=1.0\linewidth]{../www/mnist_all_mlp_ens_revise.png}
+  \caption{Counterfactuals for \textit{MNIST} data generated by \textit{REVISE}.
+The underlying model is an \textit{MLP} ensemble.
+Original images are shown on the diagonal with the corresponding counterfactuals plotted across rows.}\label{fig:mnist-revise-mlp-ens}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=1.0\linewidth]{../www/mnist_all_jem_revise.png}
+  \caption{Counterfactuals for \textit{MNIST} data generated by \textit{REVISE}.
+The underlying model is a \textit{JEM}.
+Original images are shown on the diagonal with the corresponding counterfactuals plotted across rows.}\label{fig:mnist-revise-jem}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=1.0\linewidth]{../www/mnist_all_jem_ens_revise.png}
+  \caption{Counterfactuals for \textit{MNIST} data generated by \textit{REVISE}.
+The underlying model is a \textit{JEM} ensemble.
+Original images are shown on the diagonal with the corresponding counterfactuals plotted across rows.}\label{fig:mnist-revise-jem-ens}
 \end{figure}

### 2023-09-26 paper/appendix.tex
almost

@@
 While the results for \textit{REVISE} look fairly poor here, we have observed better results for optizers with higher step sizes.
+Note that the seemingly poor performance by \textit{REVISE} upon visual inspection is not driven by a weak surrogate VAE: Figure~\ref{fig:vae-reg} shows image reconstructions generated by the VAE.
+
+% ECCCo params λ=[0.1,0.5,1.0], reg_strength=0.0
 
@@
 
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=1.0\linewidth]{../www/vae_rec.png}
+  \caption{Randomly drawn \textit{MNIST} images and their reconstructions generated by the VAE used by \textit{REVISE}.}\label{fig:vae-rec}
+\end{figure}
+
 Tables~\ref{tab:results-linearly-separable} to~\ref{tab:results-fashion-mnist} reports all of the evaluation metrics we have computed.

### 2023-09-26 paper/body.tex
uh
almost

@@
 
-For the predictive modelling tasks, we use multi-layer perceptrons (\textit{MLP}), deep ensembles, joint energy models (\textit{JEM}) and convolutional neural networks (\textit{CNN}).
+For the predictive modelling tasks, we use multi-layer perceptrons (\textit{MLP}), deep ensembles, joint energy models (\textit{JEM}) and convolutional neural networks (LeNet-5 \textit{CNN}).
 Both joint-energy modelling and ensembling have been associated with improved generative properties and adversarial robustness~\citep{grathwohl2020your,lakshminarayanan2016simple}, so we expect this to be positively correlated with the plausibility of \textit{ECCCo}.
@@
 Looking at the benchmark results presented in Tables~\ref{tab:results-tabular} and~\ref{tab:results-vision} we firstly note that although \textit{REVISE} generally performs best, \textit{ECCCo} and in particular \textit{ECCCo+} often approach SOTA performance.
+Upon visual inspection of the generated images we actually find that \textit{ECCCo+} performs much better than \textit{REVISE} (see appendix).
 Zooming in on the details we observe that \textit{ECCCo} and its variations do particularly well, whenever the underlying model has been explicitly trained to learn plausible representations of the data.
@@
 
-This work leverages recent advances in energy-based modelling and conformal prediction in the context of Explainable Artificial Intelligence.
+This work leverages ideas from energy-based modelling and conformal prediction in the context of counterfactual explanations.
 We have proposed a new way to generate counterfactuals that are maximally faithful to the black-box model they aim to explain.
-Our proposed generator, \textit{ECCCo}, produces plausible counterfactuals if and only if the black-box model itself has learned realistic explanations for the data, which we have demonstrated through rigorous empirical analysis.
+Our proposed generator, \textit{ECCCo}, produces plausible counterfactuals iff the black-box model itself has learned realistic explanations for the data, which we have demonstrated through rigorous empirical analysis.
 This should enable researchers and practitioners to use counterfactuals in order to discern trustworthy models from unreliable ones.

### 2023-09-26 paper/body.tex
done

@@
 
-For the predictive modelling tasks, we use multi-layer perceptrons (\textit{MLP}), deep ensembles, joint energy models (\textit{JEM}) and convolutional neural networks (LeNet-5 \textit{CNN}).
+For the predictive modelling tasks, we use multi-layer perceptrons (\textit{MLP}), deep ensembles, joint energy models (\textit{JEM}) and convolutional neural networks (LeNet-5 \textit{CNN}~\citep{lecun1998gradient}).
 Both joint-energy modelling and ensembling have been associated with improved generative properties and adversarial robustness~\citep{grathwohl2020your,lakshminarayanan2016simple}, so we expect this to be positively correlated with the plausibility of \textit{ECCCo}.

### 2023-09-26 paper/appendix.tex
done

@@
 
+In our experiments we always generate multiple counterfactuals for each model and generator.
+Each time the factual and target class is drawn randomly.
+For each generator and model we choose $n_f=100$ factuals for all of our synthetic and vision data.
+For tabular data we choose $n_f=25$ because larger values made grid search computationally prohibitive.
+For vision data, grid search was computationally prohibitve in any case, so hyperparameters were tuned manually.
+For all other datasets grid search was performed over different combinations of penalty strengths and optimizer steps sizes.
+
 Table~\ref{tab:params} provides an overview of all parameters related to our experiments.

### 2023-09-26 experiments/post_processing/plotting.jl

@@ -164,3 +164,41 @@ function plot_all_mnist(gen, model, data=load_mnist_test(); img_height=150, seed
     return plt
 
 end
+
+using MLDatasets
+using MosaicViews
+function vae_reconstructions(seed=123)
+
+    # Set seed:
+    if !isnothing(seed)
+        Random.seed!(seed)
+    end
+
+    counterfactual_data = load_mnist()
+    counterfactual_data.generative_model = CounterfactualExplanations.Models.load_mnist_vae()
+    X = counterfactual_data.X
+    y = counterfactual_data.output_encoder.y  
+    images = []
+    rec_images = []
+    for i in 0:9
+        j = 0
+        while j < 10
+            x = X[:,rand(findall(y .== i))]
+            x̂ = CounterfactualExplanations.GenerativeModels.reconstruct(vae, x)[1] |> 
+                x̂ -> clamp.((x̂ .+ 1.0) ./ 2.0, 0.0, 1.0) |>
+                x̂ -> reshape(x̂, 28,28) |>
+                x̂ -> MLDatasets.convert2image(MNIST, x̂)
+            x = clamp.((x .+ 1.0) ./ 2.0, 0.0, 1.0) |> 
+                x -> reshape(x, 28,28) |>
+                x -> MLDatasets.convert2image(MNIST, x)
+            push!(images, x)
+            push!(rec_images, x̂)
+            j += 1
+        end
+    end
+    p1 = plot(mosaic(images..., ncol=10), title="Images")
+    p2 = plot(mosaic(rec_images..., ncol=10), title="Reconstructions")
+    plt = plot(p1, p2, axis=false, size=(800,375))
+
+    return plt
+end

### 2023-09-26 paper/aaai/author_response.qmd

@@
 Another common concern was that *ECCCo* primiarly achieved good results for JEMs.
+This has been addressed by introducing *ECCCo+* for situations when plausibility is crucial.
+We find that *ECCCo+* achieves good plausibility and faithfulness across the board.
+We have added convolutional neural networks to our analysis and find that *ECCCo+* achieves results for these models that are at least on par with the results for JEMs.
 

### 2023-09-26 paper/contents/table_gen_params.tex

\begin{table}

\caption{Generator hyperparameters. \label{tab:genparams} \newline}
\centering
\fontsize{8}{10}\selectfont
\begin{tabular}[t]{rrrrr}
\toprule
Dataset & $\eta$ & $\lambda_1$ & $\lambda_2$ & $\lambda_3$\\
\midrule
Linearly Separable & 0.01 & 0.10 & 0.10 & 0.05\\
Moons & 0.01 & 0.10 & 0.10 & 0.50\\
Circles & 0.05 & 0.10 & 0.10 & 0.05\\
California Housing & 0.05 & 0.10 & 0.10 & 0.10\\
GMSC & 0.05 & 0.10 & 0.10 & 0.10\\
German Credit & 0.05 & 0.20 & 0.20 & 0.20\\
MNIST & 0.10 & 0.01 & 0.25 & 0.25\\
Fashion MNIST & 0.10 & 0.01 & 0.25 & 0.25\\
\bottomrule
\end{tabular}
\end{table}

### 2023-09-26 paper/tables.Rmd

@@
 Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (\\textit{Wachter}).
-\\label{tab:results-%s} \\newline",
+\\label{tab:results-%s-valid} \\newline",
     name,

### 2023-09-26 paper/appendix.tex
done

@@
 For vision data, grid search was computationally prohibitve in any case, so hyperparameters were tuned manually.
-For all other datasets grid search was performed over different combinations of penalty strengths and optimizer steps sizes.
+For all other datasets grid search was performed over different combinations of penalty strengths and optimizer steps sizes (details of which can be found in the code).
+To select the final hyperparameter setting we used the unfaithfulness metric as our criterion.
 
@@
 While the results for \textit{REVISE} look fairly poor here, we have observed better results for optizers with higher step sizes.
-Note that the seemingly poor performance by \textit{REVISE} upon visual inspection is not driven by a weak surrogate VAE: Figure~\ref{fig:vae-reg} shows image reconstructions generated by the VAE.
+Note that the seemingly poor performance by \textit{REVISE} upon visual inspection is not driven by a weak surrogate VAE: Figure~\ref{fig:vae-rec} shows image reconstructions generated by the VAE.
 

### 2023-09-26 paper/aaai/author_response.qmd
updated response

@@
 
-Related work on plausibility of counterfactuals has largely relied on small image datasets like *MNIST* [@dhurandhar2018explanations;@schut2021generating;@delaney2023counterfactual].
+Related work on plausibility of counterfactuals has largely relied on small image datasets like *MNIST* [@dhurandhar2018explanations;@schut2021generating].
 This may be due to the fact that generating counterfactuals for high-dimensional input data is computationally very challenging.
@@
 That same reviewer also raised concern about our definitions of plausibility and faithfulness that rely on distributional properties.
-We have extensively argued our case during the rebuttal and pointed to a potential reviwer misunderstanding in this context.
+In particular, the reviewer argued that "[...] the class-condition distribution $p(\mathbf{x}|\mathbf{y^{+}})$ is existed but unknown and learning this distribution is very challenging especially for structural data".
+We have extensively argued our case during the rebuttal and pointed to a potential reviewer misunderstanding in this context.
+In particular, we argued:
+
+> We do not see this as a weakness of our paper.
+While we agree that learning this distribution is not always trivial, we note that this task is at the very core of Generative Modelling and AI&mdash;a field that has recently enjoyed success, especially in the context of large unstructured data like images and language.
+Learning the generative task is also at the core of related approaches mentioned in the paper like REVISE: as we mention in line 89, the authors of REVISE "propose using a generative model such as a Variational Autoencoder (VAE)" to learn $p(\mathbf{x})$.
+We also point to other related approaches towards plausibility that all centre around learning the data-generating process of the inputs $X$ (lines 85 to 104).
+Learning $p(\mathbf{x}|\mathbf{y^{+}})$ should generally be easier than learning the unconditional distribution $p(\mathbf{x})$, because the information contained in labels can be leveraged in the latter case.
+
 None of the other reviewers found any issue with our definitions and we have made no changes in this regard.
@@
 All of our code will be open-sourced as a package and we hope that it will be as useful to the community as was to us during our research.
-
-## References

### 2023-09-26 paper/appendix.tex
ss

@@
   \centering
-  \includegraphics[width=0.75\linewidth]{../artifacts/results/images/mnist_generated_JEM Ensemble.png}
+  \includegraphics[width=0.75\linewidth]{../www/mnist_generated_JEM Ensemble.png}
   \caption{Conditionally generated \textit{MNIST} images for our JEM Ensemble.}\label{fig:mnist-gen}
@@
   \centering
-  \includegraphics[width=0.5\linewidth]{../artifacts/results/images/moons_generated_JEM.png}
+  \includegraphics[width=0.5\linewidth]{../www/moons_generated_JEM.png}
   \caption{Conditionally generated samples (stars) for our \textit{Moons} data using a JEM.}\label{fig:moons-gen}
@@
   \centering
-  \includegraphics[width=1.0\linewidth]{../artifacts/results/images/poc_set_size.png}
+  \includegraphics[width=1.0\linewidth]{../www/poc_set_size.png}
   \caption{Prediction set size (left), smooth set size loss (centre) and configurable classification loss (right) for a JEM trained on our \textit{Linearly Separable} data.}\label{fig:cp-diff}

### 2023-09-26 paper/contents/table_gen_params.tex
uh

@@
 
-\caption{Generator hyperparameters. \label{tab:genparams} \newline}
+\caption{Generator hyperparameters: the optimiser step size ($\eta$); penalty strengths where $\lambda_1$ applies to all generators but \textit{Schut} and the other parameter are specific to \textit{ECCCo}; finally, the strength for the Ridge penalty on energy for \textit{ECCCo}.\label{tab:genparams} \newline}
 \centering
 \fontsize{8}{10}\selectfont
-\begin{tabular}[t]{rrrrr}
+\begin{tabular}[t]{rrrrrr}
 \toprule
-Dataset & $\eta$ & $\lambda_1$ & $\lambda_2$ & $\lambda_3$\\
+Dataset & $\eta$ & $\lambda_1$ & $\lambda_2$ & $\lambda_3$ & Ridge penalty\\
 \midrule
-Linearly Separable & 0.01 & 0.10 & 0.10 & 0.05\\
-Moons & 0.01 & 0.10 & 0.10 & 0.50\\
-Circles & 0.05 & 0.10 & 0.10 & 0.05\\
-California Housing & 0.05 & 0.10 & 0.10 & 0.10\\
-GMSC & 0.05 & 0.10 & 0.10 & 0.10\\
-German Credit & 0.05 & 0.20 & 0.20 & 0.20\\
-MNIST & 0.10 & 0.01 & 0.25 & 0.25\\
-Fashion MNIST & 0.10 & 0.01 & 0.25 & 0.25\\
+Linearly Separable & 0.01 & 0.10 & 0.1 & 0.05 & 0.0\\
+Moons & 0.01 & 0.10 & 0.1 & 0.50 & 0.0\\
+Circles & 0.05 & 0.10 & 0.1 & 0.05 & 1.0\\
+California Housing & 0.05 & 0.10 & 0.1 & 0.10 & 0.0\\
+GMSC & 0.05 & 0.10 & 0.1 & 0.10 & 0.0\\
+German Credit & 0.05 & 0.20 & 0.2 & 0.20 & 0.5\\
+MNIST & 0.10 & 0.01 & 0.1 & 0.30 & 0.0\\
+Fashion MNIST & 0.10 & 0.01 & 0.1 & 0.30 & 0.0\\
 \bottomrule

### 2023-09-27 paper/tables.Rmd

@@
 dt <- Reduce(function(x,y) {rbind(x,y, fill=TRUE)}, dt)
-dt <- dt[,.(dataname,eta,lambda_1,lambda_2,lambda_3)]
+dt[is.na(reg_strength),reg_strength:=0]
+dt <- dt[,.(dataname,eta,lambda_1_Δ,lambda_2_Δ,lambda_3_Δ,reg_strength)]
 dt[,dataname:=factor(dataname, levels=dataset_order)]
@@
   "Dataset",
-  "$\\eta$", "$\\lambda_1$", "$\\lambda_2$", "$\\lambda_3$"
+  "$\\eta$", "$\\lambda_1$", "$\\lambda_2$", "$\\lambda_3$", "Ridge penalty"
 )
 kbl(
-  dt, caption = "Generator hyperparameters. \\label{tab:genparams} \\newline", 
+  dt, caption = "Generator hyperparameters: the optimiser step size ($\\eta$); penalty strengths where $\\lambda_1$ applies to all generators but \\textit{Schut} and the other parameter are specific to \\textit{ECCCo}; finally, the strength for the Ridge penalty on energy for \\textit{ECCCo}.\\label{tab:genparams} \\newline",
   align = "r", col.names=col_names, booktabs = T, escape=F, 

### 2023-11-06 experiments/notebooks/plots.qmd
plot reviewer 1

---
format: pdf
---

# Plots

```{julia}
using Pkg; Pkg.activate("experiments")
include("$(pwd())/experiments/setup_env.jl")
```

## Counterfactual Path - MNIST

```{julia}
outcome = Serialization.deserialize("results/linearly_separable_outcome.jls")
data = outcome.exper.counterfactual_data
```

```{julia}
factual = data.y_levels[2]
target = data.y_levels[1]
```

```{julia}
n_samp = 100
plts = []
for (mod_name, model) in outcome.model_dict
    t = get_target_index(data.y_levels, target)
    E(x) = -logits(model, x)[t, :]
    x_samp = data.X[:,rand(findall(data.output_encoder.labels.==target),n_samp)]
    x_rand = Float32.(randn(size(data.X,1), 1000))
    dist = map(eachcol(x_rand)) do x
        mean(map(y -> norm(x.-y),eachcol(x_samp)))
    end
    plt = scatter(E(x_rand), dist, alpha=0.5, label="", title=mod_name)
    push!(plts, plt)
end
plot(plts..., layout=(1,length(outcome.model_dict)), size=(1000,250))
```

### 2023-11-06 experiments/notebooks/plots.qmd
plot reviewer 1

@@
 ```{julia}
-outcome = Serialization.deserialize("results/linearly_separable_outcome.jls")
+outcome = Serialization.deserialize("results/moons_outcome.jls")
 data = outcome.exper.counterfactual_data
@@
 n_samp = 100
+n_rand = 500
+alpha = 0.2
 plts = []
@@
     x_samp = data.X[:,rand(findall(data.output_encoder.labels.==target),n_samp)]
-    x_rand = Float32.(randn(size(data.X,1), 1000))
+    x_rand = Float32.(randn(size(data.X,1), n_rand))
     dist = map(eachcol(x_rand)) do x
@@
     end
-    plt = scatter(E(x_rand), dist, alpha=0.5, label="", title=mod_name)
+    plt = scatter(
+        E(x_rand), dist;
+        fillalpha=alpha, label="", title=mod_name, smooth=:true,
+        lc=:red, lw=2
+    )
     push!(plts, plt)

### 2023-11-06 paper/aaai/rebuttal.tex
changing format

%File: anonymous-submission-latex-2024.tex
\documentclass[letterpaper]{article} % DO NOT CHANGE THIS
\usepackage[submission]{aaai24}  % DO NOT CHANGE THIS
\usepackage{times}  % DO NOT CHANGE THIS
\usepackage{helvet}  % DO NOT CHANGE THIS
\usepackage{courier}  % DO NOT CHANGE THIS
\usepackage[hyphens]{url}  % DO NOT CHANGE THIS
\usepackage{graphicx} % DO NOT CHANGE THIS
\urlstyle{rm} % DO NOT CHANGE THIS
\def\UrlFont{\rm}  % DO NOT CHANGE THIS
\usepackage{natbib}  % DO NOT CHANGE THIS AND DO NOT ADD ANY OPTIONS TO IT
\usepackage{caption} % DO NOT CHANGE THIS AND DO NOT ADD ANY OPTIONS TO IT
\frenchspacing  % DO NOT CHANGE THIS
\setlength{\pdfpagewidth}{8.5in} % DO NOT CHANGE THIS
\setlength{\pdfpageheight}{11in} % DO NOT CHANGE THIS
%
% These are recommended to typeset algorithms but not required. See the subsubsection on algorithms. Remove them if you don't have algorithms in your paper.
\usepackage{algorithm}
% \usepackage{algorithmic}

%
% These are are recommended to typeset listings but not required. See the subsubsection on listing. Remove this block if you don't have listings in your paper.
% \usepackage{newfloat}
% \usepackage{listings}
% \DeclareCaptionStyle{ruled}{labelfont=normalfont,labelsep=colon,strut=off} % DO NOT CHANGE THIS
% \lstset{%
% 	basicstyle={\footnotesize\ttfamily},% footnotesize acceptable for monospace
% 	numbers=left,numberstyle=\footnotesize,xleftmargin=2em,% show line numbers, remove this entire line if you don't want the numbers.
% 	aboveskip=0pt,belowskip=0pt,%
% 	showstringspaces=false,tabsize=2,breaklines=true}
% \floatstyle{ruled}
% \newfloat{listing}{tb}{lst}{}
% \floatname{listing}{Listing}
%
% Keep the \pdfinfo as shown here. There's no need
% for you to add the /Title and /Author tags.
\pdfinfo{
/TemplateVersion (2024.1)
}

\usepackage{amsfonts}       % blackboard math symbols
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{caption}
\usepackage{graphicx}
\usepackage{algpseudocode}
\usepackage{import}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{placeins}

% Numbered Environments:
\newtheorem{definition}{Definition}[section]
\newtheorem{question}{Research Question}[section]

% Bibliography
% \bibliographystyle{unsrtnat}
% \setcitestyle{numbers,square,comma}

% Algorithm
\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}

% DISALLOWED PACKAGES
% \usepackage{authblk} -- This package is specifically forbidden
% \usepackage{balance} -- This package is specifically forbidden
% \usepackage{color (if used in text)
% \usepackage{CJK} -- This package is specifically forbidden
% \usepackage{float} -- This package is specifically forbidden
% \usepackage{flushend} -- This package is specifically forbidden
% \usepackage{fontenc} -- This package is specifically forbidden
% \usepackage{fullpage} -- This package is specifically forbidden
% \usepackage{geometry} -- This package is specifically forbidden
% \usepackage{grffile} -- This package is specifically forbidden
% \usepackage{hyperref} -- This package is specifically forbidden
% \usepackage{navigator} -- This package is specifically forbidden
% (or any other package that embeds links such as navigator or hyperref)
% \indentfirst} -- This package is specifically forbidden
% \layout} -- This package is specifically forbidden
% \multicol} -- This package is specifically forbidden
% \nameref} -- This package is specifically forbidden
% \usepackage{savetrees} -- This package is specifically forbidden
% \usepackage{setspace} -- This package is specifically forbidden
% \usepackage{stfloats} -- This package is specifically forbidden
% \usepackage{tabu} -- This package is specifically forbidden
% \usepackage{titlesec} -- This package is specifically forbidden
% \usepackage{tocbibind} -- This package is specifically forbidden
% \usepackage{ulem} -- This package is specifically forbidden
% \usepackage{wrapfig} -- This package is specifically forbidden
% DISALLOWED COMMANDS
% \nocopyright -- Your paper will not be published if you use this command
% \addtolength -- This command may not be used
% \balance -- This command may not be used
% \baselinestretch -- Your paper will not be published if you use this command
% \clearpage -- No page breaks of any kind may be used for the final version of your paper
% \columnsep -- This command may not be used
% \newpage -- No page breaks of any kind may be used for the final version of your paper
% \pagebreak -- No page breaks of any kind may be used for the final version of your paperr
% \pagestyle -- This command may not be used
% \tiny -- This is not an acceptable font size.
% \vspace{- -- No negative value may be used in proximity of a caption, figure, table, section, subsection, subsubsection, or reference
% \vskip{- -- No negative value may be used to alter spacing above or below a caption, figure, table, section, subsection, subsubsection, or reference

\setcounter{secnumdepth}{2} %May be changed to 1 or 2 if section numbers are desired.

% The file aaai24.sty is the style file for AAAI Press
% proceedings, working notes, and technical reports.
%

% Title

% Your title must be in mixed case, not sentence case.
% That means all verbs (including short verbs like be, is, using,and go),
% nouns, adverbs, adjectives should be capitalized, including both words in hyphenated terms, while
% articles, conjunctions, and prepositions are lower case unless they
% directly follow a colon or long dash
\title{Faithful Model Explanations through\\
Energy-Constrained Conformal Counterfactuals}
\author{
    %Authors
    % All authors must be in the same font size and format.
    Written by AAAI Press Staff\textsuperscript{\rm 1}\thanks{With help from the AAAI Publications Committee.}\\
    AAAI Style Contributions by Pater Patel Schneider,
    Sunil Issar,\\
    J. Scott Penberthy,
    George Ferguson,
    Hans Guesgen,
    Francisco Cruz\equalcontrib,
    Marc Pujol-Gonzalez\equalcontrib
}
\affiliations{
    %Afiliations
    \textsuperscript{\rm 1}Association for the Advancement of Artificial Intelligence\\
    % If you have multiple authors and multiple affiliations
    % use superscripts in text and roman font to identify them.
    % For example,

    % Sunil Issar\textsuperscript{\rm 2},
    % J. Scott Penberthy\textsuperscript{\rm 3},
    % George Ferguson\textsuperscript{\rm 4},
    % Hans Guesgen\textsuperscript{\rm 5}
    % Note that the comma should be placed after the superscript

    1900 Embarcadero Road, Suite 101\\
    Palo Alto, California 94303-3310 USA\\
    % email address must be in roman text type, not monospace or sans serif
    proceedings-questions@aaai.org
%
% See more examples next
}

%Example, Single Author, ->> remove \iffalse,\fi and place them surrounding AAAI title to use it
% \iffalse
% \title{My Publication Title --- Single Author}
% \author {
%     Author Name
% }
% \affiliations{
%     Affiliation\\
%     Affiliation Line 2\\
%     name@example.com
% }
% \fi

% \iffalse
% %Example, Multiple Authors, ->> remove \iffalse,\fi and place them surrounding AAAI title to use it
% \title{My Publication Title --- Multiple Authors}
% \author {
%     % Authors
%     First Author Name\textsuperscript{\rm 1},
%     Second Author Name\textsuperscript{\rm 2},
%     Third Author Name\textsuperscript{\rm 1}
% }
% \affiliations {
%     % Affiliations
%     \textsuperscript{\rm 1}Affiliation 1\\
%     \textsuperscript{\rm 2}Affiliation 2\\
%     firstAuthor@affiliation1.com, secondAuthor@affilation2.com, thirdAuthor@affiliation1.com
% }
% \fi

\begin{document}

\subsubsection{Experiment results: linguistic explanation of results}

Following the suggestion by the reviewer, we plan to add a the following linguistic explanation in a prominent place of Section 6:

"Overall, our findings demonstrate that \textit{ECCCo} produces plausible counterfactuals if and only if the black-box model itself has learned plausible explanations for the data. Thus, \textit{ECCCo} avoids the risk of generating plausible but potentially misleading explanations for models that are highly susceptible to implausible explanations. We, therefore, believe that \textit{ECCCo} can help researchers and practitioners to generate explanations they can trust and discern unreliable from trustworthy models."

Elements of this explanation are already scattered across the paper, but we agree that it would be useful to highlight this notion in Section 6.

\subsubsection{Core innovation: need more visualizations}

Following the reviewer's suggestion, we have plotted the distance of randomly generated MNIST images from images in the target class against their energy-constrained score. As expected, this relationship is positive: the higher the distance, the higher the corresponding generative loss. The size of this relationship appears to depend positively on the model's generative property: the observed relationships are stronger for joint energy models.

\subsubsection{Structural clarity: add a flow chart}

Adding a systematic flowchart is a great idea. Due to the limited scope, may we suggest adding the following flowchart to the appendix? Alternatively, we may swap out Figure 2 for the flowchart.

\subsubsection{Why use an embedding}

We agree that for any type of surrogate model, there is a risk of introducing bias. In exceptional cases, however, it may be necessary to accept some degree of bias in favour of plausibility. Our results for \textit{ECCCo+} demonstrate this tradeoff as we discuss in Section 6.3. In the context of PCA, the introduced bias can be explained intuitively: by constraining the counterfactual search to the space spanned by the first $n_z$ principal components, the search is sensitive only to the variation in the data explained by those components. In other words, we would expect counterfactuals to be less sensitive to small variations in features that do not typically vary much. It is therefore an intuitive finding, that \textit{ECCCo+} tends to generate less noisy counterfactual images, for example (the same is true for \textit{REVISE}). In our mind, restricting the search space to the first $n_z$ components quite literally corresponds to denoising the search space and hence the resulting counterfactuals. We will highlight this rationale in Section 6.3.

We think that the bias introduced by PCA may be acceptable in some cases, precisely because it "will not add any information on the input distribution" as the reviewer correctly points out. To maintain faithfulness, we want to avoid introducing additional information through surrogate models as much as possible. We will make this intuition clearer in Section 6.3.

Another argument in favour of using a lower-dimensional latent embedding is the reduction in computational costs, which can be prohibitive for high-dimensional input data. We will highlight this in Section 5.

\subsubsection{What is "epsilon" and "s"}

From the paper: "$\mathbf{r}_j \sim \mathcal{N}(\mathbf{0},\mathbf{I})$ is the stochastic term and the step-size $\epsilon_j$ is typically polynomially decayed. [...] To allow for faster sampling, we follow the common practice of choosing the step-size $\epsilon_j$ and the standard deviation of $\mathbf{r}_j$ separately." We go on to explain in the appendix that we use the following biased sampler

$$
\hat{\mathbf{x}}_{j+1} \leftarrow \hat{\mathbf{x}}_j - \frac{\phi}{2} \mathcal{E}_{\theta}(\hat{\mathbf{x}}_j|\mathbf{y}^+) + \sigma \mathbf{r}_j, j=1,...,J
$$

where "consistent with~\citet{grathwohl2020your}, we have specified $\phi=2$ and $\sigma=0.01$ as the default values for all of our experiments". Intuitively, $\epsilon_j$ determines the size of gradient updates and random noise in each iteration of SGLD.

Regarding $s(\cdot)$, this was an oversight, apologies. In the appendix we explain that "[the calibration dataset] is then used to compute so-called nonconformity scores: $\mathcal{S}=\{s(\mathbf{x}_i,\mathbf{y}_i)\}_{i \in \mathcal{D}_{\text{cal}}}$ where $s: (\mathcal{X},\mathcal{Y}) \mapsto \mathbb{R}$ is referred to as \textit{score function}." We will add this in Section 4.2 of the main paper.

\subsubsection{Euclidean distance}

As we mentioned in the additional author response, we investigated different distance metrics. We found that the overall qualitative results were largely independent of the exact metric. In the context of the high-dimensional image data, we still decided to report the results for a dissimilarity metric that is more appropriate in this context. All of our distance-based metrics are computed with respect to features, not latent features. This is because, as the reviewer correctly points out, we would expect certain discrepancies between distances evaluated in the feature space and distances evaluated in the latent space of the VAE, for example. Working in the feature space does come with higher computational costs, but the evaluation of counterfactuals was generally less costly than generating counterfactuals in the first place. In cases where high dimensionality leads to prohibitive computational costs, we would suggest either reducing the number of nearest neighbors or working in a lower-dimensional subspace that is independent of the underlying classifier itself (such as PCA).

\subsubsection{Faithfulness metric: is it fair?}

We have taken measures to not unfairly bias our generator with respect to the unfaithfulness metric: instead of penalizing the unfaithfulness metric directly, we penalize model energy in our preferred implementation. In contrast, \textit{Wachter} penalizes the closeness criterion directly and hence does particularly well in this regard. That being said, \textit{ECCCo} is of course designed to generate faithful explanations first and foremost and therefore has an advantage with respect to our faithfulness metric. In lieue of other established metrics to measure faithfulness, we can only point out that \textit{ECCCo} achieves strong performance for other commonly used metrics as well. With respect to \textit{validity}, for example, which as we have explained corresponds to \textit{fidelity}, \textit{ECCCo} typically outperforms \textit{REVISE} and \textit{Schut}.

Our joint energy models (JEM) are indeed explicitly trained to model $\mathcal{X}|y$ and the same quantity is used in our proposed faithfulness metric. However, the faithfulness metric itself is not computed with respect to samples generated by our JEMs. It is computed with respect to counterfactuals generated by merely constraining model energy and we would therefore argue that it is not unfairly biased. Our empirical findings support this argument: firstly, \textit{ECCCo} achieves high faithfulness also for classifiers that have not been trained to model $\mathcal{X}|y$; secondly, our additional results in the appendix for \textit{ECCCo-L1} show that if we do indeed explicitly penalize the unfaithfulness metric, we achieve even better results in this regard.

\subsubsection{Test with unreliable models}

We would argue that the simple multi-layer perceptrons (MLPs) are unreliable, especially compared to ensembles, joint energy models and convolutional neural networks for our image datasets. Simple neural networks have been shown to be vulnerable to adversarial attacks, which makes them susceptible to implausible counterfactual explanations as we point out in Section 3. Our results support this notion, in that they demonstrate faithful model explanations only coincide with high plausibility if the model itself has been trained to be more reliable. Consistent with the idea proposed by the reviewer, we originally considered introducing "poisoned" VAEs as well, to illustrate what we identify as the key vulnerability of \textit{REVISE}. If the underlying VAE is trained on poisoned data, this could be expected to adversely affect counterfactual outcomes as well. We ultimately discarded this idea due to limited scope and because we decided that Section 3 sufficiently illustrates our thinking.

\FloatBarrier

\bibliography{../bib}

\end{document}

### 2023-11-06 experiments/notebooks/plots.qmd
changing format

--- 2023-11-06 experiments/notebooks/plots.qmd
@@
 ```{julia}
-outcome = Serialization.deserialize("results/linearly_separable_outcome.jls")
-data = outcome.exper.counterfactual_data
+data_name = "mnist"
+plt_order = ["MLP", "MLP Ensemble", "LeNet-5", "JEM", "JEM Ensemble"]
+models = Serialization.deserialize("models/$(data_name)_models.jls")
+data = eval(Meta.parse("load_$(data_name)()"))
+plt_order = plt_order[[x in collect(keys(models)) for x in plt_order]]
 ```
@@
 ```{julia}
-factual = data.y_levels[2]
-target = data.y_levels[1]
-```
-
-```{julia}
+using Plots.PlotMeasures
 n_samp = 100
+n_rand = 500
+σ = 0.1
 plts = []
-for (mod_name, model) in outcome.model_dict
+for (mod_name, model) in models
+    Δ = []
+    L = []
+    for i in 1:n_rand
+        factual = rand(data.y_levels)
+        target = rand(data.y_levels[data.y_levels .!= factual])
         t = get_target_index(data.y_levels, target)
@@
         x_samp = data.X[:,rand(findall(data.output_encoder.labels.==target),n_samp)]
-    x_rand = Float32.(randn(size(data.X,1), 1000))
-    dist = map(eachcol(x_rand)) do x
-        mean(map(y -> norm(x.-y),eachcol(x_samp)))
+        x_rand = data.X[:,rand(findall(data.output_encoder.labels.==factual),1)]
+        x_rand .+= Float32.(randn(size(x_samp,1),1)) .* σ
+        δ = mean(map(y -> norm(x_rand.-y),eachcol(x_samp)))
+        push!(Δ,δ)
+        l = E(x_rand)[1]
+        push!(L,l)
     end
-    plt = scatter(E(x_rand), dist, alpha=0.5, label="", title=mod_name)
+    plt = scatter(
+        Δ, L;
+        label="", title=mod_name, smooth=:true,
+        lc=:red, lw=2
+    )
     push!(plts, plt)
 end
-plot(plts..., layout=(1,length(outcome.model_dict)), size=(1000,250))
+plot(
+    plts[sortperm(collect(keys(models)))[invperm(sortperm(plt_order))]]..., 
+    layout=(1,length(models)), size=(1200,1100/length(models)),
+    margin=5mm
+)
 ```

### 2023-11-06 paper/aaai/rebuttal.tex
trimming down

@@
 
-Following the suggestion by the reviewer, we plan to add a the following linguistic explanation in a prominent place of Section 6:
+In Section 6, we will add the following linguistic explanation:
 
-"Overall, our findings demonstrate that \textit{ECCCo} produces plausible counterfactuals if and only if the black-box model itself has learned plausible explanations for the data.
+`Overall, our findings demonstrate that \textit{ECCCo} produces plausible counterfactuals if and only if the black-box model itself has learned plausible explanations for the data.
 Thus, \textit{ECCCo} avoids the risk of generating plausible but potentially misleading explanations for models that are highly susceptible to implausible explanations.
-We, therefore, believe that \textit{ECCCo} can help researchers and practitioners to generate explanations they can trust and discern unreliable from trustworthy models."
-
-Elements of this explanation are already scattered across the paper, but we agree that it would be useful to highlight this notion in Section 6.
+We, therefore, believe that \textit{ECCCo} can help researchers and practitioners to generate explanations they can trust and discern unreliable from trustworthy models.'
 
@@
 We agree that for any type of surrogate model, there is a risk of introducing bias.
-In exceptional cases, however, it may be necessary to accept some degree of bias in favour of plausibility.
+In exceptional cases, it may be necessary to accept some degree of bias in favor of plausibility.
 Our results for \textit{ECCCo+} demonstrate this tradeoff as we discuss in Section 6.3.
 In the context of PCA, the introduced bias can be explained intuitively: by constraining the counterfactual search to the space spanned by the first $n_z$ principal components, the search is sensitive only to the variation in the data explained by those components.
-In other words, we would expect counterfactuals to be less sensitive to small variations in features that do not typically vary much.
-It is therefore an intuitive finding, that \textit{ECCCo+} tends to generate less noisy counterfactual images, for example (the same is true for \textit{REVISE}).
-In our mind, restricting the search space to the first $n_z$ components quite literally corresponds to denoising the search space and hence the resulting counterfactuals.
+It is therefore an intuitive finding, that \textit{ECCCo+} tends to generate less noisy counterfactuals.
+In our mind, restricting the search space to the first $n_z$ components quite literally corresponds to denoising the search space.
 We will highlight this rationale in Section 6.3.
 
-We think that the bias introduced by PCA may be acceptable in some cases, precisely because it "will not add any information on the input distribution" as the reviewer correctly points out.
+We think that the bias introduced by PCA may be acceptable, precisely because it `will not add any information on the input distribution' as the reviewer correctly points out.
 To maintain faithfulness, we want to avoid introducing additional information through surrogate models as much as possible.
@@
 
-Another argument in favour of using a lower-dimensional latent embedding is the reduction in computational costs, which can be prohibitive for high-dimensional input data.
-We will highlight this in Section 5.
+Another argument for using a lower-dimensional latent embedding is the reduction in computational costs, which can be prohibitive for certain data. We will highlight this in Section 5.
 
-\subsubsection{What is "epsilon" and "s"}
+\subsubsection{What is `epsilon' and `s'}
 
-From the paper: "$\mathbf{r}_j \sim \mathcal{N}(\mathbf{0},\mathbf{I})$ is the stochastic term and the step-size $\epsilon_j$ is typically polynomially decayed.
-[...] To allow for faster sampling, we follow the common practice of choosing the step-size $\epsilon_j$ and the standard deviation of $\mathbf{r}_j$ separately." We go on to explain in the appendix that we use the following biased sampler
+From the paper: `[...] the step-size $\epsilon_j$ is typically polynomially decayed.
+[...] To allow for faster sampling, we follow the common practice of choosing the step-size $\epsilon_j$ and the standard deviation of $\mathbf{r}_j$ separately.' Intuitively, $\epsilon_j$ determines the size of gradient updates and random noise in each iteration of SGLD.
 
-$$
-\hat{\mathbf{x}}_{j+1} \leftarrow \hat{\mathbf{x}}_j - \frac{\phi}{2} \mathcal{E}_{\theta}(\hat{\mathbf{x}}_j|\mathbf{y}^+) + \sigma \mathbf{r}_j, j=1,...,J
-$$
-
-where "consistent with~\citet{grathwohl2020your}, we have specified $\phi=2$ and $\sigma=0.01$ as the default values for all of our experiments".
-Intuitively, $\epsilon_j$ determines the size of gradient updates and random noise in each iteration of SGLD.
-
-Regarding $s(\cdot)$, this was an oversight, apologies.
+Regarding $s(\cdot)$, this was an oversight.
 In the appendix we explain that "[the calibration dataset] is then used to compute so-called nonconformity scores: $\mathcal{S}=\{s(\mathbf{x}_i,\mathbf{y}_i)\}_{i \in \mathcal{D}_{\text{cal}}}$ where $s: (\mathcal{X},\mathcal{Y}) \mapsto \mathbb{R}$ is referred to as \textit{score function}." We will add this in Section 4.2 of the main paper.
@@
 
-As we mentioned in the additional author response, we investigated different distance metrics.
-We found that the overall qualitative results were largely independent of the exact metric.
+As we mentioned in the additional author response, we investigated different distance metrics and found that the overall qualitative results were largely independent of the choice.
 In the context of the high-dimensional image data, we still decided to report the results for a dissimilarity metric that is more appropriate in this context.
-All of our distance-based metrics are computed with respect to features, not latent features.
-This is because, as the reviewer correctly points out, we would expect certain discrepancies between distances evaluated in the feature space and distances evaluated in the latent space of the VAE, for example.
-Working in the feature space does come with higher computational costs, but the evaluation of counterfactuals was generally less costly than generating counterfactuals in the first place.
-In cases where high dimensionality leads to prohibitive computational costs, we would suggest either reducing the number of nearest neighbors or working in a lower-dimensional subspace that is independent of the underlying classifier itself (such as PCA).
+All of our distance-based metrics are computed in the feature space.
+This is because we would indeed expect certain discrepancies between distances evaluated in the feature space and distances evaluated in the latent space of the VAE, for example.
+In cases where high dimensionality leads to prohibitive computational costs, we suggest working in a lower-dimensional subspace that is independent of the underlying classifier itself (such as PCA).
 
@@
 
-We have taken measures to not unfairly bias our generator with respect to the unfaithfulness metric: instead of penalizing the unfaithfulness metric directly, we penalize model energy in our preferred implementation.
+We have taken measures to not unfairly bias our generator for the unfaithfulness metric: instead of penalizing the unfaithfulness metric directly, we penalize model energy in our preferred implementation.
 In contrast, \textit{Wachter} penalizes the closeness criterion directly and hence does particularly well in this regard.
-That being said, \textit{ECCCo} is of course designed to generate faithful explanations first and foremost and therefore has an advantage with respect to our faithfulness metric.
-In lieue of other established metrics to measure faithfulness, we can only point out that \textit{ECCCo} achieves strong performance for other commonly used metrics as well.
-With respect to \textit{validity}, for example, which as we have explained corresponds to \textit{fidelity}, \textit{ECCCo} typically outperforms \textit{REVISE} and \textit{Schut}.
+In the lack of other established metrics to measure faithfulness, we can only point out that \textit{ECCCo} achieves strong performance for other commonly used metrics as well.
+For \textit{validity}, which as we have explained corresponds to \textit{fidelity}, \textit{ECCCo} typically performs strongly.
 
 Our joint energy models (JEM) are indeed explicitly trained to model $\mathcal{X}|y$ and the same quantity is used in our proposed faithfulness metric.
-However, the faithfulness metric itself is not computed with respect to samples generated by our JEMs.
-It is computed with respect to counterfactuals generated by merely constraining model energy and we would therefore argue that it is not unfairly biased.
+However, the faithfulness metric itself is not computed for samples generated by our JEMs.
+It is computed for counterfactuals generated by constraining model energy and we would therefore argue that it is not unfairly biased.
 Our empirical findings support this argument: firstly, \textit{ECCCo} achieves high faithfulness also for classifiers that have not been trained to model $\mathcal{X}|y$; secondly, our additional results in the appendix for \textit{ECCCo-L1} show that if we do indeed explicitly penalize the unfaithfulness metric, we achieve even better results in this regard.
@@
 We would argue that the simple multi-layer perceptrons (MLPs) are unreliable, especially compared to ensembles, joint energy models and convolutional neural networks for our image datasets.
-Simple neural networks have been shown to be vulnerable to adversarial attacks, which makes them susceptible to implausible counterfactual explanations as we point out in Section 3.
+Simple neural networks are vulnerable to adversarial attacks, which makes them susceptible to implausible counterfactual explanations as we point out in Section 3.
 Our results support this notion, in that they demonstrate faithful model explanations only coincide with high plausibility if the model itself has been trained to be more reliable.
@@
 
-\FloatBarrier
-
-\bibliography{../bib}
-
 \end{document}

### 2023-11-06 paper/aaai/rebuttal.tex
done

@@
 
-\subsubsection{Experiment results: linguistic explanation of results}
+\subsubsection{Experiment results: linguistic explanation.}
 
@@
 
-`Overall, our findings demonstrate that \textit{ECCCo} produces plausible counterfactuals if and only if the black-box model itself has learned plausible explanations for the data.
+`Our findings demonstrate that \textit{ECCCo} produces plausible counterfactuals if and only if the black-box model itself has learned plausible explanations for the data.
 Thus, \textit{ECCCo} avoids the risk of generating plausible but potentially misleading explanations for models that are highly susceptible to implausible explanations.
-We, therefore, believe that \textit{ECCCo} can help researchers and practitioners to generate explanations they can trust and discern unreliable from trustworthy models.'
+Thus, \textit{ECCCo} can be instrumental in discerning unreliable from trustworthy models.'
 
-\subsubsection{Core innovation: need more visualizations}
+\subsubsection{Core innovation: more visualization.}
 
-Following the reviewer's suggestion, we have plotted the distance of randomly generated MNIST images from images in the target class against their energy-constrained score.
-As expected, this relationship is positive: the higher the distance, the higher the corresponding generative loss.
-The size of this relationship appears to depend positively on the model's generative property: the observed relationships are stronger for joint energy models.
+Figure~\ref{fig:poc} shows the relationship between implausibility and the energy constraint for MNIST data.
+As expected, this relationship is positive and the size of the relationship depends positively on the model's generative property (the observed relationships are stronger for joint energy models).
+We will add such images for all datasets to the appendix.
 
-\subsubsection{Structural clarity: add a flow chart}
+\begin{figure}
+    \centering
+    \includegraphics[width=\linewidth]{../../www/dist_energy.png}
+    \caption{The distance of randomly drawn MNIST images from images in the target class (horizontal axis) plotted against their energy-constrained score (vertical axis).}\label{fig:poc}
+\end{figure}
 
-Adding a systematic flowchart is a great idea. Due to the limited scope, may we suggest adding the following flowchart to the appendix? Alternatively, we may swap out Figure 2 for the flowchart.
+\subsubsection{Structural clarity.}
 
-\subsubsection{Why use an embedding}
+To facilitate comprehension, we will follow the reviewer's advice and add a systematic flowchart either in the appendix or in place of Figure 2.
 
-We agree that for any type of surrogate model, there is a risk of introducing bias.
+\subsubsection{Why use an embedding?}
+
+For any type of surrogate model, there is a risk of introducing bias.
 In exceptional cases, it may be necessary to accept some degree of bias in favor of plausibility.
-Our results for \textit{ECCCo+} demonstrate this tradeoff as we discuss in Section 6.3.
-In the context of PCA, the introduced bias can be explained intuitively: by constraining the counterfactual search to the space spanned by the first $n_z$ principal components, the search is sensitive only to the variation in the data explained by those components.
+Our discussion of the results for \textit{ECCCo+} in Section 6.3 demonstrates this tradeoff.
+The bias introduced through PCA can be explained intuitively: by constraining the counterfactual search to the space spanned by the first $n_z$ principal components, the search is sensitive only to the variation explained by those components.
 It is therefore an intuitive finding, that \textit{ECCCo+} tends to generate less noisy counterfactuals.
-In our mind, restricting the search space to the first $n_z$ components quite literally corresponds to denoising the search space.
 We will highlight this rationale in Section 6.3.
@@
 
-Another argument for using a lower-dimensional latent embedding is the reduction in computational costs, which can be prohibitive for certain data. We will highlight this in Section 5.
+Another argument for using a low-dimensional latent embedding is the reduction in computational costs, which can be prohibitive for some data. We will add this in Section 5.
 
-\subsubsection{What is `epsilon' and `s'}
+\subsubsection{What is `epsilon' and `s'?}
 
-From the paper: `[...] the step-size $\epsilon_j$ is typically polynomially decayed.
-[...] To allow for faster sampling, we follow the common practice of choosing the step-size $\epsilon_j$ and the standard deviation of $\mathbf{r}_j$ separately.' Intuitively, $\epsilon_j$ determines the size of gradient updates and random noise in each iteration of SGLD.
+From the paper: `[...] the step-size $\epsilon_j$ is typically polynomially decayed.' Intuitively, $\epsilon_j$ determines the size of gradient updates and random noise in each iteration of SGLD.
 
@@
 
-\subsubsection{Euclidean distance}
+\subsubsection{Euclidean distance.}
 
@@
 In the lack of other established metrics to measure faithfulness, we can only point out that \textit{ECCCo} achieves strong performance for other commonly used metrics as well.
-For \textit{validity}, which as we have explained corresponds to \textit{fidelity}, \textit{ECCCo} typically performs strongly.
+For \textit{validity}, which corresponds to \textit{fidelity}, \textit{ECCCo} typically performs strongly.
 
@@
 
-\subsubsection{Test with unreliable models}
+\subsubsection{Add unreliable models.}
 
@@
 Our results support this notion, in that they demonstrate faithful model explanations only coincide with high plausibility if the model itself has been trained to be more reliable.
-Consistent with the idea proposed by the reviewer, we originally considered introducing "poisoned" VAEs as well, to illustrate what we identify as the key vulnerability of \textit{REVISE}.
-If the underlying VAE is trained on poisoned data, this could be expected to adversely affect counterfactual outcomes as well.
+Consistent with the idea proposed by the reviewer, we originally considered introducing `poisoned' VAEs as well, to illustrate what we identify as the key vulnerability of \textit{REVISE}: if the underlying VAE is trained on poisoned data, this will adversely affect counterfactual outcomes as well.
 We ultimately discarded this idea due to limited scope and because we decided that Section 3 sufficiently illustrates our thinking.

### 2023-11-06 paper/aaai/rebuttal.tex
minor

@@
 
-\subsubsection{Core innovation: more visualization.}
+\subsubsection{Core innovation: more visualizations.}
 
@@
 We will add such images for all datasets to the appendix.
+We note that our final benchmark results involve around 1.5 million counterfactuals per dataset (not including grid search).
 
@@
 The bias introduced through PCA can be explained intuitively: by constraining the counterfactual search to the space spanned by the first $n_z$ principal components, the search is sensitive only to the variation explained by those components.
-It is therefore an intuitive finding, that \textit{ECCCo+} tends to generate less noisy counterfactuals.
+It is an intuitive finding, that the generated counterfactuals are less noisy.
 We will highlight this rationale in Section 6.3.

### 2023-11-06 paper/aaai/rebuttal.tex
uh

@@
 
+We thank the reviewers for their thoughtful comments.
+
 \subsubsection{Experiment results: linguistic explanation.}
@@
 We will add such images for all datasets to the appendix.
-We note that our final benchmark results involve around 1.5 million counterfactuals per dataset (not including grid search).
+Qur final benchmark results involve around 1.5 million counterfactuals per dataset.
 
@@
 
-We would argue that the simple multi-layer perceptrons (MLPs) are unreliable, especially compared to ensembles, joint energy models and convolutional neural networks for our image datasets.
+We would argue that the simple multi-layer perceptrons are unreliable, especially compared to ensembles, joint energy models and convolutional neural networks.
 Simple neural networks are vulnerable to adversarial attacks, which makes them susceptible to implausible counterfactual explanations as we point out in Section 3.
-Our results support this notion, in that they demonstrate faithful model explanations only coincide with high plausibility if the model itself has been trained to be more reliable.
+Our results support this notion, in that the quality of counterfactuals produced by \textit{ECCCo} is higher for more reliable models.
 Consistent with the idea proposed by the reviewer, we originally considered introducing `poisoned' VAEs as well, to illustrate what we identify as the key vulnerability of \textit{REVISE}: if the underlying VAE is trained on poisoned data, this will adversely affect counterfactual outcomes as well.

### 2023-11-07 paper/aaai/rebuttal.tex
rebuttal done

@@
 
-We thank the reviewers for their thoughtful comments.
+We thank the reviewers for their thoughtful comments and are glad with the overall positive response.
 
-\subsubsection{Experiment results: linguistic explanation.}
+\subsection*{Reviewer \#1}
 
-In Section 6, we will add the following linguistic explanation:
+\subsubsection{1. Experiment results: linguistic explanation.}
 
-`Our findings demonstrate that \textit{ECCCo} produces plausible counterfactuals if and only if the black-box model itself has learned plausible explanations for the data.
-Thus, \textit{ECCCo} avoids the risk of generating plausible but potentially misleading explanations for models that are highly susceptible to implausible explanations.
-Thus, \textit{ECCCo} can be instrumental in discerning unreliable from trustworthy models.'
+We will add a linguistic explanation in Section 6 where we highlight that \textit{ECCCo} produces plausible counterfactuals iff the classifier itself has learned plausible explanations for the data.
+It thus avoids the risk of generating plausible but potentially misleading explanations for models that are highly susceptible to implausible explanations.
 
-\subsubsection{Core innovation: more visualizations.}
+\subsubsection{2. Core innovation: more visualizations.}
 
@@
 We will add such images for all datasets to the appendix.
-Qur final benchmark results involve around 1.5 million counterfactuals per dataset.
+We note that our final benchmark results involve around 1.5 million counterfactuals per dataset (not including grid searches).
 
-\begin{figure}
+\begin{figure}[h]
     \centering
@@
 
-\subsubsection{Structural clarity.}
+\subsubsection{3. Structural clarity.}
 
@@
 
-\subsubsection{Why use an embedding?}
+\subsection*{Reviewer \#2}
 
-For any type of surrogate model, there is a risk of introducing bias.
-In exceptional cases, it may be necessary to accept some degree of bias in favor of plausibility.
-Our discussion of the results for \textit{ECCCo+} in Section 6.3 demonstrates this tradeoff.
-The bias introduced through PCA can be explained intuitively: by constraining the counterfactual search to the space spanned by the first $n_z$ principal components, the search is sensitive only to the variation explained by those components.
-It is an intuitive finding, that the generated counterfactuals are less noisy.
-We will highlight this rationale in Section 6.3.
+\subsubsection{4. Why use an embedding?}
 
+There are two main reasons for using a low-dimensional latent embedding: firstly, to help with plausibility and, secondly, to reduce computational costs.
+The latter is not currently made explicit in the paper and we will add this in Section 5.
+The former is discussed in the context of the results for \textit{ECCCo+} in Section 6.3, but we will highlight the following rationale.
+
+There is indeed a tradeoff between plausibility and faithfulness through the introduction of bias: plausibility is improved because counterfactuals are insensitive to variation captured by higher-order principal components.
+Intuitively, the generated counterfactuals are therefore less noisy.
 We think that the bias introduced by PCA may be acceptable, precisely because it `will not add any information on the input distribution' as the reviewer correctly points out.
-To maintain faithfulness, we want to avoid introducing additional information through surrogate models as much as possible.
-We will make this intuition clearer in Section 6.3.
+To maintain faithfulness, we want to avoid adding any information through surrogate models as much as possible.
 
-Another argument for using a low-dimensional latent embedding is the reduction in computational costs, which can be prohibitive for some data. We will add this in Section 5.
-
-\subsubsection{What is `epsilon' and `s'?}
+\subsubsection{5. What is `epsilon' and `s'?}
 
@@
 Regarding $s(\cdot)$, this was an oversight.
-In the appendix we explain that "[the calibration dataset] is then used to compute so-called nonconformity scores: $\mathcal{S}=\{s(\mathbf{x}_i,\mathbf{y}_i)\}_{i \in \mathcal{D}_{\text{cal}}}$ where $s: (\mathcal{X},\mathcal{Y}) \mapsto \mathbb{R}$ is referred to as \textit{score function}." We will add this in Section 4.2 of the main paper.
+In the appendix we explain that `[the calibration dataset] is then used to compute so-called nonconformity scores: $\mathcal{S}=\{s(\mathbf{x}_i,\mathbf{y}_i)\}_{i \in \mathcal{D}_{\text{cal}}}$ where $s: (\mathcal{X},\mathcal{Y}) \mapsto \mathbb{R}$ is referred to as \textit{score function}.' We will add this in Section 4.2 of the paper.
 
-\subsubsection{Euclidean distance.}
+\subsubsection{6. Euclidean distance.}
 
-As we mentioned in the additional author response, we investigated different distance metrics and found that the overall qualitative results were largely independent of the choice.
-In the context of the high-dimensional image data, we still decided to report the results for a dissimilarity metric that is more appropriate in this context.
+As we mentioned in the additional author response, we investigated different distance metrics and found that the overall qualitative results were largely independent of the choice of metric.
+For image data, we still decided to report the results for a dissimilarity metric that is more appropriate in this context.
 All of our distance-based metrics are computed in the feature space.
-This is because we would indeed expect certain discrepancies between distances evaluated in the feature space and distances evaluated in the latent space of the VAE, for example.
-In cases where high dimensionality leads to prohibitive computational costs, we suggest working in a lower-dimensional subspace that is independent of the underlying classifier itself (such as PCA).
+This is because we would indeed expect certain discrepancies between distances evaluated in the feature space and distances evaluated in the latent space of a VAE, for example.
+In cases where high dimensionality leads to prohibitive computational costs, we suggest working in a lower-dimensional subspace that is as uninformative as possible (such as PCA).
 
-\subsubsection{Faithfulness metric: is it fair?}
+\subsubsection{7. Model fails to learn plausible explanations.}
+
+In these cases, \textit{ECCCo} generally achieves lower plausibility while maintaining faithfulness (see also points 1 and 9).
+
+\subsubsection{8. Faithfulness metric: is it fair?}
 
@@
 In contrast, \textit{Wachter} penalizes the closeness criterion directly and hence does particularly well in this regard.
-In the lack of other established metrics to measure faithfulness, we can only point out that \textit{ECCCo} achieves strong performance for other commonly used metrics as well.
-For \textit{validity}, which corresponds to \textit{fidelity}, \textit{ECCCo} typically performs strongly.
+In the absence of other established faithfulness metrics, we can only point out that \textit{ECCCo} achieves strong performance for other commonly used metrics as well.
+For \textit{validity}, which corresponds to \textit{fidelity}, \textit{ECCCo} performs strongly.
 
-Our joint energy models (JEM) are indeed explicitly trained to model $\mathcal{X}|y$ and the same quantity is used in our proposed faithfulness metric.
-However, the faithfulness metric itself is not computed for samples generated by our JEMs.
-It is computed for counterfactuals generated by constraining model energy and we would therefore argue that it is not unfairly biased.
-Our empirical findings support this argument: firstly, \textit{ECCCo} achieves high faithfulness also for classifiers that have not been trained to model $\mathcal{X}|y$; secondly, our additional results in the appendix for \textit{ECCCo-L1} show that if we do indeed explicitly penalize the unfaithfulness metric, we achieve even better results in this regard.
+Joint energy models (JEM) are indeed explicitly trained to model $\mathcal{X}|y$, but the faithfulness metric is not computed for samples generated by JEMs.
+It is computed for counterfactuals generated by constraining model energy and hence there is no obvious source of bias.
+Our empirical findings support this argument: firstly, \textit{ECCCo} achieves high faithfulness also for classifiers that have not been trained to model $\mathcal{X}|y$; secondly, our additional results in the appendix for \textit{ECCCo-L1} show that if we do indeed explicitly penalize the unfaithfulness metric, we achieve even better results in this regard (also for models not trained to model $\mathcal{X}|y$).
 
-\subsubsection{Add unreliable models.}
+\subsubsection{9. Add unreliable models.}
 
-We would argue that the simple multi-layer perceptrons are unreliable, especially compared to ensembles, joint energy models and convolutional neural networks.
-Simple neural networks are vulnerable to adversarial attacks, which makes them susceptible to implausible counterfactual explanations as we point out in Section 3.
+We would argue that the simple multi-layer perceptrons (MLP) are unreliable, especially compared to ensembles, joint energy models and convolutional neural networks.
+Simple MLPs are generally more vulnerable to adversarial attacks, which makes them susceptible to implausible counterfactual explanations as we point out in Section 3.
 Our results support this notion, in that the quality of counterfactuals produced by \textit{ECCCo} is higher for more reliable models.
-Consistent with the idea proposed by the reviewer, we originally considered introducing `poisoned' VAEs as well, to illustrate what we identify as the key vulnerability of \textit{REVISE}: if the underlying VAE is trained on poisoned data, this will adversely affect counterfactual outcomes as well.
-We ultimately discarded this idea due to limited scope and because we decided that Section 3 sufficiently illustrates our thinking.
+Consistent with the reviewer's idea, we originally considered introducing `poisoned' VAEs to illustrate what we identify as the key vulnerability of \textit{REVISE}: if the underlying VAE is misspecified, this will adversely affect counterfactual outcomes as well.
+We discarded this idea due to limited scope and because we decided that Section 3 sufficiently illustrates our line of thinking.
 

### 2023-11-07 paper/aaai/rebuttal.tex
elaborated on image description

@@
     \includegraphics[width=\linewidth]{../../www/dist_energy.png}
-    \caption{The distance of randomly drawn MNIST images from images in the target class (horizontal axis) plotted against their energy-constrained score (vertical axis).}\label{fig:poc}
+    \caption{The L2 distance of randomly drawn MNIST images with Gaussian perturbations from unperturbed images in the target class (horizontal axis) plotted against their energy-constrained score, i.e.
+target logit (vertical axis).}\label{fig:poc}
 \end{figure}

### 2023-12-11 paper/tables.Rmd
rerendering tables etc

@@
 generators <- unique(dt$generator)[sapply(unique(dt$generator), function(x) {!grepl("L1",x)})]
+
+# Aggregate by run:
 tab <- dt[
@@
     .(
-      value=sprintf("%1.2f ± %1.2f", mean(value), sd(value)),
-      val = mean(value),
-      std = sd(value)
+      val_group = mean(value),
+      std_group = sd(value)
+    ),
+    .(dataname, generator, model, variable, run, source)
+]
+
+# Aggregate accross runs:
+tab <- tab[
+    generator %in% generators,
+    .(
+      value = ifelse(
+        source != "vision",
+        sprintf("%1.2f ± %1.2f", mean(val_group), sd(val_group)),
+        sprintf("%1.3f ± %1.3f", mean(val_group), sd(val_group))
+      ),
+      val = mean(val_group),
+      std = sd(val_group)
     ),
@@
 generators <- unique(dt$generator)[sapply(unique(dt$generator), function(x) {!grepl("L1",x)})]
+
+# Aggregate by run:
 tab_valid <- dt_valid[
@@
     .(
-      value=sprintf("%1.2f ± %1.2f", mean(value), sd(value)),
-      val = mean(value),
-      std = sd(value)
+      val_group = mean(value),
+      std_group = sd(value)
+    ),
+    .(dataname, generator, model, variable, run, source)
+]
+
+# Aggregate accross runs:
+tab_valid <- tab_valid[
+    generator %in% generators,
+    .(
+      value = ifelse(
+        source != "vision",
+        sprintf("%1.2f ± %1.2f", mean(val_group), sd(val_group)),
+        sprintf("%1.3f ± %1.3f", mean(val_group), sd(val_group))
+      ),
+      val = mean(val_group),
+      std = sd(val_group)
     ),
@@
 
-tab_i <- tab
+tab_i <- tab_valid
 
@@
 ```{r}
+# Aggregate by run:
 tab <- dt[
-    ,
+    generator %in% generators,
     .(
-      value=sprintf("%1.2f ± %1.2f", mean(value), sd(value)),
-      val = mean(value),
-      std = sd(value)
+      val_group = mean(value),
+      std_group = sd(value)
+    ),
+    .(dataname, generator, model, variable, run, source)
+]
+
+# Aggregate accross runs:
+tab <- tab[
+    generator %in% generators,
+    .(
+      value =  sprintf("%1.2f ± %1.2f", mean(val_group), sd(val_group)),
+      val = mean(val_group),
+      std = sd(val_group)
     ),
@@
 ```{r}
+# Aggregate by run:
 tab_valid <- dt_valid[
-    ,
+    generator %in% generators,
     .(
-      value=sprintf("%1.2f ± %1.2f", mean(value), sd(value)),
-      val = mean(value),
-      std = sd(value)
+      val_group = mean(value),
+      std_group = sd(value)
+    ),
+    .(dataname, generator, model, variable, run, source)
+]
+
+# Aggregate accross runs:
+tab_valid <- tab_valid[
+    generator %in% generators,
+    .(
+      value = sprintf("%1.2f ± %1.2f", mean(val_group), sd(val_group)),
+      val = mean(val_group),
+      std = sd(val_group)
     ),

### 2023-12-11 paper/tables.Rmd
more updating

@@
 # Aggregate by run:
-tab <- dt[
-    generator %in% generators,
+tab <- dt[,
     .(
@@
 # Aggregate accross runs:
-tab <- tab[
-    generator %in% generators,
+tab <- tab[,
     .(
@@
 # Aggregate by run:
-tab_valid <- dt_valid[
-    generator %in% generators,
+tab_valid <- dt_valid[,
     .(
@@
 # Aggregate accross runs:
-tab_valid <- tab_valid[
-    generator %in% generators,
+tab_valid <- tab_valid[,
     .(

### 2023-12-12 paper/tables.Rmd
choices made for tables

@@
 chosen_model <- c(
-  "MLP",
+  "MLP Ensemble",
   "JEM Ensemble"
@@
 caption <- sprintf(
-  "Results for %s datasets: sample averages +/- one standard deviation across counterfactuals.
+  "Results for %s datasets: sample averages +/- one standard deviation across valid counterfactuals.
 Best outcomes are highlighted in bold.

### 2023-12-12 paper/camera-ready/altmeyer.tex
uh

--- 2023-09-12 AAAI/anonymous-submission-latex-2024.tex
@@
-%File: anonymous-submission-latex-2024.tex
+%File: formatting-instructions-latex-2024.tex
+%release 2024.0
 \documentclass[letterpaper]{article} % DO NOT CHANGE THIS
-\usepackage[submission]{aaai24}  % DO NOT CHANGE THIS
+\usepackage{aaai24}  % DO NOT CHANGE THIS
 \usepackage{times}  % DO NOT CHANGE THIS
@@
 }
-
-\usepackage[utf8]{inputenc} % allow utf-8 input
-\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
-\usepackage{hyperref}       % hyperlinks
-\usepackage{url}            % simple URL typesetting
-\usepackage{booktabs}       % professional-quality tables
-\usepackage{amsfonts}       % blackboard math symbols
-\usepackage{nicefrac}       % compact symbols for 1/2, etc.
-\usepackage{microtype}      % microtypography
-\usepackage{xcolor}         % colors
-
-\usepackage{amsmath}
-\usepackage{amsthm}
-\usepackage{caption}
-\usepackage{graphicx}
-\usepackage{algpseudocode}
-\usepackage{import}
-\usepackage{booktabs}
-\usepackage{longtable}
-\usepackage{array}
-\usepackage{multirow}
-
-% Numbered Environments:
-\newtheorem{definition}{Definition}[section]
-\newtheorem{question}{Research Question}[section]
 
@@
 % directly follow a colon or long dash
-\title{AAAI Press Anonymous Submission\\Instructions for Authors Using \LaTeX{}}
+\title{Faithful Model Explanations through\\
+Energy-Constrained Conformal Counterfactuals}
 \author{
@@
     % All authors must be in the same font size and format.
-    Written by AAAI Press Staff\textsuperscript{\rm 1}\thanks{With help from the AAAI Publications Committee.}\\
-    AAAI Style Contributions by Pater Patel Schneider,
-    Sunil Issar,\\
-    J. Scott Penberthy,
-    George Ferguson,
-    Hans Guesgen,
-    Francisco Cruz\equalcontrib,
-    Marc Pujol-Gonzalez\equalcontrib
+    Patrick Altmeyer\textsuperscript{\rm 1},
+    Mojtaba Farmanbar\textsuperscript{\rm 2},
+    Arie van Deursen\textsuperscript{\rm 1},
+    Cynthia C. S. Liem \textsuperscript{\rm 1}
 }
@@
     %Afiliations
-    \textsuperscript{\rm 1}Association for the Advancement of Artificial Intelligence\\
-    % If you have multiple authors and multiple affiliations
-    % use superscripts in text and roman font to identify them.
-    % For example,
-
-    % Sunil Issar\textsuperscript{\rm 2},
-    % J. Scott Penberthy\textsuperscript{\rm 3},
-    % George Ferguson\textsuperscript{\rm 4},
-    % Hans Guesgen\textsuperscript{\rm 5}
-    % Note that the comma should be placed after the superscript
-
-    1900 Embarcadero Road, Suite 101\\
-    Palo Alto, California 94303-3310 USA\\
-    % email address must be in roman text type, not monospace or sans serif
-    proceedings-questions@aaai.org
+    \textsuperscript{\rm 1}Delft University of Technology,\\
+    \textsuperscript{\rm 2}ING\\
 %
@@
 }
-
-%Example, Single Author, ->> remove \iffalse,\fi and place them surrounding AAAI title to use it
-\iffalse
-\title{My Publication Title --- Single Author}
-\author {
-    Author Name
-}
-\affiliations{
-    Affiliation\\
-    Affiliation Line 2\\
-    name@example.com
-}
-\fi
-
-\iffalse
-%Example, Multiple Authors, ->> remove \iffalse,\fi and place them surrounding AAAI title to use it
-\title{My Publication Title --- Multiple Authors}
-\author {
-    % Authors
-    First Author Name\textsuperscript{\rm 1},
-    Second Author Name\textsuperscript{\rm 2},
-    Third Author Name\textsuperscript{\rm 1}
-}
-\affiliations {
-    % Affiliations
-    \textsuperscript{\rm 1}Affiliation 1\\
-    \textsuperscript{\rm 2}Affiliation 2\\
-    firstAuthor@affiliation1.com, secondAuthor@affilation2.com, thirdAuthor@affiliation1.com
-}
-\fi
-
-% REMOVE THIS: bibentry
-% This is only needed to show inline citations in the guidelines document. You should not need it and can safely delete it.
-\usepackage{bibentry}
-% END REMOVE bibentry
 
@@
 \begin{abstract}
-    Counterfactual Explanations offer an intuitive and straightforward way to explain black-box models and offer Algorithmic Recourse to individuals.
+    Counterfactual explanations offer an intuitive and straightforward way to explain black-box models and offer algorithmic recourse to individuals.
 To address the need for plausible explanations, existing work has primarily relied on surrogate models to learn how the input data is distributed.
@@
 Consequently, the generated explanations may seem plausible to humans but need not necessarily describe the behaviour of the black-box model faithfully.
-We formalise this notion of faithfulness through the introduction of a tailored evaluation metric and propose a novel algorithmic framework for generating \textbf{E}nergy-\textbf{C}onstrained \textbf{C}onformal \textbf{Co}unterfactuals (ECCCos) that are only as plausible as the model permits.
-Through extensive empirical studies, we demonstrate that ECCCos reconcile the need for faithfulness and plausibility.
+We formalise this notion of faithfulness through the introduction of a tailored evaluation metric and propose a novel algorithmic framework for generating \textbf{E}nergy-\textbf{C}onstrained \textbf{C}onformal \textbf{Co}unterfactuals that are only as plausible as the model permits.
+Through extensive empirical studies, we demonstrate that \textit{ECCCo} reconciles the need for faithfulness and plausibility.
 In particular, we show that for models with gradient access, it is possible to achieve state-of-the-art performance without the need for surrogate models.
-To do so, our framework relies solely on properties defining the black-box model itself by leveraging recent advances in Energy-Based Modelling and Conformal Prediction.
-To our knowledge, this is the first venture in this direction for generating faithful Counterfactual Explanations.
-Thus, we anticipate that ECCCos can serve as a baseline for future research.
+To do so, our framework relies solely on properties defining the black-box model itself by leveraging recent advances in energy-based modelling and conformal prediction.
+To our knowledge, this is the first venture in this direction for generating faithful counterfactual explanations.
+Thus, we anticipate that \textit{ECCCo} can serve as a baseline for future research.
 We believe that our work opens avenues for researchers and practitioners seeking tools to better distinguish trustworthy from unreliable models.
@@
 
-  \section{Introduction}\label{intro}
-
-  Counterfactual Explanations (CE) provide a powerful, flexible and intuitive way to not only explain black-box models but also help affected individuals through the means of Algorithmic Recourse.
-Instead of opening the Black Box, CE works under the premise of strategically perturbing model inputs to understand model behaviour~\citep{wachter2017counterfactual}.
-Intuitively speaking, we generate explanations in this context by asking what-if questions of the following nature: `Our credit risk model currently predicts that this individual is not credit-worthy.
-What if they reduced their monthly expenditures by 10\%?'
-
-  This is typically implemented by defining a target outcome $\mathbf{y}^+ \in \mathcal{Y}$ for some individual $\mathbf{x} \in \mathcal{X}=\mathbb{R}^D$ described by $D$ attributes, for which the model $M_{\theta}:\mathcal{X}\mapsto\mathcal{Y}$ initially predicts a different outcome: $M_{\theta}(\mathbf{x})\ne \mathbf{y}^+$.
-Counterfactuals are then searched by minimizing a loss function that compares the predicted model output to the target outcome: $\text{yloss}(M_{\theta}(\mathbf{x}),\mathbf{y}^+)$.
-Since CE work directly with the black-box model, valid counterfactuals always have full local fidelity by construction where fidelity is defined as the degree to which explanations approximate the predictions of a black-box model~\citep{mothilal2020explaining,molnar2020interpretable}.
-
-  In situations where full fidelity is a requirement, CE offer a more appropriate solution to Explainable Artificial Intelligence (XAI) than other popular approaches like LIME~\citep{ribeiro2016why} and SHAP~\citep{lundberg2017unified}, which involve local surrogate models.
-But even full fidelity is not a sufficient condition for ensuring that an explanation faithfully describes the behaviour of a model.
-That is because multiple very distinct explanations can all lead to the same model prediction, especially when dealing with heavily parameterized models like deep neural networks, which are typically underspecified by the data~\citep{wilson2020case}.
-
-  In the context of CE, the idea that no two explanations are the same arises almost naturally.
-A key focus in the literature has therefore been to identify those explanations and algorithmic recourses that are most appropriate based on a myriad of desiderata such as sparsity, actionability and plausibility.
-In this work, we draw closer attention to model faithfulness rather than fidelity as a desideratum for counterfactuals.
-Our key contributions are as follows:
-
-  \begin{itemize}
-    \item We show that fidelity is an insufficient evaluation metric for counterfactuals (Section~\ref{fidelity}) and propose a definition of faithfulness that gives rise to more suitable metrics (Section~\ref{faithfulness}).
-    \item We introduce a novel algorithmic approach for generating Energy-Constrained Conformal Counterfactuals (ECCCos) in Section~\ref{meth}.
-    \item We provide extensive empirical evidence demonstrating that ECCCos faithfully explain model behaviour and attain plausibility only when appropriate (Section~\ref{emp}).
-  \end{itemize}
-
-  To our knowledge, this is the first venture in this direction for generating faithful counterfactuals.
-Thus, we anticipate that ECCCos can serve as a baseline for future research.
-We believe that our work opens avenues for researchers and practitioners seeking tools to better distinguish trustworthy from unreliable models.
-
-  \section{Background}\label{background}
-
-  While CE can also be generated for arbitrary regression models~\citep{spooner2021counterfactual}, existing work has primarily focused on classification problems.
-Let $\mathcal{Y}=(0,1)^K$ denote the one-hot-encoded output domain with $K$ classes.
-Then most counterfactual generators rely on gradient descent to optimize different flavours of the following counterfactual search objective:
-
-  \begin{equation} \label{eq:general}
-  \begin{aligned}
-  \mathbf{Z}^\prime &= \arg \min_{\mathbf{Z}^\prime \in \mathcal{Z}^L} \left\{  {\text{yloss}(M_{\theta}(f(\mathbf{Z}^\prime)),\mathbf{y}^+)}+ \lambda {\text{cost}(f(\mathbf{Z}^\prime)) }  \right\} 
-  \end{aligned} 
-  \end{equation}
-
-  Here $\text{yloss}(\cdot)$ denotes the primary loss function, $f(\cdot)$ is a function that maps from the counterfactual state space to the feature space and $\text{cost}(\cdot)$ is either a single penalty or a collection of penalties that are used to impose constraints through regularization.
-Equation~\ref{eq:general} restates the baseline approach to gradient-based counterfactual search proposed by~\citet{wachter2017counterfactual} in general form as introduced by~\citet{altmeyer2023endogenous}.
-To explicitly account for the multiplicity of explanations, $\mathbf{Z}^\prime=\{ \mathbf{z}_l\}_L$ denotes an $L$-dimensional array of counterfactual states.
-
-  The baseline approach, which we will simply refer to as \textit{Wachter}, searches a single counterfactual directly in the feature space and penalises its distance to the original factual.
-In this case, $f(\cdot)$ is simply the identity function and $\mathcal{Z}$ corresponds to the feature space itself.
-Many derivative works of~\citet{wachter2017counterfactual} have proposed new flavours of Equation~\ref{eq:general}, each of them designed to address specific \textit{desiderata} that counterfactuals ought to meet in order to properly serve both AI practitioners and individuals affected by algorithmic decision-making systems.
-The list of desiderata includes but is not limited to the following: sparsity, proximity~\citep{wachter2017counterfactual}, actionability~\citep{ustun2019actionable}, diversity~\citep{mothilal2020explaining}, plausibility~\citep{joshi2019realistic,poyiadzi2020face,schut2021generating}, robustness~\citep{upadhyay2021robust,pawelczyk2022probabilistically,altmeyer2023endogenous} and causality~\citep{karimi2021algorithmic}.
-Different counterfactual generators addressing these needs have been extensively surveyed and evaluated in various studies~\citep{verma2020counterfactual,karimi2020survey,pawelczyk2021carla,artelt2021evaluating,guidotti2022counterfactual}.
-
-  Perhaps unsurprisingly, the different desiderata are often positively correlated.
-For example, \citet{artelt2021evaluating} find that plausibility typically also leads to improved robustness.
-Similarly, plausibility has also been connected to causality in the sense that plausible counterfactuals respect causal relationships~\citep{mahajan2020preserving}.
-Consequently, the plausibility of counterfactuals has been among the primary concerns for researchers.
-Achieving plausibility is equivalent to ensuring that the generated counterfactuals comply with the true and unobserved data-generating process (DGP).
-We define plausibility formally in this work as follows:
-
-  \begin{definition}[Plausible Counterfactuals]
-    \label{def:plausible}
-    Let $\mathcal{X}|\mathbf{y}^+= p(\mathbf{x}|\mathbf{y}^+)$ denote the true conditional distribution of samples in the target class $\mathbf{y}^+$.
-Then for $\mathbf{x}^{\prime}$ to be considered a plausible counterfactual, we need: $\mathbf{x}^{\prime} \sim \mathcal{X}|\mathbf{y}^+$.
-  \end{definition}
-
-  To generate plausible counterfactuals, we need to be able to quantify the DGP: $\mathcal{X}|\mathbf{y}^+$.
-One straightforward way to do this is to use surrogate models for the task.
-\citet{joshi2019realistic}, for example, suggest that instead of searching counterfactuals in the feature space $\mathcal{X}$, we can instead traverse a latent embedding $\mathcal{Z}$ (Equation~\ref{eq:general}) that implicitly codifies the DGP.
-To learn the latent embedding, they propose using a generative model such as a Variational Autoencoder (VAE).
-Provided the surrogate model is well-specified, their proposed approach called \textit{REVISE} can yield plausible explanations.
-Others have proposed similar approaches: \citet{dombrowski2021diffeomorphic} traverse the base space of a normalizing flow to solve Equation~\ref{eq:general}; \citet{poyiadzi2020face} use density estimators ($\hat{p}: \mathcal{X} \mapsto [0,1]$) to constrain the counterfactuals to dense regions in the feature space; and, finally, \citet{karimi2021algorithmic} assume knowledge about the structural causal model that generates the data.
-
-  A competing approach towards plausibility that is also closely related to this work instead relies on the black-box model itself.
-\citet{schut2021generating} show that to meet the plausibility objective we need not explicitly model the input distribution.
-Pointing to the undesirable engineering overhead induced by surrogate models, they propose that we rely on the implicit minimisation of predictive uncertainty instead.
-Their proposed methodology, which we will refer to as \textit{Schut}, solves Equation~\ref{eq:general} by greedily applying Jacobian-Based Saliency Map Attacks (JSMA) in the feature space with cross-entropy loss and no penalty at all.
-The authors demonstrate theoretically and empirically that their approach yields counterfactuals for which the model $M_{\theta}$ predicts the target label $\mathbf{y}^+$ with high confidence.
-Provided the model is well-specified, these counterfactuals are plausible.
-This idea hinges on the assumption that the black-box model provides well-calibrated predictive uncertainty estimates.
-
-  \section{Why Fidelity is not Enough}\label{fidelity}
-
-  As discussed in the introduction, any valid counterfactual also has full fidelity by construction: solutions to Equation~\ref{eq:general} are considered valid as soon as the label predicted by the model matches the target class.
-So while fidelity always applies, counterfactuals that address the various desiderata introduced above can look vastly different from each other.
-
-  To demonstrate this with an example, we have trained a simple image classifier $M_{\theta}$ on the well-known \textit{MNIST} dataset~\citep{lecun1998mnist}: a Multi-Layer Perceptron (\textit{MLP}) with above 90 percent test accuracy.
-No measures have been taken to improve the model's adversarial robustness or its capacity for predictive uncertainty quantification.
-The far left panel of Figure ~\ref{fig:motiv} shows a random sample drawn from the dataset.
-The underlying classifier correctly predicts the label `nine' for this image.
-For the given factual image and model, we have used \textit{Wachter}, \textit{Schut} and \textit{REVISE} to generate one counterfactual each in the target class `seven'.
-The perturbed images are shown next to the factual image from left to right in Figure ~\ref{fig:motiv}.
-Captions on top of the individual images indicate the generator along with the predicted probability that the image belongs to the target class.
-In all three cases that probability is above 90 percent and yet the counterfactuals look very different from each other.
-
-  \begin{figure}
-    \centering
-    \includegraphics[width=0.8\textwidth]{../artifacts/results/images/mnist_motivation.png}
-    \caption{Counterfactuals for turning a 9 (nine) into a 7 (seven): original image (left); then from left to right the counterfactuals generated using \textit{Wachter}, \textit{Schut} and \textit{REVISE}.}\label{fig:motiv}
-  \end{figure}
-
-  Since \textit{Wachter} is only concerned with proximity, the generated counterfactual is almost indistinguishable from the factual.
-The approach by~\citet{schut2021generating} expects a well-calibrated model that can generate predictive uncertainty estimates.
-Since this is not the case, the generated counterfactual looks like an adversarial example.
-Finally, the counterfactual generated by \textit{REVISE} looks much more plausible than the other two.
-But is it also more faithful to the behaviour of our \textit{MNIST} classifier?
-That is much less clear because the surrogate used by \textit{REVISE} introduces friction: the generated explanations no longer depend exclusively on the black-box model itself.
-
-  So which of the counterfactuals most faithfully explains the behaviour of our image classifier?
-Fidelity cannot help us to make that judgement, because all of these counterfactuals have full fidelity.
-Thus, fidelity is an insufficient evaluation metric to assess the faithfulness of CE.
-
-  \section{A New Notion of Faithfulness}\label{faithfulness}
-
-  Considering the limitations of fidelity as demonstrated in the previous section, analogous to Definition~\ref{def:plausible}, we introduce a new notion of faithfulness in the context of CE:
-
-  \begin{definition}[Faithful Counterfactuals]
-    \label{def:faithful}
-    Let $\mathcal{X}_{\theta}|\mathbf{y}^+ = p_{\theta}(\mathbf{x}|\mathbf{y}^+)$ denote the conditional distribution of $\mathbf{x}$ in the target class $\mathbf{y}^+$, where $\theta$ denotes the parameters of model $M_{\theta}$.
-Then for $\mathbf{x}^{\prime}$ to be considered a faithful counterfactual, we need: $\mathbf{x}^{\prime} \sim \mathcal{X}_{\theta}|\mathbf{y}^+$.
-  \end{definition}
-
-  In doing this, we merge in and nuance the concept of plausibility (Definition~\ref{def:plausible}) where the notion of `consistent with the data' becomes `consistent with what the model has learned about the data'.
-
-  \subsection{Quantifying the Model's Generative Property}
-
-  To assess counterfactuals with respect to Definition~\ref{def:faithful}, we need a way to quantify the posterior conditional distribution $p_{\theta}(\mathbf{x}|\mathbf{y}^+)$.
-To this end, we draw on recent advances in Energy-Based Modelling (EBM), a subdomain of machine learning that is concerned with generative or hybrid modelling~\citep{grathwohl2020your,du2020implicit}.
-In particular, note that if we fix $\mathbf{y}$ to our target value $\mathbf{y}^+$, we can conditionally draw from $p_{\theta}(\mathbf{x}|\mathbf{y}^+)$ by randomly initializing $\mathbf{x}_0$ and then using Stochastic Gradient Langevin Dynamics (SGLD) as follows,
-
-  \begin{equation}\label{eq:sgld}
-    \begin{aligned}
-      \mathbf{x}_{j+1} &\leftarrow \mathbf{x}_j - \frac{\epsilon^2}{2} \mathcal{E}(\mathbf{x}_j|\mathbf{y}^+) + \epsilon \mathbf{r}_j, && j=1,...,J
-    \end{aligned}
-  \end{equation}
-
-  where $\mathbf{r}_j \sim \mathcal{N}(\mathbf{0},\mathbf{I})$ is the stochastic term and the step-size $\epsilon$ is typically polynomially decayed~\citep{welling2011bayesian}.
-The term $\mathcal{E}(\mathbf{x}_j|\mathbf{y}^+)$ denotes the model energy conditioned on the target class label $\mathbf{y}^+$ which we specify as the negative logit corresponding to the target class label $\mathbf{y}^*$.
-To allow for faster sampling, we follow the common practice of choosing the step-size $\epsilon$ and the standard deviation of $\mathbf{r}_j$ separately.
-While $\mathbf{x}_J$ is only guaranteed to distribute as $p_{\theta}(\mathbf{x}|\mathbf{y}^*)$ if $\epsilon \rightarrow 0$ and $J \rightarrow \infty$, the bias introduced for a small finite $\epsilon$ is negligible in practice \citep{murphy2023probabilistic,grathwohl2020your}.
-Appendix~\ref{app:jem} provides additional implementation details for any tasks related to energy-based modelling.
-
-  Generating multiple samples using SGLD thus yields an empirical distribution $\hat{\mathbf{X}}_{\theta,\mathbf{y}^+}$ that approximates what the model has learned about the input data.
-While in the context of EBM, this is usually done during training, we propose to repurpose this approach during inference in order to evaluate and generate faithful model explanations.
-
-  \subsection{Evaluating Plausibility and Faithfulness}
-
-  The parallels between our definitions of plausibility and faithfulness imply that we can also use similar evaluation metrics in both cases.
-Since existing work has focused heavily on plausibility, it offers a useful starting point.
-In particular,~\citet{guidotti2022counterfactual} have proposed an implausibility metric that measures the distance of the counterfactual from its nearest neighbour in the target class.
-As this distance is reduced, counterfactuals get more plausible under the assumption that the nearest neighbour itself is plausible in the sense of Definition~\ref{def:plausible}.
-In this work, we use the following adapted implausibility metric,
-
-  \begin{equation}\label{eq:impl}
-    \begin{aligned}
-      \text{impl}(\mathbf{x}^{\prime},\mathbf{X}_{\mathbf{y}^+}) = \frac{1}{\lvert\mathbf{X}_{\mathbf{y}^+}\rvert} \sum_{\mathbf{x} \in \mathbf{X}_{\mathbf{y}^+}} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
-    \end{aligned}
-  \end{equation}
-
-  where $\mathbf{x}^{\prime}$ denotes the counterfactual and $\mathbf{X}_{\mathbf{y}^+}$ is a subsample of the training data in the target class $\mathbf{y}^+$.
-By averaging over multiple samples in this manner, we avoid the risk that the nearest neighbour of $\mathbf{x}^{\prime}$ itself is not plausible according to Definition~\ref{def:plausible} (e.g an outlier).
-
-  Equation~\ref{eq:impl} gives rise to a similar evaluation metric for unfaithfulness.
-We merely swap out the subsample of individuals in the target class for a subset $\hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^+}$ of the generated conditional samples:
-
-  \begin{equation}\label{eq:faith}
-    \begin{aligned}
-      \text{unfaith}(\mathbf{x}^{\prime},\hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^+}) = \frac{1}{\lvert \hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^+}\rvert} \sum_{\mathbf{x} \in \hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^+}} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
-    \end{aligned}
-  \end{equation}
-
-  Specifically, we form this subset based on the $n_E$ generated samples with the lowest energy.
-
-  \section{Energy-Constrained Conformal Counterfactuals}\label{meth}
-
-  In this section, we describe \textit{ECCCo}, our proposed framework for generating Energy-Constrained Conformal Counterfactuals (ECCCos).
-It is based on the premise that counterfactuals should first and foremost be faithful.
-Plausibility, as a secondary concern, is then still attainable, but only to the degree that the black-box model itself has learned plausible explanations for the underlying data.
-
-  We begin by stating our proposed objective function, which involves tailored loss and penalty functions that we will explain in the following.
-In particular, we extend Equation~\ref{eq:general} as follows:
-
-  \begin{equation} \label{eq:eccco}
-    \begin{aligned}
-    \mathbf{Z}^\prime= \arg \min_{\mathbf{Z}^\prime \in \mathcal{Z}^M}  &\{  {\text{yloss}(M_{\theta}(f(\mathbf{Z}^\prime)),\mathbf{y}^+)}+ \lambda_{1} {\text{dist}(f(\mathbf{Z}^\prime),\mathbf{x}) } \\
-    &+ \lambda_2 \text{unfaith}(f(\mathbf{Z}^\prime),\hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^+}) + \lambda_3 \Omega(C_{\theta}(f(\mathbf{Z}^\prime);\alpha)) \} 
-    \end{aligned} 
-  \end{equation}
-
-  The first penalty term involving $\lambda_1$ induces proximity like in~\citet{wachter2017counterfactual}.
-Our default choice for $\text{dist}(\cdot)$ is the L1 Norm due to its sparsity-inducing properties.
-The second penalty term involving $\lambda_2$ induces faithfulness by constraining the energy of the generated counterfactual where $\text{unfaith}(\cdot)$ corresponds to the metric defined in Equation~\ref{eq:faith}.
-The third and final penalty term involving $\lambda_3$ introduces a new concept: it ensures that the generated counterfactual is associated with low predictive uncertainty.
-As mentioned above,~\citet{schut2021generating} have shown that plausible counterfactuals can be generated implicitly through predictive uncertainty minimization.
-Unfortunately, this relies on the assumption that the model itself can provide predictive uncertainty estimates, which may be too restrictive in practice.
-
-  To relax this assumption, we leverage recent advances in Conformal Prediction (CP), an approach to predictive uncertainty quantification that has recently gained popularity~\citep{angelopoulos2021gentle,manokhin2022awesome}.
-Crucially for our intended application, CP is model-agnostic and can be applied during inference without placing any restrictions on model training.
-Intuitively, CP works under the premise of turning heuristic notions of uncertainty into rigorous uncertainty estimates by repeatedly sifting through the training data or a dedicated calibration dataset.
-Conformal classifiers produce prediction sets for individual inputs that include all output labels that can be reasonably attributed to the input.
-These sets tend to be larger for inputs that do not conform with the training data and are characterized by high predictive uncertainty.
-
-  In order to generate counterfactuals that are associated with low predictive uncertainty, we use a smooth set size penalty introduced by~\citet{stutz2022learning} in the context of conformal training:
-
-  \begin{equation}\label{eq:setsize}
-    \begin{aligned}
-      \Omega(C_{\theta}(\mathbf{x};\alpha))&=\max \left(0, \sum_{\mathbf{y}\in\mathcal{Y}}C_{\theta,\mathbf{y}}(\mathbf{x}_i;\alpha) - \kappa \right)
-    \end{aligned}
-  \end{equation}
-
-  Here, $\kappa \in \{0,1\}$ is a hyper-parameter and $C_{\theta,\mathbf{y}}(\mathbf{x}_i;\alpha)$ can be interpreted as the probability of label $\mathbf{y}$ being included in the prediction set.
-In order to compute this penalty for any black-box model we merely need to perform a single calibration pass through a holdout set $\mathcal{D}_{\text{cal}}$.
-Arguably, data is typically abundant and in most applications, practitioners tend to hold out a test data set anyway.
-Consequently, CP removes the restriction on the family of predictive models, at the small cost of reserving a subset of the available data for calibration.
-This particular case of conformal prediction is referred to as Split Conformal Prediction (SCP) as it involves splitting the training data into a proper training dataset and a calibration dataset.
-In addition to the smooth set size penalty, we have also experimented with the use of a tailored function for $\text{yloss}(\cdot)$ that enforces that only the target label $\mathbf{y}^+$ is included in the prediction set ~\citet{stutz2022learning}.
-Further details are provided in Appendix~\ref{app:cp}.
-
-  \begin{figure}
-    \centering
-    \includegraphics[width=1.0\textwidth]{../artifacts/results/images/poc_gradient_fields.png}
-    \caption{Gradient fields and counterfactual paths for different generators.
-The objective is to generate a counterfual in the `blue' class for a sample from the `orange' class.
-Bright yellow stars indicate conditional samples generated through SGLD.
-The underlying classifier is a Joint Energy Model.}\label{fig:poc}
-  \end{figure}
-
-  \renewcommand{\algorithmicrequire}{\textbf{Input:}}
-  \renewcommand{\algorithmicensure}{\textbf{Output:}}
-
-  \begin{algorithm}
-    \caption{The \textit{ECCCo} generator}\label{alg:eccco}
-    \begin{algorithmic}[1]
-      \Require $\mathbf{x}, \mathbf{y}^+, M_{\theta}, f, \Lambda=[\lambda_1,\lambda_2,\lambda_3], \alpha, \mathcal{D}, T, \eta, n_{\mathcal{B}}, n_E$ where $M_{\theta}(\mathbf{x})\neq\mathbf{y}^+$
-      \Ensure $\mathbf{x}^\prime$
-      \State Initialize $\mathbf{z}^\prime \gets f^{-1}(\mathbf{x})$ \Comment{Map to counterfactual state space.}
-      \State Generate $\left\{\hat{\mathbf{x}}_{\theta,\mathbf{y}^+}\right\}_{n_{\mathcal{B}}} \gets p_{\theta}(\mathbf{x}_{\mathbf{y}^+})$ \Comment{Generate $n_{\mathcal{B}}$ samples using SGLD (Equation~\ref{eq:sgld}).}
-      \State Store $\hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^+} \gets \left\{\hat{\mathbf{x}}_{\theta,\mathbf{y}^+}\right\}_{n_{\mathcal{B}}}$ \Comment{Choose $n_E$ lowest-energy samples.}
-      \State Run \textit{SCP} for $M_{\theta}$ using $\mathcal{D}$ \Comment{Calibrate model through Split Conformal Prediction.}
-      \State Initialize $t \gets 0$
-      \While{\textit{not converged} or $t < T$} \Comment{For convergence conditions see Appendix~\ref{app:eccco}.}
-      \State $\mathbf{z}^\prime \gets \mathbf{z}^\prime - \eta \nabla_{\mathbf{z}^\prime} \mathcal{L}(\mathbf{z}^\prime,\mathbf{y}^+,\hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^+}; \Lambda, \alpha)$ \Comment{Take gradient step of size $\eta$.}
-      \State $t \gets t+1$
-      \EndWhile
-      \State $\mathbf{x}^\prime \gets f(\mathbf{z}^\prime)$ \Comment{Map back to feature space.}
-    \end{algorithmic}
-  \end{algorithm}
-
-  To provide some further intuition about our objective defined in Equation~\ref{eq:eccco}, Figure~\ref{fig:poc} illustrates how the different components affect the counterfactual search for a synthetic dataset.
-The underlying classifier is a Joint Energy Model (\textit{JEM}) that was trained to predict the output class (`blue' or `orange') and generate class-conditional samples~\citep{grathwohl2020your}.
-We have used four different generator flavours to produce a counterfactual in the `blue' class for a sample from the `orange' class: \textit{Wachter}, which only uses the first penalty ($\lambda_2=\lambda_3=0$); \textit{ECCCo (no EBM)}, which does not constrain energy ($\lambda_2=0$); \textit{ECCCo (no CP)}, which involves no set size penalty ($\lambda_3=0$); and, finally, \textit{ECCCo}, which involves all penalties defined in Equation~\ref{eq:eccco}.
-Arrows indicate (negative) gradients with respect to the objective function at different points in the feature space.
-
-  While \textit{Wachter} generates a valid counterfactual, it ends up close to the original starting point consistent with its objective.
-\textit{ECCCo (no EBM)} pushes the counterfactual further into the target domain to minimize predictive uncertainty, but the outcome is still not plausible.
-The counterfactual produced by \textit{ECCCo (no CP)} is attracted by the generated samples shown in bright yellow.
-Since the \textit{JEM} has learned the conditional input distribution reasonably well in this case, the counterfactuals are both faithful and plausible.
-Finally, the outcome for \textit{ECCCo} looks similar, but the additional smooth set size penalty leads to somewhat faster convergence.
-
-  Algorithm~\ref{alg:eccco} describes how exactly \textit{ECCCo} works.
-For the sake of simplicity and without loss of generality, we limit our attention to generating a single counterfactual $\mathbf{x}^\prime=f(\mathbf{z}^\prime)$.
-The counterfactual state $\mathbf{z}^\prime$ is initialized by passing the factual $\mathbf{x}$ through a simple feature transformer $f^{-1}$.
-Next, we generate $n_{\mathcal{B}}$ conditional samples $\hat{\mathbf{x}}_{\theta,\mathbf{y}^+}$ using SGLD (Equation~\ref{eq:sgld}) and store the $n_E$ instances with the lowest energy.
-We then calibrate the model $M_{\theta}$ through Split Conformal Prediction.
-Finally, we search counterfactuals through gradient descent where $\mathcal{L}(\mathbf{z}^\prime,\mathbf{y}^+,\hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^+}; \Lambda, \alpha)$ denotes our loss function defined in Equation~\ref{eq:eccco}.
-The search terminates once the convergence criterium is met or the maximum number of iterations $T$ has been exhausted.
-Note that the choice of convergence criterium has important implications on the final counterfactual which we explain in Appendix~\ref{app:eccco}.
-
-  \section{Empirical Analysis}\label{emp}
-
-  Our goal in this section is to shed light on the following research questions:
-
-  \begin{question}[Faithfulness]\label{rq:faithfulness}
-    Are ECCCos more faithful than counterfactuals produced by our benchmark generators?
-  \end{question}
-
-  \begin{question}[Balancing Objectives]\label{rq:plausibility}
-    Compared to our benchmark generators, how do ECCCos balance the two key objectives of faithfulness and plausibility?
-  \end{question}
-
-  The second question is motivated by the intuition that faithfulness and plausibility should coincide for models that have learned plausible explanations of the data.
-Next, we first briefly describe our experimental setup before presenting our main results.
-
-  \subsection{Experimental Setup}
-
-  To assess and benchmark the performance of our proposed generator against the state of the art, we generate multiple counterfactuals for different models and datasets.
-In particular, we compare \textit{ECCCo} and its variants to the following counterfactual generators that were introduced above: firstly; \textit{Schut}, which works under the premise of minimizing predictive uncertainty; secondly, \textit{REVISE}, which is state-of-the-art with respect to plausibility; and, finally, \textit{Wachter}, which serves as our baseline.
-
-  We use both synthetic and real-world datasets from different domains, all of which are publicly available and commonly used to train and benchmark classification algorithms.
-We synthetically generate a dataset containing two \textit{Linearly Separable} Gaussian clusters ($n=1000$), as well as the well-known \textit{Circles} ($n=1000$) and \textit{Moons} ($n=2500$) data.
-Since these data are generated by distributions of varying degrees of complexity, they allow us to assess how the generators and our proposed evaluation metrics handle this.
-
-  As for real-world data, we follow~\citet{schut2021generating} and use the \textit{MNIST}~\citep{lecun1998mnist} dataset containing images of handwritten digits such as the example shown above in Figure~\ref{fig:motiv}.
-From the social sciences domain, we include Give Me Some Credit (\textit{GMSC})~\citep{kaggle2011give}: a tabular dataset that has been studied extensively in the literature on Algorithmic Recourse~\citep{pawelczyk2021carla}.
-It consists of 11 numeric features that can be used to predict the binary outcome variable indicating whether retail borrowers experience financial distress.
-
-  For the predictive modelling tasks, we use simple neural networks (\textit{MLP}) and Joint Energy Models (\textit{JEM}).
-For the more complex real-world datasets we also use ensembling in each case.
-Both joint-energy modelling and ensembling have been associated with improved generative properties and adversarial robustness~\citep{grathwohl2020your,lakshminarayanan2016simple}, so we expect this to be positively correlated with the plausibility of ECCCos.
-To account for stochasticity, we generate multiple counterfactuals for each target class, generator, model and dataset.
-Specifically, we randomly sample $n^{-}$ times from the subset of individuals for which the given model predicts the non-target class $\mathbf{y}^{-}$ given the current target.
-We set $n^{-}=25$ for all of our synthetic datasets, $n^{-}=10$ for \textit{GMSC} and $n^{-}=5$ for \textit{MNIST}.
-Full details concerning our parameter choices, training procedures and model performance can be found in Appendix~\ref{app:setup}.
-
-  \subsection{Results for Synthetic Data}
-
-  Table~\ref{tab:results-synthetic} shows the key results for the synthetic datasets separated by model (first column) and generator (second column).
-The numerical columns show sample averages and standard deviations of our key evaluation metrics computed across all counterfactuals.
-We have highlighted the best outcome for each model and metric in bold.
-To provide some sense of effect sizes, we have added asterisks to indicate that a given value is at least one ($*$) or two ($**$) standard deviations lower than the baseline (\textit{Wachter}).
-
-  Starting with the high-level results for our \textit{Linearly Separable} data, we find that \textit{ECCCo} produces the most faithful counterfactuals for both black-box models.
-This is consistent with our design since \textit{ECCCo} directly enforces faithfulness through regularization.
-Crucially though, \textit{ECCCo} also produces the most plausible counterfactuals for both models.
-This dataset is so simple that even the \textit{MLP} has learned plausible explanations of the input data.
-Zooming in on the granular details for the \textit{Linearly Separable} data, the results for \textit{ECCCo (no CP)} and \textit{ECCCo (no EBM)} indicate that the positive results are dominated by the effect of quantifying and leveraging the model's generative property (EBM).
-Conformal Prediction alone only leads to marginally improved faithfulness and plausibility.
-
-  The findings for the \textit{Moons} dataset are broadly in line with the findings so far: for the \textit{JEM}, \textit{ECCCo} yields substantially more faithful and plausible counterfactuals than all other generators.
-For the \textit{MLP}, faithfulness is maintained but counterfactuals are not plausible.
-This high-level pattern is broadly consistent with other more complex datasets and supportive of our narrative, so it is worth highlighting: ECCCos consistently achieve high faithfulness, which---subject to the quality of the model itself---coincides with high plausibility.
-By comparison, \textit{REVISE} yields the most plausible counterfactuals for the \textit{MLP}, but it does so at the cost of faithfulness.
-We also observe that the best results for \textit{ECCCo} are achieved when using both penalties.
-Once again though, the generative component (EBM) has a stronger impact on the positive results for the \textit{JEM}.
-
-  For the \textit{Circles} data, it appears that \textit{REVISE} performs well, but we note that it generates valid counterfactuals only half of the time (see Appendix~\ref{app:results} for a complete overview including additional common evaluation metrics).
-The underlying VAE with default parameters has not adequately learned the data-generating process.
-Of course, it is possible to improve generative performance through hyperparameter tuning but this example serves to illustrate that \textit{REVISE} depends on the quality of its surrogate.
-Independent of the outcome for \textit{REVISE}, however, the results do not seem to indicate that \textit{ECCCo} substantially improves faithfulness and plausibility for the \textit{Circles} data.
-We think this points to a limitation of our evaluation metrics rather than \textit{ECCCo} itself: computing average distances fails to account for the `wraparound' effect associated with circular data~\citep{gill2010circular}.
-
-  \import{contents/}{table-synthetic.tex}
-
-  \subsection{Results for Real-World Data}
-
-  The results for our real-world datasets are shown in Table~\ref{tab:results-real-world}.
-Once again the findings indicate that the plausibility of ECCCos is positively correlated with the capacity of the black-box model to distinguish plausible from implausible inputs.
-The case is very clear for \textit{MNIST}: ECCCos are consistently more faithful than the counterfactuals produced by our benchmark generators and their plausibility gradually improves through ensembling and joint-energy modelling.
-Interestingly, faithfulness also gradually improves for \textit{REVISE}.
-This indicates that as our models improve, their generative capacity approaches that of the surrogate VAE used by \textit{REVISE}.
-The VAE still outperforms our classifiers in this regard, as evident from the fact that \textit{ECCCo} never quite reaches the same level of plausibility as \textit{REVISE}.
-With reference to Appendix~\ref{app:results} we note that the results for \textit{Schut} need to be discounted as it rarely produces valid counterfactuals for \textit{MNIST}.
-Relatedly, we find that \textit{ECCCo} is the only generator that consistently achieves full validity.
-Finally, it is worth noting that \textit{ECCCo} produces counterfactual images with the lowest average predictive uncertainty for all models.
-
-  For the tabular credit dataset (\textit{GMSC}) it is inherently challenging to use deep neural networks in order to achieve good discriminative performance~\citep{borisov2021deep,grinsztajn2022why} and generative performance~\citep{liu2023goggle}, respectively.
-In order to achieve high plausibility, \textit{ECCCo} effectively requires classifiers to achieve good performance for both tasks.
-Since this is a challenging task even for Joint Energy Models, it is not surprising to find that even though \textit{ECCCo} once again achieves state-of-the-art faithfulness, it is outperformed by \textit{REVISE} and \textit{Schut} with respect to plausibility.
-
-  \subsection{Key Takeways}
-
-  To conclude this section, we summarize our findings with reference to the opening questions.
-The results clearly demonstrate that \textit{ECCCo} consistently achieves state-of-the-art faithfulness, as it was designed to do (Research Question~\ref{rq:faithfulness}).
-A related important finding is that \textit{ECCCo} yields highly plausible explanations provided that they faithfully describe model behaviour (Research Question~\ref{rq:plausibility}).
-\textit{ECCCo} achieves this result primarily by leveraging the model's generative property.
-
-  \import{contents/}{table-real-world.tex}
-
-  \section{Limitations}
-
-  Even though we have taken considerable measures to study our proposed methodology carefully, limitations can still be identified.
-In particular, we have found that the performance of \textit{ECCCo} is sensitive to hyperparameter choices.
-In order to achieve faithfulness, we generally had to penalise the distance from generated samples slightly more than the distance from factual values.
-
-  Conversely, we have not found that strongly penalising prediction set sizes had any discernable effect.
-Our results indicate that CP alone is often not sufficient to achieve faithfulness and plausibility, although we acknowledge that this needs to be investigated more thoroughly through future work.
-
-  While our approach is readily applicable to models with gradient access like deep neural networks, more work is needed to generalise it to other machine learning models such as decision trees.
-Relatedly, common challenges associated with Energy-Based Modelling including sensitivity to scale, training instabilities and sensitivity to hyperparameters also apply to \textit{ECCCo}.
-
-  \section{Conclusion}
-
-  This work leverages recent advances in Energy-Based Modelling and Conformal Prediction in the context of Explainable Artificial Intelligence.
-We have proposed a new way to generate counterfactuals that are maximally faithful to the black-box model they aim to explain.
-Our proposed generator, \textit{ECCCo}, produces plausible counterfactuals if and only if the black-box model itself has learned realistic explanations for the data, which we have demonstrated through rigorous empirical analysis.
-This should enable researchers and practitioners to use counterfactuals in order to discern trustworthy models from unreliable ones.
-While the scope of this work limits its generalizability, we believe that \textit{ECCCo} offers a solid baseline for future work on faithful Counterfactual Explanations.
-
-  \section{Acknowledgments}
-
-  Some of the members of TU Delft were partially funded by ICAI AI for Fintech Research, an ING — TU Delft
-  collaboration.
-
-  \bibliography{bib}
+\bibliography{aaai24}
 

### 2023-12-12 paper/camera-ready/altmeyer.tex
first version rendering

@@
 
+% Additional Packages:
+\usepackage{amsfonts}       % blackboard math symbols
+\usepackage{amsmath}
+\usepackage{amsthm}
+\usepackage{caption}
+\usepackage{graphicx}
+\usepackage{algpseudocode}
+\usepackage{import}
+\usepackage{booktabs}
+\usepackage{longtable}
+\usepackage{array}
+\usepackage{multirow}
+\usepackage{placeins}
+
+% Numbered Environments:
+\newtheorem{definition}{Definition}[section]
+\newtheorem{question}{Research Question}[section]
+
 \setcounter{secnumdepth}{0} %May be changed to 1 or 2 if section numbers are desired.
@@
 % directly follow a colon or long dash
+
 \title{Faithful Model Explanations through\\
@@
 
+\section{Introduction}\label{intro}
+
+Counterfactual explanations provide a powerful, flexible and intuitive way to not only explain black-box models but also offer the possibility of algorithmic recourse to affected individuals.
+Instead of opening the black box, counterfactual explanations work under the premise of strategically perturbing model inputs to understand model behaviour~\citep{wachter2017counterfactual}.
+Intuitively speaking, we generate explanations in this context by asking what-if questions of the following nature: `Our credit risk model currently predicts that this individual is not credit-worthy.
+What if they reduced their monthly expenditures by 10\%?'
+
+This is typically implemented by defining a target outcome $\mathbf{y}^+ \in \mathcal{Y}$ for some individual $\mathbf{x} \in \mathcal{X}=\mathbb{R}^D$ described by $D$ attributes, for which the model $M_{\theta}:\mathcal{X}\mapsto\mathcal{Y}$ initially predicts a different outcome: $M_{\theta}(\mathbf{x})\ne \mathbf{y}^+$.
+Counterfactuals are then searched by minimizing a loss function that compares the predicted model output to the target outcome: $\text{yloss}(M_{\theta}(\mathbf{x}),\mathbf{y}^+)$.
+Since counterfactual explanations work directly with the black-box model, valid counterfactuals always have full local fidelity by construction where fidelity is defined as the degree to which explanations approximate the predictions of a black-box model~\citep{molnar2022interpretable}.
+
+In situations where full fidelity is a requirement, counterfactual explanations offer a more appropriate solution to Explainable Artificial Intelligence (XAI) than other popular approaches like LIME~\citep{ribeiro2016why} and SHAP~\citep{lundberg2017unified}, which involve local surrogate models.
+But even full fidelity is not a sufficient condition for ensuring that an explanation \textit{faithfully} describes the behaviour of a model.
+That is because multiple distinct explanations can lead to the same model prediction, especially when dealing with heavily parameterized models like deep neural networks, which are underspecified by the data~\citep{wilson2020case}.
+In the context of counterfactuals, the idea that no two explanations are the same arises almost naturally.
+A key focus in the literature has therefore been to identify those explanations that are most appropriate based on a myriad of desiderata such as closeness~\citep{wachter2017counterfactual}, sparsity~\citep{schut2021generating}, actionability~\citep{ustun2019actionable} and plausibility~\citep{joshi2019realistic}.
+
+In this work, we draw closer attention to model faithfulness rather than fidelity as a desideratum for counterfactuals.
+We define faithfulness as the degree to which counterfactuals are consistent with what the model has learned about the data.
+Our key contributions are as follows: first, we show that fidelity is an insufficient evaluation metric for counterfactuals (Section~\ref{fidelity}) and propose a definition of faithfulness that gives rise to more suitable metrics (Section~\ref{faithfulness}).
+Next, we introduce a \textit{ECCCo}: a novel algorithmic approach aimed at generating energy-constrained conformal counterfactuals that faithfully explain model behaviour in Section~\ref{meth}.
+Finally, we provide extensive empirical evidence demonstrating that \textit{ECCCo} faithfully explains model behaviour and attains plausibility only when appropriate (Section~\ref{emp}).
+
+To our knowledge, this is the first venture in this direction for generating faithful counterfactuals.
+Thus, we anticipate that \textit{ECCCo} can serve as a baseline for future research.
+We believe that our work opens avenues for researchers and practitioners seeking tools to better distinguish trustworthy from unreliable models.
+
+\section{Background}\label{background}
+
+While counterfactual explanations (CE) can also be generated for arbitrary regression models~\citep{spooner2021counterfactual}, existing work has primarily focused on classification problems.
+Let $\mathcal{Y}=(0,1)^K$ denote the one-hot-encoded output domain with $K$ classes.
+Then most counterfactual generators rely on gradient descent to optimize different flavours of the following counterfactual search objective:
+
+\begin{equation} \label{eq:general}
+\begin{aligned}
+\mathbf{Z}^\prime &= \arg \min_{\mathbf{Z}^\prime \in \mathcal{Z}^L} \left\{  {\text{yloss}(M_{\theta}(f(\mathbf{Z}^\prime)),\mathbf{y}^+)}+ \lambda {\text{cost}(f(\mathbf{Z}^\prime)) }  \right\} 
+\end{aligned} 
+\end{equation}
+
+Here $\text{yloss}(\cdot)$ denotes the primary loss function, $f(\cdot)$ is a function that maps from the counterfactual state space to the feature space and $\text{cost}(\cdot)$ is either a single penalty or a collection of penalties that are used to impose constraints through regularization.
+Equation~\ref{eq:general} restates the baseline approach to gradient-based counterfactual search proposed by~\citet{wachter2017counterfactual} in general form as introduced by~\citet{altmeyer2023endogenous}.
+To explicitly account for the multiplicity of explanations, $\mathbf{Z}^\prime=\{ \mathbf{z}_l\}_L$ denotes an $L$-dimensional array of counterfactual states.
+
+The baseline approach, which we will simply refer to as \textit{Wachter}, searches a single counterfactual directly in the feature space and penalises its distance to the original factual.
+In this case, $f(\cdot)$ is simply the identity function and $\mathcal{Z}$ corresponds to the feature space itself.
+Many derivative works of~\citet{wachter2017counterfactual} have proposed new flavours of Equation~\ref{eq:general}, each of them designed to address specific \textit{desiderata} that counterfactuals ought to meet in order to properly serve both AI practitioners and individuals affected by algorithmic decision-making systems.
+The list of desiderata includes but is not limited to the following: sparsity, closeness~\citep{wachter2017counterfactual}, actionability~\citep{ustun2019actionable}, diversity~\citep{mothilal2020explaining}, plausibility~\citep{joshi2019realistic,poyiadzi2020face,schut2021generating}, robustness~\citep{upadhyay2021robust,pawelczyk2022probabilistically,altmeyer2023endogenous} and causality~\citep{karimi2021algorithmic}.
+Different counterfactual generators addressing these needs have been extensively surveyed and evaluated in various studies~\citep{verma2020counterfactual,karimi2020survey,pawelczyk2021carla,artelt2021evaluating,guidotti2022counterfactual}.
+
+The notion of plausibility is central to all of the desiderata.
+For example, \citet{artelt2021evaluating} find that plausibility typically also leads to improved robustness.
+Similarly, plausibility has also been connected to causality in the sense that plausible counterfactuals respect causal relationships~\citep{mahajan2019preserving}.
+Consequently, the plausibility of counterfactuals has been among the primary concerns for researchers.
+Achieving plausibility is equivalent to ensuring that the generated counterfactuals comply with the true and unobserved data-generating process (DGP).
+We define plausibility formally in this work as follows:
+
+\begin{definition}[Plausible Counterfactuals]
+  \label{def:plausible}
+  Let $\mathcal{X}|\mathbf{y}^+= p(\mathbf{x}|\mathbf{y}^+)$ denote the true conditional distribution of samples in the target class $\mathbf{y}^+$.
+Then for $\mathbf{x}^{\prime}$ to be considered a plausible counterfactual, we need: $\mathbf{x}^{\prime} \sim \mathcal{X}|\mathbf{y}^+$.
+\end{definition}
+
+To generate plausible counterfactuals, we first need to quantify the conditional distribution of samples in the target class ($\mathcal{X}|\mathbf{y}^+$).
+We can then ensure that we generate counterfactuals that comply with that distribution.
+
+One straightforward way to do this is to use surrogate models for the task.
+\citet{joshi2019realistic}, for example, suggest that instead of searching counterfactuals in the feature space $\mathcal{X}$, we can instead traverse a latent embedding $\mathcal{Z}$ (Equation~\ref{eq:general}) that implicitly codifies the DGP.
+To learn the latent embedding, they propose using a generative model such as a Variational Autoencoder (VAE).
+Provided the surrogate model is well-specified, their proposed approach \textit{REVISE} can yield plausible explanations.
+Others have proposed similar approaches: \citet{dombrowski2021diffeomorphic} traverse the base space of a normalizing flow to solve Equation~\ref{eq:general}; \citet{poyiadzi2020face} use density estimators ($\hat{p}: \mathcal{X} \mapsto [0,1]$) to constrain the counterfactuals to dense regions in the feature space; and, finally, \citet{karimi2021algorithmic} assume knowledge about the structural causal model that generates the data.
+
+A competing approach towards plausibility that is also closely related to this work instead relies on the black-box model itself.
+\citet{schut2021generating} show that to meet the plausibility objective we need not explicitly model the input distribution.
+Pointing to the undesirable engineering overhead induced by surrogate models, they propose that we rely on the implicit minimisation of predictive uncertainty instead.
+Their proposed methodology, which we will refer to as \textit{Schut}, solves Equation~\ref{eq:general} by greedily applying Jacobian-Based Saliency Map Attacks (JSMA) in the feature space with cross-entropy loss and no penalty at all.
+The authors demonstrate theoretically and empirically that their approach yields counterfactuals for which the model $M_{\theta}$ predicts the target label $\mathbf{y}^+$ with high confidence.
+Provided the model is well-specified, these counterfactuals are plausible.
+This idea hinges on the assumption that the black-box model provides well-calibrated predictive uncertainty estimates.
+
+\section{Why Fidelity is not Enough: A Motivational Example}\label{fidelity}
+
+As discussed in the introduction, any valid counterfactual also has full fidelity by construction: solutions to Equation~\ref{eq:general} are considered valid as soon as the label predicted by the model matches the target class.
+So while fidelity always applies, counterfactuals that address the various desiderata introduced above can look vastly different from each other.
+
+To demonstrate this with an example, we have trained a simple image classifier $M_{\theta}$ on the well-known \textit{MNIST} dataset~\citep{lecun1998mnist}: a Multi-Layer Perceptron (\textit{MLP}) with test set accuracy $> 0.9$.
+No measures have been taken to improve the model's adversarial robustness or its capacity for predictive uncertainty quantification.
+The far left panel of Figure ~\ref{fig:motiv} shows a random sample drawn from the dataset.
+The underlying classifier correctly predicts the label `nine' for this image.
+For the given factual image and model, we have used \textit{Wachter}, \textit{Schut} and \textit{REVISE} to generate one counterfactual each in the target class `seven'.
+The perturbed images are shown next to the factual image from left to right in Figure ~\ref{fig:motiv}.
+Captions on top of the images indicate the generator along with the predicted probability that the image belongs to the target class.
+In all cases, that probability is very high, while the counterfactuals look very different.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.8\linewidth]{figures/mnist_motivation.png}
+  \caption{Counterfactuals for turning a 9 (nine) into a 7 (seven): original image (left), then the counterfactuals generated using \textit{Wachter}, \textit{Schut} and \textit{REVISE}.}\label{fig:motiv}
+\end{figure}
+
+Since \textit{Wachter} is only concerned with closeness, the generated counterfactual is almost indistinguishable from the factual.
+The approach by~\citet{schut2021generating} expects a well-calibrated model that can generate predictive uncertainty estimates.
+Since this is not the case, the generated counterfactual looks like an adversarial example.
+Finally, the counterfactual generated by \textit{REVISE} looks much more plausible than the other two.
+But is it also more faithful to the behaviour of our \textit{MNIST} classifier?
+That is much less clear because the surrogate used by \textit{REVISE} introduces friction: the generated explanations no longer depend exclusively on the black-box model itself.
+
+So which of the counterfactuals most faithfully explains the behaviour of our image classifier?
+Fidelity cannot help us to make that judgement, because all of these counterfactuals have full fidelity.
+Thus, fidelity is an insufficient evaluation metric to assess the faithfulness of CE.
+
+\section{Faithful first, Plausible second}\label{faithfulness}
+
+Considering the limitations of fidelity as demonstrated in the previous section, analogous to Definition~\ref{def:plausible}, we introduce a new notion of faithfulness in the context of CE:
+
+\begin{definition}[Faithful Counterfactuals]
+  \label{def:faithful}
+  Let $\mathcal{X}_{\theta}|\mathbf{y}^+ = p_{\theta}(\mathbf{x}|\mathbf{y}^+)$ denote the conditional distribution of $\mathbf{x}$ in the target class $\mathbf{y}^+$, where $\theta$ denotes the parameters of model $M_{\theta}$.
+Then for $\mathbf{x}^{\prime}$ to be considered a faithful counterfactual, we need: $\mathbf{x}^{\prime} \sim \mathcal{X}_{\theta}|\mathbf{y}^+$.
+\end{definition}
+
+In doing this, we merge in and nuance the concept of plausibility (Definition~\ref{def:plausible}) where the notion of `consistent with the data' becomes `consistent with what the model has learned about the data'.
+
+\subsection{Quantifying the Model's Generative Property}
+
+To assess counterfactuals with respect to Definition~\ref{def:faithful}, we need a way to quantify the posterior conditional distribution $p_{\theta}(\mathbf{x}|\mathbf{y}^+)$.
+To this end, we draw on ideas from energy-based modelling (EBM), a subdomain of machine learning that is concerned with generative or hybrid modelling~\citep{grathwohl2020your,du2019implicit}.
+In particular, note that if we fix $\mathbf{y}$ to our target value $\mathbf{y}^+$, we can conditionally draw from $p_{\theta}(\mathbf{x}|\mathbf{y}^+)$ by randomly initializing $\mathbf{x}_0$ and then using Stochastic Gradient Langevin Dynamics (SGLD) as follows,
+
+\begin{equation}\label{eq:sgld}
+  \begin{aligned}
+    \mathbf{x}_{j+1} &\leftarrow \mathbf{x}_j - \frac{\epsilon_j^2}{2} \mathcal{E}_{\theta}(\mathbf{x}_j|\mathbf{y}^+) + \epsilon_j \mathbf{r}_j, && j=1,...,J
+  \end{aligned}
+\end{equation}
+
+where $\mathbf{r}_j \sim \mathcal{N}(\mathbf{0},\mathbf{I})$ is the stochastic term and the step-size $\epsilon_j$ is typically polynomially decayed~\citep{welling2011bayesian}.
+The term $\mathcal{E}_{\theta}(\mathbf{x}_j|\mathbf{y}^+)$ denotes the model energy conditioned on the target class label $\mathbf{y}^+$ which we specify as the negative logit corresponding to the target class label $\mathbf{y}^{+}$.
+To allow for faster sampling, we follow the common practice of choosing the step-size $\epsilon_j$ and the standard deviation of $\mathbf{r}_j$ separately.
+While $\mathbf{x}_J$ is only guaranteed to distribute as $p_{\theta}(\mathbf{x}|\mathbf{y}^{+})$ if $\epsilon \rightarrow 0$ and $J \rightarrow \infty$, the bias introduced for a small finite $\epsilon$ is negligible in practice \citep{murphy2023probabilistic}.
+
+Generating multiple samples using SGLD thus yields an empirical distribution $\widehat{\mathbf{X}}_{\theta,\mathbf{y}^+}$ that approximates what the model has learned about the input data.
+While in the context of EBM, this is usually done during training, we propose to repurpose this approach during inference in order to evaluate the faithfulness of model explanations.
+The technical appendix provides additional implementation details for any tasks related to energy-based modelling.
+
+\subsection{Quantifying the Model's Predictive Uncertainty}
+
+Faithful counterfactuals can be expected to also be plausible if the learned conditional distribution $\mathcal{X}_{\theta}|\mathbf{y}^+$ (Defintion~\ref{def:faithful}) is close to the true conditional distribution $\mathcal{X}|\mathbf{y}^+$ (Definition~\ref{def:plausible}).
+We can further improve the plausibility of counterfactuals without the need for surrogate models that may interfere with faithfulness by minimizing predictive uncertainty~\citep{schut2021generating}.
+Unfortunately, this idea relies on the assumption that the model itself provides predictive uncertainty estimates, which may be too restrictive in practice.
+
+To relax this assumption, we use conformal prediction (CP), an approach to predictive uncertainty quantification that has recently gained popularity~\citep{angelopoulos2021gentle,manokhin2022awesome}.
+Crucially for our intended application, CP is model-agnostic and can be applied during inference without placing any restrictions on model training.
+It works under the premise of turning heuristic notions of uncertainty into rigorous estimates by repeatedly sifting through the training data or a dedicated calibration dataset.
+
+Conformal classifiers produce prediction sets for individual inputs that include all output labels that can be reasonably attributed to the input. These sets are formed as follows,
+
+\begin{equation}\label{eq:scp}
+  \begin{aligned}
+    C_{\theta}(\mathbf{x}_i;\alpha)=\{\mathbf{y}: s(\mathbf{x}_i,\mathbf{y}) \le \hat{q}\}
+  \end{aligned}
+\end{equation}
+
+where $\hat{q}$ denotes the $(1-\alpha)$-quantile of $\mathcal{S}$ and $\alpha$ is a predetermined error rate.
+These sets tend to be larger for inputs that do not conform with the training data and are characterized by high predictive uncertainty.
+To leverage this notion of predictive uncertainty in the context of gradient-based counterfactual search, we use a smooth set size penalty introduced by~\citet{stutz2022learning}:
+
+\begin{equation}\label{eq:setsize}
+  \begin{aligned}
+    \Omega(C_{\theta}(\mathbf{x};\alpha))&=\max \left(0, \sum_{\mathbf{y}\in\mathcal{Y}}C_{\theta,\mathbf{y}}(\mathbf{x}_i;\alpha) - \kappa \right)
+  \end{aligned}
+\end{equation}
+
+Here, $\kappa \in \{0,1\}$ is a hyper-parameter and $C_{\theta,\mathbf{y}}(\mathbf{x}_i;\alpha)$ can be interpreted as the probability of label $\mathbf{y}$ being included in the prediction set (see appendix for details).
+In order to compute this penalty for any black-box model, we merely need to perform a single calibration pass through a holdout set $\mathcal{D}_{\text{cal}}$.
+Arguably, data is typically abundant and in most applications, practitioners tend to hold out a test data set anyway.
+Consequently, CP removes the restriction on the family of predictive models, at the small cost of reserving a subset of the available data for calibration.
+This particular case of conformal prediction is referred to as \textit{split conformal prediction} (SCP) as it involves splitting the training data into a proper training dataset and a calibration dataset.
+
+\subsection{Evaluating Plausibility and Faithfulness}
+
+The parallels between our definitions of plausibility and faithfulness imply that we can also use similar evaluation metrics in both cases.
+Since existing work has focused heavily on plausibility, it offers a useful starting point.
+In particular,~\citet{guidotti2022counterfactual} have proposed an implausibility metric that measures the distance of the counterfactual from its nearest neighbour in the target class.
+As this distance is reduced, counterfactuals get more plausible under the assumption that the nearest neighbour itself is plausible in the sense of Definition~\ref{def:plausible}.
+In this work, we use the following adapted implausibility metric,
+
+\begin{equation}\label{eq:impl}
+  \begin{aligned}
+    \text{impl}(\mathbf{x}^{\prime},\mathbf{X}_{\mathbf{y}^+}) = \frac{1}{\lvert\mathbf{X}_{\mathbf{y}^+}\rvert} \sum_{\mathbf{x} \in \mathbf{X}_{\mathbf{y}^+}} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
+  \end{aligned}
+\end{equation}
+
+where $\mathbf{x}^{\prime}$ denotes the counterfactual and $\mathbf{X}_{\mathbf{y}^+}$ is a subsample of the training data in the target class $\mathbf{y}^+$.
+By averaging over multiple samples in this manner, we avoid the risk that the nearest neighbour of $\mathbf{x}^{\prime}$ itself is not plausible according to Definition~\ref{def:plausible} (e.g an outlier).
+
+Equation~\ref{eq:impl} gives rise to a similar evaluation metric for unfaithfulness.
+We swap out the subsample of observed individuals in the target class for the set of samples generated through SGLD ($\widehat{\mathbf{X}}_{\mathbf{y}^+}$):
+
+\begin{equation}\label{eq:faith}
+  \begin{aligned}
+    \text{unfaith}(\mathbf{x}^{\prime},\widehat{\mathbf{X}}_{\theta,\mathbf{y}^+}) = \frac{1}{\lvert \widehat{\mathbf{X}}_{\theta,\mathbf{y}^+} \rvert} \sum_{\mathbf{x} \in \widehat{\mathbf{X}}_{\theta,\mathbf{y}^+}} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
+  \end{aligned}
+\end{equation}
+
+Our default choice for the $\text{dist}(\cdot)$ function in both cases is the Euclidean Norm.
+Depending on the type of input data other choices may be more adequate, which we discuss further in Section~\ref{emp:setup}.
+
+\section{Energy-Constrained Conformal Counterfactuals}\label{meth}
+
+Given our proposed notion of faithfulness, we now describe \textit{ECCCo}, our proposed framework for generating Energy-Constrained Conformal Counterfactuals.
+It is based on the premise that counterfactuals should first and foremost be faithful.
+Plausibility, as a secondary concern, is then still attainable to the degree that the black-box model itself has learned plausible explanations for the underlying data.
+
+We begin by substituting the loss function in Equation~\ref{eq:general},
+
+\begin{equation} \label{eq:eccco-start}
+  \begin{aligned}
+  \mathbf{Z}^\prime =& \arg \min_{\mathbf{Z}^\prime \in \mathcal{Z}^L} \{  {L_{\text{JEM}}(f(\mathbf{Z}^\prime);M_{\theta},\mathbf{y}^+)}+ \lambda {\text{cost}(f(\mathbf{Z}^\prime)) } \} 
+  \end{aligned} 
+\end{equation}
+
+where $L_{\text{JEM}}(f(\mathbf{Z}^\prime);M_{\theta},\mathbf{y}^+)$ is a hybrid loss function used in joint-energy modelling evaluated at a given counterfactual state for a given model and target outcome:
+
+\begin{equation}
+  \begin{aligned}
+    L_{\text{JEM}}(f(\mathbf{Z}^\prime); \cdot) = L_{\text{clf}}(f(\mathbf{Z}^\prime); \cdot) + L_{\text{gen}}(f(\mathbf{Z}^\prime); \cdot)
+  \end{aligned}
+\end{equation}
+
+The first term, $L_{\text{clf}}$, is any standard classification loss function such as cross-entropy loss.
+The second term, $L_{\text{gen}}$, is used to measure loss with respect to the generative task\footnote{In practice, regularization loss is typically also added.
+We follow this convention but have omitted the term here for simplicity.}.
+In the context of joint-energy training, $L_{\text{gen}}$ induces changes in model parameters $\theta$ that decrease the energy of observed samples and increase the energy of samples generated through SGLD~\citep{du2019implicit}.
+
+The key observation in our context is that we can rely solely on decreasing the energy of the counterfactual itself.
+This is sufficient to capture the generative property of the underlying model since it is implicitly captured by its parameters $\theta$.
+Importantly, this means that we do not need to generate conditional samples through SGLD during our counterfactual search at all as we explain in the technical appendix.
+
+This observation leads to the following simple objective function for \textit{ECCCo}:
+
+\begin{equation} \label{eq:eccco}
+  \begin{aligned}
+  \mathbf{Z}^\prime =& \arg \min_{\mathbf{Z}^\prime \in \mathcal{Z}^L} \{  {L_{\text{clf}}(f(\mathbf{Z}^\prime);M_{\theta},\mathbf{y}^+)}+ \lambda_1 {\text{cost}(f(\mathbf{Z}^\prime)) } \\
+  &+ \lambda_2 \mathcal{E}_{\theta}(f(\mathbf{Z}^\prime)|\mathbf{y}^+) + \lambda_3 \Omega(C_{\theta}(f(\mathbf{Z}^\prime);\alpha)) \} 
+  \end{aligned} 
+\end{equation}
+
+The first penalty term involving $\lambda_1$ induces closeness like in~\citet{wachter2017counterfactual}.
+The second penalty term involving $\lambda_2$ induces faithfulness by constraining the energy of the generated counterfactual.
+The third and final penalty term involving $\lambda_3$ ensures that the generated counterfactual is associated with low predictive uncertainty.
+To tune theses hyperparameters we have relied on grid search.
+
+Concerning feature autoencoding ($f: \mathcal{Z} \mapsto \mathcal{X}$), \textit{ECCCo} does not rely on latent space search to achieve its primary objective of faithfulness.
+By default, we choose $f(\cdot)$ to be the identity function as in \textit{Wachter}.
+This is generally also enough to achieve plausibility, provided the model has learned plausible explanations for the data.
+In some cases, plausibility can be improved further by mapping counterfactuals to a lower-dimensional latent space.
+In the following, we refer to this approach as \textit{ECCCo+}: that is, \textit{ECCCo} plus dimensionality reduction.
+
+\begin{figure*}
+  \centering
+  \includegraphics[width=0.75\linewidth]{figures/poc_gradient_fields.png}
+  \caption{Gradient fields and counterfactual paths for different generators.
+The objective is to generate a counterfactual in the blue class for a sample from the orange class.
+Bright yellow stars indicate conditional samples generated through SGLD.
+The underlying classifier is a Joint Energy Model.}\label{fig:poc}
+\end{figure*}
+
+Figure~\ref{fig:poc} illustrates how the different components in Equation~\ref{eq:eccco} affect the counterfactual search for a synthetic dataset.
+The underlying classifier is a Joint Energy Model (\textit{JEM}) that was trained to predict the output class (blue or orange) and generate class-conditional samples~\citep{grathwohl2020your}.
+We have used four different generator flavours to produce a counterfactual in the blue class for a sample from the orange class: \textit{Wachter}, which only uses the first penalty ($\lambda_2=\lambda_3=0$); \textit{ECCCo (no EBM)}, which does not constrain energy ($\lambda_2=0$); \textit{ECCCo (no CP)}, which involves no set size penalty ($\lambda_3=0$); and, finally, \textit{ECCCo}, which involves all penalties defined in Equation~\ref{eq:eccco}.
+Arrows indicate (negative) gradients with respect to the objective function at different points in the feature space.
+
+While \textit{Wachter} generates a valid counterfactual, it ends up close to the original starting point consistent with its objective.
+\textit{ECCCo (no EBM)} pushes the counterfactual further into the target domain to minimize predictive uncertainty, but the outcome is still not plausible.
+The counterfactual produced by \textit{ECCCo (no CP)} is energy-constrained.
+Since the \textit{JEM} has learned the conditional input distribution reasonably well in this case, the counterfactuals are both faithful and plausible.
+Finally, the outcome for \textit{ECCCo} looks similar, but the additional smooth set size penalty leads to somewhat faster convergence.
+
+\section{Empirical Analysis}\label{emp}
+
+Our goal in this section is to shed light on the following research questions:
+
+\begin{question}[Faithfulness]\label{rq:faithfulness}
+  To what extent are counterfactuals generated by \textit{ECCCo} more faithful than those produced by state-of-the-art generators?
+\end{question}
+
+\begin{question}[Balancing Desiderata]\label{rq:plausibility}
+  Compared to state-of-the-art generators, how does \textit{ECCCo} balance the two key objectives of faithfulness and plausibility?
+\end{question}
+
+The second question is motivated by the intuition that faithfulness and plausibility should coincide for models that have learned plausible explanations of the data.
+
+\subsection{Experimental Setup}\label{emp:setup}
+
+To assess and benchmark the performance of our proposed generator against the state of the art, we generate multiple counterfactuals for different models and datasets.
+In particular, we compare \textit{ECCCo} and its variants to the following counterfactual generators that were introduced above: firstly; \textit{Schut}, which works under the premise of minimizing predictive uncertainty; secondly, \textit{REVISE}, which is state-of-the-art (SOTA) with respect to plausibility; and, finally, \textit{Wachter}, which serves as our baseline.
+In the case of \textit{ECCCo+}, we use principal component analysis (PCA) for dimensionality reduction: the latent space $\mathcal{Z}$ is spanned by the first $n_z$ principal components where we choose $n_z$ to be equal to the latent dimension of the VAE used by \textit{REVISE}.
+
+For the predictive modelling tasks, we use multi-layer perceptrons (\textit{MLP}), deep ensembles, joint energy models (\textit{JEM}) and convolutional neural networks (LeNet-5 \textit{CNN}~\citep{lecun1998gradient}).
+Both joint-energy modelling and ensembling have been associated with improved generative properties and adversarial robustness~\citep{grathwohl2020your,lakshminarayanan2016simple}, so we expect this to be positively correlated with the plausibility of \textit{ECCCo}.
+To account for stochasticity, we generate multiple counterfactuals for each target class, generator, model and dataset.
+Full details concerning our parameter choices, training procedures and model performance can be found in the appendix.
+
+We perform benchmarks on eight datasets from different domains.
+From the credit and finance domain we include three tabular datasets: Give Me Some Credit (\textit{GMSC})~\citep{kaggle2011give}, \textit{German Credit}~\citet{hoffman1994german} and \textit{California Housing}~\citet{pace1997sparse}.
+All of these are commonly used in the related literature~\citep{karimi2020survey,altmeyer2023endogenous,pawelczyk2021carla}.
+Following related literature~\citep{schut2021generating,dhurandhar2018explanations} we also include two image datasets: \textit{MNIST}~\citep{lecun1998mnist} and \textit{Fashion MNIST}~\citep{xiao2017fashion}.
+Detailed descriptions and results for all datasets can be found in the appendix.
+
+In the following, we will focus on the most relevant results highlighted in Tables~\ref{tab:results-tabular} and~\ref{tab:results-vision}.
+The tables show sample averages along with standard deviations for our key evaluation metrics for the \textit{California Housing} and \textit{GMSC} datasets (Table~\ref{tab:results-tabular}) and the \textit{MNIST} dataset (Table~\ref{tab:results-vision}).
+For each metric, the best outcomes are highlighted in bold.
+Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (Wachter).
+For the tabular datasets, we use the default Euclidian distance to measure unfaithfulness and implausibility as defined in Equations~\ref{eq:faith} and~\ref{eq:impl}, respectively.
+The third metric presented~\ref{tab:results-tabular} in Table quantifies the predictive uncertainty of the counterfactual as measured by Equation~\ref{eq:setsize}.
+For the vision datasets, we rely on measuring the structural dissimilarity between images for our unfaithfulness and implausibility metrics~\citep{wang2003multiscale}.
+
+\subsection{Faithfulness}
+
+Overall, we find strong empirical evidence suggesting that \textit{ECCCo} consistently achieves state-of-the-art faithfulness.
+Across all models and datasets highlighted here, all variations of \textit{ECCCo} consistently outperform all other generators with respect to faithfulness, in many cases substantially.
+This pattern is mostly robust across all other benchmark datasets (Tables~\ref{tab:results-linearly-separable} to~\ref{tab:results-fashion-mnist} in the technical appendix).
+
+In particular, we note that the best results are generally obtained when using the full \textit{ECCCo} objective (Equation~\ref{eq:eccco}).
+In other words, constraining both energy and predictive uncertainty typically yields the most faithful counterfactuals.
+We expected the former to play a more significant role in this context and that is typically what we find across all datasets.
+For example, the results for \textit{GMSC} in Table~\ref{tab:results-tabular} indicate that faithfulness can be improved substantially by relying solely on the energy constraint (\textit{ECCCo (no CP)}).
+In some cases though, as for the \textit{California Housing} dataset, \textit{ECCCo (no EBM)} actually outperforms \textit{ECCCo (no CP)}.
+This indicates that predictive uncertainty minimization plays an important role in achieving faithfulness.
+
+We also generally find that the highest degree of faithfulness is obtained when the counterfactual search is performed directly in the feature space $\mathcal{X}$.
+While \textit{ECCCo+} typically attains high levels of faithfulness compared to most other generators, it is consistently outperformed by \textit{ECCCo}.
+The case is even stronger for \textit{REVISE}, which performs worst out of all generators for faithfulness on the \textit{GMSC} dataset and better only than \textit{Wachter} on \textit{California Housing}.
+
+These findings are consistent with the notion that surrogate models may inhibit faithfulness.
+Even though dimensionality reduction through PCA in the case of \textit{ECCCo+} can be considered a relatively mild form of intervention, the first $n_z$ principal components fail to capture some of the variation in the data, that the underlying model itself may be sensitive to.
+This notion is illustrated nicely in Figure~\ref{fig:mnist-bmk}, where the counterfactual produced by \textit{ECCCo} is somewhat noisier and grainier than the one produced by \textit{ECCCo+}.
+
+In conclusion, we recommend in light of the findings here to use the full \textit{ECCCo} search objective whenever model faithfulness is a key priority.
+
+\subsection{Balancing Desiderata}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=1.0\linewidth]{figures/mnist_benchmark.png}
+  \caption{Counterfactuals for turning a 3 into a 5: factual (left), then the counterfactuals generated by \textit{ECCCo}, \textit{ECCCo+}, \textit{REVISE}, \textit{Schut} and \textit{Wachter}.}\label{fig:mnist-bmk}
+\end{figure}
+
+% Sample UUID of factual: UUID("1ceb06f0-5949-11ee-0e9c-dd49ccfec8c3")
+
+Overall, we find strong empirical evidence suggesting that \textit{ECCCo} can achieve near state-of-the-art plausibility without sacrificing faithfulness.
+Figure~\ref{fig:mnist-bmk} shows one such example taken from the \textit{MNIST} benchmark where the objective is to turn the factual three (far left) into a five.
+The underlying model is a LeNet-5 \textit{CNN}.
+The different images show the counterfactuals produced by the generators, of which all but the one produced by \textit{Schut} are valid.
+Both variations of \textit{ECCCo} produce plausible counterfactuals.
+
+Looking at the benchmark results presented in Tables~\ref{tab:results-tabular} and~\ref{tab:results-vision} we firstly note that although \textit{REVISE} generally performs best, \textit{ECCCo} and in particular \textit{ECCCo+} often approach SOTA performance.
+Upon visual inspection of the generated images we actually find that \textit{ECCCo+} performs much better than \textit{REVISE} (see appendix).
+Zooming in on the details we observe that \textit{ECCCo} and its variations do particularly well, whenever the underlying model has been explicitly trained to learn plausible representations of the data.
+For both tabular datasets in Table~\ref{tab:results-tabular}, \textit{ECCCo} improves plausibility substantially compared to the baseline.
+This broad pattern is mostly consistent for all other datasets, although there are notable exceptions for which \textit{ECCCo} takes the lead on both plausibility and faithfulness (see, for example, Tables~\ref{tab:results-moons} and~\ref{tab:results-german-credit} in the appendix).
+
+While we maintain that generally speaking plausibility should hinge on the quality of the model, our results also indicate that it is possible to trade off some degree of faithfulness for plausibility if needed: \textit{ECCCo+} generally outperforms other variants of \textit{ECCCo} in this context at the small cost of slightly reduced faithfulness.
+For the vision datasets especially, we find that  \textit{ECCCo+} is consistently second only to \textit{REVISE} for all models and regularly substantially better than the baseline.
+Looking at the \textit{California Housing} data, latent space search markedly improves plausibility without sacrificing faithfulness: for the \textit{JEM} Ensemble, \textit{ECCCo+} performs substantially better than the baseline and only marginally worse than \textit{REVISE}.
+Importantly, \textit{ECCCo+} does not attain plausibility at all costs: for the MLP, plausibility is still very low but this seems to faithfully represent what the model has learned.
+
+We conclude that \textit{ECCCo} offers us a way to balance the objectives of faithfulness and plausibility. \textit{ECCCo+} can be used to tilt the scale in favour of plausibility if needed.
+
+\subsection{Additional Desiderata}
+
+While we have deliberately focused on our key metrics of interest so far, it is worth briefly considering other common desiderata for counterfactuals.
+With reference to the right-most columns for each dataset in Table~\ref{tab:results-tabular}, we firstly note that \textit{ECCCo} typically reduces predictive uncertainty as intended.
+Consistent with its design, \textit{Schut} performs well on this metric even though it does not explicitly address uncertainty as measured by conformal prediction set sizes.
+
+Another commonly discussed desideratum is closeness~\citep{wachter2017counterfactual}: counterfactuals that are closer to their factuals are associated with smaller costs to individuals in the context of algorithmic recourse.
+As evident from the additional tables in the appendix, the closeness desideratum tends to be negatively correlated with plausibility and faithfulness.
+Consequently, both \textit{REVISE} and \textit{ECCCo} generally yield more costly counterfactuals than the baseline.
+Nonetheless, \textit{ECCCo} does not seem to stretch costs unnecessarily: in Figure~\ref{fig:mnist-bmk} useful parts of the factual three are clearly retained.
+
+\section{Limitations}
+
+Despite having taken considerable measures to study our methodology carefully, limitations can still be identified.
+
+Firstly, we recognise that our proposed distance-based evaluation metrics for plausibility and faithfulness may not be universally applicable to all types of data.
+In any case, they depend on choosing a distance metric on a case-by-case basis, as we have done in this work.
+Arguably, commonly used metrics for measuring other desiderata such as closeness suffer from the same pitfall.
+We therefore think that future work on counterfactual explanations could benefit from defining universal evaluation metrics.
+
+Relatedly, we note that our proposed metric for measuring faithfulness depends on the availability of samples generated through SGLD, which in turn requires gradient access for models.
+This means it cannot be used to evaluate non-differentiable classifiers.
+Consequently, we also have not applied \textit{ECCCo} to some machine learning models commonly used for classification such as decision trees.
+Since \textit{ECCCo} itself does not rely on SGLD, its defining penalty functions are indeed applicable to gradient-free counterfactual generators.
+This is an interesting avenue for future research.
+
+Next, common challenges associated with energy-based modelling including sensitivity to scale, training instabilities and sensitivity to hyperparameters also apply to \textit{ECCCo} to some extent.
+In grid searches for optimal hyperparameters, we have noticed that unless properly regularized, \textit{ECCCo} is sometimes prone to overshoot for the energy constraint.
+
+Finally, while we have used ablation to understand the roles of the different components of \textit{ECCCo}, the scope of this work has prevented us from investigating the role of conformal prediction in this context more thoroughly.
+We have exclusively relied on split conformal prediction and have used fixed values for the predetermined error rate and other hyperparameters.
+Future work could benefit from more extensive ablation studies that tune hyperparameters and investigate different approaches to conformal prediction.
+
+\section{Conclusion}
+
+This work leverages ideas from energy-based modelling and conformal prediction in the context of counterfactual explanations.
+We have proposed a new way to generate counterfactuals that are maximally faithful to the black-box model they aim to explain.
+Our proposed generator, \textit{ECCCo}, produces plausible counterfactuals iff the black-box model itself has learned realistic explanations for the data, which we have demonstrated through rigorous empirical analysis.
+This should enable researchers and practitioners to use counterfactuals in order to discern trustworthy models from unreliable ones.
+While the scope of this work limits its generalizability, we believe that \textit{ECCCo} offers a solid base for future work on faithful counterfactual explanations.
+
 \bibliography{aaai24}

### 2023-12-13 paper/camera-ready/altmeyer.tex
sorted out references
tables added in

@@
 
-\setcounter{secnumdepth}{0} %May be changed to 1 or 2 if section numbers are desired.
+\setcounter{secnumdepth}{1} %May be changed to 1 or 2 if section numbers are desired.
 
@@
 
+\begin{table*}
+\centering
+\resizebox{\linewidth}{!}{
+\begin{tabular}[t]{llcccccc}
+\toprule
+\multicolumn{2}{c}{ } & \multicolumn{3}{c}{California Housing} & \multicolumn{3}{c}{GMSC} \\
+\cmidrule(l{3pt}r{3pt}){3-5} \cmidrule(l{3pt}r{3pt}){6-8}
+Model & Generator & Unfaithfulness ↓ & Implausibility ↓ & Uncertainty ↓ & Unfaithfulness ↓ & Implausibility ↓ & Uncertainty ↓\\
+\midrule
+    & ECCCo & \textbf{3.69 ± 0.08}** & 1.94 ± 0.13\hphantom{*}\hphantom{*} & \textbf{0.09 ± 0.01}** & 3.84 ± 0.07** & 2.13 ± 0.08\hphantom{*}\hphantom{*} & \textbf{0.23 ± 0.01}**\\
+
+    & ECCCo+ & 3.88 ± 0.07** & 1.20 ± 0.09\hphantom{*}\hphantom{*} & 0.15 ± 0.02\hphantom{*}\hphantom{*} & \textbf{3.79 ± 0.05}** & 1.81 ± 0.05\hphantom{*}\hphantom{*} & 0.30 ± 0.01*\hphantom{*}\\
+
+    & ECCCo (no CP) & 3.70 ± 0.08** & 1.94 ± 0.13\hphantom{*}\hphantom{*} & 0.10 ± 0.01** & 3.85 ± 0.07** & 2.13 ± 0.08\hphantom{*}\hphantom{*} & 0.23 ± 0.01**\\
+
+    & ECCCo (no EBM) & 4.03 ± 0.07\hphantom{*}\hphantom{*} & 1.12 ± 0.12\hphantom{*}\hphantom{*} & 0.14 ± 0.01** & 4.08 ± 0.06\hphantom{*}\hphantom{*} & 0.97 ± 0.08\hphantom{*}\hphantom{*} & 0.31 ± 0.01*\hphantom{*}\\
+
+    & REVISE & 3.96 ± 0.07*\hphantom{*} & \textbf{0.58 ± 0.03}** & 0.17 ± 0.03\hphantom{*}\hphantom{*} & 4.09 ± 0.07\hphantom{*}\hphantom{*} & \textbf{0.63 ± 0.02}** & 0.33 ± 0.06\hphantom{*}\hphantom{*}\\
+
+    & Schut & 4.00 ± 0.06\hphantom{*}\hphantom{*} & 1.15 ± 0.12\hphantom{*}\hphantom{*} & 0.10 ± 0.01** & 4.04 ± 0.08\hphantom{*}\hphantom{*} & 1.21 ± 0.08\hphantom{*}\hphantom{*} & 0.30 ± 0.01*\hphantom{*}\\
+
+\multirow{-7}{*}{\raggedright\arraybackslash MLP Ensemble} & Wachter & 4.04 ± 0.07\hphantom{*}\hphantom{*} & 1.13 ± 0.12\hphantom{*}\hphantom{*} & 0.16 ± 0.01\hphantom{*}\hphantom{*} & 4.10 ± 0.07\hphantom{*}\hphantom{*} & 0.95 ± 0.08\hphantom{*}\hphantom{*} & 0.32 ± 0.01\hphantom{*}\hphantom{*}\\
+\cmidrule{1-8}
+    & ECCCo & 1.40 ± 0.08** & 0.69 ± 0.05** & 0.11 ± 0.00** & 1.20 ± 0.06*\hphantom{*} & 0.78 ± 0.07** & 0.38 ± 0.01\hphantom{*}\hphantom{*}\\
+
+    & ECCCo+ & \textbf{1.28 ± 0.08}** & 0.60 ± 0.04** & 0.11 ± 0.00** & \textbf{1.01 ± 0.07}** & 0.70 ± 0.07** & 0.37 ± 0.01\hphantom{*}\hphantom{*}\\
+
+    & ECCCo (no CP) & 1.39 ± 0.08** & 0.69 ± 0.05** & 0.11 ± 0.00** & 1.21 ± 0.07*\hphantom{*} & 0.77 ± 0.07** & 0.39 ± 0.01\hphantom{*}\hphantom{*}\\
+
+    & ECCCo (no EBM) & 1.70 ± 0.09\hphantom{*}\hphantom{*} & 0.99 ± 0.08\hphantom{*}\hphantom{*} & 0.14 ± 0.00*\hphantom{*} & 1.31 ± 0.07\hphantom{*}\hphantom{*} & 0.97 ± 0.10\hphantom{*}\hphantom{*} & 0.32 ± 0.01**\\
+
+    & REVISE & 1.39 ± 0.15** & \textbf{0.59 ± 0.04}** & 0.25 ± 0.07\hphantom{*}\hphantom{*} & 1.01 ± 0.07** & \textbf{0.63 ± 0.04}** & 0.33 ± 0.07\hphantom{*}\hphantom{*}\\
+
+    & Schut & 1.59 ± 0.10*\hphantom{*} & 1.10 ± 0.06\hphantom{*}\hphantom{*} & \textbf{0.09 ± 0.00}** & 1.34 ± 0.07\hphantom{*}\hphantom{*} & 1.21 ± 0.10\hphantom{*}\hphantom{*} & \textbf{0.26 ± 0.01}**\\
+
+\multirow{-7}{*}{\raggedright\arraybackslash JEM Ensemble} & Wachter & 1.71 ± 0.09\hphantom{*}\hphantom{*} & 0.99 ± 0.08\hphantom{*}\hphantom{*} & 0.14 ± 0.00\hphantom{*}\hphantom{*} & 1.31 ± 0.08\hphantom{*}\hphantom{*} & 0.95 ± 0.10\hphantom{*}\hphantom{*} & 0.33 ± 0.01\hphantom{*}\hphantom{*}\\
+\bottomrule
+\end{tabular}}
+\caption{Results for tabular datasets: sample averages +/- one standard deviation across valid counterfactuals.
+Best outcomes are highlighted in bold.
+Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (\textit{Wachter}).
+\label{tab:results-tabular}}
+\end{table*}
+
 Overall, we find strong empirical evidence suggesting that \textit{ECCCo} consistently achieves state-of-the-art faithfulness.
@@
 % Sample UUID of factual: UUID("1ceb06f0-5949-11ee-0e9c-dd49ccfec8c3")
+
+\begin{table}
+\centering
+\resizebox{\linewidth}{!}{
+\begin{tabular}[t]{llcc}
+\toprule
+\multicolumn{2}{c}{ } & \multicolumn{2}{c}{MNIST} \\
+\cmidrule(l{3pt}r{3pt}){3-4}
+Model & Generator & Unfaithfulness ↓ & Implausibility ↓\\
+\midrule
+    & ECCCo & \textbf{0.243 ± 0.000}** & 0.420 ± 0.001\hphantom{*}\hphantom{*}\\
+
+    & ECCCo+ & 0.246 ± 0.000*\hphantom{*} & 0.306 ± 0.001**\\
+
+    & REVISE & 0.248 ± 0.000\hphantom{*}\hphantom{*} & \textbf{0.301 ± 0.004}**\\
+
+    & Schut & 0.247 ± 0.001\hphantom{*}\hphantom{*} & 0.303 ± 0.008**\\
+
+\multirow{-5}{*}{\raggedright\arraybackslash MLP} & Wachter & 0.247 ± 0.000\hphantom{*}\hphantom{*} & 0.344 ± 0.002\hphantom{*}\hphantom{*}\\
+\cmidrule{1-4}
+    & ECCCo & 0.248 ± 0.000** & 0.387 ± 0.002\hphantom{*}\hphantom{*}\\
+
+    & ECCCo+ & \textbf{0.248 ± 0.000}** & 0.310 ± 0.002**\\
+
+    & REVISE & 0.248 ± 0.000** & 0.301 ± 0.002**\\
+
+    & Schut & 0.250 ± 0.002\hphantom{*}\hphantom{*} & \textbf{0.289 ± 0.024}*\hphantom{*}\\
+
+\multirow{-5}{*}{\raggedright\arraybackslash LeNet-5} & Wachter & 0.249 ± 0.000\hphantom{*}\hphantom{*} & 0.335 ± 0.002\hphantom{*}\hphantom{*}\\
+\bottomrule
+\end{tabular}}
+\caption{Results for vision dataset. Formatting details are the same as in Table~\ref{tab:results-tabular}. \label{tab:results-vision}}
+\end{table}
 
@@
 
-\bibliography{aaai24}
+\bibliography{aaai24,bib}
 

### 2023-12-13 experiments/notebooks/poc.qmd
acknowledgements

--- 2023-09-24 notebooks/poc.qmd
@@
 ```{julia}
-include("$(pwd())/notebooks/setup.jl")
-eval(setup_notebooks)
+using Pkg; Pkg.activate("experiments")
+include("$(pwd())/experiments/setup_env.jl")
+using TaijaPlotting
 ```
@@
 input_dim, nobs = size(X)
-batch_size = Int(round(nobs/10))
+_batch_size = Int(round(nobs/10))
 epochs = 100
@@
     ),
-    batch_size=batch_size,
-    finalizer=Flux.softmax,
+    batch_size=_batch_size,
+    finaliser=Flux.softmax,
     loss=Flux.Losses.crossentropy,
@@
 jem = mach.model.model.jem
-batch_size = mach.model.model.batch_size
+_batch_size = mach.model.model.batch_size
 X = Float32.(matrix(X))
@@
     for target in 1:2
-        X̂ = generate_conditional_samples(jem, batch_size, target; niter=niter) 
+        X̂ = generate_conditional_samples(jem, _batch_size, target; niter=niter) 
         ex = extrema(hcat(X,X̂), dims=2)

### 2023-12-13 paper/camera-ready/altmeyer.tex
acknowledgements

@@
 
+\section*{Acknowledgements}
+
+Some of the members of TU Delft were partially funded by ICAI AI for Fintech Research, an ING---TU Delft
+collaboration.
+Research reported in this work was partially or completely facilitated by computational resources and support of the DelftBlue~\citep{DHPC2022} and the Delft AI Cluster (DAIC: https://doc.daic.tudelft.nl/) at TU Delft.
+The authors would like to thank Azza Ahmed, in particular, for her tremendous help with running Julia jobs on the cluster.
+The work remains the sole responsibility of the authors.
+
 \bibliography{aaai24,bib}

### 2023-12-13 experiments/notebooks/poc.qmd
rerunning POC chart

@@
 ```
+
+## Data
 
@@
 ```
+
+## Model
 
@@
 fit!(mach)
-Serialization.serialize(joinpath(output_path,"poc_model.jls"), mach)
 ```
+
+## Conditional Samples
 
@@
 #| echo: false
+
+using MLJBase
 niter = 1000
@@
 _batch_size = mach.model.model.batch_size
-X = Float32.(matrix(X))
+X = Float32.(MLJBase.matrix(X))
 if typeof(jem.sampler) <: ConditionalSampler
@@
 
+## Losses and Penalties
+
 ```{julia}
@@
 p1 = Plots.contourf(mach.model, mach.fitresult, permutedims(X), labels; plot_set_loss=true, zoom=0, temp=temp)
-p2 = Plots.contourf(mach.model, mach.fitresult, permutedims(X), labels; plot_classification_loss=true, zoom=0, temp=temp, clim=nothing, loss_matrix=ones(2,2))
-plt = display(Plots.plot(p0, p1, p2, size=(1400,320), layout=(1,3)))
-savefig(joinpath(output_images_path, "poc_set_size.png"))
+plt = display(Plots.plot(p0, p1, size=(800,320), layout=(1,2)))
 ```
 
+## Counterfactuals
+
+Compared to other generators:
+
 ```{julia}
+using OrderedCollections
 Random.seed!(1234)
@@
 λ₂ = 0.4
-λ₃ = 0.4
+λ₃ = 5.0
 Λ = [λ₁, λ₂, λ₃]
@@
 display(plt)
-savefig(plt, joinpath(output_images_path, "poc.png"))
 ```
@@
 using Colors
+panel_height = 200
 col_pal = palette(:seaborn_colorblind)
@@
 
-λ₁ = 0.1
+λ₁ = 0.2
 λ₂ = 0.5
-λ₃ = 0.5
+λ₃ = 3.0
 Λ = [λ₁, λ₂, λ₃]
@@
 generator_dict = OrderedDict(
-    "Wachter" => WachterGenerator(λ = 0.3, opt=opt),
+    "Wachter" => WachterGenerator(λ = 0.7, opt=opt),
     "ECCCo (no EBM)" => ECCCoGenerator(λ = [λ₁,λ₂,0.0], opt=opt),
@@
 display(plt)
-savefig(plt, joinpath(output_images_path, "poc_gradient_fields.png"))
 ```
-
-```{julia}
-ce = ces["ECCCo"]
-using CounterfactualExplanations.Generators: ℓ
-
-# loss
-function f1(x)
-    _ce = deepcopy(ce)
-    _ce.s′ = x
-    return - ℓ(_ce.generator, _ce)
-end
-
-# cost
-function f2(x)
-    _ce = deepcopy(ce)
-    _ce.s′ = x
-    λ = _ce.generator.λ[1]
-    _loss = _ce.generator.penalty[1]
-    return - λ * _loss(_ce)
-end
-
-# set size
-function f3(x)
-    _ce = deepcopy(ce)
-    _ce.s′ = x
-    λ = _ce.generator.λ[2]
-    _loss = _ce.generator.penalty[2]
-    return - λ * _loss(_ce)
-end
-
-# distance from energy
-function f4(x)
-    _ce = deepcopy(ce)
-    _ce.s′ = x
-    λ = _ce.generator.λ[3]
-    _loss = _ce.generator.penalty[3]
-    return - λ * _loss(_ce)
-end
-
-# Helper function:
-meshgrid(x, y) = (repeat(x, outer=length(y)), repeat(y, inner=length(x)))
-
-xlims, ylims = extrema(X, dims=2)
-xrange = range(xlims..., length=10)
-yrange = range(ylims..., length=10)
-
-x1, x2 = meshgrid(xrange, yrange)
-inputs = zip(x1, x2)
-
-u = []
-v = []
-
-for (x, y) in inputs
-    push!(u, scale * gradient(f1, [x, y][:,:])[1][1])
-    push!(v, scale * gradient(f1, [x, y][:,:])[1][2])
-end
-
-Plots.plot(xlims=xlims, ylims=ylims)
-Plots.scatter!(counterfactual_data)
-Plots.quiver!(x1, x2, quiver=(u, v))
-```

### 2023-12-13 experiments/notebooks/annex.qmd
poc chart done

---
format: pdf
---

# Plots

```{julia}
using Pkg; Pkg.activate("experiments")
include("$(pwd())/experiments/setup_env.jl")
```

## Counterfactual Path - MNIST

```{julia}
data_name = "mnist"
plt_order = ["MLP", "MLP Ensemble", "LeNet-5", "JEM", "JEM Ensemble"]
fixed_factual = 3
fixed_target = 9
models = Serialization.deserialize("models/$(data_name)_models.jls")
data = eval(Meta.parse("load_$(data_name)()"))
plt_order = plt_order[[x in collect(keys(models)) for x in plt_order]]
```

```{julia}
using Plots.PlotMeasures
n_samp = 100
n_rand = 500
σ = 0.1
plts = []
for (mod_name, model) in models
    Δ = []
    L = []
    for i in 1:n_rand
        factual = isnothing(fixed_factual) ? rand(data.y_levels) : fixed_factual
        target = isnothing(fixed_target) ? rand(data.y_levels[data.y_levels .!= factual]) : fixed_target
        t = get_target_index(data.y_levels, target)
        E(x) = -logits(model, x)[t, :]
        x_samp = data.X[:,rand(findall(data.output_encoder.labels.==target),n_samp)]
        x_rand = data.X[:,rand(findall(data.output_encoder.labels.==factual),1)]
        x_rand .+= Float32.(randn(size(x_samp,1),1)) .* σ
        δ = mean(map(y -> norm(x_rand.-y),eachcol(x_samp)))
        push!(Δ,δ)
        l = E(x_rand)[1]
        push!(L,l)
    end
    plt = scatter(
        Δ, L;
        label="", title=mod_name, smooth=:true,
        lc=:red, lw=2
    )
    push!(plts, plt)
end
width = 1000
plt = plot(
    plts[sortperm(collect(keys(models)))[invperm(sortperm(plt_order))]]..., 
    layout=(1,length(models)), size=(width,round(1.0width)/length(models)),
    ticks=false
)
savefig(plt, "www/dist_energy.png")
```

### 2023-12-13 experiments/notebooks/poc.qmd
poc chart done

@@
 
-temp = 0.3
+temp = 0.1
 p0 = Plots.contourf(mach.model, mach.fitresult, permutedims(X), labels; plot_set_size=true, zoom=0, temp=temp)
 p1 = Plots.contourf(mach.model, mach.fitresult, permutedims(X), labels; plot_set_loss=true, zoom=0, temp=temp)
-plt = display(Plots.plot(p0, p1, size=(800,320), layout=(1,2)))
+p2 = Plots.contourf(mach.model, mach.fitresult, permutedims(X), labels; plot_classification_loss=true, zoom=0, temp=temp, clim=nothing, loss_matrix=ones(2,2))
+plt = display(Plots.plot(p0, p1, p2, size=(1400,320), layout=(1,3)))
+# savefig(joinpath(output_images_path, "poc_set_size.png"))
 ```
@@
 λ₁ = 0.2
-λ₂ = 0.5
+λ₂ = 1.5
 λ₃ = 3.0

### 2023-12-13 paper/camera-ready/altmeyer.tex
poc chart done

@@
 While in the context of EBM, this is usually done during training, we propose to repurpose this approach during inference in order to evaluate the faithfulness of model explanations.
-The technical appendix provides additional implementation details for any tasks related to energy-based modelling.
+The appendix provides additional implementation details for any tasks related to energy-based modeling.
 
@@
 This is sufficient to capture the generative property of the underlying model since it is implicitly captured by its parameters $\theta$.
-Importantly, this means that we do not need to generate conditional samples through SGLD during our counterfactual search at all as we explain in the technical appendix.
+Importantly, this means that we do not need to generate conditional samples through SGLD during our counterfactual search at all as we explain in the appendix.
 
@@
 Across all models and datasets highlighted here, all variations of \textit{ECCCo} consistently outperform all other generators with respect to faithfulness, in many cases substantially.
-This pattern is mostly robust across all other benchmark datasets (Tables~\ref{tab:results-linearly-separable} to~\ref{tab:results-fashion-mnist} in the technical appendix).
+This pattern is mostly robust across all other benchmark datasets (additional results can be found in the appendix).
 
@@
 For both tabular datasets in Table~\ref{tab:results-tabular}, \textit{ECCCo} improves plausibility substantially compared to the baseline.
-This broad pattern is mostly consistent for all other datasets, although there are notable exceptions for which \textit{ECCCo} takes the lead on both plausibility and faithfulness (see, for example, Tables~\ref{tab:results-moons} and~\ref{tab:results-german-credit} in the appendix).
+This broad pattern is mostly consistent for all other datasets, although there are notable exceptions for which \textit{ECCCo} takes the lead on both plausibility and faithfulness.
 

### 2023-12-13 experiments/jobscripts/extra/mnist_with_ce.sh
mnist with images

@@ -1,0 +1,16 @@
+#!/bin/bash
+
+#SBATCH --job-name="MNIST w/ CE"
+#SBATCH --time=03:00:00
+#SBATCH --ntasks=10
+#SBATCH --cpus-per-task=5
+#SBATCH --partition=compute
+#SBATCH --mem-per-cpu=8GB
+#SBATCH --account=research-eemcs-insy
+#SBATCH --mail-type=END     # Set mail type to 'END' to receive a mail when the job finishes. 
+
+module load 2023r1 openmpi
+
+source experiments/slurm_header.sh
+
+srun julia --project=experiments --threads $SLURM_CPUS_PER_TASK experiments/run_experiments.jl -- data=mnist output_path=results mpi threaded n_individuals=50 n_runs=1 vertical_splits=100 store_ce > experiments/logs/mnist.log

### 2023-12-13 experiments/notebooks/mnist_examples.qmd
most charts updates

--- 2023-09-05 notebooks/mnist.qmd
@@
 ```{julia}
-include("$(pwd())/notebooks/setup.jl")
-eval(setup_notebooks)
+using Pkg; Pkg.activate("experiments")
+include("$(pwd())/experiments/setup_env.jl")
+include("$(pwd())/experiments/notebooks/setup.jl")
 ```
 
-# MNIST
+## Motivation
 
-## Anecdotal Evidence
-
-### Examples in Introduction
-
-#### Wachter and JSMA
+### Wachter and JSMA
 
@@
 factual = predict_label(M, counterfactual_data, x_factual)[1]
-γ = 0.9
+γ = 0.90
 
@@
 # Search:
-generic_generator = WachterGenerator()
+generic_generator = WachterGenerator(opt=Flux.Adam(0.25))
 ce_wachter = generate_counterfactual(
@@
 vae_weak = load_mnist_vae(;strong=false)
-Serialization.serialize(joinpath(output_path,"mnist_classifier.jls"), M)
-Serialization.serialize(joinpath(output_path,"mnist_vae.jls"), vae)
-Serialization.serialize(joinpath(output_path,"mnist_vae_weak.jls"), vae_weak)
 ```
@@
 revise_generator = REVISEGenerator(
-    opt = Flux.Optimise.Descent(0.25),
-    λ=0.0,
+    opt = Flux.Optimise.Adam(0.25),
+    λ=0.1,
 )
@@
 display(plt)
-savefig(plt, joinpath(output_images_path, "mnist_motivation.png"))
+savefig(plt, joinpath(paper_figpath, "mnist_motivation.png"))
 ```
 
-### ECCCo
+## Comparison with ECCCo
 
 ```{julia}
-function pre_process(x; noise::Float32=0.03f0)
-    ϵ = Float32.(randn(size(x)) * noise)
-    x += ϵ
-    return x
-end
+outcome = Serialization.deserialize("results_extra/mnist_outcome.jls")
 ```
-
-```{julia}
-# Hyper:
-_retrain = true
-_regen = true
-
-# Data:
-n_obs = 10000
-counterfactual_data = load_mnist(n_obs)
-counterfactual_data.X = pre_process.(counterfactual_data.X)
-counterfactual_data.generative_model = vae
-X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
-X = table(permutedims(X))
-x_factual = reshape(pre_process(x_factual, noise=0.0f0), input_dim, 1)
-labels = counterfactual_data.output_encoder.labels
-input_dim, n_obs = size(counterfactual_data.X)
-n_digits = Int(sqrt(input_dim))
-output_dim = length(unique(labels))
-```
-
-First, let's create a couple of image classifier architectures:
-
-```{julia}
-# Model parameters:
-epochs = 25
-batch_size = minimum([Int(round(n_obs/10)), 128])
-n_hidden = 128
-activation = Flux.swish
-builder = MLJFlux.@builder Flux.Chain(
-    Dense(n_in, n_hidden, activation),
-    Dense(n_hidden, n_out),
-)
-n_ens = 5                                   # number of models in ensemble
-_loss = Flux.Losses.crossentropy       # loss function
-_finaliser = Flux.softmax                        # finalizer function
-```
-
-```{julia}
-# JEM parameters:
-𝒟x = Uniform(-1.0,1.0)
-𝒟y = Categorical(ones(output_dim) ./ output_dim)
-sampler = ConditionalSampler(
-    𝒟x, 𝒟y, 
-    input_size=(input_dim,), 
-    batch_size=10,
-)
-α = [1.0,1.0,1e-2]      # penalty strengths
-```
-
-```{julia}
-# Simple MLP:
-mlp = NeuralNetworkClassifier(
-    builder=builder, 
-    epochs=epochs,
-    batch_size=batch_size,
-    finalizer=_finaliser,
-    loss=_loss,
-)
-
-# Deep Ensemble:
-mlp_ens = EnsembleModel(model=mlp, n=n_ens)
-
-# Joint Energy Model:
-jem = JointEnergyClassifier(
-    sampler;
-    builder=builder,
-    epochs=epochs,
-    batch_size=batch_size,
-    finalizer=_finaliser,
-    loss=_loss,
-    jem_training_params=(
-        α=α,verbosity=10,
-    ),
-    sampling_steps=25,
-)
-
-# JEM with adversarial training:
-jem_adv = deepcopy(jem)
-# jem_adv.adv_training = true
-
-# Deep Ensemble of Joint Energy Models:
-jem_ens = EnsembleModel(model=jem, n=n_ens)
-
-# Deep Ensemble of Joint Energy Models with adversarial training:
-# jem_ens_plus = EnsembleModel(model=jem_adv, n=n_ens)
-
-# Dictionary of models:
-models = Dict(
-    "MLP" => mlp,
-    "MLP Ensemble" => mlp_ens,
-    "JEM" => jem,
-    "JEM Ensemble" => jem_ens,
-    # "JEM Ensemble+" => jem_ens_plus,
-)
-
-Serialization.serialize(joinpath(output_path,"mnist_architectures.jls"), models)
-```
-
-```{julia}
-# Train models:
-function _train(model, X=X, y=labels; cov=.95, method=:simple_inductive, mod_name="model")
-    conf_model = conformal_model(model; method=method, coverage=cov)
-    mach = machine(conf_model, X, y)
-    @info "Begin training $mod_name."
-    fit!(mach)
-    @info "Finished training $mod_name."
-    M = ECCCo.ConformalModel(mach.model, mach.fitresult)
-    return M
-end
-if _retrain
-    model_dict = Dict(mod_name => _train(mod; mod_name=mod_name) for (mod_name, mod) in models)
-    Serialization.serialize(joinpath(output_path,"mnist_models.jls"), model_dict)
-else
-    model_dict = Serialization.deserialize(joinpath(output_path,"mnist_models.jls"))
-end
-```
-
-```{julia}
-params = DataFrame(
-    Dict(
-        :n_obs => Int.(round(n_obs/10)*10),
-        :epochs => epochs,
-        :batch_size => batch_size,
-        :n_hidden => n_hidden,
-        :n_layers => length(model_dict["MLP"].fitresult[1][1])-1,
-        :activation => string(activation),
-        :n_ens => n_ens,
-        :lambda => string(α[3]),
-        :jem_sampling_steps => jem.sampling_steps,
-        :sgld_batch_size => sampler.batch_size,
-        :dataname => "MNIST",
-    )
-)
-CSV.write(joinpath(params_path, "mnist.csv"), params)
-```
-
-```{julia}
-# Plot generated samples:
-n_regen = 500
-if _regen 
-    for (mod_name, mod) in model_dict
-        K = length(counterfactual_data.y_levels)
-        input_size = size(selectdim(counterfactual_data.X, ndims(counterfactual_data.X), 1))
-        𝒟x = Uniform(extrema(counterfactual_data.X)...)
-        𝒟y = Categorical(ones(K) ./ K)
-        sampler = ConditionalSampler(𝒟x, 𝒟y; input_size=input_size, prob_buffer=0.0)
-        opt = ImproperSGLD()
-        f(x) = logits(mod, x)
-
-        _w = 1500
-        plts = []
-        neach = 10
-        for i in 1:10
-            x = sampler(f, opt; niter=n_regen, n_samples=neach, y=i)
-            plts_i = []
-            for j in 1:size(x, 2)
-                xj = x[:,j]
-                xj = reshape(xj, (n_digits, n_digits))
-                plts_i = [plts_i..., Plots.heatmap(rotl90(xj), axis=nothing, cb=false)]
-            end
-            plt = Plots.plot(plts_i..., size=(_w,0.10*_w), layout=(1,10))
-            plts = [plts..., plt]
-        end
-        plt = Plots.plot(plts..., size=(_w,_w), layout=(10,1), plot_title=mod_name)
-        savefig(plt, joinpath(output_images_path, "mnist_generated_$(mod_name).png"))
-        display(plt)
-    end
-end
-```
-
-```{julia}
-# Evaluate models:
-
-measure = Dict(
-    :f1score => multiclass_f1score, 
-    :acc => accuracy, 
-    :precision => multiclass_precision
-)
-model_performance = DataFrame()
-for (mod_name, mod) in model_dict
-    # Test performance:
-    test_data = load_mnist_test()
-    _perf = CounterfactualExplanations.Models.model_evaluation(mod, test_data, measure=collect(values(measure)))
-    _perf = DataFrame([[p] for p in _perf], collect(keys(measure)))
-    _perf.mod_name .= mod_name
-    _perf.dataname .= "MNIST"
-    model_performance = vcat(model_performance, _perf)
-end
-Serialization.serialize(joinpath(output_path,"mnist_model_performance.jls"), model_performance)
-CSV.write(joinpath(output_path, "mnist_model_performance.csv"), model_performance)
-model_performance
-```
-
-### Different Models
-
-```{julia}
-function plot_mnist(ce; size=(img_height, img_height), kwrgs...)
-    x = CounterfactualExplanations.counterfactual(ce)
-    phat = target_probs(ce)
-    plt = Plots.plot(
-        convert2image(MNIST, reshape(x,28,28));
-        axis=([], false), 
-        size=size,
-        kwrgs...,
-    )
-end
-
-function _plot_eccco_mnist(
-    x::Union{AbstractArray, Int}=x_factual, target::Int=target;
-    λ=[0.1,0.25,0.25],
-    temp=0.1,
-    plt_order = ["MLP", "MLP Ensemble", "JEM", "JEM Ensemble"],
-    opt = nothing,
-    rng::Union{Int,AbstractRNG}=1234,
-    T::Int = 100,
-    use_class_loss::Bool = true,
-    model_dict=model_dict,
-    wide::Bool = false,
-    img_height::Int = img_height,
-    plot_factual::Bool = false,
-    generator::Union{Nothing,CounterfactualExplanations.AbstractGenerator}=nothing,
-    test_data::Bool = false,
-    use_energy_delta::Bool = false,
-    kwrgs...,
-)
-
-    # Setup:
-    Random.seed!(rng)
-    if x isa Int
-        x_fact = counterfactual_data.X[:,rand(findall(labels.==x))][:,:]
-    else
-        x_fact = x
-    end
-
-    if isnothing(generator)
-        # Generate counterfactuals using ECCCo generator:
-        generator = ECCCoGenerator(
-            λ=λ, 
-            temp=temp, 
-            opt=opt,
-            use_class_loss=use_class_loss,
-            nsamples=10,
-            nmin=10,
-            use_energy_delta=use_energy_delta,
-        )
-    end
-
-    if test_data
-        data = load_mnist_test()
-    else
-        data = counterfactual_data
-    end
-
-    ces = Dict()
-    for (mod_name, mod) in model_dict
-        ce = generate_counterfactual(
-            x_fact, target, data, mod, generator; 
-            decision_threshold=γ, max_iter=T,
-            initialization=:identity,
-            converge_when=:generator_conditions,
-        )
-        ces[mod_name] = ce
-    end
-    _plt_order = map(x -> findall(collect(keys(model_dict)) .== x)[1], plt_order)
-
-    # Plot:
-    p1 = Plots.plot(
-        convert2image(MNIST, reshape(x_fact,28,28)),
-        axis=([], false),
-        size=(img_height, img_height),
-        title="Factual"
-    )
-
-    if plot_factual
-        plts = [p1]
-    else
-        plts = []
-    end
-    letters = collect('a':'z')[1:length(ces)]
-    _count = 1
-
-    for (_name,ce) in collect(ces)[_plt_order]
-        _x = CounterfactualExplanations.counterfactual(ce)
-        _phat = target_probs(ce)
-        _title = "($(letters[_count]))"
-        plt = Plots.plot(
-            convert2image(MNIST, reshape(_x,28,28)),
-            axis=([], false), 
-            size=(img_height, img_height),
-            title=_title
-        )
-        plts = [plts..., plt]
-        _count += 1
-    end
-    if wide
-        plt = Plots.plot(plts...; size=(img_height*length(plts),img_height), layout=(1,length(plts)), kwrgs...)
-    else
-        plt = Plots.plot(plts...; size=(img_height,img_height), kwrgs...)
-    end
-
-    return plt, generator, ces
-end
-```
-
-```{julia}
-plt, eccco_generator, ces = _plot_eccco_mnist()
-display(plt)
-savefig(plt, joinpath(output_images_path, "mnist_eccco.png"))
-```
-
-#### Energy Delta (not in paper)
-
-```{julia}
-plt, gen_delta, ces = _plot_eccco_mnist(λ = [0.1,0.1,3.0], use_energy_delta=true)
-display(plt)
-savefig(plt, joinpath(output_images_path, "mnist_eccco_energy_delta.png"))
-```
-
-```{julia}
-λ_delta = [0.1,0.1,2.5]
-λ = [0.1,0.25,0.25]
-plts = []
-for i in 0:9
-    plt, _, _ = _plot_eccco_mnist(x_factual, i; λ = λ, plot_title="Distance")
-    plt_delta, _, _ = _plot_eccco_mnist(x_factual, i; λ = λ_delta, use_energy_delta=true, plot_title="Energy Delta")
-    plt = Plots.plot(plt, plt_delta; size=(img_height*2,img_height), layout=(1,2))
-    display(plt)
-    push!(plts, plt)
-end
-```
-
-#### Additional Models (not in paper)
-
-LeNet-5:
-
-```{julia}
-mutable struct LeNetBuilder
-	filter_size::Int
-	channels1::Int
-	channels2::Int
-end
-
-preproc(X) = reshape(X, (28, 28, 1, :))
-
-function MLJFlux.build(b::LeNetBuilder, rng, n_in, n_out)
-
-    _n_in = Int(sqrt(n_in))
-
-	k, c1, c2 = b.filter_size, b.channels1, b.channels2
-
-	mod(k, 2) == 1 || error("`filter_size` must be odd. ")
-
-	# padding to preserve image size on convolution:
-	p = div(k - 1, 2)
-
-    preproc(x) = reshape(x, (_n_in, _n_in, 1, :))
-
-	front = Flux.Chain(
-        Conv((k, k), 1 => c1, pad=(p, p), relu),
-        MaxPool((2, 2)),
-        Conv((k, k), c1 => c2, pad=(p, p), relu),
-        MaxPool((2, 2)),
-        Flux.flatten
-    )
-	d = Flux.outputsize(front, (_n_in, _n_in, 1, 1)) |> first
-    back = Flux.Chain(
-        Dense(d, 120, relu),
-        Dense(120, 84, relu),
-        Dense(84, n_out),
-    )
-
-    chain = Flux.Chain(preproc, front, back)
-
-	return chain
-end
-
-# Final model:
-lenet = NeuralNetworkClassifier(
-    builder=LeNetBuilder(5, 6, 16),
-    epochs=50,
-    batch_size=batch_size,
-    finalizer=_finaliser,
-    loss=_loss,
-)
-```
-
-Training all of them:
-
-```{julia}
-add_retrain = true
-
-# Deep Ensemble:
-mlp_large_ens = EnsembleModel(model=mlp, n=50)
-
-# CNN Ensemble:
-lenet_ens = EnsembleModel(model=lenet, n=5)
-
-add_models = Dict(
-    "LeNet-5" => lenet,
-    # "LeNet-5 Ensemble" => lenet_ens,
-    # "RobustNet" => rob_net,
-    "Large Ensemble (n=50)" => mlp_large_ens,
-)
-
-if add_retrain
-    add_model_dict = Dict(mod_name => _train(mod; mod_name=mod_name) for (mod_name, mod) in add_models)
-    large_model_dict = merge(model_dict, add_model_dict)
-    Serialization.serialize(joinpath(output_path,"mnist_models_large.jls"), large_model_dict)
-else
-    large_model_dict = Serialization.deserialize(joinpath(output_path,"mnist_models_large.jls"))
-end
-```
-
-```{julia}
-# Evaluate models:
-
-measure = Dict(
-    :f1score => multiclass_f1score, 
-    :acc => accuracy, 
-    :precision => multiclass_precision
-)
-model_performance = DataFrame()
-for (mod_name, mod) in large_model_dict
-    # Test performance:
-    test_data = load_mnist_test()
-    _perf = CounterfactualExplanations.Models.model_evaluation(mod, test_data, measure=collect(values(measure)))
-    _perf = DataFrame([[p] for p in _perf], collect(keys(measure)))
-    _perf.mod_name .= mod_name
-    _perf.dataname .= "MNIST"
-    model_performance = vcat(model_performance, _perf)
-end
-Serialization.serialize(joinpath(output_path,"mnist_model_performance_additional.jls"), model_performance)
-CSV.write(joinpath(output_path, "mnist_model_performance_additional.csv"), model_performance)
-model_performance
-```
-
-```{julia}
-_plt_order = [
-    "MLP", 
-    "MLP Ensemble", 
-    "Large Ensemble (n=50)", 
-    "LeNet-5",
-    # "LeNet-5 Ensemble",
-    # "RobustNet",
-    "JEM", 
-    "JEM Ensemble",
-]
-plt_additional_models, _, _ces_ = _plot_eccco_mnist(
-    λ = [0.1,0.21,0.21],
-    plt_order = _plt_order,
-    model_dict=large_model_dict,
-    wide = true,
-    plot_factual = true,
-    img_height = 150,
-)
-display(plt_additional_models)
-# savefig(plt_additional_models, joinpath(output_images_path, "mnist_eccco_additional.png"))
-```
-
-```{julia}
-Random.seed!(123)
-
-λ = [0.1,0.22,0.22]
-
-wachter = WachterGenerator(
-    λ=λ[1], 
-    opt=eccco_generator.opt
-)
-
-combos = [
-    (9,7),
-    (7,2),
-    (6,0),
-    (4,1),
-    (2,3),
-    (5,8),
-    (6,5),
-    (1,7),
-    (1,4),
-    (0,3)
-]
-
-n_each = 3
-combos = reduce(vcat, [fill(c, n_each) for c in combos])
-
-plts_eccco = []
-plts_wachter = []
-ces_eccco = []
-ces_wachter = []
-rngs = []
-
-for (factual, target) in combos
-
-    rng = rand(1:10000)
-
-    # ECCCo:
-    plt, _, ces = _plot_eccco_mnist(
-        factual, target; 
-        λ = λ,
-        plt_order = _plt_order,
-        model_dict = large_model_dict,
-        wide = true,
-        plot_factual = true,
-        rng = rng,
-        img_height = 150
-    )
-    display(plt)
-    push!(plts_eccco, plt)
-    push!(ces_eccco, reduce(hcat, target_probs.(values(ces))))
-
-    # Wachter:
-    plt, _, ces = _plot_eccco_mnist(
-        factual, target; 
-        plt_order = _plt_order,
-        model_dict = large_model_dict,
-        wide = true,
-        plot_factual = true,
-        rng = rng,
-        img_height = 150,
-        generator = wachter,
-    )
-    display(plt)
-    push!(plts_wachter, plt)
-    push!(ces_wachter, reduce(hcat, target_probs.(values(ces))))
-
-    push!(rngs, rng)
-
-end
-```
-
-```{julia}
-final_plts = []
-for (i, (factual, target)) in enumerate(combos)
-    p1 = plts_eccco[i]
-    p2 = plts_wachter[i]
-    plt = Plots.plot(
-        p1, p2, layout=(2,1),
-        size = (1200, 400),
-    )
-    display(plt)
-    savefig(plt, joinpath("dev/rebuttal/www", "mnist_$(factual)to$(target)_$(i).png"))
-    push!(final_plts, plt)
-end
-```
-
-### All digits
-
-```{julia}
-function plot_mnist(
-    factual::Int,target::Int;
-    generator::AbstractGenerator,
-    model::AbstractFittedModel=model_dict["JEM Ensemble"],
-    data::CounterfactualData=counterfactual_data,
-    rng::Union{Int,AbstractRNG}=Random.GLOBAL_RNG,
-    _plot_title::Bool=true,
-    show_factual::Bool=false,
-    img_height::Int=180,
-    kwargs...,
-)
-    Random.seed!(rng)
-
-    decision_threshold = !isdefined(kwargs, :decision_threshold) ? 0.9 : decision_threshold
-    max_iter = !isdefined(kwargs, :max_iter) ? 100 : max_iter
-    initialization = !isdefined(kwargs, :initialization) ? :identity : initialization
-    converge_when = !isdefined(kwargs, :converge_when) ? :generator_conditions : converge_when
-
-    x = reshape(data.X[:,rand(findall(predict_label(model, data).==factual))],input_dim,1)
-    ce = generate_counterfactual(
-        x, target, data, model, generator; 
-        decision_threshold=decision_threshold, max_iter=max_iter,
-        initialization=initialization,
-        converge_when=converge_when,
-        kwargs...
-    )
-
-    _title = _plot_title ? "$(factual) -> $(target)" : ""
-
-    _x = CounterfactualExplanations.counterfactual(ce)
-    plt = Plots.plot(
-        convert2image(MNIST, reshape(_x,28,28)),
-        axis=([], false),
-        size=(img_height, img_height),
-        title=_title
-    )
-    if show_factual
-        plt_factual = Plots.plot(
-            convert2image(MNIST, reshape(x,28,28)),
-            axis=([], false),
-            size=(img_height, img_height),
-            title="Factual"
-        )
-        plt = Plots.plot(plt_factual, plt; size=(img_height*2,img_height), layout=(1,2))
-    end
-
-    return plt
-end
-```
-
-```{julia}
-_regen_all_digits = false
-if _regen_all_digits
-    function plot_all_digits(rng=123;verbose=true,img_height=180,kwargs...)
-        plts = []
-        for i in 0:9
-            for j in 0:9
-                @info "Generating counterfactual for $(i) -> $(j)"
-                plt = plot_mnist(i,j;kwargs...,rng=rng, img_height=img_height)
-                !verbose || display(plt)
-                plts = [plts..., plt]
-            end
-        end
-        plt = Plots.plot(plts...; size=(img_height*10,img_height*10), layout=(10,10), dpi=300)
-        return plt
-    end
-    plt = plot_all_digits(generator=eccco_generator)
-    savefig(plt, joinpath(output_images_path, "mnist_eccco_all_digits.png"))
-end
-```
-
-#### Energy Delta (not in paper)
-
-```{julia}
-_regen_all_digits = true
-if _regen_all_digits
-    function plot_all_digits(rng=123;verbose=true,img_height=180,kwargs...)
-        plts = []
-        for i in 0:9
-            for j in 0:9
-                @info "Generating counterfactual for $(i) -> $(j)"
-                plt = plot_mnist(i,j;kwargs...,rng=rng, img_height=img_height)
-                !verbose || display(plt)
-                plts = [plts..., plt]
-            end
-        end
-        plt = Plots.plot(plts...; size=(img_height*10,img_height*10), layout=(10,10), dpi=300)
-        return plt
-    end
-    plt = plot_all_digits(generator=gen_delta)
-    savefig(plt, joinpath(output_images_path, "mnist_eccco_all_digits-delta.png"))
-end
-```
-
-## Benchmark
-
-```{julia}
-Λ = eccco_generator.λ
-
-# Benchmark generators:
-generator_dict = Dict(
-    "Wachter" => WachterGenerator(λ=Λ[1], opt=eccco_generator.opt),
-    "REVISE" => REVISEGenerator(λ=Λ[1], opt=eccco_generator.opt),
-    "Schut" => greedy_generator,
-    "ECCCo" => eccco_generator,
-)
-```
-
-```{julia}
-generator_params = DataFrame(
-    Dict(
-        :λ1 => Λ[1],
-        :λ2 => Λ[2],
-        :λ3 => Λ[3],
-        :opt => string(typeof(eccco_generator.opt)),
-        :eta => eccco_generator.opt.eta,
-        :dataname => "MNIST",
-    )
-)
-CSV.write(joinpath(params_path, "generator/mnist.csv"), generator_params)
-```
-
-```{julia}
-# Measures:
-measures = [
-    CounterfactualExplanations.distance,
-    ECCCo.distance_from_energy,
-    ECCCo.distance_from_targets,
-    CounterfactualExplanations.Evaluation.validity,
-    CounterfactualExplanations.Evaluation.redundancy,
-    ECCCo.set_size_penalty,
-]
-
-bmks = []
-for target in sort(unique(labels))
-    for factual in sort(unique(labels))
-        if factual == target
-            continue
-        end
-        bmk = benchmark(
-            counterfactual_data; 
-            models=model_dict, 
-            generators=generator_dict, 
-            measure=measures,
-            suppress_training=true, dataname="MNIST",
-            n_individuals=5,
-            target=target, factual=factual,
-            initialization=:identity,
-            converge_when=:generator_conditions,
-        )
-        push!(bmks, bmk)
-    end
-end
-bmk = reduce(vcat, bmks)
-
-CSV.write(joinpath(output_path, "mnist_benchmark.csv"), bmk())
-```
-
-```{julia}
-df = @chain bmk() begin
-    @mutate(variable = ifelse.(variable .== "distance_from_energy", "Non-Conformity", variable))
-    @mutate(variable = ifelse.(variable .== "distance_from_targets", "Implausibility", variable))
-    @mutate(variable = ifelse.(variable .== "distance", "Cost", variable))
-    @mutate(variable = ifelse.(variable .== "redundancy", "Redundancy", variable))
-    @mutate(variable = ifelse.(variable .== "Validity", "Validity", variable))
-end
-plt = AlgebraOfGraphics.data(df) * visual(BoxPlot) * 
-    mapping(:generator, :value, row=:variable, col=:model, color=:generator)
-plt = draw(
-    plt, axis=(xlabel="", xticksvisible=false, xticklabelsvisible=false, width=150, height=120), 
-    facet=(; linkyaxes=:minimal)
-)   
-display(plt)
-save(joinpath(output_images_path, "mnist_benchmark.png"), plt, px_per_unit=5)
-```

### 2023-12-13 experiments/notebooks/poc.qmd
mnist with images
most charts updates

--- 2023-12-13 experiments/notebooks/poc.qmd
@@
 include("$(pwd())/experiments/setup_env.jl")
-using TaijaPlotting
+include("$(pwd())/experiments/notebooks/setup.jl")
 ```
@@
 
-temp = 0.3
+temp = 0.1
 p0 = Plots.contourf(mach.model, mach.fitresult, permutedims(X), labels; plot_set_size=true, zoom=0, temp=temp)
 p1 = Plots.contourf(mach.model, mach.fitresult, permutedims(X), labels; plot_set_loss=true, zoom=0, temp=temp)
-plt = display(Plots.plot(p0, p1, size=(800,320), layout=(1,2)))
+p2 = Plots.contourf(mach.model, mach.fitresult, permutedims(X), labels; plot_classification_loss=true, zoom=0, temp=temp, clim=nothing, loss_matrix=ones(2,2))
+plt = display(Plots.plot(p0, p1, p2, size=(1400,320), layout=(1,3)))
+savefig(joinpath(output_images_path, "poc_set_size.png"))
 ```
@@
 display(plt)
+savefig(plt, joinpath(output_images_path, "poc.png"))
 ```
@@
 col_pal = palette(:seaborn_colorblind)
-Random.seed!(1234)
+Random.seed!(1234) # 1234
 using CounterfactualExplanations.Generators: ∇
@@
 λ₁ = 0.2
-λ₂ = 0.5
+λ₂ = 2.0
 λ₃ = 3.0
@@
 η = 0.01
+max_iter = 100
 
@@
 generator_dict = OrderedDict(
-    "Wachter" => WachterGenerator(λ = 0.7, opt=opt),
+    "Wachter" => WachterGenerator(λ = 1.0, opt=opt),
     "ECCCo (no EBM)" => ECCCoGenerator(λ = [λ₁,λ₂,0.0], opt=opt),
@@
         converge_when=:generator_conditions,
+        max_iter=max_iter,
     )
@@
 plt = Plots.plot(plts...; size=(panel_height*length(plts),panel_height), layout=(1,length(plts)), dpi=300)
-# plt = Plots.plot(plts..., size=(1000,250), layout=(1,4), dpi=300)
 display(plt)
+savefig(plt, joinpath(paper_figpath, "poc_gradient_fields.png"))
 ```

### 2023-12-13 paper/camera-ready/altmeyer.tex
mnist with images
most charts updates

@@
     \textsuperscript{\rm 1}Delft University of Technology,\\
-    \textsuperscript{\rm 2}ING\\
+    Faculty of Electrical Engineering, Mathematics and Computer Science\\
+    \textsuperscript{\rm 2}ING Bank\\
+    p.altmeyer@tudelft.nl
 %
@@
 While in the context of EBM, this is usually done during training, we propose to repurpose this approach during inference in order to evaluate the faithfulness of model explanations.
-The appendix provides additional implementation details for any tasks related to energy-based modeling.
+The appendix provides additional implementation details for any tasks related to energy-based modelling.
 
@@
 While \textit{Wachter} generates a valid counterfactual, it ends up close to the original starting point consistent with its objective.
-\textit{ECCCo (no EBM)} pushes the counterfactual further into the target domain to minimize predictive uncertainty, but the outcome is still not plausible.
+\textit{ECCCo (no EBM)} avoids regions of high predictive uncertainty near the decision boundary, but the outcome is still not plausible.
 The counterfactual produced by \textit{ECCCo (no CP)} is energy-constrained.
-Since the \textit{JEM} has learned the conditional input distribution reasonably well in this case, the counterfactuals are both faithful and plausible.
+Since the \textit{JEM} has learned the conditional input distribution reasonably well in this case, the counterfactual is both faithful and plausible.
 Finally, the outcome for \textit{ECCCo} looks similar, but the additional smooth set size penalty leads to somewhat faster convergence.
@@
 Research reported in this work was partially or completely facilitated by computational resources and support of the DelftBlue~\citep{DHPC2022} and the Delft AI Cluster (DAIC: https://doc.daic.tudelft.nl/) at TU Delft.
+Detailed information about the utilized computing resources can be found in the appendix.
 The authors would like to thank Azza Ahmed, in particular, for her tremendous help with running Julia jobs on the cluster.

### 2023-12-13 experiments/notebooks/setup.jl

@@ -1,0 +1,6 @@
+using TaijaPlotting
+Plots.theme(:wong)
+img_height = 300
+panel_height = 250
+paper_figpath = "paper/camera-ready/figures"
+output_images_path = "www/"

### 2023-12-14 experiments/notebooks/figures1and3.qmd
all charts in paper recreated

--- 2023-12-13 experiments/notebooks/mnist_examples.qmd
@@
 ```
+
+This notebook can be used to recreate Figures 1 and 3 in the main paper.
 
@@
 
+The code below generates Figure 1 in the main paper.
+
 ```{julia}
@@
 
+The code below generates Figure 3 in the main paper. This uses the results from this [release](https://github.com/pat-alt/ECCCo.jl/releases/tag/results-camera-ready).
+
 ```{julia}
-outcome = Serialization.deserialize("results_extra/mnist_outcome.jls")
+# Using results from: https://github.com/pat-alt/ECCCo.jl/releases/tag/results-camera-ready
+outcome = Serialization.deserialize("results/mnist_outcome.jls") # results from 
+factual = 3
+target = 5
+seed = 3333 
+plt, ces = compare_counterfactuals(
+    outcome; factual=factual, target=target,
+    seed = seed
+)
+savefig(plt, joinpath(paper_figpath, "mnist_benchmark.png"))
+CounterfactualExplanations.target_probs.(ces)
 ```

### 2023-12-14 paper/camera-ready/altmeyer.tex
resolve overflow issue
text updated

@@
 \begin{aligned}
-\mathbf{Z}^\prime &= \arg \min_{\mathbf{Z}^\prime \in \mathcal{Z}^L} \left\{  {\text{yloss}(M_{\theta}(f(\mathbf{Z}^\prime)),\mathbf{y}^+)}+ \lambda {\text{cost}(f(\mathbf{Z}^\prime)) }  \right\} 
+\min_{\mathbf{Z}^\prime \in \mathcal{Z}^L} \left\{  {\text{yloss}(M_{\theta}(f(\mathbf{Z}^\prime)),\mathbf{y}^+)}+ \lambda {\text{cost}(f(\mathbf{Z}^\prime)) }  \right\} 
 \end{aligned} 
@@
 It works under the premise of turning heuristic notions of uncertainty into rigorous estimates by repeatedly sifting through the training data or a dedicated calibration dataset.
+Calibration data is used to compute so-called nonconformity scores: $\mathcal{S}=\{s(\mathbf{x}_i,\mathbf{y}_i)\}_{i \in \mathcal{D}_{\text{cal}}}$ where $s: (\mathcal{X},\mathcal{Y}) \mapsto \mathbb{R}$ is referred to as \textit{score function} (see appendix for details).
 
@@
 This is sufficient to capture the generative property of the underlying model since it is implicitly captured by its parameters $\theta$.
-Importantly, this means that we do not need to generate conditional samples through SGLD during our counterfactual search at all as we explain in the appendix.
+Importantly, this means that we do not need to generate conditional samples through SGLD during our counterfactual search at all (see appendix for details).
 
@@
 To account for stochasticity, we generate multiple counterfactuals for each target class, generator, model and dataset.
-Full details concerning our parameter choices, training procedures and model performance can be found in the appendix.
 
@@
 Following related literature~\citep{schut2021generating,dhurandhar2018explanations} we also include two image datasets: \textit{MNIST}~\citep{lecun1998mnist} and \textit{Fashion MNIST}~\citep{xiao2017fashion}.
-Detailed descriptions and results for all datasets can be found in the appendix.
 
+Full details concerning model training as well as detailed descriptions and results for all datasets can be found in the appendix.
 In the following, we will focus on the most relevant results highlighted in Tables~\ref{tab:results-tabular} and~\ref{tab:results-vision}.
-The tables show sample averages along with standard deviations for our key evaluation metrics for the \textit{California Housing} and \textit{GMSC} datasets (Table~\ref{tab:results-tabular}) and the \textit{MNIST} dataset (Table~\ref{tab:results-vision}).
+The tables show sample averages along with standard deviations across multiple runs for our key evaluation metrics for the \textit{California Housing} and \textit{GMSC} datasets (Table~\ref{tab:results-tabular}) and the \textit{MNIST} dataset (Table~\ref{tab:results-vision}).
 For each metric, the best outcomes are highlighted in bold.
@@
 Overall, we find strong empirical evidence suggesting that \textit{ECCCo} consistently achieves state-of-the-art faithfulness.
-Across all models and datasets highlighted here, all variations of \textit{ECCCo} consistently outperform all other generators with respect to faithfulness, in many cases substantially.
-This pattern is mostly robust across all other benchmark datasets (additional results can be found in the appendix).
+Across all models and datasets highlighted here, all variations of \textit{ECCCo} consistently outperform other generators with respect to faithfulness, in many cases substantially.
+This pattern is mostly robust across all other datasets.
 
@@
 We expected the former to play a more significant role in this context and that is typically what we find across all datasets.
-For example, the results for \textit{GMSC} in Table~\ref{tab:results-tabular} indicate that faithfulness can be improved substantially by relying solely on the energy constraint (\textit{ECCCo (no CP)}).
-In some cases though, as for the \textit{California Housing} dataset, \textit{ECCCo (no EBM)} actually outperforms \textit{ECCCo (no CP)}.
+The results in Table~\ref{tab:results-tabular} indicate that faithfulness can be improved substantially by relying solely on the energy constraint (\textit{ECCCo (no CP)}).
+In most cases, however, the full objective yields the most faithful counterfactuals.
 This indicates that predictive uncertainty minimization plays an important role in achieving faithfulness.
 
-We also generally find that the highest degree of faithfulness is obtained when the counterfactual search is performed directly in the feature space $\mathcal{X}$.
-While \textit{ECCCo+} typically attains high levels of faithfulness compared to most other generators, it is consistently outperformed by \textit{ECCCo}.
-The case is even stronger for \textit{REVISE}, which performs worst out of all generators for faithfulness on the \textit{GMSC} dataset and better only than \textit{Wachter} on \textit{California Housing}.
+We also generally find that latent space search does not impede faithfulness for \textit{ECCCo}.
+In most cases \textit{ECCCo+} is either on par with \textit{ECCCo} or even outperforms it.
+There are some notable exceptions though.
+Cases in which \textit{ECCCo} achieves substantially better faithfulness without latent space search tend to involve more vulnerable models like the simple MLP for MNIST in Table~\ref{tab:results-vision}.
+We explain this finding as follows: even though dimensionality reduction through PCA in the case of \textit{ECCCo+} can be considered a relatively mild form of intervention, the first $n_z$ principal components fail to capture some of the variation in the data.
+More vulnerable models may be particularly sensitive to this residual variation in the data.
 
-These findings are consistent with the notion that surrogate models may inhibit faithfulness.
-Even though dimensionality reduction through PCA in the case of \textit{ECCCo+} can be considered a relatively mild form of intervention, the first $n_z$ principal components fail to capture some of the variation in the data, that the underlying model itself may be sensitive to.
-This notion is illustrated nicely in Figure~\ref{fig:mnist-bmk}, where the counterfactual produced by \textit{ECCCo} is somewhat noisier and grainier than the one produced by \textit{ECCCo+}.
-
-In conclusion, we recommend in light of the findings here to use the full \textit{ECCCo} search objective whenever model faithfulness is a key priority.
+Consistent with this finding, we also observe that \textit{REVISE} ranks higher for faithfulness, if the model itself has learned more plausible representations of the underlying data: \textit{REVISE} generates more faithful counterfactuals than the baseline for the \textit{JEM} Ensemble in Table~\ref{tab:results-tabular} and the LeNet-5 \textit{CNN} in Table~\ref{tab:results-vision}.
+This demonstrates that the two desiderata---faithfulness and plausibility---are not mutually exclusive.
 
@@
 Overall, we find strong empirical evidence suggesting that \textit{ECCCo} can achieve near state-of-the-art plausibility without sacrificing faithfulness.
-Figure~\ref{fig:mnist-bmk} shows one such example taken from the \textit{MNIST} benchmark where the objective is to turn the factual three (far left) into a five.
+Figure~\ref{fig:mnist-bmk} shows one such example taken from the \textit{MNIST} benchmark where the objective is to turn the factual `three' (far left) into a `five'.
 The underlying model is a LeNet-5 \textit{CNN}.
@@
 
-While we maintain that generally speaking plausibility should hinge on the quality of the model, our results also indicate that it is possible to trade off some degree of faithfulness for plausibility if needed: \textit{ECCCo+} generally outperforms other variants of \textit{ECCCo} in this context at the small cost of slightly reduced faithfulness.
+While we maintain that generally speaking plausibility should hinge on the quality of the model, our results also indicate that it is possible to balance faithfulness and plausibility if needed: \textit{ECCCo+} generally outperforms other variants of \textit{ECCCo} in this context at the small cost of slightly reduced faithfulness.
 For the vision datasets especially, we find that  \textit{ECCCo+} is consistently second only to \textit{REVISE} for all models and regularly substantially better than the baseline.
 Looking at the \textit{California Housing} data, latent space search markedly improves plausibility without sacrificing faithfulness: for the \textit{JEM} Ensemble, \textit{ECCCo+} performs substantially better than the baseline and only marginally worse than \textit{REVISE}.
-Importantly, \textit{ECCCo+} does not attain plausibility at all costs: for the MLP, plausibility is still very low but this seems to faithfully represent what the model has learned.
+Importantly, \textit{ECCCo+} does not attain plausibility at all costs: for the \textit{MLP} Ensemble, plausibility is still very low but this seems to faithfully represent what the model has learned.
 
-We conclude that \textit{ECCCo} offers us a way to balance the objectives of faithfulness and plausibility. \textit{ECCCo+} can be used to tilt the scale in favour of plausibility if needed.
+We conclude from the findings presented thus far that \textit{ECCCo} enables us to reconcile the objectives of faithfulness and plausibility.
+It produces plausible counterfactuals if and only if the model itself has learned plausible explanations for the data.
+It thus avoids the risk of generating plausible but potentially misleading explanations for models that are highly susceptible to implausible explanations.
 
@@
 Consequently, both \textit{REVISE} and \textit{ECCCo} generally yield more costly counterfactuals than the baseline.
-Nonetheless, \textit{ECCCo} does not seem to stretch costs unnecessarily: in Figure~\ref{fig:mnist-bmk} useful parts of the factual three are clearly retained.
+Nonetheless, \textit{ECCCo} does not seem to stretch costs unnecessarily: in Figure~\ref{fig:mnist-bmk} useful parts of the factual `three' are clearly retained.
 

### 2023-12-14 paper/camera-ready/altmeyer.tex
done

@@
 collaboration.
+
 Research reported in this work was partially or completely facilitated by computational resources and support of the DelftBlue~\citep{DHPC2022} and the Delft AI Cluster (DAIC: https://doc.daic.tudelft.nl/) at TU Delft.
@@
 
+Patrick would also like to express his gratitude to the group of students who have recently contributed to the development of CounterfactualExplanations.jl~\citep{altmeyer2023explaining}, the Julia package that was used for this analysis: Rauno Arike, Simon Kasdorp, Lauri Kesküll, Mariusz Kicior, Vincent Pikand.
+
 \bibliography{aaai24,bib}

### 2023-12-15 paper/camera-ready/altmeyer.tex
version with footnote for appendix

--- 2023-12-14 paper/camera-ready/altmeyer.tex
@@
     Arie van Deursen\textsuperscript{\rm 1},
-    Cynthia C. S. Liem \textsuperscript{\rm 1}
+    Cynthia C. S. Liem\textsuperscript{\rm 1}
 }
@@
 One straightforward way to do this is to use surrogate models for the task.
-\citet{joshi2019realistic}, for example, suggest that instead of searching counterfactuals in the feature space $\mathcal{X}$, we can instead traverse a latent embedding $\mathcal{Z}$ (Equation~\ref{eq:general}) that implicitly codifies the DGP.
+\citet{joshi2019realistic}, for example, suggest that instead of searching counterfactuals in the feature space $\mathcal{X}$, we can traverse a latent embedding $\mathcal{Z}$ (Equation~\ref{eq:general}) that implicitly codifies the DGP.
 To learn the latent embedding, they propose using a generative model such as a Variational Autoencoder (VAE).
 Provided the surrogate model is well-specified, their proposed approach \textit{REVISE} can yield plausible explanations.
-Others have proposed similar approaches: \citet{dombrowski2021diffeomorphic} traverse the base space of a normalizing flow to solve Equation~\ref{eq:general}; \citet{poyiadzi2020face} use density estimators ($\hat{p}: \mathcal{X} \mapsto [0,1]$) to constrain the counterfactuals to dense regions in the feature space; and, finally, \citet{karimi2021algorithmic} assume knowledge about the structural causal model that generates the data.
+Others have proposed similar approaches: \citet{dombrowski2021diffeomorphic} traverse the base space of a normalizing flow to solve Equation~\ref{eq:general}; \citet{poyiadzi2020face} use density estimators ($\hat{p}: \mathcal{X} \mapsto [0,1]$) to constrain the counterfactuals to dense regions in the feature space; finally, \citet{karimi2021algorithmic} assume knowledge about the causal graph that generates the data.
 
@@
 \citet{schut2021generating} show that to meet the plausibility objective we need not explicitly model the input distribution.
-Pointing to the undesirable engineering overhead induced by surrogate models, they propose that we rely on the implicit minimisation of predictive uncertainty instead.
+Pointing to the undesirable engineering overhead induced by surrogate models, they propose to rely on the implicit minimization of predictive uncertainty instead.
 Their proposed methodology, which we will refer to as \textit{Schut}, solves Equation~\ref{eq:general} by greedily applying Jacobian-Based Saliency Map Attacks (JSMA) in the feature space with cross-entropy loss and no penalty at all.
@@
 Since \textit{Wachter} is only concerned with closeness, the generated counterfactual is almost indistinguishable from the factual.
-The approach by~\citet{schut2021generating} expects a well-calibrated model that can generate predictive uncertainty estimates.
+\textit{Schut} expects a well-calibrated model that can generate predictive uncertainty estimates.
 Since this is not the case, the generated counterfactual looks like an adversarial example.
@@
 But is it also more faithful to the behaviour of our \textit{MNIST} classifier?
-That is much less clear because the surrogate used by \textit{REVISE} introduces friction: the generated explanations no longer depend exclusively on the black-box model itself.
+That is much less clear because the surrogate used by \textit{REVISE} introduces friction: explanations no longer depend exclusively on the black-box model itself.
 
@@
 where $\mathbf{r}_j \sim \mathcal{N}(\mathbf{0},\mathbf{I})$ is the stochastic term and the step-size $\epsilon_j$ is typically polynomially decayed~\citep{welling2011bayesian}.
-The term $\mathcal{E}_{\theta}(\mathbf{x}_j|\mathbf{y}^+)$ denotes the model energy conditioned on the target class label $\mathbf{y}^+$ which we specify as the negative logit corresponding to the target class label $\mathbf{y}^{+}$.
+The term $\mathcal{E}_{\theta}(\mathbf{x}_j|\mathbf{y}^+)$ denotes the model energy conditioned on the target class label $\mathbf{y}^+$ which we specify as the negative logit corresponding to $\mathbf{y}^{+}$.
 To allow for faster sampling, we follow the common practice of choosing the step-size $\epsilon_j$ and the standard deviation of $\mathbf{r}_j$ separately.
@@
 While in the context of EBM, this is usually done during training, we propose to repurpose this approach during inference in order to evaluate the faithfulness of model explanations.
-The appendix provides additional implementation details for any tasks related to energy-based modelling.
+The appendix provides additional implementation details for any tasks related to energy-based modelling\footnote{The supplementary appendix can be found here: https://github.com/pat-alt/ECCCo.jl.}.
 
@@
 
-Our default choice for the $\text{dist}(\cdot)$ function in both cases is the Euclidean Norm.
-Depending on the type of input data other choices may be more adequate, which we discuss further in Section~\ref{emp:setup}.
+Our default choice for the $\text{dist}(\cdot)$ function in both cases is the Euclidean Norm. Depending on the type of input data other choices may be more adequate (see Section~\ref{emp:setup}).
 
@@
 The third and final penalty term involving $\lambda_3$ ensures that the generated counterfactual is associated with low predictive uncertainty.
-To tune theses hyperparameters we have relied on grid search.
+These hyperparameters have been tuned through grid search.
 
@@
 The third metric presented~\ref{tab:results-tabular} in Table quantifies the predictive uncertainty of the counterfactual as measured by Equation~\ref{eq:setsize}.
-For the vision datasets, we rely on measuring the structural dissimilarity between images for our unfaithfulness and implausibility metrics~\citep{wang2003multiscale}.
+For vision datasets, we rely on measuring the structural dissimilarity between images for our unfaithfulness and implausibility metrics~\citep{wang2003multiscale}.
 
@@
 
+We would also like to express our gratitude to the group of students who have recently contributed to the development of CounterfactualExplanations.jl~\citep{altmeyer2023explaining}, the Julia package that was used for this analysis: Rauno Arike, Simon Kasdorp, Lauri Kesküll, Mariusz Kicior, Vincent Pikand.
+
+All code used for the analysis in this paper can be found here: https://github.com/pat-alt/ECCCo.jl.
+
 \bibliography{aaai24,bib}

### 2023-12-15 paper/appendix.tex
rerendered appendix

@@
 
-To enable others to easily replicate our experiments, we have chosen to work with small neural network architectures and randomly undersampled the \textit{MNIST} dataset (maintaining class balance).
-All of our final benchmarks could then be run locally on a personal machine, but grid searches and the sheer number of datasets required us to move to high-performance computing clusters to conduct our experiments efficiently.
+Research reported in this work was partially or completely facilitated by computational resources and support of the DelftBlue~\citep{DHPC2022} and the Delft AI Cluster (DAIC: https://doc.daic.tudelft.nl/) at TU Delft.
 
-\subsubsection{Local runs}
-
-The longest runtimes we experienced for model training and counterfactual benchmarking for a single benchmark were on the order of 12-24 hours (\textit{MNIST} data).
-For the synthetic data, single benchmarks could be completed in less than an hour.
-We have summarised our system information below:
-
-\textbf{Software}:
-
-\begin{itemize}
-  \item System Version: macOS 13.3.1
-  \item Kernel Version: Darwin 22.4.0
-\end{itemize}
-
-\textbf{Hardware}:
-
-\begin{itemize}
-  \item Model Name: MacBook Pro
-  \item Model Identifier: MacBookPro16,1
-  \item Processor Name: 8-Core Intel Core i9
-  \item Processor Speed: 2.3 GHz
-  \item Number of Processors: 1
-  \item Total Number of Cores: 8
-  \item L2 Cache (per Core): 256 KB
-  \item L3 Cache: 16 MB
-  \item Hyper-Threading Technology: Enabled
-  \item Memory: 32 GB
-\end{itemize}
-
-\subsubsection{HPC}
-
-We used two large clusters, which we do not cite or mention by name at this point to not interfer with the double-blind review process.
-All of our grid searches were multi-processed on 100 CPUs at 4-8GB or memory each.
+For grid search, we used 300 CPUs for tabular real-world datasets ($<$1.5 hours each), 150 CPUs for synthetic datasets ($\approx$1 hour each) and 100 CPUs for vision datasets ($<$4 hours each), where we ran a smaller grid search for the latter.
+To generate the final results reported in the tables we used 300 CPUs for all datasets ($<$1.5 hours each) except the vision datasets.
+For the latter, we used 50 CPUs for smaller final experiments and at longer run times ($\approx$5 hours each).
 

### 2023-12-15 paper/camera-ready/altmeyer.tex
rerendered appendix

@@
 The third and final penalty term involving $\lambda_3$ ensures that the generated counterfactual is associated with low predictive uncertainty.
-These hyperparameters have been tuned through grid search.
+To tune theses hyperparameters we have relied on grid search.
 
@@
 The third metric presented~\ref{tab:results-tabular} in Table quantifies the predictive uncertainty of the counterfactual as measured by Equation~\ref{eq:setsize}.
-For vision datasets, we rely on measuring the structural dissimilarity between images for our unfaithfulness and implausibility metrics~\citep{wang2003multiscale}.
+For the vision datasets, we rely on measuring the structural dissimilarity between images for our unfaithfulness and implausibility metrics~\citep{wang2003multiscale}.
 

### 2023-12-16 paper/camera-ready/altmeyer.tex
final things on main paper

@@
 Equation~\ref{eq:impl} gives rise to a similar evaluation metric for unfaithfulness.
-We swap out the subsample of observed individuals in the target class for the set of samples generated through SGLD ($\widehat{\mathbf{X}}_{\mathbf{y}^+}$):
+We swap out the subsample of observed individuals in the target class for the set of samples generated through SGLD ($\widehat{\mathbf{X}}_{\theta,\mathbf{y}^+}$):
 
@@
   \begin{aligned}
-  \mathbf{Z}^\prime =& \arg \min_{\mathbf{Z}^\prime \in \mathcal{Z}^L} \{  {L_{\text{JEM}}(f(\mathbf{Z}^\prime);M_{\theta},\mathbf{y}^+)}+ \lambda {\text{cost}(f(\mathbf{Z}^\prime)) } \} 
+  \min_{\mathbf{Z}^\prime \in \mathcal{Z}^L} \{  {L_{\text{JEM}}(f(\mathbf{Z}^\prime);M_{\theta},\mathbf{y}^+)}+ \lambda {\text{cost}(f(\mathbf{Z}^\prime)) } \} 
   \end{aligned} 
@@
   \begin{aligned}
-  \mathbf{Z}^\prime =& \arg \min_{\mathbf{Z}^\prime \in \mathcal{Z}^L} \{  {L_{\text{clf}}(f(\mathbf{Z}^\prime);M_{\theta},\mathbf{y}^+)}+ \lambda_1 {\text{cost}(f(\mathbf{Z}^\prime)) } \\
+  & \min_{\mathbf{Z}^\prime \in \mathcal{Z}^L} \{  {L_{\text{clf}}(f(\mathbf{Z}^\prime);M_{\theta},\mathbf{y}^+)}+ \lambda_1 {\text{cost}(f(\mathbf{Z}^\prime)) } \\
   &+ \lambda_2 \mathcal{E}_{\theta}(f(\mathbf{Z}^\prime)|\mathbf{y}^+) + \lambda_3 \Omega(C_{\theta}(f(\mathbf{Z}^\prime);\alpha)) \} 
@@
 To assess and benchmark the performance of our proposed generator against the state of the art, we generate multiple counterfactuals for different models and datasets.
-In particular, we compare \textit{ECCCo} and its variants to the following counterfactual generators that were introduced above: firstly; \textit{Schut}, which works under the premise of minimizing predictive uncertainty; secondly, \textit{REVISE}, which is state-of-the-art (SOTA) with respect to plausibility; and, finally, \textit{Wachter}, which serves as our baseline.
+In particular, we compare \textit{ECCCo} and its variants to the following counterfactual generators that were introduced above: firstly, \textit{Schut}, which works under the premise of minimizing predictive uncertainty; secondly, \textit{REVISE}, which is state-of-the-art (SOTA) with respect to plausibility; and, finally, \textit{Wachter}, which serves as our baseline.
 In the case of \textit{ECCCo+}, we use principal component analysis (PCA) for dimensionality reduction: the latent space $\mathcal{Z}$ is spanned by the first $n_z$ principal components where we choose $n_z$ to be equal to the latent dimension of the VAE used by \textit{REVISE}.
@@
 Both joint-energy modelling and ensembling have been associated with improved generative properties and adversarial robustness~\citep{grathwohl2020your,lakshminarayanan2016simple}, so we expect this to be positively correlated with the plausibility of \textit{ECCCo}.
-To account for stochasticity, we generate multiple counterfactuals for each target class, generator, model and dataset.
+To account for stochasticity, we generate many counterfactuals for each target class, generator, model and dataset over multiple runs.
 
 We perform benchmarks on eight datasets from different domains.
-From the credit and finance domain we include three tabular datasets: Give Me Some Credit (\textit{GMSC})~\citep{kaggle2011give}, \textit{German Credit}~\citet{hoffman1994german} and \textit{California Housing}~\citet{pace1997sparse}.
+From the credit and finance domain we include three tabular datasets: Give Me Some Credit (\textit{GMSC})~\citep{kaggle2011give}, \textit{German Credit}~\citep{hoffman1994german} and \textit{California Housing}~\citep{pace1997sparse}.
 All of these are commonly used in the related literature~\citep{karimi2020survey,altmeyer2023endogenous,pawelczyk2021carla}.
@@
 For the tabular datasets, we use the default Euclidian distance to measure unfaithfulness and implausibility as defined in Equations~\ref{eq:faith} and~\ref{eq:impl}, respectively.
-The third metric presented~\ref{tab:results-tabular} in Table quantifies the predictive uncertainty of the counterfactual as measured by Equation~\ref{eq:setsize}.
+The third metric presented in Table~\ref{tab:results-tabular} quantifies the predictive uncertainty of the counterfactual as measured by Equation~\ref{eq:setsize}.
 For the vision datasets, we rely on measuring the structural dissimilarity between images for our unfaithfulness and implausibility metrics~\citep{wang2003multiscale}.
@@
 Overall, we find strong empirical evidence suggesting that \textit{ECCCo} consistently achieves state-of-the-art faithfulness.
-Across all models and datasets highlighted here, all variations of \textit{ECCCo} consistently outperform other generators with respect to faithfulness, in many cases substantially.
+Across all models and datasets highlighted here, different variations of \textit{ECCCo} consistently outperform other generators with respect to faithfulness, in many cases substantially.
 This pattern is mostly robust across all other datasets.
@@
 
-While we maintain that generally speaking plausibility should hinge on the quality of the model, our results also indicate that it is possible to balance faithfulness and plausibility if needed: \textit{ECCCo+} generally outperforms other variants of \textit{ECCCo} in this context at the small cost of slightly reduced faithfulness.
+While we maintain that generally speaking plausibility should hinge on the quality of the model, our results also indicate that it is possible to balance faithfulness and plausibility if needed: \textit{ECCCo+} generally outperforms other variants of \textit{ECCCo} in this context, occasionally at the small cost of slightly reduced faithfulness.
 For the vision datasets especially, we find that  \textit{ECCCo+} is consistently second only to \textit{REVISE} for all models and regularly substantially better than the baseline.

### 2023-12-16 paper/camera-ready/altmeyer.tex
final spelling stuff

@@
 Consequently, the generated explanations may seem plausible to humans but need not necessarily describe the behaviour of the black-box model faithfully.
-We formalise this notion of faithfulness through the introduction of a tailored evaluation metric and propose a novel algorithmic framework for generating \textbf{E}nergy-\textbf{C}onstrained \textbf{C}onformal \textbf{Co}unterfactuals that are only as plausible as the model permits.
+We formalize this notion of faithfulness through the introduction of a tailored evaluation metric and propose a novel algorithmic framework for generating \textbf{E}nergy-\textbf{C}onstrained \textbf{C}onformal \textbf{Co}unterfactuals that are only as plausible as the model permits.
 Through extensive empirical studies, we demonstrate that \textit{ECCCo} reconciles the need for faithfulness and plausibility.
@@
 
-In this work, we draw closer attention to model faithfulness rather than fidelity as a desideratum for counterfactuals.
+In this work, we draw closer attention to modelling faithfulness rather than fidelity as a desideratum for counterfactuals.
 We define faithfulness as the degree to which counterfactuals are consistent with what the model has learned about the data.
@@
 
-The baseline approach, which we will simply refer to as \textit{Wachter}, searches a single counterfactual directly in the feature space and penalises its distance to the original factual.
+The baseline approach, which we will simply refer to as \textit{Wachter}, searches a single counterfactual directly in the feature space and penalizes its distance to the original factual.
 In this case, $f(\cdot)$ is simply the identity function and $\mathcal{Z}$ corresponds to the feature space itself.
@@
 
-A competing approach towards plausibility that is also closely related to this work instead relies on the black-box model itself.
-\citet{schut2021generating} show that to meet the plausibility objective we need not explicitly model the input distribution.
+A competing approach towards plausibility that is also closely related to this work instead relies on the black-box model itself.~\citet{schut2021generating} show that to meet the plausibility objective we need not explicitly model the input distribution.
 Pointing to the undesirable engineering overhead induced by surrogate models, they propose to rely on the implicit minimization of predictive uncertainty instead.
@@
 No measures have been taken to improve the model's adversarial robustness or its capacity for predictive uncertainty quantification.
-The far left panel of Figure ~\ref{fig:motiv} shows a random sample drawn from the dataset.
+The far left panel of Figure~\ref{fig:motiv} shows a random sample drawn from the dataset.
 The underlying classifier correctly predicts the label `nine' for this image.
 For the given factual image and model, we have used \textit{Wachter}, \textit{Schut} and \textit{REVISE} to generate one counterfactual each in the target class `seven'.
-The perturbed images are shown next to the factual image from left to right in Figure ~\ref{fig:motiv}.
+The perturbed images are shown next to the factual image from left to right in Figure~\ref{fig:motiv}.
 Captions on top of the images indicate the generator along with the predicted probability that the image belongs to the target class.
@@
 where $\mathbf{x}^{\prime}$ denotes the counterfactual and $\mathbf{X}_{\mathbf{y}^+}$ is a subsample of the training data in the target class $\mathbf{y}^+$.
-By averaging over multiple samples in this manner, we avoid the risk that the nearest neighbour of $\mathbf{x}^{\prime}$ itself is not plausible according to Definition~\ref{def:plausible} (e.g an outlier).
+By averaging over multiple samples in this manner, we avoid the risk that the nearest neighbour of $\mathbf{x}^{\prime}$ itself is not plausible according to Definition~\ref{def:plausible} (e.g.
+an outlier).
 
@@
 The third and final penalty term involving $\lambda_3$ ensures that the generated counterfactual is associated with low predictive uncertainty.
-To tune theses hyperparameters we have relied on grid search.
+To tune these hyperparameters we have relied on grid search.
 
@@
 \caption{Results for tabular datasets: sample averages +/- one standard deviation across valid counterfactuals.
-Best outcomes are highlighted in bold.
+The best outcomes are highlighted in bold.
 Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (\textit{Wachter}).
@@
 \end{figure}
-
-% Sample UUID of factual: UUID("1ceb06f0-5949-11ee-0e9c-dd49ccfec8c3")
 
@@
 
-Firstly, we recognise that our proposed distance-based evaluation metrics for plausibility and faithfulness may not be universally applicable to all types of data.
+Firstly, we recognize that our proposed distance-based evaluation metrics for plausibility and faithfulness may not be universally applicable to all types of data.
 In any case, they depend on choosing a distance metric on a case-by-case basis, as we have done in this work.

### 2023-12-16 paper/paper.qmd
final spelling stuff

---
title: "Faithful Model Explanations through Energy-Constrained Conformal Counterfactuals"
authors:
    - name: Patrick Altmeyer
      orcid: 0000-0003-4726-8613
      affiliation: Delft University of Technology
    - name: Arie van Deursen
      affiliation: Delft University of Technology
    - name: Cynthia C. S. Liem
      affiliation: Delft University of Technology
format:
    arxiv-pdf: 
        include-in-header:
        - text: |
            \usepackage{listings}
        keep-tex: true
abstract: |
    We present [`CounterfactualExplanations.jl`](https://github.com/JuliaTrustworthyAI/CounterfactualExplanations.jl): a package for generating Counterfactual Explanations (CE) and Algorithmic Recourse (AR) for black-box models in Julia. CE explain how inputs into a model need to change to yield specific model predictions. Explanations that involve realistic and actionable changes can be used to provide AR: a set of proposed actions for individuals to change an undesirable outcome for the better. In this article, we discuss the usefulness of CE for Explainable Artificial Intelligence and demonstrate the functionality of our package. The package is straightforward to use and designed with a focus on customization and extensibility. We envision it to one day be the go-to place for explaining arbitrary predictive models in Julia through a diverse suite of counterfactual generators.
keywords:
    - Julia
    - Explainable Artificial Intelligence
    - Counterfactual Explanations
    - Algorithmic Recourse
date: today
---

### 2023-12-16 paper/paper.tex
just using separate file

@@
-\documentclass{article}
+%File: formatting-instructions-latex-2024.tex
+%release 2024.0
+\documentclass[letterpaper]{article} % DO NOT CHANGE THIS
+\usepackage{aaai24}  % DO NOT CHANGE THIS
+\usepackage{times}  % DO NOT CHANGE THIS
+\usepackage{helvet}  % DO NOT CHANGE THIS
+\usepackage{courier}  % DO NOT CHANGE THIS
+\usepackage[hyphens]{url}  % DO NOT CHANGE THIS
+\usepackage{graphicx} % DO NOT CHANGE THIS
+\urlstyle{rm} % DO NOT CHANGE THIS
+\def\UrlFont{\rm}  % DO NOT CHANGE THIS
+\usepackage{natbib}  % DO NOT CHANGE THIS AND DO NOT ADD ANY OPTIONS TO IT
+\usepackage{caption} % DO NOT CHANGE THIS AND DO NOT ADD ANY OPTIONS TO IT
+\frenchspacing  % DO NOT CHANGE THIS
+\setlength{\pdfpagewidth}{8.5in}  % DO NOT CHANGE THIS
+\setlength{\pdfpageheight}{11in}  % DO NOT CHANGE THIS
+%
+% These are recommended to typeset algorithms but not required. See the subsubsection on algorithms. Remove them if you don't have algorithms in your paper.
+\usepackage{algorithm}
+% \usepackage{algorithmic}
 
-% if you need to pass options to natbib, use, e.g.:
-%     \PassOptionsToPackage{numbers, compress}{natbib}
-% before loading neurips_2023
+%
+% These are are recommended to typeset listings but not required. See the subsubsection on listing. Remove this block if you don't have listings in your paper.
+\usepackage{newfloat}
+\usepackage{listings}
+\DeclareCaptionStyle{ruled}{labelfont=normalfont,labelsep=colon,strut=off} % DO NOT CHANGE THIS
+\lstset{%
+	basicstyle={\footnotesize\ttfamily},% footnotesize acceptable for monospace
+	numbers=left,numberstyle=\footnotesize,xleftmargin=2em,% show line numbers, remove this entire line if you don't want the numbers.
+	aboveskip=0pt,belowskip=0pt,%
+	showstringspaces=false,tabsize=2,breaklines=true}
+\floatstyle{ruled}
+\newfloat{listing}{tb}{lst}{}
+\floatname{listing}{Listing}
+%
+% Keep the \pdfinfo as shown here. There's no need
+% for you to add the /Title and /Author tags.
+\pdfinfo{
+/TemplateVersion (2024.1)
+}
 
-% ready for submission
-\usepackage{neurips_2023}
+% DISALLOWED PACKAGES
+% \usepackage{authblk} -- This package is specifically forbidden
+% \usepackage{balance} -- This package is specifically forbidden
+% \usepackage{color (if used in text)
+% \usepackage{CJK} -- This package is specifically forbidden
+% \usepackage{float} -- This package is specifically forbidden
+% \usepackage{flushend} -- This package is specifically forbidden
+% \usepackage{fontenc} -- This package is specifically forbidden
+% \usepackage{fullpage} -- This package is specifically forbidden
+% \usepackage{geometry} -- This package is specifically forbidden
+% \usepackage{grffile} -- This package is specifically forbidden
+% \usepackage{hyperref} -- This package is specifically forbidden
+% \usepackage{navigator} -- This package is specifically forbidden
+% (or any other package that embeds links such as navigator or hyperref)
+% \indentfirst} -- This package is specifically forbidden
+% \layout} -- This package is specifically forbidden
+% \multicol} -- This package is specifically forbidden
+% \nameref} -- This package is specifically forbidden
+% \usepackage{savetrees} -- This package is specifically forbidden
+% \usepackage{setspace} -- This package is specifically forbidden
+% \usepackage{stfloats} -- This package is specifically forbidden
+% \usepackage{tabu} -- This package is specifically forbidden
+% \usepackage{titlesec} -- This package is specifically forbidden
+% \usepackage{tocbibind} -- This package is specifically forbidden
+% \usepackage{ulem} -- This package is specifically forbidden
+% \usepackage{wrapfig} -- This package is specifically forbidden
+% DISALLOWED COMMANDS
+% \nocopyright -- Your paper will not be published if you use this command
+% \addtolength -- This command may not be used
+% \balance -- This command may not be used
+% \baselinestretch -- Your paper will not be published if you use this command
+% \clearpage -- No page breaks of any kind may be used for the final version of your paper
+% \columnsep -- This command may not be used
+% \newpage -- No page breaks of any kind may be used for the final version of your paper
+% \pagebreak -- No page breaks of any kind may be used for the final version of your paperr
+% \pagestyle -- This command may not be used
+% \tiny -- This is not an acceptable font size.
+% \vspace{- -- No negative value may be used in proximity of a caption, figure, table, section, subsection, subsubsection, or reference
+% \vskip{- -- No negative value may be used to alter spacing above or below a caption, figure, table, section, subsection, subsubsection, or reference
 
-% to compile a preprint version, e.g., for submission to arXiv, add add the
-% [preprint] option:
-% \usepackage[preprint]{neurips_2023}
-
-% to compile a camera-ready version, add the [final] option, e.g.:
-%     \usepackage[final]{neurips_2023}
-
-% to avoid loading the natbib package, add option nonatbib:
-%    \usepackage[nonatbib]{neurips_2023}
-
-\usepackage[utf8]{inputenc} % allow utf-8 input
-\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
-\usepackage{hyperref}       % hyperlinks
-\usepackage{url}            % simple URL typesetting
-\usepackage{booktabs}       % professional-quality tables
+% Additional Packages:
 \usepackage{amsfonts}       % blackboard math symbols
-\usepackage{nicefrac}       % compact symbols for 1/2, etc.
-\usepackage{microtype}      % microtypography
-\usepackage{xcolor}         % colors
-
 \usepackage{amsmath}
@@
 \usepackage{multirow}
-
-% Bibliography
-\bibliographystyle{unsrtnat}
-\setcitestyle{numbers,square,comma}
+\usepackage{placeins}
 
@@
 
-\title{ECCCos from the Black Box:\\
-Faithful Explanations through\\
+\setcounter{secnumdepth}{1} %May be changed to 1 or 2 if section numbers are desired.
+
+% The file aaai24.sty is the style file for AAAI Press
+% proceedings, working notes, and technical reports.
+%
+
+% Title
+
+% Your title must be in mixed case, not sentence case.
+% That means all verbs (including short verbs like be, is, using,and go),
+% nouns, adverbs, adjectives should be capitalized, including both words in hyphenated terms, while
+% articles, conjunctions, and prepositions are lower case unless they
+% directly follow a colon or long dash
+
+\title{Faithful Model Explanations through\\
 Energy-Constrained Conformal Counterfactuals}
-
-% The \author macro works with any number of authors. There are two commands
-% used to separate the names and addresses of multiple authors: \And and \AND.
+\author{
+    %Authors
+    % All authors must be in the same font size and format.
+    Patrick Altmeyer\textsuperscript{\rm 1},
+    Mojtaba Farmanbar\textsuperscript{\rm 2},
+    Arie van Deursen\textsuperscript{\rm 1},
+    Cynthia C. S. Liem\textsuperscript{\rm 1}
+}
+\affiliations{
+    %Afiliations
+    \textsuperscript{\rm 1}Delft University of Technology,\\
+    Faculty of Electrical Engineering, Mathematics and Computer Science\\
+    \textsuperscript{\rm 2}ING Bank\\
+    p.altmeyer@tudelft.nl
 %
-% Using \And between authors leaves it to LaTeX to determine where to break the
-% lines. Using \AND forces a line break at that point. So, if LaTeX puts 3 of 4
-% authors names on the first line, and the last on the second line, try using
-% \AND instead of \And before the third author name.
-
-\author{%
-  Anonymous Author\thanks{See also: } \\
-  Faculty \\
-  University \\
-  Address \\
-  \texttt{email} \\
-  \And
-  Anonymous Author\thanks{See also: } \\
-  Faculty \\
-  University \\
-  Address \\
-  \texttt{email} \\
+% See more examples next
 }
@@
 
-\section{Introduction}\label{intro}
+% Body of the paper
+\import{partials/}{body.tex}
 
-Counterfactual Explanations (CE) provide a powerful, flexible and intuitive way to not only explain black-box models but also help affected individuals through the means of Algorithmic Recourse.
-Instead of opening the Black Box, CE works under the premise of strategically perturbing model inputs to understand model behaviour~\citep{wachter2017counterfactual}.
-Intuitively speaking, we generate explanations in this context by asking what-if questions of the following nature: `Our credit risk model currently predicts that this individual is not credit-worthy.
-What if they reduced their monthly expenditures by 10\%?'
+\FloatBarrier
 
-This is typically implemented by defining a target outcome $\mathbf{y}^+ \in \mathcal{Y}$ for some individual $\mathbf{x} \in \mathcal{X}=\mathbb{R}^D$ described by $D$ attributes, for which the model $M_{\theta}:\mathcal{X}\mapsto\mathcal{Y}$ initially predicts a different outcome: $M_{\theta}(\mathbf{x})\ne \mathbf{y}^+$.
-Counterfactuals are then searched by minimizing a loss function that compares the predicted model output to the target outcome: $\text{yloss}(M_{\theta}(\mathbf{x}),\mathbf{y}^+)$.
-Since CE work directly with the black-box model, valid counterfactuals always have full local fidelity by construction where fidelity is defined as the degree to which explanations approximate the predictions of a black-box model~\citep{mothilal2020explaining,molnar2020interpretable}.
+\bibliography{aaai24,bib}
 
-In situations where full fidelity is a requirement, CE offer a more appropriate solution to Explainable Artificial Intelligence (XAI) than other popular approaches like LIME~\citep{ribeiro2016why} and SHAP~\citep{lundberg2017unified}, which involve local surrogate models.
-But even full fidelity is not a sufficient condition for ensuring that an explanation faithfully describes the behaviour of a model.
-That is because multiple very distinct explanations can all lead to the same model prediction, especially when dealing with heavily parameterized models like deep neural networks, which are typically underspecified by the data~\citep{wilson2020case}.
+\onecolumn
 
-In the context of CE, the idea that no two explanations are the same arises almost naturally.
-A key focus in the literature has therefore been to identify those explanations and algorithmic recourses that are most appropriate based on a myriad of desiderata such as sparsity, actionability and plausibility.
-In this work, we draw closer attention to model faithfulness rather than fidelity as a desideratum for counterfactuals.
-Our key contributions are as follows:
-
-\begin{itemize}
-  \item We show that fidelity is an insufficient evaluation metric for counterfactuals (Section~\ref{fidelity}) and propose a definition of faithfulness that gives rise to more suitable metrics (Section~\ref{faithfulness}).
-  \item We introduce a novel algorithmic approach for generating Energy-Constrained Conformal Counterfactuals (ECCCos) in Section~\ref{meth}.
-  \item We provide extensive empirical evidence demonstrating that ECCCos faithfully explain model behaviour and attain plausibility only when appropriate (Section~\ref{emp}).
-\end{itemize}
-
-To our knowledge, this is the first venture in this direction for generating faithful counterfactuals.
-Thus, we anticipate that ECCCos can serve as a baseline for future research.
-We believe that our work opens avenues for researchers and practitioners seeking tools to better distinguish trustworthy from unreliable models.
-
-\section{Background}\label{background}
-
-While CE can also be generated for arbitrary regression models~\citep{spooner2021counterfactual}, existing work has primarily focused on classification problems.
-Let $\mathcal{Y}=(0,1)^K$ denote the one-hot-encoded output domain with $K$ classes.
-Then most counterfactual generators rely on gradient descent to optimize different flavours of the following counterfactual search objective:
-
-\begin{equation} \label{eq:general}
-\begin{aligned}
-\mathbf{Z}^\prime &= \arg \min_{\mathbf{Z}^\prime \in \mathcal{Z}^L} \left\{  {\text{yloss}(M_{\theta}(f(\mathbf{Z}^\prime)),\mathbf{y}^+)}+ \lambda {\text{cost}(f(\mathbf{Z}^\prime)) }  \right\} 
-\end{aligned} 
-\end{equation}
-
-Here $\text{yloss}(\cdot)$ denotes the primary loss function, $f(\cdot)$ is a function that maps from the counterfactual state space to the feature space and $\text{cost}(\cdot)$ is either a single penalty or a collection of penalties that are used to impose constraints through regularization.
-Equation~\ref{eq:general} restates the baseline approach to gradient-based counterfactual search proposed by~\citet{wachter2017counterfactual} in general form as introduced by~\citet{altmeyer2023endogenous}.
-To explicitly account for the multiplicity of explanations, $\mathbf{Z}^\prime=\{ \mathbf{z}_l\}_L$ denotes an $L$-dimensional array of counterfactual states.
-
-The baseline approach, which we will simply refer to as \textit{Wachter}, searches a single counterfactual directly in the feature space and penalises its distance to the original factual.
-In this case, $f(\cdot)$ is simply the identity function and $\mathcal{Z}$ corresponds to the feature space itself.
-Many derivative works of~\citet{wachter2017counterfactual} have proposed new flavours of Equation~\ref{eq:general}, each of them designed to address specific \textit{desiderata} that counterfactuals ought to meet in order to properly serve both AI practitioners and individuals affected by algorithmic decision-making systems.
-The list of desiderata includes but is not limited to the following: sparsity, proximity~\citep{wachter2017counterfactual}, actionability~\citep{ustun2019actionable}, diversity~\citep{mothilal2020explaining}, plausibility~\citep{joshi2019realistic,poyiadzi2020face,schut2021generating}, robustness~\citep{upadhyay2021robust,pawelczyk2022probabilistically,altmeyer2023endogenous} and causality~\citep{karimi2021algorithmic}.
-Different counterfactual generators addressing these needs have been extensively surveyed and evaluated in various studies~\citep{verma2020counterfactual,karimi2020survey,pawelczyk2021carla,artelt2021evaluating,guidotti2022counterfactual}.
-
-Perhaps unsurprisingly, the different desiderata are often positively correlated.
-For example, \citet{artelt2021evaluating} find that plausibility typically also leads to improved robustness.
-Similarly, plausibility has also been connected to causality in the sense that plausible counterfactuals respect causal relationships~\citep{mahajan2020preserving}.
-Consequently, the plausibility of counterfactuals has been among the primary concerns for researchers.
-Achieving plausibility is equivalent to ensuring that the generated counterfactuals comply with the true and unobserved data-generating process (DGP).
-We define plausibility formally in this work as follows:
-
-\begin{definition}[Plausible Counterfactuals]
-  \label{def:plausible}
-  Let $\mathcal{X}|\mathbf{y}^+= p(\mathbf{x}|\mathbf{y}^+)$ denote the true conditional distribution of samples in the target class $\mathbf{y}^+$.
-Then for $\mathbf{x}^{\prime}$ to be considered a plausible counterfactual, we need: $\mathbf{x}^{\prime} \sim \mathcal{X}|\mathbf{y}^+$.
-\end{definition}
-
-To generate plausible counterfactuals, we need to be able to quantify the DGP: $\mathcal{X}|\mathbf{y}^+$.
-One straightforward way to do this is to use surrogate models for the task.
-\citet{joshi2019realistic}, for example, suggest that instead of searching counterfactuals in the feature space $\mathcal{X}$, we can instead traverse a latent embedding $\mathcal{Z}$ (Equation~\ref{eq:general}) that implicitly codifies the DGP.
-To learn the latent embedding, they propose using a generative model such as a Variational Autoencoder (VAE).
-Provided the surrogate model is well-specified, their proposed approach called \textit{REVISE} can yield plausible explanations.
-Others have proposed similar approaches: \citet{dombrowski2021diffeomorphic} traverse the base space of a normalizing flow to solve Equation~\ref{eq:general}; \citet{poyiadzi2020face} use density estimators ($\hat{p}: \mathcal{X} \mapsto [0,1]$) to constrain the counterfactuals to dense regions in the feature space; and, finally, \citet{karimi2021algorithmic} assume knowledge about the structural causal model that generates the data.
-
-A competing approach towards plausibility that is also closely related to this work instead relies on the black-box model itself.
-\citet{schut2021generating} show that to meet the plausibility objective we need not explicitly model the input distribution.
-Pointing to the undesirable engineering overhead induced by surrogate models, they propose that we rely on the implicit minimisation of predictive uncertainty instead.
-Their proposed methodology, which we will refer to as \textit{Schut}, solves Equation~\ref{eq:general} by greedily applying Jacobian-Based Saliency Map Attacks (JSMA) in the feature space with cross-entropy loss and no penalty at all.
-The authors demonstrate theoretically and empirically that their approach yields counterfactuals for which the model $M_{\theta}$ predicts the target label $\mathbf{y}^+$ with high confidence.
-Provided the model is well-specified, these counterfactuals are plausible.
-This idea hinges on the assumption that the black-box model provides well-calibrated predictive uncertainty estimates.
-
-\section{Why Fidelity is not Enough}\label{fidelity}
-
-As discussed in the introduction, any valid counterfactual also has full fidelity by construction: solutions to Equation~\ref{eq:general} are considered valid as soon as the label predicted by the model matches the target class.
-So while fidelity always applies, counterfactuals that address the various desiderata introduced above can look vastly different from each other.
-
-To demonstrate this with an example, we have trained a simple image classifier $M_{\theta}$ on the well-known \textit{MNIST} dataset~\citep{lecun1998mnist}: a Multi-Layer Perceptron (\textit{MLP}) with above 90 percent test accuracy.
-No measures have been taken to improve the model's adversarial robustness or its capacity for predictive uncertainty quantification.
-The far left panel of Figure ~\ref{fig:motiv} shows a random sample drawn from the dataset.
-The underlying classifier correctly predicts the label `nine' for this image.
-For the given factual image and model, we have used \textit{Wachter}, \textit{Schut} and \textit{REVISE} to generate one counterfactual each in the target class `seven'.
-The perturbed images are shown next to the factual image from left to right in Figure ~\ref{fig:motiv}.
-Captions on top of the individual images indicate the generator along with the predicted probability that the image belongs to the target class.
-In all three cases that probability is above 90 percent and yet the counterfactuals look very different from each other.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=0.8\textwidth]{../artifacts/results/images/mnist_motivation.png}
-  \caption{Counterfactuals for turning a 9 (nine) into a 7 (seven): original image (left); then from left to right the counterfactuals generated using \textit{Wachter}, \textit{Schut} and \textit{REVISE}.}\label{fig:motiv}
-\end{figure}
-
-Since \textit{Wachter} is only concerned with proximity, the generated counterfactual is almost indistinguishable from the factual.
-The approach by~\citet{schut2021generating} expects a well-calibrated model that can generate predictive uncertainty estimates.
-Since this is not the case, the generated counterfactual looks like an adversarial example.
-Finally, the counterfactual generated by \textit{REVISE} looks much more plausible than the other two.
-But is it also more faithful to the behaviour of our \textit{MNIST} classifier?
-That is much less clear because the surrogate used by \textit{REVISE} introduces friction: the generated explanations no longer depend exclusively on the black-box model itself.
-
-So which of the counterfactuals most faithfully explains the behaviour of our image classifier?
-Fidelity cannot help us to make that judgement, because all of these counterfactuals have full fidelity.
-Thus, fidelity is an insufficient evaluation metric to assess the faithfulness of CE.
-
-\section{A New Notion of Faithfulness}\label{faithfulness}
-
-Considering the limitations of fidelity as demonstrated in the previous section, analogous to Definition~\ref{def:plausible}, we introduce a new notion of faithfulness in the context of CE:
-
-\begin{definition}[Faithful Counterfactuals]
-  \label{def:faithful}
-  Let $\mathcal{X}_{\theta}|\mathbf{y}^+ = p_{\theta}(\mathbf{x}|\mathbf{y}^+)$ denote the conditional distribution of $\mathbf{x}$ in the target class $\mathbf{y}^+$, where $\theta$ denotes the parameters of model $M_{\theta}$.
-Then for $\mathbf{x}^{\prime}$ to be considered a faithful counterfactual, we need: $\mathbf{x}^{\prime} \sim \mathcal{X}_{\theta}|\mathbf{y}^+$.
-\end{definition}
-
-In doing this, we merge in and nuance the concept of plausibility (Definition~\ref{def:plausible}) where the notion of `consistent with the data' becomes `consistent with what the model has learned about the data'.
-
-\subsection{Quantifying the Model's Generative Property}
-
-To assess counterfactuals with respect to Definition~\ref{def:faithful}, we need a way to quantify the posterior conditional distribution $p_{\theta}(\mathbf{x}|\mathbf{y}^+)$.
-To this end, we draw on recent advances in Energy-Based Modelling (EBM), a subdomain of machine learning that is concerned with generative or hybrid modelling~\citep{grathwohl2020your,du2020implicit}.
-In particular, note that if we fix $\mathbf{y}$ to our target value $\mathbf{y}^+$, we can conditionally draw from $p_{\theta}(\mathbf{x}|\mathbf{y}^+)$ by randomly initializing $\mathbf{x}_0$ and then using Stochastic Gradient Langevin Dynamics (SGLD) as follows,
-
-\begin{equation}\label{eq:sgld}
-  \begin{aligned}
-    \mathbf{x}_{j+1} &\leftarrow \mathbf{x}_j - \frac{\epsilon^2}{2} \mathcal{E}(\mathbf{x}_j|\mathbf{y}^+) + \epsilon \mathbf{r}_j, && j=1,...,J
-  \end{aligned}
-\end{equation}
-
-where $\mathbf{r}_j \sim \mathcal{N}(\mathbf{0},\mathbf{I})$ is the stochastic term and the step-size $\epsilon$ is typically polynomially decayed~\citep{welling2011bayesian}.
-The term $\mathcal{E}(\mathbf{x}_j|\mathbf{y}^+)$ denotes the model energy conditioned on the target class label $\mathbf{y}^+$ which we specify as the negative logit corresponding to the target class label $\mathbf{y}^*$.
-To allow for faster sampling, we follow the common practice of choosing the step-size $\epsilon$ and the standard deviation of $\mathbf{r}_j$ separately.
-While $\mathbf{x}_J$ is only guaranteed to distribute as $p_{\theta}(\mathbf{x}|\mathbf{y}^*)$ if $\epsilon \rightarrow 0$ and $J \rightarrow \infty$, the bias introduced for a small finite $\epsilon$ is negligible in practice \citep{murphy2023probabilistic,grathwohl2020your}.
-Appendix~\ref{app:jem} provides additional implementation details for any tasks related to energy-based modelling.
-
-Generating multiple samples using SGLD thus yields an empirical distribution $\hat{\mathbf{X}}_{\theta,\mathbf{y}^+}$ that approximates what the model has learned about the input data.
-While in the context of EBM, this is usually done during training, we propose to repurpose this approach during inference in order to evaluate and generate faithful model explanations.
-
-\subsection{Evaluating Plausibility and Faithfulness}
-
-The parallels between our definitions of plausibility and faithfulness imply that we can also use similar evaluation metrics in both cases.
-Since existing work has focused heavily on plausibility, it offers a useful starting point.
-In particular,~\citet{guidotti2022counterfactual} have proposed an implausibility metric that measures the distance of the counterfactual from its nearest neighbour in the target class.
-As this distance is reduced, counterfactuals get more plausible under the assumption that the nearest neighbour itself is plausible in the sense of Definition~\ref{def:plausible}.
-In this work, we use the following adapted implausibility metric,
-
-\begin{equation}\label{eq:impl}
-  \begin{aligned}
-    \text{impl}(\mathbf{x}^{\prime},\mathbf{X}_{\mathbf{y}^+}) = \frac{1}{\lvert\mathbf{X}_{\mathbf{y}^+}\rvert} \sum_{\mathbf{x} \in \mathbf{X}_{\mathbf{y}^+}} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
-  \end{aligned}
-\end{equation}
-
-where $\mathbf{x}^{\prime}$ denotes the counterfactual and $\mathbf{X}_{\mathbf{y}^+}$ is a subsample of the training data in the target class $\mathbf{y}^+$.
-By averaging over multiple samples in this manner, we avoid the risk that the nearest neighbour of $\mathbf{x}^{\prime}$ itself is not plausible according to Definition~\ref{def:plausible} (e.g an outlier).
-
-Equation~\ref{eq:impl} gives rise to a similar evaluation metric for unfaithfulness.
-We merely swap out the subsample of individuals in the target class for a subset $\hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^+}$ of the generated conditional samples:
-
-\begin{equation}\label{eq:faith}
-  \begin{aligned}
-    \text{unfaith}(\mathbf{x}^{\prime},\hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^+}) = \frac{1}{\lvert \hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^+}\rvert} \sum_{\mathbf{x} \in \hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^+}} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
-  \end{aligned}
-\end{equation}
-
-Specifically, we form this subset based on the $n_E$ generated samples with the lowest energy.
-
-\section{Energy-Constrained Conformal Counterfactuals}\label{meth}
-
-In this section, we describe \textit{ECCCo}, our proposed framework for generating Energy-Constrained Conformal Counterfactuals (ECCCos).
-It is based on the premise that counterfactuals should first and foremost be faithful.
-Plausibility, as a secondary concern, is then still attainable, but only to the degree that the black-box model itself has learned plausible explanations for the underlying data.
-
-We begin by stating our proposed objective function, which involves tailored loss and penalty functions that we will explain in the following.
-In particular, we extend Equation~\ref{eq:general} as follows:
-
-\begin{equation} \label{eq:eccco}
-  \begin{aligned}
-  \mathbf{Z}^\prime= \arg \min_{\mathbf{Z}^\prime \in \mathcal{Z}^M}  &\{  {\text{yloss}(M_{\theta}(f(\mathbf{Z}^\prime)),\mathbf{y}^+)}+ \lambda_{1} {\text{dist}(f(\mathbf{Z}^\prime),\mathbf{x}) } \\
-  &+ \lambda_2 \text{unfaith}(f(\mathbf{Z}^\prime),\hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^+}) + \lambda_3 \Omega(C_{\theta}(f(\mathbf{Z}^\prime);\alpha)) \} 
-  \end{aligned} 
-\end{equation}
-
-The first penalty term involving $\lambda_1$ induces proximity like in~\citet{wachter2017counterfactual}.
-Our default choice for $\text{dist}(\cdot)$ is the L1 Norm due to its sparsity-inducing properties.
-The second penalty term involving $\lambda_2$ induces faithfulness by constraining the energy of the generated counterfactual where $\text{unfaith}(\cdot)$ corresponds to the metric defined in Equation~\ref{eq:faith}.
-The third and final penalty term involving $\lambda_3$ introduces a new concept: it ensures that the generated counterfactual is associated with low predictive uncertainty.
-As mentioned above,~\citet{schut2021generating} have shown that plausible counterfactuals can be generated implicitly through predictive uncertainty minimization.
-Unfortunately, this relies on the assumption that the model itself can provide predictive uncertainty estimates, which may be too restrictive in practice.
-
-To relax this assumption, we leverage recent advances in Conformal Prediction (CP), an approach to predictive uncertainty quantification that has recently gained popularity~\citep{angelopoulos2021gentle,manokhin2022awesome}.
-Crucially for our intended application, CP is model-agnostic and can be applied during inference without placing any restrictions on model training.
-Intuitively, CP works under the premise of turning heuristic notions of uncertainty into rigorous uncertainty estimates by repeatedly sifting through the training data or a dedicated calibration dataset.
-Conformal classifiers produce prediction sets for individual inputs that include all output labels that can be reasonably attributed to the input.
-These sets tend to be larger for inputs that do not conform with the training data and are characterized by high predictive uncertainty.
-
-In order to generate counterfactuals that are associated with low predictive uncertainty, we use a smooth set size penalty introduced by~\citet{stutz2022learning} in the context of conformal training:
-
-\begin{equation}\label{eq:setsize}
-  \begin{aligned}
-    \Omega(C_{\theta}(\mathbf{x};\alpha))&=\max \left(0, \sum_{\mathbf{y}\in\mathcal{Y}}C_{\theta,\mathbf{y}}(\mathbf{x}_i;\alpha) - \kappa \right)
-  \end{aligned}
-\end{equation}
-
-Here, $\kappa \in \{0,1\}$ is a hyper-parameter and $C_{\theta,\mathbf{y}}(\mathbf{x}_i;\alpha)$ can be interpreted as the probability of label $\mathbf{y}$ being included in the prediction set.
-In order to compute this penalty for any black-box model we merely need to perform a single calibration pass through a holdout set $\mathcal{D}_{\text{cal}}$.
-Arguably, data is typically abundant and in most applications, practitioners tend to hold out a test data set anyway.
-Consequently, CP removes the restriction on the family of predictive models, at the small cost of reserving a subset of the available data for calibration.
-This particular case of conformal prediction is referred to as Split Conformal Prediction (SCP) as it involves splitting the training data into a proper training dataset and a calibration dataset.
-In addition to the smooth set size penalty, we have also experimented with the use of a tailored function for $\text{yloss}(\cdot)$ that enforces that only the target label $\mathbf{y}^+$ is included in the prediction set ~\citet{stutz2022learning}.
-Further details are provided in Appendix~\ref{app:cp}.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=1.0\textwidth]{../artifacts/results/images/poc_gradient_fields.png}
-  \caption{Gradient fields and counterfactual paths for different generators.
-The objective is to generate a counterfual in the `blue' class for a sample from the `orange' class.
-Bright yellow stars indicate conditional samples generated through SGLD.
-The underlying classifier is a Joint Energy Model.}\label{fig:poc}
-\end{figure}
-
-\renewcommand{\algorithmicrequire}{\textbf{Input:}}
-\renewcommand{\algorithmicensure}{\textbf{Output:}}
-
-\begin{algorithm}
-  \caption{The \textit{ECCCo} generator}\label{alg:eccco}
-  \begin{algorithmic}[1]
-    \Require $\mathbf{x}, \mathbf{y}^+, M_{\theta}, f, \Lambda=[\lambda_1,\lambda_2,\lambda_3], \alpha, \mathcal{D}, T, \eta, n_{\mathcal{B}}, n_E$ where $M_{\theta}(\mathbf{x})\neq\mathbf{y}^+$
-    \Ensure $\mathbf{x}^\prime$
-    \State Initialize $\mathbf{z}^\prime \gets f^{-1}(\mathbf{x})$ \Comment{Map to counterfactual state space.}
-    \State Generate $\left\{\hat{\mathbf{x}}_{\theta,\mathbf{y}^+}\right\}_{n_{\mathcal{B}}} \gets p_{\theta}(\mathbf{x}_{\mathbf{y}^+})$ \Comment{Generate $n_{\mathcal{B}}$ samples using SGLD (Equation~\ref{eq:sgld}).}
-    \State Store $\hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^+} \gets \left\{\hat{\mathbf{x}}_{\theta,\mathbf{y}^+}\right\}_{n_{\mathcal{B}}}$ \Comment{Choose $n_E$ lowest-energy samples.}
-    \State Run \textit{SCP} for $M_{\theta}$ using $\mathcal{D}$ \Comment{Calibrate model through Split Conformal Prediction.}
-    \State Initialize $t \gets 0$
-    \While{\textit{not converged} or $t < T$} \Comment{For convergence conditions see Appendix~\ref{app:eccco}.}
-    \State $\mathbf{z}^\prime \gets \mathbf{z}^\prime - \eta \nabla_{\mathbf{z}^\prime} \mathcal{L}(\mathbf{z}^\prime,\mathbf{y}^+,\hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^+}; \Lambda, \alpha)$ \Comment{Take gradient step of size $\eta$.}
-    \State $t \gets t+1$
-    \EndWhile
-    \State $\mathbf{x}^\prime \gets f(\mathbf{z}^\prime)$ \Comment{Map back to feature space.}
-  \end{algorithmic}
-\end{algorithm}
-
-To provide some further intuition about our objective defined in Equation~\ref{eq:eccco}, Figure~\ref{fig:poc} illustrates how the different components affect the counterfactual search for a synthetic dataset.
-The underlying classifier is a Joint Energy Model (\textit{JEM}) that was trained to predict the output class (`blue' or `orange') and generate class-conditional samples~\citep{grathwohl2020your}.
-We have used four different generator flavours to produce a counterfactual in the `blue' class for a sample from the `orange' class: \textit{Wachter}, which only uses the first penalty ($\lambda_2=\lambda_3=0$); \textit{ECCCo (no EBM)}, which does not constrain energy ($\lambda_2=0$); \textit{ECCCo (no CP)}, which involves no set size penalty ($\lambda_3=0$); and, finally, \textit{ECCCo}, which involves all penalties defined in Equation~\ref{eq:eccco}.
-Arrows indicate (negative) gradients with respect to the objective function at different points in the feature space.
-
-While \textit{Wachter} generates a valid counterfactual, it ends up close to the original starting point consistent with its objective.
-\textit{ECCCo (no EBM)} pushes the counterfactual further into the target domain to minimize predictive uncertainty, but the outcome is still not plausible.
-The counterfactual produced by \textit{ECCCo (no CP)} is attracted by the generated samples shown in bright yellow.
-Since the \textit{JEM} has learned the conditional input distribution reasonably well in this case, the counterfactuals are both faithful and plausible.
-Finally, the outcome for \textit{ECCCo} looks similar, but the additional smooth set size penalty leads to somewhat faster convergence.
-
-Algorithm~\ref{alg:eccco} describes how exactly \textit{ECCCo} works.
-For the sake of simplicity and without loss of generality, we limit our attention to generating a single counterfactual $\mathbf{x}^\prime=f(\mathbf{z}^\prime)$.
-The counterfactual state $\mathbf{z}^\prime$ is initialized by passing the factual $\mathbf{x}$ through a simple feature transformer $f^{-1}$.
-Next, we generate $n_{\mathcal{B}}$ conditional samples $\hat{\mathbf{x}}_{\theta,\mathbf{y}^+}$ using SGLD (Equation~\ref{eq:sgld}) and store the $n_E$ instances with the lowest energy.
-We then calibrate the model $M_{\theta}$ through Split Conformal Prediction.
-Finally, we search counterfactuals through gradient descent where $\mathcal{L}(\mathbf{z}^\prime,\mathbf{y}^+,\hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^+}; \Lambda, \alpha)$ denotes our loss function defined in Equation~\ref{eq:eccco}.
-The search terminates once the convergence criterium is met or the maximum number of iterations $T$ has been exhausted.
-Note that the choice of convergence criterium has important implications on the final counterfactual which we explain in Appendix~\ref{app:eccco}.
-
-\section{Empirical Analysis}\label{emp}
-
-Our goal in this section is to shed light on the following research questions:
-
-\begin{question}[Faithfulness]\label{rq:faithfulness}
-  Are ECCCos more faithful than counterfactuals produced by our benchmark generators?
-\end{question}
-
-\begin{question}[Balancing Objectives]\label{rq:plausibility}
-  Compared to our benchmark generators, how do ECCCos balance the two key objectives of faithfulness and plausibility?
-\end{question}
-
-The second question is motivated by the intuition that faithfulness and plausibility should coincide for models that have learned plausible explanations of the data.
-Next, we first briefly describe our experimental setup before presenting our main results.
-
-\subsection{Experimental Setup}
-
-To assess and benchmark the performance of our proposed generator against the state of the art, we generate multiple counterfactuals for different models and datasets.
-In particular, we compare \textit{ECCCo} and its variants to the following counterfactual generators that were introduced above: firstly; \textit{Schut}, which works under the premise of minimizing predictive uncertainty; secondly, \textit{REVISE}, which is state-of-the-art with respect to plausibility; and, finally, \textit{Wachter}, which serves as our baseline.
-
-We use both synthetic and real-world datasets from different domains, all of which are publicly available and commonly used to train and benchmark classification algorithms.
-We synthetically generate a dataset containing two \textit{Linearly Separable} Gaussian clusters ($n=1000$), as well as the well-known \textit{Circles} ($n=1000$) and \textit{Moons} ($n=2500$) data.
-Since these data are generated by distributions of varying degrees of complexity, they allow us to assess how the generators and our proposed evaluation metrics handle this.
-
-As for real-world data, we follow~\citet{schut2021generating} and use the \textit{MNIST}~\citep{lecun1998mnist} dataset containing images of handwritten digits such as the example shown above in Figure~\ref{fig:motiv}.
-From the social sciences domain, we include Give Me Some Credit (\textit{GMSC})~\citep{kaggle2011give}: a tabular dataset that has been studied extensively in the literature on Algorithmic Recourse~\citep{pawelczyk2021carla}.
-It consists of 11 numeric features that can be used to predict the binary outcome variable indicating whether retail borrowers experience financial distress.
-
-For the predictive modelling tasks, we use simple neural networks (\textit{MLP}) and Joint Energy Models (\textit{JEM}).
-For the more complex real-world datasets we also use ensembling in each case.
-Both joint-energy modelling and ensembling have been associated with improved generative properties and adversarial robustness~\citep{grathwohl2020your,lakshminarayanan2016simple}, so we expect this to be positively correlated with the plausibility of ECCCos.
-To account for stochasticity, we generate multiple counterfactuals for each target class, generator, model and dataset.
-Specifically, we randomly sample $n^{-}$ times from the subset of individuals for which the given model predicts the non-target class $\mathbf{y}^{-}$ given the current target.
-We set $n^{-}=25$ for all of our synthetic datasets, $n^{-}=10$ for \textit{GMSC} and $n^{-}=5$ for \textit{MNIST}.
-Full details concerning our parameter choices, training procedures and model performance can be found in Appendix~\ref{app:setup}.
-
-\subsection{Results for Synthetic Data}
-
-Table~\ref{tab:results-synthetic} shows the key results for the synthetic datasets separated by model (first column) and generator (second column).
-The numerical columns show sample averages and standard deviations of our key evaluation metrics computed across all counterfactuals.
-We have highlighted the best outcome for each model and metric in bold.
-To provide some sense of effect sizes, we have added asterisks to indicate that a given value is at least one ($*$) or two ($**$) standard deviations lower than the baseline (\textit{Wachter}).
-
-Starting with the high-level results for our \textit{Linearly Separable} data, we find that \textit{ECCCo} produces the most faithful counterfactuals for both black-box models.
-This is consistent with our design since \textit{ECCCo} directly enforces faithfulness through regularization.
-Crucially though, \textit{ECCCo} also produces the most plausible counterfactuals for both models.
-This dataset is so simple that even the \textit{MLP} has learned plausible explanations of the input data.
-Zooming in on the granular details for the \textit{Linearly Separable} data, the results for \textit{ECCCo (no CP)} and \textit{ECCCo (no EBM)} indicate that the positive results are dominated by the effect of quantifying and leveraging the model's generative property (EBM).
-Conformal Prediction alone only leads to marginally improved faithfulness and plausibility.
-
-The findings for the \textit{Moons} dataset are broadly in line with the findings so far: for the \textit{JEM}, \textit{ECCCo} yields substantially more faithful and plausible counterfactuals than all other generators.
-For the \textit{MLP}, faithfulness is maintained but counterfactuals are not plausible.
-This high-level pattern is broadly consistent with other more complex datasets and supportive of our narrative, so it is worth highlighting: ECCCos consistently achieve high faithfulness, which---subject to the quality of the model itself---coincides with high plausibility.
-By comparison, \textit{REVISE} yields the most plausible counterfactuals for the \textit{MLP}, but it does so at the cost of faithfulness.
-We also observe that the best results for \textit{ECCCo} are achieved when using both penalties.
-Once again though, the generative component (EBM) has a stronger impact on the positive results for the \textit{JEM}.
-
-For the \textit{Circles} data, it appears that \textit{REVISE} performs well, but we note that it generates valid counterfactuals only half of the time (see Appendix~\ref{app:results} for a complete overview including additional common evaluation metrics).
-The underlying VAE with default parameters has not adequately learned the data-generating process.
-Of course, it is possible to improve generative performance through hyperparameter tuning but this example serves to illustrate that \textit{REVISE} depends on the quality of its surrogate.
-Independent of the outcome for \textit{REVISE}, however, the results do not seem to indicate that \textit{ECCCo} substantially improves faithfulness and plausibility for the \textit{Circles} data.
-We think this points to a limitation of our evaluation metrics rather than \textit{ECCCo} itself: computing average distances fails to account for the `wraparound' effect associated with circular data~\citep{gill2010circular}.
-
-\import{contents/}{table-synthetic.tex}
-
-\subsection{Results for Real-World Data}
-
-The results for our real-world datasets are shown in Table~\ref{tab:results-real-world}.
-Once again the findings indicate that the plausibility of ECCCos is positively correlated with the capacity of the black-box model to distinguish plausible from implausible inputs.
-The case is very clear for \textit{MNIST}: ECCCos are consistently more faithful than the counterfactuals produced by our benchmark generators and their plausibility gradually improves through ensembling and joint-energy modelling.
-Interestingly, faithfulness also gradually improves for \textit{REVISE}.
-This indicates that as our models improve, their generative capacity approaches that of the surrogate VAE used by \textit{REVISE}.
-The VAE still outperforms our classifiers in this regard, as evident from the fact that \textit{ECCCo} never quite reaches the same level of plausibility as \textit{REVISE}.
-With reference to Appendix~\ref{app:results} we note that the results for \textit{Schut} need to be discounted as it rarely produces valid counterfactuals for \textit{MNIST}.
-Relatedly, we find that \textit{ECCCo} is the only generator that consistently achieves full validity.
-Finally, it is worth noting that \textit{ECCCo} produces counterfactual images with the lowest average predictive uncertainty for all models.
-
-For the tabular credit dataset (\textit{GMSC}) it is inherently challenging to use deep neural networks in order to achieve good discriminative performance~\citep{borisov2021deep,grinsztajn2022why} and generative performance~\citep{liu2023goggle}, respectively.
-In order to achieve high plausibility, \textit{ECCCo} effectively requires classifiers to achieve good performance for both tasks.
-Since this is a challenging task even for Joint Energy Models, it is not surprising to find that even though \textit{ECCCo} once again achieves state-of-the-art faithfulness, it is outperformed by \textit{REVISE} and \textit{Schut} with respect to plausibility.
-
-\subsection{Key Takeways}
-
-To conclude this section, we summarize our findings with reference to the opening questions.
-The results clearly demonstrate that \textit{ECCCo} consistently achieves state-of-the-art faithfulness, as it was designed to do (Research Question~\ref{rq:faithfulness}).
-A related important finding is that \textit{ECCCo} yields highly plausible explanations provided that they faithfully describe model behaviour (Research Question~\ref{rq:plausibility}).
-\textit{ECCCo} achieves this result primarily by leveraging the model's generative property.
-
-\import{contents/}{table-real-world.tex}
-
-\section{Limitations}
-
-Even though we have taken considerable measures to study our proposed methodology carefully, limitations can still be identified.
-In particular, we have found that the performance of \textit{ECCCo} is sensitive to hyperparameter choices.
-In order to achieve faithfulness, we generally had to penalise the distance from generated samples slightly more than the distance from factual values.
-
-Conversely, we have not found that strongly penalising prediction set sizes had any discernable effect.
-Our results indicate that CP alone is often not sufficient to achieve faithfulness and plausibility, although we acknowledge that this needs to be investigated more thoroughly through future work.
-
-While our approach is readily applicable to models with gradient access like deep neural networks, more work is needed to generalise it to other machine learning models such as decision trees.
-Relatedly, common challenges associated with Energy-Based Modelling including sensitivity to scale, training instabilities and sensitivity to hyperparameters also apply to \textit{ECCCo}.
-
-\section{Conclusion}
-
-This work leverages recent advances in Energy-Based Modelling and Conformal Prediction in the context of Explainable Artificial Intelligence.
-We have proposed a new way to generate counterfactuals that are maximally faithful to the black-box model they aim to explain.
-Our proposed generator, \textit{ECCCo}, produces plausible counterfactuals if and only if the black-box model itself has learned realistic explanations for the data, which we have demonstrated through rigorous empirical analysis.
-This should enable researchers and practitioners to use counterfactuals in order to discern trustworthy models from unreliable ones.
-While the scope of this work limits its generalizability, we believe that \textit{ECCCo} offers a solid baseline for future work on faithful Counterfactual Explanations.
-
-\begin{ack}
-
-Some of the members of TU Delft were partially funded by ICAI AI for Fintech Research, an ING — TU Delft
-collaboration.
-
-\end{ack}
-
-\bibliography{bib}
-
-\pagebreak
-
-\appendix
-\section*{Appendices}
-\renewcommand{\thesubsection}{\Alph{subsection}}
-
-The following appendices provide additional details that are relevant to the paper.
-Appendices~\ref{app:jem} and~\ref{app:cp} explain any tasks related to Energy-Based Modelling and Predictive Uncertainty Quantification through Conformal Prediction, respectively.
-Appendix~\ref{app:eccco} provides additional technical and implementation details about our proposed generator, \textit{ECCCo}, including references to our open-sourced code base.
-A complete overview of our experimental setup detailing our parameter choices, training procedures and initial black-box model performance can be found in Appendix~\ref{app:setup}.
-Finally, Appendix~\ref{app:results} reports all of our experimental results in more detail.
-
-\subsection{Energy-Based Modelling}\label{app:jem}
-
-Since we were not able to identify any existing open-source software for Energy-Based Modelling that would be flexible enough to cater to our needs, we have developed a \texttt{Julia} package from scratch.
-The package has been open-sourced, but to avoid compromising the double-blind review process, we refrain from providing more information at this stage.
-In our development we have heavily drawn on the existing literature:~\citet{du2020implicit} describe best practices for using EBM for generative modelling;~\citet{grathwohl2020your} explain how EBM can be used to train classifiers jointly for the discriminative and generative tasks.
-We have used the same package for training and inference, but there are some important differences between the two cases that are worth highlighting here.
-
-\subsubsection{Training: Joint Energy Models}
-
-To train our Joint Energy Models we broadly follow the approach outlined in~\citet{grathwohl2020your}.
-These models are trained to optimize a hybrid objective that involves a standard classification loss component $L_{\text{clf}}(\theta)=-\log p_{\theta}(\mathbf{y}|\mathbf{x})$ (e.g.
-cross-entropy loss) as well as a generative loss component $L_{\text{gen}}(\theta)=-\log p_{\theta}(\mathbf{x})$.
-
-To draw samples from $p_{\theta}(\mathbf{x})$, we rely exclusively on the conditional sampling approach described in~\citet{grathwohl2020your} for both training and inference: we first draw $\mathbf{y}\sim p(\mathbf{y})$ and then sample $\mathbf{x} \sim p_{\theta}(\mathbf{x}|\mathbf{y})$~\citep{grathwohl2020your} via Equation~\ref{eq:sgld} with energy $\mathcal{E}(\mathbf{x}|\mathbf{y})=\mu_{\theta}(\mathbf{x})[\mathbf{y}]$ where $\mu_{\theta}: \mathcal{X} \mapsto \mathbb{R}^K$ returns the linear predictions (logits) of our classifier $M_{\theta}$.
-While our package also supports unconditional sampling, we found conditional sampling to work well.
-It is also well aligned with CE, since in this context we are interested in conditioning on the target class.
-
-As mentioned in the body of the paper, we rely on a biased sampler involving separately specified values for the step size $\epsilon$ and the standard deviation $\sigma$ of the stochastic term involving $\mathbf{r}$.
-Formally, our biased sampler performs updates as follows:
-
-\begin{equation}\label{eq:biased-sgld}
-  \begin{aligned}
-    \hat{\mathbf{x}}_{j+1} &\leftarrow \hat{\mathbf{x}}_j - \frac{\epsilon}{2} \mathcal{E}(\hat{\mathbf{x}}_j|\mathbf{y}^+) + \sigma \mathbf{r}_j, && j=1,...,J
-  \end{aligned}
-\end{equation}
-
-Consistent with~\citet{grathwohl2020your}, we have specified $\epsilon=2$ and $\sigma=0.01$ as the default values for all of our experiments.
-The number of total SGLD steps $J$ varies by dataset (Table~\ref{tab:ebmparams}).
-Following best practices, we initialize $\mathbf{x}_0$ randomly in 5\% of all cases and sample from a buffer in all other cases.
-The buffer itself is randomly initialised and gradually grows to a maximum of 10,000 samples during training as $\hat{\mathbf{x}}_{J}$ is stored in each epoch~\citep{du2020implicit,grathwohl2020your}.
-
-It is important to realise that sampling is done during each training epoch, which makes training Joint Energy Models significantly harder than conventional neural classifiers.
-In each epoch the generated (batch of) sample(s) $\hat{\mathbf{x}}_{J}$ is used as part of the generative loss component, which compares its energy to that of observed samples $\mathbf{x}$: $L_{\text{gen}}(\theta)=\mu_{\theta}(\mathbf{x})[\mathbf{y}]-\mu_{\theta}(\hat{\mathbf{x}}_{J})[\mathbf{y}]$.
-Our full training objective can be summarized as follows,
-
-\begin{equation}\label{eq:jem-loss}
-  \begin{aligned}
-    L(\theta) &= L_{\text{clf}}(\theta) + L_{\text{gen}}(\theta) + \lambda L_{\text{reg}}(\theta) 
-  \end{aligned}
-\end{equation}
-
-where $L_{\text{reg}}(\theta)$ is a Ridge penalty (L2 norm) that regularises energy magnitudes for both observed and generated samples~\citep{du2020implicit}.
-We have used varying degrees of regularization depending on the dataset ($\lambda$ in Table~\ref{tab:ebmparams}).
-
-Contrary to existing work, we have not typically used the entire minibatch of training data for the generative loss component but found that using a subset of the minibatch was often sufficient in attaining decent generative performance (Table~\ref{tab:ebmparams}).
-This has helped to reduce the computational burden for our models, which should make it easier for others to reproduce our findings.
-Figures~\ref{fig:mnist-gen} and~\ref{fig:moons-gen} show generated samples for our \textit{MNIST} and \textit{Moons} data, to provide a sense of their generative property.
-
-\import{contents/}{table_ebm_params.tex}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=0.75\textwidth]{../artifacts/results/images/mnist_generated_JEM Ensemble.png}
-  \caption{Conditionally generated \textit{MNIST} images for our JEM Ensemble.}\label{fig:mnist-gen}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=0.5\textwidth]{../artifacts/results/images/moons_generated_JEM.png}
-  \caption{Conditionally generated samples (stars) for our \textit{Moons} data using a JEM.}\label{fig:moons-gen}
-\end{figure}
-\subsubsection{Inference: Quantifying Models' Generative Property}
-
-At inference time, we assume no prior knowledge about the model's generative property.
-This means that we do not tab into the existing buffer of generated samples for our Joint Energy Models, but instead generate conditional samples from scratch.
-While we have relied on the default values $\epsilon=2$ and $\sigma=0.01$ also during inference, the number of total SGLD steps was set to $J=500$ in all cases, so significantly higher than during training.
-For all of our synthetic datasets and models, we generated 50 conditional samples and then formed subsets containing the $n_{E}=25$ lowest-energy samples.
-While in practice it would be sufficient to do this once for each model and dataset, we have chosen to perform sampling separately for each individual counterfactual in our experiments to account for stochasticity.
-To help reduce the computational burden for our real-world datasets we have generated only 10 conditional samples each time and used all of them in our counterfactual search.
-Using more samples, as we originally did, had no substantial impact on our results.
-
-\subsection{Conformal Prediction}\label{app:cp}
-
-In this Appendix~\ref{app:cp} we provide some more background on CP and explain in some more detail how we have used recent advances in Conformal Training for our purposes.
-
-\subsubsection{Background on CP}
-
-Intuitively, CP works under the premise of turning heuristic notions of uncertainty into rigorous uncertainty estimates by repeatedly sifting through the data.
-It can be used to generate prediction intervals for regression models and prediction sets for classification models.
-Since the literature on CE and AR is typically concerned with classification problems, we focus on the latter.
-A particular variant of CP called Split Conformal Prediction (SCP) is well-suited for our purposes, because it imposes only minimal restrictions on model training.
-
-Specifically, SCP involves splitting the data $\mathcal{D}_n=\{(\mathbf{x}_i,\mathbf{y}_i)\}_{i=1,...,n}$ into a proper training set $\mathcal{D}_{\text{train}}$ and a calibration set $\mathcal{D}_{\text{cal}}$.
-The former is used to train the classifier in any conventional fashion.
-The latter is then used to compute so-called nonconformity scores: $\mathcal{S}=\{s(\mathbf{x}_i,\mathbf{y}_i)\}_{i \in \mathcal{D}_{\text{cal}}}$ where $s: (\mathcal{X},\mathcal{Y}) \mapsto \mathbb{R}$ is referred to as \textit{score function}.
-In the context of classification, a common choice for the score function is just $s_i=1-M_{\theta}(\mathbf{x}_i)[\mathbf{y}_i]$, that is one minus the softmax output corresponding to the observed label $\mathbf{y}_i$~\citep{angelopoulos2021gentle}.
-
-Finally, classification sets are formed as follows,
-
-\begin{equation}\label{eq:scp}
-  \begin{aligned}
-    C_{\theta}(\mathbf{x}_i;\alpha)=\{\mathbf{y}: s(\mathbf{x}_i,\mathbf{y}) \le \hat{q}\}
-  \end{aligned}
-\end{equation}
-
-where $\hat{q}$ denotes the $(1-\alpha)$-quantile of $\mathcal{S}$ and $\alpha$ is a predetermined error rate.
-As the size of the calibration set increases, the probability that the classification set $C(\mathbf{x}_{\text{test}})$ for a newly arrived sample $\mathbf{x}_{\text{test}}$ does not cover the true test label $\mathbf{y}_{\text{test}}$ approaches $\alpha$~\citep{angelopoulos2021gentle}.
-
-Observe from Equation~\ref{eq:scp} that Conformal Prediction works on an instance-level basis, much like CE are local.
-The prediction set for an individual instance $\mathbf{x}_i$ depends only on the characteristics of that sample and the specified error rate.
-Intuitively, the set is more likely to include multiple labels for samples that are difficult to classify, so the set size is indicative of predictive uncertainty.
-To see why this effect is exacerbated by small choices for $\alpha$ consider the case of $\alpha=0$, which requires that the true label is covered by the prediction set with probability equal to 1.
-
-\subsubsection{Differentiability}
-
-The fact that conformal classifiers produce set-valued predictions introduces a challenge: it is not immediately obvious how to use such classifiers in the context of gradient-based counterfactual search.
-Put differently, it is not clear how to use prediction sets in Equation~\ref{eq:general}.
-Fortunately, \citet{stutz2022learning} have recently proposed a framework for Conformal Training that also hinges on differentiability.
-Specifically, they show how Stochastic Gradient Descent can be used to train classifiers not only for the discriminative task but also for additional objectives related to Conformal Prediction.
-One such objective is \textit{efficiency}: for a given target error rate $\alpha$, the efficiency of a conformal classifier improves as its average prediction set size decreases.
-To this end, the authors introduce a smooth set size penalty defined in Equation~\ref{eq:setsize} in the body of this paper.
-Formally, it is defined as $C_{\theta,\mathbf{y}}(\mathbf{x}_i;\alpha):=\sigma\left((s(\mathbf{x}_i,\mathbf{y})-\alpha) T^{-1}\right)$ for $\mathbf{y}\in\mathcal{Y}$, where $\sigma$ is the sigmoid function and $T$ is a hyper-parameter used for temperature scaling~\citep{stutz2022learning}.
-
-In addition to the smooth set size penalty,~\citet{stutz2022learning} also propose a configurable classification loss function, that can be used to enforce coverage.
-For \textit{MNIST} data, we found that using this function generally improved the visual quality of the generated counterfactuals, so we used it in our experiments involving real-world data.
-For the synthetic dataset, visual inspection of the counterfactuals showed that using the configurable loss function sometimes led to overshooting: counterfactuals would end up deep inside the target domain but far away from the observed samples.
-For this reason, we instead relied on standard cross-entropy loss for our synthetic datasets.
-As we have noted in the body of the paper, more experimental work is certainly needed in this context.
-Figure~\ref{fig:cp-diff} shows the prediction set size (left), smooth set size loss (centre) and configurable classification loss (right) for a \textit{JEM} trained on our \textit{Linearly Separable} data.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=1.0\textwidth]{../artifacts/results/images/poc_set_size.png}
-  \caption{Prediction set size (left), smooth set size loss (centre) and configurable classification loss (right) for a JEM trained on our \textit{Linearly Separable} data.}\label{fig:cp-diff}
-\end{figure}
-
-\subsection{ECCCo}\label{app:eccco}
-
-In this section, we briefly discuss convergence conditions for CE and provide details concerning the actual implementation of our framework in \texttt{Julia}.  
-\subsubsection{A Note on Convergence}
-
-Convergence is not typically discussed much in the context of CE, even though it has important implications on outcomes.
-One intuitive way to specify convergence is in terms of threshold probabilities: once the predicted probability $p(\mathbf{y}^+|\mathbf{x}^{\prime})$ exceeds some user-defined threshold $\gamma$ such that the counterfactual is valid, we could consider the search to have converged.
-In the binary case, for example, convergence could be defined as $p(\mathbf{y}^+|\mathbf{x}^{\prime})>0.5$ in this sense.
-Note, however, how this can be expected to yield counterfactuals in the proximity of the decision boundary, a region characterized by high aleatoric uncertainty.
-In other words, counterfactuals generated in this way would generally not be plausible.
-To avoid this from happening, we specify convergence in terms of gradients approaching zero for all our experiments and all of our generators.
-This is allows us to get a cleaner read on how the different counterfactual search objectives affect counterfactual outcomes.
-
-\subsubsection{\texttt{ECCCo.jl}}
-
-The core part of our code base is integrated into a larger ecosystem of \texttt{Julia} packages that we are actively developing and maintaining.
-To avoid compromising the double-blind review process, we only provide a link to an anonymized repository at this stage: \url{https://anonymous.4open.science/r/ECCCo-1252/README.md}.
-
-\subsection{Experimental Setup}\label{app:setup}
-
-Table~\ref{tab:params} provides an overview of all parameters related to our experiments.
-The \textit{GMSC} data were randomly undersampled for balancing purposes and all features were standardized.
-\textit{MNIST} data was also randomly undersampled for reasons outlined below.
-Pixel values were preprocessed to fall in the range of $[-1,1]$ and a small Gaussian noise component ($\sigma=0.03$) was added to training samples following common practice in the EBM literature.
-All of our models were trained through mini-batch training using the Adam optimiser (\citet{kingma2017adam}).
-Table~\ref{tab:perf} shows standard evaluation metrics measuring the predictive performance of our different models grouped by dataset.
-These measures were computed on test data.
-
-Table~\ref{tab:genparams} summarises our hyperparameter choices for the counterfactual generators where $\eta$ denotes the learning rate used for Stochastic Gradient Descent (SGD) and $\lambda_1$, $\lambda_2$, $\lambda_3$ represent the chosen penalty strengths (Equations~\ref{eq:general} and~\ref{eq:eccco}).
-Here $\lambda_1$ also refers to the chosen penalty for the distance from factual values that applies to both \textit{Wachter} and \textit{REVISE}, but not \textit{Schut} which is penalty-free.
-\textit{Schut} is also the only generator that uses JSMA instead of SGD for optimization.
-
-\import{contents/}{table_params.tex}
-
-\import{contents/}{table_perf.tex}
-
-\import{contents/}{table_gen_params.tex}
-
-\subsubsection{Compute}
-
-To enable others to easily replicate our experiments, we have chosen to work with small neural network architectures and randomly undersampled the \textit{MNIST} dataset (maintaining class balance).
-All of our experiments could then be run locally on a personal machine.
-The longest runtimes we experienced for model training and counterfactual benchmarking were on the order of 8-12 hours (\textit{MNIST} data).
-For the synthetic data, all experiments could be completed in less than an hour.
-
-We have summarised our system information below:
-
-\textbf{Software}:
-
-\begin{itemize}
-  \item System Version: macOS 13.3.1
-  \item Kernel Version: Darwin 22.4.0
-\end{itemize}
-
-\textbf{Hardware}:
-
-\begin{itemize}
-  \item Model Name: MacBook Pro
-  \item Model Identifier: MacBookPro16,1
-  \item Processor Name: 8-Core Intel Core i9
-  \item Processor Speed: 2.3 GHz
-  \item Number of Processors: 1
-  \item Total Number of Cores: 8
-  \item L2 Cache (per Core): 256 KB
-  \item L3 Cache: 16 MB
-  \item Hyper-Threading Technology: Enabled
-  \item Memory: 32 GB
-\end{itemize}
-
-\subsection{Results}\label{app:results}
-
-Figure~\ref{fig:mnist-eccco} shows examples of counterfactuals for \textit{MNIST} data where the underlying model is our \textit{JEM Ensemble}.
-Original images are shown on the diagonal and the corresponding counterfactuals are plotted across rows.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=0.9\textwidth]{../artifacts/results/images/mnist_eccco_all_digits.png}
-  \caption{Counterfactuals for \textit{MNIST} data and our \textit{JEM Ensemble}.
-Original images are shown on the diagonal with the corresponding counterfactuals plotted across rows.}\label{fig:mnist-eccco}
-\end{figure}
-
-Table~\ref{tab:results-full} reports all of the evaluation metrics we have computed.
-Table~\ref{tab:results-full-valid} reports the same metrics for the subset of valid counterfactuals.
-The `Unfaithfulness' and `Implausibility' metrics have been discussed extensively in the body of the paper.
-The `Cost' metric relates to the distance between the factual and the counterfactual.
-The `Redundancy' metric measures sparsity in is defined as the percentage of features that remain unperturbed (higher is better).
-The `Uncertainty' metric is just the average value of the smooth set size penalty (Equation~\ref{eq:setsize}).
-Finally, `Validity' is the percentage of valid counterfactuals.
-
-\import{contents/}{table_all.tex}
-
-\import{contents/}{table_all_valid.tex}
+\import{partials/}{appendix.tex}
 

### 2023-12-16 paper/partials/appendix.tex
just using separate file

--- 2023-12-15 paper/appendix.tex
@@
 
-\import{contents/}{table_ebm_params.tex}
+\import{../contents/}{table_ebm_params.tex}
 
@@
 
-\import{contents/}{table_params.tex}
+\import{../contents/}{table_params.tex}
 
-\import{contents/}{table_perf.tex}
+\import{../contents/}{table_perf.tex}
 
-\import{contents/}{table_gen_params.tex}
+\import{../contents/}{table_gen_params.tex}
 
@@
 
-\import{contents/}{table-linearly-separable.tex}
+\import{../contents/}{table-linearly-separable.tex}
 
-\import{contents/}{table-circles.tex}
+\import{../contents/}{table-circles.tex}
 
-\import{contents/}{table-moons.tex}
+\import{../contents/}{table-moons.tex}
 
-\import{contents/}{table-california-housing.tex}
+\import{../contents/}{table-california-housing.tex}
 
-\import{contents/}{table-gmsc.tex}
+\import{../contents/}{table-gmsc.tex}
 
-\import{contents/}{table-german-credit.tex}
+\import{../contents/}{table-german-credit.tex}
 
-\import{contents/}{table-mnist.tex}
+\import{../contents/}{table-mnist.tex}
 
-\import{contents/}{table-fashion-mnist.tex}
+\import{../contents/}{table-fashion-mnist.tex}
 
-\import{contents/}{table-linearly-separable-valid.tex}
+\import{../contents/}{table-linearly-separable-valid.tex}
 
-\import{contents/}{table-circles-valid.tex}
+\import{../contents/}{table-circles-valid.tex}
 
-\import{contents/}{table-moons-valid.tex}
+\import{../contents/}{table-moons-valid.tex}
 
-\import{contents/}{table-california-housing-valid.tex}
+\import{../contents/}{table-california-housing-valid.tex}
 
-\import{contents/}{table-gmsc-valid.tex}
+\import{../contents/}{table-gmsc-valid.tex}
 
-\import{contents/}{table-german-credit-valid.tex}
+\import{../contents/}{table-german-credit-valid.tex}
 
-\import{contents/}{table-mnist-valid.tex}
+\import{../contents/}{table-mnist-valid.tex}
 
-\import{contents/}{table-fashion-mnist-valid.tex}
+\import{../contents/}{table-fashion-mnist-valid.tex}

### 2023-12-16 paper/partials/body.tex
just using separate file

--- 2023-12-16 paper/camera-ready/altmeyer.tex
@@
-%File: formatting-instructions-latex-2024.tex
-%release 2024.0
-\documentclass[letterpaper]{article} % DO NOT CHANGE THIS
-\usepackage{aaai24}  % DO NOT CHANGE THIS
-\usepackage{times}  % DO NOT CHANGE THIS
-\usepackage{helvet}  % DO NOT CHANGE THIS
-\usepackage{courier}  % DO NOT CHANGE THIS
-\usepackage[hyphens]{url}  % DO NOT CHANGE THIS
-\usepackage{graphicx} % DO NOT CHANGE THIS
-\urlstyle{rm} % DO NOT CHANGE THIS
-\def\UrlFont{\rm}  % DO NOT CHANGE THIS
-\usepackage{natbib}  % DO NOT CHANGE THIS AND DO NOT ADD ANY OPTIONS TO IT
-\usepackage{caption} % DO NOT CHANGE THIS AND DO NOT ADD ANY OPTIONS TO IT
-\frenchspacing  % DO NOT CHANGE THIS
-\setlength{\pdfpagewidth}{8.5in}  % DO NOT CHANGE THIS
-\setlength{\pdfpageheight}{11in}  % DO NOT CHANGE THIS
-%
-% These are recommended to typeset algorithms but not required. See the subsubsection on algorithms. Remove them if you don't have algorithms in your paper.
-\usepackage{algorithm}
-% \usepackage{algorithmic}
-
-%
-% These are are recommended to typeset listings but not required. See the subsubsection on listing. Remove this block if you don't have listings in your paper.
-\usepackage{newfloat}
-\usepackage{listings}
-\DeclareCaptionStyle{ruled}{labelfont=normalfont,labelsep=colon,strut=off} % DO NOT CHANGE THIS
-\lstset{%
-	basicstyle={\footnotesize\ttfamily},% footnotesize acceptable for monospace
-	numbers=left,numberstyle=\footnotesize,xleftmargin=2em,% show line numbers, remove this entire line if you don't want the numbers.
-	aboveskip=0pt,belowskip=0pt,%
-	showstringspaces=false,tabsize=2,breaklines=true}
-\floatstyle{ruled}
-\newfloat{listing}{tb}{lst}{}
-\floatname{listing}{Listing}
-%
-% Keep the \pdfinfo as shown here. There's no need
-% for you to add the /Title and /Author tags.
-\pdfinfo{
-/TemplateVersion (2024.1)
-}
-
-% DISALLOWED PACKAGES
-% \usepackage{authblk} -- This package is specifically forbidden
-% \usepackage{balance} -- This package is specifically forbidden
-% \usepackage{color (if used in text)
-% \usepackage{CJK} -- This package is specifically forbidden
-% \usepackage{float} -- This package is specifically forbidden
-% \usepackage{flushend} -- This package is specifically forbidden
-% \usepackage{fontenc} -- This package is specifically forbidden
-% \usepackage{fullpage} -- This package is specifically forbidden
-% \usepackage{geometry} -- This package is specifically forbidden
-% \usepackage{grffile} -- This package is specifically forbidden
-% \usepackage{hyperref} -- This package is specifically forbidden
-% \usepackage{navigator} -- This package is specifically forbidden
-% (or any other package that embeds links such as navigator or hyperref)
-% \indentfirst} -- This package is specifically forbidden
-% \layout} -- This package is specifically forbidden
-% \multicol} -- This package is specifically forbidden
-% \nameref} -- This package is specifically forbidden
-% \usepackage{savetrees} -- This package is specifically forbidden
-% \usepackage{setspace} -- This package is specifically forbidden
-% \usepackage{stfloats} -- This package is specifically forbidden
-% \usepackage{tabu} -- This package is specifically forbidden
-% \usepackage{titlesec} -- This package is specifically forbidden
-% \usepackage{tocbibind} -- This package is specifically forbidden
-% \usepackage{ulem} -- This package is specifically forbidden
-% \usepackage{wrapfig} -- This package is specifically forbidden
-% DISALLOWED COMMANDS
-% \nocopyright -- Your paper will not be published if you use this command
-% \addtolength -- This command may not be used
-% \balance -- This command may not be used
-% \baselinestretch -- Your paper will not be published if you use this command
-% \clearpage -- No page breaks of any kind may be used for the final version of your paper
-% \columnsep -- This command may not be used
-% \newpage -- No page breaks of any kind may be used for the final version of your paper
-% \pagebreak -- No page breaks of any kind may be used for the final version of your paperr
-% \pagestyle -- This command may not be used
-% \tiny -- This is not an acceptable font size.
-% \vspace{- -- No negative value may be used in proximity of a caption, figure, table, section, subsection, subsubsection, or reference
-% \vskip{- -- No negative value may be used to alter spacing above or below a caption, figure, table, section, subsection, subsubsection, or reference
-
-% Additional Packages:
-\usepackage{amsfonts}       % blackboard math symbols
-\usepackage{amsmath}
-\usepackage{amsthm}
-\usepackage{caption}
-\usepackage{graphicx}
-\usepackage{algpseudocode}
-\usepackage{import}
-\usepackage{booktabs}
-\usepackage{longtable}
-\usepackage{array}
-\usepackage{multirow}
-\usepackage{placeins}
-
-% Numbered Environments:
-\newtheorem{definition}{Definition}[section]
-\newtheorem{question}{Research Question}[section]
-
-\setcounter{secnumdepth}{1} %May be changed to 1 or 2 if section numbers are desired.
-
-% The file aaai24.sty is the style file for AAAI Press
-% proceedings, working notes, and technical reports.
-%
-
-% Title
-
-% Your title must be in mixed case, not sentence case.
-% That means all verbs (including short verbs like be, is, using,and go),
-% nouns, adverbs, adjectives should be capitalized, including both words in hyphenated terms, while
-% articles, conjunctions, and prepositions are lower case unless they
-% directly follow a colon or long dash
-
-\title{Faithful Model Explanations through\\
-Energy-Constrained Conformal Counterfactuals}
-\author{
-    %Authors
-    % All authors must be in the same font size and format.
-    Patrick Altmeyer\textsuperscript{\rm 1},
-    Mojtaba Farmanbar\textsuperscript{\rm 2},
-    Arie van Deursen\textsuperscript{\rm 1},
-    Cynthia C. S. Liem\textsuperscript{\rm 1}
-}
-\affiliations{
-    %Afiliations
-    \textsuperscript{\rm 1}Delft University of Technology,\\
-    Faculty of Electrical Engineering, Mathematics and Computer Science\\
-    \textsuperscript{\rm 2}ING Bank\\
-    p.altmeyer@tudelft.nl
-%
-% See more examples next
-}
-
-\begin{document}
-
-\maketitle
-
-\begin{abstract}
-    Counterfactual explanations offer an intuitive and straightforward way to explain black-box models and offer algorithmic recourse to individuals.
-To address the need for plausible explanations, existing work has primarily relied on surrogate models to learn how the input data is distributed.
-This effectively reallocates the task of learning realistic explanations for the data from the model itself to the surrogate.
-Consequently, the generated explanations may seem plausible to humans but need not necessarily describe the behaviour of the black-box model faithfully.
-We formalize this notion of faithfulness through the introduction of a tailored evaluation metric and propose a novel algorithmic framework for generating \textbf{E}nergy-\textbf{C}onstrained \textbf{C}onformal \textbf{Co}unterfactuals that are only as plausible as the model permits.
-Through extensive empirical studies, we demonstrate that \textit{ECCCo} reconciles the need for faithfulness and plausibility.
-In particular, we show that for models with gradient access, it is possible to achieve state-of-the-art performance without the need for surrogate models.
-To do so, our framework relies solely on properties defining the black-box model itself by leveraging recent advances in energy-based modelling and conformal prediction.
-To our knowledge, this is the first venture in this direction for generating faithful counterfactual explanations.
-Thus, we anticipate that \textit{ECCCo} can serve as a baseline for future research.
-We believe that our work opens avenues for researchers and practitioners seeking tools to better distinguish trustworthy from unreliable models.
-\end{abstract}
-
 \section{Introduction}\label{intro}
@@
   \centering
-  \includegraphics[width=0.8\linewidth]{figures/mnist_motivation.png}
+  \includegraphics[width=0.8\linewidth]{../camera-ready/figures/mnist_motivation.png}
   \caption{Counterfactuals for turning a 9 (nine) into a 7 (seven): original image (left), then the counterfactuals generated using \textit{Wachter}, \textit{Schut} and \textit{REVISE}.}\label{fig:motiv}
@@
   \centering
-  \includegraphics[width=0.75\linewidth]{figures/poc_gradient_fields.png}
+  \includegraphics[width=0.75\linewidth]{../camera-ready/figures/poc_gradient_fields.png}
   \caption{Gradient fields and counterfactual paths for different generators.
@@
   \centering
-  \includegraphics[width=1.0\linewidth]{figures/mnist_benchmark.png}
+  \includegraphics[width=1.0\linewidth]{../camera-ready/figures/mnist_benchmark.png}
   \caption{Counterfactuals for turning a 3 into a 5: factual (left), then the counterfactuals generated by \textit{ECCCo}, \textit{ECCCo+}, \textit{REVISE}, \textit{Schut} and \textit{Wachter}.}\label{fig:mnist-bmk}
@@
 All code used for the analysis in this paper can be found here: https://github.com/pat-alt/ECCCo.jl.
-
-\bibliography{aaai24,bib}
-
-\end{document}

### 2023-12-16 paper/partials/appendix.tex
appendix done

@@
 
-The core part of our code base is integrated into a larger ecosystem of \texttt{Julia} packages that we are actively developing and maintaining.
-To avoid compromising the double-blind review process, we only provide a link to an anonymized repository at this stage: \url{https://anonymous.4open.science/r/ECCCo-1252/README.md}.
+We have built a small Julia package, \texttt{ECCCo.jl}, that was used to perform the analysis in this paper: https://github.com/pat-alt/ECCCo.jl.
+The package will be absorbed by \texttt{CounterfactualExplanations.jl} in the future~\citep{altmeyer2023explaining}.
 
@@
 
-In our experiments we always generate multiple counterfactuals for each model and generator.
-Each time the factual and target class is drawn randomly.
-For each generator and model we choose $n_f=100$ factuals for all of our synthetic and vision data.
-For tabular data we choose $n_f=25$ because larger values made grid search computationally prohibitive.
-For vision data, grid search was computationally prohibitve in any case, so hyperparameters were tuned manually.
-For all other datasets grid search was performed over different combinations of penalty strengths and optimizer steps sizes (details of which can be found in the code).
-To select the final hyperparameter setting we used the unfaithfulness metric as our criterion.
+In our experiments, we always generate multiple counterfactuals for each model and generator over multiple runs to account for stochasticity.
+For each generator, model and dataset, we sample a total of $n_f=100$ factuals for a single benchmark, where each time the factual and target class is drawn randomly.
+This process is repeated over 50 runs for the synthetic and tabular datasets, which yields a total of around 1.5 million counterfactuals per dataset for the final results.
+For vision datasets, we relied on 10 runs yielding a total of 300,000 counterfactuals.
+The final results presented in the paper and this appendix show the sample averages and the standard deviations of samples averages across all runs.
+For grid search, we used single runs for each parameter specification generating 100 counterfactuals per generator and model for synthetic and tabular datasets.
+For vision data, we generated 10 counterfactuals for each model, generator and parameter specification.
 

### 2023-12-16 paper/partials/body.tex
appendix done

@@
 While in the context of EBM, this is usually done during training, we propose to repurpose this approach during inference in order to evaluate the faithfulness of model explanations.
-The appendix provides additional implementation details for any tasks related to energy-based modelling\footnote{The supplementary appendix can be found here: https://github.com/pat-alt/ECCCo.jl.}.
+The appendix provides additional implementation details for any tasks related to energy-based modelling.
 

### 2023-12-17 paper/partials/appendix.tex
more minor updates

@@
 This has helped to reduce the computational burden for our models, which should make it easier for others to reproduce our findings.
-Figures~\ref{fig:mnist-gen} and~\ref{fig:moons-gen} show generated samples for our \textit{MNIST} and \textit{Moons} data, to provide a sense of their generative property.
+Figures~\ref{fig:mnist-gen} and~\ref{fig:poc-gen} show generated samples for our \textit{MNIST} and \textit{Moons} data, to provide a sense of their generative property.
 
@@
   \centering
-  \includegraphics[width=0.5\linewidth]{../www/moons_generated_JEM.png}
-  \caption{Conditionally generated samples (stars) for our \textit{Moons} data using a JEM.}\label{fig:moons-gen}
+  \includegraphics[width=0.5\linewidth]{../www/poc_conditional_samples.png}
+  \caption{Conditionally generated samples (stars) for our \textit{Linearly Separable} data using a JEM.}\label{fig:poc-gen}
 \end{figure}

### 2023-12-17 experiments/notebooks/annex.qmd
arxived

@@
 )
-savefig(plt, "www/dist_energy.png")
+savefig(plt, "paper/www/dist_energy.png")
 ```

### 2023-12-17 experiments/notebooks/setup.jl

@@ -3,4 +3,4 @@ Plots.theme(:wong)
 img_height = 300
 panel_height = 250
 paper_figpath = "paper/camera-ready/figures"
-output_images_path = "www/"
+output_images_path = "paper/www/"

### 2023-12-19 experiments/notebooks/figure2.qmd
all charts in paper recreated

--- 2023-12-13 experiments/notebooks/poc.qmd
@@
 ```
+
+This notebook can be used to recreate Figure 2 in the paper.
 
@@
 end
+savefig(plt, joinpath(output_images_path, "poc_conditional_samples.png"))
 ```
@@
 ```
+
+The code below generates Figure 2 in the paper.
 

### 2023-12-19 paper/camera-ready/altmeyer.tex

@@
 While in the context of EBM, this is usually done during training, we propose to repurpose this approach during inference in order to evaluate the faithfulness of model explanations.
-The appendix provides additional implementation details for any tasks related to energy-based modelling\footnote{The supplementary appendix can be found here: https://github.com/pat-alt/ECCCo.jl.}.
+The appendix provides additional implementation details for any tasks related to energy-based modelling\footnote{The supplementary appendix can be found here: https://arxiv.org/abs/2312.10648.}.
 

### 2023-12-19 paper/contents/table_gen_params.tex

@@
 \midrule
-Linearly Separable & 0.01 & 0.10 & 0.1 & 0.05 & 0.0\\
-Moons & 0.01 & 0.10 & 0.1 & 0.50 & 0.0\\
-Circles & 0.05 & 0.10 & 0.1 & 0.05 & 1.0\\
-California Housing & 0.05 & 0.10 & 0.1 & 0.10 & 0.0\\
-GMSC & 0.05 & 0.10 & 0.1 & 0.10 & 0.0\\
-German Credit & 0.05 & 0.20 & 0.2 & 0.20 & 0.5\\
-MNIST & 0.10 & 0.01 & 0.1 & 0.30 & 0.0\\
-Fashion MNIST & 0.10 & 0.01 & 0.1 & 0.30 & 0.0\\
+Linearly Separable & 0.05 & 0.10 & 0.1 & 0.5 & 0.0\\
+Moons & 0.05 & 0.10 & 0.1 & 0.2 & 0.0\\
+Circles & 0.05 & 0.10 & 0.1 & 0.2 & 0.0\\
+California Housing & 0.05 & 0.10 & 0.1 & 0.5 & 0.0\\
+GMSC & 0.05 & 0.10 & 0.1 & 0.5 & 0.0\\
+German Credit & 0.05 & 0.10 & 0.1 & 0.1 & 0.5\\
+MNIST & 0.05 & 0.01 & 0.1 & 0.3 & 0.0\\
+Fashion MNIST & 0.05 & 0.01 & 0.1 & 0.3 & 0.0\\
 \bottomrule

### 2023-12-19 paper/partials/appendix.tex

@@
   \centering
-  \includegraphics[width=0.75\linewidth]{../www/mnist_generated_JEM Ensemble.png}
+  \includegraphics[width=0.75\linewidth]{www/mnist_generated_JEM_Ensemble.png}
   \caption{Conditionally generated \textit{MNIST} images for our JEM Ensemble.}\label{fig:mnist-gen}
@@
   \centering
-  \includegraphics[width=0.5\linewidth]{../www/poc_conditional_samples.png}
+  \includegraphics[width=0.5\linewidth]{www/poc_conditional_samples.png}
   \caption{Conditionally generated samples (stars) for our \textit{Linearly Separable} data using a JEM.}\label{fig:poc-gen}
@@
   \centering
-  \includegraphics[width=1.0\linewidth]{../www/poc_set_size.png}
+  \includegraphics[width=1.0\linewidth]{www/poc_set_size.png}
   \caption{Prediction set size (left), smooth set size loss (centre) and configurable classification loss (right) for a JEM trained on our \textit{Linearly Separable} data.}\label{fig:cp-diff}
@@
   \centering
-  \includegraphics[width=1.0\linewidth]{../www/mnist_all_lenet_eccco.png}
+  \includegraphics[width=1.0\linewidth]{www/mnist_all_lenet_eccco.png}
   \caption{Counterfactuals for \textit{MNIST} data generated by \textit{ECCCo+}.
@@
   \centering
-  \includegraphics[width=1.0\linewidth]{../www/mnist_all_mlp_eccco.png}
+  \includegraphics[width=1.0\linewidth]{www/mnist_all_mlp_eccco.png}
   \caption{Counterfactuals for \textit{MNIST} data generated by \textit{ECCCo+}.
@@
   \centering
-  \includegraphics[width=1.0\linewidth]{../www/mnist_all_mlp_ens_eccco.png}
+  \includegraphics[width=1.0\linewidth]{www/mnist_all_mlp_ens_eccco.png}
   \caption{Counterfactuals for \textit{MNIST} data generated by \textit{ECCCo+}.
@@
   \centering
-  \includegraphics[width=1.0\linewidth]{../www/mnist_all_jem_eccco.png}
+  \includegraphics[width=1.0\linewidth]{www/mnist_all_jem_eccco.png}
   \caption{Counterfactuals for \textit{MNIST} data generated by \textit{ECCCo+}.
@@
   \centering
-  \includegraphics[width=1.0\linewidth]{../www/mnist_all_jem_ens_eccco.png}
+  \includegraphics[width=1.0\linewidth]{www/mnist_all_jem_ens_eccco.png}
   \caption{Counterfactuals for \textit{MNIST} data generated by \textit{ECCCo+}.
@@
   \centering
-  \includegraphics[width=1.0\linewidth]{../www/mnist_all_lenet_revise.png}
+  \includegraphics[width=1.0\linewidth]{www/mnist_all_lenet_revise.png}
   \caption{Counterfactuals for \textit{MNIST} data generated by \textit{REVISE}.
@@
   \centering
-  \includegraphics[width=1.0\linewidth]{../www/mnist_all_mlp_revise.png}
+  \includegraphics[width=1.0\linewidth]{www/mnist_all_mlp_revise.png}
   \caption{Counterfactuals for \textit{MNIST} data generated by \textit{REVISE}.
@@
   \centering
-  \includegraphics[width=1.0\linewidth]{../www/mnist_all_mlp_ens_revise.png}
+  \includegraphics[width=1.0\linewidth]{www/mnist_all_mlp_ens_revise.png}
   \caption{Counterfactuals for \textit{MNIST} data generated by \textit{REVISE}.
@@
   \centering
-  \includegraphics[width=1.0\linewidth]{../www/mnist_all_jem_revise.png}
+  \includegraphics[width=1.0\linewidth]{www/mnist_all_jem_revise.png}
   \caption{Counterfactuals for \textit{MNIST} data generated by \textit{REVISE}.
@@
   \centering
-  \includegraphics[width=1.0\linewidth]{../www/mnist_all_jem_ens_revise.png}
+  \includegraphics[width=1.0\linewidth]{www/mnist_all_jem_ens_revise.png}
   \caption{Counterfactuals for \textit{MNIST} data generated by \textit{REVISE}.
@@
   \centering
-  \includegraphics[width=1.0\linewidth]{../www/vae_rec.png}
+  \includegraphics[width=1.0\linewidth]{www/vae_rec.png}
   \caption{Randomly drawn \textit{MNIST} images and their reconstructions generated by the VAE used by \textit{REVISE}.}\label{fig:vae-rec}

### 2024-01-05 experiments/notebooks/figure2.qmd

@@
 X = Float32.(MLJBase.matrix(X))
+gen_samples = Dict{Any,Any}()
 if typeof(jem.sampler) <: ConditionalSampler
@@
         X̂ = generate_conditional_samples(jem, _batch_size, target; niter=niter) 
+        gen_samples[target] = X̂
         ex = extrema(hcat(X,X̂), dims=2)
@@
 ```
+
+## Simple CE
+
+```{julia}
+Random.seed!(2024)
+M = fit_model(counterfactual_data, :Linear)
+gen = WachterGenerator(opt=Flux.Descent(0.001))
+factual_label =  levels(labels)[2]
+x_factual = reshape(X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
+target =  levels(labels)[1]
+factual = predict_label(M, counterfactual_data, x_factual)[1]
+ce = generate_counterfactual(
+    x_factual, target, counterfactual_data, M, gen;
+    converge_when=:generator_conditions,
+)
+plt = plot(
+    ce, dpi=300, size=(500,500), 
+    cbar=false, axis=nothing, legend=false,
+    palette = col_pal,
+)
+savefig(plt, joinpath(output_images_path, "simple_ce.png"))
+```
+
+## True Density
+
+```{julia}
+using KernelDensity
+using StatsPlots
+
+target_cls_mask = counterfactual_data.output_encoder.labels .== 1
+X_target = counterfactual_data.X[:,target_cls_mask]
+
+x = X_target[1,:]
+y = X_target[2,:]
+
+k = kde((x, y))
+
+plt = contourf(
+    k,
+	c = :BuPu,
+	linewidth = 0,
+    size=(500,500),
+    dpi=300,
+    colorbar=false,
+    axis=nothing,
+    legend=false,
+    palette = col_pal,
+)
+
+scatter!(
+    counterfactual_data, alpha=0.3,
+    xlims = extrema(x),
+    ylims = extrema(y),
+)
+
+ce_path = hcat(path(ce)...)
+scatter!(
+    ce_path[1,:], ce_path[2,:], 
+    color=predict_label(M, counterfactual_data, ce_path), ms=10, alpha=0.5,
+    label=""
+)
+savefig(plt, joinpath(output_images_path, "density_true.png"))
+```
+
+## Posterior
+
+```{julia}
+x = gen_samples[target][1,:]
+y = gen_samples[target][2,:]
+
+k = kde(
+    (x, y); 
+    bandwidth=(0.1,0.1),
+    boundary=(extrema(X_target[1,:]),extrema(X_target[2,:]))
+)
+
+plt = contourf(
+    k,
+	c = :BuPu,
+	linewidth = 0,
+    size=(500,500),
+    dpi=300,
+    colorbar=false,
+    axis=nothing,
+    legend=false,
+    palette = col_pal,
+)
+
+scatter!(
+    counterfactual_data, alpha=0.3,
+    xlims = extrema(X_target[1,:]),
+    ylims = extrema(X_target[2,:]),
+)
+
+ce_path = hcat(path(ce)...)
+scatter!(
+    ce_path[1,:], ce_path[2,:], 
+    color=predict_label(M, counterfactual_data, ce_path), ms=10, alpha=0.5,
+    label=""
+)
+
+scatter!(
+    plt, gen_samples[1][1,:], gen_samples[1][2,:],
+    shape=:star5, ms=10, color=:yellow, alpha=0.5, label=""
+)
+
+savefig(plt, joinpath(output_images_path, "density_model.png"))
+```

### 2024-01-10 paper/camera-ready/altmeyer.tex
some touches added

@@
 \newtheorem{definition}{Definition}[section]
+\renewcommand{\thedefinition}{\arabic{definition}}
 \newtheorem{question}{Research Question}[section]
+\renewcommand{\thequestion}{\arabic{question}}
 
-\setcounter{secnumdepth}{1} %May be changed to 1 or 2 if section numbers are desired.
+% \setcounter{secnumdepth}{1} %May be changed to 1 or 2 if section numbers are desired.
 
@@
     %Afiliations
-    \textsuperscript{\rm 1}Delft University of Technology,\\
+    \textsuperscript{\rm 1}Delft University of Technology,
     Faculty of Electrical Engineering, Mathematics and Computer Science\\
@@
 
-\section{Why Fidelity is not Enough: A Motivational Example}\label{fidelity}
+\section{Why Fidelity Is Not Enough: A Motivational Example}\label{fidelity}
 
@@
 
-\section{Faithful first, Plausible second}\label{faithfulness}
+\section{Faithful First, Plausible Second}\label{faithfulness}
 
@@
 
-\subsection{Quantifying the Model's Generative Property}
+\subsection{Quantifying the Model’s Generative Property}
 

### 2024-01-10 paper/camera-ready/altmeyer.tex
tables sorted

@@
 For the predictive modelling tasks, we use multi-layer perceptrons (\textit{MLP}), deep ensembles, joint energy models (\textit{JEM}) and convolutional neural networks (LeNet-5 \textit{CNN}~\citep{lecun1998gradient}).
-Both joint-energy modelling and ensembling have been associated with improved generative properties and adversarial robustness~\citep{grathwohl2020your,lakshminarayanan2016simple}, so we expect this to be positively correlated with the plausibility of \textit{ECCCo}.
+Both joint-energy modelling and ensembling have been associated with improved generative properties and adversarial robustness~\citep{grathwohl2020your,lakshminarayanan2016simple}, so we expect this to enhance the plausibility of \textit{ECCCo}.
 To account for stochasticity, we generate many counterfactuals for each target class, generator, model and dataset over multiple runs.
@@
 Full details concerning model training as well as detailed descriptions and results for all datasets can be found in the appendix.
-In the following, we will focus on the most relevant results highlighted in Tables~\ref{tab:results-tabular} and~\ref{tab:results-vision}.
-The tables show sample averages along with standard deviations across multiple runs for our key evaluation metrics for the \textit{California Housing} and \textit{GMSC} datasets (Table~\ref{tab:results-tabular}) and the \textit{MNIST} dataset (Table~\ref{tab:results-vision}).
-For each metric, the best outcomes are highlighted in bold.
-Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (Wachter).
-For the tabular datasets, we use the default Euclidian distance to measure unfaithfulness and implausibility as defined in Equations~\ref{eq:faith} and~\ref{eq:impl}, respectively.
+In the following, we focus on the most relevant results highlighted in Tables~\ref{tab:results-tabular} and~\ref{tab:results-vision}.
+They show sample averages along with standard deviations across multiple runs for our key evaluation metrics for the \textit{California Housing} and \textit{GMSC} datasets (Table~\ref{tab:results-tabular}) and the \textit{MNIST} dataset (Table~\ref{tab:results-vision}).
+For each metric, the best outcomes are in bold.
+Asterisks indicate that a given value is more than one (*) or two (**) standard deviations away from the baseline (\textit{Wachter}).
+For the tabular datasets, we use the Euclidian distance to measure unfaithfulness and implausibility as defined in Equations~\ref{eq:faith} and~\ref{eq:impl}, respectively.
+For the vision datasets, we rely on measuring the structural dissimilarity between images for our unfaithfulness and implausibility metrics~\citep{wang2003multiscale}.
 The third metric presented in Table~\ref{tab:results-tabular} quantifies the predictive uncertainty of the counterfactual as measured by Equation~\ref{eq:setsize}.
-For the vision datasets, we rely on measuring the structural dissimilarity between images for our unfaithfulness and implausibility metrics~\citep{wang2003multiscale}.
 
@@
 \begin{table*}
+  \setlength{\tabcolsep}{4pt}
   \centering
-\resizebox{\linewidth}{!}{
+  \fontsize{9}{11}\selectfont
   \begin{tabular}[t]{llcccccc}
@@
 
-\multirow{-7}{*}{\raggedright\arraybackslash MLP Ensemble} & Wachter & 4.04 ± 0.07\hphantom{*}\hphantom{*} & 1.13 ± 0.12\hphantom{*}\hphantom{*} & 0.16 ± 0.01\hphantom{*}\hphantom{*} & 4.10 ± 0.07\hphantom{*}\hphantom{*} & 0.95 ± 0.08\hphantom{*}\hphantom{*} & 0.32 ± 0.01\hphantom{*}\hphantom{*}\\
+  \multirow{-7}{*}{\raggedright\arraybackslash MLP} & Wachter & 4.04 ± 0.07\hphantom{*}\hphantom{*} & 1.13 ± 0.12\hphantom{*}\hphantom{*} & 0.16 ± 0.01\hphantom{*}\hphantom{*} & 4.10 ± 0.07\hphantom{*}\hphantom{*} & 0.95 ± 0.08\hphantom{*}\hphantom{*} & 0.32 ± 0.01\hphantom{*}\hphantom{*}\\
   \cmidrule{1-8}
@@
 
-\multirow{-7}{*}{\raggedright\arraybackslash JEM Ensemble} & Wachter & 1.71 ± 0.09\hphantom{*}\hphantom{*} & 0.99 ± 0.08\hphantom{*}\hphantom{*} & 0.14 ± 0.00\hphantom{*}\hphantom{*} & 1.31 ± 0.08\hphantom{*}\hphantom{*} & 0.95 ± 0.10\hphantom{*}\hphantom{*} & 0.33 ± 0.01\hphantom{*}\hphantom{*}\\
+  \multirow{-7}{*}{\raggedright\arraybackslash JEM} & Wachter & 1.71 ± 0.09\hphantom{*}\hphantom{*} & 0.99 ± 0.08\hphantom{*}\hphantom{*} & 0.14 ± 0.00\hphantom{*}\hphantom{*} & 1.31 ± 0.08\hphantom{*}\hphantom{*} & 0.95 ± 0.10\hphantom{*}\hphantom{*} & 0.33 ± 0.01\hphantom{*}\hphantom{*}\\
   \bottomrule
-\end{tabular}}
+  \end{tabular}
 \caption{Results for tabular datasets: sample averages +/- one standard deviation across valid counterfactuals.
+Model ensembling was used in both cases.
 The best outcomes are highlighted in bold.
@@
 We also generally find that latent space search does not impede faithfulness for \textit{ECCCo}.
-In most cases \textit{ECCCo+} is either on par with \textit{ECCCo} or even outperforms it.
+In most cases \textit{ECCCo+} is either on par with \textit{ECCCo} or outperforms it.
 There are some notable exceptions though.
-Cases in which \textit{ECCCo} achieves substantially better faithfulness without latent space search tend to involve more vulnerable models like the simple MLP for MNIST in Table~\ref{tab:results-vision}.
+Cases in which \textit{ECCCo} achieves substantially better faithfulness without latent space search tend to involve more vulnerable models like the MLP for MNIST in Table~\ref{tab:results-vision}.
 We explain this finding as follows: even though dimensionality reduction through PCA in the case of \textit{ECCCo+} can be considered a relatively mild form of intervention, the first $n_z$ principal components fail to capture some of the variation in the data.
@@
   \centering
-\resizebox{\linewidth}{!}{
+  \fontsize{9}{11}\selectfont
   \begin{tabular}[t]{llcc}
@@
   \bottomrule
-\end{tabular}}
-\caption{Results for vision dataset. Formatting details are the same as in Table~\ref{tab:results-tabular}. \label{tab:results-vision}}
+  \end{tabular}
+\caption{Results for vision dataset. No model ensembling was used. Formatting details are the same as in Table~\ref{tab:results-tabular}. \label{tab:results-vision}}
 \end{table}
@@
 Firstly, we recognize that our proposed distance-based evaluation metrics for plausibility and faithfulness may not be universally applicable to all types of data.
-In any case, they depend on choosing a distance metric on a case-by-case basis, as we have done in this work.
+They depend on choosing a distance metric on a case-by-case basis, as we have done in this work.
 Arguably, commonly used metrics for measuring other desiderata such as closeness suffer from the same pitfall.
-We therefore think that future work on counterfactual explanations could benefit from defining universal evaluation metrics.
+We therefore think that future work could benefit from defining universal evaluation metrics.
 

### 2024-01-10 paper/tables.Rmd
tables sorted

@@
 tab_i <- dcast(tab_i, model + generator ~ dataname + variable)
+
+# Change model names to reduce column space
+levels(tab_i$model)[match("MLP Ensemble",levels(tab_i$model))] <- "MLP"
+levels(tab_i$model)[match("JEM Ensemble",levels(tab_i$model))] <- "JEM"
+
 col_names <- c(
@@
   "Results for %s datasets: sample averages +/- one standard deviation across valid counterfactuals.
+Model ensembling was used in both cases.
 Best outcomes are highlighted in bold.
@@
   kable_styling(
-    latex_options = c("scale_down")
+    font_size = 9
   ) %>%
@@
 caption <- sprintf(
-  "Results for %s dataset. Formatting details are the same as in Table~\\ref{tab:results-tabular}. \\label{tab:results-%s} \\newline",
+  "Results for %s dataset. No model ensembling was used. Formatting details are the same as in Table~\\ref{tab:results-tabular}. \\label{tab:results-%s} \\newline",
   chosen_source,
